### 2021-06-08 thesis/01-intro.Rmd
datasets upload

---
output:
  pdf_document: default
  html_document: default
---
# Introduction

<!--### Articles:

  1. "Are We Consistently Biased? Multidimensional Analysis of Biases in Distributional Word Vectors"
  https://www.aclweb.org/anthology/S19-1010.pdf

  2. Proving the harmfullness of biases in models 
[source https://poseidon01.ssrn.com/delivery.php?ID=004020091089126117028081015085009075120015077012021005105005104113074127101113104030041107053119007034112081087110119071025006053026022082043000071011123067077124004027040024008122116091025075126111098104007096110113089007070120009111119015084097017092&EXT=pdf&INDEX=TRUE]

  3. Summary what causes bias short article
  [source https://kawine.github.io/blog/nlp/2019/09/23/bias.html]-->

<!-- Men Also Like Shopping: Reducing Gender Bias Amplification using Corpus-level Constraint -->
<!-- Bias in Bios: A Case Study of Semantic Representation Bias in a High-Stakes Setting -->

<!-- !bibliografia paperdraft; jabref
4.pomysly na czwarty rozdzial - podsumowanie; 
jak mierzyc bias za pomoca jednej liczby?
IAT - test psychologiczny -->

<!-- 
    1.state the general topic and give some background
    2.provide a review of the literature related to the topic
    3.define the terms and scope of the topic
    4.outline the current situation
    5.evaluate the current situation (advantages/ disadvantages) and identify the gap
    6.identify the importance of the proposed research
    7.state the research problem/ questions
    8.state the research aims and/or research objectives
    9.state the hypotheses
    10.outline the order of information in the thesis
    11.outline the methodology
-->

\textbf{state the general topic and give some background}

Natural language processing (NLP) is a subfield of computer science that processes and analyzes language in text and speech 
with the use of modern programming methods. It has practical applications in everyday life as it concerns tasks such as email filters,
smart assistants, search results, language translations, text analytics and so on. Models used to accomplish these tasks need a lot of
data to learn from. This data originates from humans activities and historical recordings like texts, messages, speeches. It turns out
that in the learning process these models can learn implicit biases that reflect harmful stereotypical thinking still present in modern societies. One can find methods that aim at identifying hidden biases and/or try to remove them by modifying the models explicitly. 
There are many different types of models in NLP depending on a task that they ought to solve. However all of them need as an input words represented as numerical values and this is accomplished with word embedding models. The biases seem to have their primary source
in the way the words are assigned the numerical values.
\newline

\textbf{review of the literature related to the topic}

There is a bunch of literature available on the topic of bias detection and mitigation in NLP models. Bolukbasi2016Man focuses on gender
biases that may be observable while investigating the representation of job occupations and gender in terms of their assigned numerical
values. The authors apply cosine similarity measurement to investigate the phenomenon where jobs stereotypically associated with a given
gender are in fact in the model situated closer to this gender.

Caliskan2017Semantics touches upon the topic of biases regarding race and gender. They apply knowledge from well-known psychological studies like Implicit Association Test to research the relation between human stereotypical thinking and model learnt biases to
discover close relationship between these two. For the evaluation they use Word Embedding Association Test (WEAT) and the Word Embedding Factual Association Test (WEFAT).

manzini2019black proposes novel way of using a cosine similarity method to get the information on assumed resemblance between words. They investigate 
an approach that enables to measure the bias for a class (like gender, religion, race) and express the final result with only one averaged number.
\newline

<!-- Nissim2019Fair -->

\textbf{define the terms and scope of the topic}

It is worth noticing the general distinction of biases mentioned in Caliskan2017Semantics. They refer to the publication concerning Implicit Association
Test Greenwald et al., 1998) where certain baseline of bias phenomenon was introduced. Namely it seems that humans naturally exhibit some biases and
that not always bring social concern. One can imagine the intuitive associations between for example insects and flowers, and the feelings of pleasantness or unpleasantness. In general people would rather associate flowers with feeling pleasant than insects and this preference could be named a bias or
prejudice in some direction. However this type of preference does not cause an uproar and it is rather morally neutral case. Unfortunately there are other
biases and prejudices that directly influence the quality of other people's lives and therefore they should be taken care of.

One can find a bunch of various definitions trying to capture what bias and fairness actually are. With the choice of the definition, implications into the real-life applications may change as well as it was pointed out in Mehrabi2019Survey. They mark out that there exist different types of biases, the list if long but among others there are historical bias, representation bias, measurement bias. This indicates how complex the process itself is. Without the proper understanding and awareness of the problem, people are prone to unconsciously sustain this phenomenon.

In the article one can also find the distinction on different types of discrimination, some of them will be shortly described. It is worth first mentioning that protected attributes are those qualities, traits or characteristics that one cannot, according to the law, discriminated against.
Direct discrimination refers to the situation when protected attributes of individuals explicitly result in non-favorable outcomes toward them. In contrast in indirect discrimination individuals appear to be treated equally but anyway they end up being treated unjustly due to the hidden effects of biases towards their protected attributes. Systemic discrimination takes place when policies, customs or behaviors that result from certain culture or organizational structure lead to discrimination against some groups of people. Finally, very common statistical discrimination refers to using
average group statistics to judge person belonging to the group.

The topic of discrimination is entangled with another concept which is fairness. It is essential to grasp some concepts of fairness to take them into
consideration while designing implementation of some machine learning model. In Mehrabi2019Survey one may notice that depending on the context and application different definitions may be applied.
\newline

\textbf{outline the current situation}
The most popular methods focus on comparing the similarity between words from protected groups and those that are considered to be stereotypical or harmful in some way. One can find in this group methods such as euclidean distance, dot product or cosine similarity. There are also other ways to detect the effects of biases. For example through the investigation of the model performance on certain tasks that validate if the model returns some values 
independently on gender or race or not. 
\newline

\textbf{evaluate the current situation (advantages/ disadvantages) and identify the gap}

In the currently used methods (like cosine similarity) the values of similarity are often aggregated in a way that may lead to false conclusions. For 
example due to the averaging of values and the lack of confidence interval information. 
\newline

\textbf{identify the importance of the proposed research}
One can find a number of articles on negative real-life implications resulting from the presence of unaddressed biases in the machine learning models.
\newline

\textbf{state the research problem/ questions}

In the paper we indicate how current methods used to detect biases in natural language models are limited in terms of confidence interval.
\newline

\textbf{state the research aims and/or research objectives}

Our research tries to answer the question of how to enhance the current way in which the bias detection is performed to make sure that it is
methodologically valid.
\newline

\textbf{state the hypotheses}

Our hypothesis is that there can be greater understanding of data and bias implications when confidence interval and Bayesian method are applied to the
methodology.
\newline
<!-- \textbf{outline the order of information in the thesis} -->

\textbf{outline the methodology}

To discuss!

### 2021-06-08 thesis/_book/_main.tex
datasets upload

--- 2021-06-08 thesis/01-intro.Rmd
@@
----
-output:
-  pdf_document: default
-  html_document: default
----
-# Introduction
+\documentclass[]{book}
+\usepackage{lmodern}
+\usepackage{amssymb,amsmath}
+\usepackage{ifxetex,ifluatex}
+\usepackage{fixltx2e} % provides \textsubscript
+\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+\else % if luatex or xelatex
+  \ifxetex
+    \usepackage{mathspec}
+  \else
+    \usepackage{fontspec}
+  \fi
+  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
+\fi
+% use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+% use microtype if available
+\IfFileExists{microtype.sty}{%
+\usepackage[]{microtype}
+\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
+\usepackage[unicode=true]{hyperref}
+\hypersetup{
+            pdftitle={Conceptual and methodological problems with bias detection and avoidance in natural language processing},
+            pdfauthor={Alicja Dobrzeniecka},
+            pdfborder={0 0 0},
+            breaklinks=true}
+\urlstyle{same}  % don't use monospace font for urls
+\usepackage{longtable,booktabs}
+% Fix footnotes in tables (requires footnote package)
+\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{long table}}{}
+\usepackage{graphicx,grffile}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+\IfFileExists{parskip.sty}{%
+\usepackage{parskip}
+}{% else
+\setlength{\parindent}{0pt}
+\setlength{\parskip}{6pt plus 2pt minus 1pt}
+}
+\setlength{\emergencystretch}{3em}  % prevent overfull lines
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+\setcounter{secnumdepth}{5}
+% Redefines (sub)paragraphs to behave more like sections
+\ifx\paragraph\undefined\else
+\let\oldparagraph\paragraph
+\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+\let\oldsubparagraph\subparagraph
+\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
 
-<!--### Articles:
+% set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
 
-  1. "Are We Consistently Biased? Multidimensional Analysis of Biases in Distributional Word Vectors"
-  https://www.aclweb.org/anthology/S19-1010.pdf
+\title{Conceptual and methodological problems with bias detection and avoidance
+in natural language processing}
+\author{Alicja Dobrzeniecka}
+\date{2021-06-08}
 
-  2. Proving the harmfullness of biases in models 
-[source https://poseidon01.ssrn.com/delivery.php?ID=004020091089126117028081015085009075120015077012021005105005104113074127101113104030041107053119007034112081087110119071025006053026022082043000071011123067077124004027040024008122116091025075126111098104007096110113089007070120009111119015084097017092&EXT=pdf&INDEX=TRUE]
+\begin{document}
+\maketitle
 
-  3. Summary what causes bias short article
-  [source https://kawine.github.io/blog/nlp/2019/09/23/bias.html]-->
+{
+\setcounter{tocdepth}{5}
+\tableofcontents
+}
+\chapter*{Preface}\label{preface}
+\addcontentsline{toc}{chapter}{Preface}
 
-<!-- Men Also Like Shopping: Reducing Gender Bias Amplification using Corpus-level Constraint -->
-<!-- Bias in Bios: A Case Study of Semantic Representation Bias in a High-Stakes Setting -->
+Test
 
-<!-- !bibliografia paperdraft; jabref
-4.pomysly na czwarty rozdzial - podsumowanie; 
-jak mierzyc bias za pomoca jednej liczby?
-IAT - test psychologiczny -->
-
-<!-- 
-    1.state the general topic and give some background
-    2.provide a review of the literature related to the topic
-    3.define the terms and scope of the topic
-    4.outline the current situation
-    5.evaluate the current situation (advantages/ disadvantages) and identify the gap
-    6.identify the importance of the proposed research
-    7.state the research problem/ questions
-    8.state the research aims and/or research objectives
-    9.state the hypotheses
-    10.outline the order of information in the thesis
-    11.outline the methodology
--->
+\chapter{Introduction}\label{introduction}
 
@@
 
-Natural language processing (NLP) is a subfield of computer science that processes and analyzes language in text and speech 
-with the use of modern programming methods. It has practical applications in everyday life as it concerns tasks such as email filters,
-smart assistants, search results, language translations, text analytics and so on. Models used to accomplish these tasks need a lot of
-data to learn from. This data originates from humans activities and historical recordings like texts, messages, speeches. It turns out
-that in the learning process these models can learn implicit biases that reflect harmful stereotypical thinking still present in modern societies.
-One can find methods that aim at identifying hidden biases and/or try to remove them by modifying the models explicitly.
-There are many different types of models in NLP depending on a task that they ought to solve.
-However all of them need as an input words represented as numerical values and this is accomplished with word embedding models.
-The biases seem to have their primary source
-in the way the words are assigned the numerical values.
-\newline
+Natural language processing (NLP) is a subfield of computer science that
+processes and analyzes language in text and speech with the use of
+modern programming methods. It has practical applications in everyday
+life as it concerns tasks such as email filters, smart assistants,
+search results, language translations, text analytics and so on. Models
+used to accomplish these tasks need a lot of data to learn from. This
+data originates from humans activities and historical recordings like
+texts, messages, speeches. It turns out that in the learning process
+these models can learn implicit biases that reflect harmful
+stereotypical thinking still present in modern societies. One can find
+methods that aim at identifying hidden biases and/or try to remove them
+by modifying the models explicitly. There are many different types of
+models in NLP depending on a task that they ought to solve. However all
+of them need as an input words represented as numerical values and this
+is accomplished with word embedding models. The biases seem to have
+their primary source in the way the words are assigned the numerical
+values. \newline
 
@@
 
-There is a bunch of literature available on the topic of bias detection and mitigation in NLP models. Bolukbasi2016Man focuses on gender
-biases that may be observable while investigating the representation of job occupations and gender in terms of their assigned numerical
-values. The authors apply cosine similarity measurement to investigate the phenomenon where jobs stereotypically associated with a given
-gender are in fact in the model situated closer to this gender.
+There is a bunch of literature available on the topic of bias detection
+and mitigation in NLP models. Bolukbasi2016Man focuses on gender biases
+that may be observable while investigating the representation of job
+occupations and gender in terms of their assigned numerical values. The
+authors apply cosine similarity measurement to investigate the
+phenomenon where jobs stereotypically associated with a given gender are
+in fact in the model situated closer to this gender.
 
-Caliskan2017Semantics touches upon the topic of biases regarding race and gender.
-They apply knowledge from well-known psychological studies like Implicit Association Test to research the relation between human stereotypical thinking and model learnt biases to
-discover close relationship between these two. For the evaluation they use Word Embedding Association Test (WEAT) and the Word Embedding Factual Association Test (WEFAT).
+Caliskan2017Semantics touches upon the topic of biases regarding race
+and gender. They apply knowledge from well-known psychological studies
+like Implicit Association Test to research the relation between human
+stereotypical thinking and model learnt biases to discover close
+relationship between these two. For the evaluation they use Word
+Embedding Association Test (WEAT) and the Word Embedding Factual
+Association Test (WEFAT).
 
-manzini2019black proposes novel way of using a cosine similarity method to get the information on assumed resemblance between words. They investigate 
-an approach that enables to measure the bias for a class (like gender, religion, race) and express the final result with only one averaged number.
-\newline
-
-<!-- Nissim2019Fair -->
+manzini2019black proposes novel way of using a cosine similarity method
+to get the information on assumed resemblance between words. They
+investigate an approach that enables to measure the bias for a class
+(like gender, religion, race) and express the final result with only one
+averaged number. \newline
 
@@
 
-It is worth noticing the general distinction of biases mentioned in Caliskan2017Semantics. They refer to the publication concerning Implicit Association
-Test Greenwald et al., 1998) where certain baseline of bias phenomenon was introduced. Namely it seems that humans naturally exhibit some biases and
-that not always bring social concern.
-One can imagine the intuitive associations between for example insects and flowers, and the feelings of pleasantness or unpleasantness.
-In general people would rather associate flowers with feeling pleasant than insects and this preference could be named a bias or
-prejudice in some direction. However this type of preference does not cause an uproar and it is rather morally neutral case. Unfortunately there are other
-biases and prejudices that directly influence the quality of other people's lives and therefore they should be taken care of.
+It is worth noticing the general distinction of biases mentioned in
+Caliskan2017Semantics. They refer to the publication concerning Implicit
+Association Test Greenwald et al., 1998) where certain baseline of bias
+phenomenon was introduced. Namely it seems that humans naturally exhibit
+some biases and that not always bring social concern. One can imagine
+the intuitive associations between for example insects and flowers, and
+the feelings of pleasantness or unpleasantness. In general people would
+rather associate flowers with feeling pleasant than insects and this
+preference could be named a bias or prejudice in some direction. However
+this type of preference does not cause an uproar and it is rather
+morally neutral case. Unfortunately there are other biases and
+prejudices that directly influence the quality of other people's lives
+and therefore they should be taken care of.
 
-One can find a bunch of various definitions trying to capture what bias and fairness actually are.
-With the choice of the definition, implications into the real-life applications may change as well as it was pointed out in Mehrabi2019Survey.
-They mark out that there exist different types of biases, the list if long but among others there are historical bias, representation bias, measurement bias.
-This indicates how complex the process itself is.
-Without the proper understanding and awareness of the problem, people are prone to unconsciously sustain this phenomenon.
+One can find a bunch of various definitions trying to capture what bias
+and fairness actually are. With the choice of the definition,
+implications into the real-life applications may change as well as it
+was pointed out in Mehrabi2019Survey. They mark out that there exist
+different types of biases, the list if long but among others there are
+historical bias, representation bias, measurement bias. This indicates
+how complex the process itself is. Without the proper understanding and
+awareness of the problem, people are prone to unconsciously sustain this
+phenomenon.
 
-In the article one can also find the distinction on different types of discrimination, some of them will be shortly described.
-It is worth first mentioning that protected attributes are those qualities, traits or characteristics that one cannot, according to the law, discriminated against.
-Direct discrimination refers to the situation when protected attributes of individuals explicitly result in non-favorable outcomes toward them.
-In contrast in indirect discrimination individuals appear to be treated equally but anyway they end up being treated unjustly due to the hidden effects of biases towards their protected attributes.
-Systemic discrimination takes place when policies, customs or behaviors that result from certain culture or organizational structure lead to discrimination against some groups of people.
-Finally, very common statistical discrimination refers to using
-average group statistics to judge person belonging to the group.
+In the article one can also find the distinction on different types of
+discrimination, some of them will be shortly described. It is worth
+first mentioning that protected attributes are those qualities, traits
+or characteristics that one cannot, according to the law, discriminated
+against. Direct discrimination refers to the situation when protected
+attributes of individuals explicitly result in non-favorable outcomes
+toward them. In contrast in indirect discrimination individuals appear
+to be treated equally but anyway they end up being treated unjustly due
+to the hidden effects of biases towards their protected attributes.
+Systemic discrimination takes place when policies, customs or behaviors
+that result from certain culture or organizational structure lead to
+discrimination against some groups of people. Finally, very common
+statistical discrimination refers to using average group statistics to
+judge person belonging to the group.
 
-The topic of discrimination is entangled with another concept which is fairness. It is essential to grasp some concepts of fairness to take them into
-consideration while designing implementation of some machine learning model. In Mehrabi2019Survey one may notice that depending on the context and application different definitions may be applied.
+The topic of discrimination is entangled with another concept which is
+fairness. It is essential to grasp some concepts of fairness to take
+them into consideration while designing implementation of some machine
+learning model. In Mehrabi2019Survey one may notice that depending on
+the context and application different definitions may be applied.
 \newline
 
-\textbf{outline the current situation}
-The most popular methods focus on comparing the similarity between words from protected groups and those that are considered to be stereotypical or harmful in some way.
-One can find in this group methods such as euclidean distance, dot product or cosine similarity.
-There are also other ways to detect the effects of biases.
-For example through the investigation of the model performance on certain tasks that validate if the model returns some values
-independently on gender or race or not. 
-\newline
+\textbf{outline the current situation} The most popular methods focus on
+comparing the similarity between words from protected groups and those
+that are considered to be stereotypical or harmful in some way. One can
+find in this group methods such as euclidean distance, dot product or
+cosine similarity. There are also other ways to detect the effects of
+biases. For example through the investigation of the model performance
+on certain tasks that validate if the model returns some values
+independently on gender or race or not. \newline
 
@@
 
-In the currently used methods (like cosine similarity) the values of similarity are often aggregated in a way that may lead to false conclusions. For 
-example due to the averaging of values and the lack of confidence interval information. 
-\newline
+In the currently used methods (like cosine similarity) the values of
+similarity are often aggregated in a way that may lead to false
+conclusions. For example due to the averaging of values and the lack of
+confidence interval information. \newline
 
-\textbf{identify the importance of the proposed research}
-One can find a number of articles on negative real-life implications resulting from the presence of unaddressed biases in the machine learning models.
-\newline
+\textbf{identify the importance of the proposed research} One can find a
+number of articles on negative real-life implications resulting from the
+presence of unaddressed biases in the machine learning models. \newline
 
@@
 
-In the paper we indicate how current methods used to detect biases in natural language models are limited in terms of confidence interval.
+In the paper we indicate how current methods used to detect biases in
+natural language models are limited in terms of confidence interval.
 \newline
@@
 
-Our research tries to answer the question of how to enhance the current way in which the bias detection is performed to make sure that it is
-methodologically valid.
-\newline
+Our research tries to answer the question of how to enhance the current
+way in which the bias detection is performed to make sure that it is
+methodologically valid. \newline
 
@@
 
-Our hypothesis is that there can be greater understanding of data and bias implications when confidence interval and Bayesian method are applied to the
-methodology.
-\newline
+Our hypothesis is that there can be greater understanding of data and
+bias implications when confidence interval and Bayesian method are
+applied to the methodology. \newline
 <!-- \textbf{outline the order of information in the thesis} -->
@@
 To discuss!
+
+\end{document}

### 2021-06-08 thesis/index.Rmd
datasets upload

---
title: "Conceptual and methodological problems with bias detection and avoidance in natural language processing"
author: "Alicja Dobrzeniecka"
date: '`r Sys.Date()`'
output:
  bookdown::pdf_book: 
    keep_tex: yes
    toc_depth: 6
    latex_engine: xelatex
  bookdown::gitbook:
    lib_dir: "book_assets"
includes:
    in_header:
      - Rafal_latex6.sty
documentclass: book
classoptions: dvipsnames, usenames
site: bookdown::bookdown_site
bibliography: [../references/bibliographyRafal.bib]
biblio-style: 
link-citations: yes
csl: apa-5th-edition.csl
---

<!-- github-repo: rstudio/bookdown-demo -->

```{r setup, include=FALSE, echo=FALSE}
library(tidyverse)
library(knitr)

options(kableExtra.latex.load_packages = FALSE)
library(kableExtra)
```

# Preface {-}

Test

<!-- testsds -->

<!-- \begin{align}  -->
<!--   f\left(k\right) = \binom{n}{k} p^k\left(1-p\right)^{n-k} -->
<!--   (\#eq:binom) -->
<!-- \end{align}  -->

<!-- This is a citation [@diamond90] which uses keys from the bib file listed in the preamble. -->

<!-- Equation \@ref(eq:binom)^[This is a footnote containing a double citation [@diamond90; @dahlmanNakedStat2020; @Zhao2019Gender].] -->

<!-- Note that chapter files are found and compiled automatically, but the file names have to contain chapter numbers first. For instance, we used `01-intro.Rmd`, placed in the same folder. Observe how we included r code inline. -->

<!-- ```{r fig-margin, fig.margin=TRUE} -->
<!-- plot(cars) -->
<!-- ``` -->

### 2021-06-08 thesis/03-bayesian.Rmd
started adding chapter 3

# Bayesian analysis of cosine-based bias

We start with loading the libraries needed for the analysis.

\footnotesize
```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%", warning = FALSE, message = FALSE}
library(ggplot2)
library(ggthemes)
library(rethinking)
library(tidyverse)
library(ggpubr)
library(kableExtra)
library(dplyr)
library(ggExtra)
library(cowplot)
```
\normalsize

# Walkthrough with the religion dataset

We will use the choice of protected words and stereotypical predicates used in REF. This is a decent point of departure,  not only we want to compare our method to that of REF, but also because this data format is fairly general (as contrasted, say, with a set up for binary stereotypes). Note also that the method we develop here  can fairly easily be  run  for different stereotypization patterns.  Let's start  with  explaining the method and its deployment using a dataset obtained for the religion-related protected words.

Let's load, clean a bit and inspect the head of the religion dataset we prepared.  In order to obtain this  dataset, we  calculated the cosine distance between each protected word and each word from both the bias-related attribute groups, which were used in the original study, and to neutral and human control attributes which we added  as control groups.  For instance, for religion, the bias-related predicates (coming from the original study in REF) include muslim bias attributes, jew bias attributes, christian bias attributes (see a list in the APPENDIX).

We decided to add control groups in the form of two classes --- neutral words and human-related words. Without a proper control group it is quite hard to compare the resulting cosine distances and decide on their significance in bias detection. We prepared approximately 300\todo{FIX later} more or less neutral words to double-check the prima-facie neutral hypothesis that their cosine similarity to the protected words will oscillate around 0 (that is, the distances will be around 1). This provides us with a  more reliable point of reference. Moreover, we added human attributes that are associated with people in general to investigate  whether the smaller cosine distance between protected words and stereotypes can result  simply from the fact that the stereotype predicates  are associated with humans. For two control groups, we have randomly drawn 300 words that do not express any property usually attributed to humans, and human attributes. \todo{describe once the words are selected}

<!-- In other words, we took each protected word and in nested iteration calculated all of the cosine distances between the word and all available harmful stereotypes, neutral words, and human ones, which resulted with approximately 333 cosine distances per each protected word. -->

\vspace{1mm}
\footnotesize

```{r religionTableHead,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
religion <- read.csv("../datasets/religionReddit.csv")[-1]
colnames(religion) <- c("protectedWord","wordToCompare","wordClass",
                        "cosineDistance","cosineSimilarity","connection")
levels(religion$wordClass) <- c("christian","human","jewish","muslim","neutral")
head(religion)  %>%  kable(format = "latex",booktabs=T,
                      linesep = "",  escape = FALSE, 
                      caption = "Head of the religion dataset.") %>%
                      kable_styling(latex_options=c("scale_down"))
```
\normalsize

 The `protectedWord` column contains words from a protected class that (in a perfect world according to the assumptions of the orignal study) should not be associated with  harmful stereotypes. `wordToCompare` contains attributes, including stereotypes and control group words. For each row we compute the cosine distances between a given protected  word and a given attribute word. `wordClass` tells us  which class an attribute is supposed to be stereotypically associated with, that is, whether the  word from `wordToCompare` is associated stereotypically with jews, christians or muslims, or whether it belongs to a control group. `cosineDistance` is simply a calculation of the cosine distance between protected word and atrribute.  `cosineSimilarity`  contains the result of substracting cosine distance from 1. `connection` contains information about  the relation type between a protected word and an attribute. If the attribute is e.g. a harmful jewish stereotype and the protected word is also from the judaism group,  the connection has value `associated`.  If the attribute is still stereotypically jewish, but the protected word comes from another religion, the connection is labelled as  `different`. If the attribute belongs to  a neutral group then the connection is labelled as `none` and if an attribute belongs to the `human` class, then the connection is labelled as `human`.

First let's take a look at the empirical distribution of distances by the connection type, initially ignoring the human control class for now.

\vspace{1mm}
\footnotesize
```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
religionNoHumans <- religion[religion$connection !="human",]

ggplot(religionNoHumans, aes(x =  cosineDistance, fill = connection, color = connection ))+geom_density(alpha=0.6,size = .2)+theme_tufte()+ggtitle("Empirical distribution of cosine distances (religion), no human attributes.")+ scale_fill_manual(values = c("orangered4","chartreuse4","gray"))+scale_x_continuous(breaks = seq(0.3,1.5, by = 0.1))+xlab("cosine distance")+ scale_color_manual(values = c("orangered4","chartreuse4","gray"))
```
\normalsize

The first impression is that while  there is a  shift for  associated words towards smaller cosine distances as compared to the neutral words, slightly surprisingly a slightly weaker shift in the same direction is visible for attributes associated with different stereotypes. Moreover,  the empirical distributions overlap to alarge extent and the  means grouped by connection type do not seem too far from each other. In fact, as there is a lot of  variety in the consine distances (as we will soon see), abd we need to gauge the uncertaintly involved, and  to look more carefully at individual protected words to get a better idea of how the cosine distance distribution changes for different attribute groups and different protected classes.   Now, let's add the human attributes to the picture:

\vspace{1mm}
\footnotesize
```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
ggplot(religion, aes(x =  cosineDistance, fill = connection, color = connection))+geom_density(alpha=0.6,size = .2)+theme_tufte()+ggtitle("Empirical distribution of cosine distances (religion)")+ scale_fill_manual(values = c("orangered4","chartreuse4", "skyblue", "gray"))+scale_x_continuous(breaks = seq(0.3,1.5, by = 0.1))+xlab("cosine distance")+ scale_color_manual(values = c("orangered4","chartreuse4","skyblue","gray"))
```
\normalsize

\noindent Notice that the distribution for `human` (even though we did our best not to include in it any stereotype-related atributes) is left-skewed, with much overlap with `associated` and `different`, which illustrates the need to take being associated with humans as an important predictor.

Our focus lies in  `connection` as a predictor. Morever, later on we'll be interested in looking at the protected words separately, and at protected words split by connection. For technical reasons it is useful to represent these factors as integer vectors.

\vspace{1mm}
\footnotesize
```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
religion$con <- as.integer(religion$connection)
religion$pw <- as.integer(religion$protectedWord)
religion$pwFactor <- factor(paste0(religion$protectedWord, religion$connection))
religion$pwIndex <- as.integer(religion$pwFactor)
```
\normalsize

<!-- For technical reasons it is useful to represent an $n$-level factor with $n-1$ binary vectors. `connection` in this case has four levels. -->

<!-- \vspace{1mm} -->
<!-- \footnotesize -->
<!-- ```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"} -->
<!-- religion$associated <- ifelse(religion$connection == "associated", 1, 0) -->
<!-- religion$different <- ifelse(religion$connection == "different", 1, 0) -->
<!-- religion$human <- ifelse(religion$connection == "human" |religion$connection == "associated"| -->
<!--                            religion$connection == "different", 1, 0)   -->
<!-- ```   -->
<!-- \normalsize -->

A short script, `cleanDataset` to make this faster, so equivalently:

\vspace{1mm}
\footnotesize
```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
source("../functions/cleanDataset.R")
religion <- read.csv("../datasets/religionReddit.csv")[-1]
religion <- cleanDataset(religion,c("christian","human","jewish","muslim","neutral"))
```
\normalsize

For now, let's focus on five protected words related to islam ("imam", "islam", "mosque", "muslim", and "quran"). The word list associates with islam  four stereotypical attributes ("violent", "terrorist", "uneducated" and "dirty").  First, we select and plot the empirical distributions for these protected words.

\vspace{1mm}
\footnotesize
```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%",message =FALSE, warning = FALSE}
library(tidyverse)
muslimWords <- c("imam","islam","mosque","muslim","quran")
muslim <- religion %>% filter(protectedWord %in% muslimWords)
ggplot(muslim, aes(x =  cosineDistance, fill = connection, color = connection))+
  geom_density(alpha=0.6,size = .2)+ 
  scale_fill_manual(values = c("orangered4","chartreuse4", "skyblue", "gray"))+
  scale_x_continuous(breaks = seq(0.3,1.5, by = 0.1))+xlab("cosine distance")+
  scale_color_manual(values = c("orangered4","chartreuse4","skyblue","gray"))+
  theme_tufte()+ggtitle("Empirical distribution of distances (muslim)")
```
\normalsize

\noindent Once we focus on words related to islam, the associated bias seems to be stronger than in the whole dataset.  This is a  step towards illustrating that the distribution of bias is uneven.

Now, say we want to look at a single protected word. Since the dataset also contains comparison multiple control neutral and  human attributes, we randomly select only 5 from `none` and 5 from `human` control groups of those for the visualisation purposes.

\vspace{1mm}
\footnotesize
```{r tableMuslimActive,echo=TRUE,eval=TRUE,fig.align = "center",cache=FALSE, fig.show = "hold", out.width = "100%"}
library(tidyverse)
muslimClass <- muslim %>% filter(protectedWord == "muslim")
neutralSample <- sample_n(filter(muslimClass,connection == "none"), 5)
humanSample <- sample_n(filter(muslimClass,connection == "human"), 5)
muslimVis <- muslimClass %>% filter(connection != "none" & connection !="human")
muslimVis <- rbind(muslimVis,neutralSample,humanSample)

#we plug in our visualisation script
source("../functions/visualisationTools.R")
#two arguments: dataset and protected word
visualiseProtected(muslimVis,"muslim")
```
\normalsize

Note that the distance between the grey point and the other points is proportional to cosine distance, the non-grey point size is proportional to cosine similarity to the protected word, and color groups by the connection type.  So for `muslim` it seems that the stereotypes coming from the word list are fairly well visible. To give you some taste of how uneven the dataset is, compare this to what happens with `priest`.

\vspace{1mm}
\footnotesize
```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
library(tidyverse)
priestClass <- religion %>% filter(protectedWord == "priest")
neutralSample <- sample_n(filter(priestClass,connection == "none"), 5)
humanSample <- sample_n(filter(priestClass,connection == "human"), 5)
priestVis <- priestClass %>% filter(connection != "none" & connection !="human")
priestVis <- rbind(priestVis,neutralSample,humanSample)

#we plug in our visualisation script
source("../functions/visualisationTools.R")
#two arguments: dataset and protected word
visualiseProtected(priestVis,"priest")
```
\normalsize

\noindent Here you can see that some human attributes are closer than stereotype attributes, and that there is no clear reason to claim that `associated` attributes are closer than `different`  or `human` attributes. This, again, illustrates the need of case-by-case analysis with  control groups.

The general idea now is that the word lists provided in different pieces of research are just samples of attributes associates with various stereotypes and should be treated as such: the uncertaintly involved and the sample sizes should have clear impact on our estimates.

We will now think of cosine distance as the output variable, and 
will build a few bayesian models to compare. First, we just build a baseline model which estimates cosine distance to the attributes separately for each protected word. The underlying idea is that different protected words migh in general have different relations to all the attributes and this relations should be our point of departure.

Here is the intution behind the mathematical Bayesian model involved. Our outcome variable is `cosine difference`, which we take to me normally distributed around the predicted mean for a given protected word (that is, we assume the residuals are normally distributed). The simplest  model specification is:

\begin{align}
cosineDistance_i  & \sim dnorm(\mu_i, \sigma) \\
\mu_i & = m_{pw} \\
m_{pw} & ~ dnorm(1,.5) \\
\sigma &\sim  dcauchy(0,1)
\end{align}

That is, we assume the estimated means might be different for diferent protected words and our prior for the mean and the overal standard deviation are normal with mean 1 and sd=.5 and half-cauchy with parameters `0,1`. Further on we'll also estimate additional impact the connection type may have. For this impact we take a slightly skeptical prior centered around 0 distributed normally with sd = 1. These are fairly weak and slightly skeptical regularizing priors, which can be illustrated as follows:

\vspace{2mm}

```{r priorsVis,echo=FALSE,eval=TRUE,fig.align = "center",cache=FALSE, fig.show = "hold", out.width = "100%", fig.caption = "Regularizing priors for the bayesian models."}
library(ggpubr)
sim <- seq(-1,1,by = 0.001)
dis <- 1-sim

p <- ggplot(data = data.frame(distance = dis), 
            mapping = aes(x = distance))+theme_tufte()

none <- function(x) dnorm(x,1,.5)
cau <- function(x) dcauchy(x,0,1)
par <- function(x) dnorm(x,0,1)

noneG <- p + stat_function(fun = none)+xlim(c(-0.5,2.5))+ggtitle("Prior for mean distances")
parG <- p + stat_function(fun = par)+xlim(c(-2,2))+xlab("distance change")+ggtitle("Prior for coefficients")
cauG <- p + stat_function(fun = cau)+xlim(c(0,2))+ggtitle("Prior for standard deviation")+xlab("distance change")
ggarrange(noneG,cauG, parG, ncol = 1)
```

Now we can define and compile the baseline model. Its  parameters will have a posterior distribution obtained using either Hamiltionian Monte Carlo methods (STAN) available through the `rethinking` package.

### 2021-06-08 thesis/_book/_main.tex
started adding chapter 3

@@
 \urlstyle{same}  % don't use monospace font for urls
+\usepackage{color}
+\usepackage{fancyvrb}
+\newcommand{\VerbBar}{|}
+\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+% Add ',fontsize=\small' for more characters per line
+\usepackage{framed}
+\definecolor{shadecolor}{RGB}{248,248,248}
+\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\ImportTok}[1]{#1}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
+\newcommand{\BuiltInTok}[1]{#1}
+\newcommand{\ExtensionTok}[1]{#1}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
+\newcommand{\RegionMarkerTok}[1]{#1}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
+\newcommand{\NormalTok}[1]{#1}
 \usepackage{longtable,booktabs}
@@
 \makeatother
+
+\usepackage{todonotes}
 
@@
 
+\chapter{Bayesian analysis of cosine-based
+bias}\label{bayesian-analysis-of-cosine-based-bias}
+
+We start with loading the libraries needed for the analysis.
+
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(ggplot2)}
+\KeywordTok{library}\NormalTok{(ggthemes)}
+\KeywordTok{library}\NormalTok{(rethinking)}
+\KeywordTok{library}\NormalTok{(tidyverse)}
+\KeywordTok{library}\NormalTok{(ggpubr)}
+\KeywordTok{library}\NormalTok{(kableExtra)}
+\KeywordTok{library}\NormalTok{(dplyr)}
+\KeywordTok{library}\NormalTok{(ggExtra)}
+\KeywordTok{library}\NormalTok{(cowplot)}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+\chapter{Walkthrough with the religion
+dataset}\label{walkthrough-with-the-religion-dataset}
+
+We will use the choice of protected words and stereotypical predicates
+used in REF. This is a decent point of departure, not only we want to
+compare our method to that of REF, but also because this data format is
+fairly general (as contrasted, say, with a set up for binary
+stereotypes). Note also that the method we develop here can fairly
+easily be run for different stereotypization patterns. Let's start with
+explaining the method and its deployment using a dataset obtained for
+the religion-related protected words.
+
+Let's load, clean a bit and inspect the head of the religion dataset we
+prepared. In order to obtain this dataset, we calculated the cosine
+distance between each protected word and each word from both the
+bias-related attribute groups, which were used in the original study,
+and to neutral and human control attributes which we added as control
+groups. For instance, for religion, the bias-related predicates (coming
+from the original study in REF) include muslim bias attributes, jew bias
+attributes, christian bias attributes (see a list in the APPENDIX).
+
+We decided to add control groups in the form of two classes --- neutral
+words and human-related words. Without a proper control group it is
+quite hard to compare the resulting cosine distances and decide on their
+significance in bias detection. We prepared approximately
+300\todo{FIX later} more or less neutral words to double-check the
+prima-facie neutral hypothesis that their cosine similarity to the
+protected words will oscillate around 0 (that is, the distances will be
+around 1). This provides us with a more reliable point of reference.
+Moreover, we added human attributes that are associated with people in
+general to investigate whether the smaller cosine distance between
+protected words and stereotypes can result simply from the fact that the
+stereotype predicates are associated with humans. For two control
+groups, we have randomly drawn 300 words that do not express any
+property usually attributed to humans, and human attributes.
+\todo{describe once the words are selected}
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{religion <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
+\KeywordTok{colnames}\NormalTok{(religion) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
+                        \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
+\KeywordTok{levels}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
+\KeywordTok{head}\NormalTok{(religion)  }\OperatorTok{%>%}\StringTok{  }\KeywordTok{kable}\NormalTok{(}\DataTypeTok{format =} \StringTok{"latex"}\NormalTok{,}\DataTypeTok{booktabs=}\NormalTok{T,}
+                      \DataTypeTok{linesep =} \StringTok{""}\NormalTok{,  }\DataTypeTok{escape =} \OtherTok{FALSE}\NormalTok{, }
+                      \DataTypeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\OperatorTok{%>%}
+\StringTok{                      }\KeywordTok{kable_styling}\NormalTok{(}\DataTypeTok{latex_options=}\KeywordTok{c}\NormalTok{(}\StringTok{"scale_down"}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{table}
+
+\caption{\label{tab:religionTableHead}Head of the religion dataset.}
+\centering
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{lllrrl}
+\toprule
+protectedWord & wordToCompare & wordClass & cosineDistance & cosineSimilarity & connection\\
+\midrule
+judaism & violent & muslim & 0.7141939 & 0.2858061 & different\\
+judaism & terrorist & muslim & 0.7461333 & 0.2538667 & different\\
+judaism & dirty & muslim & 1.2002599 & -0.2002599 & different\\
+judaism & uneducated & muslim & 0.7885469 & 0.2114531 & different\\
+judaism & greedy & jewish & 1.0026172 & -0.0026172 & associated\\
+judaism & cheap & jewish & 1.2323229 & -0.2323229 & associated\\
+\bottomrule
+\end{tabular}}
+\end{table}
+
+\normalsize
+
+The \texttt{protectedWord} column contains words from a protected class
+that (in a perfect world according to the assumptions of the orignal
+study) should not be associated with harmful stereotypes.
+\texttt{wordToCompare} contains attributes, including stereotypes and
+control group words. For each row we compute the cosine distances
+between a given protected word and a given attribute word.
+\texttt{wordClass} tells us which class an attribute is supposed to be
+stereotypically associated with, that is, whether the word from
+\texttt{wordToCompare} is associated stereotypically with jews,
+christians or muslims, or whether it belongs to a control group.
+\texttt{cosineDistance} is simply a calculation of the cosine distance
+between protected word and atrribute. \texttt{cosineSimilarity} contains
+the result of substracting cosine distance from 1. \texttt{connection}
+contains information about the relation type between a protected word
+and an attribute. If the attribute is e.g.~a harmful jewish stereotype
+and the protected word is also from the judaism group, the connection
+has value \texttt{associated}. If the attribute is still stereotypically
+jewish, but the protected word comes from another religion, the
+connection is labelled as \texttt{different}. If the attribute belongs
+to a neutral group then the connection is labelled as \texttt{none} and
+if an attribute belongs to the \texttt{human} class, then the connection
+is labelled as \texttt{human}.
+
+First let's take a look at the empirical distribution of distances by
+the connection type, initially ignoring the human control class for now.
+
+\vspace{1mm} \footnotesize
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-2-1} \end{center}
+
+\normalsize
+
+The first impression is that while there is a shift for associated words
+towards smaller cosine distances as compared to the neutral words,
+slightly surprisingly a slightly weaker shift in the same direction is
+visible for attributes associated with different stereotypes. Moreover,
+the empirical distributions overlap to alarge extent and the means
+grouped by connection type do not seem too far from each other. In fact,
+as there is a lot of variety in the consine distances (as we will soon
+see), abd we need to gauge the uncertaintly involved, and to look more
+carefully at individual protected words to get a better idea of how the
+cosine distance distribution changes for different attribute groups and
+different protected classes. Now, let's add the human attributes to the
+picture:
+
+\vspace{1mm} \footnotesize
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-3-1} \end{center}
+
+\normalsize
+
+\noindent Notice that the distribution for \texttt{human} (even though
+we did our best not to include in it any stereotype-related atributes)
+is left-skewed, with much overlap with \texttt{associated} and
+\texttt{different}, which illustrates the need to take being associated
+with humans as an important predictor.
+
+Our focus lies in \texttt{connection} as a predictor. Morever, later on
+we'll be interested in looking at the protected words separately, and at
+protected words split by connection. For technical reasons it is useful
+to represent these factors as integer vectors.
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{religion}\OperatorTok{$}\NormalTok{con <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(religion}\OperatorTok{$}\NormalTok{connection)}
+\NormalTok{religion}\OperatorTok{$}\NormalTok{pw <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(religion}\OperatorTok{$}\NormalTok{protectedWord)}
+\NormalTok{religion}\OperatorTok{$}\NormalTok{pwFactor <-}\StringTok{ }\KeywordTok{factor}\NormalTok{(}\KeywordTok{paste0}\NormalTok{(religion}\OperatorTok{$}\NormalTok{protectedWord, religion}\OperatorTok{$}\NormalTok{connection))}
+\NormalTok{religion}\OperatorTok{$}\NormalTok{pwIndex <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(religion}\OperatorTok{$}\NormalTok{pwFactor)}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+A short script, \texttt{cleanDataset} to make this faster, so
+equivalently:
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{source}\NormalTok{(}\StringTok{"../functions/cleanDataset.R"}\NormalTok{)}
+\NormalTok{religion <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
+\NormalTok{religion <-}\StringTok{ }\KeywordTok{cleanDataset}\NormalTok{(religion,}\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+For now, let's focus on five protected words related to islam (``imam'',
+``islam'', ``mosque'', ``muslim'', and ``quran''). The word list
+associates with islam four stereotypical attributes (``violent'',
+``terrorist'', ``uneducated'' and ``dirty''). First, we select and plot
+the empirical distributions for these protected words.
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(tidyverse)}
+\NormalTok{muslimWords <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"imam"}\NormalTok{,}\StringTok{"islam"}\NormalTok{,}\StringTok{"mosque"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"quran"}\NormalTok{)}
+\NormalTok{muslim <-}\StringTok{ }\NormalTok{religion }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(protectedWord }\OperatorTok{%in%}\StringTok{ }\NormalTok{muslimWords)}
+\KeywordTok{ggplot}\NormalTok{(muslim, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{  cosineDistance, }\DataTypeTok{fill =}\NormalTok{ connection, }\DataTypeTok{color =}\NormalTok{ connection))}\OperatorTok{+}
+\StringTok{  }\KeywordTok{geom_density}\NormalTok{(}\DataTypeTok{alpha=}\FloatTok{0.6}\NormalTok{,}\DataTypeTok{size =} \FloatTok{.2}\NormalTok{)}\OperatorTok{+}\StringTok{ }
+\StringTok{  }\KeywordTok{scale_fill_manual}\NormalTok{(}\DataTypeTok{values =} \KeywordTok{c}\NormalTok{(}\StringTok{"orangered4"}\NormalTok{,}\StringTok{"chartreuse4"}\NormalTok{, }\StringTok{"skyblue"}\NormalTok{, }\StringTok{"gray"}\NormalTok{))}\OperatorTok{+}
+\StringTok{  }\KeywordTok{scale_x_continuous}\NormalTok{(}\DataTypeTok{breaks =} \KeywordTok{seq}\NormalTok{(}\FloatTok{0.3}\NormalTok{,}\FloatTok{1.5}\NormalTok{, }\DataTypeTok{by =} \FloatTok{0.1}\NormalTok{))}\OperatorTok{+}\KeywordTok{xlab}\NormalTok{(}\StringTok{"cosine distance"}\NormalTok{)}\OperatorTok{+}
+\StringTok{  }\KeywordTok{scale_color_manual}\NormalTok{(}\DataTypeTok{values =} \KeywordTok{c}\NormalTok{(}\StringTok{"orangered4"}\NormalTok{,}\StringTok{"chartreuse4"}\NormalTok{,}\StringTok{"skyblue"}\NormalTok{,}\StringTok{"gray"}\NormalTok{))}\OperatorTok{+}
+\StringTok{  }\KeywordTok{theme_tufte}\NormalTok{()}\OperatorTok{+}\KeywordTok{ggtitle}\NormalTok{(}\StringTok{"Empirical distribution of distances (muslim)"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-6-1} \end{center}
+
+\normalsize
+
+\noindent Once we focus on words related to islam, the associated bias
+seems to be stronger than in the whole dataset. This is a step towards
+illustrating that the distribution of bias is uneven.
+
+Now, say we want to look at a single protected word. Since the dataset
+also contains comparison multiple control neutral and human attributes,
+we randomly select only 5 from \texttt{none} and 5 from \texttt{human}
+control groups of those for the visualisation purposes.
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(tidyverse)}
+\NormalTok{muslimClass <-}\StringTok{ }\NormalTok{muslim }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(protectedWord }\OperatorTok{==}\StringTok{ "muslim"}\NormalTok{)}
+\NormalTok{neutralSample <-}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\KeywordTok{filter}\NormalTok{(muslimClass,connection }\OperatorTok{==}\StringTok{ "none"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
+\NormalTok{humanSample <-}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\KeywordTok{filter}\NormalTok{(muslimClass,connection }\OperatorTok{==}\StringTok{ "human"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
+\NormalTok{muslimVis <-}\StringTok{ }\NormalTok{muslimClass }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(connection }\OperatorTok{!=}\StringTok{ "none"} \OperatorTok{&}\StringTok{ }\NormalTok{connection }\OperatorTok{!=}\StringTok{"human"}\NormalTok{)}
+\NormalTok{muslimVis <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(muslimVis,neutralSample,humanSample)}
+
+\CommentTok{#we plug in our visualisation script}
+\KeywordTok{source}\NormalTok{(}\StringTok{"../functions/visualisationTools.R"}\NormalTok{)}
+\CommentTok{#two arguments: dataset and protected word}
+\KeywordTok{visualiseProtected}\NormalTok{(muslimVis,}\StringTok{"muslim"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/tableMuslimActive-1} \end{center}
+
+\normalsize
+
+Note that the distance between the grey point and the other points is
+proportional to cosine distance, the non-grey point size is proportional
+to cosine similarity to the protected word, and color groups by the
+connection type. So for \texttt{muslim} it seems that the stereotypes
+coming from the word list are fairly well visible. To give you some
+taste of how uneven the dataset is, compare this to what happens with
+\texttt{priest}.
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(tidyverse)}
+\NormalTok{priestClass <-}\StringTok{ }\NormalTok{religion }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(protectedWord }\OperatorTok{==}\StringTok{ "priest"}\NormalTok{)}
+\NormalTok{neutralSample <-}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\KeywordTok{filter}\NormalTok{(priestClass,connection }\OperatorTok{==}\StringTok{ "none"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
+\NormalTok{humanSample <-}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\KeywordTok{filter}\NormalTok{(priestClass,connection }\OperatorTok{==}\StringTok{ "human"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
+\NormalTok{priestVis <-}\StringTok{ }\NormalTok{priestClass }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(connection }\OperatorTok{!=}\StringTok{ "none"} \OperatorTok{&}\StringTok{ }\NormalTok{connection }\OperatorTok{!=}\StringTok{"human"}\NormalTok{)}
+\NormalTok{priestVis <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(priestVis,neutralSample,humanSample)}
+
+\CommentTok{#we plug in our visualisation script}
+\KeywordTok{source}\NormalTok{(}\StringTok{"../functions/visualisationTools.R"}\NormalTok{)}
+\CommentTok{#two arguments: dataset and protected word}
+\KeywordTok{visualiseProtected}\NormalTok{(priestVis,}\StringTok{"priest"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-7-1} \end{center}
+
+\normalsize
+
+\noindent Here you can see that some human attributes are closer than
+stereotype attributes, and that there is no clear reason to claim that
+\texttt{associated} attributes are closer than \texttt{different} or
+\texttt{human} attributes. This, again, illustrates the need of
+case-by-case analysis with control groups.
+
+The general idea now is that the word lists provided in different pieces
+of research are just samples of attributes associates with various
+stereotypes and should be treated as such: the uncertaintly involved and
+the sample sizes should have clear impact on our estimates.
+
+We will now think of cosine distance as the output variable, and will
+build a few bayesian models to compare. First, we just build a baseline
+model which estimates cosine distance to the attributes separately for
+each protected word. The underlying idea is that different protected
+words migh in general have different relations to all the attributes and
+this relations should be our point of departure.
+
+Here is the intution behind the mathematical Bayesian model involved.
+Our outcome variable is \texttt{cosine\ difference}, which we take to me
+normally distributed around the predicted mean for a given protected
+word (that is, we assume the residuals are normally distributed). The
+simplest model specification is:
+
+\begin{align}
+cosineDistance_i  & \sim dnorm(\mu_i, \sigma) \\
+\mu_i & = m_{pw} \\
+m_{pw} & ~ dnorm(1,.5) \\
+\sigma &\sim  dcauchy(0,1)
+\end{align}
+
+That is, we assume the estimated means might be different for diferent
+protected words and our prior for the mean and the overal standard
+deviation are normal with mean 1 and sd=.5 and half-cauchy with
+parameters \texttt{0,1}. Further on we'll also estimate additional
+impact the connection type may have. For this impact we take a slightly
+skeptical prior centered around 0 distributed normally with sd = 1.
+These are fairly weak and slightly skeptical regularizing priors, which
+can be illustrated as follows:
+
+\vspace{2mm}
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/priorsVis-1} \end{center}
+
+Now we can define and compile the baseline model. Its parameters will
+have a posterior distribution obtained using either Hamiltionian Monte
+Carlo methods (STAN) available through the \texttt{rethinking} package.
+
 \end{document}

### 2021-06-08 thesis/index.Rmd
started adding chapter 3

@@
     latex_engine: xelatex
+    extra_dependencies: ["todonotes"]
   bookdown::gitbook:

### 2021-06-08 functions/checkMAPweights.R

@@ -1,0 +1,113 @@
+library(ggplot2)
+library(ggthemes)
+library(rethinking)
+
+
+
+setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
+religion <- read.csv("../datasets/religion_cosine_table.csv")
+
+
+dataset <- religion
+
+dataset$associated <- ifelse(dataset$connection == "associated", 1, 0)
+dataset$different <- ifelse(dataset$connection == "different", 1, 0)
+
+#build models
+null  <- map(
+  alist(
+    cosine_distance ~ dnorm(mu,sigma),
+    mu ~  dnorm(1,.5),
+    sigma  ~ dcauchy(0,1)
+  ),
+  data = dataset,
+  start=list(mu = 1, sigma= .3)
+)
+
+
+associated  <- map(
+  alist(
+    cosine_distance ~ dnorm(mu,sigma),
+    mu <- no + a * associated,
+    no ~ dnorm(1,.5),
+    a ~ dnorm(0,1),
+    sigma ~ dcauchy(0,1)
+  ),
+  data = dataset,
+  start=list(no = 1, a = 0, sigma= .3)
+)
+
+
+type  <- map(
+  alist(
+    cosine_distance ~ dnorm(mu,sigma),
+    mu <- no + a * associated  + d * different,
+    no ~ dnorm(1,.5),
+    a ~ dnorm(0,1),
+    d ~ dnorm(0,1),
+    sigma ~ dcauchy(0,1)
+  ),
+  data = dataset,
+  start=list(no = 1, a = 0, d = 0, sigma= .3)
+)
+
+
+compare(null, associated, type)
+
+
+
+
+associated.st  <- map2stan(
+  alist(
+    cosine_distance ~ dnorm(mu,sigma),
+    mu <- no + a * associated,
+    no ~ dnorm(1,.5),
+    a ~ dnorm(0,1),
+    sigma ~ dcauchy(0,1)
+  ),
+  data = dataset,
+  chains=2 , iter=4000 , warmup=1000,
+  start=list(no = 1, a = 0, sigma= .3)
+)
+
+
+type.st  <- map2stan(
+  alist(
+    cosine_distance ~ dnorm(mu,sigma),
+    mu <- no + a * associated  + d * different,
+    no ~ dnorm(1,.5),
+    a ~ dnorm(0,1),
+    d ~ dnorm(0,1),
+    sigma ~ dcauchy(0,1)
+  ),
[…]

### 2021-06-08 functions/evaluateProtected.R

@@ -1,0 +1,106 @@
+library(ggplot2)
+library(ggthemes)
+
+library(tidyverse)
+library(dplyr)
+library(rethinking)
+
+options(buildtools.check = function(action) TRUE )
+
+#setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
+source("cosineAnalysis/functions/cleanDataset.R")
+
+
+extractPars <- function(typeModel) {
+  return(data.frame(n = precis(typeModel)["no","mean"], 
+                    nlo = precis(typeModel)["no","5.5%"],
+                    nhigh = precis(typeModel)["no","94.5%"],
+                    a = precis(typeModel)["a","mean"],
+                    alo = precis(typeModel)["a","5.5%"],
+                    ahigh =  precis(typeModel)["a","94.5%"],
+                    d = precis(typeModel)["d","mean"],
+                    dlo = precis(typeModel)["d","5.5%"],
+                    dhigh =  precis(typeModel)["d","94.5%"],
+                    h = precis(typeModel)["h","mean"],
+                    hlo = precis(typeModel)["h","5.5%"],
+                    hhigh = precis(typeModel)["h","94.5%"])
+  )
+}
+
+
+
+dataset <- religion
+
+parTable <- function(dataset){
+
+type.full.st  <- map2stan(
+  alist(
+    cosineDistance ~ dnorm(mu,sigma),
+    mu <- no + a * associated  + d * different + h * human,
+    no ~ dnorm(1,.5),
+    a ~ dnorm(0,1),
+    d ~ dnorm(0,1),
+    h ~ dnorm(0,1),
+    sigma ~ dcauchy(0,1)), data = dataset,
+  chains=2 , iter=6000 , warmup=1000,
+  start=list(no = 1, a = 0, d = 0, sigma= .3),
+  control = list(max_treedepth = 11)
+)
+
+
+full.stats <- extractPars(type.full.st)
+rownames(full.stats) <- "full dataset"
+
+protectedWords <-unique(dataset$protectedWord)
+
+stats <- list()
+
+for(word in 1:length(protectedWords)){
+  
+  bunch <- dataset[dataset$protectedWord == protectedWords[word],]
+  type  <- rethinking::map(
+    alist(
+      cosineDistance ~ dnorm(mu,sigma),
+      mu <- no + a * associated  + d * different + h * human,
[…]

### 2021-06-08 functions/installRethinkingProperly.R

@@ -1,0 +1,4 @@
+devtools::install_github("stan-dev/cmdstanr")
+install.packages(c("coda","mvtnorm","devtools","loo","dagitty"))
+devtools::install_github("rmcelreath/rethinking")
+devtools::install_github("rmcelreath/rethinking", force = TRUE)

### 2021-06-08 thesis/03-bayesian.Rmd
working on ch 3

@@
 Its  parameters will have a posterior distribution obtained using either Hamiltionian Monte Carlo methods (STAN) available through the `rethinking` package.
+
+\vspace{1mm}
+\footnotesize
+```{r religionBaseline,echo=TRUE,eval=FALSE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%", results= "hide"}
+library(rethinking)
+options(buildtools.check = function(action) TRUE )
+religionBaseline <- ulam(
+  alist(
+    cosineDistance ~ dnorm(mu,sigma),
+    mu <- m[pw],
+    m[pw] ~ dnorm(1,.5),
+    sigma ~ dcauchy(0,1)
+  ),
+  data = religion,
+  chains=2 , iter=4000 , warmup=1000,
+  start= list(mu = 1, co = 0, sigma= .3),
+  log_lik = TRUE, cores=4
+)
+#saving
+#saveRDS(religionBaseline, 
+#file = "cosineAnalysis/models/religionBaseline.rds")
+```
+
+The only reason we need it is the evaluation of connection as a predictor.
+Does including it in o the model improve the situation?
+To investigate this, let's now build a model according to the following specification:
+
+\begin{align}
+cosineDistance_i  & \sim dnorm(\mu_i, \sigma) \\
+\mu_i & = m_{pw} + co_{con}\\
+m_{pw} & ~ dnorm(1,.5) \\
+co_{con} & ~ dnorm(0,1) \\
+\sigma &\sim  dcauchy(0,1)
+\end{align}
+
+\noindent The idea now is that each connection type comes with its own coefficient $co$ that has impact on mean distances for protected words taken separately.
+
+\vspace{1mm}
+\footnotesize
+```{r Coefsmodel,echo=TRUE,eval=FALSE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+library(rethinking)
+options(buildtools.check = function(action) TRUE )
+religionCoefs <- ulam(
+  alist(
+    cosineDistance ~ dnorm(mu,sigma),
+    mu <- m[pw] + co[con],
+    m[pw] ~ dnorm(1,.5),
+    co[con] ~dnorm(0,.5),
+    sigma ~ dcauchy(0,1)
+  ),
+  data = religion,
+  chains=2 , iter=8000 , warmup=1000, 
+  log_lik = TRUE
+)
+```
+\normalsize
+
+\noindent First, let's see if this model is really better in terms of the Widely Acceptable Information Criterion (WAIC):
+
+\vspace{1mm}
+\footnotesize
+```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+#the calculation requires models which are large, we prepared the table
+#compareBaseline <- print(round(compare(religionBaseline, religionCoefs)), 3)
+compareBaseline <- readRDS("../datasets/compareBaseline.rds")
+compareBaseline
+```
+\normalsize

### 2021-06-08 thesis/_book/_main.tex
working on ch 3

@@
 
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(rethinking)}
+\KeywordTok{options}\NormalTok{(}\DataTypeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\OtherTok{TRUE}\NormalTok{ )}
+\NormalTok{religionBaseline <-}\StringTok{ }\KeywordTok{ulam}\NormalTok{(}
+  \KeywordTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu <-}\StringTok{ }\NormalTok{m[pw],}
+\NormalTok{    m[pw] }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\OperatorTok{~}\StringTok{ }\KeywordTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\NormalTok{  ),}
+  \DataTypeTok{data =}\NormalTok{ religion,}
+  \DataTypeTok{chains=}\DecValTok{2}\NormalTok{ , }\DataTypeTok{iter=}\DecValTok{4000}\NormalTok{ , }\DataTypeTok{warmup=}\DecValTok{1000}\NormalTok{,}
+  \DataTypeTok{start=} \KeywordTok{list}\NormalTok{(}\DataTypeTok{mu =} \DecValTok{1}\NormalTok{, }\DataTypeTok{co =} \DecValTok{0}\NormalTok{, }\DataTypeTok{sigma=} \FloatTok{.3}\NormalTok{),}
+  \DataTypeTok{log_lik =} \OtherTok{TRUE}\NormalTok{, }\DataTypeTok{cores=}\DecValTok{4}
+\NormalTok{)}
+\CommentTok{#saving}
+\CommentTok{#saveRDS(religionBaseline, }
+\CommentTok{#file = "cosineAnalysis/models/religionBaseline.rds")}
+\end{Highlighting}
+\end{Shaded}
+
+The only reason we need it is the evaluation of connection as a
+predictor. Does including it in o the model improve the situation? To
+investigate this, let's now build a model according to the following
+specification:
+
+\begin{align}
+cosineDistance_i  & \sim dnorm(\mu_i, \sigma) \\
+\mu_i & = m_{pw} + co_{con}\\
+m_{pw} & ~ dnorm(1,.5) \\
+co_{con} & ~ dnorm(0,1) \\
+\sigma &\sim  dcauchy(0,1)
+\end{align}
+
+\noindent The idea now is that each connection type comes with its own
+coefficient \(co\) that has impact on mean distances for protected words
+taken separately.
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(rethinking)}
+\KeywordTok{options}\NormalTok{(}\DataTypeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\OtherTok{TRUE}\NormalTok{ )}
+\NormalTok{religionCoefs <-}\StringTok{ }\KeywordTok{ulam}\NormalTok{(}
+  \KeywordTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu <-}\StringTok{ }\NormalTok{m[pw] }\OperatorTok{+}\StringTok{ }\NormalTok{co[con],}
+\NormalTok{    m[pw] }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    co[con] }\OperatorTok{~}\KeywordTok{dnorm}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\OperatorTok{~}\StringTok{ }\KeywordTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\NormalTok{  ),}
+  \DataTypeTok{data =}\NormalTok{ religion,}
+  \DataTypeTok{chains=}\DecValTok{2}\NormalTok{ , }\DataTypeTok{iter=}\DecValTok{8000}\NormalTok{ , }\DataTypeTok{warmup=}\DecValTok{1000}\NormalTok{, }
+  \DataTypeTok{log_lik =} \OtherTok{TRUE}
+\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+\noindent First, let's see if this model is really better in terms of
+the Widely Acceptable Information Criterion (WAIC):
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{#the calculation requires models which are large, we prepared the table}
+\CommentTok{#compareBaseline <- print(round(compare(religionBaseline, religionCoefs)), 3)}
+\NormalTok{compareBaseline <-}\StringTok{ }\KeywordTok{readRDS}\NormalTok{(}\StringTok{"../datasets/compareBaseline.rds"}\NormalTok{)}
+\NormalTok{compareBaseline}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+##                   WAIC SE dWAIC dSE pWAIC weight
+## religionCoefs    -2328 93     0  NA    20      1
+## religionBaseline -2283 95    45  17    16      0
+\end{verbatim}
+
+\normalsize
+
 \end{document}

### 2021-06-08 thesis/03-bayesian.Rmd
writing chapter 3

@@
 \normalsize
+
+Clearly, it should be given weight 1 as compared to the baseline model. So far, we've learned that the connection type actually has predictive value. Let's take a look at the coefficient estimates:
+
+ \vspace{1mm}
+ \footnotesize
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+#precisCoefs <- print(precis(religionCoefs, depth =2, pars= "co" ))
+#saveRDS(precisCoefs, "../datasets/precisCoefs.rds")
+precisCoefs <- readRDS("../datasets/precisCoefs.rds")
+precisCoefs
+```
+ \normalsize
+
+\noindent Let's plot them together with their highest posterior density invervals, for the three topic groups.
+
+\begin{center}
+\begin{figure}[!htb]\centering
+   \begin{minipage}{0.55\textwidth}
+  \includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
+   \end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/religionGoogleCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/genderCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/genderGoogleCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/raceCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/raceGoogleCoeffs.jpeg}
+   \end{minipage}
+\end{figure}
+
+\end{center}
+
+\noindent What should strike us is that while the mean estimates of the coefficients indeed do differ a bit, usually the highest posterior density invervals all include zero, and so we do not have strong reasons to say that, say, as far as the whole religion dataset is involved, being associated indeed is connected  with lower cosine distance.
+A second striking observation is that the estimated impact for associated stereotypes is quite often not too different from the estimated impact of attributes associated with different stereotypes, and both are sometimes not too far from the estimated impact for simply human attributes.
+In general, once the uncertainty involved is taken seriously by using control groups and statistical uncertainty estimation that does not dispose of pointwise data, the picture which focuses only on differences between means of means is too simplistic.
+
+But this doesn't mean important differences for some protected words are not there.
+For one thing, if you start with a word list that is very uneven, the actually not so bad status of some of the protected words might mask  a pretty bad situation in which some other protected words are.
+For comparison, let's see what a model focused on words related to islam tells us.
+
+\vspace{1mm}
+\footnotesize
+```{r,echo=TRUE,eval=FALSE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+#this is how we build the model
+religion <- read.csv("cosineAnalysis/datasets/religionReddit.csv")[-1]
+colnames(religion) <- c("protectedWord","wordToCompare","wordClass",
+                        "cosineDistance","cosineSimilarity","connection")
+levels(religion$wordClass) <- c("christian","human","jewish","muslim","neutral")
+muslimWords <- c("imam","islam","mosque","muslim","quran")
+muslim <- religion %>% filter(protectedWord %in% muslimWords)
+muslim$protectedWord <- droplevels(muslim$protectedWord)
+muslim$pw <- as.integer(muslim$protectedWord)
+muslim$con <- as.integer(muslim$connection)
+muslim$pwFactor <- factor(paste0(muslim$protectedWord, muslim$connection))
+muslim$pwIndex <- as.integer(muslim$pwFactor)
+
+islamCoefs <- ulam(
+  alist(
+    cosineDistance ~ dnorm(mu,sigma),
+    mu <- m[pw] + co[con],
+    m[pw] ~ dnorm(1,.5),
+    co[con] ~dnorm(0,.5),
+    sigma ~ dcauchy(0,1)
+  ),
+  data = muslim,
+  chains=2 , iter=10000 , warmup=1000, cores = 4,
+  log_lik = TRUE
+)
+```
+\normalsize
+
+Let's take a look at the coefficients:
+
+\vspace{1mm}
+\footnotesize
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+islamCoefs <- readRDS("../datasets/islamCoefs.rds")
+islamCoefs
+```
+\normalsize
+
+\begin{center}
+\includegraphics[width=8cm]{../images/islamCoeffs.jpeg}
+\end{center}
+
+\normalsize
+
+While muslim words were unusual in the sense that the disparity between associated attributes and others is stronger, the evidence is still not conclusive.
+This is because the variation even within islam-related words is large enough (and sample sizes sufficiently small) for all the highest posterior density invervals to still include zeros.
+
+So, it seems, taking a closer look does seem to make a difference. The question is, what happens if we do take a close look at the level of protected words?
+
+#  Protected-word level analysis
+
+Let's turn then to data analysis that takes a look at protected words separately.
+This time for each combination of a protected word and a connection status we will have a separate mean cosine distance estimate, each coming with its own highest posterior density interval.
+This means we will use indices that are result from all such combinations (and then we will split them up in the model precis to build visualisation, feel free to look at the `visualiseStats.R` script for details).
+
+\vspace{1mm}
+\footnotesize
+```{r build religionSeparate,echo=TRUE,eval=FALSE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+options(buildtools.check = function(action) TRUE ) #removes install pop-up request
+religion$pwFactor <- factor(paste0(religion$protectedWord, "-", religion$connection))
+religion$pwIndex <- as.integer(religion$pwFactor)
+
+religionSeparate <- ulam(
+  alist(
+    cosineDistance ~ dnorm(mu,sigma),
+    mu <- c[pwIndex],
+    c[pwIndex] ~ dnorm(1,.5),
+    sigma ~ dcauchy(0,1)
+  ),
+  data = religion,
+  chains=2 , iter=10000 , warmup=1000,
+  start=list(no = 1, a = 0, d = 0, sigma= .3), log_lik =  TRUE
+)
+```
+\normalsize
+
+\noindent Let's see if the individualized model does better than the previous models in light of WAIC which does add penalty for the number of parameters.
+
+\vspace{1mm}
+\footnotesize
+```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+compareBaselineCoefsSeparate<- readRDS("../datasets/compareBaselineCoefsSeparate.rds")
+compareBaselineCoefsSeparate
+```
+\normalsize
+
+Great, it seems that we do want to prefer this model, despite its relative complication.
+Now, what does it tell us about the protected words?
+Let's visualise the predicted means together with 89\% highest posterior density intervals.
+
+\includegraphics[width=12cm]{../images/visReligionReddit.png}

### 2021-06-08 thesis/_book/_main.tex
writing chapter 3

@@
 
-Caliskan2017Semantics touches upon the topic of biases regarding race
-and gender. They apply knowledge from well-known psychological studies
-like Implicit Association Test to research the relation between human
-stereotypical thinking and model learnt biases to discover close
-relationship between these two. For the evaluation they use Word
-Embedding Association Test (WEAT) and the Word Embedding Factual
-Association Test (WEFAT).
+Islam, Bryson, \& Narayanan
+(\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) touches upon the
+topic of biases regarding race and gender. They apply knowledge from
+well-known psychological studies like Implicit Association Test to
+research the relation between human stereotypical thinking and model
+learnt biases to discover close relationship between these two. For the
+evaluation they use Word Embedding Association Test (WEAT) and the Word
+Embedding Factual Association Test (WEFAT).
 
@@
 
+Clearly, it should be given weight 1 as compared to the baseline model.
+So far, we've learned that the connection type actually has predictive
+value. Let's take a look at the coefficient estimates:
+
+\vspace{1mm} \footnotesize
+
+\begin{verbatim}
+##              mean        sd       5.5%      94.5%    n_eff    Rhat4
+## co[1] -0.14956420 0.1151675 -0.3261650 0.03741930 294.2033 1.001449
+## co[2] -0.09880543 0.1145024 -0.2736985 0.08813271 291.5044 1.001564
+## co[3] -0.07282752 0.1133894 -0.2447778 0.11158986 287.7820 1.001627
+## co[4] -0.03103179 0.1131420 -0.2034442 0.15268770 286.8283 1.001606
+\end{verbatim}
+
+\normalsize
+
+\noindent Let's plot them together with their highest posterior density
+invervals, for the three topic groups.
+
+\begin{center}
+\begin{figure}[!htb]\centering
+   \begin{minipage}{0.55\textwidth}
+  \includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
+   \end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/religionGoogleCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/genderCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/genderGoogleCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/raceCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/raceGoogleCoeffs.jpeg}
+   \end{minipage}
+\end{figure}
+
+\end{center}
+
+\noindent What should strike us is that while the mean estimates of the
+coefficients indeed do differ a bit, usually the highest posterior
+density invervals all include zero, and so we do not have strong reasons
+to say that, say, as far as the whole religion dataset is involved,
+being associated indeed is connected with lower cosine distance. A
+second striking observation is that the estimated impact for associated
+stereotypes is quite often not too different from the estimated impact
+of attributes associated with different stereotypes, and both are
+sometimes not too far from the estimated impact for simply human
+attributes. In general, once the uncertainty involved is taken seriously
+by using control groups and statistical uncertainty estimation that does
+not dispose of pointwise data, the picture which focuses only on
+differences between means of means is too simplistic.
+
+But this doesn't mean important differences for some protected words are
+not there. For one thing, if you start with a word list that is very
+uneven, the actually not so bad status of some of the protected words
+might mask a pretty bad situation in which some other protected words
+are. For comparison, let's see what a model focused on words related to
+islam tells us.
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{#this is how we build the model}
+\NormalTok{religion <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"cosineAnalysis/datasets/religionReddit.csv"}\NormalTok{)[}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
+\KeywordTok{colnames}\NormalTok{(religion) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
+                        \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
+\KeywordTok{levels}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
+\NormalTok{muslimWords <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"imam"}\NormalTok{,}\StringTok{"islam"}\NormalTok{,}\StringTok{"mosque"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"quran"}\NormalTok{)}
+\NormalTok{muslim <-}\StringTok{ }\NormalTok{religion }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(protectedWord }\OperatorTok{%in%}\StringTok{ }\NormalTok{muslimWords)}
+\NormalTok{muslim}\OperatorTok{$}\NormalTok{protectedWord <-}\StringTok{ }\KeywordTok{droplevels}\NormalTok{(muslim}\OperatorTok{$}\NormalTok{protectedWord)}
+\NormalTok{muslim}\OperatorTok{$}\NormalTok{pw <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(muslim}\OperatorTok{$}\NormalTok{protectedWord)}
+\NormalTok{muslim}\OperatorTok{$}\NormalTok{con <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(muslim}\OperatorTok{$}\NormalTok{connection)}
+\NormalTok{muslim}\OperatorTok{$}\NormalTok{pwFactor <-}\StringTok{ }\KeywordTok{factor}\NormalTok{(}\KeywordTok{paste0}\NormalTok{(muslim}\OperatorTok{$}\NormalTok{protectedWord, muslim}\OperatorTok{$}\NormalTok{connection))}
+\NormalTok{muslim}\OperatorTok{$}\NormalTok{pwIndex <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(muslim}\OperatorTok{$}\NormalTok{pwFactor)}
+
+\NormalTok{islamCoefs <-}\StringTok{ }\KeywordTok{ulam}\NormalTok{(}
+  \KeywordTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu <-}\StringTok{ }\NormalTok{m[pw] }\OperatorTok{+}\StringTok{ }\NormalTok{co[con],}
+\NormalTok{    m[pw] }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    co[con] }\OperatorTok{~}\KeywordTok{dnorm}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\OperatorTok{~}\StringTok{ }\KeywordTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\NormalTok{  ),}
+  \DataTypeTok{data =}\NormalTok{ muslim,}
+  \DataTypeTok{chains=}\DecValTok{2}\NormalTok{ , }\DataTypeTok{iter=}\DecValTok{10000}\NormalTok{ , }\DataTypeTok{warmup=}\DecValTok{1000}\NormalTok{, }\DataTypeTok{cores =} \DecValTok{4}\NormalTok{,}
+  \DataTypeTok{log_lik =} \OtherTok{TRUE}
+\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+Let's take a look at the coefficients:
+
+\vspace{1mm} \footnotesize
+
+\begin{verbatim}
+##                mean        sd       5.5%      94.5%    n_eff    Rhat4
+## co[1] -0.1979930035 0.1708785 -0.4682696 0.07634977 1789.894 1.003445
+## co[2] -0.0334215769 0.1687587 -0.3021954 0.23720938 1738.720 1.003575
+## co[3] -0.0192492753 0.1675754 -0.2840596 0.24860974 1732.907 1.003755
+## co[4] -0.0003911363 0.1670610 -0.2661047 0.26815172 1723.758 1.003837
+\end{verbatim}
+
+\normalsize
+
+\begin{center}
+\includegraphics[width=8cm]{../images/islamCoeffs.jpeg}
+\end{center}
+
+\normalsize
+
+While muslim words were unusual in the sense that the disparity between
+associated attributes and others is stronger, the evidence is still not
+conclusive. This is because the variation even within islam-related
+words is large enough (and sample sizes sufficiently small) for all the
+highest posterior density invervals to still include zeros.
+
+So, it seems, taking a closer look does seem to make a difference. The
+question is, what happens if we do take a close look at the level of
+protected words?
+
+\chapter{Protected-word level
+analysis}\label{protected-word-level-analysis}
+
+Let's turn then to data analysis that takes a look at protected words
+separately. This time for each combination of a protected word and a
+connection status we will have a separate mean cosine distance estimate,
+each coming with its own highest posterior density interval. This means
+we will use indices that are result from all such combinations (and then
+we will split them up in the model precis to build visualisation, feel
+free to look at the \texttt{visualiseStats.R} script for details).
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{options}\NormalTok{(}\DataTypeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\OtherTok{TRUE}\NormalTok{ ) }\CommentTok{#removes install pop-up request}
+\NormalTok{religion}\OperatorTok{$}\NormalTok{pwFactor <-}\StringTok{ }\KeywordTok{factor}\NormalTok{(}\KeywordTok{paste0}\NormalTok{(religion}\OperatorTok{$}\NormalTok{protectedWord, }\StringTok{"-"}\NormalTok{, religion}\OperatorTok{$}\NormalTok{connection))}
+\NormalTok{religion}\OperatorTok{$}\NormalTok{pwIndex <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(religion}\OperatorTok{$}\NormalTok{pwFactor)}
+
+\NormalTok{religionSeparate <-}\StringTok{ }\KeywordTok{ulam}\NormalTok{(}
+  \KeywordTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu <-}\StringTok{ }\NormalTok{c[pwIndex],}
+\NormalTok{    c[pwIndex] }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\OperatorTok{~}\StringTok{ }\KeywordTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\NormalTok{  ),}
+  \DataTypeTok{data =}\NormalTok{ religion,}
+  \DataTypeTok{chains=}\DecValTok{2}\NormalTok{ , }\DataTypeTok{iter=}\DecValTok{10000}\NormalTok{ , }\DataTypeTok{warmup=}\DecValTok{1000}\NormalTok{,}
+  \DataTypeTok{start=}\KeywordTok{list}\NormalTok{(}\DataTypeTok{no =} \DecValTok{1}\NormalTok{, }\DataTypeTok{a =} \DecValTok{0}\NormalTok{, }\DataTypeTok{d =} \DecValTok{0}\NormalTok{, }\DataTypeTok{sigma=} \FloatTok{.3}\NormalTok{), }\DataTypeTok{log_lik =}  \OtherTok{TRUE}
+\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+\noindent Let's see if the individualized model does better than the
+previous models in light of WAIC which does add penalty for the number
+of parameters.
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{compareBaselineCoefsSeparate<-}\StringTok{ }\KeywordTok{readRDS}\NormalTok{(}\StringTok{"../datasets/compareBaselineCoefsSeparate.rds"}\NormalTok{)}
+\NormalTok{compareBaselineCoefsSeparate}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+##                   WAIC SE dWAIC dSE pWAIC weight
+## religionSeparate -2400 93     0  NA    60      1
+## religionCoefs    -2328 93    72  29    20      0
+## religionBaseline -2283 95   117  37    16      0
+\end{verbatim}
+
+\normalsize
+
+Great, it seems that we do want to prefer this model, despite its
+relative complication. Now, what does it tell us about the protected
+words? Let's visualise the predicted means together with 89\% highest
+posterior density intervals.
+
+\includegraphics[width=12cm]{../images/visReligionReddit.png}
+
+\hypertarget{refs}{}
+\hypertarget{ref-Caliskan2017Semantics}{}
+Islam, A. C., Bryson, J. J., \& Narayanan, A. (2016). Semantics derived
+automatically from language corpora necessarily contain human biases.
+\emph{CoRR}, \emph{abs/1608.07187}. Retrieved from
+\url{http://arxiv.org/abs/1608.07187}
+
 \end{document}

### 2021-06-08 thesis/03-bayesian.Rmd
finished bayesian analysis

@@
-# Bayesian analysis of cosine-based bias
-
 We start with loading the libraries needed for the analysis.
@@
 
-```{r priorsVis,echo=FALSE,eval=TRUE,fig.align = "center",cache=FALSE, fig.show = "hold", out.width = "100%", fig.caption = "Regularizing priors for the bayesian models."}
+```{r priorsVis,echo=FALSE,eval=TRUE,fig.align = "center",cache=FALSE, out.width = "100%", fig.caption = "Regularizing priors for the bayesian models."}
 library(ggpubr)
@@
 \footnotesize
-```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
 #the calculation requires models which are large, we prepared the table
@@
 
-Great, it seems that we do want to prefer this model, despite its relative complication.
+It seems that we do want to prefer this model, despite its relative complication.
 Now, what does it tell us about the protected words?
@@
 
-\includegraphics[width=12cm]{../images/visReligionReddit.png}
+\includegraphics[width=14cm]{../images/visReligionReddit.png}
+
+Before we move on, let's perform analogous analyses for the remaining types of supposed bias: gender and race (the model building is analogous).
+
+\includegraphics[width=14cm]{../images/visGenderReddit.png}
+
+\includegraphics[width=14cm]{../images/visRaceReddit.png}
+
+\todo{Explain where you took the embedding from, cite publications first announcing it.}
+
+\includegraphics[width=14cm]{../images/visReligionGoogle.png}
+
+\includegraphics[width=14cm]{../images/visGenderGoogle.png}
+
+\includegraphics[width=14cm]{../images/visRaceGoogle.png}
+
+# The role of debiasing
+
+\todo{describe which debiasing you used and how it was advertised in the paper}
+
+First, let's look at coefficient estimated for the whole datasets, as compared to their estimation prior to debiasing:
+
+\begin{center}
+\begin{figure}[!htb]\centering
+   \begin{minipage}{0.55\textwidth}
+  \includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
+   \end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/debiasedReligionRedditCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/genderCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/debiasedGenderRedditCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/raceCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/debiasedRaceRedditCoeffs.jpeg}
+   \end{minipage}
+\end{figure}
+
+\end{center}
+
+\todo{comment on this}
+
+Some points to discuss:
+
+- humans in religion got closer (should they?)
+
+- change in Gender is really minor, still zero out of HPDI range, note how different is very similar
+
+- small improvement in race, at the price of moving neutral terms closer
+
+Now, perhaps, the effects of debiasing will be better appreciated if we look at the level of protected words.
+After all, the hope is, the situation of extremely ill-positioned protected words have improved?
+
+\includegraphics[width=14cm]{../images/visDebReligionReddit.png}
+
+\includegraphics[width=14cm]{../images/visDebGenderReddit.png}
+
+\includegraphics[width=14cm]{../images/visDebRaceReddit.png}

### 2021-06-08 thesis/_book/_main.tex
finished bayesian analysis

@@
-\documentclass[]{book}
+\documentclass[12pt,]{book}
 \usepackage{lmodern}
+\usepackage{setspace}
+\setstretch{1.5}
 \usepackage{amssymb,amsmath}
@@
 \urlstyle{same}  % don't use monospace font for urls
+\usepackage[left=3.5cm, right=2.5cm, top=2.5cm, bottom=2.5cm, asymmetric,
+includeheadfoot]{geometry}
 \usepackage{color}
@@
 }
-\chapter*{Preface}\label{preface}
-\addcontentsline{toc}{chapter}{Preface}
-
-Test
-
 \chapter{Introduction}\label{introduction}
@@
 To discuss!
-
-\chapter{Bayesian analysis of cosine-based
-bias}\label{bayesian-analysis-of-cosine-based-bias}
 
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{#the calculation requires models which are large, we prepared the table}
-\CommentTok{#compareBaseline <- print(round(compare(religionBaseline, religionCoefs)), 3)}
-\NormalTok{compareBaseline <-}\StringTok{ }\KeywordTok{readRDS}\NormalTok{(}\StringTok{"../datasets/compareBaseline.rds"}\NormalTok{)}
-\NormalTok{compareBaseline}
-\end{Highlighting}
-\end{Shaded}
-
 \begin{verbatim}
@@
 
-Great, it seems that we do want to prefer this model, despite its
-relative complication. Now, what does it tell us about the protected
-words? Let's visualise the predicted means together with 89\% highest
-posterior density intervals.
+It seems that we do want to prefer this model, despite its relative
+complication. Now, what does it tell us about the protected words? Let's
+visualise the predicted means together with 89\% highest posterior
+density intervals.
 
-\includegraphics[width=12cm]{../images/visReligionReddit.png}
+\includegraphics[width=14cm]{../images/visReligionReddit.png}
+
+Before we move on, let's perform analogous analyses for the remaining
+types of supposed bias: gender and race (the model building is
+analogous).
+
+\includegraphics[width=14cm]{../images/visGenderReddit.png}
+
+\includegraphics[width=14cm]{../images/visRaceReddit.png}
+
+\todo{Explain where you took the embedding from, cite publications first announcing it.}
+
+\includegraphics[width=14cm]{../images/visReligionGoogle.png}
+
+\includegraphics[width=14cm]{../images/visGenderGoogle.png}
+
+\includegraphics[width=14cm]{../images/visRaceGoogle.png}
+
+\chapter{The role of debiasing}\label{the-role-of-debiasing}
+
+\todo{describe which debiasing you used and how it was advertised in the paper}
+
+First, let's look at coefficient estimated for the whole datasets, as
+compared to their estimation prior to debiasing:
+
+\begin{center}
+\begin{figure}[!htb]\centering
+   \begin{minipage}{0.55\textwidth}
+  \includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
+   \end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/debiasedReligionRedditCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/genderCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/debiasedGenderRedditCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/raceCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/debiasedRaceRedditCoeffs.jpeg}
+   \end{minipage}
+\end{figure}
+
+\end{center}
+
+\todo{comment on this}
+
+Some points to discuss:
+
+\begin{itemize}
+\item
+  humans in religion got closer (should they?)
+\item
+  change in Gender is really minor, still zero out of HPDI range, note
+  how different is very similar
+\item
+  small improvement in race, at the price of moving neutral terms closer
+\end{itemize}
+
+Now, perhaps, the effects of debiasing will be better appreciated if we
+look at the level of protected words. After all, the hope is, the
+situation of extremely ill-positioned protected words have improved?
+
+\includegraphics[width=14cm]{../images/visDebReligionReddit.png}
+
+\includegraphics[width=14cm]{../images/visDebGenderReddit.png}
+
+\includegraphics[width=14cm]{../images/visDebRaceReddit.png}
 

### 2021-06-08 thesis/02-cosine.Rmd
added chapter 2

# Cosine similarity and bias detection

## Word embeddings

To understand what cosine similarity measurement is, one first needs to grasp the concept of translating words to a computer-readable form. In the field of natural language processsing there are two main types of words representation --- localist and distributed. One-hot encoding is an example of a method used to achieve a localist representation of words. Here each vector contains information only about a single data point, this is achieved by first mapping categorical values (words) to integers and then to each integers a binary vector is assigned which contains only 0s except for the index of the integer, which is assigned 1. An example of a localist representation is:

word | 1 | 2 | 3 | 4 | 5  
|---|---|---|---|---|---|
woman | 1 | 0 | 0 | 0 | 0 
man | 0 | 1 | 0 | 0 | 0 
girl | 0 | 0 | 1 | 0 | 0 
boy | 0 | 0 | 0 | 1 | 0 
monarch | 0 | 0 | 0 | 0 | 1

In the example above it is clear that the length of the vectors increases with the number of words in a vocabulary. It is not a very computationally efficient representation. It has other flaws as well. For example, it is unable to capture the resemblance between words appearing in similar contexts.

In contrast to the localist representation, a distributed representation returns vectors that contain continuous values instead of discrete 1s and 0s. Word embeddings are a class of various techniques that allow one to represent words as distributed vectors. Such learned representations of text have certain properties. At least prima facie, they store similar (or at least co-occurring) words close to each other in a vector space. An example of distributed representation is:

word | 1 | 2 | 3 | 4   
|---|---|---|---|---|
woman | 0.456 | 0.267 | 0.675 | 0.131
man | 0.451 | 0.897 | 0.472 | 0.088
girl | 0.604 | 0.262 | 0.414 | 0.706 
boy | 0.279 | 0.172 | 0.475 | 0.010 
monarch | 0.565 | 0.678 | 0.463 | 0.975

One of the advantages of using a distributed representation is that one is able to represent an enormous number of concepts with a smaller number of units. It is also possible to better capture similarities as words of similar meanings can have similar numeric vectors.

Word embeddings have many applications in natural language processing. They are handy in document search and information retrieval. They also play their part in improving automatic translations. Well learned word representations may also contribute to the improvement of sentiment analysis or spam detection.

The numbers occurring in such representations are not random. They are learned in a process that uses a very shallow neural network. There are various types of techniques used for learning the vectors representations. One of the most straightforward ones is a skip-gram model.
Given a word the models tries to predict its neighboring words from the sentence. The mathematics behind the process relies on the idea that the prediction concerns the conditional probability of the adjacent words. The algorithm tries to minimize the loss function, which penalizes the system for discrepancy with actual co-occurrence frequencies in the corpus. One can choose various parameters of the model, such as the window size that determines how many surrounding words the model should predict. After preparing such a fitted model one takes only the learned weights from a neural network, and uses them as vectors in a word embeddings representation.

## Cosine similarity and distance
<!-- [based on https://deepai.org/machine-learning-glossary-and-terms/cosine-similarity] -->

Cosine similarity is often used as a method of finding out whether vector representations for two words suggest that they are similar or somehow connected. Cosine similarity is the cosine of the angle between two vectors: the result of dividing their inner product (dot product usually) by the product of their magnitudes.
\begin{align} \tag{Sim}
\mathsf{cosineSimilarity}(A,B) & = \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert}
\end{align}
<!-- add information about inner product role -->
Cosine similarity is considered a proper tool for this operation as its result has a clear connection to geometry and at least for a low number of dimensions may be easily interpreted. Using this scale, one can compare vector similarities in a fairly clear manner. When the vectors are aligned perpendicularly to each other, their similarity equals 0 (which is the same as the cosine of 90 degrees). This tells us that the similarity between the vectors is small. As the angle between vectors decreases, cosine similarity approaches one, which stands for the greatest similarity.

<!-- todo: plotting exemplary vectors in 2d -->
One of the limitations of this measure is that it informs us only about similarities between vectors in terms of their orientation. However, it is often argued that in comparing words in terms of this metric, the magnitude of vectors may be treated as irrelevant, as the most important information pertains to direction. <!-- todo: find reference--> In what follows, it is important to distinguish between cosine similarity and cosine distance, defined as:
\begin{align} \tag{Sim}
\mathsf{cosineDistance}(A,B) &  = 1 - \mathsf{cosineSimilarity}(A,B)\\
 &  = 1 - \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert} \nonumber
\end{align}

The greater the similarity between two vectors, the smaller the distance between them. The cosine distance ranges between 0 and 2. If the vectors are in an opposite direction to each other, the cosine distance is 2. And if the vectors are extremely similar then the cosine distance is very close to 0.

It is worth mentioning one more point concerning cosine similarity. After the vectors are normalized to have length equal to 1, inner product itself (often dot product) is used to measure the similarity.

## Cosine distance in a one-class bias detection
<!-- https://arxiv.org/pdf/1607.06520.pdf -->
<!-- source of plot https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings -->

@Bolukbasi2016Man  define similarity between words as the inner product of their normalized vectors (that is, they use cosine similarity).  They focus on examining what the geometry of word embedding is in regard to "he" and "she" words. In other words, whether the similarity between those concepts and other words reflects expected gender stereotypes. They test this hypothesis by investigating whether there is a connection between word embeddings representing certain professions and words referring to gender. They also evaluate whether automatically produced analogies between words reflect the stereotypes as well.

A very vivid way to follow their method of arguing that bias in word embeddings is real is to plot the values of inner product of chosen words. The plot below does not originate from the original paper (it comes from [https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings](https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings)) but  a similar visualization may be found there. Data used to create our plot is as follows.
\newline

Occupations associated with feminine: \textbf{"homemaker", "nurse", "receptionist", "librarian", "socialite", "hairdresser", "nanny", "bookkeeper", "stylist", "housekeeper", "interior designer", "guidance counselor"}
\newline

Occupations associated with masculine: \textbf{"maestro", "skipper", "protege", "philosopher", "captain", "architect", "financier", "warrior", "broadcaster", "magician", "fighter pilot", "boss"}
\newline

<!-- ![gender bias in professions as measured by inner product](gender_professions.png) -->

\vspace{1mm}
\footnotesize
```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
genderProjections <-  read.csv("../datasets/genderProjections.csv")[,-1]
colnames(genderProjections)[4] <- "Stereotype"
levels(genderProjections$Stereotype) <- c("male occupation", "female occupation")
ggplot(genderProjections, aes(x=he,y=she, color = Stereotype, shape = Stereotype))+geom_point(size = 2, alpha = 0.75)+theme_tufte()+labs(title ="Stereotypical professions projected on the he-she axis (dot product, unnormalized)", subtitle ="GloVe embeddings trained on the Wikipedia 2014 and Gigaword 5th Edition corpus")+scale_color_manual(values =c("orangered4","chartreuse4"))
```
\normalsize

The points in the plot above result from the calculation of the inner product of a chosen vector for a profession word and a vector for a gender word (she or he). Inner product of two vectors expresses similarity between words. This assumption originates from the geometry and properties of a vector space.

<!-- todo: shortly describe what they do later and how they verify if it was removed - analogies etc. -->

## Cosine distance in a multi-class bias detection
<!-- [article https://arxiv.org/pdf/1904.04047.pdf] -->

In the article
@manzini2019black
the authors present different approach towards finding similarities between classes of words. The authors claim that texts available online are full of direct or indirect human stereotypes. As a result, word embeddings are prone to learn and maybe amplify those biases and propagate them further into AI models that are used for various applications. Cosine distance is used in the article as a measure to first argue for the existence of multi-class bias and then to show how through bias mitigation techniques the bias may be decreased. The main steps in the procedure are as follows.

<!-- The methodology used in the paper  -->

As the code has been provided by the authors we were able to reconstruct their results. Let us go through an example that refers to the process of hard debiasing on religious attributes.

The example refers to the process of hard debiasing on religious attributes.

1. First we load word embeddings from reddit.US.txt.tok.clean.cleanedforw2v.w2v dataset
  - The word embeddings have only 50 dimensions
  - The number of individual words from the dataset is 44895

2. The protected group the authors assume should ideally not have high cosine similarity to stereotypical words. The word embeddings geometry should not place this group close to harmful stereotypes, if it is to be bias-free.

For instance, let's look at the religion-related words.

Protected words by religion type:

\textbf{"jew"       : ["judaism", "jew", "synagogue", "torah", "rabbi"]}
\newline
\textbf{"christian" : ["christianity", "christian", "church", "bible", "priest"]}
\newline
\textbf{"muslim"    : ["islam", "muslim", "mosque", "quran", "imam"]}
\newline

Stereotypical words by religion type:

\textbf{"jew"       : ["greedy", "cheap", "hairy", "liberal"]}
\newline
\textbf{"christian"       : ["judgemental", "conservative", "familial"]}
\newline
\textbf{"muslim"       : ["violent", "terrorist", "dirty", "uneducated"]}
\newline

We have prepared a dataset including the values of cosine distance for each protected word with each attribute (stereotype), its head is displayed in the table.

\footnotesize 
```{r religionTableHeadEarly,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
religion <- read.csv("../datasets/religionReddit.csv")[-1]
colnames(religion) <- c("protectedWord","wordToCompare","wordClass",
                        "cosineDistance","cosineSimilarity","connection")
levels(religion$wordClass) <- c("christian","human","jewish","muslim","neutral")
head(religion)  %>%  kable(format = "latex",booktabs=T,
                      linesep = "",  escape = FALSE, 
                      caption = "Head of the religion dataset.") %>%
                      kable_styling(latex_options=c("scale_down"))
```
\normalsize

\pagebreak

In the article there was no analysis of individual distances but the general look at the data through the usage of mean. The authors introduced a metric that tries to generalize the presence of bias through the classification of multi-class bias in groups of words connected with gender, religion or race.

In the process they first take the mean of cosine distances between a given protected word and each stereotype. They do not differentiate between stereotypes associated with a word and stereotypes associated with different words (in case of religion stereotypes characteristic for christianity has also cosine distance measured with judaism or islam). Then after collecting the list of mean cosine distances, they average the list to get one final value represeting whole group, in this example religion. In case of religion, the final mean of all mean distances is equal to 0.859.

3. In the article the authors also try to remove previously defined biases from word embedding.

First they identify the bias subspace using Principal Component Analysis (PCA) which is a technique for dimensionality reduction. It is applied here to choose the subspace that contains the greatest amount of information. There can be many subspaces found in a given group, for example in terms of religion one can identify at least a few sets that are to grasp the concept of religion in general:
\newline
\textbf{["judaism", "christianity", "islam"]}
\newline
\textbf{["jew", "christian", "muslim"]}
\newline
\textbf{["synagogue", "church", "mosque"]}
\newline

The idea is to find a set that provides enough information to create from it a vector representing the concept of religion among words. This strategy is based on the idea that different dimensions of vectors contain different types of information and in some words in vector layers (subspaces) the information about religiousness is implicitly conveyed. In some cases this knowledge is useful but in case of harmful stereotypes one does not want to have religion concept in stereotypical words.

After finding bias subspace they use it to modify the vectors values individually so that their cosine distances towards certain words are changed. In case of stereotypes the aim is to make the cosine distances larger so that the association between protected word and harmful stereotype is smaller.

<!-- go through it again -->

4. Finally there is a multi-class evaluation. They check each protected attribute with each group of debiased words. Then they do each time mean from the cosine distances from protected word to neutral words. Then they do a mean from all of these means to return one value. If the returned value of cosine distance is greater than the one calculated at the very beginning then it means that the bias was removed as the similarity between stereotypical word and the protected one decreased.

How are the multi-class biases actually defined? The authors take a list of alleged biases towards protected groups like gender, race, religion from various scientific reports. The multi-class bias existence is further proved by providing the reader with the value of a mean from the cosine similarity between protected word and list of of previously defined biases.

<!-- add more information on the source of words used -->

# Limitations of the approach

## Selection and number of attributes

<!-- TODO: Universal ways to decide on the quality and type of attributes so that the research is significant? -->

The attributes are taken from different sources, therefore there is no one justification for their choice.

<!-- TODO: Universal ways to decide on the number of attributes so that proving bias is significant? -->

<!-- https://cs.stanford.edu/people/wmorgan/sigtest.pdf -->

In the article there is no mention of methodology for deciding on the number of attributes necessary to prove a hypothesis on the given size of dataset. There are however some ways to estimate how many samples we need to make sure that the result is significant.

## No control groups

\textbf{Control group}
\todo{Tutaj raczej powiedz ze nie maj agrup a ze my bedziemy mieli}

How to properly prepare control group in terms of quality and quantity?

## Means of means

The authors use the mean average cosine similarity to check on multi-class similarity between protected word and harmful stereotypes. Details of the process are described as follows.

When one calculates individual cosine similarity between some chosen word like "christianity" and neutral attributes that are supposed to be biased are actually often having very small and negative similarity value. As a result the measurement cosine distance is for such pair of words greater than 1. The similarity interpretation for that would suggest that the words are not similar as their cosine distance is so enormous.

Doing a mean hides this issue and as there are pairs having negative and small similarities and there are those that have similarity equal to 0.5, the resulting calculation seems to be in norm. Additionally in such method the uncertainty interval is also not included which makes it even more difficult to give reasonable interpretations of the results. In the original paper words from all three religions were checked with all of the stereotypes which means that there was no distinction on whether the stereotype is associated with given religion or with the other. Not all of the stereotypical words should be considered as harmful for all of the religions. In our research we distinguished between associated stereotypes, not associated (stereotype is valid but for different religion than currently checked) and none (meaning neutral words).
\newline

## Interpretability issues

Assuming for a moment that the value of multi-class cosine distance is correct, one may question the results' interpretation. In [Black is to Criminal as Caucasian is to Police:Detecting and Removing Multiclass Bias in Word Embeddings] Table 2, there are summarized the averages of cosine distance per group (gender, race, religion). I would like to focus  now on analyzing the values relating to religious biases. Here is the fragment of table that refers to that:

Religion Debiasing  | MAC
------------- | -------------
Biased        | 0.859
Hard Debiased | 0.934
Soft Debiased ($\lambda$ = 0.2) | 0.894

MAC stand for mean average cosine similarity although in reality the values of cosine distance are there stored. What may attract attention is the fact that the value of cosine distance in "Biased" category is already quite high even before "debiasing". High cosine distance indicates low cosine similarity between values. One could think that average cosine similarity equal to approximately 0.141 is not significant enough to consider it as biased. However the authors aim to mitigate "biases" in vectors with such great distance to make it even larger. Methodologically there is a question on what bases this small similarity is still considered as a proof of bias presence.

This is in general the problem of scale and the lack of universal intervals that could be applied to know whether the cosine similarity is high enough to considered given words more similar that if we chose them on random. In other metrics like Pearson coefficient we are provided with artificial interval that tries to make the interpretations more objective with a point of reference. It seems like similar practic could be beneficial for metrics aimed at measuring similarity between words.

##  The curse of dimensionality

<!-- TODO: -->
<!-- 1. https://www.cs.princeton.edu/courses/archive/fall13/cos521/lecnotes/lec11.pdf -->
<!-- 2. https://stats.stackexchange.com/questions/341535/curse-of-dimensionality-does-cosine-similarity-work-better-and-if-so-why -->
<!-- 3. https://stats.stackexchange.com/questions/21547/distance-metric-and-curse-of-dimensions -->
<!-- + check if it may be proved in the code -->

Curse of dimensionality may take place when there is an increase in volume of data that results in adding extra dimensions to the Euclidean space. According to the article 
"https://analyticsindiamag.com/curse-of-dimensionality-and-what-beginners-should-do-to-overcome-it/"
as the number of features increases, it may be harder and harder to get useful information from the data with the usage of available algorithms. One may notice that more data should contribute to greater amount of information but more information also means greater risk of noise and distractions in data. At the same time, many times modern solutions are adapted to smaller dimensions and their results in higher ones are not intuitive or may be prone to be mistaken.

Using cosine similarity in high dimensions in word embeddings may also be prone to the curse of dimensionality. According to this article "https://www.researchgate.net/publication/327498046_The_Curse_of_Dimensionality_Inside_Out"
there are reasons to consider this phenomenon when searching for word similarities in higher dimensions.

<!-- Add more information on the experiment -->

In the article an experiment is conducted that aims at showing how the similarity values and variation change as the number of dimensions increases. The hypothesis made in the paper states that two things will happen as the number of dimensions increase, the first one is that effort required to measure cosine similarity will be greater and the second one is that the similarity between data will blur out and have less variation. In details, the authors generate random points with increasing number of dimensions where each dimension of a data point is given a value between 0 and 1. Then they pick one vector on random from each dimension class and calculate cosine similarity between chosen vector and the rest of the data. Then they check how the variation of values changes as the number of dimensions increases. It seems like the more dimensions there are, the smaller the variance and therefore it is less obvious how to interpret the resulting cosine similarities. Maybe the scale should be adjusted to the number of dimensions and variance so that it still gives us sensible information about data. According to some publishers the cosine similarity in high dimensions is not reliable enough to trust it as it may be the case that choosing words on random may result in getting similar values as when picking them consciously. 
\newline

![curse of dimensionality,  number of dimensions on the x axis,  standard deviation of similarity on the y axis](../images/curseOfDimensionality.png)

<!-- \textbf{Verifying other similarity meaures} -->

<!-- Besides cosine similarity, there are other methods used to find the similarity between vectors.   -->

<!-- a) TS-SS -->
<!-- <!-- todo: check if it makes sense to calculate for all of the datasets; interpretations etc. -->  -->

<!-- <!-- https://www.researchgate.net/publication/303513110_A_Hybrid_Geometric_Approach_for_Measuring_Similarity_Level_Among_Documents_and_Document_Clustering --> -->

<!-- b) WEAT test  -->
<!-- <!-- https://docs.responsibly.ai/word-embedding-bias.html#module-responsibly.we.weat --> -->

<!-- <!-- [to check https://github.com/taki0112/Vector_Similarity] --> -->

<!-- <!-- TODO: check in code if they make more sense --> -->

<!-- ## Bibliography
- https://www.amazon.com/Language-Processing-Synthesis-Lectures-Technologies/dp/1627052984/ref=as_li_ss_tl?ie=UTF8&qid=1502062931&sr=8-1&keywords=Neural+Network+Methods+in+Natural+Language+Processing&linkCode=sl1&tag=inspiredalgor-20&linkId=d63df073fea3ebe2d405820570b3ff03 ->

<!-- ## *** Additional -->

<!-- ### Man is to Computer Programmer as Woman is to Homemaker? Debiasing Word Embeddings" -->
<!-- [article https://arxiv.org/pdf/1607.06520.pdf] -->

<!-- Assumptions: -->
<!-- "We assume there is a set of gender neutral words $N\subseteq W$ such as flight attendant or shoes, which, by definition, are not specific to any gender." -->

<!-- TODO: check if it is useful for the analysis of black is .. -->

<!-- ### Identifying the bias subspace  -->

<!-- Interesting analysis about limits o subspace and research [source https://kawine.github.io/blog/nlp/2019/09/23/bias.html] -->

<!-- Resources to analyze: -->
<!-- 1. https://arxiv.org/pdf/2005.00965.pdf -->
<!-- 2. https://arxiv.org/pdf/2009.09435.pdf -->
<!-- 3. https://arxiv.org/pdf/1904.04047.pdf -->
<!-- 4. https://www.aclweb.org/anthology/P19-1166.pdf -->

<!-- ### How is cosine similarity entangled with analogy topic in NLP (is it??) -->

<!-- article:  https://arxiv.org/pdf/1905.09866.pdf -->

<!-- Critique of the usage of analogies as a proof for gender biases -->

<!-- TODO: check if there is other argument than that algorithm does not allow it -->

<!-- general thoughts 
https://www.researchgate.net/publication/349408535_Robustness_and_Reliability_of_Gender_Bias_Assessment_in_Word_Embeddings_The_Role_of_Base_Pairs

-
-
-
-->

### 2021-06-08 thesis/_book/_main.tex
added chapter 2

@@
 
+\chapter{Cosine similarity and bias
+detection}\label{cosine-similarity-and-bias-detection}
+
+\section{Word embeddings}\label{word-embeddings}
+
+To understand what cosine similarity measurement is, one first needs to
+grasp the concept of translating words to a computer-readable form. In
+the field of natural language processsing there are two main types of
+words representation --- localist and distributed. One-hot encoding is
+an example of a method used to achieve a localist representation of
+words. Here each vector contains information only about a single data
+point, this is achieved by first mapping categorical values (words) to
+integers and then to each integers a binary vector is assigned which
+contains only 0s except for the index of the integer, which is assigned
+1. An example of a localist representation is:
+
+\begin{longtable}[]{@{}llllll@{}}
+\toprule
+word & 1 & 2 & 3 & 4 & 5\tabularnewline
+\midrule
+\endhead
+woman & 1 & 0 & 0 & 0 & 0\tabularnewline
+man & 0 & 1 & 0 & 0 & 0\tabularnewline
+girl & 0 & 0 & 1 & 0 & 0\tabularnewline
+boy & 0 & 0 & 0 & 1 & 0\tabularnewline
+monarch & 0 & 0 & 0 & 0 & 1\tabularnewline
+\bottomrule
+\end{longtable}
+
+In the example above it is clear that the length of the vectors
+increases with the number of words in a vocabulary. It is not a very
+computationally efficient representation. It has other flaws as well.
+For example, it is unable to capture the resemblance between words
+appearing in similar contexts.
+
+In contrast to the localist representation, a distributed representation
+returns vectors that contain continuous values instead of discrete 1s
+and 0s. Word embeddings are a class of various techniques that allow one
+to represent words as distributed vectors. Such learned representations
+of text have certain properties. At least prima facie, they store
+similar (or at least co-occurring) words close to each other in a vector
+space. An example of distributed representation is:
+
+\begin{longtable}[]{@{}lllll@{}}
+\toprule
+word & 1 & 2 & 3 & 4\tabularnewline
+\midrule
+\endhead
+woman & 0.456 & 0.267 & 0.675 & 0.131\tabularnewline
+man & 0.451 & 0.897 & 0.472 & 0.088\tabularnewline
+girl & 0.604 & 0.262 & 0.414 & 0.706\tabularnewline
+boy & 0.279 & 0.172 & 0.475 & 0.010\tabularnewline
+monarch & 0.565 & 0.678 & 0.463 & 0.975\tabularnewline
+\bottomrule
+\end{longtable}
+
+One of the advantages of using a distributed representation is that one
+is able to represent an enormous number of concepts with a smaller
+number of units. It is also possible to better capture similarities as
+words of similar meanings can have similar numeric vectors.
+
+Word embeddings have many applications in natural language processing.
+They are handy in document search and information retrieval. They also
+play their part in improving automatic translations. Well learned word
+representations may also contribute to the improvement of sentiment
+analysis or spam detection.
+
+The numbers occurring in such representations are not random. They are
+learned in a process that uses a very shallow neural network. There are
+various types of techniques used for learning the vectors
+representations. One of the most straightforward ones is a skip-gram
+model. Given a word the models tries to predict its neighboring words
+from the sentence. The mathematics behind the process relies on the idea
+that the prediction concerns the conditional probability of the adjacent
+words. The algorithm tries to minimize the loss function, which
+penalizes the system for discrepancy with actual co-occurrence
+frequencies in the corpus. One can choose various parameters of the
+model, such as the window size that determines how many surrounding
+words the model should predict. After preparing such a fitted model one
+takes only the learned weights from a neural network, and uses them as
+vectors in a word embeddings representation.
+
+\section{Cosine similarity and
+distance}\label{cosine-similarity-and-distance}
+
+Cosine similarity is often used as a method of finding out whether
+vector representations for two words suggest that they are similar or
+somehow connected. Cosine similarity is the cosine of the angle between
+two vectors: the result of dividing their inner product (dot product
+usually) by the product of their magnitudes.
+
+\begin{align} \tag{Sim}
+\mathsf{cosineSimilarity}(A,B) & = \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert}
+\end{align}
+
+Cosine similarity is considered a proper tool for this operation as its
+result has a clear connection to geometry and at least for a low number
+of dimensions may be easily interpreted. Using this scale, one can
+compare vector similarities in a fairly clear manner. When the vectors
+are aligned perpendicularly to each other, their similarity equals 0
+(which is the same as the cosine of 90 degrees). This tells us that the
+similarity between the vectors is small. As the angle between vectors
+decreases, cosine similarity approaches one, which stands for the
+greatest similarity.
+
+One of the limitations of this measure is that it informs us only about
+similarities between vectors in terms of their orientation. However, it
+is often argued that in comparing words in terms of this metric, the
+magnitude of vectors may be treated as irrelevant, as the most important
+information pertains to direction. In what follows, it is important to
+distinguish between cosine similarity and cosine distance, defined as:
+
+\begin{align} \tag{Sim}
+\mathsf{cosineDistance}(A,B) &  = 1 - \mathsf{cosineSimilarity}(A,B)\\
+ &  = 1 - \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert} \nonumber
+\end{align}
+
+The greater the similarity between two vectors, the smaller the distance
+between them. The cosine distance ranges between 0 and 2. If the vectors
+are in an opposite direction to each other, the cosine distance is 2.
+And if the vectors are extremely similar then the cosine distance is
+very close to 0.
+
+It is worth mentioning one more point concerning cosine similarity.
+After the vectors are normalized to have length equal to 1, inner
+product itself (often dot product) is used to measure the similarity.
+
+\section{Cosine distance in a one-class bias
+detection}\label{cosine-distance-in-a-one-class-bias-detection}
+
+Bolukbasi, Chang, Zou, Saligrama, \& Kalai
+(\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) define similarity
+between words as the inner product of their normalized vectors (that is,
+they use cosine similarity). They focus on examining what the geometry
+of word embedding is in regard to ``he'' and ``she'' words. In other
+words, whether the similarity between those concepts and other words
+reflects expected gender stereotypes. They test this hypothesis by
+investigating whether there is a connection between word embeddings
+representing certain professions and words referring to gender. They
+also evaluate whether automatically produced analogies between words
+reflect the stereotypes as well.
+
+A very vivid way to follow their method of arguing that bias in word
+embeddings is real is to plot the values of inner product of chosen
+words. The plot below does not originate from the original paper (it
+comes from
+\url{https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings}) but
+a similar visualization may be found there. Data used to create our plot
+is as follows. \newline
+
+Occupations associated with feminine:
+\textbf{"homemaker", "nurse", "receptionist", "librarian", "socialite", "hairdresser", "nanny", "bookkeeper", "stylist", "housekeeper", "interior designer", "guidance counselor"}
+\newline
+
+Occupations associated with masculine:
+\textbf{"maestro", "skipper", "protege", "philosopher", "captain", "architect", "financier", "warrior", "broadcaster", "magician", "fighter pilot", "boss"}
+\newline
+
+\vspace{1mm} \footnotesize
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-1-1} \end{center}
+
+\normalsize
+
+The points in the plot above result from the calculation of the inner
+product of a chosen vector for a profession word and a vector for a
+gender word (she or he). Inner product of two vectors expresses
+similarity between words. This assumption originates from the geometry
+and properties of a vector space.
+
+\section{Cosine distance in a multi-class bias
+detection}\label{cosine-distance-in-a-multi-class-bias-detection}
+
+In the article Manzini, Lim, Tsvetkov, \& Black
+(\protect\hyperlink{ref-manzini2019black}{2019}) the authors present
+different approach towards finding similarities between classes of
+words. The authors claim that texts available online are full of direct
+or indirect human stereotypes. As a result, word embeddings are prone to
+learn and maybe amplify those biases and propagate them further into AI
+models that are used for various applications. Cosine distance is used
+in the article as a measure to first argue for the existence of
+multi-class bias and then to show how through bias mitigation techniques
+the bias may be decreased. The main steps in the procedure are as
+follows.
+
+As the code has been provided by the authors we were able to reconstruct
+their results. Let us go through an example that refers to the process
+of hard debiasing on religious attributes.
+
+The example refers to the process of hard debiasing on religious
+attributes.
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  First we load word embeddings from
+  reddit.US.txt.tok.clean.cleanedforw2v.w2v dataset
+\end{enumerate}
+
+\begin{itemize}
+\tightlist
+\item
+  The word embeddings have only 50 dimensions
+\item
+  The number of individual words from the dataset is 44895
+\end{itemize}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{1}
+\tightlist
+\item
+  The protected group the authors assume should ideally not have high
+  cosine similarity to stereotypical words. The word embeddings geometry
+  should not place this group close to harmful stereotypes, if it is to
+  be bias-free.
+\end{enumerate}
+
+For instance, let's look at the religion-related words.
+
+Protected words by religion type:
+
+\textbf{"jew"       : ["judaism", "jew", "synagogue", "torah", "rabbi"]}
+\newline
+\textbf{"christian" : ["christianity", "christian", "church", "bible", "priest"]}
+\newline
+\textbf{"muslim"    : ["islam", "muslim", "mosque", "quran", "imam"]}
+\newline
+
+Stereotypical words by religion type:
+
+\textbf{"jew"       : ["greedy", "cheap", "hairy", "liberal"]} \newline
+\textbf{"christian"       : ["judgemental", "conservative", "familial"]}
+\newline
+\textbf{"muslim"       : ["violent", "terrorist", "dirty", "uneducated"]}
+\newline
+
+We have prepared a dataset including the values of cosine distance for
+each protected word with each attribute (stereotype), its head is
+displayed in the table.
+
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{religion <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
+\KeywordTok{colnames}\NormalTok{(religion) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
+                        \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
+\KeywordTok{levels}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
+\KeywordTok{head}\NormalTok{(religion)  }\OperatorTok{%>%}\StringTok{  }\KeywordTok{kable}\NormalTok{(}\DataTypeTok{format =} \StringTok{"latex"}\NormalTok{,}\DataTypeTok{booktabs=}\NormalTok{T,}
+                      \DataTypeTok{linesep =} \StringTok{""}\NormalTok{,  }\DataTypeTok{escape =} \OtherTok{FALSE}\NormalTok{, }
+                      \DataTypeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\OperatorTok{%>%}
+\StringTok{                      }\KeywordTok{kable_styling}\NormalTok{(}\DataTypeTok{latex_options=}\KeywordTok{c}\NormalTok{(}\StringTok{"scale_down"}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{table}
+
+\caption{\label{tab:religionTableHeadEarly}Head of the religion dataset.}
+\centering
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{lllrrl}
+\toprule
+protectedWord & wordToCompare & wordClass & cosineDistance & cosineSimilarity & connection\\
+\midrule
+judaism & violent & muslim & 0.7141939 & 0.2858061 & different\\
+judaism & terrorist & muslim & 0.7461333 & 0.2538667 & different\\
+judaism & dirty & muslim & 1.2002599 & -0.2002599 & different\\
+judaism & uneducated & muslim & 0.7885469 & 0.2114531 & different\\
+judaism & greedy & jewish & 1.0026172 & -0.0026172 & associated\\
+judaism & cheap & jewish & 1.2323229 & -0.2323229 & associated\\
+\bottomrule
+\end{tabular}}
+\end{table}
+
+\normalsize
+
+\pagebreak
+
+In the article there was no analysis of individual distances but the
+general look at the data through the usage of mean. The authors
+introduced a metric that tries to generalize the presence of bias
+through the classification of multi-class bias in groups of words
+connected with gender, religion or race.
+
+In the process they first take the mean of cosine distances between a
+given protected word and each stereotype. They do not differentiate
+between stereotypes associated with a word and stereotypes associated
+with different words (in case of religion stereotypes characteristic for
+christianity has also cosine distance measured with judaism or islam).
+Then after collecting the list of mean cosine distances, they average
+the list to get one final value represeting whole group, in this example
+religion. In case of religion, the final mean of all mean distances is
+equal to 0.859.
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{2}
+\tightlist
+\item
+  In the article the authors also try to remove previously defined
+  biases from word embedding.
+\end{enumerate}
+
+First they identify the bias subspace using Principal Component Analysis
+(PCA) which is a technique for dimensionality reduction. It is applied
+here to choose the subspace that contains the greatest amount of
+information. There can be many subspaces found in a given group, for
+example in terms of religion one can identify at least a few sets that
+are to grasp the concept of religion in general: \newline
+\textbf{["judaism", "christianity", "islam"]} \newline
+\textbf{["jew", "christian", "muslim"]} \newline
+\textbf{["synagogue", "church", "mosque"]} \newline
+
+The idea is to find a set that provides enough information to create
+from it a vector representing the concept of religion among words. This
+strategy is based on the idea that different dimensions of vectors
+contain different types of information and in some words in vector
+layers (subspaces) the information about religiousness is implicitly
+conveyed. In some cases this knowledge is useful but in case of harmful
+stereotypes one does not want to have religion concept in stereotypical
+words.
+
+After finding bias subspace they use it to modify the vectors values
+individually so that their cosine distances towards certain words are
+changed. In case of stereotypes the aim is to make the cosine distances
+larger so that the association between protected word and harmful
+stereotype is smaller.
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{3}
+\tightlist
+\item
+  Finally there is a multi-class evaluation. They check each protected
+  attribute with each group of debiased words. Then they do each time
+  mean from the cosine distances from protected word to neutral words.
+  Then they do a mean from all of these means to return one value. If
+  the returned value of cosine distance is greater than the one
+  calculated at the very beginning then it means that the bias was
+  removed as the similarity between stereotypical word and the protected
+  one decreased.
+\end{enumerate}
+
+How are the multi-class biases actually defined? The authors take a list
+of alleged biases towards protected groups like gender, race, religion
+from various scientific reports. The multi-class bias existence is
+further proved by providing the reader with the value of a mean from the
+cosine similarity between protected word and list of of previously
+defined biases.
+
+\chapter{Limitations of the approach}\label{limitations-of-the-approach}
+
+\section{Selection and number of
+attributes}\label{selection-and-number-of-attributes}
+
+The attributes are taken from different sources, therefore there is no
+one justification for their choice.
+
+In the article there is no mention of methodology for deciding on the
+number of attributes necessary to prove a hypothesis on the given size
+of dataset. There are however some ways to estimate how many samples we
+need to make sure that the result is significant.
+
+\section{No control groups}\label{no-control-groups}
+
+\textbf{Control group}
+\todo{Tutaj raczej powiedz ze nie maj agrup a ze my bedziemy mieli}
+
+How to properly prepare control group in terms of quality and quantity?
+
+\section{Means of means}\label{means-of-means}
+
+The authors use the mean average cosine similarity to check on
+multi-class similarity between protected word and harmful stereotypes.
+Details of the process are described as follows.
+
+When one calculates individual cosine similarity between some chosen
+word like ``christianity'' and neutral attributes that are supposed to
+be biased are actually often having very small and negative similarity
+value. As a result the measurement cosine distance is for such pair of
+words greater than 1. The similarity interpretation for that would
+suggest that the words are not similar as their cosine distance is so
+enormous.
+
+Doing a mean hides this issue and as there are pairs having negative and
+small similarities and there are those that have similarity equal to
+0.5, the resulting calculation seems to be in norm. Additionally in such
+method the uncertainty interval is also not included which makes it even
+more difficult to give reasonable interpretations of the results. In the
+original paper words from all three religions were checked with all of
+the stereotypes which means that there was no distinction on whether the
+stereotype is associated with given religion or with the other. Not all
+of the stereotypical words should be considered as harmful for all of
+the religions. In our research we distinguished between associated
+stereotypes, not associated (stereotype is valid but for different
+religion than currently checked) and none (meaning neutral words).
+\newline
+
+\section{Interpretability issues}\label{interpretability-issues}
+
+Assuming for a moment that the value of multi-class cosine distance is
+correct, one may question the results' interpretation. In {[}Black is to
+Criminal as Caucasian is to Police:Detecting and Removing Multiclass
+Bias in Word Embeddings{]} Table 2, there are summarized the averages of
+cosine distance per group (gender, race, religion). I would like to
+focus now on analyzing the values relating to religious biases. Here is
+the fragment of table that refers to that:
+
+\begin{longtable}[]{@{}ll@{}}
+\toprule
+Religion Debiasing & MAC\tabularnewline
+\midrule
+\endhead
+Biased & 0.859\tabularnewline
+Hard Debiased & 0.934\tabularnewline
+Soft Debiased (\(\lambda\) = 0.2) & 0.894\tabularnewline
+\bottomrule
+\end{longtable}
+
+MAC stand for mean average cosine similarity although in reality the
+values of cosine distance are there stored. What may attract attention
+is the fact that the value of cosine distance in ``Biased'' category is
+already quite high even before ``debiasing''. High cosine distance
+indicates low cosine similarity between values. One could think that
+average cosine similarity equal to approximately 0.141 is not
+significant enough to consider it as biased. However the authors aim to
+mitigate ``biases'' in vectors with such great distance to make it even
+larger. Methodologically there is a question on what bases this small
+similarity is still considered as a proof of bias presence.
+
+This is in general the problem of scale and the lack of universal
+intervals that could be applied to know whether the cosine similarity is
+high enough to considered given words more similar that if we chose them
+on random. In other metrics like Pearson coefficient we are provided
+with artificial interval that tries to make the interpretations more
+objective with a point of reference. It seems like similar practic could
+be beneficial for metrics aimed at measuring similarity between words.
+
+\section{The curse of dimensionality}\label{the-curse-of-dimensionality}
+
+Curse of dimensionality may take place when there is an increase in
+volume of data that results in adding extra dimensions to the Euclidean
+space. According to the article
+``\url{https://analyticsindiamag.com/curse-of-dimensionality-and-what-beginners-should-do-to-overcome-it/}''
+as the number of features increases, it may be harder and harder to get
+useful information from the data with the usage of available algorithms.
+One may notice that more data should contribute to greater amount of
+information but more information also means greater risk of noise and
+distractions in data. At the same time, many times modern solutions are
+adapted to smaller dimensions and their results in higher ones are not
+intuitive or may be prone to be mistaken.
+
+Using cosine similarity in high dimensions in word embeddings may also
+be prone to the curse of dimensionality. According to this article
+``\url{https://www.researchgate.net/publication/327498046_The_Curse_of_Dimensionality_Inside_Out}''
+there are reasons to consider this phenomenon when searching for word
+similarities in higher dimensions.
+
+In the article an experiment is conducted that aims at showing how the
+similarity values and variation change as the number of dimensions
+increases. The hypothesis made in the paper states that two things will
+happen as the number of dimensions increase, the first one is that
+effort required to measure cosine similarity will be greater and the
+second one is that the similarity between data will blur out and have
+less variation. In details, the authors generate random points with
+increasing number of dimensions where each dimension of a data point is
+given a value between 0 and 1. Then they pick one vector on random from
+each dimension class and calculate cosine similarity between chosen
+vector and the rest of the data. Then they check how the variation of
+values changes as the number of dimensions increases. It seems like the
+more dimensions there are, the smaller the variance and therefore it is
+less obvious how to interpret the resulting cosine similarities. Maybe
+the scale should be adjusted to the number of dimensions and variance so
+that it still gives us sensible information about data. According to
+some publishers the cosine similarity in high dimensions is not reliable
+enough to trust it as it may be the case that choosing words on random
+may result in getting similar values as when picking them consciously.
+\newline
+
+\begin{figure}
+\centering
+\includegraphics{../images/curseOfDimensionality.png}
+\caption{curse of dimensionality, number of dimensions on the x axis,
+standard deviation of similarity on the y axis}
+\end{figure}
+
+--\textgreater{}
+
+--\textgreater{}
+
+--\textgreater{}
+
+--\textgreater{}
+
+--\textgreater{}
+
+\chapter{Walkthrough with the religion
+dataset}\label{walkthrough-with-the-religion-dataset}
+
+\section{Loading and understanding the
+dataset}\label{loading-and-understanding-the-dataset}
+
 We start with loading the libraries needed for the analysis.
@@
 \normalsize
-
-\chapter{Walkthrough with the religion
-dataset}\label{walkthrough-with-the-religion-dataset}
 
@@
 
+\section{First look at the empirical
+distributions}\label{first-look-at-the-empirical-distributions}
+
 First let's take a look at the empirical distribution of distances by
@@
 
-\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-2-1} \end{center}
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-3-1} \end{center}
 
@@
 
-\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-3-1} \end{center}
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-4-1} \end{center}
 
@@
 \normalsize
+
+\section{Looking at the islam-related
+words}\label{looking-at-the-islam-related-words}
 
@@
 
-\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-6-1} \end{center}
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-7-1} \end{center}
 
@@
 
-\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-7-1} \end{center}
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-8-1} \end{center}
 
@@
 the sample sizes should have clear impact on our estimates.
+
+\section{Bayesian model structure and
+assumptions}\label{bayesian-model-structure-and-assumptions}
 
@@
 \begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/priorsVis-1} \end{center}
+
+\section{Choosing predictors}\label{choosing-predictors}
 
@@
 
+\section{Dataset-level coefficients}\label{dataset-level-coefficients}
+
 \noindent Let's plot them together with their highest posterior density
@@
 
+\section{Model structure and
+assumptions}\label{model-structure-and-assumptions}
+
 Let's turn then to data analysis that takes a look at protected words
@@
 
+\section{Protected classes in Reddit and Google
+embeddings}\label{protected-classes-in-reddit-and-google-embeddings}
+
 It seems that we do want to prefer this model, despite its relative
@@
 \todo{describe which debiasing you used and how it was advertised in the paper}
+
+\section{Dataset-level coefficients after
+debiasing}\label{dataset-level-coefficients-after-debiasing}
 
@@
 
+\section{Protected classes after
+debiasing}\label{protected-classes-after-debiasing}
+
 Now, perhaps, the effects of debiasing will be better appreciated if we
@@
 \hypertarget{refs}{}
+\hypertarget{ref-Bolukbasi2016Man}{}
+Bolukbasi, T., Chang, K., Zou, J. Y., Saligrama, V., \& Kalai, A.
+(2016). Man is to computer programmer as woman is to homemaker?
+Debiasing word embeddings. \emph{CoRR}, \emph{abs/1607.06520}. Retrieved
+from \url{http://arxiv.org/abs/1607.06520}
+
 \hypertarget{ref-Caliskan2017Semantics}{}
@@
 
+\hypertarget{ref-manzini2019black}{}
+Manzini, T., Lim, Y. C., Tsvetkov, Y., \& Black, A. W. (2019). Black is
+to criminal as caucasian is to police: Detecting and removing multiclass
+bias in word embeddings.
+
 \end{document}

### 2021-06-08 thesis/01-intro.Rmd
started adding chapter 3
writing chapter 3
Adds introduction

@@
 
-\textbf{state the general topic and give some background}
+<!-- \textbf{state the general topic and give some background}-->
 
@@
 smart assistants, search results, language translations, text analytics and so on. Models used to accomplish these tasks need a lot of
-data to learn from. This data originates from humans activities and historical recordings like texts, messages, speeches. It turns out
+data to learn from. This data originates from humans activities and historical recordings such as texts, messages or speeches. It turns out
 that in the learning process these models can learn implicit biases that reflect harmful stereotypical thinking still present in modern societies.
-One can find methods that aim at identifying hidden biases and/or try to remove them by modifying the models explicitly.
-There are many different types of models in NLP depending on a task that they ought to solve.
-However all of them need as an input words represented as numerical values and this is accomplished with word embedding models.
-The biases seem to have their primary source
-in the way the words are assigned the numerical values.
-\newline
+One can find methods that aim at identifying and measuring hidden biases and/or try to remove them by modifying the models.
+There are many different types of models in NLP depending on a task that they are supposed to solve.
+However, all of them need as an input words represented by means of numbers and this is accomplished with word embedding models.
+The models usually assign the values based on the context in which the words appear.
+It means that the input data can have enormous influence on the outcome.
+The biases seem to have their primary source in the way the words are assigned the numerical values.
 
-\textbf{review of the literature related to the topic}
+<!-- \textbf{review of the literature related to the topic}-->
 
-There is a bunch of literature available on the topic of bias detection and mitigation in NLP models. Bolukbasi2016Man focuses on gender
-biases that may be observable while investigating the representation of job occupations and gender in terms of their assigned numerical
-values. The authors apply cosine similarity measurement to investigate the phenomenon where jobs stereotypically associated with a given
-gender are in fact in the model situated closer to this gender.
+There is considerable amount of literature available on the topic of bias detection and mitigation in NLP models.
+@Bolukbasi2016Man focuses on gender biases that may be observable while investigating the representation of job occupations and gender in terms of their assigned numerical
+values.
+The authors apply cosine similarity measurement to investigate the phenomenon where (the vectors corresponding to) words related to jobs that are stereotypically associated with a given gender are in fact in the model situated closer to this gender.
+They also use analogy tasks to evaluate if the bias is present in the word embedding model.
+They check analogies by comparing pairs of word vectors, for example they search for the word complementing the puzzle: man is to doctor as woman is to ...?
+First they subtract word "man" from word "woman" and then
+they search for the ranked list of other words pairs that have similar vectors' difference.
+They also include in the formula a threshold to ensure that the resulting pairs could not be randomly picked.
 
-Caliskan2017Semantics touches upon the topic of biases regarding race and gender.
+However, as in @Nissim2019Fair it is pointed out, there are some limitations of this approach.
+According to the authors in practice most of analogies implementations do not return any input words.
+This means that it does not make sense to expect the algorithm to return the same profession for both woman and man.
+Therefore this method seems to be limited in terms of bias detection.
+The other problems regard for example the
+choice of pairs and words that are used to detect the presence of discrimination as it is often subjective and without proper justification.
+Additionally the choice of parameter set in @Bolukbasi2016Man formula to ensure that word pairs are not picked by random, is also not justified and changing it drastically
+influences the results.
+
+@Caliskan2017Semantics touches upon the topic of biases regarding race and gender.
 They apply knowledge from well-known psychological studies like Implicit Association Test to research the relation between human stereotypical thinking and model learnt biases to
@@
 
-manzini2019black proposes novel way of using a cosine similarity method to get the information on assumed resemblance between words. They investigate 
-an approach that enables to measure the bias for a class (like gender, religion, race) and express the final result with only one averaged number.
-\newline
+@manzini2019black proposes a novel way of using cosine similarity to obtain the information on assumed resemblance between words. They investigate 
+an approach that enables them to measure the bias for a class (like gender, religion, race) and express the final result with a single metric.
 
@@
 
-\textbf{define the terms and scope of the topic}
+<!-- \textbf{define the terms and scope of the topic}-->
 
-It is worth noticing the general distinction of biases mentioned in Caliskan2017Semantics. They refer to the publication concerning Implicit Association
+It is worth noticing the general distinction of biases mentioned in @Caliskan2017Semantics. They refer to the publication concerning Implicit Association
 Test Greenwald et al., 1998) where certain baseline of bias phenomenon was introduced. Namely it seems that humans naturally exhibit some biases and
@@
 One can find a bunch of various definitions trying to capture what bias and fairness actually are.
-With the choice of the definition, implications into the real-life applications may change as well as it was pointed out in Mehrabi2019Survey.
+With the choice of the definition, implications into the real-life applications may change as well as it was pointed out in @Mehrabi2019Survey.
 They mark out that there exist different types of biases, the list if long but among others there are historical bias, representation bias, measurement bias.
@@
 consideration while designing implementation of some machine learning model. In Mehrabi2019Survey one may notice that depending on the context and application different definitions may be applied.
-\newline
 
-\textbf{outline the current situation}
+<!--\textbf{outline the current situation} -->
+
 The most popular methods focus on comparing the similarity between words from protected groups and those that are considered to be stereotypical or harmful in some way.
@@
 independently on gender or race or not.
-\newline
 
-\textbf{evaluate the current situation (advantages/ disadvantages) and identify the gap}
+<!-- \textbf{evaluate the current situation (advantages/ disadvantages) and identify the gap}-->
 
@@
 example due to the averaging of values and the lack of confidence interval information.
-\newline
 
-\textbf{identify the importance of the proposed research}
+<!-- \textbf{identify the importance of the proposed research}-->
 One can find a number of articles on negative real-life implications resulting from the presence of unaddressed biases in the machine learning models.
-\newline
 
-\textbf{state the research problem/ questions}
+<!-- \textbf{state the research problem/ questions} -->
 
 In the paper we indicate how current methods used to detect biases in natural language models are limited in terms of confidence interval.
-\newline
 
-\textbf{state the research aims and/or research objectives}
+<!-- \textbf{state the research aims and/or research objectives} -->
 
@@
 methodologically valid.
-\newline
 
-\textbf{state the hypotheses}
+<!-- \textbf{state the hypotheses} -->
 
@@
 methodology.
-\newline
+
 <!-- \textbf{outline the order of information in the thesis} -->
-
-\textbf{outline the methodology}
-
-To discuss!

### 2021-06-08 thesis/_book/_main.tex
Adds introduction

--- 2021-06-08 thesis/_book/_main.tex
@@
-\documentclass[]{book}
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+%
+\documentclass[
+]{book}
+\usepackage{amsmath,amssymb}
 \usepackage{lmodern}
-\usepackage{amssymb,amsmath}
 \usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
 \ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
@@
   \usepackage[utf8]{inputenc}
-\else % if luatex or xelatex
-  \ifxetex
-    \usepackage{mathspec}
-  \else
-    \usepackage{fontspec}
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
+  \usepackage{unicode-math}
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
 \fi
-  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
-\fi
-% use upquote if available, for straight quotes in verbatim environments
+% Use upquote if available, for straight quotes in verbatim environments
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-% use microtype if available
-\IfFileExists{microtype.sty}{%
+\IfFileExists{microtype.sty}{% use microtype if available
   \usepackage[]{microtype}
@@
 }{}
-\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
-\usepackage[unicode=true]{hyperref}
+\makeatletter
+\@ifundefined{KOMAClassName}{% if non-KOMA class
+  \IfFileExists{parskip.sty}{%
+    \usepackage{parskip}
+  }{% else
+    \setlength{\parindent}{0pt}
+    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
+}{% if KOMA class
+  \KOMAoptions{parskip=half}}
+\makeatother
+\usepackage{xcolor}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
 \hypersetup{
@@
   pdfauthor={Alicja Dobrzeniecka},
-            pdfborder={0 0 0},
-            breaklinks=true}
-\urlstyle{same}  % don't use monospace font for urls
-\usepackage{color}
-\usepackage{fancyvrb}
-\newcommand{\VerbBar}{|}
-\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
-\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
-% Add ',fontsize=\small' for more characters per line
-\usepackage{framed}
-\definecolor{shadecolor}{RGB}{248,248,248}
-\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
-\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\ImportTok}[1]{#1}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
-\newcommand{\BuiltInTok}[1]{#1}
-\newcommand{\ExtensionTok}[1]{#1}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
-\newcommand{\RegionMarkerTok}[1]{#1}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
-\newcommand{\NormalTok}[1]{#1}
-\usepackage{longtable,booktabs}
-% Fix footnotes in tables (requires footnote package)
-\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{long table}}{}
-\usepackage{graphicx,grffile}
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
+\urlstyle{same} % disable monospaced font for URLs
+\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
+\usepackage{graphicx}
 \makeatletter
@@
 \setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-\IfFileExists{parskip.sty}{%
-\usepackage{parskip}
-}{% else
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
-}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
 \setlength{\emergencystretch}{3em} % prevent overfull lines
@@
 \setcounter{secnumdepth}{5}
-% Redefines (sub)paragraphs to behave more like sections
-\ifx\paragraph\undefined\else
-\let\oldparagraph\paragraph
-\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+\usepackage{todonotes}
+\ifluatex
+  \usepackage{selnolig}  % disable illegal ligatures
 \fi
-\ifx\subparagraph\undefined\else
-\let\oldsubparagraph\subparagraph
-\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
+  % set entry spacing
+  \ifnum #2 > 0
+  \setlength{\parskip}{#2\baselineskip}
   \fi
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
-% set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-
-\usepackage{todonotes}
-
-\title{Conceptual and methodological problems with bias detection and avoidance
-in natural language processing}
+\title{Conceptual and methodological problems with bias detection and avoidance in natural language processing}
 \author{Alicja Dobrzeniecka}
@@
 }
-\chapter*{Preface}\label{preface}
+\hypertarget{preface}{%
+\chapter*{Preface}\label{preface}}
 \addcontentsline{toc}{chapter}{Preface}
 
-Test
+Placeholder
 
-\chapter{Introduction}\label{introduction}
+\hypertarget{introduction}{%
+\chapter{Introduction}\label{introduction}}
 
-\textbf{state the general topic and give some background}
+Natural language processing (NLP) is a subfield of computer science that processes and analyzes language in text and speech
+with the use of modern programming methods. It has practical applications in everyday life as it concerns tasks such as email filters,
+smart assistants, search results, language translations, text analytics and so on. Models used to accomplish these tasks need a lot of
+data to learn from. This data originates from humans activities and historical recordings such as texts, messages or speeches. It turns out
+that in the learning process these models can learn implicit biases that reflect harmful stereotypical thinking still present in modern societies.
+One can find methods that aim at identifying and measuring hidden biases and/or try to remove them by modifying the models.
+There are many different types of models in NLP depending on a task that they are supposed to solve.
+However, all of them need as an input words represented by means of numbers and this is accomplished with word embedding models.
+The models usually assign the values based on the context in which the words appear.
+It means that the input data can have enormous influence on the outcome.
+The biases seem to have their primary source in the way the words are assigned the numerical values.
 
-Natural language processing (NLP) is a subfield of computer science that
-processes and analyzes language in text and speech with the use of
-modern programming methods. It has practical applications in everyday
-life as it concerns tasks such as email filters, smart assistants,
-search results, language translations, text analytics and so on. Models
-used to accomplish these tasks need a lot of data to learn from. This
-data originates from humans activities and historical recordings like
-texts, messages, speeches. It turns out that in the learning process
-these models can learn implicit biases that reflect harmful
-stereotypical thinking still present in modern societies. One can find
-methods that aim at identifying hidden biases and/or try to remove them
-by modifying the models explicitly. There are many different types of
-models in NLP depending on a task that they ought to solve. However all
-of them need as an input words represented as numerical values and this
-is accomplished with word embedding models. The biases seem to have
-their primary source in the way the words are assigned the numerical
-values. \newline
+There is considerable amount of literature available on the topic of bias detection and mitigation in NLP models.
+\protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) focuses on gender biases that may be observable while investigating the representation of job occupations and gender in terms of their assigned numerical
+values. The authors apply cosine similarity measurement to investigate the phenomenon where jobs stereotypically associated with a given gender are in fact in the model situated closer to this gender.
 
-\textbf{review of the literature related to the topic}
+\protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) touches upon the topic of biases regarding race and gender.
+They apply knowledge from well-known psychological studies like Implicit Association Test to research the relation between human stereotypical thinking and model learnt biases to
+discover close relationship between these two. For the evaluation they use Word Embedding Association Test (WEAT) and the Word Embedding Factual Association Test (WEFAT).
 
-There is a bunch of literature available on the topic of bias detection
-and mitigation in NLP models. Bolukbasi2016Man focuses on gender biases
-that may be observable while investigating the representation of job
-occupations and gender in terms of their assigned numerical values. The
-authors apply cosine similarity measurement to investigate the
-phenomenon where jobs stereotypically associated with a given gender are
-in fact in the model situated closer to this gender.
+\protect\hyperlink{ref-manzini2019black}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-manzini2019black}{2019}) proposes novel way of using a cosine similarity method to get the information on assumed resemblance between words.
+They investigate
+an approach that enables to measure the bias for a class (like gender, religion, race) and express the final result with only one averaged number.
 
-Islam, Bryson, \& Narayanan
-(\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) touches upon the
-topic of biases regarding race and gender. They apply knowledge from
-well-known psychological studies like Implicit Association Test to
-research the relation between human stereotypical thinking and model
-learnt biases to discover close relationship between these two. For the
-evaluation they use Word Embedding Association Test (WEAT) and the Word
-Embedding Factual Association Test (WEFAT).
+It is worth noticing the general distinction of biases mentioned in \protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}).
+They refer to the publication concerning Implicit Association
+Test Greenwald et al., 1998) where certain baseline of bias phenomenon was introduced. Namely it seems that humans naturally exhibit some biases and
+that not always bring social concern.
+One can imagine the intuitive associations between for example insects and flowers, and the feelings of pleasantness or unpleasantness.
+In general people would rather associate flowers with feeling pleasant than insects and this preference could be named a bias or
+prejudice in some direction. However this type of preference does not cause an uproar and it is rather morally neutral case. Unfortunately there are other
+biases and prejudices that directly influence the quality of other people's lives and therefore they should be taken care of.
 
-manzini2019black proposes novel way of using a cosine similarity method
-to get the information on assumed resemblance between words. They
-investigate an approach that enables to measure the bias for a class
-(like gender, religion, race) and express the final result with only one
-averaged number. \newline
+One can find a bunch of various definitions trying to capture what bias and fairness actually are.
+With the choice of the definition, implications into the real-life applications may change as well as it was pointed out in \protect\hyperlink{ref-Mehrabi2019Survey}{Mehrabi, Morstatter, Saxena, Lerman, \& Galstyan} (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}).
+They mark out that there exist different types of biases, the list if long but among others there are historical bias, representation bias, measurement bias.
+This indicates how complex the process itself is.
+Without the proper understanding and awareness of the problem, people are prone to unconsciously sustain this phenomenon.
 
-\textbf{define the terms and scope of the topic}
+In the article one can also find the distinction on different types of discrimination, some of them will be shortly described.
+It is worth first mentioning that protected attributes are those qualities, traits or characteristics that one cannot, according to the law, discriminated against.
+Direct discrimination refers to the situation when protected attributes of individuals explicitly result in non-favorable outcomes toward them.
+In contrast in indirect discrimination individuals appear to be treated equally but anyway they end up being treated unjustly due to the hidden effects of biases towards their protected attributes.
+Systemic discrimination takes place when policies, customs or behaviors that result from certain culture or organizational structure lead to discrimination against some groups of people.
+Finally, very common statistical discrimination refers to using
+average group statistics to judge person belonging to the group.
 
-It is worth noticing the general distinction of biases mentioned in
-Caliskan2017Semantics. They refer to the publication concerning Implicit
-Association Test Greenwald et al., 1998) where certain baseline of bias
-phenomenon was introduced. Namely it seems that humans naturally exhibit
-some biases and that not always bring social concern. One can imagine
-the intuitive associations between for example insects and flowers, and
-the feelings of pleasantness or unpleasantness. In general people would
-rather associate flowers with feeling pleasant than insects and this
-preference could be named a bias or prejudice in some direction. However
-this type of preference does not cause an uproar and it is rather
-morally neutral case. Unfortunately there are other biases and
-prejudices that directly influence the quality of other people's lives
-and therefore they should be taken care of.
+The topic of discrimination is entangled with another concept which is fairness. It is essential to grasp some concepts of fairness to take them into
+consideration while designing implementation of some machine learning model. In Mehrabi2019Survey one may notice that depending on the context and application different definitions may be applied.
 
-One can find a bunch of various definitions trying to capture what bias
-and fairness actually are. With the choice of the definition,
-implications into the real-life applications may change as well as it
-was pointed out in Mehrabi2019Survey. They mark out that there exist
-different types of biases, the list if long but among others there are
-historical bias, representation bias, measurement bias. This indicates
-how complex the process itself is. Without the proper understanding and
-awareness of the problem, people are prone to unconsciously sustain this
-phenomenon.
+The most popular methods focus on comparing the similarity between words from protected groups and those that are considered to be stereotypical or harmful in some way.
+One can find in this group methods such as euclidean distance, dot product or cosine similarity.
+There are also other ways to detect the effects of biases.
+For example through the investigation of the model performance on certain tasks that validate if the model returns some values
+independently on gender or race or not.
 
-In the article one can also find the distinction on different types of
-discrimination, some of them will be shortly described. It is worth
-first mentioning that protected attributes are those qualities, traits
-or characteristics that one cannot, according to the law, discriminated
-against. Direct discrimination refers to the situation when protected
-attributes of individuals explicitly result in non-favorable outcomes
-toward them. In contrast in indirect discrimination individuals appear
-to be treated equally but anyway they end up being treated unjustly due
-to the hidden effects of biases towards their protected attributes.
-Systemic discrimination takes place when policies, customs or behaviors
-that result from certain culture or organizational structure lead to
-discrimination against some groups of people. Finally, very common
-statistical discrimination refers to using average group statistics to
-judge person belonging to the group.
+In the currently used methods (like cosine similarity) the values of similarity are often aggregated in a way that may lead to false conclusions. For
+example due to the averaging of values and the lack of confidence interval information.
 
-The topic of discrimination is entangled with another concept which is
-fairness. It is essential to grasp some concepts of fairness to take
-them into consideration while designing implementation of some machine
-learning model. In Mehrabi2019Survey one may notice that depending on
-the context and application different definitions may be applied.
-\newline
+One can find a number of articles on negative real-life implications resulting from the presence of unaddressed biases in the machine learning models.
 
-\textbf{outline the current situation} The most popular methods focus on
-comparing the similarity between words from protected groups and those
-that are considered to be stereotypical or harmful in some way. One can
-find in this group methods such as euclidean distance, dot product or
-cosine similarity. There are also other ways to detect the effects of
-biases. For example through the investigation of the model performance
-on certain tasks that validate if the model returns some values
-independently on gender or race or not. \newline
+In the paper we indicate how current methods used to detect biases in natural language models are limited in terms of confidence interval.
 
-\textbf{evaluate the current situation (advantages/ disadvantages) and identify the gap}
+Our research tries to answer the question of how to enhance the current way in which the bias detection is performed to make sure that it is
+methodologically valid.
 
-In the currently used methods (like cosine similarity) the values of
-similarity are often aggregated in a way that may lead to false
-conclusions. For example due to the averaging of values and the lack of
-confidence interval information. \newline
+Our hypothesis is that there can be greater understanding of data and bias implications when confidence interval and Bayesian method are applied to the
+methodology.
 
-\textbf{identify the importance of the proposed research} One can find a
-number of articles on negative real-life implications resulting from the
-presence of unaddressed biases in the machine learning models. \newline
+\hypertarget{bayesian-analysis-of-cosine-based-bias}{%
+\chapter{Bayesian analysis of cosine-based bias}\label{bayesian-analysis-of-cosine-based-bias}}
 
-\textbf{state the research problem/ questions}
-
-In the paper we indicate how current methods used to detect biases in
-natural language models are limited in terms of confidence interval.
-\newline
-
-\textbf{state the research aims and/or research objectives}
-
-Our research tries to answer the question of how to enhance the current
-way in which the bias detection is performed to make sure that it is
-methodologically valid. \newline
-
-\textbf{state the hypotheses}
-
-Our hypothesis is that there can be greater understanding of data and
-bias implications when confidence interval and Bayesian method are
-applied to the methodology. \newline
-<!-- \textbf{outline the order of information in the thesis} -->
-
-\textbf{outline the methodology}
-
-To discuss!
-
-\chapter{Bayesian analysis of cosine-based
-bias}\label{bayesian-analysis-of-cosine-based-bias}
-
-We start with loading the libraries needed for the analysis.
-
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{(ggplot2)}
-\KeywordTok{library}\NormalTok{(ggthemes)}
-\KeywordTok{library}\NormalTok{(rethinking)}
-\KeywordTok{library}\NormalTok{(tidyverse)}
-\KeywordTok{library}\NormalTok{(ggpubr)}
-\KeywordTok{library}\NormalTok{(kableExtra)}
-\KeywordTok{library}\NormalTok{(dplyr)}
-\KeywordTok{library}\NormalTok{(ggExtra)}
-\KeywordTok{library}\NormalTok{(cowplot)}
-\end{Highlighting}
-\end{Shaded}
-
-\normalsize
-
-\chapter{Walkthrough with the religion
-dataset}\label{walkthrough-with-the-religion-dataset}
-
-We will use the choice of protected words and stereotypical predicates
-used in REF. This is a decent point of departure, not only we want to
-compare our method to that of REF, but also because this data format is
-fairly general (as contrasted, say, with a set up for binary
-stereotypes). Note also that the method we develop here can fairly
-easily be run for different stereotypization patterns. Let's start with
-explaining the method and its deployment using a dataset obtained for
-the religion-related protected words.
-
-Let's load, clean a bit and inspect the head of the religion dataset we
-prepared. In order to obtain this dataset, we calculated the cosine
-distance between each protected word and each word from both the
-bias-related attribute groups, which were used in the original study,
-and to neutral and human control attributes which we added as control
-groups. For instance, for religion, the bias-related predicates (coming
-from the original study in REF) include muslim bias attributes, jew bias
-attributes, christian bias attributes (see a list in the APPENDIX).
-
-We decided to add control groups in the form of two classes --- neutral
-words and human-related words. Without a proper control group it is
-quite hard to compare the resulting cosine distances and decide on their
-significance in bias detection. We prepared approximately
-300\todo{FIX later} more or less neutral words to double-check the
-prima-facie neutral hypothesis that their cosine similarity to the
-protected words will oscillate around 0 (that is, the distances will be
-around 1). This provides us with a more reliable point of reference.
-Moreover, we added human attributes that are associated with people in
-general to investigate whether the smaller cosine distance between
-protected words and stereotypes can result simply from the fact that the
-stereotype predicates are associated with humans. For two control
-groups, we have randomly drawn 300 words that do not express any
-property usually attributed to humans, and human attributes.
-\todo{describe once the words are selected}
-
-\vspace{1mm} \footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{religion <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
-\KeywordTok{colnames}\NormalTok{(religion) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
-                        \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
-\KeywordTok{levels}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
-\KeywordTok{head}\NormalTok{(religion)  }\OperatorTok{%>%}\StringTok{  }\KeywordTok{kable}\NormalTok{(}\DataTypeTok{format =} \StringTok{"latex"}\NormalTok{,}\DataTypeTok{booktabs=}\NormalTok{T,}
-                      \DataTypeTok{linesep =} \StringTok{""}\NormalTok{,  }\DataTypeTok{escape =} \OtherTok{FALSE}\NormalTok{, }
-                      \DataTypeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\OperatorTok{%>%}
-\StringTok{                      }\KeywordTok{kable_styling}\NormalTok{(}\DataTypeTok{latex_options=}\KeywordTok{c}\NormalTok{(}\StringTok{"scale_down"}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{table}
-
-\caption{\label{tab:religionTableHead}Head of the religion dataset.}
-\centering
-\resizebox{\linewidth}{!}{
-\begin{tabular}[t]{lllrrl}
-\toprule
-protectedWord & wordToCompare & wordClass & cosineDistance & cosineSimilarity & connection\\
-\midrule
-judaism & violent & muslim & 0.7141939 & 0.2858061 & different\\
-judaism & terrorist & muslim & 0.7461333 & 0.2538667 & different\\
-judaism & dirty & muslim & 1.2002599 & -0.2002599 & different\\
-judaism & uneducated & muslim & 0.7885469 & 0.2114531 & different\\
-judaism & greedy & jewish & 1.0026172 & -0.0026172 & associated\\
-judaism & cheap & jewish & 1.2323229 & -0.2323229 & associated\\
-\bottomrule
-\end{tabular}}
-\end{table}
-
-\normalsize
-
-The \texttt{protectedWord} column contains words from a protected class
-that (in a perfect world according to the assumptions of the orignal
-study) should not be associated with harmful stereotypes.
-\texttt{wordToCompare} contains attributes, including stereotypes and
-control group words. For each row we compute the cosine distances
-between a given protected word and a given attribute word.
-\texttt{wordClass} tells us which class an attribute is supposed to be
-stereotypically associated with, that is, whether the word from
-\texttt{wordToCompare} is associated stereotypically with jews,
-christians or muslims, or whether it belongs to a control group.
-\texttt{cosineDistance} is simply a calculation of the cosine distance
-between protected word and atrribute. \texttt{cosineSimilarity} contains
-the result of substracting cosine distance from 1. \texttt{connection}
-contains information about the relation type between a protected word
-and an attribute. If the attribute is e.g.~a harmful jewish stereotype
-and the protected word is also from the judaism group, the connection
-has value \texttt{associated}. If the attribute is still stereotypically
-jewish, but the protected word comes from another religion, the
-connection is labelled as \texttt{different}. If the attribute belongs
-to a neutral group then the connection is labelled as \texttt{none} and
-if an attribute belongs to the \texttt{human} class, then the connection
-is labelled as \texttt{human}.
-
-First let's take a look at the empirical distribution of distances by
-the connection type, initially ignoring the human control class for now.
-
-\vspace{1mm} \footnotesize
-
-\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-2-1} \end{center}
-
-\normalsize
-
-The first impression is that while there is a shift for associated words
-towards smaller cosine distances as compared to the neutral words,
-slightly surprisingly a slightly weaker shift in the same direction is
-visible for attributes associated with different stereotypes. Moreover,
-the empirical distributions overlap to alarge extent and the means
-grouped by connection type do not seem too far from each other. In fact,
-as there is a lot of variety in the consine distances (as we will soon
-see), abd we need to gauge the uncertaintly involved, and to look more
-carefully at individual protected words to get a better idea of how the
-cosine distance distribution changes for different attribute groups and
-different protected classes. Now, let's add the human attributes to the
-picture:
-
-\vspace{1mm} \footnotesize
-
-\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-3-1} \end{center}
-
-\normalsize
-
-\noindent Notice that the distribution for \texttt{human} (even though
-we did our best not to include in it any stereotype-related atributes)
-is left-skewed, with much overlap with \texttt{associated} and
-\texttt{different}, which illustrates the need to take being associated
-with humans as an important predictor.
-
-Our focus lies in \texttt{connection} as a predictor. Morever, later on
-we'll be interested in looking at the protected words separately, and at
-protected words split by connection. For technical reasons it is useful
-to represent these factors as integer vectors.
-
-\vspace{1mm} \footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{religion}\OperatorTok{$}\NormalTok{con <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(religion}\OperatorTok{$}\NormalTok{connection)}
-\NormalTok{religion}\OperatorTok{$}\NormalTok{pw <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(religion}\OperatorTok{$}\NormalTok{protectedWord)}
-\NormalTok{religion}\OperatorTok{$}\NormalTok{pwFactor <-}\StringTok{ }\KeywordTok{factor}\NormalTok{(}\KeywordTok{paste0}\NormalTok{(religion}\OperatorTok{$}\NormalTok{protectedWord, religion}\OperatorTok{$}\NormalTok{connection))}
-\NormalTok{religion}\OperatorTok{$}\NormalTok{pwIndex <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(religion}\OperatorTok{$}\NormalTok{pwFactor)}
-\end{Highlighting}
-\end{Shaded}
-
-\normalsize
-
-A short script, \texttt{cleanDataset} to make this faster, so
-equivalently:
-
-\vspace{1mm} \footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{source}\NormalTok{(}\StringTok{"../functions/cleanDataset.R"}\NormalTok{)}
-\NormalTok{religion <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
-\NormalTok{religion <-}\StringTok{ }\KeywordTok{cleanDataset}\NormalTok{(religion,}\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\normalsize
-
-For now, let's focus on five protected words related to islam (``imam'',
-``islam'', ``mosque'', ``muslim'', and ``quran''). The word list
-associates with islam four stereotypical attributes (``violent'',
-``terrorist'', ``uneducated'' and ``dirty''). First, we select and plot
-the empirical distributions for these protected words.
-
-\vspace{1mm} \footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{(tidyverse)}
-\NormalTok{muslimWords <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"imam"}\NormalTok{,}\StringTok{"islam"}\NormalTok{,}\StringTok{"mosque"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"quran"}\NormalTok{)}
-\NormalTok{muslim <-}\StringTok{ }\NormalTok{religion }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(protectedWord }\OperatorTok{%in%}\StringTok{ }\NormalTok{muslimWords)}
-\KeywordTok{ggplot}\NormalTok{(muslim, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{  cosineDistance, }\DataTypeTok{fill =}\NormalTok{ connection, }\DataTypeTok{color =}\NormalTok{ connection))}\OperatorTok{+}
-\StringTok{  }\KeywordTok{geom_density}\NormalTok{(}\DataTypeTok{alpha=}\FloatTok{0.6}\NormalTok{,}\DataTypeTok{size =} \FloatTok{.2}\NormalTok{)}\OperatorTok{+}\StringTok{ }
-\StringTok{  }\KeywordTok{scale_fill_manual}\NormalTok{(}\DataTypeTok{values =} \KeywordTok{c}\NormalTok{(}\StringTok{"orangered4"}\NormalTok{,}\StringTok{"chartreuse4"}\NormalTok{, }\StringTok{"skyblue"}\NormalTok{, }\StringTok{"gray"}\NormalTok{))}\OperatorTok{+}
-\StringTok{  }\KeywordTok{scale_x_continuous}\NormalTok{(}\DataTypeTok{breaks =} \KeywordTok{seq}\NormalTok{(}\FloatTok{0.3}\NormalTok{,}\FloatTok{1.5}\NormalTok{, }\DataTypeTok{by =} \FloatTok{0.1}\NormalTok{))}\OperatorTok{+}\KeywordTok{xlab}\NormalTok{(}\StringTok{"cosine distance"}\NormalTok{)}\OperatorTok{+}
-\StringTok{  }\KeywordTok{scale_color_manual}\NormalTok{(}\DataTypeTok{values =} \KeywordTok{c}\NormalTok{(}\StringTok{"orangered4"}\NormalTok{,}\StringTok{"chartreuse4"}\NormalTok{,}\StringTok{"skyblue"}\NormalTok{,}\StringTok{"gray"}\NormalTok{))}\OperatorTok{+}
-\StringTok{  }\KeywordTok{theme_tufte}\NormalTok{()}\OperatorTok{+}\KeywordTok{ggtitle}\NormalTok{(}\StringTok{"Empirical distribution of distances (muslim)"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-6-1} \end{center}
-
-\normalsize
-
-\noindent Once we focus on words related to islam, the associated bias
-seems to be stronger than in the whole dataset. This is a step towards
-illustrating that the distribution of bias is uneven.
-
-Now, say we want to look at a single protected word. Since the dataset
-also contains comparison multiple control neutral and human attributes,
-we randomly select only 5 from \texttt{none} and 5 from \texttt{human}
-control groups of those for the visualisation purposes.
-
-\vspace{1mm} \footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{(tidyverse)}
-\NormalTok{muslimClass <-}\StringTok{ }\NormalTok{muslim }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(protectedWord }\OperatorTok{==}\StringTok{ "muslim"}\NormalTok{)}
-\NormalTok{neutralSample <-}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\KeywordTok{filter}\NormalTok{(muslimClass,connection }\OperatorTok{==}\StringTok{ "none"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
-\NormalTok{humanSample <-}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\KeywordTok{filter}\NormalTok{(muslimClass,connection }\OperatorTok{==}\StringTok{ "human"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
-\NormalTok{muslimVis <-}\StringTok{ }\NormalTok{muslimClass }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(connection }\OperatorTok{!=}\StringTok{ "none"} \OperatorTok{&}\StringTok{ }\NormalTok{connection }\OperatorTok{!=}\StringTok{"human"}\NormalTok{)}
-\NormalTok{muslimVis <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(muslimVis,neutralSample,humanSample)}
-
-\CommentTok{#we plug in our visualisation script}
-\KeywordTok{source}\NormalTok{(}\StringTok{"../functions/visualisationTools.R"}\NormalTok{)}
-\CommentTok{#two arguments: dataset and protected word}
-\KeywordTok{visualiseProtected}\NormalTok{(muslimVis,}\StringTok{"muslim"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/tableMuslimActive-1} \end{center}
-
-\normalsize
-
-Note that the distance between the grey point and the other points is
-proportional to cosine distance, the non-grey point size is proportional
-to cosine similarity to the protected word, and color groups by the
-connection type. So for \texttt{muslim} it seems that the stereotypes
-coming from the word list are fairly well visible. To give you some
-taste of how uneven the dataset is, compare this to what happens with
-\texttt{priest}.
-
-\vspace{1mm} \footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{(tidyverse)}
-\NormalTok{priestClass <-}\StringTok{ }\NormalTok{religion }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(protectedWord }\OperatorTok{==}\StringTok{ "priest"}\NormalTok{)}
-\NormalTok{neutralSample <-}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\KeywordTok{filter}\NormalTok{(priestClass,connection }\OperatorTok{==}\StringTok{ "none"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
-\NormalTok{humanSample <-}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\KeywordTok{filter}\NormalTok{(priestClass,connection }\OperatorTok{==}\StringTok{ "human"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
-\NormalTok{priestVis <-}\StringTok{ }\NormalTok{priestClass }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(connection }\OperatorTok{!=}\StringTok{ "none"} \OperatorTok{&}\StringTok{ }\NormalTok{connection }\OperatorTok{!=}\StringTok{"human"}\NormalTok{)}
-\NormalTok{priestVis <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(priestVis,neutralSample,humanSample)}
-
-\CommentTok{#we plug in our visualisation script}
-\KeywordTok{source}\NormalTok{(}\StringTok{"../functions/visualisationTools.R"}\NormalTok{)}
-\CommentTok{#two arguments: dataset and protected word}
-\KeywordTok{visualiseProtected}\NormalTok{(priestVis,}\StringTok{"priest"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-7-1} \end{center}
-
-\normalsize
-
-\noindent Here you can see that some human attributes are closer than
-stereotype attributes, and that there is no clear reason to claim that
-\texttt{associated} attributes are closer than \texttt{different} or
-\texttt{human} attributes. This, again, illustrates the need of
-case-by-case analysis with control groups.
-
-The general idea now is that the word lists provided in different pieces
-of research are just samples of attributes associates with various
-stereotypes and should be treated as such: the uncertaintly involved and
-the sample sizes should have clear impact on our estimates.
-
-We will now think of cosine distance as the output variable, and will
-build a few bayesian models to compare. First, we just build a baseline
-model which estimates cosine distance to the attributes separately for
-each protected word. The underlying idea is that different protected
-words migh in general have different relations to all the attributes and
-this relations should be our point of departure.
-
-Here is the intution behind the mathematical Bayesian model involved.
-Our outcome variable is \texttt{cosine\ difference}, which we take to me
-normally distributed around the predicted mean for a given protected
-word (that is, we assume the residuals are normally distributed). The
-simplest model specification is:
-
-\begin{align}
-cosineDistance_i  & \sim dnorm(\mu_i, \sigma) \\
-\mu_i & = m_{pw} \\
-m_{pw} & ~ dnorm(1,.5) \\
-\sigma &\sim  dcauchy(0,1)
-\end{align}
-
-That is, we assume the estimated means might be different for diferent
-protected words and our prior for the mean and the overal standard
-deviation are normal with mean 1 and sd=.5 and half-cauchy with
-parameters \texttt{0,1}. Further on we'll also estimate additional
-impact the connection type may have. For this impact we take a slightly
-skeptical prior centered around 0 distributed normally with sd = 1.
-These are fairly weak and slightly skeptical regularizing priors, which
-can be illustrated as follows:
-
-\vspace{2mm}
-
-\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/priorsVis-1} \end{center}
-
-Now we can define and compile the baseline model. Its parameters will
-have a posterior distribution obtained using either Hamiltionian Monte
-Carlo methods (STAN) available through the \texttt{rethinking} package.
-
-\vspace{1mm} \footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{(rethinking)}
-\KeywordTok{options}\NormalTok{(}\DataTypeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\OtherTok{TRUE}\NormalTok{ )}
-\NormalTok{religionBaseline <-}\StringTok{ }\KeywordTok{ulam}\NormalTok{(}
-  \KeywordTok{alist}\NormalTok{(}
-\NormalTok{    cosineDistance }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(mu,sigma),}
-\NormalTok{    mu <-}\StringTok{ }\NormalTok{m[pw],}
-\NormalTok{    m[pw] }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
-\NormalTok{    sigma }\OperatorTok{~}\StringTok{ }\KeywordTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
-\NormalTok{  ),}
-  \DataTypeTok{data =}\NormalTok{ religion,}
-  \DataTypeTok{chains=}\DecValTok{2}\NormalTok{ , }\DataTypeTok{iter=}\DecValTok{4000}\NormalTok{ , }\DataTypeTok{warmup=}\DecValTok{1000}\NormalTok{,}
-  \DataTypeTok{start=} \KeywordTok{list}\NormalTok{(}\DataTypeTok{mu =} \DecValTok{1}\NormalTok{, }\DataTypeTok{co =} \DecValTok{0}\NormalTok{, }\DataTypeTok{sigma=} \FloatTok{.3}\NormalTok{),}
-  \DataTypeTok{log_lik =} \OtherTok{TRUE}\NormalTok{, }\DataTypeTok{cores=}\DecValTok{4}
-\NormalTok{)}
-\CommentTok{#saving}
-\CommentTok{#saveRDS(religionBaseline, }
-\CommentTok{#file = "cosineAnalysis/models/religionBaseline.rds")}
-\end{Highlighting}
-\end{Shaded}
-
-The only reason we need it is the evaluation of connection as a
-predictor. Does including it in o the model improve the situation? To
-investigate this, let's now build a model according to the following
-specification:
-
-\begin{align}
-cosineDistance_i  & \sim dnorm(\mu_i, \sigma) \\
-\mu_i & = m_{pw} + co_{con}\\
-m_{pw} & ~ dnorm(1,.5) \\
-co_{con} & ~ dnorm(0,1) \\
-\sigma &\sim  dcauchy(0,1)
-\end{align}
-
-\noindent The idea now is that each connection type comes with its own
-coefficient \(co\) that has impact on mean distances for protected words
-taken separately.
-
-\vspace{1mm} \footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{(rethinking)}
-\KeywordTok{options}\NormalTok{(}\DataTypeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\OtherTok{TRUE}\NormalTok{ )}
-\NormalTok{religionCoefs <-}\StringTok{ }\KeywordTok{ulam}\NormalTok{(}
-  \KeywordTok{alist}\NormalTok{(}
-\NormalTok{    cosineDistance }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(mu,sigma),}
-\NormalTok{    mu <-}\StringTok{ }\NormalTok{m[pw] }\OperatorTok{+}\StringTok{ }\NormalTok{co[con],}
-\NormalTok{    m[pw] }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
-\NormalTok{    co[con] }\OperatorTok{~}\KeywordTok{dnorm}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
-\NormalTok{    sigma }\OperatorTok{~}\StringTok{ }\KeywordTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
-\NormalTok{  ),}
-  \DataTypeTok{data =}\NormalTok{ religion,}
-  \DataTypeTok{chains=}\DecValTok{2}\NormalTok{ , }\DataTypeTok{iter=}\DecValTok{8000}\NormalTok{ , }\DataTypeTok{warmup=}\DecValTok{1000}\NormalTok{, }
-  \DataTypeTok{log_lik =} \OtherTok{TRUE}
-\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\normalsize
-
-\noindent First, let's see if this model is really better in terms of
-the Widely Acceptable Information Criterion (WAIC):
-
-\vspace{1mm} \footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{#the calculation requires models which are large, we prepared the table}
-\CommentTok{#compareBaseline <- print(round(compare(religionBaseline, religionCoefs)), 3)}
-\NormalTok{compareBaseline <-}\StringTok{ }\KeywordTok{readRDS}\NormalTok{(}\StringTok{"../datasets/compareBaseline.rds"}\NormalTok{)}
-\NormalTok{compareBaseline}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-##                   WAIC SE dWAIC dSE pWAIC weight
-## religionCoefs    -2328 93     0  NA    20      1
-## religionBaseline -2283 95    45  17    16      0
-\end{verbatim}
-
-\normalsize
-
-Clearly, it should be given weight 1 as compared to the baseline model.
-So far, we've learned that the connection type actually has predictive
-value. Let's take a look at the coefficient estimates:
-
-\vspace{1mm} \footnotesize
-
-\begin{verbatim}
-##              mean        sd       5.5%      94.5%    n_eff    Rhat4
-## co[1] -0.14956420 0.1151675 -0.3261650 0.03741930 294.2033 1.001449
-## co[2] -0.09880543 0.1145024 -0.2736985 0.08813271 291.5044 1.001564
-## co[3] -0.07282752 0.1133894 -0.2447778 0.11158986 287.7820 1.001627
-## co[4] -0.03103179 0.1131420 -0.2034442 0.15268770 286.8283 1.001606
-\end{verbatim}
-
-\normalsize
-
-\noindent Let's plot them together with their highest posterior density
-invervals, for the three topic groups.
-
-\begin{center}
-\begin{figure}[!htb]\centering
-   \begin{minipage}{0.55\textwidth}
-  \includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
-   \end{minipage}
-   \begin {minipage}{0.43\textwidth}
-    \includegraphics[width=7cm]{../images/religionGoogleCoeffs.jpeg}
-   \end{minipage}
-
-  \begin{minipage}{0.55\textwidth}
-\includegraphics[width=7cm]{../images/genderCoeffs.jpeg}
-\end{minipage}
-   \begin {minipage}{0.43\textwidth}
-    \includegraphics[width=7cm]{../images/genderGoogleCoeffs.jpeg}
-   \end{minipage}
-
-  \begin{minipage}{0.55\textwidth}
-\includegraphics[width=7cm]{../images/raceCoeffs.jpeg}
-\end{minipage}
-   \begin {minipage}{0.43\textwidth}
-    \includegraphics[width=7cm]{../images/raceGoogleCoeffs.jpeg}
-   \end{minipage}
-\end{figure}
-
-\end{center}
-
-\noindent What should strike us is that while the mean estimates of the
-coefficients indeed do differ a bit, usually the highest posterior
-density invervals all include zero, and so we do not have strong reasons
-to say that, say, as far as the whole religion dataset is involved,
-being associated indeed is connected with lower cosine distance. A
-second striking observation is that the estimated impact for associated
-stereotypes is quite often not too different from the estimated impact
-of attributes associated with different stereotypes, and both are
-sometimes not too far from the estimated impact for simply human
-attributes. In general, once the uncertainty involved is taken seriously
-by using control groups and statistical uncertainty estimation that does
-not dispose of pointwise data, the picture which focuses only on
-differences between means of means is too simplistic.
-
-But this doesn't mean important differences for some protected words are
-not there. For one thing, if you start with a word list that is very
-uneven, the actually not so bad status of some of the protected words
-might mask a pretty bad situation in which some other protected words
-are. For comparison, let's see what a model focused on words related to
-islam tells us.
-
-\vspace{1mm} \footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{#this is how we build the model}
-\NormalTok{religion <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"cosineAnalysis/datasets/religionReddit.csv"}\NormalTok{)[}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
-\KeywordTok{colnames}\NormalTok{(religion) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
-                        \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
-\KeywordTok{levels}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
-\NormalTok{muslimWords <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"imam"}\NormalTok{,}\StringTok{"islam"}\NormalTok{,}\StringTok{"mosque"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"quran"}\NormalTok{)}
-\NormalTok{muslim <-}\StringTok{ }\NormalTok{religion }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(protectedWord }\OperatorTok{%in%}\StringTok{ }\NormalTok{muslimWords)}
-\NormalTok{muslim}\OperatorTok{$}\NormalTok{protectedWord <-}\StringTok{ }\KeywordTok{droplevels}\NormalTok{(muslim}\OperatorTok{$}\NormalTok{protectedWord)}
-\NormalTok{muslim}\OperatorTok{$}\NormalTok{pw <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(muslim}\OperatorTok{$}\NormalTok{protectedWord)}
-\NormalTok{muslim}\OperatorTok{$}\NormalTok{con <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(muslim}\OperatorTok{$}\NormalTok{connection)}
-\NormalTok{muslim}\OperatorTok{$}\NormalTok{pwFactor <-}\StringTok{ }\KeywordTok{factor}\NormalTok{(}\KeywordTok{paste0}\NormalTok{(muslim}\OperatorTok{$}\NormalTok{protectedWord, muslim}\OperatorTok{$}\NormalTok{connection))}
-\NormalTok{muslim}\OperatorTok{$}\NormalTok{pwIndex <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(muslim}\OperatorTok{$}\NormalTok{pwFactor)}
-
-\NormalTok{islamCoefs <-}\StringTok{ }\KeywordTok{ulam}\NormalTok{(}
-  \KeywordTok{alist}\NormalTok{(}
-\NormalTok{    cosineDistance }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(mu,sigma),}
-\NormalTok{    mu <-}\StringTok{ }\NormalTok{m[pw] }\OperatorTok{+}\StringTok{ }\NormalTok{co[con],}
-\NormalTok{    m[pw] }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
-\NormalTok{    co[con] }\OperatorTok{~}\KeywordTok{dnorm}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
-\NormalTok{    sigma }\OperatorTok{~}\StringTok{ }\KeywordTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
-\NormalTok{  ),}
-  \DataTypeTok{data =}\NormalTok{ muslim,}
-  \DataTypeTok{chains=}\DecValTok{2}\NormalTok{ , }\DataTypeTok{iter=}\DecValTok{10000}\NormalTok{ , }\DataTypeTok{warmup=}\DecValTok{1000}\NormalTok{, }\DataTypeTok{cores =} \DecValTok{4}\NormalTok{,}
-  \DataTypeTok{log_lik =} \OtherTok{TRUE}
-\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\normalsize
-
-Let's take a look at the coefficients:
-
-\vspace{1mm} \footnotesize
-
-\begin{verbatim}
-##                mean        sd       5.5%      94.5%    n_eff    Rhat4
-## co[1] -0.1979930035 0.1708785 -0.4682696 0.07634977 1789.894 1.003445
-## co[2] -0.0334215769 0.1687587 -0.3021954 0.23720938 1738.720 1.003575
-## co[3] -0.0192492753 0.1675754 -0.2840596 0.24860974 1732.907 1.003755
-## co[4] -0.0003911363 0.1670610 -0.2661047 0.26815172 1723.758 1.003837
-\end{verbatim}
-
-\normalsize
-
-\begin{center}
-\includegraphics[width=8cm]{../images/islamCoeffs.jpeg}
-\end{center}
-
-\normalsize
-
-While muslim words were unusual in the sense that the disparity between
-associated attributes and others is stronger, the evidence is still not
-conclusive. This is because the variation even within islam-related
-words is large enough (and sample sizes sufficiently small) for all the
-highest posterior density invervals to still include zeros.
-
-So, it seems, taking a closer look does seem to make a difference. The
-question is, what happens if we do take a close look at the level of
-protected words?
-
-\chapter{Protected-word level
-analysis}\label{protected-word-level-analysis}
-
-Let's turn then to data analysis that takes a look at protected words
-separately. This time for each combination of a protected word and a
-connection status we will have a separate mean cosine distance estimate,
-each coming with its own highest posterior density interval. This means
-we will use indices that are result from all such combinations (and then
-we will split them up in the model precis to build visualisation, feel
-free to look at the \texttt{visualiseStats.R} script for details).
-
-\vspace{1mm} \footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{options}\NormalTok{(}\DataTypeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\OtherTok{TRUE}\NormalTok{ ) }\CommentTok{#removes install pop-up request}
-\NormalTok{religion}\OperatorTok{$}\NormalTok{pwFactor <-}\StringTok{ }\KeywordTok{factor}\NormalTok{(}\KeywordTok{paste0}\NormalTok{(religion}\OperatorTok{$}\NormalTok{protectedWord, }\StringTok{"-"}\NormalTok{, religion}\OperatorTok{$}\NormalTok{connection))}
-\NormalTok{religion}\OperatorTok{$}\NormalTok{pwIndex <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(religion}\OperatorTok{$}\NormalTok{pwFactor)}
-
-\NormalTok{religionSeparate <-}\StringTok{ }\KeywordTok{ulam}\NormalTok{(}
-  \KeywordTok{alist}\NormalTok{(}
-\NormalTok{    cosineDistance }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(mu,sigma),}
-\NormalTok{    mu <-}\StringTok{ }\NormalTok{c[pwIndex],}
-\NormalTok{    c[pwIndex] }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
-\NormalTok{    sigma }\OperatorTok{~}\StringTok{ }\KeywordTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
-\NormalTok{  ),}
-  \DataTypeTok{data =}\NormalTok{ religion,}
-  \DataTypeTok{chains=}\DecValTok{2}\NormalTok{ , }\DataTypeTok{iter=}\DecValTok{10000}\NormalTok{ , }\DataTypeTok{warmup=}\DecValTok{1000}\NormalTok{,}
-  \DataTypeTok{start=}\KeywordTok{list}\NormalTok{(}\DataTypeTok{no =} \DecValTok{1}\NormalTok{, }\DataTypeTok{a =} \DecValTok{0}\NormalTok{, }\DataTypeTok{d =} \DecValTok{0}\NormalTok{, }\DataTypeTok{sigma=} \FloatTok{.3}\NormalTok{), }\DataTypeTok{log_lik =}  \OtherTok{TRUE}
-\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\normalsize
-
-\noindent Let's see if the individualized model does better than the
-previous models in light of WAIC which does add penalty for the number
-of parameters.
-
-\vspace{1mm} \footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{compareBaselineCoefsSeparate<-}\StringTok{ }\KeywordTok{readRDS}\NormalTok{(}\StringTok{"../datasets/compareBaselineCoefsSeparate.rds"}\NormalTok{)}
-\NormalTok{compareBaselineCoefsSeparate}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-##                   WAIC SE dWAIC dSE pWAIC weight
-## religionSeparate -2400 93     0  NA    60      1
-## religionCoefs    -2328 93    72  29    20      0
-## religionBaseline -2283 95   117  37    16      0
-\end{verbatim}
-
-\normalsize
-
-Great, it seems that we do want to prefer this model, despite its
-relative complication. Now, what does it tell us about the protected
-words? Let's visualise the predicted means together with 89\% highest
-posterior density intervals.
-
-\includegraphics[width=12cm]{../images/visReligionReddit.png}
+Placeholder
 
 \hypertarget{refs}{}
-\hypertarget{ref-Caliskan2017Semantics}{}
-Islam, A. C., Bryson, J. J., \& Narayanan, A. (2016). Semantics derived
-automatically from language corpora necessarily contain human biases.
-\emph{CoRR}, \emph{abs/1608.07187}. Retrieved from
-\url{http://arxiv.org/abs/1608.07187}
+\begin{CSLReferences}{1}{0}
+\leavevmode\hypertarget{ref-Bolukbasi2016Man}{}%
+Bolukbasi, T., Chang, K.-W., Zou, J.
+Y., Saligrama, V., \& Kalai, A.
+(2016).
+Man is to computer programmer as woman is to homemaker?
+Debiasing word embeddings.
+\emph{CoRR}, \emph{abs/1607.06520}.
+Retrieved from \url{http://arxiv.org/abs/1607.06520}
+
+\leavevmode\hypertarget{ref-Caliskan2017Semantics}{}%
+Islam, A.
+C., Bryson, J.
+J., \& Narayanan, A.
+(2016).
+Semantics derived automatically from language corpora necessarily contain human biases.
+\emph{CoRR}, \emph{abs/1608.07187}.
+Retrieved from \url{http://arxiv.org/abs/1608.07187}
+
+\leavevmode\hypertarget{ref-manzini2019black}{}%
+Manzini, T., Lim, Y.
+C., Tsvetkov, Y., \& Black, A.
+W.
+(2019).
+Black is to criminal as caucasian is to police: Detecting and removing multiclass bias in word embeddings.
+Retrieved from \url{http://arxiv.org/abs/1904.04047}
+
+\leavevmode\hypertarget{ref-Mehrabi2019Survey}{}%
+Mehrabi, N., Morstatter, F., Saxena, N., Lerman, K., \& Galstyan, A.
+(2019).
+A survey on bias and fairness in machine learning.
+\emph{CoRR}, \emph{abs/1908.09635}.
+Retrieved from \url{http://arxiv.org/abs/1908.09635}
+
+\end{CSLReferences}
 

### 2021-06-08 thesis/03-bayesian.Rmd

@@
+# Walkthrough with the religion dataset
+
+## Loading and understanding the dataset
+
 We start with loading the libraries needed for the analysis.
@@
 \normalsize
-
-# Walkthrough with the religion dataset
 
@@
 
+## First look at the empirical distributions
+
 First let's take a look at the empirical distribution of distances by the connection type, initially ignoring the human control class for now.
@@
 
-<!-- For technical reasons it is useful to represent an $n$-level factor with $n-1$ binary vectors. `connection` in this case has four levels. -->
-
-<!-- \vspace{1mm} -->
-<!-- \footnotesize -->
-<!-- ```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"} -->
-<!-- religion$associated <- ifelse(religion$connection == "associated", 1, 0) -->
-<!-- religion$different <- ifelse(religion$connection == "different", 1, 0) -->
-<!-- religion$human <- ifelse(religion$connection == "human" |religion$connection == "associated"| -->
-<!--                            religion$connection == "different", 1, 0)   -->
-<!-- ```   -->
-<!-- \normalsize -->
-
 A short script, `cleanDataset` to make this faster, so equivalently:
@@
 \normalsize
+
+## Looking at the islam-related words
 
@@
 
+## Bayesian model structure and assumptions
+
 We will now think of cosine distance as the output variable, and 
@@
 ```
+
+## Choosing predictors
 
@@
  \normalsize
+
+## Dataset-level coefficients
 
@@
 
+## Model structure and assumptions
+
 Let's turn then to data analysis that takes a look at protected words separately.
@@
 
+## Protected classes in Reddit and Google embeddings
+
 It seems that we do want to prefer this model, despite its relative complication.
@@
 \todo{describe which debiasing you used and how it was advertised in the paper}
+
+## Dataset-level coefficients after debiasing
 
@@
 
+## Protected classes  after debiasing
+
 Now, perhaps, the effects of debiasing will be better appreciated if we look at the level of protected words.

### 2021-06-08 thesis/_book/_main.tex

--- 2021-06-08 thesis/_book/_main.tex
@@
 \usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
 \ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
@@
 \setcounter{secnumdepth}{5}
-% Redefines (sub)paragraphs to behave more like sections
-\ifx\paragraph\undefined\else
-\let\oldparagraph\paragraph
-\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+\usepackage{todonotes}
+\ifluatex
+  \usepackage{selnolig}  % disable illegal ligatures
 \fi
-\ifx\subparagraph\undefined\else
-\let\oldsubparagraph\subparagraph
-\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
+  % set entry spacing
+  \ifnum #2 > 0
+  \setlength{\parskip}{#2\baselineskip}
   \fi
-
-% set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-
-\usepackage{todonotes}
-
-\title{Conceptual and methodological problems with bias detection and avoidance
-in natural language processing}
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+
+\title{Conceptual and methodological problems with bias detection and avoidance in natural language processing}
 \author{Alicja Dobrzeniecka}
@@
 
-\noindent Notice that the distribution for \texttt{human} (even though
-we did our best not to include in it any stereotype-related atributes)
-is left-skewed, with much overlap with \texttt{associated} and
-\texttt{different}, which illustrates the need to take being associated
-with humans as an important predictor.
-
-Our focus lies in \texttt{connection} as a predictor. Morever, later on
-we'll be interested in looking at the protected words separately, and at
-protected words split by connection. For technical reasons it is useful
-to represent these factors as integer vectors.
-
-\vspace{1mm} \footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{religion}\OperatorTok{$}\NormalTok{con <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(religion}\OperatorTok{$}\NormalTok{connection)}
-\NormalTok{religion}\OperatorTok{$}\NormalTok{pw <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(religion}\OperatorTok{$}\NormalTok{protectedWord)}
-\NormalTok{religion}\OperatorTok{$}\NormalTok{pwFactor <-}\StringTok{ }\KeywordTok{factor}\NormalTok{(}\KeywordTok{paste0}\NormalTok{(religion}\OperatorTok{$}\NormalTok{protectedWord, religion}\OperatorTok{$}\NormalTok{connection))}
-\NormalTok{religion}\OperatorTok{$}\NormalTok{pwIndex <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(religion}\OperatorTok{$}\NormalTok{pwFactor)}
-\end{Highlighting}
-\end{Shaded}
-
-\normalsize
-
-A short script, \texttt{cleanDataset} to make this faster, so
-equivalently:
-
-\vspace{1mm} \footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{source}\NormalTok{(}\StringTok{"../functions/cleanDataset.R"}\NormalTok{)}
-\NormalTok{religion <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
-\NormalTok{religion <-}\StringTok{ }\KeywordTok{cleanDataset}\NormalTok{(religion,}\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\normalsize
+\hypertarget{introduction}{%
+\chapter{Introduction}\label{introduction}}
+
+Natural language processing (NLP) is a subfield of computer science that processes and analyzes language in text and speech
+with the use of modern programming methods. It has practical applications in everyday life as it concerns tasks such as email filters,
+smart assistants, search results, language translations, text analytics and so on. Models used to accomplish these tasks need a lot of
+data to learn from. This data originates from humans activities and historical recordings such as texts, messages or speeches. It turns out
+that in the learning process these models can learn implicit biases that reflect harmful stereotypical thinking still present in modern societies.
+One can find methods that aim at identifying and measuring hidden biases and/or try to remove them by modifying the models.
+There are many different types of models in NLP depending on a task that they are supposed to solve.
+However, all of them need as an input words represented by means of numbers and this is accomplished with word embedding models.
+The models usually assign the values based on the context in which the words appear.
+It means that the input data can have enormous influence on the outcome.
+The biases seem to have their primary source in the way the words are assigned the numerical values.
+
+There is considerable amount of literature available on the topic of bias detection and mitigation in NLP models.
+\protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) focuses on gender biases that may be observable while investigating the representation of job occupations and gender in terms of their assigned numerical
+values. The authors apply cosine similarity measurement to investigate the phenomenon where jobs stereotypically associated with a given gender are in fact in the model situated closer to this gender.
+
+\protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) touches upon the topic of biases regarding race and gender.
+They apply knowledge from well-known psychological studies like Implicit Association Test to research the relation between human stereotypical thinking and model learnt biases to
+discover close relationship between these two. For the evaluation they use Word Embedding Association Test (WEAT) and the Word Embedding Factual Association Test (WEFAT).
+
+\protect\hyperlink{ref-manzini2019black}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-manzini2019black}{2019}) proposes novel way of using a cosine similarity method to get the information on assumed resemblance between words.
+They investigate
+an approach that enables to measure the bias for a class (like gender, religion, race) and express the final result with only one averaged number.
+
+It is worth noticing the general distinction of biases mentioned in \protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}).
+They refer to the publication concerning Implicit Association
+Test Greenwald et al., 1998) where certain baseline of bias phenomenon was introduced. Namely it seems that humans naturally exhibit some biases and
+that not always bring social concern.
+One can imagine the intuitive associations between for example insects and flowers, and the feelings of pleasantness or unpleasantness.
+In general people would rather associate flowers with feeling pleasant than insects and this preference could be named a bias or
+prejudice in some direction. However this type of preference does not cause an uproar and it is rather morally neutral case. Unfortunately there are other
+biases and prejudices that directly influence the quality of other people's lives and therefore they should be taken care of.
+
+One can find a bunch of various definitions trying to capture what bias and fairness actually are.
+With the choice of the definition, implications into the real-life applications may change as well as it was pointed out in \protect\hyperlink{ref-Mehrabi2019Survey}{Mehrabi, Morstatter, Saxena, Lerman, \& Galstyan} (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}).
+They mark out that there exist different types of biases, the list if long but among others there are historical bias, representation bias, measurement bias.
+This indicates how complex the process itself is.
+Without the proper understanding and awareness of the problem, people are prone to unconsciously sustain this phenomenon.
+
+In the article one can also find the distinction on different types of discrimination, some of them will be shortly described.
+It is worth first mentioning that protected attributes are those qualities, traits or characteristics that one cannot, according to the law, discriminated against.
+Direct discrimination refers to the situation when protected attributes of individuals explicitly result in non-favorable outcomes toward them.
+In contrast in indirect discrimination individuals appear to be treated equally but anyway they end up being treated unjustly due to the hidden effects of biases towards their protected attributes.
+Systemic discrimination takes place when policies, customs or behaviors that result from certain culture or organizational structure lead to discrimination against some groups of people.
+Finally, very common statistical discrimination refers to using
+average group statistics to judge person belonging to the group.
+
+The topic of discrimination is entangled with another concept which is fairness. It is essential to grasp some concepts of fairness to take them into
+consideration while designing implementation of some machine learning model. In Mehrabi2019Survey one may notice that depending on the context and application different definitions may be applied.
 
@@
 
-\vspace{1mm} \footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{(tidyverse)}
-\NormalTok{muslimWords <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"imam"}\NormalTok{,}\StringTok{"islam"}\NormalTok{,}\StringTok{"mosque"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"quran"}\NormalTok{)}
-\NormalTok{muslim <-}\StringTok{ }\NormalTok{religion }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(protectedWord }\OperatorTok{%in%}\StringTok{ }\NormalTok{muslimWords)}
-\KeywordTok{ggplot}\NormalTok{(muslim, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{  cosineDistance, }\DataTypeTok{fill =}\NormalTok{ connection, }\DataTypeTok{color =}\NormalTok{ connection))}\OperatorTok{+}
-\StringTok{  }\KeywordTok{geom_density}\NormalTok{(}\DataTypeTok{alpha=}\FloatTok{0.6}\NormalTok{,}\DataTypeTok{size =} \FloatTok{.2}\NormalTok{)}\OperatorTok{+}\StringTok{ }
-\StringTok{  }\KeywordTok{scale_fill_manual}\NormalTok{(}\DataTypeTok{values =} \KeywordTok{c}\NormalTok{(}\StringTok{"orangered4"}\NormalTok{,}\StringTok{"chartreuse4"}\NormalTok{, }\StringTok{"skyblue"}\NormalTok{, }\StringTok{"gray"}\NormalTok{))}\OperatorTok{+}
-\StringTok{  }\KeywordTok{scale_x_continuous}\NormalTok{(}\DataTypeTok{breaks =} \KeywordTok{seq}\NormalTok{(}\FloatTok{0.3}\NormalTok{,}\FloatTok{1.5}\NormalTok{, }\DataTypeTok{by =} \FloatTok{0.1}\NormalTok{))}\OperatorTok{+}\KeywordTok{xlab}\NormalTok{(}\StringTok{"cosine distance"}\NormalTok{)}\OperatorTok{+}
-\StringTok{  }\KeywordTok{scale_color_manual}\NormalTok{(}\DataTypeTok{values =} \KeywordTok{c}\NormalTok{(}\StringTok{"orangered4"}\NormalTok{,}\StringTok{"chartreuse4"}\NormalTok{,}\StringTok{"skyblue"}\NormalTok{,}\StringTok{"gray"}\NormalTok{))}\OperatorTok{+}
-\StringTok{  }\KeywordTok{theme_tufte}\NormalTok{()}\OperatorTok{+}\KeywordTok{ggtitle}\NormalTok{(}\StringTok{"Empirical distribution of distances (muslim)"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
+In the currently used methods (like cosine similarity) the values of similarity are often aggregated in a way that may lead to false conclusions. For
+example due to the averaging of values and the lack of confidence interval information.
+
+One can find a number of articles on negative real-life implications resulting from the presence of unaddressed biases in the machine learning models.
 
@@
 
-\normalsize
-
-\noindent Once we focus on words related to islam, the associated bias
-seems to be stronger than in the whole dataset. This is a step towards
-illustrating that the distribution of bias is uneven.
-
-Now, say we want to look at a single protected word. Since the dataset
-also contains comparison multiple control neutral and human attributes,
-we randomly select only 5 from \texttt{none} and 5 from \texttt{human}
-control groups of those for the visualisation purposes.
-
-\vspace{1mm} \footnotesize
+Our research tries to answer the question of how to enhance the current way in which the bias detection is performed to make sure that it is
+methodologically valid.
+
+Our hypothesis is that there can be greater understanding of data and bias implications when confidence interval and Bayesian method are applied to the
+methodology.
+
+\hypertarget{bayesian-analysis-of-cosine-based-bias}{%
+\chapter{Bayesian analysis of cosine-based bias}\label{bayesian-analysis-of-cosine-based-bias}}
+
+Placeholder
 
@@
 
-\normalsize
+\leavevmode\hypertarget{ref-Caliskan2017Semantics}{}%
+Islam, A.
+C., Bryson, J.
+J., \& Narayanan, A.
+(2016).
+Semantics derived automatically from language corpora necessarily contain human biases.
+\emph{CoRR}, \emph{abs/1608.07187}.
+Retrieved from \url{http://arxiv.org/abs/1608.07187}
 

### 2021-06-08 thesis/_book/_main.tex
revised point 2

--- 2021-06-08 thesis/_book/_main.tex
@@
 
-\textbf{state the general topic and give some background}
-
 Natural language processing (NLP) is a subfield of computer science that
@@
 used to accomplish these tasks need a lot of data to learn from. This
-data originates from humans activities and historical recordings like
-texts, messages, speeches. It turns out that in the learning process
+data originates from humans activities and historical recordings such as
+texts, messages or speeches. It turns out that in the learning process
 these models can learn implicit biases that reflect harmful
 stereotypical thinking still present in modern societies. One can find
-methods that aim at identifying hidden biases and/or try to remove them
-by modifying the models explicitly. There are many different types of
-models in NLP depending on a task that they ought to solve. However all
-of them need as an input words represented as numerical values and this
-is accomplished with word embedding models. The biases seem to have
-their primary source in the way the words are assigned the numerical
-values. \newline
-
-\textbf{review of the literature related to the topic}
-
-There is a bunch of literature available on the topic of bias detection
-and mitigation in NLP models. Bolukbasi2016Man focuses on gender biases
-that may be observable while investigating the representation of job
-occupations and gender in terms of their assigned numerical values. The
-authors apply cosine similarity measurement to investigate the
-phenomenon where jobs stereotypically associated with a given gender are
-in fact in the model situated closer to this gender.
+methods that aim at identifying and measuring hidden biases and/or try
+to remove them by modifying the models. There are many different types
+of models in NLP depending on a task that they are supposed to solve.
+However, all of them need as an input words represented by means of
+numbers and this is accomplished with word embedding models. The models
+usually assign the values based on the context in which the words
+appear. It means that the input data can have enormous influence on the
+outcome. The biases seem to have their primary source in the way the
+words are assigned the numerical values.
+
+There is considerable amount of literature available on the topic of
+bias detection and mitigation in NLP models. Bolukbasi, Chang, Zou,
+Saligrama, \& Kalai (\protect\hyperlink{ref-Bolukbasi2016Man}{2016})
+focuses on gender biases that may be observable while investigating the
+representation of job occupations and gender in terms of their assigned
+numerical values. The authors apply cosine similarity measurement to
+investigate the phenomenon where (the vectors corresponding to) words
+related to jobs that are stereotypically associated with a given gender
+are in fact in the model situated closer to this gender. They also use
+analogy tasks to evaluate if the bias is present in the word embedding
+model. They check analogies by comparing pairs of word vectors, for
+example they search for the word complementing the puzzle: man is to
+doctor as woman is to \ldots{}? First they subtract word ``man'' from
+word ``woman'' and then they search for the ranked list of other words
+pairs that have similar vectors' difference. They also include in the
+formula a threshold to ensure that the resulting pairs could not be
+randomly picked.
+
+However, as in Nissim, Noord, \& Goot
+(\protect\hyperlink{ref-Nissim2019Fair}{2019}) it is pointed out, there
+are some limitations of this approach. According to the authors in
+practice most of analogies implementations do not return any input
+words. This means that it does not make sense to expect the algorithm to
+return the same profession for both woman and man. Therefore this method
+seems to be limited in terms of bias detection. The other problems
+regard for example the choice of pairs and words that are used to detect
+the presence of discrimination as it is often subjective and without
+proper justification. Additionally the choice of parameter set in
+Bolukbasi et al. (\protect\hyperlink{ref-Bolukbasi2016Man}{2016})
+formula to ensure that word pairs are not picked by random, is also not
+justified and changing it drastically influences the results.
 
@@
 
-manzini2019black proposes novel way of using a cosine similarity method
-to get the information on assumed resemblance between words. They
-investigate an approach that enables to measure the bias for a class
-(like gender, religion, race) and express the final result with only one
-averaged number. \newline
-
-\textbf{define the terms and scope of the topic}
+Manzini, Lim, Tsvetkov, \& Black
+(\protect\hyperlink{ref-manzini2019black}{2019}) proposes a novel way of
+using cosine similarity to obtain the information on assumed resemblance
+between words. They investigate an approach that enables them to measure
+the bias for a class (like gender, religion, race) and express the final
+result with a single metric.
 
 It is worth noticing the general distinction of biases mentioned in
-Caliskan2017Semantics. They refer to the publication concerning Implicit
-Association Test Greenwald et al., 1998) where certain baseline of bias
-phenomenon was introduced. Namely it seems that humans naturally exhibit
-some biases and that not always bring social concern. One can imagine
-the intuitive associations between for example insects and flowers, and
-the feelings of pleasantness or unpleasantness. In general people would
-rather associate flowers with feeling pleasant than insects and this
-preference could be named a bias or prejudice in some direction. However
-this type of preference does not cause an uproar and it is rather
-morally neutral case. Unfortunately there are other biases and
-prejudices that directly influence the quality of other people's lives
-and therefore they should be taken care of.
+Islam et al. (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}). They
+refer to the publication concerning Implicit Association Test Greenwald
+et al., 1998) where certain baseline of bias phenomenon was introduced.
+Namely it seems that humans naturally exhibit some biases and that not
+always bring social concern. One can imagine the intuitive associations
+between for example insects and flowers, and the feelings of
+pleasantness or unpleasantness. In general people would rather associate
+flowers with feeling pleasant than insects and this preference could be
+named a bias or prejudice in some direction. However this type of
+preference does not cause an uproar and it is rather morally neutral
+case. Unfortunately there are other biases and prejudices that directly
+influence the quality of other people's lives and therefore they should
+be taken care of.
 
@@
 implications into the real-life applications may change as well as it
-was pointed out in Mehrabi2019Survey. They mark out that there exist
-different types of biases, the list if long but among others there are
-historical bias, representation bias, measurement bias. This indicates
-how complex the process itself is. Without the proper understanding and
-awareness of the problem, people are prone to unconsciously sustain this
-phenomenon.
+was pointed out in Mehrabi, Morstatter, Saxena, Lerman, \& Galstyan
+(\protect\hyperlink{ref-Mehrabi2019Survey}{2019}). They mark out that
+there exist different types of biases, the list if long but among others
+there are historical bias, representation bias, measurement bias. This
+indicates how complex the process itself is. Without the proper
+understanding and awareness of the problem, people are prone to
+unconsciously sustain this phenomenon.
 
@@
 the context and application different definitions may be applied.
-\newline
-
-\textbf{outline the current situation} The most popular methods focus on
-comparing the similarity between words from protected groups and those
-that are considered to be stereotypical or harmful in some way. One can
-find in this group methods such as euclidean distance, dot product or
-cosine similarity. There are also other ways to detect the effects of
-biases. For example through the investigation of the model performance
-on certain tasks that validate if the model returns some values
-independently on gender or race or not. \newline
-
-\textbf{evaluate the current situation (advantages/ disadvantages) and identify the gap}
+
+The most popular methods focus on comparing the similarity between words
+from protected groups and those that are considered to be stereotypical
+or harmful in some way. One can find in this group methods such as
+euclidean distance, dot product or cosine similarity. There are also
+other ways to detect the effects of biases. For example through the
+investigation of the model performance on certain tasks that validate if
+the model returns some values independently on gender or race or not.
 
@@
 conclusions. For example due to the averaging of values and the lack of
-confidence interval information. \newline
-
-\textbf{identify the importance of the proposed research} One can find a
-number of articles on negative real-life implications resulting from the
-presence of unaddressed biases in the machine learning models. \newline
-
-\textbf{state the research problem/ questions}
+confidence interval information.
+
+One can find a number of articles on negative real-life implications
+resulting from the presence of unaddressed biases in the machine
+learning models.
 
@@
 natural language models are limited in terms of confidence interval.
-\newline
-
-\textbf{state the research aims and/or research objectives}
 
@@
 way in which the bias detection is performed to make sure that it is
-methodologically valid. \newline
-
-\textbf{state the hypotheses}
+methodologically valid.
 
@@
 bias implications when confidence interval and Bayesian method are
-applied to the methodology. \newline
-<!-- \textbf{outline the order of information in the thesis} -->
-
-\textbf{outline the methodology}
-
-To discuss!
+applied to the methodology.
 
@@
 
-Bolukbasi, Chang, Zou, Saligrama, \& Kalai
-(\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) define similarity
-between words as the inner product of their normalized vectors (that is,
-they use cosine similarity). They focus on examining what the geometry
-of word embedding is in regard to ``he'' and ``she'' words. In other
-words, whether the similarity between those concepts and other words
-reflects expected gender stereotypes. They test this hypothesis by
-investigating whether there is a connection between word embeddings
-representing certain professions and words referring to gender. They
-also evaluate whether automatically produced analogies between words
-reflect the stereotypes as well.
+Bolukbasi et al. (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) define
+similarity between words as the inner product of their normalized
+vectors (that is, they use cosine similarity). They focus on examining
+what the geometry of word embedding is in regard to ``he'' and ``she''
+words. In other words, whether the similarity between those concepts and
+other words reflects expected gender stereotypes. They test this
+hypothesis by investigating whether there is a connection between word
+embeddings representing certain professions and words referring to
+gender. They also evaluate whether automatically produced analogies
+between words reflect the stereotypes as well.
 
@@
 
-In the article Manzini, Lim, Tsvetkov, \& Black
+In the article Manzini et al.
 (\protect\hyperlink{ref-manzini2019black}{2019}) the authors present
@@
 
+\hypertarget{ref-Mehrabi2019Survey}{}
+Mehrabi, N., Morstatter, F., Saxena, N., Lerman, K., \& Galstyan, A.
+(2019). A survey on bias and fairness in machine learning. \emph{CoRR},
+\emph{abs/1908.09635}. Retrieved from
+\url{http://arxiv.org/abs/1908.09635}
+
+\hypertarget{ref-Nissim2019Fair}{}
+Nissim, M., Noord, R. van, \& Goot, R. van der. (2019). Fair is better
+than sensational: Man is to doctor as woman is to doctor. \emph{CoRR},
+\emph{abs/1905.09866}. Retrieved from
+\url{http://arxiv.org/abs/1905.09866}
+
 \end{document}

### 2021-06-08 thesis/index.Rmd
Merge branch 'main' of github.com:rfl-urbaniak/BayesBias

--- 2021-06-08 thesis/index.Rmd
@@
 link-citations: yes
+fontsize: 12pt
+linestretch: 1.5
+geometry: left=3.5cm, right=2.5cm, top=2.5cm, bottom=2.5cm, asymmetric, includeheadfoot
 csl: apa-5th-edition.csl
@@
 
-# Preface {-}
+<!-- # Preface {-} -->
 
-Test
+<!-- Test  -->
 

### 2021-06-09 thesis/01-intro.Rmd
Merge branch 'main' of github.com:rfl-urbaniak/BayesBias
Adds updates on cosine

@@
 @Caliskan2017Semantics touches upon the topic of biases regarding race and gender.
-They apply knowledge from well-known psychological studies like Implicit Association Test to research the relation between human stereotypical thinking and model learnt biases to
+They apply knowledge from well-known psychological studies such as Implicit Association Test to research the relation between human stereotypical thinking and model learnt biases to
 discover close relationship between these two. For the evaluation they use Word Embedding Association Test (WEAT) and the Word Embedding Factual Association Test (WEFAT).
@@
 It is worth noticing the general distinction of biases mentioned in @Caliskan2017Semantics. They refer to the publication concerning Implicit Association
-Test Greenwald et al., 1998) where certain baseline of bias phenomenon was introduced. Namely it seems that humans naturally exhibit some biases and
+Test (Greenwald et al., 1998) where certain baseline of bias phenomenon was introduced. Namely it seems that humans naturally exhibit some biases and
 that not always bring social concern.

### 2021-06-09 thesis/02-cosine.Rmd
Adds updates on cosine

@@
 ## Word embeddings
+<!-- [based on https://machinelearningmastery.com/what-are-word-embeddings/] -->
 
@@
 
-Word embeddings have many applications in natural language processing.
-They are handy in document search and information retrieval.
-They also play their part in improving automatic translations.
-Well learned word representations may also contribute to the improvement of sentiment analysis or spam detection.
-
 The numbers occurring in such representations are not random.
@@
 After preparing such a fitted model one takes only the learned weights from a neural network, and uses them as vectors in a word embeddings representation.
+
+Word embeddings have many applications in natural language processing.
+They are handy in document search and information retrieval.
+They also play their part in improving automatic translations.
+Well learned word representations may also contribute to the improvement of sentiment analysis or spam detection.
 
@@
 However, it is often argued that in comparing words in terms of this metric, the magnitude of vectors may be treated as irrelevant, as the most important information pertains to direction.
-<!-- todo: find reference--> In what follows, it is important to distinguish between cosine similarity and cosine distance, defined as:
+<!-- todo: find reference-->
+
+In what follows, it is important to distinguish between cosine similarity and cosine distance, defined as:
 \begin{align} \tag{Sim}
@@
 
-@Bolukbasi2016Man  define similarity between words as the inner product of their normalized vectors (that is, they use cosine similarity).
+@Bolukbasi2016Man
+ define similarity between words as the outcome inner product of their normalized vectors. 
 They focus on examining what the geometry of word embedding is in regard to "he" and "she" words.
@@
 A very vivid way to follow their method of arguing that bias in word embeddings is real is to plot the values of inner product of chosen words.
-The plot below does not originate from the original paper (it comes from [https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings](https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings)) but  a similar visualization may be found there.
+The plot below does not originate from the original paper (it is from https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings) but similar visualization may be found there.
 Data used to create our plot is as follows.
@@
 
-<!-- ![gender bias in professions as measured by inner product](gender_professions.png) -->
-
-\vspace{1mm}
-\footnotesize
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
-genderProjections <-  read.csv("../datasets/genderProjections.csv")[,-1]
-colnames(genderProjections)[4] <- "Stereotype"
-levels(genderProjections$Stereotype) <- c("male occupation", "female occupation")
-ggplot(genderProjections, aes(x=he,y=she, color = Stereotype, shape = Stereotype))+geom_point(size = 2, alpha = 0.75)+theme_tufte()+labs(title ="Stereotypical professions projected on the he-she axis (dot product, unnormalized)", subtitle ="GloVe embeddings trained on the Wikipedia 2014 and Gigaword 5th Edition corpus")+scale_color_manual(values =c("orangered4","chartreuse4"))
-```
-\normalsize
+![gender bias in professions as measured by inner product](gender_professions.png)
 
@@
 
-In the article
-@manzini2019black
-the authors present different approach towards finding similarities between classes of words.
+@manzini2019black present a different approach towards finding similarities between classes of words.
 The authors claim that texts available online are full of direct or indirect human stereotypes.
@@
 Cosine distance is used in the article as a measure to first argue for the existence of multi-class bias and then to show how through bias mitigation techniques the bias may be decreased.
-The main steps in the procedure are as follows.
 
@@
 
-As the code has been provided by the authors we were able to reconstruct their results. Let us go through an example that refers to the process of hard debiasing on religious attributes.
+As the code has been provided by the authors we were able to reconstruct their results.
+The main steps in the procedure are as follows.
+Let us go through an example that refers to the process of hard debiasing on religious attributes.
 
@@
 The word embeddings geometry should not place this group close to harmful stereotypes, if it is to be bias-free.
+\newline
 
@@
 
-We have prepared a dataset including the values of cosine distance for each protected word with each attribute (stereotype), its head is displayed in the table.
+We have prepared a table presenting the values of cosine distance for each protected word with each attribute (stereotype). The part of the results is shown below.
 
-\footnotesize 
-```{r religionTableHeadEarly,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
-religion <- read.csv("../datasets/religionReddit.csv")[-1]
-colnames(religion) <- c("protectedWord","wordToCompare","wordClass",
-                        "cosineDistance","cosineSimilarity","connection")
-levels(religion$wordClass) <- c("christian","human","jewish","muslim","neutral")
-head(religion)  %>%  kable(format = "latex",booktabs=T,
-                      linesep = "",  escape = FALSE, 
-                      caption = "Head of the religion dataset.") %>%
-                      kable_styling(latex_options=c("scale_down"))
+```{python}
+import pandas as pd
+data = pd.read_csv("religion_cosine_table.csv")
+print(data.head())
+
 ```
-\normalsize
-
-\pagebreak
 
@@
 The authors introduced a metric that tries to generalize the presence of bias through the classification of multi-class bias in groups of words connected with gender, religion or race.
-
-In the process they first take the mean of cosine distances between a given protected word and each stereotype.
-They do not differentiate between stereotypes associated with a word and stereotypes associated with different words (in case of religion stereotypes characteristic for christianity has also cosine distance measured with judaism or islam).
-Then after collecting the list of mean cosine distances, they average the list to get one final value represeting whole group, in this example religion.
-In case of religion, the final mean of all mean distances is equal to 0.859.
+In the process they first take the mean of cosine distances between a given protected word and attributes assigned to each stereotype.
+They do not differentiate between stereotypes associated with a word and stereotypes associated with different words (in case of religion, stereotypes characteristic for Christianity has also cosine distance measured with for instance, Judaism or Islam).
+Then, after collecting the list of mean cosine distances, they average the list to obtain one final value representing the whole group, in this example religion, for which the final mean of all mean distances is equal to 0.859.
 
@@
 This strategy is based on the idea that different dimensions of vectors contain different types of information and in some words in vector layers (subspaces) the information about religiousness is implicitly conveyed.
-In some cases this knowledge is useful but in case of harmful stereotypes one does not want to have religion concept in stereotypical words.
+In some cases this knowledge is useful but in the case of harmful stereotypes one does not want to have religion concept in stereotypical words.
 
-After finding bias subspace they use it to modify the vectors values individually so that their cosine distances towards certain words are changed.
-In case of stereotypes the aim is to make the cosine distances larger so that the association between protected word and harmful stereotype is smaller.
+After finding bias subspace, they use it to modify the vector values individually so that their cosine distances towards certain words are changed.
+In the case of stereotypes the aim is to make the cosine distances larger so that the association between protected word and harmful stereotype is smaller.
 
@@
 
-4.
-Finally there is a multi-class evaluation.
-They check each protected attribute with each group of debiased words.
-Then they do each time mean from the cosine distances from protected word to neutral words.
-Then they do a mean from all of these means to return one value.
-If the returned value of cosine distance is greater than the one calculated at the very beginning then it means that the bias was removed as the similarity between stereotypical word and the protected one decreased.
-
-How are the multi-class biases actually defined?
-The authors take a list of alleged biases towards protected groups like gender, race, religion from various scientific reports.
-The multi-class bias existence is further proved by providing the reader with the value of a mean from the cosine similarity between protected word and list of of previously defined biases.
+4. In the final step there is an evaluation of the results. The cosine distances are calculated again but this time using debiased vocabulary. After
+taking the mean of all distances one final value is obtained and then it is compared with the average value from the beginning.
+If the cosine distances are on average greater than before then it leads the authors to the conclusion that improvement was made.
+As the cosine distance increases it is assumed that the association between protected and stereotypical words decreases.
 
@@
 
-# Limitations of the approach
+## Limitations of the approach
 
-## Selection and number of attributes
+\textbf{1. Selection of attributes}
 
@@
 
-The attributes are taken from different sources, therefore there is no one justification for their choice.
+The attributes are taken from different sources, there is no principled justification for their choice.
 
-<!-- TODO: Universal ways to decide on the number of attributes so that proving bias is significant? -->
+There is no mention of methodology for deciding on the number of attributes necessary to prove a hypothesis on the given size of dataset.
+There are however some ways to estimate how many samples we need to make sure that the result is significant.
+\newline
 
-<!-- https://cs.stanford.edu/people/wmorgan/sigtest.pdf -->
-
-In the article there is no mention of methodology for deciding on the number of attributes necessary to prove a hypothesis on the given size of dataset.
-There are however some ways to estimate how many samples we need to make sure that the result is significant.
-
-## No control groups
-
-\textbf{Control group}
-\todo{Tutaj raczej powiedz ze nie maj agrup a ze my bedziemy mieli}
-
-How to properly prepare control group in terms of quality and quantity?
-
-## Means of means
+\textbf{Taking the mean of cosines}
 
@@
 
-## Interpretability issues
+\textbf{Interpreting the results}
 
@@
 It seems like similar practic could be beneficial for metrics aimed at measuring similarity between words.
+\newline
+<!-- TODO: numerical test if this distance is significant? check mean mean cosine distances of some random words? -->
 
-##  The curse of dimensionality
+\textbf{The curse of dimensionality}
 
@@
 
-![curse of dimensionality,  number of dimensions on the x axis,  standard deviation of similarity on the y axis](../images/curseOfDimensionality.png)
+![curse of dimensionality, on x-axis number of dimensions, on y-axis standard deviation of similarity](curseOfDimensionality.png)
 
-<!-- \textbf{Verifying other similarity meaures} -->
+\textbf{Control group}
 
-<!-- Besides cosine similarity, there are other methods used to find the similarity between vectors.   -->
+How to properly prepare control group in terms of quality and quantity?
+\newline
+<!-- + check if it may be proved in the code -->
 
-<!-- a) TS-SS -->
-<!-- <!-- todo: check if it makes sense to calculate for all of the datasets; interpretations etc. -->  -->
+\textbf{Verifying other similarity meaures}
 
-<!-- <!-- https://www.researchgate.net/publication/303513110_A_Hybrid_Geometric_Approach_for_Measuring_Similarity_Level_Among_Documents_and_Document_Clustering --> -->
+Besides cosine similarity, there are other methods used to find the similarity between vectors.
 
-<!-- b) WEAT test  -->
-<!-- <!-- https://docs.responsibly.ai/word-embedding-bias.html#module-responsibly.we.weat --> -->
+a) TS-SS
+<!-- todo: check if it makes sense to calculate for all of the datasets; interpretations etc. -->
 
-<!-- <!-- [to check https://github.com/taki0112/Vector_Similarity] --> -->
+<!-- https://www.researchgate.net/publication/303513110_A_Hybrid_Geometric_Approach_for_Measuring_Similarity_Level_Among_Documents_and_Document_Clustering -->
 
-<!-- <!-- TODO: check in code if they make more sense --> -->
+b) WEAT test 
+<!-- https://docs.responsibly.ai/word-embedding-bias.html#module-responsibly.we.weat -->
+
+<!-- [to check https://github.com/taki0112/Vector_Similarity] -->
+
+<!-- TODO: check in code if they make more sense -->
 

### 2021-06-09 thesis/_book/_main.tex
Adds updates on cosine

--- 2021-06-08 thesis/_book/_main.tex
@@
 \documentclass[
+  12pt,
 ]{book}
@@
 \usepackage{lmodern}
+\usepackage{setspace}
 \usepackage{ifxetex,ifluatex}
@@
 \urlstyle{same} % disable monospaced font for URLs
+\usepackage[left=3.5cm, right=2.5cm, top=2.5cm, bottom=2.5cm, asymmetric, includeheadfoot]{geometry}
 \usepackage{longtable,booktabs,array}
@@
 \author{Alicja Dobrzeniecka}
-\date{2021-06-08}
+\date{2021-06-09}
 
@@
 }
-\hypertarget{preface}{%
-\chapter*{Preface}\label{preface}}
-\addcontentsline{toc}{chapter}{Preface}
-
-Placeholder
-
+\setstretch{1.5}
 \hypertarget{introduction}{%
@@
 \protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) focuses on gender biases that may be observable while investigating the representation of job occupations and gender in terms of their assigned numerical
-values. The authors apply cosine similarity measurement to investigate the phenomenon where jobs stereotypically associated with a given gender are in fact in the model situated closer to this gender.
+values.
+The authors apply cosine similarity measurement to investigate the phenomenon where (the vectors corresponding to) words related to jobs that are stereotypically associated with a given gender are in fact in the model situated closer to this gender.
+They also use analogy tasks to evaluate if the bias is present in the word embedding model.
+They check analogies by comparing pairs of word vectors, for example they search for the word complementing the puzzle: man is to doctor as woman is to \ldots?
+First they subtract word ``man'' from word ``woman'' and then
+they search for the ranked list of other words pairs that have similar vectors' difference.
+They also include in the formula a threshold to ensure that the resulting pairs could not be randomly picked.
+
+However, as in \protect\hyperlink{ref-Nissim2019Fair}{Nissim, Noord, \& Goot} (\protect\hyperlink{ref-Nissim2019Fair}{2019}) it is pointed out, there are some limitations of this approach.
+According to the authors in practice most of analogies implementations do not return any input words.
+This means that it does not make sense to expect the algorithm to return the same profession for both woman and man.
+Therefore this method seems to be limited in terms of bias detection.
+The other problems regard for example the
+choice of pairs and words that are used to detect the presence of discrimination as it is often subjective and without proper justification.
+Additionally the choice of parameter set in \protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) formula to ensure that word pairs are not picked by random, is also not justified and changing it drastically
+influences the results.
 
@@
 
-\protect\hyperlink{ref-manzini2019black}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-manzini2019black}{2019}) proposes novel way of using a cosine similarity method to get the information on assumed resemblance between words.
+\protect\hyperlink{ref-manzini2019black}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-manzini2019black}{2019}) proposes a novel way of using cosine similarity to obtain the information on assumed resemblance between words.
 They investigate
-an approach that enables to measure the bias for a class (like gender, religion, race) and express the final result with only one averaged number.
+an approach that enables them to measure the bias for a class (like gender, religion, race) and express the final result with a single metric.
 
@@
 
-\hypertarget{bayesian-analysis-of-cosine-based-bias}{%
-\chapter{Bayesian analysis of cosine-based bias}\label{bayesian-analysis-of-cosine-based-bias}}
+\hypertarget{cosine-similarity-and-bias-detection}{%
+\chapter{Cosine similarity and bias detection}\label{cosine-similarity-and-bias-detection}}
 
 Placeholder
+
+\hypertarget{word-embeddings}{%
+\section{Word embeddings}\label{word-embeddings}}
+
+\hypertarget{cosine-similarity-and-distance}{%
+\section{Cosine similarity and distance}\label{cosine-similarity-and-distance}}
+
+\hypertarget{cosine-distance-in-a-one-class-bias-detection}{%
+\section{Cosine distance in a one-class bias detection}\label{cosine-distance-in-a-one-class-bias-detection}}
+
+\hypertarget{cosine-distance-in-a-multi-class-bias-detection}{%
+\section{Cosine distance in a multi-class bias detection}\label{cosine-distance-in-a-multi-class-bias-detection}}
+
+\hypertarget{selection-and-number-of-attributes}{%
+\section{Selection and number of attributes}\label{selection-and-number-of-attributes}}
+
+\hypertarget{no-control-groups}{%
+\section{No control groups}\label{no-control-groups}}
+
+\hypertarget{means-of-means}{%
+\section{Means of means}\label{means-of-means}}
+
+\hypertarget{interpretability-issues}{%
+\section{Interpretability issues}\label{interpretability-issues}}
+
+\hypertarget{the-curse-of-dimensionality}{%
+\section{The curse of dimensionality}\label{the-curse-of-dimensionality}}
+
+\hypertarget{walkthrough-with-the-religion-dataset}{%
+\chapter{Walkthrough with the religion dataset}\label{walkthrough-with-the-religion-dataset}}
+
+Placeholder
+
+\hypertarget{loading-and-understanding-the-dataset}{%
+\section{Loading and understanding the dataset}\label{loading-and-understanding-the-dataset}}
+
+\hypertarget{first-look-at-the-empirical-distributions}{%
+\section{First look at the empirical distributions}\label{first-look-at-the-empirical-distributions}}
+
+\hypertarget{looking-at-the-islam-related-words}{%
+\section{Looking at the islam-related words}\label{looking-at-the-islam-related-words}}
+
+\hypertarget{bayesian-model-structure-and-assumptions}{%
+\section{Bayesian model structure and assumptions}\label{bayesian-model-structure-and-assumptions}}
+
+\hypertarget{choosing-predictors}{%
+\section{Choosing predictors}\label{choosing-predictors}}
+
+\hypertarget{dataset-level-coefficients}{%
+\section{Dataset-level coefficients}\label{dataset-level-coefficients}}
+
+\hypertarget{model-structure-and-assumptions}{%
+\section{Model structure and assumptions}\label{model-structure-and-assumptions}}
+
+\hypertarget{protected-classes-in-reddit-and-google-embeddings}{%
+\section{Protected classes in Reddit and Google embeddings}\label{protected-classes-in-reddit-and-google-embeddings}}
+
+\hypertarget{dataset-level-coefficients-after-debiasing}{%
+\section{Dataset-level coefficients after debiasing}\label{dataset-level-coefficients-after-debiasing}}
+
+\hypertarget{protected-classes-after-debiasing}{%
+\section{Protected classes after debiasing}\label{protected-classes-after-debiasing}}
 
@@
 
+\leavevmode\hypertarget{ref-Nissim2019Fair}{}%
+Nissim, M., Noord, R.
+van, \& Goot, R.
+van der.
+(2019).
+Fair is better than sensational: Man is to doctor as woman is to doctor.
+\emph{CoRR}, \emph{abs/1905.09866}.
+Retrieved from \url{http://arxiv.org/abs/1905.09866}
+
 \end{CSLReferences}

### 2021-06-09 thesis/01-intro.Rmd
New updates

@@
 
-It is worth noticing the general distinction of biases mentioned in @Caliskan2017Semantics. They refer to the publication concerning Implicit Association
-Test (Greenwald et al., 1998) where certain baseline of bias phenomenon was introduced. Namely it seems that humans naturally exhibit some biases and
-that not always bring social concern.
+It is worth noticing the general distinction of biases mentioned in @Caliskan2017Semantics.
+They refer to the publication concerning Implicit Association Test (Greenwald et al., 1998) that measures the strength of associations between concepts or stereotypes by calculating the time of reaction for special tasks.
+It is worth noticing that humans naturally exhibit some biases and that they do not always cause social concern.
 One can imagine the intuitive associations between for example insects and flowers, and the feelings of pleasantness or unpleasantness.
-In general people would rather associate flowers with feeling pleasant than insects and this preference could be named a bias or
-prejudice in some direction. However this type of preference does not cause an uproar and it is rather morally neutral case. Unfortunately there are other
+In general, people would rather associate flowers with feeling pleasant than insects, and this preference could be named a bias or
+prejudice in some direction. However, this type of preference does not cause an uproar and is a rather morally neutral case. Unfortunately, there are other
 biases and prejudices that directly influence the quality of other people's lives and therefore they should be taken care of.
 
-One can find a bunch of various definitions trying to capture what bias and fairness actually are.
-With the choice of the definition, implications into the real-life applications may change as well as it was pointed out in @Mehrabi2019Survey.
-They mark out that there exist different types of biases, the list if long but among others there are historical bias, representation bias, measurement bias.
-This indicates how complex the process itself is.
-Without the proper understanding and awareness of the problem, people are prone to unconsciously sustain this phenomenon.
+One can find various definitions trying to capture what bias and fairness actually are.
+With the choice of the definition, implications into the real-life applications may change as well.
+@Mehrabi2019Survey mark out that there exist different types of biases such as historical bias, representation bias, measurement bias (the list is long).
+This indicates how complex the issue of bias is.
+Without the proper understanding and awareness of the problem, people are prone to unconsciously sustain the bias existence.
 
-In the article one can also find the distinction on different types of discrimination, some of them will be shortly described.
-It is worth first mentioning that protected attributes are those qualities, traits or characteristics that one cannot, according to the law, discriminated against.
-Direct discrimination refers to the situation when protected attributes of individuals explicitly result in non-favorable outcomes toward them.
+@Mehrabi2019Survey also distinguish different types of discrimination, some of them will be briefly described.
+By protected attributes we mean those qualities, traits or characteristics that one cannot legally discriminate against.
+Direct discrimination occurs when protected attributes of individuals explicitly result in non-favorable outcomes toward them.
 In contrast in indirect discrimination individuals appear to be treated equally but anyway they end up being treated unjustly due to the hidden effects of biases towards their protected attributes.
@@
 The most popular methods focus on comparing the similarity between words from protected groups and those that are considered to be stereotypical or harmful in some way.
-One can find in this group methods such as euclidean distance, dot product or cosine similarity.
+One can find in this group methods such as euclidean distance or cosine similarity (which is equivalent to dot product if the
+vectors are normalized).
 There are also other ways to detect the effects of biases.
@@
 
-In the currently used methods (like cosine similarity) the values of similarity are often aggregated in a way that may lead to false conclusions. For 
-example due to the averaging of values and the lack of confidence interval information.
+The currently used methods (such as cosine similarity) make the similarity values often aggregated in a way that may lead to hasty conclusions.
+The averaging of values and the lack of uncertainty may lead to the incomplete picture of the bias situation in the vocabulary.
 
@@
 
-In the paper we indicate how current methods used to detect biases in natural language models are limited in terms of confidence interval.
+In the paper we indicate how current methods used to detect biases in natural language models are limited from the perspective of Bayesian analysis.
 
@@
 
-Our research tries to answer the question of how to enhance the current way in which the bias detection is performed to make sure that it is
+Our research enhances the current way in which the bias detection is performed to make sure that it is
 methodologically valid.
@@
 
-Our hypothesis is that there can be greater understanding of data and bias implications when confidence interval and Bayesian method are applied to the
-methodology.
+The key hypothesis is that greater understanding of data and bias implications can be achieved when Bayesian methods are applied to issue.
 

### 2021-06-09 thesis/_book/_main.tex
New updates

@@
 \protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) touches upon the topic of biases regarding race and gender.
-They apply knowledge from well-known psychological studies like Implicit Association Test to research the relation between human stereotypical thinking and model learnt biases to
+They apply knowledge from well-known psychological studies such as Implicit Association Test to research the relation between human stereotypical thinking and model learnt biases to
 discover close relationship between these two. For the evaluation they use Word Embedding Association Test (WEAT) and the Word Embedding Factual Association Test (WEFAT).
@@
 It is worth noticing the general distinction of biases mentioned in \protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}).
-They refer to the publication concerning Implicit Association
-Test Greenwald et al., 1998) where certain baseline of bias phenomenon was introduced. Namely it seems that humans naturally exhibit some biases and
-that not always bring social concern.
+They refer to the publication concerning Implicit Association Test (Greenwald et al., 1998) that measures the strength of associations between concepts or stereotypes by calculating the time of reaction for special tasks.
+It is worth noticing that humans naturally exhibit some biases and that they do not always cause social concern.
 One can imagine the intuitive associations between for example insects and flowers, and the feelings of pleasantness or unpleasantness.
-In general people would rather associate flowers with feeling pleasant than insects and this preference could be named a bias or
-prejudice in some direction. However this type of preference does not cause an uproar and it is rather morally neutral case. Unfortunately there are other
+In general, people would rather associate flowers with feeling pleasant than insects, and this preference could be named a bias or
+prejudice in some direction. However, this type of preference does not cause an uproar and is a rather morally neutral case. Unfortunately there are other
 biases and prejudices that directly influence the quality of other people's lives and therefore they should be taken care of.
@@
 
-\hypertarget{selection-and-number-of-attributes}{%
-\section{Selection and number of attributes}\label{selection-and-number-of-attributes}}
-
-\hypertarget{no-control-groups}{%
-\section{No control groups}\label{no-control-groups}}
-
-\hypertarget{means-of-means}{%
-\section{Means of means}\label{means-of-means}}
-
-\hypertarget{interpretability-issues}{%
-\section{Interpretability issues}\label{interpretability-issues}}
-
-\hypertarget{the-curse-of-dimensionality}{%
-\section{The curse of dimensionality}\label{the-curse-of-dimensionality}}
+\hypertarget{limitations-of-the-approach}{%
+\section{Limitations of the approach}\label{limitations-of-the-approach}}
 

### 2021-06-09 thesis/_main.Rmd
New updates

--- 2021-06-09 thesis/02-cosine.Rmd
@@
+---
+title: "Conceptual and methodological problems with bias detection and avoidance in natural language processing"
+author: "Alicja Dobrzeniecka"
+date: '`r Sys.Date()`'
+output:
+  bookdown::pdf_book: 
+    keep_tex: yes
+    toc_depth: 6
+    latex_engine: xelatex
+    extra_dependencies: ["todonotes"]
+  bookdown::gitbook:
+    lib_dir: "book_assets"
+includes:
+    in_header:
+      - Rafal_latex6.sty
+documentclass: book
+classoptions: dvipsnames, usenames
+site: bookdown::bookdown_site
+bibliography: [../references/bibliographyRafal.bib]
+biblio-style: 
+link-citations: yes
+fontsize: 12pt
+linestretch: 1.5
+geometry: left=3.5cm, right=2.5cm, top=2.5cm, bottom=2.5cm, asymmetric, includeheadfoot
+csl: apa-5th-edition.csl
+---
+
+<!--chapter:end:index.Rmd-->
+
+# Introduction
+
+Placeholder
+
+<!--chapter:end:01-intro.Rmd-->
+
+---
+output:
+  pdf_document: default
+  html_document: default
+---
 # Cosine similarity and bias detection
@@
 
-![gender bias in professions as measured by inner product](gender_professions.png)
+\vspace{1mm}
+\footnotesize
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+genderProjections <-  read.csv("../datasets/genderProjections.csv")[,-1]
+colnames(genderProjections)[4] <- "Stereotype"
+levels(genderProjections$Stereotype) <- c("male occupation", "female occupation")
+ggplot(genderProjections, aes(x=he,y=she, color = Stereotype, shape = Stereotype))+geom_point(size = 2, alpha = 0.75)+theme_tufte()+labs(title ="Stereotypical professions projected on the he-she axis (dot product, unnormalized)", subtitle ="GloVe embeddings trained on the Wikipedia 2014 and Gigaword 5th Edition corpus")+scale_color_manual(values =c("orangered4","chartreuse4"))
+```
+\normalsize
 
@@
 
-```{python}
-import pandas as pd
-data = pd.read_csv("religion_cosine_table.csv")
-print(data.head())
+\footnotesize 
+```{r religionTableHeadEarly,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+religion <- read.csv("../datasets/religionReddit.csv")[-1]
+colnames(religion) <- c("protectedWord","wordToCompare","wordClass",
+                        "cosineDistance","cosineSimilarity","connection")
+levels(religion$wordClass) <- c("christian","human","jewish","muslim","neutral")
+head(religion)  %>%  kable(format = "latex",booktabs=T,
+                      linesep = "",  escape = FALSE, 
+                      caption = "Head of the religion dataset.") %>%
+                      kable_styling(latex_options=c("scale_down"))
+```
+\normalsize
 
-```
+\pagebreak
 
@@
 
-The attributes are taken from different sources, there is no principled justification for their choice.
+The attributes are taken from different sources, there is no principled justification for their choice. From our analysis it will become clear that
+the list is rather uneven.
 
-There is no mention of methodology for deciding on the number of attributes necessary to prove a hypothesis on the given size of dataset.
+There is no mention of methodology for deciding on the number of attributes necessary to decide a hypothesis on the given size of dataset.
 There are however some ways to estimate how many samples we need to make sure that the result is significant.
-\newline
+Our research will show that the numbers used
+are rather insufficient.
 
-\textbf{Taking the mean of cosines}
+\textbf{2. No control while taking the mean of cosines}
 
-The authors use the mean average cosine similarity to check on multi-class similarity between protected word and harmful stereotypes. Details of the process are described as follows.
+The authors use the mean average cosine similarity to check on multi-class similarity between protected word and harmful stereotypes. 
+They average the results until they obtain one final value to represent the mean cosine distance between protected word from a given class and the
+attributes of that class. As there is no control of the individual values of cosine distances, one can observe values that have cosine distances
+greater than 1 which would suggest that these compare words do not have significant similarity. However, as there is a lack of control, even such
+words are taken into the account for calculations.
 
-When one calculates individual cosine similarity between some chosen word like "christianity" and neutral attributes that are supposed to be biased are actually often having very small and negative similarity value.
-As a result the measurement cosine distance is for such pair of words greater than 1.
-The similarity interpretation for that would suggest that the words are not similar as their cosine distance is so enormous.
+\textbf{3. The lack of uncertainty}
 
-Doing a mean hides this issue and as there are pairs having negative and small similarities and there are those that have similarity equal to 0.5, the resulting calculation seems to be in norm.
-Additionally in such method the uncertainty interval is also not included which makes it even more difficult to give reasonable interpretations of the results.
+A mean hides this issue and as there are pairs having negative and small similarities and there are those that have similarity equal to 0.5, the resulting calculation seems to be in norm.
+Additionally in such method the uncertainty is also not included which makes it even more difficult to give reasonable interpretations of the results.
+We propose the use of Bayesian method to obtain some understanding of the influence the uncertainty has on the interpretation of final results.
+
+\textbf{4. No word class distinction}
+
 In the original paper words from all three religions were checked with all of the stereotypes which means that there was no distinction on whether the stereotype is associated with given religion or with the other.
 Not all of the stereotypical words should be considered as harmful for all of the religions.
-In our research we distinguished between associated stereotypes, not associated (stereotype is valid but for different religion than currently checked) and none (meaning neutral words).
-\newline
+In our analysis we distinguished between associated stereotypes, not associated (stereotype is valid but for different religion than currently checked) and none (meaning neutral words).
 
-\textbf{Interpreting the results}
+\textbf{5. Interpreting the results}
 
 Assuming for a moment that the value of multi-class cosine distance is correct, one may question the results' interpretation.
-In [Black is to Criminal as Caucasian is to Police:Detecting and Removing Multiclass Bias in Word Embeddings] Table 2, there are summarized the averages of cosine distance per group (gender, race, religion).
+In @manzini2019black Table 2, there are summarized the averages of cosine distance per group (gender, race, religion).
 I would like to focus  now on analyzing the values relating to religious biases.
@@
 
-MAC stand for mean average cosine similarity although in reality the values of cosine distance are there stored.
+MAC stand for mean average cosine similarity although in reality the values of cosine distance are stored there.
 What may attract attention is the fact that the value of cosine distance in "Biased" category is already quite high even before "debiasing".
@@
 However the authors aim to mitigate "biases" in vectors with such great distance to make it even larger.
-Methodologically there is a question on what bases this small similarity is still considered as a proof of bias presence.
+Methodologically there is a question on what basis this small similarity is still considered as a proof of bias presence.
 
 This is in general the problem of scale and the lack of universal intervals that could be applied to know whether the cosine similarity is high enough to considered given words more similar that if we chose them on random.
-In other metrics like Pearson coefficient we are provided with artificial interval that tries to make the interpretations more objective with a point of reference.
-It seems like similar practic could be beneficial for metrics aimed at measuring similarity between words.
-\newline
+It seems like similar practice could be beneficial for metrics aimed at measuring similarity between words.
+
 <!-- TODO: numerical test if this distance is significant? check mean mean cosine distances of some random words? -->
 
-\textbf{The curse of dimensionality}
+\textbf{6. The curse of dimensionality}
 
@@
 "https://analyticsindiamag.com/curse-of-dimensionality-and-what-beginners-should-do-to-overcome-it/"
-as the number of features increases, it may be harder and harder to get useful information from the data with the usage of available algorithms.
+as the number of features increases, it may be harder and harder to obtain useful information from the data with the usage of available algorithms.
 One may notice that more data should contribute to greater amount of information but more information also means greater risk of noise and distractions in data.
@@
 In details, the authors generate random points with increasing number of dimensions where each dimension of a data point is given a value between 0 and 1.
-Then they pick one vector on random from each dimension class and calculate cosine similarity between chosen vector and the rest of the data.
+Then they pick one vector at random from each dimension class and calculate cosine similarity between the chosen vector and the rest of the data.
 Then they check how the variation of values changes as the number of dimensions increases.
@@
 Maybe the scale should be adjusted to the number of dimensions and variance so that it still gives us sensible information about data.
-According to some publishers the cosine similarity in high dimensions is not reliable enough to trust it as it may be the case that choosing words on random may result in getting similar values as when picking them consciously.
-\newline
+According to some researches the cosine similarity in high dimensions is not reliable enough to trust it as it may be the case that choosing words on random may result in getting similar values as when picking them consciously.
 
-![curse of dimensionality, on x-axis number of dimensions, on y-axis standard deviation of similarity](curseOfDimensionality.png)
+![curse of dimensionality,  number of dimensions on the x axis,  standard deviation of similarity on the y axis](../images/curseOfDimensionality.png)
 
-\textbf{Control group}
+<!-- textbf{Control group}
 
 How to properly prepare control group in terms of quality and quantity?
-\newline
 <!-- + check if it may be proved in the code -->
 
-\textbf{Verifying other similarity meaures}
+<!--  \textbf{Verifying other similarity meaures}
 
@@
 a) TS-SS 
-<!-- todo: check if it makes sense to calculate for all of the datasets; interpretations etc. -->
-
-<!-- https://www.researchgate.net/publication/303513110_A_Hybrid_Geometric_Approach_for_Measuring_Similarity_Level_Among_Documents_and_Document_Clustering -->
+todo: check if it makes sense to calculate for all of the datasets; interpretations etc.
+https://www.researchgate.net/publication/303513110_A_Hybrid_Geometric_Approach_for_Measuring_Similarity_Level_Among_Documents_and_Document_Clustering
 
@@
 -->
+
+<!--chapter:end:02-cosine.Rmd-->
+
+# Walkthrough with the religion dataset
+
+Placeholder
+
+## Loading and understanding the dataset
+## First look at the empirical distributions
+## Looking at the islam-related words
+## Bayesian model structure and assumptions
+## Choosing predictors
+## Dataset-level coefficients
+## Model structure and assumptions
+## Protected classes in Reddit and Google embeddings
+## Dataset-level coefficients after debiasing
+## Protected classes  after debiasing
+
+<!--chapter:end:03-bayesian.Rmd-->

### 2021-06-10 paperDraft/paperDraft.Rmd
added paper draft and texts

---
title: "A bayesian method of cosine-based word2vec bias"
author: "Alicja Dobrzeniecka and Rafal Urbaniak"
output:
  pdf_document:
    number_sections: true
    df_print: kable 
    keep_tex: true
    includes:
      in_header:
        - Rafal_latex6.sty
classoption: dvipsnames,enabledeprecatedfontcommands
fontsize: 10pt
documentclass: scrartcl
urlcolor: blue
bibliography: cosineReferences1.bib
csl: apa-5th-edition.csl
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = '../')

library(ggplot2)
library(ggthemes)
library(gridExtra)
```

test

# Cosine-based measures of bias

One of the first measures in the discussion has been developed by @bolukbasi2016man. There, the gender bias of a word $w$ is understood as its projection on the gender direction $\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})$ (the gender direction is the top principal compontent of ten gender pair difference vectors). The underlying idea is that no bias is present if non-explicitly gendered words are in equal distance to both elements in all explicitly gendered pairs. Given the (ideally) gender netural words $N$ and the gender direction $g$ the direct direct gender bias is defined as the average distance of the words in $N$ from $g$ ($c$ is a parameter determining how strict we want to be):
\begin{align}
\mathsf{directBias_c(N,g)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},g)\vert^c}{\vert N \vert }
\end{align}

The use of projections has been ciriticized for instance by @Gonen2019lipstick, who point out that while gender-direction might be an indicator of bias, it is only one possible manifestation of it, and reducing a projection of words might be insufficient. For instance, "math" and "delicate" might be in equal distance to both explicitly gendered words while being closer to quite different stereotypical attribute words. Further, the authors point out that most word pairs preserve similarity under debiasing meant to minimize projection-based bias.^[@bolukbasi2016man use also another method which involves analogies and their evaluations by human users on Mechanical Turk. It is discussed and criticized in [@Nissim2020fair].]

To measure bias in word embeddings, @Caliskan2017semanticsBiases proposed the Word Embedding Association Test (WEAT). The idea is that the measure of  biases between two sets of target words, $X$ and $Y$, (we call them protected words) should be quantified in terms of the cosine similarity  between the protected words and attribute words coming from  two sets of stereotype attribute words, $A$ and $B$ (we'll call them attributes). For instance, $X$ might be a set of male names, $Y$ a set of female names, $A$ might contain stereotypically male-related career words, and $B$ stereotypically female-related family words. WEAT is a modification  of the Implicit Association Test (IAT) [@Nosek2002harvesting] used in psychology and uses almost the same word sets, allowing for a \emph{prima facie} sensible comparison with bias in humans.
Let $f$ be a distance or similarity measure (usually, cosine similarity). The association difference for a term $t$ is:
\begin{align}
s(t,A,B) & = \frac{\sum_{a\in A}f(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}f(t,b)}{\vert B\vert}
\end{align}
\noindent then, the association difference between $A$ a $B$ is:
\begin{align}
s(X,Y,A,B) & = \sum_{x\in X} s(x,A,B) -  \sum_{y\in Y} s(y,A,B)
\end{align}
\noindent
$s(X,Y,A,B)$ is the statistic used in the signifcance test, and the $p$-value obtained by bootstrapping: it is the frequency of $s(X_i,Y_i,A,B)>s(X,Y,A,B)$ for all equally sized partitions $X_i, Y_i$ of $X\cup Y$. The effect size is computed by normalizing the difference in means as follows:
\begin{align}
bias(A,B) & = \frac{
\mu(\{s(x,A,B)\}_{x\in X}) -\mu(\{s(y,A,B)\}_{y\in Y}) 
}{
\sigma(\{s(w,A,B)\}_{w\in X\cup Y})
}
\end{align}

@Caliskan2017semanticsBiases show that significant biases---thus measured--- similar to the ones discovered by IAT can be discovered in word embeddings. @Lauscher2019multidimensional extended the methodology to a multilingual and cross-lingual setting, arguing that using Euclidean distance instead of similarty does not make much difference, while the bias effects vary greatly accross embedding models (interestingly, with social media-text trained embeddings being less biased than those based on Wikipedia).

A similar methodology is employed by @Garg2018years, who employ word embeddings trained od corpora from different decades to study the shifts in various biases. For instance, to compute the occupational embeddings bias for women the authors first compute the average vector of vector emeddings of words that represent women (e.g. "she", "female"), then calculate the Euclidean distance between this mean vector  and words for occupations. Then they take the mean of these distances and subtract from it the analogously obtained mean for the average vector of vector embeddings of words that represent men. Formally they take the relative norm distance between $X$ and $Y$ to be:
\begin{align}
\textsf{relative norm distance} & = \sum_{v_m\in M} \vert \vert v_m - v_X\vert \vert_2 - \vert v_m - v_Y\vert \vert_2
\end{align}
\noindent where the norm used is Euclidean, and $v_X$ and $x_Y$ are average vectors for sets $X$ and $Y$ respectively.

There is a problem with taking means of means in the calculations, though. By pre-averaging data we throw away information about sample size. For instance, think about proportions: 10 out of 20 and 2 out of 4 give the same mean, but you would obtain more information by making the former observation rather than by making the latter. Connectedly, when we pre-average, we remove variation, and so pre-averaging  tend to manufacture false confidence.

To illustrate let's employ the formulas used by @Caliskan2017semanticsBiases and go through an example. Conceptually, all such tests come up with rather short lists of protected words and rather short lists of stereotypical attributes. Clearly, these are not complete list. So let's treat them as  samples from richer pools of stereotypical predicates and let's take the uncertainty and variation involved seriously.

Consider a situation in which there are two protected classes, $X=\{t_1,t_2\}$ and $Y=\{t_3,t_4\}$ and two five-element sets $A$ and $B$.

First, we play around with a scenario in which all the protected terms are on average equally distant from both sets ($\mu =1$) with standard deviation of $.05$. First, let's randomly draw distances and plot the results with group means plotted as vertical lines.

\footnotesize
```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
set.seed(123)
t1 <- data.frame(A  = rnorm(5,1,0.05), B = rnorm(5,1,0.05))
t2 <- data.frame(A  = rnorm(5,1,0.05), B = rnorm(5,1,0.05))
t3 <- data.frame(A  = rnorm(5,1,0.05), B = rnorm(5,1,0.05))
t4 <- data.frame(A  = rnorm(5,1,0.05), B = rnorm(5,1,0.05))
```
\normalsize

```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
t1Plot <- ggplot(t1)+geom_point(aes(y = 0, x = A), size = 2, alpha = 0.7)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 2, alpha = 0.7)+theme_tufte()+xlab("distance")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Distances for term t1 in X")

t2Plot <- ggplot(t2)+geom_point(aes(y = 0, x = A), size = 2, alpha = 0.7)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 2, alpha = 0.7)+theme_tufte()+xlab("distance")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Distances for term t2 in X")

t3Plot <- ggplot(t3)+geom_point(aes(y = 0, x = A), size = 2, alpha = 0.7)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 2, alpha = 0.7)+theme_tufte()+xlab("distance")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Distances for term t3 in Y")

t4Plot <- ggplot(t4)+geom_point(aes(y = 0, x = A), size = 2, alpha = 0.7)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 2, alpha = 0.7)+theme_tufte()+xlab("distance")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Distances for term t4 in Y")

grid.arrange(t1Plot,t2Plot, t3Plot, t4Plot, ncol=2)
```

When you look at the datapoints, do you have the impression of a strong bias here? We wouldn't think so.  But now let's run the calculations from [@Caliskan2017semanticsBiases].

\vspace{1mm}
\footnotesize
```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
s <- function (table){ mean(table$A) - mean(table$B)}
factor <- sd(c(s(t1),s(t2),s(t3),s(t4)))
numerator <-  mean(s(t1),s(t2)) - mean(s(t3),s(t4))
bias <- numerator / factor
bias
```
\normalsize

This should make us pause. We know these were points randomly drawn from distributions where there is no difference in means. Yet, the calculated effect size is 1.82, whereas the largest effect size reported by @Caliskan2017semanticsBiases is 1.81!

Now, let's try sampling from distributions where there in fact is a difference in means. Terms from $X$ are on average .8 away from $A$ (and $1$ from $B$), while terms from $Y$ are $.8$ away to $B$ (and 1 to $A$). The standard deviation is 0.1 in all the cases. There is a  clear difference between $X$ and $Y$ and quick visual inspection should tell us so.

\vspace{1mm}
\footnotesize
```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
set.seed(123)
t1 <- data.frame(A  = rnorm(5,.8,0.1), B = rnorm(5,1,0.1))
t2 <- data.frame(A  = rnorm(5,.8,0.1), B = rnorm(5,1,0.1))
t3 <- data.frame(A  = rnorm(5,1,0.1), B = rnorm(5,.8,0.1))
t4 <- data.frame(A  = rnorm(5,1,0.1), B = rnorm(5,.8,0.1))
```
\normalsize

\vspace{1mm}
\footnotesize
```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
t1Plot2 <- ggplot(t1)+geom_point(aes(y = 0, x = A), size = 2, alpha = 0.7)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 2, alpha = 0.7)+theme_tufte()+xlab("distance")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Distances for term t1 in X")

t2Plot2 <- ggplot(t2)+geom_point(aes(y = 0, x = A), size = 2, alpha = 0.7)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 2, alpha = 0.7)+theme_tufte()+xlab("distance")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Distances for term t2 in X")

t3Plot2 <- ggplot(t3)+geom_point(aes(y = 0, x = A), size = 2, alpha = 0.7)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 2, alpha = 0.7)+theme_tufte()+xlab("distance")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Distances for term t3 in Y")

t4Plot2 <- ggplot(t4)+geom_point(aes(y = 0, x = A), size = 2, alpha = 0.7)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 2, alpha = 0.7)+theme_tufte()+xlab("distance")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Distances for term t4 in Y")

grid.arrange(t1Plot2,t2Plot2, t3Plot2, t4Plot2, ncol=2)
```
\normalsize

Is this clear difference mirrored in the calculations?

\footnotesize
```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
factor <- sd(c(s(t1),s(t2),s(t3),s(t4)))
numerator <-  mean(s(t1),s(t2)) - mean(s(t3),s(t4))
bias <- numerator / factor
bias
```
\normalsize

The absolute value of the effect size is smaller than in the previous case!

\newpage

\noindent \huge  \textbf{Appendix} \normalsize

\noindent \Large \textbf{Word lists, including human and neutral predicates} \normalsize

#  References {-}

\vspace{-3mm}

### 2021-06-10 paperDraft/paperDraft.tex
added paper draft and texts

\documentclass[10pt,dvipsnames,enabledeprecatedfontcommands]{scrartcl}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage[unicode=true]{hyperref}
\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
\hypersetup{
            pdftitle={A bayesian method of cosine-based word2vec bias},
            pdfauthor={Alicja Dobrzeniecka and Rafal Urbaniak},
            colorlinks=true,
            linkcolor=Maroon,
            citecolor=Blue,
            urlcolor=blue,
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

%\documentclass{article}

% %packages
 \usepackage{booktabs}

\usepackage{multirow}

\usepackage{graphicx}
\usepackage{longtable}
\usepackage{ragged2e}
\usepackage{etex}
%\usepackage{yfonts}
\usepackage{marvosym}
\usepackage[notextcomp]{kpfonts}
\usepackage{nicefrac}
\newcommand*{\QED}{\hfill \footnotesize {\sc Q.e.d.}}
\usepackage{floatrow}

\usepackage[textsize=footnotesize]{todonotes}
%\linespread{1.5}

\setlength{\parindent}{10pt}
\setlength{\parskip}{1pt}

%language
\usepackage{times}
\usepackage{t1enc}
%\usepackage[utf8x]{inputenc}
%\usepackage[polish]{babel}
%\usepackage{polski}

\usepackage{mathptmx}
\usepackage[scaled=0.88]{helvet}

%AMS
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{mathtools}

\usepackage{geometry}
 \geometry{a4paper,left=35mm,top=20mm,}

%environments
\newtheorem{fact}{Fact}

%abbreviations
\newcommand{\ra}{\rangle}
\newcommand{\la}{\langle}
\newcommand{\n}{\neg}
\newcommand{\et}{\wedge}
\newcommand{\jt}{\rightarrow}
\newcommand{\ko}[1]{\forall  #1\,}
\newcommand{\ro}{\leftrightarrow}
\newcommand{\exi}[1]{\exists\, {_{#1}}}
\newcommand{\pr}[1]{\mathsf{P}(#1)}
\newcommand{\cost}{\mathsf{cost}}

\newcommand{\odds}{\mathsf{Odds}}
\newcommand{\ind}{\mathsf{Ind}}
\newcommand{\nf}[2]{\nicefrac{#1\,}{#2}}
\newcommand{\R}[1]{\texttt{#1}}
\newcommand{\prr}[1]{\mbox{$\mathtt{P}_{prior}(#1)$}}
\newcommand{\prp}[1]{\mbox{$\mathtt{P}_{posterior}(#1)$}}

\newtheorem{q}{\color{blue}Question}
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}

%technical intermezzo
%---------------------

\newcommand{\intermezzoa}{
	\begin{minipage}[c]{13cm}
	\begin{center}\rule{10cm}{0.4pt}

	\tiny{\sc Optional Content Starts}

	\vspace{-1mm}

	\rule{10cm}{0.4pt}\end{center}
	\end{minipage}\nopagebreak 
	}

\newcommand{\intermezzob}{\nopagebreak 
	\begin{minipage}[c]{13cm}
	\begin{center}\rule{10cm}{0.4pt}

	\tiny{\sc Optional Content Ends}

	\vspace{-1mm}

	\rule{10cm}{0.4pt}\end{center}
	\end{minipage}
	}
%--------------------

\newtheorem*{reply*}{Reply}
\usepackage{enumitem}
\newcommand{\question}[1]{\begin{enumerate}[resume,leftmargin=0cm,labelsep=0cm,align=left]
\item #1
\end{enumerate}}

\usepackage{float}

% \setbeamertemplate{blocks}[rounded][shadow=true]
% \setbeamertemplate{itemize items}[ball]
% \AtBeginPart{}
% \AtBeginSection{}
% \AtBeginSubsection{}
% \AtBeginSubsubsection{}
% \setlength{\emergencystretch}{0em}
% \setlength{\parskip}{0pt}

\usepackage[authoryear]{natbib}

%\bibliographystyle{apalike}

\title{A bayesian method of cosine-based word2vec bias}
\author{Alicja Dobrzeniecka and Rafal Urbaniak}
\date{}

\begin{document}
\maketitle

test

\section{Cosine-based measures of
bias}\label{cosine-based-measures-of-bias}

One of the first measures in the discussion has been developed by
Bolukbasi, Chang, Zou, Saligrama, \& Kalai (2016). There, the gender
bias of a word \(w\) is understood as its projection on the gender
direction \(\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})\)
(the gender direction is the top principal compontent of ten gender pair
difference vectors). The underlying idea is that no bias is present if
non-explicitly gendered words are in equal distance to both elements in
all explicitly gendered pairs. Given the (ideally) gender netural words
\(N\) and the gender direction \(g\) the direct direct gender bias is
defined as the average distance of the words in \(N\) from \(g\) (\(c\)
is a parameter determining how strict we want to be):

\begin{align}
\mathsf{directBias_c(N,g)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},g)\vert^c}{\vert N \vert }
\end{align}

The use of projections has been ciriticized for instance by Gonen \&
Goldberg (2019), who point out that while gender-direction might be an
indicator of bias, it is only one possible manifestation of it, and
reducing a projection of words might be insufficient. For instance,
``math'' and ``delicate'' might be in equal distance to both explicitly
gendered words while being closer to quite different stereotypical
attribute words. Further, the authors point out that most word pairs
preserve similarity under debiasing meant to minimize projection-based
bias.\footnote{Bolukbasi et al. (2016) use also another method which
  involves analogies and their evaluations by human users on Mechanical
  Turk. It is discussed and criticized in (Nissim, Noord, \& Goot,
  2020).}

To measure bias in word embeddings, Caliskan, Bryson, \& Narayanan
(2017) proposed the Word Embedding Association Test (WEAT). The idea is
that the measure of biases between two sets of target words, \(X\) and
\(Y\), (we call them protected words) should be quantified in terms of
the cosine similarity between the protected words and attribute words
coming from two sets of stereotype attribute words, \(A\) and \(B\)
(we'll call them attributes). For instance, \(X\) might be a set of male
names, \(Y\) a set of female names, \(A\) might contain stereotypically
male-related career words, and \(B\) stereotypically female-related
family words. WEAT is a modification of the Implicit Association Test
(IAT) (Nosek, Banaji, \& Greenwald, 2002) used in psychology and uses
almost the same word sets, allowing for a \emph{prima facie} sensible
comparison with bias in humans. Let \(f\) be a distance or similarity
measure (usually, cosine similarity). The association difference for a
term \(t\) is:

\begin{align}
s(t,A,B) & = \frac{\sum_{a\in A}f(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}f(t,b)}{\vert B\vert}
\end{align}

\noindent then, the association difference between \(A\) a \(B\) is:

\begin{align}
s(X,Y,A,B) & = \sum_{x\in X} s(x,A,B) -  \sum_{y\in Y} s(y,A,B)
\end{align}

\noindent
\(s(X,Y,A,B)\) is the statistic used in the signifcance test, and the
\(p\)-value obtained by bootstrapping: it is the frequency of
\(s(X_i,Y_i,A,B)>s(X,Y,A,B)\) for all equally sized partitions
\(X_i, Y_i\) of \(X\cup Y\). The effect size is computed by normalizing
the difference in means as follows:

\begin{align}
bias(A,B) & = \frac{
\mu(\{s(x,A,B)\}_{x\in X}) -\mu(\{s(y,A,B)\}_{y\in Y}) 
}{
\sigma(\{s(w,A,B)\}_{w\in X\cup Y})
}
\end{align}

Caliskan et al. (2017) show that significant biases---thus measured---
similar to the ones discovered by IAT can be discovered in word
embeddings. Lauscher \& Glavas (2019) extended the methodology to a
multilingual and cross-lingual setting, arguing that using Euclidean
distance instead of similarty does not make much difference, while the
bias effects vary greatly accross embedding models (interestingly, with
social media-text trained embeddings being less biased than those based
on Wikipedia).

A similar methodology is employed by Garg, Schiebinger, Jurafsky, \& Zou
(2018), who employ word embeddings trained od corpora from different
decades to study the shifts in various biases. For instance, to compute
the occupational embeddings bias for women the authors first compute the
average vector of vector emeddings of words that represent women (e.g.
``she'', ``female''), then calculate the Euclidean distance between this
mean vector and words for occupations. Then they take the mean of these
distances and subtract from it the analogously obtained mean for the
average vector of vector embeddings of words that represent men.
Formally they take the relative norm distance between \(X\) and \(Y\) to
be:

\begin{align}
\textsf{relative norm distance} & = \sum_{v_m\in M} \vert \vert v_m - v_X\vert \vert_2 - \vert v_m - v_Y\vert \vert_2
\end{align}

\noindent where the norm used is Euclidean, and \(v_X\) and \(x_Y\) are
average vectors for sets \(X\) and \(Y\) respectively.

There is a problem with taking means of means in the calculations,
though. By pre-averaging data we throw away information about sample
size. For instance, think about proportions: 10 out of 20 and 2 out of 4
give the same mean, but you would obtain more information by making the
former observation rather than by making the latter. Connectedly, when
we pre-average, we remove variation, and so pre-averaging tend to
manufacture false confidence.

To illustrate let's employ the formulas used by Caliskan et al. (2017)
and go through an example. Conceptually, all such tests come up with
rather short lists of protected words and rather short lists of
stereotypical attributes. Clearly, these are not complete list. So let's
treat them as samples from richer pools of stereotypical predicates and
let's take the uncertainty and variation involved seriously.

Consider a situation in which there are two protected classes,
\(X=\{t_1,t_2\}\) and \(Y=\{t_3,t_4\}\) and two five-element sets \(A\)
and \(B\).

First, we play around with a scenario in which all the protected terms
are on average equally distant from both sets (\(\mu =1\)) with standard
deviation of \(.05\). First, let's randomly draw distances and plot the
results with group means plotted as vertical lines.

\footnotesize

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{123}\NormalTok{)}
\NormalTok{t1 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
\NormalTok{t2 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
\NormalTok{t3 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
\NormalTok{t4 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\normalsize

\begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/unnamed-chunk-2-1} \end{center}

When you look at the datapoints, do you have the impression of a strong
bias here? We wouldn't think so. But now let's run the calculations from
(Caliskan et al., 2017).

\vspace{1mm} \footnotesize

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{s <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{ (table)\{ }\KeywordTok{mean}\NormalTok{(table}\OperatorTok{$}\NormalTok{A) }\OperatorTok{-}\StringTok{ }\KeywordTok{mean}\NormalTok{(table}\OperatorTok{$}\NormalTok{B)\}}
\NormalTok{factor <-}\StringTok{ }\KeywordTok{sd}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1),}\KeywordTok{s}\NormalTok{(t2),}\KeywordTok{s}\NormalTok{(t3),}\KeywordTok{s}\NormalTok{(t4)))}
\NormalTok{numerator <-}\StringTok{  }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1),}\KeywordTok{s}\NormalTok{(t2)) }\OperatorTok{-}\StringTok{ }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t3),}\KeywordTok{s}\NormalTok{(t4))}
\NormalTok{bias <-}\StringTok{ }\NormalTok{numerator }\OperatorTok{/}\StringTok{ }\NormalTok{factor}
\NormalTok{bias}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 1.825005
\end{verbatim}

\normalsize

This should make us pause. We know these were points randomly drawn from
distributions where there is no difference in means. Yet, the calculated
effect size is 1.82, whereas the largest effect size reported by
Caliskan et al. (2017) is 1.81!

Now, let's try sampling from distributions where there in fact is a
difference in means. Terms from \(X\) are on average .8 away from \(A\)
(and \(1\) from \(B\)), while terms from \(Y\) are \(.8\) away to \(B\)
(and 1 to \(A\)). The standard deviation is 0.1 in all the cases. There
is a clear difference between \(X\) and \(Y\) and quick visual
inspection should tell us so.

\vspace{1mm} \footnotesize

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{123}\NormalTok{)}
\NormalTok{t1 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{8}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
\NormalTok{t2 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{8}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
\NormalTok{t3 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{8}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
\NormalTok{t4 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{8}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\normalsize

\vspace{1mm} \footnotesize

\begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/unnamed-chunk-5-1} \end{center}

\normalsize

Is this clear difference mirrored in the calculations?

\footnotesize

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{factor <-}\StringTok{ }\KeywordTok{sd}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1),}\KeywordTok{s}\NormalTok{(t2),}\KeywordTok{s}\NormalTok{(t3),}\KeywordTok{s}\NormalTok{(t4)))}
\NormalTok{numerator <-}\StringTok{  }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1),}\KeywordTok{s}\NormalTok{(t2)) }\OperatorTok{-}\StringTok{ }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t3),}\KeywordTok{s}\NormalTok{(t4))}
\NormalTok{bias <-}\StringTok{ }\NormalTok{numerator }\OperatorTok{/}\StringTok{ }\NormalTok{factor}
\NormalTok{bias}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] -1.444026
\end{verbatim}

\normalsize

The absolute value of the effect size is smaller than in the previous
case!

\newpage

\noindent \huge  \textbf{Appendix} \normalsize

\noindent \Large \textbf{Word lists, including human and neutral predicates}
\normalsize

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\vspace{-3mm}

\hypertarget{refs}{}
\hypertarget{ref-bolukbasi2016man}{}
Bolukbasi, T., Chang, K.-W., Zou, J., Saligrama, V., \& Kalai, A.
(2016). Man is to computer programmer as woman is to homemaker?
Debiasing word embeddings.

\hypertarget{ref-Caliskan2017semanticsBiases}{}
Caliskan, A., Bryson, J. J., \& Narayanan, A. (2017). Semantics derived
automatically from language corpora contain human-like biases.
\emph{Science}, \emph{356}(6334), 183--186. American Association for the
Advancement of Science (AAAS). Retrieved from
\url{https://doi.org/10.1126/science.aal4230}

\hypertarget{ref-Garg2018years}{}
Garg, N., Schiebinger, L., Jurafsky, D., \& Zou, J. (2018). Word
embeddings quantify 100 years of gender and ethnic stereotypes.
\emph{Proceedings of the National Academy of Sciences}, \emph{115}(16),
E3635--E3644. Proceedings of the National Academy of Sciences. Retrieved
from \url{https://doi.org/10.1073/pnas.1720347115}

\hypertarget{ref-Gonen2019lipstick}{}
Gonen, H., \& Goldberg, Y. (2019). Lipstick on a pig: Debiasing methods
cover up systematic gender biases in word embeddings but do not remove
them. In \emph{Proceedings of the 2019 conference of the north American
chapter of the association for computational linguistics: Human language
technologies, volume 1 (long and short papers)} (pp. 609--614).
Minneapolis, Minnesota: Association for Computational Linguistics.
Retrieved from \url{https://www.aclweb.org/anthology/N19-1061}

\hypertarget{ref-Lauscher2019multidimensional}{}
Lauscher, A., \& Glavas, G. (2019). Are we consistently biased?
Multidimensional analysis of biases in distributional word vectors.
\emph{CoRR}, \emph{abs/1904.11783}. Retrieved from
\url{http://arxiv.org/abs/1904.11783}

\hypertarget{ref-Nissim2020fair}{}
Nissim, M., Noord, R. van, \& Goot, R. van der. (2020). Fair is better
than sensational: Man is to doctor as woman is to doctor.
\emph{Computational Linguistics}, \emph{46}(2), 487--497. MIT Press -
Journals. Retrieved from \url{https://doi.org/10.1162/coli_a_00379}

\hypertarget{ref-Nosek2002harvesting}{}
Nosek, B. A., Banaji, M. R., \& Greenwald, A. G. (2002). Harvesting
implicit group attitudes and beliefs from a demonstration web site.
\emph{Group Dynamics: Theory, Research, and Practice}, \emph{6}(1),
101--115. American Psychological Association (APA). Retrieved from
\url{https://doi.org/10.1037/1089-2699.6.1.101}

\end{document}

### 2021-06-10 paperDraft/paperDraft.Rmd
finished drafting means of means

@@
 
-test
-
 # Cosine-based measures of bias
@@
 WEAT is a modification  of the Implicit Association Test (IAT) [@Nosek2002harvesting] used in psychology and uses almost the same word sets, allowing for a \emph{prima facie} sensible comparison with bias in humans.
-Let $f$ be a distance or similarity measure (usually, cosine similarity). The association difference for a term $t$ is:
+Let $f$ be a similarity measure (usually, cosine similarity). The association difference for a term $t$ is:
 \begin{align}
@@
 
-There is a problem with taking means of means in the calculations, though.
-By pre-averaging data we throw away information about sample size.
-For instance, think about proportions: 10 out of 20 and 2 out of 4 give the same mean, but you would obtain more information by making the former observation rather than by making the latter.
-Connectedly, when we pre-average, we remove variation, and so pre-averaging  tend to manufacture false confidence.
+@Manzini2019blackToCriminal modify WEAT to a multi-class setting, introducing Mean Average Cosine similarity as a measure of bias (in fact, in the paper  they report distances rather than similarities).
+Let $T = \{t_1, \dots, t_k\}$ be a class of protected word embeddings, and let each $A_j\in A$ be a set of attributes stereotypically associated with a protected word).
+Then:
+\begin{align}
+S(t_i, A_j) & = \frac{1}{\vert A_j\vert}\sum_{a\in A_j}\mathsf{cos}(t,a) \\
+MAC(T,A) & = \frac{1}{\vert T \vert \,\vert A\vert}\sum_{t_i \in T }\sum_{A_j \in A} S(t_i,A_j)
+\end{align}
+That is, for each protected word $T and each attribute class, they first take the mean for this protected word and all attributes in a given attribute class, and then take the mean of thus obtained means for all the protected words.
+The t-tests they employ are run on average cosines used to calculate MAC.
 
-To illustrate let's employ the formulas used by @Caliskan2017semanticsBiases and go through an example.
+## The problem with means of means
+
+The measures described all calculate means of means and their authors run statistical tests on sets of means.
+This, however, is problematic for two related reasons.
+One, by pre-averaging data we throw away information about sample sizes.
+For the former point, think about proportions: 10 out of 20 and 2 out of 4 give the same mean, but you would obtain more information by making the former observation rather than by making the latter.
+Two, when we pre-average, we remove variation, and so pre-averaging  tend to manufacture false confidence.
+We will have more to say about whether this happens in the case of applications of MAC, for now let's go over a simple example to make the conceptual point clear.
+
+To illustrate let's employ the formulas used by @Caliskan2017semanticsBiases in a simple example.
 Conceptually, all such tests come up with rather short lists of protected words and rather short lists of stereotypical attributes.
@@
 
-Consider a situation in which there are two protected classes, $X=\{t_1,t_2\}$ and $Y=\{t_3,t_4\}$ and two five-element sets $A$ and $B$.
+Consider a simple situation in which there are two protected classes, $X=\{t_1,t_2\}$ and $Y=\{t_3,t_4\}$ and two five-element attribute sets $A$ and $B$.
 
-First, we play around with a scenario in which all the protected terms are on average equally distant from both sets ($\mu =1$) with standard deviation of $.05$.
-First, let's randomly draw distances and plot the results with group means plotted as vertical lines.
+First, we play around with a scenario in which all the protected terms are on average equally unsimilar to both sets ($\mu =0$) with standard deviation of $.05$.
+Let's randomly draw similarity scores and plot the results with group means plotted as vertical lines.
 
@@
 set.seed(123)
-t1 <- data.frame(A  = rnorm(5,1,0.05), B = rnorm(5,1,0.05))
-t2 <- data.frame(A  = rnorm(5,1,0.05), B = rnorm(5,1,0.05))
-t3 <- data.frame(A  = rnorm(5,1,0.05), B = rnorm(5,1,0.05))
-t4 <- data.frame(A  = rnorm(5,1,0.05), B = rnorm(5,1,0.05))
+t1 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
+t2 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
+t3 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
+t4 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
 ```
@@
 ```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
-t1Plot <- ggplot(t1)+geom_point(aes(y = 0, x = A), size = 2, alpha = 0.7)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 2, alpha = 0.7)+theme_tufte()+xlab("distance")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Distances for term t1 in X")
+xlimits <- c(-.3,.3)
 
-t2Plot <- ggplot(t2)+geom_point(aes(y = 0, x = A), size = 2, alpha = 0.7)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 2, alpha = 0.7)+theme_tufte()+xlab("distance")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Distances for term t2 in X")
+t1Plot <- ggplot(t1)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t1 in X")+xlim(xlimits)
 
-t3Plot <- ggplot(t3)+geom_point(aes(y = 0, x = A), size = 2, alpha = 0.7)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 2, alpha = 0.7)+theme_tufte()+xlab("distance")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Distances for term t3 in Y")
+t2Plot <- ggplot(t2)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t2 in X")+xlim(xlimits)
 
-t4Plot <- ggplot(t4)+geom_point(aes(y = 0, x = A), size = 2, alpha = 0.7)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 2, alpha = 0.7)+theme_tufte()+xlab("distance")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Distances for term t4 in Y")
+t3Plot <- ggplot(t3)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t3 in Y")+xlim(xlimits)
+
+t4Plot <- ggplot(t4)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t4 in Y")+xlim(xlimits)
 
@@
 
-When you look at the datapoints, do you have the impression of a strong bias here? We wouldn't think so.  But now let's run the calculations from [@Caliskan2017semanticsBiases].
+\noindent When you look at the datapoints, do you have the impression of a strong bias here? We wouldn't think so.  But now let's run the calculations from [@Caliskan2017semanticsBiases].
 
@@
 numerator <-  mean(s(t1),s(t2)) - mean(s(t3),s(t4))
-bias <- numerator / factor
-bias
+print(list(factor = factor,numerator = numerator, bias = numerator / factor))
 ```
@@
 
-This should make us pause.
+\noindent This should make us pause.
 We know these were points randomly drawn from distributions where there is no difference in means.
@@
 
+Interestingly, if we repeat the drawing 10000 times, each time calculating the  bias, it turns out that with this variance and sample size, pretty much anything can happen.
+
+\vspace{1mm}
+\footnotesize
+```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+
+biasesNull <- numeric(10000)
+for(i in 1:10000){
+t1 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
+t2 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
+t3 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
+t4 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
+
+factor <- sd(c(s(t1),s(t2),s(t3),s(t4)))
+numerator <-  mean(s(t1),s(t2)) - mean(s(t3),s(t4))
+biasesNull[i]  <- numerator / factor
+}
+ggplot()+geom_histogram(aes(x=biasesNull, y = ..density..), alpha = 0.6, bins=50)+
+  theme_tufte()+labs(title="10k biases for identical means and sd =.05")+ xlab("bias")
+```
+\normalsize
+
+Now, let's simulate a situation where the means are identical but the standard deviation is much smaller, .001.
+
+\footnotesize
+```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+set.seed(124)
+t1v <- data.frame(A  = rnorm(5,0,0.001), B = rnorm(5,0,0.001))
+t2v <- data.frame(A  = rnorm(5,0,0.001), B = rnorm(5,0,0.001))
+t3v <- data.frame(A  = rnorm(5,0,0.001), B = rnorm(5,0,0.001))
+t4v <- data.frame(A  = rnorm(5,0,0.001), B = rnorm(5,0,0.001))
+```
+\normalsize
+
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+xlimits <- c(-.3,.3)
+
+t1vPlot <- ggplot(t1v)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t1 in X")+xlim(xlimits)
+
+t2vPlot <- ggplot(t2v)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t2 in X")+xlim(xlimits)
+
+t3vPlot <- ggplot(t3v)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t3 in Y")+xlim(xlimits)
+
+t4vPlot <- ggplot(t4v)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t4 in Y")+xlim(xlimits)
+
+grid.arrange(t1vPlot,t2vPlot, t3vPlot, t4vPlot, ncol=2)
+```
+
+\noindent When you look at the datapoints, do you have the impression of a strong bias here? We wouldn't think so.  But now let's run the calculations from [@Caliskan2017semanticsBiases].
+
+\vspace{1mm}
+\footnotesize
+```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+factorV <- sd(c(s(t1v),s(t2v),s(t3v),s(t4v)))
+numeratorV <-  mean(s(t1v),s(t2v)) - mean(s(t3v),s(t4v))
+print(list(factor = factorV,numerator = numeratorV, bias = numeratorV / factorV))
+```
+\normalsize
+
+While the numerator and the factors changed a lot, the bias actually increased.
+One reason bias increases is that once the standard deviation goes down, so does the factor used in the calculation of bias.
+
+Again, to see whether this metric would provide us with meaningful information, let's simulate 10000 drawings.
+
+\vspace{1mm}
+\footnotesize
+```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+set.seed(124)
+
+biasesLowVariance <- numeric(10000)
+for(i in 1:10000){
+t1v <- data.frame(A  = rnorm(5,0,0.001), B = rnorm(5,0,0.001))
+t2v <- data.frame(A  = rnorm(5,0,0.001), B = rnorm(5,0,0.001))
+t3v <- data.frame(A  = rnorm(5,0,0.001), B = rnorm(5,0,0.001))
+t4v <- data.frame(A  = rnorm(5,0,0.001), B = rnorm(5,0,0.001))
+
+factorV <- sd(c(s(t1v),s(t2v),s(t3v),s(t4v)))
+
+numeratorV <-  mean(s(t1v),s(t2v)) - mean(s(t3v),s(t4v))
+
+biasesLowVariance[i] <- numeratorV / factorV
+}
+ggplot()+geom_histogram(aes(x=biasesLowVariance, y = ..density..), alpha = 0.6, bins=50)+
+  theme_tufte()+labs(title="10k biases for identical means and sd =.001")+ xlab("bias")
+```
+\normalsize
+
+Again, not so informative.
 Now, let's try sampling from distributions where there in fact is a difference in means.
-Terms from $X$ are on average .8 away from $A$ (and $1$ from $B$), while terms from $Y$ are $.8$ away to $B$ (and 1 to $A$).
-The standard deviation is 0.1 in all the cases.
+Terms from $X$ are on average .1 similar to $A$ (and still $0$ to $B$), while terms from $Y$ are $.1$ similar to $B$ (and 0 to $A$).
+The standard deviation is 0.05 in all the cases.
 There is a  clear difference between $X$ and $Y$ and quick visual inspection should tell us so.
@@
 \footnotesize
-```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
-set.seed(123)
-t1 <- data.frame(A  = rnorm(5,.8,0.1), B = rnorm(5,1,0.1))
-t2 <- data.frame(A  = rnorm(5,.8,0.1), B = rnorm(5,1,0.1))
-t3 <- data.frame(A  = rnorm(5,1,0.1), B = rnorm(5,.8,0.1))
-t4 <- data.frame(A  = rnorm(5,1,0.1), B = rnorm(5,.8,0.1))
+```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=FALSE, fig.show = "hold", out.width = "100%"}
+set.seed(766)
+t1d2 <- data.frame(A  = rnorm(5,.1,0.05), B = rnorm(5,0,0.05))
+t2d2 <- data.frame(A  = rnorm(5,.1,0.05), B = rnorm(5,0,0.05))
+t3d2 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,.1,0.05))
+t4d2 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,.1,0.05))
 ```
@@
 ```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
-t1Plot2 <- ggplot(t1)+geom_point(aes(y = 0, x = A), size = 2, alpha = 0.7)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 2, alpha = 0.7)+theme_tufte()+xlab("distance")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Distances for term t1 in X")
+library(grid)
+library(gridExtra)
+t1d2Plot <- ggplot(t1d2)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t1 in X")+xlim(xlimits)
 
-t2Plot2 <- ggplot(t2)+geom_point(aes(y = 0, x = A), size = 2, alpha = 0.7)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 2, alpha = 0.7)+theme_tufte()+xlab("distance")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Distances for term t2 in X")
+t2d2Plot <- ggplot(t2d2)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t2 in X")+xlim(xlimits)
 
-t3Plot2 <- ggplot(t3)+geom_point(aes(y = 0, x = A), size = 2, alpha = 0.7)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 2, alpha = 0.7)+theme_tufte()+xlab("distance")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Distances for term t3 in Y")
+t3d2Plot <- ggplot(t3d2)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t3 in Y")+xlim(xlimits)
 
-t4Plot2 <- ggplot(t4)+geom_point(aes(y = 0, x = A), size = 2, alpha = 0.7)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 2, alpha = 0.7)+theme_tufte()+xlab("distance")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Distances for term t4 in Y")
+t4d2Plot <- ggplot(t4d2)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t4 in Y")+xlim(xlimits)
 
-grid.arrange(t1Plot2,t2Plot2, t3Plot2, t4Plot2, ncol=2)
+grid.arrange(t1d2Plot,t2d2Plot, t3d2Plot, t4d2Plot, ncol=2,   
+      top = textGrob("Different means, sd=0.05",gp=gpar(fontsize=15,font=1)))
 ```
@@
 
-Is this clear difference mirrored in the calculations?
+\noindent  Is this clear difference mirrored in the calculations?
 
 \footnotesize
-```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
-factor <- sd(c(s(t1),s(t2),s(t3),s(t4)))
-numerator <-  mean(s(t1),s(t2)) - mean(s(t3),s(t4))
-bias <- numerator / factor
-bias
+```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=FALSE, fig.show = "hold", out.width = "100%"}
+factorD2 <- sd(c(s(t1d2),s(t2d2),s(t3d2),s(t4d2)))
+numeratorD2 <-  mean(s(t1d2),s(t2d2)) - mean(s(t3d2),s(t4d2))
+biasD2 <- numeratorD2 / factorD2
+biasD2
 ```
@@
 
-The absolute value of the effect size is smaller than in the previous case!
+\noindent The absolute value of the effect size is smaller than in the null case with the same standard deviation.  Let's simulate 10000 drawings:
+
+\vspace{1mm}
+\footnotesize
+```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+biasesD2 <- numeric(10000)
+for(i in 1:10000){
+t1d2 <- data.frame(A  = rnorm(5,.1,0.05), B = rnorm(5,0,0.05))
+t2d2 <- data.frame(A  = rnorm(5,.1,0.05), B = rnorm(5,0,0.05))
+t3d2 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,.1,0.05))
+t4d2 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,.1,0.05))
+
+factorD2 <- sd(c(s(t1d2),s(t2d2),s(t3d2),s(t4d2)))
+numeratorD2 <-  mean(s(t1d2),s(t2d2)) - mean(s(t3d2),s(t4d2))
+biasesD2[i] <- numeratorD2/factorD2
+}
+
+ggplot()+geom_histogram(aes(x=biasesD2, y = ..density..), alpha = 0.6, bins=50)+
+  theme_tufte()+labs(title="10k biases for different means and sd =.05")+ xlab("bias")
+```
+\normalsize
+
+\noindent Now suppose we keep the means the same but increase the standard deviation to .1.
+
+\vspace{1mm}
+\footnotesize
+```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=FALSE, fig.show = "hold", out.width = "100%"}
+set.seed(766)
+t1d1 <- data.frame(A  = rnorm(5,.1,0.1), B = rnorm(5,0,0.1))
+t2d1 <- data.frame(A  = rnorm(5,.1,0.1), B = rnorm(5,0,0.1))
+t3d1 <- data.frame(A  = rnorm(5,0,0.1), B = rnorm(5,.1,0.1))
+t4d1 <- data.frame(A  = rnorm(5,0,0.1), B = rnorm(5,.1,0.1))
+```
+\normalsize
+
+\vspace{1mm}
+\footnotesize
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+t1d1Plot <- ggplot(t1d1)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t1 in X")+xlim(xlimits)
+
+t2d1Plot <- ggplot(t2d1)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t2 in X")+xlim(xlimits)
+
+t3d1Plot <- ggplot(t3d1)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t3 in Y")+xlim(xlimits)
+
+t4d1Plot <- ggplot(t4d1)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t4 in Y")+xlim(xlimits)
+
+grid.arrange(t1d1Plot,t2d1Plot, t3d1Plot, t4d1Plot, ncol=2)
+```
+\normalsize
+
+\noindent  Is this clear difference mirrored in the calculations?
+
+\footnotesize
+```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=FALSE, fig.show = "hold", out.width = "100%"}
+factorD1 <- sd(c(s(t1d1),s(t2d1),s(t3d1),s(t4d1)))
+numeratorD1 <-  mean(s(t1d1),s(t2d1)) - mean(s(t3d1),s(t4d1))
+biasD1 <- numeratorD1 / factorD1
+biasD1
+```
+\normalsize
+
+\noindent The absolute value of the effect size is smaller than in the previous case!
+
+\vspace{1mm}
+\footnotesize
+```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+biasesD1 <- numeric(10000)
+
+for(i in 1:10000){
+t1d1 <- data.frame(A  = rnorm(5,.1,0.1), B = rnorm(5,0,0.1))
+t2d1 <- data.frame(A  = rnorm(5,.1,0.1), B = rnorm(5,0,0.1))
+t3d1 <- data.frame(A  = rnorm(5,0,0.1), B = rnorm(5,.1,0.1))
+t4d1 <- data.frame(A  = rnorm(5,0,0.1), B = rnorm(5,.1,0.1))
+
+factorD1 <- sd(c(s(t1d1),s(t2d1),s(t3d1),s(t4d1)))
+numeratorD1 <-  mean(s(t1d1),s(t2d1)) - mean(s(t3d1),s(t4d1))
+biasesD1[i] <- numeratorD1/factorD1
+}
+
+ggplot()+geom_histogram(aes(x=biasesD1, y = ..density..), alpha = 0.6, bins=50)+
+  theme_tufte()+labs(title="10k biases for different means and sd =.001")+ xlab("bias")
+```
+\normalsize
+
+This is is a bit better, but still quite some uncertainty is involved, far from what systematically low mean-based p-values reported in the papers might suggest.
+Of course, this is a bit of a caricature, as our word lists were short (four protected words and 10 attributes).
+But the word lists used in the actually papers are not much longer.
+In such a set-up the key observations are as follows:
+
+- Seemingly high bias measures might arise even if the underlying processes actually have the same mean.
+
+- Even if the mean remains the same, non-negligible changes can result from a shift in the standard deviation of the original process, and the change might go in the opposite direction than a visualisation of datapoints might suggest, because with the decrease of standard deviation, the factor decreases and the bias increases.
+
+- Even if there is a difference in means, the bias metric can be lower, and the uncertainty about it needs to be gauged.
+
+- Even if the difference in means is the same, but the variation is different, the bias metric in the long run could tend toward a different value:
+
+\vspace{1mm}
+\footnotesize
+```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+mean(biasesD1); mean(biasesD2)
+median(biasesD1); median(biasesD2)
+```
+
+\normalsize
+and so the point estimations of bias are sensitive to factors other than the underlying process means.
 

### 2021-06-10 paperDraft/paperDraft.tex
finished drafting means of means

@@
 
-test
-
 \section{Cosine-based measures of
@@
 almost the same word sets, allowing for a \emph{prima facie} sensible
-comparison with bias in humans. Let \(f\) be a distance or similarity
-measure (usually, cosine similarity). The association difference for a
-term \(t\) is:
+comparison with bias in humans. Let \(f\) be a similarity measure
+(usually, cosine similarity). The association difference for a term
+\(t\) is:
 
@@
 
-There is a problem with taking means of means in the calculations,
-though. By pre-averaging data we throw away information about sample
-size. For instance, think about proportions: 10 out of 20 and 2 out of 4
-give the same mean, but you would obtain more information by making the
-former observation rather than by making the latter. Connectedly, when
-we pre-average, we remove variation, and so pre-averaging tend to
-manufacture false confidence.
+({\textbf{???}}) modify WEAT to a multi-class setting, introducing Mean
+Average Cosine similarity as a measure of bias (in fact, in the paper
+they report distances rather than similarities). Let
+\(T = \{t_1, \dots, t_k\}\) be a class of protected word embeddings, and
+let each \(A_j\in A\) be a set of attributes stereotypically associated
+with a protected word). Then:
+
+\begin{align}
+S(t_i, A_j) & = \frac{1}{\vert A_j\vert}\sum_{a\in A_j}\mathsf{cos}(t,a) \\
+MAC(T,A) & = \frac{1}{\vert T \vert \,\vert A\vert}\sum_{t_i \in T }\sum_{A_j \in A} S(t_i,A_j)
+\end{align}
+
+That is, for each protected word \$T and each attribute class, they
+first take the mean for this protected word and all attributes in a
+given attribute class, and then take the mean of thus obtained means for
+all the protected words. The t-tests they employ are run on average
+cosines used to calculate MAC.
+
+\subsection{The problem with means of
+means}\label{the-problem-with-means-of-means}
+
+The measures described all calculate means of means and their authors
+run statistical tests on sets of means. This, however, is problematic
+for two related reasons. One, by pre-averaging data we throw away
+information about sample sizes. For the former point, think about
+proportions: 10 out of 20 and 2 out of 4 give the same mean, but you
+would obtain more information by making the former observation rather
+than by making the latter. Two, when we pre-average, we remove
+variation, and so pre-averaging tend to manufacture false confidence. We
+will have more to say about whether this happens in the case of
+applications of MAC, for now let's go over a simple example to make the
+conceptual point clear.
 
 To illustrate let's employ the formulas used by Caliskan et al. (2017)
-and go through an example. Conceptually, all such tests come up with
-rather short lists of protected words and rather short lists of
-stereotypical attributes. Clearly, these are not complete list. So let's
-treat them as samples from richer pools of stereotypical predicates and
-let's take the uncertainty and variation involved seriously.
+in a simple example. Conceptually, all such tests come up with rather
+short lists of protected words and rather short lists of stereotypical
+attributes. Clearly, these are not complete list. So let's treat them as
+samples from richer pools of stereotypical predicates and let's take the
+uncertainty and variation involved seriously.
 
-Consider a situation in which there are two protected classes,
-\(X=\{t_1,t_2\}\) and \(Y=\{t_3,t_4\}\) and two five-element sets \(A\)
-and \(B\).
+Consider a simple situation in which there are two protected classes,
+\(X=\{t_1,t_2\}\) and \(Y=\{t_3,t_4\}\) and two five-element attribute
+sets \(A\) and \(B\).
 
 First, we play around with a scenario in which all the protected terms
-are on average equally distant from both sets (\(\mu =1\)) with standard
-deviation of \(.05\). First, let's randomly draw distances and plot the
+are on average equally unsimilar to both sets (\(\mu =0\)) with standard
+deviation of \(.05\). Let's randomly draw similarity scores and plot the
 results with group means plotted as vertical lines.
@@
 \KeywordTok{set.seed}\NormalTok{(}\DecValTok{123}\NormalTok{)}
-\NormalTok{t1 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
-\NormalTok{t2 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
-\NormalTok{t3 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
-\NormalTok{t4 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\NormalTok{t1 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\NormalTok{t2 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\NormalTok{t3 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\NormalTok{t4 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
 \end{Highlighting}
@@
 
-When you look at the datapoints, do you have the impression of a strong
-bias here? We wouldn't think so. But now let's run the calculations from
-(Caliskan et al., 2017).
+\noindent When you look at the datapoints, do you have the impression of
+a strong bias here? We wouldn't think so. But now let's run the
+calculations from (Caliskan et al., 2017).
 
@@
 \NormalTok{numerator <-}\StringTok{  }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1),}\KeywordTok{s}\NormalTok{(t2)) }\OperatorTok{-}\StringTok{ }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t3),}\KeywordTok{s}\NormalTok{(t4))}
-\NormalTok{bias <-}\StringTok{ }\NormalTok{numerator }\OperatorTok{/}\StringTok{ }\NormalTok{factor}
-\NormalTok{bias}
+\KeywordTok{print}\NormalTok{(}\KeywordTok{list}\NormalTok{(}\DataTypeTok{factor =}\NormalTok{ factor,}\DataTypeTok{numerator =}\NormalTok{ numerator, }\DataTypeTok{bias =}\NormalTok{ numerator }\OperatorTok{/}\StringTok{ }\NormalTok{factor))}
 \end{Highlighting}
@@
 \begin{verbatim}
+## $factor
+## [1] 0.02342637
+## 
+## $numerator
+## [1] 0.04275325
+## 
+## $bias
 ## [1] 1.825005
@@
 
-This should make us pause. We know these were points randomly drawn from
-distributions where there is no difference in means. Yet, the calculated
-effect size is 1.82, whereas the largest effect size reported by
-Caliskan et al. (2017) is 1.81!
+\noindent This should make us pause. We know these were points randomly
+drawn from distributions where there is no difference in means. Yet, the
+calculated effect size is 1.82, whereas the largest effect size reported
+by Caliskan et al. (2017) is 1.81!
 
-Now, let's try sampling from distributions where there in fact is a
-difference in means. Terms from \(X\) are on average .8 away from \(A\)
-(and \(1\) from \(B\)), while terms from \(Y\) are \(.8\) away to \(B\)
-(and 1 to \(A\)). The standard deviation is 0.1 in all the cases. There
-is a clear difference between \(X\) and \(Y\) and quick visual
-inspection should tell us so.
+Interestingly, if we repeat the drawing 10000 times, each time
+calculating the bias, it turns out that with this variance and sample
+size, pretty much anything can happen.
 
@@
 \begin{Highlighting}[]
-\KeywordTok{set.seed}\NormalTok{(}\DecValTok{123}\NormalTok{)}
-\NormalTok{t1 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{8}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
-\NormalTok{t2 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{8}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
-\NormalTok{t3 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{8}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
-\NormalTok{t4 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{8}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
+\NormalTok{biasesNull <-}\StringTok{ }\KeywordTok{numeric}\NormalTok{(}\DecValTok{10000}\NormalTok{)}
+\ControlFlowTok{for}\NormalTok{(i }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\DecValTok{10000}\NormalTok{)\{}
+\NormalTok{t1 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\NormalTok{t2 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\NormalTok{t3 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\NormalTok{t4 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+
+\NormalTok{factor <-}\StringTok{ }\KeywordTok{sd}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1),}\KeywordTok{s}\NormalTok{(t2),}\KeywordTok{s}\NormalTok{(t3),}\KeywordTok{s}\NormalTok{(t4)))}
+\NormalTok{numerator <-}\StringTok{  }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1),}\KeywordTok{s}\NormalTok{(t2)) }\OperatorTok{-}\StringTok{ }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t3),}\KeywordTok{s}\NormalTok{(t4))}
+\NormalTok{biasesNull[i]  <-}\StringTok{ }\NormalTok{numerator }\OperatorTok{/}\StringTok{ }\NormalTok{factor}
+\NormalTok{\}}
+\KeywordTok{ggplot}\NormalTok{()}\OperatorTok{+}\KeywordTok{geom_histogram}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x=}\NormalTok{biasesNull, }\DataTypeTok{y =}\NormalTok{ ..density..), }\DataTypeTok{alpha =} \FloatTok{0.6}\NormalTok{, }\DataTypeTok{bins=}\DecValTok{50}\NormalTok{)}\OperatorTok{+}
+\StringTok{  }\KeywordTok{theme_tufte}\NormalTok{()}\OperatorTok{+}\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title=}\StringTok{"10k biases for identical means and sd =.05"}\NormalTok{)}\OperatorTok{+}\StringTok{ }\KeywordTok{xlab}\NormalTok{(}\StringTok{"bias"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/unnamed-chunk-4-1} \end{center}
+
+\normalsize
+
+Now, let's simulate a situation where the means are identical but the
+standard deviation is much smaller, .001.
+
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{124}\NormalTok{)}
+\NormalTok{t1v <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
+\NormalTok{t2v <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
+\NormalTok{t3v <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
+\NormalTok{t4v <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+\begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/unnamed-chunk-6-1} \end{center}
+
+\noindent When you look at the datapoints, do you have the impression of
+a strong bias here? We wouldn't think so. But now let's run the
+calculations from (Caliskan et al., 2017).
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{factorV <-}\StringTok{ }\KeywordTok{sd}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1v),}\KeywordTok{s}\NormalTok{(t2v),}\KeywordTok{s}\NormalTok{(t3v),}\KeywordTok{s}\NormalTok{(t4v)))}
+\NormalTok{numeratorV <-}\StringTok{  }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1v),}\KeywordTok{s}\NormalTok{(t2v)) }\OperatorTok{-}\StringTok{ }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t3v),}\KeywordTok{s}\NormalTok{(t4v))}
+\KeywordTok{print}\NormalTok{(}\KeywordTok{list}\NormalTok{(}\DataTypeTok{factor =}\NormalTok{ factorV,}\DataTypeTok{numerator =}\NormalTok{ numeratorV, }\DataTypeTok{bias =}\NormalTok{ numeratorV }\OperatorTok{/}\StringTok{ }\NormalTok{factorV))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## $factor
+## [1] 0.0006402666
+## 
+## $numerator
+## [1] -0.001237637
+## 
+## $bias
+## [1] -1.933003
+\end{verbatim}
+
+\normalsize
+
+While the numerator and the factors changed a lot, the bias actually
+increased. One reason bias increases is that once the standard deviation
+goes down, so does the factor used in the calculation of bias.
+
+Again, to see whether this metric would provide us with meaningful
+information, let's simulate 10000 drawings.
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{124}\NormalTok{)}
+
+\NormalTok{biasesLowVariance <-}\StringTok{ }\KeywordTok{numeric}\NormalTok{(}\DecValTok{10000}\NormalTok{)}
+\ControlFlowTok{for}\NormalTok{(i }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\DecValTok{10000}\NormalTok{)\{}
+\NormalTok{t1v <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
+\NormalTok{t2v <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
+\NormalTok{t3v <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
+\NormalTok{t4v <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
+
+\NormalTok{factorV <-}\StringTok{ }\KeywordTok{sd}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1v),}\KeywordTok{s}\NormalTok{(t2v),}\KeywordTok{s}\NormalTok{(t3v),}\KeywordTok{s}\NormalTok{(t4v)))}
+
+\NormalTok{numeratorV <-}\StringTok{  }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1v),}\KeywordTok{s}\NormalTok{(t2v)) }\OperatorTok{-}\StringTok{ }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t3v),}\KeywordTok{s}\NormalTok{(t4v))}
+
+\NormalTok{biasesLowVariance[i] <-}\StringTok{ }\NormalTok{numeratorV }\OperatorTok{/}\StringTok{ }\NormalTok{factorV}
+\NormalTok{\}}
+\KeywordTok{ggplot}\NormalTok{()}\OperatorTok{+}\KeywordTok{geom_histogram}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x=}\NormalTok{biasesLowVariance, }\DataTypeTok{y =}\NormalTok{ ..density..), }\DataTypeTok{alpha =} \FloatTok{0.6}\NormalTok{, }\DataTypeTok{bins=}\DecValTok{50}\NormalTok{)}\OperatorTok{+}
+\StringTok{  }\KeywordTok{theme_tufte}\NormalTok{()}\OperatorTok{+}\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title=}\StringTok{"10k biases for identical means and sd =.001"}\NormalTok{)}\OperatorTok{+}\StringTok{ }\KeywordTok{xlab}\NormalTok{(}\StringTok{"bias"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/unnamed-chunk-8-1} \end{center}
+
+\normalsize
+
+Again, not so informative. Now, let's try sampling from distributions
+where there in fact is a difference in means. Terms from \(X\) are on
+average .1 similar to \(A\) (and still \(0\) to \(B\)), while terms from
+\(Y\) are \(.1\) similar to \(B\) (and 0 to \(A\)). The standard
+deviation is 0.05 in all the cases. There is a clear difference between
+\(X\) and \(Y\) and quick visual inspection should tell us so.
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{766}\NormalTok{)}
+\NormalTok{t1d2 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\NormalTok{t2d2 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\NormalTok{t3d2 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\NormalTok{t4d2 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
 \end{Highlighting}
@@
 
-\begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/unnamed-chunk-5-1} \end{center}
+\begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/unnamed-chunk-10-1} \end{center}
 
@@
 
-Is this clear difference mirrored in the calculations?
+\noindent  Is this clear difference mirrored in the calculations?
 
@@
 \begin{Highlighting}[]
-\NormalTok{factor <-}\StringTok{ }\KeywordTok{sd}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1),}\KeywordTok{s}\NormalTok{(t2),}\KeywordTok{s}\NormalTok{(t3),}\KeywordTok{s}\NormalTok{(t4)))}
-\NormalTok{numerator <-}\StringTok{  }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1),}\KeywordTok{s}\NormalTok{(t2)) }\OperatorTok{-}\StringTok{ }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t3),}\KeywordTok{s}\NormalTok{(t4))}
-\NormalTok{bias <-}\StringTok{ }\NormalTok{numerator }\OperatorTok{/}\StringTok{ }\NormalTok{factor}
-\NormalTok{bias}
+\NormalTok{factorD2 <-}\StringTok{ }\KeywordTok{sd}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1d2),}\KeywordTok{s}\NormalTok{(t2d2),}\KeywordTok{s}\NormalTok{(t3d2),}\KeywordTok{s}\NormalTok{(t4d2)))}
+\NormalTok{numeratorD2 <-}\StringTok{  }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1d2),}\KeywordTok{s}\NormalTok{(t2d2)) }\OperatorTok{-}\StringTok{ }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t3d2),}\KeywordTok{s}\NormalTok{(t4d2))}
+\NormalTok{biasD2 <-}\StringTok{ }\NormalTok{numeratorD2 }\OperatorTok{/}\StringTok{ }\NormalTok{factorD2}
+\NormalTok{biasD2}
 \end{Highlighting}
@@
 \begin{verbatim}
-## [1] -1.444026
+## [1] 1.490014
 \end{verbatim}
@@
 
-The absolute value of the effect size is smaller than in the previous
-case!
+\noindent The absolute value of the effect size is smaller than in the
+null case with the same standard deviation. Let's simulate 10000
+drawings:
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{biasesD2 <-}\StringTok{ }\KeywordTok{numeric}\NormalTok{(}\DecValTok{10000}\NormalTok{)}
+\ControlFlowTok{for}\NormalTok{(i }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\DecValTok{10000}\NormalTok{)\{}
+\NormalTok{t1d2 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\NormalTok{t2d2 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\NormalTok{t3d2 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\NormalTok{t4d2 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+
+\NormalTok{factorD2 <-}\StringTok{ }\KeywordTok{sd}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1d2),}\KeywordTok{s}\NormalTok{(t2d2),}\KeywordTok{s}\NormalTok{(t3d2),}\KeywordTok{s}\NormalTok{(t4d2)))}
+\NormalTok{numeratorD2 <-}\StringTok{  }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1d2),}\KeywordTok{s}\NormalTok{(t2d2)) }\OperatorTok{-}\StringTok{ }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t3d2),}\KeywordTok{s}\NormalTok{(t4d2))}
+\NormalTok{biasesD2[i] <-}\StringTok{ }\NormalTok{numeratorD2}\OperatorTok{/}\NormalTok{factorD2}
+\NormalTok{\}}
+
+\KeywordTok{ggplot}\NormalTok{()}\OperatorTok{+}\KeywordTok{geom_histogram}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x=}\NormalTok{biasesD2, }\DataTypeTok{y =}\NormalTok{ ..density..), }\DataTypeTok{alpha =} \FloatTok{0.6}\NormalTok{, }\DataTypeTok{bins=}\DecValTok{50}\NormalTok{)}\OperatorTok{+}
+\StringTok{  }\KeywordTok{theme_tufte}\NormalTok{()}\OperatorTok{+}\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title=}\StringTok{"10k biases for different means and sd =.05"}\NormalTok{)}\OperatorTok{+}\StringTok{ }\KeywordTok{xlab}\NormalTok{(}\StringTok{"bias"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/unnamed-chunk-12-1} \end{center}
+
+\normalsize
+
+\noindent Now suppose we keep the means the same but increase the
+standard deviation to .1.
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{766}\NormalTok{)}
+\NormalTok{t1d1 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
+\NormalTok{t2d1 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
+\NormalTok{t3d1 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
+\NormalTok{t4d1 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+\vspace{1mm} \footnotesize
+
+\begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/unnamed-chunk-14-1} \end{center}
+
+\normalsize
+
+\noindent  Is this clear difference mirrored in the calculations?
+
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{factorD1 <-}\StringTok{ }\KeywordTok{sd}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1d1),}\KeywordTok{s}\NormalTok{(t2d1),}\KeywordTok{s}\NormalTok{(t3d1),}\KeywordTok{s}\NormalTok{(t4d1)))}
+\NormalTok{numeratorD1 <-}\StringTok{  }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1d1),}\KeywordTok{s}\NormalTok{(t2d1)) }\OperatorTok{-}\StringTok{ }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t3d1),}\KeywordTok{s}\NormalTok{(t4d1))}
+\NormalTok{biasD1 <-}\StringTok{ }\NormalTok{numeratorD1 }\OperatorTok{/}\StringTok{ }\NormalTok{factorD1}
+\NormalTok{biasD1}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## [1] 1.223023
+\end{verbatim}
+
+\normalsize
+
+\noindent The absolute value of the effect size is smaller than in the
+previous case!
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{biasesD1 <-}\StringTok{ }\KeywordTok{numeric}\NormalTok{(}\DecValTok{10000}\NormalTok{)}
+
+\ControlFlowTok{for}\NormalTok{(i }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\DecValTok{10000}\NormalTok{)\{}
+\NormalTok{t1d1 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
+\NormalTok{t2d1 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
+\NormalTok{t3d1 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
+\NormalTok{t4d1 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
+
+\NormalTok{factorD1 <-}\StringTok{ }\KeywordTok{sd}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1d1),}\KeywordTok{s}\NormalTok{(t2d1),}\KeywordTok{s}\NormalTok{(t3d1),}\KeywordTok{s}\NormalTok{(t4d1)))}
+\NormalTok{numeratorD1 <-}\StringTok{  }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1d1),}\KeywordTok{s}\NormalTok{(t2d1)) }\OperatorTok{-}\StringTok{ }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t3d1),}\KeywordTok{s}\NormalTok{(t4d1))}
+\NormalTok{biasesD1[i] <-}\StringTok{ }\NormalTok{numeratorD1}\OperatorTok{/}\NormalTok{factorD1}
+\NormalTok{\}}
+
+\KeywordTok{ggplot}\NormalTok{()}\OperatorTok{+}\KeywordTok{geom_histogram}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x=}\NormalTok{biasesD1, }\DataTypeTok{y =}\NormalTok{ ..density..), }\DataTypeTok{alpha =} \FloatTok{0.6}\NormalTok{, }\DataTypeTok{bins=}\DecValTok{50}\NormalTok{)}\OperatorTok{+}
+\StringTok{  }\KeywordTok{theme_tufte}\NormalTok{()}\OperatorTok{+}\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title=}\StringTok{"10k biases for different means and sd =.001"}\NormalTok{)}\OperatorTok{+}\StringTok{ }\KeywordTok{xlab}\NormalTok{(}\StringTok{"bias"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/unnamed-chunk-16-1} \end{center}
+
+\normalsize
+
+This is is a bit better, but still quite some uncertainty is involved,
+far from what systematically low mean-based p-values reported in the
+papers might suggest. Of course, this is a bit of a caricature, as our
+word lists were short (four protected words and 10 attributes). But the
+word lists used in the actually papers are not much longer. In such a
+set-up the key observations are as follows:
+
+\begin{itemize}
+\item
+  Seemingly high bias measures might arise even if the underlying
+  processes actually have the same mean.
+\item
+  Even if the mean remains the same, non-negligible changes can result
+  from a shift in the standard deviation of the original process, and
+  the change might go in the opposite direction than a visualisation of
+  datapoints might suggest, because with the decrease of standard
+  deviation, the factor decreases and the bias increases.
+\item
+  Even if there is a difference in means, the bias metric can be lower,
+  and the uncertainty about it needs to be gauged.
+\item
+  Even if the difference in means is the same, but the variation is
+  different, the bias metric in the long run could tend toward a
+  different value:
+\end{itemize}
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{mean}\NormalTok{(biasesD1); }\KeywordTok{mean}\NormalTok{(biasesD2)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## [1] 1.563319
+\end{verbatim}
+
+\begin{verbatim}
+## [1] 1.690664
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{median}\NormalTok{(biasesD1); }\KeywordTok{median}\NormalTok{(biasesD2)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## [1] 1.638136
+\end{verbatim}
+
+\begin{verbatim}
+## [1] 1.707511
+\end{verbatim}
+
+\normalsize
+and so the point estimations of bias are sensitive to factors other than
+the underlying process means.
 

### 2021-06-10 paperDraft/paperDraft.Rmd

@@
 
-- Even if there is a difference in means, the bias metric can be lower, and the uncertainty about it needs to be gauged.
-
-- Even if the difference in means is the same, but the variation is different, the bias metric in the long run could tend toward a different value:
+- Even if the underlying means are the same, but the variation is different, the bias metric in the long run could tend toward a different value:
 
@@
 
+- Even if there is a difference in means, the bias metric can be lower, and the uncertainty about it needs to be gauged.
+
+- The uncertainty resulting from including the raw datapoint variance into considerations is more extensive than the one suggested by the low p-values obtained from taking means as datapoints.
+
+\vspace{1mm}
+\footnotesize
+```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+quantile(biasesD2, probs = c(0.275,0.975))
+quantile(biasesD1, probs = c(0.275,0.975))
+```
+\normalsize
+
 \newpage

### 2021-06-10 paperDraft/paperDraft.tex

@@
 \item
-  Even if there is a difference in means, the bias metric can be lower,
-  and the uncertainty about it needs to be gauged.
-\item
-  Even if the difference in means is the same, but the variation is
+  Even if the underlying means are the same, but the variation is
   different, the bias metric in the long run could tend toward a
@@
 the underlying process means.
+
+\begin{itemize}
+\item
+  Even if there is a difference in means, the bias metric can be lower,
+  and the uncertainty about it needs to be gauged.
+\item
+  The uncertainty resulting from including the raw datapoint variance
+  into considerations is more extensive than the one suggested by the
+  low p-values obtained from taking means as datapoints.
+\end{itemize}
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{quantile}\NormalTok{(biasesD2, }\DataTypeTok{probs =} \KeywordTok{c}\NormalTok{(}\FloatTok{0.275}\NormalTok{,}\FloatTok{0.975}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+##    27.5%    97.5% 
+## 1.539149 2.166488
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{quantile}\NormalTok{(biasesD1, }\DataTypeTok{probs =} \KeywordTok{c}\NormalTok{(}\FloatTok{0.275}\NormalTok{,}\FloatTok{0.975}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+##    27.5%    97.5% 
+## 1.300648 2.356815
+\end{verbatim}
+
+\normalsize
 

### 2021-06-10 thesis/_book/_main.tex
addds logs

@@
 \fi
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
-  % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
@@
 \author{Alicja Dobrzeniecka}
-\date{2021-06-09}
+\date{2021-06-10}
 
@@
 
-Natural language processing (NLP) is a subfield of computer science that processes and analyzes language in text and speech
-with the use of modern programming methods. It has practical applications in everyday life as it concerns tasks such as email filters,
-smart assistants, search results, language translations, text analytics and so on. Models used to accomplish these tasks need a lot of
-data to learn from. This data originates from humans activities and historical recordings such as texts, messages or speeches. It turns out
-that in the learning process these models can learn implicit biases that reflect harmful stereotypical thinking still present in modern societies.
-One can find methods that aim at identifying and measuring hidden biases and/or try to remove them by modifying the models.
-There are many different types of models in NLP depending on a task that they are supposed to solve.
-However, all of them need as an input words represented by means of numbers and this is accomplished with word embedding models.
-The models usually assign the values based on the context in which the words appear.
-It means that the input data can have enormous influence on the outcome.
-The biases seem to have their primary source in the way the words are assigned the numerical values.
-
-There is considerable amount of literature available on the topic of bias detection and mitigation in NLP models.
-\protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) focuses on gender biases that may be observable while investigating the representation of job occupations and gender in terms of their assigned numerical
-values.
-The authors apply cosine similarity measurement to investigate the phenomenon where (the vectors corresponding to) words related to jobs that are stereotypically associated with a given gender are in fact in the model situated closer to this gender.
-They also use analogy tasks to evaluate if the bias is present in the word embedding model.
-They check analogies by comparing pairs of word vectors, for example they search for the word complementing the puzzle: man is to doctor as woman is to \ldots?
-First they subtract word ``man'' from word ``woman'' and then
-they search for the ranked list of other words pairs that have similar vectors' difference.
-They also include in the formula a threshold to ensure that the resulting pairs could not be randomly picked.
-
-However, as in \protect\hyperlink{ref-Nissim2019Fair}{Nissim, Noord, \& Goot} (\protect\hyperlink{ref-Nissim2019Fair}{2019}) it is pointed out, there are some limitations of this approach.
-According to the authors in practice most of analogies implementations do not return any input words.
-This means that it does not make sense to expect the algorithm to return the same profession for both woman and man.
-Therefore this method seems to be limited in terms of bias detection.
-The other problems regard for example the
-choice of pairs and words that are used to detect the presence of discrimination as it is often subjective and without proper justification.
-Additionally the choice of parameter set in \protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) formula to ensure that word pairs are not picked by random, is also not justified and changing it drastically
-influences the results.
-
-\protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) touches upon the topic of biases regarding race and gender.
-They apply knowledge from well-known psychological studies such as Implicit Association Test to research the relation between human stereotypical thinking and model learnt biases to
-discover close relationship between these two. For the evaluation they use Word Embedding Association Test (WEAT) and the Word Embedding Factual Association Test (WEFAT).
-
-\protect\hyperlink{ref-manzini2019black}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-manzini2019black}{2019}) proposes a novel way of using cosine similarity to obtain the information on assumed resemblance between words.
-They investigate
-an approach that enables them to measure the bias for a class (like gender, religion, race) and express the final result with a single metric.
-
-It is worth noticing the general distinction of biases mentioned in \protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}).
-They refer to the publication concerning Implicit Association Test (Greenwald et al., 1998) that measures the strength of associations between concepts or stereotypes by calculating the time of reaction for special tasks.
-It is worth noticing that humans naturally exhibit some biases and that they do not always cause social concern.
-One can imagine the intuitive associations between for example insects and flowers, and the feelings of pleasantness or unpleasantness.
-In general, people would rather associate flowers with feeling pleasant than insects, and this preference could be named a bias or
-prejudice in some direction. However, this type of preference does not cause an uproar and is a rather morally neutral case. Unfortunately there are other
-biases and prejudices that directly influence the quality of other people's lives and therefore they should be taken care of.
-
-One can find a bunch of various definitions trying to capture what bias and fairness actually are.
-With the choice of the definition, implications into the real-life applications may change as well as it was pointed out in \protect\hyperlink{ref-Mehrabi2019Survey}{Mehrabi, Morstatter, Saxena, Lerman, \& Galstyan} (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}).
-They mark out that there exist different types of biases, the list if long but among others there are historical bias, representation bias, measurement bias.
-This indicates how complex the process itself is.
-Without the proper understanding and awareness of the problem, people are prone to unconsciously sustain this phenomenon.
-
-In the article one can also find the distinction on different types of discrimination, some of them will be shortly described.
-It is worth first mentioning that protected attributes are those qualities, traits or characteristics that one cannot, according to the law, discriminated against.
-Direct discrimination refers to the situation when protected attributes of individuals explicitly result in non-favorable outcomes toward them.
-In contrast in indirect discrimination individuals appear to be treated equally but anyway they end up being treated unjustly due to the hidden effects of biases towards their protected attributes.
-Systemic discrimination takes place when policies, customs or behaviors that result from certain culture or organizational structure lead to discrimination against some groups of people.
-Finally, very common statistical discrimination refers to using
-average group statistics to judge person belonging to the group.
-
-The topic of discrimination is entangled with another concept which is fairness. It is essential to grasp some concepts of fairness to take them into
-consideration while designing implementation of some machine learning model. In Mehrabi2019Survey one may notice that depending on the context and application different definitions may be applied.
-
-The most popular methods focus on comparing the similarity between words from protected groups and those that are considered to be stereotypical or harmful in some way.
-One can find in this group methods such as euclidean distance, dot product or cosine similarity.
-There are also other ways to detect the effects of biases.
-For example through the investigation of the model performance on certain tasks that validate if the model returns some values
-independently on gender or race or not.
-
-In the currently used methods (like cosine similarity) the values of similarity are often aggregated in a way that may lead to false conclusions. For
-example due to the averaging of values and the lack of confidence interval information.
-
-One can find a number of articles on negative real-life implications resulting from the presence of unaddressed biases in the machine learning models.
-
-In the paper we indicate how current methods used to detect biases in natural language models are limited in terms of confidence interval.
-
-Our research tries to answer the question of how to enhance the current way in which the bias detection is performed to make sure that it is
-methodologically valid.
-
-Our hypothesis is that there can be greater understanding of data and bias implications when confidence interval and Bayesian method are applied to the
-methodology.
+Placeholder
 
@@
 
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\hypertarget{ref-Bolukbasi2016Man}{}%
-Bolukbasi, T., Chang, K.-W., Zou, J.
-Y., Saligrama, V., \& Kalai, A.
-(2016).
-Man is to computer programmer as woman is to homemaker?
-Debiasing word embeddings.
-\emph{CoRR}, \emph{abs/1607.06520}.
-Retrieved from \url{http://arxiv.org/abs/1607.06520}
-
-\leavevmode\hypertarget{ref-Caliskan2017Semantics}{}%
-Islam, A.
-C., Bryson, J.
-J., \& Narayanan, A.
-(2016).
-Semantics derived automatically from language corpora necessarily contain human biases.
-\emph{CoRR}, \emph{abs/1608.07187}.
-Retrieved from \url{http://arxiv.org/abs/1608.07187}
-
-\leavevmode\hypertarget{ref-manzini2019black}{}%
-Manzini, T., Lim, Y.
-C., Tsvetkov, Y., \& Black, A.
-W.
-(2019).
-Black is to criminal as caucasian is to police: Detecting and removing multiclass bias in word embeddings.
-Retrieved from \url{http://arxiv.org/abs/1904.04047}
-
-\leavevmode\hypertarget{ref-Mehrabi2019Survey}{}%
-Mehrabi, N., Morstatter, F., Saxena, N., Lerman, K., \& Galstyan, A.
-(2019).
-A survey on bias and fairness in machine learning.
-\emph{CoRR}, \emph{abs/1908.09635}.
-Retrieved from \url{http://arxiv.org/abs/1908.09635}
-
-\leavevmode\hypertarget{ref-Nissim2019Fair}{}%
-Nissim, M., Noord, R.
-van, \& Goot, R.
-van der.
-(2019).
-Fair is better than sensational: Man is to doctor as woman is to doctor.
-\emph{CoRR}, \emph{abs/1905.09866}.
-Retrieved from \url{http://arxiv.org/abs/1905.09866}
-
-\end{CSLReferences}
-
 \end{document}

### 2021-06-10 thesis/_main.Rmd
addds logs

@@
 
+<!-- github-repo: rstudio/bookdown-demo -->
+
+```{r setup, include=FALSE, echo=FALSE}
+library(tidyverse)
+library(knitr)
+
+options(kableExtra.latex.load_packages = FALSE)
+library(kableExtra)
+```
+
+<!-- # Preface {-} -->
+
+<!-- Test  -->
+
+<!-- testsds -->
+
+<!-- \begin{align}  -->
+<!--   f\left(k\right) = \binom{n}{k} p^k\left(1-p\right)^{n-k} -->
+<!--   (\#eq:binom) -->
+<!-- \end{align}  -->
+
+<!-- This is a citation [@diamond90] which uses keys from the bib file listed in the preamble. -->
+
+<!-- Equation \@ref(eq:binom)^[This is a footnote containing a double citation [@diamond90; @dahlmanNakedStat2020; @Zhao2019Gender].] -->
+
+<!-- Note that chapter files are found and compiled automatically, but the file names have to contain chapter numbers first.
+For instance, we used `01-intro.Rmd`, placed in the same folder.
+Observe how we included r code inline.
+-->
+
+<!-- ```{r fig-margin, fig.margin=TRUE} -->
+<!-- plot(cars) -->
+<!-- ``` -->
+
 <!--chapter:end:index.Rmd-->
 
+---
+output:
+  pdf_document: default
+  html_document: default
+---
 # Introduction
 
-Placeholder
+<!--### Articles:
+
+  1. "Are We Consistently Biased? Multidimensional Analysis of Biases in Distributional Word Vectors"
+  https://www.aclweb.org/anthology/S19-1010.pdf
+
+  2. Proving the harmfullness of biases in models 
+[source https://poseidon01.ssrn.com/delivery.php?ID=004020091089126117028081015085009075120015077012021005105005104113074127101113104030041107053119007034112081087110119071025006053026022082043000071011123067077124004027040024008122116091025075126111098104007096110113089007070120009111119015084097017092&EXT=pdf&INDEX=TRUE]
+
+  3. Summary what causes bias short article
+  [source https://kawine.github.io/blog/nlp/2019/09/23/bias.html]-->
+
+<!-- Men Also Like Shopping: Reducing Gender Bias Amplification using Corpus-level Constraint -->
+<!-- Bias in Bios: A Case Study of Semantic Representation Bias in a High-Stakes Setting -->
+
+<!-- !bibliografia paperdraft; jabref
+4.pomysly na czwarty rozdzial - podsumowanie; 
+jak mierzyc bias za pomoca jednej liczby?
+IAT - test psychologiczny -->
+
+<!-- 
+    1.state the general topic and give some background
+    2.provide a review of the literature related to the topic
+    3.define the terms and scope of the topic
+    4.outline the current situation
+    5.evaluate the current situation (advantages/ disadvantages) and identify the gap
+    6.identify the importance of the proposed research
+    7.state the research problem/ questions
+    8.state the research aims and/or research objectives
+    9.state the hypotheses
+    10.outline the order of information in the thesis
+    11.outline the methodology
+-->
+
+<!-- \textbf{state the general topic and give some background}-->
+
+Natural language processing (NLP) is a subfield of computer science that processes and analyzes language in text and speech 
+with the use of modern programming methods. It has practical applications in everyday life as it concerns tasks such as email filters,
+smart assistants, search results, language translations, text analytics and so on. Models used to accomplish these tasks need a lot of
+data to learn from. This data originates from humans activities and historical recordings such as texts, messages or speeches. It turns out
+that in the learning process these models can learn implicit biases that reflect harmful stereotypical thinking still present in modern societies.
+One can find methods that aim at identifying and measuring hidden biases and/or try to remove them by modifying the models.
+There are many different types of models in NLP depending on a task that they are supposed to solve.
+However, all of them need as an input words represented by means of numbers and this is accomplished with word embedding models.
+The models usually assign the values based on the context in which the words appear.
+It means that the input data can have enormous influence on the outcome.
+The biases seem to have their primary source in the way the words are assigned the numerical values.
+
+<!-- \textbf{review of the literature related to the topic}-->
+
+There is considerable amount of literature available on the topic of bias detection and mitigation in NLP models.
+@Bolukbasi2016Man focuses on gender biases that may be observable while investigating the representation of job occupations and gender in terms of their assigned numerical
+values.
+The authors apply cosine similarity measurement to investigate the phenomenon where (the vectors corresponding to) words related to jobs that are stereotypically associated with a given gender are in fact in the model situated closer to this gender.
+They also use analogy tasks to evaluate if the bias is present in the word embedding model.
+They check analogies by comparing pairs of word vectors, for example they search for the word complementing the puzzle: man is to doctor as woman is to ...?
+First they subtract word "man" from word "woman" and then
+they search for the ranked list of other words pairs that have similar vectors' difference.
+They also include in the formula a threshold to ensure that the resulting pairs could not be randomly picked.
+
+However, as in @Nissim2019Fair it is pointed out, there are some limitations of this approach.
+According to the authors in practice most of analogies implementations do not return any input words.
+This means that it does not make sense to expect the algorithm to return the same profession for both woman and man.
+Therefore this method seems to be limited in terms of bias detection.
+The other problems regard for example the
+choice of pairs and words that are used to detect the presence of discrimination as it is often subjective and without proper justification.
+Additionally the choice of parameter set in @Bolukbasi2016Man formula to ensure that word pairs are not picked by random, is also not justified and changing it drastically
+influences the results.
+
+@Caliskan2017Semantics touches upon the topic of biases regarding race and gender.
+They apply knowledge from well-known psychological studies such as Implicit Association Test to research the relation between human stereotypical thinking and model learnt biases to
+discover close relationship between these two. For the evaluation they use Word Embedding Association Test (WEAT) and the Word Embedding Factual Association Test (WEFAT).
+
+@manzini2019black proposes a novel way of using cosine similarity to obtain the information on assumed resemblance between words. They investigate 
+an approach that enables them to measure the bias for a class (like gender, religion, race) and express the final result with a single metric.
+
+<!-- Nissim2019Fair -->
+
+<!-- \textbf{define the terms and scope of the topic}-->
+
+It is worth noticing the general distinction of biases mentioned in @Caliskan2017Semantics.
+They refer to the publication concerning Implicit Association Test (Greenwald et al., 1998) that measures the strength of associations between concepts or stereotypes by calculating the time of reaction for special tasks.
+It is worth noticing that humans naturally exhibit some biases and that they do not always cause social concern.
+One can imagine the intuitive associations between for example insects and flowers, and the feelings of pleasantness or unpleasantness.
+In general, people would rather associate flowers with feeling pleasant than insects, and this preference could be named a bias or
+prejudice in some direction. However, this type of preference does not cause an uproar and is a rather morally neutral case. Unfortunately, there are other
+biases and prejudices that directly influence the quality of other people's lives and therefore they should be taken care of.
+
+One can find various definitions trying to capture what bias and fairness actually are.
+With the choice of the definition, implications into the real-life applications may change as well.
+@Mehrabi2019Survey mark out that there exist different types of biases such as historical bias, representation bias, measurement bias (the list is long).
+This indicates how complex the issue of bias is.
+Without the proper understanding and awareness of the problem, people are prone to unconsciously sustain the bias existence.
+
+@Mehrabi2019Survey also distinguish different types of discrimination, some of them will be briefly described.
+By protected attributes we mean those qualities, traits or characteristics that one cannot legally discriminate against.
+Direct discrimination occurs when protected attributes of individuals explicitly result in non-favorable outcomes toward them.
+In contrast in indirect discrimination individuals appear to be treated equally but anyway they end up being treated unjustly due to the hidden effects of biases towards their protected attributes.
+Systemic discrimination takes place when policies, customs or behaviors that result from certain culture or organizational structure lead to discrimination against some groups of people.
+Finally, very common statistical discrimination refers to using
+average group statistics to judge person belonging to the group.
+
+The topic of discrimination is entangled with another concept which is fairness. It is essential to grasp some concepts of fairness to take them into
+consideration while designing implementation of some machine learning model. In Mehrabi2019Survey one may notice that depending on the context and application different definitions may be applied.
+
+<!--\textbf{outline the current situation} -->
+
+The most popular methods focus on comparing the similarity between words from protected groups and those that are considered to be stereotypical or harmful in some way.
+One can find in this group methods such as euclidean distance or cosine similarity (which is equivalent to dot product if the
+vectors are normalized).
+There are also other ways to detect the effects of biases.
+For example through the investigation of the model performance on certain tasks that validate if the model returns some values
+independently on gender or race or not.
+
+<!-- \textbf{evaluate the current situation (advantages/ disadvantages) and identify the gap}-->
+
+The currently used methods (such as cosine similarity) make the similarity values often aggregated in a way that may lead to hasty conclusions.
+The averaging of values and the lack of uncertainty may lead to the incomplete picture of the bias situation in the vocabulary.
+
+<!-- \textbf{identify the importance of the proposed research}-->
+One can find a number of articles on negative real-life implications resulting from the presence of unaddressed biases in the machine learning models.
+
+<!-- \textbf{state the research problem/ questions} -->
+
+In the paper we indicate how current methods used to detect biases in natural language models are limited from the perspective of Bayesian analysis.
+
+<!-- \textbf{state the research aims and/or research objectives} -->
+
+Our research enhances the current way in which the bias detection is performed to make sure that it is
+methodologically valid.
+
+<!-- \textbf{state the hypotheses} -->
+
+The key hypothesis is that greater understanding of data and bias implications can be achieved when Bayesian methods are applied to issue.
+
+<!-- \textbf{outline the order of information in the thesis} -->
 
@@
 
-Placeholder
+## Loading and understanding the dataset
 
-## Loading and understanding the dataset
+We start with loading the libraries needed for the analysis.
+
+\footnotesize
+```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%", warning = FALSE, message = FALSE}
+library(ggplot2)
+library(ggthemes)
+library(rethinking)
+library(tidyverse)
+library(ggpubr)
+library(kableExtra)
+library(dplyr)
+library(ggExtra)
+library(cowplot)
+```
+\normalsize
+
+We will use the choice of protected words and stereotypical predicates used in REF.
+This is a decent point of departure,  not only we want to compare our method to that of REF, but also because this data format is fairly general (as contrasted, say, with a set up for binary stereotypes).
+Note also that the method we develop here  can fairly easily be  run  for different stereotypization patterns.
+Let's start  with  explaining the method and its deployment using a dataset obtained for the religion-related protected words.
+
+Let's load, clean a bit and inspect the head of the religion dataset we prepared.
+In order to obtain this  dataset, we  calculated the cosine distance between each protected word and each word from both the bias-related attribute groups, which were used in the original study, and to neutral and human control attributes which we added  as control groups.
+For instance, for religion, the bias-related predicates (coming from the original study in REF) include muslim bias attributes, jew bias attributes, christian bias attributes (see a list in the APPENDIX).
+
+We decided to add control groups in the form of two classes --- neutral words and human-related words.
+Without a proper control group it is quite hard to compare the resulting cosine distances and decide on their significance in bias detection.
+We prepared approximately 300\todo{FIX later} more or less neutral words to double-check the prima-facie neutral hypothesis that their cosine similarity to the protected words will oscillate around 0 (that is, the distances will be around 1).
+This provides us with a  more reliable point of reference.
+Moreover, we added human attributes that are associated with people in general to investigate  whether the smaller cosine distance between protected words and stereotypes can result  simply from the fact that the stereotype predicates  are associated with humans.
+For two control groups, we have randomly drawn 300 words that do not express any property usually attributed to humans, and human attributes.
+\todo{describe once the words are selected}
+
+<!-- In other words, we took each protected word and in nested iteration calculated all of the cosine distances between the word and all available harmful stereotypes, neutral words, and human ones, which resulted with approximately 333 cosine distances per each protected word.
+-->
+
+\vspace{1mm}
+\footnotesize
+
+```{r religionTableHead,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+religion <- read.csv("../datasets/religionReddit.csv")[-1]
+colnames(religion) <- c("protectedWord","wordToCompare","wordClass",
+                        "cosineDistance","cosineSimilarity","connection")
+levels(religion$wordClass) <- c("christian","human","jewish","muslim","neutral")
+head(religion)  %>%  kable(format = "latex",booktabs=T,
+                      linesep = "",  escape = FALSE, 
+                      caption = "Head of the religion dataset.") %>%
+                      kable_styling(latex_options=c("scale_down"))
+```
+\normalsize
+
+ The `protectedWord` column contains words from a protected class that (in a perfect world according to the assumptions of the orignal study) should not be associated with  harmful stereotypes.
+`wordToCompare` contains attributes, including stereotypes and control group words.
+For each row we compute the cosine distances between a given protected  word and a given attribute word.
+`wordClass` tells us  which class an attribute is supposed to be stereotypically associated with, that is, whether the  word from `wordToCompare` is associated stereotypically with jews, christians or muslims, or whether it belongs to a control group.
+`cosineDistance` is simply a calculation of the cosine distance between protected word and atrribute.
+`cosineSimilarity`  contains the result of substracting cosine distance from 1.
+`connection` contains information about  the relation type between a protected word and an attribute.
+If the attribute is e.g.
+a harmful jewish stereotype and the protected word is also from the judaism group,  the connection has value `associated`.
+If the attribute is still stereotypically jewish, but the protected word comes from another religion, the connection is labelled as  `different`.
+If the attribute belongs to  a neutral group then the connection is labelled as `none` and if an attribute belongs to the `human` class, then the connection is labelled as `human`.
+
 ## First look at the empirical distributions
+
+First let's take a look at the empirical distribution of distances by the connection type, initially ignoring the human control class for now.
+
+\vspace{1mm}
+\footnotesize
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+religionNoHumans <- religion[religion$connection !="human",]
+
+ggplot(religionNoHumans, aes(x =  cosineDistance, fill = connection, color = connection ))+geom_density(alpha=0.6,size = .2)+theme_tufte()+ggtitle("Empirical distribution of cosine distances (religion), no human attributes.")+ scale_fill_manual(values = c("orangered4","chartreuse4","gray"))+scale_x_continuous(breaks = seq(0.3,1.5, by = 0.1))+xlab("cosine distance")+ scale_color_manual(values = c("orangered4","chartreuse4","gray"))
+```
+\normalsize
+
+The first impression is that while  there is a  shift for  associated words towards smaller cosine distances as compared to the neutral words, slightly surprisingly a slightly weaker shift in the same direction is visible for attributes associated with different stereotypes.
+Moreover,  the empirical distributions overlap to alarge extent and the  means grouped by connection type do not seem too far from each other.
+In fact, as there is a lot of  variety in the consine distances (as we will soon see), abd we need to gauge the uncertaintly involved, and  to look more carefully at individual protected words to get a better idea of how the cosine distance distribution changes for different attribute groups and different protected classes.
+Now, let's add the human attributes to the picture:
+
+\vspace{1mm}
+\footnotesize
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+ggplot(religion, aes(x =  cosineDistance, fill = connection, color = connection))+geom_density(alpha=0.6,size = .2)+theme_tufte()+ggtitle("Empirical distribution of cosine distances (religion)")+ scale_fill_manual(values = c("orangered4","chartreuse4", "skyblue", "gray"))+scale_x_continuous(breaks = seq(0.3,1.5, by = 0.1))+xlab("cosine distance")+ scale_color_manual(values = c("orangered4","chartreuse4","skyblue","gray"))
+```
+\normalsize
+
+\noindent Notice that the distribution for `human` (even though we did our best not to include in it any stereotype-related atributes) is left-skewed, with much overlap with `associated` and `different`, which illustrates the need to take being associated with humans as an important predictor.
+
+Our focus lies in  `connection` as a predictor.
+Morever, later on we'll be interested in looking at the protected words separately, and at protected words split by connection.
+For technical reasons it is useful to represent these factors as integer vectors.
+
+\vspace{1mm}
+\footnotesize
+```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+religion$con <- as.integer(religion$connection)
+religion$pw <- as.integer(religion$protectedWord)
+religion$pwFactor <- factor(paste0(religion$protectedWord, religion$connection))
+religion$pwIndex <- as.integer(religion$pwFactor)
+```
+\normalsize
+
+A short script, `cleanDataset` to make this faster, so equivalently:
+
+\vspace{1mm}
+\footnotesize
+```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+source("../functions/cleanDataset.R")
+religion <- read.csv("../datasets/religionReddit.csv")[-1]
+religion <- cleanDataset(religion,c("christian","human","jewish","muslim","neutral"))
+```
+\normalsize
+
 ## Looking at the islam-related words
+
+For now, let's focus on five protected words related to islam ("imam", "islam", "mosque", "muslim", and "quran").
+The word list associates with islam  four stereotypical attributes ("violent", "terrorist", "uneducated" and "dirty").
+First, we select and plot the empirical distributions for these protected words.
+
+\vspace{1mm}
+\footnotesize
+```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%",message =FALSE, warning = FALSE}
+library(tidyverse)
+muslimWords <- c("imam","islam","mosque","muslim","quran")
+muslim <- religion %>% filter(protectedWord %in% muslimWords)
+ggplot(muslim, aes(x =  cosineDistance, fill = connection, color = connection))+
+  geom_density(alpha=0.6,size = .2)+ 
+  scale_fill_manual(values = c("orangered4","chartreuse4", "skyblue", "gray"))+
+  scale_x_continuous(breaks = seq(0.3,1.5, by = 0.1))+xlab("cosine distance")+
+  scale_color_manual(values = c("orangered4","chartreuse4","skyblue","gray"))+
+  theme_tufte()+ggtitle("Empirical distribution of distances (muslim)")
+```
+\normalsize
+
+\noindent Once we focus on words related to islam, the associated bias seems to be stronger than in the whole dataset.  This is a  step towards illustrating that the distribution of bias is uneven.
+
+Now, say we want to look at a single protected word.
+Since the dataset also contains comparison multiple control neutral and  human attributes, we randomly select only 5 from `none` and 5 from `human` control groups of those for the visualisation purposes.
+
+\vspace{1mm}
+\footnotesize
+```{r tableMuslimActive,echo=TRUE,eval=TRUE,fig.align = "center",cache=FALSE, fig.show = "hold", out.width = "100%"}
+library(tidyverse)
+muslimClass <- muslim %>% filter(protectedWord == "muslim")
+neutralSample <- sample_n(filter(muslimClass,connection == "none"), 5)
+humanSample <- sample_n(filter(muslimClass,connection == "human"), 5)
+muslimVis <- muslimClass %>% filter(connection != "none" & connection !="human")
+muslimVis <- rbind(muslimVis,neutralSample,humanSample)
+
+#we plug in our visualisation script
+source("../functions/visualisationTools.R")
+#two arguments: dataset and protected word
+visualiseProtected(muslimVis,"muslim")
+```
+\normalsize
+
+Note that the distance between the grey point and the other points is proportional to cosine distance, the non-grey point size is proportional to cosine similarity to the protected word, and color groups by the connection type.
+So for `muslim` it seems that the stereotypes coming from the word list are fairly well visible.
+To give you some taste of how uneven the dataset is, compare this to what happens with `priest`.
+
+\vspace{1mm}
+\footnotesize
+```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+library(tidyverse)
+priestClass <- religion %>% filter(protectedWord == "priest")
+neutralSample <- sample_n(filter(priestClass,connection == "none"), 5)
+humanSample <- sample_n(filter(priestClass,connection == "human"), 5)
+priestVis <- priestClass %>% filter(connection != "none" & connection !="human")
+priestVis <- rbind(priestVis,neutralSample,humanSample)
+
+#we plug in our visualisation script
+source("../functions/visualisationTools.R")
+#two arguments: dataset and protected word
+visualiseProtected(priestVis,"priest")
+```
+\normalsize
+
+\noindent Here you can see that some human attributes are closer than stereotype attributes, and that there is no clear reason to claim that `associated` attributes are closer than `different`  or `human` attributes.
+This, again, illustrates the need of case-by-case analysis with  control groups.
+
+The general idea now is that the word lists provided in different pieces of research are just samples of attributes associates with various stereotypes and should be treated as such: the uncertaintly involved and the sample sizes should have clear impact on our estimates.
+
 ## Bayesian model structure and assumptions
+
+We will now think of cosine distance as the output variable, and 
+will build a few bayesian models to compare.
+First, we just build a baseline model which estimates cosine distance to the attributes separately for each protected word.
+The underlying idea is that different protected words migh in general have different relations to all the attributes and this relations should be our point of departure.
+
+Here is the intution behind the mathematical Bayesian model involved.
+Our outcome variable is `cosine difference`, which we take to me normally distributed around the predicted mean for a given protected word (that is, we assume the residuals are normally distributed).
+The simplest  model specification is:
+
+\begin{align}
+cosineDistance_i  & \sim dnorm(\mu_i, \sigma) \\
+\mu_i & = m_{pw} \\
+m_{pw} & ~ dnorm(1,.5) \\
+\sigma &\sim  dcauchy(0,1)
+\end{align}
+
+That is, we assume the estimated means might be different for diferent protected words and our prior for the mean and the overal standard deviation are normal with mean 1 and sd=.5 and half-cauchy with parameters `0,1`.
+Further on we'll also estimate additional impact the connection type may have.
+For this impact we take a slightly skeptical prior centered around 0 distributed normally with sd = 1.
+These are fairly weak and slightly skeptical regularizing priors, which can be illustrated as follows:
+
+\vspace{2mm}
+
+```{r priorsVis,echo=FALSE,eval=TRUE,fig.align = "center",cache=FALSE, out.width = "100%", fig.caption = "Regularizing priors for the bayesian models."}
+library(ggpubr)
+sim <- seq(-1,1,by = 0.001)
+dis <- 1-sim
+
+p <- ggplot(data = data.frame(distance = dis), 
+            mapping = aes(x = distance))+theme_tufte()
+
+none <- function(x) dnorm(x,1,.5)
+cau <- function(x) dcauchy(x,0,1)
+par <- function(x) dnorm(x,0,1)
+
+noneG <- p + stat_function(fun = none)+xlim(c(-0.5,2.5))+ggtitle("Prior for mean distances")
+parG <- p + stat_function(fun = par)+xlim(c(-2,2))+xlab("distance change")+ggtitle("Prior for coefficients")
+cauG <- p + stat_function(fun = cau)+xlim(c(0,2))+ggtitle("Prior for standard deviation")+xlab("distance change")
+ggarrange(noneG,cauG, parG, ncol = 1)
+```
+
 ## Choosing predictors
+
+Now we can define and compile the baseline model.
+Its  parameters will have a posterior distribution obtained using either Hamiltionian Monte Carlo methods (STAN) available through the `rethinking` package.
+
+\vspace{1mm}
+\footnotesize
+```{r religionBaseline,echo=TRUE,eval=FALSE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%", results= "hide"}
+library(rethinking)
+options(buildtools.check = function(action) TRUE )
+religionBaseline <- ulam(
+  alist(
+    cosineDistance ~ dnorm(mu,sigma),
+    mu <- m[pw],
+    m[pw] ~ dnorm(1,.5),
+    sigma ~ dcauchy(0,1)
+  ),
+  data = religion,
+  chains=2 , iter=4000 , warmup=1000,
+  start= list(mu = 1, co = 0, sigma= .3),
+  log_lik = TRUE, cores=4
+)
+#saving
+#saveRDS(religionBaseline, 
+#file = "cosineAnalysis/models/religionBaseline.rds")
+```
+
+The only reason we need it is the evaluation of connection as a predictor.
+Does including it in o the model improve the situation?
+To investigate this, let's now build a model according to the following specification:
+
+\begin{align}
+cosineDistance_i  & \sim dnorm(\mu_i, \sigma) \\
+\mu_i & = m_{pw} + co_{con}\\
+m_{pw} & ~ dnorm(1,.5) \\
+co_{con} & ~ dnorm(0,1) \\
+\sigma &\sim  dcauchy(0,1)
+\end{align}
+
+\noindent The idea now is that each connection type comes with its own coefficient $co$ that has impact on mean distances for protected words taken separately.
+
+\vspace{1mm}
+\footnotesize
+```{r Coefsmodel,echo=TRUE,eval=FALSE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+library(rethinking)
+options(buildtools.check = function(action) TRUE )
+religionCoefs <- ulam(
+  alist(
+    cosineDistance ~ dnorm(mu,sigma),
+    mu <- m[pw] + co[con],
+    m[pw] ~ dnorm(1,.5),
+    co[con] ~dnorm(0,.5),
+    sigma ~ dcauchy(0,1)
+  ),
+  data = religion,
+  chains=2 , iter=8000 , warmup=1000, 
+  log_lik = TRUE
+)
+```
+\normalsize
+
+\noindent First, let's see if this model is really better in terms of the Widely Acceptable Information Criterion (WAIC):
+
+\vspace{1mm}
+\footnotesize
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+#the calculation requires models which are large, we prepared the table
+#compareBaseline <- print(round(compare(religionBaseline, religionCoefs)), 3)
+compareBaseline <- readRDS("../datasets/compareBaseline.rds")
+compareBaseline
+```
+\normalsize
+
+Clearly, it should be given weight 1 as compared to the baseline model. So far, we've learned that the connection type actually has predictive value. Let's take a look at the coefficient estimates:
+
+ \vspace{1mm}
+ \footnotesize
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+#precisCoefs <- print(precis(religionCoefs, depth =2, pars= "co" ))
+#saveRDS(precisCoefs, "../datasets/precisCoefs.rds")
+precisCoefs <- readRDS("../datasets/precisCoefs.rds")
+precisCoefs
+```
+ \normalsize
+
 ## Dataset-level coefficients
+
+\noindent Let's plot them together with their highest posterior density invervals, for the three topic groups.
+
+\begin{center}
+\begin{figure}[!htb]\centering
+   \begin{minipage}{0.55\textwidth}
+  \includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
+   \end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/religionGoogleCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/genderCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/genderGoogleCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/raceCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/raceGoogleCoeffs.jpeg}
+   \end{minipage}
+\end{figure}
+
+\end{center}
+
+\noindent What should strike us is that while the mean estimates of the coefficients indeed do differ a bit, usually the highest posterior density invervals all include zero, and so we do not have strong reasons to say that, say, as far as the whole religion dataset is involved, being associated indeed is connected  with lower cosine distance.
+A second striking observation is that the estimated impact for associated stereotypes is quite often not too different from the estimated impact of attributes associated with different stereotypes, and both are sometimes not too far from the estimated impact for simply human attributes.
+In general, once the uncertainty involved is taken seriously by using control groups and statistical uncertainty estimation that does not dispose of pointwise data, the picture which focuses only on differences between means of means is too simplistic.
+
+But this doesn't mean important differences for some protected words are not there.
+For one thing, if you start with a word list that is very uneven, the actually not so bad status of some of the protected words might mask  a pretty bad situation in which some other protected words are.
+For comparison, let's see what a model focused on words related to islam tells us.
+
+\vspace{1mm}
+\footnotesize
+```{r,echo=TRUE,eval=FALSE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+#this is how we build the model
+religion <- read.csv("cosineAnalysis/datasets/religionReddit.csv")[-1]
+colnames(religion) <- c("protectedWord","wordToCompare","wordClass",
+                        "cosineDistance","cosineSimilarity","connection")
+levels(religion$wordClass) <- c("christian","human","jewish","muslim","neutral")
+muslimWords <- c("imam","islam","mosque","muslim","quran")
+muslim <- religion %>% filter(protectedWord %in% muslimWords)
+muslim$protectedWord <- droplevels(muslim$protectedWord)
+muslim$pw <- as.integer(muslim$protectedWord)
+muslim$con <- as.integer(muslim$connection)
+muslim$pwFactor <- factor(paste0(muslim$protectedWord, muslim$connection))
+muslim$pwIndex <- as.integer(muslim$pwFactor)
+
+islamCoefs <- ulam(
+  alist(
+    cosineDistance ~ dnorm(mu,sigma),
+    mu <- m[pw] + co[con],
+    m[pw] ~ dnorm(1,.5),
+    co[con] ~dnorm(0,.5),
+    sigma ~ dcauchy(0,1)
+  ),
+  data = muslim,
+  chains=2 , iter=10000 , warmup=1000, cores = 4,
+  log_lik = TRUE
+)
+```
+\normalsize
+
+Let's take a look at the coefficients:
+
+\vspace{1mm}
+\footnotesize
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+islamCoefs <- readRDS("../datasets/islamCoefs.rds")
+islamCoefs
+```
+\normalsize
+
+\begin{center}
+\includegraphics[width=8cm]{../images/islamCoeffs.jpeg}
+\end{center}
+
+\normalsize
+
+While muslim words were unusual in the sense that the disparity between associated attributes and others is stronger, the evidence is still not conclusive.
+This is because the variation even within islam-related words is large enough (and sample sizes sufficiently small) for all the highest posterior density invervals to still include zeros.
+
+So, it seems, taking a closer look does seem to make a difference. The question is, what happens if we do take a close look at the level of protected words?
+
+#  Protected-word level analysis
+
 ## Model structure and assumptions
+
+Let's turn then to data analysis that takes a look at protected words separately.
+This time for each combination of a protected word and a connection status we will have a separate mean cosine distance estimate, each coming with its own highest posterior density interval.
+This means we will use indices that are result from all such combinations (and then we will split them up in the model precis to build visualisation, feel free to look at the `visualiseStats.R` script for details).
+
+\vspace{1mm}
+\footnotesize
+```{r build religionSeparate,echo=TRUE,eval=FALSE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+options(buildtools.check = function(action) TRUE ) #removes install pop-up request
+religion$pwFactor <- factor(paste0(religion$protectedWord, "-", religion$connection))
+religion$pwIndex <- as.integer(religion$pwFactor)
+
+religionSeparate <- ulam(
+  alist(
+    cosineDistance ~ dnorm(mu,sigma),
+    mu <- c[pwIndex],
+    c[pwIndex] ~ dnorm(1,.5),
+    sigma ~ dcauchy(0,1)
+  ),
+  data = religion,
+  chains=2 , iter=10000 , warmup=1000,
+  start=list(no = 1, a = 0, d = 0, sigma= .3), log_lik =  TRUE
+)
+```
+\normalsize
+
+\noindent Let's see if the individualized model does better than the previous models in light of WAIC which does add penalty for the number of parameters.
+
+\vspace{1mm}
+\footnotesize
+```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+compareBaselineCoefsSeparate<- readRDS("../datasets/compareBaselineCoefsSeparate.rds")
+compareBaselineCoefsSeparate
+```
+\normalsize
+
 ## Protected classes in Reddit and Google embeddings
+
+It seems that we do want to prefer this model, despite its relative complication.
+Now, what does it tell us about the protected words?
+Let's visualise the predicted means together with 89\% highest posterior density intervals.
+
+\includegraphics[width=14cm]{../images/visReligionReddit.png}
+
+Before we move on, let's perform analogous analyses for the remaining types of supposed bias: gender and race (the model building is analogous).
+
+\includegraphics[width=14cm]{../images/visGenderReddit.png}
+
+\includegraphics[width=14cm]{../images/visRaceReddit.png}
+
+\todo{Explain where you took the embedding from, cite publications first announcing it.}
+
+\includegraphics[width=14cm]{../images/visReligionGoogle.png}
+
+\includegraphics[width=14cm]{../images/visGenderGoogle.png}
+
+\includegraphics[width=14cm]{../images/visRaceGoogle.png}
+
+# The role of debiasing
+
+\todo{describe which debiasing you used and how it was advertised in the paper}
+
 ## Dataset-level coefficients after debiasing
+
+First, let's look at coefficient estimated for the whole datasets, as compared to their estimation prior to debiasing:
+
+\begin{center}
+\begin{figure}[!htb]\centering
+   \begin{minipage}{0.55\textwidth}
+  \includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
+   \end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/debiasedReligionRedditCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/genderCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/debiasedGenderRedditCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/raceCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/debiasedRaceRedditCoeffs.jpeg}
+   \end{minipage}
+\end{figure}
+
+\end{center}
+
+\todo{comment on this}
+
+Some points to discuss:
+
+- humans in religion got closer (should they?)
+
+- change in Gender is really minor, still zero out of HPDI range, note how different is very similar
+
+- small improvement in race, at the price of moving neutral terms closer
+
 ## Protected classes  after debiasing
 
+Now, perhaps, the effects of debiasing will be better appreciated if we look at the level of protected words.
+After all, the hope is, the situation of extremely ill-positioned protected words have improved?
+
+\includegraphics[width=14cm]{../images/visDebReligionReddit.png}
+
+\includegraphics[width=14cm]{../images/visDebGenderReddit.png}
+
+\includegraphics[width=14cm]{../images/visDebRaceReddit.png}
+
 <!--chapter:end:03-bayesian.Rmd-->

### 2021-06-10 thesis/_main.tex
addds logs

--- 2021-06-09 thesis/_main.Rmd
@@
----
-title: "Conceptual and methodological problems with bias detection and avoidance in natural language processing"
-author: "Alicja Dobrzeniecka"
-date: '`r Sys.Date()`'
-output:
-  bookdown::pdf_book: 
-    keep_tex: yes
-    toc_depth: 6
-    latex_engine: xelatex
-    extra_dependencies: ["todonotes"]
-  bookdown::gitbook:
-    lib_dir: "book_assets"
-includes:
-    in_header:
-      - Rafal_latex6.sty
-documentclass: book
-classoptions: dvipsnames, usenames
-site: bookdown::bookdown_site
-bibliography: [../references/bibliographyRafal.bib]
-biblio-style: 
-link-citations: yes
-fontsize: 12pt
-linestretch: 1.5
-geometry: left=3.5cm, right=2.5cm, top=2.5cm, bottom=2.5cm, asymmetric, includeheadfoot
-csl: apa-5th-edition.csl
----
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+%
+\documentclass[
+  12pt,
+]{book}
+\usepackage{amsmath,amssymb}
+\usepackage{lmodern}
+\usepackage{setspace}
+\usepackage{ifxetex,ifluatex}
+\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
+  \usepackage{unicode-math}
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+% Use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+\IfFileExists{microtype.sty}{% use microtype if available
+  \usepackage[]{microtype}
+  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\makeatletter
+\@ifundefined{KOMAClassName}{% if non-KOMA class
+  \IfFileExists{parskip.sty}{%
+    \usepackage{parskip}
+  }{% else
+    \setlength{\parindent}{0pt}
+    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
+}{% if KOMA class
+  \KOMAoptions{parskip=half}}
+\makeatother
+\usepackage{xcolor}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\hypersetup{
+  pdftitle={Conceptual and methodological problems with bias detection and avoidance in natural language processing},
+  pdfauthor={Alicja Dobrzeniecka},
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
+\urlstyle{same} % disable monospaced font for URLs
+\usepackage[left=3.5cm, right=2.5cm, top=2.5cm, bottom=2.5cm, asymmetric, includeheadfoot]{geometry}
+\usepackage{color}
+\usepackage{fancyvrb}
+\newcommand{\VerbBar}{|}
+\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+% Add ',fontsize=\small' for more characters per line
+\usepackage{framed}
+\definecolor{shadecolor}{RGB}{248,248,248}
+\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\BuiltInTok}[1]{#1}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
+\newcommand{\ExtensionTok}[1]{#1}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ImportTok}[1]{#1}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\NormalTok}[1]{#1}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\RegionMarkerTok}[1]{#1}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
+\usepackage{graphicx}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+\setcounter{secnumdepth}{5}
+\usepackage{todonotes}
+\ifluatex
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
+  % set entry spacing
+  \ifnum #2 > 0
+  \setlength{\parskip}{#2\baselineskip}
+  \fi
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
-<!--chapter:end:index.Rmd-->
+\title{Conceptual and methodological problems with bias detection and avoidance in natural language processing}
+\author{Alicja Dobrzeniecka}
+\date{2021-06-10}
 
-# Introduction
+\begin{document}
+\maketitle
 
-Placeholder
+{
+\setcounter{tocdepth}{5}
+\tableofcontents
+}
+\setstretch{1.5}
+\hypertarget{introduction}{%
+\chapter{Introduction}\label{introduction}}
 
-<!--chapter:end:01-intro.Rmd-->
+Natural language processing (NLP) is a subfield of computer science that processes and analyzes language in text and speech
+with the use of modern programming methods. It has practical applications in everyday life as it concerns tasks such as email filters,
+smart assistants, search results, language translations, text analytics and so on. Models used to accomplish these tasks need a lot of
+data to learn from. This data originates from humans activities and historical recordings such as texts, messages or speeches. It turns out
+that in the learning process these models can learn implicit biases that reflect harmful stereotypical thinking still present in modern societies.
+One can find methods that aim at identifying and measuring hidden biases and/or try to remove them by modifying the models.
+There are many different types of models in NLP depending on a task that they are supposed to solve.
+However, all of them need as an input words represented by means of numbers and this is accomplished with word embedding models.
+The models usually assign the values based on the context in which the words appear.
+It means that the input data can have enormous influence on the outcome.
+The biases seem to have their primary source in the way the words are assigned the numerical values.
 
----
-output:
-  pdf_document: default
-  html_document: default
----
-# Cosine similarity and bias detection
+There is considerable amount of literature available on the topic of bias detection and mitigation in NLP models.
+\protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) focuses on gender biases that may be observable while investigating the representation of job occupations and gender in terms of their assigned numerical
+values.
+The authors apply cosine similarity measurement to investigate the phenomenon where (the vectors corresponding to) words related to jobs that are stereotypically associated with a given gender are in fact in the model situated closer to this gender.
+They also use analogy tasks to evaluate if the bias is present in the word embedding model.
+They check analogies by comparing pairs of word vectors, for example they search for the word complementing the puzzle: man is to doctor as woman is to \ldots?
+First they subtract word ``man'' from word ``woman'' and then
+they search for the ranked list of other words pairs that have similar vectors' difference.
+They also include in the formula a threshold to ensure that the resulting pairs could not be randomly picked.
 
-## Word embeddings
-<!-- [based on https://machinelearningmastery.com/what-are-word-embeddings/] -->
+However, as in \protect\hyperlink{ref-Nissim2019Fair}{Nissim, Noord, \& Goot} (\protect\hyperlink{ref-Nissim2019Fair}{2019}) it is pointed out, there are some limitations of this approach.
+According to the authors in practice most of analogies implementations do not return any input words.
+This means that it does not make sense to expect the algorithm to return the same profession for both woman and man.
+Therefore this method seems to be limited in terms of bias detection.
+The other problems regard for example the
+choice of pairs and words that are used to detect the presence of discrimination as it is often subjective and without proper justification.
+Additionally the choice of parameter set in \protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) formula to ensure that word pairs are not picked by random, is also not justified and changing it drastically
+influences the results.
+
+\protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) touches upon the topic of biases regarding race and gender.
+They apply knowledge from well-known psychological studies such as Implicit Association Test to research the relation between human stereotypical thinking and model learnt biases to
+discover close relationship between these two. For the evaluation they use Word Embedding Association Test (WEAT) and the Word Embedding Factual Association Test (WEFAT).
+
+\protect\hyperlink{ref-manzini2019black}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-manzini2019black}{2019}) proposes a novel way of using cosine similarity to obtain the information on assumed resemblance between words.
+They investigate
+an approach that enables them to measure the bias for a class (like gender, religion, race) and express the final result with a single metric.
+
+It is worth noticing the general distinction of biases mentioned in \protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}).
+They refer to the publication concerning Implicit Association Test (Greenwald et al., 1998) that measures the strength of associations between concepts or stereotypes by calculating the time of reaction for special tasks.
+It is worth noticing that humans naturally exhibit some biases and that they do not always cause social concern.
+One can imagine the intuitive associations between for example insects and flowers, and the feelings of pleasantness or unpleasantness.
+In general, people would rather associate flowers with feeling pleasant than insects, and this preference could be named a bias or
+prejudice in some direction. However, this type of preference does not cause an uproar and is a rather morally neutral case. Unfortunately, there are other
+biases and prejudices that directly influence the quality of other people's lives and therefore they should be taken care of.
+
+One can find various definitions trying to capture what bias and fairness actually are.
+With the choice of the definition, implications into the real-life applications may change as well.
+\protect\hyperlink{ref-Mehrabi2019Survey}{Mehrabi, Morstatter, Saxena, Lerman, \& Galstyan} (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) mark out that there exist different types of biases such as historical bias, representation bias, measurement bias (the list is long).
+This indicates how complex the issue of bias is.
+Without the proper understanding and awareness of the problem, people are prone to unconsciously sustain the bias existence.
+
+\protect\hyperlink{ref-Mehrabi2019Survey}{Mehrabi, Morstatter, Saxena, Lerman, \& Galstyan} (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) also distinguish different types of discrimination, some of them will be briefly described.
+By protected attributes we mean those qualities, traits or characteristics that one cannot legally discriminate against.
+Direct discrimination occurs when protected attributes of individuals explicitly result in non-favorable outcomes toward them.
+In contrast in indirect discrimination individuals appear to be treated equally but anyway they end up being treated unjustly due to the hidden effects of biases towards their protected attributes.
+Systemic discrimination takes place when policies, customs or behaviors that result from certain culture or organizational structure lead to discrimination against some groups of people.
+Finally, very common statistical discrimination refers to using
+average group statistics to judge person belonging to the group.
+
+The topic of discrimination is entangled with another concept which is fairness. It is essential to grasp some concepts of fairness to take them into
+consideration while designing implementation of some machine learning model. In Mehrabi2019Survey one may notice that depending on the context and application different definitions may be applied.
+
+The most popular methods focus on comparing the similarity between words from protected groups and those that are considered to be stereotypical or harmful in some way.
+One can find in this group methods such as euclidean distance or cosine similarity (which is equivalent to dot product if the
+vectors are normalized).
+There are also other ways to detect the effects of biases.
+For example through the investigation of the model performance on certain tasks that validate if the model returns some values
+independently on gender or race or not.
+
+The currently used methods (such as cosine similarity) make the similarity values often aggregated in a way that may lead to hasty conclusions.
+The averaging of values and the lack of uncertainty may lead to the incomplete picture of the bias situation in the vocabulary.
+
+One can find a number of articles on negative real-life implications resulting from the presence of unaddressed biases in the machine learning models.
+
+In the paper we indicate how current methods used to detect biases in natural language models are limited from the perspective of Bayesian analysis.
+
+Our research enhances the current way in which the bias detection is performed to make sure that it is
+methodologically valid.
+
+The key hypothesis is that greater understanding of data and bias implications can be achieved when Bayesian methods are applied to issue.
+
+\hypertarget{cosine-similarity-and-bias-detection}{%
+\chapter{Cosine similarity and bias detection}\label{cosine-similarity-and-bias-detection}}
+
+\hypertarget{word-embeddings}{%
+\section{Word embeddings}\label{word-embeddings}}
 
@@
 
-word | 1 | 2 | 3 | 4 | 5  
-|---|---|---|---|---|---|
-woman | 1 | 0 | 0 | 0 | 0 
-man | 0 | 1 | 0 | 0 | 0 
-girl | 0 | 0 | 1 | 0 | 0 
-boy | 0 | 0 | 0 | 1 | 0 
-monarch | 0 | 0 | 0 | 0 | 1
+\begin{longtable}[]{@{}llllll@{}}
+\toprule
+word & 1 & 2 & 3 & 4 & 5 \\
+\midrule
+\endhead
+woman & 1 & 0 & 0 & 0 & 0 \\
+man & 0 & 1 & 0 & 0 & 0 \\
+girl & 0 & 0 & 1 & 0 & 0 \\
+boy & 0 & 0 & 0 & 1 & 0 \\
+monarch & 0 & 0 & 0 & 0 & 1 \\
+\bottomrule
+\end{longtable}
 
@@
 
-word | 1 | 2 | 3 | 4   
-|---|---|---|---|---|
-woman | 0.456 | 0.267 | 0.675 | 0.131
-man | 0.451 | 0.897 | 0.472 | 0.088
-girl | 0.604 | 0.262 | 0.414 | 0.706 
-boy | 0.279 | 0.172 | 0.475 | 0.010 
-monarch | 0.565 | 0.678 | 0.463 | 0.975
+\begin{longtable}[]{@{}lllll@{}}
+\toprule
+word & 1 & 2 & 3 & 4 \\
+\midrule
+\endhead
+woman & 0.456 & 0.267 & 0.675 & 0.131 \\
+man & 0.451 & 0.897 & 0.472 & 0.088 \\
+girl & 0.604 & 0.262 & 0.414 & 0.706 \\
+boy & 0.279 & 0.172 & 0.475 & 0.010 \\
+monarch & 0.565 & 0.678 & 0.463 & 0.975 \\
+\bottomrule
+\end{longtable}
 
@@
 
-## Cosine similarity and distance
-<!-- [based on https://deepai.org/machine-learning-glossary-and-terms/cosine-similarity] -->
+\hypertarget{cosine-similarity-and-distance}{%
+\section{Cosine similarity and distance}\label{cosine-similarity-and-distance}}
 
@@
 \end{align}
-<!-- add information about inner product role -->
 Cosine similarity is considered a proper tool for this operation as its result has a clear connection to geometry and at least for a low number of dimensions may be easily interpreted.
@@
 
-<!-- todo: plotting exemplary vectors in 2d -->
 One of the limitations of this measure is that it informs us only about similarities between vectors in terms of their orientation.
 However, it is often argued that in comparing words in terms of this metric, the magnitude of vectors may be treated as irrelevant, as the most important information pertains to direction.
-<!-- todo: find reference-->
 
@@
 
-## Cosine distance in a one-class bias detection
-<!-- https://arxiv.org/pdf/1607.06520.pdf -->
-<!-- source of plot https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings -->
+\hypertarget{cosine-distance-in-a-one-class-bias-detection}{%
+\section{Cosine distance in a one-class bias detection}\label{cosine-distance-in-a-one-class-bias-detection}}
 
-@Bolukbasi2016Man
+\protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016})
 define similarity between words as the outcome inner product of their normalized vectors.
-They focus on examining what the geometry of word embedding is in regard to "he" and "she" words.
+They focus on examining what the geometry of word embedding is in regard to ``he'' and ``she'' words.
 In other words, whether the similarity between those concepts and other words reflects expected gender stereotypes.
@@
 A very vivid way to follow their method of arguing that bias in word embeddings is real is to plot the values of inner product of chosen words.
-The plot below does not originate from the original paper (it is from https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings) but similar visualization may be found there.
+The plot below does not originate from the original paper (it is from \url{https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings}) but similar visualization may be found there.
 Data used to create our plot is as follows.
@@
 \footnotesize
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
-genderProjections <-  read.csv("../datasets/genderProjections.csv")[,-1]
-colnames(genderProjections)[4] <- "Stereotype"
-levels(genderProjections$Stereotype) <- c("male occupation", "female occupation")
-ggplot(genderProjections, aes(x=he,y=she, color = Stereotype, shape = Stereotype))+geom_point(size = 2, alpha = 0.75)+theme_tufte()+labs(title ="Stereotypical professions projected on the he-she axis (dot product, unnormalized)", subtitle ="GloVe embeddings trained on the Wikipedia 2014 and Gigaword 5th Edition corpus")+scale_color_manual(values =c("orangered4","chartreuse4"))
-```
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-1-1} \end{center}
 \normalsize
@@
 
-<!-- todo: shortly describe what they do later and how they verify if it was removed - analogies etc. -->
+\hypertarget{cosine-distance-in-a-multi-class-bias-detection}{%
+\section{Cosine distance in a multi-class bias detection}\label{cosine-distance-in-a-multi-class-bias-detection}}
 
-## Cosine distance in a multi-class bias detection
-<!-- [article https://arxiv.org/pdf/1904.04047.pdf] -->
-
-@manzini2019black present a different approach towards finding similarities between classes of words.
+\protect\hyperlink{ref-manzini2019black}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-manzini2019black}{2019}) present a different approach towards finding similarities between classes of words.
 The authors claim that texts available online are full of direct or indirect human stereotypes.
@@
 Cosine distance is used in the article as a measure to first argue for the existence of multi-class bias and then to show how through bias mitigation techniques the bias may be decreased.
-
-<!-- The methodology used in the paper  -->
 
@@
 
-1. First we load word embeddings from reddit.US.txt.tok.clean.cleanedforw2v.w2v dataset
-  - The word embeddings have only 50 dimensions
-  - The number of individual words from the dataset is 44895
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  First we load word embeddings from reddit.US.txt.tok.clean.cleanedforw2v.w2v dataset
+\end{enumerate}
 
-2.
+\begin{itemize}
+\tightlist
+\item
+  The word embeddings have only 50 dimensions
+\item
+  The number of individual words from the dataset is 44895
+\end{itemize}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{1}
+\tightlist
+\item
   The protected group the authors assume should ideally not have high cosine similarity to stereotypical words.
@@
   \newline
+\end{enumerate}
 
@@
 \footnotesize
-```{r religionTableHeadEarly,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
-religion <- read.csv("../datasets/religionReddit.csv")[-1]
-colnames(religion) <- c("protectedWord","wordToCompare","wordClass",
-                        "cosineDistance","cosineSimilarity","connection")
-levels(religion$wordClass) <- c("christian","human","jewish","muslim","neutral")
-head(religion)  %>%  kable(format = "latex",booktabs=T,
-                      linesep = "",  escape = FALSE, 
-                      caption = "Head of the religion dataset.") %>%
-                      kable_styling(latex_options=c("scale_down"))
-```
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{religion }\OtherTok{\textless{}{-}} \FunctionTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{]}
+\FunctionTok{colnames}\NormalTok{(religion) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
+                        \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
+\FunctionTok{levels}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{wordClass) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
+\FunctionTok{head}\NormalTok{(religion)  }\SpecialCharTok{\%\textgreater{}\%}  \FunctionTok{kable}\NormalTok{(}\AttributeTok{format =} \StringTok{"latex"}\NormalTok{,}\AttributeTok{booktabs=}\NormalTok{T,}
+                      \AttributeTok{linesep =} \StringTok{""}\NormalTok{,  }\AttributeTok{escape =} \ConstantTok{FALSE}\NormalTok{, }
+                      \AttributeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+                      \FunctionTok{kable\_styling}\NormalTok{(}\AttributeTok{latex\_options=}\FunctionTok{c}\NormalTok{(}\StringTok{"scale\_down"}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{table}
+
+\caption{\label{tab:religionTableHeadEarly}Head of the religion dataset.}
+\centering
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{lllrrl}
+\toprule
+protectedWord & wordToCompare & wordClass & cosineDistance & cosineSimilarity & connection\\
+\midrule
+judaism & violent & muslim_stereotype & 0.7141939 & 0.2858061 & different\\
+judaism & terrorist & muslim_stereotype & 0.7461333 & 0.2538667 & different\\
+judaism & dirty & muslim_stereotype & 1.2002599 & -0.2002599 & different\\
+judaism & uneducated & muslim_stereotype & 0.7885469 & 0.2114531 & different\\
+judaism & greedy & jew_stereotype & 1.0026172 & -0.0026172 & associated\\
+judaism & cheap & jew_stereotype & 1.2323229 & -0.2323229 & associated\\
+\bottomrule
+\end{tabular}}
+\end{table}
 \normalsize
@@
 
-3. In the article the authors also try to remove previously defined biases from word embedding.
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{2}
+\tightlist
+\item
+  In the article the authors also try to remove previously defined biases from word embedding.
+\end{enumerate}
 
@@
 
-<!-- go through it again -->
-
-4. In the final step there is an evaluation of the results. The cosine distances are calculated again but this time using debiased vocabulary. After
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{3}
+\tightlist
+\item
+  In the final step there is an evaluation of the results. The cosine distances are calculated again but this time using debiased vocabulary. After
   taking the mean of all distances one final value is obtained and then it is compared with the average value from the beginning.
@@
 As the cosine distance increases it is assumed that the association between protected and stereotypical words decreases.
+\end{enumerate}
 
-<!-- add more information on the source of words used -->
-
-## Limitations of the approach
+\hypertarget{limitations-of-the-approach}{%
+\section{Limitations of the approach}\label{limitations-of-the-approach}}
 
 \textbf{1. Selection of attributes}
-
-<!-- TODO: Universal ways to decide on the quality and type of attributes so that the research is significant? -->
 
@@
 Assuming for a moment that the value of multi-class cosine distance is correct, one may question the results' interpretation.
-In @manzini2019black Table 2, there are summarized the averages of cosine distance per group (gender, race, religion).
+In \protect\hyperlink{ref-manzini2019black}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-manzini2019black}{2019}) Table 2, there are summarized the averages of cosine distance per group (gender, race, religion).
 I would like to focus now on analyzing the values relating to religious biases.
@@
 
-Religion Debiasing  | MAC
-------------- | -------------
-Biased        | 0.859
-Hard Debiased | 0.934
-Soft Debiased ($\lambda$ = 0.2) | 0.894
+\begin{longtable}[]{@{}ll@{}}
+\toprule
+Religion Debiasing & MAC \\
+\midrule
+\endhead
+Biased & 0.859 \\
+Hard Debiased & 0.934 \\
+Soft Debiased (\(\lambda\) = 0.2) & 0.894 \\
+\bottomrule
+\end{longtable}
 
 MAC stand for mean average cosine similarity although in reality the values of cosine distance are stored there.
-What may attract attention is the fact that the value of cosine distance in "Biased" category is already quite high even before "debiasing".
-High cosine distance indicates low cosine similarity between values.
+What may attract attention is the fact that the value of cosine distance in ``Biased'' category is already quite high even before ``debiasing.'' High cosine distance indicates low cosine similarity between values.
 One could think that average cosine similarity equal to approximately 0.141 is not significant enough to consider it as biased.
-However the authors aim to mitigate "biases" in vectors with such great distance to make it even larger.
+However the authors aim to mitigate ``biases'' in vectors with such great distance to make it even larger.
 Methodologically there is a question on what basis this small similarity is still considered as a proof of bias presence.
@@
 
-<!-- TODO: numerical test if this distance is significant? check mean mean cosine distances of some random words? -->
-
 \textbf{6. The curse of dimensionality}
 
-<!-- TODO: -->
-<!-- 1. https://www.cs.princeton.edu/courses/archive/fall13/cos521/lecnotes/lec11.pdf -->
-<!-- 2. https://stats.stackexchange.com/questions/341535/curse-of-dimensionality-does-cosine-similarity-work-better-and-if-so-why -->
-<!-- 3. https://stats.stackexchange.com/questions/21547/distance-metric-and-curse-of-dimensions -->
-<!-- + check if it may be proved in the code -->
-
 Curse of dimensionality may take place when there is an increase in volume of data that results in adding extra dimensions to the Euclidean space. According to the article
-"https://analyticsindiamag.com/curse-of-dimensionality-and-what-beginners-should-do-to-overcome-it/"
+``\url{https://analyticsindiamag.com/curse-of-dimensionality-and-what-beginners-should-do-to-overcome-it/}''
 as the number of features increases, it may be harder and harder to obtain useful information from the data with the usage of available algorithms.
@@
 Using cosine similarity in high dimensions in word embeddings may also be prone to the curse of dimensionality.
-According to this article "https://www.researchgate.net/publication/327498046_The_Curse_of_Dimensionality_Inside_Out"
+According to this article ``\url{https://www.researchgate.net/publication/327498046_The_Curse_of_Dimensionality_Inside_Out}''
 there are reasons to consider this phenomenon when searching for word similarities in higher dimensions.
-
-<!-- Add more information on the experiment -->
 
@@
 
-![curse of dimensionality,  number of dimensions on the x axis,  standard deviation of similarity on the y axis](../images/curseOfDimensionality.png)
+\begin{figure}
+\centering
+\includegraphics{../images/curseOfDimensionality.png}
+\caption{curse of dimensionality, number of dimensions on the x axis, standard deviation of similarity on the y axis}
+\end{figure}
 
-<!-- textbf{Control group}
+\hypertarget{walkthrough-with-the-religion-dataset}{%
+\chapter{Walkthrough with the religion dataset}\label{walkthrough-with-the-religion-dataset}}
 
-How to properly prepare control group in terms of quality and quantity?
-<!-- + check if it may be proved in the code -->
+\hypertarget{loading-and-understanding-the-dataset}{%
+\section{Loading and understanding the dataset}\label{loading-and-understanding-the-dataset}}
 
-<!--  \textbf{Verifying other similarity meaures}
+We start with loading the libraries needed for the analysis.
 
-Besides cosine similarity, there are other methods used to find the similarity between vectors.
+\footnotesize
 
-a) TS-SS 
-todo: check if it makes sense to calculate for all of the datasets; interpretations etc.
-https://www.researchgate.net/publication/303513110_A_Hybrid_Geometric_Approach_for_Measuring_Similarity_Level_Among_Documents_and_Document_Clustering
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{library}\NormalTok{(ggplot2)}
+\FunctionTok{library}\NormalTok{(ggthemes)}
+\FunctionTok{library}\NormalTok{(rethinking)}
+\FunctionTok{library}\NormalTok{(tidyverse)}
+\FunctionTok{library}\NormalTok{(ggpubr)}
+\FunctionTok{library}\NormalTok{(kableExtra)}
+\FunctionTok{library}\NormalTok{(dplyr)}
+\FunctionTok{library}\NormalTok{(ggExtra)}
+\FunctionTok{library}\NormalTok{(cowplot)}
+\end{Highlighting}
+\end{Shaded}
 
-b) WEAT test 
-<!-- https://docs.responsibly.ai/word-embedding-bias.html#module-responsibly.we.weat -->
+\normalsize
 
-<!-- [to check https://github.com/taki0112/Vector_Similarity] -->
+We will use the choice of protected words and stereotypical predicates used in REF.
+This is a decent point of departure, not only we want to compare our method to that of REF, but also because this data format is fairly general (as contrasted, say, with a set up for binary stereotypes).
+Note also that the method we develop here can fairly easily be run for different stereotypization patterns.
+Let's start with explaining the method and its deployment using a dataset obtained for the religion-related protected words.
 
-<!-- TODO: check in code if they make more sense -->
+Let's load, clean a bit and inspect the head of the religion dataset we prepared.
+In order to obtain this dataset, we calculated the cosine distance between each protected word and each word from both the bias-related attribute groups, which were used in the original study, and to neutral and human control attributes which we added as control groups.
+For instance, for religion, the bias-related predicates (coming from the original study in REF) include muslim bias attributes, jew bias attributes, christian bias attributes (see a list in the APPENDIX).
 
-<!-- ## Bibliography
-- https://www.amazon.com/Language-Processing-Synthesis-Lectures-Technologies/dp/1627052984/ref=as_li_ss_tl?ie=UTF8&qid=1502062931&sr=8-1&keywords=Neural+Network+Methods+in+Natural+Language+Processing&linkCode=sl1&tag=inspiredalgor-20&linkId=d63df073fea3ebe2d405820570b3ff03 ->
+We decided to add control groups in the form of two classes --- neutral words and human-related words.
+Without a proper control group it is quite hard to compare the resulting cosine distances and decide on their significance in bias detection.
+We prepared approximately 300\todo{FIX later} more or less neutral words to double-check the prima-facie neutral hypothesis that their cosine similarity to the protected words will oscillate around 0 (that is, the distances will be around 1).
+This provides us with a more reliable point of reference.
+Moreover, we added human attributes that are associated with people in general to investigate whether the smaller cosine distance between protected words and stereotypes can result simply from the fact that the stereotype predicates are associated with humans.
+For two control groups, we have randomly drawn 300 words that do not express any property usually attributed to humans, and human attributes.
+\todo{describe once the words are selected}
 
-<!-- ## *** Additional -->
+\vspace{1mm}
+\footnotesize
 
-<!-- ### Man is to Computer Programmer as Woman is to Homemaker? Debiasing Word Embeddings" -->
-<!-- [article https://arxiv.org/pdf/1607.06520.pdf] -->
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{religion }\OtherTok{\textless{}{-}} \FunctionTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{]}
+\FunctionTok{colnames}\NormalTok{(religion) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
+                        \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
+\FunctionTok{levels}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{wordClass) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
+\FunctionTok{head}\NormalTok{(religion)  }\SpecialCharTok{\%\textgreater{}\%}  \FunctionTok{kable}\NormalTok{(}\AttributeTok{format =} \StringTok{"latex"}\NormalTok{,}\AttributeTok{booktabs=}\NormalTok{T,}
+                      \AttributeTok{linesep =} \StringTok{""}\NormalTok{,  }\AttributeTok{escape =} \ConstantTok{FALSE}\NormalTok{, }
+                      \AttributeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+                      \FunctionTok{kable\_styling}\NormalTok{(}\AttributeTok{latex\_options=}\FunctionTok{c}\NormalTok{(}\StringTok{"scale\_down"}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
 
-<!-- Assumptions: -->
-<!-- "We assume there is a set of gender neutral words $N\subseteq W$ such as flight attendant or shoes, which, by definition, are not specific to any gender." -->
+\begin{table}
 
-<!-- TODO: check if it is useful for the analysis of black is .. -->
+\caption{\label{tab:religionTableHead}Head of the religion dataset.}
+\centering
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{lllrrl}
+\toprule
+protectedWord & wordToCompare & wordClass & cosineDistance & cosineSimilarity & connection\\
+\midrule
+judaism & violent & muslim_stereotype & 0.7141939 & 0.2858061 & different\\
+judaism & terrorist & muslim_stereotype & 0.7461333 & 0.2538667 & different\\
+judaism & dirty & muslim_stereotype & 1.2002599 & -0.2002599 & different\\
+judaism & uneducated & muslim_stereotype & 0.7885469 & 0.2114531 & different\\
+judaism & greedy & jew_stereotype & 1.0026172 & -0.0026172 & associated\\
+judaism & cheap & jew_stereotype & 1.2323229 & -0.2323229 & associated\\
+\bottomrule
+\end{tabular}}
+\end{table}
+\normalsize
 
-<!-- ### Identifying the bias subspace  -->
+The \texttt{protectedWord} column contains words from a protected class that (in a perfect world according to the assumptions of the orignal study) should not be associated with harmful stereotypes.
+\texttt{wordToCompare} contains attributes, including stereotypes and control group words.
+For each row we compute the cosine distances between a given protected word and a given attribute word.
+\texttt{wordClass} tells us which class an attribute is supposed to be stereotypically associated with, that is, whether the word from \texttt{wordToCompare} is associated stereotypically with jews, christians or muslims, or whether it belongs to a control group.
+\texttt{cosineDistance} is simply a calculation of the cosine distance between protected word and atrribute.
+\texttt{cosineSimilarity} contains the result of substracting cosine distance from 1.
+\texttt{connection} contains information about the relation type between a protected word and an attribute.
+If the attribute is e.g.~a harmful jewish stereotype and the protected word is also from the judaism group, the connection has value \texttt{associated}.
+If the attribute is still stereotypically jewish, but the protected word comes from another religion, the connection is labelled as \texttt{different}.
+If the attribute belongs to a neutral group then the connection is labelled as \texttt{none} and if an attribute belongs to the \texttt{human} class, then the connection is labelled as \texttt{human}.
 
-<!-- Interesting analysis about limits o subspace and research [source https://kawine.github.io/blog/nlp/2019/09/23/bias.html] -->
+\hypertarget{first-look-at-the-empirical-distributions}{%
+\section{First look at the empirical distributions}\label{first-look-at-the-empirical-distributions}}
 
-<!-- Resources to analyze: -->
-<!-- 1. https://arxiv.org/pdf/2005.00965.pdf -->
-<!-- 2. https://arxiv.org/pdf/2009.09435.pdf -->
-<!-- 3. https://arxiv.org/pdf/1904.04047.pdf -->
-<!-- 4. https://www.aclweb.org/anthology/P19-1166.pdf -->
+First let's take a look at the empirical distribution of distances by the connection type, initially ignoring the human control class for now.
 
-<!-- ### How is cosine similarity entangled with analogy topic in NLP (is it??) -->
+\vspace{1mm}
+\footnotesize
 
-<!-- article:  https://arxiv.org/pdf/1905.09866.pdf -->
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-3-1} \end{center}
+\normalsize
 
-<!-- Critique of the usage of analogies as a proof for gender biases -->
+The first impression is that while there is a shift for associated words towards smaller cosine distances as compared to the neutral words, slightly surprisingly a slightly weaker shift in the same direction is visible for attributes associated with different stereotypes.
+Moreover, the empirical distributions overlap to alarge extent and the means grouped by connection type do not seem too far from each other.
+In fact, as there is a lot of variety in the consine distances (as we will soon see), abd we need to gauge the uncertaintly involved, and to look more carefully at individual protected words to get a better idea of how the cosine distance distribution changes for different attribute groups and different protected classes.
+Now, let's add the human attributes to the picture:
 
-<!-- TODO: check if there is other argument than that algorithm does not allow it -->
+\vspace{1mm}
+\footnotesize
 
-<!-- general thoughts 
-https://www.researchgate.net/publication/349408535_Robustness_and_Reliability_of_Gender_Bias_Assessment_in_Word_Embeddings_The_Role_of_Base_Pairs
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-4-1} \end{center}
+\normalsize
 
--
--
--
--->
+\noindent Notice that the distribution for \texttt{human} (even though we did our best not to include in it any stereotype-related atributes) is left-skewed, with much overlap with \texttt{associated} and \texttt{different}, which illustrates the need to take being associated with humans as an important predictor.
 
-<!--chapter:end:02-cosine.Rmd-->
+Our focus lies in \texttt{connection} as a predictor.
+Morever, later on we'll be interested in looking at the protected words separately, and at protected words split by connection.
+For technical reasons it is useful to represent these factors as integer vectors.
 
-# Walkthrough with the religion dataset
+\vspace{1mm}
+\footnotesize
 
-Placeholder
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{religion}\SpecialCharTok{$}\NormalTok{con }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{connection)}
+\end{Highlighting}
+\end{Shaded}
 
-## Loading and understanding the dataset
-## First look at the empirical distributions
-## Looking at the islam-related words
-## Bayesian model structure and assumptions
-## Choosing predictors
-## Dataset-level coefficients
-## Model structure and assumptions
-## Protected classes in Reddit and Google embeddings
-## Dataset-level coefficients after debiasing
-## Protected classes  after debiasing
+\begin{verbatim}
+## Warning: NAs introduced by coercion
+\end{verbatim}
 
-<!--chapter:end:03-bayesian.Rmd-->
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{religion}\SpecialCharTok{$}\NormalTok{pw }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{protectedWord)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning: NAs introduced by coercion
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{religion}\SpecialCharTok{$}\NormalTok{pwFactor }\OtherTok{\textless{}{-}} \FunctionTok{factor}\NormalTok{(}\FunctionTok{paste0}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{protectedWord, religion}\SpecialCharTok{$}\NormalTok{connection))}
+\NormalTok{religion}\SpecialCharTok{$}\NormalTok{pwIndex }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{pwFactor)}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+A short script, \texttt{cleanDataset} to make this faster, so equivalently:
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{source}\NormalTok{(}\StringTok{"../functions/cleanDataset.R"}\NormalTok{)}
+\NormalTok{religion }\OtherTok{\textless{}{-}} \FunctionTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{]}
+\NormalTok{religion }\OtherTok{\textless{}{-}} \FunctionTok{cleanDataset}\NormalTok{(religion,}\FunctionTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning in cleanDataset(religion, c("christian", "human", "jewish", "muslim", :
+## NAs introduced by coercion
+
+## Warning in cleanDataset(religion, c("christian", "human", "jewish", "muslim", :
+## NAs introduced by coercion
+\end{verbatim}
+
+\normalsize
+
+\hypertarget{looking-at-the-islam-related-words}{%
+\section{Looking at the islam-related words}\label{looking-at-the-islam-related-words}}
+
+For now, let's focus on five protected words related to islam (``imam,'' ``islam,'' ``mosque,'' ``muslim,'' and ``quran'').
+The word list associates with islam four stereotypical attributes (``violent,'' ``terrorist,'' ``uneducated'' and ``dirty'').
+First, we select and plot the empirical distributions for these protected words.
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{library}\NormalTok{(tidyverse)}
+\NormalTok{muslimWords }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"imam"}\NormalTok{,}\StringTok{"islam"}\NormalTok{,}\StringTok{"mosque"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"quran"}\NormalTok{)}
+\NormalTok{muslim }\OtherTok{\textless{}{-}}\NormalTok{ religion }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(protectedWord }\SpecialCharTok{\%in\%}\NormalTok{ muslimWords)}
+\FunctionTok{ggplot}\NormalTok{(muslim, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{  cosineDistance, }\AttributeTok{fill =}\NormalTok{ connection, }\AttributeTok{color =}\NormalTok{ connection))}\SpecialCharTok{+}
+  \FunctionTok{geom\_density}\NormalTok{(}\AttributeTok{alpha=}\FloatTok{0.6}\NormalTok{,}\AttributeTok{size =}\NormalTok{ .}\DecValTok{2}\NormalTok{)}\SpecialCharTok{+} 
+  \FunctionTok{scale\_fill\_manual}\NormalTok{(}\AttributeTok{values =} \FunctionTok{c}\NormalTok{(}\StringTok{"orangered4"}\NormalTok{,}\StringTok{"chartreuse4"}\NormalTok{, }\StringTok{"skyblue"}\NormalTok{, }\StringTok{"gray"}\NormalTok{))}\SpecialCharTok{+}
+  \FunctionTok{scale\_x\_continuous}\NormalTok{(}\AttributeTok{breaks =} \FunctionTok{seq}\NormalTok{(}\FloatTok{0.3}\NormalTok{,}\FloatTok{1.5}\NormalTok{, }\AttributeTok{by =} \FloatTok{0.1}\NormalTok{))}\SpecialCharTok{+}\FunctionTok{xlab}\NormalTok{(}\StringTok{"cosine distance"}\NormalTok{)}\SpecialCharTok{+}
+  \FunctionTok{scale\_color\_manual}\NormalTok{(}\AttributeTok{values =} \FunctionTok{c}\NormalTok{(}\StringTok{"orangered4"}\NormalTok{,}\StringTok{"chartreuse4"}\NormalTok{,}\StringTok{"skyblue"}\NormalTok{,}\StringTok{"gray"}\NormalTok{))}\SpecialCharTok{+}
+  \FunctionTok{theme\_tufte}\NormalTok{()}\SpecialCharTok{+}\FunctionTok{ggtitle}\NormalTok{(}\StringTok{"Empirical distribution of distances (muslim)"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-7-1} \end{center}
+\normalsize
+
+\noindent Once we focus on words related to islam, the associated bias seems to be stronger than in the whole dataset. This is a step towards illustrating that the distribution of bias is uneven.
+
+Now, say we want to look at a single protected word.
+Since the dataset also contains comparison multiple control neutral and human attributes, we randomly select only 5 from \texttt{none} and 5 from \texttt{human} control groups of those for the visualisation purposes.
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{library}\NormalTok{(tidyverse)}
+\NormalTok{muslimClass }\OtherTok{\textless{}{-}}\NormalTok{ muslim }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(protectedWord }\SpecialCharTok{==} \StringTok{"muslim"}\NormalTok{)}
+\NormalTok{neutralSample }\OtherTok{\textless{}{-}} \FunctionTok{sample\_n}\NormalTok{(}\FunctionTok{filter}\NormalTok{(muslimClass,connection }\SpecialCharTok{==} \StringTok{"none"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
+\NormalTok{humanSample }\OtherTok{\textless{}{-}} \FunctionTok{sample\_n}\NormalTok{(}\FunctionTok{filter}\NormalTok{(muslimClass,connection }\SpecialCharTok{==} \StringTok{"human"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
+\NormalTok{muslimVis }\OtherTok{\textless{}{-}}\NormalTok{ muslimClass }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(connection }\SpecialCharTok{!=} \StringTok{"none"} \SpecialCharTok{\&}\NormalTok{ connection }\SpecialCharTok{!=}\StringTok{"human"}\NormalTok{)}
+\NormalTok{muslimVis }\OtherTok{\textless{}{-}} \FunctionTok{rbind}\NormalTok{(muslimVis,neutralSample,humanSample)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning in `[<-.factor`(`*tmp*`, ri, value = structure(c("muslim_stereotype", :
+## invalid factor level, NA generated
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#we plug in our visualisation script}
+\FunctionTok{source}\NormalTok{(}\StringTok{"../functions/visualisationTools.R"}\NormalTok{)}
+\CommentTok{\#two arguments: dataset and protected word}
+\FunctionTok{visualiseProtected}\NormalTok{(muslimVis,}\StringTok{"muslim"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/tableMuslimActive-1} \end{center}
+\normalsize
+
+Note that the distance between the grey point and the other points is proportional to cosine distance, the non-grey point size is proportional to cosine similarity to the protected word, and color groups by the connection type.
+So for \texttt{muslim} it seems that the stereotypes coming from the word list are fairly well visible.
+To give you some taste of how uneven the dataset is, compare this to what happens with \texttt{priest}.
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{library}\NormalTok{(tidyverse)}
+\NormalTok{priestClass }\OtherTok{\textless{}{-}}\NormalTok{ religion }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(protectedWord }\SpecialCharTok{==} \StringTok{"priest"}\NormalTok{)}
+\NormalTok{neutralSample }\OtherTok{\textless{}{-}} \FunctionTok{sample\_n}\NormalTok{(}\FunctionTok{filter}\NormalTok{(priestClass,connection }\SpecialCharTok{==} \StringTok{"none"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
+\NormalTok{humanSample }\OtherTok{\textless{}{-}} \FunctionTok{sample\_n}\NormalTok{(}\FunctionTok{filter}\NormalTok{(priestClass,connection }\SpecialCharTok{==} \StringTok{"human"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
+\NormalTok{priestVis }\OtherTok{\textless{}{-}}\NormalTok{ priestClass }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(connection }\SpecialCharTok{!=} \StringTok{"none"} \SpecialCharTok{\&}\NormalTok{ connection }\SpecialCharTok{!=}\StringTok{"human"}\NormalTok{)}
+\NormalTok{priestVis }\OtherTok{\textless{}{-}} \FunctionTok{rbind}\NormalTok{(priestVis,neutralSample,humanSample)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning in `[<-.factor`(`*tmp*`, ri, value = structure(c("muslim_stereotype", :
+## invalid factor level, NA generated
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#we plug in our visualisation script}
+\FunctionTok{source}\NormalTok{(}\StringTok{"../functions/visualisationTools.R"}\NormalTok{)}
+\CommentTok{\#two arguments: dataset and protected word}
+\FunctionTok{visualiseProtected}\NormalTok{(priestVis,}\StringTok{"priest"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-8-1} \end{center}
+\normalsize
+
+\noindent Here you can see that some human attributes are closer than stereotype attributes, and that there is no clear reason to claim that \texttt{associated} attributes are closer than \texttt{different} or \texttt{human} attributes.
+This, again, illustrates the need of case-by-case analysis with control groups.
+
+The general idea now is that the word lists provided in different pieces of research are just samples of attributes associates with various stereotypes and should be treated as such: the uncertaintly involved and the sample sizes should have clear impact on our estimates.
+
+\hypertarget{bayesian-model-structure-and-assumptions}{%
+\section{Bayesian model structure and assumptions}\label{bayesian-model-structure-and-assumptions}}
+
+We will now think of cosine distance as the output variable, and
+will build a few bayesian models to compare.
+First, we just build a baseline model which estimates cosine distance to the attributes separately for each protected word.
+The underlying idea is that different protected words migh in general have different relations to all the attributes and this relations should be our point of departure.
+
+Here is the intution behind the mathematical Bayesian model involved.
+Our outcome variable is \texttt{cosine\ difference}, which we take to me normally distributed around the predicted mean for a given protected word (that is, we assume the residuals are normally distributed).
+The simplest model specification is:
+
+\begin{align}
+cosineDistance_i  & \sim dnorm(\mu_i, \sigma) \\
+\mu_i & = m_{pw} \\
+m_{pw} & ~ dnorm(1,.5) \\
+\sigma &\sim  dcauchy(0,1)
+\end{align}
+
+That is, we assume the estimated means might be different for diferent protected words and our prior for the mean and the overal standard deviation are normal with mean 1 and sd=.5 and half-cauchy with parameters \texttt{0,1}.
+Further on we'll also estimate additional impact the connection type may have.
+For this impact we take a slightly skeptical prior centered around 0 distributed normally with sd = 1.
+These are fairly weak and slightly skeptical regularizing priors, which can be illustrated as follows:
+
+\vspace{2mm}
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/priorsVis-1} \end{center}
+
+\hypertarget{choosing-predictors}{%
+\section{Choosing predictors}\label{choosing-predictors}}
+
+Now we can define and compile the baseline model.
+Its parameters will have a posterior distribution obtained using either Hamiltionian Monte Carlo methods (STAN) available through the \texttt{rethinking} package.
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{library}\NormalTok{(rethinking)}
+\FunctionTok{options}\NormalTok{(}\AttributeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\ConstantTok{TRUE}\NormalTok{ )}
+\NormalTok{religionBaseline }\OtherTok{\textless{}{-}} \FunctionTok{ulam}\NormalTok{(}
+  \FunctionTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu }\OtherTok{\textless{}{-}}\NormalTok{ m[pw],}
+\NormalTok{    m[pw] }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\NormalTok{  ),}
+  \AttributeTok{data =}\NormalTok{ religion,}
+  \AttributeTok{chains=}\DecValTok{2}\NormalTok{ , }\AttributeTok{iter=}\DecValTok{4000}\NormalTok{ , }\AttributeTok{warmup=}\DecValTok{1000}\NormalTok{,}
+  \AttributeTok{start=} \FunctionTok{list}\NormalTok{(}\AttributeTok{mu =} \DecValTok{1}\NormalTok{, }\AttributeTok{co =} \DecValTok{0}\NormalTok{, }\AttributeTok{sigma=}\NormalTok{ .}\DecValTok{3}\NormalTok{),}
+  \AttributeTok{log\_lik =} \ConstantTok{TRUE}\NormalTok{, }\AttributeTok{cores=}\DecValTok{4}
+\NormalTok{)}
+\CommentTok{\#saving}
+\CommentTok{\#saveRDS(religionBaseline, }
+\CommentTok{\#file = "cosineAnalysis/models/religionBaseline.rds")}
+\end{Highlighting}
+\end{Shaded}
+
+The only reason we need it is the evaluation of connection as a predictor.
+Does including it in o the model improve the situation?
+To investigate this, let's now build a model according to the following specification:
+
+\begin{align}
+cosineDistance_i  & \sim dnorm(\mu_i, \sigma) \\
+\mu_i & = m_{pw} + co_{con}\\
+m_{pw} & ~ dnorm(1,.5) \\
+co_{con} & ~ dnorm(0,1) \\
+\sigma &\sim  dcauchy(0,1)
+\end{align}
+
+\noindent The idea now is that each connection type comes with its own coefficient \(co\) that has impact on mean distances for protected words taken separately.
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{library}\NormalTok{(rethinking)}
+\FunctionTok{options}\NormalTok{(}\AttributeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\ConstantTok{TRUE}\NormalTok{ )}
+\NormalTok{religionCoefs }\OtherTok{\textless{}{-}} \FunctionTok{ulam}\NormalTok{(}
+  \FunctionTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu }\OtherTok{\textless{}{-}}\NormalTok{ m[pw] }\SpecialCharTok{+}\NormalTok{ co[con],}
+\NormalTok{    m[pw] }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    co[con] }\SpecialCharTok{\textasciitilde{}}\FunctionTok{dnorm}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\NormalTok{  ),}
+  \AttributeTok{data =}\NormalTok{ religion,}
+  \AttributeTok{chains=}\DecValTok{2}\NormalTok{ , }\AttributeTok{iter=}\DecValTok{8000}\NormalTok{ , }\AttributeTok{warmup=}\DecValTok{1000}\NormalTok{, }
+  \AttributeTok{log\_lik =} \ConstantTok{TRUE}
+\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+\noindent First, let's see if this model is really better in terms of the Widely Acceptable Information Criterion (WAIC):
+
+\vspace{1mm}
+\footnotesize
+
+\begin{verbatim}
+##                   WAIC SE dWAIC dSE pWAIC weight
+## religionCoefs    -2328 93     0  NA    20      1
+## religionBaseline -2283 95    45  17    16      0
+\end{verbatim}
+
+\normalsize
+
+Clearly, it should be given weight 1 as compared to the baseline model. So far, we've learned that the connection type actually has predictive value. Let's take a look at the coefficient estimates:
+
+\vspace{1mm}
+\footnotesize
+
+\begin{verbatim}
+##              mean        sd       5.5%      94.5%    n_eff    Rhat4
+## co[1] -0.14956420 0.1151675 -0.3261650 0.03741930 294.2033 1.001449
+## co[2] -0.09880543 0.1145024 -0.2736985 0.08813271 291.5044 1.001564
+## co[3] -0.07282752 0.1133894 -0.2447778 0.11158986 287.7820 1.001627
+## co[4] -0.03103179 0.1131420 -0.2034442 0.15268770 286.8283 1.001606
+\end{verbatim}
+
+\normalsize
+
+\hypertarget{dataset-level-coefficients}{%
+\section{Dataset-level coefficients}\label{dataset-level-coefficients}}
+
+\noindent Let's plot them together with their highest posterior density invervals, for the three topic groups.
+
+\begin{center}
+\begin{figure}[!htb]\centering
+   \begin{minipage}{0.55\textwidth}
+  \includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
+   \end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/religionGoogleCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/genderCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/genderGoogleCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/raceCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/raceGoogleCoeffs.jpeg}
+   \end{minipage}
+\end{figure}
+
+\end{center}
+
+\noindent What should strike us is that while the mean estimates of the coefficients indeed do differ a bit, usually the highest posterior density invervals all include zero, and so we do not have strong reasons to say that, say, as far as the whole religion dataset is involved, being associated indeed is connected with lower cosine distance.
+A second striking observation is that the estimated impact for associated stereotypes is quite often not too different from the estimated impact of attributes associated with different stereotypes, and both are sometimes not too far from the estimated impact for simply human attributes.
+In general, once the uncertainty involved is taken seriously by using control groups and statistical uncertainty estimation that does not dispose of pointwise data, the picture which focuses only on differences between means of means is too simplistic.
+
+But this doesn't mean important differences for some protected words are not there.
+For one thing, if you start with a word list that is very uneven, the actually not so bad status of some of the protected words might mask a pretty bad situation in which some other protected words are.
+For comparison, let's see what a model focused on words related to islam tells us.
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#this is how we build the model}
+\NormalTok{religion }\OtherTok{\textless{}{-}} \FunctionTok{read.csv}\NormalTok{(}\StringTok{"cosineAnalysis/datasets/religionReddit.csv"}\NormalTok{)[}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{]}
+\FunctionTok{colnames}\NormalTok{(religion) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
+                        \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
+\FunctionTok{levels}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{wordClass) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
+\NormalTok{muslimWords }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"imam"}\NormalTok{,}\StringTok{"islam"}\NormalTok{,}\StringTok{"mosque"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"quran"}\NormalTok{)}
+\NormalTok{muslim }\OtherTok{\textless{}{-}}\NormalTok{ religion }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(protectedWord }\SpecialCharTok{\%in\%}\NormalTok{ muslimWords)}
+\NormalTok{muslim}\SpecialCharTok{$}\NormalTok{protectedWord }\OtherTok{\textless{}{-}} \FunctionTok{droplevels}\NormalTok{(muslim}\SpecialCharTok{$}\NormalTok{protectedWord)}
+\NormalTok{muslim}\SpecialCharTok{$}\NormalTok{pw }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(muslim}\SpecialCharTok{$}\NormalTok{protectedWord)}
+\NormalTok{muslim}\SpecialCharTok{$}\NormalTok{con }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(muslim}\SpecialCharTok{$}\NormalTok{connection)}
+\NormalTok{muslim}\SpecialCharTok{$}\NormalTok{pwFactor }\OtherTok{\textless{}{-}} \FunctionTok{factor}\NormalTok{(}\FunctionTok{paste0}\NormalTok{(muslim}\SpecialCharTok{$}\NormalTok{protectedWord, muslim}\SpecialCharTok{$}\NormalTok{connection))}
+\NormalTok{muslim}\SpecialCharTok{$}\NormalTok{pwIndex }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(muslim}\SpecialCharTok{$}\NormalTok{pwFactor)}
+
+\NormalTok{islamCoefs }\OtherTok{\textless{}{-}} \FunctionTok{ulam}\NormalTok{(}
+  \FunctionTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu }\OtherTok{\textless{}{-}}\NormalTok{ m[pw] }\SpecialCharTok{+}\NormalTok{ co[con],}
+\NormalTok{    m[pw] }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    co[con] }\SpecialCharTok{\textasciitilde{}}\FunctionTok{dnorm}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\NormalTok{  ),}
+  \AttributeTok{data =}\NormalTok{ muslim,}
+  \AttributeTok{chains=}\DecValTok{2}\NormalTok{ , }\AttributeTok{iter=}\DecValTok{10000}\NormalTok{ , }\AttributeTok{warmup=}\DecValTok{1000}\NormalTok{, }\AttributeTok{cores =} \DecValTok{4}\NormalTok{,}
+  \AttributeTok{log\_lik =} \ConstantTok{TRUE}
+\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+Let's take a look at the coefficients:
+
+\vspace{1mm}
+\footnotesize
+
+\begin{verbatim}
+##                mean        sd       5.5%      94.5%    n_eff    Rhat4
+## co[1] -0.1979930035 0.1708785 -0.4682696 0.07634977 1789.894 1.003445
+## co[2] -0.0334215769 0.1687587 -0.3021954 0.23720938 1738.720 1.003575
+## co[3] -0.0192492753 0.1675754 -0.2840596 0.24860974 1732.907 1.003755
+## co[4] -0.0003911363 0.1670610 -0.2661047 0.26815172 1723.758 1.003837
+\end{verbatim}
+
+\normalsize
+
+\begin{center}
+\includegraphics[width=8cm]{../images/islamCoeffs.jpeg}
+\end{center}
+
+\normalsize
+
+While muslim words were unusual in the sense that the disparity between associated attributes and others is stronger, the evidence is still not conclusive.
+This is because the variation even within islam-related words is large enough (and sample sizes sufficiently small) for all the highest posterior density invervals to still include zeros.
+
+So, it seems, taking a closer look does seem to make a difference. The question is, what happens if we do take a close look at the level of protected words?
+
+\hypertarget{protected-word-level-analysis}{%
+\chapter{Protected-word level analysis}\label{protected-word-level-analysis}}
+
+\hypertarget{model-structure-and-assumptions}{%
+\section{Model structure and assumptions}\label{model-structure-and-assumptions}}
+
+Let's turn then to data analysis that takes a look at protected words separately.
+This time for each combination of a protected word and a connection status we will have a separate mean cosine distance estimate, each coming with its own highest posterior density interval.
+This means we will use indices that are result from all such combinations (and then we will split them up in the model precis to build visualisation, feel free to look at the \texttt{visualiseStats.R} script for details).
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{options}\NormalTok{(}\AttributeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\ConstantTok{TRUE}\NormalTok{ ) }\CommentTok{\#removes install pop{-}up request}
+\NormalTok{religion}\SpecialCharTok{$}\NormalTok{pwFactor }\OtherTok{\textless{}{-}} \FunctionTok{factor}\NormalTok{(}\FunctionTok{paste0}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{protectedWord, }\StringTok{"{-}"}\NormalTok{, religion}\SpecialCharTok{$}\NormalTok{connection))}
+\NormalTok{religion}\SpecialCharTok{$}\NormalTok{pwIndex }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{pwFactor)}
+
+\NormalTok{religionSeparate }\OtherTok{\textless{}{-}} \FunctionTok{ulam}\NormalTok{(}
+  \FunctionTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu }\OtherTok{\textless{}{-}}\NormalTok{ c[pwIndex],}
+\NormalTok{    c[pwIndex] }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\NormalTok{  ),}
+  \AttributeTok{data =}\NormalTok{ religion,}
+  \AttributeTok{chains=}\DecValTok{2}\NormalTok{ , }\AttributeTok{iter=}\DecValTok{10000}\NormalTok{ , }\AttributeTok{warmup=}\DecValTok{1000}\NormalTok{,}
+  \AttributeTok{start=}\FunctionTok{list}\NormalTok{(}\AttributeTok{no =} \DecValTok{1}\NormalTok{, }\AttributeTok{a =} \DecValTok{0}\NormalTok{, }\AttributeTok{d =} \DecValTok{0}\NormalTok{, }\AttributeTok{sigma=}\NormalTok{ .}\DecValTok{3}\NormalTok{), }\AttributeTok{log\_lik =}  \ConstantTok{TRUE}
+\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+\noindent Let's see if the individualized model does better than the previous models in light of WAIC which does add penalty for the number of parameters.
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{compareBaselineCoefsSeparate}\OtherTok{\textless{}{-}} \FunctionTok{readRDS}\NormalTok{(}\StringTok{"../datasets/compareBaselineCoefsSeparate.rds"}\NormalTok{)}
+\NormalTok{compareBaselineCoefsSeparate}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+##                   WAIC SE dWAIC dSE pWAIC weight
+## religionSeparate -2400 93     0  NA    60      1
+## religionCoefs    -2328 93    72  29    20      0
+## religionBaseline -2283 95   117  37    16      0
+\end{verbatim}
+
+\normalsize
+
+\hypertarget{protected-classes-in-reddit-and-google-embeddings}{%
+\section{Protected classes in Reddit and Google embeddings}\label{protected-classes-in-reddit-and-google-embeddings}}
+
+It seems that we do want to prefer this model, despite its relative complication.
+Now, what does it tell us about the protected words?
+Let's visualise the predicted means together with 89\% highest posterior density intervals.
+
+\includegraphics[width=14cm]{../images/visReligionReddit.png}
+
+Before we move on, let's perform analogous analyses for the remaining types of supposed bias: gender and race (the model building is analogous).
+
+\includegraphics[width=14cm]{../images/visGenderReddit.png}
+
+\includegraphics[width=14cm]{../images/visRaceReddit.png}
+
+\todo{Explain where you took the embedding from, cite publications first announcing it.}
+
+\includegraphics[width=14cm]{../images/visReligionGoogle.png}
+
+\includegraphics[width=14cm]{../images/visGenderGoogle.png}
+
+\includegraphics[width=14cm]{../images/visRaceGoogle.png}
+
+\hypertarget{the-role-of-debiasing}{%
+\chapter{The role of debiasing}\label{the-role-of-debiasing}}
+
+\todo{describe which debiasing you used and how it was advertised in the paper}
+
+\hypertarget{dataset-level-coefficients-after-debiasing}{%
+\section{Dataset-level coefficients after debiasing}\label{dataset-level-coefficients-after-debiasing}}
+
+First, let's look at coefficient estimated for the whole datasets, as compared to their estimation prior to debiasing:
+
+\begin{center}
+\begin{figure}[!htb]\centering
+   \begin{minipage}{0.55\textwidth}
+  \includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
+   \end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/debiasedReligionRedditCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/genderCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/debiasedGenderRedditCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/raceCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/debiasedRaceRedditCoeffs.jpeg}
+   \end{minipage}
+\end{figure}
+
+\end{center}
+
+\todo{comment on this}
+
+Some points to discuss:
+
+\begin{itemize}
+\item
+  humans in religion got closer (should they?)
+\item
+  change in Gender is really minor, still zero out of HPDI range, note how different is very similar
+\item
+  small improvement in race, at the price of moving neutral terms closer
+\end{itemize}
+
+\hypertarget{protected-classes-after-debiasing}{%
+\section{Protected classes after debiasing}\label{protected-classes-after-debiasing}}
+
+Now, perhaps, the effects of debiasing will be better appreciated if we look at the level of protected words.
+After all, the hope is, the situation of extremely ill-positioned protected words have improved?
+
+\includegraphics[width=14cm]{../images/visDebReligionReddit.png}
+
+\includegraphics[width=14cm]{../images/visDebGenderReddit.png}
+
+\includegraphics[width=14cm]{../images/visDebRaceReddit.png}
+
+\hypertarget{refs}{}
+\begin{CSLReferences}{1}{0}
+\leavevmode\hypertarget{ref-Bolukbasi2016Man}{}%
+Bolukbasi, T., Chang, K.-W., Zou, J.
+Y., Saligrama, V., \& Kalai, A.
+(2016).
+Man is to computer programmer as woman is to homemaker?
+Debiasing word embeddings.
+\emph{CoRR}, \emph{abs/1607.06520}.
+Retrieved from \url{http://arxiv.org/abs/1607.06520}
+
+\leavevmode\hypertarget{ref-Caliskan2017Semantics}{}%
+Islam, A.
+C., Bryson, J.
+J., \& Narayanan, A.
+(2016).
+Semantics derived automatically from language corpora necessarily contain human biases.
+\emph{CoRR}, \emph{abs/1608.07187}.
+Retrieved from \url{http://arxiv.org/abs/1608.07187}
+
+\leavevmode\hypertarget{ref-manzini2019black}{}%
+Manzini, T., Lim, Y.
+C., Tsvetkov, Y., \& Black, A.
+W.
+(2019).
+Black is to criminal as caucasian is to police: Detecting and removing multiclass bias in word embeddings.
+Retrieved from \url{http://arxiv.org/abs/1904.04047}
+
+\leavevmode\hypertarget{ref-Mehrabi2019Survey}{}%
+Mehrabi, N., Morstatter, F., Saxena, N., Lerman, K., \& Galstyan, A.
+(2019).
+A survey on bias and fairness in machine learning.
+\emph{CoRR}, \emph{abs/1908.09635}.
+Retrieved from \url{http://arxiv.org/abs/1908.09635}
+
+\leavevmode\hypertarget{ref-Nissim2019Fair}{}%
+Nissim, M., Noord, R.
+van, \& Goot, R.
+van der.
+(2019).
+Fair is better than sensational: Man is to doctor as woman is to doctor.
+\emph{CoRR}, \emph{abs/1905.09866}.
+Retrieved from \url{http://arxiv.org/abs/1905.09866}
+
+\end{CSLReferences}
+
+\end{document}

### 2021-06-10 thesis/_book/_main.tex
fixing table

--- 2021-06-09 thesis/_book/_main.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-%
-\documentclass[
-  12pt,
-]{book}
-\usepackage{amsmath,amssymb}
+\documentclass[12pt,]{book}
 \usepackage{lmodern}
 \usepackage{setspace}
+\setstretch{1.5}
+\usepackage{amssymb,amsmath}
 \usepackage{ifxetex,ifluatex}
+\usepackage{fixltx2e} % provides \textsubscript
 \ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
@@
   \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\else % if luatex or xelatex
+  \ifxetex
+    \usepackage{mathspec}
+  \else
+    \usepackage{fontspec}
   \fi
-% Use upquote if available, for straight quotes in verbatim environments
+  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
+\fi
+% use upquote if available, for straight quotes in verbatim environments
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
+% use microtype if available
+\IfFileExists{microtype.sty}{%
 \usepackage[]{microtype}
@@
 }{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
+\usepackage[unicode=true]{hyperref}
 \hypersetup{
@@
             pdfauthor={Alicja Dobrzeniecka},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage[left=3.5cm, right=2.5cm, top=2.5cm, bottom=2.5cm, asymmetric, includeheadfoot]{geometry}
-\usepackage{longtable,booktabs,array}
-\usepackage{calc} % for calculating minipage widths
-% Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-\makeatother
-% Allow footnotes in longtable head/foot
-\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
-\makesavenoteenv{longtable}
-\usepackage{graphicx}
+            pdfborder={0 0 0},
+            breaklinks=true}
+\urlstyle{same}  % don't use monospace font for urls
+\usepackage[left=3.5cm, right=2.5cm, top=2.5cm, bottom=2.5cm, asymmetric,
+includeheadfoot]{geometry}
+\usepackage{color}
+\usepackage{fancyvrb}
+\newcommand{\VerbBar}{|}
+\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+% Add ',fontsize=\small' for more characters per line
+\usepackage{framed}
+\definecolor{shadecolor}{RGB}{248,248,248}
+\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\ImportTok}[1]{#1}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
+\newcommand{\BuiltInTok}[1]{#1}
+\newcommand{\ExtensionTok}[1]{#1}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
+\newcommand{\RegionMarkerTok}[1]{#1}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
+\newcommand{\NormalTok}[1]{#1}
+\usepackage{longtable,booktabs}
+% Fix footnotes in tables (requires footnote package)
+\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{long table}}{}
+\usepackage{graphicx,grffile}
 \makeatletter
@@
 \setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
+\IfFileExists{parskip.sty}{%
+\usepackage{parskip}
+}{% else
+\setlength{\parindent}{0pt}
+\setlength{\parskip}{6pt plus 2pt minus 1pt}
+}
 \setlength{\emergencystretch}{3em}  % prevent overfull lines
@@
 \setcounter{secnumdepth}{5}
+% Redefines (sub)paragraphs to behave more like sections
+\ifx\paragraph\undefined\else
+\let\oldparagraph\paragraph
+\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+\let\oldsubparagraph\subparagraph
+\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
+
+% set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
+
 \usepackage{todonotes}
-\ifluatex
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
-  % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
-\title{Conceptual and methodological problems with bias detection and avoidance in natural language processing}
+\title{Conceptual and methodological problems with bias detection and avoidance
+in natural language processing}
 \author{Alicja Dobrzeniecka}
-\date{2021-06-09}
+\date{2021-06-10}
 
@@
 }
-\setstretch{1.5}
-\hypertarget{introduction}{%
-\chapter{Introduction}\label{introduction}}
+\chapter{Introduction}\label{introduction}
 
-Natural language processing (NLP) is a subfield of computer science that processes and analyzes language in text and speech
-with the use of modern programming methods. It has practical applications in everyday life as it concerns tasks such as email filters,
-smart assistants, search results, language translations, text analytics and so on. Models used to accomplish these tasks need a lot of
-data to learn from. This data originates from humans activities and historical recordings such as texts, messages or speeches. It turns out
-that in the learning process these models can learn implicit biases that reflect harmful stereotypical thinking still present in modern societies.
-One can find methods that aim at identifying and measuring hidden biases and/or try to remove them by modifying the models.
-There are many different types of models in NLP depending on a task that they are supposed to solve.
-However, all of them need as an input words represented by means of numbers and this is accomplished with word embedding models.
-The models usually assign the values based on the context in which the words appear.
-It means that the input data can have enormous influence on the outcome.
-The biases seem to have their primary source in the way the words are assigned the numerical values.
+Natural language processing (NLP) is a subfield of computer science that
+processes and analyzes language in text and speech with the use of
+modern programming methods. It has practical applications in everyday
+life as it concerns tasks such as email filters, smart assistants,
+search results, language translations, text analytics and so on. Models
+used to accomplish these tasks need a lot of data to learn from. This
+data originates from humans activities and historical recordings such as
+texts, messages or speeches. It turns out that in the learning process
+these models can learn implicit biases that reflect harmful
+stereotypical thinking still present in modern societies. One can find
+methods that aim at identifying and measuring hidden biases and/or try
+to remove them by modifying the models. There are many different types
+of models in NLP depending on a task that they are supposed to solve.
+However, all of them need as an input words represented by means of
+numbers and this is accomplished with word embedding models. The models
+usually assign the values based on the context in which the words
+appear. It means that the input data can have enormous influence on the
+outcome. The biases seem to have their primary source in the way the
+words are assigned the numerical values.
 
-There is considerable amount of literature available on the topic of bias detection and mitigation in NLP models.
-\protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) focuses on gender biases that may be observable while investigating the representation of job occupations and gender in terms of their assigned numerical
-values.
-The authors apply cosine similarity measurement to investigate the phenomenon where (the vectors corresponding to) words related to jobs that are stereotypically associated with a given gender are in fact in the model situated closer to this gender.
-They also use analogy tasks to evaluate if the bias is present in the word embedding model.
-They check analogies by comparing pairs of word vectors, for example they search for the word complementing the puzzle: man is to doctor as woman is to \ldots?
-First they subtract word ``man'' from word ``woman'' and then
-they search for the ranked list of other words pairs that have similar vectors' difference.
-They also include in the formula a threshold to ensure that the resulting pairs could not be randomly picked.
+There is considerable amount of literature available on the topic of
+bias detection and mitigation in NLP models. Bolukbasi, Chang, Zou,
+Saligrama, \& Kalai (\protect\hyperlink{ref-Bolukbasi2016Man}{2016})
+focuses on gender biases that may be observable while investigating the
+representation of job occupations and gender in terms of their assigned
+numerical values. The authors apply cosine similarity measurement to
+investigate the phenomenon where (the vectors corresponding to) words
+related to jobs that are stereotypically associated with a given gender
+are in fact in the model situated closer to this gender. They also use
+analogy tasks to evaluate if the bias is present in the word embedding
+model. They check analogies by comparing pairs of word vectors, for
+example they search for the word complementing the puzzle: man is to
+doctor as woman is to \ldots{}? First they subtract word ``man'' from
+word ``woman'' and then they search for the ranked list of other words
+pairs that have similar vectors' difference. They also include in the
+formula a threshold to ensure that the resulting pairs could not be
+randomly picked.
 
-However, as in \protect\hyperlink{ref-Nissim2019Fair}{Nissim, Noord, \& Goot} (\protect\hyperlink{ref-Nissim2019Fair}{2019}) it is pointed out, there are some limitations of this approach.
-According to the authors in practice most of analogies implementations do not return any input words.
-This means that it does not make sense to expect the algorithm to return the same profession for both woman and man.
-Therefore this method seems to be limited in terms of bias detection.
-The other problems regard for example the
-choice of pairs and words that are used to detect the presence of discrimination as it is often subjective and without proper justification.
-Additionally the choice of parameter set in \protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) formula to ensure that word pairs are not picked by random, is also not justified and changing it drastically
-influences the results.
+However, as in Nissim, Noord, \& Goot
+(\protect\hyperlink{ref-Nissim2019Fair}{2019}) it is pointed out, there
+are some limitations of this approach. According to the authors in
+practice most of analogies implementations do not return any input
+words. This means that it does not make sense to expect the algorithm to
+return the same profession for both woman and man. Therefore this method
+seems to be limited in terms of bias detection. The other problems
+regard for example the choice of pairs and words that are used to detect
+the presence of discrimination as it is often subjective and without
+proper justification. Additionally the choice of parameter set in
+Bolukbasi et al. (\protect\hyperlink{ref-Bolukbasi2016Man}{2016})
+formula to ensure that word pairs are not picked by random, is also not
+justified and changing it drastically influences the results.
 
-\protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) touches upon the topic of biases regarding race and gender.
-They apply knowledge from well-known psychological studies such as Implicit Association Test to research the relation between human stereotypical thinking and model learnt biases to
-discover close relationship between these two. For the evaluation they use Word Embedding Association Test (WEAT) and the Word Embedding Factual Association Test (WEFAT).
+Islam, Bryson, \& Narayanan
+(\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) touches upon the
+topic of biases regarding race and gender. They apply knowledge from
+well-known psychological studies such as Implicit Association Test to
+research the relation between human stereotypical thinking and model
+learnt biases to discover close relationship between these two. For the
+evaluation they use Word Embedding Association Test (WEAT) and the Word
+Embedding Factual Association Test (WEFAT).
 
-\protect\hyperlink{ref-manzini2019black}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-manzini2019black}{2019}) proposes a novel way of using cosine similarity to obtain the information on assumed resemblance between words.
-They investigate
-an approach that enables them to measure the bias for a class (like gender, religion, race) and express the final result with a single metric.
+Manzini, Lim, Tsvetkov, \& Black
+(\protect\hyperlink{ref-manzini2019black}{2019}) proposes a novel way of
+using cosine similarity to obtain the information on assumed resemblance
+between words. They investigate an approach that enables them to measure
+the bias for a class (like gender, religion, race) and express the final
+result with a single metric.
 
-It is worth noticing the general distinction of biases mentioned in \protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}).
-They refer to the publication concerning Implicit Association Test (Greenwald et al., 1998) that measures the strength of associations between concepts or stereotypes by calculating the time of reaction for special tasks.
-It is worth noticing that humans naturally exhibit some biases and that they do not always cause social concern.
-One can imagine the intuitive associations between for example insects and flowers, and the feelings of pleasantness or unpleasantness.
-In general, people would rather associate flowers with feeling pleasant than insects, and this preference could be named a bias or
-prejudice in some direction. However, this type of preference does not cause an uproar and is a rather morally neutral case. Unfortunately there are other
-biases and prejudices that directly influence the quality of other people's lives and therefore they should be taken care of.
+It is worth noticing the general distinction of biases mentioned in
+Islam et al. (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}). They
+refer to the publication concerning Implicit Association Test (Greenwald
+et al., 1998) that measures the strength of associations between
+concepts or stereotypes by calculating the time of reaction for special
+tasks. It is worth noticing that humans naturally exhibit some biases
+and that they do not always cause social concern. One can imagine the
+intuitive associations between for example insects and flowers, and the
+feelings of pleasantness or unpleasantness. In general, people would
+rather associate flowers with feeling pleasant than insects, and this
+preference could be named a bias or prejudice in some direction.
+However, this type of preference does not cause an uproar and is a
+rather morally neutral case. Unfortunately, there are other biases and
+prejudices that directly influence the quality of other people's lives
+and therefore they should be taken care of.
 
-One can find a bunch of various definitions trying to capture what bias and fairness actually are.
-With the choice of the definition, implications into the real-life applications may change as well as it was pointed out in \protect\hyperlink{ref-Mehrabi2019Survey}{Mehrabi, Morstatter, Saxena, Lerman, \& Galstyan} (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}).
-They mark out that there exist different types of biases, the list if long but among others there are historical bias, representation bias, measurement bias.
-This indicates how complex the process itself is.
-Without the proper understanding and awareness of the problem, people are prone to unconsciously sustain this phenomenon.
+One can find various definitions trying to capture what bias and
+fairness actually are. With the choice of the definition, implications
+into the real-life applications may change as well. Mehrabi, Morstatter,
+Saxena, Lerman, \& Galstyan
+(\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) mark out that there
+exist different types of biases such as historical bias, representation
+bias, measurement bias (the list is long). This indicates how complex
+the issue of bias is. Without the proper understanding and awareness of
+the problem, people are prone to unconsciously sustain the bias
+existence.
 
-In the article one can also find the distinction on different types of discrimination, some of them will be shortly described.
-It is worth first mentioning that protected attributes are those qualities, traits or characteristics that one cannot, according to the law, discriminated against.
-Direct discrimination refers to the situation when protected attributes of individuals explicitly result in non-favorable outcomes toward them.
-In contrast in indirect discrimination individuals appear to be treated equally but anyway they end up being treated unjustly due to the hidden effects of biases towards their protected attributes.
-Systemic discrimination takes place when policies, customs or behaviors that result from certain culture or organizational structure lead to discrimination against some groups of people.
-Finally, very common statistical discrimination refers to using
-average group statistics to judge person belonging to the group.
+Mehrabi et al. (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) also
+distinguish different types of discrimination, some of them will be
+briefly described. By protected attributes we mean those qualities,
+traits or characteristics that one cannot legally discriminate against.
+Direct discrimination occurs when protected attributes of individuals
+explicitly result in non-favorable outcomes toward them. In contrast in
+indirect discrimination individuals appear to be treated equally but
+anyway they end up being treated unjustly due to the hidden effects of
+biases towards their protected attributes. Systemic discrimination takes
+place when policies, customs or behaviors that result from certain
+culture or organizational structure lead to discrimination against some
+groups of people. Finally, very common statistical discrimination refers
+to using average group statistics to judge person belonging to the
+group.
 
-The topic of discrimination is entangled with another concept which is fairness. It is essential to grasp some concepts of fairness to take them into
-consideration while designing implementation of some machine learning model. In Mehrabi2019Survey one may notice that depending on the context and application different definitions may be applied.
+The topic of discrimination is entangled with another concept which is
+fairness. It is essential to grasp some concepts of fairness to take
+them into consideration while designing implementation of some machine
+learning model. In Mehrabi2019Survey one may notice that depending on
+the context and application different definitions may be applied.
 
-The most popular methods focus on comparing the similarity between words from protected groups and those that are considered to be stereotypical or harmful in some way.
-One can find in this group methods such as euclidean distance, dot product or cosine similarity.
-There are also other ways to detect the effects of biases.
-For example through the investigation of the model performance on certain tasks that validate if the model returns some values
-independently on gender or race or not.
+The most popular methods focus on comparing the similarity between words
+from protected groups and those that are considered to be stereotypical
+or harmful in some way. One can find in this group methods such as
+euclidean distance or cosine similarity (which is equivalent to dot
+product if the vectors are normalized). There are also other ways to
+detect the effects of biases. For example through the investigation of
+the model performance on certain tasks that validate if the model
+returns some values independently on gender or race or not.
 
-In the currently used methods (like cosine similarity) the values of similarity are often aggregated in a way that may lead to false conclusions. For
-example due to the averaging of values and the lack of confidence interval information.
+The currently used methods (such as cosine similarity) make the
+similarity values often aggregated in a way that may lead to hasty
+conclusions. The averaging of values and the lack of uncertainty may
+lead to the incomplete picture of the bias situation in the vocabulary.
 
-One can find a number of articles on negative real-life implications resulting from the presence of unaddressed biases in the machine learning models.
+One can find a number of articles on negative real-life implications
+resulting from the presence of unaddressed biases in the machine
+learning models.
 
-In the paper we indicate how current methods used to detect biases in natural language models are limited in terms of confidence interval.
+In the paper we indicate how current methods used to detect biases in
+natural language models are limited from the perspective of Bayesian
+analysis.
 
-Our research tries to answer the question of how to enhance the current way in which the bias detection is performed to make sure that it is
-methodologically valid.
+Our research enhances the current way in which the bias detection is
+performed to make sure that it is methodologically valid.
 
-Our hypothesis is that there can be greater understanding of data and bias implications when confidence interval and Bayesian method are applied to the
-methodology.
+The key hypothesis is that greater understanding of data and bias
+implications can be achieved when Bayesian methods are applied to issue.
 
-\hypertarget{cosine-similarity-and-bias-detection}{%
-\chapter{Cosine similarity and bias detection}\label{cosine-similarity-and-bias-detection}}
+\chapter{Cosine similarity and bias
+detection}\label{cosine-similarity-and-bias-detection}
 
-Placeholder
+\section{Word embeddings}\label{word-embeddings}
 
-\hypertarget{word-embeddings}{%
-\section{Word embeddings}\label{word-embeddings}}
+To understand what cosine similarity measurement is, one first needs to
+grasp the concept of translating words to a computer-readable form. In
+the field of natural language processsing there are two main types of
+words representation --- localist and distributed. One-hot encoding is
+an example of a method used to achieve a localist representation of
+words. Here each vector contains information only about a single data
+point, this is achieved by first mapping categorical values (words) to
+integers and then to each integers a binary vector is assigned which
+contains only 0s except for the index of the integer, which is assigned
+1. An example of a localist representation is:
 
-\hypertarget{cosine-similarity-and-distance}{%
-\section{Cosine similarity and distance}\label{cosine-similarity-and-distance}}
+\begin{longtable}[]{@{}llllll@{}}
+\toprule
+word & 1 & 2 & 3 & 4 & 5\tabularnewline
+\midrule
+\endhead
+woman & 1 & 0 & 0 & 0 & 0\tabularnewline
+man & 0 & 1 & 0 & 0 & 0\tabularnewline
+girl & 0 & 0 & 1 & 0 & 0\tabularnewline
+boy & 0 & 0 & 0 & 1 & 0\tabularnewline
+monarch & 0 & 0 & 0 & 0 & 1\tabularnewline
+\bottomrule
+\end{longtable}
 
-\hypertarget{cosine-distance-in-a-one-class-bias-detection}{%
-\section{Cosine distance in a one-class bias detection}\label{cosine-distance-in-a-one-class-bias-detection}}
+In the example above it is clear that the length of the vectors
+increases with the number of words in a vocabulary. It is not a very
+computationally efficient representation. It has other flaws as well.
+For example, it is unable to capture the resemblance between words
+appearing in similar contexts.
 
-\hypertarget{cosine-distance-in-a-multi-class-bias-detection}{%
-\section{Cosine distance in a multi-class bias detection}\label{cosine-distance-in-a-multi-class-bias-detection}}
+In contrast to the localist representation, a distributed representation
+returns vectors that contain continuous values instead of discrete 1s
+and 0s. Word embeddings are a class of various techniques that allow one
+to represent words as distributed vectors. Such learned representations
+of text have certain properties. At least prima facie, they store
+similar (or at least co-occurring) words close to each other in a vector
+space. An example of distributed representation is:
 
-\hypertarget{limitations-of-the-approach}{%
-\section{Limitations of the approach}\label{limitations-of-the-approach}}
+\begin{longtable}[]{@{}lllll@{}}
+\toprule
+word & 1 & 2 & 3 & 4\tabularnewline
+\midrule
+\endhead
+woman & 0.456 & 0.267 & 0.675 & 0.131\tabularnewline
+man & 0.451 & 0.897 & 0.472 & 0.088\tabularnewline
+girl & 0.604 & 0.262 & 0.414 & 0.706\tabularnewline
+boy & 0.279 & 0.172 & 0.475 & 0.010\tabularnewline
+monarch & 0.565 & 0.678 & 0.463 & 0.975\tabularnewline
+\bottomrule
+\end{longtable}
 
-\hypertarget{walkthrough-with-the-religion-dataset}{%
-\chapter{Walkthrough with the religion dataset}\label{walkthrough-with-the-religion-dataset}}
+One of the advantages of using a distributed representation is that one
+is able to represent an enormous number of concepts with a smaller
+number of units. It is also possible to better capture similarities as
+words of similar meanings can have similar numeric vectors.
 
-Placeholder
+The numbers occurring in such representations are not random. They are
+learned in a process that uses a very shallow neural network. There are
+various types of techniques used for learning the vectors
+representations. One of the most straightforward ones is a skip-gram
+model. Given a word the models tries to predict its neighboring words
+from the sentence. The mathematics behind the process relies on the idea
+that the prediction concerns the conditional probability of the adjacent
+words. The algorithm tries to minimize the loss function, which
+penalizes the system for discrepancy with actual co-occurrence
+frequencies in the corpus. One can choose various parameters of the
+model, such as the window size that determines how many surrounding
+words the model should predict. After preparing such a fitted model one
+takes only the learned weights from a neural network, and uses them as
+vectors in a word embeddings representation.
 
-\hypertarget{loading-and-understanding-the-dataset}{%
-\section{Loading and understanding the dataset}\label{loading-and-understanding-the-dataset}}
+Word embeddings have many applications in natural language processing.
+They are handy in document search and information retrieval. They also
+play their part in improving automatic translations. Well learned word
+representations may also contribute to the improvement of sentiment
+analysis or spam detection.
 
-\hypertarget{first-look-at-the-empirical-distributions}{%
-\section{First look at the empirical distributions}\label{first-look-at-the-empirical-distributions}}
+\section{Cosine similarity and
+distance}\label{cosine-similarity-and-distance}
 
-\hypertarget{looking-at-the-islam-related-words}{%
-\section{Looking at the islam-related words}\label{looking-at-the-islam-related-words}}
+Cosine similarity is often used as a method of finding out whether
+vector representations for two words suggest that they are similar or
+somehow connected. Cosine similarity is the cosine of the angle between
+two vectors: the result of dividing their inner product (dot product
+usually) by the product of their magnitudes.
 
-\hypertarget{bayesian-model-structure-and-assumptions}{%
-\section{Bayesian model structure and assumptions}\label{bayesian-model-structure-and-assumptions}}
+\begin{align} \tag{Sim}
+\mathsf{cosineSimilarity}(A,B) & = \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert}
+\end{align}
 
-\hypertarget{choosing-predictors}{%
-\section{Choosing predictors}\label{choosing-predictors}}
+Cosine similarity is considered a proper tool for this operation as its
+result has a clear connection to geometry and at least for a low number
+of dimensions may be easily interpreted. Using this scale, one can
+compare vector similarities in a fairly clear manner. When the vectors
+are aligned perpendicularly to each other, their similarity equals 0
+(which is the same as the cosine of 90 degrees). This tells us that the
+similarity between the vectors is small. As the angle between vectors
+decreases, cosine similarity approaches one, which stands for the
+greatest similarity.
 
-\hypertarget{dataset-level-coefficients}{%
-\section{Dataset-level coefficients}\label{dataset-level-coefficients}}
+One of the limitations of this measure is that it informs us only about
+similarities between vectors in terms of their orientation. However, it
+is often argued that in comparing words in terms of this metric, the
+magnitude of vectors may be treated as irrelevant, as the most important
+information pertains to direction.
 
-\hypertarget{model-structure-and-assumptions}{%
-\section{Model structure and assumptions}\label{model-structure-and-assumptions}}
+In what follows, it is important to distinguish between cosine
+similarity and cosine distance, defined as:
 
-\hypertarget{protected-classes-in-reddit-and-google-embeddings}{%
-\section{Protected classes in Reddit and Google embeddings}\label{protected-classes-in-reddit-and-google-embeddings}}
+\begin{align} \tag{Sim}
+\mathsf{cosineDistance}(A,B) &  = 1 - \mathsf{cosineSimilarity}(A,B)\\
+ &  = 1 - \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert} \nonumber
+\end{align}
 
-\hypertarget{dataset-level-coefficients-after-debiasing}{%
-\section{Dataset-level coefficients after debiasing}\label{dataset-level-coefficients-after-debiasing}}
+The greater the similarity between two vectors, the smaller the distance
+between them. The cosine distance ranges between 0 and 2. If the vectors
+are in an opposite direction to each other, the cosine distance is 2.
+And if the vectors are extremely similar then the cosine distance is
+very close to 0.
 
-\hypertarget{protected-classes-after-debiasing}{%
-\section{Protected classes after debiasing}\label{protected-classes-after-debiasing}}
+It is worth mentioning one more point concerning cosine similarity.
+After the vectors are normalized to have length equal to 1, inner
+product itself (often dot product) is used to measure the similarity.
+
+\section{Cosine distance in a one-class bias
+detection}\label{cosine-distance-in-a-one-class-bias-detection}
+
+Bolukbasi et al. (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) define
+similarity between words as the outcome inner product of their
+normalized vectors. They focus on examining what the geometry of word
+embedding is in regard to ``he'' and ``she'' words. In other words,
+whether the similarity between those concepts and other words reflects
+expected gender stereotypes. They test this hypothesis by investigating
+whether there is a connection between word embeddings representing
+certain professions and words referring to gender. They also evaluate
+whether automatically produced analogies between words reflect the
+stereotypes as well.
+
+A very vivid way to follow their method of arguing that bias in word
+embeddings is real is to plot the values of inner product of chosen
+words. The plot below does not originate from the original paper (it is
+from
+\url{https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings}) but
+similar visualization may be found there. Data used to create our plot
+is as follows. \newline
+
+Occupations associated with feminine:
+\textbf{"homemaker", "nurse", "receptionist", "librarian", "socialite", "hairdresser", "nanny", "bookkeeper", "stylist", "housekeeper", "interior designer", "guidance counselor"}
+\newline
+
+Occupations associated with masculine:
+\textbf{"maestro", "skipper", "protege", "philosopher", "captain", "architect", "financier", "warrior", "broadcaster", "magician", "fighter pilot", "boss"}
+\newline
+
+\vspace{1mm} \footnotesize
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-1-1} \end{center}
+
+\normalsize
+
+The points in the plot above result from the calculation of the inner
+product of a chosen vector for a profession word and a vector for a
+gender word (she or he). Inner product of two vectors expresses
+similarity between words. This assumption originates from the geometry
+and properties of a vector space.
+
+\section{Cosine distance in a multi-class bias
+detection}\label{cosine-distance-in-a-multi-class-bias-detection}
+
+Manzini et al. (\protect\hyperlink{ref-manzini2019black}{2019}) present
+a different approach towards finding similarities between classes of
+words. The authors claim that texts available online are full of direct
+or indirect human stereotypes. As a result, word embeddings are prone to
+learn and maybe amplify those biases and propagate them further into AI
+models that are used for various applications. Cosine distance is used
+in the article as a measure to first argue for the existence of
+multi-class bias and then to show how through bias mitigation techniques
+the bias may be decreased.
+
+As the code has been provided by the authors we were able to reconstruct
+their results. The main steps in the procedure are as follows. Let us go
+through an example that refers to the process of hard debiasing on
+religious attributes.
+
+The example refers to the process of hard debiasing on religious
+attributes.
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  First we load word embeddings from
+  reddit.US.txt.tok.clean.cleanedforw2v.w2v dataset
+\end{enumerate}
+
+\begin{itemize}
+\tightlist
+\item
+  The word embeddings have only 50 dimensions
+\item
+  The number of individual words from the dataset is 44895
+\end{itemize}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{1}
+\tightlist
+\item
+  The protected group the authors assume should ideally not have high
+  cosine similarity to stereotypical words. The word embeddings geometry
+  should not place this group close to harmful stereotypes, if it is to
+  be bias-free. \newline
+\end{enumerate}
+
+For instance, let's look at the religion-related words.
+
+Protected words by religion type:
+
+\textbf{"jew"       : ["judaism", "jew", "synagogue", "torah", "rabbi"]}
+\newline
+\textbf{"christian" : ["christianity", "christian", "church", "bible", "priest"]}
+\newline
+\textbf{"muslim"    : ["islam", "muslim", "mosque", "quran", "imam"]}
+\newline
+
+Stereotypical words by religion type:
+
+\textbf{"jew"       : ["greedy", "cheap", "hairy", "liberal"]} \newline
+\textbf{"christian"       : ["judgemental", "conservative", "familial"]}
+\newline
+\textbf{"muslim"       : ["violent", "terrorist", "dirty", "uneducated"]}
+\newline
+
+We have prepared a table presenting the values of cosine distance for
+each protected word with each attribute (stereotype). The part of the
+results is shown below.
+
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{religion <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
+\KeywordTok{colnames}\NormalTok{(religion) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
+                        \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
+\KeywordTok{levels}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
+\KeywordTok{head}\NormalTok{(religion)  }\OperatorTok{%>%}\StringTok{  }\KeywordTok{kable}\NormalTok{(}\DataTypeTok{format =} \StringTok{"latex"}\NormalTok{,}\DataTypeTok{booktabs=}\NormalTok{T,}
+                      \DataTypeTok{linesep =} \StringTok{""}\NormalTok{,  }\DataTypeTok{escape =} \OtherTok{FALSE}\NormalTok{, }
+                      \DataTypeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\OperatorTok{%>%}
+\StringTok{                      }\KeywordTok{kable_styling}\NormalTok{(}\DataTypeTok{latex_options=}\KeywordTok{c}\NormalTok{(}\StringTok{"scale_down"}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{table}
+
+\caption{\label{tab:religionTableHeadEarly}Head of the religion dataset.}
+\centering
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{lllrrl}
+\toprule
+protectedWord & wordToCompare & wordClass & cosineDistance & cosineSimilarity & connection\\
+\midrule
+judaism & violent & muslim & 0.7141939 & 0.2858061 & different\\
+judaism & terrorist & muslim & 0.7461333 & 0.2538667 & different\\
+judaism & dirty & muslim & 1.2002599 & -0.2002599 & different\\
+judaism & uneducated & muslim & 0.7885469 & 0.2114531 & different\\
+judaism & greedy & jewish & 1.0026172 & -0.0026172 & associated\\
+judaism & cheap & jewish & 1.2323229 & -0.2323229 & associated\\
+\bottomrule
+\end{tabular}}
+\end{table}
+
+\normalsize
+
+\pagebreak
+
+In the article there was no analysis of individual distances but the
+general look at the data through the usage of mean. The authors
+introduced a metric that tries to generalize the presence of bias
+through the classification of multi-class bias in groups of words
+connected with gender, religion or race. In the process they first take
+the mean of cosine distances between a given protected word and
+attributes assigned to each stereotype. They do not differentiate
+between stereotypes associated with a word and stereotypes associated
+with different words (in case of religion, stereotypes characteristic
+for Christianity has also cosine distance measured with for instance,
+Judaism or Islam). Then, after collecting the list of mean cosine
+distances, they average the list to obtain one final value representing
+the whole group, in this example religion, for which the final mean of
+all mean distances is equal to 0.859.
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{2}
+\tightlist
+\item
+  In the article the authors also try to remove previously defined
+  biases from word embedding.
+\end{enumerate}
+
+First they identify the bias subspace using Principal Component Analysis
+(PCA) which is a technique for dimensionality reduction. It is applied
+here to choose the subspace that contains the greatest amount of
+information. There can be many subspaces found in a given group, for
+example in terms of religion one can identify at least a few sets that
+are to grasp the concept of religion in general: \newline
+\textbf{["judaism", "christianity", "islam"]} \newline
+\textbf{["jew", "christian", "muslim"]} \newline
+\textbf{["synagogue", "church", "mosque"]} \newline
+
+The idea is to find a set that provides enough information to create
+from it a vector representing the concept of religion among words. This
+strategy is based on the idea that different dimensions of vectors
+contain different types of information and in some words in vector
+layers (subspaces) the information about religiousness is implicitly
+conveyed. In some cases this knowledge is useful but in the case of
+harmful stereotypes one does not want to have religion concept in
+stereotypical words.
+
+After finding bias subspace, they use it to modify the vector values
+individually so that their cosine distances towards certain words are
+changed. In the case of stereotypes the aim is to make the cosine
+distances larger so that the association between protected word and
+harmful stereotype is smaller.
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{3}
+\tightlist
+\item
+  In the final step there is an evaluation of the results. The cosine
+  distances are calculated again but this time using debiased
+  vocabulary. After taking the mean of all distances one final value is
+  obtained and then it is compared with the average value from the
+  beginning. If the cosine distances are on average greater than before
+  then it leads the authors to the conclusion that improvement was made.
+  As the cosine distance increases it is assumed that the association
+  between protected and stereotypical words decreases.
+\end{enumerate}
+
+\section{Limitations of the approach}\label{limitations-of-the-approach}
+
+\textbf{1. Selection of attributes}
+
+The attributes are taken from different sources, there is no principled
+justification for their choice. From our analysis it will become clear
+that the list is rather uneven.
+
+There is no mention of methodology for deciding on the number of
+attributes necessary to decide a hypothesis on the given size of
+dataset. There are however some ways to estimate how many samples we
+need to make sure that the result is significant. Our research will show
+that the numbers used are rather insufficient.
+
+\textbf{2. No control while taking the mean of cosines}
+
+The authors use the mean average cosine similarity to check on
+multi-class similarity between protected word and harmful stereotypes.
+They average the results until they obtain one final value to represent
+the mean cosine distance between protected word from a given class and
+the attributes of that class. As there is no control of the individual
+values of cosine distances, one can observe values that have cosine
+distances greater than 1 which would suggest that these compare words do
+not have significant similarity. However, as there is a lack of control,
+even such words are taken into the account for calculations.
+
+\textbf{3. The lack of uncertainty}
+
+A mean hides this issue and as there are pairs having negative and small
+similarities and there are those that have similarity equal to 0.5, the
+resulting calculation seems to be in norm. Additionally in such method
+the uncertainty is also not included which makes it even more difficult
+to give reasonable interpretations of the results. We propose the use of
+Bayesian method to obtain some understanding of the influence the
+uncertainty has on the interpretation of final results.
+
+\textbf{4. No word class distinction}
+
+In the original paper words from all three religions were checked with
+all of the stereotypes which means that there was no distinction on
+whether the stereotype is associated with given religion or with the
+other. Not all of the stereotypical words should be considered as
+harmful for all of the religions. In our analysis we distinguished
+between associated stereotypes, not associated (stereotype is valid but
+for different religion than currently checked) and none (meaning neutral
+words).
+
+\textbf{5. Interpreting the results}
+
+Assuming for a moment that the value of multi-class cosine distance is
+correct, one may question the results' interpretation. In Manzini et al.
+(\protect\hyperlink{ref-manzini2019black}{2019}) Table 2, there are
+summarized the averages of cosine distance per group (gender, race,
+religion). I would like to focus now on analyzing the values relating to
+religious biases. Here is the fragment of table that refers to that:
+
+\begin{longtable}[]{@{}ll@{}}
+\toprule
+Religion Debiasing & MAC\tabularnewline
+\midrule
+\endhead
+Biased & 0.859\tabularnewline
+Hard Debiased & 0.934\tabularnewline
+Soft Debiased (\(\lambda\) = 0.2) & 0.894\tabularnewline
+\bottomrule
+\end{longtable}
+
+MAC stand for mean average cosine similarity although in reality the
+values of cosine distance are stored there. What may attract attention
+is the fact that the value of cosine distance in ``Biased'' category is
+already quite high even before ``debiasing''. High cosine distance
+indicates low cosine similarity between values. One could think that
+average cosine similarity equal to approximately 0.141 is not
+significant enough to consider it as biased. However the authors aim to
+mitigate ``biases'' in vectors with such great distance to make it even
+larger. Methodologically there is a question on what basis this small
+similarity is still considered as a proof of bias presence.
+
+This is in general the problem of scale and the lack of universal
+intervals that could be applied to know whether the cosine similarity is
+high enough to considered given words more similar that if we chose them
+on random. It seems like similar practice could be beneficial for
+metrics aimed at measuring similarity between words.
+
+\textbf{6. The curse of dimensionality}
+
+Curse of dimensionality may take place when there is an increase in
+volume of data that results in adding extra dimensions to the Euclidean
+space. According to the article
+``\url{https://analyticsindiamag.com/curse-of-dimensionality-and-what-beginners-should-do-to-overcome-it/}''
+as the number of features increases, it may be harder and harder to
+obtain useful information from the data with the usage of available
+algorithms. One may notice that more data should contribute to greater
+amount of information but more information also means greater risk of
+noise and distractions in data. At the same time, many times modern
+solutions are adapted to smaller dimensions and their results in higher
+ones are not intuitive or may be prone to be mistaken.
+
+Using cosine similarity in high dimensions in word embeddings may also
+be prone to the curse of dimensionality. According to this article
+``\url{https://www.researchgate.net/publication/327498046_The_Curse_of_Dimensionality_Inside_Out}''
+there are reasons to consider this phenomenon when searching for word
+similarities in higher dimensions.
+
+In the article an experiment is conducted that aims at showing how the
+similarity values and variation change as the number of dimensions
+increases. The hypothesis made in the paper states that two things will
+happen as the number of dimensions increase, the first one is that
+effort required to measure cosine similarity will be greater and the
+second one is that the similarity between data will blur out and have
+less variation. In details, the authors generate random points with
+increasing number of dimensions where each dimension of a data point is
+given a value between 0 and 1. Then they pick one vector at random from
+each dimension class and calculate cosine similarity between the chosen
+vector and the rest of the data. Then they check how the variation of
+values changes as the number of dimensions increases. It seems like the
+more dimensions there are, the smaller the variance and therefore it is
+less obvious how to interpret the resulting cosine similarities. Maybe
+the scale should be adjusted to the number of dimensions and variance so
+that it still gives us sensible information about data. According to
+some researches the cosine similarity in high dimensions is not reliable
+enough to trust it as it may be the case that choosing words on random
+may result in getting similar values as when picking them consciously.
+
+\begin{figure}
+\centering
+\includegraphics{../images/curseOfDimensionality.png}
+\caption{curse of dimensionality, number of dimensions on the x axis,
+standard deviation of similarity on the y axis}
+\end{figure}
+
+\chapter{Walkthrough with the religion
+dataset}\label{walkthrough-with-the-religion-dataset}
+
+\section{Loading and understanding the
+dataset}\label{loading-and-understanding-the-dataset}
+
+We start with loading the libraries needed for the analysis.
+
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(ggplot2)}
+\KeywordTok{library}\NormalTok{(ggthemes)}
+\KeywordTok{library}\NormalTok{(rethinking)}
+\KeywordTok{library}\NormalTok{(tidyverse)}
+\KeywordTok{library}\NormalTok{(ggpubr)}
+\KeywordTok{library}\NormalTok{(kableExtra)}
+\KeywordTok{library}\NormalTok{(dplyr)}
+\KeywordTok{library}\NormalTok{(ggExtra)}
+\KeywordTok{library}\NormalTok{(cowplot)}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+We will use the choice of protected words and stereotypical predicates
+used in REF. This is a decent point of departure, not only we want to
+compare our method to that of REF, but also because this data format is
+fairly general (as contrasted, say, with a set up for binary
+stereotypes). Note also that the method we develop here can fairly
+easily be run for different stereotypization patterns. Let's start with
+explaining the method and its deployment using a dataset obtained for
+the religion-related protected words.
+
+Let's load, clean a bit and inspect the head of the religion dataset we
+prepared. In order to obtain this dataset, we calculated the cosine
+distance between each protected word and each word from both the
+bias-related attribute groups, which were used in the original study,
+and to neutral and human control attributes which we added as control
+groups. For instance, for religion, the bias-related predicates (coming
+from the original study in REF) include muslim bias attributes, jew bias
+attributes, christian bias attributes (see a list in the APPENDIX).
+
+We decided to add control groups in the form of two classes --- neutral
+words and human-related words. Without a proper control group it is
+quite hard to compare the resulting cosine distances and decide on their
+significance in bias detection. We prepared approximately
+300\todo{FIX later} more or less neutral words to double-check the
+prima-facie neutral hypothesis that their cosine similarity to the
+protected words will oscillate around 0 (that is, the distances will be
+around 1). This provides us with a more reliable point of reference.
+Moreover, we added human attributes that are associated with people in
+general to investigate whether the smaller cosine distance between
+protected words and stereotypes can result simply from the fact that the
+stereotype predicates are associated with humans. For two control
+groups, we have randomly drawn 300 words that do not express any
+property usually attributed to humans, and human attributes.
+\todo{describe once the words are selected}
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{religion <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
+\KeywordTok{colnames}\NormalTok{(religion) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
+                        \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
+\KeywordTok{library}\NormalTok{(plyr)}
+\NormalTok{religion}\OperatorTok{$}\NormalTok{wordClass <-}\StringTok{ }\KeywordTok{mapvalues}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass, }\DataTypeTok{from =} \KeywordTok{levels}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass), }\DataTypeTok{to =} \KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{))}
+\CommentTok{#levels(religion$wordClass) <- c("christian","human","jewish","muslim","neutral")}
+\KeywordTok{head}\NormalTok{(religion)  }\OperatorTok{%>%}\StringTok{  }\KeywordTok{kable}\NormalTok{(}\DataTypeTok{format =} \StringTok{"latex"}\NormalTok{,}\DataTypeTok{booktabs=}\NormalTok{T,}
+                      \DataTypeTok{linesep =} \StringTok{""}\NormalTok{,  }\DataTypeTok{escape =} \OtherTok{FALSE}\NormalTok{, }
+                      \DataTypeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\OperatorTok{%>%}\StringTok{      }\KeywordTok{kable_styling}\NormalTok{(}\DataTypeTok{latex_options=}\KeywordTok{c}\NormalTok{(}\StringTok{"scale_down"}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{table}
+
+\caption{\label{tab:religionTableHead}Head of the religion dataset.}
+\centering
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{lllrrl}
+\toprule
+protectedWord & wordToCompare & wordClass & cosineDistance & cosineSimilarity & connection\\
+\midrule
+judaism & violent & muslim & 0.7141939 & 0.2858061 & different\\
+judaism & terrorist & muslim & 0.7461333 & 0.2538667 & different\\
+judaism & dirty & muslim & 1.2002599 & -0.2002599 & different\\
+judaism & uneducated & muslim & 0.7885469 & 0.2114531 & different\\
+judaism & greedy & jewish & 1.0026172 & -0.0026172 & associated\\
+judaism & cheap & jewish & 1.2323229 & -0.2323229 & associated\\
+\bottomrule
+\end{tabular}}
+\end{table}
+
+\normalsize
+
+The \texttt{protectedWord} column contains words from a protected class
+that (in a perfect world according to the assumptions of the orignal
+study) should not be associated with harmful stereotypes.
+\texttt{wordToCompare} contains attributes, including stereotypes and
+control group words. For each row we compute the cosine distances
+between a given protected word and a given attribute word.
+\texttt{wordClass} tells us which class an attribute is supposed to be
+stereotypically associated with, that is, whether the word from
+\texttt{wordToCompare} is associated stereotypically with jews,
+christians or muslims, or whether it belongs to a control group.
+\texttt{cosineDistance} is simply a calculation of the cosine distance
+between protected word and atrribute. \texttt{cosineSimilarity} contains
+the result of substracting cosine distance from 1. \texttt{connection}
+contains information about the relation type between a protected word
+and an attribute. If the attribute is e.g.~a harmful jewish stereotype
+and the protected word is also from the judaism group, the connection
+has value \texttt{associated}. If the attribute is still stereotypically
+jewish, but the protected word comes from another religion, the
+connection is labelled as \texttt{different}. If the attribute belongs
+to a neutral group then the connection is labelled as \texttt{none} and
+if an attribute belongs to the \texttt{human} class, then the connection
+is labelled as \texttt{human}.
+
+\section{First look at the empirical
+distributions}\label{first-look-at-the-empirical-distributions}
+
+First let's take a look at the empirical distribution of distances by
+the connection type, initially ignoring the human control class for now.
+
+\vspace{1mm} \footnotesize
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-3-1} \end{center}
+
+\normalsize
+
+The first impression is that while there is a shift for associated words
+towards smaller cosine distances as compared to the neutral words,
+slightly surprisingly a slightly weaker shift in the same direction is
+visible for attributes associated with different stereotypes. Moreover,
+the empirical distributions overlap to alarge extent and the means
+grouped by connection type do not seem too far from each other. In fact,
+as there is a lot of variety in the consine distances (as we will soon
+see), abd we need to gauge the uncertaintly involved, and to look more
+carefully at individual protected words to get a better idea of how the
+cosine distance distribution changes for different attribute groups and
+different protected classes. Now, let's add the human attributes to the
+picture:
+
+\vspace{1mm} \footnotesize
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-4-1} \end{center}
+
+\normalsize
+
+\noindent Notice that the distribution for \texttt{human} (even though
+we did our best not to include in it any stereotype-related atributes)
+is left-skewed, with much overlap with \texttt{associated} and
+\texttt{different}, which illustrates the need to take being associated
+with humans as an important predictor.
+
+Our focus lies in \texttt{connection} as a predictor. Morever, later on
+we'll be interested in looking at the protected words separately, and at
+protected words split by connection. For technical reasons it is useful
+to represent these factors as integer vectors.
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{religion}\OperatorTok{$}\NormalTok{con <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(religion}\OperatorTok{$}\NormalTok{connection)}
+\NormalTok{religion}\OperatorTok{$}\NormalTok{pw <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(religion}\OperatorTok{$}\NormalTok{protectedWord)}
+\NormalTok{religion}\OperatorTok{$}\NormalTok{pwFactor <-}\StringTok{ }\KeywordTok{factor}\NormalTok{(}\KeywordTok{paste0}\NormalTok{(religion}\OperatorTok{$}\NormalTok{protectedWord, religion}\OperatorTok{$}\NormalTok{connection))}
+\NormalTok{religion}\OperatorTok{$}\NormalTok{pwIndex <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(religion}\OperatorTok{$}\NormalTok{pwFactor)}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+A short script, \texttt{cleanDataset} to make this faster, so
+equivalently:
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{source}\NormalTok{(}\StringTok{"../functions/cleanDataset.R"}\NormalTok{)}
+\NormalTok{religion <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
+\NormalTok{religion <-}\StringTok{ }\KeywordTok{cleanDataset}\NormalTok{(religion,}\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+\section{Looking at the islam-related
+words}\label{looking-at-the-islam-related-words}
+
+For now, let's focus on five protected words related to islam (``imam'',
+``islam'', ``mosque'', ``muslim'', and ``quran''). The word list
+associates with islam four stereotypical attributes (``violent'',
+``terrorist'', ``uneducated'' and ``dirty''). First, we select and plot
+the empirical distributions for these protected words.
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(tidyverse)}
+\NormalTok{muslimWords <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"imam"}\NormalTok{,}\StringTok{"islam"}\NormalTok{,}\StringTok{"mosque"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"quran"}\NormalTok{)}
+\NormalTok{muslim <-}\StringTok{ }\NormalTok{religion }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(protectedWord }\OperatorTok{%in%}\StringTok{ }\NormalTok{muslimWords)}
+\KeywordTok{ggplot}\NormalTok{(muslim, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{  cosineDistance, }\DataTypeTok{fill =}\NormalTok{ connection, }\DataTypeTok{color =}\NormalTok{ connection))}\OperatorTok{+}
+\StringTok{  }\KeywordTok{geom_density}\NormalTok{(}\DataTypeTok{alpha=}\FloatTok{0.6}\NormalTok{,}\DataTypeTok{size =} \FloatTok{.2}\NormalTok{)}\OperatorTok{+}\StringTok{ }
+\StringTok{  }\KeywordTok{scale_fill_manual}\NormalTok{(}\DataTypeTok{values =} \KeywordTok{c}\NormalTok{(}\StringTok{"orangered4"}\NormalTok{,}\StringTok{"chartreuse4"}\NormalTok{, }\StringTok{"skyblue"}\NormalTok{, }\StringTok{"gray"}\NormalTok{))}\OperatorTok{+}
+\StringTok{  }\KeywordTok{scale_x_continuous}\NormalTok{(}\DataTypeTok{breaks =} \KeywordTok{seq}\NormalTok{(}\FloatTok{0.3}\NormalTok{,}\FloatTok{1.5}\NormalTok{, }\DataTypeTok{by =} \FloatTok{0.1}\NormalTok{))}\OperatorTok{+}\KeywordTok{xlab}\NormalTok{(}\StringTok{"cosine distance"}\NormalTok{)}\OperatorTok{+}
+\StringTok{  }\KeywordTok{scale_color_manual}\NormalTok{(}\DataTypeTok{values =} \KeywordTok{c}\NormalTok{(}\StringTok{"orangered4"}\NormalTok{,}\StringTok{"chartreuse4"}\NormalTok{,}\StringTok{"skyblue"}\NormalTok{,}\StringTok{"gray"}\NormalTok{))}\OperatorTok{+}
+\StringTok{  }\KeywordTok{theme_tufte}\NormalTok{()}\OperatorTok{+}\KeywordTok{ggtitle}\NormalTok{(}\StringTok{"Empirical distribution of distances (muslim)"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-7-1} \end{center}
+
+\normalsize
+
+\noindent Once we focus on words related to islam, the associated bias
+seems to be stronger than in the whole dataset. This is a step towards
+illustrating that the distribution of bias is uneven.
+
+Now, say we want to look at a single protected word. Since the dataset
+also contains comparison multiple control neutral and human attributes,
+we randomly select only 5 from \texttt{none} and 5 from \texttt{human}
+control groups of those for the visualisation purposes.
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(tidyverse)}
+\NormalTok{muslimClass <-}\StringTok{ }\NormalTok{muslim }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(protectedWord }\OperatorTok{==}\StringTok{ "muslim"}\NormalTok{)}
+\NormalTok{neutralSample <-}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\KeywordTok{filter}\NormalTok{(muslimClass,connection }\OperatorTok{==}\StringTok{ "none"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
+\NormalTok{humanSample <-}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\KeywordTok{filter}\NormalTok{(muslimClass,connection }\OperatorTok{==}\StringTok{ "human"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
+\NormalTok{muslimVis <-}\StringTok{ }\NormalTok{muslimClass }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(connection }\OperatorTok{!=}\StringTok{ "none"} \OperatorTok{&}\StringTok{ }\NormalTok{connection }\OperatorTok{!=}\StringTok{"human"}\NormalTok{)}
+\NormalTok{muslimVis <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(muslimVis,neutralSample,humanSample)}
+
+\CommentTok{#we plug in our visualisation script}
+\KeywordTok{source}\NormalTok{(}\StringTok{"../functions/visualisationTools.R"}\NormalTok{)}
+\CommentTok{#two arguments: dataset and protected word}
+\KeywordTok{visualiseProtected}\NormalTok{(muslimVis,}\StringTok{"muslim"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/tableMuslimActive-1} \end{center}
+
+\normalsize
+
+Note that the distance between the grey point and the other points is
+proportional to cosine distance, the non-grey point size is proportional
+to cosine similarity to the protected word, and color groups by the
+connection type. So for \texttt{muslim} it seems that the stereotypes
+coming from the word list are fairly well visible. To give you some
+taste of how uneven the dataset is, compare this to what happens with
+\texttt{priest}.
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(tidyverse)}
+\NormalTok{priestClass <-}\StringTok{ }\NormalTok{religion }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(protectedWord }\OperatorTok{==}\StringTok{ "priest"}\NormalTok{)}
+\NormalTok{neutralSample <-}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\KeywordTok{filter}\NormalTok{(priestClass,connection }\OperatorTok{==}\StringTok{ "none"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
+\NormalTok{humanSample <-}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\KeywordTok{filter}\NormalTok{(priestClass,connection }\OperatorTok{==}\StringTok{ "human"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
+\NormalTok{priestVis <-}\StringTok{ }\NormalTok{priestClass }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(connection }\OperatorTok{!=}\StringTok{ "none"} \OperatorTok{&}\StringTok{ }\NormalTok{connection }\OperatorTok{!=}\StringTok{"human"}\NormalTok{)}
+\NormalTok{priestVis <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(priestVis,neutralSample,humanSample)}
+
+\CommentTok{#we plug in our visualisation script}
+\KeywordTok{source}\NormalTok{(}\StringTok{"../functions/visualisationTools.R"}\NormalTok{)}
+\CommentTok{#two arguments: dataset and protected word}
+\KeywordTok{visualiseProtected}\NormalTok{(priestVis,}\StringTok{"priest"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-8-1} \end{center}
+
+\normalsize
+
+\noindent Here you can see that some human attributes are closer than
+stereotype attributes, and that there is no clear reason to claim that
+\texttt{associated} attributes are closer than \texttt{different} or
+\texttt{human} attributes. This, again, illustrates the need of
+case-by-case analysis with control groups.
+
+The general idea now is that the word lists provided in different pieces
+of research are just samples of attributes associates with various
+stereotypes and should be treated as such: the uncertaintly involved and
+the sample sizes should have clear impact on our estimates.
+
+\section{Bayesian model structure and
+assumptions}\label{bayesian-model-structure-and-assumptions}
+
+We will now think of cosine distance as the output variable, and will
+build a few bayesian models to compare. First, we just build a baseline
+model which estimates cosine distance to the attributes separately for
+each protected word. The underlying idea is that different protected
+words migh in general have different relations to all the attributes and
+this relations should be our point of departure.
+
+Here is the intution behind the mathematical Bayesian model involved.
+Our outcome variable is \texttt{cosine\ difference}, which we take to me
+normally distributed around the predicted mean for a given protected
+word (that is, we assume the residuals are normally distributed). The
+simplest model specification is:
+
+\begin{align}
+cosineDistance_i  & \sim dnorm(\mu_i, \sigma) \\
+\mu_i & = m_{pw} \\
+m_{pw} & ~ dnorm(1,.5) \\
+\sigma &\sim  dcauchy(0,1)
+\end{align}
+
+That is, we assume the estimated means might be different for diferent
+protected words and our prior for the mean and the overal standard
+deviation are normal with mean 1 and sd=.5 and half-cauchy with
+parameters \texttt{0,1}. Further on we'll also estimate additional
+impact the connection type may have. For this impact we take a slightly
+skeptical prior centered around 0 distributed normally with sd = 1.
+These are fairly weak and slightly skeptical regularizing priors, which
+can be illustrated as follows:
+
+\vspace{2mm}
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/priorsVis-1} \end{center}
+
+\section{Choosing predictors}\label{choosing-predictors}
+
+Now we can define and compile the baseline model. Its parameters will
+have a posterior distribution obtained using either Hamiltionian Monte
+Carlo methods (STAN) available through the \texttt{rethinking} package.
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(rethinking)}
+\KeywordTok{options}\NormalTok{(}\DataTypeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\OtherTok{TRUE}\NormalTok{ )}
+\NormalTok{religionBaseline <-}\StringTok{ }\KeywordTok{ulam}\NormalTok{(}
+  \KeywordTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu <-}\StringTok{ }\NormalTok{m[pw],}
+\NormalTok{    m[pw] }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\OperatorTok{~}\StringTok{ }\KeywordTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\NormalTok{  ),}
+  \DataTypeTok{data =}\NormalTok{ religion,}
+  \DataTypeTok{chains=}\DecValTok{2}\NormalTok{ , }\DataTypeTok{iter=}\DecValTok{4000}\NormalTok{ , }\DataTypeTok{warmup=}\DecValTok{1000}\NormalTok{,}
+  \DataTypeTok{start=} \KeywordTok{list}\NormalTok{(}\DataTypeTok{mu =} \DecValTok{1}\NormalTok{, }\DataTypeTok{co =} \DecValTok{0}\NormalTok{, }\DataTypeTok{sigma=} \FloatTok{.3}\NormalTok{),}
+  \DataTypeTok{log_lik =} \OtherTok{TRUE}\NormalTok{, }\DataTypeTok{cores=}\DecValTok{4}
+\NormalTok{)}
+\CommentTok{#saving}
+\CommentTok{#saveRDS(religionBaseline, }
+\CommentTok{#file = "cosineAnalysis/models/religionBaseline.rds")}
+\end{Highlighting}
+\end{Shaded}
+
+The only reason we need it is the evaluation of connection as a
+predictor. Does including it in o the model improve the situation? To
+investigate this, let's now build a model according to the following
+specification:
+
+\begin{align}
+cosineDistance_i  & \sim dnorm(\mu_i, \sigma) \\
+\mu_i & = m_{pw} + co_{con}\\
+m_{pw} & ~ dnorm(1,.5) \\
+co_{con} & ~ dnorm(0,1) \\
+\sigma &\sim  dcauchy(0,1)
+\end{align}
+
+\noindent The idea now is that each connection type comes with its own
+coefficient \(co\) that has impact on mean distances for protected words
+taken separately.
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(rethinking)}
+\KeywordTok{options}\NormalTok{(}\DataTypeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\OtherTok{TRUE}\NormalTok{ )}
+\NormalTok{religionCoefs <-}\StringTok{ }\KeywordTok{ulam}\NormalTok{(}
+  \KeywordTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu <-}\StringTok{ }\NormalTok{m[pw] }\OperatorTok{+}\StringTok{ }\NormalTok{co[con],}
+\NormalTok{    m[pw] }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    co[con] }\OperatorTok{~}\KeywordTok{dnorm}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\OperatorTok{~}\StringTok{ }\KeywordTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\NormalTok{  ),}
+  \DataTypeTok{data =}\NormalTok{ religion,}
+  \DataTypeTok{chains=}\DecValTok{2}\NormalTok{ , }\DataTypeTok{iter=}\DecValTok{8000}\NormalTok{ , }\DataTypeTok{warmup=}\DecValTok{1000}\NormalTok{, }
+  \DataTypeTok{log_lik =} \OtherTok{TRUE}
+\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+\noindent First, let's see if this model is really better in terms of
+the Widely Acceptable Information Criterion (WAIC):
+
+\vspace{1mm} \footnotesize
+
+\begin{verbatim}
+##                   WAIC SE dWAIC dSE pWAIC weight
+## religionCoefs    -2328 93     0  NA    20      1
+## religionBaseline -2283 95    45  17    16      0
+\end{verbatim}
+
+\normalsize
+
+Clearly, it should be given weight 1 as compared to the baseline model.
+So far, we've learned that the connection type actually has predictive
+value. Let's take a look at the coefficient estimates:
+
+\vspace{1mm} \footnotesize
+
+\begin{verbatim}
+##              mean        sd       5.5%      94.5%    n_eff    Rhat4
+## co[1] -0.14956420 0.1151675 -0.3261650 0.03741930 294.2033 1.001449
+## co[2] -0.09880543 0.1145024 -0.2736985 0.08813271 291.5044 1.001564
+## co[3] -0.07282752 0.1133894 -0.2447778 0.11158986 287.7820 1.001627
+## co[4] -0.03103179 0.1131420 -0.2034442 0.15268770 286.8283 1.001606
+\end{verbatim}
+
+\normalsize
+
+\section{Dataset-level coefficients}\label{dataset-level-coefficients}
+
+\noindent Let's plot them together with their highest posterior density
+invervals, for the three topic groups.
+
+\begin{center}
+\begin{figure}[!htb]\centering
+   \begin{minipage}{0.55\textwidth}
+  \includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
+   \end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/religionGoogleCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/genderCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/genderGoogleCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/raceCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/raceGoogleCoeffs.jpeg}
+   \end{minipage}
+\end{figure}
+
+\end{center}
+
+\noindent What should strike us is that while the mean estimates of the
+coefficients indeed do differ a bit, usually the highest posterior
+density invervals all include zero, and so we do not have strong reasons
+to say that, say, as far as the whole religion dataset is involved,
+being associated indeed is connected with lower cosine distance. A
+second striking observation is that the estimated impact for associated
+stereotypes is quite often not too different from the estimated impact
+of attributes associated with different stereotypes, and both are
+sometimes not too far from the estimated impact for simply human
+attributes. In general, once the uncertainty involved is taken seriously
+by using control groups and statistical uncertainty estimation that does
+not dispose of pointwise data, the picture which focuses only on
+differences between means of means is too simplistic.
+
+But this doesn't mean important differences for some protected words are
+not there. For one thing, if you start with a word list that is very
+uneven, the actually not so bad status of some of the protected words
+might mask a pretty bad situation in which some other protected words
+are. For comparison, let's see what a model focused on words related to
+islam tells us.
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{#this is how we build the model}
+\NormalTok{religion <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"cosineAnalysis/datasets/religionReddit.csv"}\NormalTok{)[}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
+\KeywordTok{colnames}\NormalTok{(religion) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
+                        \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
+\KeywordTok{levels}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
+\NormalTok{muslimWords <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"imam"}\NormalTok{,}\StringTok{"islam"}\NormalTok{,}\StringTok{"mosque"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"quran"}\NormalTok{)}
+\NormalTok{muslim <-}\StringTok{ }\NormalTok{religion }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(protectedWord }\OperatorTok{%in%}\StringTok{ }\NormalTok{muslimWords)}
+\NormalTok{muslim}\OperatorTok{$}\NormalTok{protectedWord <-}\StringTok{ }\KeywordTok{droplevels}\NormalTok{(muslim}\OperatorTok{$}\NormalTok{protectedWord)}
+\NormalTok{muslim}\OperatorTok{$}\NormalTok{pw <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(muslim}\OperatorTok{$}\NormalTok{protectedWord)}
+\NormalTok{muslim}\OperatorTok{$}\NormalTok{con <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(muslim}\OperatorTok{$}\NormalTok{connection)}
+\NormalTok{muslim}\OperatorTok{$}\NormalTok{pwFactor <-}\StringTok{ }\KeywordTok{factor}\NormalTok{(}\KeywordTok{paste0}\NormalTok{(muslim}\OperatorTok{$}\NormalTok{protectedWord, muslim}\OperatorTok{$}\NormalTok{connection))}
+\NormalTok{muslim}\OperatorTok{$}\NormalTok{pwIndex <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(muslim}\OperatorTok{$}\NormalTok{pwFactor)}
+
+\NormalTok{islamCoefs <-}\StringTok{ }\KeywordTok{ulam}\NormalTok{(}
+  \KeywordTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu <-}\StringTok{ }\NormalTok{m[pw] }\OperatorTok{+}\StringTok{ }\NormalTok{co[con],}
+\NormalTok{    m[pw] }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    co[con] }\OperatorTok{~}\KeywordTok{dnorm}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\OperatorTok{~}\StringTok{ }\KeywordTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\NormalTok{  ),}
+  \DataTypeTok{data =}\NormalTok{ muslim,}
+  \DataTypeTok{chains=}\DecValTok{2}\NormalTok{ , }\DataTypeTok{iter=}\DecValTok{10000}\NormalTok{ , }\DataTypeTok{warmup=}\DecValTok{1000}\NormalTok{, }\DataTypeTok{cores =} \DecValTok{4}\NormalTok{,}
+  \DataTypeTok{log_lik =} \OtherTok{TRUE}
+\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+Let's take a look at the coefficients:
+
+\vspace{1mm} \footnotesize
+
+\begin{verbatim}
+##                mean        sd       5.5%      94.5%    n_eff    Rhat4
+## co[1] -0.1979930035 0.1708785 -0.4682696 0.07634977 1789.894 1.003445
+## co[2] -0.0334215769 0.1687587 -0.3021954 0.23720938 1738.720 1.003575
+## co[3] -0.0192492753 0.1675754 -0.2840596 0.24860974 1732.907 1.003755
+## co[4] -0.0003911363 0.1670610 -0.2661047 0.26815172 1723.758 1.003837
+\end{verbatim}
+
+\normalsize
+
+\begin{center}
+\includegraphics[width=8cm]{../images/islamCoeffs.jpeg}
+\end{center}
+
+\normalsize
+
+While muslim words were unusual in the sense that the disparity between
+associated attributes and others is stronger, the evidence is still not
+conclusive. This is because the variation even within islam-related
+words is large enough (and sample sizes sufficiently small) for all the
+highest posterior density invervals to still include zeros.
+
+So, it seems, taking a closer look does seem to make a difference. The
+question is, what happens if we do take a close look at the level of
+protected words?
+
+\chapter{Protected-word level
+analysis}\label{protected-word-level-analysis}
+
+\section{Model structure and
+assumptions}\label{model-structure-and-assumptions}
+
+Let's turn then to data analysis that takes a look at protected words
+separately. This time for each combination of a protected word and a
+connection status we will have a separate mean cosine distance estimate,
+each coming with its own highest posterior density interval. This means
+we will use indices that are result from all such combinations (and then
+we will split them up in the model precis to build visualisation, feel
+free to look at the \texttt{visualiseStats.R} script for details).
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{options}\NormalTok{(}\DataTypeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\OtherTok{TRUE}\NormalTok{ ) }\CommentTok{#removes install pop-up request}
+\NormalTok{religion}\OperatorTok{$}\NormalTok{pwFactor <-}\StringTok{ }\KeywordTok{factor}\NormalTok{(}\KeywordTok{paste0}\NormalTok{(religion}\OperatorTok{$}\NormalTok{protectedWord, }\StringTok{"-"}\NormalTok{, religion}\OperatorTok{$}\NormalTok{connection))}
+\NormalTok{religion}\OperatorTok{$}\NormalTok{pwIndex <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(religion}\OperatorTok{$}\NormalTok{pwFactor)}
+
+\NormalTok{religionSeparate <-}\StringTok{ }\KeywordTok{ulam}\NormalTok{(}
+  \KeywordTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu <-}\StringTok{ }\NormalTok{c[pwIndex],}
+\NormalTok{    c[pwIndex] }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\OperatorTok{~}\StringTok{ }\KeywordTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\NormalTok{  ),}
+  \DataTypeTok{data =}\NormalTok{ religion,}
+  \DataTypeTok{chains=}\DecValTok{2}\NormalTok{ , }\DataTypeTok{iter=}\DecValTok{10000}\NormalTok{ , }\DataTypeTok{warmup=}\DecValTok{1000}\NormalTok{,}
+  \DataTypeTok{start=}\KeywordTok{list}\NormalTok{(}\DataTypeTok{no =} \DecValTok{1}\NormalTok{, }\DataTypeTok{a =} \DecValTok{0}\NormalTok{, }\DataTypeTok{d =} \DecValTok{0}\NormalTok{, }\DataTypeTok{sigma=} \FloatTok{.3}\NormalTok{), }\DataTypeTok{log_lik =}  \OtherTok{TRUE}
+\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+\noindent Let's see if the individualized model does better than the
+previous models in light of WAIC which does add penalty for the number
+of parameters.
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{compareBaselineCoefsSeparate<-}\StringTok{ }\KeywordTok{readRDS}\NormalTok{(}\StringTok{"../datasets/compareBaselineCoefsSeparate.rds"}\NormalTok{)}
+\NormalTok{compareBaselineCoefsSeparate}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+##                   WAIC SE dWAIC dSE pWAIC weight
+## religionSeparate -2400 93     0  NA    60      1
+## religionCoefs    -2328 93    72  29    20      0
+## religionBaseline -2283 95   117  37    16      0
+\end{verbatim}
+
+\normalsize
+
+\section{Protected classes in Reddit and Google
+embeddings}\label{protected-classes-in-reddit-and-google-embeddings}
+
+It seems that we do want to prefer this model, despite its relative
+complication. Now, what does it tell us about the protected words? Let's
+visualise the predicted means together with 89\% highest posterior
+density intervals.
+
+\includegraphics[width=14cm]{../images/visReligionReddit.png}
+
+Before we move on, let's perform analogous analyses for the remaining
+types of supposed bias: gender and race (the model building is
+analogous).
+
+\includegraphics[width=14cm]{../images/visGenderReddit.png}
+
+\includegraphics[width=14cm]{../images/visRaceReddit.png}
+
+\todo{Explain where you took the embedding from, cite publications first announcing it.}
+
+\includegraphics[width=14cm]{../images/visReligionGoogle.png}
+
+\includegraphics[width=14cm]{../images/visGenderGoogle.png}
+
+\includegraphics[width=14cm]{../images/visRaceGoogle.png}
+
+\chapter{The role of debiasing}\label{the-role-of-debiasing}
+
+\todo{describe which debiasing you used and how it was advertised in the paper}
+
+\section{Dataset-level coefficients after
+debiasing}\label{dataset-level-coefficients-after-debiasing}
+
+First, let's look at coefficient estimated for the whole datasets, as
+compared to their estimation prior to debiasing:
+
+\begin{center}
+\begin{figure}[!htb]\centering
+   \begin{minipage}{0.55\textwidth}
+  \includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
+   \end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/debiasedReligionRedditCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/genderCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/debiasedGenderRedditCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/raceCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/debiasedRaceRedditCoeffs.jpeg}
+   \end{minipage}
+\end{figure}
+
+\end{center}
+
+\todo{comment on this}
+
+Some points to discuss:
+
+\begin{itemize}
+\item
+  humans in religion got closer (should they?)
+\item
+  change in Gender is really minor, still zero out of HPDI range, note
+  how different is very similar
+\item
+  small improvement in race, at the price of moving neutral terms closer
+\end{itemize}
+
+\section{Protected classes after
+debiasing}\label{protected-classes-after-debiasing}
+
+Now, perhaps, the effects of debiasing will be better appreciated if we
+look at the level of protected words. After all, the hope is, the
+situation of extremely ill-positioned protected words have improved?
+
+\includegraphics[width=14cm]{../images/visDebReligionReddit.png}
+
+\includegraphics[width=14cm]{../images/visDebGenderReddit.png}
+
+\includegraphics[width=14cm]{../images/visDebRaceReddit.png}
 
 \hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\hypertarget{ref-Bolukbasi2016Man}{}%
-Bolukbasi, T., Chang, K.-W., Zou, J.
-Y., Saligrama, V., \& Kalai, A.
-(2016).
-Man is to computer programmer as woman is to homemaker?
-Debiasing word embeddings.
-\emph{CoRR}, \emph{abs/1607.06520}.
-Retrieved from \url{http://arxiv.org/abs/1607.06520}
+\hypertarget{ref-Bolukbasi2016Man}{}
+Bolukbasi, T., Chang, K., Zou, J. Y., Saligrama, V., \& Kalai, A.
+(2016). Man is to computer programmer as woman is to homemaker?
+Debiasing word embeddings. \emph{CoRR}, \emph{abs/1607.06520}. Retrieved
+from \url{http://arxiv.org/abs/1607.06520}
 
-\leavevmode\hypertarget{ref-Caliskan2017Semantics}{}%
-Islam, A.
-C., Bryson, J.
-J., \& Narayanan, A.
-(2016).
-Semantics derived automatically from language corpora necessarily contain human biases.
-\emph{CoRR}, \emph{abs/1608.07187}.
-Retrieved from \url{http://arxiv.org/abs/1608.07187}
+\hypertarget{ref-Caliskan2017Semantics}{}
+Islam, A. C., Bryson, J. J., \& Narayanan, A. (2016). Semantics derived
+automatically from language corpora necessarily contain human biases.
+\emph{CoRR}, \emph{abs/1608.07187}. Retrieved from
+\url{http://arxiv.org/abs/1608.07187}
 
-\leavevmode\hypertarget{ref-manzini2019black}{}%
-Manzini, T., Lim, Y.
-C., Tsvetkov, Y., \& Black, A.
-W.
-(2019).
-Black is to criminal as caucasian is to police: Detecting and removing multiclass bias in word embeddings.
-Retrieved from \url{http://arxiv.org/abs/1904.04047}
+\hypertarget{ref-manzini2019black}{}
+Manzini, T., Lim, Y. C., Tsvetkov, Y., \& Black, A. W. (2019). Black is
+to criminal as caucasian is to police: Detecting and removing multiclass
+bias in word embeddings.
 
-\leavevmode\hypertarget{ref-Mehrabi2019Survey}{}%
+\hypertarget{ref-Mehrabi2019Survey}{}
 Mehrabi, N., Morstatter, F., Saxena, N., Lerman, K., \& Galstyan, A.
-(2019).
-A survey on bias and fairness in machine learning.
-\emph{CoRR}, \emph{abs/1908.09635}.
-Retrieved from \url{http://arxiv.org/abs/1908.09635}
+(2019). A survey on bias and fairness in machine learning. \emph{CoRR},
+\emph{abs/1908.09635}. Retrieved from
+\url{http://arxiv.org/abs/1908.09635}
 
-\leavevmode\hypertarget{ref-Nissim2019Fair}{}%
-Nissim, M., Noord, R.
-van, \& Goot, R.
-van der.
-(2019).
-Fair is better than sensational: Man is to doctor as woman is to doctor.
-\emph{CoRR}, \emph{abs/1905.09866}.
-Retrieved from \url{http://arxiv.org/abs/1905.09866}
-
-\end{CSLReferences}
+\hypertarget{ref-Nissim2019Fair}{}
+Nissim, M., Noord, R. van, \& Goot, R. van der. (2019). Fair is better
+than sensational: Man is to doctor as woman is to doctor. \emph{CoRR},
+\emph{abs/1905.09866}. Retrieved from
+\url{http://arxiv.org/abs/1905.09866}
 

### 2021-06-10 thesis/03-bayesian.Rmd

@@
 
-```{r religionTableHead,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+```{r religionTableHead,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%", warning = FALSE, message = FALSE}
 religion <- read.csv("../datasets/religionReddit.csv")[-1]
@@
                         "cosineDistance","cosineSimilarity","connection")
-levels(religion$wordClass) <- c("christian","human","jewish","muslim","neutral")
+library(plyr)
+religion$wordClass <- mapvalues(religion$wordClass, 
+                          from = levels(religion$wordClass),
+                          to = c("christian","human","jewish","muslim","neutral"))
+#levels(religion$wordClass) <- c("christian","human","jewish","muslim","neutral")
 head(religion)  %>%  kable(format = "latex",booktabs=T,
                       linesep = "",  escape = FALSE, 
-                      caption = "Head of the religion dataset.") %>%
-                      kable_styling(latex_options=c("scale_down"))
+                      caption = "Head of the religion dataset.") %>%      kable_styling(latex_options=c("scale_down"))
 ```

### 2021-06-10 thesis/02-cosine.Rmd
test
adds fixes

@@
+---
+output:
+  pdf_document: default
+  html_document: default
+---
 # Cosine similarity and bias detection
@@
 
-```{python}
-import pandas as pd
-data = pd.read_csv("religion_cosine_table.csv")
-print(data.head())
+\footnotesize 
+```{r religionTableHeadEarly,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+religion <- read.csv("../datasets/religionReddit.csv")[-1]
+colnames(religion) <- c("protectedWord","wordToCompare","wordClass",
+                        "cosineDistance","cosineSimilarity","connection")
+religion$wordClass <- as.factor(religion$wordClass)
+levels(religion$wordClass) <- c("christian","human","jewish","muslim","neutral")
+head(religion)  %>%  kable(format = "latex",booktabs=T,
+                      linesep = "",  escape = FALSE, 
+                      caption = "Head of the religion dataset.") %>%
+                      kable_styling(latex_options=c("scale_down"))
+```
+\normalsize
 
-```
+\pagebreak
 

### 2021-06-10 thesis/03-bayesian.Rmd
adds fixes

--- 2021-06-08 thesis/03-bayesian.Rmd
@@
                         "cosineDistance","cosineSimilarity","connection")
+#head(religion)
+#library(plyr)
+religion$wordClass <- as.factor(religion$wordClass)
+#levels(religion$wordClass)
+#religion$wordClass <- mapvalues(religion$wordClass, 
+#                          from = levels(religion$wordClass),
+#                          to = c("christian","human","jewish","muslim","neutral"))
 levels(religion$wordClass) <- c("christian","human","jewish","muslim","neutral")
@@
                       linesep = "",  escape = FALSE, 
-                      caption = "Head of the religion dataset.") %>%
-                      kable_styling(latex_options=c("scale_down"))
+                      caption = "Head of the religion dataset.") %>%      kable_styling(latex_options=c("scale_down"))
 ```

### 2021-06-10 thesis/_book/_main.tex
adds fixes

@@
-\documentclass[12pt,]{book}
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+%
+\documentclass[
+  12pt,
+]{book}
+\usepackage{amsmath,amssymb}
 \usepackage{lmodern}
 \usepackage{setspace}
-\setstretch{1.5}
-\usepackage{amssymb,amsmath}
 \usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
 \ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
@@
   pdfauthor={Alicja Dobrzeniecka},
-            pdfborder={0 0 0},
-            breaklinks=true}
-\urlstyle{same}  % don't use monospace font for urls
-\usepackage[left=3.5cm, right=2.5cm, top=2.5cm, bottom=2.5cm, asymmetric,
-includeheadfoot]{geometry}
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
+\urlstyle{same} % disable monospaced font for URLs
+\usepackage[left=3.5cm, right=2.5cm, top=2.5cm, bottom=2.5cm, asymmetric, includeheadfoot]{geometry}
 \usepackage{color}
@@
 \newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\BuiltInTok}[1]{#1}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
 \newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
 \newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
+\newcommand{\ExtensionTok}[1]{#1}
 \newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ImportTok}[1]{#1}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\NormalTok}[1]{#1}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\RegionMarkerTok}[1]{#1}
 \newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
 \newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
 \newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\ImportTok}[1]{#1}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
-\newcommand{\BuiltInTok}[1]{#1}
-\newcommand{\ExtensionTok}[1]{#1}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
-\newcommand{\RegionMarkerTok}[1]{#1}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
 \newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
-\newcommand{\NormalTok}[1]{#1}
-\usepackage{longtable,booktabs}
-% Fix footnotes in tables (requires footnote package)
-\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{long table}}{}
-\usepackage{graphicx,grffile}
+\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
+\usepackage{graphicx}
 \makeatletter
@@
 }
-\chapter{Introduction}\label{introduction}
-
-Natural language processing (NLP) is a subfield of computer science that
-processes and analyzes language in text and speech with the use of
-modern programming methods. It has practical applications in everyday
-life as it concerns tasks such as email filters, smart assistants,
-search results, language translations, text analytics and so on. Models
-used to accomplish these tasks need a lot of data to learn from. This
-data originates from humans activities and historical recordings such as
-texts, messages or speeches. It turns out that in the learning process
-these models can learn implicit biases that reflect harmful
-stereotypical thinking still present in modern societies. One can find
-methods that aim at identifying and measuring hidden biases and/or try
-to remove them by modifying the models. There are many different types
-of models in NLP depending on a task that they are supposed to solve.
-However, all of them need as an input words represented by means of
-numbers and this is accomplished with word embedding models. The models
-usually assign the values based on the context in which the words
-appear. It means that the input data can have enormous influence on the
-outcome. The biases seem to have their primary source in the way the
-words are assigned the numerical values.
-
-There is considerable amount of literature available on the topic of
-bias detection and mitigation in NLP models. Bolukbasi, Chang, Zou,
-Saligrama, \& Kalai (\protect\hyperlink{ref-Bolukbasi2016Man}{2016})
-focuses on gender biases that may be observable while investigating the
-representation of job occupations and gender in terms of their assigned
-numerical values. The authors apply cosine similarity measurement to
-investigate the phenomenon where (the vectors corresponding to) words
-related to jobs that are stereotypically associated with a given gender
-are in fact in the model situated closer to this gender. They also use
-analogy tasks to evaluate if the bias is present in the word embedding
-model. They check analogies by comparing pairs of word vectors, for
-example they search for the word complementing the puzzle: man is to
-doctor as woman is to \ldots{}? First they subtract word ``man'' from
-word ``woman'' and then they search for the ranked list of other words
-pairs that have similar vectors' difference. They also include in the
-formula a threshold to ensure that the resulting pairs could not be
-randomly picked.
-
-However, as in Nissim, Noord, \& Goot
-(\protect\hyperlink{ref-Nissim2019Fair}{2019}) it is pointed out, there
-are some limitations of this approach. According to the authors in
-practice most of analogies implementations do not return any input
-words. This means that it does not make sense to expect the algorithm to
-return the same profession for both woman and man. Therefore this method
-seems to be limited in terms of bias detection. The other problems
-regard for example the choice of pairs and words that are used to detect
-the presence of discrimination as it is often subjective and without
-proper justification. Additionally the choice of parameter set in
-Bolukbasi et al. (\protect\hyperlink{ref-Bolukbasi2016Man}{2016})
-formula to ensure that word pairs are not picked by random, is also not
-justified and changing it drastically influences the results.
-
-Islam, Bryson, \& Narayanan
-(\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) touches upon the
-topic of biases regarding race and gender. They apply knowledge from
-well-known psychological studies such as Implicit Association Test to
-research the relation between human stereotypical thinking and model
-learnt biases to discover close relationship between these two. For the
-evaluation they use Word Embedding Association Test (WEAT) and the Word
-Embedding Factual Association Test (WEFAT).
-
-Manzini, Lim, Tsvetkov, \& Black
-(\protect\hyperlink{ref-manzini2019black}{2019}) proposes a novel way of
-using cosine similarity to obtain the information on assumed resemblance
-between words. They investigate an approach that enables them to measure
-the bias for a class (like gender, religion, race) and express the final
-result with a single metric.
-
-It is worth noticing the general distinction of biases mentioned in
-Islam et al. (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}). They
-refer to the publication concerning Implicit Association Test (Greenwald
-et al., 1998) that measures the strength of associations between
-concepts or stereotypes by calculating the time of reaction for special
-tasks. It is worth noticing that humans naturally exhibit some biases
-and that they do not always cause social concern. One can imagine the
-intuitive associations between for example insects and flowers, and the
-feelings of pleasantness or unpleasantness. In general, people would
-rather associate flowers with feeling pleasant than insects, and this
-preference could be named a bias or prejudice in some direction.
-However, this type of preference does not cause an uproar and is a
-rather morally neutral case. Unfortunately, there are other biases and
-prejudices that directly influence the quality of other people's lives
-and therefore they should be taken care of.
-
-One can find various definitions trying to capture what bias and
-fairness actually are. With the choice of the definition, implications
-into the real-life applications may change as well. Mehrabi, Morstatter,
-Saxena, Lerman, \& Galstyan
-(\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) mark out that there
-exist different types of biases such as historical bias, representation
-bias, measurement bias (the list is long). This indicates how complex
-the issue of bias is. Without the proper understanding and awareness of
-the problem, people are prone to unconsciously sustain the bias
-existence.
-
-Mehrabi et al. (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) also
-distinguish different types of discrimination, some of them will be
-briefly described. By protected attributes we mean those qualities,
-traits or characteristics that one cannot legally discriminate against.
-Direct discrimination occurs when protected attributes of individuals
-explicitly result in non-favorable outcomes toward them. In contrast in
-indirect discrimination individuals appear to be treated equally but
-anyway they end up being treated unjustly due to the hidden effects of
-biases towards their protected attributes. Systemic discrimination takes
-place when policies, customs or behaviors that result from certain
-culture or organizational structure lead to discrimination against some
-groups of people. Finally, very common statistical discrimination refers
-to using average group statistics to judge person belonging to the
-group.
-
-The topic of discrimination is entangled with another concept which is
-fairness. It is essential to grasp some concepts of fairness to take
-them into consideration while designing implementation of some machine
-learning model. In Mehrabi2019Survey one may notice that depending on
-the context and application different definitions may be applied.
-
-The most popular methods focus on comparing the similarity between words
-from protected groups and those that are considered to be stereotypical
-or harmful in some way. One can find in this group methods such as
-euclidean distance or cosine similarity (which is equivalent to dot
-product if the vectors are normalized). There are also other ways to
-detect the effects of biases. For example through the investigation of
-the model performance on certain tasks that validate if the model
-returns some values independently on gender or race or not.
-
-The currently used methods (such as cosine similarity) make the
-similarity values often aggregated in a way that may lead to hasty
-conclusions. The averaging of values and the lack of uncertainty may
-lead to the incomplete picture of the bias situation in the vocabulary.
-
-One can find a number of articles on negative real-life implications
-resulting from the presence of unaddressed biases in the machine
-learning models.
-
-In the paper we indicate how current methods used to detect biases in
-natural language models are limited from the perspective of Bayesian
-analysis.
-
-Our research enhances the current way in which the bias detection is
-performed to make sure that it is methodologically valid.
-
-The key hypothesis is that greater understanding of data and bias
-implications can be achieved when Bayesian methods are applied to issue.
-
-\chapter{Cosine similarity and bias
-detection}\label{cosine-similarity-and-bias-detection}
-
-\section{Word embeddings}\label{word-embeddings}
-
-To understand what cosine similarity measurement is, one first needs to
-grasp the concept of translating words to a computer-readable form. In
-the field of natural language processsing there are two main types of
-words representation --- localist and distributed. One-hot encoding is
-an example of a method used to achieve a localist representation of
-words. Here each vector contains information only about a single data
-point, this is achieved by first mapping categorical values (words) to
-integers and then to each integers a binary vector is assigned which
-contains only 0s except for the index of the integer, which is assigned
-1. An example of a localist representation is:
+\setstretch{1.5}
+\hypertarget{introduction}{%
+\chapter{Introduction}\label{introduction}}
+
+Natural language processing (NLP) is a subfield of computer science that processes and analyzes language in text and speech
+with the use of modern programming methods. It has practical applications in everyday life as it concerns tasks such as email filters,
+smart assistants, search results, language translations, text analytics and so on. Models used to accomplish these tasks need a lot of
+data to learn from. This data originates from humans activities and historical recordings such as texts, messages or speeches. It turns out
+that in the learning process these models can learn implicit biases that reflect harmful stereotypical thinking still present in modern societies.
+One can find methods that aim at identifying and measuring hidden biases and/or try to remove them by modifying the models.
+There are many different types of models in NLP depending on a task that they are supposed to solve.
+However, all of them need as an input words represented by means of numbers and this is accomplished with word embedding models.
+The models usually assign the values based on the context in which the words appear.
+It means that the input data can have enormous influence on the outcome.
+The biases seem to have their primary source in the way the words are assigned the numerical values.
+
+There is considerable amount of literature available on the topic of bias detection and mitigation in NLP models.
+\protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) focuses on gender biases that may be observable while investigating the representation of job occupations and gender in terms of their assigned numerical
+values.
+The authors apply cosine similarity measurement to investigate the phenomenon where (the vectors corresponding to) words related to jobs that are stereotypically associated with a given gender are in fact in the model situated closer to this gender.
+They also use analogy tasks to evaluate if the bias is present in the word embedding model.
+They check analogies by comparing pairs of word vectors, for example they search for the word complementing the puzzle: man is to doctor as woman is to \ldots?
+First they subtract word ``man'' from word ``woman'' and then
+they search for the ranked list of other words pairs that have similar vectors' difference.
+They also include in the formula a threshold to ensure that the resulting pairs could not be randomly picked.
+
+However, as in \protect\hyperlink{ref-Nissim2019Fair}{Nissim, Noord, \& Goot} (\protect\hyperlink{ref-Nissim2019Fair}{2019}) it is pointed out, there are some limitations of this approach.
+According to the authors in practice most of analogies implementations do not return any input words.
+This means that it does not make sense to expect the algorithm to return the same profession for both woman and man.
+Therefore this method seems to be limited in terms of bias detection.
+The other problems regard for example the
+choice of pairs and words that are used to detect the presence of discrimination as it is often subjective and without proper justification.
+Additionally the choice of parameter set in \protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) formula to ensure that word pairs are not picked by random, is also not justified and changing it drastically
+influences the results.
+
+\protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) touches upon the topic of biases regarding race and gender.
+They apply knowledge from well-known psychological studies such as Implicit Association Test to research the relation between human stereotypical thinking and model learnt biases to
+discover close relationship between these two. For the evaluation they use Word Embedding Association Test (WEAT) and the Word Embedding Factual Association Test (WEFAT).
+
+\protect\hyperlink{ref-manzini2019black}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-manzini2019black}{2019}) proposes a novel way of using cosine similarity to obtain the information on assumed resemblance between words.
+They investigate
+an approach that enables them to measure the bias for a class (like gender, religion, race) and express the final result with a single metric.
+
+It is worth noticing the general distinction of biases mentioned in \protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}).
+They refer to the publication concerning Implicit Association Test (Greenwald et al., 1998) that measures the strength of associations between concepts or stereotypes by calculating the time of reaction for special tasks.
+It is worth noticing that humans naturally exhibit some biases and that they do not always cause social concern.
+One can imagine the intuitive associations between for example insects and flowers, and the feelings of pleasantness or unpleasantness.
+In general, people would rather associate flowers with feeling pleasant than insects, and this preference could be named a bias or
+prejudice in some direction. However, this type of preference does not cause an uproar and is a rather morally neutral case. Unfortunately, there are other
+biases and prejudices that directly influence the quality of other people's lives and therefore they should be taken care of.
+
+One can find various definitions trying to capture what bias and fairness actually are.
+With the choice of the definition, implications into the real-life applications may change as well.
+\protect\hyperlink{ref-Mehrabi2019Survey}{Mehrabi, Morstatter, Saxena, Lerman, \& Galstyan} (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) mark out that there exist different types of biases such as historical bias, representation bias, measurement bias (the list is long).
+This indicates how complex the issue of bias is.
+Without the proper understanding and awareness of the problem, people are prone to unconsciously sustain the bias existence.
+
+\protect\hyperlink{ref-Mehrabi2019Survey}{Mehrabi, Morstatter, Saxena, Lerman, \& Galstyan} (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) also distinguish different types of discrimination, some of them will be briefly described.
+By protected attributes we mean those qualities, traits or characteristics that one cannot legally discriminate against.
+Direct discrimination occurs when protected attributes of individuals explicitly result in non-favorable outcomes toward them.
+In contrast in indirect discrimination individuals appear to be treated equally but anyway they end up being treated unjustly due to the hidden effects of biases towards their protected attributes.
+Systemic discrimination takes place when policies, customs or behaviors that result from certain culture or organizational structure lead to discrimination against some groups of people.
+Finally, very common statistical discrimination refers to using
+average group statistics to judge person belonging to the group.
+
+The topic of discrimination is entangled with another concept which is fairness. It is essential to grasp some concepts of fairness to take them into
+consideration while designing implementation of some machine learning model. In Mehrabi2019Survey one may notice that depending on the context and application different definitions may be applied.
+
+The most popular methods focus on comparing the similarity between words from protected groups and those that are considered to be stereotypical or harmful in some way.
+One can find in this group methods such as euclidean distance or cosine similarity (which is equivalent to dot product if the
+vectors are normalized).
+There are also other ways to detect the effects of biases.
+For example through the investigation of the model performance on certain tasks that validate if the model returns some values
+independently on gender or race or not.
+
+The currently used methods (such as cosine similarity) make the similarity values often aggregated in a way that may lead to hasty conclusions.
+The averaging of values and the lack of uncertainty may lead to the incomplete picture of the bias situation in the vocabulary.
+
+One can find a number of articles on negative real-life implications resulting from the presence of unaddressed biases in the machine learning models.
+
+In the paper we indicate how current methods used to detect biases in natural language models are limited from the perspective of Bayesian analysis.
+
+Our research enhances the current way in which the bias detection is performed to make sure that it is
+methodologically valid.
+
+The key hypothesis is that greater understanding of data and bias implications can be achieved when Bayesian methods are applied to issue.
+
+\hypertarget{cosine-similarity-and-bias-detection}{%
+\chapter{Cosine similarity and bias detection}\label{cosine-similarity-and-bias-detection}}
+
+\hypertarget{word-embeddings}{%
+\section{Word embeddings}\label{word-embeddings}}
+
+To understand what cosine similarity measurement is, one first needs to grasp the concept of translating words to a computer-readable form.
+In the field of natural language processsing there are two main types of words representation --- localist and distributed.
+One-hot encoding is an example of a method used to achieve a localist representation of words.
+Here each vector contains information only about a single data point, this is achieved by first mapping categorical values (words) to integers and then to each integers a binary vector is assigned which contains only 0s except for the index of the integer, which is assigned 1.
+An example of a localist representation is:
 
@@
 \toprule
-word & 1 & 2 & 3 & 4 & 5\tabularnewline
+word & 1 & 2 & 3 & 4 & 5 \\
 \midrule
 \endhead
-woman & 1 & 0 & 0 & 0 & 0\tabularnewline
-man & 0 & 1 & 0 & 0 & 0\tabularnewline
-girl & 0 & 0 & 1 & 0 & 0\tabularnewline
-boy & 0 & 0 & 0 & 1 & 0\tabularnewline
-monarch & 0 & 0 & 0 & 0 & 1\tabularnewline
+woman & 1 & 0 & 0 & 0 & 0 \\
+man & 0 & 1 & 0 & 0 & 0 \\
+girl & 0 & 0 & 1 & 0 & 0 \\
+boy & 0 & 0 & 0 & 1 & 0 \\
+monarch & 0 & 0 & 0 & 0 & 1 \\
 \bottomrule
@@
 
-In the example above it is clear that the length of the vectors
-increases with the number of words in a vocabulary. It is not a very
-computationally efficient representation. It has other flaws as well.
-For example, it is unable to capture the resemblance between words
-appearing in similar contexts.
-
-In contrast to the localist representation, a distributed representation
-returns vectors that contain continuous values instead of discrete 1s
-and 0s. Word embeddings are a class of various techniques that allow one
-to represent words as distributed vectors. Such learned representations
-of text have certain properties. At least prima facie, they store
-similar (or at least co-occurring) words close to each other in a vector
-space. An example of distributed representation is:
+In the example above it is clear that the length of the vectors increases with the number of words in a vocabulary.
+It is not a very computationally efficient representation.
+It has other flaws as well.
+For example, it is unable to capture the resemblance between words appearing in similar contexts.
+
+In contrast to the localist representation, a distributed representation returns vectors that contain continuous values instead of discrete 1s and 0s.
+Word embeddings are a class of various techniques that allow one to represent words as distributed vectors.
+Such learned representations of text have certain properties.
+At least prima facie, they store similar (or at least co-occurring) words close to each other in a vector space.
+An example of distributed representation is:
 
@@
 \toprule
-word & 1 & 2 & 3 & 4\tabularnewline
+word & 1 & 2 & 3 & 4 \\
 \midrule
 \endhead
-woman & 0.456 & 0.267 & 0.675 & 0.131\tabularnewline
-man & 0.451 & 0.897 & 0.472 & 0.088\tabularnewline
-girl & 0.604 & 0.262 & 0.414 & 0.706\tabularnewline
-boy & 0.279 & 0.172 & 0.475 & 0.010\tabularnewline
-monarch & 0.565 & 0.678 & 0.463 & 0.975\tabularnewline
+woman & 0.456 & 0.267 & 0.675 & 0.131 \\
+man & 0.451 & 0.897 & 0.472 & 0.088 \\
+girl & 0.604 & 0.262 & 0.414 & 0.706 \\
+boy & 0.279 & 0.172 & 0.475 & 0.010 \\
+monarch & 0.565 & 0.678 & 0.463 & 0.975 \\
 \bottomrule
@@
 
-One of the advantages of using a distributed representation is that one
-is able to represent an enormous number of concepts with a smaller
-number of units. It is also possible to better capture similarities as
-words of similar meanings can have similar numeric vectors.
-
-The numbers occurring in such representations are not random. They are
-learned in a process that uses a very shallow neural network. There are
-various types of techniques used for learning the vectors
-representations. One of the most straightforward ones is a skip-gram
-model. Given a word the models tries to predict its neighboring words
-from the sentence. The mathematics behind the process relies on the idea
-that the prediction concerns the conditional probability of the adjacent
-words. The algorithm tries to minimize the loss function, which
-penalizes the system for discrepancy with actual co-occurrence
-frequencies in the corpus. One can choose various parameters of the
-model, such as the window size that determines how many surrounding
-words the model should predict. After preparing such a fitted model one
-takes only the learned weights from a neural network, and uses them as
-vectors in a word embeddings representation.
+One of the advantages of using a distributed representation is that one is able to represent an enormous number of concepts with a smaller number of units.
+It is also possible to better capture similarities as words of similar meanings can have similar numeric vectors.
+
+The numbers occurring in such representations are not random.
+They are learned in a process that uses a very shallow neural network.
+There are various types of techniques used for learning the vectors representations.
+One of the most straightforward ones is a skip-gram model.
+Given a word the models tries to predict its neighboring words from the sentence.
+The mathematics behind the process relies on the idea that the prediction concerns the conditional probability of the adjacent words.
+The algorithm tries to minimize the loss function, which penalizes the system for discrepancy with actual co-occurrence frequencies in the corpus.
+One can choose various parameters of the model, such as the window size that determines how many surrounding words the model should predict.
+After preparing such a fitted model one takes only the learned weights from a neural network, and uses them as vectors in a word embeddings representation.
 
 Word embeddings have many applications in natural language processing.
-They are handy in document search and information retrieval. They also
-play their part in improving automatic translations. Well learned word
-representations may also contribute to the improvement of sentiment
-analysis or spam detection.
-
-\section{Cosine similarity and
-distance}\label{cosine-similarity-and-distance}
-
-Cosine similarity is often used as a method of finding out whether
-vector representations for two words suggest that they are similar or
-somehow connected. Cosine similarity is the cosine of the angle between
-two vectors: the result of dividing their inner product (dot product
-usually) by the product of their magnitudes.
-
+They are handy in document search and information retrieval.
+They also play their part in improving automatic translations.
+Well learned word representations may also contribute to the improvement of sentiment analysis or spam detection.
+
+\hypertarget{cosine-similarity-and-distance}{%
+\section{Cosine similarity and distance}\label{cosine-similarity-and-distance}}
+
+Cosine similarity is often used as a method of finding out whether vector representations for two words suggest that they are similar or somehow connected.
+Cosine similarity is the cosine of the angle between two vectors: the result of dividing their inner product (dot product usually) by the product of their magnitudes.
 \begin{align} \tag{Sim}
@@
 \end{align}
-
-Cosine similarity is considered a proper tool for this operation as its
-result has a clear connection to geometry and at least for a low number
-of dimensions may be easily interpreted. Using this scale, one can
-compare vector similarities in a fairly clear manner. When the vectors
-are aligned perpendicularly to each other, their similarity equals 0
-(which is the same as the cosine of 90 degrees). This tells us that the
-similarity between the vectors is small. As the angle between vectors
-decreases, cosine similarity approaches one, which stands for the
-greatest similarity.
-
-One of the limitations of this measure is that it informs us only about
-similarities between vectors in terms of their orientation. However, it
-is often argued that in comparing words in terms of this metric, the
-magnitude of vectors may be treated as irrelevant, as the most important
-information pertains to direction.
-
-In what follows, it is important to distinguish between cosine
-similarity and cosine distance, defined as:
-
+Cosine similarity is considered a proper tool for this operation as its result has a clear connection to geometry and at least for a low number of dimensions may be easily interpreted.
+Using this scale, one can compare vector similarities in a fairly clear manner.
+When the vectors are aligned perpendicularly to each other, their similarity equals 0 (which is the same as the cosine of 90 degrees).
+This tells us that the similarity between the vectors is small.
+As the angle between vectors decreases, cosine similarity approaches one, which stands for the greatest similarity.
+
+One of the limitations of this measure is that it informs us only about similarities between vectors in terms of their orientation.
+However, it is often argued that in comparing words in terms of this metric, the magnitude of vectors may be treated as irrelevant, as the most important information pertains to direction.
+
+In what follows, it is important to distinguish between cosine similarity and cosine distance, defined as:
 \begin{align} \tag{Sim}
@@
 
-The greater the similarity between two vectors, the smaller the distance
-between them. The cosine distance ranges between 0 and 2. If the vectors
-are in an opposite direction to each other, the cosine distance is 2.
-And if the vectors are extremely similar then the cosine distance is
-very close to 0.
+The greater the similarity between two vectors, the smaller the distance between them.
+The cosine distance ranges between 0 and 2.
+If the vectors are in an opposite direction to each other, the cosine distance is 2.
+And if the vectors are extremely similar then the cosine distance is very close to 0.
 
 It is worth mentioning one more point concerning cosine similarity.
-After the vectors are normalized to have length equal to 1, inner
-product itself (often dot product) is used to measure the similarity.
-
-\section{Cosine distance in a one-class bias
-detection}\label{cosine-distance-in-a-one-class-bias-detection}
-
-Bolukbasi et al. (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) define
-similarity between words as the outcome inner product of their
-normalized vectors. They focus on examining what the geometry of word
-embedding is in regard to ``he'' and ``she'' words. In other words,
-whether the similarity between those concepts and other words reflects
-expected gender stereotypes. They test this hypothesis by investigating
-whether there is a connection between word embeddings representing
-certain professions and words referring to gender. They also evaluate
-whether automatically produced analogies between words reflect the
-stereotypes as well.
-
-A very vivid way to follow their method of arguing that bias in word
-embeddings is real is to plot the values of inner product of chosen
-words. The plot below does not originate from the original paper (it is
-from
-\url{https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings}) but
-similar visualization may be found there. Data used to create our plot
-is as follows. \newline
-
-Occupations associated with feminine:
-\textbf{"homemaker", "nurse", "receptionist", "librarian", "socialite", "hairdresser", "nanny", "bookkeeper", "stylist", "housekeeper", "interior designer", "guidance counselor"}
-\newline
-
-Occupations associated with masculine:
-\textbf{"maestro", "skipper", "protege", "philosopher", "captain", "architect", "financier", "warrior", "broadcaster", "magician", "fighter pilot", "boss"}
-\newline
-
-\vspace{1mm} \footnotesize
+After the vectors are normalized to have length equal to 1, inner product itself (often dot product) is used to measure the similarity.
+
+\hypertarget{cosine-distance-in-a-one-class-bias-detection}{%
+\section{Cosine distance in a one-class bias detection}\label{cosine-distance-in-a-one-class-bias-detection}}
+
+\protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016})
+define similarity between words as the outcome inner product of their normalized vectors.
+They focus on examining what the geometry of word embedding is in regard to ``he'' and ``she'' words.
+In other words, whether the similarity between those concepts and other words reflects expected gender stereotypes.
+They test this hypothesis by investigating whether there is a connection between word embeddings representing certain professions and words referring to gender.
+They also evaluate whether automatically produced analogies between words reflect the stereotypes as well.
+
+A very vivid way to follow their method of arguing that bias in word embeddings is real is to plot the values of inner product of chosen words.
+The plot below does not originate from the original paper (it is from \url{https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings}) but similar visualization may be found there.
+Data used to create our plot is as follows.
+\newline
+
+Occupations associated with feminine: \textbf{"homemaker", "nurse", "receptionist", "librarian", "socialite", "hairdresser", "nanny", "bookkeeper", "stylist", "housekeeper", "interior designer", "guidance counselor"}
+\newline
+
+Occupations associated with masculine: \textbf{"maestro", "skipper", "protege", "philosopher", "captain", "architect", "financier", "warrior", "broadcaster", "magician", "fighter pilot", "boss"}
+\newline
+
+\vspace{1mm}
+\footnotesize
 
 \begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-1-1} \end{center}
-
-\normalsize
-
-The points in the plot above result from the calculation of the inner
-product of a chosen vector for a profession word and a vector for a
-gender word (she or he). Inner product of two vectors expresses
-similarity between words. This assumption originates from the geometry
-and properties of a vector space.
-
-\section{Cosine distance in a multi-class bias
-detection}\label{cosine-distance-in-a-multi-class-bias-detection}
-
-Manzini et al. (\protect\hyperlink{ref-manzini2019black}{2019}) present
-a different approach towards finding similarities between classes of
-words. The authors claim that texts available online are full of direct
-or indirect human stereotypes. As a result, word embeddings are prone to
-learn and maybe amplify those biases and propagate them further into AI
-models that are used for various applications. Cosine distance is used
-in the article as a measure to first argue for the existence of
-multi-class bias and then to show how through bias mitigation techniques
-the bias may be decreased.
-
-As the code has been provided by the authors we were able to reconstruct
-their results. The main steps in the procedure are as follows. Let us go
-through an example that refers to the process of hard debiasing on
-religious attributes.
-
-The example refers to the process of hard debiasing on religious
-attributes.
+\normalsize
+
+The points in the plot above result from the calculation of the inner product of a chosen vector for a profession word and a vector for a gender word (she or he).
+Inner product of two vectors expresses similarity between words.
+This assumption originates from the geometry and properties of a vector space.
+
+\hypertarget{cosine-distance-in-a-multi-class-bias-detection}{%
+\section{Cosine distance in a multi-class bias detection}\label{cosine-distance-in-a-multi-class-bias-detection}}
+
+\protect\hyperlink{ref-manzini2019black}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-manzini2019black}{2019}) present a different approach towards finding similarities between classes of words.
+The authors claim that texts available online are full of direct or indirect human stereotypes.
+As a result, word embeddings are prone to learn and maybe amplify those biases and propagate them further into AI models that are used for various applications.
+Cosine distance is used in the article as a measure to first argue for the existence of multi-class bias and then to show how through bias mitigation techniques the bias may be decreased.
+
+As the code has been provided by the authors we were able to reconstruct their results.
+The main steps in the procedure are as follows.
+Let us go through an example that refers to the process of hard debiasing on religious attributes.
+
+The example refers to the process of hard debiasing on religious attributes.
 
@@
 \item
-  First we load word embeddings from
-  reddit.US.txt.tok.clean.cleanedforw2v.w2v dataset
+  First we load word embeddings from reddit.US.txt.tok.clean.cleanedforw2v.w2v dataset
 \end{enumerate}
@@
 \item
-  The protected group the authors assume should ideally not have high
-  cosine similarity to stereotypical words. The word embeddings geometry
-  should not place this group close to harmful stereotypes, if it is to
-  be bias-free. \newline
+  The protected group the authors assume should ideally not have high cosine similarity to stereotypical words.
+The word embeddings geometry should not place this group close to harmful stereotypes, if it is to be bias-free.
+  \newline
 \end{enumerate}
@@
 
-\textbf{"jew"       : ["greedy", "cheap", "hairy", "liberal"]} \newline
+\textbf{"jew"       : ["greedy", "cheap", "hairy", "liberal"]}
+\newline
 \textbf{"christian"       : ["judgemental", "conservative", "familial"]}
@@
 
-We have prepared a table presenting the values of cosine distance for
-each protected word with each attribute (stereotype). The part of the
-results is shown below.
-
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{religion <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
-\KeywordTok{colnames}\NormalTok{(religion) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
+We have prepared a table presenting the values of cosine distance for each protected word with each attribute (stereotype). The part of the results is shown below.
+
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{religion }\OtherTok{\textless{}{-}} \FunctionTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{]}
+\FunctionTok{colnames}\NormalTok{(religion) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
                         \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
-\KeywordTok{levels}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
-\KeywordTok{head}\NormalTok{(religion)  }\OperatorTok{%>%}\StringTok{  }\KeywordTok{kable}\NormalTok{(}\DataTypeTok{format =} \StringTok{"latex"}\NormalTok{,}\DataTypeTok{booktabs=}\NormalTok{T,}
-                      \DataTypeTok{linesep =} \StringTok{""}\NormalTok{,  }\DataTypeTok{escape =} \OtherTok{FALSE}\NormalTok{, }
-                      \DataTypeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\OperatorTok{%>%}
-\StringTok{                      }\KeywordTok{kable_styling}\NormalTok{(}\DataTypeTok{latex_options=}\KeywordTok{c}\NormalTok{(}\StringTok{"scale_down"}\NormalTok{))}
+\NormalTok{religion}\SpecialCharTok{$}\NormalTok{wordClass }\OtherTok{\textless{}{-}} \FunctionTok{as.factor}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{wordClass)}
+\FunctionTok{levels}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{wordClass) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
+\FunctionTok{head}\NormalTok{(religion)  }\SpecialCharTok{\%\textgreater{}\%}  \FunctionTok{kable}\NormalTok{(}\AttributeTok{format =} \StringTok{"latex"}\NormalTok{,}\AttributeTok{booktabs=}\NormalTok{T,}
+                      \AttributeTok{linesep =} \StringTok{""}\NormalTok{,  }\AttributeTok{escape =} \ConstantTok{FALSE}\NormalTok{, }
+                      \AttributeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+                      \FunctionTok{kable\_styling}\NormalTok{(}\AttributeTok{latex\_options=}\FunctionTok{c}\NormalTok{(}\StringTok{"scale\_down"}\NormalTok{))}
 \end{Highlighting}
@@
 \end{table}
-
 \normalsize
@@
 
-In the article there was no analysis of individual distances but the
-general look at the data through the usage of mean. The authors
-introduced a metric that tries to generalize the presence of bias
-through the classification of multi-class bias in groups of words
-connected with gender, religion or race. In the process they first take
-the mean of cosine distances between a given protected word and
-attributes assigned to each stereotype. They do not differentiate
-between stereotypes associated with a word and stereotypes associated
-with different words (in case of religion, stereotypes characteristic
-for Christianity has also cosine distance measured with for instance,
-Judaism or Islam). Then, after collecting the list of mean cosine
-distances, they average the list to obtain one final value representing
-the whole group, in this example religion, for which the final mean of
-all mean distances is equal to 0.859.
+In the article there was no analysis of individual distances but the general look at the data through the usage of mean.
+The authors introduced a metric that tries to generalize the presence of bias through the classification of multi-class bias in groups of words connected with gender, religion or race.
+In the process they first take the mean of cosine distances between a given protected word and attributes assigned to each stereotype.
+They do not differentiate between stereotypes associated with a word and stereotypes associated with different words (in case of religion, stereotypes characteristic for Christianity has also cosine distance measured with for instance, Judaism or Islam).
+Then, after collecting the list of mean cosine distances, they average the list to obtain one final value representing the whole group, in this example religion, for which the final mean of all mean distances is equal to 0.859.
 
@@
 \item
-  In the article the authors also try to remove previously defined
-  biases from word embedding.
+  In the article the authors also try to remove previously defined biases from word embedding.
 \end{enumerate}
 
-First they identify the bias subspace using Principal Component Analysis
-(PCA) which is a technique for dimensionality reduction. It is applied
-here to choose the subspace that contains the greatest amount of
-information. There can be many subspaces found in a given group, for
-example in terms of religion one can identify at least a few sets that
-are to grasp the concept of religion in general: \newline
-\textbf{["judaism", "christianity", "islam"]} \newline
-\textbf{["jew", "christian", "muslim"]} \newline
-\textbf{["synagogue", "church", "mosque"]} \newline
-
-The idea is to find a set that provides enough information to create
-from it a vector representing the concept of religion among words. This
-strategy is based on the idea that different dimensions of vectors
-contain different types of information and in some words in vector
-layers (subspaces) the information about religiousness is implicitly
-conveyed. In some cases this knowledge is useful but in the case of
-harmful stereotypes one does not want to have religion concept in
-stereotypical words.
-
-After finding bias subspace, they use it to modify the vector values
-individually so that their cosine distances towards certain words are
-changed. In the case of stereotypes the aim is to make the cosine
-distances larger so that the association between protected word and
-harmful stereotype is smaller.
+First they identify the bias subspace using Principal Component Analysis (PCA) which is a technique for dimensionality reduction.
+It is applied here to choose the subspace that contains the greatest amount of information.
+There can be many subspaces found in a given group, for example in terms of religion one can identify at least a few sets that are to grasp the concept of religion in general:
+\newline
+\textbf{["judaism", "christianity", "islam"]}
+\newline
+\textbf{["jew", "christian", "muslim"]}
+\newline
+\textbf{["synagogue", "church", "mosque"]}
+\newline
+
+The idea is to find a set that provides enough information to create from it a vector representing the concept of religion among words.
+This strategy is based on the idea that different dimensions of vectors contain different types of information and in some words in vector layers (subspaces) the information about religiousness is implicitly conveyed.
+In some cases this knowledge is useful but in the case of harmful stereotypes one does not want to have religion concept in stereotypical words.
+
+After finding bias subspace, they use it to modify the vector values individually so that their cosine distances towards certain words are changed.
+In the case of stereotypes the aim is to make the cosine distances larger so that the association between protected word and harmful stereotype is smaller.
 
@@
 \item
-  In the final step there is an evaluation of the results. The cosine
-  distances are calculated again but this time using debiased
-  vocabulary. After taking the mean of all distances one final value is
-  obtained and then it is compared with the average value from the
-  beginning. If the cosine distances are on average greater than before
-  then it leads the authors to the conclusion that improvement was made.
-  As the cosine distance increases it is assumed that the association
-  between protected and stereotypical words decreases.
+  In the final step there is an evaluation of the results. The cosine distances are calculated again but this time using debiased vocabulary. After
+  taking the mean of all distances one final value is obtained and then it is compared with the average value from the beginning.
+If the cosine distances are on average greater than before then it leads the authors to the conclusion that improvement was made.
+As the cosine distance increases it is assumed that the association between protected and stereotypical words decreases.
 \end{enumerate}
 
-\section{Limitations of the approach}\label{limitations-of-the-approach}
+\hypertarget{limitations-of-the-approach}{%
+\section{Limitations of the approach}\label{limitations-of-the-approach}}
 
@@
 
-The attributes are taken from different sources, there is no principled
-justification for their choice. From our analysis it will become clear
-that the list is rather uneven.
-
-There is no mention of methodology for deciding on the number of
-attributes necessary to decide a hypothesis on the given size of
-dataset. There are however some ways to estimate how many samples we
-need to make sure that the result is significant. Our research will show
-that the numbers used are rather insufficient.
+The attributes are taken from different sources, there is no principled justification for their choice. From our analysis it will become clear that
+the list is rather uneven.
+
+There is no mention of methodology for deciding on the number of attributes necessary to decide a hypothesis on the given size of dataset.
+There are however some ways to estimate how many samples we need to make sure that the result is significant.
+Our research will show that the numbers used
+are rather insufficient.
 
@@
 
-The authors use the mean average cosine similarity to check on
-multi-class similarity between protected word and harmful stereotypes.
-They average the results until they obtain one final value to represent
-the mean cosine distance between protected word from a given class and
-the attributes of that class. As there is no control of the individual
-values of cosine distances, one can observe values that have cosine
-distances greater than 1 which would suggest that these compare words do
-not have significant similarity. However, as there is a lack of control,
-even such words are taken into the account for calculations.
+The authors use the mean average cosine similarity to check on multi-class similarity between protected word and harmful stereotypes.
+They average the results until they obtain one final value to represent the mean cosine distance between protected word from a given class and the
+attributes of that class. As there is no control of the individual values of cosine distances, one can observe values that have cosine distances
+greater than 1 which would suggest that these compare words do not have significant similarity. However, as there is a lack of control, even such
+words are taken into the account for calculations.
 
@@
 
-A mean hides this issue and as there are pairs having negative and small
-similarities and there are those that have similarity equal to 0.5, the
-resulting calculation seems to be in norm. Additionally in such method
-the uncertainty is also not included which makes it even more difficult
-to give reasonable interpretations of the results. We propose the use of
-Bayesian method to obtain some understanding of the influence the
-uncertainty has on the interpretation of final results.
+A mean hides this issue and as there are pairs having negative and small similarities and there are those that have similarity equal to 0.5, the resulting calculation seems to be in norm.
+Additionally in such method the uncertainty is also not included which makes it even more difficult to give reasonable interpretations of the results.
+We propose the use of Bayesian method to obtain some understanding of the influence the uncertainty has on the interpretation of final results.
 
@@
 
-In the original paper words from all three religions were checked with
-all of the stereotypes which means that there was no distinction on
-whether the stereotype is associated with given religion or with the
-other. Not all of the stereotypical words should be considered as
-harmful for all of the religions. In our analysis we distinguished
-between associated stereotypes, not associated (stereotype is valid but
-for different religion than currently checked) and none (meaning neutral
-words).
+In the original paper words from all three religions were checked with all of the stereotypes which means that there was no distinction on whether the stereotype is associated with given religion or with the other.
+Not all of the stereotypical words should be considered as harmful for all of the religions.
+In our analysis we distinguished between associated stereotypes, not associated (stereotype is valid but for different religion than currently checked) and none (meaning neutral words).
 
@@
 
-Assuming for a moment that the value of multi-class cosine distance is
-correct, one may question the results' interpretation. In Manzini et al.
-(\protect\hyperlink{ref-manzini2019black}{2019}) Table 2, there are
-summarized the averages of cosine distance per group (gender, race,
-religion). I would like to focus now on analyzing the values relating to
-religious biases. Here is the fragment of table that refers to that:
+Assuming for a moment that the value of multi-class cosine distance is correct, one may question the results' interpretation.
+In \protect\hyperlink{ref-manzini2019black}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-manzini2019black}{2019}) Table 2, there are summarized the averages of cosine distance per group (gender, race, religion).
+I would like to focus now on analyzing the values relating to religious biases.
+Here is the fragment of table that refers to that:
 
@@
 \toprule
-Religion Debiasing & MAC\tabularnewline
+Religion Debiasing & MAC \\
 \midrule
 \endhead
-Biased & 0.859\tabularnewline
-Hard Debiased & 0.934\tabularnewline
-Soft Debiased (\(\lambda\) = 0.2) & 0.894\tabularnewline
+Biased & 0.859 \\
+Hard Debiased & 0.934 \\
+Soft Debiased (\(\lambda\) = 0.2) & 0.894 \\
 \bottomrule
@@
 
-MAC stand for mean average cosine similarity although in reality the
-values of cosine distance are stored there. What may attract attention
-is the fact that the value of cosine distance in ``Biased'' category is
-already quite high even before ``debiasing''. High cosine distance
-indicates low cosine similarity between values. One could think that
-average cosine similarity equal to approximately 0.141 is not
-significant enough to consider it as biased. However the authors aim to
-mitigate ``biases'' in vectors with such great distance to make it even
-larger. Methodologically there is a question on what basis this small
-similarity is still considered as a proof of bias presence.
-
-This is in general the problem of scale and the lack of universal
-intervals that could be applied to know whether the cosine similarity is
-high enough to considered given words more similar that if we chose them
-on random. It seems like similar practice could be beneficial for
-metrics aimed at measuring similarity between words.
+MAC stand for mean average cosine similarity although in reality the values of cosine distance are stored there.
+What may attract attention is the fact that the value of cosine distance in ``Biased'' category is already quite high even before ``debiasing.'' High cosine distance indicates low cosine similarity between values.
+One could think that average cosine similarity equal to approximately 0.141 is not significant enough to consider it as biased.
+However the authors aim to mitigate ``biases'' in vectors with such great distance to make it even larger.
+Methodologically there is a question on what basis this small similarity is still considered as a proof of bias presence.
+
+This is in general the problem of scale and the lack of universal intervals that could be applied to know whether the cosine similarity is high enough to considered given words more similar that if we chose them on random.
+It seems like similar practice could be beneficial for metrics aimed at measuring similarity between words.
 
@@
 
-Curse of dimensionality may take place when there is an increase in
-volume of data that results in adding extra dimensions to the Euclidean
-space. According to the article
+Curse of dimensionality may take place when there is an increase in volume of data that results in adding extra dimensions to the Euclidean space. According to the article
 ``\url{https://analyticsindiamag.com/curse-of-dimensionality-and-what-beginners-should-do-to-overcome-it/}''
-as the number of features increases, it may be harder and harder to
-obtain useful information from the data with the usage of available
-algorithms. One may notice that more data should contribute to greater
-amount of information but more information also means greater risk of
-noise and distractions in data. At the same time, many times modern
-solutions are adapted to smaller dimensions and their results in higher
-ones are not intuitive or may be prone to be mistaken.
-
-Using cosine similarity in high dimensions in word embeddings may also
-be prone to the curse of dimensionality. According to this article
-``\url{https://www.researchgate.net/publication/327498046_The_Curse_of_Dimensionality_Inside_Out}''
-there are reasons to consider this phenomenon when searching for word
-similarities in higher dimensions.
-
-In the article an experiment is conducted that aims at showing how the
-similarity values and variation change as the number of dimensions
-increases. The hypothesis made in the paper states that two things will
-happen as the number of dimensions increase, the first one is that
-effort required to measure cosine similarity will be greater and the
-second one is that the similarity between data will blur out and have
-less variation. In details, the authors generate random points with
-increasing number of dimensions where each dimension of a data point is
-given a value between 0 and 1. Then they pick one vector at random from
-each dimension class and calculate cosine similarity between the chosen
-vector and the rest of the data. Then they check how the variation of
-values changes as the number of dimensions increases. It seems like the
-more dimensions there are, the smaller the variance and therefore it is
-less obvious how to interpret the resulting cosine similarities. Maybe
-the scale should be adjusted to the number of dimensions and variance so
-that it still gives us sensible information about data. According to
-some researches the cosine similarity in high dimensions is not reliable
-enough to trust it as it may be the case that choosing words on random
-may result in getting similar values as when picking them consciously.
+as the number of features increases, it may be harder and harder to obtain useful information from the data with the usage of available algorithms.
+One may notice that more data should contribute to greater amount of information but more information also means greater risk of noise and distractions in data.
+At the same time, many times modern solutions are adapted to smaller dimensions and their results in higher ones are not intuitive or may be prone to be mistaken.
+
+Using cosine similarity in high dimensions in word embeddings may also be prone to the curse of dimensionality.
+According to this article ``\url{https://www.researchgate.net/publication/327498046_The_Curse_of_Dimensionality_Inside_Out}''
+there are reasons to consider this phenomenon when searching for word similarities in higher dimensions.
+
+In the article an experiment is conducted that aims at showing how the similarity values and variation change as the number of dimensions increases.
+The hypothesis made in the paper states that two things will happen as the number of dimensions increase, the first one is that effort required to measure cosine similarity will be greater and the second one is that the similarity between data will blur out and have less variation.
+In details, the authors generate random points with increasing number of dimensions where each dimension of a data point is given a value between 0 and 1.
+Then they pick one vector at random from each dimension class and calculate cosine similarity between the chosen vector and the rest of the data.
+Then they check how the variation of values changes as the number of dimensions increases.
+It seems like the more dimensions there are, the smaller the variance and therefore it is less obvious how to interpret the resulting cosine similarities.
+Maybe the scale should be adjusted to the number of dimensions and variance so that it still gives us sensible information about data.
+According to some researches the cosine similarity in high dimensions is not reliable enough to trust it as it may be the case that choosing words on random may result in getting similar values as when picking them consciously.
 
@@
 \includegraphics{../images/curseOfDimensionality.png}
-\caption{curse of dimensionality, number of dimensions on the x axis,
-standard deviation of similarity on the y axis}
+\caption{curse of dimensionality, number of dimensions on the x axis, standard deviation of similarity on the y axis}
 \end{figure}
 
-\chapter{Walkthrough with the religion
-dataset}\label{walkthrough-with-the-religion-dataset}
-
-\section{Loading and understanding the
-dataset}\label{loading-and-understanding-the-dataset}
+\hypertarget{walkthrough-with-the-religion-dataset}{%
+\chapter{Walkthrough with the religion dataset}\label{walkthrough-with-the-religion-dataset}}
+
+\hypertarget{loading-and-understanding-the-dataset}{%
+\section{Loading and understanding the dataset}\label{loading-and-understanding-the-dataset}}
 
@@
 \begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{(ggplot2)}
-\KeywordTok{library}\NormalTok{(ggthemes)}
-\KeywordTok{library}\NormalTok{(rethinking)}
-\KeywordTok{library}\NormalTok{(tidyverse)}
-\KeywordTok{library}\NormalTok{(ggpubr)}
-\KeywordTok{library}\NormalTok{(kableExtra)}
-\KeywordTok{library}\NormalTok{(dplyr)}
-\KeywordTok{library}\NormalTok{(ggExtra)}
-\KeywordTok{library}\NormalTok{(cowplot)}
-\end{Highlighting}
-\end{Shaded}
-
-\normalsize
-
-We will use the choice of protected words and stereotypical predicates
-used in REF. This is a decent point of departure, not only we want to
-compare our method to that of REF, but also because this data format is
-fairly general (as contrasted, say, with a set up for binary
-stereotypes). Note also that the method we develop here can fairly
-easily be run for different stereotypization patterns. Let's start with
-explaining the method and its deployment using a dataset obtained for
-the religion-related protected words.
-
-Let's load, clean a bit and inspect the head of the religion dataset we
-prepared. In order to obtain this dataset, we calculated the cosine
-distance between each protected word and each word from both the
-bias-related attribute groups, which were used in the original study,
-and to neutral and human control attributes which we added as control
-groups. For instance, for religion, the bias-related predicates (coming
-from the original study in REF) include muslim bias attributes, jew bias
-attributes, christian bias attributes (see a list in the APPENDIX).
-
-We decided to add control groups in the form of two classes --- neutral
-words and human-related words. Without a proper control group it is
-quite hard to compare the resulting cosine distances and decide on their
-significance in bias detection. We prepared approximately
-300\todo{FIX later} more or less neutral words to double-check the
-prima-facie neutral hypothesis that their cosine similarity to the
-protected words will oscillate around 0 (that is, the distances will be
-around 1). This provides us with a more reliable point of reference.
-Moreover, we added human attributes that are associated with people in
-general to investigate whether the smaller cosine distance between
-protected words and stereotypes can result simply from the fact that the
-stereotype predicates are associated with humans. For two control
-groups, we have randomly drawn 300 words that do not express any
-property usually attributed to humans, and human attributes.
+\FunctionTok{library}\NormalTok{(ggplot2)}
+\FunctionTok{library}\NormalTok{(ggthemes)}
+\FunctionTok{library}\NormalTok{(rethinking)}
+\FunctionTok{library}\NormalTok{(tidyverse)}
+\FunctionTok{library}\NormalTok{(ggpubr)}
+\FunctionTok{library}\NormalTok{(kableExtra)}
+\FunctionTok{library}\NormalTok{(dplyr)}
+\FunctionTok{library}\NormalTok{(ggExtra)}
+\FunctionTok{library}\NormalTok{(cowplot)}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+We will use the choice of protected words and stereotypical predicates used in REF.
+This is a decent point of departure, not only we want to compare our method to that of REF, but also because this data format is fairly general (as contrasted, say, with a set up for binary stereotypes).
+Note also that the method we develop here can fairly easily be run for different stereotypization patterns.
+Let's start with explaining the method and its deployment using a dataset obtained for the religion-related protected words.
+
+Let's load, clean a bit and inspect the head of the religion dataset we prepared.
+In order to obtain this dataset, we calculated the cosine distance between each protected word and each word from both the bias-related attribute groups, which were used in the original study, and to neutral and human control attributes which we added as control groups.
+For instance, for religion, the bias-related predicates (coming from the original study in REF) include muslim bias attributes, jew bias attributes, christian bias attributes (see a list in the APPENDIX).
+
+We decided to add control groups in the form of two classes --- neutral words and human-related words.
+Without a proper control group it is quite hard to compare the resulting cosine distances and decide on their significance in bias detection.
+We prepared approximately 300\todo{FIX later} more or less neutral words to double-check the prima-facie neutral hypothesis that their cosine similarity to the protected words will oscillate around 0 (that is, the distances will be around 1).
+This provides us with a more reliable point of reference.
+Moreover, we added human attributes that are associated with people in general to investigate whether the smaller cosine distance between protected words and stereotypes can result simply from the fact that the stereotype predicates are associated with humans.
+For two control groups, we have randomly drawn 300 words that do not express any property usually attributed to humans, and human attributes.
 \todo{describe once the words are selected}
 
-\vspace{1mm} \footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{religion <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
-\KeywordTok{colnames}\NormalTok{(religion) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{religion }\OtherTok{\textless{}{-}} \FunctionTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{]}
+\FunctionTok{colnames}\NormalTok{(religion) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
                         \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
-\KeywordTok{library}\NormalTok{(plyr)}
-\NormalTok{religion}\OperatorTok{$}\NormalTok{wordClass <-}\StringTok{ }\KeywordTok{mapvalues}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass, }\DataTypeTok{from =} \KeywordTok{levels}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass), }\DataTypeTok{to =} \KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{))}
-\CommentTok{#levels(religion$wordClass) <- c("christian","human","jewish","muslim","neutral")}
-\KeywordTok{head}\NormalTok{(religion)  }\OperatorTok{%>%}\StringTok{  }\KeywordTok{kable}\NormalTok{(}\DataTypeTok{format =} \StringTok{"latex"}\NormalTok{,}\DataTypeTok{booktabs=}\NormalTok{T,}
-                      \DataTypeTok{linesep =} \StringTok{""}\NormalTok{,  }\DataTypeTok{escape =} \OtherTok{FALSE}\NormalTok{, }
-                      \DataTypeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\OperatorTok{%>%}\StringTok{      }\KeywordTok{kable_styling}\NormalTok{(}\DataTypeTok{latex_options=}\KeywordTok{c}\NormalTok{(}\StringTok{"scale_down"}\NormalTok{))}
+\CommentTok{\#head(religion)}
+\CommentTok{\#library(plyr)}
+\NormalTok{religion}\SpecialCharTok{$}\NormalTok{wordClass }\OtherTok{\textless{}{-}} \FunctionTok{as.factor}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{wordClass)}
+\CommentTok{\#levels(religion$wordClass)}
+\CommentTok{\#religion$wordClass \textless{}{-} mapvalues(religion$wordClass, }
+\CommentTok{\#                          from = levels(religion$wordClass),}
+\CommentTok{\#                          to = c("christian","human","jewish","muslim","neutral"))}
+\FunctionTok{levels}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{wordClass) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
+\FunctionTok{head}\NormalTok{(religion)  }\SpecialCharTok{\%\textgreater{}\%}  \FunctionTok{kable}\NormalTok{(}\AttributeTok{format =} \StringTok{"latex"}\NormalTok{,}\AttributeTok{booktabs=}\NormalTok{T,}
+                      \AttributeTok{linesep =} \StringTok{""}\NormalTok{,  }\AttributeTok{escape =} \ConstantTok{FALSE}\NormalTok{, }
+                      \AttributeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}      \FunctionTok{kable\_styling}\NormalTok{(}\AttributeTok{latex\_options=}\FunctionTok{c}\NormalTok{(}\StringTok{"scale\_down"}\NormalTok{))}
 \end{Highlighting}
@@
 \end{table}
-
-\normalsize
-
-The \texttt{protectedWord} column contains words from a protected class
-that (in a perfect world according to the assumptions of the orignal
-study) should not be associated with harmful stereotypes.
-\texttt{wordToCompare} contains attributes, including stereotypes and
-control group words. For each row we compute the cosine distances
-between a given protected word and a given attribute word.
-\texttt{wordClass} tells us which class an attribute is supposed to be
-stereotypically associated with, that is, whether the word from
-\texttt{wordToCompare} is associated stereotypically with jews,
-christians or muslims, or whether it belongs to a control group.
-\texttt{cosineDistance} is simply a calculation of the cosine distance
-between protected word and atrribute. \texttt{cosineSimilarity} contains
-the result of substracting cosine distance from 1. \texttt{connection}
-contains information about the relation type between a protected word
-and an attribute. If the attribute is e.g.~a harmful jewish stereotype
-and the protected word is also from the judaism group, the connection
-has value \texttt{associated}. If the attribute is still stereotypically
-jewish, but the protected word comes from another religion, the
-connection is labelled as \texttt{different}. If the attribute belongs
-to a neutral group then the connection is labelled as \texttt{none} and
-if an attribute belongs to the \texttt{human} class, then the connection
-is labelled as \texttt{human}.
-
-\section{First look at the empirical
-distributions}\label{first-look-at-the-empirical-distributions}
-
-First let's take a look at the empirical distribution of distances by
-the connection type, initially ignoring the human control class for now.
-
-\vspace{1mm} \footnotesize
+\normalsize
+
+The \texttt{protectedWord} column contains words from a protected class that (in a perfect world according to the assumptions of the orignal study) should not be associated with harmful stereotypes.
+\texttt{wordToCompare} contains attributes, including stereotypes and control group words.
+For each row we compute the cosine distances between a given protected word and a given attribute word.
+\texttt{wordClass} tells us which class an attribute is supposed to be stereotypically associated with, that is, whether the word from \texttt{wordToCompare} is associated stereotypically with jews, christians or muslims, or whether it belongs to a control group.
+\texttt{cosineDistance} is simply a calculation of the cosine distance between protected word and atrribute.
+\texttt{cosineSimilarity} contains the result of substracting cosine distance from 1.
+\texttt{connection} contains information about the relation type between a protected word and an attribute.
+If the attribute is e.g.~a harmful jewish stereotype and the protected word is also from the judaism group, the connection has value \texttt{associated}.
+If the attribute is still stereotypically jewish, but the protected word comes from another religion, the connection is labelled as \texttt{different}.
+If the attribute belongs to a neutral group then the connection is labelled as \texttt{none} and if an attribute belongs to the \texttt{human} class, then the connection is labelled as \texttt{human}.
+
+\hypertarget{first-look-at-the-empirical-distributions}{%
+\section{First look at the empirical distributions}\label{first-look-at-the-empirical-distributions}}
+
+First let's take a look at the empirical distribution of distances by the connection type, initially ignoring the human control class for now.
+
+\vspace{1mm}
+\footnotesize
 
 \begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-3-1} \end{center}
-
-\normalsize
-
-The first impression is that while there is a shift for associated words
-towards smaller cosine distances as compared to the neutral words,
-slightly surprisingly a slightly weaker shift in the same direction is
-visible for attributes associated with different stereotypes. Moreover,
-the empirical distributions overlap to alarge extent and the means
-grouped by connection type do not seem too far from each other. In fact,
-as there is a lot of variety in the consine distances (as we will soon
-see), abd we need to gauge the uncertaintly involved, and to look more
-carefully at individual protected words to get a better idea of how the
-cosine distance distribution changes for different attribute groups and
-different protected classes. Now, let's add the human attributes to the
-picture:
-
-\vspace{1mm} \footnotesize
+\normalsize
+
+The first impression is that while there is a shift for associated words towards smaller cosine distances as compared to the neutral words, slightly surprisingly a slightly weaker shift in the same direction is visible for attributes associated with different stereotypes.
+Moreover, the empirical distributions overlap to alarge extent and the means grouped by connection type do not seem too far from each other.
+In fact, as there is a lot of variety in the consine distances (as we will soon see), abd we need to gauge the uncertaintly involved, and to look more carefully at individual protected words to get a better idea of how the cosine distance distribution changes for different attribute groups and different protected classes.
+Now, let's add the human attributes to the picture:
+
+\vspace{1mm}
+\footnotesize
 
 \begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-4-1} \end{center}
-
-\normalsize
-
-\noindent Notice that the distribution for \texttt{human} (even though
-we did our best not to include in it any stereotype-related atributes)
-is left-skewed, with much overlap with \texttt{associated} and
-\texttt{different}, which illustrates the need to take being associated
-with humans as an important predictor.
-
-Our focus lies in \texttt{connection} as a predictor. Morever, later on
-we'll be interested in looking at the protected words separately, and at
-protected words split by connection. For technical reasons it is useful
-to represent these factors as integer vectors.
-
-\vspace{1mm} \footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{religion}\OperatorTok{$}\NormalTok{con <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(religion}\OperatorTok{$}\NormalTok{connection)}
-\NormalTok{religion}\OperatorTok{$}\NormalTok{pw <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(religion}\OperatorTok{$}\NormalTok{protectedWord)}
-\NormalTok{religion}\OperatorTok{$}\NormalTok{pwFactor <-}\StringTok{ }\KeywordTok{factor}\NormalTok{(}\KeywordTok{paste0}\NormalTok{(religion}\OperatorTok{$}\NormalTok{protectedWord, religion}\OperatorTok{$}\NormalTok{connection))}
-\NormalTok{religion}\OperatorTok{$}\NormalTok{pwIndex <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(religion}\OperatorTok{$}\NormalTok{pwFactor)}
-\end{Highlighting}
-\end{Shaded}
-
-\normalsize
-
-A short script, \texttt{cleanDataset} to make this faster, so
-equivalently:
-
-\vspace{1mm} \footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{source}\NormalTok{(}\StringTok{"../functions/cleanDataset.R"}\NormalTok{)}
-\NormalTok{religion <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
-\NormalTok{religion <-}\StringTok{ }\KeywordTok{cleanDataset}\NormalTok{(religion,}\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\normalsize
-
-\section{Looking at the islam-related
-words}\label{looking-at-the-islam-related-words}
-
-For now, let's focus on five protected words related to islam (``imam'',
-``islam'', ``mosque'', ``muslim'', and ``quran''). The word list
-associates with islam four stereotypical attributes (``violent'',
-``terrorist'', ``uneducated'' and ``dirty''). First, we select and plot
-the empirical distributions for these protected words.
-
-\vspace{1mm} \footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{(tidyverse)}
-\NormalTok{muslimWords <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"imam"}\NormalTok{,}\StringTok{"islam"}\NormalTok{,}\StringTok{"mosque"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"quran"}\NormalTok{)}
-\NormalTok{muslim <-}\StringTok{ }\NormalTok{religion }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(protectedWord }\OperatorTok{%in%}\StringTok{ }\NormalTok{muslimWords)}
-\KeywordTok{ggplot}\NormalTok{(muslim, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{  cosineDistance, }\DataTypeTok{fill =}\NormalTok{ connection, }\DataTypeTok{color =}\NormalTok{ connection))}\OperatorTok{+}
-\StringTok{  }\KeywordTok{geom_density}\NormalTok{(}\DataTypeTok{alpha=}\FloatTok{0.6}\NormalTok{,}\DataTypeTok{size =} \FloatTok{.2}\NormalTok{)}\OperatorTok{+}\StringTok{ }
-\StringTok{  }\KeywordTok{scale_fill_manual}\NormalTok{(}\DataTypeTok{values =} \KeywordTok{c}\NormalTok{(}\StringTok{"orangered4"}\NormalTok{,}\StringTok{"chartreuse4"}\NormalTok{, }\StringTok{"skyblue"}\NormalTok{, }\StringTok{"gray"}\NormalTok{))}\OperatorTok{+}
-\StringTok{  }\KeywordTok{scale_x_continuous}\NormalTok{(}\DataTypeTok{breaks =} \KeywordTok{seq}\NormalTok{(}\FloatTok{0.3}\NormalTok{,}\FloatTok{1.5}\NormalTok{, }\DataTypeTok{by =} \FloatTok{0.1}\NormalTok{))}\OperatorTok{+}\KeywordTok{xlab}\NormalTok{(}\StringTok{"cosine distance"}\NormalTok{)}\OperatorTok{+}
-\StringTok{  }\KeywordTok{scale_color_manual}\NormalTok{(}\DataTypeTok{values =} \KeywordTok{c}\NormalTok{(}\StringTok{"orangered4"}\NormalTok{,}\StringTok{"chartreuse4"}\NormalTok{,}\StringTok{"skyblue"}\NormalTok{,}\StringTok{"gray"}\NormalTok{))}\OperatorTok{+}
-\StringTok{  }\KeywordTok{theme_tufte}\NormalTok{()}\OperatorTok{+}\KeywordTok{ggtitle}\NormalTok{(}\StringTok{"Empirical distribution of distances (muslim)"}\NormalTok{)}
+\normalsize
+
+\noindent Notice that the distribution for \texttt{human} (even though we did our best not to include in it any stereotype-related atributes) is left-skewed, with much overlap with \texttt{associated} and \texttt{different}, which illustrates the need to take being associated with humans as an important predictor.
+
+Our focus lies in \texttt{connection} as a predictor.
+Morever, later on we'll be interested in looking at the protected words separately, and at protected words split by connection.
+For technical reasons it is useful to represent these factors as integer vectors.
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{religion}\SpecialCharTok{$}\NormalTok{con }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{connection)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning: NAs introduced by coercion
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{religion}\SpecialCharTok{$}\NormalTok{pw }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{protectedWord)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning: NAs introduced by coercion
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{religion}\SpecialCharTok{$}\NormalTok{pwFactor }\OtherTok{\textless{}{-}} \FunctionTok{factor}\NormalTok{(}\FunctionTok{paste0}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{protectedWord, religion}\SpecialCharTok{$}\NormalTok{connection))}
+\NormalTok{religion}\SpecialCharTok{$}\NormalTok{pwIndex }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{pwFactor)}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+A short script, \texttt{cleanDataset} to make this faster, so equivalently:
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{source}\NormalTok{(}\StringTok{"../functions/cleanDataset.R"}\NormalTok{)}
+\NormalTok{religion }\OtherTok{\textless{}{-}} \FunctionTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{]}
+\NormalTok{religion }\OtherTok{\textless{}{-}} \FunctionTok{cleanDataset}\NormalTok{(religion,}\FunctionTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning in cleanDataset(religion, c("christian", "human", "jewish", "muslim", :
+## NAs introduced by coercion
+
+## Warning in cleanDataset(religion, c("christian", "human", "jewish", "muslim", :
+## NAs introduced by coercion
+\end{verbatim}
+
+\normalsize
+
+\hypertarget{looking-at-the-islam-related-words}{%
+\section{Looking at the islam-related words}\label{looking-at-the-islam-related-words}}
+
+For now, let's focus on five protected words related to islam (``imam,'' ``islam,'' ``mosque,'' ``muslim,'' and ``quran'').
+The word list associates with islam four stereotypical attributes (``violent,'' ``terrorist,'' ``uneducated'' and ``dirty'').
+First, we select and plot the empirical distributions for these protected words.
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{library}\NormalTok{(tidyverse)}
+\NormalTok{muslimWords }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"imam"}\NormalTok{,}\StringTok{"islam"}\NormalTok{,}\StringTok{"mosque"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"quran"}\NormalTok{)}
+\NormalTok{muslim }\OtherTok{\textless{}{-}}\NormalTok{ religion }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(protectedWord }\SpecialCharTok{\%in\%}\NormalTok{ muslimWords)}
+\FunctionTok{ggplot}\NormalTok{(muslim, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{  cosineDistance, }\AttributeTok{fill =}\NormalTok{ connection, }\AttributeTok{color =}\NormalTok{ connection))}\SpecialCharTok{+}
+  \FunctionTok{geom\_density}\NormalTok{(}\AttributeTok{alpha=}\FloatTok{0.6}\NormalTok{,}\AttributeTok{size =}\NormalTok{ .}\DecValTok{2}\NormalTok{)}\SpecialCharTok{+} 
+  \FunctionTok{scale\_fill\_manual}\NormalTok{(}\AttributeTok{values =} \FunctionTok{c}\NormalTok{(}\StringTok{"orangered4"}\NormalTok{,}\StringTok{"chartreuse4"}\NormalTok{, }\StringTok{"skyblue"}\NormalTok{, }\StringTok{"gray"}\NormalTok{))}\SpecialCharTok{+}
+  \FunctionTok{scale\_x\_continuous}\NormalTok{(}\AttributeTok{breaks =} \FunctionTok{seq}\NormalTok{(}\FloatTok{0.3}\NormalTok{,}\FloatTok{1.5}\NormalTok{, }\AttributeTok{by =} \FloatTok{0.1}\NormalTok{))}\SpecialCharTok{+}\FunctionTok{xlab}\NormalTok{(}\StringTok{"cosine distance"}\NormalTok{)}\SpecialCharTok{+}
+  \FunctionTok{scale\_color\_manual}\NormalTok{(}\AttributeTok{values =} \FunctionTok{c}\NormalTok{(}\StringTok{"orangered4"}\NormalTok{,}\StringTok{"chartreuse4"}\NormalTok{,}\StringTok{"skyblue"}\NormalTok{,}\StringTok{"gray"}\NormalTok{))}\SpecialCharTok{+}
+  \FunctionTok{theme\_tufte}\NormalTok{()}\SpecialCharTok{+}\FunctionTok{ggtitle}\NormalTok{(}\StringTok{"Empirical distribution of distances (muslim)"}\NormalTok{)}
 \end{Highlighting}
@@
 \begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-7-1} \end{center}
-
-\normalsize
-
-\noindent Once we focus on words related to islam, the associated bias
-seems to be stronger than in the whole dataset. This is a step towards
-illustrating that the distribution of bias is uneven.
-
-Now, say we want to look at a single protected word. Since the dataset
-also contains comparison multiple control neutral and human attributes,
-we randomly select only 5 from \texttt{none} and 5 from \texttt{human}
-control groups of those for the visualisation purposes.
-
-\vspace{1mm} \footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{(tidyverse)}
-\NormalTok{muslimClass <-}\StringTok{ }\NormalTok{muslim }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(protectedWord }\OperatorTok{==}\StringTok{ "muslim"}\NormalTok{)}
-\NormalTok{neutralSample <-}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\KeywordTok{filter}\NormalTok{(muslimClass,connection }\OperatorTok{==}\StringTok{ "none"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
-\NormalTok{humanSample <-}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\KeywordTok{filter}\NormalTok{(muslimClass,connection }\OperatorTok{==}\StringTok{ "human"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
-\NormalTok{muslimVis <-}\StringTok{ }\NormalTok{muslimClass }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(connection }\OperatorTok{!=}\StringTok{ "none"} \OperatorTok{&}\StringTok{ }\NormalTok{connection }\OperatorTok{!=}\StringTok{"human"}\NormalTok{)}
-\NormalTok{muslimVis <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(muslimVis,neutralSample,humanSample)}
-
-\CommentTok{#we plug in our visualisation script}
-\KeywordTok{source}\NormalTok{(}\StringTok{"../functions/visualisationTools.R"}\NormalTok{)}
-\CommentTok{#two arguments: dataset and protected word}
-\KeywordTok{visualiseProtected}\NormalTok{(muslimVis,}\StringTok{"muslim"}\NormalTok{)}
+\normalsize
+
+\noindent Once we focus on words related to islam, the associated bias seems to be stronger than in the whole dataset. This is a step towards illustrating that the distribution of bias is uneven.
+
+Now, say we want to look at a single protected word.
+Since the dataset also contains comparison multiple control neutral and human attributes, we randomly select only 5 from \texttt{none} and 5 from \texttt{human} control groups of those for the visualisation purposes.
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{library}\NormalTok{(tidyverse)}
+\NormalTok{muslimClass }\OtherTok{\textless{}{-}}\NormalTok{ muslim }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(protectedWord }\SpecialCharTok{==} \StringTok{"muslim"}\NormalTok{)}
+\NormalTok{neutralSample }\OtherTok{\textless{}{-}} \FunctionTok{sample\_n}\NormalTok{(}\FunctionTok{filter}\NormalTok{(muslimClass,connection }\SpecialCharTok{==} \StringTok{"none"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
+\NormalTok{humanSample }\OtherTok{\textless{}{-}} \FunctionTok{sample\_n}\NormalTok{(}\FunctionTok{filter}\NormalTok{(muslimClass,connection }\SpecialCharTok{==} \StringTok{"human"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
+\NormalTok{muslimVis }\OtherTok{\textless{}{-}}\NormalTok{ muslimClass }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(connection }\SpecialCharTok{!=} \StringTok{"none"} \SpecialCharTok{\&}\NormalTok{ connection }\SpecialCharTok{!=}\StringTok{"human"}\NormalTok{)}
+\NormalTok{muslimVis }\OtherTok{\textless{}{-}} \FunctionTok{rbind}\NormalTok{(muslimVis,neutralSample,humanSample)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning in `[<-.factor`(`*tmp*`, ri, value = structure(c("muslim_stereotype", :
+## invalid factor level, NA generated
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#we plug in our visualisation script}
+\FunctionTok{source}\NormalTok{(}\StringTok{"../functions/visualisationTools.R"}\NormalTok{)}
+\CommentTok{\#two arguments: dataset and protected word}
+\FunctionTok{visualiseProtected}\NormalTok{(muslimVis,}\StringTok{"muslim"}\NormalTok{)}
 \end{Highlighting}
@@
 \begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/tableMuslimActive-1} \end{center}
-
-\normalsize
-
-Note that the distance between the grey point and the other points is
-proportional to cosine distance, the non-grey point size is proportional
-to cosine similarity to the protected word, and color groups by the
-connection type. So for \texttt{muslim} it seems that the stereotypes
-coming from the word list are fairly well visible. To give you some
-taste of how uneven the dataset is, compare this to what happens with
-\texttt{priest}.
-
-\vspace{1mm} \footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{(tidyverse)}
-\NormalTok{priestClass <-}\StringTok{ }\NormalTok{religion }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(protectedWord }\OperatorTok{==}\StringTok{ "priest"}\NormalTok{)}
-\NormalTok{neutralSample <-}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\KeywordTok{filter}\NormalTok{(priestClass,connection }\OperatorTok{==}\StringTok{ "none"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
-\NormalTok{humanSample <-}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\KeywordTok{filter}\NormalTok{(priestClass,connection }\OperatorTok{==}\StringTok{ "human"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
-\NormalTok{priestVis <-}\StringTok{ }\NormalTok{priestClass }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(connection }\OperatorTok{!=}\StringTok{ "none"} \OperatorTok{&}\StringTok{ }\NormalTok{connection }\OperatorTok{!=}\StringTok{"human"}\NormalTok{)}
-\NormalTok{priestVis <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(priestVis,neutralSample,humanSample)}
-
-\CommentTok{#we plug in our visualisation script}
-\KeywordTok{source}\NormalTok{(}\StringTok{"../functions/visualisationTools.R"}\NormalTok{)}
-\CommentTok{#two arguments: dataset and protected word}
-\KeywordTok{visualiseProtected}\NormalTok{(priestVis,}\StringTok{"priest"}\NormalTok{)}
+\normalsize
+
+Note that the distance between the grey point and the other points is proportional to cosine distance, the non-grey point size is proportional to cosine similarity to the protected word, and color groups by the connection type.
+So for \texttt{muslim} it seems that the stereotypes coming from the word list are fairly well visible.
+To give you some taste of how uneven the dataset is, compare this to what happens with \texttt{priest}.
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{library}\NormalTok{(tidyverse)}
+\NormalTok{priestClass }\OtherTok{\textless{}{-}}\NormalTok{ religion }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(protectedWord }\SpecialCharTok{==} \StringTok{"priest"}\NormalTok{)}
+\NormalTok{neutralSample }\OtherTok{\textless{}{-}} \FunctionTok{sample\_n}\NormalTok{(}\FunctionTok{filter}\NormalTok{(priestClass,connection }\SpecialCharTok{==} \StringTok{"none"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
+\NormalTok{humanSample }\OtherTok{\textless{}{-}} \FunctionTok{sample\_n}\NormalTok{(}\FunctionTok{filter}\NormalTok{(priestClass,connection }\SpecialCharTok{==} \StringTok{"human"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
+\NormalTok{priestVis }\OtherTok{\textless{}{-}}\NormalTok{ priestClass }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(connection }\SpecialCharTok{!=} \StringTok{"none"} \SpecialCharTok{\&}\NormalTok{ connection }\SpecialCharTok{!=}\StringTok{"human"}\NormalTok{)}
+\NormalTok{priestVis }\OtherTok{\textless{}{-}} \FunctionTok{rbind}\NormalTok{(priestVis,neutralSample,humanSample)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning in `[<-.factor`(`*tmp*`, ri, value = structure(c("muslim_stereotype", :
+## invalid factor level, NA generated
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#we plug in our visualisation script}
+\FunctionTok{source}\NormalTok{(}\StringTok{"../functions/visualisationTools.R"}\NormalTok{)}
+\CommentTok{\#two arguments: dataset and protected word}
+\FunctionTok{visualiseProtected}\NormalTok{(priestVis,}\StringTok{"priest"}\NormalTok{)}
 \end{Highlighting}
@@
 \begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-8-1} \end{center}
-
-\normalsize
-
-\noindent Here you can see that some human attributes are closer than
-stereotype attributes, and that there is no clear reason to claim that
-\texttt{associated} attributes are closer than \texttt{different} or
-\texttt{human} attributes. This, again, illustrates the need of
-case-by-case analysis with control groups.
-
-The general idea now is that the word lists provided in different pieces
-of research are just samples of attributes associates with various
-stereotypes and should be treated as such: the uncertaintly involved and
-the sample sizes should have clear impact on our estimates.
-
-\section{Bayesian model structure and
-assumptions}\label{bayesian-model-structure-and-assumptions}
-
-We will now think of cosine distance as the output variable, and will
-build a few bayesian models to compare. First, we just build a baseline
-model which estimates cosine distance to the attributes separately for
-each protected word. The underlying idea is that different protected
-words migh in general have different relations to all the attributes and
-this relations should be our point of departure.
+\normalsize
+
+\noindent Here you can see that some human attributes are closer than stereotype attributes, and that there is no clear reason to claim that \texttt{associated} attributes are closer than \texttt{different} or \texttt{human} attributes.
+This, again, illustrates the need of case-by-case analysis with control groups.
+
+The general idea now is that the word lists provided in different pieces of research are just samples of attributes associates with various stereotypes and should be treated as such: the uncertaintly involved and the sample sizes should have clear impact on our estimates.
+
+\hypertarget{bayesian-model-structure-and-assumptions}{%
+\section{Bayesian model structure and assumptions}\label{bayesian-model-structure-and-assumptions}}
+
+We will now think of cosine distance as the output variable, and
+will build a few bayesian models to compare.
+First, we just build a baseline model which estimates cosine distance to the attributes separately for each protected word.
+The underlying idea is that different protected words migh in general have different relations to all the attributes and this relations should be our point of departure.
 
 Here is the intution behind the mathematical Bayesian model involved.
-Our outcome variable is \texttt{cosine\ difference}, which we take to me
-normally distributed around the predicted mean for a given protected
-word (that is, we assume the residuals are normally distributed). The
-simplest model specification is:
+Our outcome variable is \texttt{cosine\ difference}, which we take to me normally distributed around the predicted mean for a given protected word (that is, we assume the residuals are normally distributed).
+The simplest model specification is:
 
@@
 
-That is, we assume the estimated means might be different for diferent
-protected words and our prior for the mean and the overal standard
-deviation are normal with mean 1 and sd=.5 and half-cauchy with
-parameters \texttt{0,1}. Further on we'll also estimate additional
-impact the connection type may have. For this impact we take a slightly
-skeptical prior centered around 0 distributed normally with sd = 1.
-These are fairly weak and slightly skeptical regularizing priors, which
-can be illustrated as follows:
+That is, we assume the estimated means might be different for diferent protected words and our prior for the mean and the overal standard deviation are normal with mean 1 and sd=.5 and half-cauchy with parameters \texttt{0,1}.
+Further on we'll also estimate additional impact the connection type may have.
+For this impact we take a slightly skeptical prior centered around 0 distributed normally with sd = 1.
+These are fairly weak and slightly skeptical regularizing priors, which can be illustrated as follows:
 
@@
 
-\section{Choosing predictors}\label{choosing-predictors}
-
-Now we can define and compile the baseline model. Its parameters will
-have a posterior distribution obtained using either Hamiltionian Monte
-Carlo methods (STAN) available through the \texttt{rethinking} package.
-
-\vspace{1mm} \footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{(rethinking)}
-\KeywordTok{options}\NormalTok{(}\DataTypeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\OtherTok{TRUE}\NormalTok{ )}
-\NormalTok{religionBaseline <-}\StringTok{ }\KeywordTok{ulam}\NormalTok{(}
-  \KeywordTok{alist}\NormalTok{(}
-\NormalTok{    cosineDistance }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(mu,sigma),}
-\NormalTok{    mu <-}\StringTok{ }\NormalTok{m[pw],}
-\NormalTok{    m[pw] }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
-\NormalTok{    sigma }\OperatorTok{~}\StringTok{ }\KeywordTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\hypertarget{choosing-predictors}{%
+\section{Choosing predictors}\label{choosing-predictors}}
+
+Now we can define and compile the baseline model.
+Its parameters will have a posterior distribution obtained using either Hamiltionian Monte Carlo methods (STAN) available through the \texttt{rethinking} package.
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{library}\NormalTok{(rethinking)}
+\FunctionTok{options}\NormalTok{(}\AttributeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\ConstantTok{TRUE}\NormalTok{ )}
+\NormalTok{religionBaseline }\OtherTok{\textless{}{-}} \FunctionTok{ulam}\NormalTok{(}
+  \FunctionTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu }\OtherTok{\textless{}{-}}\NormalTok{ m[pw],}
+\NormalTok{    m[pw] }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
 \NormalTok{  ),}
-  \DataTypeTok{data =}\NormalTok{ religion,}
-  \DataTypeTok{chains=}\DecValTok{2}\NormalTok{ , }\DataTypeTok{iter=}\DecValTok{4000}\NormalTok{ , }\DataTypeTok{warmup=}\DecValTok{1000}\NormalTok{,}
-  \DataTypeTok{start=} \KeywordTok{list}\NormalTok{(}\DataTypeTok{mu =} \DecValTok{1}\NormalTok{, }\DataTypeTok{co =} \DecValTok{0}\NormalTok{, }\DataTypeTok{sigma=} \FloatTok{.3}\NormalTok{),}
-  \DataTypeTok{log_lik =} \OtherTok{TRUE}\NormalTok{, }\DataTypeTok{cores=}\DecValTok{4}
+  \AttributeTok{data =}\NormalTok{ religion,}
+  \AttributeTok{chains=}\DecValTok{2}\NormalTok{ , }\AttributeTok{iter=}\DecValTok{4000}\NormalTok{ , }\AttributeTok{warmup=}\DecValTok{1000}\NormalTok{,}
+  \AttributeTok{start=} \FunctionTok{list}\NormalTok{(}\AttributeTok{mu =} \DecValTok{1}\NormalTok{, }\AttributeTok{co =} \DecValTok{0}\NormalTok{, }\AttributeTok{sigma=}\NormalTok{ .}\DecValTok{3}\NormalTok{),}
+  \AttributeTok{log\_lik =} \ConstantTok{TRUE}\NormalTok{, }\AttributeTok{cores=}\DecValTok{4}
 \NormalTok{)}
-\CommentTok{#saving}
-\CommentTok{#saveRDS(religionBaseline, }
-\CommentTok{#file = "cosineAnalysis/models/religionBaseline.rds")}
-\end{Highlighting}
-\end{Shaded}
-
-The only reason we need it is the evaluation of connection as a
-predictor. Does including it in o the model improve the situation? To
-investigate this, let's now build a model according to the following
-specification:
+\CommentTok{\#saving}
+\CommentTok{\#saveRDS(religionBaseline, }
+\CommentTok{\#file = "cosineAnalysis/models/religionBaseline.rds")}
+\end{Highlighting}
+\end{Shaded}
+
+The only reason we need it is the evaluation of connection as a predictor.
+Does including it in o the model improve the situation?
+To investigate this, let's now build a model according to the following specification:
 
@@
 
-\noindent The idea now is that each connection type comes with its own
-coefficient \(co\) that has impact on mean distances for protected words
-taken separately.
-
-\vspace{1mm} \footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{(rethinking)}
-\KeywordTok{options}\NormalTok{(}\DataTypeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\OtherTok{TRUE}\NormalTok{ )}
-\NormalTok{religionCoefs <-}\StringTok{ }\KeywordTok{ulam}\NormalTok{(}
-  \KeywordTok{alist}\NormalTok{(}
-\NormalTok{    cosineDistance }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(mu,sigma),}
-\NormalTok{    mu <-}\StringTok{ }\NormalTok{m[pw] }\OperatorTok{+}\StringTok{ }\NormalTok{co[con],}
-\NormalTok{    m[pw] }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
-\NormalTok{    co[con] }\OperatorTok{~}\KeywordTok{dnorm}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
-\NormalTok{    sigma }\OperatorTok{~}\StringTok{ }\KeywordTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\noindent The idea now is that each connection type comes with its own coefficient \(co\) that has impact on mean distances for protected words taken separately.
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{library}\NormalTok{(rethinking)}
+\FunctionTok{options}\NormalTok{(}\AttributeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\ConstantTok{TRUE}\NormalTok{ )}
+\NormalTok{religionCoefs }\OtherTok{\textless{}{-}} \FunctionTok{ulam}\NormalTok{(}
+  \FunctionTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu }\OtherTok{\textless{}{-}}\NormalTok{ m[pw] }\SpecialCharTok{+}\NormalTok{ co[con],}
+\NormalTok{    m[pw] }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    co[con] }\SpecialCharTok{\textasciitilde{}}\FunctionTok{dnorm}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
 \NormalTok{  ),}
-  \DataTypeTok{data =}\NormalTok{ religion,}
-  \DataTypeTok{chains=}\DecValTok{2}\NormalTok{ , }\DataTypeTok{iter=}\DecValTok{8000}\NormalTok{ , }\DataTypeTok{warmup=}\DecValTok{1000}\NormalTok{, }
-  \DataTypeTok{log_lik =} \OtherTok{TRUE}
+  \AttributeTok{data =}\NormalTok{ religion,}
+  \AttributeTok{chains=}\DecValTok{2}\NormalTok{ , }\AttributeTok{iter=}\DecValTok{8000}\NormalTok{ , }\AttributeTok{warmup=}\DecValTok{1000}\NormalTok{, }
+  \AttributeTok{log\_lik =} \ConstantTok{TRUE}
 \NormalTok{)}
@@
 
-\noindent First, let's see if this model is really better in terms of
-the Widely Acceptable Information Criterion (WAIC):
-
-\vspace{1mm} \footnotesize
+\noindent First, let's see if this model is really better in terms of the Widely Acceptable Information Criterion (WAIC):
+
+\vspace{1mm}
+\footnotesize
 
@@
 
-Clearly, it should be given weight 1 as compared to the baseline model.
-So far, we've learned that the connection type actually has predictive
-value. Let's take a look at the coefficient estimates:
-
-\vspace{1mm} \footnotesize
+Clearly, it should be given weight 1 as compared to the baseline model. So far, we've learned that the connection type actually has predictive value. Let's take a look at the coefficient estimates:
+
+\vspace{1mm}
+\footnotesize
 
@@
 
-\section{Dataset-level coefficients}\label{dataset-level-coefficients}
-
-\noindent Let's plot them together with their highest posterior density
-invervals, for the three topic groups.
+\hypertarget{dataset-level-coefficients}{%
+\section{Dataset-level coefficients}\label{dataset-level-coefficients}}
+
+\noindent Let's plot them together with their highest posterior density invervals, for the three topic groups.
 
@@
 
-\noindent What should strike us is that while the mean estimates of the
-coefficients indeed do differ a bit, usually the highest posterior
-density invervals all include zero, and so we do not have strong reasons
-to say that, say, as far as the whole religion dataset is involved,
-being associated indeed is connected with lower cosine distance. A
-second striking observation is that the estimated impact for associated
-stereotypes is quite often not too different from the estimated impact
-of attributes associated with different stereotypes, and both are
-sometimes not too far from the estimated impact for simply human
-attributes. In general, once the uncertainty involved is taken seriously
-by using control groups and statistical uncertainty estimation that does
-not dispose of pointwise data, the picture which focuses only on
-differences between means of means is too simplistic.
-
-But this doesn't mean important differences for some protected words are
-not there. For one thing, if you start with a word list that is very
-uneven, the actually not so bad status of some of the protected words
-might mask a pretty bad situation in which some other protected words
-are. For comparison, let's see what a model focused on words related to
-islam tells us.
-
-\vspace{1mm} \footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{#this is how we build the model}
-\NormalTok{religion <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"cosineAnalysis/datasets/religionReddit.csv"}\NormalTok{)[}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
-\KeywordTok{colnames}\NormalTok{(religion) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
+\noindent What should strike us is that while the mean estimates of the coefficients indeed do differ a bit, usually the highest posterior density invervals all include zero, and so we do not have strong reasons to say that, say, as far as the whole religion dataset is involved, being associated indeed is connected with lower cosine distance.
+A second striking observation is that the estimated impact for associated stereotypes is quite often not too different from the estimated impact of attributes associated with different stereotypes, and both are sometimes not too far from the estimated impact for simply human attributes.
+In general, once the uncertainty involved is taken seriously by using control groups and statistical uncertainty estimation that does not dispose of pointwise data, the picture which focuses only on differences between means of means is too simplistic.
+
+But this doesn't mean important differences for some protected words are not there.
+For one thing, if you start with a word list that is very uneven, the actually not so bad status of some of the protected words might mask a pretty bad situation in which some other protected words are.
+For comparison, let's see what a model focused on words related to islam tells us.
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#this is how we build the model}
+\NormalTok{religion }\OtherTok{\textless{}{-}} \FunctionTok{read.csv}\NormalTok{(}\StringTok{"cosineAnalysis/datasets/religionReddit.csv"}\NormalTok{)[}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{]}
+\FunctionTok{colnames}\NormalTok{(religion) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
                         \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
-\KeywordTok{levels}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
-\NormalTok{muslimWords <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"imam"}\NormalTok{,}\StringTok{"islam"}\NormalTok{,}\StringTok{"mosque"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"quran"}\NormalTok{)}
-\NormalTok{muslim <-}\StringTok{ }\NormalTok{religion }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(protectedWord }\OperatorTok{%in%}\StringTok{ }\NormalTok{muslimWords)}
-\NormalTok{muslim}\OperatorTok{$}\NormalTok{protectedWord <-}\StringTok{ }\KeywordTok{droplevels}\NormalTok{(muslim}\OperatorTok{$}\NormalTok{protectedWord)}
-\NormalTok{muslim}\OperatorTok{$}\NormalTok{pw <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(muslim}\OperatorTok{$}\NormalTok{protectedWord)}
-\NormalTok{muslim}\OperatorTok{$}\NormalTok{con <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(muslim}\OperatorTok{$}\NormalTok{connection)}
-\NormalTok{muslim}\OperatorTok{$}\NormalTok{pwFactor <-}\StringTok{ }\KeywordTok{factor}\NormalTok{(}\KeywordTok{paste0}\NormalTok{(muslim}\OperatorTok{$}\NormalTok{protectedWord, muslim}\OperatorTok{$}\NormalTok{connection))}
-\NormalTok{muslim}\OperatorTok{$}\NormalTok{pwIndex <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(muslim}\OperatorTok{$}\NormalTok{pwFactor)}
-
-\NormalTok{islamCoefs <-}\StringTok{ }\KeywordTok{ulam}\NormalTok{(}
-  \KeywordTok{alist}\NormalTok{(}
-\NormalTok{    cosineDistance }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(mu,sigma),}
-\NormalTok{    mu <-}\StringTok{ }\NormalTok{m[pw] }\OperatorTok{+}\StringTok{ }\NormalTok{co[con],}
-\NormalTok{    m[pw] }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
-\NormalTok{    co[con] }\OperatorTok{~}\KeywordTok{dnorm}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
-\NormalTok{    sigma }\OperatorTok{~}\StringTok{ }\KeywordTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\FunctionTok{levels}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{wordClass) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
+\NormalTok{muslimWords }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"imam"}\NormalTok{,}\StringTok{"islam"}\NormalTok{,}\StringTok{"mosque"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"quran"}\NormalTok{)}
+\NormalTok{muslim }\OtherTok{\textless{}{-}}\NormalTok{ religion }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(protectedWord }\SpecialCharTok{\%in\%}\NormalTok{ muslimWords)}
+\NormalTok{muslim}\SpecialCharTok{$}\NormalTok{protectedWord }\OtherTok{\textless{}{-}} \FunctionTok{droplevels}\NormalTok{(muslim}\SpecialCharTok{$}\NormalTok{protectedWord)}
+\NormalTok{muslim}\SpecialCharTok{$}\NormalTok{pw }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(muslim}\SpecialCharTok{$}\NormalTok{protectedWord)}
+\NormalTok{muslim}\SpecialCharTok{$}\NormalTok{con }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(muslim}\SpecialCharTok{$}\NormalTok{connection)}
+\NormalTok{muslim}\SpecialCharTok{$}\NormalTok{pwFactor }\OtherTok{\textless{}{-}} \FunctionTok{factor}\NormalTok{(}\FunctionTok{paste0}\NormalTok{(muslim}\SpecialCharTok{$}\NormalTok{protectedWord, muslim}\SpecialCharTok{$}\NormalTok{connection))}
+\NormalTok{muslim}\SpecialCharTok{$}\NormalTok{pwIndex }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(muslim}\SpecialCharTok{$}\NormalTok{pwFactor)}
+
+\NormalTok{islamCoefs }\OtherTok{\textless{}{-}} \FunctionTok{ulam}\NormalTok{(}
+  \FunctionTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu }\OtherTok{\textless{}{-}}\NormalTok{ m[pw] }\SpecialCharTok{+}\NormalTok{ co[con],}
+\NormalTok{    m[pw] }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    co[con] }\SpecialCharTok{\textasciitilde{}}\FunctionTok{dnorm}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
 \NormalTok{  ),}
-  \DataTypeTok{data =}\NormalTok{ muslim,}
-  \DataTypeTok{chains=}\DecValTok{2}\NormalTok{ , }\DataTypeTok{iter=}\DecValTok{10000}\NormalTok{ , }\DataTypeTok{warmup=}\DecValTok{1000}\NormalTok{, }\DataTypeTok{cores =} \DecValTok{4}\NormalTok{,}
-  \DataTypeTok{log_lik =} \OtherTok{TRUE}
+  \AttributeTok{data =}\NormalTok{ muslim,}
+  \AttributeTok{chains=}\DecValTok{2}\NormalTok{ , }\AttributeTok{iter=}\DecValTok{10000}\NormalTok{ , }\AttributeTok{warmup=}\DecValTok{1000}\NormalTok{, }\AttributeTok{cores =} \DecValTok{4}\NormalTok{,}
+  \AttributeTok{log\_lik =} \ConstantTok{TRUE}
 \NormalTok{)}
@@
 
-\vspace{1mm} \footnotesize
+\vspace{1mm}
+\footnotesize
 
@@
 
-While muslim words were unusual in the sense that the disparity between
-associated attributes and others is stronger, the evidence is still not
-conclusive. This is because the variation even within islam-related
-words is large enough (and sample sizes sufficiently small) for all the
-highest posterior density invervals to still include zeros.
-
-So, it seems, taking a closer look does seem to make a difference. The
-question is, what happens if we do take a close look at the level of
-protected words?
-
-\chapter{Protected-word level
-analysis}\label{protected-word-level-analysis}
-
-\section{Model structure and
-assumptions}\label{model-structure-and-assumptions}
-
-Let's turn then to data analysis that takes a look at protected words
-separately. This time for each combination of a protected word and a
-connection status we will have a separate mean cosine distance estimate,
-each coming with its own highest posterior density interval. This means
-we will use indices that are result from all such combinations (and then
-we will split them up in the model precis to build visualisation, feel
-free to look at the \texttt{visualiseStats.R} script for details).
-
-\vspace{1mm} \footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{options}\NormalTok{(}\DataTypeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\OtherTok{TRUE}\NormalTok{ ) }\CommentTok{#removes install pop-up request}
-\NormalTok{religion}\OperatorTok{$}\NormalTok{pwFactor <-}\StringTok{ }\KeywordTok{factor}\NormalTok{(}\KeywordTok{paste0}\NormalTok{(religion}\OperatorTok{$}\NormalTok{protectedWord, }\StringTok{"-"}\NormalTok{, religion}\OperatorTok{$}\NormalTok{connection))}
-\NormalTok{religion}\OperatorTok{$}\NormalTok{pwIndex <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(religion}\OperatorTok{$}\NormalTok{pwFactor)}
-
-\NormalTok{religionSeparate <-}\StringTok{ }\KeywordTok{ulam}\NormalTok{(}
-  \KeywordTok{alist}\NormalTok{(}
-\NormalTok{    cosineDistance }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(mu,sigma),}
-\NormalTok{    mu <-}\StringTok{ }\NormalTok{c[pwIndex],}
-\NormalTok{    c[pwIndex] }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
-\NormalTok{    sigma }\OperatorTok{~}\StringTok{ }\KeywordTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+While muslim words were unusual in the sense that the disparity between associated attributes and others is stronger, the evidence is still not conclusive.
+This is because the variation even within islam-related words is large enough (and sample sizes sufficiently small) for all the highest posterior density invervals to still include zeros.
+
+So, it seems, taking a closer look does seem to make a difference. The question is, what happens if we do take a close look at the level of protected words?
+
+\hypertarget{protected-word-level-analysis}{%
+\chapter{Protected-word level analysis}\label{protected-word-level-analysis}}
+
+\hypertarget{model-structure-and-assumptions}{%
+\section{Model structure and assumptions}\label{model-structure-and-assumptions}}
+
+Let's turn then to data analysis that takes a look at protected words separately.
+This time for each combination of a protected word and a connection status we will have a separate mean cosine distance estimate, each coming with its own highest posterior density interval.
+This means we will use indices that are result from all such combinations (and then we will split them up in the model precis to build visualisation, feel free to look at the \texttt{visualiseStats.R} script for details).
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{options}\NormalTok{(}\AttributeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\ConstantTok{TRUE}\NormalTok{ ) }\CommentTok{\#removes install pop{-}up request}
+\NormalTok{religion}\SpecialCharTok{$}\NormalTok{pwFactor }\OtherTok{\textless{}{-}} \FunctionTok{factor}\NormalTok{(}\FunctionTok{paste0}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{protectedWord, }\StringTok{"{-}"}\NormalTok{, religion}\SpecialCharTok{$}\NormalTok{connection))}
+\NormalTok{religion}\SpecialCharTok{$}\NormalTok{pwIndex }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{pwFactor)}
+
+\NormalTok{religionSeparate }\OtherTok{\textless{}{-}} \FunctionTok{ulam}\NormalTok{(}
+  \FunctionTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu }\OtherTok{\textless{}{-}}\NormalTok{ c[pwIndex],}
+\NormalTok{    c[pwIndex] }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
 \NormalTok{  ),}
-  \DataTypeTok{data =}\NormalTok{ religion,}
-  \DataTypeTok{chains=}\DecValTok{2}\NormalTok{ , }\DataTypeTok{iter=}\DecValTok{10000}\NormalTok{ , }\DataTypeTok{warmup=}\DecValTok{1000}\NormalTok{,}
-  \DataTypeTok{start=}\KeywordTok{list}\NormalTok{(}\DataTypeTok{no =} \DecValTok{1}\NormalTok{, }\DataTypeTok{a =} \DecValTok{0}\NormalTok{, }\DataTypeTok{d =} \DecValTok{0}\NormalTok{, }\DataTypeTok{sigma=} \FloatTok{.3}\NormalTok{), }\DataTypeTok{log_lik =}  \OtherTok{TRUE}
+  \AttributeTok{data =}\NormalTok{ religion,}
+  \AttributeTok{chains=}\DecValTok{2}\NormalTok{ , }\AttributeTok{iter=}\DecValTok{10000}\NormalTok{ , }\AttributeTok{warmup=}\DecValTok{1000}\NormalTok{,}
+  \AttributeTok{start=}\FunctionTok{list}\NormalTok{(}\AttributeTok{no =} \DecValTok{1}\NormalTok{, }\AttributeTok{a =} \DecValTok{0}\NormalTok{, }\AttributeTok{d =} \DecValTok{0}\NormalTok{, }\AttributeTok{sigma=}\NormalTok{ .}\DecValTok{3}\NormalTok{), }\AttributeTok{log\_lik =}  \ConstantTok{TRUE}
 \NormalTok{)}
@@
 
-\noindent Let's see if the individualized model does better than the
-previous models in light of WAIC which does add penalty for the number
-of parameters.
-
-\vspace{1mm} \footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{compareBaselineCoefsSeparate<-}\StringTok{ }\KeywordTok{readRDS}\NormalTok{(}\StringTok{"../datasets/compareBaselineCoefsSeparate.rds"}\NormalTok{)}
+\noindent Let's see if the individualized model does better than the previous models in light of WAIC which does add penalty for the number of parameters.
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{compareBaselineCoefsSeparate}\OtherTok{\textless{}{-}} \FunctionTok{readRDS}\NormalTok{(}\StringTok{"../datasets/compareBaselineCoefsSeparate.rds"}\NormalTok{)}
 \NormalTok{compareBaselineCoefsSeparate}
@@
 
-\section{Protected classes in Reddit and Google
-embeddings}\label{protected-classes-in-reddit-and-google-embeddings}
-
-It seems that we do want to prefer this model, despite its relative
-complication. Now, what does it tell us about the protected words? Let's
-visualise the predicted means together with 89\% highest posterior
-density intervals.
+\hypertarget{protected-classes-in-reddit-and-google-embeddings}{%
+\section{Protected classes in Reddit and Google embeddings}\label{protected-classes-in-reddit-and-google-embeddings}}
+
+It seems that we do want to prefer this model, despite its relative complication.
+Now, what does it tell us about the protected words?
+Let's visualise the predicted means together with 89\% highest posterior density intervals.
 
@@
 
-Before we move on, let's perform analogous analyses for the remaining
-types of supposed bias: gender and race (the model building is
-analogous).
+Before we move on, let's perform analogous analyses for the remaining types of supposed bias: gender and race (the model building is analogous).
 
@@
 
-\chapter{The role of debiasing}\label{the-role-of-debiasing}
+\hypertarget{the-role-of-debiasing}{%
+\chapter{The role of debiasing}\label{the-role-of-debiasing}}
 
@@
 
-\section{Dataset-level coefficients after
-debiasing}\label{dataset-level-coefficients-after-debiasing}
-
-First, let's look at coefficient estimated for the whole datasets, as
-compared to their estimation prior to debiasing:
+\hypertarget{dataset-level-coefficients-after-debiasing}{%
+\section{Dataset-level coefficients after debiasing}\label{dataset-level-coefficients-after-debiasing}}
+
+First, let's look at coefficient estimated for the whole datasets, as compared to their estimation prior to debiasing:
 
@@
 \item
-  change in Gender is really minor, still zero out of HPDI range, note
-  how different is very similar
+  change in Gender is really minor, still zero out of HPDI range, note how different is very similar
 \item
@@
 
-\section{Protected classes after
-debiasing}\label{protected-classes-after-debiasing}
-
-Now, perhaps, the effects of debiasing will be better appreciated if we
-look at the level of protected words. After all, the hope is, the
-situation of extremely ill-positioned protected words have improved?
+\hypertarget{protected-classes-after-debiasing}{%
+\section{Protected classes after debiasing}\label{protected-classes-after-debiasing}}
+
+Now, perhaps, the effects of debiasing will be better appreciated if we look at the level of protected words.
+After all, the hope is, the situation of extremely ill-positioned protected words have improved?
 
@@
 \hypertarget{refs}{}
-\hypertarget{ref-Bolukbasi2016Man}{}
-Bolukbasi, T., Chang, K., Zou, J. Y., Saligrama, V., \& Kalai, A.
-(2016). Man is to computer programmer as woman is to homemaker?
-Debiasing word embeddings. \emph{CoRR}, \emph{abs/1607.06520}. Retrieved
-from \url{http://arxiv.org/abs/1607.06520}
-
-\hypertarget{ref-Caliskan2017Semantics}{}
-Islam, A. C., Bryson, J. J., \& Narayanan, A. (2016). Semantics derived
-automatically from language corpora necessarily contain human biases.
-\emph{CoRR}, \emph{abs/1608.07187}. Retrieved from
-\url{http://arxiv.org/abs/1608.07187}
-
-\hypertarget{ref-manzini2019black}{}
-Manzini, T., Lim, Y. C., Tsvetkov, Y., \& Black, A. W. (2019). Black is
-to criminal as caucasian is to police: Detecting and removing multiclass
-bias in word embeddings.
-
-\hypertarget{ref-Mehrabi2019Survey}{}
+\begin{CSLReferences}{1}{0}
+\leavevmode\hypertarget{ref-Bolukbasi2016Man}{}%
+Bolukbasi, T., Chang, K.-W., Zou, J.
+Y., Saligrama, V., \& Kalai, A.
+(2016).
+Man is to computer programmer as woman is to homemaker?
+Debiasing word embeddings.
+\emph{CoRR}, \emph{abs/1607.06520}.
+Retrieved from \url{http://arxiv.org/abs/1607.06520}
+
+\leavevmode\hypertarget{ref-Caliskan2017Semantics}{}%
+Islam, A.
+C., Bryson, J.
+J., \& Narayanan, A.
+(2016).
+Semantics derived automatically from language corpora necessarily contain human biases.
+\emph{CoRR}, \emph{abs/1608.07187}.
+Retrieved from \url{http://arxiv.org/abs/1608.07187}
+
+\leavevmode\hypertarget{ref-manzini2019black}{}%
+Manzini, T., Lim, Y.
+C., Tsvetkov, Y., \& Black, A.
+W.
+(2019).
+Black is to criminal as caucasian is to police: Detecting and removing multiclass bias in word embeddings.
+Retrieved from \url{http://arxiv.org/abs/1904.04047}
+
+\leavevmode\hypertarget{ref-Mehrabi2019Survey}{}%
 Mehrabi, N., Morstatter, F., Saxena, N., Lerman, K., \& Galstyan, A.
-(2019). A survey on bias and fairness in machine learning. \emph{CoRR},
-\emph{abs/1908.09635}. Retrieved from
-\url{http://arxiv.org/abs/1908.09635}
-
-\hypertarget{ref-Nissim2019Fair}{}
-Nissim, M., Noord, R. van, \& Goot, R. van der. (2019). Fair is better
-than sensational: Man is to doctor as woman is to doctor. \emph{CoRR},
-\emph{abs/1905.09866}. Retrieved from
-\url{http://arxiv.org/abs/1905.09866}
+(2019).
+A survey on bias and fairness in machine learning.
+\emph{CoRR}, \emph{abs/1908.09635}.
+Retrieved from \url{http://arxiv.org/abs/1908.09635}
+
+\leavevmode\hypertarget{ref-Nissim2019Fair}{}%
+Nissim, M., Noord, R.
+van, \& Goot, R.
+van der.
+(2019).
+Fair is better than sensational: Man is to doctor as woman is to doctor.
+\emph{CoRR}, \emph{abs/1905.09866}.
+Retrieved from \url{http://arxiv.org/abs/1905.09866}
+
+\end{CSLReferences}
 

### 2021-06-10 paperDraft/paperDraft.Rmd
Adds

@@
 
-A similar methodology is employed by @Garg2018years, who employ word embeddings trained od corpora from different decades to study the shifts in various biases.
+A similar methodology is employed by @Garg2018years, who employ word embeddings trained on corpora from different decades to study the shifts in various biases.
 For instance, to compute the occupational embeddings bias for women the authors first compute the average vector of vector emeddings of words that represent women (e.g.

### 2021-06-10 paperDraft/paperDraft.Rmd
Paperdraft small fixes

@@
 WEAT is a modification  of the Implicit Association Test (IAT) [@Nosek2002harvesting] used in psychology and uses almost the same word sets, allowing for a \emph{prima facie} sensible comparison with bias in humans.
+If the person's attitude towards given pair of concept is to be interpreted as neutral, the final value from the formula should be around 0.
 Let $f$ be a similarity measure (usually, cosine similarity). The association difference for a term $t$ is:
@@
 @Caliskan2017semanticsBiases show that significant biases---thus measured--- similar to the ones discovered by IAT can be discovered in word embeddings.
-@Lauscher2019multidimensional extended the methodology to a multilingual and cross-lingual setting, arguing that using Euclidean distance instead of similarty does not make much difference, while the bias effects vary greatly accross embedding models (interestingly, with social media-text trained embeddings being less biased than those based on Wikipedia).
+@Lauscher2019multidimensional extended the methodology to a multilingual and cross-lingual setting, arguing that using Euclidean distance instead of similarity does not make much difference, while the bias effects vary greatly across embedding models (interestingly, with social media-text trained embeddings being less biased than those based on Wikipedia).
 
@@
 
+- The lack of control group in the paper and our analysis indicates that without neutral baseline it is difficult
+to interpret the effectiveness of the metric.
+
 \vspace{1mm}

### 2021-06-10 paperDraft/paperDraft.tex

--- 2021-06-10 paperDraft/paperDraft.tex
@@
-\documentclass[10pt,dvipsnames,enabledeprecatedfontcommands]{scrartcl}
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}
+%
+\documentclass[
+  10pt,
+  dvipsnames,enabledeprecatedfontcommands]{scrartcl}
+\usepackage{amsmath,amssymb}
 \usepackage{lmodern}
-\usepackage{amssymb,amsmath}
 \usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
 \ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
@@
 }{}
-\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
-\usepackage[unicode=true]{hyperref}
-\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
+\makeatletter
+\@ifundefined{KOMAClassName}{% if non-KOMA class
+  \IfFileExists{parskip.sty}{%
+    \usepackage{parskip}
+  }{% else
+    \setlength{\parindent}{0pt}
+    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
+}{% if KOMA class
+  \KOMAoptions{parskip=half}}
+\makeatother
+\usepackage{xcolor}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
 \hypersetup{
@@
   linkcolor=Maroon,
+  filecolor=Maroon,
   citecolor=Blue,
   urlcolor=blue,
-            breaklinks=true}
-\urlstyle{same}  % don't use monospace font for urls
+  pdfcreator={LaTeX via pandoc}}
+\urlstyle{same} % disable monospaced font for URLs
 \usepackage{color}
@@
 \newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\BuiltInTok}[1]{#1}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
 \newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
 \newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
+\newcommand{\ExtensionTok}[1]{#1}
 \newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ImportTok}[1]{#1}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\NormalTok}[1]{#1}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\RegionMarkerTok}[1]{#1}
 \newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
 \newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
 \newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\ImportTok}[1]{#1}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
-\newcommand{\BuiltInTok}[1]{#1}
-\newcommand{\ExtensionTok}[1]{#1}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
-\newcommand{\RegionMarkerTok}[1]{#1}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
 \newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
-\newcommand{\NormalTok}[1]{#1}
-\usepackage{graphicx,grffile}
+\usepackage{graphicx}
 \makeatletter
@@
 \setcounter{secnumdepth}{5}
-% Redefines (sub)paragraphs to behave more like sections
-\ifx\paragraph\undefined\else
-\let\oldparagraph\paragraph
-\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-\let\oldsubparagraph\subparagraph
-\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-
-% set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-
 %\documentclass{article}
@@
 %\bibliographystyle{apalike}
+\ifluatex
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
+  % set entry spacing
+  \ifnum #2 > 0
+  \setlength{\parskip}{#2\baselineskip}
+  \fi
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
@@
 
-test
-
+\hypertarget{cosine-based-measures-of-bias}{%
 \section{Cosine-based measures of
-bias}\label{cosine-based-measures-of-bias}
+bias}\label{cosine-based-measures-of-bias}}
 
@@
 defined as the average distance of the words in \(N\) from \(g\) (\(c\)
-is a parameter determining how strict we want to be):
-
-\begin{align}
+is a parameter determining how strict we want to be): \begin{align}
 \mathsf{directBias_c(N,g)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},g)\vert^c}{\vert N \vert }
@@
 preserve similarity under debiasing meant to minimize projection-based
-bias.\footnote{Bolukbasi et al. (2016) use also another method which
-  involves analogies and their evaluations by human users on Mechanical
-  Turk. It is discussed and criticized in (Nissim, Noord, \& Goot,
-  2020).}
+bias.\footnote{Bolukbasi, Chang, Zou, Saligrama, \& Kalai (2016) use
+  also another method which involves analogies and their evaluations by
+  human users on Mechanical Turk. It is discussed and criticized in
+  (Nissim, Noord, \& Goot, 2020).}
 
@@
 almost the same word sets, allowing for a \emph{prima facie} sensible
-comparison with bias in humans. Let \(f\) be a distance or similarity
-measure (usually, cosine similarity). The association difference for a
-term \(t\) is:
-
-\begin{align}
+comparison with bias in humans. Let \(f\) be a similarity measure
+(usually, cosine similarity). The association difference for a term
+\(t\) is: \begin{align}
 s(t,A,B) & = \frac{\sum_{a\in A}f(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}f(t,b)}{\vert B\vert}
-\end{align}
-
-\noindent then, the association difference between \(A\) a \(B\) is:
-
-\begin{align}
+\end{align} \noindent then, the association difference between \(A\) a
+\(B\) is: \begin{align}
 s(X,Y,A,B) & = \sum_{x\in X} s(x,A,B) -  \sum_{y\in Y} s(y,A,B)
-\end{align}
-
-\noindent
-\(s(X,Y,A,B)\) is the statistic used in the signifcance test, and the
-\(p\)-value obtained by bootstrapping: it is the frequency of
-\(s(X_i,Y_i,A,B)>s(X,Y,A,B)\) for all equally sized partitions
-\(X_i, Y_i\) of \(X\cup Y\). The effect size is computed by normalizing
-the difference in means as follows:
-
-\begin{align}
+\end{align} \noindent \(s(X,Y,A,B)\) is the statistic used in the
+signifcance test, and the \(p\)-value obtained by bootstrapping: it is
+the frequency of \(s(X_i,Y_i,A,B)>s(X,Y,A,B)\) for all equally sized
+partitions \(X_i, Y_i\) of \(X\cup Y\). The effect size is computed by
+normalizing the difference in means as follows: \begin{align}
 bias(A,B) & = \frac{
@@
 
-Caliskan et al. (2017) show that significant biases---thus measured---
-similar to the ones discovered by IAT can be discovered in word
-embeddings. Lauscher \& Glavas (2019) extended the methodology to a
-multilingual and cross-lingual setting, arguing that using Euclidean
-distance instead of similarty does not make much difference, while the
-bias effects vary greatly accross embedding models (interestingly, with
-social media-text trained embeddings being less biased than those based
-on Wikipedia).
+Caliskan, Bryson, \& Narayanan (2017) show that significant
+biases---thus measured--- similar to the ones discovered by IAT can be
+discovered in word embeddings. Lauscher \& Glavas (2019) extended the
+methodology to a multilingual and cross-lingual setting, arguing that
+using Euclidean distance instead of similarty does not make much
+difference, while the bias effects vary greatly accross embedding models
+(interestingly, with social media-text trained embeddings being less
+biased than those based on Wikipedia).
 
 A similar methodology is employed by Garg, Schiebinger, Jurafsky, \& Zou
-(2018), who employ word embeddings trained od corpora from different
+(2018), who employ word embeddings trained on corpora from different
 decades to study the shifts in various biases. For instance, to compute
 the occupational embeddings bias for women the authors first compute the
-average vector of vector emeddings of words that represent women (e.g.
-``she'', ``female''), then calculate the Euclidean distance between this
-mean vector and words for occupations. Then they take the mean of these
-distances and subtract from it the analogously obtained mean for the
-average vector of vector embeddings of words that represent men.
-Formally they take the relative norm distance between \(X\) and \(Y\) to
-be:
+average vector of vector emeddings of words that represent women
+(e.g.~``she,'' ``female''), then calculate the Euclidean distance
+between this mean vector and words for occupations. Then they take the
+mean of these distances and subtract from it the analogously obtained
+mean for the average vector of vector embeddings of words that represent
+men. Formally they take the relative norm distance between \(X\) and
+\(Y\) to be: \begin{align}
+\textsf{relative norm distance} & = \sum_{v_m\in M} \vert \vert v_m - v_X\vert \vert_2 - \vert v_m - v_Y\vert \vert_2
+\end{align} \noindent where the norm used is Euclidean, and \(v_X\) and
+\(x_Y\) are average vectors for sets \(X\) and \(Y\) respectively.
 
-\begin{align}
-\textsf{relative norm distance} & = \sum_{v_m\in M} \vert \vert v_m - v_X\vert \vert_2 - \vert v_m - v_Y\vert \vert_2
-\end{align}
+(\textbf{Manzini2019blackToCriminal?}) modify WEAT to a multi-class
+setting, introducing Mean Average Cosine similarity as a measure of bias
+(in fact, in the paper they report distances rather than similarities).
+Let \(T = \{t_1, \dots, t_k\}\) be a class of protected word embeddings,
+and let each \(A_j\in A\) be a set of attributes stereotypically
+associated with a protected word). Then: \begin{align}
+S(t_i, A_j) & = \frac{1}{\vert A_j\vert}\sum_{a\in A_j}\mathsf{cos}(t,a) \\
+MAC(T,A) & = \frac{1}{\vert T \vert \,\vert A\vert}\sum_{t_i \in T }\sum_{A_j \in A} S(t_i,A_j)
+\end{align} That is, for each protected word \$T and each attribute
+class, they first take the mean for this protected word and all
+attributes in a given attribute class, and then take the mean of thus
+obtained means for all the protected words. The t-tests they employ are
+run on average cosines used to calculate MAC.
 
-\noindent where the norm used is Euclidean, and \(v_X\) and \(x_Y\) are
-average vectors for sets \(X\) and \(Y\) respectively.
+\hypertarget{the-problem-with-means-of-means}{%
+\subsection{The problem with means of
+means}\label{the-problem-with-means-of-means}}
 
-There is a problem with taking means of means in the calculations,
-though. By pre-averaging data we throw away information about sample
-size. For instance, think about proportions: 10 out of 20 and 2 out of 4
-give the same mean, but you would obtain more information by making the
-former observation rather than by making the latter. Connectedly, when
-we pre-average, we remove variation, and so pre-averaging tend to
-manufacture false confidence.
+The measures described all calculate means of means and their authors
+run statistical tests on sets of means. This, however, is problematic
+for two related reasons. One, by pre-averaging data we throw away
+information about sample sizes. For the former point, think about
+proportions: 10 out of 20 and 2 out of 4 give the same mean, but you
+would obtain more information by making the former observation rather
+than by making the latter. Two, when we pre-average, we remove
+variation, and so pre-averaging tend to manufacture false confidence. We
+will have more to say about whether this happens in the case of
+applications of MAC, for now let's go over a simple example to make the
+conceptual point clear.
 
-To illustrate let's employ the formulas used by Caliskan et al. (2017)
-and go through an example. Conceptually, all such tests come up with
-rather short lists of protected words and rather short lists of
+To illustrate let's employ the formulas used by Caliskan, Bryson, \&
+Narayanan (2017) in a simple example. Conceptually, all such tests come
+up with rather short lists of protected words and rather short lists of
 stereotypical attributes. Clearly, these are not complete list. So let's
@@
 
-Consider a situation in which there are two protected classes,
-\(X=\{t_1,t_2\}\) and \(Y=\{t_3,t_4\}\) and two five-element sets \(A\)
-and \(B\).
+Consider a simple situation in which there are two protected classes,
+\(X=\{t_1,t_2\}\) and \(Y=\{t_3,t_4\}\) and two five-element attribute
+sets \(A\) and \(B\).
 
 First, we play around with a scenario in which all the protected terms
-are on average equally distant from both sets (\(\mu =1\)) with standard
-deviation of \(.05\). First, let's randomly draw distances and plot the
+are on average equally unsimilar to both sets (\(\mu =0\)) with standard
+deviation of \(.05\). Let's randomly draw similarity scores and plot the
 results with group means plotted as vertical lines.
@@
 \begin{Highlighting}[]
-\KeywordTok{set.seed}\NormalTok{(}\DecValTok{123}\NormalTok{)}
-\NormalTok{t1 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
-\NormalTok{t2 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
-\NormalTok{t3 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
-\NormalTok{t4 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\FunctionTok{set.seed}\NormalTok{(}\DecValTok{123}\NormalTok{)}
+\NormalTok{t1 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\NormalTok{t2 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\NormalTok{t3 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\NormalTok{t4 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
 \end{Highlighting}
@@
 
-When you look at the datapoints, do you have the impression of a strong
-bias here? We wouldn't think so. But now let's run the calculations from
-(Caliskan et al., 2017).
+\noindent When you look at the datapoints, do you have the impression of
+a strong bias here? We wouldn't think so. But now let's run the
+calculations from (Caliskan, Bryson, \& Narayanan, 2017).
 
-\vspace{1mm} \footnotesize
+\vspace{1mm}
+\footnotesize
 
@@
 \begin{Highlighting}[]
-\NormalTok{s <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{ (table)\{ }\KeywordTok{mean}\NormalTok{(table}\OperatorTok{$}\NormalTok{A) }\OperatorTok{-}\StringTok{ }\KeywordTok{mean}\NormalTok{(table}\OperatorTok{$}\NormalTok{B)\}}
-\NormalTok{factor <-}\StringTok{ }\KeywordTok{sd}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1),}\KeywordTok{s}\NormalTok{(t2),}\KeywordTok{s}\NormalTok{(t3),}\KeywordTok{s}\NormalTok{(t4)))}
-\NormalTok{numerator <-}\StringTok{  }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1),}\KeywordTok{s}\NormalTok{(t2)) }\OperatorTok{-}\StringTok{ }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t3),}\KeywordTok{s}\NormalTok{(t4))}
-\NormalTok{bias <-}\StringTok{ }\NormalTok{numerator }\OperatorTok{/}\StringTok{ }\NormalTok{factor}
-\NormalTok{bias}
+\NormalTok{s }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{ (table)\{ }\FunctionTok{mean}\NormalTok{(table}\SpecialCharTok{$}\NormalTok{A) }\SpecialCharTok{{-}} \FunctionTok{mean}\NormalTok{(table}\SpecialCharTok{$}\NormalTok{B)\}}
+\NormalTok{factor }\OtherTok{\textless{}{-}} \FunctionTok{sd}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1),}\FunctionTok{s}\NormalTok{(t2),}\FunctionTok{s}\NormalTok{(t3),}\FunctionTok{s}\NormalTok{(t4)))}
+\NormalTok{numerator }\OtherTok{\textless{}{-}}  \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1),}\FunctionTok{s}\NormalTok{(t2)) }\SpecialCharTok{{-}} \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t3),}\FunctionTok{s}\NormalTok{(t4))}
+\FunctionTok{print}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{factor =}\NormalTok{ factor,}\AttributeTok{numerator =}\NormalTok{ numerator, }\AttributeTok{bias =}\NormalTok{ numerator }\SpecialCharTok{/}\NormalTok{ factor))}
 \end{Highlighting}
@@
 
-This should make us pause. We know these were points randomly drawn from
-distributions where there is no difference in means. Yet, the calculated
-effect size is 1.82, whereas the largest effect size reported by
-Caliskan et al. (2017) is 1.81!
+\noindent This should make us pause. We know these were points randomly
+drawn from distributions where there is no difference in means. Yet, the
+calculated effect size is 1.82, whereas the largest effect size reported
+by Caliskan, Bryson, \& Narayanan (2017) is 1.81!
 
-Now, let's try sampling from distributions where there in fact is a
-difference in means. Terms from \(X\) are on average .8 away from \(A\)
-(and \(1\) from \(B\)), while terms from \(Y\) are \(.8\) away to \(B\)
-(and 1 to \(A\)). The standard deviation is 0.1 in all the cases. There
-is a clear difference between \(X\) and \(Y\) and quick visual
-inspection should tell us so.
+Interestingly, if we repeat the drawing 10000 times, each time
+calculating the bias, it turns out that with this variance and sample
+size, pretty much anything can happen.
 
-\vspace{1mm} \footnotesize
+\vspace{1mm}
+\footnotesize
 
@@
 \begin{Highlighting}[]
-\KeywordTok{set.seed}\NormalTok{(}\DecValTok{123}\NormalTok{)}
-\NormalTok{t1 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{8}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
-\NormalTok{t2 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{8}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
-\NormalTok{t3 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{8}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
-\NormalTok{t4 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{8}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
+\NormalTok{biasesNull }\OtherTok{\textless{}{-}} \FunctionTok{numeric}\NormalTok{(}\DecValTok{10000}\NormalTok{)}
+\ControlFlowTok{for}\NormalTok{(i }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\DecValTok{10000}\NormalTok{)\{}
+\NormalTok{t1 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\NormalTok{t2 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\NormalTok{t3 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\NormalTok{t4 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+
+\NormalTok{factor }\OtherTok{\textless{}{-}} \FunctionTok{sd}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1),}\FunctionTok{s}\NormalTok{(t2),}\FunctionTok{s}\NormalTok{(t3),}\FunctionTok{s}\NormalTok{(t4)))}
+\NormalTok{numerator }\OtherTok{\textless{}{-}}  \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1),}\FunctionTok{s}\NormalTok{(t2)) }\SpecialCharTok{{-}} \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t3),}\FunctionTok{s}\NormalTok{(t4))}
+\NormalTok{biasesNull[i]  }\OtherTok{\textless{}{-}}\NormalTok{ numerator }\SpecialCharTok{/}\NormalTok{ factor}
+\NormalTok{\}}
+\FunctionTok{ggplot}\NormalTok{()}\SpecialCharTok{+}\FunctionTok{geom\_histogram}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{biasesNull, }\AttributeTok{y =}\NormalTok{ ..density..), }\AttributeTok{alpha =} \FloatTok{0.6}\NormalTok{, }\AttributeTok{bins=}\DecValTok{50}\NormalTok{)}\SpecialCharTok{+}
+  \FunctionTok{theme\_tufte}\NormalTok{()}\SpecialCharTok{+}\FunctionTok{labs}\NormalTok{(}\AttributeTok{title=}\StringTok{"10k biases for identical means and sd =.05"}\NormalTok{)}\SpecialCharTok{+} \FunctionTok{xlab}\NormalTok{(}\StringTok{"bias"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/unnamed-chunk-4-1} \end{center}
+\normalsize
+
+Now, let's simulate a situation where the means are identical but the
+standard deviation is much smaller, .001.
+
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{set.seed}\NormalTok{(}\DecValTok{124}\NormalTok{)}
+\NormalTok{t1v }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
+\NormalTok{t2v }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
+\NormalTok{t3v }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
+\NormalTok{t4v }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
 \end{Highlighting}
@@
 
-\vspace{1mm} \footnotesize
+\begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/unnamed-chunk-6-1} \end{center}
 
-\begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/unnamed-chunk-5-1} \end{center}
+\noindent When you look at the datapoints, do you have the impression of
+a strong bias here? We wouldn't think so. But now let's run the
+calculations from (Caliskan, Bryson, \& Narayanan, 2017).
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{factorV }\OtherTok{\textless{}{-}} \FunctionTok{sd}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1v),}\FunctionTok{s}\NormalTok{(t2v),}\FunctionTok{s}\NormalTok{(t3v),}\FunctionTok{s}\NormalTok{(t4v)))}
+\NormalTok{numeratorV }\OtherTok{\textless{}{-}}  \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1v),}\FunctionTok{s}\NormalTok{(t2v)) }\SpecialCharTok{{-}} \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t3v),}\FunctionTok{s}\NormalTok{(t4v))}
+\FunctionTok{print}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{factor =}\NormalTok{ factorV,}\AttributeTok{numerator =}\NormalTok{ numeratorV, }\AttributeTok{bias =}\NormalTok{ numeratorV }\SpecialCharTok{/}\NormalTok{ factorV))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## $factor
+## [1] 0.0006402666
+## 
+## $numerator
+## [1] -0.001237637
+## 
+## $bias
+## [1] -1.933003
+\end{verbatim}
 
@@
 
-Is this clear difference mirrored in the calculations?
+While the numerator and the factors changed a lot, the bias actually
+increased. One reason bias increases is that once the standard deviation
+goes down, so does the factor used in the calculation of bias.
+
+Again, to see whether this metric would provide us with meaningful
+information, let's simulate 10000 drawings.
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{set.seed}\NormalTok{(}\DecValTok{124}\NormalTok{)}
+
+\NormalTok{biasesLowVariance }\OtherTok{\textless{}{-}} \FunctionTok{numeric}\NormalTok{(}\DecValTok{10000}\NormalTok{)}
+\ControlFlowTok{for}\NormalTok{(i }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\DecValTok{10000}\NormalTok{)\{}
+\NormalTok{t1v }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
+\NormalTok{t2v }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
+\NormalTok{t3v }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
+\NormalTok{t4v }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
+
+\NormalTok{factorV }\OtherTok{\textless{}{-}} \FunctionTok{sd}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1v),}\FunctionTok{s}\NormalTok{(t2v),}\FunctionTok{s}\NormalTok{(t3v),}\FunctionTok{s}\NormalTok{(t4v)))}
+
+\NormalTok{numeratorV }\OtherTok{\textless{}{-}}  \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1v),}\FunctionTok{s}\NormalTok{(t2v)) }\SpecialCharTok{{-}} \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t3v),}\FunctionTok{s}\NormalTok{(t4v))}
+
+\NormalTok{biasesLowVariance[i] }\OtherTok{\textless{}{-}}\NormalTok{ numeratorV }\SpecialCharTok{/}\NormalTok{ factorV}
+\NormalTok{\}}
+\FunctionTok{ggplot}\NormalTok{()}\SpecialCharTok{+}\FunctionTok{geom\_histogram}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{biasesLowVariance, }\AttributeTok{y =}\NormalTok{ ..density..), }\AttributeTok{alpha =} \FloatTok{0.6}\NormalTok{, }\AttributeTok{bins=}\DecValTok{50}\NormalTok{)}\SpecialCharTok{+}
+  \FunctionTok{theme\_tufte}\NormalTok{()}\SpecialCharTok{+}\FunctionTok{labs}\NormalTok{(}\AttributeTok{title=}\StringTok{"10k biases for identical means and sd =.001"}\NormalTok{)}\SpecialCharTok{+} \FunctionTok{xlab}\NormalTok{(}\StringTok{"bias"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/unnamed-chunk-8-1} \end{center}
+\normalsize
+
+Again, not so informative. Now, let's try sampling from distributions
+where there in fact is a difference in means. Terms from \(X\) are on
+average .1 similar to \(A\) (and still \(0\) to \(B\)), while terms from
+\(Y\) are \(.1\) similar to \(B\) (and 0 to \(A\)). The standard
+deviation is 0.05 in all the cases. There is a clear difference between
+\(X\) and \(Y\) and quick visual inspection should tell us so.
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{set.seed}\NormalTok{(}\DecValTok{766}\NormalTok{)}
+\NormalTok{t1d2 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\NormalTok{t2d2 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\NormalTok{t3d2 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\NormalTok{t4d2 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+\vspace{1mm}
+\footnotesize
+
+\begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/unnamed-chunk-10-1} \end{center}
+\normalsize
+
+\noindent  Is this clear difference mirrored in the calculations?
 
@@
 \begin{Highlighting}[]
-\NormalTok{factor <-}\StringTok{ }\KeywordTok{sd}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1),}\KeywordTok{s}\NormalTok{(t2),}\KeywordTok{s}\NormalTok{(t3),}\KeywordTok{s}\NormalTok{(t4)))}
-\NormalTok{numerator <-}\StringTok{  }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1),}\KeywordTok{s}\NormalTok{(t2)) }\OperatorTok{-}\StringTok{ }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t3),}\KeywordTok{s}\NormalTok{(t4))}
-\NormalTok{bias <-}\StringTok{ }\NormalTok{numerator }\OperatorTok{/}\StringTok{ }\NormalTok{factor}
-\NormalTok{bias}
+\NormalTok{factorD2 }\OtherTok{\textless{}{-}} \FunctionTok{sd}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1d2),}\FunctionTok{s}\NormalTok{(t2d2),}\FunctionTok{s}\NormalTok{(t3d2),}\FunctionTok{s}\NormalTok{(t4d2)))}
+\NormalTok{numeratorD2 }\OtherTok{\textless{}{-}}  \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1d2),}\FunctionTok{s}\NormalTok{(t2d2)) }\SpecialCharTok{{-}} \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t3d2),}\FunctionTok{s}\NormalTok{(t4d2))}
+\NormalTok{biasD2 }\OtherTok{\textless{}{-}}\NormalTok{ numeratorD2 }\SpecialCharTok{/}\NormalTok{ factorD2}
+\NormalTok{biasD2}
 \end{Highlighting}
@@
 
-The absolute value of the effect size is smaller than in the previous
-case!
+\noindent The absolute value of the effect size is smaller than in the
+null case with the same standard deviation. Let's simulate 10000
+drawings:
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{biasesD2 }\OtherTok{\textless{}{-}} \FunctionTok{numeric}\NormalTok{(}\DecValTok{10000}\NormalTok{)}
+\ControlFlowTok{for}\NormalTok{(i }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\DecValTok{10000}\NormalTok{)\{}
+\NormalTok{t1d2 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\NormalTok{t2d2 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\NormalTok{t3d2 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\NormalTok{t4d2 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+
+\NormalTok{factorD2 }\OtherTok{\textless{}{-}} \FunctionTok{sd}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1d2),}\FunctionTok{s}\NormalTok{(t2d2),}\FunctionTok{s}\NormalTok{(t3d2),}\FunctionTok{s}\NormalTok{(t4d2)))}
+\NormalTok{numeratorD2 }\OtherTok{\textless{}{-}}  \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1d2),}\FunctionTok{s}\NormalTok{(t2d2)) }\SpecialCharTok{{-}} \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t3d2),}\FunctionTok{s}\NormalTok{(t4d2))}
+\NormalTok{biasesD2[i] }\OtherTok{\textless{}{-}}\NormalTok{ numeratorD2}\SpecialCharTok{/}\NormalTok{factorD2}
+\NormalTok{\}}
+
+\FunctionTok{ggplot}\NormalTok{()}\SpecialCharTok{+}\FunctionTok{geom\_histogram}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{biasesD2, }\AttributeTok{y =}\NormalTok{ ..density..), }\AttributeTok{alpha =} \FloatTok{0.6}\NormalTok{, }\AttributeTok{bins=}\DecValTok{50}\NormalTok{)}\SpecialCharTok{+}
+  \FunctionTok{theme\_tufte}\NormalTok{()}\SpecialCharTok{+}\FunctionTok{labs}\NormalTok{(}\AttributeTok{title=}\StringTok{"10k biases for different means and sd =.05"}\NormalTok{)}\SpecialCharTok{+} \FunctionTok{xlab}\NormalTok{(}\StringTok{"bias"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/unnamed-chunk-12-1} \end{center}
+\normalsize
+
+\noindent Now suppose we keep the means the same but increase the
+standard deviation to .1.
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{set.seed}\NormalTok{(}\DecValTok{766}\NormalTok{)}
+\NormalTok{t1d1 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
+\NormalTok{t2d1 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
+\NormalTok{t3d1 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
+\NormalTok{t4d1 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+\vspace{1mm}
+\footnotesize
+
+\begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/unnamed-chunk-14-1} \end{center}
+\normalsize
+
+\noindent  Is this clear difference mirrored in the calculations?
+
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{factorD1 }\OtherTok{\textless{}{-}} \FunctionTok{sd}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1d1),}\FunctionTok{s}\NormalTok{(t2d1),}\FunctionTok{s}\NormalTok{(t3d1),}\FunctionTok{s}\NormalTok{(t4d1)))}
+\NormalTok{numeratorD1 }\OtherTok{\textless{}{-}}  \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1d1),}\FunctionTok{s}\NormalTok{(t2d1)) }\SpecialCharTok{{-}} \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t3d1),}\FunctionTok{s}\NormalTok{(t4d1))}
+\NormalTok{biasD1 }\OtherTok{\textless{}{-}}\NormalTok{ numeratorD1 }\SpecialCharTok{/}\NormalTok{ factorD1}
+\NormalTok{biasD1}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## [1] 1.223023
+\end{verbatim}
+
+\normalsize
+
+\noindent The absolute value of the effect size is smaller than in the
+previous case!
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{biasesD1 }\OtherTok{\textless{}{-}} \FunctionTok{numeric}\NormalTok{(}\DecValTok{10000}\NormalTok{)}
+
+\ControlFlowTok{for}\NormalTok{(i }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\DecValTok{10000}\NormalTok{)\{}
+\NormalTok{t1d1 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
+\NormalTok{t2d1 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
+\NormalTok{t3d1 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
+\NormalTok{t4d1 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
+
+\NormalTok{factorD1 }\OtherTok{\textless{}{-}} \FunctionTok{sd}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1d1),}\FunctionTok{s}\NormalTok{(t2d1),}\FunctionTok{s}\NormalTok{(t3d1),}\FunctionTok{s}\NormalTok{(t4d1)))}
+\NormalTok{numeratorD1 }\OtherTok{\textless{}{-}}  \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1d1),}\FunctionTok{s}\NormalTok{(t2d1)) }\SpecialCharTok{{-}} \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t3d1),}\FunctionTok{s}\NormalTok{(t4d1))}
+\NormalTok{biasesD1[i] }\OtherTok{\textless{}{-}}\NormalTok{ numeratorD1}\SpecialCharTok{/}\NormalTok{factorD1}
+\NormalTok{\}}
+
+\FunctionTok{ggplot}\NormalTok{()}\SpecialCharTok{+}\FunctionTok{geom\_histogram}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{biasesD1, }\AttributeTok{y =}\NormalTok{ ..density..), }\AttributeTok{alpha =} \FloatTok{0.6}\NormalTok{, }\AttributeTok{bins=}\DecValTok{50}\NormalTok{)}\SpecialCharTok{+}
+  \FunctionTok{theme\_tufte}\NormalTok{()}\SpecialCharTok{+}\FunctionTok{labs}\NormalTok{(}\AttributeTok{title=}\StringTok{"10k biases for different means and sd =.001"}\NormalTok{)}\SpecialCharTok{+} \FunctionTok{xlab}\NormalTok{(}\StringTok{"bias"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/unnamed-chunk-16-1} \end{center}
+\normalsize
+
+This is is a bit better, but still quite some uncertainty is involved,
+far from what systematically low mean-based p-values reported in the
+papers might suggest. Of course, this is a bit of a caricature, as our
+word lists were short (four protected words and 10 attributes). But the
+word lists used in the actually papers are not much longer. In such a
+set-up the key observations are as follows:
+
+\begin{itemize}
+\item
+  Seemingly high bias measures might arise even if the underlying
+  processes actually have the same mean.
+\item
+  Even if the mean remains the same, non-negligible changes can result
+  from a shift in the standard deviation of the original process, and
+  the change might go in the opposite direction than a visualisation of
+  datapoints might suggest, because with the decrease of standard
+  deviation, the factor decreases and the bias increases.
+\item
+  Even if the underlying means are the same, but the variation is
+  different, the bias metric in the long run could tend toward a
+  different value:
+\end{itemize}
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{mean}\NormalTok{(biasesD1); }\FunctionTok{mean}\NormalTok{(biasesD2)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## [1] 1.563319
+\end{verbatim}
+
+\begin{verbatim}
+## [1] 1.690664
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{median}\NormalTok{(biasesD1); }\FunctionTok{median}\NormalTok{(biasesD2)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## [1] 1.638136
+\end{verbatim}
+
+\begin{verbatim}
+## [1] 1.707511
+\end{verbatim}
+
+\normalsize
+
+and so the point estimations of bias are sensitive to factors other than
+the underlying process means.
+
+\begin{itemize}
+\item
+  Even if there is a difference in means, the bias metric can be lower,
+  and the uncertainty about it needs to be gauged.
+\item
+  The uncertainty resulting from including the raw datapoint variance
+  into considerations is more extensive than the one suggested by the
+  low p-values obtained from taking means as datapoints.
+\end{itemize}
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{quantile}\NormalTok{(biasesD2, }\AttributeTok{probs =} \FunctionTok{c}\NormalTok{(}\FloatTok{0.275}\NormalTok{,}\FloatTok{0.975}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+##    27.5%    97.5% 
+## 1.539149 2.166488
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{quantile}\NormalTok{(biasesD1, }\AttributeTok{probs =} \FunctionTok{c}\NormalTok{(}\FloatTok{0.275}\NormalTok{,}\FloatTok{0.975}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+##    27.5%    97.5% 
+## 1.300648 2.356815
+\end{verbatim}
+
+\normalsize
 
@@
 
-\section*{References}\label{references}
+\hypertarget{references}{%
+\section*{References}\label{references}}
 \addcontentsline{toc}{section}{References}
@@
 \hypertarget{refs}{}
-\hypertarget{ref-bolukbasi2016man}{}
+\begin{CSLReferences}{1}{0}
+\leavevmode\hypertarget{ref-bolukbasi2016man}{}%
 Bolukbasi, T., Chang, K.-W., Zou, J., Saligrama, V., \& Kalai, A.
 (2016). Man is to computer programmer as woman is to homemaker?
-Debiasing word embeddings.
+Debiasing word embeddings. Retrieved from
+\url{http://arxiv.org/abs/1607.06520}
 
-\hypertarget{ref-Caliskan2017semanticsBiases}{}
+\leavevmode\hypertarget{ref-Caliskan2017semanticsBiases}{}%
 Caliskan, A., Bryson, J. J., \& Narayanan, A. (2017). Semantics derived
@@
 \emph{Science}, \emph{356}(6334), 183--186. American Association for the
-Advancement of Science (AAAS). Retrieved from
+Advancement of Science ({AAAS}). Retrieved from
 \url{https://doi.org/10.1126/science.aal4230}
 
-\hypertarget{ref-Garg2018years}{}
+\leavevmode\hypertarget{ref-Garg2018years}{}%
 Garg, N., Schiebinger, L., Jurafsky, D., \& Zou, J. (2018). Word
@@
 
-\hypertarget{ref-Gonen2019lipstick}{}
-Gonen, H., \& Goldberg, Y. (2019). Lipstick on a pig: Debiasing methods
-cover up systematic gender biases in word embeddings but do not remove
-them. In \emph{Proceedings of the 2019 conference of the north American
-chapter of the association for computational linguistics: Human language
-technologies, volume 1 (long and short papers)} (pp. 609--614).
-Minneapolis, Minnesota: Association for Computational Linguistics.
-Retrieved from \url{https://www.aclweb.org/anthology/N19-1061}
+\leavevmode\hypertarget{ref-Gonen2019lipstick}{}%
+Gonen, H., \& Goldberg, Y. (2019). Lipstick on a pig: {D}ebiasing
+methods cover up systematic gender biases in word embeddings but do not
+remove them. \emph{Proceedings of the 2019 conference of the north
+{A}merican chapter of the association for computational linguistics:
+Human language technologies, volume 1 (long and short papers)} (pp.
+609--614). Minneapolis, Minnesota: Association for Computational
+Linguistics. Retrieved from
+\url{https://www.aclweb.org/anthology/N19-1061}
 
-\hypertarget{ref-Lauscher2019multidimensional}{}
+\leavevmode\hypertarget{ref-Lauscher2019multidimensional}{}%
 Lauscher, A., \& Glavas, G. (2019). Are we consistently biased?
@@
 
-\hypertarget{ref-Nissim2020fair}{}
+\leavevmode\hypertarget{ref-Nissim2020fair}{}%
 Nissim, M., Noord, R. van, \& Goot, R. van der. (2020). Fair is better
 than sensational: Man is to doctor as woman is to doctor.
-\emph{Computational Linguistics}, \emph{46}(2), 487--497. MIT Press -
+\emph{Computational Linguistics}, \emph{46}(2), 487--497. {MIT} Press -
 Journals. Retrieved from \url{https://doi.org/10.1162/coli_a_00379}
 
-\hypertarget{ref-Nosek2002harvesting}{}
+\leavevmode\hypertarget{ref-Nosek2002harvesting}{}%
 Nosek, B. A., Banaji, M. R., \& Greenwald, A. G. (2002). Harvesting
@@
 \emph{Group Dynamics: Theory, Research, and Practice}, \emph{6}(1),
-101--115. American Psychological Association (APA). Retrieved from
+101--115. American Psychological Association ({APA}). Retrieved from
 \url{https://doi.org/10.1037/1089-2699.6.1.101}
 
+\end{CSLReferences}
+
 \end{document}

### 2021-06-10 paperDraft/paperDraft.tex

@@
 almost the same word sets, allowing for a \emph{prima facie} sensible
-comparison with bias in humans. Let \(f\) be a similarity measure
-(usually, cosine similarity). The association difference for a term
-\(t\) is: \begin{align}
+comparison with bias in humans. If the attitude towards given pair of
+concept is to be interpreted as neutral, the final value from the
+formula should be around 0. Let \(f\) be a similarity measure (usually,
+cosine similarity). The association difference for a term \(t\) is:
+\begin{align}
 s(t,A,B) & = \frac{\sum_{a\in A}f(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}f(t,b)}{\vert B\vert}
@@
 methodology to a multilingual and cross-lingual setting, arguing that
-using Euclidean distance instead of similarty does not make much
-difference, while the bias effects vary greatly accross embedding models
+using Euclidean distance instead of similarity does not make much
+difference, while the bias effects vary greatly across embedding models
 (interestingly, with social media-text trained embeddings being less

### 2021-06-10 abstractESSLLI21/abstractESSLLI1.Rmd
first draft of abstract

---
title: ""
author: ""
output:
  pdf_document:
    number_sections: true
    df_print: kable 
    keep_tex: true
    includes:
      in_header:
        - Rafal_latex6.sty
classoption: dvipsnames,enabledeprecatedfontcommands
fontsize: 12pt
documentclass: scrartcl
urlcolor: blue
bibliography: [../references/cosineReferences1.bib]
csl: transactions-on-computational-logic.csl
---

```{r, setup, include=FALSE}
library(ggplot2)
library(ggthemes)
library(gridExtra)
library(kableExtra)
```

\begin{center}
\Large \textbf{A bayesian method of cosine-based word2vec bias estimation}
\end{center}

\vspace{1mm}

A considerable amount of literature exists on  bias detection and mitigation in NLP models, especially word2vec embeddings, which represent words as number vectors (see e.g. @bolukbasi2016man and @Gonen2019lipstick and references therein). The most common method used compares  cosine similarity between words from protected groups and attributes that are considered to be stereotypical or harmful in some way, and this method will be in our focus.

In one well-known approach, @Caliskan2017semanticsBiases proposed the Word Embedding Association Test (WEAT). The idea is that the measure of  biases between two sets of target words, $X$ and $Y$, (we call them protected words) should be quantified in terms of the cosine similarity  between the protected words and attribute words coming from  two sets of stereotype attribute words, $A$ and $B$ (we'll call them attributes). For instance, $X$ might be a set of male names, $Y$ a set of female names, $A$ might contain stereotypically male-related career words, and $B$ stereotypically female-related family words. WEAT is a modification  of the Implicit Association Test (IAT) [@Nosek2002harvesting] used in psychology and uses almost the same word sets, allowing for a \emph{prima facie} sensible comparison with bias in humans.  The association difference for a term $t$ is $s(t,A,B)$, and the effect size is computed by normalizing the difference in means as follows:

\vspace{-3mm}

\footnotesize \begin{align}
\mathsf{s}(t,A,B)  = \frac{\sum_{a\in A}f(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}f(t,b)}{\vert B\vert}
& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,
\mathsf{bias}(A,B)  = \frac{
\mu(\{s(x,A,B)\}_{x\in X}) -\mu(\{s(y,A,B)\}_{y\in Y}) 
}{
\sigma(\{s(w,A,B)\}_{w\in X\cup Y})
} \tag{WEAT}
\end{align}
\normalsize
\noindent @Caliskan2017semanticsBiases show that significant biases---thus measured--- similar to the ones discovered by IAT can be discovered in word embeddings. @Lauscher2019multidimensional extends the methodology to a multilingual and cross-lingual setting; a similar methodology is employed by @Garg2018years, who employ word embeddings trained on corpora from different decades. @Manzini2019blackToCriminal modify WEAT to a multi-class setting, introducing Mean Average Cosine (MAC) similarity as a measure of bias (in fact, in the paper  they report distances rather than similarities). Let $T = \{t_1, \dots, t_k\}$ be a class of protected word embeddings, and let each $A_j\in A$ be a set of attributes stereotypically associated with a protected word). Then:

\vspace{-2mm}

\footnotesize
\begin{align}
\mathsf{S}(t_i, A_j)  = \frac{1}{\vert A_j\vert}\sum_{a\in A_j}\mathsf{cos}(t,a) & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,
\mathsf{MAC}(T,A)  = \frac{1}{\vert T \vert \,\vert A\vert}\sum_{t_i \in T }\sum_{A_j \in A} S(t_i,A_j) \tag{MAC}
\end{align}
\normalsize 
\noindent That is, for each protected word  and each attribute class, they first take the mean for this protected word and all attributes in a given attribute class, and then take the mean of thus obtained means for all the protected words.

Such methods are statistically problematic. One issue, specific to @Manzini2019blackToCriminal, is that no distinction is made based on whether given a protected word and a  class of attributes is stereotypically associated with this  protected word or a different protected word. A related, general problem, is that all the authors ignore the step of comparing their results with control groups, especially control groups of stereotype-neutral human attributes. That such a move is important is suggested for instance by Figure 1 prepared using the original word list for religion-related protected words extended with control attributes, where such human attributes are also closer to protected classes than 1.

Another, also serious problem, is that  all the authors calculate means of means and their authors run statistical tests on sets of means. Unforutnately, by pre-averaging the data they throw away information about sample sizes, and  they remove variation, and so pre-averaging  tends to manufacture false confidence.

All such approaches come up with rather short lists of protected words and rather short lists of stereotypical attributes. Clearly, these are not complete list. So let's treat them as  samples from richer pools of stereotypical predicates and let's take the uncertainty and variation involved seriously. To illustrate, let's employ the formulas used by @Caliskan2017semanticsBiases in a simple simulation.

\noindent
\begin{tabular}{ll}
\begin{minipage}[c]{0.35\linewidth}
\scalebox{0.65}{
\begin{tabular}{cccccc}\toprule
$X$ to $A$  & $X$ to $B$ & $Y$ to $A$ & $Y$ to $B$ & $\sigma$  & WEAT \\
\midrule
0 & 0 & 0 & 0 & 0.05 & 1.82 \\
0 & 0 & 0 & 0 & 0.001 & -1.93\\
.1 & 0 & 0 & .1 & 0.05 & 1.49 \\
.1 & 0 & 0 & .1 & .1 & 1.22\\
\bottomrule
\end{tabular}
}
\end{minipage}& \begin{minipage}[c]{0.6\linewidth}
Two protected classes, $X=\{t_1,t_2\}$ and $Y=\{t_3,t_4\}$, and two five-element attribute sets $A$ and $B$. One simple simulation draws normally distributed values for such combinations for two cases in which the underlying mean similarities are in fact equal, two cases in which they are different, and the underlying variances are different (code available upon request).
\end{minipage}
\end{tabular}

\vspace{1mm}

The following observations are worthwhile. (1) For points randomly drawn from distributions where there is no difference in means the calculated effect size can easily be  1.82, whereas the largest effect size reported by @Caliskan2017semanticsBiases is 1.81. (2) For samples from distributions where the means are different, the (absolute) effect sizes can easily be lower than in the first two simulations.  As Figure 2. illustrates,  quite some uncertainty is involved, far from what systematically low mean-based p-values reported in the papers might suggest. Part of the problem is random variation unaccounted for in the original approach (see Figures 3 and 4 for an example), and part of the problem is that  that  non-negligible changes in effect size can result from a shift in the standard deviation of the original process, because with the decrease of standard deviation the numerator in (WEAT) decreases.

To improve on the situation, we  build  bayesian models to estimate the biases involved using the raw datapoints, actually using control groups,  distinguishing the connection types,  and taking the uncertainty involved seriously.
For the general impact of being associated, we build  models using STAN according to the following specification:

\footnotesize

\vspace{-5mm}

\begin{align}
cosineDistance_i  & \sim dnorm(\mu_i, \sigma) \\
\mu_i & = m_{pw} + co_{con}\\
m_{pw} & \sim dnorm(1,.5) \\
co_{con} & \sim dnorm(0,1) \\
\sigma &\sim  dcauchy(0,1)
\end{align}
\normalsize

The resulting coefficients for the religion dataset based on Reddit embeddings are in Figure 5. While there is some difference in the means, the 89\% highest posterior density invervals are quite wide and include 0s for all the coefficients.  Then, motivated by large differences between the status of different protected words, we build bayesian models with separate mean estimates, with the same regularizing priors as before. The results for the religion dataset based on Reddit embeddings are in Figure 6.

We build analogous models for other topic groups (race, gender), and different embeddings (Google, and hard-debiased Reddit). The general conclusion is that once we proceed this way, the situation is much less obvious. The word list sizes are small, sample sizes are small, and so posterior density intervals are wide. Some stronger bias can be discerned in the gender case. Moreover, sometimes the difference between associated, different and human, are not very impressive. The cost of debiasing turns out sometimes to be that   neutral human predicates get closer to protected classes (in religion),  change in gender after debiasing is really minor with   zero still out out of HPDI range, and debiasing improvement for race is a small improvement   achieved at the price of moving neutral terms closer to protected words (there is no space to visualise these results in this abstract).

The bottom line is that if we want to take bias seriously, so should we approach the uncertainty involved in our estimations. There is no replacement for proper statistical evaluation that does not discard information about the uncertaintly involved, larger word lists are needed, and visualisation of the results particular protected classes provides much better guidance than chasing a single metric based on a means of means.

<!-- Let's take a look at the head of the religion dataset used in their calculations: -->

```{r religionTableHeadEarly,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
religion <- read.csv("../datasets/religionReddit.csv")[-1]
colnames(religion) <- c("protectedWord","wordToCompare","wordClass",
                        "cosineDistance","cosineSimilarity","connection")
religion$wordClass <- as.factor(religion$wordClass)
levels(religion$wordClass) <- c("christian","human","jewish","muslim","neutral")
#head(religion)  %>%  kable(format = "latex",booktabs=T,
#                      linesep = "",  escape = FALSE,
#                      caption = "Head of the religion dataset.") %>%
#                      kable_styling(latex_options=c("scale_down"))
```

<!-- \pagebreak  -->

\pagebreak

\begin{center}
\begin{figure}[!htb]\centering
   \begin{minipage}{0.48\textwidth}
```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
ggplot(religion, aes(x =  cosineDistance, fill = connection, color = connection))+geom_density(alpha=0.6,size = .2)+theme_tufte()+labs(title  = "Figure 1.", subtitle = "Empirical distribution of cosine distances (religion, Reddit)")+ scale_fill_manual(values = c("orangered4","chartreuse4", "skyblue", "gray"))+scale_x_continuous(breaks = seq(0.3,1.5, by = 0.1))+xlab("cosine distance")+ scale_color_manual(values = c("orangered4","chartreuse4","skyblue","gray"))
```
   \end{minipage}
   \begin {minipage}{0.48\textwidth}
```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
s <- function (table){ mean(table$A) - mean(table$B)}
biasesD2 <- numeric(10000)
for(i in 1:10000){
t1d2 <- data.frame(A  = rnorm(5,.1,0.05), B = rnorm(5,0,0.05))
t2d2 <- data.frame(A  = rnorm(5,.1,0.05), B = rnorm(5,0,0.05))
t3d2 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,.1,0.05))
t4d2 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,.1,0.05))

factorD2 <- sd(c(s(t1d2),s(t2d2),s(t3d2),s(t4d2)))
numeratorD2 <-  mean(s(t1d2),s(t2d2)) - mean(s(t3d2),s(t4d2))
biasesD2[i] <- numeratorD2/factorD2
}

ggplot()+geom_histogram(aes(x=biasesD2, y = ..density..), alpha = 0.6, bins=50)+
  theme_tufte()+labs(title="Figure 2.", subtitle = "10k biases for different means and sd =.05")+ xlab("bias")
```
   \end{minipage}

  \begin{minipage}{0.48\textwidth}
```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%",message =FALSE, warning = FALSE}
library(tidyverse)
muslimWords <- c("imam","islam","mosque","muslim","quran")
muslim <- religion %>% filter(protectedWord %in% muslimWords)
muslimClass <- muslim %>% filter(protectedWord == "muslim")
neutralSample <- sample_n(filter(muslimClass,connection == "none"), 5)
humanSample <- sample_n(filter(muslimClass,connection == "human"), 5)
muslimVis <- muslimClass %>% filter(connection != "none" & connection !="human")
muslimVis <- rbind(muslimVis,neutralSample,humanSample)
#we plug in our visualisation script
source("../functions/visualisationTools.R")
#two arguments: dataset and protected word
visualiseProtected(muslimVis,"muslim")+labs(title = "Figure 3.",
subtitle = "Cosine distances for active attributes (musim, Reddit)")
```
\end{minipage}
   \begin {minipage}{0.48\textwidth}
```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
library(tidyverse)
priestClass <- religion %>% filter(protectedWord == "priest")
neutralSample <- sample_n(filter(priestClass,connection == "none"), 5)
humanSample <- sample_n(filter(priestClass,connection == "human"), 5)
priestVis <- priestClass %>% filter(connection != "none" & connection !="human")
priestVis <- rbind(priestVis,neutralSample,humanSample)

#we plug in our visualisation script
source("../functions/visualisationTools.R")
#two arguments: dataset and protected word
visualiseProtected(priestVis,"priest")+labs(title = "Figure 4.",
subtitle = "Cosine distances for active attributes (priest, Reddit)")
```
   \end{minipage}

  \begin{minipage}{0.38\textwidth}
\includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
\end{minipage}
   \begin {minipage}{0.58\textwidth}
\includegraphics{../images/visReligionReddit.png}
   \end{minipage}
\end{figure}

\end{center}

\scriptsize

\vspace{-4mm}

### 2021-06-10 abstractESSLLI21/abstractESSLLI1.tex
first draft of abstract

\documentclass[12pt,dvipsnames,enabledeprecatedfontcommands]{scrartcl}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage[unicode=true]{hyperref}
\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
\hypersetup{
            colorlinks=true,
            linkcolor=Maroon,
            citecolor=Blue,
            urlcolor=blue,
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

%\documentclass{article}

% %packages
 \usepackage{booktabs}

\usepackage{multirow}
\usepackage{multicol}

\usepackage{graphicx}
\usepackage{longtable}
\usepackage{ragged2e}
\usepackage{etex}
%\usepackage{yfonts}
\usepackage{marvosym}
\usepackage[notextcomp]{kpfonts}
\usepackage{nicefrac}
\newcommand*{\QED}{\hfill \footnotesize {\sc Q.e.d.}}
\usepackage{floatrow}

\usepackage[textsize=scriptsize, textwidth = 1.5cm]{todonotes}
%\linespread{1.5}

\setlength{\parindent}{10pt}
\setlength{\parskip}{1pt}

%language
\usepackage{times}
\usepackage{t1enc}
%\usepackage[utf8x]{inputenc}
%\usepackage[polish]{babel}
%\usepackage{polski}

\usepackage{mathptmx}
\usepackage[scaled=0.88]{helvet}

%AMS
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{mathtools}

\usepackage{geometry}
 \geometry{a4paper,left=20mm,top=15mm,bottom = 20mm, right = 20mm}

%environments
\newtheorem{fact}{Fact}

%abbreviations
\newcommand{\ra}{\rangle}
\newcommand{\la}{\langle}
\newcommand{\n}{\neg}
\newcommand{\et}{\wedge}
\newcommand{\jt}{\rightarrow}
\newcommand{\ko}[1]{\forall  #1\,}
\newcommand{\ro}{\leftrightarrow}
\newcommand{\exi}[1]{\exists\, {_{#1}}}
\newcommand{\pr}[1]{\mathsf{P}(#1)}
\newcommand{\cost}{\mathsf{cost}}

\newcommand{\odds}{\mathsf{Odds}}
\newcommand{\ind}{\mathsf{Ind}}
\newcommand{\nf}[2]{\nicefrac{#1\,}{#2}}
\newcommand{\R}[1]{\texttt{#1}}
\newcommand{\prr}[1]{\mbox{$\mathtt{P}_{prior}(#1)$}}
\newcommand{\prp}[1]{\mbox{$\mathtt{P}_{posterior}(#1)$}}

\newtheorem{q}{\color{blue}Question}
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}

%technical intermezzo
%---------------------

\newcommand{\intermezzoa}{
	\begin{minipage}[c]{13cm}
	\begin{center}\rule{10cm}{0.4pt}

	\tiny{\sc Optional Content Starts}

	\vspace{-1mm}

	\rule{10cm}{0.4pt}\end{center}
	\end{minipage}\nopagebreak 
	}

\newcommand{\intermezzob}{\nopagebreak 
	\begin{minipage}[c]{13cm}
	\begin{center}\rule{10cm}{0.4pt}

	\tiny{\sc Optional Content Ends}

	\vspace{-1mm}

	\rule{10cm}{0.4pt}\end{center}
	\end{minipage}
	}
%--------------------

\DeclareUnicodeCharacter{0301}{*************************************}

\newtheorem*{reply*}{Reply}
\usepackage{enumitem}
\newcommand{\question}[1]{\begin{enumerate}[resume,leftmargin=0cm,labelsep=0cm,align=left]
\item #1
\end{enumerate}}

\usepackage{float}

% \setbeamertemplate{blocks}[rounded][shadow=true]
% \setbeamertemplate{itemize items}[ball]
% \AtBeginPart{}
% \AtBeginSection{}
% \AtBeginSubsection{}
% \AtBeginSubsubsection{}
% \setlength{\emergencystretch}{0em}
% \setlength{\parskip}{0pt}

\usepackage[authoryear]{natbib}

%\bibliographystyle{apalike}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\author{}
\date{\vspace{-2.5em}}

\begin{document}

\begin{center}
\Large \textbf{A bayesian method of cosine-based word2vec bias estimation}
\end{center}

\vspace{1mm}

A considerable amount of literature exists on bias detection and
mitigation in NLP models, especially word2vec embeddings, which
represent words as number vectors (see e.g. {[}1{]} and {[}4{]} and
references therein). The most common method used compares cosine
similarity between words from protected groups and attributes that are
considered to be stereotypical or harmful in some way, and this method
will be in our focus.

In one well-known approach, {[}2{]} proposed the Word Embedding
Association Test (WEAT). The idea is that the measure of biases between
two sets of target words, \(X\) and \(Y\), (we call them protected
words) should be quantified in terms of the cosine similarity between
the protected words and attribute words coming from two sets of
stereotype attribute words, \(A\) and \(B\) (we'll call them
attributes). For instance, \(X\) might be a set of male names, \(Y\) a
set of female names, \(A\) might contain stereotypically male-related
career words, and \(B\) stereotypically female-related family words.
WEAT is a modification of the Implicit Association Test (IAT) {[}7{]}
used in psychology and uses almost the same word sets, allowing for a
\emph{prima facie} sensible comparison with bias in humans. The
association difference for a term \(t\) is \(s(t,A,B)\), and the effect
size is computed by normalizing the difference in means as follows:

\vspace{-3mm}

\footnotesize

\begin{align}
\mathsf{s}(t,A,B)  = \frac{\sum_{a\in A}f(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}f(t,b)}{\vert B\vert}
& \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,
\mathsf{bias}(A,B)  = \frac{
\mu(\{s(x,A,B)\}_{x\in X}) -\mu(\{s(y,A,B)\}_{y\in Y}) 
}{
\sigma(\{s(w,A,B)\}_{w\in X\cup Y})
} \tag{WEAT}
\end{align}

\normalsize
\noindent {[}2{]} show that significant biases---thus measured---
similar to the ones discovered by IAT can be discovered in word
embeddings. {[}5{]} extends the methodology to a multilingual and
cross-lingual setting; a similar methodology is employed by {[}3{]}, who
employ word embeddings trained on corpora from different decades.
{[}6{]} modify WEAT to a multi-class setting, introducing Mean Average
Cosine (MAC) similarity as a measure of bias (in fact, in the paper they
report distances rather than similarities). Let
\(T = \{t_1, \dots, t_k\}\) be a class of protected word embeddings, and
let each \(A_j\in A\) be a set of attributes stereotypically associated
with a protected word). Then:

\vspace{-2mm}

\footnotesize

\begin{align}
\mathsf{S}(t_i, A_j)  = \frac{1}{\vert A_j\vert}\sum_{a\in A_j}\mathsf{cos}(t,a) & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,
\mathsf{MAC}(T,A)  = \frac{1}{\vert T \vert \,\vert A\vert}\sum_{t_i \in T }\sum_{A_j \in A} S(t_i,A_j) \tag{MAC}
\end{align}

\normalsize  \noindent That is, for each protected word and each
attribute class, they first take the mean for this protected word and
all attributes in a given attribute class, and then take the mean of
thus obtained means for all the protected words.

Such methods are statistically problematic. One issue, specific to
{[}6{]}, is that no distinction is made based on whether given a
protected word and a class of attributes is stereotypically associated
with this protected word or a different protected word. A related,
general problem, is that all the authors ignore the step of comparing
their results with control groups, especially control groups of
stereotype-neutral human attributes. That such a move is important is
suggested for instance by Figure 1 prepared using the original word list
for religion-related protected words extended with control attributes,
where such human attributes are also closer to protected classes than 1.

Another, also serious problem, is that all the authors calculate means
of means and their authors run statistical tests on sets of means.
Unforutnately, by pre-averaging the data they throw away information
about sample sizes, and they remove variation, and so pre-averaging
tends to manufacture false confidence.

All such approaches come up with rather short lists of protected words
and rather short lists of stereotypical attributes. Clearly, these are
not complete list. So let's treat them as samples from richer pools of
stereotypical predicates and let's take the uncertainty and variation
involved seriously. To illustrate, let's employ the formulas used by
{[}2{]} in a simple simulation.

\noindent

\begin{tabular}{ll}
\begin{minipage}[c]{0.35\linewidth}
\scalebox{0.65}{
\begin{tabular}{cccccc}\toprule
$X$ to $A$  & $X$ to $B$ & $Y$ to $A$ & $Y$ to $B$ & $\sigma$  & WEAT \\
\midrule
0 & 0 & 0 & 0 & 0.05 & 1.82 \\
0 & 0 & 0 & 0 & 0.001 & -1.93\\
.1 & 0 & 0 & .1 & 0.05 & 1.49 \\
.1 & 0 & 0 & .1 & .1 & 1.22\\
\bottomrule
\end{tabular}
}
\end{minipage}& \begin{minipage}[c]{0.6\linewidth}
Two protected classes, $X=\{t_1,t_2\}$ and $Y=\{t_3,t_4\}$, and two five-element attribute sets $A$ and $B$. One simple simulation draws normally distributed values for such combinations for two cases in which the underlying mean similarities are in fact equal, two cases in which they are different, and the underlying variances are different (code available upon request).
\end{minipage}
\end{tabular}

\vspace{1mm}

The following observations are worthwhile. (1) For points randomly drawn
from distributions where there is no difference in means the calculated
effect size can easily be 1.82, whereas the largest effect size reported
by {[}2{]} is 1.81. (2) For samples from distributions where the means
are different, the (absolute) effect sizes can easily be lower than in
the first two simulations. As Figure 2. illustrates, quite some
uncertainty is involved, far from what systematically low mean-based
p-values reported in the papers might suggest. Part of the problem is
random variation unaccounted for in the original approach (see Figures 3
and 4 for an example), and part of the problem is that that
non-negligible changes in effect size can result from a shift in the
standard deviation of the original process, because with the decrease of
standard deviation the numerator in (WEAT) decreases.

To improve on the situation, we build bayesian models to estimate the
biases involved using the raw datapoints, actually using control groups,
distinguishing the connection types, and taking the uncertainty involved
seriously. For the general impact of being associated, we build models
using STAN according to the following specification:

\footnotesize

\vspace{-5mm}

\begin{align}
cosineDistance_i  & \sim dnorm(\mu_i, \sigma) \\
\mu_i & = m_{pw} + co_{con}\\
m_{pw} & \sim dnorm(1,.5) \\
co_{con} & \sim dnorm(0,1) \\
\sigma &\sim  dcauchy(0,1)
\end{align}

\normalsize

The resulting coefficients for the religion dataset based on Reddit
embeddings are in Figure 5. While there is some difference in the means,
the 89\% highest posterior density invervals are quite wide and include
0s for all the coefficients. Then, motivated by large differences
between the status of different protected words, we build bayesian
models with separate mean estimates, with the same regularizing priors
as before. The results for the religion dataset based on Reddit
embeddings are in Figure 6.

We build analogous models for other topic groups (race, gender), and
different embeddings (Google, and hard-debiased Reddit). The general
conclusion is that once we proceed this way, the situation is much less
obvious. The word list sizes are small, sample sizes are small, and so
posterior density intervals are wide. Some stronger bias can be
discerned in the gender case. Moreover, sometimes the difference between
associated, different and human, are not very impressive. The cost of
debiasing turns out sometimes to be that neutral human predicates get
closer to protected classes (in religion), change in gender after
debiasing is really minor with zero still out out of HPDI range, and
debiasing improvement for race is a small improvement achieved at the
price of moving neutral terms closer to protected words (there is no
space to visualise these results in this abstract).

The bottom line is that if we want to take bias seriously, so should we
approach the uncertainty involved in our estimations. There is no
replacement for proper statistical evaluation that does not discard
information about the uncertaintly involved, larger word lists are
needed, and visualisation of the results particular protected classes
provides much better guidance than chasing a single metric based on a
means of means.

\pagebreak

\begin{center}
\begin{figure}[!htb]\centering
   \begin{minipage}{0.48\textwidth}

\begin{center}\includegraphics[width=1\linewidth]{abstractESSLLI1_files/figure-latex/unnamed-chunk-1-1} \end{center}
   \end{minipage}
   \begin {minipage}{0.48\textwidth}

\begin{center}\includegraphics[width=1\linewidth]{abstractESSLLI1_files/figure-latex/unnamed-chunk-2-1} \end{center}
   \end{minipage}

  \begin{minipage}{0.48\textwidth}

\begin{center}\includegraphics[width=1\linewidth]{abstractESSLLI1_files/figure-latex/unnamed-chunk-3-1} \end{center}
\end{minipage}
   \begin {minipage}{0.48\textwidth}

\begin{center}\includegraphics[width=1\linewidth]{abstractESSLLI1_files/figure-latex/unnamed-chunk-4-1} \end{center}
   \end{minipage}

  \begin{minipage}{0.38\textwidth}
\includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
\end{minipage}
   \begin {minipage}{0.58\textwidth}
\includegraphics{../images/visReligionReddit.png}
   \end{minipage}
\end{figure}

\end{center}

\scriptsize

\vspace{-4mm}

\hypertarget{refs}{}
\hypertarget{ref-bolukbasi2016man}{}
{[}1{]} Tolga Bolukbasi, Kai-Wei Chang, James Zou, Venkatesh Saligrama,
and Adam Kalai. 2016. Man is to computer programmer as woman is to
homemaker? Debiasing word embeddings.

\hypertarget{ref-Caliskan2017semanticsBiases}{}
{[}2{]} Aylin Caliskan, Joanna J. Bryson, and Arvind Narayanan. 2017.
Semantics derived automatically from language corpora contain human-like
biases. \emph{Science} 356, 6334 (April 2017), 183--186.
DOI:\url{https://doi.org/10.1126/science.aal4230}

\hypertarget{ref-Garg2018years}{}
{[}3{]} Nikhil Garg, Londa Schiebinger, Dan Jurafsky, and James Zou.
2018. Word embeddings quantify 100 years of gender and ethnic
stereotypes. \emph{Proceedings of the National Academy of Sciences} 115,
16 (April 2018), E3635--E3644.
DOI:\url{https://doi.org/10.1073/pnas.1720347115}

\hypertarget{ref-Gonen2019lipstick}{}
{[}4{]} Hila Gonen and Yoav Goldberg. 2019. Lipstick on a pig: Debiasing
methods cover up systematic gender biases in word embeddings but do not
remove them. In \emph{Proceedings of the 2019 conference of the north
American chapter of the association for computational linguistics: Human
language technologies, volume 1 (long and short papers)}, Association
for Computational Linguistics, Minneapolis, Minnesota, 609--614.
DOI:\url{https://doi.org/10.18653/v1/N19-1061}

\hypertarget{ref-Lauscher2019multidimensional}{}
{[}5{]} Anne Lauscher and Goran Glavas. 2019. Are we consistently
biased? Multidimensional analysis of biases in distributional word
vectors. \emph{CoRR} abs/1904.11783, (2019). Retrieved from
\url{http://arxiv.org/abs/1904.11783}

\hypertarget{ref-Manzini2019blackToCriminal}{}
{[}6{]} Thomas Manzini, Yao Chong Lim, Yulia Tsvetkov, and Alan W Black.
2019. Black is to criminal as caucasian is to police: Detecting and
removing multiclass bias in word embeddings.

\hypertarget{ref-Nosek2002harvesting}{}
{[}7{]} Brian A. Nosek, Mahzarin R. Banaji, and Anthony G. Greenwald.
2002. Harvesting implicit group attitudes and beliefs from a
demonstration web site. \emph{Group Dynamics: Theory, Research, and
Practice} 6, 1 (2002), 101--115.
DOI:\url{https://doi.org/10.1037/1089-2699.6.1.101}

\end{document}

### 2021-06-10 paperDraft/paperDraft.Rmd
first draft of abstract

@@
 WEAT is a modification  of the Implicit Association Test (IAT) [@Nosek2002harvesting] used in psychology and uses almost the same word sets, allowing for a \emph{prima facie} sensible comparison with bias in humans.
-If the person's attitude towards given pair of concept is to be interpreted as neutral, the final value from the formula should be around 0.
+If the person's attitude towards given pair of concept is to be interpreted as neutral, there should be no noticeable task completion time difference, and the final value from the formula should be around 0.
 Let $f$ be a similarity measure (usually, cosine similarity). The association difference for a term $t$ is:

### 2021-06-10 paperDraft/paperDraft.tex
first draft of abstract

--- 2021-06-10 paperDraft/paperDraft.tex
@@
 almost the same word sets, allowing for a \emph{prima facie} sensible
-comparison with bias in humans. Let \(f\) be a similarity measure
-(usually, cosine similarity). The association difference for a term
-\(t\) is:
+comparison with bias in humans. If the person's attitude towards given
+pair of concept is to be interpreted as neutral, there should be no
+noticeable task completion time difference, and the final value from the
+formula should be around 0. Let \(f\) be a similarity measure (usually,
+cosine similarity). The association difference for a term \(t\) is:
 
@@
 multilingual and cross-lingual setting, arguing that using Euclidean
-distance instead of similarty does not make much difference, while the
-bias effects vary greatly accross embedding models (interestingly, with
+distance instead of similarity does not make much difference, while the
+bias effects vary greatly across embedding models (interestingly, with
 social media-text trained embeddings being less biased than those based
@@
 A similar methodology is employed by Garg, Schiebinger, Jurafsky, \& Zou
-(2018), who employ word embeddings trained od corpora from different
+(2018), who employ word embeddings trained on corpora from different
 decades to study the shifts in various biases. For instance, to compute
@@
   different value:
+\item
+  The lack of control group in the paper and our analysis indicates that
+  without neutral baseline it is difficult to interpret the
+  effectiveness of the metric.
 \end{itemize}

### 2021-06-10 thesis/_book/_main.tex
first draft of abstract

--- 2021-06-10 thesis/_book/_main.tex
@@
                         \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
+\NormalTok{religion}\OperatorTok{$}\NormalTok{wordClass <-}\StringTok{ }\KeywordTok{as.factor}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass)}
 \KeywordTok{levels}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
@@
                         \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
-\KeywordTok{library}\NormalTok{(plyr)}
-\NormalTok{religion}\OperatorTok{$}\NormalTok{wordClass <-}\StringTok{ }\KeywordTok{mapvalues}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass, }\DataTypeTok{from =} \KeywordTok{levels}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass), }\DataTypeTok{to =} \KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{))}
-\CommentTok{#levels(religion$wordClass) <- c("christian","human","jewish","muslim","neutral")}
+\CommentTok{#head(religion)}
+\CommentTok{#library(plyr)}
+\NormalTok{religion}\OperatorTok{$}\NormalTok{wordClass <-}\StringTok{ }\KeywordTok{as.factor}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass)}
+\CommentTok{#levels(religion$wordClass)}
+\CommentTok{#religion$wordClass <- mapvalues(religion$wordClass, }
+\CommentTok{#                          from = levels(religion$wordClass),}
+\CommentTok{#                          to = c("christian","human","jewish","muslim","neutral"))}
+\KeywordTok{levels}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
 \KeywordTok{head}\NormalTok{(religion)  }\OperatorTok{%>%}\StringTok{  }\KeywordTok{kable}\NormalTok{(}\DataTypeTok{format =} \StringTok{"latex"}\NormalTok{,}\DataTypeTok{booktabs=}\NormalTok{T,}
@@
 
+\chapter{Discussion}\label{discussion}
+
 \hypertarget{refs}{}

### 2021-06-10 abstractESSLLI21/abstractESSLLI1.Rmd
Adds abstract review

@@
 
+\todo{czy konieczne jest pisanie "number vectors"? jeżeli tak, to w licencjacie musze pozmieniac, bo tam czasami jest samo vectors, a czasami "they are assigned numerical values"}
 A considerable amount of literature exists on  bias detection and mitigation in NLP models, especially word2vec embeddings, which represent words as number vectors (see e.g.
@@
 A related, general problem, is that all the authors ignore the step of comparing their results with control groups, especially control groups of stereotype-neutral human attributes.
-That such a move is important is suggested for instance by Figure 1 prepared using the original word list for religion-related protected words extended with control attributes, where such human attributes are also closer to protected classes than 1.
+\todo{nie wiem, czy nie przeredagowałabym tego zdania, chodzi mi o końcówkę, "closer than 1", moze zamiast than 1 powiedziec, "than neutral non-human words"?
+albo jakoś podobnie} That such a move is important is suggested for instance by Figure 1 prepared using the original word list for religion-related protected words extended with control attributes, where such human attributes are also closer to protected classes than 1.
 
-Another, also serious problem, is that  all the authors calculate means of means and their authors run statistical tests on sets of means.
-Unforutnately, by pre-averaging the data they throw away information about sample sizes, and  they remove variation, and so pre-averaging  tends to manufacture false confidence.
+Another, also serious problem, is that  all the authors calculate means of means and their authors \todo{jak rozumiec drugi raz powtorzone their authors?} run statistical tests on sets of means.
+Unfortunately, by pre-averaging the data they throw away information about sample sizes, and  they remove variation, and so pre-averaging  tends to manufacture false confidence.
 
@@
 \end{minipage}& \begin{minipage}[c]{0.6\linewidth}
+\todo{początek zdania "two protected classes" bez wstepu typu "We created..", nie wiem czy tak miało być?}
 Two protected classes, $X=\{t_1,t_2\}$ and $Y=\{t_3,t_4\}$, and two five-element attribute sets $A$ and $B$.
@@
 illustrates,  quite some uncertainty is involved, far from what systematically low mean-based p-values reported in the papers might suggest.
-Part of the problem is random variation unaccounted for in the original approach (see Figures 3 and 4 for an example), and part of the problem is that  that  non-negligible changes in effect size can result from a shift in the standard deviation of the original process, because with the decrease of standard deviation the numerator in (WEAT) decreases.
+Part of the problem is random variation unaccounted for in the original approach (see Figures 3 and 4 for an example), and part of the problem is that non-negligible changes in effect size can result from a shift in the standard deviation of the original process, because with the decrease of standard deviation the numerator in (WEAT) decreases.
 
@@
 Some stronger bias can be discerned in the gender case.
-Moreover, sometimes the difference between associated, different and human, are not very impressive.
+Moreover, sometimes the differences between associated, different and human, are not very impressive.
 The cost of debiasing turns out sometimes to be that   neutral human predicates get closer to protected classes (in religion),  change in gender after debiasing is really minor with   zero still out out of HPDI range, and debiasing improvement for race is a small improvement   achieved at the price of moving neutral terms closer to protected words (there is no space to visualise these results in this abstract).

### 2021-06-10 abstractESSLLI21/abstractESSLLI1.tex
Adds abstract review

@@
-\documentclass[12pt,dvipsnames,enabledeprecatedfontcommands]{scrartcl}
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}
+%
+\documentclass[
+  12pt,
+  dvipsnames,enabledeprecatedfontcommands]{scrartcl}
+\usepackage{amsmath,amssymb}
 \usepackage{lmodern}
-\usepackage{amssymb,amsmath}
 \usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
 \ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
@@
   linkcolor=Maroon,
+  filecolor=Maroon,
   citecolor=Blue,
   urlcolor=blue,
-            breaklinks=true}
-\urlstyle{same}  % don't use monospace font for urls
-\usepackage{graphicx,grffile}
+  pdfcreator={LaTeX via pandoc}}
+\urlstyle{same} % disable monospaced font for URLs
+\usepackage{graphicx}
 \makeatletter
@@
 \usepackage{xcolor}
+\ifluatex
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
+  % set entry spacing
+  \ifnum #2 > 0
+  \setlength{\parskip}{#2\baselineskip}
+  \fi
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
@@
 \vspace{1mm}
+
+\todo{czy konieczne jest pisanie "number vectors"? jeżeli tak, to w licencjacie musze pozmieniac, bo tam czasami jest samo vectors, a czasami "they are assigned numerical values"}
 
@@
 
-\footnotesize
-
-\begin{align}
+\footnotesize \begin{align}
 \mathsf{s}(t,A,B)  = \frac{\sum_{a\in A}f(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}f(t,b)}{\vert B\vert}
@@
 } \tag{WEAT}
-\end{align}
-
-\normalsize
-\noindent {[}2{]} show that significant biases---thus measured---
-similar to the ones discovered by IAT can be discovered in word
-embeddings. {[}5{]} extends the methodology to a multilingual and
-cross-lingual setting; a similar methodology is employed by {[}3{]}, who
-employ word embeddings trained on corpora from different decades.
-{[}6{]} modify WEAT to a multi-class setting, introducing Mean Average
-Cosine (MAC) similarity as a measure of bias (in fact, in the paper they
-report distances rather than similarities). Let
-\(T = \{t_1, \dots, t_k\}\) be a class of protected word embeddings, and
-let each \(A_j\in A\) be a set of attributes stereotypically associated
-with a protected word). Then:
+\end{align} \normalsize \noindent {[}2{]} show that significant
+biases---thus measured--- similar to the ones discovered by IAT can be
+discovered in word embeddings. {[}5{]} extends the methodology to a
+multilingual and cross-lingual setting; a similar methodology is
+employed by {[}3{]}, who employ word embeddings trained on corpora from
+different decades. {[}6{]} modify WEAT to a multi-class setting,
+introducing Mean Average Cosine (MAC) similarity as a measure of bias
+(in fact, in the paper they report distances rather than similarities).
+Let \(T = \{t_1, \dots, t_k\}\) be a class of protected word embeddings,
+and let each \(A_j\in A\) be a set of attributes stereotypically
+associated with a protected word). Then:
 
@@
 \mathsf{MAC}(T,A)  = \frac{1}{\vert T \vert \,\vert A\vert}\sum_{t_i \in T }\sum_{A_j \in A} S(t_i,A_j) \tag{MAC}
-\end{align}
-
-\normalsize  \noindent That is, for each protected word and each
-attribute class, they first take the mean for this protected word and
-all attributes in a given attribute class, and then take the mean of
+\end{align} \normalsize  \noindent That is, for each protected word and
+each attribute class, they first take the mean for this protected word
+and all attributes in a given attribute class, and then take the mean of
 thus obtained means for all the protected words.
@@
 their results with control groups, especially control groups of
-stereotype-neutral human attributes. That such a move is important is
-suggested for instance by Figure 1 prepared using the original word list
-for religion-related protected words extended with control attributes,
-where such human attributes are also closer to protected classes than 1.
+stereotype-neutral human attributes.
+\todo{nie wiem, czy nie przeredagowałabym tego zdania, chodzi mi o końcówkę, "closer than 1", moze zamiast than 1 powiedziec, "than neutral non-human words"? albo jakoś podobnie}
+That such a move is important is suggested for instance by Figure 1
+prepared using the original word list for religion-related protected
+words extended with control attributes, where such human attributes are
+also closer to protected classes than 1.
 
 Another, also serious problem, is that all the authors calculate means
-of means and their authors run statistical tests on sets of means.
-Unforutnately, by pre-averaging the data they throw away information
-about sample sizes, and they remove variation, and so pre-averaging
-tends to manufacture false confidence.
+of means and their authors
+\todo{jak rozumiec drugi raz powtorzone their authors?} run statistical
+tests on sets of means. Unfortunately, by pre-averaging the data they
+throw away information about sample sizes, and they remove variation,
+and so pre-averaging tends to manufacture false confidence.
 
@@
 random variation unaccounted for in the original approach (see Figures 3
-and 4 for an example), and part of the problem is that that
-non-negligible changes in effect size can result from a shift in the
-standard deviation of the original process, because with the decrease of
-standard deviation the numerator in (WEAT) decreases.
+and 4 for an example), and part of the problem is that non-negligible
+changes in effect size can result from a shift in the standard deviation
+of the original process, because with the decrease of standard deviation
+the numerator in (WEAT) decreases.
 
@@
 \sigma &\sim  dcauchy(0,1)
-\end{align}
-
-\normalsize
+\end{align} \normalsize
 
@@
 posterior density intervals are wide. Some stronger bias can be
-discerned in the gender case. Moreover, sometimes the difference between
-associated, different and human, are not very impressive. The cost of
-debiasing turns out sometimes to be that neutral human predicates get
-closer to protected classes (in religion), change in gender after
-debiasing is really minor with zero still out out of HPDI range, and
-debiasing improvement for race is a small improvement achieved at the
-price of moving neutral terms closer to protected words (there is no
-space to visualise these results in this abstract).
+discerned in the gender case. Moreover, sometimes the differences
+between associated, different and human, are not very impressive. The
+cost of debiasing turns out sometimes to be that neutral human
+predicates get closer to protected classes (in religion), change in
+gender after debiasing is really minor with zero still out out of HPDI
+range, and debiasing improvement for race is a small improvement
+achieved at the price of moving neutral terms closer to protected words
+(there is no space to visualise these results in this abstract).
 
@@
 \hypertarget{refs}{}
-\hypertarget{ref-bolukbasi2016man}{}
-{[}1{]} Tolga Bolukbasi, Kai-Wei Chang, James Zou, Venkatesh Saligrama,
-and Adam Kalai. 2016. Man is to computer programmer as woman is to
-homemaker? Debiasing word embeddings.
+\begin{CSLReferences}{0}{0}
+\leavevmode\hypertarget{ref-bolukbasi2016man}{}%
+\CSLLeftMargin{{[}1{]} }
+\CSLRightInline{Tolga Bolukbasi, Kai-Wei Chang, James Zou, Venkatesh
+Saligrama, and Adam Kalai. 2016. Man is to computer programmer as woman
+is to homemaker? Debiasing word embeddings. Retrieved from
+\url{http://arxiv.org/abs/1607.06520}}
 
-\hypertarget{ref-Caliskan2017semanticsBiases}{}
-{[}2{]} Aylin Caliskan, Joanna J. Bryson, and Arvind Narayanan. 2017.
-Semantics derived automatically from language corpora contain human-like
-biases. \emph{Science} 356, 6334 (April 2017), 183--186.
-DOI:\url{https://doi.org/10.1126/science.aal4230}
+\leavevmode\hypertarget{ref-Caliskan2017semanticsBiases}{}%
+\CSLLeftMargin{{[}2{]} }
+\CSLRightInline{Aylin Caliskan, Joanna J. Bryson, and Arvind Narayanan.
+2017. Semantics derived automatically from language corpora contain
+human-like biases. \emph{Science} 356, 6334 (April 2017), 183--186.
+DOI:https://doi.org/\href{https://doi.org/10.1126/science.aal4230}{10.1126/science.aal4230}}
 
-\hypertarget{ref-Garg2018years}{}
-{[}3{]} Nikhil Garg, Londa Schiebinger, Dan Jurafsky, and James Zou.
-2018. Word embeddings quantify 100 years of gender and ethnic
+\leavevmode\hypertarget{ref-Garg2018years}{}%
+\CSLLeftMargin{{[}3{]} }
+\CSLRightInline{Nikhil Garg, Londa Schiebinger, Dan Jurafsky, and James
+Zou. 2018. Word embeddings quantify 100 years of gender and ethnic
 stereotypes. \emph{Proceedings of the National Academy of Sciences} 115,
 16 (April 2018), E3635--E3644.
-DOI:\url{https://doi.org/10.1073/pnas.1720347115}
+DOI:https://doi.org/\href{https://doi.org/10.1073/pnas.1720347115}{10.1073/pnas.1720347115}}
 
-\hypertarget{ref-Gonen2019lipstick}{}
-{[}4{]} Hila Gonen and Yoav Goldberg. 2019. Lipstick on a pig: Debiasing
-methods cover up systematic gender biases in word embeddings but do not
-remove them. In \emph{Proceedings of the 2019 conference of the north
-American chapter of the association for computational linguistics: Human
-language technologies, volume 1 (long and short papers)}, Association
-for Computational Linguistics, Minneapolis, Minnesota, 609--614.
-DOI:\url{https://doi.org/10.18653/v1/N19-1061}
+\leavevmode\hypertarget{ref-Gonen2019lipstick}{}%
+\CSLLeftMargin{{[}4{]} }
+\CSLRightInline{Hila Gonen and Yoav Goldberg. 2019. Lipstick on a pig:
+{D}ebiasing methods cover up systematic gender biases in word embeddings
+but do not remove them. In \emph{Proceedings of the 2019 conference of
+the north {A}merican chapter of the association for computational
+linguistics: Human language technologies, volume 1 (long and short
+papers)}, Association for Computational Linguistics, Minneapolis,
+Minnesota, 609--614.
+DOI:https://doi.org/\href{https://doi.org/10.18653/v1/N19-1061}{10.18653/v1/N19-1061}}
 
-\hypertarget{ref-Lauscher2019multidimensional}{}
-{[}5{]} Anne Lauscher and Goran Glavas. 2019. Are we consistently
-biased? Multidimensional analysis of biases in distributional word
-vectors. \emph{CoRR} abs/1904.11783, (2019). Retrieved from
-\url{http://arxiv.org/abs/1904.11783}
+\leavevmode\hypertarget{ref-Lauscher2019multidimensional}{}%
+\CSLLeftMargin{{[}5{]} }
+\CSLRightInline{Anne Lauscher and Goran Glavas. 2019. Are we
+consistently biased? Multidimensional analysis of biases in
+distributional word vectors. \emph{CoRR} abs/1904.11783, (2019).
+Retrieved from \url{http://arxiv.org/abs/1904.11783}}
 
-\hypertarget{ref-Manzini2019blackToCriminal}{}
-{[}6{]} Thomas Manzini, Yao Chong Lim, Yulia Tsvetkov, and Alan W Black.
-2019. Black is to criminal as caucasian is to police: Detecting and
-removing multiclass bias in word embeddings.
+\leavevmode\hypertarget{ref-Manzini2019blackToCriminal}{}%
+\CSLLeftMargin{{[}6{]} }
+\CSLRightInline{Thomas Manzini, Yao Chong Lim, Yulia Tsvetkov, and Alan
+W Black. 2019. Black is to criminal as caucasian is to police: Detecting
+and removing multiclass bias in word embeddings. Retrieved from
+\url{http://arxiv.org/abs/1904.04047}}
 
-\hypertarget{ref-Nosek2002harvesting}{}
-{[}7{]} Brian A. Nosek, Mahzarin R. Banaji, and Anthony G. Greenwald.
-2002. Harvesting implicit group attitudes and beliefs from a
+\leavevmode\hypertarget{ref-Nosek2002harvesting}{}%
+\CSLLeftMargin{{[}7{]} }
+\CSLRightInline{Brian A. Nosek, Mahzarin R. Banaji, and Anthony G.
+Greenwald. 2002. Harvesting implicit group attitudes and beliefs from a
 demonstration web site. \emph{Group Dynamics: Theory, Research, and
 Practice} 6, 1 (2002), 101--115.
-DOI:\url{https://doi.org/10.1037/1089-2699.6.1.101}
+DOI:https://doi.org/\href{https://doi.org/10.1037/1089-2699.6.1.101}{10.1037/1089-2699.6.1.101}}
+
+\end{CSLReferences}
 

### 2021-06-10 thesis/_book/_main.tex
Adds abstract review

@@
                         \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
-\NormalTok{religion}\OperatorTok{$}\NormalTok{wordClass <-}\StringTok{ }\KeywordTok{as.factor}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass)}
 \KeywordTok{levels}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
@@
                         \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
-\CommentTok{#head(religion)}
-\CommentTok{#library(plyr)}
-\NormalTok{religion}\OperatorTok{$}\NormalTok{wordClass <-}\StringTok{ }\KeywordTok{as.factor}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass)}
-\CommentTok{#levels(religion$wordClass)}
-\CommentTok{#religion$wordClass <- mapvalues(religion$wordClass, }
-\CommentTok{#                          from = levels(religion$wordClass),}
-\CommentTok{#                          to = c("christian","human","jewish","muslim","neutral"))}
 \KeywordTok{levels}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
@@
                       \DataTypeTok{linesep =} \StringTok{""}\NormalTok{,  }\DataTypeTok{escape =} \OtherTok{FALSE}\NormalTok{, }
-                      \DataTypeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\OperatorTok{%>%}\StringTok{      }\KeywordTok{kable_styling}\NormalTok{(}\DataTypeTok{latex_options=}\KeywordTok{c}\NormalTok{(}\StringTok{"scale_down"}\NormalTok{))}
+                      \DataTypeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\OperatorTok{%>%}
+\StringTok{                      }\KeywordTok{kable_styling}\NormalTok{(}\DataTypeTok{latex_options=}\KeywordTok{c}\NormalTok{(}\StringTok{"scale_down"}\NormalTok{))}
 \end{Highlighting}
@@
 
-\chapter{Discussion}\label{discussion}
-
 \hypertarget{refs}{}

### 2021-06-10 abstractESSLLI21/abstractESSLLI1.Rmd
adds review fix

@@
 To illustrate, let's employ the formulas used by @Caliskan2017semanticsBiases in a simple simulation.
+\todo{początek zdania "two protected classes" bez wstepu typu "We created..", nie wiem czy tak miało być?}
 
@@
 \end{minipage}& \begin{minipage}[c]{0.6\linewidth}
-\todo{początek zdania "two protected classes" bez wstepu typu "We created..", nie wiem czy tak miało być?}
+
 Two protected classes, $X=\{t_1,t_2\}$ and $Y=\{t_3,t_4\}$, and two five-element attribute sets $A$ and $B$.

### 2021-06-10 abstractESSLLI21/abstractESSLLI1.tex
adds review fix

@@
 {[}2{]} in a simple simulation.
+\todo{początek zdania "two protected classes" bez wstepu typu "We created..", nie wiem czy tak miało być?}
 

### 2021-06-10 abstractESSLLI21/abstractESSLLI1.Rmd
proofread abstract ESSLLI1

--- 2021-06-10 abstractESSLLI21/abstractESSLLI1.Rmd
@@
 WEAT is a modification  of the Implicit Association Test (IAT) [@Nosek2002harvesting] used in psychology and uses almost the same word sets, allowing for a \emph{prima facie} sensible comparison with bias in humans.
-The association difference for a term $t$ is $s(t,A,B)$, and the effect size is computed by normalizing the difference in means as follows:
+The association difference for a term $t$ is $s(t,A,B)$, and the effect size is computed by normalizing the difference in means as follows ($f$ is a similarity measure, such as cosine similarity):
 
@@
 \noindent @Caliskan2017semanticsBiases show that significant biases---thus measured--- similar to the ones discovered by IAT can be discovered in word embeddings.
-@Lauscher2019multidimensional extends the methodology to a multilingual and cross-lingual setting; a similar methodology is employed by @Garg2018years, who employ word embeddings trained on corpora from different decades.
+@Lauscher2019multidimensional extends the methodology to a multilingual and cross-lingual setting; a similar methodology is employed by @Garg2018years, who use word embeddings trained on corpora from different decades.
 @Manzini2019blackToCriminal modify WEAT to a multi-class setting, introducing Mean Average Cosine (MAC) similarity as a measure of bias (in fact, in the paper  they report distances rather than similarities).
@@
 Such methods are statistically problematic.
-One issue, specific to @Manzini2019blackToCriminal, is that no distinction is made based on whether given a protected word and a  class of attributes is stereotypically associated with this  protected word or a different protected word.
+One issue, specific to @Manzini2019blackToCriminal, is that no distinction is made based on whether given a protected word a  class of attributes is stereotypically associated with this  protected word or with a different protected word.
 A related, general problem, is that all the authors ignore the step of comparing their results with control groups, especially control groups of stereotype-neutral human attributes.
@@
 \end{minipage}& \begin{minipage}[c]{0.6\linewidth}
-Two protected classes, $X=\{t_1,t_2\}$ and $Y=\{t_3,t_4\}$, and two five-element attribute sets $A$ and $B$.
-One simple simulation draws normally distributed values for such combinations for two cases in which the underlying mean similarities are in fact equal, two cases in which they are different, and the underlying variances are different (code available upon request).
+Randomly drawn protected classes, $X=\{t_1,t_2\}$ and $Y=\{t_3,t_4\}$, and two five-element attribute sets $A$ and $B$.
+One simple simulation draws normally distributed values for such combinations for two cases in which the underlying mean similarities are in fact equal to zero, two cases in which they are different, with a range of choices of standard devation  (code available upon request).
 \end{minipage}
@@
 As Figure 2.
-illustrates,  quite some uncertainty is involved, far from what systematically low mean-based p-values reported in the papers might suggest.
-Part of the problem is random variation unaccounted for in the original approach (see Figures 3 and 4 for an example), and part of the problem is that  that  non-negligible changes in effect size can result from a shift in the standard deviation of the original process, because with the decrease of standard deviation the numerator in (WEAT) decreases.
+illustrates,  quite some uncertainty is involved, far more than what systematically low mean-based p-values reported in the papers might suggest.
+Part of the problem is random variation unaccounted for in the original approach (see Figures 3 and 4 for an example), and part of the problem is that    non-negligible changes in effect size can result from a shift in the standard deviation of the original process, because with the decrease of standard deviation the numerator in (WEAT) decreases.
 
-To improve on the situation, we  build  bayesian models to estimate the biases involved using the raw datapoints, actually using control groups,  distinguishing the connection types,  and taking the uncertainty involved seriously.
-For the general impact of being associated, we build  models using STAN according to the following specification:
+To improve on the situation, we  build  Bayesian models to estimate the biases involved using the raw datapoints, actually using control groups,  distinguishing the connection types,  and taking the uncertainty involved seriously.
+For the general impact of being associated, we build  models using STAN according to the following specification (linear categorical model and the priors):
 
@@
 While there is some difference in the means, the 89\% highest posterior density invervals are quite wide and include 0s for all the coefficients.
-Then, motivated by large differences between the status of different protected words, we build bayesian models with separate mean estimates, with the same regularizing priors as before.
+Then, motivated by large differences between the states of different protected words, we build bayesian models with separate mean estimates, with the same regularizing priors as before.
 The results for the religion dataset based on Reddit embeddings are in Figure 6.
@@
 The bottom line is that if we want to take bias seriously, so should we approach the uncertainty involved in our estimations.
-There is no replacement for proper statistical evaluation that does not discard information about the uncertaintly involved, larger word lists are needed, and visualisation of the results particular protected classes provides much better guidance than chasing a single metric based on a means of means.
+There is no replacement for proper statistical evaluation that does not discard information about the uncertaintly involved, larger word lists are needed, and visualisation of the results for particular protected classes provides much better guidance than chasing a single metric based on a means of means.
 

### 2021-06-10 abstractESSLLI21/abstractESSLLI1.tex
proofread abstract ESSLLI1

--- 2021-06-10 abstractESSLLI21/abstractESSLLI1.tex
@@
+\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
+\PassOptionsToPackage{hyphens}{url}
+\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}
+%
 \documentclass[12pt,dvipsnames,enabledeprecatedfontcommands]{scrartcl}
@@
   \usepackage[utf8]{inputenc}
+  \usepackage{textcomp} % provides euro and other symbols
 \else % if luatex or xelatex
-  \ifxetex
-    \usepackage{mathspec}
-  \else
-    \usepackage{fontspec}
-  \fi
+  \usepackage{unicode-math}
   \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
@@
 }{}
-\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
-\usepackage[unicode=true]{hyperref}
-\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
+\IfFileExists{parskip.sty}{%
+\usepackage{parskip}
+}{% else
+\setlength{\parindent}{0pt}
+\setlength{\parskip}{6pt plus 2pt minus 1pt}
+}
+\usepackage{xcolor}
+\usepackage{hyperref}
 \hypersetup{
@@
             linkcolor=Maroon,
+            filecolor=Maroon,
             citecolor=Blue,
@@
 \setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-\IfFileExists{parskip.sty}{%
-\usepackage{parskip}
-}{% else
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
-}
 \setlength{\emergencystretch}{3em}  % prevent overfull lines
@@
 mitigation in NLP models, especially word2vec embeddings, which
-represent words as number vectors (see e.g. {[}1{]} and {[}4{]} and
+represent words as number vectors (see e.g.~{[}1{]} and {[}4{]} and
 references therein). The most common method used compares cosine
@@
 association difference for a term \(t\) is \(s(t,A,B)\), and the effect
-size is computed by normalizing the difference in means as follows:
+size is computed by normalizing the difference in means as follows
+(\(f\) is a similarity measure, such as cosine similarity):
 
@@
 } \tag{WEAT}
-\end{align}
-
-\normalsize
-\noindent {[}2{]} show that significant biases---thus measured---
-similar to the ones discovered by IAT can be discovered in word
-embeddings. {[}5{]} extends the methodology to a multilingual and
-cross-lingual setting; a similar methodology is employed by {[}3{]}, who
-employ word embeddings trained on corpora from different decades.
-{[}6{]} modify WEAT to a multi-class setting, introducing Mean Average
-Cosine (MAC) similarity as a measure of bias (in fact, in the paper they
-report distances rather than similarities). Let
-\(T = \{t_1, \dots, t_k\}\) be a class of protected word embeddings, and
-let each \(A_j\in A\) be a set of attributes stereotypically associated
-with a protected word). Then:
+\end{align} \normalsize \noindent {[}2{]} show that significant
+biases---thus measured--- similar to the ones discovered by IAT can be
+discovered in word embeddings. {[}5{]} extends the methodology to a
+multilingual and cross-lingual setting; a similar methodology is
+employed by {[}3{]}, who use word embeddings trained on corpora from
+different decades. {[}6{]} modify WEAT to a multi-class setting,
+introducing Mean Average Cosine (MAC) similarity as a measure of bias
+(in fact, in the paper they report distances rather than similarities).
+Let \(T = \{t_1, \dots, t_k\}\) be a class of protected word embeddings,
+and let each \(A_j\in A\) be a set of attributes stereotypically
+associated with a protected word). Then:
 
@@
 {[}6{]}, is that no distinction is made based on whether given a
-protected word and a class of attributes is stereotypically associated
-with this protected word or a different protected word. A related,
+protected word a class of attributes is stereotypically associated with
+this protected word or with a different protected word. A related,
 general problem, is that all the authors ignore the step of comparing
@@
 the first two simulations. As Figure 2. illustrates, quite some
-uncertainty is involved, far from what systematically low mean-based
-p-values reported in the papers might suggest. Part of the problem is
-random variation unaccounted for in the original approach (see Figures 3
-and 4 for an example), and part of the problem is that that
+uncertainty is involved, far more than what systematically low
+mean-based p-values reported in the papers might suggest. Part of the
+problem is random variation unaccounted for in the original approach
+(see Figures 3 and 4 for an example), and part of the problem is that
 non-negligible changes in effect size can result from a shift in the
@@
 
-To improve on the situation, we build bayesian models to estimate the
+To improve on the situation, we build Bayesian models to estimate the
 biases involved using the raw datapoints, actually using control groups,
@@
 seriously. For the general impact of being associated, we build models
-using STAN according to the following specification:
+using STAN according to the following specification (linear categorical
+model and the priors):
 
@@
 0s for all the coefficients. Then, motivated by large differences
-between the status of different protected words, we build bayesian
+between the states of different protected words, we build bayesian
 models with separate mean estimates, with the same regularizing priors
@@
 information about the uncertaintly involved, larger word lists are
-needed, and visualisation of the results particular protected classes
-provides much better guidance than chasing a single metric based on a
-means of means.
+needed, and visualisation of the results for particular protected
+classes provides much better guidance than chasing a single metric based
+on a means of means.
 
@@
 \hypertarget{refs}{}
-\hypertarget{ref-bolukbasi2016man}{}
+\leavevmode\hypertarget{ref-bolukbasi2016man}{}%
 {[}1{]} Tolga Bolukbasi, Kai-Wei Chang, James Zou, Venkatesh Saligrama,
 and Adam Kalai. 2016. Man is to computer programmer as woman is to
-homemaker? Debiasing word embeddings.
+homemaker? Debiasing word embeddings. Retrieved from
+\url{http://arxiv.org/abs/1607.06520}
 
-\hypertarget{ref-Caliskan2017semanticsBiases}{}
+\leavevmode\hypertarget{ref-Caliskan2017semanticsBiases}{}%
 {[}2{]} Aylin Caliskan, Joanna J. Bryson, and Arvind Narayanan. 2017.
@@
 
-\hypertarget{ref-Garg2018years}{}
+\leavevmode\hypertarget{ref-Garg2018years}{}%
 {[}3{]} Nikhil Garg, Londa Schiebinger, Dan Jurafsky, and James Zou.
@@
 
-\hypertarget{ref-Gonen2019lipstick}{}
+\leavevmode\hypertarget{ref-Gonen2019lipstick}{}%
 {[}4{]} Hila Gonen and Yoav Goldberg. 2019. Lipstick on a pig: Debiasing
@@
 
-\hypertarget{ref-Lauscher2019multidimensional}{}
+\leavevmode\hypertarget{ref-Lauscher2019multidimensional}{}%
 {[}5{]} Anne Lauscher and Goran Glavas. 2019. Are we consistently
@@
 
-\hypertarget{ref-Manzini2019blackToCriminal}{}
+\leavevmode\hypertarget{ref-Manzini2019blackToCriminal}{}%
 {[}6{]} Thomas Manzini, Yao Chong Lim, Yulia Tsvetkov, and Alan W Black.
 2019. Black is to criminal as caucasian is to police: Detecting and
-removing multiclass bias in word embeddings.
+removing multiclass bias in word embeddings. Retrieved from
+\url{http://arxiv.org/abs/1904.04047}
 
-\hypertarget{ref-Nosek2002harvesting}{}
+\leavevmode\hypertarget{ref-Nosek2002harvesting}{}%
 {[}7{]} Brian A. Nosek, Mahzarin R. Banaji, and Anthony G. Greenwald.

### 2021-06-10 abstractESSLLI21/abstractESSLLI1.Rmd
Fixes abstract merge

--- 2021-06-10 abstractESSLLI21/abstractESSLLI1.Rmd
@@
 
-Two protected classes, $X=\{t_1,t_2\}$ and $Y=\{t_3,t_4\}$, and two five-element attribute sets $A$ and $B$.
-One simple simulation draws normally distributed values for such combinations for two cases in which the underlying mean similarities are in fact equal, two cases in which they are different, and the underlying variances are different (code available upon request).
+Randomly drawn protected classes, $X=\{t_1,t_2\}$ and $Y=\{t_3,t_4\}$, and two five-element attribute sets $A$ and $B$.
+One simple simulation draws normally distributed values for such combinations for two cases in which the underlying mean similarities are in fact equal to zero, two cases in which they are different, with a range of choices of standard deviation  (code available upon request).
+
 \end{minipage}
@@
 As Figure 2.
-illustrates,  quite some uncertainty is involved, far from what systematically low mean-based p-values reported in the papers might suggest.
+illustrates,  quite some uncertainty is involved, far more than what systematically low mean-based p-values reported in the papers might suggest.
 Part of the problem is random variation unaccounted for in the original approach (see Figures 3 and 4 for an example), and part of the problem is that    non-negligible changes in effect size can result from a shift in the standard deviation of the original process, because with the decrease of standard deviation the numerator in (WEAT) decreases.
 
-To improve on the situation, we  build  bayesian models to estimate the biases involved using the raw datapoints, actually using control groups,  distinguishing the connection types,  and taking the uncertainty involved seriously.
-For the general impact of being associated, we build  models using STAN according to the following specification:
+To improve on the situation, we  build  Bayesian models to estimate the biases involved using the raw datapoints, actually using control groups,  distinguishing the connection types,  and taking the uncertainty involved seriously.
+For the general impact of being associated, we build  models using STAN according to the following specification (linear categorical model and the priors):
 

### 2021-06-10 abstractESSLLI21/abstractESSLLI1.tex
Fixes abstract merge

--- 2021-06-10 abstractESSLLI21/abstractESSLLI1.tex
@@
 multilingual and cross-lingual setting; a similar methodology is
-employed by {[}3{]}, who employ word embeddings trained on corpora from
+employed by {[}3{]}, who use word embeddings trained on corpora from
 different decades. {[}6{]} modify WEAT to a multi-class setting,
@@
 the first two simulations. As Figure 2. illustrates, quite some
-uncertainty is involved, far from what systematically low mean-based
-p-values reported in the papers might suggest. Part of the problem is
-random variation unaccounted for in the original approach (see Figures 3
-and 4 for an example), and part of the problem is that non-negligible
-changes in effect size can result from a shift in the standard deviation
-of the original process, because with the decrease of standard deviation
-the numerator in (WEAT) decreases.
+uncertainty is involved, far more than what systematically low
+mean-based p-values reported in the papers might suggest. Part of the
+problem is random variation unaccounted for in the original approach
+(see Figures 3 and 4 for an example), and part of the problem is that
+non-negligible changes in effect size can result from a shift in the
+standard deviation of the original process, because with the decrease of
+standard deviation the numerator in (WEAT) decreases.
 
-To improve on the situation, we build bayesian models to estimate the
+To improve on the situation, we build Bayesian models to estimate the
 biases involved using the raw datapoints, actually using control groups,

### 2021-06-11 abstractESSLLI21/abstractESSLLI1.Rmd
proofread abstract, small changes

@@
 
-\todo{czy konieczne jest pisanie "number vectors"? jeżeli tak, to w licencjacie musze pozmieniac, bo tam czasami jest samo vectors, a czasami "they are assigned numerical values"}
-A considerable amount of literature exists on  bias detection and mitigation in NLP models, especially word2vec embeddings, which represent words as number vectors (see e.g.
+A considerable amount of literature exists on  bias detection and mitigation in NLP models, especially word2vec embeddings, which represent words as  vectors of real numbers (see e.g.
 @bolukbasi2016man and @Gonen2019lipstick and references therein).
@@
 Such methods are statistically problematic.
-One issue, specific to @Manzini2019blackToCriminal, is that no distinction is made based on whether given a protected word and a  class of attributes is stereotypically associated with this  protected word or a different protected word.
-A related, general problem, is that all the authors ignore the step of comparing their results with control groups, especially control groups of stereotype-neutral human attributes.
-\todo{nie wiem, czy nie przeredagowałabym tego zdania, chodzi mi o końcówkę, "closer than 1", moze zamiast than 1 powiedziec, "than neutral non-human words"?
-albo jakoś podobnie} That such a move is important is suggested for instance by Figure 1 prepared using the original word list for religion-related protected words extended with control attributes, where such human attributes are also closer to protected classes than 1.
+Two issues, specific to @Manzini2019blackToCriminal, are that (i) no distinction is made based on whether given a protected word and a  class of attributes is stereotypically associated with this  protected word or a different protected word, and (ii) no re-scaling by standard deviation (typical in effect size measures such as  Cohen's $d$) is used, so MAC is not a good measure of effect size.
+A more general problem, is that all the authors ignore the step of comparing their results with control groups, especially control groups of stereotype-neutral human attributes.
+That such a move is important is suggested for instance by Figure 1 prepared using the original word list for religion-related protected words extended with control attributes, where such human attributes are also closer to protected classes than neutral non-human words.
 
-Another, also serious problem, is that  all the authors calculate means of means and their authors \todo{jak rozumiec drugi raz powtorzone their authors?} run statistical tests on sets of means.
+Another, also serious problem, is that  all the authors calculate means of means and  run statistical tests on sets of means.
 Unfortunately, by pre-averaging the data they throw away information about sample sizes, and  they remove variation, and so pre-averaging  tends to manufacture false confidence.
@@
 To illustrate, let's employ the formulas used by @Caliskan2017semanticsBiases in a simple simulation.
-\todo{początek zdania "two protected classes" bez wstepu typu "We created..", nie wiem czy tak miało być?}
 
@@
 
-Randomly drawn protected classes, $X=\{t_1,t_2\}$ and $Y=\{t_3,t_4\}$, and two five-element attribute sets $A$ and $B$.
-One simple simulation draws normally distributed values for such combinations for two cases in which the underlying mean similarities are in fact equal to zero, two cases in which they are different, with a range of choices of standard deviation  (code available upon request).
+We have randomly drawn protected classes, $X=\{t_1,t_2\}$ and $Y=\{t_3,t_4\}$, and two five-element attribute sets $A$ and $B$.
+One simulation draws normally distributed values  for two situations in which the underlying mean similarities are in fact equal to zero, in two other cases  the means are different, with a range of choices of common standard deviation  accross the cases (code available upon request).
 

### 2021-06-11 abstractESSLLI21/abstractESSLLI1.tex
proofread abstract, small changes

--- 2021-06-10 abstractESSLLI21/abstractESSLLI1.tex
@@
-\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
-\PassOptionsToPackage{hyphens}{url}
-\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}
-%
 \documentclass[12pt,dvipsnames,enabledeprecatedfontcommands]{scrartcl}
@@
   \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provides euro and other symbols
 \else % if luatex or xelatex
-  \usepackage{unicode-math}
+  \ifxetex
+    \usepackage{mathspec}
+  \else
+    \usepackage{fontspec}
+  \fi
   \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
@@
 }{}
-\IfFileExists{parskip.sty}{%
-\usepackage{parskip}
-}{% else
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
-}
-\usepackage{xcolor}
-\usepackage{hyperref}
+\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
+\usepackage[unicode=true]{hyperref}
+\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
 \hypersetup{
@@
             linkcolor=Maroon,
-            filecolor=Maroon,
             citecolor=Blue,
@@
 \setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+\IfFileExists{parskip.sty}{%
+\usepackage{parskip}
+}{% else
+\setlength{\parindent}{0pt}
+\setlength{\parskip}{6pt plus 2pt minus 1pt}
+}
 \setlength{\emergencystretch}{3em}  % prevent overfull lines
@@
 mitigation in NLP models, especially word2vec embeddings, which
-represent words as number vectors (see e.g.~{[}1{]} and {[}4{]} and
-references therein). The most common method used compares cosine
+represent words as vectors of real numbers (see e.g. {[}1{]} and {[}4{]}
+and references therein). The most common method used compares cosine
 similarity between words from protected groups and attributes that are
@@
 
-\footnotesize \begin{align}
+\footnotesize
+
+\begin{align}
 \mathsf{s}(t,A,B)  = \frac{\sum_{a\in A}f(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}f(t,b)}{\vert B\vert}
@@
 } \tag{WEAT}
-\end{align} \normalsize \noindent {[}2{]} show that significant
-biases---thus measured--- similar to the ones discovered by IAT can be
-discovered in word embeddings. {[}5{]} extends the methodology to a
-multilingual and cross-lingual setting; a similar methodology is
-employed by {[}3{]}, who use word embeddings trained on corpora from
-different decades. {[}6{]} modify WEAT to a multi-class setting,
-introducing Mean Average Cosine (MAC) similarity as a measure of bias
-(in fact, in the paper they report distances rather than similarities).
-Let \(T = \{t_1, \dots, t_k\}\) be a class of protected word embeddings,
-and let each \(A_j\in A\) be a set of attributes stereotypically
-associated with a protected word). Then:
+\end{align}
+
+\normalsize
+\noindent {[}2{]} show that significant biases---thus measured---
+similar to the ones discovered by IAT can be discovered in word
+embeddings. {[}5{]} extends the methodology to a multilingual and
+cross-lingual setting; a similar methodology is employed by {[}3{]}, who
+use word embeddings trained on corpora from different decades. {[}6{]}
+modify WEAT to a multi-class setting, introducing Mean Average Cosine
+(MAC) similarity as a measure of bias (in fact, in the paper they report
+distances rather than similarities). Let \(T = \{t_1, \dots, t_k\}\) be
+a class of protected word embeddings, and let each \(A_j\in A\) be a set
+of attributes stereotypically associated with a protected word). Then:
 
@@
 \mathsf{MAC}(T,A)  = \frac{1}{\vert T \vert \,\vert A\vert}\sum_{t_i \in T }\sum_{A_j \in A} S(t_i,A_j) \tag{MAC}
-\end{align} \normalsize  \noindent That is, for each protected word and
-each attribute class, they first take the mean for this protected word
-and all attributes in a given attribute class, and then take the mean of
+\end{align}
+
+\normalsize  \noindent That is, for each protected word and each
+attribute class, they first take the mean for this protected word and
+all attributes in a given attribute class, and then take the mean of
 thus obtained means for all the protected words.
 
-Such methods are statistically problematic. One issue, specific to
-{[}6{]}, is that no distinction is made based on whether given a
-protected word a class of attributes is stereotypically associated with
-this protected word or with a different protected word. A related,
-general problem, is that all the authors ignore the step of comparing
-their results with control groups, especially control groups of
-stereotype-neutral human attributes. That such a move is important is
+Such methods are statistically problematic. Two issues, specific to
+{[}6{]}, are that (i) no distinction is made based on whether given a
+protected word and a class of attributes is stereotypically associated
+with this protected word or a different protected word, and (ii) no
+re-scaling by standard deviation (typical in effect size measures such
+as Cohen's \(d\)) is used, so MAC is not a good measure of effect size.
+A more general problem, is that all the authors ignore the step of
+comparing their results with control groups, especially control groups
+of stereotype-neutral human attributes. That such a move is important is
 suggested for instance by Figure 1 prepared using the original word list
 for religion-related protected words extended with control attributes,
-where such human attributes are also closer to protected classes than 1.
+where such human attributes are also closer to protected classes than
+neutral non-human words.
 
 Another, also serious problem, is that all the authors calculate means
-of means and their authors run statistical tests on sets of means.
-Unforutnately, by pre-averaging the data they throw away information
-about sample sizes, and they remove variation, and so pre-averaging
-tends to manufacture false confidence.
+of means and run statistical tests on sets of means. Unfortunately, by
+pre-averaging the data they throw away information about sample sizes,
+and they remove variation, and so pre-averaging tends to manufacture
+false confidence.
 
@@
 \end{minipage}& \begin{minipage}[c]{0.6\linewidth}
-Randomly drawn protected classes, $X=\{t_1,t_2\}$ and $Y=\{t_3,t_4\}$, and two five-element attribute sets $A$ and $B$.
-One simple simulation draws normally distributed values for such combinations for two cases in which the underlying mean similarities are in fact equal to zero, two cases in which they are different, with a range of choices of standard devation  (code available upon request).
+
+We have randomly drawn protected classes, $X=\{t_1,t_2\}$ and $Y=\{t_3,t_4\}$, and two five-element attribute sets $A$ and $B$.
+One simulation draws normally distributed values  for two situations in which the underlying mean similarities are in fact equal to zero, in two other cases  the means are different, with a range of choices of common standard deviation  accross the cases (code available upon request).
+
 \end{minipage}
@@
 \sigma &\sim  dcauchy(0,1)
-\end{align} \normalsize
+\end{align}
+
+\normalsize
 
@@
 \hypertarget{refs}{}
-\leavevmode\hypertarget{ref-bolukbasi2016man}{}%
+\hypertarget{ref-bolukbasi2016man}{}
 {[}1{]} Tolga Bolukbasi, Kai-Wei Chang, James Zou, Venkatesh Saligrama,
 and Adam Kalai. 2016. Man is to computer programmer as woman is to
-homemaker? Debiasing word embeddings. Retrieved from
-\url{http://arxiv.org/abs/1607.06520}
+homemaker? Debiasing word embeddings.
 
-\leavevmode\hypertarget{ref-Caliskan2017semanticsBiases}{}%
+\hypertarget{ref-Caliskan2017semanticsBiases}{}
 {[}2{]} Aylin Caliskan, Joanna J. Bryson, and Arvind Narayanan. 2017.
@@
 
-\leavevmode\hypertarget{ref-Garg2018years}{}%
+\hypertarget{ref-Garg2018years}{}
 {[}3{]} Nikhil Garg, Londa Schiebinger, Dan Jurafsky, and James Zou.
@@
 
-\leavevmode\hypertarget{ref-Gonen2019lipstick}{}%
+\hypertarget{ref-Gonen2019lipstick}{}
 {[}4{]} Hila Gonen and Yoav Goldberg. 2019. Lipstick on a pig: Debiasing
@@
 
-\leavevmode\hypertarget{ref-Lauscher2019multidimensional}{}%
+\hypertarget{ref-Lauscher2019multidimensional}{}
 {[}5{]} Anne Lauscher and Goran Glavas. 2019. Are we consistently
@@
 
-\leavevmode\hypertarget{ref-Manzini2019blackToCriminal}{}%
+\hypertarget{ref-Manzini2019blackToCriminal}{}
 {[}6{]} Thomas Manzini, Yao Chong Lim, Yulia Tsvetkov, and Alan W Black.
 2019. Black is to criminal as caucasian is to police: Detecting and
-removing multiclass bias in word embeddings. Retrieved from
-\url{http://arxiv.org/abs/1904.04047}
+removing multiclass bias in word embeddings.
 
-\leavevmode\hypertarget{ref-Nosek2002harvesting}{}%
+\hypertarget{ref-Nosek2002harvesting}{}
 {[}7{]} Brian A. Nosek, Mahzarin R. Banaji, and Anthony G. Greenwald.

### 2021-06-12 abstractESSLLI21/abstractESSLLI1.tex
Adds models names

--- 2021-06-10 abstractESSLLI21/abstractESSLLI1.tex
@@
 
-\todo{czy konieczne jest pisanie "number vectors"? jeżeli tak, to w licencjacie musze pozmieniac, bo tam czasami jest samo vectors, a czasami "they are assigned numerical values"}
-
 A considerable amount of literature exists on bias detection and
 mitigation in NLP models, especially word2vec embeddings, which
-represent words as number vectors (see e.g. {[}1{]} and {[}4{]} and
-references therein). The most common method used compares cosine
+represent words as vectors of real numbers (see e.g. {[}1{]} and {[}4{]}
+and references therein). The most common method used compares cosine
 similarity between words from protected groups and attributes that are
@@
 
-Such methods are statistically problematic. One issue, specific to
-{[}6{]}, is that no distinction is made based on whether given a
+Such methods are statistically problematic. Two issues, specific to
+{[}6{]}, are that (i) no distinction is made based on whether given a
 protected word and a class of attributes is stereotypically associated
-with this protected word or a different protected word. A related,
-general problem, is that all the authors ignore the step of comparing
-their results with control groups, especially control groups of
-stereotype-neutral human attributes.
-\todo{nie wiem, czy nie przeredagowałabym tego zdania, chodzi mi o końcówkę, "closer than 1", moze zamiast than 1 powiedziec, "than neutral non-human words"? albo jakoś podobnie}
-That such a move is important is suggested for instance by Figure 1
-prepared using the original word list for religion-related protected
-words extended with control attributes, where such human attributes are
-also closer to protected classes than 1.
+with this protected word or a different protected word, and (ii) no
+re-scaling by standard deviation (typical in effect size measures such
+as Cohen's \(d\)) is used, so MAC is not a good measure of effect size.
+A more general problem, is that all the authors ignore the step of
+comparing their results with control groups, especially control groups
+of stereotype-neutral human attributes. That such a move is important is
+suggested for instance by Figure 1 prepared using the original word list
+for religion-related protected words extended with control attributes,
+where such human attributes are also closer to protected classes than
+neutral non-human words.
 
 Another, also serious problem, is that all the authors calculate means
-of means and their authors
-\todo{jak rozumiec drugi raz powtorzone their authors?} run statistical
-tests on sets of means. Unfortunately, by pre-averaging the data they
-throw away information about sample sizes, and they remove variation,
-and so pre-averaging tends to manufacture false confidence.
+of means and run statistical tests on sets of means. Unfortunately, by
+pre-averaging the data they throw away information about sample sizes,
+and they remove variation, and so pre-averaging tends to manufacture
+false confidence.
 
@@
 {[}2{]} in a simple simulation.
-\todo{początek zdania "two protected classes" bez wstepu typu "We created..", nie wiem czy tak miało być?}
 
@@
 We build analogous models for other topic groups (race, gender), and
-different embeddings (Google, and hard-debiased Reddit). The general
+different embeddings (we compare the results of two models:
+GoogleNews-vectors-negative300, and Reddit Corpus). The general
 conclusion is that once we proceed this way, the situation is much less

### 2021-06-14 abstractESSLLI21/abstractESSLLI1.Rmd

@@
 
-We build analogous models for other topic groups (race, gender), and different embeddings (Google, and hard-debiased Reddit).
+We build analogous models for other topic groups (race, gender), and different embeddings (we compare the results of two models: GoogleNews-vectors-negative300, and Reddit Corpus).
 The general conclusion is that once we proceed this way, the situation is much less obvious.

### 2021-06-14 abstractESSLLI21/abstractESSLLI1.Rmd
revised abstract

--- 2021-06-11 abstractESSLLI21/abstractESSLLI1.Rmd
@@
 To improve on the situation, we  build  Bayesian models to estimate the biases involved using the raw datapoints, actually using control groups,  distinguishing the connection types,  and taking the uncertainty involved seriously.
-For the general impact of being associated, we build  models using STAN according to the following specification (linear categorical model and the priors):
+For the general impact of being associated, we build  models using hamiltonian Monte Carlo (STAN) according to the following specification (linear categorical model and the priors):
 
@@
 \begin{align}
-cosineDistance_i  & \sim dnorm(\mu_i, \sigma) \\
-\mu_i & = m_{pw} + co_{con}\\
-m_{pw} & \sim dnorm(1,.5) \\
-co_{con} & \sim dnorm(0,1) \\
-\sigma &\sim  dcauchy(0,1)
+cosineDistance_i  & \sim dnorm(\mu_i, \sigma) &
+\mu_i & = m_{pw} + co_{con} \tag{Model}\\
+m_{pw} & \sim dnorm(1,.5)  & 
+co_{con} & \sim dnorm(0,1)  \,\,\,\,\,
+ & \sigma  \sim  dcauchy(0,1) \tag{Priors}
 \end{align} 
@@
 
+\vspace{-2mm}
+
+\noindent That is, given a protected word we assume its cosine distance to  attributes from a given group is normally distributed around a mean, which is determined by the mean assigned to this protected word and by the connection type of that group of attributes (stereotypically associated/ associated with a different predicate/neutral human attribute/random word), and we use weak regularizing priors for the parameters.
+The task of the model is to estimate the mean for a given protected word and the coefficients of change in that mean corresponding to different connection types.
+
 The resulting coefficients for the religion dataset based on Reddit embeddings are in Figure 5.
 While there is some difference in the means, the 89\% highest posterior density invervals are quite wide and include 0s for all the coefficients.
-Then, motivated by large differences between the states of different protected words, we build bayesian models with separate mean estimates, with the same regularizing priors as before.
+Then, motivated by large differences between the states of different protected words, we build bayesian models with separate coefficients for protected classes, with analogous regularizing priors.
 The results for the religion dataset based on Reddit embeddings are in Figure 6.
 
-We build analogous models for other topic groups (race, gender), and different embeddings (Google, and hard-debiased Reddit).
-The general conclusion is that once we proceed this way, the situation is much less obvious.
+We build analogous models for other topic groups (race, gender), and different embeddings (we use  three embeddings: GoogleNews-vectors-negative300, and Reddit Corpus, and the hard-debiased Reddit embedding used by @Manzini2019blackToCriminal).
+The general conclusion is that from this bayesian perpective the situation is much less obvious.
 The word list sizes are small, sample sizes are small, and so posterior density intervals are wide.
@@
 Moreover, sometimes the differences between associated, different and human, are not very impressive.
-The cost of debiasing turns out sometimes to be that   neutral human predicates get closer to protected classes (in religion),  change in gender after debiasing is really minor with   zero still out out of HPDI range, and debiasing improvement for race is a small improvement   achieved at the price of moving neutral terms closer to protected words (there is no space to visualise these results in this abstract).
+The cost of debiasing turns out sometimes to be that   neutral human predicates get closer to protected classes (in religion),  change in gender after debiasing is really minor with   zero still out out of HPDI range, and debiasing improvement for race is a small improvement   achieved at the price of moving neutral terms closer to protected words (there is no space to include these results in this abstract).
 
@@
 ggplot()+geom_histogram(aes(x=biasesD2, y = ..density..), alpha = 0.6, bins=50)+
-  theme_tufte()+labs(title="Figure 2.", subtitle = "10k biases for different means and sd =.05")+ xlab("bias")
+  theme_tufte()+labs(title="Figure 2.", subtitle = "10k biases for different means  and sd =.05 (case 3 in the table)")+ xlab("bias")
 ```

### 2021-06-14 abstractESSLLI21/abstractESSLLI1.tex
revised abstract

--- 2021-06-11 abstractESSLLI21/abstractESSLLI1.tex
@@
 seriously. For the general impact of being associated, we build models
-using STAN according to the following specification (linear categorical
-model and the priors):
+using hamiltonian Monte Carlo (STAN) according to the following
+specification (linear categorical model and the priors):
 
@@
 \begin{align}
-cosineDistance_i  & \sim dnorm(\mu_i, \sigma) \\
-\mu_i & = m_{pw} + co_{con}\\
-m_{pw} & \sim dnorm(1,.5) \\
-co_{con} & \sim dnorm(0,1) \\
-\sigma &\sim  dcauchy(0,1)
+cosineDistance_i  & \sim dnorm(\mu_i, \sigma) &
+\mu_i & = m_{pw} + co_{con} \tag{Model}\\
+m_{pw} & \sim dnorm(1,.5)  & 
+co_{con} & \sim dnorm(0,1)  \,\,\,\,\,
+ & \sigma  \sim  dcauchy(0,1) \tag{Priors}
 \end{align}
@@
 \normalsize
+
+\vspace{-2mm}
+
+\noindent That is, given a protected word we assume its cosine distance
+to attributes from a given group is normally distributed around a mean,
+which is determined by the mean assigned to this protected word and by
+the connection type of that group of attributes (stereotypically
+associated/ associated with a different predicate/neutral human
+attribute/random word), and we use weak regularizing priors for the
+parameters. The task of the model is to estimate the mean for a given
+protected word and the coefficients of change in that mean corresponding
+to different connection types.
 
@@
 between the states of different protected words, we build bayesian
-models with separate mean estimates, with the same regularizing priors
-as before. The results for the religion dataset based on Reddit
-embeddings are in Figure 6.
+models with separate coefficients for protected classes, with analogous
+regularizing priors. The results for the religion dataset based on
+Reddit embeddings are in Figure 6.
 
 We build analogous models for other topic groups (race, gender), and
-different embeddings (Google, and hard-debiased Reddit). The general
-conclusion is that once we proceed this way, the situation is much less
-obvious. The word list sizes are small, sample sizes are small, and so
-posterior density intervals are wide. Some stronger bias can be
-discerned in the gender case. Moreover, sometimes the differences
-between associated, different and human, are not very impressive. The
-cost of debiasing turns out sometimes to be that neutral human
-predicates get closer to protected classes (in religion), change in
-gender after debiasing is really minor with zero still out out of HPDI
-range, and debiasing improvement for race is a small improvement
-achieved at the price of moving neutral terms closer to protected words
-(there is no space to visualise these results in this abstract).
+different embeddings (we use three embeddings:
+GoogleNews-vectors-negative300, and Reddit Corpus, and the hard-debiased
+Reddit embedding used by {[}6{]}). The general conclusion is that from
+this bayesian perpective the situation is much less obvious. The word
+list sizes are small, sample sizes are small, and so posterior density
+intervals are wide. Some stronger bias can be discerned in the gender
+case. Moreover, sometimes the differences between associated, different
+and human, are not very impressive. The cost of debiasing turns out
+sometimes to be that neutral human predicates get closer to protected
+classes (in religion), change in gender after debiasing is really minor
+with zero still out out of HPDI range, and debiasing improvement for
+race is a small improvement achieved at the price of moving neutral
+terms closer to protected words (there is no space to include these
+results in this abstract).
 

### 2021-06-14 abstractESSLLI21/abstractESSLLI1.Rmd
revised abstract

@@
 \normalsize 
-\noindent That is, for each protected word  and each attribute class, they first take the mean for this protected word and all attributes in a given attribute class, and then take the mean of thus obtained means for all the protected words.
+\noindent That is, for each protected word  and each attribute class, they first take the mean for this protected word and all attributes in a given attribute class, and then they take the mean of thus obtained means for all the protected words.
 
 Such methods are statistically problematic.
-Two issues, specific to @Manzini2019blackToCriminal, are that (i) no distinction is made based on whether given a protected word and a  class of attributes is stereotypically associated with this  protected word or a different protected word, and (ii) no re-scaling by standard deviation (typical in effect size measures such as  Cohen's $d$) is used, so MAC is not a good measure of effect size.
+Two issues, specific to @Manzini2019blackToCriminal, are that (i) no distinction is made based on whether  a  class of attributes is stereotypically associated with a given   protected word or with a different protected word, and (ii) no re-scaling by standard deviation (typical in effect size measures such as  Cohen's $d$) is used, so MAC is not a good measure of effect size.
 A more general problem, is that all the authors ignore the step of comparing their results with control groups, especially control groups of stereotype-neutral human attributes.
@@
 
-Another, also serious problem, is that  all the authors calculate means of means and  run statistical tests on sets of means.
+Another  serious problem is that  all the authors calculate means of means and  run statistical tests on sets of means.
 Unfortunately, by pre-averaging the data they throw away information about sample sizes, and  they remove variation, and so pre-averaging  tends to manufacture false confidence.
 
+To illustrate, let's employ the formulas used by @Caliskan2017semanticsBiases in a simple simulation. 
 All such approaches come up with rather short lists of protected words and rather short lists of stereotypical attributes.
@@
 So let's treat them as  samples from richer pools of stereotypical predicates and let's take the uncertainty and variation involved seriously.
-To illustrate, let's employ the formulas used by @Caliskan2017semanticsBiases in a simple simulation.
 

### 2021-06-14 abstractESSLLI21/abstractESSLLI1.tex
revised abstract

@@
 attribute class, they first take the mean for this protected word and
-all attributes in a given attribute class, and then take the mean of
-thus obtained means for all the protected words.
+all attributes in a given attribute class, and then they take the mean
+of thus obtained means for all the protected words.
 
 Such methods are statistically problematic. Two issues, specific to
-{[}6{]}, are that (i) no distinction is made based on whether given a
-protected word and a class of attributes is stereotypically associated
-with this protected word or a different protected word, and (ii) no
-re-scaling by standard deviation (typical in effect size measures such
-as Cohen's \(d\)) is used, so MAC is not a good measure of effect size.
-A more general problem, is that all the authors ignore the step of
-comparing their results with control groups, especially control groups
-of stereotype-neutral human attributes. That such a move is important is
+{[}6{]}, are that (i) no distinction is made based on whether a class of
+attributes is stereotypically associated with a given protected word or
+with a different protected word, and (ii) no re-scaling by standard
+deviation (typical in effect size measures such as Cohen's \(d\)) is
+used, so MAC is not a good measure of effect size. A more general
+problem, is that all the authors ignore the step of comparing their
+results with control groups, especially control groups of
+stereotype-neutral human attributes. That such a move is important is
 suggested for instance by Figure 1 prepared using the original word list
@@
 
-Another, also serious problem, is that all the authors calculate means
-of means and run statistical tests on sets of means. Unfortunately, by
+Another serious problem is that all the authors calculate means of means
+and run statistical tests on sets of means. Unfortunately, by
 pre-averaging the data they throw away information about sample sizes,
@@
 
-All such approaches come up with rather short lists of protected words
-and rather short lists of stereotypical attributes. Clearly, these are
-not complete list. So let's treat them as samples from richer pools of
-stereotypical predicates and let's take the uncertainty and variation
-involved seriously. To illustrate, let's employ the formulas used by
-{[}2{]} in a simple simulation.
+To illustrate, let's employ the formulas used by {[}2{]} in a simple
+simulation. All such approaches come up with rather short lists of
+protected words and rather short lists of stereotypical attributes.
+Clearly, these are not complete list. So let's treat them as samples
+from richer pools of stereotypical predicates and let's take the
+uncertainty and variation involved seriously.
 

### 2021-06-14 abstractESSLLI21/abstractESSLLI1.tex
abstract

@@
 similar to the ones discovered by IAT can be discovered in word
-embeddings. {[}5{]} extends the methodology to a multilingual and
+embeddings. {[}5{]} extend the methodology to a multilingual and
 cross-lingual setting; a similar methodology is employed by {[}3{]}, who
@@
 classes (in religion), change in gender after debiasing is really minor
-with zero still out out of HPDI range, and debiasing improvement for
-race is a small improvement achieved at the price of moving neutral
-terms closer to protected words (there is no space to include these
-results in this abstract).
+with zero still out of HPDI range, and debiasing improvement for race is
+a small improvement achieved at the price of moving neutral terms closer
+to protected words (there is no space to include these results in this
+abstract).
 

### 2021-06-14 abstractESSLLI21/abstractESSLLI1.tex
Adds files

--- 2021-06-14 abstractESSLLI21/abstractESSLLI1.tex
@@
 } \tag{WEAT}
-\end{align}
-
-\normalsize
-\noindent {[}2{]} show that significant biases---thus measured---
-similar to the ones discovered by IAT can be discovered in word
-embeddings. {[}5{]} extends the methodology to a multilingual and
-cross-lingual setting; a similar methodology is employed by {[}3{]}, who
-use word embeddings trained on corpora from different decades. {[}6{]}
-modify WEAT to a multi-class setting, introducing Mean Average Cosine
-(MAC) similarity as a measure of bias (in fact, in the paper they report
-distances rather than similarities). Let \(T = \{t_1, \dots, t_k\}\) be
-a class of protected word embeddings, and let each \(A_j\in A\) be a set
-of attributes stereotypically associated with a protected word). Then:
+\end{align} \normalsize \noindent {[}2{]} show that significant
+biases---thus measured--- similar to the ones discovered by IAT can be
+discovered in word embeddings. {[}5{]} extends the methodology to a
+multilingual and cross-lingual setting; a similar methodology is
+employed by {[}3{]}, who use word embeddings trained on corpora from
+different decades. {[}6{]} modify WEAT to a multi-class setting,
+introducing Mean Average Cosine (MAC) similarity as a measure of bias
+(in fact, in the paper they report distances rather than similarities).
+Let \(T = \{t_1, \dots, t_k\}\) be a class of protected word embeddings,
+and let each \(A_j\in A\) be a set of attributes stereotypically
+associated with a protected word). Then:
 
@@
 \mathsf{MAC}(T,A)  = \frac{1}{\vert T \vert \,\vert A\vert}\sum_{t_i \in T }\sum_{A_j \in A} S(t_i,A_j) \tag{MAC}
-\end{align}
-
-\normalsize  \noindent That is, for each protected word and each
-attribute class, they first take the mean for this protected word and
-all attributes in a given attribute class, and then they take the mean
-of thus obtained means for all the protected words.
+\end{align} \normalsize  \noindent That is, for each protected word and
+each attribute class, they first take the mean for this protected word
+and all attributes in a given attribute class, and then they take the
+mean of thus obtained means for all the protected words.
 
@@
  & \sigma  \sim  dcauchy(0,1) \tag{Priors}
-\end{align}
-
-\normalsize
+\end{align} \normalsize
 

### 2021-06-14 abstractESSLLI21/abstractESSLLI1.Rmd

@@
 \noindent @Caliskan2017semanticsBiases show that significant biases---thus measured--- similar to the ones discovered by IAT can be discovered in word embeddings.
-@Lauscher2019multidimensional extends the methodology to a multilingual and cross-lingual setting; a similar methodology is employed by @Garg2018years, who use word embeddings trained on corpora from different decades.
+@Lauscher2019multidimensional extend the methodology to a multilingual and cross-lingual setting; a similar methodology is employed by @Garg2018years, who use word embeddings trained on corpora from different decades.
 @Manzini2019blackToCriminal modify WEAT to a multi-class setting, introducing Mean Average Cosine (MAC) similarity as a measure of bias (in fact, in the paper  they report distances rather than similarities).
@@
 Moreover, sometimes the differences between associated, different and human, are not very impressive.
-The cost of debiasing turns out sometimes to be that   neutral human predicates get closer to protected classes (in religion),  change in gender after debiasing is really minor with   zero still out out of HPDI range, and debiasing improvement for race is a small improvement   achieved at the price of moving neutral terms closer to protected words (there is no space to include these results in this abstract).
+The cost of debiasing turns out sometimes to be that   neutral human predicates get closer to protected classes (in religion),  change in gender after debiasing is really minor with   zero still out  of HPDI range, and debiasing improvement for race is a small improvement   achieved at the price of moving neutral terms closer to protected words (there is no space to include these results in this abstract).
 

### 2021-06-14 abstractESSLLI21/abstractESSLLI1.tex

@@
 } \tag{WEAT}
-\end{align} \normalsize \noindent {[}2{]} show that significant
-biases---thus measured--- similar to the ones discovered by IAT can be
-discovered in word embeddings. {[}5{]} extends the methodology to a
-multilingual and cross-lingual setting; a similar methodology is
-employed by {[}3{]}, who use word embeddings trained on corpora from
-different decades. {[}6{]} modify WEAT to a multi-class setting,
-introducing Mean Average Cosine (MAC) similarity as a measure of bias
-(in fact, in the paper they report distances rather than similarities).
-Let \(T = \{t_1, \dots, t_k\}\) be a class of protected word embeddings,
-and let each \(A_j\in A\) be a set of attributes stereotypically
-associated with a protected word). Then:
+\end{align}
+
+\normalsize
+\noindent {[}2{]} show that significant biases---thus measured---
+similar to the ones discovered by IAT can be discovered in word
+embeddings. {[}5{]} extend the methodology to a multilingual and
+cross-lingual setting; a similar methodology is employed by {[}3{]}, who
+use word embeddings trained on corpora from different decades. {[}6{]}
+modify WEAT to a multi-class setting, introducing Mean Average Cosine
+(MAC) similarity as a measure of bias (in fact, in the paper they report
+distances rather than similarities). Let \(T = \{t_1, \dots, t_k\}\) be
+a class of protected word embeddings, and let each \(A_j\in A\) be a set
+of attributes stereotypically associated with a protected word). Then:
 

### 2021-06-14 thesis/01-intro.Rmd
Adds intro review fixes

@@
 
-Natural language processing (NLP) is a subfield of computer science that processes and analyzes language in text and speech 
-with the use of modern programming methods. It has practical applications in everyday life as it concerns tasks such as email filters,
-smart assistants, search results, language translations, text analytics and so on. Models used to accomplish these tasks need a lot of
-data to learn from. This data originates from humans activities and historical recordings such as texts, messages or speeches. It turns out
-that in the learning process these models can learn implicit biases that reflect harmful stereotypical thinking still present in modern societies.
+Natural language processing (NLP) is a subfield of computer science that processes and analyzes language in text and speech with the use of modern programming methods.
+It has practical applications in everyday life as it concerns tasks such as email filters, smart assistants, search results, language translations, text analytics and so on.
+Models used to accomplish these tasks need a lot of data to learn from.
+This data originates from human activities and historical recordings such as texts, messages or speeches.
+It turns out that in the learning process these models can learn implicit biases that reflect harmful stereotypical thinking still present in modern societies.
 One can find methods that aim at identifying and measuring hidden biases and/or try to remove them by modifying the models.
@@
 There is considerable amount of literature available on the topic of bias detection and mitigation in NLP models.
-@Bolukbasi2016Man focuses on gender biases that may be observable while investigating the representation of job occupations and gender in terms of their assigned numerical
+@Bolukbasi2016Man focus on gender biases that may be observable while investigating the representation of job occupations and gender in terms of their assigned numerical
 values.
@@
 They check analogies by comparing pairs of word vectors, for example they search for the word complementing the puzzle: man is to doctor as woman is to ...?
-First they subtract word "man" from word "woman" and then
-they search for the ranked list of other words pairs that have similar vectors' difference.
-They also include in the formula a threshold to ensure that the resulting pairs could not be randomly picked.
+First they subtract the word "man" from the word "woman" and then
+they search for the ranked list of other words pairs that have similar vector difference. They also include in the formula a threshold to ensure that the resulting pairs could not be randomly picked.
 
-However, as in @Nissim2019Fair it is pointed out, there are some limitations of this approach.
-According to the authors in practice most of analogies implementations do not return any input words.
+However, as @Nissim2019Fair points out, there are some limitations to this approach.
+In practice, most of analogies implementations do not return any input words.
 This means that it does not make sense to expect the algorithm to return the same profession for both woman and man.
-Therefore this method seems to be limited in terms of bias detection.
-The other problems regard for example the
-choice of pairs and words that are used to detect the presence of discrimination as it is often subjective and without proper justification.
-Additionally the choice of parameter set in @Bolukbasi2016Man formula to ensure that word pairs are not picked by random, is also not justified and changing it drastically
+Therefore, this method seems to be limited in terms of bias detection.
+Other problems are related for example to the choice of pairs and words that are used to detect the presence of discrimination as it is often subjective and without proper justification.
+Additionally the choice of the parameter used in @Bolukbasi2016Man formula to ensure that word pairs are not picked by random, is also not justified and changing it drastically
 influences the results.
 
-@Caliskan2017Semantics touches upon the topic of biases regarding race and gender.
+@Caliskan2017Semantics touch upon the topic of biases regarding race and gender.
 They apply knowledge from well-known psychological studies such as Implicit Association Test to research the relation between human stereotypical thinking and model learnt biases to
@@
 
-@manzini2019black proposes a novel way of using cosine similarity to obtain the information on assumed resemblance between words. They investigate 
+@manzini2019black propose a novel way of using cosine similarity to obtain information on assumed resemblance between words. They investigate 
 an approach that enables them to measure the bias for a class (like gender, religion, race) and express the final result with a single metric.
@@
 
-It is worth noticing the general distinction of biases mentioned in @Caliskan2017Semantics.
-They refer to the publication concerning Implicit Association Test (Greenwald et al., 1998) that measures the strength of associations between concepts or stereotypes by calculating the time of reaction for special tasks.
-It is worth noticing that humans naturally exhibit some biases and that they do not always cause social concern.
+@Caliskan2017Semantics refer to Implicit Association Test (Greenwald et al., 1998) that measures the strength of associations between concepts or stereotypes by calculating the time of reaction for special tasks.
+Humans naturally exhibit some biases which do not always cause social concern.
 One can imagine the intuitive associations between for example insects and flowers, and the feelings of pleasantness or unpleasantness.
-In general, people would rather associate flowers with feeling pleasant than insects, and this preference could be named a bias or
-prejudice in some direction. However, this type of preference does not cause an uproar and is a rather morally neutral case. Unfortunately, there are other
-biases and prejudices that directly influence the quality of other people's lives and therefore they should be taken care of.
+In general, people would rather associate flowers with feeling pleasant than insects, and this preference in a sense is a bias or
+prejudice in some direction.
+However, this type of preference does not cause an uproar and is a rather morally neutral case.
+This example is rather used to show that the methodology used in IAT makes sense.
+Unfortunately, the further studies discover other biases and prejudices that directly influence the quality of other people's lives and therefore they should be taken care of.
 
@@
 With the choice of the definition, implications into the real-life applications may change as well.
-@Mehrabi2019Survey mark out that there exist different types of biases such as historical bias, representation bias, measurement bias (the list is long).
+@Mehrabi2019Survey mark out that there exist different types of biases such as historical bias, representation bias or measurement bias (the list is long).
 This indicates how complex the issue of bias is.
-Without the proper understanding and awareness of the problem, people are prone to unconsciously sustain the bias existence.
+Without the proper understanding and awareness of the problem, people are prone to unconsciously sustain the bias.
 
 @Mehrabi2019Survey also distinguish different types of discrimination, some of them will be briefly described.
-By protected attributes we mean those qualities, traits or characteristics that one cannot legally discriminate against.
+By protected attributes we mean those qualities, traits or characteristics that one should not discriminate against.
 Direct discrimination occurs when protected attributes of individuals explicitly result in non-favorable outcomes toward them.
+Such discrimination takes place for example when one uses an information about someone's religion belief to evaluate whether they are competent for the work they apply for.
 In contrast in indirect discrimination individuals appear to be treated equally but anyway they end up being treated unjustly due to the hidden effects of biases towards their protected attributes.
+An example of such situation is when in a job advert there is a requirement of 10 years experience instead of the list of a specific type of experience and knowledge.
+This job advert can discriminate indirectly against young people who can have equivalent skills but not yet so much work experience.
 Systemic discrimination takes place when policies, customs or behaviors that result from certain culture or organizational structure lead to discrimination against some groups of people.
-Finally, very common statistical discrimination refers to using
+Finally, statistical discrimination consists in to using
 average group statistics to judge person belonging to the group.
 
-The topic of discrimination is entangled with another concept which is fairness. It is essential to grasp some concepts of fairness to take them into
-consideration while designing implementation of some machine learning model. In Mehrabi2019Survey one may notice that depending on the context and application different definitions may be applied.
+The topic of discrimination is entangled with another concept; fairness. It is essential to grasp some concepts of fairness to take them into
+consideration while designing implementation of some machine learning model. In @Mehrabi2019Survey one may notice that depending on the context and application different definitions may be applied.
 
@@
 
-The most popular methods focus on comparing the similarity between words from protected groups and those that are considered to be stereotypical or harmful in some way.
+The most common methods in the literature focus on comparing the similarity between words from protected groups and those that are considered to be stereotypical or harmful in some way.
 One can find in this group methods such as euclidean distance or cosine similarity (which is equivalent to dot product if the
@@
 There are also other ways to detect the effects of biases.
-For example through the investigation of the model performance on certain tasks that validate if the model returns some values
-independently on gender or race or not.
+One can investigate the model performance on certain tasks and validate if the model uses the gender or race information to complete the task.
 
@@
 The currently used methods (such as cosine similarity) make the similarity values often aggregated in a way that may lead to hasty conclusions.
-The averaging of values and the lack of uncertainty may lead to the incomplete picture of the bias situation in the vocabulary.
-
-<!-- \textbf{identify the importance of the proposed research}-->
-One can find a number of articles on negative real-life implications resulting from the presence of unaddressed biases in the machine learning models.
+The averaging of values and the lack of uncertainty may lead to the incomplete picture of the situation.
 
@@
 
-In the paper we indicate how current methods used to detect biases in natural language models are limited from the perspective of Bayesian analysis.
+In the paper we indicate how current methods used to detect biases in natural language models are limited, if we investigate them from the perspective of Bayesian analysis.
 

### 2021-06-14 thesis/_book/_main.tex
Adds intro review fixes

--- 2021-06-10 thesis/_book/_main.tex
@@
 \usepackage[left=3.5cm, right=2.5cm, top=2.5cm, bottom=2.5cm, asymmetric, includeheadfoot]{geometry}
-\usepackage{color}
-\usepackage{fancyvrb}
-\newcommand{\VerbBar}{|}
-\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
-\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
-% Add ',fontsize=\small' for more characters per line
-\usepackage{framed}
-\definecolor{shadecolor}{RGB}{248,248,248}
-\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\BuiltInTok}[1]{#1}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
-\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
-\newcommand{\ExtensionTok}[1]{#1}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ImportTok}[1]{#1}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\NormalTok}[1]{#1}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\RegionMarkerTok}[1]{#1}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
 \usepackage{longtable,booktabs,array}
@@
 \author{Alicja Dobrzeniecka}
-\date{2021-06-10}
+\date{2021-06-14}
 
@@
 
-Natural language processing (NLP) is a subfield of computer science that processes and analyzes language in text and speech
-with the use of modern programming methods. It has practical applications in everyday life as it concerns tasks such as email filters,
-smart assistants, search results, language translations, text analytics and so on. Models used to accomplish these tasks need a lot of
-data to learn from. This data originates from humans activities and historical recordings such as texts, messages or speeches. It turns out
-that in the learning process these models can learn implicit biases that reflect harmful stereotypical thinking still present in modern societies.
+Natural language processing (NLP) is a subfield of computer science that processes and analyzes language in text and speech with the use of modern programming methods.
+It has practical applications in everyday life as it concerns tasks such as email filters, smart assistants, search results, language translations, text analytics and so on.
+Models used to accomplish these tasks need a lot of data to learn from.
+This data originates from human activities and historical recordings such as texts, messages or speeches.
+It turns out that in the learning process these models can learn implicit biases that reflect harmful stereotypical thinking still present in modern societies.
 One can find methods that aim at identifying and measuring hidden biases and/or try to remove them by modifying the models.
@@
 There is considerable amount of literature available on the topic of bias detection and mitigation in NLP models.
-\protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) focuses on gender biases that may be observable while investigating the representation of job occupations and gender in terms of their assigned numerical
+\protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) focus on gender biases that may be observable while investigating the representation of job occupations and gender in terms of their assigned numerical
 values.
@@
 They check analogies by comparing pairs of word vectors, for example they search for the word complementing the puzzle: man is to doctor as woman is to \ldots?
-First they subtract word ``man'' from word ``woman'' and then
-they search for the ranked list of other words pairs that have similar vectors' difference.
-They also include in the formula a threshold to ensure that the resulting pairs could not be randomly picked.
+First they subtract the word ``man'' from the word ``woman'' and then
+they search for the ranked list of other words pairs that have similar vector difference. They also include in the formula a threshold to ensure that the resulting pairs could not be randomly picked.
 
-However, as in \protect\hyperlink{ref-Nissim2019Fair}{Nissim, Noord, \& Goot} (\protect\hyperlink{ref-Nissim2019Fair}{2019}) it is pointed out, there are some limitations of this approach.
-According to the authors in practice most of analogies implementations do not return any input words.
+However, as \protect\hyperlink{ref-Nissim2019Fair}{Nissim, Noord, \& Goot} (\protect\hyperlink{ref-Nissim2019Fair}{2019}) points out, there are some limitations to this approach.
+In practice, most of analogies implementations do not return any input words.
 This means that it does not make sense to expect the algorithm to return the same profession for both woman and man.
-Therefore this method seems to be limited in terms of bias detection.
-The other problems regard for example the
-choice of pairs and words that are used to detect the presence of discrimination as it is often subjective and without proper justification.
-Additionally the choice of parameter set in \protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) formula to ensure that word pairs are not picked by random, is also not justified and changing it drastically
+Therefore, this method seems to be limited in terms of bias detection.
+Other problems are related for example to the choice of pairs and words that are used to detect the presence of discrimination as it is often subjective and without proper justification.
+Additionally the choice of the parameter used in \protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) formula to ensure that word pairs are not picked by random, is also not justified and changing it drastically
 influences the results.
 
-\protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) touches upon the topic of biases regarding race and gender.
+\protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) touch upon the topic of biases regarding race and gender.
 They apply knowledge from well-known psychological studies such as Implicit Association Test to research the relation between human stereotypical thinking and model learnt biases to
@@
 
-\protect\hyperlink{ref-manzini2019black}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-manzini2019black}{2019}) proposes a novel way of using cosine similarity to obtain the information on assumed resemblance between words.
-They investigate
+(\protect\hyperlink{ref-manzini2019black}{\textbf{manzini2019black?}}) propose a novel way of using cosine similarity to obtain information on assumed resemblance between words. They investigate
 an approach that enables them to measure the bias for a class (like gender, religion, race) and express the final result with a single metric.
 
-It is worth noticing the general distinction of biases mentioned in \protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}).
-They refer to the publication concerning Implicit Association Test (Greenwald et al., 1998) that measures the strength of associations between concepts or stereotypes by calculating the time of reaction for special tasks.
-It is worth noticing that humans naturally exhibit some biases and that they do not always cause social concern.
+\protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) refer to Implicit Association Test (Greenwald et al., 1998) that measures the strength of associations between concepts or stereotypes by calculating the time of reaction for special tasks.
+Humans naturally exhibit some biases which do not always cause social concern.
 One can imagine the intuitive associations between for example insects and flowers, and the feelings of pleasantness or unpleasantness.
-In general, people would rather associate flowers with feeling pleasant than insects, and this preference could be named a bias or
-prejudice in some direction. However, this type of preference does not cause an uproar and is a rather morally neutral case. Unfortunately, there are other
-biases and prejudices that directly influence the quality of other people's lives and therefore they should be taken care of.
+In general, people would rather associate flowers with feeling pleasant than insects, and this preference in a sense is a bias or
+prejudice in some direction.
+However, this type of preference does not cause an uproar and is a rather morally neutral case.
+This example is rather used to show that the methodology used in IAT makes sense.
+Unfortunately, the further studies discover other biases and prejudices that directly influence the quality of other people's lives and therefore they should be taken care of.
 
@@
 With the choice of the definition, implications into the real-life applications may change as well.
-\protect\hyperlink{ref-Mehrabi2019Survey}{Mehrabi, Morstatter, Saxena, Lerman, \& Galstyan} (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) mark out that there exist different types of biases such as historical bias, representation bias, measurement bias (the list is long).
+\protect\hyperlink{ref-Mehrabi2019Survey}{Mehrabi, Morstatter, Saxena, Lerman, \& Galstyan} (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) mark out that there exist different types of biases such as historical bias, representation bias or measurement bias (the list is long).
 This indicates how complex the issue of bias is.
-Without the proper understanding and awareness of the problem, people are prone to unconsciously sustain the bias existence.
+Without the proper understanding and awareness of the problem, people are prone to unconsciously sustain the bias.
 
 \protect\hyperlink{ref-Mehrabi2019Survey}{Mehrabi, Morstatter, Saxena, Lerman, \& Galstyan} (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) also distinguish different types of discrimination, some of them will be briefly described.
-By protected attributes we mean those qualities, traits or characteristics that one cannot legally discriminate against.
+By protected attributes we mean those qualities, traits or characteristics that one should not discriminate against.
 Direct discrimination occurs when protected attributes of individuals explicitly result in non-favorable outcomes toward them.
+Such discrimination takes place for example when one uses an information about someone's religion belief to evaluate whether they are competent for the work they apply for.
 In contrast in indirect discrimination individuals appear to be treated equally but anyway they end up being treated unjustly due to the hidden effects of biases towards their protected attributes.
+An example of such situation is when in a job advert there is a requirement of 10 years experience instead of the list of a specific type of experience and knowledge.
+This job advert can discriminate indirectly against young people who can have equivalent skills but not yet so much work experience.
 Systemic discrimination takes place when policies, customs or behaviors that result from certain culture or organizational structure lead to discrimination against some groups of people.
-Finally, very common statistical discrimination refers to using
+Finally, statistical discrimination consists in to using
 average group statistics to judge person belonging to the group.
 
-The topic of discrimination is entangled with another concept which is fairness. It is essential to grasp some concepts of fairness to take them into
-consideration while designing implementation of some machine learning model. In Mehrabi2019Survey one may notice that depending on the context and application different definitions may be applied.
+The topic of discrimination is entangled with another concept; fairness. It is essential to grasp some concepts of fairness to take them into
+consideration while designing implementation of some machine learning model.
+In \protect\hyperlink{ref-Mehrabi2019Survey}{Mehrabi, Morstatter, Saxena, Lerman, \& Galstyan} (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) one may notice that depending on the context and application different definitions may be applied.
 
-The most popular methods focus on comparing the similarity between words from protected groups and those that are considered to be stereotypical or harmful in some way.
+The most common methods in the literature focus on comparing the similarity between words from protected groups and those that are considered to be stereotypical or harmful in some way.
 One can find in this group methods such as euclidean distance or cosine similarity (which is equivalent to dot product if the
@@
 There are also other ways to detect the effects of biases.
-For example through the investigation of the model performance on certain tasks that validate if the model returns some values
-independently on gender or race or not.
+One can investigate the model performance on certain tasks and validate if the model uses the gender or race information to complete the task.
 
 The currently used methods (such as cosine similarity) make the similarity values often aggregated in a way that may lead to hasty conclusions.
-The averaging of values and the lack of uncertainty may lead to the incomplete picture of the bias situation in the vocabulary.
+The averaging of values and the lack of uncertainty may lead to the incomplete picture of the situation.
 
-One can find a number of articles on negative real-life implications resulting from the presence of unaddressed biases in the machine learning models.
-
-In the paper we indicate how current methods used to detect biases in natural language models are limited from the perspective of Bayesian analysis.
+In the paper we indicate how current methods used to detect biases in natural language models are limited, if we investigate them from the perspective of Bayesian analysis.
 
@@
 
+Placeholder
+
 \hypertarget{word-embeddings}{%
 \section{Word embeddings}\label{word-embeddings}}
-
-To understand what cosine similarity measurement is, one first needs to grasp the concept of translating words to a computer-readable form.
-In the field of natural language processsing there are two main types of words representation --- localist and distributed.
-One-hot encoding is an example of a method used to achieve a localist representation of words.
-Here each vector contains information only about a single data point, this is achieved by first mapping categorical values (words) to integers and then to each integers a binary vector is assigned which contains only 0s except for the index of the integer, which is assigned 1.
-An example of a localist representation is:
-
-\begin{longtable}[]{@{}llllll@{}}
-\toprule
-word & 1 & 2 & 3 & 4 & 5 \\
-\midrule
-\endhead
-woman & 1 & 0 & 0 & 0 & 0 \\
-man & 0 & 1 & 0 & 0 & 0 \\
-girl & 0 & 0 & 1 & 0 & 0 \\
-boy & 0 & 0 & 0 & 1 & 0 \\
-monarch & 0 & 0 & 0 & 0 & 1 \\
-\bottomrule
-\end{longtable}
-
-In the example above it is clear that the length of the vectors increases with the number of words in a vocabulary.
-It is not a very computationally efficient representation.
-It has other flaws as well.
-For example, it is unable to capture the resemblance between words appearing in similar contexts.
-
-In contrast to the localist representation, a distributed representation returns vectors that contain continuous values instead of discrete 1s and 0s.
-Word embeddings are a class of various techniques that allow one to represent words as distributed vectors.
-Such learned representations of text have certain properties.
-At least prima facie, they store similar (or at least co-occurring) words close to each other in a vector space.
-An example of distributed representation is:
-
-\begin{longtable}[]{@{}lllll@{}}
-\toprule
-word & 1 & 2 & 3 & 4 \\
-\midrule
-\endhead
-woman & 0.456 & 0.267 & 0.675 & 0.131 \\
-man & 0.451 & 0.897 & 0.472 & 0.088 \\
-girl & 0.604 & 0.262 & 0.414 & 0.706 \\
-boy & 0.279 & 0.172 & 0.475 & 0.010 \\
-monarch & 0.565 & 0.678 & 0.463 & 0.975 \\
-\bottomrule
-\end{longtable}
-
-One of the advantages of using a distributed representation is that one is able to represent an enormous number of concepts with a smaller number of units.
-It is also possible to better capture similarities as words of similar meanings can have similar numeric vectors.
-
-The numbers occurring in such representations are not random.
-They are learned in a process that uses a very shallow neural network.
-There are various types of techniques used for learning the vectors representations.
-One of the most straightforward ones is a skip-gram model.
-Given a word the models tries to predict its neighboring words from the sentence.
-The mathematics behind the process relies on the idea that the prediction concerns the conditional probability of the adjacent words.
-The algorithm tries to minimize the loss function, which penalizes the system for discrepancy with actual co-occurrence frequencies in the corpus.
-One can choose various parameters of the model, such as the window size that determines how many surrounding words the model should predict.
-After preparing such a fitted model one takes only the learned weights from a neural network, and uses them as vectors in a word embeddings representation.
-
-Word embeddings have many applications in natural language processing.
-They are handy in document search and information retrieval.
-They also play their part in improving automatic translations.
-Well learned word representations may also contribute to the improvement of sentiment analysis or spam detection.
 
@@
 
-Cosine similarity is often used as a method of finding out whether vector representations for two words suggest that they are similar or somehow connected.
-Cosine similarity is the cosine of the angle between two vectors: the result of dividing their inner product (dot product usually) by the product of their magnitudes.
-\begin{align} \tag{Sim}
-\mathsf{cosineSimilarity}(A,B) & = \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert}
-\end{align}
-Cosine similarity is considered a proper tool for this operation as its result has a clear connection to geometry and at least for a low number of dimensions may be easily interpreted.
-Using this scale, one can compare vector similarities in a fairly clear manner.
-When the vectors are aligned perpendicularly to each other, their similarity equals 0 (which is the same as the cosine of 90 degrees).
-This tells us that the similarity between the vectors is small.
-As the angle between vectors decreases, cosine similarity approaches one, which stands for the greatest similarity.
-
-One of the limitations of this measure is that it informs us only about similarities between vectors in terms of their orientation.
-However, it is often argued that in comparing words in terms of this metric, the magnitude of vectors may be treated as irrelevant, as the most important information pertains to direction.
-
-In what follows, it is important to distinguish between cosine similarity and cosine distance, defined as:
-\begin{align} \tag{Sim}
-\mathsf{cosineDistance}(A,B) &  = 1 - \mathsf{cosineSimilarity}(A,B)\\
- &  = 1 - \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert} \nonumber
-\end{align}
-
-The greater the similarity between two vectors, the smaller the distance between them.
-The cosine distance ranges between 0 and 2.
-If the vectors are in an opposite direction to each other, the cosine distance is 2.
-And if the vectors are extremely similar then the cosine distance is very close to 0.
-
-It is worth mentioning one more point concerning cosine similarity.
-After the vectors are normalized to have length equal to 1, inner product itself (often dot product) is used to measure the similarity.
-
 \hypertarget{cosine-distance-in-a-one-class-bias-detection}{%
 \section{Cosine distance in a one-class bias detection}\label{cosine-distance-in-a-one-class-bias-detection}}
-
-\protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016})
-define similarity between words as the outcome inner product of their normalized vectors.
-They focus on examining what the geometry of word embedding is in regard to ``he'' and ``she'' words.
-In other words, whether the similarity between those concepts and other words reflects expected gender stereotypes.
-They test this hypothesis by investigating whether there is a connection between word embeddings representing certain professions and words referring to gender.
-They also evaluate whether automatically produced analogies between words reflect the stereotypes as well.
-
-A very vivid way to follow their method of arguing that bias in word embeddings is real is to plot the values of inner product of chosen words.
-The plot below does not originate from the original paper (it is from \url{https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings}) but similar visualization may be found there.
-Data used to create our plot is as follows.
-\newline
-
-Occupations associated with feminine: \textbf{"homemaker", "nurse", "receptionist", "librarian", "socialite", "hairdresser", "nanny", "bookkeeper", "stylist", "housekeeper", "interior designer", "guidance counselor"}
-\newline
-
-Occupations associated with masculine: \textbf{"maestro", "skipper", "protege", "philosopher", "captain", "architect", "financier", "warrior", "broadcaster", "magician", "fighter pilot", "boss"}
-\newline
-
-\vspace{1mm}
-\footnotesize
-
-\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-1-1} \end{center}
-\normalsize
-
-The points in the plot above result from the calculation of the inner product of a chosen vector for a profession word and a vector for a gender word (she or he).
-Inner product of two vectors expresses similarity between words.
-This assumption originates from the geometry and properties of a vector space.
 
@@
 
-\protect\hyperlink{ref-manzini2019black}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-manzini2019black}{2019}) present a different approach towards finding similarities between classes of words.
-The authors claim that texts available online are full of direct or indirect human stereotypes.
-As a result, word embeddings are prone to learn and maybe amplify those biases and propagate them further into AI models that are used for various applications.
-Cosine distance is used in the article as a measure to first argue for the existence of multi-class bias and then to show how through bias mitigation techniques the bias may be decreased.
-
-As the code has been provided by the authors we were able to reconstruct their results.
-The main steps in the procedure are as follows.
-Let us go through an example that refers to the process of hard debiasing on religious attributes.
-
-The example refers to the process of hard debiasing on religious attributes.
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  First we load word embeddings from reddit.US.txt.tok.clean.cleanedforw2v.w2v dataset
-\end{enumerate}
-
-\begin{itemize}
-\tightlist
-\item
-  The word embeddings have only 50 dimensions
-\item
-  The number of individual words from the dataset is 44895
-\end{itemize}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{1}
-\tightlist
-\item
-  The protected group the authors assume should ideally not have high cosine similarity to stereotypical words.
-The word embeddings geometry should not place this group close to harmful stereotypes, if it is to be bias-free.
-  \newline
-\end{enumerate}
-
-For instance, let's look at the religion-related words.
-
-Protected words by religion type:
-
-\textbf{"jew"       : ["judaism", "jew", "synagogue", "torah", "rabbi"]}
-\newline
-\textbf{"christian" : ["christianity", "christian", "church", "bible", "priest"]}
-\newline
-\textbf{"muslim"    : ["islam", "muslim", "mosque", "quran", "imam"]}
-\newline
-
-Stereotypical words by religion type:
-
-\textbf{"jew"       : ["greedy", "cheap", "hairy", "liberal"]}
-\newline
-\textbf{"christian"       : ["judgemental", "conservative", "familial"]}
-\newline
-\textbf{"muslim"       : ["violent", "terrorist", "dirty", "uneducated"]}
-\newline
-
-We have prepared a table presenting the values of cosine distance for each protected word with each attribute (stereotype). The part of the results is shown below.
-
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{religion }\OtherTok{\textless{}{-}} \FunctionTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{]}
-\FunctionTok{colnames}\NormalTok{(religion) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
-                        \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
-\NormalTok{religion}\SpecialCharTok{$}\NormalTok{wordClass }\OtherTok{\textless{}{-}} \FunctionTok{as.factor}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{wordClass)}
-\FunctionTok{levels}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{wordClass) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
-\FunctionTok{head}\NormalTok{(religion)  }\SpecialCharTok{\%\textgreater{}\%}  \FunctionTok{kable}\NormalTok{(}\AttributeTok{format =} \StringTok{"latex"}\NormalTok{,}\AttributeTok{booktabs=}\NormalTok{T,}
-                      \AttributeTok{linesep =} \StringTok{""}\NormalTok{,  }\AttributeTok{escape =} \ConstantTok{FALSE}\NormalTok{, }
-                      \AttributeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
-                      \FunctionTok{kable\_styling}\NormalTok{(}\AttributeTok{latex\_options=}\FunctionTok{c}\NormalTok{(}\StringTok{"scale\_down"}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{table}
-
-\caption{\label{tab:religionTableHeadEarly}Head of the religion dataset.}
-\centering
-\resizebox{\linewidth}{!}{
-\begin{tabular}[t]{lllrrl}
-\toprule
-protectedWord & wordToCompare & wordClass & cosineDistance & cosineSimilarity & connection\\
-\midrule
-judaism & violent & muslim & 0.7141939 & 0.2858061 & different\\
-judaism & terrorist & muslim & 0.7461333 & 0.2538667 & different\\
-judaism & dirty & muslim & 1.2002599 & -0.2002599 & different\\
-judaism & uneducated & muslim & 0.7885469 & 0.2114531 & different\\
-judaism & greedy & jewish & 1.0026172 & -0.0026172 & associated\\
-judaism & cheap & jewish & 1.2323229 & -0.2323229 & associated\\
-\bottomrule
-\end{tabular}}
-\end{table}
-\normalsize
-
-\pagebreak
-
-In the article there was no analysis of individual distances but the general look at the data through the usage of mean.
-The authors introduced a metric that tries to generalize the presence of bias through the classification of multi-class bias in groups of words connected with gender, religion or race.
-In the process they first take the mean of cosine distances between a given protected word and attributes assigned to each stereotype.
-They do not differentiate between stereotypes associated with a word and stereotypes associated with different words (in case of religion, stereotypes characteristic for Christianity has also cosine distance measured with for instance, Judaism or Islam).
-Then, after collecting the list of mean cosine distances, they average the list to obtain one final value representing the whole group, in this example religion, for which the final mean of all mean distances is equal to 0.859.
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{2}
-\tightlist
-\item
-  In the article the authors also try to remove previously defined biases from word embedding.
-\end{enumerate}
-
-First they identify the bias subspace using Principal Component Analysis (PCA) which is a technique for dimensionality reduction.
-It is applied here to choose the subspace that contains the greatest amount of information.
-There can be many subspaces found in a given group, for example in terms of religion one can identify at least a few sets that are to grasp the concept of religion in general:
-\newline
-\textbf{["judaism", "christianity", "islam"]}
-\newline
-\textbf{["jew", "christian", "muslim"]}
-\newline
-\textbf{["synagogue", "church", "mosque"]}
-\newline
-
-The idea is to find a set that provides enough information to create from it a vector representing the concept of religion among words.
-This strategy is based on the idea that different dimensions of vectors contain different types of information and in some words in vector layers (subspaces) the information about religiousness is implicitly conveyed.
-In some cases this knowledge is useful but in the case of harmful stereotypes one does not want to have religion concept in stereotypical words.
-
-After finding bias subspace, they use it to modify the vector values individually so that their cosine distances towards certain words are changed.
-In the case of stereotypes the aim is to make the cosine distances larger so that the association between protected word and harmful stereotype is smaller.
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{3}
-\tightlist
-\item
-  In the final step there is an evaluation of the results. The cosine distances are calculated again but this time using debiased vocabulary. After
-  taking the mean of all distances one final value is obtained and then it is compared with the average value from the beginning.
-If the cosine distances are on average greater than before then it leads the authors to the conclusion that improvement was made.
-As the cosine distance increases it is assumed that the association between protected and stereotypical words decreases.
-\end{enumerate}
-
 \hypertarget{limitations-of-the-approach}{%
 \section{Limitations of the approach}\label{limitations-of-the-approach}}
-
-\textbf{1. Selection of attributes}
-
-The attributes are taken from different sources, there is no principled justification for their choice. From our analysis it will become clear that
-the list is rather uneven.
-
-There is no mention of methodology for deciding on the number of attributes necessary to decide a hypothesis on the given size of dataset.
-There are however some ways to estimate how many samples we need to make sure that the result is significant.
-Our research will show that the numbers used
-are rather insufficient.
-
-\textbf{2. No control while taking the mean of cosines}
-
-The authors use the mean average cosine similarity to check on multi-class similarity between protected word and harmful stereotypes.
-They average the results until they obtain one final value to represent the mean cosine distance between protected word from a given class and the
-attributes of that class. As there is no control of the individual values of cosine distances, one can observe values that have cosine distances
-greater than 1 which would suggest that these compare words do not have significant similarity. However, as there is a lack of control, even such
-words are taken into the account for calculations.
-
-\textbf{3. The lack of uncertainty}
-
-A mean hides this issue and as there are pairs having negative and small similarities and there are those that have similarity equal to 0.5, the resulting calculation seems to be in norm.
-Additionally in such method the uncertainty is also not included which makes it even more difficult to give reasonable interpretations of the results.
-We propose the use of Bayesian method to obtain some understanding of the influence the uncertainty has on the interpretation of final results.
-
-\textbf{4. No word class distinction}
-
-In the original paper words from all three religions were checked with all of the stereotypes which means that there was no distinction on whether the stereotype is associated with given religion or with the other.
-Not all of the stereotypical words should be considered as harmful for all of the religions.
-In our analysis we distinguished between associated stereotypes, not associated (stereotype is valid but for different religion than currently checked) and none (meaning neutral words).
-
-\textbf{5. Interpreting the results}
-
-Assuming for a moment that the value of multi-class cosine distance is correct, one may question the results' interpretation.
-In \protect\hyperlink{ref-manzini2019black}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-manzini2019black}{2019}) Table 2, there are summarized the averages of cosine distance per group (gender, race, religion).
-I would like to focus now on analyzing the values relating to religious biases.
-Here is the fragment of table that refers to that:
-
-\begin{longtable}[]{@{}ll@{}}
-\toprule
-Religion Debiasing & MAC \\
-\midrule
-\endhead
-Biased & 0.859 \\
-Hard Debiased & 0.934 \\
-Soft Debiased (\(\lambda\) = 0.2) & 0.894 \\
-\bottomrule
-\end{longtable}
-
-MAC stand for mean average cosine similarity although in reality the values of cosine distance are stored there.
-What may attract attention is the fact that the value of cosine distance in ``Biased'' category is already quite high even before ``debiasing.'' High cosine distance indicates low cosine similarity between values.
-One could think that average cosine similarity equal to approximately 0.141 is not significant enough to consider it as biased.
-However the authors aim to mitigate ``biases'' in vectors with such great distance to make it even larger.
-Methodologically there is a question on what basis this small similarity is still considered as a proof of bias presence.
-
-This is in general the problem of scale and the lack of universal intervals that could be applied to know whether the cosine similarity is high enough to considered given words more similar that if we chose them on random.
-It seems like similar practice could be beneficial for metrics aimed at measuring similarity between words.
-
-\textbf{6. The curse of dimensionality}
-
-Curse of dimensionality may take place when there is an increase in volume of data that results in adding extra dimensions to the Euclidean space. According to the article
-``\url{https://analyticsindiamag.com/curse-of-dimensionality-and-what-beginners-should-do-to-overcome-it/}''
-as the number of features increases, it may be harder and harder to obtain useful information from the data with the usage of available algorithms.
-One may notice that more data should contribute to greater amount of information but more information also means greater risk of noise and distractions in data.
-At the same time, many times modern solutions are adapted to smaller dimensions and their results in higher ones are not intuitive or may be prone to be mistaken.
-
-Using cosine similarity in high dimensions in word embeddings may also be prone to the curse of dimensionality.
-According to this article ``\url{https://www.researchgate.net/publication/327498046_The_Curse_of_Dimensionality_Inside_Out}''
-there are reasons to consider this phenomenon when searching for word similarities in higher dimensions.
-
-In the article an experiment is conducted that aims at showing how the similarity values and variation change as the number of dimensions increases.
-The hypothesis made in the paper states that two things will happen as the number of dimensions increase, the first one is that effort required to measure cosine similarity will be greater and the second one is that the similarity between data will blur out and have less variation.
-In details, the authors generate random points with increasing number of dimensions where each dimension of a data point is given a value between 0 and 1.
-Then they pick one vector at random from each dimension class and calculate cosine similarity between the chosen vector and the rest of the data.
-Then they check how the variation of values changes as the number of dimensions increases.
-It seems like the more dimensions there are, the smaller the variance and therefore it is less obvious how to interpret the resulting cosine similarities.
-Maybe the scale should be adjusted to the number of dimensions and variance so that it still gives us sensible information about data.
-According to some researches the cosine similarity in high dimensions is not reliable enough to trust it as it may be the case that choosing words on random may result in getting similar values as when picking them consciously.
-
-\begin{figure}
-\centering
-\includegraphics{../images/curseOfDimensionality.png}
-\caption{curse of dimensionality, number of dimensions on the x axis, standard deviation of similarity on the y axis}
-\end{figure}
 
@@
 
+Placeholder
+
 \hypertarget{loading-and-understanding-the-dataset}{%
 \section{Loading and understanding the dataset}\label{loading-and-understanding-the-dataset}}
-
-We start with loading the libraries needed for the analysis.
-
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{library}\NormalTok{(ggplot2)}
-\FunctionTok{library}\NormalTok{(ggthemes)}
-\FunctionTok{library}\NormalTok{(rethinking)}
-\FunctionTok{library}\NormalTok{(tidyverse)}
-\FunctionTok{library}\NormalTok{(ggpubr)}
-\FunctionTok{library}\NormalTok{(kableExtra)}
-\FunctionTok{library}\NormalTok{(dplyr)}
-\FunctionTok{library}\NormalTok{(ggExtra)}
-\FunctionTok{library}\NormalTok{(cowplot)}
-\end{Highlighting}
-\end{Shaded}
-
-\normalsize
-
-We will use the choice of protected words and stereotypical predicates used in REF.
-This is a decent point of departure, not only we want to compare our method to that of REF, but also because this data format is fairly general (as contrasted, say, with a set up for binary stereotypes).
-Note also that the method we develop here can fairly easily be run for different stereotypization patterns.
-Let's start with explaining the method and its deployment using a dataset obtained for the religion-related protected words.
-
-Let's load, clean a bit and inspect the head of the religion dataset we prepared.
-In order to obtain this dataset, we calculated the cosine distance between each protected word and each word from both the bias-related attribute groups, which were used in the original study, and to neutral and human control attributes which we added as control groups.
-For instance, for religion, the bias-related predicates (coming from the original study in REF) include muslim bias attributes, jew bias attributes, christian bias attributes (see a list in the APPENDIX).
-
-We decided to add control groups in the form of two classes --- neutral words and human-related words.
-Without a proper control group it is quite hard to compare the resulting cosine distances and decide on their significance in bias detection.
-We prepared approximately 300\todo{FIX later} more or less neutral words to double-check the prima-facie neutral hypothesis that their cosine similarity to the protected words will oscillate around 0 (that is, the distances will be around 1).
-This provides us with a more reliable point of reference.
-Moreover, we added human attributes that are associated with people in general to investigate whether the smaller cosine distance between protected words and stereotypes can result simply from the fact that the stereotype predicates are associated with humans.
-For two control groups, we have randomly drawn 300 words that do not express any property usually attributed to humans, and human attributes.
-\todo{describe once the words are selected}
-
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{religion }\OtherTok{\textless{}{-}} \FunctionTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{]}
-\FunctionTok{colnames}\NormalTok{(religion) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
-                        \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
-\CommentTok{\#head(religion)}
-\CommentTok{\#library(plyr)}
-\NormalTok{religion}\SpecialCharTok{$}\NormalTok{wordClass }\OtherTok{\textless{}{-}} \FunctionTok{as.factor}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{wordClass)}
-\CommentTok{\#levels(religion$wordClass)}
-\CommentTok{\#religion$wordClass \textless{}{-} mapvalues(religion$wordClass, }
-\CommentTok{\#                          from = levels(religion$wordClass),}
-\CommentTok{\#                          to = c("christian","human","jewish","muslim","neutral"))}
-\FunctionTok{levels}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{wordClass) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
-\FunctionTok{head}\NormalTok{(religion)  }\SpecialCharTok{\%\textgreater{}\%}  \FunctionTok{kable}\NormalTok{(}\AttributeTok{format =} \StringTok{"latex"}\NormalTok{,}\AttributeTok{booktabs=}\NormalTok{T,}
-                      \AttributeTok{linesep =} \StringTok{""}\NormalTok{,  }\AttributeTok{escape =} \ConstantTok{FALSE}\NormalTok{, }
-                      \AttributeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}      \FunctionTok{kable\_styling}\NormalTok{(}\AttributeTok{latex\_options=}\FunctionTok{c}\NormalTok{(}\StringTok{"scale\_down"}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{table}
-
-\caption{\label{tab:religionTableHead}Head of the religion dataset.}
-\centering
-\resizebox{\linewidth}{!}{
-\begin{tabular}[t]{lllrrl}
-\toprule
-protectedWord & wordToCompare & wordClass & cosineDistance & cosineSimilarity & connection\\
-\midrule
-judaism & violent & muslim & 0.7141939 & 0.2858061 & different\\
-judaism & terrorist & muslim & 0.7461333 & 0.2538667 & different\\
-judaism & dirty & muslim & 1.2002599 & -0.2002599 & different\\
-judaism & uneducated & muslim & 0.7885469 & 0.2114531 & different\\
-judaism & greedy & jewish & 1.0026172 & -0.0026172 & associated\\
-judaism & cheap & jewish & 1.2323229 & -0.2323229 & associated\\
-\bottomrule
-\end{tabular}}
-\end{table}
-\normalsize
-
-The \texttt{protectedWord} column contains words from a protected class that (in a perfect world according to the assumptions of the orignal study) should not be associated with harmful stereotypes.
-\texttt{wordToCompare} contains attributes, including stereotypes and control group words.
-For each row we compute the cosine distances between a given protected word and a given attribute word.
-\texttt{wordClass} tells us which class an attribute is supposed to be stereotypically associated with, that is, whether the word from \texttt{wordToCompare} is associated stereotypically with jews, christians or muslims, or whether it belongs to a control group.
-\texttt{cosineDistance} is simply a calculation of the cosine distance between protected word and atrribute.
-\texttt{cosineSimilarity} contains the result of substracting cosine distance from 1.
-\texttt{connection} contains information about the relation type between a protected word and an attribute.
-If the attribute is e.g.~a harmful jewish stereotype and the protected word is also from the judaism group, the connection has value \texttt{associated}.
-If the attribute is still stereotypically jewish, but the protected word comes from another religion, the connection is labelled as \texttt{different}.
-If the attribute belongs to a neutral group then the connection is labelled as \texttt{none} and if an attribute belongs to the \texttt{human} class, then the connection is labelled as \texttt{human}.
 
@@
 
-First let's take a look at the empirical distribution of distances by the connection type, initially ignoring the human control class for now.
-
-\vspace{1mm}
-\footnotesize
-
-\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-3-1} \end{center}
-\normalsize
-
-The first impression is that while there is a shift for associated words towards smaller cosine distances as compared to the neutral words, slightly surprisingly a slightly weaker shift in the same direction is visible for attributes associated with different stereotypes.
-Moreover, the empirical distributions overlap to alarge extent and the means grouped by connection type do not seem too far from each other.
-In fact, as there is a lot of variety in the consine distances (as we will soon see), abd we need to gauge the uncertaintly involved, and to look more carefully at individual protected words to get a better idea of how the cosine distance distribution changes for different attribute groups and different protected classes.
-Now, let's add the human attributes to the picture:
-
-\vspace{1mm}
-\footnotesize
-
-\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-4-1} \end{center}
-\normalsize
-
-\noindent Notice that the distribution for \texttt{human} (even though we did our best not to include in it any stereotype-related atributes) is left-skewed, with much overlap with \texttt{associated} and \texttt{different}, which illustrates the need to take being associated with humans as an important predictor.
-
-Our focus lies in \texttt{connection} as a predictor.
-Morever, later on we'll be interested in looking at the protected words separately, and at protected words split by connection.
-For technical reasons it is useful to represent these factors as integer vectors.
-
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{religion}\SpecialCharTok{$}\NormalTok{con }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{connection)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Warning: NAs introduced by coercion
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{religion}\SpecialCharTok{$}\NormalTok{pw }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{protectedWord)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Warning: NAs introduced by coercion
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{religion}\SpecialCharTok{$}\NormalTok{pwFactor }\OtherTok{\textless{}{-}} \FunctionTok{factor}\NormalTok{(}\FunctionTok{paste0}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{protectedWord, religion}\SpecialCharTok{$}\NormalTok{connection))}
-\NormalTok{religion}\SpecialCharTok{$}\NormalTok{pwIndex }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{pwFactor)}
-\end{Highlighting}
-\end{Shaded}
-
-\normalsize
-
-A short script, \texttt{cleanDataset} to make this faster, so equivalently:
-
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{source}\NormalTok{(}\StringTok{"../functions/cleanDataset.R"}\NormalTok{)}
-\NormalTok{religion }\OtherTok{\textless{}{-}} \FunctionTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{]}
-\NormalTok{religion }\OtherTok{\textless{}{-}} \FunctionTok{cleanDataset}\NormalTok{(religion,}\FunctionTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Warning in cleanDataset(religion, c("christian", "human", "jewish", "muslim", :
-## NAs introduced by coercion
-
-## Warning in cleanDataset(religion, c("christian", "human", "jewish", "muslim", :
-## NAs introduced by coercion
-\end{verbatim}
-
-\normalsize
-
 \hypertarget{looking-at-the-islam-related-words}{%
 \section{Looking at the islam-related words}\label{looking-at-the-islam-related-words}}
-
-For now, let's focus on five protected words related to islam (``imam,'' ``islam,'' ``mosque,'' ``muslim,'' and ``quran'').
-The word list associates with islam four stereotypical attributes (``violent,'' ``terrorist,'' ``uneducated'' and ``dirty'').
-First, we select and plot the empirical distributions for these protected words.
-
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{library}\NormalTok{(tidyverse)}
-\NormalTok{muslimWords }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"imam"}\NormalTok{,}\StringTok{"islam"}\NormalTok{,}\StringTok{"mosque"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"quran"}\NormalTok{)}
-\NormalTok{muslim }\OtherTok{\textless{}{-}}\NormalTok{ religion }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(protectedWord }\SpecialCharTok{\%in\%}\NormalTok{ muslimWords)}
-\FunctionTok{ggplot}\NormalTok{(muslim, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{  cosineDistance, }\AttributeTok{fill =}\NormalTok{ connection, }\AttributeTok{color =}\NormalTok{ connection))}\SpecialCharTok{+}
-  \FunctionTok{geom\_density}\NormalTok{(}\AttributeTok{alpha=}\FloatTok{0.6}\NormalTok{,}\AttributeTok{size =}\NormalTok{ .}\DecValTok{2}\NormalTok{)}\SpecialCharTok{+} 
-  \FunctionTok{scale\_fill\_manual}\NormalTok{(}\AttributeTok{values =} \FunctionTok{c}\NormalTok{(}\StringTok{"orangered4"}\NormalTok{,}\StringTok{"chartreuse4"}\NormalTok{, }\StringTok{"skyblue"}\NormalTok{, }\StringTok{"gray"}\NormalTok{))}\SpecialCharTok{+}
-  \FunctionTok{scale\_x\_continuous}\NormalTok{(}\AttributeTok{breaks =} \FunctionTok{seq}\NormalTok{(}\FloatTok{0.3}\NormalTok{,}\FloatTok{1.5}\NormalTok{, }\AttributeTok{by =} \FloatTok{0.1}\NormalTok{))}\SpecialCharTok{+}\FunctionTok{xlab}\NormalTok{(}\StringTok{"cosine distance"}\NormalTok{)}\SpecialCharTok{+}
-  \FunctionTok{scale\_color\_manual}\NormalTok{(}\AttributeTok{values =} \FunctionTok{c}\NormalTok{(}\StringTok{"orangered4"}\NormalTok{,}\StringTok{"chartreuse4"}\NormalTok{,}\StringTok{"skyblue"}\NormalTok{,}\StringTok{"gray"}\NormalTok{))}\SpecialCharTok{+}
-  \FunctionTok{theme\_tufte}\NormalTok{()}\SpecialCharTok{+}\FunctionTok{ggtitle}\NormalTok{(}\StringTok{"Empirical distribution of distances (muslim)"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-7-1} \end{center}
-\normalsize
-
-\noindent Once we focus on words related to islam, the associated bias seems to be stronger than in the whole dataset. This is a step towards illustrating that the distribution of bias is uneven.
-
-Now, say we want to look at a single protected word.
-Since the dataset also contains comparison multiple control neutral and human attributes, we randomly select only 5 from \texttt{none} and 5 from \texttt{human} control groups of those for the visualisation purposes.
-
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{library}\NormalTok{(tidyverse)}
-\NormalTok{muslimClass }\OtherTok{\textless{}{-}}\NormalTok{ muslim }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(protectedWord }\SpecialCharTok{==} \StringTok{"muslim"}\NormalTok{)}
-\NormalTok{neutralSample }\OtherTok{\textless{}{-}} \FunctionTok{sample\_n}\NormalTok{(}\FunctionTok{filter}\NormalTok{(muslimClass,connection }\SpecialCharTok{==} \StringTok{"none"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
-\NormalTok{humanSample }\OtherTok{\textless{}{-}} \FunctionTok{sample\_n}\NormalTok{(}\FunctionTok{filter}\NormalTok{(muslimClass,connection }\SpecialCharTok{==} \StringTok{"human"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
-\NormalTok{muslimVis }\OtherTok{\textless{}{-}}\NormalTok{ muslimClass }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(connection }\SpecialCharTok{!=} \StringTok{"none"} \SpecialCharTok{\&}\NormalTok{ connection }\SpecialCharTok{!=}\StringTok{"human"}\NormalTok{)}
-\NormalTok{muslimVis }\OtherTok{\textless{}{-}} \FunctionTok{rbind}\NormalTok{(muslimVis,neutralSample,humanSample)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Warning in `[<-.factor`(`*tmp*`, ri, value = structure(c("muslim_stereotype", :
-## invalid factor level, NA generated
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\#we plug in our visualisation script}
-\FunctionTok{source}\NormalTok{(}\StringTok{"../functions/visualisationTools.R"}\NormalTok{)}
-\CommentTok{\#two arguments: dataset and protected word}
-\FunctionTok{visualiseProtected}\NormalTok{(muslimVis,}\StringTok{"muslim"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/tableMuslimActive-1} \end{center}
-\normalsize
-
-Note that the distance between the grey point and the other points is proportional to cosine distance, the non-grey point size is proportional to cosine similarity to the protected word, and color groups by the connection type.
-So for \texttt{muslim} it seems that the stereotypes coming from the word list are fairly well visible.
-To give you some taste of how uneven the dataset is, compare this to what happens with \texttt{priest}.
-
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{library}\NormalTok{(tidyverse)}
-\NormalTok{priestClass }\OtherTok{\textless{}{-}}\NormalTok{ religion }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(protectedWord }\SpecialCharTok{==} \StringTok{"priest"}\NormalTok{)}
-\NormalTok{neutralSample }\OtherTok{\textless{}{-}} \FunctionTok{sample\_n}\NormalTok{(}\FunctionTok{filter}\NormalTok{(priestClass,connection }\SpecialCharTok{==} \StringTok{"none"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
-\NormalTok{humanSample }\OtherTok{\textless{}{-}} \FunctionTok{sample\_n}\NormalTok{(}\FunctionTok{filter}\NormalTok{(priestClass,connection }\SpecialCharTok{==} \StringTok{"human"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
-\NormalTok{priestVis }\OtherTok{\textless{}{-}}\NormalTok{ priestClass }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(connection }\SpecialCharTok{!=} \StringTok{"none"} \SpecialCharTok{\&}\NormalTok{ connection }\SpecialCharTok{!=}\StringTok{"human"}\NormalTok{)}
-\NormalTok{priestVis }\OtherTok{\textless{}{-}} \FunctionTok{rbind}\NormalTok{(priestVis,neutralSample,humanSample)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Warning in `[<-.factor`(`*tmp*`, ri, value = structure(c("muslim_stereotype", :
-## invalid factor level, NA generated
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\#we plug in our visualisation script}
-\FunctionTok{source}\NormalTok{(}\StringTok{"../functions/visualisationTools.R"}\NormalTok{)}
-\CommentTok{\#two arguments: dataset and protected word}
-\FunctionTok{visualiseProtected}\NormalTok{(priestVis,}\StringTok{"priest"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-8-1} \end{center}
-\normalsize
-
-\noindent Here you can see that some human attributes are closer than stereotype attributes, and that there is no clear reason to claim that \texttt{associated} attributes are closer than \texttt{different} or \texttt{human} attributes.
-This, again, illustrates the need of case-by-case analysis with control groups.
-
-The general idea now is that the word lists provided in different pieces of research are just samples of attributes associates with various stereotypes and should be treated as such: the uncertaintly involved and the sample sizes should have clear impact on our estimates.
 
@@
 
-We will now think of cosine distance as the output variable, and
-will build a few bayesian models to compare.
-First, we just build a baseline model which estimates cosine distance to the attributes separately for each protected word.
-The underlying idea is that different protected words migh in general have different relations to all the attributes and this relations should be our point of departure.
-
-Here is the intution behind the mathematical Bayesian model involved.
-Our outcome variable is \texttt{cosine\ difference}, which we take to me normally distributed around the predicted mean for a given protected word (that is, we assume the residuals are normally distributed).
-The simplest model specification is:
-
-\begin{align}
-cosineDistance_i  & \sim dnorm(\mu_i, \sigma) \\
-\mu_i & = m_{pw} \\
-m_{pw} & ~ dnorm(1,.5) \\
-\sigma &\sim  dcauchy(0,1)
-\end{align}
-
-That is, we assume the estimated means might be different for diferent protected words and our prior for the mean and the overal standard deviation are normal with mean 1 and sd=.5 and half-cauchy with parameters \texttt{0,1}.
-Further on we'll also estimate additional impact the connection type may have.
-For this impact we take a slightly skeptical prior centered around 0 distributed normally with sd = 1.
-These are fairly weak and slightly skeptical regularizing priors, which can be illustrated as follows:
-
-\vspace{2mm}
-
-\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/priorsVis-1} \end{center}
-
 \hypertarget{choosing-predictors}{%
 \section{Choosing predictors}\label{choosing-predictors}}
-
-Now we can define and compile the baseline model.
-Its parameters will have a posterior distribution obtained using either Hamiltionian Monte Carlo methods (STAN) available through the \texttt{rethinking} package.
-
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{library}\NormalTok{(rethinking)}
-\FunctionTok{options}\NormalTok{(}\AttributeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\ConstantTok{TRUE}\NormalTok{ )}
-\NormalTok{religionBaseline }\OtherTok{\textless{}{-}} \FunctionTok{ulam}\NormalTok{(}
-  \FunctionTok{alist}\NormalTok{(}
-\NormalTok{    cosineDistance }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(mu,sigma),}
-\NormalTok{    mu }\OtherTok{\textless{}{-}}\NormalTok{ m[pw],}
-\NormalTok{    m[pw] }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
-\NormalTok{    sigma }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
-\NormalTok{  ),}
-  \AttributeTok{data =}\NormalTok{ religion,}
-  \AttributeTok{chains=}\DecValTok{2}\NormalTok{ , }\AttributeTok{iter=}\DecValTok{4000}\NormalTok{ , }\AttributeTok{warmup=}\DecValTok{1000}\NormalTok{,}
-  \AttributeTok{start=} \FunctionTok{list}\NormalTok{(}\AttributeTok{mu =} \DecValTok{1}\NormalTok{, }\AttributeTok{co =} \DecValTok{0}\NormalTok{, }\AttributeTok{sigma=}\NormalTok{ .}\DecValTok{3}\NormalTok{),}
-  \AttributeTok{log\_lik =} \ConstantTok{TRUE}\NormalTok{, }\AttributeTok{cores=}\DecValTok{4}
-\NormalTok{)}
-\CommentTok{\#saving}
-\CommentTok{\#saveRDS(religionBaseline, }
-\CommentTok{\#file = "cosineAnalysis/models/religionBaseline.rds")}
-\end{Highlighting}
-\end{Shaded}
-
-The only reason we need it is the evaluation of connection as a predictor.
-Does including it in o the model improve the situation?
-To investigate this, let's now build a model according to the following specification:
-
-\begin{align}
-cosineDistance_i  & \sim dnorm(\mu_i, \sigma) \\
-\mu_i & = m_{pw} + co_{con}\\
-m_{pw} & ~ dnorm(1,.5) \\
-co_{con} & ~ dnorm(0,1) \\
-\sigma &\sim  dcauchy(0,1)
-\end{align}
-
-\noindent The idea now is that each connection type comes with its own coefficient \(co\) that has impact on mean distances for protected words taken separately.
-
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{library}\NormalTok{(rethinking)}
-\FunctionTok{options}\NormalTok{(}\AttributeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\ConstantTok{TRUE}\NormalTok{ )}
-\NormalTok{religionCoefs }\OtherTok{\textless{}{-}} \FunctionTok{ulam}\NormalTok{(}
-  \FunctionTok{alist}\NormalTok{(}
-\NormalTok{    cosineDistance }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(mu,sigma),}
-\NormalTok{    mu }\OtherTok{\textless{}{-}}\NormalTok{ m[pw] }\SpecialCharTok{+}\NormalTok{ co[con],}
-\NormalTok{    m[pw] }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
-\NormalTok{    co[con] }\SpecialCharTok{\textasciitilde{}}\FunctionTok{dnorm}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
-\NormalTok{    sigma }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
-\NormalTok{  ),}
-  \AttributeTok{data =}\NormalTok{ religion,}
-  \AttributeTok{chains=}\DecValTok{2}\NormalTok{ , }\AttributeTok{iter=}\DecValTok{8000}\NormalTok{ , }\AttributeTok{warmup=}\DecValTok{1000}\NormalTok{, }
-  \AttributeTok{log\_lik =} \ConstantTok{TRUE}
-\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\normalsize
-
-\noindent First, let's see if this model is really better in terms of the Widely Acceptable Information Criterion (WAIC):
-
-\vspace{1mm}
-\footnotesize
-
-\begin{verbatim}
-##                   WAIC SE dWAIC dSE pWAIC weight
-## religionCoefs    -2328 93     0  NA    20      1
-## religionBaseline -2283 95    45  17    16      0
-\end{verbatim}
-
-\normalsize
-
-Clearly, it should be given weight 1 as compared to the baseline model. So far, we've learned that the connection type actually has predictive value. Let's take a look at the coefficient estimates:
-
-\vspace{1mm}
-\footnotesize
-
-\begin{verbatim}
-##              mean        sd       5.5%      94.5%    n_eff    Rhat4
-## co[1] -0.14956420 0.1151675 -0.3261650 0.03741930 294.2033 1.001449
-## co[2] -0.09880543 0.1145024 -0.2736985 0.08813271 291.5044 1.001564
-## co[3] -0.07282752 0.1133894 -0.2447778 0.11158986 287.7820 1.001627
-## co[4] -0.03103179 0.1131420 -0.2034442 0.15268770 286.8283 1.001606
-\end{verbatim}
-
-\normalsize
 
@@
 
-\noindent Let's plot them together with their highest posterior density invervals, for the three topic groups.
-
-\begin{center}
-\begin{figure}[!htb]\centering
-   \begin{minipage}{0.55\textwidth}
-  \includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
-   \end{minipage}
-   \begin {minipage}{0.43\textwidth}
-    \includegraphics[width=7cm]{../images/religionGoogleCoeffs.jpeg}
-   \end{minipage}
-
-  \begin{minipage}{0.55\textwidth}
-\includegraphics[width=7cm]{../images/genderCoeffs.jpeg}
-\end{minipage}
-   \begin {minipage}{0.43\textwidth}
-    \includegraphics[width=7cm]{../images/genderGoogleCoeffs.jpeg}
-   \end{minipage}
-
-  \begin{minipage}{0.55\textwidth}
-\includegraphics[width=7cm]{../images/raceCoeffs.jpeg}
-\end{minipage}
-   \begin {minipage}{0.43\textwidth}
-    \includegraphics[width=7cm]{../images/raceGoogleCoeffs.jpeg}
-   \end{minipage}
-\end{figure}
-
-\end{center}
-
-\noindent What should strike us is that while the mean estimates of the coefficients indeed do differ a bit, usually the highest posterior density invervals all include zero, and so we do not have strong reasons to say that, say, as far as the whole religion dataset is involved, being associated indeed is connected with lower cosine distance.
-A second striking observation is that the estimated impact for associated stereotypes is quite often not too different from the estimated impact of attributes associated with different stereotypes, and both are sometimes not too far from the estimated impact for simply human attributes.
-In general, once the uncertainty involved is taken seriously by using control groups and statistical uncertainty estimation that does not dispose of pointwise data, the picture which focuses only on differences between means of means is too simplistic.
-
-But this doesn't mean important differences for some protected words are not there.
-For one thing, if you start with a word list that is very uneven, the actually not so bad status of some of the protected words might mask a pretty bad situation in which some other protected words are.
-For comparison, let's see what a model focused on words related to islam tells us.
-
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\#this is how we build the model}
-\NormalTok{religion }\OtherTok{\textless{}{-}} \FunctionTok{read.csv}\NormalTok{(}\StringTok{"cosineAnalysis/datasets/religionReddit.csv"}\NormalTok{)[}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{]}
-\FunctionTok{colnames}\NormalTok{(religion) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
-                        \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
-\FunctionTok{levels}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{wordClass) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
-\NormalTok{muslimWords }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"imam"}\NormalTok{,}\StringTok{"islam"}\NormalTok{,}\StringTok{"mosque"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"quran"}\NormalTok{)}
-\NormalTok{muslim }\OtherTok{\textless{}{-}}\NormalTok{ religion }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(protectedWord }\SpecialCharTok{\%in\%}\NormalTok{ muslimWords)}
-\NormalTok{muslim}\SpecialCharTok{$}\NormalTok{protectedWord }\OtherTok{\textless{}{-}} \FunctionTok{droplevels}\NormalTok{(muslim}\SpecialCharTok{$}\NormalTok{protectedWord)}
-\NormalTok{muslim}\SpecialCharTok{$}\NormalTok{pw }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(muslim}\SpecialCharTok{$}\NormalTok{protectedWord)}
-\NormalTok{muslim}\SpecialCharTok{$}\NormalTok{con }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(muslim}\SpecialCharTok{$}\NormalTok{connection)}
-\NormalTok{muslim}\SpecialCharTok{$}\NormalTok{pwFactor }\OtherTok{\textless{}{-}} \FunctionTok{factor}\NormalTok{(}\FunctionTok{paste0}\NormalTok{(muslim}\SpecialCharTok{$}\NormalTok{protectedWord, muslim}\SpecialCharTok{$}\NormalTok{connection))}
-\NormalTok{muslim}\SpecialCharTok{$}\NormalTok{pwIndex }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(muslim}\SpecialCharTok{$}\NormalTok{pwFactor)}
-
-\NormalTok{islamCoefs }\OtherTok{\textless{}{-}} \FunctionTok{ulam}\NormalTok{(}
-  \FunctionTok{alist}\NormalTok{(}
-\NormalTok{    cosineDistance }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(mu,sigma),}
-\NormalTok{    mu }\OtherTok{\textless{}{-}}\NormalTok{ m[pw] }\SpecialCharTok{+}\NormalTok{ co[con],}
-\NormalTok{    m[pw] }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
-\NormalTok{    co[con] }\SpecialCharTok{\textasciitilde{}}\FunctionTok{dnorm}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
-\NormalTok{    sigma }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
-\NormalTok{  ),}
-  \AttributeTok{data =}\NormalTok{ muslim,}
-  \AttributeTok{chains=}\DecValTok{2}\NormalTok{ , }\AttributeTok{iter=}\DecValTok{10000}\NormalTok{ , }\AttributeTok{warmup=}\DecValTok{1000}\NormalTok{, }\AttributeTok{cores =} \DecValTok{4}\NormalTok{,}
-  \AttributeTok{log\_lik =} \ConstantTok{TRUE}
-\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\normalsize
-
-Let's take a look at the coefficients:
-
-\vspace{1mm}
-\footnotesize
-
-\begin{verbatim}
-##                mean        sd       5.5%      94.5%    n_eff    Rhat4
-## co[1] -0.1979930035 0.1708785 -0.4682696 0.07634977 1789.894 1.003445
-## co[2] -0.0334215769 0.1687587 -0.3021954 0.23720938 1738.720 1.003575
-## co[3] -0.0192492753 0.1675754 -0.2840596 0.24860974 1732.907 1.003755
-## co[4] -0.0003911363 0.1670610 -0.2661047 0.26815172 1723.758 1.003837
-\end{verbatim}
-
-\normalsize
-
-\begin{center}
-\includegraphics[width=8cm]{../images/islamCoeffs.jpeg}
-\end{center}
-
-\normalsize
-
-While muslim words were unusual in the sense that the disparity between associated attributes and others is stronger, the evidence is still not conclusive.
-This is because the variation even within islam-related words is large enough (and sample sizes sufficiently small) for all the highest posterior density invervals to still include zeros.
-
-So, it seems, taking a closer look does seem to make a difference. The question is, what happens if we do take a close look at the level of protected words?
-
-\hypertarget{protected-word-level-analysis}{%
-\chapter{Protected-word level analysis}\label{protected-word-level-analysis}}
-
 \hypertarget{model-structure-and-assumptions}{%
 \section{Model structure and assumptions}\label{model-structure-and-assumptions}}
-
-Let's turn then to data analysis that takes a look at protected words separately.
-This time for each combination of a protected word and a connection status we will have a separate mean cosine distance estimate, each coming with its own highest posterior density interval.
-This means we will use indices that are result from all such combinations (and then we will split them up in the model precis to build visualisation, feel free to look at the \texttt{visualiseStats.R} script for details).
-
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{options}\NormalTok{(}\AttributeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\ConstantTok{TRUE}\NormalTok{ ) }\CommentTok{\#removes install pop{-}up request}
-\NormalTok{religion}\SpecialCharTok{$}\NormalTok{pwFactor }\OtherTok{\textless{}{-}} \FunctionTok{factor}\NormalTok{(}\FunctionTok{paste0}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{protectedWord, }\StringTok{"{-}"}\NormalTok{, religion}\SpecialCharTok{$}\NormalTok{connection))}
-\NormalTok{religion}\SpecialCharTok{$}\NormalTok{pwIndex }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{pwFactor)}
-
-\NormalTok{religionSeparate }\OtherTok{\textless{}{-}} \FunctionTok{ulam}\NormalTok{(}
-  \FunctionTok{alist}\NormalTok{(}
-\NormalTok{    cosineDistance }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(mu,sigma),}
-\NormalTok{    mu }\OtherTok{\textless{}{-}}\NormalTok{ c[pwIndex],}
-\NormalTok{    c[pwIndex] }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
-\NormalTok{    sigma }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
-\NormalTok{  ),}
-  \AttributeTok{data =}\NormalTok{ religion,}
-  \AttributeTok{chains=}\DecValTok{2}\NormalTok{ , }\AttributeTok{iter=}\DecValTok{10000}\NormalTok{ , }\AttributeTok{warmup=}\DecValTok{1000}\NormalTok{,}
-  \AttributeTok{start=}\FunctionTok{list}\NormalTok{(}\AttributeTok{no =} \DecValTok{1}\NormalTok{, }\AttributeTok{a =} \DecValTok{0}\NormalTok{, }\AttributeTok{d =} \DecValTok{0}\NormalTok{, }\AttributeTok{sigma=}\NormalTok{ .}\DecValTok{3}\NormalTok{), }\AttributeTok{log\_lik =}  \ConstantTok{TRUE}
-\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\normalsize
-
-\noindent Let's see if the individualized model does better than the previous models in light of WAIC which does add penalty for the number of parameters.
-
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{compareBaselineCoefsSeparate}\OtherTok{\textless{}{-}} \FunctionTok{readRDS}\NormalTok{(}\StringTok{"../datasets/compareBaselineCoefsSeparate.rds"}\NormalTok{)}
-\NormalTok{compareBaselineCoefsSeparate}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-##                   WAIC SE dWAIC dSE pWAIC weight
-## religionSeparate -2400 93     0  NA    60      1
-## religionCoefs    -2328 93    72  29    20      0
-## religionBaseline -2283 95   117  37    16      0
-\end{verbatim}
-
-\normalsize
 
@@
 
-It seems that we do want to prefer this model, despite its relative complication.
-Now, what does it tell us about the protected words?
-Let's visualise the predicted means together with 89\% highest posterior density intervals.
-
-\includegraphics[width=14cm]{../images/visReligionReddit.png}
-
-Before we move on, let's perform analogous analyses for the remaining types of supposed bias: gender and race (the model building is analogous).
-
-\includegraphics[width=14cm]{../images/visGenderReddit.png}
-
-\includegraphics[width=14cm]{../images/visRaceReddit.png}
-
-\todo{Explain where you took the embedding from, cite publications first announcing it.}
-
-\includegraphics[width=14cm]{../images/visReligionGoogle.png}
-
-\includegraphics[width=14cm]{../images/visGenderGoogle.png}
-
-\includegraphics[width=14cm]{../images/visRaceGoogle.png}
-
-\hypertarget{the-role-of-debiasing}{%
-\chapter{The role of debiasing}\label{the-role-of-debiasing}}
-
-\todo{describe which debiasing you used and how it was advertised in the paper}
-
 \hypertarget{dataset-level-coefficients-after-debiasing}{%
 \section{Dataset-level coefficients after debiasing}\label{dataset-level-coefficients-after-debiasing}}
-
-First, let's look at coefficient estimated for the whole datasets, as compared to their estimation prior to debiasing:
-
-\begin{center}
-\begin{figure}[!htb]\centering
-   \begin{minipage}{0.55\textwidth}
-  \includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
-   \end{minipage}
-   \begin {minipage}{0.43\textwidth}
-    \includegraphics[width=7cm]{../images/debiasedReligionRedditCoeffs.jpeg}
-   \end{minipage}
-
-  \begin{minipage}{0.55\textwidth}
-\includegraphics[width=7cm]{../images/genderCoeffs.jpeg}
-\end{minipage}
-   \begin {minipage}{0.43\textwidth}
-    \includegraphics[width=7cm]{../images/debiasedGenderRedditCoeffs.jpeg}
-   \end{minipage}
-
-  \begin{minipage}{0.55\textwidth}
-\includegraphics[width=7cm]{../images/raceCoeffs.jpeg}
-\end{minipage}
-   \begin {minipage}{0.43\textwidth}
-    \includegraphics[width=7cm]{../images/debiasedRaceRedditCoeffs.jpeg}
-   \end{minipage}
-\end{figure}
-
-\end{center}
-
-\todo{comment on this}
-
-Some points to discuss:
-
-\begin{itemize}
-\item
-  humans in religion got closer (should they?)
-\item
-  change in Gender is really minor, still zero out of HPDI range, note how different is very similar
-\item
-  small improvement in race, at the price of moving neutral terms closer
-\end{itemize}
 
@@
 
-Now, perhaps, the effects of debiasing will be better appreciated if we look at the level of protected words.
-After all, the hope is, the situation of extremely ill-positioned protected words have improved?
-
-\includegraphics[width=14cm]{../images/visDebReligionReddit.png}
-
-\includegraphics[width=14cm]{../images/visDebGenderReddit.png}
-
-\includegraphics[width=14cm]{../images/visDebRaceReddit.png}
+\hypertarget{discussion}{%
+\chapter{Discussion}\label{discussion}}
 
@@
 
-\leavevmode\hypertarget{ref-manzini2019black}{}%
-Manzini, T., Lim, Y.
-C., Tsvetkov, Y., \& Black, A.
-W.
-(2019).
-Black is to criminal as caucasian is to police: Detecting and removing multiclass bias in word embeddings.
-Retrieved from \url{http://arxiv.org/abs/1904.04047}
-
 \leavevmode\hypertarget{ref-Mehrabi2019Survey}{}%

### 2021-06-15 thesis/02-cosine.Rmd
Small fix in chapter 2

@@
 One of the most straightforward ones is a skip-gram model.
-Given a word the models tries to predict its neighboring words from the sentence.
+Given a word the model tries to predict its neighboring words from the sentence.
 The mathematics behind the process relies on the idea that the prediction concerns the conditional probability of the adjacent words.
@@
 One can choose various parameters of the model, such as the window size that determines how many surrounding words the model should predict.
-After preparing such a fitted model one takes only the learned weights from a neural network, and uses them as vectors in a word embeddings representation.
+After preparing such a fitted modeln=, one takes only the learned weights from a neural network, and uses them as vectors in a word embeddings representation.
 
@@
 Cosine similarity is the cosine of the angle between two vectors: the result of dividing their inner product (dot product usually) by the product of their magnitudes.
+
+It is worth mentioning one more point concerning cosine similarity.
+After the vectors are normalized to have length equal to 1, inner product itself (often dot product) is used to measure the similarity.
+
 \begin{align} \tag{Sim}
@@
 And if the vectors are extremely similar then the cosine distance is very close to 0.
-
-It is worth mentioning one more point concerning cosine similarity.
-After the vectors are normalized to have length equal to 1, inner product itself (often dot product) is used to measure the similarity.
 

### 2021-06-15 thesis/03-bayesian.Rmd
Develops chapters

@@
+---
+output:
+  pdf_document: default
+  html_document: default
+---
+
 # Walkthrough with the religion dataset
@@
 Without a proper control group it is quite hard to compare the resulting cosine distances and decide on their significance in bias detection.
-We prepared approximately 300\todo{FIX later} more or less neutral words to double-check the prima-facie neutral hypothesis that their cosine similarity to the protected words will oscillate around 0 (that is, the distances will be around 1).
+We prepared approximately 230 neutral words to double-check the prima-facie neutral hypothesis that their cosine similarity to the protected words will oscillate around 0 (that is, the distances will be around 1).
 This provides us with a  more reliable point of reference.
 Moreover, we added human attributes that are associated with people in general to investigate  whether the smaller cosine distance between protected words and stereotypes can result  simply from the fact that the stereotype predicates  are associated with humans.
-For two control groups, we have randomly drawn 300 words that do not express any property usually attributed to humans, and human attributes.
-\todo{describe once the words are selected}
+For two control groups, we have randomly drawn 230 words that do not express any property usually attributed to humans, and human related attributes.
 
@@
 Moreover,  the empirical distributions overlap to alarge extent and the  means grouped by connection type do not seem too far from each other.
-In fact, as there is a lot of  variety in the consine distances (as we will soon see), abd we need to gauge the uncertaintly involved, and  to look more carefully at individual protected words to get a better idea of how the cosine distance distribution changes for different attribute groups and different protected classes.
+In fact, as there is a lot of  variety in the cosine distances (as we will soon see), we need to gauge the uncertainty involved, and  to look more carefully at individual protected words to get a better idea of how the cosine distance distribution changes for different attribute groups and different protected classes.
 Now, let's add the human attributes to the picture:
@@
 
-The general idea now is that the word lists provided in different pieces of research are just samples of attributes associates with various stereotypes and should be treated as such: the uncertaintly involved and the sample sizes should have clear impact on our estimates.
+The general idea now is that the word lists provided in different pieces of research are just samples of attributes associates with various stereotypes and should be treated as such: the uncertainty involved and the sample sizes should have clear impact on our estimates.
 
@@
 We will now think of cosine distance as the output variable, and 
-will build a few bayesian models to compare.
+will build a few Bayesian models to compare.
 First, we just build a baseline model which estimates cosine distance to the attributes separately for each protected word.
-The underlying idea is that different protected words migh in general have different relations to all the attributes and this relations should be our point of departure.
+The underlying idea is that different protected words might in general have different relations to all the attributes and these relations should be our point of departure.
 
-Here is the intution behind the mathematical Bayesian model involved.
+Here is the intuition behind the mathematical Bayesian model involved.
 Our outcome variable is `cosine difference`, which we take to me normally distributed around the predicted mean for a given protected word (that is, we assume the residuals are normally distributed).
@@
 
-\todo{comment on this}
-
-Some points to discuss:
-
-- humans in religion got closer (should they?)
-
-- change in Gender is really minor, still zero out of HPDI range, note how different is very similar
-
-- small improvement in race, at the price of moving neutral terms closer
-
 ## Protected classes  after debiasing

### 2021-06-15 thesis/04-discussion.Rmd
Develops chapters

--- 2021-06-09 thesis/02-cosine.Rmd
@@
-# Cosine similarity and bias detection
+# Discussion
 
-## Word embeddings
-<!-- [based on https://machinelearningmastery.com/what-are-word-embeddings/] -->
+Some points to discuss:
 
-To understand what cosine similarity measurement is, one first needs to grasp the concept of translating words to a computer-readable form.
-In the field of natural language processsing there are two main types of words representation --- localist and distributed.
-One-hot encoding is an example of a method used to achieve a localist representation of words.
-Here each vector contains information only about a single data point, this is achieved by first mapping categorical values (words) to integers and then to each integers a binary vector is assigned which contains only 0s except for the index of the integer, which is assigned 1.
-An example of a localist representation is:
+- humans in religion got closer (should they?)
 
-word | 1 | 2 | 3 | 4 | 5  
-|---|---|---|---|---|---|
-woman | 1 | 0 | 0 | 0 | 0 
-man | 0 | 1 | 0 | 0 | 0 
-girl | 0 | 0 | 1 | 0 | 0 
-boy | 0 | 0 | 0 | 1 | 0 
-monarch | 0 | 0 | 0 | 0 | 1
+- change in Gender is really minor, still zero out of HPDI range, note how different is very similar
 
-In the example above it is clear that the length of the vectors increases with the number of words in a vocabulary.
-It is not a very computationally efficient representation.
-It has other flaws as well.
-For example, it is unable to capture the resemblance between words appearing in similar contexts.
+- small improvement in race, at the price of moving neutral terms closer
 
-In contrast to the localist representation, a distributed representation returns vectors that contain continuous values instead of discrete 1s and 0s.
-Word embeddings are a class of various techniques that allow one to represent words as distributed vectors.
-Such learned representations of text have certain properties.
-At least prima facie, they store similar (or at least co-occurring) words close to each other in a vector space.
-An example of distributed representation is:
-
-word | 1 | 2 | 3 | 4   
-|---|---|---|---|---|
-woman | 0.456 | 0.267 | 0.675 | 0.131
-man | 0.451 | 0.897 | 0.472 | 0.088
-girl | 0.604 | 0.262 | 0.414 | 0.706 
-boy | 0.279 | 0.172 | 0.475 | 0.010 
-monarch | 0.565 | 0.678 | 0.463 | 0.975
-
-One of the advantages of using a distributed representation is that one is able to represent an enormous number of concepts with a smaller number of units.
-It is also possible to better capture similarities as words of similar meanings can have similar numeric vectors.
-
-The numbers occurring in such representations are not random.
-They are learned in a process that uses a very shallow neural network.
-There are various types of techniques used for learning the vectors representations.
-One of the most straightforward ones is a skip-gram model.
-Given a word the models tries to predict its neighboring words from the sentence.
-The mathematics behind the process relies on the idea that the prediction concerns the conditional probability of the adjacent words.
-The algorithm tries to minimize the loss function, which penalizes the system for discrepancy with actual co-occurrence frequencies in the corpus.
-One can choose various parameters of the model, such as the window size that determines how many surrounding words the model should predict.
-After preparing such a fitted model one takes only the learned weights from a neural network, and uses them as vectors in a word embeddings representation.
-
-Word embeddings have many applications in natural language processing.
-They are handy in document search and information retrieval.
-They also play their part in improving automatic translations.
-Well learned word representations may also contribute to the improvement of sentiment analysis or spam detection.
-
-## Cosine similarity and distance
-<!-- [based on https://deepai.org/machine-learning-glossary-and-terms/cosine-similarity] -->
-
-Cosine similarity is often used as a method of finding out whether vector representations for two words suggest that they are similar or somehow connected.
-Cosine similarity is the cosine of the angle between two vectors: the result of dividing their inner product (dot product usually) by the product of their magnitudes.
-\begin{align} \tag{Sim}
-\mathsf{cosineSimilarity}(A,B) & = \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert}
-\end{align}
-<!-- add information about inner product role -->
-Cosine similarity is considered a proper tool for this operation as its result has a clear connection to geometry and at least for a low number of dimensions may be easily interpreted.
-Using this scale, one can compare vector similarities in a fairly clear manner.
-When the vectors are aligned perpendicularly to each other, their similarity equals 0 (which is the same as the cosine of 90 degrees).
-This tells us that the similarity between the vectors is small.
-As the angle between vectors decreases, cosine similarity approaches one, which stands for the greatest similarity.
-
-<!-- todo: plotting exemplary vectors in 2d -->
-One of the limitations of this measure is that it informs us only about similarities between vectors in terms of their orientation.
-However, it is often argued that in comparing words in terms of this metric, the magnitude of vectors may be treated as irrelevant, as the most important information pertains to direction.
-<!-- todo: find reference-->
-
-In what follows, it is important to distinguish between cosine similarity and cosine distance, defined as:
-\begin{align} \tag{Sim}
-\mathsf{cosineDistance}(A,B) &  = 1 - \mathsf{cosineSimilarity}(A,B)\\
- &  = 1 - \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert} \nonumber
-\end{align}
-
-The greater the similarity between two vectors, the smaller the distance between them.
-The cosine distance ranges between 0 and 2.
-If the vectors are in an opposite direction to each other, the cosine distance is 2.
-And if the vectors are extremely similar then the cosine distance is very close to 0.
-
-It is worth mentioning one more point concerning cosine similarity.
-After the vectors are normalized to have length equal to 1, inner product itself (often dot product) is used to measure the similarity.
-
-## Cosine distance in a one-class bias detection
-<!-- https://arxiv.org/pdf/1607.06520.pdf -->
-<!-- source of plot https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings -->
-
-@Bolukbasi2016Man
- define similarity between words as the outcome inner product of their normalized vectors. 
-They focus on examining what the geometry of word embedding is in regard to "he" and "she" words.
-In other words, whether the similarity between those concepts and other words reflects expected gender stereotypes.
-They test this hypothesis by investigating whether there is a connection between word embeddings representing certain professions and words referring to gender.
-They also evaluate whether automatically produced analogies between words reflect the stereotypes as well.
-
-A very vivid way to follow their method of arguing that bias in word embeddings is real is to plot the values of inner product of chosen words.
-The plot below does not originate from the original paper (it is from https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings) but similar visualization may be found there.
-Data used to create our plot is as follows.
-\newline
-
-Occupations associated with feminine: \textbf{"homemaker", "nurse", "receptionist", "librarian", "socialite", "hairdresser", "nanny", "bookkeeper", "stylist", "housekeeper", "interior designer", "guidance counselor"}
-\newline
-
-Occupations associated with masculine: \textbf{"maestro", "skipper", "protege", "philosopher", "captain", "architect", "financier", "warrior", "broadcaster", "magician", "fighter pilot", "boss"}
-\newline
-
-![gender bias in professions as measured by inner product](gender_professions.png)
-
-The points in the plot above result from the calculation of the inner product of a chosen vector for a profession word and a vector for a gender word (she or he).
-Inner product of two vectors expresses similarity between words.
-This assumption originates from the geometry and properties of a vector space.
-
-<!-- todo: shortly describe what they do later and how they verify if it was removed - analogies etc. -->
-
-## Cosine distance in a multi-class bias detection
-<!-- [article https://arxiv.org/pdf/1904.04047.pdf] -->
-
-@manzini2019black present a different approach towards finding similarities between classes of words.
-The authors claim that texts available online are full of direct or indirect human stereotypes.
-As a result, word embeddings are prone to learn and maybe amplify those biases and propagate them further into AI models that are used for various applications.
-Cosine distance is used in the article as a measure to first argue for the existence of multi-class bias and then to show how through bias mitigation techniques the bias may be decreased.
-
-<!-- The methodology used in the paper  -->
-
-As the code has been provided by the authors we were able to reconstruct their results.
-The main steps in the procedure are as follows.
-Let us go through an example that refers to the process of hard debiasing on religious attributes.
-
-The example refers to the process of hard debiasing on religious attributes.
-
-1. First we load word embeddings from reddit.US.txt.tok.clean.cleanedforw2v.w2v dataset
-  - The word embeddings have only 50 dimensions
-  - The number of individual words from the dataset is 44895
-
-2.
-The protected group the authors assume should ideally not have high cosine similarity to stereotypical words.
-The word embeddings geometry should not place this group close to harmful stereotypes, if it is to be bias-free.
-\newline
-
-For instance, let's look at the religion-related words.
-
-Protected words by religion type:
-
-\textbf{"jew"       : ["judaism", "jew", "synagogue", "torah", "rabbi"]}
-\newline
-\textbf{"christian" : ["christianity", "christian", "church", "bible", "priest"]}
-\newline
-\textbf{"muslim"    : ["islam", "muslim", "mosque", "quran", "imam"]}
-\newline
-
-Stereotypical words by religion type:
-
-\textbf{"jew"       : ["greedy", "cheap", "hairy", "liberal"]}
-\newline
-\textbf{"christian"       : ["judgemental", "conservative", "familial"]}
-\newline
-\textbf{"muslim"       : ["violent", "terrorist", "dirty", "uneducated"]}
-\newline
-
-We have prepared a table presenting the values of cosine distance for each protected word with each attribute (stereotype). The part of the results is shown below.
-
-```{python}
-import pandas as pd
-data = pd.read_csv("religion_cosine_table.csv")
-print(data.head())
-
-```
-
-In the article there was no analysis of individual distances but the general look at the data through the usage of mean.
-The authors introduced a metric that tries to generalize the presence of bias through the classification of multi-class bias in groups of words connected with gender, religion or race.
-In the process they first take the mean of cosine distances between a given protected word and attributes assigned to each stereotype.
-They do not differentiate between stereotypes associated with a word and stereotypes associated with different words (in case of religion, stereotypes characteristic for Christianity has also cosine distance measured with for instance, Judaism or Islam).
-Then, after collecting the list of mean cosine distances, they average the list to obtain one final value representing the whole group, in this example religion, for which the final mean of all mean distances is equal to 0.859.
-
-3. In the article the authors also try to remove previously defined biases from word embedding.
-
-First they identify the bias subspace using Principal Component Analysis (PCA) which is a technique for dimensionality reduction.
-It is applied here to choose the subspace that contains the greatest amount of information.
-There can be many subspaces found in a given group, for example in terms of religion one can identify at least a few sets that are to grasp the concept of religion in general:
-\newline
-\textbf{["judaism", "christianity", "islam"]}
-\newline
-\textbf{["jew", "christian", "muslim"]}
-\newline
-\textbf{["synagogue", "church", "mosque"]}
-\newline
-
-The idea is to find a set that provides enough information to create from it a vector representing the concept of religion among words.
-This strategy is based on the idea that different dimensions of vectors contain different types of information and in some words in vector layers (subspaces) the information about religiousness is implicitly conveyed.
-In some cases this knowledge is useful but in the case of harmful stereotypes one does not want to have religion concept in stereotypical words.
-
-After finding bias subspace, they use it to modify the vector values individually so that their cosine distances towards certain words are changed.
-In the case of stereotypes the aim is to make the cosine distances larger so that the association between protected word and harmful stereotype is smaller.
-
-<!-- go through it again -->
-
-4. In the final step there is an evaluation of the results. The cosine distances are calculated again but this time using debiased vocabulary. After
-taking the mean of all distances one final value is obtained and then it is compared with the average value from the beginning.
-If the cosine distances are on average greater than before then it leads the authors to the conclusion that improvement was made.
-As the cosine distance increases it is assumed that the association between protected and stereotypical words decreases.
-
-<!-- add more information on the source of words used -->
-
-## Limitations of the approach
-
-\textbf{1. Selection of attributes}
-
-<!-- TODO: Universal ways to decide on the quality and type of attributes so that the research is significant? -->
-
-The attributes are taken from different sources, there is no principled justification for their choice.
-
-There is no mention of methodology for deciding on the number of attributes necessary to prove a hypothesis on the given size of dataset.
-There are however some ways to estimate how many samples we need to make sure that the result is significant.
-\newline
-
-\textbf{Taking the mean of cosines}
-
-The authors use the mean average cosine similarity to check on multi-class similarity between protected word and harmful stereotypes. Details of the process are described as follows.
-
-When one calculates individual cosine similarity between some chosen word like "christianity" and neutral attributes that are supposed to be biased are actually often having very small and negative similarity value.
-As a result the measurement cosine distance is for such pair of words greater than 1.
-The similarity interpretation for that would suggest that the words are not similar as their cosine distance is so enormous.
-
-Doing a mean hides this issue and as there are pairs having negative and small similarities and there are those that have similarity equal to 0.5, the resulting calculation seems to be in norm.
-Additionally in such method the uncertainty interval is also not included which makes it even more difficult to give reasonable interpretations of the results.
-In the original paper words from all three religions were checked with all of the stereotypes which means that there was no distinction on whether the stereotype is associated with given religion or with the other.
-Not all of the stereotypical words should be considered as harmful for all of the religions.
-In our research we distinguished between associated stereotypes, not associated (stereotype is valid but for different religion than currently checked) and none (meaning neutral words).
-\newline
-
-\textbf{Interpreting the results}
-
-Assuming for a moment that the value of multi-class cosine distance is correct, one may question the results' interpretation.
-In [Black is to Criminal as Caucasian is to Police:Detecting and Removing Multiclass Bias in Word Embeddings] Table 2, there are summarized the averages of cosine distance per group (gender, race, religion).
-I would like to focus  now on analyzing the values relating to religious biases.
-Here is the fragment of table that refers to that:
-
-Religion Debiasing  | MAC
-------------- | -------------
-Biased        | 0.859
-Hard Debiased | 0.934
-Soft Debiased ($\lambda$ = 0.2) | 0.894
-
-MAC stand for mean average cosine similarity although in reality the values of cosine distance are there stored.
-What may attract attention is the fact that the value of cosine distance in "Biased" category is already quite high even before "debiasing".
-High cosine distance indicates low cosine similarity between values.
-One could think that average cosine similarity equal to approximately 0.141 is not significant enough to consider it as biased.
-However the authors aim to mitigate "biases" in vectors with such great distance to make it even larger.
-Methodologically there is a question on what bases this small similarity is still considered as a proof of bias presence.
-
-This is in general the problem of scale and the lack of universal intervals that could be applied to know whether the cosine similarity is high enough to considered given words more similar that if we chose them on random.
-In other metrics like Pearson coefficient we are provided with artificial interval that tries to make the interpretations more objective with a point of reference.
-It seems like similar practic could be beneficial for metrics aimed at measuring similarity between words.
-\newline
-<!-- TODO: numerical test if this distance is significant? check mean mean cosine distances of some random words? -->
-
-\textbf{The curse of dimensionality}
-
-<!-- TODO: -->
-<!-- 1. https://www.cs.princeton.edu/courses/archive/fall13/cos521/lecnotes/lec11.pdf -->
-<!-- 2. https://stats.stackexchange.com/questions/341535/curse-of-dimensionality-does-cosine-similarity-work-better-and-if-so-why -->
-<!-- 3. https://stats.stackexchange.com/questions/21547/distance-metric-and-curse-of-dimensions -->
-<!-- + check if it may be proved in the code -->
-
-Curse of dimensionality may take place when there is an increase in volume of data that results in adding extra dimensions to the Euclidean space. According to the article 
-"https://analyticsindiamag.com/curse-of-dimensionality-and-what-beginners-should-do-to-overcome-it/"
-as the number of features increases, it may be harder and harder to get useful information from the data with the usage of available algorithms.
-One may notice that more data should contribute to greater amount of information but more information also means greater risk of noise and distractions in data.
-At the same time, many times modern solutions are adapted to smaller dimensions and their results in higher ones are not intuitive or may be prone to be mistaken.
-
-Using cosine similarity in high dimensions in word embeddings may also be prone to the curse of dimensionality.
-According to this article "https://www.researchgate.net/publication/327498046_The_Curse_of_Dimensionality_Inside_Out"
-there are reasons to consider this phenomenon when searching for word similarities in higher dimensions.
-
-<!-- Add more information on the experiment -->
-
-In the article an experiment is conducted that aims at showing how the similarity values and variation change as the number of dimensions increases.
-The hypothesis made in the paper states that two things will happen as the number of dimensions increase, the first one is that effort required to measure cosine similarity will be greater and the second one is that the similarity between data will blur out and have less variation.
-In details, the authors generate random points with increasing number of dimensions where each dimension of a data point is given a value between 0 and 1.
-Then they pick one vector on random from each dimension class and calculate cosine similarity between chosen vector and the rest of the data.
-Then they check how the variation of values changes as the number of dimensions increases.
-It seems like the more dimensions there are, the smaller the variance and therefore it is less obvious how to interpret the resulting cosine similarities.
-Maybe the scale should be adjusted to the number of dimensions and variance so that it still gives us sensible information about data.
-According to some publishers the cosine similarity in high dimensions is not reliable enough to trust it as it may be the case that choosing words on random may result in getting similar values as when picking them consciously.
-\newline
-
-![curse of dimensionality, on x-axis number of dimensions, on y-axis standard deviation of similarity](curseOfDimensionality.png)
-
-\textbf{Control group}
-
-How to properly prepare control group in terms of quality and quantity?
-\newline
-<!-- + check if it may be proved in the code -->
-
-\textbf{Verifying other similarity meaures}
-
-Besides cosine similarity, there are other methods used to find the similarity between vectors.
-
-a) TS-SS
-<!-- todo: check if it makes sense to calculate for all of the datasets; interpretations etc. -->
-
-<!-- https://www.researchgate.net/publication/303513110_A_Hybrid_Geometric_Approach_for_Measuring_Similarity_Level_Among_Documents_and_Document_Clustering -->
-
-b) WEAT test 
-<!-- https://docs.responsibly.ai/word-embedding-bias.html#module-responsibly.we.weat -->
-
-<!-- [to check https://github.com/taki0112/Vector_Similarity] -->
-
-<!-- TODO: check in code if they make more sense -->
-
-<!-- ## Bibliography
-- https://www.amazon.com/Language-Processing-Synthesis-Lectures-Technologies/dp/1627052984/ref=as_li_ss_tl?ie=UTF8&qid=1502062931&sr=8-1&keywords=Neural+Network+Methods+in+Natural+Language+Processing&linkCode=sl1&tag=inspiredalgor-20&linkId=d63df073fea3ebe2d405820570b3ff03 ->
-
-<!-- ## *** Additional -->
-
-<!-- ### Man is to Computer Programmer as Woman is to Homemaker? Debiasing Word Embeddings" -->
-<!-- [article https://arxiv.org/pdf/1607.06520.pdf] -->
-
-<!-- Assumptions: -->
-<!-- "We assume there is a set of gender neutral words $N\subseteq W$ such as flight attendant or shoes, which, by definition, are not specific to any gender." -->
-
-<!-- TODO: check if it is useful for the analysis of black is .. -->
-
-<!-- ### Identifying the bias subspace  -->
-
-<!-- Interesting analysis about limits o subspace and research [source https://kawine.github.io/blog/nlp/2019/09/23/bias.html] -->
-
-<!-- Resources to analyze: -->
-<!-- 1. https://arxiv.org/pdf/2005.00965.pdf -->
-<!-- 2. https://arxiv.org/pdf/2009.09435.pdf -->
-<!-- 3. https://arxiv.org/pdf/1904.04047.pdf -->
-<!-- 4. https://www.aclweb.org/anthology/P19-1166.pdf -->
-
-<!-- ### How is cosine similarity entangled with analogy topic in NLP (is it??) -->
-
-<!-- article:  https://arxiv.org/pdf/1905.09866.pdf -->
-
-<!-- Critique of the usage of analogies as a proof for gender biases -->
-
-<!-- TODO: check if there is other argument than that algorithm does not allow it -->
-
-<!-- general thoughts 
-https://www.researchgate.net/publication/349408535_Robustness_and_Reliability_of_Gender_Bias_Assessment_in_Word_Embeddings_The_Role_of_Base_Pairs
-
--
--
--
--->
+- 89% highest posterior density intervals

### 2021-06-15 thesis/_book/_main.tex
Develops chapters

@@
 \usepackage[left=3.5cm, right=2.5cm, top=2.5cm, bottom=2.5cm, asymmetric, includeheadfoot]{geometry}
+\usepackage{color}
+\usepackage{fancyvrb}
+\newcommand{\VerbBar}{|}
+\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+% Add ',fontsize=\small' for more characters per line
+\usepackage{framed}
+\definecolor{shadecolor}{RGB}{248,248,248}
+\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\BuiltInTok}[1]{#1}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
+\newcommand{\ExtensionTok}[1]{#1}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ImportTok}[1]{#1}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\NormalTok}[1]{#1}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\RegionMarkerTok}[1]{#1}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
 \usepackage{longtable,booktabs,array}
@@
 \author{Alicja Dobrzeniecka}
-\date{2021-06-14}
+\date{2021-06-15}
 
@@
 
-Natural language processing (NLP) is a subfield of computer science that processes and analyzes language in text and speech with the use of modern programming methods.
-It has practical applications in everyday life as it concerns tasks such as email filters, smart assistants, search results, language translations, text analytics and so on.
-Models used to accomplish these tasks need a lot of data to learn from.
-This data originates from human activities and historical recordings such as texts, messages or speeches.
-It turns out that in the learning process these models can learn implicit biases that reflect harmful stereotypical thinking still present in modern societies.
-One can find methods that aim at identifying and measuring hidden biases and/or try to remove them by modifying the models.
-There are many different types of models in NLP depending on a task that they are supposed to solve.
-However, all of them need as an input words represented by means of numbers and this is accomplished with word embedding models.
-The models usually assign the values based on the context in which the words appear.
-It means that the input data can have enormous influence on the outcome.
-The biases seem to have their primary source in the way the words are assigned the numerical values.
-
-There is considerable amount of literature available on the topic of bias detection and mitigation in NLP models.
-\protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) focus on gender biases that may be observable while investigating the representation of job occupations and gender in terms of their assigned numerical
-values.
-The authors apply cosine similarity measurement to investigate the phenomenon where (the vectors corresponding to) words related to jobs that are stereotypically associated with a given gender are in fact in the model situated closer to this gender.
-They also use analogy tasks to evaluate if the bias is present in the word embedding model.
-They check analogies by comparing pairs of word vectors, for example they search for the word complementing the puzzle: man is to doctor as woman is to \ldots?
-First they subtract the word ``man'' from the word ``woman'' and then
-they search for the ranked list of other words pairs that have similar vector difference. They also include in the formula a threshold to ensure that the resulting pairs could not be randomly picked.
-
-However, as \protect\hyperlink{ref-Nissim2019Fair}{Nissim, Noord, \& Goot} (\protect\hyperlink{ref-Nissim2019Fair}{2019}) points out, there are some limitations to this approach.
-In practice, most of analogies implementations do not return any input words.
-This means that it does not make sense to expect the algorithm to return the same profession for both woman and man.
-Therefore, this method seems to be limited in terms of bias detection.
-Other problems are related for example to the choice of pairs and words that are used to detect the presence of discrimination as it is often subjective and without proper justification.
-Additionally the choice of the parameter used in \protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) formula to ensure that word pairs are not picked by random, is also not justified and changing it drastically
-influences the results.
-
-\protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) touch upon the topic of biases regarding race and gender.
-They apply knowledge from well-known psychological studies such as Implicit Association Test to research the relation between human stereotypical thinking and model learnt biases to
-discover close relationship between these two. For the evaluation they use Word Embedding Association Test (WEAT) and the Word Embedding Factual Association Test (WEFAT).
-
-(\protect\hyperlink{ref-manzini2019black}{\textbf{manzini2019black?}}) propose a novel way of using cosine similarity to obtain information on assumed resemblance between words. They investigate
-an approach that enables them to measure the bias for a class (like gender, religion, race) and express the final result with a single metric.
-
-\protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) refer to Implicit Association Test (Greenwald et al., 1998) that measures the strength of associations between concepts or stereotypes by calculating the time of reaction for special tasks.
-Humans naturally exhibit some biases which do not always cause social concern.
-One can imagine the intuitive associations between for example insects and flowers, and the feelings of pleasantness or unpleasantness.
-In general, people would rather associate flowers with feeling pleasant than insects, and this preference in a sense is a bias or
-prejudice in some direction.
-However, this type of preference does not cause an uproar and is a rather morally neutral case.
-This example is rather used to show that the methodology used in IAT makes sense.
-Unfortunately, the further studies discover other biases and prejudices that directly influence the quality of other people's lives and therefore they should be taken care of.
-
-One can find various definitions trying to capture what bias and fairness actually are.
-With the choice of the definition, implications into the real-life applications may change as well.
-\protect\hyperlink{ref-Mehrabi2019Survey}{Mehrabi, Morstatter, Saxena, Lerman, \& Galstyan} (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) mark out that there exist different types of biases such as historical bias, representation bias or measurement bias (the list is long).
-This indicates how complex the issue of bias is.
-Without the proper understanding and awareness of the problem, people are prone to unconsciously sustain the bias.
-
-\protect\hyperlink{ref-Mehrabi2019Survey}{Mehrabi, Morstatter, Saxena, Lerman, \& Galstyan} (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) also distinguish different types of discrimination, some of them will be briefly described.
-By protected attributes we mean those qualities, traits or characteristics that one should not discriminate against.
-Direct discrimination occurs when protected attributes of individuals explicitly result in non-favorable outcomes toward them.
-Such discrimination takes place for example when one uses an information about someone's religion belief to evaluate whether they are competent for the work they apply for.
-In contrast in indirect discrimination individuals appear to be treated equally but anyway they end up being treated unjustly due to the hidden effects of biases towards their protected attributes.
-An example of such situation is when in a job advert there is a requirement of 10 years experience instead of the list of a specific type of experience and knowledge.
-This job advert can discriminate indirectly against young people who can have equivalent skills but not yet so much work experience.
-Systemic discrimination takes place when policies, customs or behaviors that result from certain culture or organizational structure lead to discrimination against some groups of people.
-Finally, statistical discrimination consists in to using
-average group statistics to judge person belonging to the group.
-
-The topic of discrimination is entangled with another concept; fairness. It is essential to grasp some concepts of fairness to take them into
-consideration while designing implementation of some machine learning model.
-In \protect\hyperlink{ref-Mehrabi2019Survey}{Mehrabi, Morstatter, Saxena, Lerman, \& Galstyan} (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) one may notice that depending on the context and application different definitions may be applied.
-
-The most common methods in the literature focus on comparing the similarity between words from protected groups and those that are considered to be stereotypical or harmful in some way.
-One can find in this group methods such as euclidean distance or cosine similarity (which is equivalent to dot product if the
-vectors are normalized).
-There are also other ways to detect the effects of biases.
-One can investigate the model performance on certain tasks and validate if the model uses the gender or race information to complete the task.
-
-The currently used methods (such as cosine similarity) make the similarity values often aggregated in a way that may lead to hasty conclusions.
-The averaging of values and the lack of uncertainty may lead to the incomplete picture of the situation.
-
-In the paper we indicate how current methods used to detect biases in natural language models are limited, if we investigate them from the perspective of Bayesian analysis.
-
-Our research enhances the current way in which the bias detection is performed to make sure that it is
-methodologically valid.
-
-The key hypothesis is that greater understanding of data and bias implications can be achieved when Bayesian methods are applied to issue.
+Placeholder
 
@@
 
-Placeholder
-
 \hypertarget{loading-and-understanding-the-dataset}{%
 \section{Loading and understanding the dataset}\label{loading-and-understanding-the-dataset}}
+
+We start with loading the libraries needed for the analysis.
+
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{library}\NormalTok{(ggplot2)}
+\FunctionTok{library}\NormalTok{(ggthemes)}
+\FunctionTok{library}\NormalTok{(rethinking)}
+\FunctionTok{library}\NormalTok{(tidyverse)}
+\FunctionTok{library}\NormalTok{(ggpubr)}
+\FunctionTok{library}\NormalTok{(kableExtra)}
+\FunctionTok{library}\NormalTok{(dplyr)}
+\FunctionTok{library}\NormalTok{(ggExtra)}
+\FunctionTok{library}\NormalTok{(cowplot)}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+We will use the choice of protected words and stereotypical predicates used in REF.
+This is a decent point of departure, not only we want to compare our method to that of REF, but also because this data format is fairly general (as contrasted, say, with a set up for binary stereotypes).
+Note also that the method we develop here can fairly easily be run for different stereotypization patterns.
+Let's start with explaining the method and its deployment using a dataset obtained for the religion-related protected words.
+
+Let's load, clean a bit and inspect the head of the religion dataset we prepared.
+In order to obtain this dataset, we calculated the cosine distance between each protected word and each word from both the bias-related attribute groups, which were used in the original study, and to neutral and human control attributes which we added as control groups.
+For instance, for religion, the bias-related predicates (coming from the original study in REF) include muslim bias attributes, jew bias attributes, christian bias attributes (see a list in the APPENDIX).
+
+We decided to add control groups in the form of two classes --- neutral words and human-related words.
+Without a proper control group it is quite hard to compare the resulting cosine distances and decide on their significance in bias detection.
+We prepared approximately 230 neutral words to double-check the prima-facie neutral hypothesis that their cosine similarity to the protected words will oscillate around 0 (that is, the distances will be around 1).
+This provides us with a more reliable point of reference.
+Moreover, we added human attributes that are associated with people in general to investigate whether the smaller cosine distance between protected words and stereotypes can result simply from the fact that the stereotype predicates are associated with humans.
+For two control groups, we have randomly drawn 230 words that do not express any property usually attributed to humans, and human related attributes.
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{religion }\OtherTok{\textless{}{-}} \FunctionTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{]}
+\FunctionTok{colnames}\NormalTok{(religion) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
+                        \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
+\CommentTok{\#head(religion)}
+\CommentTok{\#library(plyr)}
+\NormalTok{religion}\SpecialCharTok{$}\NormalTok{wordClass }\OtherTok{\textless{}{-}} \FunctionTok{as.factor}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{wordClass)}
+\CommentTok{\#levels(religion$wordClass)}
+\CommentTok{\#religion$wordClass \textless{}{-} mapvalues(religion$wordClass, }
+\CommentTok{\#                          from = levels(religion$wordClass),}
+\CommentTok{\#                          to = c("christian","human","jewish","muslim","neutral"))}
+\FunctionTok{levels}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{wordClass) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
+\FunctionTok{head}\NormalTok{(religion)  }\SpecialCharTok{\%\textgreater{}\%}  \FunctionTok{kable}\NormalTok{(}\AttributeTok{format =} \StringTok{"latex"}\NormalTok{,}\AttributeTok{booktabs=}\NormalTok{T,}
+                      \AttributeTok{linesep =} \StringTok{""}\NormalTok{,  }\AttributeTok{escape =} \ConstantTok{FALSE}\NormalTok{, }
+                      \AttributeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}      \FunctionTok{kable\_styling}\NormalTok{(}\AttributeTok{latex\_options=}\FunctionTok{c}\NormalTok{(}\StringTok{"scale\_down"}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{table}
+
+\caption{\label{tab:religionTableHead}Head of the religion dataset.}
+\centering
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{lllrrl}
+\toprule
+protectedWord & wordToCompare & wordClass & cosineDistance & cosineSimilarity & connection\\
+\midrule
+judaism & violent & muslim & 0.7141939 & 0.2858061 & different\\
+judaism & terrorist & muslim & 0.7461333 & 0.2538667 & different\\
+judaism & dirty & muslim & 1.2002599 & -0.2002599 & different\\
+judaism & uneducated & muslim & 0.7885469 & 0.2114531 & different\\
+judaism & greedy & jewish & 1.0026172 & -0.0026172 & associated\\
+judaism & cheap & jewish & 1.2323229 & -0.2323229 & associated\\
+\bottomrule
+\end{tabular}}
+\end{table}
+\normalsize
+
+The \texttt{protectedWord} column contains words from a protected class that (in a perfect world according to the assumptions of the orignal study) should not be associated with harmful stereotypes.
+\texttt{wordToCompare} contains attributes, including stereotypes and control group words.
+For each row we compute the cosine distances between a given protected word and a given attribute word.
+\texttt{wordClass} tells us which class an attribute is supposed to be stereotypically associated with, that is, whether the word from \texttt{wordToCompare} is associated stereotypically with jews, christians or muslims, or whether it belongs to a control group.
+\texttt{cosineDistance} is simply a calculation of the cosine distance between protected word and atrribute.
+\texttt{cosineSimilarity} contains the result of substracting cosine distance from 1.
+\texttt{connection} contains information about the relation type between a protected word and an attribute.
+If the attribute is e.g.~a harmful jewish stereotype and the protected word is also from the judaism group, the connection has value \texttt{associated}.
+If the attribute is still stereotypically jewish, but the protected word comes from another religion, the connection is labelled as \texttt{different}.
+If the attribute belongs to a neutral group then the connection is labelled as \texttt{none} and if an attribute belongs to the \texttt{human} class, then the connection is labelled as \texttt{human}.
 
@@
 
+First let's take a look at the empirical distribution of distances by the connection type, initially ignoring the human control class for now.
+
+\vspace{1mm}
+\footnotesize
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-2-1} \end{center}
+\normalsize
+
+The first impression is that while there is a shift for associated words towards smaller cosine distances as compared to the neutral words, slightly surprisingly a slightly weaker shift in the same direction is visible for attributes associated with different stereotypes.
+Moreover, the empirical distributions overlap to alarge extent and the means grouped by connection type do not seem too far from each other.
+In fact, as there is a lot of variety in the consine distances (as we will soon see), abd we need to gauge the uncertaintly involved, and to look more carefully at individual protected words to get a better idea of how the cosine distance distribution changes for different attribute groups and different protected classes.
+Now, let's add the human attributes to the picture:
+
+\vspace{1mm}
+\footnotesize
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-3-1} \end{center}
+\normalsize
+
+\noindent Notice that the distribution for \texttt{human} (even though we did our best not to include in it any stereotype-related atributes) is left-skewed, with much overlap with \texttt{associated} and \texttt{different}, which illustrates the need to take being associated with humans as an important predictor.
+
+Our focus lies in \texttt{connection} as a predictor.
+Morever, later on we'll be interested in looking at the protected words separately, and at protected words split by connection.
+For technical reasons it is useful to represent these factors as integer vectors.
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{religion}\SpecialCharTok{$}\NormalTok{con }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{connection)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning: NAs introduced by coercion
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{religion}\SpecialCharTok{$}\NormalTok{pw }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{protectedWord)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning: NAs introduced by coercion
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{religion}\SpecialCharTok{$}\NormalTok{pwFactor }\OtherTok{\textless{}{-}} \FunctionTok{factor}\NormalTok{(}\FunctionTok{paste0}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{protectedWord, religion}\SpecialCharTok{$}\NormalTok{connection))}
+\NormalTok{religion}\SpecialCharTok{$}\NormalTok{pwIndex }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{pwFactor)}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+A short script, \texttt{cleanDataset} to make this faster, so equivalently:
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{source}\NormalTok{(}\StringTok{"../functions/cleanDataset.R"}\NormalTok{)}
+\NormalTok{religion }\OtherTok{\textless{}{-}} \FunctionTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{]}
+\NormalTok{religion }\OtherTok{\textless{}{-}} \FunctionTok{cleanDataset}\NormalTok{(religion,}\FunctionTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning in cleanDataset(religion, c("christian", "human", "jewish", "muslim", :
+## NAs introduced by coercion
+
+## Warning in cleanDataset(religion, c("christian", "human", "jewish", "muslim", :
+## NAs introduced by coercion
+\end{verbatim}
+
+\normalsize
+
 \hypertarget{looking-at-the-islam-related-words}{%
 \section{Looking at the islam-related words}\label{looking-at-the-islam-related-words}}
+
+For now, let's focus on five protected words related to islam (``imam,'' ``islam,'' ``mosque,'' ``muslim,'' and ``quran'').
+The word list associates with islam four stereotypical attributes (``violent,'' ``terrorist,'' ``uneducated'' and ``dirty'').
+First, we select and plot the empirical distributions for these protected words.
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{library}\NormalTok{(tidyverse)}
+\NormalTok{muslimWords }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"imam"}\NormalTok{,}\StringTok{"islam"}\NormalTok{,}\StringTok{"mosque"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"quran"}\NormalTok{)}
+\NormalTok{muslim }\OtherTok{\textless{}{-}}\NormalTok{ religion }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(protectedWord }\SpecialCharTok{\%in\%}\NormalTok{ muslimWords)}
+\FunctionTok{ggplot}\NormalTok{(muslim, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{  cosineDistance, }\AttributeTok{fill =}\NormalTok{ connection, }\AttributeTok{color =}\NormalTok{ connection))}\SpecialCharTok{+}
+  \FunctionTok{geom\_density}\NormalTok{(}\AttributeTok{alpha=}\FloatTok{0.6}\NormalTok{,}\AttributeTok{size =}\NormalTok{ .}\DecValTok{2}\NormalTok{)}\SpecialCharTok{+} 
+  \FunctionTok{scale\_fill\_manual}\NormalTok{(}\AttributeTok{values =} \FunctionTok{c}\NormalTok{(}\StringTok{"orangered4"}\NormalTok{,}\StringTok{"chartreuse4"}\NormalTok{, }\StringTok{"skyblue"}\NormalTok{, }\StringTok{"gray"}\NormalTok{))}\SpecialCharTok{+}
+  \FunctionTok{scale\_x\_continuous}\NormalTok{(}\AttributeTok{breaks =} \FunctionTok{seq}\NormalTok{(}\FloatTok{0.3}\NormalTok{,}\FloatTok{1.5}\NormalTok{, }\AttributeTok{by =} \FloatTok{0.1}\NormalTok{))}\SpecialCharTok{+}\FunctionTok{xlab}\NormalTok{(}\StringTok{"cosine distance"}\NormalTok{)}\SpecialCharTok{+}
+  \FunctionTok{scale\_color\_manual}\NormalTok{(}\AttributeTok{values =} \FunctionTok{c}\NormalTok{(}\StringTok{"orangered4"}\NormalTok{,}\StringTok{"chartreuse4"}\NormalTok{,}\StringTok{"skyblue"}\NormalTok{,}\StringTok{"gray"}\NormalTok{))}\SpecialCharTok{+}
+  \FunctionTok{theme\_tufte}\NormalTok{()}\SpecialCharTok{+}\FunctionTok{ggtitle}\NormalTok{(}\StringTok{"Empirical distribution of distances (muslim)"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-6-1} \end{center}
+\normalsize
+
+\noindent Once we focus on words related to islam, the associated bias seems to be stronger than in the whole dataset. This is a step towards illustrating that the distribution of bias is uneven.
+
+Now, say we want to look at a single protected word.
+Since the dataset also contains comparison multiple control neutral and human attributes, we randomly select only 5 from \texttt{none} and 5 from \texttt{human} control groups of those for the visualisation purposes.
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{library}\NormalTok{(tidyverse)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## -- Attaching packages --------------------------------------- tidyverse 1.3.1 --
+\end{verbatim}
+
+\begin{verbatim}
+## v ggplot2 3.3.3     v purrr   0.3.4
+## v tibble  3.1.2     v dplyr   1.0.6
+## v tidyr   1.1.3     v stringr 1.4.0
+## v readr   1.4.0     v forcats 0.5.1
+\end{verbatim}
+
+\begin{verbatim}
+## -- Conflicts ------------------------------------------ tidyverse_conflicts() --
+## x dplyr::filter() masks stats::filter()
+## x dplyr::lag()    masks stats::lag()
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{muslimClass }\OtherTok{\textless{}{-}}\NormalTok{ muslim }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(protectedWord }\SpecialCharTok{==} \StringTok{"muslim"}\NormalTok{)}
+\NormalTok{neutralSample }\OtherTok{\textless{}{-}} \FunctionTok{sample\_n}\NormalTok{(}\FunctionTok{filter}\NormalTok{(muslimClass,connection }\SpecialCharTok{==} \StringTok{"none"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
+\NormalTok{humanSample }\OtherTok{\textless{}{-}} \FunctionTok{sample\_n}\NormalTok{(}\FunctionTok{filter}\NormalTok{(muslimClass,connection }\SpecialCharTok{==} \StringTok{"human"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
+\NormalTok{muslimVis }\OtherTok{\textless{}{-}}\NormalTok{ muslimClass }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(connection }\SpecialCharTok{!=} \StringTok{"none"} \SpecialCharTok{\&}\NormalTok{ connection }\SpecialCharTok{!=}\StringTok{"human"}\NormalTok{)}
+\NormalTok{muslimVis }\OtherTok{\textless{}{-}} \FunctionTok{rbind}\NormalTok{(muslimVis,neutralSample,humanSample)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning in `[<-.factor`(`*tmp*`, ri, value = structure(c("muslim_stereotype", :
+## invalid factor level, NA generated
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#we plug in our visualisation script}
+\FunctionTok{source}\NormalTok{(}\StringTok{"../functions/visualisationTools.R"}\NormalTok{)}
+\CommentTok{\#two arguments: dataset and protected word}
+\FunctionTok{visualiseProtected}\NormalTok{(muslimVis,}\StringTok{"muslim"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/tableMuslimActive-1} \end{center}
+\normalsize
+
+Note that the distance between the grey point and the other points is proportional to cosine distance, the non-grey point size is proportional to cosine similarity to the protected word, and color groups by the connection type.
+So for \texttt{muslim} it seems that the stereotypes coming from the word list are fairly well visible.
+To give you some taste of how uneven the dataset is, compare this to what happens with \texttt{priest}.
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{library}\NormalTok{(tidyverse)}
+\NormalTok{priestClass }\OtherTok{\textless{}{-}}\NormalTok{ religion }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(protectedWord }\SpecialCharTok{==} \StringTok{"priest"}\NormalTok{)}
+\NormalTok{neutralSample }\OtherTok{\textless{}{-}} \FunctionTok{sample\_n}\NormalTok{(}\FunctionTok{filter}\NormalTok{(priestClass,connection }\SpecialCharTok{==} \StringTok{"none"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
+\NormalTok{humanSample }\OtherTok{\textless{}{-}} \FunctionTok{sample\_n}\NormalTok{(}\FunctionTok{filter}\NormalTok{(priestClass,connection }\SpecialCharTok{==} \StringTok{"human"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
+\NormalTok{priestVis }\OtherTok{\textless{}{-}}\NormalTok{ priestClass }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(connection }\SpecialCharTok{!=} \StringTok{"none"} \SpecialCharTok{\&}\NormalTok{ connection }\SpecialCharTok{!=}\StringTok{"human"}\NormalTok{)}
+\NormalTok{priestVis }\OtherTok{\textless{}{-}} \FunctionTok{rbind}\NormalTok{(priestVis,neutralSample,humanSample)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning in `[<-.factor`(`*tmp*`, ri, value = structure(c("muslim_stereotype", :
+## invalid factor level, NA generated
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#we plug in our visualisation script}
+\FunctionTok{source}\NormalTok{(}\StringTok{"../functions/visualisationTools.R"}\NormalTok{)}
+\CommentTok{\#two arguments: dataset and protected word}
+\FunctionTok{visualiseProtected}\NormalTok{(priestVis,}\StringTok{"priest"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-7-1} \end{center}
+\normalsize
+
+\noindent Here you can see that some human attributes are closer than stereotype attributes, and that there is no clear reason to claim that \texttt{associated} attributes are closer than \texttt{different} or \texttt{human} attributes.
+This, again, illustrates the need of case-by-case analysis with control groups.
+
+The general idea now is that the word lists provided in different pieces of research are just samples of attributes associates with various stereotypes and should be treated as such: the uncertaintly involved and the sample sizes should have clear impact on our estimates.
 
@@
 
+We will now think of cosine distance as the output variable, and
+will build a few bayesian models to compare.
+First, we just build a baseline model which estimates cosine distance to the attributes separately for each protected word.
+The underlying idea is that different protected words migh in general have different relations to all the attributes and this relations should be our point of departure.
+
+Here is the intution behind the mathematical Bayesian model involved.
+Our outcome variable is \texttt{cosine\ difference}, which we take to me normally distributed around the predicted mean for a given protected word (that is, we assume the residuals are normally distributed).
+The simplest model specification is:
+
+\begin{align}
+cosineDistance_i  & \sim dnorm(\mu_i, \sigma) \\
+\mu_i & = m_{pw} \\
+m_{pw} & ~ dnorm(1,.5) \\
+\sigma &\sim  dcauchy(0,1)
+\end{align}
+
+That is, we assume the estimated means might be different for diferent protected words and our prior for the mean and the overal standard deviation are normal with mean 1 and sd=.5 and half-cauchy with parameters \texttt{0,1}.
+Further on we'll also estimate additional impact the connection type may have.
+For this impact we take a slightly skeptical prior centered around 0 distributed normally with sd = 1.
+These are fairly weak and slightly skeptical regularizing priors, which can be illustrated as follows:
+
+\vspace{2mm}
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/priorsVis-1} \end{center}
+
 \hypertarget{choosing-predictors}{%
 \section{Choosing predictors}\label{choosing-predictors}}
+
+Now we can define and compile the baseline model.
+Its parameters will have a posterior distribution obtained using either Hamiltionian Monte Carlo methods (STAN) available through the \texttt{rethinking} package.
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{library}\NormalTok{(rethinking)}
+\FunctionTok{options}\NormalTok{(}\AttributeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\ConstantTok{TRUE}\NormalTok{ )}
+\NormalTok{religionBaseline }\OtherTok{\textless{}{-}} \FunctionTok{ulam}\NormalTok{(}
+  \FunctionTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu }\OtherTok{\textless{}{-}}\NormalTok{ m[pw],}
+\NormalTok{    m[pw] }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\NormalTok{  ),}
+  \AttributeTok{data =}\NormalTok{ religion,}
+  \AttributeTok{chains=}\DecValTok{2}\NormalTok{ , }\AttributeTok{iter=}\DecValTok{4000}\NormalTok{ , }\AttributeTok{warmup=}\DecValTok{1000}\NormalTok{,}
+  \AttributeTok{start=} \FunctionTok{list}\NormalTok{(}\AttributeTok{mu =} \DecValTok{1}\NormalTok{, }\AttributeTok{co =} \DecValTok{0}\NormalTok{, }\AttributeTok{sigma=}\NormalTok{ .}\DecValTok{3}\NormalTok{),}
+  \AttributeTok{log\_lik =} \ConstantTok{TRUE}\NormalTok{, }\AttributeTok{cores=}\DecValTok{4}
+\NormalTok{)}
+\CommentTok{\#saving}
+\CommentTok{\#saveRDS(religionBaseline, }
+\CommentTok{\#file = "cosineAnalysis/models/religionBaseline.rds")}
+\end{Highlighting}
+\end{Shaded}
+
+The only reason we need it is the evaluation of connection as a predictor.
+Does including it in o the model improve the situation?
+To investigate this, let's now build a model according to the following specification:
+
+\begin{align}
+cosineDistance_i  & \sim dnorm(\mu_i, \sigma) \\
+\mu_i & = m_{pw} + co_{con}\\
+m_{pw} & ~ dnorm(1,.5) \\
+co_{con} & ~ dnorm(0,1) \\
+\sigma &\sim  dcauchy(0,1)
+\end{align}
+
+\noindent The idea now is that each connection type comes with its own coefficient \(co\) that has impact on mean distances for protected words taken separately.
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{library}\NormalTok{(rethinking)}
+\FunctionTok{options}\NormalTok{(}\AttributeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\ConstantTok{TRUE}\NormalTok{ )}
+\NormalTok{religionCoefs }\OtherTok{\textless{}{-}} \FunctionTok{ulam}\NormalTok{(}
+  \FunctionTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu }\OtherTok{\textless{}{-}}\NormalTok{ m[pw] }\SpecialCharTok{+}\NormalTok{ co[con],}
+\NormalTok{    m[pw] }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    co[con] }\SpecialCharTok{\textasciitilde{}}\FunctionTok{dnorm}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\NormalTok{  ),}
+  \AttributeTok{data =}\NormalTok{ religion,}
+  \AttributeTok{chains=}\DecValTok{2}\NormalTok{ , }\AttributeTok{iter=}\DecValTok{8000}\NormalTok{ , }\AttributeTok{warmup=}\DecValTok{1000}\NormalTok{, }
+  \AttributeTok{log\_lik =} \ConstantTok{TRUE}
+\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+\noindent First, let's see if this model is really better in terms of the Widely Acceptable Information Criterion (WAIC):
+
+\vspace{1mm}
+\footnotesize
+
+\begin{verbatim}
+##                   WAIC SE dWAIC dSE pWAIC weight
+## religionCoefs    -2328 93     0  NA    20      1
+## religionBaseline -2283 95    45  17    16      0
+\end{verbatim}
+
+\normalsize
+
+Clearly, it should be given weight 1 as compared to the baseline model. So far, we've learned that the connection type actually has predictive value. Let's take a look at the coefficient estimates:
+
+\vspace{1mm}
+\footnotesize
+
+\begin{verbatim}
+##              mean        sd       5.5%      94.5%    n_eff    Rhat4
+## co[1] -0.14956420 0.1151675 -0.3261650 0.03741930 294.2033 1.001449
+## co[2] -0.09880543 0.1145024 -0.2736985 0.08813271 291.5044 1.001564
+## co[3] -0.07282752 0.1133894 -0.2447778 0.11158986 287.7820 1.001627
+## co[4] -0.03103179 0.1131420 -0.2034442 0.15268770 286.8283 1.001606
+\end{verbatim}
+
+\normalsize
 
@@
 
+\noindent Let's plot them together with their highest posterior density invervals, for the three topic groups.
+
+\begin{center}
+\begin{figure}[!htb]\centering
+   \begin{minipage}{0.55\textwidth}
+  \includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
+   \end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/religionGoogleCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/genderCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/genderGoogleCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/raceCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/raceGoogleCoeffs.jpeg}
+   \end{minipage}
+\end{figure}
+
+\end{center}
+
+\noindent What should strike us is that while the mean estimates of the coefficients indeed do differ a bit, usually the highest posterior density invervals all include zero, and so we do not have strong reasons to say that, say, as far as the whole religion dataset is involved, being associated indeed is connected with lower cosine distance.
+A second striking observation is that the estimated impact for associated stereotypes is quite often not too different from the estimated impact of attributes associated with different stereotypes, and both are sometimes not too far from the estimated impact for simply human attributes.
+In general, once the uncertainty involved is taken seriously by using control groups and statistical uncertainty estimation that does not dispose of pointwise data, the picture which focuses only on differences between means of means is too simplistic.
+
+But this doesn't mean important differences for some protected words are not there.
+For one thing, if you start with a word list that is very uneven, the actually not so bad status of some of the protected words might mask a pretty bad situation in which some other protected words are.
+For comparison, let's see what a model focused on words related to islam tells us.
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#this is how we build the model}
+\NormalTok{religion }\OtherTok{\textless{}{-}} \FunctionTok{read.csv}\NormalTok{(}\StringTok{"cosineAnalysis/datasets/religionReddit.csv"}\NormalTok{)[}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{]}
+\FunctionTok{colnames}\NormalTok{(religion) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
+                        \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
+\FunctionTok{levels}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{wordClass) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
+\NormalTok{muslimWords }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"imam"}\NormalTok{,}\StringTok{"islam"}\NormalTok{,}\StringTok{"mosque"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"quran"}\NormalTok{)}
+\NormalTok{muslim }\OtherTok{\textless{}{-}}\NormalTok{ religion }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(protectedWord }\SpecialCharTok{\%in\%}\NormalTok{ muslimWords)}
+\NormalTok{muslim}\SpecialCharTok{$}\NormalTok{protectedWord }\OtherTok{\textless{}{-}} \FunctionTok{droplevels}\NormalTok{(muslim}\SpecialCharTok{$}\NormalTok{protectedWord)}
+\NormalTok{muslim}\SpecialCharTok{$}\NormalTok{pw }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(muslim}\SpecialCharTok{$}\NormalTok{protectedWord)}
+\NormalTok{muslim}\SpecialCharTok{$}\NormalTok{con }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(muslim}\SpecialCharTok{$}\NormalTok{connection)}
+\NormalTok{muslim}\SpecialCharTok{$}\NormalTok{pwFactor }\OtherTok{\textless{}{-}} \FunctionTok{factor}\NormalTok{(}\FunctionTok{paste0}\NormalTok{(muslim}\SpecialCharTok{$}\NormalTok{protectedWord, muslim}\SpecialCharTok{$}\NormalTok{connection))}
+\NormalTok{muslim}\SpecialCharTok{$}\NormalTok{pwIndex }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(muslim}\SpecialCharTok{$}\NormalTok{pwFactor)}
+
+\NormalTok{islamCoefs }\OtherTok{\textless{}{-}} \FunctionTok{ulam}\NormalTok{(}
+  \FunctionTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu }\OtherTok{\textless{}{-}}\NormalTok{ m[pw] }\SpecialCharTok{+}\NormalTok{ co[con],}
+\NormalTok{    m[pw] }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    co[con] }\SpecialCharTok{\textasciitilde{}}\FunctionTok{dnorm}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\NormalTok{  ),}
+  \AttributeTok{data =}\NormalTok{ muslim,}
+  \AttributeTok{chains=}\DecValTok{2}\NormalTok{ , }\AttributeTok{iter=}\DecValTok{10000}\NormalTok{ , }\AttributeTok{warmup=}\DecValTok{1000}\NormalTok{, }\AttributeTok{cores =} \DecValTok{4}\NormalTok{,}
+  \AttributeTok{log\_lik =} \ConstantTok{TRUE}
+\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+Let's take a look at the coefficients:
+
+\vspace{1mm}
+\footnotesize
+
+\begin{verbatim}
+##                mean        sd       5.5%      94.5%    n_eff    Rhat4
+## co[1] -0.1979930035 0.1708785 -0.4682696 0.07634977 1789.894 1.003445
+## co[2] -0.0334215769 0.1687587 -0.3021954 0.23720938 1738.720 1.003575
+## co[3] -0.0192492753 0.1675754 -0.2840596 0.24860974 1732.907 1.003755
+## co[4] -0.0003911363 0.1670610 -0.2661047 0.26815172 1723.758 1.003837
+\end{verbatim}
+
+\normalsize
+
+\begin{center}
+\includegraphics[width=8cm]{../images/islamCoeffs.jpeg}
+\end{center}
+
+\normalsize
+
+While muslim words were unusual in the sense that the disparity between associated attributes and others is stronger, the evidence is still not conclusive.
+This is because the variation even within islam-related words is large enough (and sample sizes sufficiently small) for all the highest posterior density invervals to still include zeros.
+
+So, it seems, taking a closer look does seem to make a difference. The question is, what happens if we do take a close look at the level of protected words?
+
+\hypertarget{protected-word-level-analysis}{%
+\chapter{Protected-word level analysis}\label{protected-word-level-analysis}}
+
 \hypertarget{model-structure-and-assumptions}{%
 \section{Model structure and assumptions}\label{model-structure-and-assumptions}}
+
+Let's turn then to data analysis that takes a look at protected words separately.
+This time for each combination of a protected word and a connection status we will have a separate mean cosine distance estimate, each coming with its own highest posterior density interval.
+This means we will use indices that are result from all such combinations (and then we will split them up in the model precis to build visualisation, feel free to look at the \texttt{visualiseStats.R} script for details).
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{options}\NormalTok{(}\AttributeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\ConstantTok{TRUE}\NormalTok{ ) }\CommentTok{\#removes install pop{-}up request}
+\NormalTok{religion}\SpecialCharTok{$}\NormalTok{pwFactor }\OtherTok{\textless{}{-}} \FunctionTok{factor}\NormalTok{(}\FunctionTok{paste0}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{protectedWord, }\StringTok{"{-}"}\NormalTok{, religion}\SpecialCharTok{$}\NormalTok{connection))}
+\NormalTok{religion}\SpecialCharTok{$}\NormalTok{pwIndex }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{pwFactor)}
+
+\NormalTok{religionSeparate }\OtherTok{\textless{}{-}} \FunctionTok{ulam}\NormalTok{(}
+  \FunctionTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu }\OtherTok{\textless{}{-}}\NormalTok{ c[pwIndex],}
+\NormalTok{    c[pwIndex] }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\NormalTok{  ),}
+  \AttributeTok{data =}\NormalTok{ religion,}
+  \AttributeTok{chains=}\DecValTok{2}\NormalTok{ , }\AttributeTok{iter=}\DecValTok{10000}\NormalTok{ , }\AttributeTok{warmup=}\DecValTok{1000}\NormalTok{,}
+  \AttributeTok{start=}\FunctionTok{list}\NormalTok{(}\AttributeTok{no =} \DecValTok{1}\NormalTok{, }\AttributeTok{a =} \DecValTok{0}\NormalTok{, }\AttributeTok{d =} \DecValTok{0}\NormalTok{, }\AttributeTok{sigma=}\NormalTok{ .}\DecValTok{3}\NormalTok{), }\AttributeTok{log\_lik =}  \ConstantTok{TRUE}
+\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+\noindent Let's see if the individualized model does better than the previous models in light of WAIC which does add penalty for the number of parameters.
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{compareBaselineCoefsSeparate}\OtherTok{\textless{}{-}} \FunctionTok{readRDS}\NormalTok{(}\StringTok{"../datasets/compareBaselineCoefsSeparate.rds"}\NormalTok{)}
+\NormalTok{compareBaselineCoefsSeparate}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+##                   WAIC SE dWAIC dSE pWAIC weight
+## religionSeparate -2400 93     0  NA    60      1
+## religionCoefs    -2328 93    72  29    20      0
+## religionBaseline -2283 95   117  37    16      0
+\end{verbatim}
+
+\normalsize
 
@@
 
+It seems that we do want to prefer this model, despite its relative complication.
+Now, what does it tell us about the protected words?
+Let's visualise the predicted means together with 89\% highest posterior density intervals.
+
+\includegraphics[width=14cm]{../images/visReligionReddit.png}
+
+Before we move on, let's perform analogous analyses for the remaining types of supposed bias: gender and race (the model building is analogous).
+
+\includegraphics[width=14cm]{../images/visGenderReddit.png}
+
+\includegraphics[width=14cm]{../images/visRaceReddit.png}
+
+\todo{Explain where you took the embedding from, cite publications first announcing it.}
+
+\includegraphics[width=14cm]{../images/visReligionGoogle.png}
+
+\includegraphics[width=14cm]{../images/visGenderGoogle.png}
+
+\includegraphics[width=14cm]{../images/visRaceGoogle.png}
+
+\hypertarget{the-role-of-debiasing}{%
+\chapter{The role of debiasing}\label{the-role-of-debiasing}}
+
+\todo{describe which debiasing you used and how it was advertised in the paper}
+
 \hypertarget{dataset-level-coefficients-after-debiasing}{%
 \section{Dataset-level coefficients after debiasing}\label{dataset-level-coefficients-after-debiasing}}
+
+First, let's look at coefficient estimated for the whole datasets, as compared to their estimation prior to debiasing:
+
+\begin{center}
+\begin{figure}[!htb]\centering
+   \begin{minipage}{0.55\textwidth}
+  \includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
+   \end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/debiasedReligionRedditCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/genderCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/debiasedGenderRedditCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/raceCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/debiasedRaceRedditCoeffs.jpeg}
+   \end{minipage}
+\end{figure}
+
+\end{center}
+
+\todo{comment on this}
+
+Some points to discuss:
+
+\begin{itemize}
+\item
+  humans in religion got closer (should they?)
+\item
+  change in Gender is really minor, still zero out of HPDI range, note how different is very similar
+\item
+  small improvement in race, at the price of moving neutral terms closer
+\end{itemize}
 
@@
 
+Now, perhaps, the effects of debiasing will be better appreciated if we look at the level of protected words.
+After all, the hope is, the situation of extremely ill-positioned protected words have improved?
+
+\includegraphics[width=14cm]{../images/visDebReligionReddit.png}
+
+\includegraphics[width=14cm]{../images/visDebGenderReddit.png}
+
+\includegraphics[width=14cm]{../images/visDebRaceReddit.png}
+
 \hypertarget{discussion}{%
@@
 
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\hypertarget{ref-Bolukbasi2016Man}{}%
-Bolukbasi, T., Chang, K.-W., Zou, J.
-Y., Saligrama, V., \& Kalai, A.
-(2016).
-Man is to computer programmer as woman is to homemaker?
-Debiasing word embeddings.
-\emph{CoRR}, \emph{abs/1607.06520}.
-Retrieved from \url{http://arxiv.org/abs/1607.06520}
-
-\leavevmode\hypertarget{ref-Caliskan2017Semantics}{}%
-Islam, A.
-C., Bryson, J.
-J., \& Narayanan, A.
-(2016).
-Semantics derived automatically from language corpora necessarily contain human biases.
-\emph{CoRR}, \emph{abs/1608.07187}.
-Retrieved from \url{http://arxiv.org/abs/1608.07187}
-
-\leavevmode\hypertarget{ref-Mehrabi2019Survey}{}%
-Mehrabi, N., Morstatter, F., Saxena, N., Lerman, K., \& Galstyan, A.
-(2019).
-A survey on bias and fairness in machine learning.
-\emph{CoRR}, \emph{abs/1908.09635}.
-Retrieved from \url{http://arxiv.org/abs/1908.09635}
-
-\leavevmode\hypertarget{ref-Nissim2019Fair}{}%
-Nissim, M., Noord, R.
-van, \& Goot, R.
-van der.
-(2019).
-Fair is better than sensational: Man is to doctor as woman is to doctor.
-\emph{CoRR}, \emph{abs/1905.09866}.
-Retrieved from \url{http://arxiv.org/abs/1905.09866}
-
-\end{CSLReferences}
-
 \end{document}

### 2021-06-15 thesis/03-bayesian.Rmd
Just saving file
Adds fixes

@@
 
-\todo{Explain where you took the embedding from, cite publications first announcing it.}
+We first encountered the GoogleNews-vectors-negative300 word embedding model in @Bolukbasi2016Man.
+They use it for their calculations as mentioned in "https://github.com/tolga-b/debiaswe".
+We decided to compare the results obtained with the use of Reddit model and the ones that we got by GoogleNews model.
+The details of GoogleNews model can be found here "https://code.google.com/archive/p/word2vec/".
+One of the main differences between the models is that Reddit word embeggings have 50 dimensions and GoogleNews word embeddings have 300 dimensions.
+As the dimension is increasing, the vector can capture much more information but this is not always the case.
+According to some researchers the amount of dimensions should be chosen by taking into account the type of corpus and its features.
+In our case both Reddit and GoogleNews models are already trained and ready to use so we do not analyse further the choice of hyperparameters to train them.
 
@@
 \includegraphics[width=14cm]{../images/visDebRaceReddit.png}
+
+- humans in religion got closer (should they?)
+
+- change in Gender is really minor, still zero out of HPDI range, note how different is very similar
+
+- small improvement in race, at the price of moving neutral terms closer

### 2021-06-15 thesis/04-discussion.Rmd
Adds fixes

@@
 
-Some points to discuss:
-
-- humans in religion got closer (should they?)
-
-- change in Gender is really minor, still zero out of HPDI range, note how different is very similar
-
-- small improvement in race, at the price of moving neutral terms closer
+It is worth diving deeper into some observations that arise while analyzing the results.
 

### 2021-06-15 thesis/_book/_main.tex
Adds fixes

@@
 \fi
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
+  % set entry spacing
+  \ifnum #2 > 0
+  \setlength{\parskip}{#2\baselineskip}
+  \fi
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
@@
 Moreover, the empirical distributions overlap to alarge extent and the means grouped by connection type do not seem too far from each other.
-In fact, as there is a lot of variety in the consine distances (as we will soon see), abd we need to gauge the uncertaintly involved, and to look more carefully at individual protected words to get a better idea of how the cosine distance distribution changes for different attribute groups and different protected classes.
+In fact, as there is a lot of variety in the cosine distances (as we will soon see), we need to gauge the uncertainty involved, and to look more carefully at individual protected words to get a better idea of how the cosine distance distribution changes for different attribute groups and different protected classes.
 Now, let's add the human attributes to the picture:
@@
 We will now think of cosine distance as the output variable, and
-will build a few bayesian models to compare.
+will build a few Bayesian models to compare.
 First, we just build a baseline model which estimates cosine distance to the attributes separately for each protected word.
-The underlying idea is that different protected words migh in general have different relations to all the attributes and this relations should be our point of departure.
+The underlying idea is that different protected words might in general have different relations to all the attributes and these relations should be our point of departure.
 
-Here is the intution behind the mathematical Bayesian model involved.
+Here is the intuition behind the mathematical Bayesian model involved.
 Our outcome variable is \texttt{cosine\ difference}, which we take to me normally distributed around the predicted mean for a given protected word (that is, we assume the residuals are normally distributed).
@@
 
-\todo{Explain where you took the embedding from, cite publications first announcing it.}
+We first encountered the GoogleNews-vectors-negative300 word embedding model in \protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}).
+They use it for their calculations as mentioned in \url{https://github.com/tolga-b/debiaswe}.
+We decided to compare the results obtained with the use of Reddit model and the ones that we got by GoogleNews model.
+The details of GoogleNews model can be found here \url{https://code.google.com/archive/p/word2vec/}.
+One of the main differences between the models is that Reddit word embeggings have 50 dimensions and GoogleNews word embeddings have 300 dimensions.
+As the dimension is increasing, the vector can capture much more information but this is not always the case.
+According to some researchers the amount of dimensions should be chosen by taking into account the type of corpus and its features.
+In our case both Reddit and GoogleNews models are already trained and ready to use so we do not analyse further the choice of hyperparameters to train them.
 
@@
 
-\todo{comment on this}
-
-Some points to discuss:
-
-\begin{itemize}
-\item
-  humans in religion got closer (should they?)
-\item
-  change in Gender is really minor, still zero out of HPDI range, note how different is very similar
-\item
-  small improvement in race, at the price of moving neutral terms closer
-\end{itemize}
-
 \hypertarget{protected-classes-after-debiasing}{%
@@
 
+\begin{itemize}
+\item
+  humans in religion got closer (should they?)
+\item
+  change in Gender is really minor, still zero out of HPDI range, note how different is very similar
+\item
+  small improvement in race, at the price of moving neutral terms closer
+\end{itemize}
+
 \hypertarget{discussion}{%
@@
 
+It is worth diving deeper into some observations that arise while analyzing the results.
+
+\begin{itemize}
+\tightlist
+\item
+  89\% highest posterior density intervals
+\end{itemize}
+
+\hypertarget{refs}{}
+\begin{CSLReferences}{1}{0}
+\leavevmode\hypertarget{ref-Bolukbasi2016Man}{}%
+Bolukbasi, T., Chang, K.-W., Zou, J.
+Y., Saligrama, V., \& Kalai, A.
+(2016).
+Man is to computer programmer as woman is to homemaker?
+Debiasing word embeddings.
+\emph{CoRR}, \emph{abs/1607.06520}.
+Retrieved from \url{http://arxiv.org/abs/1607.06520}
+
+\end{CSLReferences}
+
 \end{document}

### 2021-06-15 thesis/03-bayesian.Rmd
Adds up-to-date changes

--- 2021-06-15 thesis/03-bayesian.Rmd
@@
 
-\todo{Explain where you took the embedding from, cite publications first announcing it.}
+We first encountered the GoogleNews word embedding model in @Bolukbasi2016Man.
+They use it for their calculations as written in [Tolga Code](https://github.com/tolga-b/debiaswe).
+We decided to compare the results obtained with the use of Reddit L2 model and the ones that we got by GoogleNews model.
+The details of the model can be found here [Google News model](https://code.google.com/archive/p/word2vec/).
+One of the main differences between the models is that Reddit word embeggings have 50 dimensions and GoogleNews word embeddings have 300 dimensions.
+As the dimension is increasing, the vector can capture much more information but this is not always the case.
+According to some researchers the amount of dimensions should be chosen by taking into account the type of corpus and its features.
+In our case both Reddit and GoogleNews models are already trained and ready to use so we do not analyse further the choice of their hyperparameters.
 
@@
 
-\todo{describe which debiasing you used and how it was advertised in the paper}
+We also applied to the word embeddings hard debiasing method found in @manzini2019black.
+After debiasing the Reddit word embeddings we calculated the cosine distance again to see if there are any significant changes in terms of the similarity between protected words and the attributes.
 
@@
 
+One may observe very slight changes in the estimated coeﬀicients.
+It seems that in religion dataset the changes are the most significant as the mean moves towards zero which stands for no similarity between words.
+
 ## Protected classes  after debiasing
@@
 \includegraphics[width=14cm]{../images/visDebRaceReddit.png}
+
+One may notice that in religion data set
+
+It is worth paying attention to gender debiased dataset. The change is really minor, still zero out of HPDI range. Note how `different` attributes have high similarity value.
+
+At the same time, there is a small improvement in race, at the price of moving neutral terms closer
+
+In @manzini2019black it is mentioned that debiasing with this method that is based on bias component removal may be insufficient.
+The reason is that words containing similar biases may still cluster together even after debiasing.

### 2021-06-15 thesis/04-discussion.Rmd
Adds up-to-date changes

@@
 
-- 89% highest posterior density intervals
+- the distributions of cosine distances of whole population and individual words can vary enormously
+- including uncertainty shows how in some cases it is not clear why a word is according to this method
+classified as anomalous
+- debiasing with the method provided by @manzini2019black does not solve the bias issues
+
+What we propose and how it can help:
+-
+-
+
+# Summary
+
+Future work:
+-
+
+<!-- 
+remarks:
+- delete papers from github/hide them
+-
+
+-->

### 2021-06-15 thesis/_book/_main.tex
Adds up-to-date changes

@@
 
-We first encountered the GoogleNews-vectors-negative300 word embedding model in \protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}).
-They use it for their calculations as mentioned in \url{https://github.com/tolga-b/debiaswe}.
-We decided to compare the results obtained with the use of Reddit model and the ones that we got by GoogleNews model.
-The details of GoogleNews model can be found here \url{https://code.google.com/archive/p/word2vec/}.
+We first encountered the GoogleNews word embedding model in \protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}).
+They use it for their calculations as mentioned in \href{https://github.com/tolga-b/debiaswe}{Tolga Code}.
+We decided to compare the results obtained with the use of Reddit L2 model and the ones that we got by GoogleNews model.
+The details of GoogleNews model can be found here \href{https://code.google.com/archive/p/word2vec/}{Google News model}.
 One of the main differences between the models is that Reddit word embeggings have 50 dimensions and GoogleNews word embeddings have 300 dimensions.
@@
 According to some researchers the amount of dimensions should be chosen by taking into account the type of corpus and its features.
-In our case both Reddit and GoogleNews models are already trained and ready to use so we do not analyse further the choice of hyperparameters to train them.
+In our case both Reddit and GoogleNews models are already trained and ready to use so we do not analyse further the choice of their hyperparameters.
 
@@
 
-\todo{describe which debiasing you used and how it was advertised in the paper}
+We also applied to the word embeddings hard debiasing method found in (\protect\hyperlink{ref-manzini2019black}{\textbf{manzini2019black?}}).
+After debiasing the Reddit word embeddings we calculated the cosine distance again to see if there are any significant changes in terms of the similarity between protected words and the attributes.
 
@@
 
+One may observe very slight changes in the estimated coeﬀicients.
+It seems that in religion dataset the changes are the most significant as the mean moves towards zero which stands for no similarity between words.
+
 \hypertarget{protected-classes-after-debiasing}{%
@@
 
+One may notice
+
 \begin{itemize}
@@
 
+In (\protect\hyperlink{ref-manzini2019black}{\textbf{manzini2019black?}}) it is mentioned that debiasing with this method that is based on bias component removal may be insufficient.
+The reason is that words containing similar biases may still cluster together even after debiasing.
+
 \hypertarget{discussion}{%
@@
 \item
-  89\% highest posterior density intervals
+  the distributions of cosine distances of whole population and individual words can vary enormously
+\item
+  including uncertainty shows how in some cases it is not clear why a word is according to this method
+  classified as anomalous
+\item
+  debiasing with the method provided by (\protect\hyperlink{ref-manzini2019black}{\textbf{manzini2019black?}}) does not solve the bias issues
 \end{itemize}
+
+\hypertarget{summary}{%
+\chapter{Summary}\label{summary}}
+
+\hypertarget{future-work}{%
+\section{Future work:}\label{future-work}}
 

### 2021-06-15 thesis/02-cosine.Rmd
Adds up-to-date changes

@@
 
-@manzini2019black present a different approach towards finding similarities between classes of words.
+@Manzini2019blackToCriminal present a different approach towards finding similarities between classes of words.
 The authors claim that texts available online are full of direct or indirect human stereotypes.
@@
 Assuming for a moment that the value of multi-class cosine distance is correct, one may question the results' interpretation.
-In @manzini2019black Table 2, there are summarized the averages of cosine distance per group (gender, race, religion).
+In @Manzini2019blackToCriminal Table 2, there are summarized the averages of cosine distance per group (gender, race, religion).
 I would like to focus  now on analyzing the values relating to religious biases.

### 2021-06-15 thesis/03-bayesian.Rmd
Adds up-to-date changes

@@
 
-We also applied to the word embeddings hard debiasing method found in @manzini2019black.
+We also applied to the word embeddings hard debiasing method found in @Manzini2019blackToCriminal.
+As it was mentioned before, debiasing consists of two components: identifying the bias subspace and then removing this vector subspace from the chosen embeddings.
+The aim is to increase the cosine distance between protected words and the set of attributes.
 After debiasing the Reddit word embeddings we calculated the cosine distance again to see if there are any significant changes in terms of the similarity between protected words and the attributes.
@@
 
-One may notice that in religion data set
+One may notice that in a religion data set most of the distances for `associated` words slightly increased.
+Some of the distances from `different` class are still smaller than for the `associated` class.
+The distances for all classes seem in most cases to cluster together even after debiasing.
+The reason may be that uncertainty is quite high in `associated` and `different` class because the number of samples is so small.
+Additionally in some cases distances from `human` class are smaller than for the `assosiated` class which also makes it less clear how to interpret the results, if the distances are really able to catch the bias presence.
 
-It is worth paying attention to gender debiased dataset. The change is really minor, still zero out of HPDI range. Note how `different` attributes have high similarity value.
+It is worth paying attention to the gender debiased dataset.
+The change is really minor, still zero out of HPDI range.
+The `different` attributes still have high similarity value.
+It seems that the distances for gender protected words are still clustered together which suggests that this method is not able to catch the complex bias presence.
+One could conclude that co-occurrence of certain terms does not always mean direct connection in terms of terms associations.
+The fact that some of stereotypically associated with female attributes still have high similarity with male protected words suggest that other techniques should be applied to detect and remove the bias.
 
-At the same time, there is a small improvement in race, at the price of moving neutral terms closer
+At the same time, there is a small improvement in race dataset.
+In some cases the distances just cluster together.
+In general distances for `assosiated` class are still higher than those for the rest of classes.
+One may assume that there reason for so small change is that either the metric does not indicate properly biases or that the issue is more complex and subtracting subspaces is not enough.
 
-In @manzini2019black it is mentioned that debiasing with this method that is based on bias component removal may be insufficient.
+In @Manzini2019blackToCriminal it is mentioned that debiasing with this method that is based on bias component removal may be insufficient.
 The reason is that words containing similar biases may still cluster together even after debiasing.

### 2021-06-15 thesis/04-discussion.Rmd
Adds up-to-date changes

--- 2021-06-15 thesis/04-discussion.Rmd
@@
 
-- 89% highest posterior density intervals
+- the distributions of cosine distances of whole population and individual words can vary enormously
+- including uncertainty shows how in some cases it is not clear why a word is according to this method
+classified as anomalous
+- debiasing with the method provided by @Manzini2019blackToCriminal does not solve the bias issues
+
+One should remember that there is no baseline for the interpretation of the averaged cosine distances in @Manzini2019blackToCriminalToCriminal.
+One may assume that if the cosine distance is close to 1 then it is a desired outcome as it means, according to cosine similarity assumptions, that there is almost no similarity between the words.
+However what does it mean to be close to 0?
+If the averaged cosine similarity is equal to 80, then should we still debiase it?
+It is unclear what the criteria are.
+On one hand, it seems to be beneficial when the outcome is simplified as it is easier to compare results with one value per set.
+On the other hand, it is prone to misunderstanding of how to interpret the results and what threshold to assume.
+
+We propose the use of Bayesian method to introduce the uncertainty measure in bias detection.
+
+# Summary
+
+The aim of this paper was to show the analysis of bias detection methods used in modern natural language processing projects. 
+In the further research it is worth to dive deeper into some issues or to examine new ideas that occurred during our work.
+
+<!-- 
+remarks:
+- delete papers from github/hide them
+-
+
+-->

### 2021-06-15 thesis/_book/_main.tex
Adds up-to-date changes

@@
 We first encountered the GoogleNews word embedding model in \protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}).
-They use it for their calculations as mentioned in \href{https://github.com/tolga-b/debiaswe}{Tolga Code}.
+They use it for their calculations as written in \href{https://github.com/tolga-b/debiaswe}{Tolga Code}.
 We decided to compare the results obtained with the use of Reddit L2 model and the ones that we got by GoogleNews model.
-The details of GoogleNews model can be found here \href{https://code.google.com/archive/p/word2vec/}{Google News model}.
+The details of the model can be found here \href{https://code.google.com/archive/p/word2vec/}{Google News model}.
 One of the main differences between the models is that Reddit word embeggings have 50 dimensions and GoogleNews word embeddings have 300 dimensions.
@@
 
-We also applied to the word embeddings hard debiasing method found in (\protect\hyperlink{ref-manzini2019black}{\textbf{manzini2019black?}}).
+We also applied to the word embeddings hard debiasing method found in \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}).
+As it was mentioned before, debiasing consists of two components: identifying the bias subspace and then removing this vector subspace from the chosen embeddings.
+The aim is to increase the cosine distance between protected words and the set of attributes.
 After debiasing the Reddit word embeddings we calculated the cosine distance again to see if there are any significant changes in terms of the similarity between protected words and the attributes.
@@
 
-One may notice
+One may notice that in a religion data set most of the distances for \texttt{associated} words slightly increased.
+Some of the distances from \texttt{different} class are still smaller than for the \texttt{associated} class.
+The distances for all classes seem in most cases to cluster together even after debiasing.
+The reason may be that uncertainty is quite high in \texttt{associated} and \texttt{different} class because the number of samples is so small.
+Additionally in some cases distances from \texttt{human} class are smaller than for the \texttt{assosiated} class which also makes it less clear how to interpret the results, if the distances are really able to catch the bias presence.
 
-\begin{itemize}
-\item
-  humans in religion got closer (should they?)
-\item
-  change in Gender is really minor, still zero out of HPDI range, note how different is very similar
-\item
-  small improvement in race, at the price of moving neutral terms closer
-\end{itemize}
+It is worth paying attention to the gender debiased dataset.
+The change is really minor, still zero out of HPDI range.
+The \texttt{different} attributes still have high similarity value.
+It seems that the distances for gender protected words are still clustered together which suggests that this method is not able to catch the complex bias presence.
+One could conclude that co-occurrence of certain terms does not always mean direct connection in terms of terms associations.
+The fact that some of stereotypically associated with female attributes still have high similarity with male protected words suggest that other techniques should be applied to detect and remove the bias.
 
-In (\protect\hyperlink{ref-manzini2019black}{\textbf{manzini2019black?}}) it is mentioned that debiasing with this method that is based on bias component removal may be insufficient.
+At the same time, there is a small improvement in race dataset.
+In some cases the distances just cluster together.
+In general distances for \texttt{assosiated} class are still higher than those for the rest of classes.
+One may assume that there reason for so small change is that either the metric does not indicate properly biases or that the issue is more complex and subtracting subspaces is not enough.
+
+In \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) it is mentioned that debiasing with this method that is based on bias component removal may be insufficient.
 The reason is that words containing similar biases may still cluster together even after debiasing.
@@
 \item
-  debiasing with the method provided by (\protect\hyperlink{ref-manzini2019black}{\textbf{manzini2019black?}}) does not solve the bias issues
+  debiasing with the method provided by \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) does not solve the bias issues
+\end{itemize}
+
+One should remember that there is no baseline for the interpretation of the averaged cosine distances in (\protect\hyperlink{ref-Manzini2019blackToCriminalToCriminal}{\textbf{Manzini2019blackToCriminalToCriminal?}}).
+One may assume that if the cosine distance is close to 1 then it is a desired outcome as it means, according to cosine similarity assumptions, that there is almost no similarity between the words.
+However what does it mean to be close to 0?
+If the averaged cosine similarity is equal to 80, then should we still debiase it?
+It is unclear what the criteria are.
+On one hand, it seems to be beneficial when the outcome is simplified as it is easier to compare results with one value per set.
+On the other hand, it is prone to misunderstanding of how to interpret the results and what threshold to assume.
+
+\hypertarget{what-we-propose-and-how-it-can-help}{%
+\section{What we propose and how it can help:}\label{what-we-propose-and-how-it-can-help}}
+
+\begin{itemize}
+\tightlist
+\item
 \end{itemize}
@@
 
+\leavevmode\hypertarget{ref-Manzini2019blackToCriminal}{}%
+Manzini, T., Lim, Y.
+C., Tsvetkov, Y., \& Black, A.
+W.
+(2019).
+Black is to criminal as caucasian is to police: Detecting and removing multiclass bias in word embeddings.
+Retrieved from \url{http://arxiv.org/abs/1904.04047}
+
 \end{CSLReferences}

### 2021-06-16 paperDraft/paperDraft.Rmd
Adds chapter two revies

@@
 There, the gender bias of a word $w$ is understood as its projection on the gender direction $\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})$ (the gender direction is the top principal compontent of ten gender pair difference vectors).
-The underlying idea is that no bias is present if non-explicitly gendered words are in equal distance to both elements in all explicitly gendered pairs.
-Given the (ideally) gender netural words $N$ and the gender direction $g$ the direct direct gender bias is defined as the average distance of the words in $N$ from $g$ ($c$ is a parameter determining how strict we want to be):
+The underlying idea is that no bias is present if non-explicitly gendered words are in equal distance to both elements in all explicitly gender pairs.
+Given the (ideally) gender neutral words $N$ and the gender direction $g$ the direct direct gender bias is defined as the average distance of the words in $N$ from $g$ ($c$ is a parameter determining how strict we want to be):
 \begin{align}

### 2021-06-16 paperDraft/paperDraft.tex
Adds chapter two revies

--- 2021-06-10 paperDraft/paperDraft.tex
@@
 almost the same word sets, allowing for a \emph{prima facie} sensible
-comparison with bias in humans. If the attitude towards given pair of
-concept is to be interpreted as neutral, the final value from the
+comparison with bias in humans. If the person's attitude towards given
+pair of concept is to be interpreted as neutral, there should be no
+noticeable task completion time difference, and the final value from the
 formula should be around 0. Let \(f\) be a similarity measure (usually,

### 2021-06-16 thesis/02-cosine.Rmd
Adds chapter two revies

@@
 
-@Bolukbasi2016Man
- define similarity between words as the outcome inner product of their normalized vectors. 
+@Bolukbasi2016Man define similarity between words as the outcome inner product of their normalized vectors. 
 They focus on examining what the geometry of word embedding is in regard to "he" and "she" words.
@@
 
+Here the gender bias of a word $w$ is understood as its projection on the gender direction $\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})$ (the gender direction is the top principal compontent of ten gender pair difference vectors).
+The underlying idea is that no bias is present if non-explicitly gender words are in equal distance to both elements in all explicitly gender pairs.
+Given the (ideally) gender neutral words $N$ and the gender direction $g$ the direct direct gender bias is defined as the average distance of the words in $N$ from $g$ ($c$ is a parameter determining how strict we want to be):
+\begin{align}
+\mathsf{directBias_c(N,g)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},g)\vert^c}{\vert N \vert }
+\end{align}
+
 A very vivid way to follow their method of arguing that bias in word embeddings is real is to plot the values of inner product of chosen words.
@@
 Data used to create our plot is as follows.
-\newline
 
 Occupations associated with feminine: \textbf{"homemaker", "nurse", "receptionist", "librarian", "socialite", "hairdresser", "nanny", "bookkeeper", "stylist", "housekeeper", "interior designer", "guidance counselor"}
-\newline
 
 Occupations associated with masculine: \textbf{"maestro", "skipper", "protege", "philosopher", "captain", "architect", "financier", "warrior", "broadcaster", "magician", "fighter pilot", "boss"}
-\newline
 
@@
 
-<!-- The methodology used in the paper  -->
+They modify WEAT to a multi-class setting, introducing Mean Average Cosine similarity as a measure of bias (in fact, in the paper  they report distances rather than similarities).
+Let $T = \{t_1, \dots, t_k\}$ be a class of protected word embeddings, and let each $A_j\in A$ be a set of attributes stereotypically associated with a protected word).
+Then:
+\begin{align}
+S(t_i, A_j) & = \frac{1}{\vert A_j\vert}\sum_{a\in A_j}\mathsf{cos}(t,a) \\
+MAC(T,A) & = \frac{1}{\vert T \vert \,\vert A\vert}\sum_{t_i \in T }\sum_{A_j \in A} S(t_i,A_j)
+\end{align}
 
@@
 
-The example refers to the process of hard debiasing on religious attributes.
-
-1. First we load word embeddings from reddit.US.txt.tok.clean.cleanedforw2v.w2v dataset
-  - The word embeddings have only 50 dimensions
-  - The number of individual words from the dataset is 44895
-
-2.
-The protected group the authors assume should ideally not have high cosine similarity to stereotypical words.
+First we load word embeddings from reddit.US.txt.tok.clean.cleanedforw2v.w2v dataset.
+The word embeddings have only 50 dimensions and the number of individual words from the dataset is 44895.
+The authors assume that the protected group  should ideally not have high cosine similarity to stereotypical words.
 The word embeddings geometry should not place this group close to harmful stereotypes, if it is to be bias-free.
-\newline
-
 For instance, let's look at the religion-related words.
@@
 \textbf{"jew"       : ["judaism", "jew", "synagogue", "torah", "rabbi"]}
-\newline
+
 \textbf{"christian" : ["christianity", "christian", "church", "bible", "priest"]}
-\newline
+
 \textbf{"muslim"    : ["islam", "muslim", "mosque", "quran", "imam"]}
-\newline
 
@@
 \textbf{"jew"       : ["greedy", "cheap", "hairy", "liberal"]}
-\newline
+
 \textbf{"christian"       : ["judgemental", "conservative", "familial"]}
-\newline
+
 \textbf{"muslim"       : ["violent", "terrorist", "dirty", "uneducated"]}
-\newline
 
@@
 
-3. In the article the authors also try to remove previously defined biases from word embedding.
-
+In the article the authors also try to remove previously defined biases from word embedding.
 First they identify the bias subspace using Principal Component Analysis (PCA) which is a technique for dimensionality reduction.
@@
 There can be many subspaces found in a given group, for example in terms of religion one can identify at least a few sets that are to grasp the concept of religion in general:
-\newline
+
 \textbf{["judaism", "christianity", "islam"]}
-\newline
+
 \textbf{["jew", "christian", "muslim"]}
-\newline
+
 \textbf{["synagogue", "church", "mosque"]}
-\newline
 
@@
 This strategy is based on the idea that different dimensions of vectors contain different types of information and in some words in vector layers (subspaces) the information about religiousness is implicitly conveyed.
-In some cases this knowledge is useful but in the case of harmful stereotypes one does not want to have religion concept in stereotypical words.
+In some cases this knowledge is useful but in the case of harmful stereotypes one does not want to include the concept of religion in stereotypical words.
 
-After finding bias subspace, they use it to modify the vector values individually so that their cosine distances towards certain words are changed.
+After finding the bias subspace, they use it to modify the vector values individually so that their cosine distances towards certain words are changed.
 In the case of stereotypes the aim is to make the cosine distances larger so that the association between protected word and harmful stereotype is smaller.
 
-<!-- go through it again -->
-
-4. In the final step there is an evaluation of the results. The cosine distances are calculated again but this time using debiased vocabulary. After
-taking the mean of all distances one final value is obtained and then it is compared with the average value from the beginning.
-If the cosine distances are on average greater than before then it leads the authors to the conclusion that improvement was made.
+In the final step there is an evaluation of the results.
+The cosine distances are calculated again but this time using debiased vocabulary.
+After taking the mean of all distances one final value is obtained and then it is compared with the average value from the beginning.
+If the cosine distances are on average greater than before then it leads the authors to the conclusion that improvement has been achieved.
 As the cosine distance increases it is assumed that the association between protected and stereotypical words decreases.
-
-<!-- add more information on the source of words used -->
 
@@
 The authors use the mean average cosine similarity to check on multi-class similarity between protected word and harmful stereotypes. 
-They average the results until they obtain one final value to represent the mean cosine distance between protected word from a given class and the
-attributes of that class. As there is no control of the individual values of cosine distances, one can observe values that have cosine distances
-greater than 1 which would suggest that these compare words do not have significant similarity. However, as there is a lack of control, even such
-words are taken into the account for calculations.
+They average the results until they obtain one final value to represent the mean cosine distance between protected word from a given class and the attributes of that class.
 
-\textbf{3. The lack of uncertainty}
+\textbf{3. Hiding the impact of uncertainty}
 
-A mean hides this issue and as there are pairs having negative and small similarities and there are those that have similarity equal to 0.5, the resulting calculation seems to be in norm.
+A mean hides issue and as there are pairs having negative and small similarities and there are those that have similarity equal to 0.5, the resulting calculation seems to be in norm.
 Additionally in such method the uncertainty is also not included which makes it even more difficult to give reasonable interpretations of the results.
@@
 
-In the original paper words from all three religions were checked with all of the stereotypes which means that there was no distinction on whether the stereotype is associated with given religion or with the other.
-Not all of the stereotypical words should be considered as harmful for all of the religions.
-In our analysis we distinguished between associated stereotypes, not associated (stereotype is valid but for different religion than currently checked) and none (meaning neutral words).
+In the original paper words from all three religions were checked with all of the stereotypes which means that there was no distinction between classes in which the stereotype is associated with a given religion, as opposed to the situation in which it is associated with another one.
+Not all of the stereotypical words have to be considered as harmful for all of the religions.
+In our analysis we distinguished between stereotypes associated with a given group, stereotype associated with different groups and control groups; neutral words and stereotypes-free human predicates.
 
@@
 
-Assuming for a moment that the value of multi-class cosine distance is correct, one may question the results' interpretation.
-In @Manzini2019blackToCriminal Table 2, there are summarized the averages of cosine distance per group (gender, race, religion).
-I would like to focus  now on analyzing the values relating to religious biases.
-Here is the fragment of table that refers to that:
+Assuming for a moment that the value of multi-class cosine distance is correct, one may question the interpretation.
+@Manzini2019blackToCriminal summarize the averages of cosine distance per group (gender, race, religion).
+We would like to focus now on analyzing the values relating to religious biases.
+Here is the relevant fragment of table:
 
@@
 
-MAC stand for mean average cosine similarity although in reality the values of cosine distance are stored there.
+MAC stand for mean average cosine similarity, although in reality the values of cosine distance are given.
 What may attract attention is the fact that the value of cosine distance in "Biased" category is already quite high even before "debiasing".
 High cosine distance indicates low cosine similarity between values.
-One could think that average cosine similarity equal to approximately 0.141 is not significant enough to consider it as biased.
-However the authors aim to mitigate "biases" in vectors with such great distance to make it even larger.
+One could think that average cosine similarity equal to approximately 0.141 is not significant enough to consider it as bias.
+However the authors aim to mitigate "biases" in vectors with such great distance to make it somewhat larger.
 Methodologically there is a question on what basis this small similarity is still considered as a proof of bias presence.

### 2021-06-16 thesis/_book/_main.tex
Adds chapter two revies

@@
 \author{Alicja Dobrzeniecka}
-\date{2021-06-15}
+\date{2021-06-16}
 
@@
 
-Placeholder
+Natural language processing (NLP) is a subfield of computer science that processes and analyzes language in text and speech with the use of modern programming methods.
+It has practical applications in everyday life as it concerns tasks such as email filters, smart assistants, search results, language translations, text analytics and so on.
+Models used to accomplish these tasks need a lot of data to learn from.
+This data originates from human activities and historical recordings such as texts, messages or speeches.
+It turns out that in the learning process these models can learn implicit biases that reflect harmful stereotypical thinking still present in modern societies.
+One can find methods that aim at identifying and measuring hidden biases and/or try to remove them by modifying the models.
+There are many different types of models in NLP depending on a task that they are supposed to solve.
+However, all of them need as an input words represented by means of numbers and this is accomplished with word embedding models.
+The models usually assign the values based on the context in which the words appear.
+It means that the input data can have enormous influence on the outcome.
+The biases seem to have their primary source in the way the words are assigned the numerical values.
+
+There is considerable amount of literature available on the topic of bias detection and mitigation in NLP models.
+\protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) focus on gender biases that may be observable while investigating the representation of job occupations and gender in terms of their assigned numerical
+values.
+The authors apply cosine similarity measurement to investigate the phenomenon where (the vectors corresponding to) words related to jobs that are stereotypically associated with a given gender are in fact in the model situated closer to this gender.
+They also use analogy tasks to evaluate if the bias is present in the word embedding model.
+They check analogies by comparing pairs of word vectors, for example they search for the word complementing the puzzle: man is to doctor as woman is to \ldots?
+First they subtract the word ``man'' from the word ``woman'' and then
+they search for the ranked list of other words pairs that have similar vector difference. They also include in the formula a threshold to ensure that the resulting pairs could not be randomly picked.
+
+However, as \protect\hyperlink{ref-Nissim2019Fair}{Nissim, Noord, \& Goot} (\protect\hyperlink{ref-Nissim2019Fair}{2019}) points out, there are some limitations to this approach.
+In practice, most of analogies implementations do not return any input words.
+This means that it does not make sense to expect the algorithm to return the same profession for both woman and man.
+Therefore, this method seems to be limited in terms of bias detection.
+Other problems are related for example to the choice of pairs and words that are used to detect the presence of discrimination as it is often subjective and without proper justification.
+Additionally the choice of the parameter used in \protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) formula to ensure that word pairs are not picked by random, is also not justified and changing it drastically
+influences the results.
+
+\protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) touch upon the topic of biases regarding race and gender.
+They apply knowledge from well-known psychological studies such as Implicit Association Test to research the relation between human stereotypical thinking and model learnt biases to
+discover close relationship between these two. For the evaluation they use Word Embedding Association Test (WEAT) and the Word Embedding Factual Association Test (WEFAT).
+
+(\protect\hyperlink{ref-manzini2019black}{\textbf{manzini2019black?}}) propose a novel way of using cosine similarity to obtain information on assumed resemblance between words. They investigate
+an approach that enables them to measure the bias for a class (like gender, religion, race) and express the final result with a single metric.
+
+\protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) refer to Implicit Association Test (Greenwald et al., 1998) that measures the strength of associations between concepts or stereotypes by calculating the time of reaction for special tasks.
+Humans naturally exhibit some biases which do not always cause social concern.
+One can imagine the intuitive associations between for example insects and flowers, and the feelings of pleasantness or unpleasantness.
+In general, people would rather associate flowers with feeling pleasant than insects, and this preference in a sense is a bias or
+prejudice in some direction.
+However, this type of preference does not cause an uproar and is a rather morally neutral case.
+This example is rather used to show that the methodology used in IAT makes sense.
+Unfortunately, the further studies discover other biases and prejudices that directly influence the quality of other people's lives and therefore they should be taken care of.
+
+One can find various definitions trying to capture what bias and fairness actually are.
+With the choice of the definition, implications into the real-life applications may change as well.
+\protect\hyperlink{ref-Mehrabi2019Survey}{Mehrabi, Morstatter, Saxena, Lerman, \& Galstyan} (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) mark out that there exist different types of biases such as historical bias, representation bias or measurement bias (the list is long).
+This indicates how complex the issue of bias is.
+Without the proper understanding and awareness of the problem, people are prone to unconsciously sustain the bias.
+
+\protect\hyperlink{ref-Mehrabi2019Survey}{Mehrabi, Morstatter, Saxena, Lerman, \& Galstyan} (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) also distinguish different types of discrimination, some of them will be briefly described.
+By protected attributes we mean those qualities, traits or characteristics that one should not discriminate against.
+Direct discrimination occurs when protected attributes of individuals explicitly result in non-favorable outcomes toward them.
+Such discrimination takes place for example when one uses an information about someone's religion belief to evaluate whether they are competent for the work they apply for.
+In contrast in indirect discrimination individuals appear to be treated equally but anyway they end up being treated unjustly due to the hidden effects of biases towards their protected attributes.
+An example of such situation is when in a job advert there is a requirement of 10 years experience instead of the list of a specific type of experience and knowledge.
+This job advert can discriminate indirectly against young people who can have equivalent skills but not yet so much work experience.
+Systemic discrimination takes place when policies, customs or behaviors that result from certain culture or organizational structure lead to discrimination against some groups of people.
+Finally, statistical discrimination consists in to using
+average group statistics to judge person belonging to the group.
+
+The topic of discrimination is entangled with another concept; fairness. It is essential to grasp some concepts of fairness to take them into
+consideration while designing implementation of some machine learning model.
+In \protect\hyperlink{ref-Mehrabi2019Survey}{Mehrabi, Morstatter, Saxena, Lerman, \& Galstyan} (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) one may notice that depending on the context and application different definitions may be applied.
+
+The most common methods in the literature focus on comparing the similarity between words from protected groups and those that are considered to be stereotypical or harmful in some way.
+One can find in this group methods such as euclidean distance or cosine similarity (which is equivalent to dot product if the
+vectors are normalized).
+There are also other ways to detect the effects of biases.
+One can investigate the model performance on certain tasks and validate if the model uses the gender or race information to complete the task.
+
+The currently used methods (such as cosine similarity) make the similarity values often aggregated in a way that may lead to hasty conclusions.
+The averaging of values and the lack of uncertainty may lead to the incomplete picture of the situation.
+
+In the paper we indicate how current methods used to detect biases in natural language models are limited, if we investigate them from the perspective of Bayesian analysis.
+
+Our research enhances the current way in which the bias detection is performed to make sure that it is
+methodologically valid.
+
+The key hypothesis is that greater understanding of data and bias implications can be achieved when Bayesian methods are applied to issue.
 
@@
 
-Placeholder
-
 \hypertarget{word-embeddings}{%
@@
 
+To understand what cosine similarity measurement is, one first needs to grasp the concept of translating words to a computer-readable form.
+In the field of natural language processsing there are two main types of words representation --- localist and distributed.
+One-hot encoding is an example of a method used to achieve a localist representation of words.
+Here each vector contains information only about a single data point, this is achieved by first mapping categorical values (words) to integers and then to each integers a binary vector is assigned which contains only 0s except for the index of the integer, which is assigned 1.
+An example of a localist representation is:
+
+\begin{longtable}[]{@{}llllll@{}}
+\toprule
+word & 1 & 2 & 3 & 4 & 5 \\
+\midrule
+\endhead
+woman & 1 & 0 & 0 & 0 & 0 \\
+man & 0 & 1 & 0 & 0 & 0 \\
+girl & 0 & 0 & 1 & 0 & 0 \\
+boy & 0 & 0 & 0 & 1 & 0 \\
+monarch & 0 & 0 & 0 & 0 & 1 \\
+\bottomrule
+\end{longtable}
+
+In the example above it is clear that the length of the vectors increases with the number of words in a vocabulary.
+It is not a very computationally efficient representation.
+It has other flaws as well.
+For example, it is unable to capture the resemblance between words appearing in similar contexts.
+
+In contrast to the localist representation, a distributed representation returns vectors that contain continuous values instead of discrete 1s and 0s.
+Word embeddings are a class of various techniques that allow one to represent words as distributed vectors.
+Such learned representations of text have certain properties.
+At least prima facie, they store similar (or at least co-occurring) words close to each other in a vector space.
+An example of distributed representation is:
+
+\begin{longtable}[]{@{}lllll@{}}
+\toprule
+word & 1 & 2 & 3 & 4 \\
+\midrule
+\endhead
+woman & 0.456 & 0.267 & 0.675 & 0.131 \\
+man & 0.451 & 0.897 & 0.472 & 0.088 \\
+girl & 0.604 & 0.262 & 0.414 & 0.706 \\
+boy & 0.279 & 0.172 & 0.475 & 0.010 \\
+monarch & 0.565 & 0.678 & 0.463 & 0.975 \\
+\bottomrule
+\end{longtable}
+
+One of the advantages of using a distributed representation is that one is able to represent an enormous number of concepts with a smaller number of units.
+It is also possible to better capture similarities as words of similar meanings can have similar numeric vectors.
+
+The numbers occurring in such representations are not random.
+They are learned in a process that uses a very shallow neural network.
+There are various types of techniques used for learning the vectors representations.
+One of the most straightforward ones is a skip-gram model.
+Given a word the model tries to predict its neighboring words from the sentence.
+The mathematics behind the process relies on the idea that the prediction concerns the conditional probability of the adjacent words.
+The algorithm tries to minimize the loss function, which penalizes the system for discrepancy with actual co-occurrence frequencies in the corpus.
+One can choose various parameters of the model, such as the window size that determines how many surrounding words the model should predict.
+After preparing such a fitted modeln=, one takes only the learned weights from a neural network, and uses them as vectors in a word embeddings representation.
+
+Word embeddings have many applications in natural language processing.
+They are handy in document search and information retrieval.
+They also play their part in improving automatic translations.
+Well learned word representations may also contribute to the improvement of sentiment analysis or spam detection.
+
 \hypertarget{cosine-similarity-and-distance}{%
@@
 
+Cosine similarity is often used as a method of finding out whether vector representations for two words suggest that they are similar or somehow connected.
+Cosine similarity is the cosine of the angle between two vectors: the result of dividing their inner product (dot product usually) by the product of their magnitudes.
+
+It is worth mentioning one more point concerning cosine similarity.
+After the vectors are normalized to have length equal to 1, inner product itself (often dot product) is used to measure the similarity.
+
+\begin{align} \tag{Sim}
+\mathsf{cosineSimilarity}(A,B) & = \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert}
+\end{align}
+Cosine similarity is considered a proper tool for this operation as its result has a clear connection to geometry and at least for a low number of dimensions may be easily interpreted.
+Using this scale, one can compare vector similarities in a fairly clear manner.
+When the vectors are aligned perpendicularly to each other, their similarity equals 0 (which is the same as the cosine of 90 degrees).
+This tells us that the similarity between the vectors is small.
+As the angle between vectors decreases, cosine similarity approaches one, which stands for the greatest similarity.
+
+One of the limitations of this measure is that it informs us only about similarities between vectors in terms of their orientation.
+However, it is often argued that in comparing words in terms of this metric, the magnitude of vectors may be treated as irrelevant, as the most important information pertains to direction.
+
+In what follows, it is important to distinguish between cosine similarity and cosine distance, defined as:
+\begin{align} \tag{Sim}
+\mathsf{cosineDistance}(A,B) &  = 1 - \mathsf{cosineSimilarity}(A,B)\\
+ &  = 1 - \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert} \nonumber
+\end{align}
+
+The greater the similarity between two vectors, the smaller the distance between them.
+The cosine distance ranges between 0 and 2.
+If the vectors are in an opposite direction to each other, the cosine distance is 2.
+And if the vectors are extremely similar then the cosine distance is very close to 0.
+
 \hypertarget{cosine-distance-in-a-one-class-bias-detection}{%
@@
 
+\protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) define similarity between words as the outcome inner product of their normalized vectors.
+They focus on examining what the geometry of word embedding is in regard to ``he'' and ``she'' words.
+In other words, whether the similarity between those concepts and other words reflects expected gender stereotypes.
+They test this hypothesis by investigating whether there is a connection between word embeddings representing certain professions and words referring to gender.
+They also evaluate whether automatically produced analogies between words reflect the stereotypes as well.
+
+Here the gender bias of a word \(w\) is understood as its projection on the gender direction \(\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})\) (the gender direction is the top principal compontent of ten gender pair difference vectors).
+The underlying idea is that no bias is present if non-explicitly gender words are in equal distance to both elements in all explicitly gender pairs.
+Given the (ideally) gender neutral words \(N\) and the gender direction \(g\) the direct direct gender bias is defined as the average distance of the words in \(N\) from \(g\) (\(c\) is a parameter determining how strict we want to be):
+\begin{align}
+\mathsf{directBias_c(N,g)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},g)\vert^c}{\vert N \vert }
+\end{align}
+
+A very vivid way to follow their method of arguing that bias in word embeddings is real is to plot the values of inner product of chosen words.
+The plot below does not originate from the original paper (it is from \url{https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings}) but similar visualization may be found there.
+Data used to create our plot is as follows.
+
+Occupations associated with feminine: \textbf{"homemaker", "nurse", "receptionist", "librarian", "socialite", "hairdresser", "nanny", "bookkeeper", "stylist", "housekeeper", "interior designer", "guidance counselor"}
+
+Occupations associated with masculine: \textbf{"maestro", "skipper", "protege", "philosopher", "captain", "architect", "financier", "warrior", "broadcaster", "magician", "fighter pilot", "boss"}
+
+\vspace{1mm}
+\footnotesize
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-1-1} \end{center}
+\normalsize
+
+The points in the plot above result from the calculation of the inner product of a chosen vector for a profession word and a vector for a gender word (she or he).
+Inner product of two vectors expresses similarity between words.
+This assumption originates from the geometry and properties of a vector space.
+
 \hypertarget{cosine-distance-in-a-multi-class-bias-detection}{%
@@
 
+\protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) present a different approach towards finding similarities between classes of words.
+The authors claim that texts available online are full of direct or indirect human stereotypes.
+As a result, word embeddings are prone to learn and maybe amplify those biases and propagate them further into AI models that are used for various applications.
+Cosine distance is used in the article as a measure to first argue for the existence of multi-class bias and then to show how through bias mitigation techniques the bias may be decreased.
+
+They modify WEAT to a multi-class setting, introducing Mean Average Cosine similarity as a measure of bias (in fact, in the paper they report distances rather than similarities).
+Let \(T = \{t_1, \dots, t_k\}\) be a class of protected word embeddings, and let each \(A_j\in A\) be a set of attributes stereotypically associated with a protected word).
+Then:
+\begin{align}
+S(t_i, A_j) & = \frac{1}{\vert A_j\vert}\sum_{a\in A_j}\mathsf{cos}(t,a) \\
+MAC(T,A) & = \frac{1}{\vert T \vert \,\vert A\vert}\sum_{t_i \in T }\sum_{A_j \in A} S(t_i,A_j)
+\end{align}
+
+As the code has been provided by the authors we were able to reconstruct their results.
+The main steps in the procedure are as follows.
+Let us go through an example that refers to the process of hard debiasing on religious attributes.
+
+First we load word embeddings from reddit.US.txt.tok.clean.cleanedforw2v.w2v dataset.
+The word embeddings have only 50 dimensions and the number of individual words from the dataset is 44895.
+The authors assume that the protected group should ideally not have high cosine similarity to stereotypical words.
+The word embeddings geometry should not place this group close to harmful stereotypes, if it is to be bias-free.
+For instance, let's look at the religion-related words.
+
+Protected words by religion type:
+
+\textbf{"jew"       : ["judaism", "jew", "synagogue", "torah", "rabbi"]}
+
+\textbf{"christian" : ["christianity", "christian", "church", "bible", "priest"]}
+
+\textbf{"muslim"    : ["islam", "muslim", "mosque", "quran", "imam"]}
+
+Stereotypical words by religion type:
+
+\textbf{"jew"       : ["greedy", "cheap", "hairy", "liberal"]}
+
+\textbf{"christian"       : ["judgemental", "conservative", "familial"]}
+
+\textbf{"muslim"       : ["violent", "terrorist", "dirty", "uneducated"]}
+
+We have prepared a table presenting the values of cosine distance for each protected word with each attribute (stereotype). The part of the results is shown below.
+
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{religion }\OtherTok{\textless{}{-}} \FunctionTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{]}
+\FunctionTok{colnames}\NormalTok{(religion) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
+                        \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
+\NormalTok{religion}\SpecialCharTok{$}\NormalTok{wordClass }\OtherTok{\textless{}{-}} \FunctionTok{as.factor}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{wordClass)}
+\FunctionTok{levels}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{wordClass) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
+\FunctionTok{head}\NormalTok{(religion)  }\SpecialCharTok{\%\textgreater{}\%}  \FunctionTok{kable}\NormalTok{(}\AttributeTok{format =} \StringTok{"latex"}\NormalTok{,}\AttributeTok{booktabs=}\NormalTok{T,}
+                      \AttributeTok{linesep =} \StringTok{""}\NormalTok{,  }\AttributeTok{escape =} \ConstantTok{FALSE}\NormalTok{, }
+                      \AttributeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+                      \FunctionTok{kable\_styling}\NormalTok{(}\AttributeTok{latex\_options=}\FunctionTok{c}\NormalTok{(}\StringTok{"scale\_down"}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{table}
+
+\caption{\label{tab:religionTableHeadEarly}Head of the religion dataset.}
+\centering
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{lllrrl}
+\toprule
+protectedWord & wordToCompare & wordClass & cosineDistance & cosineSimilarity & connection\\
+\midrule
+judaism & violent & muslim & 0.7141939 & 0.2858061 & different\\
+judaism & terrorist & muslim & 0.7461333 & 0.2538667 & different\\
+judaism & dirty & muslim & 1.2002599 & -0.2002599 & different\\
+judaism & uneducated & muslim & 0.7885469 & 0.2114531 & different\\
+judaism & greedy & jewish & 1.0026172 & -0.0026172 & associated\\
+judaism & cheap & jewish & 1.2323229 & -0.2323229 & associated\\
+\bottomrule
+\end{tabular}}
+\end{table}
+\normalsize
+
+\pagebreak
+
+In the article there was no analysis of individual distances but the general look at the data through the usage of mean.
+The authors introduced a metric that tries to generalize the presence of bias through the classification of multi-class bias in groups of words connected with gender, religion or race.
+In the process they first take the mean of cosine distances between a given protected word and attributes assigned to each stereotype.
+They do not differentiate between stereotypes associated with a word and stereotypes associated with different words (in case of religion, stereotypes characteristic for Christianity has also cosine distance measured with for instance, Judaism or Islam).
+Then, after collecting the list of mean cosine distances, they average the list to obtain one final value representing the whole group, in this example religion, for which the final mean of all mean distances is equal to 0.859.
+
+In the article the authors also try to remove previously defined biases from word embedding.
+First they identify the bias subspace using Principal Component Analysis (PCA) which is a technique for dimensionality reduction.
+It is applied here to choose the subspace that contains the greatest amount of information.
+There can be many subspaces found in a given group, for example in terms of religion one can identify at least a few sets that are to grasp the concept of religion in general:
+
+\textbf{["judaism", "christianity", "islam"]}
+
+\textbf{["jew", "christian", "muslim"]}
+
+\textbf{["synagogue", "church", "mosque"]}
+
+The idea is to find a set that provides enough information to create from it a vector representing the concept of religion among words.
+This strategy is based on the idea that different dimensions of vectors contain different types of information and in some words in vector layers (subspaces) the information about religiousness is implicitly conveyed.
+In some cases this knowledge is useful but in the case of harmful stereotypes one does not want to have religion concept in stereotypical words.
+
+After finding bias subspace, they use it to modify the vector values individually so that their cosine distances towards certain words are changed.
+In the case of stereotypes the aim is to make the cosine distances larger so that the association between protected word and harmful stereotype is smaller.
+
+In the final step there is an evaluation of the results.
+The cosine distances are calculated again but this time using debiased vocabulary.
+After taking the mean of all distances one final value is obtained and then it is compared with the average value from the beginning.
+If the cosine distances are on average greater than before then it leads the authors to the conclusion that improvement was made.
+As the cosine distance increases it is assumed that the association between protected and stereotypical words decreases.
+
 \hypertarget{limitations-of-the-approach}{%
 \section{Limitations of the approach}\label{limitations-of-the-approach}}
+
+\textbf{1. Selection of attributes}
+
+The attributes are taken from different sources, there is no principled justification for their choice. From our analysis it will become clear that
+the list is rather uneven.
+
+There is no mention of methodology for deciding on the number of attributes necessary to decide a hypothesis on the given size of dataset.
+There are however some ways to estimate how many samples we need to make sure that the result is significant.
+Our research will show that the numbers used
+are rather insufficient.
+
+\textbf{2. No control while taking the mean of cosines}
+
+The authors use the mean average cosine similarity to check on multi-class similarity between protected word and harmful stereotypes.
+They average the results until they obtain one final value to represent the mean cosine distance between protected word from a given class and the
+attributes of that class. As there is no control of the individual values of cosine distances, one can observe values that have cosine distances
+greater than 1 which would suggest that these compare words do not have significant similarity. However, as there is a lack of control, even such
+words are taken into the account for calculations.
+
+\textbf{3. The lack of uncertainty}
+
+A mean hides this issue and as there are pairs having negative and small similarities and there are those that have similarity equal to 0.5, the resulting calculation seems to be in norm.
+Additionally in such method the uncertainty is also not included which makes it even more difficult to give reasonable interpretations of the results.
+We propose the use of Bayesian method to obtain some understanding of the influence the uncertainty has on the interpretation of final results.
+
+\textbf{4. No word class distinction}
+
+In the original paper words from all three religions were checked with all of the stereotypes which means that there was no distinction on whether the stereotype is associated with given religion or with the other.
+Not all of the stereotypical words should be considered as harmful for all of the religions.
+In our analysis we distinguished between associated stereotypes, not associated (stereotype is valid but for different religion than currently checked) and none (meaning neutral words).
+
+\textbf{5. Interpreting the results}
+
+Assuming for a moment that the value of multi-class cosine distance is correct, one may question the results' interpretation.
+In \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) Table 2, there are summarized the averages of cosine distance per group (gender, race, religion).
+I would like to focus now on analyzing the values relating to religious biases.
+Here is the fragment of table that refers to that:
+
+\begin{longtable}[]{@{}ll@{}}
+\toprule
+Religion Debiasing & MAC \\
+\midrule
+\endhead
+Biased & 0.859 \\
+Hard Debiased & 0.934 \\
+Soft Debiased (\(\lambda\) = 0.2) & 0.894 \\
+\bottomrule
+\end{longtable}
+
+MAC stand for mean average cosine similarity although in reality the values of cosine distance are stored there.
+What may attract attention is the fact that the value of cosine distance in ``Biased'' category is already quite high even before ``debiasing.'' High cosine distance indicates low cosine similarity between values.
+One could think that average cosine similarity equal to approximately 0.141 is not significant enough to consider it as biased.
+However the authors aim to mitigate ``biases'' in vectors with such great distance to make it even larger.
+Methodologically there is a question on what basis this small similarity is still considered as a proof of bias presence.
+
+This is in general the problem of scale and the lack of universal intervals that could be applied to know whether the cosine similarity is high enough to considered given words more similar that if we chose them on random.
+It seems like similar practice could be beneficial for metrics aimed at measuring similarity between words.
+
+\textbf{6. The curse of dimensionality}
+
+Curse of dimensionality may take place when there is an increase in volume of data that results in adding extra dimensions to the Euclidean space. According to the article
+``\url{https://analyticsindiamag.com/curse-of-dimensionality-and-what-beginners-should-do-to-overcome-it/}''
+as the number of features increases, it may be harder and harder to obtain useful information from the data with the usage of available algorithms.
+One may notice that more data should contribute to greater amount of information but more information also means greater risk of noise and distractions in data.
+At the same time, many times modern solutions are adapted to smaller dimensions and their results in higher ones are not intuitive or may be prone to be mistaken.
+
+Using cosine similarity in high dimensions in word embeddings may also be prone to the curse of dimensionality.
+According to this article ``\url{https://www.researchgate.net/publication/327498046_The_Curse_of_Dimensionality_Inside_Out}''
+there are reasons to consider this phenomenon when searching for word similarities in higher dimensions.
+
+In the article an experiment is conducted that aims at showing how the similarity values and variation change as the number of dimensions increases.
+The hypothesis made in the paper states that two things will happen as the number of dimensions increase, the first one is that effort required to measure cosine similarity will be greater and the second one is that the similarity between data will blur out and have less variation.
+In details, the authors generate random points with increasing number of dimensions where each dimension of a data point is given a value between 0 and 1.
+Then they pick one vector at random from each dimension class and calculate cosine similarity between the chosen vector and the rest of the data.
+Then they check how the variation of values changes as the number of dimensions increases.
+It seems like the more dimensions there are, the smaller the variance and therefore it is less obvious how to interpret the resulting cosine similarities.
+Maybe the scale should be adjusted to the number of dimensions and variance so that it still gives us sensible information about data.
+According to some researches the cosine similarity in high dimensions is not reliable enough to trust it as it may be the case that choosing words on random may result in getting similar values as when picking them consciously.
+
+\begin{figure}
+\centering
+\includegraphics{../images/curseOfDimensionality.png}
+\caption{curse of dimensionality, number of dimensions on the x axis, standard deviation of similarity on the y axis}
+\end{figure}
 
@@
 
-\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-2-1} \end{center}
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-3-1} \end{center}
 \normalsize
@@
 
-\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-3-1} \end{center}
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-4-1} \end{center}
 \normalsize
@@
 
-\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-6-1} \end{center}
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-7-1} \end{center}
 \normalsize
@@
 \FunctionTok{library}\NormalTok{(tidyverse)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## -- Attaching packages --------------------------------------- tidyverse 1.3.1 --
-\end{verbatim}
-
-\begin{verbatim}
-## v ggplot2 3.3.3     v purrr   0.3.4
-## v tibble  3.1.2     v dplyr   1.0.6
-## v tidyr   1.1.3     v stringr 1.4.0
-## v readr   1.4.0     v forcats 0.5.1
-\end{verbatim}
-
-\begin{verbatim}
-## -- Conflicts ------------------------------------------ tidyverse_conflicts() --
-## x dplyr::filter() masks stats::filter()
-## x dplyr::lag()    masks stats::lag()
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
 \NormalTok{muslimClass }\OtherTok{\textless{}{-}}\NormalTok{ muslim }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(protectedWord }\SpecialCharTok{==} \StringTok{"muslim"}\NormalTok{)}
@@
 
-\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-7-1} \end{center}
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-8-1} \end{center}
 \normalsize
@@
 
-\hypertarget{what-we-propose-and-how-it-can-help}{%
-\section{What we propose and how it can help:}\label{what-we-propose-and-how-it-can-help}}
-
-\begin{itemize}
-\tightlist
-\item
-\end{itemize}
+We propose the use of Bayesian method to introduce the uncertainty measure in bias detection.
 
@@
 
-\hypertarget{future-work}{%
-\section{Future work:}\label{future-work}}
+The aim of this paper was to show the analysis of bias detection methods used in modern natural language processing projects.
+In the further research it is worth to dive deeper into some issues or to examine new ideas that occurred during our work.
 
@@
 
+\leavevmode\hypertarget{ref-Caliskan2017Semantics}{}%
+Islam, A.
+C., Bryson, J.
+J., \& Narayanan, A.
+(2016).
+Semantics derived automatically from language corpora necessarily contain human biases.
+\emph{CoRR}, \emph{abs/1608.07187}.
+Retrieved from \url{http://arxiv.org/abs/1608.07187}
+
 \leavevmode\hypertarget{ref-Manzini2019blackToCriminal}{}%
@@
 
+\leavevmode\hypertarget{ref-Mehrabi2019Survey}{}%
+Mehrabi, N., Morstatter, F., Saxena, N., Lerman, K., \& Galstyan, A.
+(2019).
+A survey on bias and fairness in machine learning.
+\emph{CoRR}, \emph{abs/1908.09635}.
+Retrieved from \url{http://arxiv.org/abs/1908.09635}
+
+\leavevmode\hypertarget{ref-Nissim2019Fair}{}%
+Nissim, M., Noord, R.
+van, \& Goot, R.
+van der.
+(2019).
+Fair is better than sensational: Man is to doctor as woman is to doctor.
+\emph{CoRR}, \emph{abs/1905.09866}.
+Retrieved from \url{http://arxiv.org/abs/1905.09866}
+
 \end{CSLReferences}

### 2021-06-16 thesis/02-cosine.Rmd
Adds fixed chapter 2 and discussion

@@
 And if the vectors are extremely similar then the cosine distance is very close to 0.
+
+One should note that cosine distance is not exactly a distance measure while it does not meet triangle inequality requirements.
+The triangle inequality formula says that for any triangle, the sum of the lengths of any two sides must be greater than or equal to the length of the remaining side.
+As shown in [Statexchange.com](https://stats.stackexchange.com/questions/198080/proving-that-cosine-distance-function-defined-by-cosine-similarity-between-two-u) in the case of cosine distance it would have to fulfill this equation 1+cos-sim(𝐴,𝐶) < cos-sim(𝐴,𝐵) + cos-sim(𝐵,𝐶).
+If one chooses specific unit vectors it is easy to demonstrate that the triangle inequality is not preserved.
 
@@
 In the process they first take the mean of cosine distances between a given protected word and attributes assigned to each stereotype.
-They do not differentiate between stereotypes associated with a word and stereotypes associated with different words (in case of religion, stereotypes characteristic for Christianity has also cosine distance measured with for instance, Judaism or Islam).
+They do not differentiate between stereotypes associated with a word and stereotypes associated with different words (in the case of religion, stereotypes characteristic for Christianity has also cosine distance measured with for instance, Judaism or Islam).
 Then, after collecting the list of mean cosine distances, they average the list to obtain one final value representing the whole group, in this example religion, for which the final mean of all mean distances is equal to 0.859.
@@
 MAC stand for mean average cosine similarity, although in reality the values of cosine distance are given.
-What may attract attention is the fact that the value of cosine distance in "Biased" category is already quite high even before "debiasing".
+What may attract attention is the fact that the value of cosine distance in "Biased" category is already quite high even before debiasing.
 High cosine distance indicates low cosine similarity between values.
@@
 However the authors aim to mitigate "biases" in vectors with such great distance to make it somewhat larger.
-Methodologically there is a question on what basis this small similarity is still considered as a proof of bias presence.
-
-This is in general the problem of scale and the lack of universal intervals that could be applied to know whether the cosine similarity is high enough to considered given words more similar that if we chose them on random.
-It seems like similar practice could be beneficial for metrics aimed at measuring similarity between words.
-
-<!-- TODO: numerical test if this distance is significant? check mean mean cosine distances of some random words? -->
+Methodologically the question is, on what basis is this small similarity still considered as a proof of a bias presence, and whether these small changes are meaningful.
+This is in general the problem of scale and the lack of universal intervals.
+Proper intervals could be used to decide whether given cosine similarity is high enough to consider the words to be more similar than if we choose them at random.
 
@@
 
-<!-- TODO: -->
-<!-- 1. https://www.cs.princeton.edu/courses/archive/fall13/cos521/lecnotes/lec11.pdf -->
-<!-- 2. https://stats.stackexchange.com/questions/341535/curse-of-dimensionality-does-cosine-similarity-work-better-and-if-so-why -->
-<!-- 3. https://stats.stackexchange.com/questions/21547/distance-metric-and-curse-of-dimensions -->
-<!-- + check if it may be proved in the code -->
-
-Curse of dimensionality may take place when there is an increase in volume of data that results in adding extra dimensions to the Euclidean space. According to the article 
-"https://analyticsindiamag.com/curse-of-dimensionality-and-what-beginners-should-do-to-overcome-it/"
-as the number of features increases, it may be harder and harder to obtain useful information from the data with the usage of available algorithms.
+In our case, the curse of dimensionality may take place when there is an increase in volume of data that results in adding extra dimensions to the Euclidean space. According to the article 
+[Analyticsindiamag Curse of dimensionality](https://analyticsindiamag.com/curse-of-dimensionality-and-what-beginners-should-do-to-overcome-it/) as the number of features increases, it may be harder and harder to obtain useful information from the data with the usage of available algorithms.
 One may notice that more data should contribute to greater amount of information but more information also means greater risk of noise and distractions in data.
@@
 Using cosine similarity in high dimensions in word embeddings may also be prone to the curse of dimensionality.
-According to this article "https://www.researchgate.net/publication/327498046_The_Curse_of_Dimensionality_Inside_Out"
-there are reasons to consider this phenomenon when searching for word similarities in higher dimensions.
-
-<!-- Add more information on the experiment -->
-
-In the article an experiment is conducted that aims at showing how the similarity values and variation change as the number of dimensions increases.
-The hypothesis made in the paper states that two things will happen as the number of dimensions increase, the first one is that effort required to measure cosine similarity will be greater and the second one is that the similarity between data will blur out and have less variation.
-In details, the authors generate random points with increasing number of dimensions where each dimension of a data point is given a value between 0 and 1.
-Then they pick one vector at random from each dimension class and calculate cosine similarity between the chosen vector and the rest of the data.
+According to @Venkat2018Curse there are reasons to consider this phenomenon when searching for word similarities in higher dimensions.
+An experiment is conducted that aims at showing how the similarity values and variation change as the number of dimensions increases.
+The hypothesis made in the paper states that two things will happen as the number of dimensions increase.
+First, the effort required to measure cosine similarity will be greater, and two, the similarity between data will blur out and have less variation.
+The authors generate random points with increasing number of dimensions where each dimension of a data point is given a value between 0 and 1.
+Then they pick one vector at random from each dimension class and calculate the cosine similarity between the chosen vector and the rest of the data.
 Then they check how the variation of values changes as the number of dimensions increases.
@@
 ![curse of dimensionality,  number of dimensions on the x axis,  standard deviation of similarity on the y axis](../images/curseOfDimensionality.png)
-
-<!-- textbf{Control group}
-
-How to properly prepare control group in terms of quality and quantity?
-<!-- + check if it may be proved in the code -->
-
-<!--  \textbf{Verifying other similarity meaures}
-
-Besides cosine similarity, there are other methods used to find the similarity between vectors.
-
-a) TS-SS 
-todo: check if it makes sense to calculate for all of the datasets; interpretations etc.
-https://www.researchgate.net/publication/303513110_A_Hybrid_Geometric_Approach_for_Measuring_Similarity_Level_Among_Documents_and_Document_Clustering
-
-b) WEAT test 
-<!-- https://docs.responsibly.ai/word-embedding-bias.html#module-responsibly.we.weat -->
-
-<!-- [to check https://github.com/taki0112/Vector_Similarity] -->
-
-<!-- TODO: check in code if they make more sense -->
-
-<!-- ## Bibliography
-- https://www.amazon.com/Language-Processing-Synthesis-Lectures-Technologies/dp/1627052984/ref=as_li_ss_tl?ie=UTF8&qid=1502062931&sr=8-1&keywords=Neural+Network+Methods+in+Natural+Language+Processing&linkCode=sl1&tag=inspiredalgor-20&linkId=d63df073fea3ebe2d405820570b3ff03 ->
-
-<!-- ## *** Additional -->
-
-<!-- ### Man is to Computer Programmer as Woman is to Homemaker? Debiasing Word Embeddings" -->
-<!-- [article https://arxiv.org/pdf/1607.06520.pdf] -->
-
-<!-- Assumptions: -->
-<!-- "We assume there is a set of gender neutral words $N\subseteq W$ such as flight attendant or shoes, which, by definition, are not specific to any gender." -->
-
-<!-- TODO: check if it is useful for the analysis of black is .. -->
-
-<!-- ### Identifying the bias subspace  -->
-
-<!-- Interesting analysis about limits o subspace and research [source https://kawine.github.io/blog/nlp/2019/09/23/bias.html] -->
-
-<!-- Resources to analyze: -->
-<!-- 1. https://arxiv.org/pdf/2005.00965.pdf -->
-<!-- 2. https://arxiv.org/pdf/2009.09435.pdf -->
-<!-- 3. https://arxiv.org/pdf/1904.04047.pdf -->
-<!-- 4. https://www.aclweb.org/anthology/P19-1166.pdf -->
-
-<!-- ### How is cosine similarity entangled with analogy topic in NLP (is it??) -->
-
-<!-- article:  https://arxiv.org/pdf/1905.09866.pdf -->
-
-<!-- Critique of the usage of analogies as a proof for gender biases -->
-
-<!-- TODO: check if there is other argument than that algorithm does not allow it -->
-
-<!-- general thoughts 
-https://www.researchgate.net/publication/349408535_Robustness_and_Reliability_of_Gender_Bias_Assessment_in_Word_Embeddings_The_Role_of_Base_Pairs
-
--
--
--
--->

### 2021-06-16 thesis/04-discussion.Rmd
Adds fixed chapter 2 and discussion

@@
+---
+output:
+  pdf_document: default
+  html_document: default
+---
 # Discussion
@@
 
-- the distributions of cosine distances of whole population and individual words can vary enormously
-- including uncertainty shows how in some cases it is not clear why a word is according to this method
-classified as anomalous
-- debiasing with the method provided by @Manzini2019blackToCriminal does not solve the bias issues
+<!-- Looking into specific values -->
+As we pointed out, if there is no control of what cosine distance values individual word pairs have, then there is less
+understanding of the exact relations between words.
+If the protected words that we chose do not have high similarity with harmful stereotypes, then one should consider at least two scenarios.
+The first one is that the choice of protected words and attributes may be corrupted.
+The second one is that the metric is not able to catch the hidden bias properly.
+In both cases it is essential to take a look at the individual values before averaging them or aggregating in other ways.
 
+<!-- Problem with gender words --> 
+One has to remember also about the specific situation present in gender dataset.
+There surprisingly one could observe high cosine distances values between some female stereotypical professions and male protected words.
+This leads to new ideas of how to understand the bias origins and its detection.
+To start with, cosine distance seems to catch the information from the vectors of words co-occurrence and not the semantic similarity.
+If word stereotypically associated with females has high cosine distance value for male protected words, then the metric is not proper to establish the bias presence as it can be misleading.
+What we mean is that in same cases it can unnecessarily label words as biased while omitting other more harmful associations.
+What is more, this highlights again the need for control group.
+One could check each stereotypical attribute with each protected word but not in order to average the values.
+One should rather plot the distributions and investigate how the metric measures vectors geometry and hidden semantic information.
+
+<!-- debiasing and uncertainty -->
+Debiasing with the method provided by @Manzini2019blackToCriminal shows how it is unpredictable in terms of the results. 
+In some cases the final cosine distance is indeed close to neutral words, but in other cases the change is minor.
+What is more, including uncertainty shows how in some cases it is not clear why a word is according to this method
+classified as anomalous.
+The fact that uncertainty for `assosiated` and `different` class is so high leads to the situation where in the boundaries of an uncertainty of some attributes cosine distance, one can find the neutral words as well.
+It makes the usage and interpretation of this metric extremely complicated and unclear.
+
+<!-- debiasing --> 
+Additionally, one cannot be sure if the bias is still preserved after the debiasing.
+The fact that all of the cosine distances for protected words and harmful attributes moved to the right, does not mean that the bias is removed.
+One could argue that maybe it is not clear how to measure bias presence and removal but the method for debiasing works fine.
+Notwithstanding, it is showed in researches such as @Gonen2019Lipstick, that the bias can hide in the vector geometry and preserve even after applying popular debiasing methods.
+Therefore it seems to not be justified to claim that the method works properly without precise methods to verify this.
+
+<!-- no baseline for the interpretation of the results -->
 One should remember that there is no baseline for the interpretation of the averaged cosine distances in @Manzini2019blackToCriminalToCriminal.
@@
 
+<!-- Baysian method --> 
 We propose the use of Bayesian method to introduce the uncertainty measure in bias detection.
+
+// see the chapter 3 to write more
 
@@
 The aim of this paper was to show the analysis of bias detection methods used in modern natural language processing projects.
+
 In the further research it is worth to dive deeper into some issues or to examine new ideas that occurred during our work.
@@
 - delete papers from github/hide them
--
+- should we quote "debiasing" or without quoting
 
 -->
+
+<!-- textbf{Control group}
+
+How to properly prepare control group in terms of quality and quantity?
+<!-- + check if it may be proved in the code -->
+
+<!--  \textbf{Verifying other similarity meaures}
+
+Besides cosine similarity, there are other methods used to find the similarity between vectors.
+
+a) TS-SS 
+todo: check if it makes sense to calculate for all of the datasets; interpretations etc.
+https://www.researchgate.net/publication/303513110_A_Hybrid_Geometric_Approach_for_Measuring_Similarity_Level_Among_Documents_and_Document_Clustering
+
+b) WEAT test 
+<!-- https://docs.responsibly.ai/word-embedding-bias.html#module-responsibly.we.weat -->
+
+<!-- [to check https://github.com/taki0112/Vector_Similarity] -->
+
+<!-- TODO: check in code if they make more sense -->
+
+<!-- ## Bibliography
+- https://www.amazon.com/Language-Processing-Synthesis-Lectures-Technologies/dp/1627052984/ref=as_li_ss_tl?ie=UTF8&qid=1502062931&sr=8-1&keywords=Neural+Network+Methods+in+Natural+Language+Processing&linkCode=sl1&tag=inspiredalgor-20&linkId=d63df073fea3ebe2d405820570b3ff03 ->
+
+<!-- ## *** Additional -->
+
+<!-- ### Man is to Computer Programmer as Woman is to Homemaker? Debiasing Word Embeddings" -->
+<!-- [article https://arxiv.org/pdf/1607.06520.pdf] -->
+
+<!-- Assumptions: -->
+<!-- "We assume there is a set of gender neutral words $N\subseteq W$ such as flight attendant or shoes, which, by definition, are not specific to any gender." -->
+
+<!-- TODO: check if it is useful for the analysis of black is .. -->
+
+<!-- ### Identifying the bias subspace  -->
+
+<!-- Interesting analysis about limits o subspace and research [source https://kawine.github.io/blog/nlp/2019/09/23/bias.html] -->
+
+<!-- Resources to analyze: -->
+<!-- 1. https://arxiv.org/pdf/2005.00965.pdf -->
+<!-- 2. https://arxiv.org/pdf/2009.09435.pdf -->
+<!-- 3. https://arxiv.org/pdf/1904.04047.pdf -->
+<!-- 4. https://www.aclweb.org/anthology/P19-1166.pdf -->
+
+<!-- ### How is cosine similarity entangled with analogy topic in NLP (is it??) -->
+
+<!-- article:  https://arxiv.org/pdf/1905.09866.pdf -->
+
+<!-- Critique of the usage of analogies as a proof for gender biases -->
+
+<!-- TODO: check if there is other argument than that algorithm does not allow it -->
+
+<!-- general thoughts 
+https://www.researchgate.net/publication/349408535_Robustness_and_Reliability_of_Gender_Bias_Assessment_in_Word_Embeddings_The_Role_of_Base_Pairs
+
+-
+-
+-
+-->

### 2021-06-16 thesis/_book/_main.tex
Adds fixed chapter 2 and discussion

--- 2021-06-15 thesis/_book/_main.tex
@@
 \usepackage{todonotes}
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+\usepackage{xcolor}
 \ifluatex
@@
 
-It is worth diving deeper into some observations that arise while analyzing the results.
-
-\begin{itemize}
-\tightlist
-\item
-  the distributions of cosine distances of whole population and individual words can vary enormously
-\item
-  including uncertainty shows how in some cases it is not clear why a word is according to this method
-  classified as anomalous
-\item
-  debiasing with the method provided by \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) does not solve the bias issues
-\end{itemize}
-
-One should remember that there is no baseline for the interpretation of the averaged cosine distances in (\protect\hyperlink{ref-Manzini2019blackToCriminalToCriminal}{\textbf{Manzini2019blackToCriminalToCriminal?}}).
-One may assume that if the cosine distance is close to 1 then it is a desired outcome as it means, according to cosine similarity assumptions, that there is almost no similarity between the words.
-However what does it mean to be close to 0?
-If the averaged cosine similarity is equal to 80, then should we still debiase it?
-It is unclear what the criteria are.
-On one hand, it seems to be beneficial when the outcome is simplified as it is easier to compare results with one value per set.
-On the other hand, it is prone to misunderstanding of how to interpret the results and what threshold to assume.
-
-\hypertarget{what-we-propose-and-how-it-can-help}{%
-\section{What we propose and how it can help:}\label{what-we-propose-and-how-it-can-help}}
-
-\begin{itemize}
-\tightlist
-\item
-\end{itemize}
-
-\hypertarget{summary}{%
-\chapter{Summary}\label{summary}}
-
-\hypertarget{future-work}{%
-\section{Future work:}\label{future-work}}
+Placeholder
 

### 2021-06-16 thesis/04-discussion.Rmd
Adds fixed discussion

@@
 It is worth diving deeper into some observations that arise while analyzing the results. 
-
 <!-- Looking into specific values -->
@@
 <!-- Problem with gender words --> 
-One has to remember also about the specific situation present in gender dataset.
-There surprisingly one could observe high cosine distances values between some female stereotypical professions and male protected words.
-This leads to new ideas of how to understand the bias origins and its detection.
-To start with, cosine distance seems to catch the information from the vectors of words co-occurrence and not the semantic similarity.
-If word stereotypically associated with females has high cosine distance value for male protected words, then the metric is not proper to establish the bias presence as it can be misleading.
+One has to also remember about the specific situation present in gender dataset.
+Surprisingly one could observe there high cosine distances values between some female stereotypical professions and male protected words.
+This leads to new ideas of how to understand the bias origins and its proper detection.
+To start with, cosine distance seems to catch the information regarding words co-occurrence and not the semantic similarity.
+If a word stereotypically associated with females has high cosine distance value for male protected words, then the metric is not proper to establish the bias presence as it can be misleading.
 What we mean is that in same cases it can unnecessarily label words as biased while omitting other more harmful associations.
@@
 In some cases the final cosine distance is indeed close to neutral words, but in other cases the change is minor.
-What is more, including uncertainty shows how in some cases it is not clear why a word is according to this method
+What is more, the inclusion of the uncertainty shows how in some cases it is not clear why a word is according to this method
 classified as anomalous.
-The fact that uncertainty for `assosiated` and `different` class is so high leads to the situation where in the boundaries of an uncertainty of some attributes cosine distance, one can find the neutral words as well.
-It makes the usage and interpretation of this metric extremely complicated and unclear.
+The fact that uncertainty for `assosiated` and `different` class is so high leads to the situation where in the boundaries of an uncertainty of some attributes, one can find the neutral words as well.
+It makes the usage and interpretation of this dataset and metric extremely complicated and unclear.
 
@@
 One should remember that there is no baseline for the interpretation of the averaged cosine distances in @Manzini2019blackToCriminalToCriminal.
-One may assume that if the cosine distance is close to 1 then it is a desired outcome as it means, according to cosine similarity assumptions, that there is almost no similarity between the words.
+One may assume that if the cosine distance is close to 1 then it is a desired outcome as it means, according to cosine distance assumptions, that there is almost no similarity between the words.
 However what does it mean to be close to 0?
-If the averaged cosine similarity is equal to 80, then should we still debiase it?
+If the averaged cosine distance is equal to 80, then should we still debiase it?
 It is unclear what the criteria are.
@@
 We propose the use of Bayesian method to introduce the uncertainty measure in bias detection.
+There are a few advantages when including this method in the bias analysis.
+As noted before in our example with religion dataset, one may obtain new insight into the bias issue after analyzing the mean estimates of the coeﬀicients.
+There seems to be a minor difference for the mean concerning the `assosiated` class.
+This can lead to the conclusions that we do not have strong reasons to perceive lower cosine distance as significantly related to the fact that the attribute is associated.
 
-// see the chapter 3 to write more
+One may dive deeper into this idea by investigating the connection coefficients for individual words with the use of the visualization.
+There one observes that indeed most of the words are clustered together independently of the class that they belong to.
+It is of course partly due to the fact that we include uncertainty now.
+This example shows how understanding of the bias measurement may differ when switching to Bayesian method.
+One could argue that after this analysis one obtains more information on the dataset and can pursue with the further research more cautiously.
 

### 2021-06-16 thesis/_book/_main.tex
Adds fixed discussion

@@
 \usepackage{todonotes}
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{wrapfig}
-\usepackage{float}
-\usepackage{colortbl}
-\usepackage{pdflscape}
-\usepackage{tabu}
-\usepackage{threeparttable}
-\usepackage{threeparttablex}
-\usepackage[normalem]{ulem}
-\usepackage{makecell}
-\usepackage{xcolor}
 \ifluatex
@@
 
+Placeholder
+
 \hypertarget{loading-and-understanding-the-dataset}{%
 \section{Loading and understanding the dataset}\label{loading-and-understanding-the-dataset}}
-
-We start with loading the libraries needed for the analysis.
-
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{library}\NormalTok{(ggplot2)}
-\FunctionTok{library}\NormalTok{(ggthemes)}
-\FunctionTok{library}\NormalTok{(rethinking)}
-\FunctionTok{library}\NormalTok{(tidyverse)}
-\FunctionTok{library}\NormalTok{(ggpubr)}
-\FunctionTok{library}\NormalTok{(kableExtra)}
-\FunctionTok{library}\NormalTok{(dplyr)}
-\FunctionTok{library}\NormalTok{(ggExtra)}
-\FunctionTok{library}\NormalTok{(cowplot)}
-\end{Highlighting}
-\end{Shaded}
-
-\normalsize
-
-We will use the choice of protected words and stereotypical predicates used in REF.
-This is a decent point of departure, not only we want to compare our method to that of REF, but also because this data format is fairly general (as contrasted, say, with a set up for binary stereotypes).
-Note also that the method we develop here can fairly easily be run for different stereotypization patterns.
-Let's start with explaining the method and its deployment using a dataset obtained for the religion-related protected words.
-
-Let's load, clean a bit and inspect the head of the religion dataset we prepared.
-In order to obtain this dataset, we calculated the cosine distance between each protected word and each word from both the bias-related attribute groups, which were used in the original study, and to neutral and human control attributes which we added as control groups.
-For instance, for religion, the bias-related predicates (coming from the original study in REF) include muslim bias attributes, jew bias attributes, christian bias attributes (see a list in the APPENDIX).
-
-We decided to add control groups in the form of two classes --- neutral words and human-related words.
-Without a proper control group it is quite hard to compare the resulting cosine distances and decide on their significance in bias detection.
-We prepared approximately 230 neutral words to double-check the prima-facie neutral hypothesis that their cosine similarity to the protected words will oscillate around 0 (that is, the distances will be around 1).
-This provides us with a more reliable point of reference.
-Moreover, we added human attributes that are associated with people in general to investigate whether the smaller cosine distance between protected words and stereotypes can result simply from the fact that the stereotype predicates are associated with humans.
-For two control groups, we have randomly drawn 230 words that do not express any property usually attributed to humans, and human related attributes.
-
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{religion }\OtherTok{\textless{}{-}} \FunctionTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{]}
-\FunctionTok{colnames}\NormalTok{(religion) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
-                        \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
-\CommentTok{\#head(religion)}
-\CommentTok{\#library(plyr)}
-\NormalTok{religion}\SpecialCharTok{$}\NormalTok{wordClass }\OtherTok{\textless{}{-}} \FunctionTok{as.factor}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{wordClass)}
-\CommentTok{\#levels(religion$wordClass)}
-\CommentTok{\#religion$wordClass \textless{}{-} mapvalues(religion$wordClass, }
-\CommentTok{\#                          from = levels(religion$wordClass),}
-\CommentTok{\#                          to = c("christian","human","jewish","muslim","neutral"))}
-\FunctionTok{levels}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{wordClass) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
-\FunctionTok{head}\NormalTok{(religion)  }\SpecialCharTok{\%\textgreater{}\%}  \FunctionTok{kable}\NormalTok{(}\AttributeTok{format =} \StringTok{"latex"}\NormalTok{,}\AttributeTok{booktabs=}\NormalTok{T,}
-                      \AttributeTok{linesep =} \StringTok{""}\NormalTok{,  }\AttributeTok{escape =} \ConstantTok{FALSE}\NormalTok{, }
-                      \AttributeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}      \FunctionTok{kable\_styling}\NormalTok{(}\AttributeTok{latex\_options=}\FunctionTok{c}\NormalTok{(}\StringTok{"scale\_down"}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{table}
-
-\caption{\label{tab:religionTableHead}Head of the religion dataset.}
-\centering
-\resizebox{\linewidth}{!}{
-\begin{tabular}[t]{lllrrl}
-\toprule
-protectedWord & wordToCompare & wordClass & cosineDistance & cosineSimilarity & connection\\
-\midrule
-judaism & violent & muslim & 0.7141939 & 0.2858061 & different\\
-judaism & terrorist & muslim & 0.7461333 & 0.2538667 & different\\
-judaism & dirty & muslim & 1.2002599 & -0.2002599 & different\\
-judaism & uneducated & muslim & 0.7885469 & 0.2114531 & different\\
-judaism & greedy & jewish & 1.0026172 & -0.0026172 & associated\\
-judaism & cheap & jewish & 1.2323229 & -0.2323229 & associated\\
-\bottomrule
-\end{tabular}}
-\end{table}
-\normalsize
-
-The \texttt{protectedWord} column contains words from a protected class that (in a perfect world according to the assumptions of the orignal study) should not be associated with harmful stereotypes.
-\texttt{wordToCompare} contains attributes, including stereotypes and control group words.
-For each row we compute the cosine distances between a given protected word and a given attribute word.
-\texttt{wordClass} tells us which class an attribute is supposed to be stereotypically associated with, that is, whether the word from \texttt{wordToCompare} is associated stereotypically with jews, christians or muslims, or whether it belongs to a control group.
-\texttt{cosineDistance} is simply a calculation of the cosine distance between protected word and atrribute.
-\texttt{cosineSimilarity} contains the result of substracting cosine distance from 1.
-\texttt{connection} contains information about the relation type between a protected word and an attribute.
-If the attribute is e.g.~a harmful jewish stereotype and the protected word is also from the judaism group, the connection has value \texttt{associated}.
-If the attribute is still stereotypically jewish, but the protected word comes from another religion, the connection is labelled as \texttt{different}.
-If the attribute belongs to a neutral group then the connection is labelled as \texttt{none} and if an attribute belongs to the \texttt{human} class, then the connection is labelled as \texttt{human}.
 
@@
 
-First let's take a look at the empirical distribution of distances by the connection type, initially ignoring the human control class for now.
-
-\vspace{1mm}
-\footnotesize
-
-\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-2-1} \end{center}
-\normalsize
-
-The first impression is that while there is a shift for associated words towards smaller cosine distances as compared to the neutral words, slightly surprisingly a slightly weaker shift in the same direction is visible for attributes associated with different stereotypes.
-Moreover, the empirical distributions overlap to alarge extent and the means grouped by connection type do not seem too far from each other.
-In fact, as there is a lot of variety in the cosine distances (as we will soon see), we need to gauge the uncertainty involved, and to look more carefully at individual protected words to get a better idea of how the cosine distance distribution changes for different attribute groups and different protected classes.
-Now, let's add the human attributes to the picture:
-
-\vspace{1mm}
-\footnotesize
-
-\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-3-1} \end{center}
-\normalsize
-
-\noindent Notice that the distribution for \texttt{human} (even though we did our best not to include in it any stereotype-related atributes) is left-skewed, with much overlap with \texttt{associated} and \texttt{different}, which illustrates the need to take being associated with humans as an important predictor.
-
-Our focus lies in \texttt{connection} as a predictor.
-Morever, later on we'll be interested in looking at the protected words separately, and at protected words split by connection.
-For technical reasons it is useful to represent these factors as integer vectors.
-
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{religion}\SpecialCharTok{$}\NormalTok{con }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{connection)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Warning: NAs introduced by coercion
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{religion}\SpecialCharTok{$}\NormalTok{pw }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{protectedWord)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Warning: NAs introduced by coercion
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{religion}\SpecialCharTok{$}\NormalTok{pwFactor }\OtherTok{\textless{}{-}} \FunctionTok{factor}\NormalTok{(}\FunctionTok{paste0}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{protectedWord, religion}\SpecialCharTok{$}\NormalTok{connection))}
-\NormalTok{religion}\SpecialCharTok{$}\NormalTok{pwIndex }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{pwFactor)}
-\end{Highlighting}
-\end{Shaded}
-
-\normalsize
-
-A short script, \texttt{cleanDataset} to make this faster, so equivalently:
-
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{source}\NormalTok{(}\StringTok{"../functions/cleanDataset.R"}\NormalTok{)}
-\NormalTok{religion }\OtherTok{\textless{}{-}} \FunctionTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{]}
-\NormalTok{religion }\OtherTok{\textless{}{-}} \FunctionTok{cleanDataset}\NormalTok{(religion,}\FunctionTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Warning in cleanDataset(religion, c("christian", "human", "jewish", "muslim", :
-## NAs introduced by coercion
-
-## Warning in cleanDataset(religion, c("christian", "human", "jewish", "muslim", :
-## NAs introduced by coercion
-\end{verbatim}
-
-\normalsize
-
 \hypertarget{looking-at-the-islam-related-words}{%
 \section{Looking at the islam-related words}\label{looking-at-the-islam-related-words}}
-
-For now, let's focus on five protected words related to islam (``imam,'' ``islam,'' ``mosque,'' ``muslim,'' and ``quran'').
-The word list associates with islam four stereotypical attributes (``violent,'' ``terrorist,'' ``uneducated'' and ``dirty'').
-First, we select and plot the empirical distributions for these protected words.
-
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{library}\NormalTok{(tidyverse)}
-\NormalTok{muslimWords }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"imam"}\NormalTok{,}\StringTok{"islam"}\NormalTok{,}\StringTok{"mosque"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"quran"}\NormalTok{)}
-\NormalTok{muslim }\OtherTok{\textless{}{-}}\NormalTok{ religion }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(protectedWord }\SpecialCharTok{\%in\%}\NormalTok{ muslimWords)}
-\FunctionTok{ggplot}\NormalTok{(muslim, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{  cosineDistance, }\AttributeTok{fill =}\NormalTok{ connection, }\AttributeTok{color =}\NormalTok{ connection))}\SpecialCharTok{+}
-  \FunctionTok{geom\_density}\NormalTok{(}\AttributeTok{alpha=}\FloatTok{0.6}\NormalTok{,}\AttributeTok{size =}\NormalTok{ .}\DecValTok{2}\NormalTok{)}\SpecialCharTok{+} 
-  \FunctionTok{scale\_fill\_manual}\NormalTok{(}\AttributeTok{values =} \FunctionTok{c}\NormalTok{(}\StringTok{"orangered4"}\NormalTok{,}\StringTok{"chartreuse4"}\NormalTok{, }\StringTok{"skyblue"}\NormalTok{, }\StringTok{"gray"}\NormalTok{))}\SpecialCharTok{+}
-  \FunctionTok{scale\_x\_continuous}\NormalTok{(}\AttributeTok{breaks =} \FunctionTok{seq}\NormalTok{(}\FloatTok{0.3}\NormalTok{,}\FloatTok{1.5}\NormalTok{, }\AttributeTok{by =} \FloatTok{0.1}\NormalTok{))}\SpecialCharTok{+}\FunctionTok{xlab}\NormalTok{(}\StringTok{"cosine distance"}\NormalTok{)}\SpecialCharTok{+}
-  \FunctionTok{scale\_color\_manual}\NormalTok{(}\AttributeTok{values =} \FunctionTok{c}\NormalTok{(}\StringTok{"orangered4"}\NormalTok{,}\StringTok{"chartreuse4"}\NormalTok{,}\StringTok{"skyblue"}\NormalTok{,}\StringTok{"gray"}\NormalTok{))}\SpecialCharTok{+}
-  \FunctionTok{theme\_tufte}\NormalTok{()}\SpecialCharTok{+}\FunctionTok{ggtitle}\NormalTok{(}\StringTok{"Empirical distribution of distances (muslim)"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-6-1} \end{center}
-\normalsize
-
-\noindent Once we focus on words related to islam, the associated bias seems to be stronger than in the whole dataset. This is a step towards illustrating that the distribution of bias is uneven.
-
-Now, say we want to look at a single protected word.
-Since the dataset also contains comparison multiple control neutral and human attributes, we randomly select only 5 from \texttt{none} and 5 from \texttt{human} control groups of those for the visualisation purposes.
-
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{library}\NormalTok{(tidyverse)}
-\NormalTok{muslimClass }\OtherTok{\textless{}{-}}\NormalTok{ muslim }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(protectedWord }\SpecialCharTok{==} \StringTok{"muslim"}\NormalTok{)}
-\NormalTok{neutralSample }\OtherTok{\textless{}{-}} \FunctionTok{sample\_n}\NormalTok{(}\FunctionTok{filter}\NormalTok{(muslimClass,connection }\SpecialCharTok{==} \StringTok{"none"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
-\NormalTok{humanSample }\OtherTok{\textless{}{-}} \FunctionTok{sample\_n}\NormalTok{(}\FunctionTok{filter}\NormalTok{(muslimClass,connection }\SpecialCharTok{==} \StringTok{"human"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
-\NormalTok{muslimVis }\OtherTok{\textless{}{-}}\NormalTok{ muslimClass }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(connection }\SpecialCharTok{!=} \StringTok{"none"} \SpecialCharTok{\&}\NormalTok{ connection }\SpecialCharTok{!=}\StringTok{"human"}\NormalTok{)}
-\NormalTok{muslimVis }\OtherTok{\textless{}{-}} \FunctionTok{rbind}\NormalTok{(muslimVis,neutralSample,humanSample)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Warning in `[<-.factor`(`*tmp*`, ri, value = structure(c("muslim_stereotype", :
-## invalid factor level, NA generated
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\#we plug in our visualisation script}
-\FunctionTok{source}\NormalTok{(}\StringTok{"../functions/visualisationTools.R"}\NormalTok{)}
-\CommentTok{\#two arguments: dataset and protected word}
-\FunctionTok{visualiseProtected}\NormalTok{(muslimVis,}\StringTok{"muslim"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/tableMuslimActive-1} \end{center}
-\normalsize
-
-Note that the distance between the grey point and the other points is proportional to cosine distance, the non-grey point size is proportional to cosine similarity to the protected word, and color groups by the connection type.
-So for \texttt{muslim} it seems that the stereotypes coming from the word list are fairly well visible.
-To give you some taste of how uneven the dataset is, compare this to what happens with \texttt{priest}.
-
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{library}\NormalTok{(tidyverse)}
-\NormalTok{priestClass }\OtherTok{\textless{}{-}}\NormalTok{ religion }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(protectedWord }\SpecialCharTok{==} \StringTok{"priest"}\NormalTok{)}
-\NormalTok{neutralSample }\OtherTok{\textless{}{-}} \FunctionTok{sample\_n}\NormalTok{(}\FunctionTok{filter}\NormalTok{(priestClass,connection }\SpecialCharTok{==} \StringTok{"none"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
-\NormalTok{humanSample }\OtherTok{\textless{}{-}} \FunctionTok{sample\_n}\NormalTok{(}\FunctionTok{filter}\NormalTok{(priestClass,connection }\SpecialCharTok{==} \StringTok{"human"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
-\NormalTok{priestVis }\OtherTok{\textless{}{-}}\NormalTok{ priestClass }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(connection }\SpecialCharTok{!=} \StringTok{"none"} \SpecialCharTok{\&}\NormalTok{ connection }\SpecialCharTok{!=}\StringTok{"human"}\NormalTok{)}
-\NormalTok{priestVis }\OtherTok{\textless{}{-}} \FunctionTok{rbind}\NormalTok{(priestVis,neutralSample,humanSample)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Warning in `[<-.factor`(`*tmp*`, ri, value = structure(c("muslim_stereotype", :
-## invalid factor level, NA generated
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\#we plug in our visualisation script}
-\FunctionTok{source}\NormalTok{(}\StringTok{"../functions/visualisationTools.R"}\NormalTok{)}
-\CommentTok{\#two arguments: dataset and protected word}
-\FunctionTok{visualiseProtected}\NormalTok{(priestVis,}\StringTok{"priest"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-7-1} \end{center}
-\normalsize
-
-\noindent Here you can see that some human attributes are closer than stereotype attributes, and that there is no clear reason to claim that \texttt{associated} attributes are closer than \texttt{different} or \texttt{human} attributes.
-This, again, illustrates the need of case-by-case analysis with control groups.
-
-The general idea now is that the word lists provided in different pieces of research are just samples of attributes associates with various stereotypes and should be treated as such: the uncertainty involved and the sample sizes should have clear impact on our estimates.
 
@@
 
-We will now think of cosine distance as the output variable, and
-will build a few Bayesian models to compare.
-First, we just build a baseline model which estimates cosine distance to the attributes separately for each protected word.
-The underlying idea is that different protected words might in general have different relations to all the attributes and these relations should be our point of departure.
-
-Here is the intuition behind the mathematical Bayesian model involved.
-Our outcome variable is \texttt{cosine\ difference}, which we take to me normally distributed around the predicted mean for a given protected word (that is, we assume the residuals are normally distributed).
-The simplest model specification is:
-
-\begin{align}
-cosineDistance_i  & \sim dnorm(\mu_i, \sigma) \\
-\mu_i & = m_{pw} \\
-m_{pw} & ~ dnorm(1,.5) \\
-\sigma &\sim  dcauchy(0,1)
-\end{align}
-
-That is, we assume the estimated means might be different for diferent protected words and our prior for the mean and the overal standard deviation are normal with mean 1 and sd=.5 and half-cauchy with parameters \texttt{0,1}.
-Further on we'll also estimate additional impact the connection type may have.
-For this impact we take a slightly skeptical prior centered around 0 distributed normally with sd = 1.
-These are fairly weak and slightly skeptical regularizing priors, which can be illustrated as follows:
-
-\vspace{2mm}
-
-\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/priorsVis-1} \end{center}
-
 \hypertarget{choosing-predictors}{%
 \section{Choosing predictors}\label{choosing-predictors}}
-
-Now we can define and compile the baseline model.
-Its parameters will have a posterior distribution obtained using either Hamiltionian Monte Carlo methods (STAN) available through the \texttt{rethinking} package.
-
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{library}\NormalTok{(rethinking)}
-\FunctionTok{options}\NormalTok{(}\AttributeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\ConstantTok{TRUE}\NormalTok{ )}
-\NormalTok{religionBaseline }\OtherTok{\textless{}{-}} \FunctionTok{ulam}\NormalTok{(}
-  \FunctionTok{alist}\NormalTok{(}
-\NormalTok{    cosineDistance }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(mu,sigma),}
-\NormalTok{    mu }\OtherTok{\textless{}{-}}\NormalTok{ m[pw],}
-\NormalTok{    m[pw] }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
-\NormalTok{    sigma }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
-\NormalTok{  ),}
-  \AttributeTok{data =}\NormalTok{ religion,}
-  \AttributeTok{chains=}\DecValTok{2}\NormalTok{ , }\AttributeTok{iter=}\DecValTok{4000}\NormalTok{ , }\AttributeTok{warmup=}\DecValTok{1000}\NormalTok{,}
-  \AttributeTok{start=} \FunctionTok{list}\NormalTok{(}\AttributeTok{mu =} \DecValTok{1}\NormalTok{, }\AttributeTok{co =} \DecValTok{0}\NormalTok{, }\AttributeTok{sigma=}\NormalTok{ .}\DecValTok{3}\NormalTok{),}
-  \AttributeTok{log\_lik =} \ConstantTok{TRUE}\NormalTok{, }\AttributeTok{cores=}\DecValTok{4}
-\NormalTok{)}
-\CommentTok{\#saving}
-\CommentTok{\#saveRDS(religionBaseline, }
-\CommentTok{\#file = "cosineAnalysis/models/religionBaseline.rds")}
-\end{Highlighting}
-\end{Shaded}
-
-The only reason we need it is the evaluation of connection as a predictor.
-Does including it in o the model improve the situation?
-To investigate this, let's now build a model according to the following specification:
-
-\begin{align}
-cosineDistance_i  & \sim dnorm(\mu_i, \sigma) \\
-\mu_i & = m_{pw} + co_{con}\\
-m_{pw} & ~ dnorm(1,.5) \\
-co_{con} & ~ dnorm(0,1) \\
-\sigma &\sim  dcauchy(0,1)
-\end{align}
-
-\noindent The idea now is that each connection type comes with its own coefficient \(co\) that has impact on mean distances for protected words taken separately.
-
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{library}\NormalTok{(rethinking)}
-\FunctionTok{options}\NormalTok{(}\AttributeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\ConstantTok{TRUE}\NormalTok{ )}
-\NormalTok{religionCoefs }\OtherTok{\textless{}{-}} \FunctionTok{ulam}\NormalTok{(}
-  \FunctionTok{alist}\NormalTok{(}
-\NormalTok{    cosineDistance }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(mu,sigma),}
-\NormalTok{    mu }\OtherTok{\textless{}{-}}\NormalTok{ m[pw] }\SpecialCharTok{+}\NormalTok{ co[con],}
-\NormalTok{    m[pw] }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
-\NormalTok{    co[con] }\SpecialCharTok{\textasciitilde{}}\FunctionTok{dnorm}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
-\NormalTok{    sigma }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
-\NormalTok{  ),}
-  \AttributeTok{data =}\NormalTok{ religion,}
-  \AttributeTok{chains=}\DecValTok{2}\NormalTok{ , }\AttributeTok{iter=}\DecValTok{8000}\NormalTok{ , }\AttributeTok{warmup=}\DecValTok{1000}\NormalTok{, }
-  \AttributeTok{log\_lik =} \ConstantTok{TRUE}
-\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\normalsize
-
-\noindent First, let's see if this model is really better in terms of the Widely Acceptable Information Criterion (WAIC):
-
-\vspace{1mm}
-\footnotesize
-
-\begin{verbatim}
-##                   WAIC SE dWAIC dSE pWAIC weight
-## religionCoefs    -2328 93     0  NA    20      1
-## religionBaseline -2283 95    45  17    16      0
-\end{verbatim}
-
-\normalsize
-
-Clearly, it should be given weight 1 as compared to the baseline model. So far, we've learned that the connection type actually has predictive value. Let's take a look at the coefficient estimates:
-
-\vspace{1mm}
-\footnotesize
-
-\begin{verbatim}
-##              mean        sd       5.5%      94.5%    n_eff    Rhat4
-## co[1] -0.14956420 0.1151675 -0.3261650 0.03741930 294.2033 1.001449
-## co[2] -0.09880543 0.1145024 -0.2736985 0.08813271 291.5044 1.001564
-## co[3] -0.07282752 0.1133894 -0.2447778 0.11158986 287.7820 1.001627
-## co[4] -0.03103179 0.1131420 -0.2034442 0.15268770 286.8283 1.001606
-\end{verbatim}
-
-\normalsize
 
@@
 
-It seems that we do want to prefer this model, despite its relative complication.
-Now, what does it tell us about the protected words?
-Let's visualise the predicted means together with 89\% highest posterior density intervals.
-
-\includegraphics[width=14cm]{../images/visReligionReddit.png}
-
-Before we move on, let's perform analogous analyses for the remaining types of supposed bias: gender and race (the model building is analogous).
-
-\includegraphics[width=14cm]{../images/visGenderReddit.png}
-
-\includegraphics[width=14cm]{../images/visRaceReddit.png}
-
-We first encountered the GoogleNews word embedding model in \protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}).
-They use it for their calculations as written in \href{https://github.com/tolga-b/debiaswe}{Tolga Code}.
-We decided to compare the results obtained with the use of Reddit L2 model and the ones that we got by GoogleNews model.
-The details of the model can be found here \href{https://code.google.com/archive/p/word2vec/}{Google News model}.
-One of the main differences between the models is that Reddit word embeggings have 50 dimensions and GoogleNews word embeddings have 300 dimensions.
-As the dimension is increasing, the vector can capture much more information but this is not always the case.
-According to some researchers the amount of dimensions should be chosen by taking into account the type of corpus and its features.
-In our case both Reddit and GoogleNews models are already trained and ready to use so we do not analyse further the choice of their hyperparameters.
-
-\includegraphics[width=14cm]{../images/visReligionGoogle.png}
-
-\includegraphics[width=14cm]{../images/visGenderGoogle.png}
-
-\includegraphics[width=14cm]{../images/visRaceGoogle.png}
-
-\hypertarget{the-role-of-debiasing}{%
-\chapter{The role of debiasing}\label{the-role-of-debiasing}}
-
-We also applied to the word embeddings hard debiasing method found in \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}).
-As it was mentioned before, debiasing consists of two components: identifying the bias subspace and then removing this vector subspace from the chosen embeddings.
-The aim is to increase the cosine distance between protected words and the set of attributes.
-After debiasing the Reddit word embeddings we calculated the cosine distance again to see if there are any significant changes in terms of the similarity between protected words and the attributes.
-
 \hypertarget{dataset-level-coefficients-after-debiasing}{%
 \section{Dataset-level coefficients after debiasing}\label{dataset-level-coefficients-after-debiasing}}
-
-First, let's look at coefficient estimated for the whole datasets, as compared to their estimation prior to debiasing:
-
-\begin{center}
-\begin{figure}[!htb]\centering
-   \begin{minipage}{0.55\textwidth}
-  \includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
-   \end{minipage}
-   \begin {minipage}{0.43\textwidth}
-    \includegraphics[width=7cm]{../images/debiasedReligionRedditCoeffs.jpeg}
-   \end{minipage}
-
-  \begin{minipage}{0.55\textwidth}
-\includegraphics[width=7cm]{../images/genderCoeffs.jpeg}
-\end{minipage}
-   \begin {minipage}{0.43\textwidth}
-    \includegraphics[width=7cm]{../images/debiasedGenderRedditCoeffs.jpeg}
-   \end{minipage}
-
-  \begin{minipage}{0.55\textwidth}
-\includegraphics[width=7cm]{../images/raceCoeffs.jpeg}
-\end{minipage}
-   \begin {minipage}{0.43\textwidth}
-    \includegraphics[width=7cm]{../images/debiasedRaceRedditCoeffs.jpeg}
-   \end{minipage}
-\end{figure}
-
-\end{center}
-
-One may observe very slight changes in the estimated coeﬀicients.
-It seems that in religion dataset the changes are the most significant as the mean moves towards zero which stands for no similarity between words.
 
@@
 
-Now, perhaps, the effects of debiasing will be better appreciated if we look at the level of protected words.
-After all, the hope is, the situation of extremely ill-positioned protected words have improved?
-
-\includegraphics[width=14cm]{../images/visDebReligionReddit.png}
-
-\includegraphics[width=14cm]{../images/visDebGenderReddit.png}
-
-\includegraphics[width=14cm]{../images/visDebRaceReddit.png}
-
-One may notice that in a religion data set most of the distances for \texttt{associated} words slightly increased.
-Some of the distances from \texttt{different} class are still smaller than for the \texttt{associated} class.
-The distances for all classes seem in most cases to cluster together even after debiasing.
-The reason may be that uncertainty is quite high in \texttt{associated} and \texttt{different} class because the number of samples is so small.
-Additionally in some cases distances from \texttt{human} class are smaller than for the \texttt{assosiated} class which also makes it less clear how to interpret the results, if the distances are really able to catch the bias presence.
-
-It is worth paying attention to the gender debiased dataset.
-The change is really minor, still zero out of HPDI range.
-The \texttt{different} attributes still have high similarity value.
-It seems that the distances for gender protected words are still clustered together which suggests that this method is not able to catch the complex bias presence.
-One could conclude that co-occurrence of certain terms does not always mean direct connection in terms of terms associations.
-The fact that some of stereotypically associated with female attributes still have high similarity with male protected words suggest that other techniques should be applied to detect and remove the bias.
-
-At the same time, there is a small improvement in race dataset.
-In some cases the distances just cluster together.
-In general distances for \texttt{assosiated} class are still higher than those for the rest of classes.
-One may assume that there reason for so small change is that either the metric does not indicate properly biases or that the issue is more complex and subtracting subspaces is not enough.
-
-In \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) it is mentioned that debiasing with this method that is based on bias component removal may be insufficient.
-The reason is that words containing similar biases may still cluster together even after debiasing.
-
 \hypertarget{discussion}{%
@@
 
-Placeholder
+It is worth diving deeper into some observations that arise while analyzing the results.
+As we pointed out, if there is no control of what cosine distance values individual word pairs have, then there is less
+understanding of the exact relations between words.
+If the protected words that we chose do not have high similarity with harmful stereotypes, then one should consider at least two scenarios.
+The first one is that the choice of protected words and attributes may be corrupted.
+The second one is that the metric is not able to catch the hidden bias properly.
+In both cases it is essential to take a look at the individual values before averaging them or aggregating in other ways.
+
+One has to also remember about the specific situation present in gender dataset.
+Surprisingly one could observe there high cosine distances values between some female stereotypical professions and male protected words.
+This leads to new ideas of how to understand the bias origins and its proper detection.
+To start with, cosine distance seems to catch the information regarding words co-occurrence and not the semantic similarity.
+If a word stereotypically associated with females has high cosine distance value for male protected words, then the metric is not proper to establish the bias presence as it can be misleading.
+What we mean is that in same cases it can unnecessarily label words as biased while omitting other more harmful associations.
+What is more, this highlights again the need for control group.
+One could check each stereotypical attribute with each protected word but not in order to average the values.
+One should rather plot the distributions and investigate how the metric measures vectors geometry and hidden semantic information.
+
+Debiasing with the method provided by \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) shows how it is unpredictable in terms of the results.
+In some cases the final cosine distance is indeed close to neutral words, but in other cases the change is minor.
+What is more, the inclusion of the uncertainty shows how in some cases it is not clear why a word is according to this method
+classified as anomalous.
+The fact that uncertainty for \texttt{assosiated} and \texttt{different} class is so high leads to the situation where in the boundaries of an uncertainty of some attributes, one can find the neutral words as well.
+It makes the usage and interpretation of this dataset and metric extremely complicated and unclear.
+
+Additionally, one cannot be sure if the bias is still preserved after the debiasing.
+The fact that all of the cosine distances for protected words and harmful attributes moved to the right, does not mean that the bias is removed.
+One could argue that maybe it is not clear how to measure bias presence and removal but the method for debiasing works fine.
+Notwithstanding, it is showed in researches such as \protect\hyperlink{ref-Gonen2019Lipstick}{Gonen \& Goldberg} (\protect\hyperlink{ref-Gonen2019Lipstick}{2019}), that the bias can hide in the vector geometry and preserve even after applying popular debiasing methods.
+Therefore it seems to not be justified to claim that the method works properly without precise methods to verify this.
+
+One should remember that there is no baseline for the interpretation of the averaged cosine distances in (\protect\hyperlink{ref-Manzini2019blackToCriminalToCriminal}{\textbf{Manzini2019blackToCriminalToCriminal?}}).
+One may assume that if the cosine distance is close to 1 then it is a desired outcome as it means, according to cosine distance assumptions, that there is almost no similarity between the words.
+However what does it mean to be close to 0?
+If the averaged cosine distance is equal to 80, then should we still debiase it?
+It is unclear what the criteria are.
+On one hand, it seems to be beneficial when the outcome is simplified as it is easier to compare results with one value per set.
+On the other hand, it is prone to misunderstanding of how to interpret the results and what threshold to assume.
+
+We propose the use of Bayesian method to introduce the uncertainty measure in bias detection.
+There are a few advantages when including this method in the bias analysis.
+As noted before in our example with religion dataset, one may obtain new insight into the bias issue after analyzing the mean estimates of the coeﬀicients.
+There seems to be a minor difference for the mean concerning the \texttt{assosiated} class.
+This can lead to the conclusions that we do not have strong reasons to perceive lower cosine distance as significantly related to the fact that the attribute is associated.
+
+One may dive deeper into this idea by investigating the connection coefficients for individual words with the use of the visualization.
+There one observe that indeed most of the words are clustered together independently of the class they belong to.
+It is of course partly due to the fact that we include uncertainty now.
+This example shows how understanding of the bias measurement may differ when switching to Bayesian method.
+One could argue that after this analysis one obtains more information on the dataset and can pursue with the further research more cautiously.
+
+\hypertarget{summary}{%
+\chapter{Summary}\label{summary}}
+
+The aim of this paper was to show the analysis of bias detection methods used in modern natural language processing projects.
+
+In the further research it is worth to dive deeper into some issues or to examine new ideas that occurred during our work.
 
@@
 \begin{CSLReferences}{1}{0}
-\leavevmode\hypertarget{ref-Bolukbasi2016Man}{}%
-Bolukbasi, T., Chang, K.-W., Zou, J.
-Y., Saligrama, V., \& Kalai, A.
-(2016).
-Man is to computer programmer as woman is to homemaker?
-Debiasing word embeddings.
-\emph{CoRR}, \emph{abs/1607.06520}.
-Retrieved from \url{http://arxiv.org/abs/1607.06520}
+\leavevmode\hypertarget{ref-Gonen2019Lipstick}{}%
+Gonen, H., \& Goldberg, Y.
+(2019).
+Lipstick on a pig: Debiasing methods cover up systematic gender biases in word embeddings but do not remove them.
+\emph{CoRR}, \emph{abs/1903.03862}.
+Retrieved from \url{http://arxiv.org/abs/1903.03862}
 

### 2021-06-16 thesis/_book/_main.tex
up-to-date

@@
 
-Placeholder
-
 \hypertarget{word-embeddings}{%
 \section{Word embeddings}\label{word-embeddings}}
+
+To understand what cosine similarity measurement is, one first needs to grasp the concept of translating words to a computer-readable form.
+In the field of natural language processsing there are two main types of words representation --- localist and distributed.
+One-hot encoding is an example of a method used to achieve a localist representation of words.
+Here each vector contains information only about a single data point, this is achieved by first mapping categorical values (words) to integers and then to each integers a binary vector is assigned which contains only 0s except for the index of the integer, which is assigned 1.
+An example of a localist representation is:
+
+\begin{longtable}[]{@{}llllll@{}}
+\toprule
+word & 1 & 2 & 3 & 4 & 5 \\
+\midrule
+\endhead
+woman & 1 & 0 & 0 & 0 & 0 \\
+man & 0 & 1 & 0 & 0 & 0 \\
+girl & 0 & 0 & 1 & 0 & 0 \\
+boy & 0 & 0 & 0 & 1 & 0 \\
+monarch & 0 & 0 & 0 & 0 & 1 \\
+\bottomrule
+\end{longtable}
+
+In the example above it is clear that the length of the vectors increases with the number of words in a vocabulary.
+It is not a very computationally efficient representation.
+It has other flaws as well.
+For example, it is unable to capture the resemblance between words appearing in similar contexts.
+
+In contrast to the localist representation, a distributed representation returns vectors that contain continuous values instead of discrete 1s and 0s.
+Word embeddings are a class of various techniques that allow one to represent words as distributed vectors.
+Such learned representations of text have certain properties.
+At least prima facie, they store similar (or at least co-occurring) words close to each other in a vector space.
+An example of distributed representation is:
+
+\begin{longtable}[]{@{}lllll@{}}
+\toprule
+word & 1 & 2 & 3 & 4 \\
+\midrule
+\endhead
+woman & 0.456 & 0.267 & 0.675 & 0.131 \\
+man & 0.451 & 0.897 & 0.472 & 0.088 \\
+girl & 0.604 & 0.262 & 0.414 & 0.706 \\
+boy & 0.279 & 0.172 & 0.475 & 0.010 \\
+monarch & 0.565 & 0.678 & 0.463 & 0.975 \\
+\bottomrule
+\end{longtable}
+
+One of the advantages of using a distributed representation is that one is able to represent an enormous number of concepts with a smaller number of units.
+It is also possible to better capture similarities as words of similar meanings can have similar numeric vectors.
+
+The numbers occurring in such representations are not random.
+They are learned in a process that uses a very shallow neural network.
+There are various types of techniques used for learning the vectors representations.
+One of the most straightforward ones is a skip-gram model.
+Given a word the model tries to predict its neighboring words from the sentence.
+The mathematics behind the process relies on the idea that the prediction concerns the conditional probability of the adjacent words.
+The algorithm tries to minimize the loss function, which penalizes the system for discrepancy with actual co-occurrence frequencies in the corpus.
+One can choose various parameters of the model, such as the window size that determines how many surrounding words the model should predict.
+After preparing such a fitted modeln=, one takes only the learned weights from a neural network, and uses them as vectors in a word embeddings representation.
+
+Word embeddings have many applications in natural language processing.
+They are handy in document search and information retrieval.
+They also play their part in improving automatic translations.
+Well learned word representations may also contribute to the improvement of sentiment analysis or spam detection.
 
@@
 
+Cosine similarity is often used as a method of finding out whether vector representations for two words suggest that they are similar or somehow connected.
+Cosine similarity is the cosine of the angle between two vectors: the result of dividing their inner product (dot product usually) by the product of their magnitudes.
+
+It is worth mentioning one more point concerning cosine similarity.
+After the vectors are normalized to have length equal to 1, inner product itself (often dot product) is used to measure the similarity.
+
+\begin{align} \tag{Sim}
+\mathsf{cosineSimilarity}(A,B) & = \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert}
+\end{align}
+Cosine similarity is considered a proper tool for this operation as its result has a clear connection to geometry and at least for a low number of dimensions may be easily interpreted.
+Using this scale, one can compare vector similarities in a fairly clear manner.
+When the vectors are aligned perpendicularly to each other, their similarity equals 0 (which is the same as the cosine of 90 degrees).
+This tells us that the similarity between the vectors is small.
+As the angle between vectors decreases, cosine similarity approaches one, which stands for the greatest similarity.
+
+One of the limitations of this measure is that it informs us only about similarities between vectors in terms of their orientation.
+However, it is often argued that in comparing words in terms of this metric, the magnitude of vectors may be treated as irrelevant, as the most important information pertains to direction.
+
+In what follows, it is important to distinguish between cosine similarity and cosine distance, defined as:
+\begin{align} \tag{Sim}
+\mathsf{cosineDistance}(A,B) &  = 1 - \mathsf{cosineSimilarity}(A,B)\\
+ &  = 1 - \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert} \nonumber
+\end{align}
+
+The greater the similarity between two vectors, the smaller the distance between them.
+The cosine distance ranges between 0 and 2.
+If the vectors are in an opposite direction to each other, the cosine distance is 2.
+And if the vectors are extremely similar then the cosine distance is very close to 0.
+
+One should note that cosine distance is not exactly a distance measure while it does not meet triangle inequality requirements.
+The triangle inequality formula says that for any triangle, the sum of the lengths of any two sides must be greater than or equal to the length of the remaining side.
+As shown in \href{https://stats.stackexchange.com/questions/198080/proving-that-cosine-distance-function-defined-by-cosine-similarity-between-two-u}{Statexchange.com} in the case of cosine distance it would have to fulfill this equation 1+cos-sim(𝐴,𝐶) \textless{} cos-sim(𝐴,𝐵) + cos-sim(𝐵,𝐶).
+If one chooses specific unit vectors it is easy to demonstrate that the triangle inequality is not preserved.
+
 \hypertarget{cosine-distance-in-a-one-class-bias-detection}{%
 \section{Cosine distance in a one-class bias detection}\label{cosine-distance-in-a-one-class-bias-detection}}
+
+\protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) define similarity between words as the outcome inner product of their normalized vectors.
+They focus on examining what the geometry of word embedding is in regard to ``he'' and ``she'' words.
+In other words, whether the similarity between those concepts and other words reflects expected gender stereotypes.
+They test this hypothesis by investigating whether there is a connection between word embeddings representing certain professions and words referring to gender.
+They also evaluate whether automatically produced analogies between words reflect the stereotypes as well.
+
+Here the gender bias of a word \(w\) is understood as its projection on the gender direction \(\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})\) (the gender direction is the top principal compontent of ten gender pair difference vectors).
+The underlying idea is that no bias is present if non-explicitly gender words are in equal distance to both elements in all explicitly gender pairs.
+Given the (ideally) gender neutral words \(N\) and the gender direction \(g\) the direct direct gender bias is defined as the average distance of the words in \(N\) from \(g\) (\(c\) is a parameter determining how strict we want to be):
+\begin{align}
+\mathsf{directBias_c(N,g)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},g)\vert^c}{\vert N \vert }
+\end{align}
+
+A very vivid way to follow their method of arguing that bias in word embeddings is real is to plot the values of inner product of chosen words.
+The plot below does not originate from the original paper (it is from \url{https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings}) but similar visualization may be found there.
+Data used to create our plot is as follows.
+
+Occupations associated with feminine: \textbf{"homemaker", "nurse", "receptionist", "librarian", "socialite", "hairdresser", "nanny", "bookkeeper", "stylist", "housekeeper", "interior designer", "guidance counselor"}
+
+Occupations associated with masculine: \textbf{"maestro", "skipper", "protege", "philosopher", "captain", "architect", "financier", "warrior", "broadcaster", "magician", "fighter pilot", "boss"}
+
+\vspace{1mm}
+\footnotesize
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-1-1} \end{center}
+\normalsize
+
+The points in the plot above result from the calculation of the inner product of a chosen vector for a profession word and a vector for a gender word (she or he).
+Inner product of two vectors expresses similarity between words.
+This assumption originates from the geometry and properties of a vector space.
 
@@
 
+\protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) present a different approach towards finding similarities between classes of words.
+The authors claim that texts available online are full of direct or indirect human stereotypes.
+As a result, word embeddings are prone to learn and maybe amplify those biases and propagate them further into AI models that are used for various applications.
+Cosine distance is used in the article as a measure to first argue for the existence of multi-class bias and then to show how through bias mitigation techniques the bias may be decreased.
+
+They modify WEAT to a multi-class setting, introducing Mean Average Cosine similarity as a measure of bias (in fact, in the paper they report distances rather than similarities).
+Let \(T = \{t_1, \dots, t_k\}\) be a class of protected word embeddings, and let each \(A_j\in A\) be a set of attributes stereotypically associated with a protected word).
+Then:
+\begin{align}
+S(t_i, A_j) & = \frac{1}{\vert A_j\vert}\sum_{a\in A_j}\mathsf{cos}(t,a) \\
+MAC(T,A) & = \frac{1}{\vert T \vert \,\vert A\vert}\sum_{t_i \in T }\sum_{A_j \in A} S(t_i,A_j)
+\end{align}
+
+As the code has been provided by the authors we were able to reconstruct their results.
+The main steps in the procedure are as follows.
+Let us go through an example that refers to the process of hard debiasing on religious attributes.
+
+First we load word embeddings from reddit.US.txt.tok.clean.cleanedforw2v.w2v dataset.
+The word embeddings have only 50 dimensions and the number of individual words from the dataset is 44895.
+The authors assume that the protected group should ideally not have high cosine similarity to stereotypical words.
+The word embeddings geometry should not place this group close to harmful stereotypes, if it is to be bias-free.
+For instance, let's look at the religion-related words.
+
+Protected words by religion type:
+
+\textbf{"jew"       : ["judaism", "jew", "synagogue", "torah", "rabbi"]}
+
+\textbf{"christian" : ["christianity", "christian", "church", "bible", "priest"]}
+
+\textbf{"muslim"    : ["islam", "muslim", "mosque", "quran", "imam"]}
+
+Stereotypical words by religion type:
+
+\textbf{"jew"       : ["greedy", "cheap", "hairy", "liberal"]}
+
+\textbf{"christian"       : ["judgemental", "conservative", "familial"]}
+
+\textbf{"muslim"       : ["violent", "terrorist", "dirty", "uneducated"]}
+
+We have prepared a table presenting the values of cosine distance for each protected word with each attribute (stereotype). The part of the results is shown below.
+
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{religion }\OtherTok{\textless{}{-}} \FunctionTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{]}
+\FunctionTok{colnames}\NormalTok{(religion) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
+                        \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
+\NormalTok{religion}\SpecialCharTok{$}\NormalTok{wordClass }\OtherTok{\textless{}{-}} \FunctionTok{as.factor}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{wordClass)}
+\FunctionTok{levels}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{wordClass) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
+\FunctionTok{head}\NormalTok{(religion)  }\SpecialCharTok{\%\textgreater{}\%}  \FunctionTok{kable}\NormalTok{(}\AttributeTok{format =} \StringTok{"latex"}\NormalTok{,}\AttributeTok{booktabs=}\NormalTok{T,}
+                      \AttributeTok{linesep =} \StringTok{""}\NormalTok{,  }\AttributeTok{escape =} \ConstantTok{FALSE}\NormalTok{, }
+                      \AttributeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+                      \FunctionTok{kable\_styling}\NormalTok{(}\AttributeTok{latex\_options=}\FunctionTok{c}\NormalTok{(}\StringTok{"scale\_down"}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{table}
+
+\caption{\label{tab:religionTableHeadEarly}Head of the religion dataset.}
+\centering
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{lllrrl}
+\toprule
+protectedWord & wordToCompare & wordClass & cosineDistance & cosineSimilarity & connection\\
+\midrule
+judaism & violent & muslim & 0.7141939 & 0.2858061 & different\\
+judaism & terrorist & muslim & 0.7461333 & 0.2538667 & different\\
+judaism & dirty & muslim & 1.2002599 & -0.2002599 & different\\
+judaism & uneducated & muslim & 0.7885469 & 0.2114531 & different\\
+judaism & greedy & jewish & 1.0026172 & -0.0026172 & associated\\
+judaism & cheap & jewish & 1.2323229 & -0.2323229 & associated\\
+\bottomrule
+\end{tabular}}
+\end{table}
+\normalsize
+
+\pagebreak
+
+In the article there was no analysis of individual distances but the general look at the data through the usage of mean.
+The authors introduced a metric that tries to generalize the presence of bias through the classification of multi-class bias in groups of words connected with gender, religion or race.
+In the process they first take the mean of cosine distances between a given protected word and attributes assigned to each stereotype.
+They do not differentiate between stereotypes associated with a word and stereotypes associated with different words (in the case of religion, stereotypes characteristic for Christianity has also cosine distance measured with for instance, Judaism or Islam).
+Then, after collecting the list of mean cosine distances, they average the list to obtain one final value representing the whole group, in this example religion, for which the final mean of all mean distances is equal to 0.859.
+
+In the article the authors also try to remove previously defined biases from word embedding.
+First they identify the bias subspace using Principal Component Analysis (PCA) which is a technique for dimensionality reduction.
+It is applied here to choose the subspace that contains the greatest amount of information.
+There can be many subspaces found in a given group, for example in terms of religion one can identify at least a few sets that are to grasp the concept of religion in general:
+
+\textbf{["judaism", "christianity", "islam"]}
+
+\textbf{["jew", "christian", "muslim"]}
+
+\textbf{["synagogue", "church", "mosque"]}
+
+The idea is to find a set that provides enough information to create from it a vector representing the concept of religion among words.
+This strategy is based on the idea that different dimensions of vectors contain different types of information and in some words in vector layers (subspaces) the information about religiousness is implicitly conveyed.
+In some cases this knowledge is useful but in the case of harmful stereotypes one does not want to include the concept of religion in stereotypical words.
+
+After finding the bias subspace, they use it to modify the vector values individually so that their cosine distances towards certain words are changed.
+In the case of stereotypes the aim is to make the cosine distances larger so that the association between protected word and harmful stereotype is smaller.
+
+In the final step there is an evaluation of the results.
+The cosine distances are calculated again but this time using debiased vocabulary.
+After taking the mean of all distances one final value is obtained and then it is compared with the average value from the beginning.
+If the cosine distances are on average greater than before then it leads the authors to the conclusion that improvement has been achieved.
+As the cosine distance increases it is assumed that the association between protected and stereotypical words decreases.
+
 \hypertarget{limitations-of-the-approach}{%
 \section{Limitations of the approach}\label{limitations-of-the-approach}}
+
+\textbf{1. Selection of attributes}
+
+The attributes are taken from different sources, there is no principled justification for their choice. From our analysis it will become clear that
+the list is rather uneven.
+
+There is no mention of methodology for deciding on the number of attributes necessary to decide a hypothesis on the given size of dataset.
+There are however some ways to estimate how many samples we need to make sure that the result is significant.
+Our research will show that the numbers used
+are rather insufficient.
+
+\textbf{2. No control while taking the mean of cosines}
+
+The authors use the mean average cosine similarity to check on multi-class similarity between protected word and harmful stereotypes.
+They average the results until they obtain one final value to represent the mean cosine distance between protected word from a given class and the attributes of that class.
+
+\textbf{3. Hiding the impact of uncertainty}
+
+A mean hides issue and as there are pairs having negative and small similarities and there are those that have similarity equal to 0.5, the resulting calculation seems to be in norm.
+Additionally in such method the uncertainty is also not included which makes it even more difficult to give reasonable interpretations of the results.
+We propose the use of Bayesian method to obtain some understanding of the influence the uncertainty has on the interpretation of final results.
+
+\textbf{4. No word class distinction}
+
+In the original paper words from all three religions were checked with all of the stereotypes which means that there was no distinction between classes in which the stereotype is associated with a given religion, as opposed to the situation in which it is associated with another one.
+Not all of the stereotypical words have to be considered as harmful for all of the religions.
+In our analysis we distinguished between stereotypes associated with a given group, stereotype associated with different groups and control groups; neutral words and stereotypes-free human predicates.
+
+\textbf{5. Interpreting the results}
+
+Assuming for a moment that the value of multi-class cosine distance is correct, one may question the interpretation.
+\protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) summarize the averages of cosine distance per group (gender, race, religion).
+We would like to focus now on analyzing the values relating to religious biases.
+Here is the relevant fragment of table:
+
+\begin{longtable}[]{@{}ll@{}}
+\toprule
+Religion Debiasing & MAC \\
+\midrule
+\endhead
+Biased & 0.859 \\
+Hard Debiased & 0.934 \\
+Soft Debiased (\(\lambda\) = 0.2) & 0.894 \\
+\bottomrule
+\end{longtable}
+
+MAC stand for mean average cosine similarity, although in reality the values of cosine distance are given.
+What may attract attention is the fact that the value of cosine distance in ``Biased'' category is already quite high even before debiasing.
+High cosine distance indicates low cosine similarity between values.
+One could think that average cosine similarity equal to approximately 0.141 is not significant enough to consider it as bias.
+However the authors aim to mitigate ``biases'' in vectors with such great distance to make it somewhat larger.
+Methodologically the question is, on what basis is this small similarity still considered as a proof of a bias presence, and whether these small changes are meaningful.
+This is in general the problem of scale and the lack of universal intervals.
+Proper intervals could be used to decide whether given cosine similarity is high enough to consider the words to be more similar than if we choose them at random.
+
+\textbf{6. The curse of dimensionality}
+
+In our case, the curse of dimensionality may take place when there is an increase in volume of data that results in adding extra dimensions to the Euclidean space. According to the article
+\href{https://analyticsindiamag.com/curse-of-dimensionality-and-what-beginners-should-do-to-overcome-it/}{Analyticsindiamag Curse of dimensionality} as the number of features increases, it may be harder and harder to obtain useful information from the data with the usage of available algorithms.
+One may notice that more data should contribute to greater amount of information but more information also means greater risk of noise and distractions in data.
+At the same time, many times modern solutions are adapted to smaller dimensions and their results in higher ones are not intuitive or may be prone to be mistaken.
+
+Using cosine similarity in high dimensions in word embeddings may also be prone to the curse of dimensionality.
+According to \protect\hyperlink{ref-Venkat2018Curse}{Venkat} (\protect\hyperlink{ref-Venkat2018Curse}{2018}) there are reasons to consider this phenomenon when searching for word similarities in higher dimensions.
+An experiment is conducted that aims at showing how the similarity values and variation change as the number of dimensions increases.
+The hypothesis made in the paper states that two things will happen as the number of dimensions increase.
+First, the effort required to measure cosine similarity will be greater, and two, the similarity between data will blur out and have less variation.
+The authors generate random points with increasing number of dimensions where each dimension of a data point is given a value between 0 and 1.
+Then they pick one vector at random from each dimension class and calculate the cosine similarity between the chosen vector and the rest of the data.
+Then they check how the variation of values changes as the number of dimensions increases.
+It seems like the more dimensions there are, the smaller the variance and therefore it is less obvious how to interpret the resulting cosine similarities.
+Maybe the scale should be adjusted to the number of dimensions and variance so that it still gives us sensible information about data.
+According to some researches the cosine similarity in high dimensions is not reliable enough to trust it as it may be the case that choosing words on random may result in getting similar values as when picking them consciously.
+
+\begin{figure}
+\centering
+\includegraphics{../images/curseOfDimensionality.png}
+\caption{curse of dimensionality, number of dimensions on the x axis, standard deviation of similarity on the y axis}
+\end{figure}
 
@@
 
-It is worth diving deeper into some observations that arise while analyzing the results.
-As we pointed out, if there is no control of what cosine distance values individual word pairs have, then there is less
-understanding of the exact relations between words.
-If the protected words that we chose do not have high similarity with harmful stereotypes, then one should consider at least two scenarios.
-The first one is that the choice of protected words and attributes may be corrupted.
-The second one is that the metric is not able to catch the hidden bias properly.
-In both cases it is essential to take a look at the individual values before averaging them or aggregating in other ways.
-
-One has to also remember about the specific situation present in gender dataset.
-Surprisingly one could observe there high cosine distances values between some female stereotypical professions and male protected words.
-This leads to new ideas of how to understand the bias origins and its proper detection.
-To start with, cosine distance seems to catch the information regarding words co-occurrence and not the semantic similarity.
-If a word stereotypically associated with females has high cosine distance value for male protected words, then the metric is not proper to establish the bias presence as it can be misleading.
-What we mean is that in same cases it can unnecessarily label words as biased while omitting other more harmful associations.
-What is more, this highlights again the need for control group.
-One could check each stereotypical attribute with each protected word but not in order to average the values.
-One should rather plot the distributions and investigate how the metric measures vectors geometry and hidden semantic information.
-
-Debiasing with the method provided by \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) shows how it is unpredictable in terms of the results.
-In some cases the final cosine distance is indeed close to neutral words, but in other cases the change is minor.
-What is more, the inclusion of the uncertainty shows how in some cases it is not clear why a word is according to this method
-classified as anomalous.
-The fact that uncertainty for \texttt{assosiated} and \texttt{different} class is so high leads to the situation where in the boundaries of an uncertainty of some attributes, one can find the neutral words as well.
-It makes the usage and interpretation of this dataset and metric extremely complicated and unclear.
-
-Additionally, one cannot be sure if the bias is still preserved after the debiasing.
-The fact that all of the cosine distances for protected words and harmful attributes moved to the right, does not mean that the bias is removed.
-One could argue that maybe it is not clear how to measure bias presence and removal but the method for debiasing works fine.
-Notwithstanding, it is showed in researches such as \protect\hyperlink{ref-Gonen2019Lipstick}{Gonen \& Goldberg} (\protect\hyperlink{ref-Gonen2019Lipstick}{2019}), that the bias can hide in the vector geometry and preserve even after applying popular debiasing methods.
-Therefore it seems to not be justified to claim that the method works properly without precise methods to verify this.
-
-One should remember that there is no baseline for the interpretation of the averaged cosine distances in (\protect\hyperlink{ref-Manzini2019blackToCriminalToCriminal}{\textbf{Manzini2019blackToCriminalToCriminal?}}).
-One may assume that if the cosine distance is close to 1 then it is a desired outcome as it means, according to cosine distance assumptions, that there is almost no similarity between the words.
-However what does it mean to be close to 0?
-If the averaged cosine distance is equal to 80, then should we still debiase it?
-It is unclear what the criteria are.
-On one hand, it seems to be beneficial when the outcome is simplified as it is easier to compare results with one value per set.
-On the other hand, it is prone to misunderstanding of how to interpret the results and what threshold to assume.
-
-We propose the use of Bayesian method to introduce the uncertainty measure in bias detection.
-There are a few advantages when including this method in the bias analysis.
-As noted before in our example with religion dataset, one may obtain new insight into the bias issue after analyzing the mean estimates of the coeﬀicients.
-There seems to be a minor difference for the mean concerning the \texttt{assosiated} class.
-This can lead to the conclusions that we do not have strong reasons to perceive lower cosine distance as significantly related to the fact that the attribute is associated.
-
-One may dive deeper into this idea by investigating the connection coefficients for individual words with the use of the visualization.
-There one observe that indeed most of the words are clustered together independently of the class they belong to.
-It is of course partly due to the fact that we include uncertainty now.
-This example shows how understanding of the bias measurement may differ when switching to Bayesian method.
-One could argue that after this analysis one obtains more information on the dataset and can pursue with the further research more cautiously.
-
-\hypertarget{summary}{%
-\chapter{Summary}\label{summary}}
-
-The aim of this paper was to show the analysis of bias detection methods used in modern natural language processing projects.
-
-In the further research it is worth to dive deeper into some issues or to examine new ideas that occurred during our work.
+Placeholder
 
@@
 \begin{CSLReferences}{1}{0}
-\leavevmode\hypertarget{ref-Gonen2019Lipstick}{}%
-Gonen, H., \& Goldberg, Y.
-(2019).
-Lipstick on a pig: Debiasing methods cover up systematic gender biases in word embeddings but do not remove them.
-\emph{CoRR}, \emph{abs/1903.03862}.
-Retrieved from \url{http://arxiv.org/abs/1903.03862}
+\leavevmode\hypertarget{ref-Bolukbasi2016Man}{}%
+Bolukbasi, T., Chang, K.-W., Zou, J.
+Y., Saligrama, V., \& Kalai, A.
+(2016).
+Man is to computer programmer as woman is to homemaker?
+Debiasing word embeddings.
+\emph{CoRR}, \emph{abs/1607.06520}.
+Retrieved from \url{http://arxiv.org/abs/1607.06520}
 
@@
 
+\leavevmode\hypertarget{ref-Venkat2018Curse}{}%
+Venkat, N. (2018). The curse of dimensionality: Inside out.
+
 \end{CSLReferences}

### 2021-06-17 thesis/_book/_main.tex
small things

@@
             pdfauthor={Alicja Dobrzeniecka},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage[left=3.5cm, right=2.5cm, top=2.5cm, bottom=2.5cm, asymmetric, includeheadfoot]{geometry}
+            pdfborder={0 0 0},
+            breaklinks=true}
+\urlstyle{same}  % don't use monospace font for urls
+\usepackage[left=3.5cm, right=2.5cm, top=2.5cm, bottom=2.5cm, asymmetric,
+includeheadfoot]{geometry}
 \usepackage{color}
@@
 \newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\BuiltInTok}[1]{#1}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
 \newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
 \newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\ImportTok}[1]{#1}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
 \newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
+\newcommand{\BuiltInTok}[1]{#1}
+\newcommand{\ExtensionTok}[1]{#1}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
+\newcommand{\RegionMarkerTok}[1]{#1}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
 \newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
-\newcommand{\ExtensionTok}[1]{#1}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ImportTok}[1]{#1}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
 \newcommand{\NormalTok}[1]{#1}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\RegionMarkerTok}[1]{#1}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\usepackage{longtable,booktabs,array}
-\usepackage{calc} % for calculating minipage widths
-% Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-\makeatother
-% Allow footnotes in longtable head/foot
-\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
-\makesavenoteenv{longtable}
-\usepackage{graphicx}
+\usepackage{longtable,booktabs}
+% Fix footnotes in tables (requires footnote package)
+\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{long table}}{}
+\usepackage{graphicx,grffile}
 \makeatletter
@@
 \setcounter{secnumdepth}{5}
+% Redefines (sub)paragraphs to behave more like sections
+\ifx\paragraph\undefined\else
+\let\oldparagraph\paragraph
+\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+\let\oldsubparagraph\subparagraph
+\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
+
+% set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
+
 \usepackage{todonotes}
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{wrapfig}
-\usepackage{float}
-\usepackage{colortbl}
-\usepackage{pdflscape}
-\usepackage{tabu}
-\usepackage{threeparttable}
-\usepackage{threeparttablex}
-\usepackage[normalem]{ulem}
-\usepackage{makecell}
-\usepackage{xcolor}
-\ifluatex
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
-  % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
-\title{Conceptual and methodological problems with bias detection and avoidance in natural language processing}
+\title{Conceptual and methodological problems with bias detection and avoidance
+in natural language processing}
 \author{Alicja Dobrzeniecka}
-\date{2021-06-16}
+\date{2021-06-17}
 
@@
 }
-\setstretch{1.5}
-\hypertarget{introduction}{%
-\chapter{Introduction}\label{introduction}}
+\chapter{Introduction}\label{introduction}
 
-Placeholder
+Natural language processing (NLP) is a subfield of computer science that
+processes and analyzes language in text and speech with the use of
+modern programming methods. It has practical applications in everyday
+life as it concerns tasks such as email filters, smart assistants,
+search results, language translations, text analytics and so on. Models
+used to accomplish these tasks need a lot of data to learn from. This
+data originates from human activities and historical recordings such as
+texts, messages or speeches. It turns out that in the learning process
+these models can learn implicit biases that reflect harmful
+stereotypical thinking still present in modern societies. One can find
+methods that aim at identifying and measuring hidden biases and/or try
+to remove them by modifying the models. There are many different types
+of models in NLP depending on a task that they are supposed to solve.
+However, all of them need as an input words represented by means of
+numbers and this is accomplished with word embedding models. The models
+usually assign the values based on the context in which the words
+appear. It means that the input data can have enormous influence on the
+outcome. The biases seem to have their primary source in the way the
+words are assigned the numerical values.
 
-\hypertarget{cosine-similarity-and-bias-detection}{%
-\chapter{Cosine similarity and bias detection}\label{cosine-similarity-and-bias-detection}}
+There is considerable amount of literature available on the topic of
+bias detection and mitigation in NLP models. Bolukbasi, Chang, Zou,
+Saligrama, \& Kalai (\protect\hyperlink{ref-Bolukbasi2016Man}{2016})
+focus on gender biases that may be observable while investigating the
+representation of job occupations and gender in terms of their assigned
+numerical values. The authors apply cosine similarity measurement to
+investigate the phenomenon where (the vectors corresponding to) words
+related to jobs that are stereotypically associated with a given gender
+are in fact in the model situated closer to this gender. They also use
+analogy tasks to evaluate if the bias is present in the word embedding
+model. They check analogies by comparing pairs of word vectors, for
+example they search for the word complementing the puzzle: man is to
+doctor as woman is to \ldots{}? First they subtract the word ``man''
+from the word ``woman'' and then they search for the ranked list of
+other words pairs that have similar vector difference. They also include
+in the formula a threshold to ensure that the resulting pairs could not
+be randomly picked.
 
-\hypertarget{word-embeddings}{%
-\section{Word embeddings}\label{word-embeddings}}
+However, as Nissim, Noord, \& Goot
+(\protect\hyperlink{ref-Nissim2019Fair}{2019}) points out, there are
+some limitations to this approach. In practice, most of analogies
+implementations do not return any input words. This means that it does
+not make sense to expect the algorithm to return the same profession for
+both woman and man. Therefore, this method seems to be limited in terms
+of bias detection. Other problems are related for example to the choice
+of pairs and words that are used to detect the presence of
+discrimination as it is often subjective and without proper
+justification. Additionally the choice of the parameter used in
+Bolukbasi et al. (\protect\hyperlink{ref-Bolukbasi2016Man}{2016})
+formula to ensure that word pairs are not picked by random, is also not
+justified and changing it drastically influences the results.
 
-To understand what cosine similarity measurement is, one first needs to grasp the concept of translating words to a computer-readable form.
-In the field of natural language processsing there are two main types of words representation --- localist and distributed.
-One-hot encoding is an example of a method used to achieve a localist representation of words.
-Here each vector contains information only about a single data point, this is achieved by first mapping categorical values (words) to integers and then to each integers a binary vector is assigned which contains only 0s except for the index of the integer, which is assigned 1.
-An example of a localist representation is:
+Islam, Bryson, \& Narayanan
+(\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) touch upon the
+topic of biases regarding race and gender. They apply knowledge from
+well-known psychological studies such as Implicit Association Test to
+research the relation between human stereotypical thinking and model
+learnt biases to discover close relationship between these two. For the
+evaluation they use Word Embedding Association Test (WEAT) and the Word
+Embedding Factual Association Test (WEFAT).
+
+({\textbf{???}}) propose a novel way of using cosine similarity to
+obtain information on assumed resemblance between words. They
+investigate an approach that enables them to measure the bias for a
+class (like gender, religion, race) and express the final result with a
+single metric.
+
+Islam et al. (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) refer
+to Implicit Association Test (Greenwald et al., 1998) that measures the
+strength of associations between concepts or stereotypes by calculating
+the time of reaction for special tasks. Humans naturally exhibit some
+biases which do not always cause social concern. One can imagine the
+intuitive associations between for example insects and flowers, and the
+feelings of pleasantness or unpleasantness. In general, people would
+rather associate flowers with feeling pleasant than insects, and this
+preference in a sense is a bias or prejudice in some direction. However,
+this type of preference does not cause an uproar and is a rather morally
+neutral case. This example is rather used to show that the methodology
+used in IAT makes sense. Unfortunately, the further studies discover
+other biases and prejudices that directly influence the quality of other
+people's lives and therefore they should be taken care of.
+
+One can find various definitions trying to capture what bias and
+fairness actually are. With the choice of the definition, implications
+into the real-life applications may change as well. Mehrabi, Morstatter,
+Saxena, Lerman, \& Galstyan
+(\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) mark out that there
+exist different types of biases such as historical bias, representation
+bias or measurement bias (the list is long). This indicates how complex
+the issue of bias is. Without the proper understanding and awareness of
+the problem, people are prone to unconsciously sustain the bias.
+
+Mehrabi et al. (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) also
+distinguish different types of discrimination, some of them will be
+briefly described. By protected attributes we mean those qualities,
+traits or characteristics that one should not discriminate against.
+Direct discrimination occurs when protected attributes of individuals
+explicitly result in non-favorable outcomes toward them. Such
+discrimination takes place for example when one uses an information
+about someone's religion belief to evaluate whether they are competent
+for the work they apply for. In contrast in indirect discrimination
+individuals appear to be treated equally but anyway they end up being
+treated unjustly due to the hidden effects of biases towards their
+protected attributes. An example of such situation is when in a job
+advert there is a requirement of 10 years experience instead of the list
+of a specific type of experience and knowledge. This job advert can
+discriminate indirectly against young people who can have equivalent
+skills but not yet so much work experience. Systemic discrimination
+takes place when policies, customs or behaviors that result from certain
+culture or organizational structure lead to discrimination against some
+groups of people. Finally, statistical discrimination consists in to
+using average group statistics to judge person belonging to the group.
+
+The topic of discrimination is entangled with another concept; fairness.
+It is essential to grasp some concepts of fairness to take them into
+consideration while designing implementation of some machine learning
+model. In Mehrabi et al.
+(\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) one may notice that
+depending on the context and application different definitions may be
+applied.
+
+The most common methods in the literature focus on comparing the
+similarity between words from protected groups and those that are
+considered to be stereotypical or harmful in some way. One can find in
+this group methods such as euclidean distance or cosine similarity
+(which is equivalent to dot product if the vectors are normalized).
+There are also other ways to detect the effects of biases. One can
+investigate the model performance on certain tasks and validate if the
+model uses the gender or race information to complete the task.
+
+The currently used methods (such as cosine similarity) make the
+similarity values often aggregated in a way that may lead to hasty
+conclusions. The averaging of values and the lack of uncertainty may
+lead to the incomplete picture of the situation.
+
+In the paper we indicate how current methods used to detect biases in
+natural language models are limited, if we investigate them from the
+perspective of Bayesian analysis.
+
+Our research enhances the current way in which the bias detection is
+performed to make sure that it is methodologically valid.
+
+The key hypothesis is that greater understanding of data and bias
+implications can be achieved when Bayesian methods are applied to issue.
+
+\chapter{Cosine similarity and bias
+detection}\label{cosine-similarity-and-bias-detection}
+
+\section{Word embeddings}\label{word-embeddings}
+
+To understand what cosine similarity measurement is, one first needs to
+grasp the concept of translating words to a computer-readable form. In
+the field of natural language processsing there are two main types of
+words representation --- localist and distributed. One-hot encoding is
+an example of a method used to achieve a localist representation of
+words. Here each vector contains information only about a single data
+point, this is achieved by first mapping categorical values (words) to
+integers and then to each integers a binary vector is assigned which
+contains only 0s except for the index of the integer, which is assigned
+1. An example of a localist representation is:
 
@@
 \toprule
-word & 1 & 2 & 3 & 4 & 5 \\
+word & 1 & 2 & 3 & 4 & 5\tabularnewline
 \midrule
 \endhead
-woman & 1 & 0 & 0 & 0 & 0 \\
-man & 0 & 1 & 0 & 0 & 0 \\
-girl & 0 & 0 & 1 & 0 & 0 \\
-boy & 0 & 0 & 0 & 1 & 0 \\
-monarch & 0 & 0 & 0 & 0 & 1 \\
+woman & 1 & 0 & 0 & 0 & 0\tabularnewline
+man & 0 & 1 & 0 & 0 & 0\tabularnewline
+girl & 0 & 0 & 1 & 0 & 0\tabularnewline
+boy & 0 & 0 & 0 & 1 & 0\tabularnewline
+monarch & 0 & 0 & 0 & 0 & 1\tabularnewline
 \bottomrule
@@
 
-In the example above it is clear that the length of the vectors increases with the number of words in a vocabulary.
-It is not a very computationally efficient representation.
-It has other flaws as well.
-For example, it is unable to capture the resemblance between words appearing in similar contexts.
+In the example above it is clear that the length of the vectors
+increases with the number of words in a vocabulary. It is not a very
+computationally efficient representation. It has other flaws as well.
+For example, it is unable to capture the resemblance between words
+appearing in similar contexts.
 
-In contrast to the localist representation, a distributed representation returns vectors that contain continuous values instead of discrete 1s and 0s.
-Word embeddings are a class of various techniques that allow one to represent words as distributed vectors.
-Such learned representations of text have certain properties.
-At least prima facie, they store similar (or at least co-occurring) words close to each other in a vector space.
-An example of distributed representation is:
+In contrast to the localist representation, a distributed representation
+returns vectors that contain continuous values instead of discrete 1s
+and 0s. Word embeddings are a class of various techniques that allow one
+to represent words as distributed vectors. Such learned representations
+of text have certain properties. At least prima facie, they store
+similar (or at least co-occurring) words close to each other in a vector
+space. An example of distributed representation is:
 
@@
 \toprule
-word & 1 & 2 & 3 & 4 \\
+word & 1 & 2 & 3 & 4\tabularnewline
 \midrule
 \endhead
-woman & 0.456 & 0.267 & 0.675 & 0.131 \\
-man & 0.451 & 0.897 & 0.472 & 0.088 \\
-girl & 0.604 & 0.262 & 0.414 & 0.706 \\
-boy & 0.279 & 0.172 & 0.475 & 0.010 \\
-monarch & 0.565 & 0.678 & 0.463 & 0.975 \\
+woman & 0.456 & 0.267 & 0.675 & 0.131\tabularnewline
+man & 0.451 & 0.897 & 0.472 & 0.088\tabularnewline
+girl & 0.604 & 0.262 & 0.414 & 0.706\tabularnewline
+boy & 0.279 & 0.172 & 0.475 & 0.010\tabularnewline
+monarch & 0.565 & 0.678 & 0.463 & 0.975\tabularnewline
 \bottomrule
@@
 
-One of the advantages of using a distributed representation is that one is able to represent an enormous number of concepts with a smaller number of units.
-It is also possible to better capture similarities as words of similar meanings can have similar numeric vectors.
+One of the advantages of using a distributed representation is that one
+is able to represent an enormous number of concepts with a smaller
+number of units. It is also possible to better capture similarities as
+words of similar meanings can have similar numeric vectors.
 
-The numbers occurring in such representations are not random.
-They are learned in a process that uses a very shallow neural network.
-There are various types of techniques used for learning the vectors representations.
-One of the most straightforward ones is a skip-gram model.
-Given a word the model tries to predict its neighboring words from the sentence.
-The mathematics behind the process relies on the idea that the prediction concerns the conditional probability of the adjacent words.
-The algorithm tries to minimize the loss function, which penalizes the system for discrepancy with actual co-occurrence frequencies in the corpus.
-One can choose various parameters of the model, such as the window size that determines how many surrounding words the model should predict.
-After preparing such a fitted modeln=, one takes only the learned weights from a neural network, and uses them as vectors in a word embeddings representation.
+The numbers occurring in such representations are not random. They are
+learned in a process that uses a very shallow neural network. There are
+various types of techniques used for learning the vectors
+representations. One of the most straightforward ones is a skip-gram
+model. Given a word the model tries to predict its neighboring words
+from the sentence. The mathematics behind the process relies on the idea
+that the prediction concerns the conditional probability of the adjacent
+words. The algorithm tries to minimize the loss function, which
+penalizes the system for discrepancy with actual co-occurrence
+frequencies in the corpus. One can choose various parameters of the
+model, such as the window size that determines how many surrounding
+words the model should predict. After preparing such a fitted modeln=,
+one takes only the learned weights from a neural network, and uses them
+as vectors in a word embeddings representation.
 
 Word embeddings have many applications in natural language processing.
-They are handy in document search and information retrieval.
-They also play their part in improving automatic translations.
-Well learned word representations may also contribute to the improvement of sentiment analysis or spam detection.
+They are handy in document search and information retrieval. They also
+play their part in improving automatic translations. Well learned word
+representations may also contribute to the improvement of sentiment
+analysis or spam detection.
 
-\hypertarget{cosine-similarity-and-distance}{%
-\section{Cosine similarity and distance}\label{cosine-similarity-and-distance}}
+\section{Cosine similarity and
+distance}\label{cosine-similarity-and-distance}
 
-Cosine similarity is often used as a method of finding out whether vector representations for two words suggest that they are similar or somehow connected.
-Cosine similarity is the cosine of the angle between two vectors: the result of dividing their inner product (dot product usually) by the product of their magnitudes.
+Cosine similarity is often used as a method of finding out whether
+vector representations for two words suggest that they are similar or
+somehow connected. Cosine similarity is the cosine of the angle between
+two vectors: the result of dividing their inner product (dot product
+usually) by the product of their magnitudes.
 
 It is worth mentioning one more point concerning cosine similarity.
-After the vectors are normalized to have length equal to 1, inner product itself (often dot product) is used to measure the similarity.
+After the vectors are normalized to have length equal to 1, inner
+product itself (often dot product) is used to measure the similarity.
 
@@
 \end{align}
-Cosine similarity is considered a proper tool for this operation as its result has a clear connection to geometry and at least for a low number of dimensions may be easily interpreted.
-Using this scale, one can compare vector similarities in a fairly clear manner.
-When the vectors are aligned perpendicularly to each other, their similarity equals 0 (which is the same as the cosine of 90 degrees).
-This tells us that the similarity between the vectors is small.
-As the angle between vectors decreases, cosine similarity approaches one, which stands for the greatest similarity.
 
-One of the limitations of this measure is that it informs us only about similarities between vectors in terms of their orientation.
-However, it is often argued that in comparing words in terms of this metric, the magnitude of vectors may be treated as irrelevant, as the most important information pertains to direction.
+Cosine similarity is considered a proper tool for this operation as its
+result has a clear connection to geometry and at least for a low number
+of dimensions may be easily interpreted. Using this scale, one can
+compare vector similarities in a fairly clear manner. When the vectors
+are aligned perpendicularly to each other, their similarity equals 0
+(which is the same as the cosine of 90 degrees). This tells us that the
+similarity between the vectors is small. As the angle between vectors
+decreases, cosine similarity approaches one, which stands for the
+greatest similarity.
 
-In what follows, it is important to distinguish between cosine similarity and cosine distance, defined as:
+One of the limitations of this measure is that it informs us only about
+similarities between vectors in terms of their orientation. However, it
+is often argued that in comparing words in terms of this metric, the
+magnitude of vectors may be treated as irrelevant, as the most important
+information pertains to direction.
+
+In what follows, it is important to distinguish between cosine
+similarity and cosine distance, defined as:
+
 \begin{align} \tag{Sim}
@@
 
-The greater the similarity between two vectors, the smaller the distance between them.
-The cosine distance ranges between 0 and 2.
-If the vectors are in an opposite direction to each other, the cosine distance is 2.
-And if the vectors are extremely similar then the cosine distance is very close to 0.
+The greater the similarity between two vectors, the smaller the distance
+between them. The cosine distance ranges between 0 and 2. If the vectors
+are in an opposite direction to each other, the cosine distance is 2.
+And if the vectors are extremely similar then the cosine distance is
+very close to 0.
 
-One should note that cosine distance is not exactly a distance measure while it does not meet triangle inequality requirements.
-The triangle inequality formula says that for any triangle, the sum of the lengths of any two sides must be greater than or equal to the length of the remaining side.
-As shown in \href{https://stats.stackexchange.com/questions/198080/proving-that-cosine-distance-function-defined-by-cosine-similarity-between-two-u}{Statexchange.com} in the case of cosine distance it would have to fulfill this equation 1+cos-sim(𝐴,𝐶) \textless{} cos-sim(𝐴,𝐵) + cos-sim(𝐵,𝐶).
-If one chooses specific unit vectors it is easy to demonstrate that the triangle inequality is not preserved.
+One should note that cosine distance is not exactly a distance measure
+while it does not meet triangle inequality requirements. The triangle
+inequality formula says that for any triangle, the sum of the lengths of
+any two sides must be greater than or equal to the length of the
+remaining side. As shown in
+\href{https://stats.stackexchange.com/questions/198080/proving-that-cosine-distance-function-defined-by-cosine-similarity-between-two-u}{Statexchange.com}
+in the case of cosine distance it would have to fulfill this equation
+1+cos-sim(𝐴,𝐶) \textless{} cos-sim(𝐴,𝐵) + cos-sim(𝐵,𝐶). If one chooses
+specific unit vectors it is easy to demonstrate that the triangle
+inequality is not preserved.
 
-\hypertarget{cosine-distance-in-a-one-class-bias-detection}{%
-\section{Cosine distance in a one-class bias detection}\label{cosine-distance-in-a-one-class-bias-detection}}
+\section{Cosine distance in a one-class bias
+detection}\label{cosine-distance-in-a-one-class-bias-detection}
 
-\protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) define similarity between words as the outcome inner product of their normalized vectors.
-They focus on examining what the geometry of word embedding is in regard to ``he'' and ``she'' words.
-In other words, whether the similarity between those concepts and other words reflects expected gender stereotypes.
-They test this hypothesis by investigating whether there is a connection between word embeddings representing certain professions and words referring to gender.
-They also evaluate whether automatically produced analogies between words reflect the stereotypes as well.
+Bolukbasi et al. (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) define
+similarity between words as the outcome inner product of their
+normalized vectors. They focus on examining what the geometry of word
+embedding is in regard to ``he'' and ``she'' words. In other words,
+whether the similarity between those concepts and other words reflects
+expected gender stereotypes. They test this hypothesis by investigating
+whether there is a connection between word embeddings representing
+certain professions and words referring to gender. They also evaluate
+whether automatically produced analogies between words reflect the
+stereotypes as well.
 
-Here the gender bias of a word \(w\) is understood as its projection on the gender direction \(\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})\) (the gender direction is the top principal compontent of ten gender pair difference vectors).
-The underlying idea is that no bias is present if non-explicitly gender words are in equal distance to both elements in all explicitly gender pairs.
-Given the (ideally) gender neutral words \(N\) and the gender direction \(g\) the direct direct gender bias is defined as the average distance of the words in \(N\) from \(g\) (\(c\) is a parameter determining how strict we want to be):
+Here the gender bias of a word \(w\) is understood as its projection on
+the gender direction
+\(\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})\) (the
+gender direction is the top principal compontent of ten gender pair
+difference vectors). The underlying idea is that no bias is present if
+non-explicitly gender words are in equal distance to both elements in
+all explicitly gender pairs. Given the (ideally) gender neutral words
+\(N\) and the gender direction \(g\) the direct direct gender bias is
+defined as the average distance of the words in \(N\) from \(g\) (\(c\)
+is a parameter determining how strict we want to be):
+
 \begin{align}
@@
 
-A very vivid way to follow their method of arguing that bias in word embeddings is real is to plot the values of inner product of chosen words.
-The plot below does not originate from the original paper (it is from \url{https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings}) but similar visualization may be found there.
-Data used to create our plot is as follows.
+A very vivid way to follow their method of arguing that bias in word
+embeddings is real is to plot the values of inner product of chosen
+words. The plot below does not originate from the original paper (it is
+from
+\url{https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings}) but
+similar visualization may be found there. Data used to create our plot
+is as follows.
 
-Occupations associated with feminine: \textbf{"homemaker", "nurse", "receptionist", "librarian", "socialite", "hairdresser", "nanny", "bookkeeper", "stylist", "housekeeper", "interior designer", "guidance counselor"}
+Occupations associated with feminine:
+\textbf{"homemaker", "nurse", "receptionist", "librarian", "socialite", "hairdresser", "nanny", "bookkeeper", "stylist", "housekeeper", "interior designer", "guidance counselor"}
 
-Occupations associated with masculine: \textbf{"maestro", "skipper", "protege", "philosopher", "captain", "architect", "financier", "warrior", "broadcaster", "magician", "fighter pilot", "boss"}
+Occupations associated with masculine:
+\textbf{"maestro", "skipper", "protege", "philosopher", "captain", "architect", "financier", "warrior", "broadcaster", "magician", "fighter pilot", "boss"}
 
-\vspace{1mm}
-\footnotesize
+\vspace{1mm} \footnotesize
 
 \begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-1-1} \end{center}
+
 \normalsize
 
-The points in the plot above result from the calculation of the inner product of a chosen vector for a profession word and a vector for a gender word (she or he).
-Inner product of two vectors expresses similarity between words.
-This assumption originates from the geometry and properties of a vector space.
+The points in the plot above result from the calculation of the inner
+product of a chosen vector for a profession word and a vector for a
+gender word (she or he). Inner product of two vectors expresses
+similarity between words. This assumption originates from the geometry
+and properties of a vector space.
 
-\hypertarget{cosine-distance-in-a-multi-class-bias-detection}{%
-\section{Cosine distance in a multi-class bias detection}\label{cosine-distance-in-a-multi-class-bias-detection}}
+\section{Cosine distance in a multi-class bias
+detection}\label{cosine-distance-in-a-multi-class-bias-detection}
 
-\protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) present a different approach towards finding similarities between classes of words.
-The authors claim that texts available online are full of direct or indirect human stereotypes.
-As a result, word embeddings are prone to learn and maybe amplify those biases and propagate them further into AI models that are used for various applications.
-Cosine distance is used in the article as a measure to first argue for the existence of multi-class bias and then to show how through bias mitigation techniques the bias may be decreased.
+Manzini, Lim, Tsvetkov, \& Black
+(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) present a
+different approach towards finding similarities between classes of
+words. The authors claim that texts available online are full of direct
+or indirect human stereotypes. As a result, word embeddings are prone to
+learn and maybe amplify those biases and propagate them further into AI
+models that are used for various applications. Cosine distance is used
+in the article as a measure to first argue for the existence of
+multi-class bias and then to show how through bias mitigation techniques
+the bias may be decreased.
 
-They modify WEAT to a multi-class setting, introducing Mean Average Cosine similarity as a measure of bias (in fact, in the paper they report distances rather than similarities).
-Let \(T = \{t_1, \dots, t_k\}\) be a class of protected word embeddings, and let each \(A_j\in A\) be a set of attributes stereotypically associated with a protected word).
-Then:
+They modify WEAT to a multi-class setting, introducing Mean Average
+Cosine similarity as a measure of bias (in fact, in the paper they
+report distances rather than similarities). Let
+\(T = \{t_1, \dots, t_k\}\) be a class of protected word embeddings, and
+let each \(A_j\in A\) be a set of attributes stereotypically associated
+with a protected word). Then:
+
 \begin{align}
@@
 
-As the code has been provided by the authors we were able to reconstruct their results.
-The main steps in the procedure are as follows.
-Let us go through an example that refers to the process of hard debiasing on religious attributes.
+As the code has been provided by the authors we were able to reconstruct
+their results. The main steps in the procedure are as follows. Let us go
+through an example that refers to the process of hard debiasing on
+religious attributes.
 
-First we load word embeddings from reddit.US.txt.tok.clean.cleanedforw2v.w2v dataset.
-The word embeddings have only 50 dimensions and the number of individual words from the dataset is 44895.
-The authors assume that the protected group should ideally not have high cosine similarity to stereotypical words.
-The word embeddings geometry should not place this group close to harmful stereotypes, if it is to be bias-free.
-For instance, let's look at the religion-related words.
+First we load word embeddings from
+reddit.US.txt.tok.clean.cleanedforw2v.w2v dataset. The word embeddings
+have only 50 dimensions and the number of individual words from the
+dataset is 44895. The authors assume that the protected group should
+ideally not have high cosine similarity to stereotypical words. The word
+embeddings geometry should not place this group close to harmful
+stereotypes, if it is to be bias-free. For instance, let's look at the
+religion-related words.
 
@@
 
-We have prepared a table presenting the values of cosine distance for each protected word with each attribute (stereotype). The part of the results is shown below.
+We have prepared a table presenting the values of cosine distance for
+each protected word with each attribute (stereotype). The part of the
+results is shown below.
 
@@
 \begin{Highlighting}[]
-\NormalTok{religion }\OtherTok{\textless{}{-}} \FunctionTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{]}
-\FunctionTok{colnames}\NormalTok{(religion) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
+\NormalTok{religion <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
+\KeywordTok{colnames}\NormalTok{(religion) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
                         \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
-\NormalTok{religion}\SpecialCharTok{$}\NormalTok{wordClass }\OtherTok{\textless{}{-}} \FunctionTok{as.factor}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{wordClass)}
-\FunctionTok{levels}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{wordClass) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
-\FunctionTok{head}\NormalTok{(religion)  }\SpecialCharTok{\%\textgreater{}\%}  \FunctionTok{kable}\NormalTok{(}\AttributeTok{format =} \StringTok{"latex"}\NormalTok{,}\AttributeTok{booktabs=}\NormalTok{T,}
-                      \AttributeTok{linesep =} \StringTok{""}\NormalTok{,  }\AttributeTok{escape =} \ConstantTok{FALSE}\NormalTok{, }
-                      \AttributeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
-                      \FunctionTok{kable\_styling}\NormalTok{(}\AttributeTok{latex\_options=}\FunctionTok{c}\NormalTok{(}\StringTok{"scale\_down"}\NormalTok{))}
+\NormalTok{religion}\OperatorTok{$}\NormalTok{wordClass <-}\StringTok{ }\KeywordTok{as.factor}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass)}
+\KeywordTok{levels}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
+\KeywordTok{head}\NormalTok{(religion)  }\OperatorTok{%>%}\StringTok{  }\KeywordTok{kable}\NormalTok{(}\DataTypeTok{format =} \StringTok{"latex"}\NormalTok{,}\DataTypeTok{booktabs=}\NormalTok{T,}
+                      \DataTypeTok{linesep =} \StringTok{""}\NormalTok{,  }\DataTypeTok{escape =} \OtherTok{FALSE}\NormalTok{, }
+                      \DataTypeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\OperatorTok{%>%}
+\StringTok{                      }\KeywordTok{kable_styling}\NormalTok{(}\DataTypeTok{latex_options=}\KeywordTok{c}\NormalTok{(}\StringTok{"scale_down"}\NormalTok{))}
 \end{Highlighting}
@@
 \end{table}
+
 \normalsize
@@
 
-In the article there was no analysis of individual distances but the general look at the data through the usage of mean.
-The authors introduced a metric that tries to generalize the presence of bias through the classification of multi-class bias in groups of words connected with gender, religion or race.
-In the process they first take the mean of cosine distances between a given protected word and attributes assigned to each stereotype.
-They do not differentiate between stereotypes associated with a word and stereotypes associated with different words (in the case of religion, stereotypes characteristic for Christianity has also cosine distance measured with for instance, Judaism or Islam).
-Then, after collecting the list of mean cosine distances, they average the list to obtain one final value representing the whole group, in this example religion, for which the final mean of all mean distances is equal to 0.859.
+In the article there was no analysis of individual distances but the
+general look at the data through the usage of mean. The authors
+introduced a metric that tries to generalize the presence of bias
+through the classification of multi-class bias in groups of words
+connected with gender, religion or race. In the process they first take
+the mean of cosine distances between a given protected word and
+attributes assigned to each stereotype. They do not differentiate
+between stereotypes associated with a word and stereotypes associated
+with different words (in the case of religion, stereotypes
+characteristic for Christianity has also cosine distance measured with
+for instance, Judaism or Islam). Then, after collecting the list of mean
+cosine distances, they average the list to obtain one final value
+representing the whole group, in this example religion, for which the
+final mean of all mean distances is equal to 0.859.
 
-In the article the authors also try to remove previously defined biases from word embedding.
-First they identify the bias subspace using Principal Component Analysis (PCA) which is a technique for dimensionality reduction.
-It is applied here to choose the subspace that contains the greatest amount of information.
-There can be many subspaces found in a given group, for example in terms of religion one can identify at least a few sets that are to grasp the concept of religion in general:
+In the article the authors also try to remove previously defined biases
+from word embedding. First they identify the bias subspace using
+Principal Component Analysis (PCA) which is a technique for
+dimensionality reduction. It is applied here to choose the subspace that
+contains the greatest amount of information. There can be many subspaces
+found in a given group, for example in terms of religion one can
+identify at least a few sets that are to grasp the concept of religion
+in general:
 
@@
 
-The idea is to find a set that provides enough information to create from it a vector representing the concept of religion among words.
-This strategy is based on the idea that different dimensions of vectors contain different types of information and in some words in vector layers (subspaces) the information about religiousness is implicitly conveyed.
-In some cases this knowledge is useful but in the case of harmful stereotypes one does not want to include the concept of religion in stereotypical words.
+The idea is to find a set that provides enough information to create
+from it a vector representing the concept of religion among words. This
+strategy is based on the idea that different dimensions of vectors
+contain different types of information and in some words in vector
+layers (subspaces) the information about religiousness is implicitly
+conveyed. In some cases this knowledge is useful but in the case of
+harmful stereotypes one does not want to include the concept of religion
+in stereotypical words.
 
-After finding the bias subspace, they use it to modify the vector values individually so that their cosine distances towards certain words are changed.
-In the case of stereotypes the aim is to make the cosine distances larger so that the association between protected word and harmful stereotype is smaller.
+After finding the bias subspace, they use it to modify the vector values
+individually so that their cosine distances towards certain words are
+changed. In the case of stereotypes the aim is to make the cosine
+distances larger so that the association between protected word and
+harmful stereotype is smaller.
 
-In the final step there is an evaluation of the results.
-The cosine distances are calculated again but this time using debiased vocabulary.
-After taking the mean of all distances one final value is obtained and then it is compared with the average value from the beginning.
-If the cosine distances are on average greater than before then it leads the authors to the conclusion that improvement has been achieved.
-As the cosine distance increases it is assumed that the association between protected and stereotypical words decreases.
+In the final step there is an evaluation of the results. The cosine
+distances are calculated again but this time using debiased vocabulary.
+After taking the mean of all distances one final value is obtained and
+then it is compared with the average value from the beginning. If the
+cosine distances are on average greater than before then it leads the
+authors to the conclusion that improvement has been achieved. As the
+cosine distance increases it is assumed that the association between
+protected and stereotypical words decreases.
 
-\hypertarget{limitations-of-the-approach}{%
-\section{Limitations of the approach}\label{limitations-of-the-approach}}
+\section{Limitations of the approach}\label{limitations-of-the-approach}
 
@@
 
-The attributes are taken from different sources, there is no principled justification for their choice. From our analysis it will become clear that
-the list is rather uneven.
+The attributes are taken from different sources, there is no principled
+justification for their choice. From our analysis it will become clear
+that the list is rather uneven.
 
-There is no mention of methodology for deciding on the number of attributes necessary to decide a hypothesis on the given size of dataset.
-There are however some ways to estimate how many samples we need to make sure that the result is significant.
-Our research will show that the numbers used
-are rather insufficient.
+There is no mention of methodology for deciding on the number of
+attributes necessary to decide a hypothesis on the given size of
+dataset. There are however some ways to estimate how many samples we
+need to make sure that the result is significant. Our research will show
+that the numbers used are rather insufficient.
 
@@
 
-The authors use the mean average cosine similarity to check on multi-class similarity between protected word and harmful stereotypes.
-They average the results until they obtain one final value to represent the mean cosine distance between protected word from a given class and the attributes of that class.
+The authors use the mean average cosine similarity to check on
+multi-class similarity between protected word and harmful stereotypes.
+They average the results until they obtain one final value to represent
+the mean cosine distance between protected word from a given class and
+the attributes of that class.
 
@@
 
-A mean hides issue and as there are pairs having negative and small similarities and there are those that have similarity equal to 0.5, the resulting calculation seems to be in norm.
-Additionally in such method the uncertainty is also not included which makes it even more difficult to give reasonable interpretations of the results.
-We propose the use of Bayesian method to obtain some understanding of the influence the uncertainty has on the interpretation of final results.
+A mean hides issue and as there are pairs having negative and small
+similarities and there are those that have similarity equal to 0.5, the
+resulting calculation seems to be in norm. Additionally in such method
+the uncertainty is also not included which makes it even more difficult
+to give reasonable interpretations of the results. We propose the use of
+Bayesian method to obtain some understanding of the influence the
+uncertainty has on the interpretation of final results.
 
@@
 
-In the original paper words from all three religions were checked with all of the stereotypes which means that there was no distinction between classes in which the stereotype is associated with a given religion, as opposed to the situation in which it is associated with another one.
-Not all of the stereotypical words have to be considered as harmful for all of the religions.
-In our analysis we distinguished between stereotypes associated with a given group, stereotype associated with different groups and control groups; neutral words and stereotypes-free human predicates.
+In the original paper words from all three religions were checked with
+all of the stereotypes which means that there was no distinction between
+classes in which the stereotype is associated with a given religion, as
+opposed to the situation in which it is associated with another one. Not
+all of the stereotypical words have to be considered as harmful for all
+of the religions. In our analysis we distinguished between stereotypes
+associated with a given group, stereotype associated with different
+groups and control groups; neutral words and stereotypes-free human
+predicates.
 
@@
 
-Assuming for a moment that the value of multi-class cosine distance is correct, one may question the interpretation.
-\protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) summarize the averages of cosine distance per group (gender, race, religion).
-We would like to focus now on analyzing the values relating to religious biases.
+Assuming for a moment that the value of multi-class cosine distance is
+correct, one may question the interpretation. Manzini et al.
+(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) summarize the
+averages of cosine distance per group (gender, race, religion). We would
+like to focus now on analyzing the values relating to religious biases.
 Here is the relevant fragment of table:
@@
 \toprule
-Religion Debiasing & MAC \\
+Religion Debiasing & MAC\tabularnewline
 \midrule
 \endhead
-Biased & 0.859 \\
-Hard Debiased & 0.934 \\
-Soft Debiased (\(\lambda\) = 0.2) & 0.894 \\
+Biased & 0.859\tabularnewline
+Hard Debiased & 0.934\tabularnewline
+Soft Debiased (\(\lambda\) = 0.2) & 0.894\tabularnewline
 \bottomrule
@@
 
-MAC stand for mean average cosine similarity, although in reality the values of cosine distance are given.
-What may attract attention is the fact that the value of cosine distance in ``Biased'' category is already quite high even before debiasing.
-High cosine distance indicates low cosine similarity between values.
-One could think that average cosine similarity equal to approximately 0.141 is not significant enough to consider it as bias.
-However the authors aim to mitigate ``biases'' in vectors with such great distance to make it somewhat larger.
-Methodologically the question is, on what basis is this small similarity still considered as a proof of a bias presence, and whether these small changes are meaningful.
-This is in general the problem of scale and the lack of universal intervals.
-Proper intervals could be used to decide whether given cosine similarity is high enough to consider the words to be more similar than if we choose them at random.
+MAC stand for mean average cosine similarity, although in reality the
+values of cosine distance are given. What may attract attention is the
+fact that the value of cosine distance in ``Biased'' category is already
+quite high even before debiasing. High cosine distance indicates low
+cosine similarity between values. One could think that average cosine
+similarity equal to approximately 0.141 is not significant enough to
+consider it as bias. However the authors aim to mitigate ``biases'' in
+vectors with such great distance to make it somewhat larger.
+Methodologically the question is, on what basis is this small similarity
+still considered as a proof of a bias presence, and whether these small
+changes are meaningful. This is in general the problem of scale and the
+lack of universal intervals. Proper intervals could be used to decide
+whether given cosine similarity is high enough to consider the words to
+be more similar than if we choose them at random.
 
@@
 
-In our case, the curse of dimensionality may take place when there is an increase in volume of data that results in adding extra dimensions to the Euclidean space. According to the article
-\href{https://analyticsindiamag.com/curse-of-dimensionality-and-what-beginners-should-do-to-overcome-it/}{Analyticsindiamag Curse of dimensionality} as the number of features increases, it may be harder and harder to obtain useful information from the data with the usage of available algorithms.
-One may notice that more data should contribute to greater amount of information but more information also means greater risk of noise and distractions in data.
-At the same time, many times modern solutions are adapted to smaller dimensions and their results in higher ones are not intuitive or may be prone to be mistaken.
+In our case, the curse of dimensionality may take place when there is an
+increase in volume of data that results in adding extra dimensions to
+the Euclidean space. According to the article
+\href{https://analyticsindiamag.com/curse-of-dimensionality-and-what-beginners-should-do-to-overcome-it/}{Analyticsindiamag
+Curse of dimensionality} as the number of features increases, it may be
+harder and harder to obtain useful information from the data with the
+usage of available algorithms. One may notice that more data should
+contribute to greater amount of information but more information also
+means greater risk of noise and distractions in data. At the same time,
+many times modern solutions are adapted to smaller dimensions and their
+results in higher ones are not intuitive or may be prone to be mistaken.
 
-Using cosine similarity in high dimensions in word embeddings may also be prone to the curse of dimensionality.
-According to \protect\hyperlink{ref-Venkat2018Curse}{Venkat} (\protect\hyperlink{ref-Venkat2018Curse}{2018}) there are reasons to consider this phenomenon when searching for word similarities in higher dimensions.
-An experiment is conducted that aims at showing how the similarity values and variation change as the number of dimensions increases.
-The hypothesis made in the paper states that two things will happen as the number of dimensions increase.
-First, the effort required to measure cosine similarity will be greater, and two, the similarity between data will blur out and have less variation.
-The authors generate random points with increasing number of dimensions where each dimension of a data point is given a value between 0 and 1.
-Then they pick one vector at random from each dimension class and calculate the cosine similarity between the chosen vector and the rest of the data.
-Then they check how the variation of values changes as the number of dimensions increases.
-It seems like the more dimensions there are, the smaller the variance and therefore it is less obvious how to interpret the resulting cosine similarities.
-Maybe the scale should be adjusted to the number of dimensions and variance so that it still gives us sensible information about data.
-According to some researches the cosine similarity in high dimensions is not reliable enough to trust it as it may be the case that choosing words on random may result in getting similar values as when picking them consciously.
+Using cosine similarity in high dimensions in word embeddings may also
+be prone to the curse of dimensionality. According to Venkat
+(\protect\hyperlink{ref-Venkat2018Curse}{2018}) there are reasons to
+consider this phenomenon when searching for word similarities in higher
+dimensions. An experiment is conducted that aims at showing how the
+similarity values and variation change as the number of dimensions
+increases. The hypothesis made in the paper states that two things will
+happen as the number of dimensions increase. First, the effort required
+to measure cosine similarity will be greater, and two, the similarity
+between data will blur out and have less variation. The authors generate
+random points with increasing number of dimensions where each dimension
+of a data point is given a value between 0 and 1. Then they pick one
+vector at random from each dimension class and calculate the cosine
+similarity between the chosen vector and the rest of the data. Then they
+check how the variation of values changes as the number of dimensions
+increases. It seems like the more dimensions there are, the smaller the
+variance and therefore it is less obvious how to interpret the resulting
+cosine similarities. Maybe the scale should be adjusted to the number of
+dimensions and variance so that it still gives us sensible information
+about data. According to some researches the cosine similarity in high
+dimensions is not reliable enough to trust it as it may be the case that
+choosing words on random may result in getting similar values as when
+picking them consciously.
 
@@
 \includegraphics{../images/curseOfDimensionality.png}
-\caption{curse of dimensionality, number of dimensions on the x axis, standard deviation of similarity on the y axis}
+\caption{curse of dimensionality, number of dimensions on the x axis,
+standard deviation of similarity on the y axis}
 \end{figure}
 
-\hypertarget{walkthrough-with-the-religion-dataset}{%
-\chapter{Walkthrough with the religion dataset}\label{walkthrough-with-the-religion-dataset}}
+\chapter{Walkthrough with the religion
+dataset}\label{walkthrough-with-the-religion-dataset}
 
-Placeholder
+\section{Loading and understanding the
+dataset}\label{loading-and-understanding-the-dataset}
 
-\hypertarget{loading-and-understanding-the-dataset}{%
-\section{Loading and understanding the dataset}\label{loading-and-understanding-the-dataset}}
+We start with loading the libraries needed for the analysis.
 
-\hypertarget{first-look-at-the-empirical-distributions}{%
-\section{First look at the empirical distributions}\label{first-look-at-the-empirical-distributions}}
+\footnotesize
 
-\hypertarget{looking-at-the-islam-related-words}{%
-\section{Looking at the islam-related words}\label{looking-at-the-islam-related-words}}
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(ggplot2)}
+\KeywordTok{library}\NormalTok{(ggthemes)}
+\KeywordTok{library}\NormalTok{(rethinking)}
+\KeywordTok{library}\NormalTok{(tidyverse)}
+\KeywordTok{library}\NormalTok{(ggpubr)}
+\KeywordTok{library}\NormalTok{(kableExtra)}
+\KeywordTok{library}\NormalTok{(dplyr)}
+\KeywordTok{library}\NormalTok{(ggExtra)}
+\KeywordTok{library}\NormalTok{(cowplot)}
+\end{Highlighting}
+\end{Shaded}
 
-\hypertarget{bayesian-model-structure-and-assumptions}{%
-\section{Bayesian model structure and assumptions}\label{bayesian-model-structure-and-assumptions}}
+\normalsize
 
-\hypertarget{choosing-predictors}{%
-\section{Choosing predictors}\label{choosing-predictors}}
+We will use the choice of protected words and stereotypical predicates
+used in REF. This is a decent point of departure, not only we want to
+compare our method to that of REF, but also because this data format is
+fairly general (as contrasted, say, with a set up for binary
+stereotypes). Note also that the method we develop here can fairly
+easily be run for different stereotypization patterns. Let's start with
+explaining the method and its deployment using a dataset obtained for
+the religion-related protected words.
 
-\hypertarget{dataset-level-coefficients}{%
-\section{Dataset-level coefficients}\label{dataset-level-coefficients}}
+Let's load, clean a bit and inspect the head of the religion dataset we
+prepared. In order to obtain this dataset, we calculated the cosine
+distance between each protected word and each word from both the
+bias-related attribute groups, which were used in the original study,
+and to neutral and human control attributes which we added as control
+groups. For instance, for religion, the bias-related predicates (coming
+from the original study in REF) include muslim bias attributes, jew bias
+attributes, christian bias attributes (see a list in the APPENDIX).
 
-\hypertarget{model-structure-and-assumptions}{%
-\section{Model structure and assumptions}\label{model-structure-and-assumptions}}
+We decided to add control groups in the form of two classes --- neutral
+words and human-related words. Without a proper control group it is
+quite hard to compare the resulting cosine distances and decide on their
+significance in bias detection. We prepared approximately 230 neutral
+words to double-check the prima-facie neutral hypothesis that their
+cosine similarity to the protected words will oscillate around 0 (that
+is, the distances will be around 1). This provides us with a more
+reliable point of reference. Moreover, we added human attributes that
+are associated with people in general to investigate whether the smaller
+cosine distance between protected words and stereotypes can result
+simply from the fact that the stereotype predicates are associated with
+humans. For two control groups, we have randomly drawn 230 words that do
+not express any property usually attributed to humans, and human related
+attributes.
 
-\hypertarget{protected-classes-in-reddit-and-google-embeddings}{%
-\section{Protected classes in Reddit and Google embeddings}\label{protected-classes-in-reddit-and-google-embeddings}}
+\vspace{1mm} \footnotesize
 
-\hypertarget{dataset-level-coefficients-after-debiasing}{%
-\section{Dataset-level coefficients after debiasing}\label{dataset-level-coefficients-after-debiasing}}
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{religion <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
+\KeywordTok{colnames}\NormalTok{(religion) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
+                        \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
+\CommentTok{#head(religion)}
+\CommentTok{#library(plyr)}
+\NormalTok{religion}\OperatorTok{$}\NormalTok{wordClass <-}\StringTok{ }\KeywordTok{as.factor}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass)}
+\CommentTok{#levels(religion$wordClass)}
+\CommentTok{#religion$wordClass <- mapvalues(religion$wordClass, }
+\CommentTok{#                          from = levels(religion$wordClass),}
+\CommentTok{#                          to = c("christian","human","jewish","muslim","neutral"))}
+\KeywordTok{levels}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
+\KeywordTok{head}\NormalTok{(religion)  }\OperatorTok{%>%}\StringTok{  }\KeywordTok{kable}\NormalTok{(}\DataTypeTok{format =} \StringTok{"latex"}\NormalTok{,}\DataTypeTok{booktabs=}\NormalTok{T,}
+                      \DataTypeTok{linesep =} \StringTok{""}\NormalTok{,  }\DataTypeTok{escape =} \OtherTok{FALSE}\NormalTok{, }
+                      \DataTypeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\OperatorTok{%>%}\StringTok{      }\KeywordTok{kable_styling}\NormalTok{(}\DataTypeTok{latex_options=}\KeywordTok{c}\NormalTok{(}\StringTok{"scale_down"}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
 
-\hypertarget{protected-classes-after-debiasing}{%
-\section{Protected classes after debiasing}\label{protected-classes-after-debiasing}}
+\begin{table}
 
-\hypertarget{discussion}{%
-\chapter{Discussion}\label{discussion}}
+\caption{\label{tab:religionTableHead}Head of the religion dataset.}
+\centering
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{lllrrl}
+\toprule
+protectedWord & wordToCompare & wordClass & cosineDistance & cosineSimilarity & connection\\
+\midrule
+judaism & violent & muslim & 0.7141939 & 0.2858061 & different\\
+judaism & terrorist & muslim & 0.7461333 & 0.2538667 & different\\
+judaism & dirty & muslim & 1.2002599 & -0.2002599 & different\\
+judaism & uneducated & muslim & 0.7885469 & 0.2114531 & different\\
+judaism & greedy & jewish & 1.0026172 & -0.0026172 & associated\\
+judaism & cheap & jewish & 1.2323229 & -0.2323229 & associated\\
+\bottomrule
+\end{tabular}}
+\end{table}
 
-Placeholder
+\normalsize
+
+The \texttt{protectedWord} column contains words from a protected class
+that (in a perfect world according to the assumptions of the orignal
+study) should not be associated with harmful stereotypes.
+\texttt{wordToCompare} contains attributes, including stereotypes and
+control group words. For each row we compute the cosine distances
+between a given protected word and a given attribute word.
+\texttt{wordClass} tells us which class an attribute is supposed to be
+stereotypically associated with, that is, whether the word from
+\texttt{wordToCompare} is associated stereotypically with jews,
+christians or muslims, or whether it belongs to a control group.
+\texttt{cosineDistance} is simply a calculation of the cosine distance
+between protected word and atrribute. \texttt{cosineSimilarity} contains
+the result of substracting cosine distance from 1. \texttt{connection}
+contains information about the relation type between a protected word
+and an attribute. If the attribute is e.g.~a harmful jewish stereotype
+and the protected word is also from the judaism group, the connection
+has value \texttt{associated}. If the attribute is still stereotypically
+jewish, but the protected word comes from another religion, the
+connection is labelled as \texttt{different}. If the attribute belongs
+to a neutral group then the connection is labelled as \texttt{none} and
+if an attribute belongs to the \texttt{human} class, then the connection
+is labelled as \texttt{human}.
+
+\section{First look at the empirical
+distributions}\label{first-look-at-the-empirical-distributions}
+
+First let's take a look at the empirical distribution of distances by
+the connection type, initially ignoring the human control class for now.
+
+\vspace{1mm} \footnotesize
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-3-1} \end{center}
+
+\normalsize
+
+The first impression is that while there is a shift for associated words
+towards smaller cosine distances as compared to the neutral words,
+slightly surprisingly a slightly weaker shift in the same direction is
+visible for attributes associated with different stereotypes. Moreover,
+the empirical distributions overlap to alarge extent and the means
+grouped by connection type do not seem too far from each other. In fact,
+as there is a lot of variety in the cosine distances (as we will soon
+see), we need to gauge the uncertainty involved, and to look more
+carefully at individual protected words to get a better idea of how the
+cosine distance distribution changes for different attribute groups and
+different protected classes. Now, let's add the human attributes to the
+picture:
+
+\vspace{1mm} \footnotesize
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-4-1} \end{center}
+
+\normalsize
+
+\noindent Notice that the distribution for \texttt{human} (even though
+we did our best not to include in it any stereotype-related atributes)
+is left-skewed, with much overlap with \texttt{associated} and
+\texttt{different}, which illustrates the need to take being associated
+with humans as an important predictor.
+
+Our focus lies in \texttt{connection} as a predictor. Morever, later on
+we'll be interested in looking at the protected words separately, and at
+protected words split by connection. For technical reasons it is useful
+to represent these factors as integer vectors.
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{religion}\OperatorTok{$}\NormalTok{con <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(religion}\OperatorTok{$}\NormalTok{connection)}
+\NormalTok{religion}\OperatorTok{$}\NormalTok{pw <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(religion}\OperatorTok{$}\NormalTok{protectedWord)}
+\NormalTok{religion}\OperatorTok{$}\NormalTok{pwFactor <-}\StringTok{ }\KeywordTok{factor}\NormalTok{(}\KeywordTok{paste0}\NormalTok{(religion}\OperatorTok{$}\NormalTok{protectedWord, religion}\OperatorTok{$}\NormalTok{connection))}
+\NormalTok{religion}\OperatorTok{$}\NormalTok{pwIndex <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(religion}\OperatorTok{$}\NormalTok{pwFactor)}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+A short script, \texttt{cleanDataset} to make this faster, so
+equivalently:
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{source}\NormalTok{(}\StringTok{"../functions/cleanDataset.R"}\NormalTok{)}
+\NormalTok{religion <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
+\NormalTok{religion <-}\StringTok{ }\KeywordTok{cleanDataset}\NormalTok{(religion,}\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+\section{Looking at the islam-related
+words}\label{looking-at-the-islam-related-words}
+
+For now, let's focus on five protected words related to islam (``imam'',
+``islam'', ``mosque'', ``muslim'', and ``quran''). The word list
+associates with islam four stereotypical attributes (``violent'',
+``terrorist'', ``uneducated'' and ``dirty''). First, we select and plot
+the empirical distributions for these protected words.
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(tidyverse)}
+\NormalTok{muslimWords <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"imam"}\NormalTok{,}\StringTok{"islam"}\NormalTok{,}\StringTok{"mosque"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"quran"}\NormalTok{)}
+\NormalTok{muslim <-}\StringTok{ }\NormalTok{religion }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(protectedWord }\OperatorTok{%in%}\StringTok{ }\NormalTok{muslimWords)}
+\KeywordTok{ggplot}\NormalTok{(muslim, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{  cosineDistance, }\DataTypeTok{fill =}\NormalTok{ connection, }\DataTypeTok{color =}\NormalTok{ connection))}\OperatorTok{+}
+\StringTok{  }\KeywordTok{geom_density}\NormalTok{(}\DataTypeTok{alpha=}\FloatTok{0.6}\NormalTok{,}\DataTypeTok{size =} \FloatTok{.2}\NormalTok{)}\OperatorTok{+}\StringTok{ }
+\StringTok{  }\KeywordTok{scale_fill_manual}\NormalTok{(}\DataTypeTok{values =} \KeywordTok{c}\NormalTok{(}\StringTok{"orangered4"}\NormalTok{,}\StringTok{"chartreuse4"}\NormalTok{, }\StringTok{"skyblue"}\NormalTok{, }\StringTok{"gray"}\NormalTok{))}\OperatorTok{+}
+\StringTok{  }\KeywordTok{scale_x_continuous}\NormalTok{(}\DataTypeTok{breaks =} \KeywordTok{seq}\NormalTok{(}\FloatTok{0.3}\NormalTok{,}\FloatTok{1.5}\NormalTok{, }\DataTypeTok{by =} \FloatTok{0.1}\NormalTok{))}\OperatorTok{+}\KeywordTok{xlab}\NormalTok{(}\StringTok{"cosine distance"}\NormalTok{)}\OperatorTok{+}
+\StringTok{  }\KeywordTok{scale_color_manual}\NormalTok{(}\DataTypeTok{values =} \KeywordTok{c}\NormalTok{(}\StringTok{"orangered4"}\NormalTok{,}\StringTok{"chartreuse4"}\NormalTok{,}\StringTok{"skyblue"}\NormalTok{,}\StringTok{"gray"}\NormalTok{))}\OperatorTok{+}
+\StringTok{  }\KeywordTok{theme_tufte}\NormalTok{()}\OperatorTok{+}\KeywordTok{ggtitle}\NormalTok{(}\StringTok{"Empirical distribution of distances (muslim)"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-7-1} \end{center}
+
+\normalsize
+
+\noindent Once we focus on words related to islam, the associated bias
+seems to be stronger than in the whole dataset. This is a step towards
+illustrating that the distribution of bias is uneven.
+
+Now, say we want to look at a single protected word. Since the dataset
+also contains comparison multiple control neutral and human attributes,
+we randomly select only 5 from \texttt{none} and 5 from \texttt{human}
+control groups of those for the visualisation purposes.
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(tidyverse)}
+\NormalTok{muslimClass <-}\StringTok{ }\NormalTok{muslim }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(protectedWord }\OperatorTok{==}\StringTok{ "muslim"}\NormalTok{)}
+\NormalTok{neutralSample <-}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\KeywordTok{filter}\NormalTok{(muslimClass,connection }\OperatorTok{==}\StringTok{ "none"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
+\NormalTok{humanSample <-}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\KeywordTok{filter}\NormalTok{(muslimClass,connection }\OperatorTok{==}\StringTok{ "human"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
+\NormalTok{muslimVis <-}\StringTok{ }\NormalTok{muslimClass }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(connection }\OperatorTok{!=}\StringTok{ "none"} \OperatorTok{&}\StringTok{ }\NormalTok{connection }\OperatorTok{!=}\StringTok{"human"}\NormalTok{)}
+\NormalTok{muslimVis <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(muslimVis,neutralSample,humanSample)}
+
+\CommentTok{#we plug in our visualisation script}
+\KeywordTok{source}\NormalTok{(}\StringTok{"../functions/visualisationTools.R"}\NormalTok{)}
+\CommentTok{#two arguments: dataset and protected word}
+\KeywordTok{visualiseProtected}\NormalTok{(muslimVis,}\StringTok{"muslim"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/tableMuslimActive-1} \end{center}
+
+\normalsize
+
+Note that the distance between the grey point and the other points is
+proportional to cosine distance, the non-grey point size is proportional
+to cosine similarity to the protected word, and color groups by the
+connection type. So for \texttt{muslim} it seems that the stereotypes
+coming from the word list are fairly well visible. To give you some
+taste of how uneven the dataset is, compare this to what happens with
+\texttt{priest}.
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(tidyverse)}
+\NormalTok{priestClass <-}\StringTok{ }\NormalTok{religion }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(protectedWord }\OperatorTok{==}\StringTok{ "priest"}\NormalTok{)}
+\NormalTok{neutralSample <-}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\KeywordTok{filter}\NormalTok{(priestClass,connection }\OperatorTok{==}\StringTok{ "none"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
+\NormalTok{humanSample <-}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\KeywordTok{filter}\NormalTok{(priestClass,connection }\OperatorTok{==}\StringTok{ "human"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
+\NormalTok{priestVis <-}\StringTok{ }\NormalTok{priestClass }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(connection }\OperatorTok{!=}\StringTok{ "none"} \OperatorTok{&}\StringTok{ }\NormalTok{connection }\OperatorTok{!=}\StringTok{"human"}\NormalTok{)}
+\NormalTok{priestVis <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(priestVis,neutralSample,humanSample)}
+
+\CommentTok{#we plug in our visualisation script}
+\KeywordTok{source}\NormalTok{(}\StringTok{"../functions/visualisationTools.R"}\NormalTok{)}
+\CommentTok{#two arguments: dataset and protected word}
+\KeywordTok{visualiseProtected}\NormalTok{(priestVis,}\StringTok{"priest"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-8-1} \end{center}
+
+\normalsize
+
+\noindent Here you can see that some human attributes are closer than
+stereotype attributes, and that there is no clear reason to claim that
+\texttt{associated} attributes are closer than \texttt{different} or
+\texttt{human} attributes. This, again, illustrates the need of
+case-by-case analysis with control groups.
+
+The general idea now is that the word lists provided in different pieces
+of research are just samples of attributes associates with various
+stereotypes and should be treated as such: the uncertainty involved and
+the sample sizes should have clear impact on our estimates.
+
+\section{Bayesian model structure and
+assumptions}\label{bayesian-model-structure-and-assumptions}
+
+We will now think of cosine distance as the output variable, and will
+build a few Bayesian models to compare. First, we just build a baseline
+model which estimates cosine distance to the attributes separately for
+each protected word. The underlying idea is that different protected
+words might in general have different relations to all the attributes
+and these relations should be our point of departure.
+
+Here is the intuition behind the mathematical Bayesian model involved.
+Our outcome variable is \texttt{cosine\ difference}, which we take to me
+normally distributed around the predicted mean for a given protected
+word (that is, we assume the residuals are normally distributed). The
+simplest model specification is:
+
+\begin{align}
+cosineDistance_i  & \sim dnorm(\mu_i, \sigma) \\
+\mu_i & = m_{pw} \\
+m_{pw} & ~ dnorm(1,.5) \\
+\sigma &\sim  dcauchy(0,1)
+\end{align}
+
+That is, we assume the estimated means might be different for diferent
+protected words and our prior for the mean and the overal standard
+deviation are normal with mean 1 and sd=.5 and half-cauchy with
+parameters \texttt{0,1}. Further on we'll also estimate additional
+impact the connection type may have. For this impact we take a slightly
+skeptical prior centered around 0 distributed normally with sd = 1.
+These are fairly weak and slightly skeptical regularizing priors, which
+can be illustrated as follows:
+
+\vspace{2mm}
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/priorsVis-1} \end{center}
+
+\section{Choosing predictors}\label{choosing-predictors}
+
+Now we can define and compile the baseline model. Its parameters will
+have a posterior distribution obtained using either Hamiltionian Monte
+Carlo methods (STAN) available through the \texttt{rethinking} package.
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(rethinking)}
+\KeywordTok{options}\NormalTok{(}\DataTypeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\OtherTok{TRUE}\NormalTok{ )}
+\NormalTok{religionBaseline <-}\StringTok{ }\KeywordTok{ulam}\NormalTok{(}
+  \KeywordTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu <-}\StringTok{ }\NormalTok{m[pw],}
+\NormalTok{    m[pw] }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\OperatorTok{~}\StringTok{ }\KeywordTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\NormalTok{  ),}
+  \DataTypeTok{data =}\NormalTok{ religion,}
+  \DataTypeTok{chains=}\DecValTok{2}\NormalTok{ , }\DataTypeTok{iter=}\DecValTok{4000}\NormalTok{ , }\DataTypeTok{warmup=}\DecValTok{1000}\NormalTok{,}
+  \DataTypeTok{start=} \KeywordTok{list}\NormalTok{(}\DataTypeTok{mu =} \DecValTok{1}\NormalTok{, }\DataTypeTok{co =} \DecValTok{0}\NormalTok{, }\DataTypeTok{sigma=} \FloatTok{.3}\NormalTok{),}
+  \DataTypeTok{log_lik =} \OtherTok{TRUE}\NormalTok{, }\DataTypeTok{cores=}\DecValTok{4}
+\NormalTok{)}
+\CommentTok{#saving}
+\CommentTok{#saveRDS(religionBaseline, }
+\CommentTok{#file = "cosineAnalysis/models/religionBaseline.rds")}
+\end{Highlighting}
+\end{Shaded}
+
+The only reason we need it is the evaluation of connection as a
+predictor. Does including it in o the model improve the situation? To
+investigate this, let's now build a model according to the following
+specification:
+
+\begin{align}
+cosineDistance_i  & \sim dnorm(\mu_i, \sigma) \\
+\mu_i & = m_{pw} + co_{con}\\
+m_{pw} & ~ dnorm(1,.5) \\
+co_{con} & ~ dnorm(0,1) \\
+\sigma &\sim  dcauchy(0,1)
+\end{align}
+
+\noindent The idea now is that each connection type comes with its own
+coefficient \(co\) that has impact on mean distances for protected words
+taken separately.
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(rethinking)}
+\KeywordTok{options}\NormalTok{(}\DataTypeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\OtherTok{TRUE}\NormalTok{ )}
+\NormalTok{religionCoefs <-}\StringTok{ }\KeywordTok{ulam}\NormalTok{(}
+  \KeywordTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu <-}\StringTok{ }\NormalTok{m[pw] }\OperatorTok{+}\StringTok{ }\NormalTok{co[con],}
+\NormalTok{    m[pw] }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    co[con] }\OperatorTok{~}\KeywordTok{dnorm}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\OperatorTok{~}\StringTok{ }\KeywordTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\NormalTok{  ),}
+  \DataTypeTok{data =}\NormalTok{ religion,}
+  \DataTypeTok{chains=}\DecValTok{2}\NormalTok{ , }\DataTypeTok{iter=}\DecValTok{8000}\NormalTok{ , }\DataTypeTok{warmup=}\DecValTok{1000}\NormalTok{, }
+  \DataTypeTok{log_lik =} \OtherTok{TRUE}
+\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+\noindent First, let's see if this model is really better in terms of
+the Widely Acceptable Information Criterion (WAIC):
+
+\vspace{1mm} \footnotesize
+
+\begin{verbatim}
+##                   WAIC SE dWAIC dSE pWAIC weight
+## religionCoefs    -2328 93     0  NA    20      1
+## religionBaseline -2283 95    45  17    16      0
+\end{verbatim}
+
+\normalsize
+
+Clearly, it should be given weight 1 as compared to the baseline model.
+So far, we've learned that the connection type actually has predictive
+value. Let's take a look at the coefficient estimates:
+
+\vspace{1mm} \footnotesize
+
+\begin{verbatim}
+##              mean        sd       5.5%      94.5%    n_eff    Rhat4
+## co[1] -0.14956420 0.1151675 -0.3261650 0.03741930 294.2033 1.001449
+## co[2] -0.09880543 0.1145024 -0.2736985 0.08813271 291.5044 1.001564
+## co[3] -0.07282752 0.1133894 -0.2447778 0.11158986 287.7820 1.001627
+## co[4] -0.03103179 0.1131420 -0.2034442 0.15268770 286.8283 1.001606
+\end{verbatim}
+
+\normalsize
+
+\section{Dataset-level coefficients}\label{dataset-level-coefficients}
+
+\noindent Let's plot them together with their highest posterior density
+invervals, for the three topic groups.
+
+\begin{center}
+\begin{figure}[!htb]\centering
+   \begin{minipage}{0.55\textwidth}
+  \includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
+   \end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/religionGoogleCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/genderCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/genderGoogleCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/raceCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/raceGoogleCoeffs.jpeg}
+   \end{minipage}
+\end{figure}
+
+\end{center}
+
+\noindent What should strike us is that while the mean estimates of the
+coefficients indeed do differ a bit, usually the highest posterior
+density invervals all include zero, and so we do not have strong reasons
+to say that, say, as far as the whole religion dataset is involved,
+being associated indeed is connected with lower cosine distance. A
+second striking observation is that the estimated impact for associated
+stereotypes is quite often not too different from the estimated impact
+of attributes associated with different stereotypes, and both are
+sometimes not too far from the estimated impact for simply human
+attributes. In general, once the uncertainty involved is taken seriously
+by using control groups and statistical uncertainty estimation that does
+not dispose of pointwise data, the picture which focuses only on
+differences between means of means is too simplistic.
+
+But this doesn't mean important differences for some protected words are
+not there. For one thing, if you start with a word list that is very
+uneven, the actually not so bad status of some of the protected words
+might mask a pretty bad situation in which some other protected words
+are. For comparison, let's see what a model focused on words related to
+islam tells us.
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{#this is how we build the model}
+\NormalTok{religion <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"cosineAnalysis/datasets/religionReddit.csv"}\NormalTok{)[}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
+\KeywordTok{colnames}\NormalTok{(religion) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
+                        \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
+\KeywordTok{levels}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
+\NormalTok{muslimWords <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"imam"}\NormalTok{,}\StringTok{"islam"}\NormalTok{,}\StringTok{"mosque"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"quran"}\NormalTok{)}
+\NormalTok{muslim <-}\StringTok{ }\NormalTok{religion }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(protectedWord }\OperatorTok{%in%}\StringTok{ }\NormalTok{muslimWords)}
+\NormalTok{muslim}\OperatorTok{$}\NormalTok{protectedWord <-}\StringTok{ }\KeywordTok{droplevels}\NormalTok{(muslim}\OperatorTok{$}\NormalTok{protectedWord)}
+\NormalTok{muslim}\OperatorTok{$}\NormalTok{pw <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(muslim}\OperatorTok{$}\NormalTok{protectedWord)}
+\NormalTok{muslim}\OperatorTok{$}\NormalTok{con <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(muslim}\OperatorTok{$}\NormalTok{connection)}
+\NormalTok{muslim}\OperatorTok{$}\NormalTok{pwFactor <-}\StringTok{ }\KeywordTok{factor}\NormalTok{(}\KeywordTok{paste0}\NormalTok{(muslim}\OperatorTok{$}\NormalTok{protectedWord, muslim}\OperatorTok{$}\NormalTok{connection))}
+\NormalTok{muslim}\OperatorTok{$}\NormalTok{pwIndex <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(muslim}\OperatorTok{$}\NormalTok{pwFactor)}
+
+\NormalTok{islamCoefs <-}\StringTok{ }\KeywordTok{ulam}\NormalTok{(}
+  \KeywordTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu <-}\StringTok{ }\NormalTok{m[pw] }\OperatorTok{+}\StringTok{ }\NormalTok{co[con],}
+\NormalTok{    m[pw] }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    co[con] }\OperatorTok{~}\KeywordTok{dnorm}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\OperatorTok{~}\StringTok{ }\KeywordTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\NormalTok{  ),}
+  \DataTypeTok{data =}\NormalTok{ muslim,}
+  \DataTypeTok{chains=}\DecValTok{2}\NormalTok{ , }\DataTypeTok{iter=}\DecValTok{10000}\NormalTok{ , }\DataTypeTok{warmup=}\DecValTok{1000}\NormalTok{, }\DataTypeTok{cores =} \DecValTok{4}\NormalTok{,}
+  \DataTypeTok{log_lik =} \OtherTok{TRUE}
+\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+Let's take a look at the coefficients:
+
+\vspace{1mm} \footnotesize
+
+\begin{verbatim}
+##                mean        sd       5.5%      94.5%    n_eff    Rhat4
+## co[1] -0.1979930035 0.1708785 -0.4682696 0.07634977 1789.894 1.003445
+## co[2] -0.0334215769 0.1687587 -0.3021954 0.23720938 1738.720 1.003575
+## co[3] -0.0192492753 0.1675754 -0.2840596 0.24860974 1732.907 1.003755
+## co[4] -0.0003911363 0.1670610 -0.2661047 0.26815172 1723.758 1.003837
+\end{verbatim}
+
+\normalsize
+
+\begin{center}
+\includegraphics[width=8cm]{../images/islamCoeffs.jpeg}
+\end{center}
+
+\normalsize
+
+While muslim words were unusual in the sense that the disparity between
+associated attributes and others is stronger, the evidence is still not
+conclusive. This is because the variation even within islam-related
+words is large enough (and sample sizes sufficiently small) for all the
+highest posterior density invervals to still include zeros.
+
+So, it seems, taking a closer look does seem to make a difference. The
+question is, what happens if we do take a close look at the level of
+protected words?
+
+\chapter{Protected-word level
+analysis}\label{protected-word-level-analysis}
+
+\section{Model structure and
+assumptions}\label{model-structure-and-assumptions}
+
+Let's turn then to data analysis that takes a look at protected words
+separately. This time for each combination of a protected word and a
+connection status we will have a separate mean cosine distance estimate,
+each coming with its own highest posterior density interval. This means
+we will use indices that are result from all such combinations (and then
+we will split them up in the model precis to build visualisation, feel
+free to look at the \texttt{visualiseStats.R} script for details).
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{options}\NormalTok{(}\DataTypeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\OtherTok{TRUE}\NormalTok{ ) }\CommentTok{#removes install pop-up request}
+\NormalTok{religion}\OperatorTok{$}\NormalTok{pwFactor <-}\StringTok{ }\KeywordTok{factor}\NormalTok{(}\KeywordTok{paste0}\NormalTok{(religion}\OperatorTok{$}\NormalTok{protectedWord, }\StringTok{"-"}\NormalTok{, religion}\OperatorTok{$}\NormalTok{connection))}
+\NormalTok{religion}\OperatorTok{$}\NormalTok{pwIndex <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(religion}\OperatorTok{$}\NormalTok{pwFactor)}
+
+\NormalTok{religionSeparate <-}\StringTok{ }\KeywordTok{ulam}\NormalTok{(}
+  \KeywordTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu <-}\StringTok{ }\NormalTok{c[pwIndex],}
+\NormalTok{    c[pwIndex] }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\OperatorTok{~}\StringTok{ }\KeywordTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\NormalTok{  ),}
+  \DataTypeTok{data =}\NormalTok{ religion,}
+  \DataTypeTok{chains=}\DecValTok{2}\NormalTok{ , }\DataTypeTok{iter=}\DecValTok{10000}\NormalTok{ , }\DataTypeTok{warmup=}\DecValTok{1000}\NormalTok{,}
+  \DataTypeTok{start=}\KeywordTok{list}\NormalTok{(}\DataTypeTok{no =} \DecValTok{1}\NormalTok{, }\DataTypeTok{a =} \DecValTok{0}\NormalTok{, }\DataTypeTok{d =} \DecValTok{0}\NormalTok{, }\DataTypeTok{sigma=} \FloatTok{.3}\NormalTok{), }\DataTypeTok{log_lik =}  \OtherTok{TRUE}
+\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+\noindent Let's see if the individualized model does better than the
+previous models in light of WAIC which does add penalty for the number
+of parameters.
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{compareBaselineCoefsSeparate<-}\StringTok{ }\KeywordTok{readRDS}\NormalTok{(}\StringTok{"../datasets/compareBaselineCoefsSeparate.rds"}\NormalTok{)}
+\NormalTok{compareBaselineCoefsSeparate}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+##                   WAIC SE dWAIC dSE pWAIC weight
+## religionSeparate -2400 93     0  NA    60      1
+## religionCoefs    -2328 93    72  29    20      0
+## religionBaseline -2283 95   117  37    16      0
+\end{verbatim}
+
+\normalsize
+
+\section{Protected classes in Reddit and Google
+embeddings}\label{protected-classes-in-reddit-and-google-embeddings}
+
+It seems that we do want to prefer this model, despite its relative
+complication. Now, what does it tell us about the protected words? Let's
+visualise the predicted means together with 89\% highest posterior
+density intervals.
+
+\includegraphics[width=14cm]{../images/visReligionReddit.png}
+
+Before we move on, let's perform analogous analyses for the remaining
+types of supposed bias: gender and race (the model building is
+analogous).
+
+\includegraphics[width=14cm]{../images/visGenderReddit.png}
+
+\includegraphics[width=14cm]{../images/visRaceReddit.png}
+
+We first encountered the GoogleNews word embedding model in Bolukbasi et
+al. (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}). They use it for
+their calculations as written in
+\href{https://github.com/tolga-b/debiaswe}{Tolga Code}. We decided to
+compare the results obtained with the use of Reddit L2 model and the
+ones that we got by GoogleNews model. The details of the model can be
+found here \href{https://code.google.com/archive/p/word2vec/}{Google
+News model}. One of the main differences between the models is that
+Reddit word embeggings have 50 dimensions and GoogleNews word embeddings
+have 300 dimensions. As the dimension is increasing, the vector can
+capture much more information but this is not always the case. According
+to some researchers the amount of dimensions should be chosen by taking
+into account the type of corpus and its features. In our case both
+Reddit and GoogleNews models are already trained and ready to use so we
+do not analyse further the choice of their hyperparameters.
+
+\includegraphics[width=14cm]{../images/visReligionGoogle.png}
+
+\includegraphics[width=14cm]{../images/visGenderGoogle.png}
+
+\includegraphics[width=14cm]{../images/visRaceGoogle.png}
+
+\chapter{The role of debiasing}\label{the-role-of-debiasing}
+
+We also applied to the word embeddings hard debiasing method found in
+Manzini et al.
+(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}). As it was
+mentioned before, debiasing consists of two components: identifying the
+bias subspace and then removing this vector subspace from the chosen
+embeddings. The aim is to increase the cosine distance between protected
+words and the set of attributes. After debiasing the Reddit word
+embeddings we calculated the cosine distance again to see if there are
+any significant changes in terms of the similarity between protected
+words and the attributes.
+
+\section{Dataset-level coefficients after
+debiasing}\label{dataset-level-coefficients-after-debiasing}
+
+First, let's look at coefficient estimated for the whole datasets, as
+compared to their estimation prior to debiasing:
+
+\begin{center}
+\begin{figure}[!htb]\centering
+   \begin{minipage}{0.55\textwidth}
+  \includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
+   \end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/debiasedReligionRedditCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/genderCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/debiasedGenderRedditCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/raceCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/debiasedRaceRedditCoeffs.jpeg}
+   \end{minipage}
+\end{figure}
+
+\end{center}
+
+One may observe very slight changes in the estimated coeﬀicients. It
+seems that in religion dataset the changes are the most significant as
+the mean moves towards zero which stands for no similarity between
+words.
+
+\section{Protected classes after
+debiasing}\label{protected-classes-after-debiasing}
+
+Now, perhaps, the effects of debiasing will be better appreciated if we
+look at the level of protected words. After all, the hope is, the
+situation of extremely ill-positioned protected words have improved?
+
+\includegraphics[width=14cm]{../images/visDebReligionReddit.png}
+
+\includegraphics[width=14cm]{../images/visDebGenderReddit.png}
+
+\includegraphics[width=14cm]{../images/visDebRaceReddit.png}
+
+One may notice that in a religion data set most of the distances for
+\texttt{associated} words slightly increased. Some of the distances from
+\texttt{different} class are still smaller than for the
+\texttt{associated} class. The distances for all classes seem in most
+cases to cluster together even after debiasing. The reason may be that
+uncertainty is quite high in \texttt{associated} and \texttt{different}
+class because the number of samples is so small. Additionally in some
+cases distances from \texttt{human} class are smaller than for the
+\texttt{assosiated} class which also makes it less clear how to
+interpret the results, if the distances are really able to catch the
+bias presence.
+
+It is worth paying attention to the gender debiased dataset. The change
+is really minor, still zero out of HPDI range. The \texttt{different}
+attributes still have high similarity value. It seems that the distances
+for gender protected words are still clustered together which suggests
+that this method is not able to catch the complex bias presence. One
+could conclude that co-occurrence of certain terms does not always mean
+direct connection in terms of terms associations. The fact that some of
+stereotypically associated with female attributes still have high
+similarity with male protected words suggest that other techniques
+should be applied to detect and remove the bias.
+
+At the same time, there is a small improvement in race dataset. In some
+cases the distances just cluster together. In general distances for
+\texttt{assosiated} class are still higher than those for the rest of
+classes. One may assume that there reason for so small change is that
+either the metric does not indicate properly biases or that the issue is
+more complex and subtracting subspaces is not enough.
+
+In Manzini et al.
+(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) it is
+mentioned that debiasing with this method that is based on bias
+component removal may be insufficient. The reason is that words
+containing similar biases may still cluster together even after
+debiasing.
+
+\chapter{Discussion}\label{discussion}
+
+It is worth diving deeper into some observations that arise while
+analyzing the results. As we pointed out, if there is no control of what
+cosine distance values individual word pairs have, then there is less
+understanding of the exact relations between words. If the protected
+words that we chose do not have high similarity with harmful
+stereotypes, then one should consider at least two scenarios. The first
+one is that the choice of protected words and attributes may be
+corrupted. The second one is that the metric is not able to catch the
+hidden bias properly. In both cases it is essential to take a look at
+the individual values before averaging them or aggregating in other
+ways.
+
+One has to also remember about the specific situation present in gender
+dataset. Surprisingly one could observe there high cosine distances
+values between some female stereotypical professions and male protected
+words. This leads to new ideas of how to understand the bias origins and
+its proper detection. To start with, cosine distance seems to catch the
+information regarding words co-occurrence and not the semantic
+similarity. If a word stereotypically associated with females has high
+cosine distance value for male protected words, then the metric is not
+proper to establish the bias presence as it can be misleading. What we
+mean is that in same cases it can unnecessarily label words as biased
+while omitting other more harmful associations. What is more, this
+highlights again the need for control group. One could check each
+stereotypical attribute with each protected word but not in order to
+average the values. One should rather plot the distributions and
+investigate how the metric measures vectors geometry and hidden semantic
+information.
+
+Debiasing with the method provided by Manzini et al.
+(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) shows how it
+is unpredictable in terms of the results. In some cases the final cosine
+distance is indeed close to neutral words, but in other cases the change
+is minor. What is more, the inclusion of the uncertainty shows how in
+some cases it is not clear why a word is according to this method
+classified as anomalous. The fact that uncertainty for
+\texttt{assosiated} and \texttt{different} class is so high leads to the
+situation where in the boundaries of an uncertainty of some attributes,
+one can find the neutral words as well. It makes the usage and
+interpretation of this dataset and metric extremely complicated and
+unclear.
+
+Additionally, one cannot be sure if the bias is still preserved after
+the debiasing. The fact that all of the cosine distances for protected
+words and harmful attributes moved to the right, does not mean that the
+bias is removed. One could argue that maybe it is not clear how to
+measure bias presence and removal but the method for debiasing works
+fine. Notwithstanding, it is showed in researches such as Gonen \&
+Goldberg (\protect\hyperlink{ref-Gonen2019Lipstick}{2019}), that the
+bias can hide in the vector geometry and preserve even after applying
+popular debiasing methods. Therefore it seems to not be justified to
+claim that the method works properly without precise methods to verify
+this.
+
+One should remember that there is no baseline for the interpretation of
+the averaged cosine distances in ({\textbf{???}}). One may assume that
+if the cosine distance is close to 1 then it is a desired outcome as it
+means, according to cosine distance assumptions, that there is almost no
+similarity between the words. However what does it mean to be close to
+0? If the averaged cosine distance is equal to 80, then should we still
+debiase it? It is unclear what the criteria are. On one hand, it seems
+to be beneficial when the outcome is simplified as it is easier to
+compare results with one value per set. On the other hand, it is prone
+to misunderstanding of how to interpret the results and what threshold
+to assume.
+
+We propose the use of Bayesian method to introduce the uncertainty
+measure in bias detection. There are a few advantages when including
+this method in the bias analysis. As noted before in our example with
+religion dataset, one may obtain new insight into the bias issue after
+analyzing the mean estimates of the coeﬀicients. There seems to be a
+minor difference for the mean concerning the \texttt{assosiated} class.
+This can lead to the conclusions that we do not have strong reasons to
+perceive lower cosine distance as significantly related to the fact that
+the attribute is associated.
+
+One may dive deeper into this idea by investigating the connection
+coefficients for individual words with the use of the visualization.
+There one observes that indeed most of the words are clustered together
+independently of the class that they belong to. It is of course partly
+due to the fact that we include uncertainty now. This example shows how
+understanding of the bias measurement may differ when switching to
+Bayesian method. One could argue that after this analysis one obtains
+more information on the dataset and can pursue with the further research
+more cautiously.
+
+\chapter{Summary}\label{summary}
+
+The aim of this paper was to show the analysis of bias detection methods
+used in modern natural language processing projects.
+
+In the further research it is worth to dive deeper into some issues or
+to examine new ideas that occurred during our work.
 
 \hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\hypertarget{ref-Bolukbasi2016Man}{}%
-Bolukbasi, T., Chang, K.-W., Zou, J.
-Y., Saligrama, V., \& Kalai, A.
-(2016).
-Man is to computer programmer as woman is to homemaker?
-Debiasing word embeddings.
-\emph{CoRR}, \emph{abs/1607.06520}.
-Retrieved from \url{http://arxiv.org/abs/1607.06520}
+\hypertarget{ref-Bolukbasi2016Man}{}
+Bolukbasi, T., Chang, K., Zou, J. Y., Saligrama, V., \& Kalai, A.
+(2016). Man is to computer programmer as woman is to homemaker?
+Debiasing word embeddings. \emph{CoRR}, \emph{abs/1607.06520}. Retrieved
+from \url{http://arxiv.org/abs/1607.06520}
 
-\leavevmode\hypertarget{ref-Manzini2019blackToCriminal}{}%
-Manzini, T., Lim, Y.
-C., Tsvetkov, Y., \& Black, A.
-W.
-(2019).
-Black is to criminal as caucasian is to police: Detecting and removing multiclass bias in word embeddings.
-Retrieved from \url{http://arxiv.org/abs/1904.04047}
+\hypertarget{ref-Gonen2019Lipstick}{}
+Gonen, H., \& Goldberg, Y. (2019). Lipstick on a pig: Debiasing methods
+cover up systematic gender biases in word embeddings but do not remove
+them. \emph{CoRR}, \emph{abs/1903.03862}. Retrieved from
+\url{http://arxiv.org/abs/1903.03862}
 
-\leavevmode\hypertarget{ref-Venkat2018Curse}{}%
+\hypertarget{ref-Caliskan2017Semantics}{}
+Islam, A. C., Bryson, J. J., \& Narayanan, A. (2016). Semantics derived
+automatically from language corpora necessarily contain human biases.
+\emph{CoRR}, \emph{abs/1608.07187}. Retrieved from
+\url{http://arxiv.org/abs/1608.07187}
+
+\hypertarget{ref-Manzini2019blackToCriminal}{}
+Manzini, T., Lim, Y. C., Tsvetkov, Y., \& Black, A. W. (2019). Black is
+to criminal as caucasian is to police: Detecting and removing multiclass
+bias in word embeddings.
+
+\hypertarget{ref-Mehrabi2019Survey}{}
+Mehrabi, N., Morstatter, F., Saxena, N., Lerman, K., \& Galstyan, A.
+(2019). A survey on bias and fairness in machine learning. \emph{CoRR},
+\emph{abs/1908.09635}. Retrieved from
+\url{http://arxiv.org/abs/1908.09635}
+
+\hypertarget{ref-Nissim2019Fair}{}
+Nissim, M., Noord, R. van, \& Goot, R. van der. (2019). Fair is better
+than sensational: Man is to doctor as woman is to doctor. \emph{CoRR},
+\emph{abs/1905.09866}. Retrieved from
+\url{http://arxiv.org/abs/1905.09866}
+
+\hypertarget{ref-Venkat2018Curse}{}
 Venkat, N. (2018). The curse of dimensionality: Inside out.
 
-\end{CSLReferences}
-
 \end{document}

### 2021-06-19 thesis/04-discussion.Rmd
Adds files

@@
 # Discussion
+
+1. Our proposition - Bayesian method (red) comment on remarks from chapter2 and then; what we do better
+- general insights - estimeated coefficients mean - general HPDI and general different similar - after debiasing the same almost for reddit
+- details from the words view - questions: analyzing table features; when human slightly closer - unexpected;
+describe the features; 3 small csvs; reddit, google reddit-debiased;general human moved closer- not in table but in sentences; 
+comments and values that refer to them: (blue and green)
+a) google scale
+b) associated females
+debiased: question about the change
+- better but not for muslims
+- corpus frequency maybe?
+- better for male
+- when in gender different and associated together maybe it is better - gender is very specific, words are about professions; 
+- when in religion it is bad different and associated
+-  in race it is not obvious - mixed sentiments
+
+black remarks then:
+- mess, change protected words; relative differences stay the same, by the way lipstick on a pig
+- word lists are strange, not enough words, large HPDI
+
+4. Summary (how abstract ends)
 
@@
 One could argue that after this analysis one obtains more information on the dataset and can pursue with the further research more cautiously.
-
-# Summary
-
-The aim of this paper was to show the analysis of bias detection methods used in modern natural language processing projects.
-
-In the further research it is worth to dive deeper into some issues or to examine new ideas that occurred during our work.
 

### 2021-06-19 thesis/_book/_main.tex
Adds files

--- 2021-06-16 thesis/_book/_main.tex
@@
 \author{Alicja Dobrzeniecka}
-\date{2021-06-16}
+\date{2021-06-17}
 
@@
 
+First let's take a look at the empirical distribution of distances by the connection type, initially ignoring the human control class for now.
+
+\vspace{1mm}
+\footnotesize
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-2-1} \end{center}
+\normalsize
+
+The first impression is that while there is a shift for associated words towards smaller cosine distances as compared to the neutral words, slightly surprisingly a slightly weaker shift in the same direction is visible for attributes associated with different stereotypes.
+Moreover, the empirical distributions overlap to alarge extent and the means grouped by connection type do not seem too far from each other.
+In fact, as there is a lot of variety in the cosine distances (as we will soon see), we need to gauge the uncertainty involved, and to look more carefully at individual protected words to get a better idea of how the cosine distance distribution changes for different attribute groups and different protected classes.
+Now, let's add the human attributes to the picture:
+
+\vspace{1mm}
+\footnotesize
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-3-1} \end{center}
+\normalsize
+
+\noindent Notice that the distribution for \texttt{human} (even though we did our best not to include in it any stereotype-related atributes) is left-skewed, with much overlap with \texttt{associated} and \texttt{different}, which illustrates the need to take being associated with humans as an important predictor.
+
+Our focus lies in \texttt{connection} as a predictor.
+Morever, later on we'll be interested in looking at the protected words separately, and at protected words split by connection.
+For technical reasons it is useful to represent these factors as integer vectors.
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{religion}\SpecialCharTok{$}\NormalTok{con }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{connection)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning: NAs introduced by coercion
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{religion}\SpecialCharTok{$}\NormalTok{pw }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{protectedWord)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning: NAs introduced by coercion
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{religion}\SpecialCharTok{$}\NormalTok{pwFactor }\OtherTok{\textless{}{-}} \FunctionTok{factor}\NormalTok{(}\FunctionTok{paste0}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{protectedWord, religion}\SpecialCharTok{$}\NormalTok{connection))}
+\NormalTok{religion}\SpecialCharTok{$}\NormalTok{pwIndex }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{pwFactor)}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+A short script, \texttt{cleanDataset} to make this faster, so equivalently:
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{source}\NormalTok{(}\StringTok{"../functions/cleanDataset.R"}\NormalTok{)}
+\NormalTok{religion }\OtherTok{\textless{}{-}} \FunctionTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{]}
+\NormalTok{religion }\OtherTok{\textless{}{-}} \FunctionTok{cleanDataset}\NormalTok{(religion,}\FunctionTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning in cleanDataset(religion, c("christian", "human", "jewish", "muslim", :
+## NAs introduced by coercion
+
+## Warning in cleanDataset(religion, c("christian", "human", "jewish", "muslim", :
+## NAs introduced by coercion
+\end{verbatim}
+
+\normalsize
+
 \hypertarget{looking-at-the-islam-related-words}{%
 \section{Looking at the islam-related words}\label{looking-at-the-islam-related-words}}
+
+For now, let's focus on five protected words related to islam (``imam,'' ``islam,'' ``mosque,'' ``muslim,'' and ``quran'').
+The word list associates with islam four stereotypical attributes (``violent,'' ``terrorist,'' ``uneducated'' and ``dirty'').
+First, we select and plot the empirical distributions for these protected words.
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{library}\NormalTok{(tidyverse)}
+\NormalTok{muslimWords }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"imam"}\NormalTok{,}\StringTok{"islam"}\NormalTok{,}\StringTok{"mosque"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"quran"}\NormalTok{)}
+\NormalTok{muslim }\OtherTok{\textless{}{-}}\NormalTok{ religion }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(protectedWord }\SpecialCharTok{\%in\%}\NormalTok{ muslimWords)}
+\FunctionTok{ggplot}\NormalTok{(muslim, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{  cosineDistance, }\AttributeTok{fill =}\NormalTok{ connection, }\AttributeTok{color =}\NormalTok{ connection))}\SpecialCharTok{+}
+  \FunctionTok{geom\_density}\NormalTok{(}\AttributeTok{alpha=}\FloatTok{0.6}\NormalTok{,}\AttributeTok{size =}\NormalTok{ .}\DecValTok{2}\NormalTok{)}\SpecialCharTok{+} 
+  \FunctionTok{scale\_fill\_manual}\NormalTok{(}\AttributeTok{values =} \FunctionTok{c}\NormalTok{(}\StringTok{"orangered4"}\NormalTok{,}\StringTok{"chartreuse4"}\NormalTok{, }\StringTok{"skyblue"}\NormalTok{, }\StringTok{"gray"}\NormalTok{))}\SpecialCharTok{+}
+  \FunctionTok{scale\_x\_continuous}\NormalTok{(}\AttributeTok{breaks =} \FunctionTok{seq}\NormalTok{(}\FloatTok{0.3}\NormalTok{,}\FloatTok{1.5}\NormalTok{, }\AttributeTok{by =} \FloatTok{0.1}\NormalTok{))}\SpecialCharTok{+}\FunctionTok{xlab}\NormalTok{(}\StringTok{"cosine distance"}\NormalTok{)}\SpecialCharTok{+}
+  \FunctionTok{scale\_color\_manual}\NormalTok{(}\AttributeTok{values =} \FunctionTok{c}\NormalTok{(}\StringTok{"orangered4"}\NormalTok{,}\StringTok{"chartreuse4"}\NormalTok{,}\StringTok{"skyblue"}\NormalTok{,}\StringTok{"gray"}\NormalTok{))}\SpecialCharTok{+}
+  \FunctionTok{theme\_tufte}\NormalTok{()}\SpecialCharTok{+}\FunctionTok{ggtitle}\NormalTok{(}\StringTok{"Empirical distribution of distances (muslim)"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-6-1} \end{center}
+\normalsize
+
+\noindent Once we focus on words related to islam, the associated bias seems to be stronger than in the whole dataset. This is a step towards illustrating that the distribution of bias is uneven.
+
+Now, say we want to look at a single protected word.
+Since the dataset also contains comparison multiple control neutral and human attributes, we randomly select only 5 from \texttt{none} and 5 from \texttt{human} control groups of those for the visualisation purposes.
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{library}\NormalTok{(tidyverse)}
+\NormalTok{muslimClass }\OtherTok{\textless{}{-}}\NormalTok{ muslim }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(protectedWord }\SpecialCharTok{==} \StringTok{"muslim"}\NormalTok{)}
+\NormalTok{neutralSample }\OtherTok{\textless{}{-}} \FunctionTok{sample\_n}\NormalTok{(}\FunctionTok{filter}\NormalTok{(muslimClass,connection }\SpecialCharTok{==} \StringTok{"none"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
+\NormalTok{humanSample }\OtherTok{\textless{}{-}} \FunctionTok{sample\_n}\NormalTok{(}\FunctionTok{filter}\NormalTok{(muslimClass,connection }\SpecialCharTok{==} \StringTok{"human"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
+\NormalTok{muslimVis }\OtherTok{\textless{}{-}}\NormalTok{ muslimClass }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(connection }\SpecialCharTok{!=} \StringTok{"none"} \SpecialCharTok{\&}\NormalTok{ connection }\SpecialCharTok{!=}\StringTok{"human"}\NormalTok{)}
+\NormalTok{muslimVis }\OtherTok{\textless{}{-}} \FunctionTok{rbind}\NormalTok{(muslimVis,neutralSample,humanSample)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning in `[<-.factor`(`*tmp*`, ri, value = structure(c("muslim_stereotype", :
+## invalid factor level, NA generated
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#we plug in our visualisation script}
+\FunctionTok{source}\NormalTok{(}\StringTok{"../functions/visualisationTools.R"}\NormalTok{)}
+\CommentTok{\#two arguments: dataset and protected word}
+\FunctionTok{visualiseProtected}\NormalTok{(muslimVis,}\StringTok{"muslim"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/tableMuslimActive-1} \end{center}
+\normalsize
+
+Note that the distance between the grey point and the other points is proportional to cosine distance, the non-grey point size is proportional to cosine similarity to the protected word, and color groups by the connection type.
+So for \texttt{muslim} it seems that the stereotypes coming from the word list are fairly well visible.
+To give you some taste of how uneven the dataset is, compare this to what happens with \texttt{priest}.
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{library}\NormalTok{(tidyverse)}
+\NormalTok{priestClass }\OtherTok{\textless{}{-}}\NormalTok{ religion }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(protectedWord }\SpecialCharTok{==} \StringTok{"priest"}\NormalTok{)}
+\NormalTok{neutralSample }\OtherTok{\textless{}{-}} \FunctionTok{sample\_n}\NormalTok{(}\FunctionTok{filter}\NormalTok{(priestClass,connection }\SpecialCharTok{==} \StringTok{"none"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
+\NormalTok{humanSample }\OtherTok{\textless{}{-}} \FunctionTok{sample\_n}\NormalTok{(}\FunctionTok{filter}\NormalTok{(priestClass,connection }\SpecialCharTok{==} \StringTok{"human"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
+\NormalTok{priestVis }\OtherTok{\textless{}{-}}\NormalTok{ priestClass }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(connection }\SpecialCharTok{!=} \StringTok{"none"} \SpecialCharTok{\&}\NormalTok{ connection }\SpecialCharTok{!=}\StringTok{"human"}\NormalTok{)}
+\NormalTok{priestVis }\OtherTok{\textless{}{-}} \FunctionTok{rbind}\NormalTok{(priestVis,neutralSample,humanSample)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning in `[<-.factor`(`*tmp*`, ri, value = structure(c("muslim_stereotype", :
+## invalid factor level, NA generated
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#we plug in our visualisation script}
+\FunctionTok{source}\NormalTok{(}\StringTok{"../functions/visualisationTools.R"}\NormalTok{)}
+\CommentTok{\#two arguments: dataset and protected word}
+\FunctionTok{visualiseProtected}\NormalTok{(priestVis,}\StringTok{"priest"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-7-1} \end{center}
+\normalsize
+
+\noindent Here you can see that some human attributes are closer than stereotype attributes, and that there is no clear reason to claim that \texttt{associated} attributes are closer than \texttt{different} or \texttt{human} attributes.
+This, again, illustrates the need of case-by-case analysis with control groups.
+
+The general idea now is that the word lists provided in different pieces of research are just samples of attributes associates with various stereotypes and should be treated as such: the uncertainty involved and the sample sizes should have clear impact on our estimates.
 
@@
 
+We will now think of cosine distance as the output variable, and
+will build a few Bayesian models to compare.
+First, we just build a baseline model which estimates cosine distance to the attributes separately for each protected word.
+The underlying idea is that different protected words might in general have different relations to all the attributes and these relations should be our point of departure.
+
+Here is the intuition behind the mathematical Bayesian model involved.
+Our outcome variable is \texttt{cosine\ difference}, which we take to me normally distributed around the predicted mean for a given protected word (that is, we assume the residuals are normally distributed).
+The simplest model specification is:
+
+\begin{align}
+cosineDistance_i  & \sim dnorm(\mu_i, \sigma) \\
+\mu_i & = m_{pw} \\
+m_{pw} & ~ dnorm(1,.5) \\
+\sigma &\sim  dcauchy(0,1)
+\end{align}
+
+That is, we assume the estimated means might be different for diferent protected words and our prior for the mean and the overal standard deviation are normal with mean 1 and sd=.5 and half-cauchy with parameters \texttt{0,1}.
+Further on we'll also estimate additional impact the connection type may have.
+For this impact we take a slightly skeptical prior centered around 0 distributed normally with sd = 1.
+These are fairly weak and slightly skeptical regularizing priors, which can be illustrated as follows:
+
+\vspace{2mm}
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/priorsVis-1} \end{center}
+
 \hypertarget{choosing-predictors}{%
 \section{Choosing predictors}\label{choosing-predictors}}
+
+Now we can define and compile the baseline model.
+Its parameters will have a posterior distribution obtained using either Hamiltionian Monte Carlo methods (STAN) available through the \texttt{rethinking} package.
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{library}\NormalTok{(rethinking)}
+\FunctionTok{options}\NormalTok{(}\AttributeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\ConstantTok{TRUE}\NormalTok{ )}
+\NormalTok{religionBaseline }\OtherTok{\textless{}{-}} \FunctionTok{ulam}\NormalTok{(}
+  \FunctionTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu }\OtherTok{\textless{}{-}}\NormalTok{ m[pw],}
+\NormalTok{    m[pw] }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\NormalTok{  ),}
+  \AttributeTok{data =}\NormalTok{ religion,}
+  \AttributeTok{chains=}\DecValTok{2}\NormalTok{ , }\AttributeTok{iter=}\DecValTok{4000}\NormalTok{ , }\AttributeTok{warmup=}\DecValTok{1000}\NormalTok{,}
+  \AttributeTok{start=} \FunctionTok{list}\NormalTok{(}\AttributeTok{mu =} \DecValTok{1}\NormalTok{, }\AttributeTok{co =} \DecValTok{0}\NormalTok{, }\AttributeTok{sigma=}\NormalTok{ .}\DecValTok{3}\NormalTok{),}
+  \AttributeTok{log\_lik =} \ConstantTok{TRUE}\NormalTok{, }\AttributeTok{cores=}\DecValTok{4}
+\NormalTok{)}
+\CommentTok{\#saving}
+\CommentTok{\#saveRDS(religionBaseline, }
+\CommentTok{\#file = "cosineAnalysis/models/religionBaseline.rds")}
+\end{Highlighting}
+\end{Shaded}
+
+The only reason we need it is the evaluation of connection as a predictor.
+Does including it in o the model improve the situation?
+To investigate this, let's now build a model according to the following specification:
+
+\begin{align}
+cosineDistance_i  & \sim dnorm(\mu_i, \sigma) \\
+\mu_i & = m_{pw} + co_{con}\\
+m_{pw} & ~ dnorm(1,.5) \\
+co_{con} & ~ dnorm(0,1) \\
+\sigma &\sim  dcauchy(0,1)
+\end{align}
+
+\noindent The idea now is that each connection type comes with its own coefficient \(co\) that has impact on mean distances for protected words taken separately.
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{library}\NormalTok{(rethinking)}
+\FunctionTok{options}\NormalTok{(}\AttributeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\ConstantTok{TRUE}\NormalTok{ )}
+\NormalTok{religionCoefs }\OtherTok{\textless{}{-}} \FunctionTok{ulam}\NormalTok{(}
+  \FunctionTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu }\OtherTok{\textless{}{-}}\NormalTok{ m[pw] }\SpecialCharTok{+}\NormalTok{ co[con],}
+\NormalTok{    m[pw] }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    co[con] }\SpecialCharTok{\textasciitilde{}}\FunctionTok{dnorm}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\NormalTok{  ),}
+  \AttributeTok{data =}\NormalTok{ religion,}
+  \AttributeTok{chains=}\DecValTok{2}\NormalTok{ , }\AttributeTok{iter=}\DecValTok{8000}\NormalTok{ , }\AttributeTok{warmup=}\DecValTok{1000}\NormalTok{, }
+  \AttributeTok{log\_lik =} \ConstantTok{TRUE}
+\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+\noindent First, let's see if this model is really better in terms of the Widely Acceptable Information Criterion (WAIC):
+
+\vspace{1mm}
+\footnotesize
+
+\begin{verbatim}
+##                   WAIC SE dWAIC dSE pWAIC weight
+## religionCoefs    -2328 93     0  NA    20      1
+## religionBaseline -2283 95    45  17    16      0
+\end{verbatim}
+
+\normalsize
+
+Clearly, it should be given weight 1 as compared to the baseline model. So far, we've learned that the connection type actually has predictive value. Let's take a look at the coefficient estimates:
+
+\vspace{1mm}
+\footnotesize
+
+\begin{verbatim}
+##              mean        sd       5.5%      94.5%    n_eff    Rhat4
+## co[1] -0.14956420 0.1151675 -0.3261650 0.03741930 294.2033 1.001449
+## co[2] -0.09880543 0.1145024 -0.2736985 0.08813271 291.5044 1.001564
+## co[3] -0.07282752 0.1133894 -0.2447778 0.11158986 287.7820 1.001627
+## co[4] -0.03103179 0.1131420 -0.2034442 0.15268770 286.8283 1.001606
+\end{verbatim}
+
+\normalsize
 
@@
 
+It seems that we do want to prefer this model, despite its relative complication.
+Now, what does it tell us about the protected words?
+Let's visualise the predicted means together with 89\% highest posterior density intervals.
+
+\includegraphics[width=14cm]{../images/visReligionReddit.png}
+
+Before we move on, let's perform analogous analyses for the remaining types of supposed bias: gender and race (the model building is analogous).
+
+\includegraphics[width=14cm]{../images/visGenderReddit.png}
+
+\includegraphics[width=14cm]{../images/visRaceReddit.png}
+
+We first encountered the GoogleNews word embedding model in \protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}).
+They use it for their calculations as written in \href{https://github.com/tolga-b/debiaswe}{Tolga Code}.
+We decided to compare the results obtained with the use of Reddit L2 model and the ones that we got by GoogleNews model.
+The details of the model can be found here \href{https://code.google.com/archive/p/word2vec/}{Google News model}.
+One of the main differences between the models is that Reddit word embeggings have 50 dimensions and GoogleNews word embeddings have 300 dimensions.
+As the dimension is increasing, the vector can capture much more information but this is not always the case.
+According to some researchers the amount of dimensions should be chosen by taking into account the type of corpus and its features.
+In our case both Reddit and GoogleNews models are already trained and ready to use so we do not analyse further the choice of their hyperparameters.
+
+\includegraphics[width=14cm]{../images/visReligionGoogle.png}
+
+\includegraphics[width=14cm]{../images/visGenderGoogle.png}
+
+\includegraphics[width=14cm]{../images/visRaceGoogle.png}
+
+\hypertarget{the-role-of-debiasing}{%
+\chapter{The role of debiasing}\label{the-role-of-debiasing}}
+
+We also applied to the word embeddings hard debiasing method found in \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}).
+As it was mentioned before, debiasing consists of two components: identifying the bias subspace and then removing this vector subspace from the chosen embeddings.
+The aim is to increase the cosine distance between protected words and the set of attributes.
+After debiasing the Reddit word embeddings we calculated the cosine distance again to see if there are any significant changes in terms of the similarity between protected words and the attributes.
+
 \hypertarget{dataset-level-coefficients-after-debiasing}{%
 \section{Dataset-level coefficients after debiasing}\label{dataset-level-coefficients-after-debiasing}}
+
+First, let's look at coefficient estimated for the whole datasets, as compared to their estimation prior to debiasing:
+
+\begin{center}
+\begin{figure}[!htb]\centering
+   \begin{minipage}{0.55\textwidth}
+  \includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
+   \end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/debiasedReligionRedditCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/genderCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/debiasedGenderRedditCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/raceCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/debiasedRaceRedditCoeffs.jpeg}
+   \end{minipage}
+\end{figure}
+
+\end{center}
+
+One may observe very slight changes in the estimated coeﬀicients.
+It seems that in religion dataset the changes are the most significant as the mean moves towards zero which stands for no similarity between words.
 
@@
 
+Now, perhaps, the effects of debiasing will be better appreciated if we look at the level of protected words.
+After all, the hope is, the situation of extremely ill-positioned protected words have improved?
+
+\includegraphics[width=14cm]{../images/visDebReligionReddit.png}
+
+\includegraphics[width=14cm]{../images/visDebGenderReddit.png}
+
+\includegraphics[width=14cm]{../images/visDebRaceReddit.png}
+
+One may notice that in a religion data set most of the distances for \texttt{associated} words slightly increased.
+Some of the distances from \texttt{different} class are still smaller than for the \texttt{associated} class.
+The distances for all classes seem in most cases to cluster together even after debiasing.
+The reason may be that uncertainty is quite high in \texttt{associated} and \texttt{different} class because the number of samples is so small.
+Additionally in some cases distances from \texttt{human} class are smaller than for the \texttt{assosiated} class which also makes it less clear how to interpret the results, if the distances are really able to catch the bias presence.
+
+It is worth paying attention to the gender debiased dataset.
+The change is really minor, still zero out of HPDI range.
+The \texttt{different} attributes still have high similarity value.
+It seems that the distances for gender protected words are still clustered together which suggests that this method is not able to catch the complex bias presence.
+One could conclude that co-occurrence of certain terms does not always mean direct connection in terms of terms associations.
+The fact that some of stereotypically associated with female attributes still have high similarity with male protected words suggest that other techniques should be applied to detect and remove the bias.
+
+At the same time, there is a small improvement in race dataset.
+In some cases the distances just cluster together.
+In general distances for \texttt{assosiated} class are still higher than those for the rest of classes.
+One may assume that there reason for so small change is that either the metric does not indicate properly biases or that the issue is more complex and subtracting subspaces is not enough.
+
+In \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) it is mentioned that debiasing with this method that is based on bias component removal may be insufficient.
+The reason is that words containing similar biases may still cluster together even after debiasing.
+
 \hypertarget{discussion}{%

### 2021-06-19 thesis/01-intro.Rmd
Applies intro remarks

@@
 # Introduction
-
-<!--### Articles:
-
-  1. "Are We Consistently Biased? Multidimensional Analysis of Biases in Distributional Word Vectors"
-  https://www.aclweb.org/anthology/S19-1010.pdf
-
-  2. Proving the harmfullness of biases in models 
-[source https://poseidon01.ssrn.com/delivery.php?ID=004020091089126117028081015085009075120015077012021005105005104113074127101113104030041107053119007034112081087110119071025006053026022082043000071011123067077124004027040024008122116091025075126111098104007096110113089007070120009111119015084097017092&EXT=pdf&INDEX=TRUE]
-
-  3. Summary what causes bias short article
-  [source https://kawine.github.io/blog/nlp/2019/09/23/bias.html]-->
-
-<!-- Men Also Like Shopping: Reducing Gender Bias Amplification using Corpus-level Constraint -->
-<!-- Bias in Bios: A Case Study of Semantic Representation Bias in a High-Stakes Setting -->
-
-<!-- !bibliografia paperdraft; jabref
-4.pomysly na czwarty rozdzial - podsumowanie; 
-jak mierzyc bias za pomoca jednej liczby?
-IAT - test psychologiczny -->
-
-<!-- 
-    1.state the general topic and give some background
-    2.provide a review of the literature related to the topic
-    3.define the terms and scope of the topic
-    4.outline the current situation
-    5.evaluate the current situation (advantages/ disadvantages) and identify the gap
-    6.identify the importance of the proposed research
-    7.state the research problem/ questions
-    8.state the research aims and/or research objectives
-    9.state the hypotheses
-    10.outline the order of information in the thesis
-    11.outline the methodology
--->
-
-<!-- \textbf{state the general topic and give some background}-->
 
@@
 The models usually assign the values based on the context in which the words appear.
-It means that the input data can have enormous influence on the outcome.
-The biases seem to have their primary source in the way the words are assigned the numerical values.
+This means that the input data can have enormous influence on the outcome.
+The biases seem to have their primary source in the way the words are assigned numerical values.
 
-<!-- \textbf{review of the literature related to the topic}-->
+One can find various definitions trying to capture what bias and fairness actually are.
+With the choice of the definition, implications into the real-life applications may change as well.
+@Mehrabi2019Survey mark out that there exist different types of biases such as historical bias, representation bias or measurement bias (the list is long).
+This indicates how complex the issue of bias is.
+Without the proper understanding and awareness of the problem, people are prone to unconsciously sustain the bias.
+@Mehrabi2019Survey also distinguish different types of discrimination, some of them will be briefly described.
+By protected attributes we mean those qualities, traits or characteristics that one should not discriminate against.
+Direct discrimination occurs when protected attributes of individuals explicitly result in non-favorable outcomes toward them.
+Such discrimination takes place for example when one uses the information about someone's religious beliefs to evaluate whether they are competent for the work they apply for.
+In contrast, in indirect discrimination individuals appear to be treated equally, but anyway they end up being treated unjustly due to the hidden effects of biases towards their protected attributes.
+An example of such situation is when in a job advert there is a requirement of 10 years experience instead of the list of a specific type of experience and knowledge.
+This job advert can discriminate indirectly against young people who can have the required skills, but not yet so much work experience.
+Systemic discrimination takes place when policies, customs or behaviors that result from certain culture or organizational structure lead to discrimination against some groups of people.
+Finally, statistical discrimination consists in relying on group statistics to judge person belonging to a given group.
+The topic of discrimination is entangled with another concept; fairness.
+It is essential to grasp some concepts of fairness to take them into consideration while designing implementation of some machine learning model.
+In @Mehrabi2019Survey one may notice that depending on the context and application different definitions may be applied.
 
-There is considerable amount of literature available on the topic of bias detection and mitigation in NLP models.
+There is a considerable amount of literature available on the topic of bias detection and mitigation in NLP models.
 @Bolukbasi2016Man focus on gender biases that may be observable while investigating the representation of job occupations and gender in terms of their assigned numerical
@@
 The authors apply cosine similarity measurement to investigate the phenomenon where (the vectors corresponding to) words related to jobs that are stereotypically associated with a given gender are in fact in the model situated closer to this gender.
-They also use analogy tasks to evaluate if the bias is present in the word embedding model.
-They check analogies by comparing pairs of word vectors, for example they search for the word complementing the puzzle: man is to doctor as woman is to ...?
-First they subtract the word "man" from the word "woman" and then
+They also use analogy tasks to evaluate if bias is present in a word embedding model.
+They check analogies by comparing pairs of word vectors.
+For example they search for the word complementing the puzzle: man is to doctor as woman is to ...?
+First they subtract the word "man" from the word "woman"^[This is pointwise subtraction of vectors, but for the simplicity we will often talk about words while really meaning vectors.
+What is meant should be clear from the context.] and then
 they search for the ranked list of other words pairs that have similar vector difference. They also include in the formula a threshold to ensure that the resulting pairs could not be randomly picked.
@@
 Therefore, this method seems to be limited in terms of bias detection.
-Other problems are related for example to the choice of pairs and words that are used to detect the presence of discrimination as it is often subjective and without proper justification.
-Additionally the choice of the parameter used in @Bolukbasi2016Man formula to ensure that word pairs are not picked by random, is also not justified and changing it drastically
+Other problems are related for example to the choice of pairs and words that are used to detect the presence of discrimination, as it is often subjective and without proper justification.
+Additionally, the choice of the parameter used in @Bolukbasi2016Man formula to ensure that word pairs are not picked by random, is also not justified and changing it drastically
 influences the results.
@@
 @Caliskan2017Semantics touch upon the topic of biases regarding race and gender.
-They apply knowledge from well-known psychological studies such as Implicit Association Test to research the relation between human stereotypical thinking and model learnt biases to
-discover close relationship between these two. For the evaluation they use Word Embedding Association Test (WEAT) and the Word Embedding Factual Association Test (WEFAT).
+They apply knowledge from well-known psychological studies such as Implicit Association Test to investigate the relation between human stereotypical thinking and model-learnt biases to discover a close relationship between these two.
+For the evaluation they use Word Embedding Association Test (WEAT) and the Word Embedding Factual Association Test (WEFAT).
 
-@manzini2019black propose a novel way of using cosine similarity to obtain information on assumed resemblance between words. They investigate 
-an approach that enables them to measure the bias for a class (like gender, religion, race) and express the final result with a single metric.
+@Manzini2019blackToCriminal propose a novel way of using cosine similarity to verify if the assumed resemblance between protected words and harmful attributes exists.
+They investigate an approach that enables them to measure the bias for a class (such as gender, religion or race) and express the final result with a single metric.
+Their approach will be analyzed in details in the next chapter.
 
-<!-- Nissim2019Fair -->
-
-<!-- \textbf{define the terms and scope of the topic}-->
-
-@Caliskan2017Semantics refer to Implicit Association Test (Greenwald et al., 1998) that measures the strength of associations between concepts or stereotypes by calculating the time of reaction for special tasks.
+@Caliskan2017Semantics refer to Implicit Association Test (Greenwald et al., 1998) that measures the strength of associations between concepts or stereotypes by measuring human reaction time for special tasks.
 Humans naturally exhibit some biases which do not always cause social concern.
@@
 This example is rather used to show that the methodology used in IAT makes sense.
-Unfortunately, the further studies discover other biases and prejudices that directly influence the quality of other people's lives and therefore they should be taken care of.
+@Caliskan2017Semantics were able to obtain similar results to the ones that measured reaction time with the use of WEAT metric.
+Unfortunately, further studies discover other biases and prejudices that directly influence the quality of other people's lives and therefore they should be taken care of.
 
-One can find various definitions trying to capture what bias and fairness actually are.
-With the choice of the definition, implications into the real-life applications may change as well.
-@Mehrabi2019Survey mark out that there exist different types of biases such as historical bias, representation bias or measurement bias (the list is long).
-This indicates how complex the issue of bias is.
-Without the proper understanding and awareness of the problem, people are prone to unconsciously sustain the bias.
-
-@Mehrabi2019Survey also distinguish different types of discrimination, some of them will be briefly described.
-By protected attributes we mean those qualities, traits or characteristics that one should not discriminate against.
-Direct discrimination occurs when protected attributes of individuals explicitly result in non-favorable outcomes toward them.
-Such discrimination takes place for example when one uses an information about someone's religion belief to evaluate whether they are competent for the work they apply for.
-In contrast in indirect discrimination individuals appear to be treated equally but anyway they end up being treated unjustly due to the hidden effects of biases towards their protected attributes.
-An example of such situation is when in a job advert there is a requirement of 10 years experience instead of the list of a specific type of experience and knowledge.
-This job advert can discriminate indirectly against young people who can have equivalent skills but not yet so much work experience.
-Systemic discrimination takes place when policies, customs or behaviors that result from certain culture or organizational structure lead to discrimination against some groups of people.
-Finally, statistical discrimination consists in to using
-average group statistics to judge person belonging to the group.
-
-The topic of discrimination is entangled with another concept; fairness. It is essential to grasp some concepts of fairness to take them into
-consideration while designing implementation of some machine learning model. In @Mehrabi2019Survey one may notice that depending on the context and application different definitions may be applied.
-
-<!--\textbf{outline the current situation} -->
-
-The most common methods in the literature focus on comparing the similarity between words from protected groups and those that are considered to be stereotypical or harmful in some way.
-One can find in this group methods such as euclidean distance or cosine similarity (which is equivalent to dot product if the
-vectors are normalized).
+The most common bias detection methods in the literature focus on comparing the similarity between words from protected groups and those that are considered to be stereotypical or harmful in some way.
+One can find in this group methods such as euclidean distance or cosine similarity (which is equivalent to dot product if the vectors are normalized).
 There are also other ways to detect the effects of biases.
-One can investigate the model performance on certain tasks and validate if the model uses the gender or race information to complete the task.
-
-<!-- \textbf{evaluate the current situation (advantages/ disadvantages) and identify the gap}-->
-
-The currently used methods (such as cosine similarity) make the similarity values often aggregated in a way that may lead to hasty conclusions.
-The averaging of values and the lack of uncertainty may lead to the incomplete picture of the situation.
-
-<!-- \textbf{state the research problem/ questions} -->
+One can investigate the model performance on certain downstream tasks and validate if the model uses the gender or race information to complete the task.
+The currently used methods employing cosine similarity, often aggregate the similarity values often aggregated in a way that may lead to hasty conclusions.
+The averaging of values and the lack of attention to uncertainty involved in such aggregation may lead to the incomplete picture of the situation.
 
 In the paper we indicate how current methods used to detect biases in natural language models are limited, if we investigate them from the perspective of Bayesian analysis.
-
-<!-- \textbf{state the research aims and/or research objectives} -->
-
-Our research enhances the current way in which the bias detection is performed to make sure that it is
-methodologically valid.
-
-<!-- \textbf{state the hypotheses} -->
-
+Our research enhances the current way in which the bias detection is performed to make sure that it is methodologically valid.
 The key hypothesis is that greater understanding of data and bias implications can be achieved when Bayesian methods are applied to issue.
-
-<!-- \textbf{outline the order of information in the thesis} -->

### 2021-06-19 thesis/04-discussion.Rmd
Applies intro remarks

@@
 -->
+
+<!--### Articles:
+
+  1. "Are We Consistently Biased? Multidimensional Analysis of Biases in Distributional Word Vectors"
+  https://www.aclweb.org/anthology/S19-1010.pdf
+
+  2. Proving the harmfullness of biases in models 
+[source https://poseidon01.ssrn.com/delivery.php?ID=004020091089126117028081015085009075120015077012021005105005104113074127101113104030041107053119007034112081087110119071025006053026022082043000071011123067077124004027040024008122116091025075126111098104007096110113089007070120009111119015084097017092&EXT=pdf&INDEX=TRUE]
+
+  3. Summary what causes bias short article
+  [source https://kawine.github.io/blog/nlp/2019/09/23/bias.html]-->
+
+<!-- Men Also Like Shopping: Reducing Gender Bias Amplification using Corpus-level Constraint -->
+<!-- Bias in Bios: A Case Study of Semantic Representation Bias in a High-Stakes Setting -->

### 2021-06-19 thesis/_book/_main.tex
Applies intro remarks

--- 2021-06-16 thesis/_book/_main.tex
@@
 \author{Alicja Dobrzeniecka}
-\date{2021-06-16}
+\date{2021-06-19}
 
@@
 
-Placeholder
+Natural language processing (NLP) is a subfield of computer science that processes and analyzes language in text and speech with the use of modern programming methods.
+It has practical applications in everyday life as it concerns tasks such as email filters, smart assistants, search results, language translations, text analytics and so on.
+Models used to accomplish these tasks need a lot of data to learn from.
+This data originates from human activities and historical recordings such as texts, messages or speeches.
+It turns out that in the learning process these models can learn implicit biases that reflect harmful stereotypical thinking still present in modern societies.
+One can find methods that aim at identifying and measuring hidden biases and/or try to remove them by modifying the models.
+There are many different types of models in NLP depending on a task that they are supposed to solve.
+However, all of them need as an input words represented by means of numbers and this is accomplished with word embedding models.
+The models usually assign the values based on the context in which the words appear.
+This means that the input data can have enormous influence on the outcome.
+The biases seem to have their primary source in the way the words are assigned numerical values.
+
+One can find various definitions trying to capture what bias and fairness actually are.
+With the choice of the definition, implications into the real-life applications may change as well.
+\protect\hyperlink{ref-Mehrabi2019Survey}{Mehrabi, Morstatter, Saxena, Lerman, \& Galstyan} (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) mark out that there exist different types of biases such as historical bias, representation bias or measurement bias (the list is long).
+This indicates how complex the issue of bias is.
+Without the proper understanding and awareness of the problem, people are prone to unconsciously sustain the bias.
+\protect\hyperlink{ref-Mehrabi2019Survey}{Mehrabi, Morstatter, Saxena, Lerman, \& Galstyan} (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) also distinguish different types of discrimination, some of them will be briefly described.
+By protected attributes we mean those qualities, traits or characteristics that one should not discriminate against.
+Direct discrimination occurs when protected attributes of individuals explicitly result in non-favorable outcomes toward them.
+Such discrimination takes place for example when one uses the information about someone's religious beliefs to evaluate whether they are competent for the work they apply for.
+In contrast, in indirect discrimination individuals appear to be treated equally, but anyway they end up being treated unjustly due to the hidden effects of biases towards their protected attributes.
+An example of such situation is when in a job advert there is a requirement of 10 years experience instead of the list of a specific type of experience and knowledge.
+This job advert can discriminate indirectly against young people who can have the required skills, but not yet so much work experience.
+Systemic discrimination takes place when policies, customs or behaviors that result from certain culture or organizational structure lead to discrimination against some groups of people.
+Finally, statistical discrimination consists in relying on group statistics to judge person belonging to a given group.
+The topic of discrimination is entangled with another concept; fairness.
+It is essential to grasp some concepts of fairness to take them into consideration while designing implementation of some machine learning model.
+In \protect\hyperlink{ref-Mehrabi2019Survey}{Mehrabi, Morstatter, Saxena, Lerman, \& Galstyan} (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) one may notice that depending on the context and application different definitions may be applied.
+
+There is a considerable amount of literature available on the topic of bias detection and mitigation in NLP models.
+\protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) focus on gender biases that may be observable while investigating the representation of job occupations and gender in terms of their assigned numerical
+values.
+The authors apply cosine similarity measurement to investigate the phenomenon where (the vectors corresponding to) words related to jobs that are stereotypically associated with a given gender are in fact in the model situated closer to this gender.
+They also use analogy tasks to evaluate if bias is present in a word embedding model.
+They check analogies by comparing pairs of word vectors.
+For example they search for the word complementing the puzzle: man is to doctor as woman is to \ldots?
+First they subtract the word ``man'' from the word ``woman''\footnote{This is pointwise subtraction of vectors, but for the simplicity we will often talk about words while really meaning vectors.
+What is meant should we clear from the context.} and then
+they search for the ranked list of other words pairs that have similar vector difference. They also include in the formula a threshold to ensure that the resulting pairs could not be randomly picked.
+
+However, as \protect\hyperlink{ref-Nissim2019Fair}{Nissim, Noord, \& Goot} (\protect\hyperlink{ref-Nissim2019Fair}{2019}) points out, there are some limitations to this approach.
+In practice, most of analogies implementations do not return any input words.
+This means that it does not make sense to expect the algorithm to return the same profession for both woman and man.
+Therefore, this method seems to be limited in terms of bias detection.
+Other problems are related for example to the choice of pairs and words that are used to detect the presence of discrimination, as it is often subjective and without proper justification.
+Additionally, the choice of the parameter used in \protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) formula to ensure that word pairs are not picked by random, is also not justified and changing it drastically
+influences the results.
+
+\protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) touch upon the topic of biases regarding race and gender.
+They apply knowledge from well-known psychological studies such as Implicit Association Test to investigate the relation between human stereotypical thinking and model-learnt biases to discover a close relationship between these two.
+For the evaluation they use Word Embedding Association Test (WEAT) and the Word Embedding Factual Association Test (WEFAT).
+
+\protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) propose a novel way of using cosine similarity to verify if the assumed resemblance between protected words and harmful attributes exists.
+They investigate an approach that enables them to measure the bias for a class (such as gender, religion or race) and express the final result with a single metric.
+
+\protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) refer to Implicit Association Test (Greenwald et al., 1998) that measures the strength of associations between concepts or stereotypes by measuring human reaction time for special tasks.
+Humans naturally exhibit some biases which do not always cause social concern.
+One can imagine the intuitive associations between for example insects and flowers, and the feelings of pleasantness or unpleasantness.
+In general, people would rather associate flowers with feeling pleasant than insects, and this preference in a sense is a bias or
+prejudice in some direction.
+However, this type of preference does not cause an uproar and is a rather morally neutral case.
+This example is rather used to show that the methodology used in IAT makes sense.
+\protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) were able to obtain similar results to the ones that measured reaction time with the use of WEAT metric.
+Unfortunately, further studies discover other biases and prejudices that directly influence the quality of other people's lives and therefore they should be taken care of.
+
+The most common bias detection methods in the literature focus on comparing the similarity between words from protected groups and those that are considered to be stereotypical or harmful in some way.
+One can find in this group methods such as euclidean distance or cosine similarity (which is equivalent to dot product if the vectors are normalized).
+There are also other ways to detect the effects of biases.
+One can investigate the model performance on certain downstream tasks and validate if the model uses the gender or race information to complete the task.
+The currently used methods employing cosine similarity, often aggregate the similarity values often aggregated in a way that may lead to hasty conclusions.
+The averaging of values and the lack of attention to uncertainty involved in such aggregation may lead to the incomplete picture of the situation.
+
+In the paper we indicate how current methods used to detect biases in natural language models are limited, if we investigate them from the perspective of Bayesian analysis.
+Our research enhances the current way in which the bias detection is performed to make sure that it is methodologically valid.
+The key hypothesis is that greater understanding of data and bias implications can be achieved when Bayesian methods are applied to issue.
 
@@
 
+Placeholder
+
 \hypertarget{word-embeddings}{%
 \section{Word embeddings}\label{word-embeddings}}
-
-To understand what cosine similarity measurement is, one first needs to grasp the concept of translating words to a computer-readable form.
-In the field of natural language processsing there are two main types of words representation --- localist and distributed.
-One-hot encoding is an example of a method used to achieve a localist representation of words.
-Here each vector contains information only about a single data point, this is achieved by first mapping categorical values (words) to integers and then to each integers a binary vector is assigned which contains only 0s except for the index of the integer, which is assigned 1.
-An example of a localist representation is:
-
-\begin{longtable}[]{@{}llllll@{}}
-\toprule
-word & 1 & 2 & 3 & 4 & 5 \\
-\midrule
-\endhead
-woman & 1 & 0 & 0 & 0 & 0 \\
-man & 0 & 1 & 0 & 0 & 0 \\
-girl & 0 & 0 & 1 & 0 & 0 \\
-boy & 0 & 0 & 0 & 1 & 0 \\
-monarch & 0 & 0 & 0 & 0 & 1 \\
-\bottomrule
-\end{longtable}
-
-In the example above it is clear that the length of the vectors increases with the number of words in a vocabulary.
-It is not a very computationally efficient representation.
-It has other flaws as well.
-For example, it is unable to capture the resemblance between words appearing in similar contexts.
-
-In contrast to the localist representation, a distributed representation returns vectors that contain continuous values instead of discrete 1s and 0s.
-Word embeddings are a class of various techniques that allow one to represent words as distributed vectors.
-Such learned representations of text have certain properties.
-At least prima facie, they store similar (or at least co-occurring) words close to each other in a vector space.
-An example of distributed representation is:
-
-\begin{longtable}[]{@{}lllll@{}}
-\toprule
-word & 1 & 2 & 3 & 4 \\
-\midrule
-\endhead
-woman & 0.456 & 0.267 & 0.675 & 0.131 \\
-man & 0.451 & 0.897 & 0.472 & 0.088 \\
-girl & 0.604 & 0.262 & 0.414 & 0.706 \\
-boy & 0.279 & 0.172 & 0.475 & 0.010 \\
-monarch & 0.565 & 0.678 & 0.463 & 0.975 \\
-\bottomrule
-\end{longtable}
-
-One of the advantages of using a distributed representation is that one is able to represent an enormous number of concepts with a smaller number of units.
-It is also possible to better capture similarities as words of similar meanings can have similar numeric vectors.
-
-The numbers occurring in such representations are not random.
-They are learned in a process that uses a very shallow neural network.
-There are various types of techniques used for learning the vectors representations.
-One of the most straightforward ones is a skip-gram model.
-Given a word the model tries to predict its neighboring words from the sentence.
-The mathematics behind the process relies on the idea that the prediction concerns the conditional probability of the adjacent words.
-The algorithm tries to minimize the loss function, which penalizes the system for discrepancy with actual co-occurrence frequencies in the corpus.
-One can choose various parameters of the model, such as the window size that determines how many surrounding words the model should predict.
-After preparing such a fitted modeln=, one takes only the learned weights from a neural network, and uses them as vectors in a word embeddings representation.
-
-Word embeddings have many applications in natural language processing.
-They are handy in document search and information retrieval.
-They also play their part in improving automatic translations.
-Well learned word representations may also contribute to the improvement of sentiment analysis or spam detection.
 
@@
 
-Cosine similarity is often used as a method of finding out whether vector representations for two words suggest that they are similar or somehow connected.
-Cosine similarity is the cosine of the angle between two vectors: the result of dividing their inner product (dot product usually) by the product of their magnitudes.
-
-It is worth mentioning one more point concerning cosine similarity.
-After the vectors are normalized to have length equal to 1, inner product itself (often dot product) is used to measure the similarity.
-
-\begin{align} \tag{Sim}
-\mathsf{cosineSimilarity}(A,B) & = \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert}
-\end{align}
-Cosine similarity is considered a proper tool for this operation as its result has a clear connection to geometry and at least for a low number of dimensions may be easily interpreted.
-Using this scale, one can compare vector similarities in a fairly clear manner.
-When the vectors are aligned perpendicularly to each other, their similarity equals 0 (which is the same as the cosine of 90 degrees).
-This tells us that the similarity between the vectors is small.
-As the angle between vectors decreases, cosine similarity approaches one, which stands for the greatest similarity.
-
-One of the limitations of this measure is that it informs us only about similarities between vectors in terms of their orientation.
-However, it is often argued that in comparing words in terms of this metric, the magnitude of vectors may be treated as irrelevant, as the most important information pertains to direction.
-
-In what follows, it is important to distinguish between cosine similarity and cosine distance, defined as:
-\begin{align} \tag{Sim}
-\mathsf{cosineDistance}(A,B) &  = 1 - \mathsf{cosineSimilarity}(A,B)\\
- &  = 1 - \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert} \nonumber
-\end{align}
-
-The greater the similarity between two vectors, the smaller the distance between them.
-The cosine distance ranges between 0 and 2.
-If the vectors are in an opposite direction to each other, the cosine distance is 2.
-And if the vectors are extremely similar then the cosine distance is very close to 0.
-
-One should note that cosine distance is not exactly a distance measure while it does not meet triangle inequality requirements.
-The triangle inequality formula says that for any triangle, the sum of the lengths of any two sides must be greater than or equal to the length of the remaining side.
-As shown in \href{https://stats.stackexchange.com/questions/198080/proving-that-cosine-distance-function-defined-by-cosine-similarity-between-two-u}{Statexchange.com} in the case of cosine distance it would have to fulfill this equation 1+cos-sim(𝐴,𝐶) \textless{} cos-sim(𝐴,𝐵) + cos-sim(𝐵,𝐶).
-If one chooses specific unit vectors it is easy to demonstrate that the triangle inequality is not preserved.
-
 \hypertarget{cosine-distance-in-a-one-class-bias-detection}{%
 \section{Cosine distance in a one-class bias detection}\label{cosine-distance-in-a-one-class-bias-detection}}
-
-\protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) define similarity between words as the outcome inner product of their normalized vectors.
-They focus on examining what the geometry of word embedding is in regard to ``he'' and ``she'' words.
-In other words, whether the similarity between those concepts and other words reflects expected gender stereotypes.
-They test this hypothesis by investigating whether there is a connection between word embeddings representing certain professions and words referring to gender.
-They also evaluate whether automatically produced analogies between words reflect the stereotypes as well.
-
-Here the gender bias of a word \(w\) is understood as its projection on the gender direction \(\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})\) (the gender direction is the top principal compontent of ten gender pair difference vectors).
-The underlying idea is that no bias is present if non-explicitly gender words are in equal distance to both elements in all explicitly gender pairs.
-Given the (ideally) gender neutral words \(N\) and the gender direction \(g\) the direct direct gender bias is defined as the average distance of the words in \(N\) from \(g\) (\(c\) is a parameter determining how strict we want to be):
-\begin{align}
-\mathsf{directBias_c(N,g)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},g)\vert^c}{\vert N \vert }
-\end{align}
-
-A very vivid way to follow their method of arguing that bias in word embeddings is real is to plot the values of inner product of chosen words.
-The plot below does not originate from the original paper (it is from \url{https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings}) but similar visualization may be found there.
-Data used to create our plot is as follows.
-
-Occupations associated with feminine: \textbf{"homemaker", "nurse", "receptionist", "librarian", "socialite", "hairdresser", "nanny", "bookkeeper", "stylist", "housekeeper", "interior designer", "guidance counselor"}
-
-Occupations associated with masculine: \textbf{"maestro", "skipper", "protege", "philosopher", "captain", "architect", "financier", "warrior", "broadcaster", "magician", "fighter pilot", "boss"}
-
-\vspace{1mm}
-\footnotesize
-
-\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-1-1} \end{center}
-\normalsize
-
-The points in the plot above result from the calculation of the inner product of a chosen vector for a profession word and a vector for a gender word (she or he).
-Inner product of two vectors expresses similarity between words.
-This assumption originates from the geometry and properties of a vector space.
 
@@
 
-\protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) present a different approach towards finding similarities between classes of words.
-The authors claim that texts available online are full of direct or indirect human stereotypes.
-As a result, word embeddings are prone to learn and maybe amplify those biases and propagate them further into AI models that are used for various applications.
-Cosine distance is used in the article as a measure to first argue for the existence of multi-class bias and then to show how through bias mitigation techniques the bias may be decreased.
-
-They modify WEAT to a multi-class setting, introducing Mean Average Cosine similarity as a measure of bias (in fact, in the paper they report distances rather than similarities).
-Let \(T = \{t_1, \dots, t_k\}\) be a class of protected word embeddings, and let each \(A_j\in A\) be a set of attributes stereotypically associated with a protected word).
-Then:
-\begin{align}
-S(t_i, A_j) & = \frac{1}{\vert A_j\vert}\sum_{a\in A_j}\mathsf{cos}(t,a) \\
-MAC(T,A) & = \frac{1}{\vert T \vert \,\vert A\vert}\sum_{t_i \in T }\sum_{A_j \in A} S(t_i,A_j)
-\end{align}
-
-As the code has been provided by the authors we were able to reconstruct their results.
-The main steps in the procedure are as follows.
-Let us go through an example that refers to the process of hard debiasing on religious attributes.
-
-First we load word embeddings from reddit.US.txt.tok.clean.cleanedforw2v.w2v dataset.
-The word embeddings have only 50 dimensions and the number of individual words from the dataset is 44895.
-The authors assume that the protected group should ideally not have high cosine similarity to stereotypical words.
-The word embeddings geometry should not place this group close to harmful stereotypes, if it is to be bias-free.
-For instance, let's look at the religion-related words.
-
-Protected words by religion type:
-
-\textbf{"jew"       : ["judaism", "jew", "synagogue", "torah", "rabbi"]}
-
-\textbf{"christian" : ["christianity", "christian", "church", "bible", "priest"]}
-
-\textbf{"muslim"    : ["islam", "muslim", "mosque", "quran", "imam"]}
-
-Stereotypical words by religion type:
-
-\textbf{"jew"       : ["greedy", "cheap", "hairy", "liberal"]}
-
-\textbf{"christian"       : ["judgemental", "conservative", "familial"]}
-
-\textbf{"muslim"       : ["violent", "terrorist", "dirty", "uneducated"]}
-
-We have prepared a table presenting the values of cosine distance for each protected word with each attribute (stereotype). The part of the results is shown below.
-
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{religion }\OtherTok{\textless{}{-}} \FunctionTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{]}
-\FunctionTok{colnames}\NormalTok{(religion) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
-                        \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
-\NormalTok{religion}\SpecialCharTok{$}\NormalTok{wordClass }\OtherTok{\textless{}{-}} \FunctionTok{as.factor}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{wordClass)}
-\FunctionTok{levels}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{wordClass) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
-\FunctionTok{head}\NormalTok{(religion)  }\SpecialCharTok{\%\textgreater{}\%}  \FunctionTok{kable}\NormalTok{(}\AttributeTok{format =} \StringTok{"latex"}\NormalTok{,}\AttributeTok{booktabs=}\NormalTok{T,}
-                      \AttributeTok{linesep =} \StringTok{""}\NormalTok{,  }\AttributeTok{escape =} \ConstantTok{FALSE}\NormalTok{, }
-                      \AttributeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
-                      \FunctionTok{kable\_styling}\NormalTok{(}\AttributeTok{latex\_options=}\FunctionTok{c}\NormalTok{(}\StringTok{"scale\_down"}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{table}
-
-\caption{\label{tab:religionTableHeadEarly}Head of the religion dataset.}
-\centering
-\resizebox{\linewidth}{!}{
-\begin{tabular}[t]{lllrrl}
-\toprule
-protectedWord & wordToCompare & wordClass & cosineDistance & cosineSimilarity & connection\\
-\midrule
-judaism & violent & muslim & 0.7141939 & 0.2858061 & different\\
-judaism & terrorist & muslim & 0.7461333 & 0.2538667 & different\\
-judaism & dirty & muslim & 1.2002599 & -0.2002599 & different\\
-judaism & uneducated & muslim & 0.7885469 & 0.2114531 & different\\
-judaism & greedy & jewish & 1.0026172 & -0.0026172 & associated\\
-judaism & cheap & jewish & 1.2323229 & -0.2323229 & associated\\
-\bottomrule
-\end{tabular}}
-\end{table}
-\normalsize
-
-\pagebreak
-
-In the article there was no analysis of individual distances but the general look at the data through the usage of mean.
-The authors introduced a metric that tries to generalize the presence of bias through the classification of multi-class bias in groups of words connected with gender, religion or race.
-In the process they first take the mean of cosine distances between a given protected word and attributes assigned to each stereotype.
-They do not differentiate between stereotypes associated with a word and stereotypes associated with different words (in the case of religion, stereotypes characteristic for Christianity has also cosine distance measured with for instance, Judaism or Islam).
-Then, after collecting the list of mean cosine distances, they average the list to obtain one final value representing the whole group, in this example religion, for which the final mean of all mean distances is equal to 0.859.
-
-In the article the authors also try to remove previously defined biases from word embedding.
-First they identify the bias subspace using Principal Component Analysis (PCA) which is a technique for dimensionality reduction.
-It is applied here to choose the subspace that contains the greatest amount of information.
-There can be many subspaces found in a given group, for example in terms of religion one can identify at least a few sets that are to grasp the concept of religion in general:
-
-\textbf{["judaism", "christianity", "islam"]}
-
-\textbf{["jew", "christian", "muslim"]}
-
-\textbf{["synagogue", "church", "mosque"]}
-
-The idea is to find a set that provides enough information to create from it a vector representing the concept of religion among words.
-This strategy is based on the idea that different dimensions of vectors contain different types of information and in some words in vector layers (subspaces) the information about religiousness is implicitly conveyed.
-In some cases this knowledge is useful but in the case of harmful stereotypes one does not want to include the concept of religion in stereotypical words.
-
-After finding the bias subspace, they use it to modify the vector values individually so that their cosine distances towards certain words are changed.
-In the case of stereotypes the aim is to make the cosine distances larger so that the association between protected word and harmful stereotype is smaller.
-
-In the final step there is an evaluation of the results.
-The cosine distances are calculated again but this time using debiased vocabulary.
-After taking the mean of all distances one final value is obtained and then it is compared with the average value from the beginning.
-If the cosine distances are on average greater than before then it leads the authors to the conclusion that improvement has been achieved.
-As the cosine distance increases it is assumed that the association between protected and stereotypical words decreases.
-
 \hypertarget{limitations-of-the-approach}{%
 \section{Limitations of the approach}\label{limitations-of-the-approach}}
-
-\textbf{1. Selection of attributes}
-
-The attributes are taken from different sources, there is no principled justification for their choice. From our analysis it will become clear that
-the list is rather uneven.
-
-There is no mention of methodology for deciding on the number of attributes necessary to decide a hypothesis on the given size of dataset.
-There are however some ways to estimate how many samples we need to make sure that the result is significant.
-Our research will show that the numbers used
-are rather insufficient.
-
-\textbf{2. No control while taking the mean of cosines}
-
-The authors use the mean average cosine similarity to check on multi-class similarity between protected word and harmful stereotypes.
-They average the results until they obtain one final value to represent the mean cosine distance between protected word from a given class and the attributes of that class.
-
-\textbf{3. Hiding the impact of uncertainty}
-
-A mean hides issue and as there are pairs having negative and small similarities and there are those that have similarity equal to 0.5, the resulting calculation seems to be in norm.
-Additionally in such method the uncertainty is also not included which makes it even more difficult to give reasonable interpretations of the results.
-We propose the use of Bayesian method to obtain some understanding of the influence the uncertainty has on the interpretation of final results.
-
-\textbf{4. No word class distinction}
-
-In the original paper words from all three religions were checked with all of the stereotypes which means that there was no distinction between classes in which the stereotype is associated with a given religion, as opposed to the situation in which it is associated with another one.
-Not all of the stereotypical words have to be considered as harmful for all of the religions.
-In our analysis we distinguished between stereotypes associated with a given group, stereotype associated with different groups and control groups; neutral words and stereotypes-free human predicates.
-
-\textbf{5. Interpreting the results}
-
-Assuming for a moment that the value of multi-class cosine distance is correct, one may question the interpretation.
-\protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) summarize the averages of cosine distance per group (gender, race, religion).
-We would like to focus now on analyzing the values relating to religious biases.
-Here is the relevant fragment of table:
-
-\begin{longtable}[]{@{}ll@{}}
-\toprule
-Religion Debiasing & MAC \\
-\midrule
-\endhead
-Biased & 0.859 \\
-Hard Debiased & 0.934 \\
-Soft Debiased (\(\lambda\) = 0.2) & 0.894 \\
-\bottomrule
-\end{longtable}
-
-MAC stand for mean average cosine similarity, although in reality the values of cosine distance are given.
-What may attract attention is the fact that the value of cosine distance in ``Biased'' category is already quite high even before debiasing.
-High cosine distance indicates low cosine similarity between values.
-One could think that average cosine similarity equal to approximately 0.141 is not significant enough to consider it as bias.
-However the authors aim to mitigate ``biases'' in vectors with such great distance to make it somewhat larger.
-Methodologically the question is, on what basis is this small similarity still considered as a proof of a bias presence, and whether these small changes are meaningful.
-This is in general the problem of scale and the lack of universal intervals.
-Proper intervals could be used to decide whether given cosine similarity is high enough to consider the words to be more similar than if we choose them at random.
-
-\textbf{6. The curse of dimensionality}
-
-In our case, the curse of dimensionality may take place when there is an increase in volume of data that results in adding extra dimensions to the Euclidean space. According to the article
-\href{https://analyticsindiamag.com/curse-of-dimensionality-and-what-beginners-should-do-to-overcome-it/}{Analyticsindiamag Curse of dimensionality} as the number of features increases, it may be harder and harder to obtain useful information from the data with the usage of available algorithms.
-One may notice that more data should contribute to greater amount of information but more information also means greater risk of noise and distractions in data.
-At the same time, many times modern solutions are adapted to smaller dimensions and their results in higher ones are not intuitive or may be prone to be mistaken.
-
-Using cosine similarity in high dimensions in word embeddings may also be prone to the curse of dimensionality.
-According to \protect\hyperlink{ref-Venkat2018Curse}{Venkat} (\protect\hyperlink{ref-Venkat2018Curse}{2018}) there are reasons to consider this phenomenon when searching for word similarities in higher dimensions.
-An experiment is conducted that aims at showing how the similarity values and variation change as the number of dimensions increases.
-The hypothesis made in the paper states that two things will happen as the number of dimensions increase.
-First, the effort required to measure cosine similarity will be greater, and two, the similarity between data will blur out and have less variation.
-The authors generate random points with increasing number of dimensions where each dimension of a data point is given a value between 0 and 1.
-Then they pick one vector at random from each dimension class and calculate the cosine similarity between the chosen vector and the rest of the data.
-Then they check how the variation of values changes as the number of dimensions increases.
-It seems like the more dimensions there are, the smaller the variance and therefore it is less obvious how to interpret the resulting cosine similarities.
-Maybe the scale should be adjusted to the number of dimensions and variance so that it still gives us sensible information about data.
-According to some researches the cosine similarity in high dimensions is not reliable enough to trust it as it may be the case that choosing words on random may result in getting similar values as when picking them consciously.
-
-\begin{figure}
-\centering
-\includegraphics{../images/curseOfDimensionality.png}
-\caption{curse of dimensionality, number of dimensions on the x axis, standard deviation of similarity on the y axis}
-\end{figure}
 
@@
 
-\leavevmode\hypertarget{ref-Venkat2018Curse}{}%
-Venkat, N. (2018). The curse of dimensionality: Inside out.
+\leavevmode\hypertarget{ref-Mehrabi2019Survey}{}%
+Mehrabi, N., Morstatter, F., Saxena, N., Lerman, K., \& Galstyan, A.
+(2019).
+A survey on bias and fairness in machine learning.
+\emph{CoRR}, \emph{abs/1908.09635}.
+Retrieved from \url{http://arxiv.org/abs/1908.09635}
+
+\leavevmode\hypertarget{ref-Nissim2019Fair}{}%
+Nissim, M., Noord, R.
+van, \& Goot, R.
+van der.
+(2019).
+Fair is better than sensational: Man is to doctor as woman is to doctor.
+\emph{CoRR}, \emph{abs/1905.09866}.
+Retrieved from \url{http://arxiv.org/abs/1905.09866}
 

### 2021-06-19 thesis/01-intro.Rmd
Adds remarks for chapter 2

@@
 It turns out that in the learning process these models can learn implicit biases that reflect harmful stereotypical thinking still present in modern societies.
-One can find methods that aim at identifying and measuring hidden biases and/or try to remove them by modifying the models.
 There are many different types of models in NLP depending on a task that they are supposed to solve.
-However, all of them need as an input words represented by means of numbers and this is accomplished with word embedding models.
-The models usually assign the values based on the context in which the words appear.
-This means that the input data can have enormous influence on the outcome.
+However, all of them need as an input words represented by means of numbers and this is accomplished with word embeddings.
+The word embeddings are usually assigned the values based on the context in which the words appear.
+This means that the input data can have an enormous influence on the outcome.
 The biases seem to have their primary source in the way the words are assigned numerical values.
+One can find methods that aim at identifying and measuring hidden biases and/or try to remove them by for example modifying the word embeddings directly.
 

### 2021-06-19 thesis/02-cosine.Rmd
Adds remarks for chapter 2

@@
 One-hot encoding is an example of a method used to achieve a localist representation of words.
-Here each vector contains information only about a single data point, this is achieved by first mapping categorical values (words) to integers and then to each integers a binary vector is assigned which contains only 0s except for the index of the integer, which is assigned 1.
+Here, each vector contains information only about a single data point.
+This is achieved by first mapping categorical values (words) to integers and then to each integers a binary vector is assigned which contains only 0s except for the index of the integer, which is assigned 1.
 An example of a localist representation is:
@@
 One can choose various parameters of the model, such as the window size that determines how many surrounding words the model should predict.
-After preparing such a fitted modeln=, one takes only the learned weights from a neural network, and uses them as vectors in a word embeddings representation.
+After preparing such a fitted model, one takes only the learned weights from a neural network, and uses them as vectors in a word embeddings representation.
 
@@
 It is worth mentioning one more point concerning cosine similarity.
-After the vectors are normalized to have length equal to 1, inner product itself (often dot product) is used to measure the similarity.
+After the vectors are normalized to have length equal to 1, the inner product itself (often dot product) is used to measure the similarity, because it then equals the cosine similarity.
 
@@
 As the angle between vectors decreases, cosine similarity approaches one, which stands for the greatest similarity.
+It is also possible to obtain negative cosine similarity, if the value approaches -1 it means that the words are contrary to each other.
 
@@
 The cosine distance ranges between 0 and 2.
-If the vectors are in an opposite direction to each other, the cosine distance is 2.
+If the vectors are in an opposite directions, the cosine distance is 2.
 And if the vectors are extremely similar then the cosine distance is very close to 0.
 
-One should note that cosine distance is not exactly a distance measure while it does not meet triangle inequality requirements.
+One should note that cosine distance is not exactly a distance measure, as it does not meet triangle inequality requirements.
 The triangle inequality formula says that for any triangle, the sum of the lengths of any two sides must be greater than or equal to the length of the remaining side.
-As shown in [Statexchange.com](https://stats.stackexchange.com/questions/198080/proving-that-cosine-distance-function-defined-by-cosine-similarity-between-two-u) in the case of cosine distance it would have to fulfill this equation 1+cos-sim(𝐴,𝐶) < cos-sim(𝐴,𝐵) + cos-sim(𝐵,𝐶).
+As shown in [discussion in stats.stackexchange.com](https://stats.stackexchange.com/questions/198080/proving-that-cosine-distance-function-defined-by-cosine-similarity-between-two-u) in the case of cosine distance it would have to fulfill this equation 1+cos-sim(𝐴,𝐶) < cos-sim(𝐴,𝐵) + cos-sim(𝐵,𝐶).
 If one chooses specific unit vectors it is easy to demonstrate that the triangle inequality is not preserved.
@@
 A very vivid way to follow their method of arguing that bias in word embeddings is real is to plot the values of inner product of chosen words.
-The plot below does not originate from the original paper (it is from https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings) but similar visualization may be found there.
+The plot below does not originate from the original paper but from [kaggle.com R notebook using data from GloVe: Global Vectors for Word Representation](https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings).
+However, similar visualization may be found there.
 Data used to create our plot is as follows.
@@
 This assumption originates from the geometry and properties of a vector space.
+In the picture one may observe a correlation between the occupation and gender.
+Stereotypical male professions are closer to the "he" word and stereotypical female professions have greater similarity with "she" word.
+One could argue that this is a hint that there is some hidden bias information stored in the words vectors.
 
@@
 
-We have prepared a table presenting the values of cosine distance for each protected word with each attribute (stereotype). The part of the results is shown below.
+We have prepared a table presenting the values of cosine distance for each protected word with each attribute (stereotype). The part of the results is shown in Table 2.3
 
@@
 
-In the article there was no analysis of individual distances but the general look at the data through the usage of mean.
-The authors introduced a metric that tries to generalize the presence of bias through the classification of multi-class bias in groups of words connected with gender, religion or race.
+In the article there was no analysis of individual distances, but the general look at the means.
+The authors introduced a metric that tries to generalize WEAT in measuring the presence of bias through the classification of multi-class bias in groups of words connected with gender, religion or race.
 In the process they first take the mean of cosine distances between a given protected word and attributes assigned to each stereotype.
@@
 
-In the article the authors also try to remove previously defined biases from word embedding.
+In the article the authors also try to remove biases, as previously defined, from a word embedding.
 First they identify the bias subspace using Principal Component Analysis (PCA) which is a technique for dimensionality reduction.
 It is applied here to choose the subspace that contains the greatest amount of information.
-There can be many subspaces found in a given group, for example in terms of religion one can identify at least a few sets that are to grasp the concept of religion in general:
+There can be many subspaces found in a given group.
+For example in terms of religion one can identify at least a few sets that are to grasp the concept of religion in general:
 
@@
 This strategy is based on the idea that different dimensions of vectors contain different types of information and in some words in vector layers (subspaces) the information about religiousness is implicitly conveyed.
-In some cases this knowledge is useful but in the case of harmful stereotypes one does not want to include the concept of religion in stereotypical words.
+In some cases this knowledge is useful, but in the case of harmful stereotypes one does not want to include the concept of religion in stereotypical words.
 
 After finding the bias subspace, they use it to modify the vector values individually so that their cosine distances towards certain words are changed.
-In the case of stereotypes the aim is to make the cosine distances larger so that the association between protected word and harmful stereotype is smaller.
+In the case of stereotypes the aim is to make the cosine distances larger so that the association between protected word and stereotype is smaller.
 
@@
 If the cosine distances are on average greater than before then it leads the authors to the conclusion that improvement has been achieved.
-As the cosine distance increases it is assumed that the association between protected and stereotypical words decreases.
+As the cosine distance increases, it is assumed that the association between protected and stereotypical words decreases.
 
@@
 
-The attributes are taken from different sources, there is no principled justification for their choice. From our analysis it will become clear that
-the list is rather uneven.
+The attributes are taken from different sources, there is no principled justification for their choice. From our analysis it will become clear that the list is rather uneven.
 
 There is no mention of methodology for deciding on the number of attributes necessary to decide a hypothesis on the given size of dataset.
-There are however some ways to estimate how many samples we need to make sure that the result is significant.
+There are however some ways to estimate sample sizes needed to ensure that the results are meaningful if the effect is present.
 Our research will show that the numbers used
@@
 
-\textbf{2. No control while taking the mean of cosines}
+\textbf{2. No attention to distributions and details}
 
-The authors use the mean average cosine similarity to check on multi-class similarity between protected word and harmful stereotypes. 
+The authors use the mean of mean average cosine similarities to measure multi-class similarity between protected word and harmful stereotypes.
 They average the results until they obtain one final value to represent the mean cosine distance between protected word from a given class and the attributes of that class.
+If one takes closer look at the individual values that are taken for the calculations it turns out that there is a bunch of outliers and surprisingly dissimilar words.
+We approach this issue by providing the tables and visualizations of individual cosine distances to make sure that we get the proper insight into the data.
 
@@
 
-A mean hides issue and as there are pairs having negative and small similarities and there are those that have similarity equal to 0.5, the resulting calculation seems to be in norm.
-Additionally in such method the uncertainty is also not included which makes it even more difficult to give reasonable interpretations of the results.
+With such a method the uncertainty is not included which makes it even more difficult to give reasonable interpretations of the results.
 We propose the use of Bayesian method to obtain some understanding of the influence the uncertainty has on the interpretation of final results.
@@
 
-In the original paper words from all three religions were checked with all of the stereotypes which means that there was no distinction between classes in which the stereotype is associated with a given religion, as opposed to the situation in which it is associated with another one.
+In the original paper, words from all three religions were compared against all of the stereotypes, which means that there was no distinction between cases in which the stereotype is associated with a given religion, as opposed to the situation in which it is associated with another one.
 Not all of the stereotypical words have to be considered as harmful for all of the religions.
-In our analysis we distinguished between stereotypes associated with a given group, stereotype associated with different groups and control groups; neutral words and stereotypes-free human predicates.
+In our analysis, we distinguish between stereotypes associated with a given group, stereotypes associated with different groups, and control groups: neutral words and stereotypes-free human predicates.
 
@@
 @Manzini2019blackToCriminal summarize the averages of cosine distance per group (gender, race, religion).
-We would like to focus now on analyzing the values relating to religious biases.
+For now let us focus now on analyzing the values relating to religious biases.
 Here is the relevant fragment of table:
@@
 
-MAC stand for mean average cosine similarity, although in reality the values of cosine distance are given.
+MAC stand for mean average cosine similarity, although in reality the the table contains mean distances.
 What may attract attention is the fact that the value of cosine distance in "Biased" category is already quite high even before debiasing.
@@
 One could think that average cosine similarity equal to approximately 0.141 is not significant enough to consider it as bias.
-However the authors aim to mitigate "biases" in vectors with such great distance to make it somewhat larger.
-Methodologically the question is, on what basis is this small similarity still considered as a proof of a bias presence, and whether these small changes are meaningful.
+However the authors aim to mitigate "biases" in vectors with such great distance to make it even larger.
+Methodologically the question is, on what basis is this small similarity still considered as a proof of the presence of bias, and whether these small changes are meaningful.
 This is in general the problem of scale and the lack of universal intervals.
-Proper intervals could be used to decide whether given cosine similarity is high enough to consider the words to be more similar than if we choose them at random.
+In contrast, proper intervals decide whether given cosine similarity is high enough to consider the words to be more similar than if we chose them at random.
+We will use highest posterior density intervals.
 
@@
 
-In our case, the curse of dimensionality may take place when there is an increase in volume of data that results in adding extra dimensions to the Euclidean space. According to the article 
-[Analyticsindiamag Curse of dimensionality](https://analyticsindiamag.com/curse-of-dimensionality-and-what-beginners-should-do-to-overcome-it/) as the number of features increases, it may be harder and harder to obtain useful information from the data with the usage of available algorithms.
-One may notice that more data should contribute to greater amount of information but more information also means greater risk of noise and distractions in data.
-At the same time, many times modern solutions are adapted to smaller dimensions and their results in higher ones are not intuitive or may be prone to be mistaken.
+In our case, the curse of dimensionality may take place when there is an increase in the volume of data that results in adding extra dimensions to the Euclidean space. According to the article 
+[Curse of dimensionality at analyticsindiamag.com](https://analyticsindiamag.com/curse-of-dimensionality-and-what-beginners-should-do-to-overcome-it/) as the number of features increases, it may be harder and harder to obtain useful information from the data using the available algorithms.
+One may notice that more data should contribute to greater amount of information, but more information also means greater risk of noise and distractions in data.
+At the same time, modern solutions are often adapted to smaller dimensions and their results in higher ones are not intuitive, or may be prone to error.
 
@@
 Maybe the scale should be adjusted to the number of dimensions and variance so that it still gives us sensible information about data.
-According to some researches the cosine similarity in high dimensions is not reliable enough to trust it as it may be the case that choosing words on random may result in getting similar values as when picking them consciously.
+Yet, according to some articles cosine similarity in high dimensions is not reliable enough as it may be the case that choosing words at random may result in similar values as when picking them consciously.
 

### 2021-06-19 thesis/04-discussion.Rmd
Develop Discussion chapter
Adds more to discussion chapter

--- 2021-06-16 thesis/04-discussion.Rmd
@@
 ---
-# Discussion
+# Discussion and summary
 
-It is worth diving deeper into some observations that arise while analyzing the results. 
-<!-- Looking into specific values -->
-As we pointed out, if there is no control of what cosine distance values individual word pairs have, then there is less
-understanding of the exact relations between words.
-If the protected words that we chose do not have high similarity with harmful stereotypes, then one should consider at least two scenarios.
+<!-- Bayesian method -->
+
+We propose the use of Bayesian methods to measure uncertainty in bias detection.
+There are a few advantages of this method.
+Including uncertainty enables one to directly observe the influence of sample sizes.
+Analyzing individual words and connection coefficients, one may notice how `neutral` words have smaller uncertainty intervals and `different` or `associated` quite the opposite.
+One of the reasons for such outcome is that we used approximately 230 neutral words and only between 11-25 (the number varies from class to class) stereotypical attributes from @Manzini2019blackToCriminal article.
+What is more, we also pay attention to distribution and details regarding anomalous values.
+With the use of simple visualizations that we introduced before, we were able to indicate suspicious cosine distance values.
+Additionally, we compare in details how the cosine distance values and uncertainty change after the debiasing.
+One can verify then how the individual vectors changed and if it what was expected.
+Our analysis with the use of Bayesian method gave us new ideas and hypothesis concerning not only the bias detection method but also the efficiency of debiasing itself.
+
+<!-- General insights - estimated coefficients mean -->
+
+We created a summary table for each of the datasets: Reddit, Reddit Debiased, and Google word embeddings.
+Let us first analyse the general observations from estimated coefficients mean introduced in 3.6.
+DATASET-LEVEL COEFFICIENTS.
+For Google embeddings the HPDI for all classes coefficients (associated, different, human, and none) has an interval that includes zero.
+This can lead one to a conclusion that the impact for associated, different, human and neutral attribute is, when averaged, quite similar.
+This indicates how including the uncertainty may change the use and interpretation of @Manzini2019blackToCriminal MAC metric.
+It seems that if one focuses only on differences between means of means, it is too simplistic.
+In case of Reddit word embeddings the situation is similar although HPDI interval is below 0 for Gender class when looking at `associated` and `different` mean coefficient.
+This can suggest that there is indeed slightly stronger impact of these attributes in cosine distance being smaller.
+One should also notice how `associated` and `different` coefficients have quite similar HPDI interval, the highest observed absolute difference is equal to only 0.1.
+This suggests that again the impact of associated attributes and difference ones is not clear at first sight when looking at averaged coefficients.
+Finally, let's compare the HPDI intervals for Reddit and Reddit debiased datasets.
+For Religion and Race dataset there is a minor shift (in absolute values the highest change is equal to approximately 0.1) of the mean coefficients towards zero.
+However for Gender dataset there is no significant change.
+This is of course the general look at the data, let's now analyze individual words.
+
+<!-- Details from the table -->
+
+In Reddit table one can observe that for Religion class the cosine distance results for the associated attributes are for approximately 60% of the words close to neutral attributes as well.
+This can suggest that in some cases words concerning humans can have higher similarity with some protected words independently if they are stereotypical or neutral-human words.
+One should also pay attention to the fact that for all of the `associated` and `different` attributes in Religion class the uncertainty interval overlaps at some point.
+What is even more surprising is that for protected words, such as `torah`  `associated` attribute has the the cosine distance slightly over 1, which means no positive similarity!
+If the protected words that we chose do not have high similarity with harmful associated attributes, then one should consider at least three scenarios.
 The first one is that the choice of protected words and attributes may be corrupted.
 The second one is that the metric is not able to catch the hidden bias properly.
-In both cases it is essential to take a look at the individual values before averaging them or aggregating in other ways.
+The third one is that there is actually no bias between the words.
+Regardless of which scenario one considers, it is essential to take a look at the individual values before averaging them or aggregating in other ways.
+It seems that using Bayesian method can enhance the process of verifying the hypotheses concerning the choice of protected words and attributes.
 
-<!-- Problem with gender words --> 
-One has to also remember about the specific situation present in gender dataset.
-Surprisingly one could observe there high cosine distances values between some female stereotypical professions and male protected words.
-This leads to new ideas of how to understand the bias origins and its proper detection.
-To start with, cosine distance seems to catch the information regarding words co-occurrence and not the semantic similarity.
-If a word stereotypically associated with females has high cosine distance value for male protected words, then the metric is not proper to establish the bias presence as it can be misleading.
+Surprisingly in Gender class one can observe high cosine similarity values between some female stereotypical professions and male protected words.
+If a word stereotypically associated with females has low cosine distance to male protected words, then one should try to figure out the reasons for that.
+Cosine distance seems to capture the information on the co-occurrence of words and not on the semantic similarity strictly speaking.
+
+In Google table ..
+
+<!-- comments and values that refer to them: (blue and green) -->
+<!-- a) google scale -->
+<!-- b) associated females -->
+
+In Reddit debiased ..
+
+<!-- debiased: question about the change -->
+<!-- - better but not for muslims -->
+<!-- - corpus frequency maybe? -->
+<!-- - better for male -->
+<!-- - when in gender different and associated together maybe it is better - gender is very specific, words are about professions;  -->
+<!-- - when in religion it is bad different and associated -->
+<!-- -  in race it is not obvious - mixed sentiments  -->
+
+Let's summarize the results of the bias detection methods analysis.
+
+<!-- black remarks then: -->
+<!-- - mess, change protected words; relative differences stay the same, by the way lipstick on a pig -->
+<!-- - word lists are strange, not enough words, large HPDI -->
+
+<!-- where to put this ? -->
+
+<!-- summary -->
+
+The bottom line is that if we want to take bias seriously, so should we approach the uncertainty involved in our estimations.
+There is no replacement for proper statistical evaluation that does not discard information about the uncertaintly involved, larger word lists are needed, and visualisation of the results for particular protected classes provides much better guidance than chasing a single metric based on a means of means.
+
 What we mean is that in same cases it can unnecessarily label words as biased while omitting other more harmful associations.
@@
 
-<!-- Baysian method --> 
-We propose the use of Bayesian method to introduce the uncertainty measure in bias detection.
-There are a few advantages when including this method in the bias analysis.
-As noted before in our example with religion dataset, one may obtain new insight into the bias issue after analyzing the mean estimates of the coeﬀicients.
-There seems to be a minor difference for the mean concerning the `assosiated` class.
-This can lead to the conclusions that we do not have strong reasons to perceive lower cosine distance as significantly related to the fact that the attribute is associated.
+<!--
 
-One may dive deeper into this idea by investigating the connection coefficients for individual words with the use of the visualization.
-There one observes that indeed most of the words are clustered together independently of the class that they belong to.
-It is of course partly due to the fact that we include uncertainty now.
-This example shows how understanding of the bias measurement may differ when switching to Bayesian method.
-One could argue that after this analysis one obtains more information on the dataset and can pursue with the further research more cautiously.
-
-# Summary
-
-The aim of this paper was to show the analysis of bias detection methods used in modern natural language processing projects.
-
-In the further research it is worth to dive deeper into some issues or to examine new ideas that occurred during our work.
-
-<!-- 
-remarks:
-- delete papers from github/hide them
-- should we quote "debiasing" or without quoting
-
+Articles:
+  1. "Are We Consistently Biased? Multidimensional Analysis of Biases in Distributional Word Vectors"
+  https://www.aclweb.org/anthology/S19-1010.pdf 
+  2. Summary what causes bias short article
+  [source https://kawine.github.io/blog/nlp/2019/09/23/bias.html]
+  3. Men Also Like Shopping: Reducing Gender Bias Amplification using Corpus-level Constraint
+  4. Bias in Bios: A Case Study of Semantic Representation Bias in a High-Stakes Setting
   -->
-
-<!-- textbf{Control group}
-
-How to properly prepare control group in terms of quality and quantity?
-<!-- + check if it may be proved in the code -->
-
-<!--  \textbf{Verifying other similarity meaures}
-
-Besides cosine similarity, there are other methods used to find the similarity between vectors.
-
-a) TS-SS 
-todo: check if it makes sense to calculate for all of the datasets; interpretations etc.
-https://www.researchgate.net/publication/303513110_A_Hybrid_Geometric_Approach_for_Measuring_Similarity_Level_Among_Documents_and_Document_Clustering
-
-b) WEAT test 
-<!-- https://docs.responsibly.ai/word-embedding-bias.html#module-responsibly.we.weat -->
-
-<!-- [to check https://github.com/taki0112/Vector_Similarity] -->
-
-<!-- TODO: check in code if they make more sense -->
-
-<!-- ## Bibliography
-- https://www.amazon.com/Language-Processing-Synthesis-Lectures-Technologies/dp/1627052984/ref=as_li_ss_tl?ie=UTF8&qid=1502062931&sr=8-1&keywords=Neural+Network+Methods+in+Natural+Language+Processing&linkCode=sl1&tag=inspiredalgor-20&linkId=d63df073fea3ebe2d405820570b3ff03 ->
-
-<!-- ## *** Additional -->
-
-<!-- ### Man is to Computer Programmer as Woman is to Homemaker? Debiasing Word Embeddings" -->
-<!-- [article https://arxiv.org/pdf/1607.06520.pdf] -->
-
-<!-- Assumptions: -->
-<!-- "We assume there is a set of gender neutral words $N\subseteq W$ such as flight attendant or shoes, which, by definition, are not specific to any gender." -->
-
-<!-- TODO: check if it is useful for the analysis of black is .. -->
-
-<!-- ### Identifying the bias subspace  -->
-
-<!-- Interesting analysis about limits o subspace and research [source https://kawine.github.io/blog/nlp/2019/09/23/bias.html] -->
-
-<!-- Resources to analyze: -->
-<!-- 1. https://arxiv.org/pdf/2005.00965.pdf -->
-<!-- 2. https://arxiv.org/pdf/2009.09435.pdf -->
-<!-- 3. https://arxiv.org/pdf/1904.04047.pdf -->
-<!-- 4. https://www.aclweb.org/anthology/P19-1166.pdf -->
-
-<!-- ### How is cosine similarity entangled with analogy topic in NLP (is it??) -->
-
-<!-- article:  https://arxiv.org/pdf/1905.09866.pdf -->
-
-<!-- Critique of the usage of analogies as a proof for gender biases -->
-
-<!-- TODO: check if there is other argument than that algorithm does not allow it -->
-
-<!-- general thoughts 
-https://www.researchgate.net/publication/349408535_Robustness_and_Reliability_of_Gender_Bias_Assessment_in_Word_Embeddings_The_Role_of_Base_Pairs
-
--
--
--
--->

### 2021-06-19 thesis/_book/_main.tex
Adds more to discussion chapter

--- 2021-06-19 thesis/_book/_main.tex
@@
 \author{Alicja Dobrzeniecka}
-\date{2021-06-17}
+\date{2021-06-19}
 
@@
 \FunctionTok{library}\NormalTok{(tidyverse)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## -- Attaching packages --------------------------------------- tidyverse 1.3.1 --
+\end{verbatim}
+
+\begin{verbatim}
+## v ggplot2 3.3.3     v purrr   0.3.4
+## v tibble  3.1.2     v dplyr   1.0.6
+## v tidyr   1.1.3     v stringr 1.4.0
+## v readr   1.4.0     v forcats 0.5.1
+\end{verbatim}
+
+\begin{verbatim}
+## -- Conflicts ------------------------------------------ tidyverse_conflicts() --
+## x dplyr::filter() masks stats::filter()
+## x dplyr::lag()    masks stats::lag()
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
 \NormalTok{muslimClass }\OtherTok{\textless{}{-}}\NormalTok{ muslim }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(protectedWord }\SpecialCharTok{==} \StringTok{"muslim"}\NormalTok{)}
@@
 
-\hypertarget{discussion}{%
-\chapter{Discussion}\label{discussion}}
+\hypertarget{discussion-and-summary}{%
+\chapter{Discussion and summary}\label{discussion-and-summary}}
 

### 2021-06-20 thesis/04-discussion.Rmd
Adds logs

@@
 <!-- - mess, change protected words; relative differences stay the same, by the way lipstick on a pig -->
+
+<!-- debiasing -->
+Additionally, one cannot be sure if the bias is still preserved after the debiasing.
+The fact that all of the cosine distances for protected words and harmful attributes moved to the right, does not mean that the bias is removed.
+It is shown in articles such as @Gonen2019Lipstick, that the bias can hide in the vector geometry and preserve even after applying popular debiasing methods.
+
 <!-- - word lists are strange, not enough words, large HPDI -->
 
-<!-- where to put this ? -->
+<!-- interpretation issues -->
+One should remember that there is no clear interpretation for the values obtained with MAC metric from @Manzini2019blackToCriminal.
+One may assume that if the cosine distance is close to 1 then it is a desired outcome as it means, according to cosine distance assumptions, that there is almost no similarity between the words.
+However what does it mean to be close to 0?
+If the averaged cosine distance is equal to 0.8, then should we still debiase it?
+It is unclear what the criteria are.
+On one hand, it seems to be beneficial when the outcome is simplified as it is easier to compare results with one value per set.
+On the other hand, it is prone to misunderstanding of how to interpret the results and what threshold to assume.
 
@@
 
-What we mean is that in same cases it can unnecessarily label words as biased while omitting other more harmful associations.
-What is more, this highlights again the need for control group.
-One could check each stereotypical attribute with each protected word but not in order to average the values.
-One should rather plot the distributions and investigate how the metric measures vectors geometry and hidden semantic information.
-
-<!-- debiasing and uncertainty -->
-Debiasing with the method provided by @Manzini2019blackToCriminal shows how it is unpredictable in terms of the results.
-In some cases the final cosine distance is indeed close to neutral words, but in other cases the change is minor.
-What is more, the inclusion of the uncertainty shows how in some cases it is not clear why a word is according to this method
-classified as anomalous.
-The fact that uncertainty for `assosiated` and `different` class is so high leads to the situation where in the boundaries of an uncertainty of some attributes, one can find the neutral words as well.
-It makes the usage and interpretation of this dataset and metric extremely complicated and unclear.
-
-<!-- debiasing -->
-Additionally, one cannot be sure if the bias is still preserved after the debiasing.
-The fact that all of the cosine distances for protected words and harmful attributes moved to the right, does not mean that the bias is removed.
-One could argue that maybe it is not clear how to measure bias presence and removal but the method for debiasing works fine.
-Notwithstanding, it is showed in researches such as @Gonen2019Lipstick, that the bias can hide in the vector geometry and preserve even after applying popular debiasing methods.
-Therefore it seems to not be justified to claim that the method works properly without precise methods to verify this.
-
-<!-- no baseline for the interpretation of the results -->
-One should remember that there is no baseline for the interpretation of the averaged cosine distances in @Manzini2019blackToCriminalToCriminal.
-One may assume that if the cosine distance is close to 1 then it is a desired outcome as it means, according to cosine distance assumptions, that there is almost no similarity between the words.
-However what does it mean to be close to 0?
-If the averaged cosine distance is equal to 80, then should we still debiase it?
-It is unclear what the criteria are.
-On one hand, it seems to be beneficial when the outcome is simplified as it is easier to compare results with one value per set.
-On the other hand, it is prone to misunderstanding of how to interpret the results and what threshold to assume.
+Bibliography:
 

### 2021-06-20 thesis/_book/_main.tex
Adds logs

--- 2021-06-19 thesis/_book/_main.tex
@@
 
-Natural language processing (NLP) is a subfield of computer science that processes and analyzes language in text and speech with the use of modern programming methods.
-It has practical applications in everyday life as it concerns tasks such as email filters, smart assistants, search results, language translations, text analytics and so on.
-Models used to accomplish these tasks need a lot of data to learn from.
-This data originates from human activities and historical recordings such as texts, messages or speeches.
-It turns out that in the learning process these models can learn implicit biases that reflect harmful stereotypical thinking still present in modern societies.
-One can find methods that aim at identifying and measuring hidden biases and/or try to remove them by modifying the models.
-There are many different types of models in NLP depending on a task that they are supposed to solve.
-However, all of them need as an input words represented by means of numbers and this is accomplished with word embedding models.
-The models usually assign the values based on the context in which the words appear.
-This means that the input data can have enormous influence on the outcome.
-The biases seem to have their primary source in the way the words are assigned numerical values.
-
-One can find various definitions trying to capture what bias and fairness actually are.
-With the choice of the definition, implications into the real-life applications may change as well.
-\protect\hyperlink{ref-Mehrabi2019Survey}{Mehrabi, Morstatter, Saxena, Lerman, \& Galstyan} (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) mark out that there exist different types of biases such as historical bias, representation bias or measurement bias (the list is long).
-This indicates how complex the issue of bias is.
-Without the proper understanding and awareness of the problem, people are prone to unconsciously sustain the bias.
-\protect\hyperlink{ref-Mehrabi2019Survey}{Mehrabi, Morstatter, Saxena, Lerman, \& Galstyan} (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) also distinguish different types of discrimination, some of them will be briefly described.
-By protected attributes we mean those qualities, traits or characteristics that one should not discriminate against.
-Direct discrimination occurs when protected attributes of individuals explicitly result in non-favorable outcomes toward them.
-Such discrimination takes place for example when one uses the information about someone's religious beliefs to evaluate whether they are competent for the work they apply for.
-In contrast, in indirect discrimination individuals appear to be treated equally, but anyway they end up being treated unjustly due to the hidden effects of biases towards their protected attributes.
-An example of such situation is when in a job advert there is a requirement of 10 years experience instead of the list of a specific type of experience and knowledge.
-This job advert can discriminate indirectly against young people who can have the required skills, but not yet so much work experience.
-Systemic discrimination takes place when policies, customs or behaviors that result from certain culture or organizational structure lead to discrimination against some groups of people.
-Finally, statistical discrimination consists in relying on group statistics to judge person belonging to a given group.
-The topic of discrimination is entangled with another concept; fairness.
-It is essential to grasp some concepts of fairness to take them into consideration while designing implementation of some machine learning model.
-In \protect\hyperlink{ref-Mehrabi2019Survey}{Mehrabi, Morstatter, Saxena, Lerman, \& Galstyan} (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) one may notice that depending on the context and application different definitions may be applied.
-
-There is a considerable amount of literature available on the topic of bias detection and mitigation in NLP models.
-\protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) focus on gender biases that may be observable while investigating the representation of job occupations and gender in terms of their assigned numerical
-values.
-The authors apply cosine similarity measurement to investigate the phenomenon where (the vectors corresponding to) words related to jobs that are stereotypically associated with a given gender are in fact in the model situated closer to this gender.
-They also use analogy tasks to evaluate if bias is present in a word embedding model.
-They check analogies by comparing pairs of word vectors.
-For example they search for the word complementing the puzzle: man is to doctor as woman is to \ldots?
-First they subtract the word ``man'' from the word ``woman''\footnote{This is pointwise subtraction of vectors, but for the simplicity we will often talk about words while really meaning vectors.
-What is meant should we clear from the context.} and then
-they search for the ranked list of other words pairs that have similar vector difference. They also include in the formula a threshold to ensure that the resulting pairs could not be randomly picked.
-
-However, as \protect\hyperlink{ref-Nissim2019Fair}{Nissim, Noord, \& Goot} (\protect\hyperlink{ref-Nissim2019Fair}{2019}) points out, there are some limitations to this approach.
-In practice, most of analogies implementations do not return any input words.
-This means that it does not make sense to expect the algorithm to return the same profession for both woman and man.
-Therefore, this method seems to be limited in terms of bias detection.
-Other problems are related for example to the choice of pairs and words that are used to detect the presence of discrimination, as it is often subjective and without proper justification.
-Additionally, the choice of the parameter used in \protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) formula to ensure that word pairs are not picked by random, is also not justified and changing it drastically
-influences the results.
-
-\protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) touch upon the topic of biases regarding race and gender.
-They apply knowledge from well-known psychological studies such as Implicit Association Test to investigate the relation between human stereotypical thinking and model-learnt biases to discover a close relationship between these two.
-For the evaluation they use Word Embedding Association Test (WEAT) and the Word Embedding Factual Association Test (WEFAT).
-
-\protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) propose a novel way of using cosine similarity to verify if the assumed resemblance between protected words and harmful attributes exists.
-They investigate an approach that enables them to measure the bias for a class (such as gender, religion or race) and express the final result with a single metric.
-
-\protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) refer to Implicit Association Test (Greenwald et al., 1998) that measures the strength of associations between concepts or stereotypes by measuring human reaction time for special tasks.
-Humans naturally exhibit some biases which do not always cause social concern.
-One can imagine the intuitive associations between for example insects and flowers, and the feelings of pleasantness or unpleasantness.
-In general, people would rather associate flowers with feeling pleasant than insects, and this preference in a sense is a bias or
-prejudice in some direction.
-However, this type of preference does not cause an uproar and is a rather morally neutral case.
-This example is rather used to show that the methodology used in IAT makes sense.
-\protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) were able to obtain similar results to the ones that measured reaction time with the use of WEAT metric.
-Unfortunately, further studies discover other biases and prejudices that directly influence the quality of other people's lives and therefore they should be taken care of.
-
-The most common bias detection methods in the literature focus on comparing the similarity between words from protected groups and those that are considered to be stereotypical or harmful in some way.
-One can find in this group methods such as euclidean distance or cosine similarity (which is equivalent to dot product if the vectors are normalized).
-There are also other ways to detect the effects of biases.
-One can investigate the model performance on certain downstream tasks and validate if the model uses the gender or race information to complete the task.
-The currently used methods employing cosine similarity, often aggregate the similarity values often aggregated in a way that may lead to hasty conclusions.
-The averaging of values and the lack of attention to uncertainty involved in such aggregation may lead to the incomplete picture of the situation.
-
-In the paper we indicate how current methods used to detect biases in natural language models are limited, if we investigate them from the perspective of Bayesian analysis.
-Our research enhances the current way in which the bias detection is performed to make sure that it is methodologically valid.
-The key hypothesis is that greater understanding of data and bias implications can be achieved when Bayesian methods are applied to issue.
+Placeholder
 
@@
 
-\hypertarget{discussion}{%
-\chapter{Discussion}\label{discussion}}
+\hypertarget{discussion-and-summary}{%
+\chapter{Discussion and summary}\label{discussion-and-summary}}
 
-Placeholder
+We propose the use of Bayesian methods to measure uncertainty in bias detection.
+There are a few advantages of this method.
+Including uncertainty enables one to directly observe the influence of sample sizes.
+Analyzing individual words and connection coefficients, one may notice how \texttt{neutral} words have smaller uncertainty intervals and \texttt{different} or \texttt{associated} quite the opposite.
+One of the reasons for such outcome is that we used approximately 230 neutral words and only between 11-25 (the number varies from class to class) stereotypical attributes from \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) article.
+What is more, we also pay attention to distribution and details regarding anomalous values.
+With the use of simple visualizations that we introduced before, we were able to indicate suspicious cosine distance values.
+Additionally, we compare in details how the cosine distance values and uncertainty change after the debiasing.
+One can verify then how the individual vectors changed and if it what was expected.
+Our analysis with the use of Bayesian method gave us new ideas and hypothesis concerning not only the bias detection method but also the efficiency of debiasing itself.
+
+We created a summary table for each of the datasets: Reddit, Reddit Debiased, and Google word embeddings.
+Let us first analyse the general observations from estimated coefficients mean introduced in 3.6.
+DATASET-LEVEL COEFFICIENTS.
+For Google embeddings the HPDI for all classes coefficients (associated, different, human, and none) has an interval that includes zero.
+This can lead one to a conclusion that the impact for associated, different, human and neutral attribute is, when averaged, quite similar.
+This indicates how including the uncertainty may change the use and interpretation of \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) MAC metric.
+It seems that if one focuses only on differences between means of means, it is too simplistic.
+In case of Reddit word embeddings the situation is similar although HPDI interval is below 0 for Gender class when looking at \texttt{associated} and \texttt{different} mean coefficient.
+This can suggest that there is indeed slightly stronger impact of these attributes in cosine distance being smaller.
+One should also notice how \texttt{associated} and \texttt{different} coefficients have quite similar HPDI interval, the highest observed absolute difference is equal to only 0.1.
+This suggests that again the impact of associated attributes and difference ones is not clear at first sight when looking at averaged coefficients.
+Finally, let's compare the HPDI intervals for Reddit and Reddit debiased datasets.
+For Religion and Race dataset there is a minor shift (in absolute values the highest change is equal to approximately 0.1) of the mean coefficients towards zero.
+However for Gender dataset there is no significant change.
+This is of course the general look at the data, let's now analyze individual words.
+
+In Reddit table one can observe that for Religion class the cosine distance results for the associated attributes are for approximately 60\% of the words close to neutral attributes as well.
+This can suggest that in some cases words concerning humans can have higher similarity with some protected words independently if they are stereotypical or neutral-human words.
+One should also pay attention to the fact that for all of the \texttt{associated} and \texttt{different} attributes in Religion class the uncertainty interval overlaps at some point.
+What is even more surprising is that for protected words, such as \texttt{torah} \texttt{associated} attribute has the the cosine distance slightly over 1, which means no positive similarity!
+If the protected words that we chose do not have high similarity with harmful associated attributes, then one should consider at least three scenarios.
+The first one is that the choice of protected words and attributes may be corrupted.
+The second one is that the metric is not able to catch the hidden bias properly.
+The third one is that there is actually no bias between the words.
+Regardless of which scenario one considers, it is essential to take a look at the individual values before averaging them or aggregating in other ways.
+It seems that using Bayesian method can enhance the process of verifying the hypotheses concerning the choice of protected words and attributes.
+
+Surprisingly in Gender class one can observe high cosine similarity values between some female stereotypical professions and male protected words.
+If a word stereotypically associated with females has low cosine distance to male protected words, then one should try to figure out the reasons for that.
+Cosine distance seems to capture the information on the co-occurrence of words and not on the semantic similarity strictly speaking.
+
+In Google table ..
+
+In Reddit debiased ..
+
+Let's summarize the results of the bias detection methods analysis.
+
+Additionally, one cannot be sure if the bias is still preserved after the debiasing.
+The fact that all of the cosine distances for protected words and harmful attributes moved to the right, does not mean that the bias is removed.
+It is shown in articles such as \protect\hyperlink{ref-Gonen2019Lipstick}{Gonen \& Goldberg} (\protect\hyperlink{ref-Gonen2019Lipstick}{2019}), that the bias can hide in the vector geometry and preserve even after applying popular debiasing methods.
+
+One should remember that there is no clear interpretation for the values obtained with MAC metric from \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}).
+One may assume that if the cosine distance is close to 1 then it is a desired outcome as it means, according to cosine distance assumptions, that there is almost no similarity between the words.
+However what does it mean to be close to 0?
+If the averaged cosine distance is equal to 0.8, then should we still debiase it?
+It is unclear what the criteria are.
+On one hand, it seems to be beneficial when the outcome is simplified as it is easier to compare results with one value per set.
+On the other hand, it is prone to misunderstanding of how to interpret the results and what threshold to assume.
+
+The bottom line is that if we want to take bias seriously, so should we approach the uncertainty involved in our estimations.
+There is no replacement for proper statistical evaluation that does not discard information about the uncertaintly involved, larger word lists are needed, and visualisation of the results for particular protected classes provides much better guidance than chasing a single metric based on a means of means.
+
+Bibliography:
 
@@
 \begin{CSLReferences}{1}{0}
-\leavevmode\hypertarget{ref-Bolukbasi2016Man}{}%
-Bolukbasi, T., Chang, K.-W., Zou, J.
-Y., Saligrama, V., \& Kalai, A.
-(2016).
-Man is to computer programmer as woman is to homemaker?
-Debiasing word embeddings.
-\emph{CoRR}, \emph{abs/1607.06520}.
-Retrieved from \url{http://arxiv.org/abs/1607.06520}
-
-\leavevmode\hypertarget{ref-Caliskan2017Semantics}{}%
-Islam, A.
-C., Bryson, J.
-J., \& Narayanan, A.
-(2016).
-Semantics derived automatically from language corpora necessarily contain human biases.
-\emph{CoRR}, \emph{abs/1608.07187}.
-Retrieved from \url{http://arxiv.org/abs/1608.07187}
+\leavevmode\hypertarget{ref-Gonen2019Lipstick}{}%
+Gonen, H., \& Goldberg, Y.
+(2019).
+Lipstick on a pig: Debiasing methods cover up systematic gender biases in word embeddings but do not remove them.
+\emph{CoRR}, \emph{abs/1903.03862}.
+Retrieved from \url{http://arxiv.org/abs/1903.03862}
 
@@
 
-\leavevmode\hypertarget{ref-Mehrabi2019Survey}{}%
-Mehrabi, N., Morstatter, F., Saxena, N., Lerman, K., \& Galstyan, A.
-(2019).
-A survey on bias and fairness in machine learning.
-\emph{CoRR}, \emph{abs/1908.09635}.
-Retrieved from \url{http://arxiv.org/abs/1908.09635}
-
-\leavevmode\hypertarget{ref-Nissim2019Fair}{}%
-Nissim, M., Noord, R.
-van, \& Goot, R.
-van der.
-(2019).
-Fair is better than sensational: Man is to doctor as woman is to doctor.
-\emph{CoRR}, \emph{abs/1905.09866}.
-Retrieved from \url{http://arxiv.org/abs/1905.09866}
-
 \end{CSLReferences}

### 2021-06-20 thesis/04-discussion.Rmd
Adds final discussion

@@
 There are a few advantages of this method.
-Including uncertainty enables one to directly observe the influence of sample sizes.
+One of them is that including uncertainty enables one to directly observe the influence of sample sizes.
 Analyzing individual words and connection coefficients, one may notice how `neutral` words have smaller uncertainty intervals and `different` or `associated` quite the opposite.
 One of the reasons for such outcome is that we used approximately 230 neutral words and only between 11-25 (the number varies from class to class) stereotypical attributes from @Manzini2019blackToCriminal article.
-What is more, we also pay attention to distribution and details regarding anomalous values.
+In our approach, we also pay attention to the distribution and details regarding anomalous values.
 With the use of simple visualizations that we introduced before, we were able to indicate suspicious cosine distance values.
 Additionally, we compare in details how the cosine distance values and uncertainty change after the debiasing.
-One can verify then how the individual vectors changed and if it what was expected.
+One can verify then how the individual vectors changed and if it is what was expected.
 Our analysis with the use of Bayesian method gave us new ideas and hypothesis concerning not only the bias detection method but also the efficiency of debiasing itself.
@@
 In case of Reddit word embeddings the situation is similar although HPDI interval is below 0 for Gender class when looking at `associated` and `different` mean coefficient.
-This can suggest that there is indeed slightly stronger impact of these attributes in cosine distance being smaller.
-One should also notice how `associated` and `different` coefficients have quite similar HPDI interval, the highest observed absolute difference is equal to only 0.1.
+This can suggest that there is indeed slightly stronger impact of these attributes.
+One should also notice how in general `associated` and `different` coefficients have quite similar HPDI range, the highest observed absolute difference is equal to only 0.1.
 This suggests that again the impact of associated attributes and difference ones is not clear at first sight when looking at averaged coefficients.
@@
 However for Gender dataset there is no significant change.
-This is of course the general look at the data, let's now analyze individual words.
+This is of course the general look at the data, let's now analyze the individual words.
 
@@
 
-In Reddit table one can observe that for Religion class the cosine distance results for the associated attributes are for approximately 60% of the words close to neutral attributes as well.
+In Reddit table one can observe that for Religion class the cosine distance results for the associated attributes are for approximately 60% of the words close to human attributes as well.
 This can suggest that in some cases words concerning humans can have higher similarity with some protected words independently if they are stereotypical or neutral-human words.
 One should also pay attention to the fact that for all of the `associated` and `different` attributes in Religion class the uncertainty interval overlaps at some point.
-What is even more surprising is that for protected words, such as `torah`  `associated` attribute has the the cosine distance slightly over 1, which means no positive similarity!
+What is even more surprising is that for protected words, such as "torah"  the `associated` attribute has the the cosine distance slightly over 1, which means no positive similarity!
 If the protected words that we chose do not have high similarity with harmful associated attributes, then one should consider at least three scenarios.
-The first one is that the choice of protected words and attributes may be corrupted.
+The first one is that the choice of the protected words and attributes may be corrupted.
 The second one is that the metric is not able to catch the hidden bias properly.
@@
 Surprisingly in Gender class one can observe high cosine similarity values between some female stereotypical professions and male protected words.
-If a word stereotypically associated with females has low cosine distance to male protected words, then one should try to figure out the reasons for that.
+If a word stereotypically associated with females has low cosine distance to male protected words, then one should investigate the issue further.
+The reasons for this unexpected cosine distance may lie in the frequency of appearance of the protected words in the raw data.
+Some of the groups (like Muslim people or females) may have less representation in the data that is taken as an input for word embeddings.
+Therefore, if we assume that the MAC detects actually co-occurrence only, it makes sense that they can have high similarity with associated attributes and lower similarity with different ones.
+At the same time male protected words and other religions can have high similarity with different attributes because they have greater representation in the dataset in general and occur close to much more concepts.
 Cosine distance seems to capture the information on the co-occurrence of words and not on the semantic similarity strictly speaking.
 
-In Google table ..
+In Google table one may notice interesting observations as well.
+Although GoogleNews dataset is larger, trained on different data sources and with the use of more dimensions, some of the results are similar to the ones obtained for the Reddit corpus.
+For Religion case almost all of the values for `associated` class have an intersection with `different` class as well.
+In the case of Race it is 100% of the available words.
+This indicates again that it is not clear how the metric should be used.
+Quite different situation takes place for Gender data where the similarity between `associated` and `different` is present mostly for the male protected words.
+This means that male words have high similarity with both male stereotypical professions and female ones.
+However, for females the similarity is high mostly only for the female stereotypical professions.
+This observation could not be make when using MAC metric only as it requires investigating the individual words.
 
-<!-- comments and values that refer to them: (blue and green) -->
-<!-- a) google scale -->
-<!-- b) associated females -->
-
-In Reddit debiased ..
-
-<!-- debiased: question about the change -->
-<!-- - better but not for muslims -->
-<!-- - corpus frequency maybe? -->
-<!-- - better for male -->
-<!-- - when in gender different and associated together maybe it is better - gender is very specific, words are about professions;  -->
-<!-- - when in religion it is bad different and associated -->
-<!-- -  in race it is not obvious - mixed sentiments  -->
+When analyzing the Reddit debiased results one should ask a question of the change of the cosine distance values.
+It seems that not all of the protected words are treated equally when performing debiasing.
+In religion data the values for `associated` class moved towards 0 for most of the words except the ones for Islam religion where still the `associated` class has quite lower cosine distance than the `different` one.
+Similar situation takes place in Gender data where female words have still much lower cosine distance values for `associated` class than for the `different` one.
+In the case of males it is mostly almost the same interval for `associated` and `different`.
+As the Gender data is quite specific as the attributes are not harmful adjectives but (in an ideal world) neutral professions, the aim (if we follow MAC assumptions) should actually be to make the cosine distance same for both female and male protected words.
+However, as we pointed out the situation after debiasing can sometimes be better only for one protected group which is not the desired outcome.
+In the case of Religion data even after debiasing, all of the protected words have intersections between `associated` and `different` class and similarity greater than 0.
+As all of the attributes in religion data are negative and harmful stereotypes, one should not aim at making the distance between protected words and `associated`, and `different` class the same but rather to move it towards 0.
 
 Let's summarize the results of the bias detection methods analysis.
-
-<!-- black remarks then: -->
-<!-- - mess, change protected words; relative differences stay the same, by the way lipstick on a pig -->
-
-<!-- debiasing -->
-Additionally, one cannot be sure if the bias is still preserved after the debiasing.
-The fact that all of the cosine distances for protected words and harmful attributes moved to the right, does not mean that the bias is removed.
+As it was presented above, using the bias detection methods without involving the uncertainty may bring the risk of limited insight into the data.
+One should be concerned with the limitations of the metric that evaluates the debiasing by the mean of the averaged mean approach.
+Additionally, one cannot be sure if the bias is still preserved after debiasing.
+The fact that all of the cosine distances for protected words and harmful attributes moved slightly to the right, does not mean that the bias is removed.
 It is shown in articles such as @Gonen2019Lipstick, that the bias can hide in the vector geometry and preserve even after applying popular debiasing methods.
-
-<!-- - word lists are strange, not enough words, large HPDI -->
-
-<!-- interpretation issues -->
+One of the reasons for that may be the fact that relative differences between words may be preserved even after debiasing.
+Another thing is that the choice of word lists used to verify the metric effectiveness is not well justified.
+The sample size is very small which leads to large HPDI.
+Some of the protected words and attributes originate from old articles and it seems that the bias phenomenon is quite dynamic and the data should be as up-to-date as it is possible.
+One could for example research the IAT examples to verify the effectiveness of MAC metric.
 One should remember that there is no clear interpretation for the values obtained with MAC metric from @Manzini2019blackToCriminal.
@@
 However what does it mean to be close to 0?
-If the averaged cosine distance is equal to 0.8, then should we still debiase it?
+If the averaged cosine distance is equal to 0.8, then should we still debias it?
 It is unclear what the criteria are.
@@
 
-<!-- summary -->
-
 The bottom line is that if we want to take bias seriously, so should we approach the uncertainty involved in our estimations.
-There is no replacement for proper statistical evaluation that does not discard information about the uncertaintly involved, larger word lists are needed, and visualisation of the results for particular protected classes provides much better guidance than chasing a single metric based on a means of means.
+There is no replacement for proper statistical evaluation that does not discard information about the uncertainty involved, larger word lists are needed, and visualization of the results for particular protected classes provides much better guidance than chasing a single metric based on a means of means.
 

### 2021-06-20 thesis/_book/_main.tex
Adds final discussion

@@
 \author{Alicja Dobrzeniecka}
-\date{2021-06-19}
+\date{2021-06-20}
 
@@
 There are a few advantages of this method.
-Including uncertainty enables one to directly observe the influence of sample sizes.
+One of them is that including uncertainty enables one to directly observe the influence of sample sizes.
 Analyzing individual words and connection coefficients, one may notice how \texttt{neutral} words have smaller uncertainty intervals and \texttt{different} or \texttt{associated} quite the opposite.
 One of the reasons for such outcome is that we used approximately 230 neutral words and only between 11-25 (the number varies from class to class) stereotypical attributes from \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) article.
-What is more, we also pay attention to distribution and details regarding anomalous values.
+In our approach, we also pay attention to the distribution and details regarding anomalous values.
 With the use of simple visualizations that we introduced before, we were able to indicate suspicious cosine distance values.
 Additionally, we compare in details how the cosine distance values and uncertainty change after the debiasing.
-One can verify then how the individual vectors changed and if it what was expected.
+One can verify then how the individual vectors changed and if it is what was expected.
 Our analysis with the use of Bayesian method gave us new ideas and hypothesis concerning not only the bias detection method but also the efficiency of debiasing itself.
@@
 In case of Reddit word embeddings the situation is similar although HPDI interval is below 0 for Gender class when looking at \texttt{associated} and \texttt{different} mean coefficient.
-This can suggest that there is indeed slightly stronger impact of these attributes in cosine distance being smaller.
-One should also notice how \texttt{associated} and \texttt{different} coefficients have quite similar HPDI interval, the highest observed absolute difference is equal to only 0.1.
+This can suggest that there is indeed slightly stronger impact of these attributes.
+One should also notice how in general \texttt{associated} and \texttt{different} coefficients have quite similar HPDI range, the highest observed absolute difference is equal to only 0.1.
 This suggests that again the impact of associated attributes and difference ones is not clear at first sight when looking at averaged coefficients.
@@
 However for Gender dataset there is no significant change.
-This is of course the general look at the data, let's now analyze individual words.
+This is of course the general look at the data, let's now analyze the individual words.
 
-In Reddit table one can observe that for Religion class the cosine distance results for the associated attributes are for approximately 60\% of the words close to neutral attributes as well.
+In Reddit table one can observe that for Religion class the cosine distance results for the associated attributes are for approximately 60\% of the words close to human attributes as well.
 This can suggest that in some cases words concerning humans can have higher similarity with some protected words independently if they are stereotypical or neutral-human words.
 One should also pay attention to the fact that for all of the \texttt{associated} and \texttt{different} attributes in Religion class the uncertainty interval overlaps at some point.
-What is even more surprising is that for protected words, such as \texttt{torah} \texttt{associated} attribute has the the cosine distance slightly over 1, which means no positive similarity!
+What is even more surprising is that for protected words, such as ``torah'' the \texttt{associated} attribute has the the cosine distance slightly over 1, which means no positive similarity!
 If the protected words that we chose do not have high similarity with harmful associated attributes, then one should consider at least three scenarios.
-The first one is that the choice of protected words and attributes may be corrupted.
+The first one is that the choice of the protected words and attributes may be corrupted.
 The second one is that the metric is not able to catch the hidden bias properly.
@@
 Surprisingly in Gender class one can observe high cosine similarity values between some female stereotypical professions and male protected words.
-If a word stereotypically associated with females has low cosine distance to male protected words, then one should try to figure out the reasons for that.
+If a word stereotypically associated with females has low cosine distance to male protected words, then one should investigate the issue further.
+The reasons for this unexpected cosine distance may lie in the frequency of appearance of the protected words in the raw data.
+Some of the groups (like Muslim people or females) may have less representation in the data that is taken as an input for word embeddings.
+Therefore, if we assume that the MAC detects actually co-occurrence only, it makes sense that they can have high similarity with associated attributes and lower similarity with different ones.
+At the same time male protected words and other religions can have high similarity with different attributes because they have greater representation in the dataset in general and occur close to much more concepts.
 Cosine distance seems to capture the information on the co-occurrence of words and not on the semantic similarity strictly speaking.
 
-In Google table ..
+In Google table one may notice interesting observations as well.
+Although GoogleNews dataset is larger, trained on different data sources and with the use of more dimensions, some of the results are similar to the ones obtained for the Reddit corpus.
+For Religion case almost all of the values for \texttt{associated} class have an intersection with \texttt{different} class as well.
+In the case of Race it is 100\% of the available words.
+This indicates again that it is not clear how the metric should be used.
+Quite different situation takes place for Gender data where the similarity between \texttt{associated} and \texttt{different} is present mostly for the male protected words.
+This means that male words have high similarity with both male stereotypical professions and female ones.
+However, for females the similarity is high mostly only for the female stereotypical professions.
+This observation could not be make when using MAC metric only as it requires investigating the individual words.
 
-In Reddit debiased ..
+When analyzing the Reddit debiased results one should ask a question of the change of the cosine distance values.
+It seems that not all of the protected words are treated equally when performing debiasing.
+In religion data the values for \texttt{associated} class moved towards 0 for most of the words except the ones for Islam religion where still the \texttt{associated} class has quite lower cosine distance than the \texttt{different} one.
+Similar situation takes place in Gender data where female words have still much lower cosine distance values for \texttt{associated} class than for the \texttt{different} one.
+In the case of males it is mostly almost the same interval for \texttt{associated} and \texttt{different}.
+As the Gender data is quite specific as the attributes are not harmful adjectives but (in an ideal world) neutral professions, the aim (if we follow MAC assumptions) should actually be to make the cosine distance same for both female and male protected words.
+However, as we pointed out the situation after debiasing can sometimes be better only for one protected group which is not the desired outcome.
+In the case of Religion data even after debiasing, all of the protected words have intersections between \texttt{associated} and \texttt{different} class and similarity greater than 0.
+As all of the attributes in religion data are negative and harmful stereotypes, one should not aim at making the distance between protected words and \texttt{associated}, and \texttt{different} class the same but rather to move it towards 0.
 
 Let's summarize the results of the bias detection methods analysis.
-
-Additionally, one cannot be sure if the bias is still preserved after the debiasing.
-The fact that all of the cosine distances for protected words and harmful attributes moved to the right, does not mean that the bias is removed.
+As it was presented above, using the bias detection methods without involving the uncertainty may bring the risk of limited insight into the data.
+One should be concerned with the limitations of the metric that evaluates the debiasing by the mean of the averaged mean approach.
+Additionally, one cannot be sure if the bias is still preserved after debiasing.
+The fact that all of the cosine distances for protected words and harmful attributes moved slightly to the right, does not mean that the bias is removed.
 It is shown in articles such as \protect\hyperlink{ref-Gonen2019Lipstick}{Gonen \& Goldberg} (\protect\hyperlink{ref-Gonen2019Lipstick}{2019}), that the bias can hide in the vector geometry and preserve even after applying popular debiasing methods.
-
+One of the reasons for that may be the fact that relative differences between words may be preserved even after debiasing.
+Another thing is that the choice of word lists used to verify the metric effectiveness is not well justified.
+The sample size is very small which leads to large HPDI.
+Some of the protected words and attributes originate from old articles and it seems that the bias phenomenon is quite dynamic and the data should be as up-to-date as it is possible.
+One could for example research the IAT examples to verify the effectiveness of MAC metric.
 One should remember that there is no clear interpretation for the values obtained with MAC metric from \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}).
@@
 The bottom line is that if we want to take bias seriously, so should we approach the uncertainty involved in our estimations.
-There is no replacement for proper statistical evaluation that does not discard information about the uncertaintly involved, larger word lists are needed, and visualisation of the results for particular protected classes provides much better guidance than chasing a single metric based on a means of means.
+There is no replacement for proper statistical evaluation that does not discard information about the uncertainty involved, larger word lists are needed, and visualization of the results for particular protected classes provides much better guidance than chasing a single metric based on a means of means.
 

### 2021-06-21 thesis/01-intro.Rmd
Adds review remarks

@@
 However, all of them need as an input words represented by means of numbers and this is accomplished with word embeddings.
-The word embeddings are usually assigned the values based on the context in which the words appear.
+Word embeddings are usually assigned the values based on the context in which the words appear.
 This means that the input data can have an enormous influence on the outcome.
-The biases seem to have their primary source in the way the words are assigned numerical values.
-One can find methods that aim at identifying and measuring hidden biases and/or try to remove them by for example modifying the word embeddings directly.
+One can find methods that aim at identifying and measuring hidden biases and/or try to remove them by, for example, modifying the word embeddings directly.
 
 One can find various definitions trying to capture what bias and fairness actually are.
-With the choice of the definition, implications into the real-life applications may change as well.
+With the choice of the definition, implications for the real-life applications may change as well.
 @Mehrabi2019Survey mark out that there exist different types of biases such as historical bias, representation bias or measurement bias (the list is long).
@@
 Systemic discrimination takes place when policies, customs or behaviors that result from certain culture or organizational structure lead to discrimination against some groups of people.
-Finally, statistical discrimination consists in relying on group statistics to judge person belonging to a given group.
+Finally, statistical discrimination consists in relying on group statistics to judge a person belonging to a given group.
 The topic of discrimination is entangled with another concept; fairness.
@@
 There is a considerable amount of literature available on the topic of bias detection and mitigation in NLP models.
-@Bolukbasi2016Man focus on gender biases that may be observable while investigating the representation of job occupations and gender in terms of their assigned numerical
+@Bolukbasi2016Man focus on gender biases that may be observable in the representation of job occupations and gender in terms of numerical
 values.
 The authors apply cosine similarity measurement to investigate the phenomenon where (the vectors corresponding to) words related to jobs that are stereotypically associated with a given gender are in fact in the model situated closer to this gender.
-They also use analogy tasks to evaluate if bias is present in a word embedding model.
+They also use analogy tasks to evaluate if a bias is present in a word embedding model.
 They check analogies by comparing pairs of word vectors.
@@
 This means that it does not make sense to expect the algorithm to return the same profession for both woman and man.
-Therefore, this method seems to be limited in terms of bias detection.
+Therefore, this method of bias detection seems limited.
 Other problems are related for example to the choice of pairs and words that are used to detect the presence of discrimination, as it is often subjective and without proper justification.
@@
 @Manzini2019blackToCriminal propose a novel way of using cosine similarity to verify if the assumed resemblance between protected words and harmful attributes exists.
-They investigate an approach that enables them to measure the bias for a class (such as gender, religion or race) and express the final result with a single metric.
+They develop an approach that enables them to measure the bias for a class (such as gender, religion or race) and express the final result with a single metric.
 Their approach will be analyzed in details in the next chapter.
@@
 One can imagine the intuitive associations between for example insects and flowers, and the feelings of pleasantness or unpleasantness.
-In general, people would rather associate flowers with feeling pleasant than insects, and this preference in a sense is a bias or
+In general, people would rather associate flowers than insects with feeling pleasant, and this preference in a sense is a bias or
 prejudice in some direction.
@@
 The most common bias detection methods in the literature focus on comparing the similarity between words from protected groups and those that are considered to be stereotypical or harmful in some way.
-One can find in this group methods such as euclidean distance or cosine similarity (which is equivalent to dot product if the vectors are normalized).
+One can find in this group methods such as Euclidean distance or cosine similarity (which is equivalent to dot product if the vectors are normalized).
 There are also other ways to detect the effects of biases.
 One can investigate the model performance on certain downstream tasks and validate if the model uses the gender or race information to complete the task.
-The currently used methods employing cosine similarity, often aggregate the similarity values often aggregated in a way that may lead to hasty conclusions.
-The averaging of values and the lack of attention to uncertainty involved in such aggregation may lead to the incomplete picture of the situation.
+Unfortunately, the currently used methods employing cosine similarity, often aggregate the similarity values in a way that may lead to hasty conclusions.
+The averaging of values and the lack of attention to uncertainty involved in such aggregation may lead to an incomplete picture of the situation.
 

### 2021-06-21 thesis/02-cosine.Rmd
Adds review remarks

@@
 One of the most straightforward ones is a skip-gram model.
-Given a word the model tries to predict its neighboring words from the sentence.
+Given a word, the model tries to predict its neighboring words from the sentence.
 The mathematics behind the process relies on the idea that the prediction concerns the conditional probability of the adjacent words.
@@
 As the angle between vectors decreases, cosine similarity approaches one, which stands for the greatest similarity.
-It is also possible to obtain negative cosine similarity, if the value approaches -1 it means that the words are contrary to each other.
+It is also possible to obtain negative cosine similarity.
+If the value approaches -1, this intuitively means that the words are contrary to each other.
 
@@
 The triangle inequality formula says that for any triangle, the sum of the lengths of any two sides must be greater than or equal to the length of the remaining side.
-As shown in [discussion in stats.stackexchange.com](https://stats.stackexchange.com/questions/198080/proving-that-cosine-distance-function-defined-by-cosine-similarity-between-two-u) in the case of cosine distance it would have to fulfill this equation 1+cos-sim(𝐴,𝐶) < cos-sim(𝐴,𝐵) + cos-sim(𝐵,𝐶).
+As shown in [discussion in stats.stackexchange.com](https://stats.stackexchange.com/questions/198080/proving-that-cosine-distance-function-defined-by-cosine-similarity-between-two-u)^[https://stats.stackexchange.com/questions/198080/proving-that-cosine-distance-function-defined-by-cosine-similarity-between-two-u] in the case of cosine distance it would have to fulfill this equation 1+cos-sim(𝐴,𝐶) < cos-sim(𝐴,𝐵) + cos-sim(𝐵,𝐶).
 If one chooses specific unit vectors it is easy to demonstrate that the triangle inequality is not preserved.
@@
 @Bolukbasi2016Man define similarity between words as the outcome inner product of their normalized vectors. 
-They focus on examining what the geometry of word embedding is in regard to "he" and "she" words.
+They focus on examining what the geometry of a word embedding is in regard to "he" and "she" words.
 In other words, whether the similarity between those concepts and other words reflects expected gender stereotypes.
@@
 @Manzini2019blackToCriminal present a different approach towards finding similarities between classes of words.
-The authors claim that texts available online are full of direct or indirect human stereotypes.
-As a result, word embeddings are prone to learn and maybe amplify those biases and propagate them further into AI models that are used for various applications.
 Cosine distance is used in the article as a measure to first argue for the existence of multi-class bias and then to show how through bias mitigation techniques the bias may be decreased.
@@
 
-In the final step there is an evaluation of the results.
-The cosine distances are calculated again but this time using debiased vocabulary.
+In the final step they evaluate the results.
+The cosine distances are calculated again but this time using the debiased vocabulary.
 After taking the mean of all distances one final value is obtained and then it is compared with the average value from the beginning.
@@
 If one takes closer look at the individual values that are taken for the calculations it turns out that there is a bunch of outliers and surprisingly dissimilar words.
-We approach this issue by providing the tables and visualizations of individual cosine distances to make sure that we get the proper insight into the data.
+We approach this issue by providing the tables and visualizations of individual cosine distances to make sure that we obtain a proper insight into the data.
 
@@
 
-With such a method the uncertainty is not included which makes it even more difficult to give reasonable interpretations of the results.
+With such a method the uncertainty involved is not really considered which makes it even more difficult to give reasonable interpretations of the results.
 We propose the use of Bayesian method to obtain some understanding of the influence the uncertainty has on the interpretation of final results.
 
-\textbf{4. No word class distinction}
+\textbf{4. No word class distinction and no control group}
 
@@
 Not all of the stereotypical words have to be considered as harmful for all of the religions.
+One should investigate the religions separately as some of them may have stronger harmful associations that others.
+One should also include control groups to have a way of comparing the stereotypical results with neutral or human-like words.
+Later in the text we will explain in details reasons for introducing control groups.
 In our analysis, we distinguish between stereotypes associated with a given group, stereotypes associated with different groups, and control groups: neutral words and stereotypes-free human predicates.
@@
 This is in general the problem of scale and the lack of universal intervals.
-In contrast, proper intervals decide whether given cosine similarity is high enough to consider the words to be more similar than if we chose them at random.
-We will use highest posterior density intervals.
+In contrast, statistical intervals will help us decide whether a given cosine similarity is high enough to consider the words to be more similar than if we chose them at random.
+We will use highest posterior density intervals, in line with Bayesian methodology.
 

### 2021-06-21 thesis/03-bayesian.Rmd
Adds review remarks

@@
 
-We will use the choice of protected words and stereotypical predicates used in REF.
-This is a decent point of departure,  not only we want to compare our method to that of REF, but also because this data format is fairly general (as contrasted, say, with a set up for binary stereotypes).
+We will use the choice of protected words and stereotypical predicates used in @Manzini2019blackToCriminal.
+This is a decent point of departure, as not only we want to compare our method to that of @Manzini2019blackToCriminal, but also because this data format is fairly general (as contrasted, say, with a set up for binary stereotypes).
 Note also that the method we develop here  can fairly easily be  run  for different stereotypization patterns.
@@
 In order to obtain this  dataset, we  calculated the cosine distance between each protected word and each word from both the bias-related attribute groups, which were used in the original study, and to neutral and human control attributes which we added  as control groups.
-For instance, for religion, the bias-related predicates (coming from the original study in REF) include muslim bias attributes, jew bias attributes, christian bias attributes (see a list in the APPENDIX).
+For instance, for religion, the bias-related predicates (coming from the original study in @Manzini2019blackToCriminal) include muslim bias attributes, jew bias attributes, christian bias attributes (see a list in the APPENDIX).
 
@@
 The first impression is that while  there is a  shift for  associated words towards smaller cosine distances as compared to the neutral words, slightly surprisingly a slightly weaker shift in the same direction is visible for attributes associated with different stereotypes.
-Moreover,  the empirical distributions overlap to alarge extent and the  means grouped by connection type do not seem too far from each other.
+Moreover,  the empirical distributions overlap to a large extent and the  means grouped by connection type do not seem too far from each other.
 In fact, as there is a lot of  variety in the cosine distances (as we will soon see), we need to gauge the uncertainty involved, and  to look more carefully at individual protected words to get a better idea of how the cosine distance distribution changes for different attribute groups and different protected classes.
@@
 
+\normalsize
 The only reason we need it is the evaluation of connection as a predictor.
@@
 Let's turn then to data analysis that takes a look at protected words separately.
-This time for each combination of a protected word and a connection status we will have a separate mean cosine distance estimate, each coming with its own highest posterior density interval.
+This time, for each combination of a protected word and a connection status we will have a separate mean cosine distance estimate, each coming with its own highest posterior density interval.
 This means we will use indices that are result from all such combinations (and then we will split them up in the model precis to build visualisation, feel free to look at the `visualiseStats.R` script for details).
@@
 We first encountered the GoogleNews word embedding model in @Bolukbasi2016Man.
-They use it for their calculations as written in [Tolga Code](https://github.com/tolga-b/debiaswe).
+They use it for their calculations as written in [Tolga Bolukbasi code](https://github.com/tolga-b/debiaswe)^[https://github.com/tolga-b/debiaswe].
 We decided to compare the results obtained with the use of Reddit L2 model and the ones that we got by GoogleNews model.
@@
 One of the main differences between the models is that Reddit word embeggings have 50 dimensions and GoogleNews word embeddings have 300 dimensions.
-As the dimension is increasing, the vector can capture much more information but this is not always the case.
-According to some researchers the amount of dimensions should be chosen by taking into account the type of corpus and its features.
+As the dimension increases, the vectors can capture much more information, but this is not always the case.
+According to some researchers, the amount of dimensions should be chosen by taking into account the type of corpus and its features.
 In our case both Reddit and GoogleNews models are already trained and ready to use so we do not analyse further the choice of their hyperparameters.
@@
 
-First, let's look at coefficient estimated for the whole datasets, as compared to their estimation prior to debiasing:
+First, let's look at coefficient estimated for the whole datasets, as compared to their estimation prior to debiasing.
+One may observe very slight changes in the estimated coeﬀicients.
+It seems that in religion dataset the changes are the most significant as the mean moves towards zero which stands for no similarity between words.
 
@@
 
-One may observe very slight changes in the estimated coeﬀicients.
-It seems that in religion dataset the changes are the most significant as the mean moves towards zero which stands for no similarity between words.
-
 ## Protected classes  after debiasing
@@
 It is worth paying attention to the gender debiased dataset.
-The change is really minor, still zero out of HPDI range.
+The change is really minor, still with zero out of the HPDI range.
 The `different` attributes still have high similarity value.
-It seems that the distances for gender protected words are still clustered together which suggests that this method is not able to catch the complex bias presence.
+It seems that the distances for gender protected words are still clustered together which suggests that this method is unable to catch the complex bias presence.
 One could conclude that co-occurrence of certain terms does not always mean direct connection in terms of terms associations.
@@
 One may assume that there reason for so small change is that either the metric does not indicate properly biases or that the issue is more complex and subtracting subspaces is not enough.
-
-In @Manzini2019blackToCriminal it is mentioned that debiasing with this method that is based on bias component removal may be insufficient.
-The reason is that words containing similar biases may still cluster together even after debiasing.

### 2021-06-21 thesis/04-discussion.Rmd
Adds review remarks

@@
 Analyzing individual words and connection coefficients, one may notice how `neutral` words have smaller uncertainty intervals and `different` or `associated` quite the opposite.
-One of the reasons for such outcome is that we used approximately 230 neutral words and only between 11-25 (the number varies from class to class) stereotypical attributes from @Manzini2019blackToCriminal article.
+One of the reasons for such an outcome is that we used approximately 230 neutral words and only between 11-25 (the number varies from class to class) stereotypical attributes from @Manzini2019blackToCriminal article.
 In our approach, we also pay attention to the distribution and details regarding anomalous values.
@@
 Additionally, we compare in details how the cosine distance values and uncertainty change after the debiasing.
-One can verify then how the individual vectors changed and if it is what was expected.
+One can investigate then how the individual vectors changed and if it is what was expected.
 Our analysis with the use of Bayesian method gave us new ideas and hypothesis concerning not only the bias detection method but also the efficiency of debiasing itself.
@@
 Let us first analyse the general observations from estimated coefficients mean introduced in 3.6.
-DATASET-LEVEL COEFFICIENTS.
-For Google embeddings the HPDI for all classes coefficients (associated, different, human, and none) has an interval that includes zero.
+Dataset-level coefficients.
+For Google embeddings the HPDIs for all class coefficients (associated, different, human, and none) include zero.
 This can lead one to a conclusion that the impact for associated, different, human and neutral attribute is, when averaged, quite similar.
 This indicates how including the uncertainty may change the use and interpretation of @Manzini2019blackToCriminal MAC metric.
-It seems that if one focuses only on differences between means of means, it is too simplistic.
-In case of Reddit word embeddings the situation is similar although HPDI interval is below 0 for Gender class when looking at `associated` and `different` mean coefficient.
+It seems that if one focuses only on differences between means of means, this is too simplistic.
+In case of Reddit word embeddings the situation is similar although HPDI is below 0 for Gender class when looking at `associated` and `different` mean coefficient.
 This can suggest that there is indeed slightly stronger impact of these attributes.
 One should also notice how in general `associated` and `different` coefficients have quite similar HPDI range, the highest observed absolute difference is equal to only 0.1.
-This suggests that again the impact of associated attributes and difference ones is not clear at first sight when looking at averaged coefficients.
-Finally, let's compare the HPDI intervals for Reddit and Reddit debiased datasets.
+This suggests that again the impact of associated attributes and different ones is not clear at first sight.
+Finally, let's compare the HPDIs for Reddit and Reddit debiased datasets.
 For Religion and Race dataset there is a minor shift (in absolute values the highest change is equal to approximately 0.1) of the mean coefficients towards zero.
 However for Gender dataset there is no significant change.
-This is of course the general look at the data, let's now analyze the individual words.
+This is of course a general look at the group coefficients.
+Let's now analyze the individual words.
 
@@
 
-In Reddit table one can observe that for Religion class the cosine distance results for the associated attributes are for approximately 60% of the words close to human attributes as well.
-This can suggest that in some cases words concerning humans can have higher similarity with some protected words independently if they are stereotypical or neutral-human words.
-One should also pay attention to the fact that for all of the `associated` and `different` attributes in Religion class the uncertainty interval overlaps at some point.
+In Reddit table one can observe that for Religion the cosine distance results for the associated attributes are for approximately 60% of the words close to human attributes as well.
+This can suggest that in some cases words concerning humans can have higher similarity with some protected words independently of whether they are stereotypical or neutral-human words.
+One should also pay attention to the fact that for all of the `associated` and `different` attributes in Religion, the uncertainty intervals overlap at some point.
 What is even more surprising is that for protected words, such as "torah"  the `associated` attribute has the the cosine distance slightly over 1, which means no positive similarity!
@@
 The first one is that the choice of the protected words and attributes may be corrupted.
-The second one is that the metric is not able to catch the hidden bias properly.
+The second one is that the metric is unable to catch the hidden bias properly.
 The third one is that there is actually no bias between the words.
@@
 
-Surprisingly in Gender class one can observe high cosine similarity values between some female stereotypical professions and male protected words.
+Surprisingly in Gender one can observe high cosine similarity values between some female stereotypical professions and male protected words.
 If a word stereotypically associated with females has low cosine distance to male protected words, then one should investigate the issue further.
@@
 
-In Google table one may notice interesting observations as well.
-Although GoogleNews dataset is larger, trained on different data sources and with the use of more dimensions, some of the results are similar to the ones obtained for the Reddit corpus.
-For Religion case almost all of the values for `associated` class have an intersection with `different` class as well.
+In Google, one may notice interesting phenomena as well.
+Although the GoogleNews dataset is larger, trained on different data sources and with the use of more dimensions, some of the results are similar to the ones obtained for the Reddit corpus.
+For Religion almost all of the values for `associated` class intersect with `different` class as well.
 In the case of Race it is 100% of the available words.
 This indicates again that it is not clear how the metric should be used.
-Quite different situation takes place for Gender data where the similarity between `associated` and `different` is present mostly for the male protected words.
+Quite a different situation takes place for Gender, where the similarity between `associated` and `different` is present mostly for the male protected words.
 This means that male words have high similarity with both male stereotypical professions and female ones.
-However, for females the similarity is high mostly only for the female stereotypical professions.
-This observation could not be make when using MAC metric only as it requires investigating the individual words.
+However, for females, the similarity is high mostly only for the female stereotypical professions.
+This observation could not be made when using MAC metric only, as it requires investigating the individual words, and providing uncertainty,
 
-When analyzing the Reddit debiased results one should ask a question of the change of the cosine distance values.
+When analyzing the Reddit debiased results one should investigate the change of the cosine distance values.
 It seems that not all of the protected words are treated equally when performing debiasing.
-In religion data the values for `associated` class moved towards 0 for most of the words except the ones for Islam religion where still the `associated` class has quite lower cosine distance than the `different` one.
-Similar situation takes place in Gender data where female words have still much lower cosine distance values for `associated` class than for the `different` one.
-In the case of males it is mostly almost the same interval for `associated` and `different`.
+In religion, the values for `associated` class moved towards 0 for most of the words except for the ones for Islam, where still the `associated` class has quite lower cosine distance than the `different` one.
+Similar situation takes place in Gender, where female words have still much lower cosine distance values for `associated` class than for the `different` one.
+In the case of male protected words it is mostly almost the same interval for `associated` and `different`.
 As the Gender data is quite specific as the attributes are not harmful adjectives but (in an ideal world) neutral professions, the aim (if we follow MAC assumptions) should actually be to make the cosine distance same for both female and male protected words.
-However, as we pointed out the situation after debiasing can sometimes be better only for one protected group which is not the desired outcome.
-In the case of Religion data even after debiasing, all of the protected words have intersections between `associated` and `different` class and similarity greater than 0.
-As all of the attributes in religion data are negative and harmful stereotypes, one should not aim at making the distance between protected words and `associated`, and `different` class the same but rather to move it towards 0.
+However, as we pointed out the situation after debiasing can sometimes be better only for one protected group, which is not the desired outcome.
+In the case of Religion, even after debiasing, all of the protected words have intersections between `associated` and `different` class and similarity greater than 0.
+As all of the attributes in religion data are negative and harmful stereotypes, one should not aim at making the distance between protected words and `associated`, and `different` class the same but rather to move it towards 1.
 
@@
 As it was presented above, using the bias detection methods without involving the uncertainty may bring the risk of limited insight into the data.
-One should be concerned with the limitations of the metric that evaluates the debiasing by the mean of the averaged mean approach.
+One should be concerned with the limitations of the metric that evaluates the debiasing by means of the averaged mean approach.
 Additionally, one cannot be sure if the bias is still preserved after debiasing.
-The fact that all of the cosine distances for protected words and harmful attributes moved slightly to the right, does not mean that the bias is removed.
-It is shown in articles such as @Gonen2019Lipstick, that the bias can hide in the vector geometry and preserve even after applying popular debiasing methods.
+The fact that all of the cosine distances for protected words and harmful attributes moved slightly to the right does not mean that the bias is removed.
+@Gonen2019Lipstick argue that the bias can hide in the vector geometry and preserve even after applying popular debiasing methods.
 One of the reasons for that may be the fact that relative differences between words may be preserved even after debiasing.
 Another thing is that the choice of word lists used to verify the metric effectiveness is not well justified.
-The sample size is very small which leads to large HPDI.
-Some of the protected words and attributes originate from old articles and it seems that the bias phenomenon is quite dynamic and the data should be as up-to-date as it is possible.
-One could for example research the IAT examples to verify the effectiveness of MAC metric.
-One should remember that there is no clear interpretation for the values obtained with MAC metric from @Manzini2019blackToCriminal.
+The sample size is very small, which leads to large HPDIs.
+Some of the protected words and attributes originate from somewhat old articles and it seems that the bias phenomenon is quite dynamic and the data should be as up-to-date as it is possible.
+Moreover, recall that there is no clear interpretation for the values obtained with MAC metric from @Manzini2019blackToCriminal.
 One may assume that if the cosine distance is close to 1 then it is a desired outcome as it means, according to cosine distance assumptions, that there is almost no similarity between the words.
 However what does it mean to be close to 0?
-If the averaged cosine distance is equal to 0.8, then should we still debias it?
+If the averaged cosine distance is equal to 0.8, then should we still debias?
 It is unclear what the criteria are.

### 2021-06-21 thesis/_book/_main.tex
Adds review remarks

--- 2021-06-19 thesis/_book/_main.tex
@@
 \author{Alicja Dobrzeniecka}
-\date{2021-06-19}
+\date{2021-06-21}
 
@@
 
-We will use the choice of protected words and stereotypical predicates used in REF.
-This is a decent point of departure, not only we want to compare our method to that of REF, but also because this data format is fairly general (as contrasted, say, with a set up for binary stereotypes).
+We will use the choice of protected words and stereotypical predicates used in \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}).
+This is a decent point of departure, as not only we want to compare our method to that of \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}), but also because this data format is fairly general (as contrasted, say, with a set up for binary stereotypes).
 Note also that the method we develop here can fairly easily be run for different stereotypization patterns.
@@
 In order to obtain this dataset, we calculated the cosine distance between each protected word and each word from both the bias-related attribute groups, which were used in the original study, and to neutral and human control attributes which we added as control groups.
-For instance, for religion, the bias-related predicates (coming from the original study in REF) include muslim bias attributes, jew bias attributes, christian bias attributes (see a list in the APPENDIX).
+For instance, for religion, the bias-related predicates (coming from the original study in \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019})) include muslim bias attributes, jew bias attributes, christian bias attributes (see a list in the APPENDIX).
 
@@
 The first impression is that while there is a shift for associated words towards smaller cosine distances as compared to the neutral words, slightly surprisingly a slightly weaker shift in the same direction is visible for attributes associated with different stereotypes.
-Moreover, the empirical distributions overlap to alarge extent and the means grouped by connection type do not seem too far from each other.
+Moreover, the empirical distributions overlap to a large extent and the means grouped by connection type do not seem too far from each other.
 In fact, as there is a lot of variety in the cosine distances (as we will soon see), we need to gauge the uncertainty involved, and to look more carefully at individual protected words to get a better idea of how the cosine distance distribution changes for different attribute groups and different protected classes.
@@
 \end{Shaded}
+
+\normalsize
 
@@
 Let's turn then to data analysis that takes a look at protected words separately.
-This time for each combination of a protected word and a connection status we will have a separate mean cosine distance estimate, each coming with its own highest posterior density interval.
+This time, for each combination of a protected word and a connection status we will have a separate mean cosine distance estimate, each coming with its own highest posterior density interval.
 This means we will use indices that are result from all such combinations (and then we will split them up in the model precis to build visualisation, feel free to look at the \texttt{visualiseStats.R} script for details).
@@
 We first encountered the GoogleNews word embedding model in \protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}).
-They use it for their calculations as written in \href{https://github.com/tolga-b/debiaswe}{Tolga Code}.
+They use it for their calculations as written in \href{https://github.com/tolga-b/debiaswe}{Tolga Bolukbasi code}\footnote{\url{https://github.com/tolga-b/debiaswe}}.
 We decided to compare the results obtained with the use of Reddit L2 model and the ones that we got by GoogleNews model.
@@
 \end{center}
-
-One may observe very slight changes in the estimated coeﬀicients.
-It seems that in religion dataset the changes are the most significant as the mean moves towards zero which stands for no similarity between words.
 

### 2021-06-22 thesis/_book/_main.tex
not much

--- 2021-06-20 thesis/_book/_main.tex
@@
 \setcounter{secnumdepth}{5}
+% Redefines (sub)paragraphs to behave more like sections
+\ifx\paragraph\undefined\else
+\let\oldparagraph\paragraph
+\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+\let\oldsubparagraph\subparagraph
+\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
+
+% set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
+
 \usepackage{todonotes}
-\ifluatex
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
-  % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
-\title{Conceptual and methodological problems with bias detection and avoidance in natural language processing}
+\title{Conceptual and methodological problems with bias detection and avoidance
+in natural language processing}
 \author{Alicja Dobrzeniecka}
-\date{2021-06-20}
+\date{2021-06-21}
 
@@
 }
-\setstretch{1.5}
-\hypertarget{introduction}{%
-\chapter{Introduction}\label{introduction}}
+\chapter{Introduction}\label{introduction}
 
-Placeholder
+Natural language processing (NLP) is a subfield of computer science that
+processes and analyzes language in text and speech with the use of
+modern programming methods. It has practical applications in everyday
+life as it concerns tasks such as email filters, smart assistants,
+search results, language translations, text analytics and so on. Models
+used to accomplish these tasks need a lot of data to learn from. This
+data originates from human activities and historical recordings such as
+texts, messages or speeches. It turns out that in the learning process
+these models can learn implicit biases that reflect harmful
+stereotypical thinking still present in modern societies. There are many
+different types of models in NLP depending on a task that they are
+supposed to solve. However, all of them need as an input words
+represented by means of numbers and this is accomplished with word
+embeddings. The word embeddings are usually assigned the values based on
+the context in which the words appear. This means that the input data
+can have an enormous influence on the outcome. The biases seem to have
+their primary source in the way the words are assigned numerical values.
+One can find methods that aim at identifying and measuring hidden biases
+and/or try to remove them by for example modifying the word embeddings
+directly.
 
-\hypertarget{cosine-similarity-and-bias-detection}{%
-\chapter{Cosine similarity and bias detection}\label{cosine-similarity-and-bias-detection}}
+One can find various definitions trying to capture what bias and
+fairness actually are. With the choice of the definition, implications
+into the real-life applications may change as well. Mehrabi, Morstatter,
+Saxena, Lerman, \& Galstyan
+(\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) mark out that there
+exist different types of biases such as historical bias, representation
+bias or measurement bias (the list is long). This indicates how complex
+the issue of bias is. Without the proper understanding and awareness of
+the problem, people are prone to unconsciously sustain the bias. Mehrabi
+et al. (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) also
+distinguish different types of discrimination, some of them will be
+briefly described. By protected attributes we mean those qualities,
+traits or characteristics that one should not discriminate against.
+Direct discrimination occurs when protected attributes of individuals
+explicitly result in non-favorable outcomes toward them. Such
+discrimination takes place for example when one uses the information
+about someone's religious beliefs to evaluate whether they are competent
+for the work they apply for. In contrast, in indirect discrimination
+individuals appear to be treated equally, but anyway they end up being
+treated unjustly due to the hidden effects of biases towards their
+protected attributes. An example of such situation is when in a job
+advert there is a requirement of 10 years experience instead of the list
+of a specific type of experience and knowledge. This job advert can
+discriminate indirectly against young people who can have the required
+skills, but not yet so much work experience. Systemic discrimination
+takes place when policies, customs or behaviors that result from certain
+culture or organizational structure lead to discrimination against some
+groups of people. Finally, statistical discrimination consists in
+relying on group statistics to judge person belonging to a given group.
+The topic of discrimination is entangled with another concept; fairness.
+It is essential to grasp some concepts of fairness to take them into
+consideration while designing implementation of some machine learning
+model. In Mehrabi et al.
+(\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) one may notice that
+depending on the context and application different definitions may be
+applied.
 
-Placeholder
+There is a considerable amount of literature available on the topic of
+bias detection and mitigation in NLP models. Bolukbasi, Chang, Zou,
+Saligrama, \& Kalai (\protect\hyperlink{ref-Bolukbasi2016Man}{2016})
+focus on gender biases that may be observable while investigating the
+representation of job occupations and gender in terms of their assigned
+numerical values. The authors apply cosine similarity measurement to
+investigate the phenomenon where (the vectors corresponding to) words
+related to jobs that are stereotypically associated with a given gender
+are in fact in the model situated closer to this gender. They also use
+analogy tasks to evaluate if bias is present in a word embedding model.
+They check analogies by comparing pairs of word vectors. For example
+they search for the word complementing the puzzle: man is to doctor as
+woman is to \ldots{}? First they subtract the word ``man'' from the word
+``woman''\footnote{This is pointwise subtraction of vectors, but for the
+  simplicity we will often talk about words while really meaning
+  vectors. What is meant should be clear from the context.} and then
+they search for the ranked list of other words pairs that have similar
+vector difference. They also include in the formula a threshold to
+ensure that the resulting pairs could not be randomly picked.
 
-\hypertarget{word-embeddings}{%
-\section{Word embeddings}\label{word-embeddings}}
+However, as Nissim, Noord, \& Goot
+(\protect\hyperlink{ref-Nissim2019Fair}{2019}) points out, there are
+some limitations to this approach. In practice, most of analogies
+implementations do not return any input words. This means that it does
+not make sense to expect the algorithm to return the same profession for
+both woman and man. Therefore, this method seems to be limited in terms
+of bias detection. Other problems are related for example to the choice
+of pairs and words that are used to detect the presence of
+discrimination, as it is often subjective and without proper
+justification. Additionally, the choice of the parameter used in
+Bolukbasi et al. (\protect\hyperlink{ref-Bolukbasi2016Man}{2016})
+formula to ensure that word pairs are not picked by random, is also not
+justified and changing it drastically influences the results.
 
-\hypertarget{cosine-similarity-and-distance}{%
-\section{Cosine similarity and distance}\label{cosine-similarity-and-distance}}
+Islam, Bryson, \& Narayanan
+(\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) touch upon the
+topic of biases regarding race and gender. They apply knowledge from
+well-known psychological studies such as Implicit Association Test to
+investigate the relation between human stereotypical thinking and
+model-learnt biases to discover a close relationship between these two.
+For the evaluation they use Word Embedding Association Test (WEAT) and
+the Word Embedding Factual Association Test (WEFAT).
 
-\hypertarget{cosine-distance-in-a-one-class-bias-detection}{%
-\section{Cosine distance in a one-class bias detection}\label{cosine-distance-in-a-one-class-bias-detection}}
+Manzini, Lim, Tsvetkov, \& Black
+(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) propose a
+novel way of using cosine similarity to verify if the assumed
+resemblance between protected words and harmful attributes exists. They
+investigate an approach that enables them to measure the bias for a
+class (such as gender, religion or race) and express the final result
+with a single metric. Their approach will be analyzed in details in the
+next chapter.
 
-\hypertarget{cosine-distance-in-a-multi-class-bias-detection}{%
-\section{Cosine distance in a multi-class bias detection}\label{cosine-distance-in-a-multi-class-bias-detection}}
+Islam et al. (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) refer
+to Implicit Association Test (Greenwald et al., 1998) that measures the
+strength of associations between concepts or stereotypes by measuring
+human reaction time for special tasks. Humans naturally exhibit some
+biases which do not always cause social concern. One can imagine the
+intuitive associations between for example insects and flowers, and the
+feelings of pleasantness or unpleasantness. In general, people would
+rather associate flowers with feeling pleasant than insects, and this
+preference in a sense is a bias or prejudice in some direction. However,
+this type of preference does not cause an uproar and is a rather morally
+neutral case. This example is rather used to show that the methodology
+used in IAT makes sense. Islam et al.
+(\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) were able to
+obtain similar results to the ones that measured reaction time with the
+use of WEAT metric. Unfortunately, further studies discover other biases
+and prejudices that directly influence the quality of other people's
+lives and therefore they should be taken care of.
 
-\hypertarget{limitations-of-the-approach}{%
-\section{Limitations of the approach}\label{limitations-of-the-approach}}
+The most common bias detection methods in the literature focus on
+comparing the similarity between words from protected groups and those
+that are considered to be stereotypical or harmful in some way. One can
+find in this group methods such as euclidean distance or cosine
+similarity (which is equivalent to dot product if the vectors are
+normalized). There are also other ways to detect the effects of biases.
+One can investigate the model performance on certain downstream tasks
+and validate if the model uses the gender or race information to
+complete the task. The currently used methods employing cosine
+similarity, often aggregate the similarity values often aggregated in a
+way that may lead to hasty conclusions. The averaging of values and the
+lack of attention to uncertainty involved in such aggregation may lead
+to the incomplete picture of the situation.
 
-\hypertarget{walkthrough-with-the-religion-dataset}{%
-\chapter{Walkthrough with the religion dataset}\label{walkthrough-with-the-religion-dataset}}
+In the paper we indicate how current methods used to detect biases in
+natural language models are limited, if we investigate them from the
+perspective of Bayesian analysis. Our research enhances the current way
+in which the bias detection is performed to make sure that it is
+methodologically valid. The key hypothesis is that greater understanding
+of data and bias implications can be achieved when Bayesian methods are
+applied to issue.
 
-Placeholder
+\chapter{Cosine similarity and bias
+detection}\label{cosine-similarity-and-bias-detection}
 
-\hypertarget{loading-and-understanding-the-dataset}{%
-\section{Loading and understanding the dataset}\label{loading-and-understanding-the-dataset}}
+\section{Word embeddings}\label{word-embeddings}
 
-\hypertarget{first-look-at-the-empirical-distributions}{%
-\section{First look at the empirical distributions}\label{first-look-at-the-empirical-distributions}}
+To understand what cosine similarity measurement is, one first needs to
+grasp the concept of translating words to a computer-readable form. In
+the field of natural language processsing there are two main types of
+words representation --- localist and distributed. One-hot encoding is
+an example of a method used to achieve a localist representation of
+words. Here, each vector contains information only about a single data
+point. This is achieved by first mapping categorical values (words) to
+integers and then to each integers a binary vector is assigned which
+contains only 0s except for the index of the integer, which is assigned
+1. An example of a localist representation is:
 
-\hypertarget{looking-at-the-islam-related-words}{%
-\section{Looking at the islam-related words}\label{looking-at-the-islam-related-words}}
+\begin{longtable}[]{@{}llllll@{}}
+\toprule
+word & 1 & 2 & 3 & 4 & 5\tabularnewline
+\midrule
+\endhead
+woman & 1 & 0 & 0 & 0 & 0\tabularnewline
+man & 0 & 1 & 0 & 0 & 0\tabularnewline
+girl & 0 & 0 & 1 & 0 & 0\tabularnewline
+boy & 0 & 0 & 0 & 1 & 0\tabularnewline
+monarch & 0 & 0 & 0 & 0 & 1\tabularnewline
+\bottomrule
+\end{longtable}
 
-\hypertarget{bayesian-model-structure-and-assumptions}{%
-\section{Bayesian model structure and assumptions}\label{bayesian-model-structure-and-assumptions}}
+In the example above it is clear that the length of the vectors
+increases with the number of words in a vocabulary. It is not a very
+computationally efficient representation. It has other flaws as well.
+For example, it is unable to capture the resemblance between words
+appearing in similar contexts.
 
-\hypertarget{choosing-predictors}{%
-\section{Choosing predictors}\label{choosing-predictors}}
+In contrast to the localist representation, a distributed representation
+returns vectors that contain continuous values instead of discrete 1s
+and 0s. Word embeddings are a class of various techniques that allow one
+to represent words as distributed vectors. Such learned representations
+of text have certain properties. At least prima facie, they store
+similar (or at least co-occurring) words close to each other in a vector
+space. An example of distributed representation is:
 
-\hypertarget{dataset-level-coefficients}{%
-\section{Dataset-level coefficients}\label{dataset-level-coefficients}}
+\begin{longtable}[]{@{}lllll@{}}
+\toprule
+word & 1 & 2 & 3 & 4\tabularnewline
+\midrule
+\endhead
+woman & 0.456 & 0.267 & 0.675 & 0.131\tabularnewline
+man & 0.451 & 0.897 & 0.472 & 0.088\tabularnewline
+girl & 0.604 & 0.262 & 0.414 & 0.706\tabularnewline
+boy & 0.279 & 0.172 & 0.475 & 0.010\tabularnewline
+monarch & 0.565 & 0.678 & 0.463 & 0.975\tabularnewline
+\bottomrule
+\end{longtable}
 
-\hypertarget{model-structure-and-assumptions}{%
-\section{Model structure and assumptions}\label{model-structure-and-assumptions}}
+One of the advantages of using a distributed representation is that one
+is able to represent an enormous number of concepts with a smaller
+number of units. It is also possible to better capture similarities as
+words of similar meanings can have similar numeric vectors.
 
-\hypertarget{protected-classes-in-reddit-and-google-embeddings}{%
-\section{Protected classes in Reddit and Google embeddings}\label{protected-classes-in-reddit-and-google-embeddings}}
+The numbers occurring in such representations are not random. They are
+learned in a process that uses a very shallow neural network. There are
+various types of techniques used for learning the vectors
+representations. One of the most straightforward ones is a skip-gram
+model. Given a word the model tries to predict its neighboring words
+from the sentence. The mathematics behind the process relies on the idea
+that the prediction concerns the conditional probability of the adjacent
+words. The algorithm tries to minimize the loss function, which
+penalizes the system for discrepancy with actual co-occurrence
+frequencies in the corpus. One can choose various parameters of the
+model, such as the window size that determines how many surrounding
+words the model should predict. After preparing such a fitted model, one
+takes only the learned weights from a neural network, and uses them as
+vectors in a word embeddings representation.
 
-\hypertarget{dataset-level-coefficients-after-debiasing}{%
-\section{Dataset-level coefficients after debiasing}\label{dataset-level-coefficients-after-debiasing}}
+Word embeddings have many applications in natural language processing.
+They are handy in document search and information retrieval. They also
+play their part in improving automatic translations. Well learned word
+representations may also contribute to the improvement of sentiment
+analysis or spam detection.
 
-\hypertarget{protected-classes-after-debiasing}{%
-\section{Protected classes after debiasing}\label{protected-classes-after-debiasing}}
+\section{Cosine similarity and
+distance}\label{cosine-similarity-and-distance}
 
-\hypertarget{discussion-and-summary}{%
-\chapter{Discussion and summary}\label{discussion-and-summary}}
+Cosine similarity is often used as a method of finding out whether
+vector representations for two words suggest that they are similar or
+somehow connected. Cosine similarity is the cosine of the angle between
+two vectors: the result of dividing their inner product (dot product
+usually) by the product of their magnitudes.
 
-We propose the use of Bayesian methods to measure uncertainty in bias detection.
-There are a few advantages of this method.
-One of them is that including uncertainty enables one to directly observe the influence of sample sizes.
-Analyzing individual words and connection coefficients, one may notice how \texttt{neutral} words have smaller uncertainty intervals and \texttt{different} or \texttt{associated} quite the opposite.
-One of the reasons for such outcome is that we used approximately 230 neutral words and only between 11-25 (the number varies from class to class) stereotypical attributes from \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) article.
-In our approach, we also pay attention to the distribution and details regarding anomalous values.
-With the use of simple visualizations that we introduced before, we were able to indicate suspicious cosine distance values.
-Additionally, we compare in details how the cosine distance values and uncertainty change after the debiasing.
-One can verify then how the individual vectors changed and if it is what was expected.
-Our analysis with the use of Bayesian method gave us new ideas and hypothesis concerning not only the bias detection method but also the efficiency of debiasing itself.
+It is worth mentioning one more point concerning cosine similarity.
+After the vectors are normalized to have length equal to 1, the inner
+product itself (often dot product) is used to measure the similarity,
+because it then equals the cosine similarity.
 
-We created a summary table for each of the datasets: Reddit, Reddit Debiased, and Google word embeddings.
-Let us first analyse the general observations from estimated coefficients mean introduced in 3.6.
-DATASET-LEVEL COEFFICIENTS.
-For Google embeddings the HPDI for all classes coefficients (associated, different, human, and none) has an interval that includes zero.
-This can lead one to a conclusion that the impact for associated, different, human and neutral attribute is, when averaged, quite similar.
-This indicates how including the uncertainty may change the use and interpretation of \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) MAC metric.
-It seems that if one focuses only on differences between means of means, it is too simplistic.
-In case of Reddit word embeddings the situation is similar although HPDI interval is below 0 for Gender class when looking at \texttt{associated} and \texttt{different} mean coefficient.
-This can suggest that there is indeed slightly stronger impact of these attributes.
-One should also notice how in general \texttt{associated} and \texttt{different} coefficients have quite similar HPDI range, the highest observed absolute difference is equal to only 0.1.
-This suggests that again the impact of associated attributes and difference ones is not clear at first sight when looking at averaged coefficients.
-Finally, let's compare the HPDI intervals for Reddit and Reddit debiased datasets.
-For Religion and Race dataset there is a minor shift (in absolute values the highest change is equal to approximately 0.1) of the mean coefficients towards zero.
-However for Gender dataset there is no significant change.
-This is of course the general look at the data, let's now analyze the individual words.
+\begin{align} \tag{Sim}
+\mathsf{cosineSimilarity}(A,B) & = \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert}
+\end{align}
 
-In Reddit table one can observe that for Religion class the cosine distance results for the associated attributes are for approximately 60\% of the words close to human attributes as well.
-This can suggest that in some cases words concerning humans can have higher similarity with some protected words independently if they are stereotypical or neutral-human words.
-One should also pay attention to the fact that for all of the \texttt{associated} and \texttt{different} attributes in Religion class the uncertainty interval overlaps at some point.
-What is even more surprising is that for protected words, such as ``torah'' the \texttt{associated} attribute has the the cosine distance slightly over 1, which means no positive similarity!
-If the protected words that we chose do not have high similarity with harmful associated attributes, then one should consider at least three scenarios.
-The first one is that the choice of the protected words and attributes may be corrupted.
-The second one is that the metric is not able to catch the hidden bias properly.
-The third one is that there is actually no bias between the words.
-Regardless of which scenario one considers, it is essential to take a look at the individual values before averaging them or aggregating in other ways.
-It seems that using Bayesian method can enhance the process of verifying the hypotheses concerning the choice of protected words and attributes.
+Cosine similarity is considered a proper tool for this operation as its
+result has a clear connection to geometry and at least for a low number
+of dimensions may be easily interpreted. Using this scale, one can
+compare vector similarities in a fairly clear manner. When the vectors
+are aligned perpendicularly to each other, their similarity equals 0
+(which is the same as the cosine of 90 degrees). This tells us that the
+similarity between the vectors is small. As the angle between vectors
+decreases, cosine similarity approaches one, which stands for the
+greatest similarity. It is also possible to obtain negative cosine
+similarity, if the value approaches -1 it means that the words are
+contrary to each other.
 
-Surprisingly in Gender class one can observe high cosine similarity values between some female stereotypical professions and male protected words.
-If a word stereotypically associated with females has low cosine distance to male protected words, then one should investigate the issue further.
-The reasons for this unexpected cosine distance may lie in the frequency of appearance of the protected words in the raw data.
-Some of the groups (like Muslim people or females) may have less representation in the data that is taken as an input for word embeddings.
-Therefore, if we assume that the MAC detects actually co-occurrence only, it makes sense that they can have high similarity with associated attributes and lower similarity with different ones.
-At the same time male protected words and other religions can have high similarity with different attributes because they have greater representation in the dataset in general and occur close to much more concepts.
-Cosine distance seems to capture the information on the co-occurrence of words and not on the semantic similarity strictly speaking.
+One of the limitations of this measure is that it informs us only about
+similarities between vectors in terms of their orientation. However, it
+is often argued that in comparing words in terms of this metric, the
+magnitude of vectors may be treated as irrelevant, as the most important
+information pertains to direction.
+
+In what follows, it is important to distinguish between cosine
+similarity and cosine distance, defined as:
+
+\begin{align} \tag{Sim}
+\mathsf{cosineDistance}(A,B) &  = 1 - \mathsf{cosineSimilarity}(A,B)\\
+ &  = 1 - \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert} \nonumber
+\end{align}
+
+The greater the similarity between two vectors, the smaller the distance
+between them. The cosine distance ranges between 0 and 2. If the vectors
+are in an opposite directions, the cosine distance is 2. And if the
+vectors are extremely similar then the cosine distance is very close to
+0.
+
+One should note that cosine distance is not exactly a distance measure,
+as it does not meet triangle inequality requirements. The triangle
+inequality formula says that for any triangle, the sum of the lengths of
+any two sides must be greater than or equal to the length of the
+remaining side. As shown in
+\href{https://stats.stackexchange.com/questions/198080/proving-that-cosine-distance-function-defined-by-cosine-similarity-between-two-u}{discussion
+in stats.stackexchange.com} in the case of cosine distance it would have
+to fulfill this equation 1+cos-sim(𝐴,𝐶) \textless{} cos-sim(𝐴,𝐵) +
+cos-sim(𝐵,𝐶). If one chooses specific unit vectors it is easy to
+demonstrate that the triangle inequality is not preserved.
+
+\section{Cosine distance in a one-class bias
+detection}\label{cosine-distance-in-a-one-class-bias-detection}
+
+Bolukbasi et al. (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) define
+similarity between words as the outcome inner product of their
+normalized vectors. They focus on examining what the geometry of word
+embedding is in regard to ``he'' and ``she'' words. In other words,
+whether the similarity between those concepts and other words reflects
+expected gender stereotypes. They test this hypothesis by investigating
+whether there is a connection between word embeddings representing
+certain professions and words referring to gender. They also evaluate
+whether automatically produced analogies between words reflect the
+stereotypes as well.
+
+Here the gender bias of a word \(w\) is understood as its projection on
+the gender direction
+\(\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})\) (the
+gender direction is the top principal compontent of ten gender pair
+difference vectors). The underlying idea is that no bias is present if
+non-explicitly gender words are in equal distance to both elements in
+all explicitly gender pairs. Given the (ideally) gender neutral words
+\(N\) and the gender direction \(g\) the direct direct gender bias is
+defined as the average distance of the words in \(N\) from \(g\) (\(c\)
+is a parameter determining how strict we want to be):
+
+\begin{align}
+\mathsf{directBias_c(N,g)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},g)\vert^c}{\vert N \vert }
+\end{align}
+
+A very vivid way to follow their method of arguing that bias in word
+embeddings is real is to plot the values of inner product of chosen
+words. The plot below does not originate from the original paper but
+from
+\href{https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings}{kaggle.com
+R notebook using data from GloVe: Global Vectors for Word
+Representation}. However, similar visualization may be found there. Data
+used to create our plot is as follows.
+
+Occupations associated with feminine:
+\textbf{"homemaker", "nurse", "receptionist", "librarian", "socialite", "hairdresser", "nanny", "bookkeeper", "stylist", "housekeeper", "interior designer", "guidance counselor"}
+
+Occupations associated with masculine:
+\textbf{"maestro", "skipper", "protege", "philosopher", "captain", "architect", "financier", "warrior", "broadcaster", "magician", "fighter pilot", "boss"}
+
+\vspace{1mm} \footnotesize
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-1-1} \end{center}
+
+\normalsize
+
+The points in the plot above result from the calculation of the inner
+product of a chosen vector for a profession word and a vector for a
+gender word (she or he). Inner product of two vectors expresses
+similarity between words. This assumption originates from the geometry
+and properties of a vector space. In the picture one may observe a
+correlation between the occupation and gender. Stereotypical male
+professions are closer to the ``he'' word and stereotypical female
+professions have greater similarity with ``she'' word. One could argue
+that this is a hint that there is some hidden bias information stored in
+the words vectors.
+
+\section{Cosine distance in a multi-class bias
+detection}\label{cosine-distance-in-a-multi-class-bias-detection}
+
+Manzini et al.
+(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) present a
+different approach towards finding similarities between classes of
+words. The authors claim that texts available online are full of direct
+or indirect human stereotypes. As a result, word embeddings are prone to
+learn and maybe amplify those biases and propagate them further into AI
+models that are used for various applications. Cosine distance is used
+in the article as a measure to first argue for the existence of
+multi-class bias and then to show how through bias mitigation techniques
+the bias may be decreased.
+
+They modify WEAT to a multi-class setting, introducing Mean Average
+Cosine similarity as a measure of bias (in fact, in the paper they
+report distances rather than similarities). Let
+\(T = \{t_1, \dots, t_k\}\) be a class of protected word embeddings, and
+let each \(A_j\in A\) be a set of attributes stereotypically associated
+with a protected word). Then:
+
+\begin{align}
+S(t_i, A_j) & = \frac{1}{\vert A_j\vert}\sum_{a\in A_j}\mathsf{cos}(t,a) \\
+MAC(T,A) & = \frac{1}{\vert T \vert \,\vert A\vert}\sum_{t_i \in T }\sum_{A_j \in A} S(t_i,A_j)
+\end{align}
+
+As the code has been provided by the authors we were able to reconstruct
+their results. The main steps in the procedure are as follows. Let us go
+through an example that refers to the process of hard debiasing on
+religious attributes.
+
+First we load word embeddings from
+reddit.US.txt.tok.clean.cleanedforw2v.w2v dataset. The word embeddings
+have only 50 dimensions and the number of individual words from the
+dataset is 44895. The authors assume that the protected group should
+ideally not have high cosine similarity to stereotypical words. The word
+embeddings geometry should not place this group close to harmful
+stereotypes, if it is to be bias-free. For instance, let's look at the
+religion-related words.
+
+Protected words by religion type:
+
+\textbf{"jew"       : ["judaism", "jew", "synagogue", "torah", "rabbi"]}
+
+\textbf{"christian" : ["christianity", "christian", "church", "bible", "priest"]}
+
+\textbf{"muslim"    : ["islam", "muslim", "mosque", "quran", "imam"]}
+
+Stereotypical words by religion type:
+
+\textbf{"jew"       : ["greedy", "cheap", "hairy", "liberal"]}
+
+\textbf{"christian"       : ["judgemental", "conservative", "familial"]}
+
+\textbf{"muslim"       : ["violent", "terrorist", "dirty", "uneducated"]}
+
+We have prepared a table presenting the values of cosine distance for
+each protected word with each attribute (stereotype). The part of the
+results is shown in Table 2.3
+
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{religion <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
+\KeywordTok{colnames}\NormalTok{(religion) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
+                        \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
+\NormalTok{religion}\OperatorTok{$}\NormalTok{wordClass <-}\StringTok{ }\KeywordTok{as.factor}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass)}
+\KeywordTok{levels}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
+\KeywordTok{head}\NormalTok{(religion)  }\OperatorTok{%>%}\StringTok{  }\KeywordTok{kable}\NormalTok{(}\DataTypeTok{format =} \StringTok{"latex"}\NormalTok{,}\DataTypeTok{booktabs=}\NormalTok{T,}
+                      \DataTypeTok{linesep =} \StringTok{""}\NormalTok{,  }\DataTypeTok{escape =} \OtherTok{FALSE}\NormalTok{, }
+                      \DataTypeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\OperatorTok{%>%}
+\StringTok{                      }\KeywordTok{kable_styling}\NormalTok{(}\DataTypeTok{latex_options=}\KeywordTok{c}\NormalTok{(}\StringTok{"scale_down"}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{table}
+
+\caption{\label{tab:religionTableHeadEarly}Head of the religion dataset.}
+\centering
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{lllrrl}
+\toprule
+protectedWord & wordToCompare & wordClass & cosineDistance & cosineSimilarity & connection\\
+\midrule
+judaism & violent & muslim & 0.7141939 & 0.2858061 & different\\
+judaism & terrorist & muslim & 0.7461333 & 0.2538667 & different\\
+judaism & dirty & muslim & 1.2002599 & -0.2002599 & different\\
+judaism & uneducated & muslim & 0.7885469 & 0.2114531 & different\\
+judaism & greedy & jewish & 1.0026172 & -0.0026172 & associated\\
+judaism & cheap & jewish & 1.2323229 & -0.2323229 & associated\\
+\bottomrule
+\end{tabular}}
+\end{table}
+
+\normalsize
+
+\pagebreak
+
+In the article there was no analysis of individual distances, but the
+general look at the means. The authors introduced a metric that tries to
+generalize WEAT in measuring the presence of bias through the
+classification of multi-class bias in groups of words connected with
+gender, religion or race. In the process they first take the mean of
+cosine distances between a given protected word and attributes assigned
+to each stereotype. They do not differentiate between stereotypes
+associated with a word and stereotypes associated with different words
+(in the case of religion, stereotypes characteristic for Christianity
+has also cosine distance measured with for instance, Judaism or Islam).
+Then, after collecting the list of mean cosine distances, they average
+the list to obtain one final value representing the whole group, in this
+example religion, for which the final mean of all mean distances is
+equal to 0.859.
+
+In the article the authors also try to remove biases, as previously
+defined, from a word embedding. First they identify the bias subspace
+using Principal Component Analysis (PCA) which is a technique for
+dimensionality reduction. It is applied here to choose the subspace that
+contains the greatest amount of information. There can be many subspaces
+found in a given group. For example in terms of religion one can
+identify at least a few sets that are to grasp the concept of religion
+in general:
+
+\textbf{["judaism", "christianity", "islam"]}
+
+\textbf{["jew", "christian", "muslim"]}
+
+\textbf{["synagogue", "church", "mosque"]}
+
+The idea is to find a set that provides enough information to create
+from it a vector representing the concept of religion among words. This
+strategy is based on the idea that different dimensions of vectors
+contain different types of information and in some words in vector
+layers (subspaces) the information about religiousness is implicitly
+conveyed. In some cases this knowledge is useful, but in the case of
+harmful stereotypes one does not want to include the concept of religion
+in stereotypical words.
+
+After finding the bias subspace, they use it to modify the vector values
+individually so that their cosine distances towards certain words are
+changed. In the case of stereotypes the aim is to make the cosine
+distances larger so that the association between protected word and
+stereotype is smaller.
+
+In the final step there is an evaluation of the results. The cosine
+distances are calculated again but this time using debiased vocabulary.
+After taking the mean of all distances one final value is obtained and
+then it is compared with the average value from the beginning. If the
+cosine distances are on average greater than before then it leads the
+authors to the conclusion that improvement has been achieved. As the
+cosine distance increases, it is assumed that the association between
+protected and stereotypical words decreases.
+
+\section{Limitations of the approach}\label{limitations-of-the-approach}
+
+\textbf{1. Selection of attributes}
+
+The attributes are taken from different sources, there is no principled
+justification for their choice. From our analysis it will become clear
+that the list is rather uneven.
+
+There is no mention of methodology for deciding on the number of
+attributes necessary to decide a hypothesis on the given size of
+dataset. There are however some ways to estimate sample sizes needed to
+ensure that the results are meaningful if the effect is present. Our
+research will show that the numbers used are rather insufficient.
+
+\textbf{2. No attention to distributions and details}
+
+The authors use the mean of mean average cosine similarities to measure
+multi-class similarity between protected word and harmful stereotypes.
+They average the results until they obtain one final value to represent
+the mean cosine distance between protected word from a given class and
+the attributes of that class. If one takes closer look at the individual
+values that are taken for the calculations it turns out that there is a
+bunch of outliers and surprisingly dissimilar words. We approach this
+issue by providing the tables and visualizations of individual cosine
+distances to make sure that we get the proper insight into the data.
+
+\textbf{3. Hiding the impact of uncertainty}
+
+With such a method the uncertainty is not included which makes it even
+more difficult to give reasonable interpretations of the results. We
+propose the use of Bayesian method to obtain some understanding of the
+influence the uncertainty has on the interpretation of final results.
+
+\textbf{4. No word class distinction}
+
+In the original paper, words from all three religions were compared
+against all of the stereotypes, which means that there was no
+distinction between cases in which the stereotype is associated with a
+given religion, as opposed to the situation in which it is associated
+with another one. Not all of the stereotypical words have to be
+considered as harmful for all of the religions. In our analysis, we
+distinguish between stereotypes associated with a given group,
+stereotypes associated with different groups, and control groups:
+neutral words and stereotypes-free human predicates.
+
+\textbf{5. Interpreting the results}
+
+Assuming for a moment that the value of multi-class cosine distance is
+correct, one may question the interpretation. Manzini et al.
+(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) summarize the
+averages of cosine distance per group (gender, race, religion). For now
+let us focus now on analyzing the values relating to religious biases.
+Here is the relevant fragment of table:
+
+\begin{longtable}[]{@{}ll@{}}
+\toprule
+Religion Debiasing & MAC\tabularnewline
+\midrule
+\endhead
+Biased & 0.859\tabularnewline
+Hard Debiased & 0.934\tabularnewline
+Soft Debiased (\(\lambda\) = 0.2) & 0.894\tabularnewline
+\bottomrule
+\end{longtable}
+
+MAC stand for mean average cosine similarity, although in reality the
+the table contains mean distances. What may attract attention is the
+fact that the value of cosine distance in ``Biased'' category is already
+quite high even before debiasing. High cosine distance indicates low
+cosine similarity between values. One could think that average cosine
+similarity equal to approximately 0.141 is not significant enough to
+consider it as bias. However the authors aim to mitigate ``biases'' in
+vectors with such great distance to make it even larger.
+Methodologically the question is, on what basis is this small similarity
+still considered as a proof of the presence of bias, and whether these
+small changes are meaningful. This is in general the problem of scale
+and the lack of universal intervals. In contrast, proper intervals
+decide whether given cosine similarity is high enough to consider the
+words to be more similar than if we chose them at random. We will use
+highest posterior density intervals.
+
+\textbf{6. The curse of dimensionality}
+
+In our case, the curse of dimensionality may take place when there is an
+increase in the volume of data that results in adding extra dimensions
+to the Euclidean space. According to the article
+\href{https://analyticsindiamag.com/curse-of-dimensionality-and-what-beginners-should-do-to-overcome-it/}{Curse
+of dimensionality at analyticsindiamag.com} as the number of features
+increases, it may be harder and harder to obtain useful information from
+the data using the available algorithms. One may notice that more data
+should contribute to greater amount of information, but more information
+also means greater risk of noise and distractions in data. At the same
+time, modern solutions are often adapted to smaller dimensions and their
+results in higher ones are not intuitive, or may be prone to error.
+
+Using cosine similarity in high dimensions in word embeddings may also
+be prone to the curse of dimensionality. According to Venkat
+(\protect\hyperlink{ref-Venkat2018Curse}{2018}) there are reasons to
+consider this phenomenon when searching for word similarities in higher
+dimensions. An experiment is conducted that aims at showing how the
+similarity values and variation change as the number of dimensions
+increases. The hypothesis made in the paper states that two things will
+happen as the number of dimensions increase. First, the effort required
+to measure cosine similarity will be greater, and two, the similarity
+between data will blur out and have less variation. The authors generate
+random points with increasing number of dimensions where each dimension
+of a data point is given a value between 0 and 1. Then they pick one
+vector at random from each dimension class and calculate the cosine
+similarity between the chosen vector and the rest of the data. Then they
+check how the variation of values changes as the number of dimensions
+increases. It seems like the more dimensions there are, the smaller the
+variance and therefore it is less obvious how to interpret the resulting
+cosine similarities. Maybe the scale should be adjusted to the number of
+dimensions and variance so that it still gives us sensible information
+about data. Yet, according to some articles cosine similarity in high
+dimensions is not reliable enough as it may be the case that choosing
+words at random may result in similar values as when picking them
+consciously.
+
+\begin{figure}
+\centering
+\includegraphics{../images/curseOfDimensionality.png}
+\caption{curse of dimensionality, number of dimensions on the x axis,
+standard deviation of similarity on the y axis}
+\end{figure}
+
+\chapter{Walkthrough with the religion
+dataset}\label{walkthrough-with-the-religion-dataset}
+
+\section{Loading and understanding the
+dataset}\label{loading-and-understanding-the-dataset}
+
+We start with loading the libraries needed for the analysis.
+
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(ggplot2)}
+\KeywordTok{library}\NormalTok{(ggthemes)}
+\KeywordTok{library}\NormalTok{(rethinking)}
+\KeywordTok{library}\NormalTok{(tidyverse)}
+\KeywordTok{library}\NormalTok{(ggpubr)}
+\KeywordTok{library}\NormalTok{(kableExtra)}
+\KeywordTok{library}\NormalTok{(dplyr)}
+\KeywordTok{library}\NormalTok{(ggExtra)}
+\KeywordTok{library}\NormalTok{(cowplot)}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+We will use the choice of protected words and stereotypical predicates
+used in REF. This is a decent point of departure, not only we want to
+compare our method to that of REF, but also because this data format is
+fairly general (as contrasted, say, with a set up for binary
+stereotypes). Note also that the method we develop here can fairly
+easily be run for different stereotypization patterns. Let's start with
+explaining the method and its deployment using a dataset obtained for
+the religion-related protected words.
+
+Let's load, clean a bit and inspect the head of the religion dataset we
+prepared. In order to obtain this dataset, we calculated the cosine
+distance between each protected word and each word from both the
+bias-related attribute groups, which were used in the original study,
+and to neutral and human control attributes which we added as control
+groups. For instance, for religion, the bias-related predicates (coming
+from the original study in REF) include muslim bias attributes, jew bias
+attributes, christian bias attributes (see a list in the APPENDIX).
+
+We decided to add control groups in the form of two classes --- neutral
+words and human-related words. Without a proper control group it is
+quite hard to compare the resulting cosine distances and decide on their
+significance in bias detection. We prepared approximately 230 neutral
+words to double-check the prima-facie neutral hypothesis that their
+cosine similarity to the protected words will oscillate around 0 (that
+is, the distances will be around 1). This provides us with a more
+reliable point of reference. Moreover, we added human attributes that
+are associated with people in general to investigate whether the smaller
+cosine distance between protected words and stereotypes can result
+simply from the fact that the stereotype predicates are associated with
+humans. For two control groups, we have randomly drawn 230 words that do
+not express any property usually attributed to humans, and human related
+attributes.
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{religion <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
+\KeywordTok{colnames}\NormalTok{(religion) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
+                        \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
+\CommentTok{#head(religion)}
+\CommentTok{#library(plyr)}
+\NormalTok{religion}\OperatorTok{$}\NormalTok{wordClass <-}\StringTok{ }\KeywordTok{as.factor}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass)}
+\CommentTok{#levels(religion$wordClass)}
+\CommentTok{#religion$wordClass <- mapvalues(religion$wordClass, }
+\CommentTok{#                          from = levels(religion$wordClass),}
+\CommentTok{#                          to = c("christian","human","jewish","muslim","neutral"))}
+\KeywordTok{levels}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
+\KeywordTok{head}\NormalTok{(religion)  }\OperatorTok{%>%}\StringTok{  }\KeywordTok{kable}\NormalTok{(}\DataTypeTok{format =} \StringTok{"latex"}\NormalTok{,}\DataTypeTok{booktabs=}\NormalTok{T,}
+                      \DataTypeTok{linesep =} \StringTok{""}\NormalTok{,  }\DataTypeTok{escape =} \OtherTok{FALSE}\NormalTok{, }
+                      \DataTypeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\OperatorTok{%>%}\StringTok{      }\KeywordTok{kable_styling}\NormalTok{(}\DataTypeTok{latex_options=}\KeywordTok{c}\NormalTok{(}\StringTok{"scale_down"}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{table}
+
+\caption{\label{tab:religionTableHead}Head of the religion dataset.}
+\centering
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{lllrrl}
+\toprule
+protectedWord & wordToCompare & wordClass & cosineDistance & cosineSimilarity & connection\\
+\midrule
+judaism & violent & muslim & 0.7141939 & 0.2858061 & different\\
+judaism & terrorist & muslim & 0.7461333 & 0.2538667 & different\\
+judaism & dirty & muslim & 1.2002599 & -0.2002599 & different\\
+judaism & uneducated & muslim & 0.7885469 & 0.2114531 & different\\
+judaism & greedy & jewish & 1.0026172 & -0.0026172 & associated\\
+judaism & cheap & jewish & 1.2323229 & -0.2323229 & associated\\
+\bottomrule
+\end{tabular}}
+\end{table}
+
+\normalsize
+
+The \texttt{protectedWord} column contains words from a protected class
+that (in a perfect world according to the assumptions of the orignal
+study) should not be associated with harmful stereotypes.
+\texttt{wordToCompare} contains attributes, including stereotypes and
+control group words. For each row we compute the cosine distances
+between a given protected word and a given attribute word.
+\texttt{wordClass} tells us which class an attribute is supposed to be
+stereotypically associated with, that is, whether the word from
+\texttt{wordToCompare} is associated stereotypically with jews,
+christians or muslims, or whether it belongs to a control group.
+\texttt{cosineDistance} is simply a calculation of the cosine distance
+between protected word and atrribute. \texttt{cosineSimilarity} contains
+the result of substracting cosine distance from 1. \texttt{connection}
+contains information about the relation type between a protected word
+and an attribute. If the attribute is e.g.~a harmful jewish stereotype
+and the protected word is also from the judaism group, the connection
+has value \texttt{associated}. If the attribute is still stereotypically
+jewish, but the protected word comes from another religion, the
+connection is labelled as \texttt{different}. If the attribute belongs
+to a neutral group then the connection is labelled as \texttt{none} and
+if an attribute belongs to the \texttt{human} class, then the connection
+is labelled as \texttt{human}.
+
+\section{First look at the empirical
+distributions}\label{first-look-at-the-empirical-distributions}
+
+First let's take a look at the empirical distribution of distances by
+the connection type, initially ignoring the human control class for now.
+
+\vspace{1mm} \footnotesize
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-3-1} \end{center}
+
+\normalsize
+
+The first impression is that while there is a shift for associated words
+towards smaller cosine distances as compared to the neutral words,
+slightly surprisingly a slightly weaker shift in the same direction is
+visible for attributes associated with different stereotypes. Moreover,
+the empirical distributions overlap to alarge extent and the means
+grouped by connection type do not seem too far from each other. In fact,
+as there is a lot of variety in the cosine distances (as we will soon
+see), we need to gauge the uncertainty involved, and to look more
+carefully at individual protected words to get a better idea of how the
+cosine distance distribution changes for different attribute groups and
+different protected classes. Now, let's add the human attributes to the
+picture:
+
+\vspace{1mm} \footnotesize
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-4-1} \end{center}
+
+\normalsize
+
+\noindent Notice that the distribution for \texttt{human} (even though
+we did our best not to include in it any stereotype-related atributes)
+is left-skewed, with much overlap with \texttt{associated} and
+\texttt{different}, which illustrates the need to take being associated
+with humans as an important predictor.
+
+Our focus lies in \texttt{connection} as a predictor. Morever, later on
+we'll be interested in looking at the protected words separately, and at
+protected words split by connection. For technical reasons it is useful
+to represent these factors as integer vectors.
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{religion}\OperatorTok{$}\NormalTok{con <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(religion}\OperatorTok{$}\NormalTok{connection)}
+\NormalTok{religion}\OperatorTok{$}\NormalTok{pw <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(religion}\OperatorTok{$}\NormalTok{protectedWord)}
+\NormalTok{religion}\OperatorTok{$}\NormalTok{pwFactor <-}\StringTok{ }\KeywordTok{factor}\NormalTok{(}\KeywordTok{paste0}\NormalTok{(religion}\OperatorTok{$}\NormalTok{protectedWord, religion}\OperatorTok{$}\NormalTok{connection))}
+\NormalTok{religion}\OperatorTok{$}\NormalTok{pwIndex <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(religion}\OperatorTok{$}\NormalTok{pwFactor)}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+A short script, \texttt{cleanDataset} to make this faster, so
+equivalently:
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{source}\NormalTok{(}\StringTok{"../functions/cleanDataset.R"}\NormalTok{)}
+\NormalTok{religion <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
+\NormalTok{religion <-}\StringTok{ }\KeywordTok{cleanDataset}\NormalTok{(religion,}\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+\section{Looking at the islam-related
+words}\label{looking-at-the-islam-related-words}
+
+For now, let's focus on five protected words related to islam (``imam'',
+``islam'', ``mosque'', ``muslim'', and ``quran''). The word list
+associates with islam four stereotypical attributes (``violent'',
+``terrorist'', ``uneducated'' and ``dirty''). First, we select and plot
+the empirical distributions for these protected words.
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(tidyverse)}
+\NormalTok{muslimWords <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"imam"}\NormalTok{,}\StringTok{"islam"}\NormalTok{,}\StringTok{"mosque"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"quran"}\NormalTok{)}
+\NormalTok{muslim <-}\StringTok{ }\NormalTok{religion }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(protectedWord }\OperatorTok{%in%}\StringTok{ }\NormalTok{muslimWords)}
+\KeywordTok{ggplot}\NormalTok{(muslim, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{  cosineDistance, }\DataTypeTok{fill =}\NormalTok{ connection, }\DataTypeTok{color =}\NormalTok{ connection))}\OperatorTok{+}
+\StringTok{  }\KeywordTok{geom_density}\NormalTok{(}\DataTypeTok{alpha=}\FloatTok{0.6}\NormalTok{,}\DataTypeTok{size =} \FloatTok{.2}\NormalTok{)}\OperatorTok{+}\StringTok{ }
+\StringTok{  }\KeywordTok{scale_fill_manual}\NormalTok{(}\DataTypeTok{values =} \KeywordTok{c}\NormalTok{(}\StringTok{"orangered4"}\NormalTok{,}\StringTok{"chartreuse4"}\NormalTok{, }\StringTok{"skyblue"}\NormalTok{, }\StringTok{"gray"}\NormalTok{))}\OperatorTok{+}
+\StringTok{  }\KeywordTok{scale_x_continuous}\NormalTok{(}\DataTypeTok{breaks =} \KeywordTok{seq}\NormalTok{(}\FloatTok{0.3}\NormalTok{,}\FloatTok{1.5}\NormalTok{, }\DataTypeTok{by =} \FloatTok{0.1}\NormalTok{))}\OperatorTok{+}\KeywordTok{xlab}\NormalTok{(}\StringTok{"cosine distance"}\NormalTok{)}\OperatorTok{+}
+\StringTok{  }\KeywordTok{scale_color_manual}\NormalTok{(}\DataTypeTok{values =} \KeywordTok{c}\NormalTok{(}\StringTok{"orangered4"}\NormalTok{,}\StringTok{"chartreuse4"}\NormalTok{,}\StringTok{"skyblue"}\NormalTok{,}\StringTok{"gray"}\NormalTok{))}\OperatorTok{+}
+\StringTok{  }\KeywordTok{theme_tufte}\NormalTok{()}\OperatorTok{+}\KeywordTok{ggtitle}\NormalTok{(}\StringTok{"Empirical distribution of distances (muslim)"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-7-1} \end{center}
+
+\normalsize
+
+\noindent Once we focus on words related to islam, the associated bias
+seems to be stronger than in the whole dataset. This is a step towards
+illustrating that the distribution of bias is uneven.
+
+Now, say we want to look at a single protected word. Since the dataset
+also contains comparison multiple control neutral and human attributes,
+we randomly select only 5 from \texttt{none} and 5 from \texttt{human}
+control groups of those for the visualisation purposes.
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(tidyverse)}
+\NormalTok{muslimClass <-}\StringTok{ }\NormalTok{muslim }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(protectedWord }\OperatorTok{==}\StringTok{ "muslim"}\NormalTok{)}
+\NormalTok{neutralSample <-}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\KeywordTok{filter}\NormalTok{(muslimClass,connection }\OperatorTok{==}\StringTok{ "none"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
+\NormalTok{humanSample <-}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\KeywordTok{filter}\NormalTok{(muslimClass,connection }\OperatorTok{==}\StringTok{ "human"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
+\NormalTok{muslimVis <-}\StringTok{ }\NormalTok{muslimClass }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(connection }\OperatorTok{!=}\StringTok{ "none"} \OperatorTok{&}\StringTok{ }\NormalTok{connection }\OperatorTok{!=}\StringTok{"human"}\NormalTok{)}
+\NormalTok{muslimVis <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(muslimVis,neutralSample,humanSample)}
+
+\CommentTok{#we plug in our visualisation script}
+\KeywordTok{source}\NormalTok{(}\StringTok{"../functions/visualisationTools.R"}\NormalTok{)}
+\CommentTok{#two arguments: dataset and protected word}
+\KeywordTok{visualiseProtected}\NormalTok{(muslimVis,}\StringTok{"muslim"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/tableMuslimActive-1} \end{center}
+
+\normalsize
+
+Note that the distance between the grey point and the other points is
+proportional to cosine distance, the non-grey point size is proportional
+to cosine similarity to the protected word, and color groups by the
+connection type. So for \texttt{muslim} it seems that the stereotypes
+coming from the word list are fairly well visible. To give you some
+taste of how uneven the dataset is, compare this to what happens with
+\texttt{priest}.
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(tidyverse)}
+\NormalTok{priestClass <-}\StringTok{ }\NormalTok{religion }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(protectedWord }\OperatorTok{==}\StringTok{ "priest"}\NormalTok{)}
+\NormalTok{neutralSample <-}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\KeywordTok{filter}\NormalTok{(priestClass,connection }\OperatorTok{==}\StringTok{ "none"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
+\NormalTok{humanSample <-}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\KeywordTok{filter}\NormalTok{(priestClass,connection }\OperatorTok{==}\StringTok{ "human"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
+\NormalTok{priestVis <-}\StringTok{ }\NormalTok{priestClass }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(connection }\OperatorTok{!=}\StringTok{ "none"} \OperatorTok{&}\StringTok{ }\NormalTok{connection }\OperatorTok{!=}\StringTok{"human"}\NormalTok{)}
+\NormalTok{priestVis <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(priestVis,neutralSample,humanSample)}
+
+\CommentTok{#we plug in our visualisation script}
+\KeywordTok{source}\NormalTok{(}\StringTok{"../functions/visualisationTools.R"}\NormalTok{)}
+\CommentTok{#two arguments: dataset and protected word}
+\KeywordTok{visualiseProtected}\NormalTok{(priestVis,}\StringTok{"priest"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-8-1} \end{center}
+
+\normalsize
+
+\noindent Here you can see that some human attributes are closer than
+stereotype attributes, and that there is no clear reason to claim that
+\texttt{associated} attributes are closer than \texttt{different} or
+\texttt{human} attributes. This, again, illustrates the need of
+case-by-case analysis with control groups.
+
+The general idea now is that the word lists provided in different pieces
+of research are just samples of attributes associates with various
+stereotypes and should be treated as such: the uncertainty involved and
+the sample sizes should have clear impact on our estimates.
+
+\section{Bayesian model structure and
+assumptions}\label{bayesian-model-structure-and-assumptions}
+
+We will now think of cosine distance as the output variable, and will
+build a few Bayesian models to compare. First, we just build a baseline
+model which estimates cosine distance to the attributes separately for
+each protected word. The underlying idea is that different protected
+words might in general have different relations to all the attributes
+and these relations should be our point of departure.
+
+Here is the intuition behind the mathematical Bayesian model involved.
+Our outcome variable is \texttt{cosine\ difference}, which we take to me
+normally distributed around the predicted mean for a given protected
+word (that is, we assume the residuals are normally distributed). The
+simplest model specification is:
+
+\begin{align}
+cosineDistance_i  & \sim dnorm(\mu_i, \sigma) \\
+\mu_i & = m_{pw} \\
+m_{pw} & ~ dnorm(1,.5) \\
+\sigma &\sim  dcauchy(0,1)
+\end{align}
+
+That is, we assume the estimated means might be different for diferent
+protected words and our prior for the mean and the overal standard
+deviation are normal with mean 1 and sd=.5 and half-cauchy with
+parameters \texttt{0,1}. Further on we'll also estimate additional
+impact the connection type may have. For this impact we take a slightly
+skeptical prior centered around 0 distributed normally with sd = 1.
+These are fairly weak and slightly skeptical regularizing priors, which
+can be illustrated as follows:
+
+\vspace{2mm}
+
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/priorsVis-1} \end{center}
+
+\section{Choosing predictors}\label{choosing-predictors}
+
+Now we can define and compile the baseline model. Its parameters will
+have a posterior distribution obtained using either Hamiltionian Monte
+Carlo methods (STAN) available through the \texttt{rethinking} package.
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(rethinking)}
+\KeywordTok{options}\NormalTok{(}\DataTypeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\OtherTok{TRUE}\NormalTok{ )}
+\NormalTok{religionBaseline <-}\StringTok{ }\KeywordTok{ulam}\NormalTok{(}
+  \KeywordTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu <-}\StringTok{ }\NormalTok{m[pw],}
+\NormalTok{    m[pw] }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\OperatorTok{~}\StringTok{ }\KeywordTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\NormalTok{  ),}
+  \DataTypeTok{data =}\NormalTok{ religion,}
+  \DataTypeTok{chains=}\DecValTok{2}\NormalTok{ , }\DataTypeTok{iter=}\DecValTok{4000}\NormalTok{ , }\DataTypeTok{warmup=}\DecValTok{1000}\NormalTok{,}
+  \DataTypeTok{start=} \KeywordTok{list}\NormalTok{(}\DataTypeTok{mu =} \DecValTok{1}\NormalTok{, }\DataTypeTok{co =} \DecValTok{0}\NormalTok{, }\DataTypeTok{sigma=} \FloatTok{.3}\NormalTok{),}
+  \DataTypeTok{log_lik =} \OtherTok{TRUE}\NormalTok{, }\DataTypeTok{cores=}\DecValTok{4}
+\NormalTok{)}
+\CommentTok{#saving}
+\CommentTok{#saveRDS(religionBaseline, }
+\CommentTok{#file = "cosineAnalysis/models/religionBaseline.rds")}
+\end{Highlighting}
+\end{Shaded}
+
+The only reason we need it is the evaluation of connection as a
+predictor. Does including it in o the model improve the situation? To
+investigate this, let's now build a model according to the following
+specification:
+
+\begin{align}
+cosineDistance_i  & \sim dnorm(\mu_i, \sigma) \\
+\mu_i & = m_{pw} + co_{con}\\
+m_{pw} & ~ dnorm(1,.5) \\
+co_{con} & ~ dnorm(0,1) \\
+\sigma &\sim  dcauchy(0,1)
+\end{align}
+
+\noindent The idea now is that each connection type comes with its own
+coefficient \(co\) that has impact on mean distances for protected words
+taken separately.
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(rethinking)}
+\KeywordTok{options}\NormalTok{(}\DataTypeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\OtherTok{TRUE}\NormalTok{ )}
+\NormalTok{religionCoefs <-}\StringTok{ }\KeywordTok{ulam}\NormalTok{(}
+  \KeywordTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu <-}\StringTok{ }\NormalTok{m[pw] }\OperatorTok{+}\StringTok{ }\NormalTok{co[con],}
+\NormalTok{    m[pw] }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    co[con] }\OperatorTok{~}\KeywordTok{dnorm}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\OperatorTok{~}\StringTok{ }\KeywordTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\NormalTok{  ),}
+  \DataTypeTok{data =}\NormalTok{ religion,}
+  \DataTypeTok{chains=}\DecValTok{2}\NormalTok{ , }\DataTypeTok{iter=}\DecValTok{8000}\NormalTok{ , }\DataTypeTok{warmup=}\DecValTok{1000}\NormalTok{, }
+  \DataTypeTok{log_lik =} \OtherTok{TRUE}
+\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+\noindent First, let's see if this model is really better in terms of
+the Widely Acceptable Information Criterion (WAIC):
+
+\vspace{1mm} \footnotesize
+
+\begin{verbatim}
+##                   WAIC SE dWAIC dSE pWAIC weight
+## religionCoefs    -2328 93     0  NA    20      1
+## religionBaseline -2283 95    45  17    16      0
+\end{verbatim}
+
+\normalsize
+
+Clearly, it should be given weight 1 as compared to the baseline model.
+So far, we've learned that the connection type actually has predictive
+value. Let's take a look at the coefficient estimates:
+
+\vspace{1mm} \footnotesize
+
+\begin{verbatim}
+##              mean        sd       5.5%      94.5%    n_eff    Rhat4
+## co[1] -0.14956420 0.1151675 -0.3261650 0.03741930 294.2033 1.001449
+## co[2] -0.09880543 0.1145024 -0.2736985 0.08813271 291.5044 1.001564
+## co[3] -0.07282752 0.1133894 -0.2447778 0.11158986 287.7820 1.001627
+## co[4] -0.03103179 0.1131420 -0.2034442 0.15268770 286.8283 1.001606
+\end{verbatim}
+
+\normalsize
+
+\section{Dataset-level coefficients}\label{dataset-level-coefficients}
+
+\noindent Let's plot them together with their highest posterior density
+invervals, for the three topic groups.
+
+\begin{center}
+\begin{figure}[!htb]\centering
+   \begin{minipage}{0.55\textwidth}
+  \includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
+   \end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/religionGoogleCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/genderCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/genderGoogleCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/raceCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/raceGoogleCoeffs.jpeg}
+   \end{minipage}
+\end{figure}
+
+\end{center}
+
+\noindent What should strike us is that while the mean estimates of the
+coefficients indeed do differ a bit, usually the highest posterior
+density invervals all include zero, and so we do not have strong reasons
+to say that, say, as far as the whole religion dataset is involved,
+being associated indeed is connected with lower cosine distance. A
+second striking observation is that the estimated impact for associated
+stereotypes is quite often not too different from the estimated impact
+of attributes associated with different stereotypes, and both are
+sometimes not too far from the estimated impact for simply human
+attributes. In general, once the uncertainty involved is taken seriously
+by using control groups and statistical uncertainty estimation that does
+not dispose of pointwise data, the picture which focuses only on
+differences between means of means is too simplistic.
+
+But this doesn't mean important differences for some protected words are
+not there. For one thing, if you start with a word list that is very
+uneven, the actually not so bad status of some of the protected words
+might mask a pretty bad situation in which some other protected words
+are. For comparison, let's see what a model focused on words related to
+islam tells us.
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{#this is how we build the model}
+\NormalTok{religion <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"cosineAnalysis/datasets/religionReddit.csv"}\NormalTok{)[}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
+\KeywordTok{colnames}\NormalTok{(religion) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
+                        \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
+\KeywordTok{levels}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
+\NormalTok{muslimWords <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"imam"}\NormalTok{,}\StringTok{"islam"}\NormalTok{,}\StringTok{"mosque"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"quran"}\NormalTok{)}
+\NormalTok{muslim <-}\StringTok{ }\NormalTok{religion }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(protectedWord }\OperatorTok{%in%}\StringTok{ }\NormalTok{muslimWords)}
+\NormalTok{muslim}\OperatorTok{$}\NormalTok{protectedWord <-}\StringTok{ }\KeywordTok{droplevels}\NormalTok{(muslim}\OperatorTok{$}\NormalTok{protectedWord)}
+\NormalTok{muslim}\OperatorTok{$}\NormalTok{pw <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(muslim}\OperatorTok{$}\NormalTok{protectedWord)}
+\NormalTok{muslim}\OperatorTok{$}\NormalTok{con <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(muslim}\OperatorTok{$}\NormalTok{connection)}
+\NormalTok{muslim}\OperatorTok{$}\NormalTok{pwFactor <-}\StringTok{ }\KeywordTok{factor}\NormalTok{(}\KeywordTok{paste0}\NormalTok{(muslim}\OperatorTok{$}\NormalTok{protectedWord, muslim}\OperatorTok{$}\NormalTok{connection))}
+\NormalTok{muslim}\OperatorTok{$}\NormalTok{pwIndex <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(muslim}\OperatorTok{$}\NormalTok{pwFactor)}
+
+\NormalTok{islamCoefs <-}\StringTok{ }\KeywordTok{ulam}\NormalTok{(}
+  \KeywordTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu <-}\StringTok{ }\NormalTok{m[pw] }\OperatorTok{+}\StringTok{ }\NormalTok{co[con],}
+\NormalTok{    m[pw] }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    co[con] }\OperatorTok{~}\KeywordTok{dnorm}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\OperatorTok{~}\StringTok{ }\KeywordTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\NormalTok{  ),}
+  \DataTypeTok{data =}\NormalTok{ muslim,}
+  \DataTypeTok{chains=}\DecValTok{2}\NormalTok{ , }\DataTypeTok{iter=}\DecValTok{10000}\NormalTok{ , }\DataTypeTok{warmup=}\DecValTok{1000}\NormalTok{, }\DataTypeTok{cores =} \DecValTok{4}\NormalTok{,}
+  \DataTypeTok{log_lik =} \OtherTok{TRUE}
+\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+Let's take a look at the coefficients:
+
+\vspace{1mm} \footnotesize
+
+\begin{verbatim}
+##                mean        sd       5.5%      94.5%    n_eff    Rhat4
+## co[1] -0.1979930035 0.1708785 -0.4682696 0.07634977 1789.894 1.003445
+## co[2] -0.0334215769 0.1687587 -0.3021954 0.23720938 1738.720 1.003575
+## co[3] -0.0192492753 0.1675754 -0.2840596 0.24860974 1732.907 1.003755
+## co[4] -0.0003911363 0.1670610 -0.2661047 0.26815172 1723.758 1.003837
+\end{verbatim}
+
+\normalsize
+
+\begin{center}
+\includegraphics[width=8cm]{../images/islamCoeffs.jpeg}
+\end{center}
+
+\normalsize
+
+While muslim words were unusual in the sense that the disparity between
+associated attributes and others is stronger, the evidence is still not
+conclusive. This is because the variation even within islam-related
+words is large enough (and sample sizes sufficiently small) for all the
+highest posterior density invervals to still include zeros.
+
+So, it seems, taking a closer look does seem to make a difference. The
+question is, what happens if we do take a close look at the level of
+protected words?
+
+\chapter{Protected-word level
+analysis}\label{protected-word-level-analysis}
+
+\section{Model structure and
+assumptions}\label{model-structure-and-assumptions}
+
+Let's turn then to data analysis that takes a look at protected words
+separately. This time for each combination of a protected word and a
+connection status we will have a separate mean cosine distance estimate,
+each coming with its own highest posterior density interval. This means
+we will use indices that are result from all such combinations (and then
+we will split them up in the model precis to build visualisation, feel
+free to look at the \texttt{visualiseStats.R} script for details).
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{options}\NormalTok{(}\DataTypeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\OtherTok{TRUE}\NormalTok{ ) }\CommentTok{#removes install pop-up request}
+\NormalTok{religion}\OperatorTok{$}\NormalTok{pwFactor <-}\StringTok{ }\KeywordTok{factor}\NormalTok{(}\KeywordTok{paste0}\NormalTok{(religion}\OperatorTok{$}\NormalTok{protectedWord, }\StringTok{"-"}\NormalTok{, religion}\OperatorTok{$}\NormalTok{connection))}
+\NormalTok{religion}\OperatorTok{$}\NormalTok{pwIndex <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(religion}\OperatorTok{$}\NormalTok{pwFactor)}
+
+\NormalTok{religionSeparate <-}\StringTok{ }\KeywordTok{ulam}\NormalTok{(}
+  \KeywordTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu <-}\StringTok{ }\NormalTok{c[pwIndex],}
+\NormalTok{    c[pwIndex] }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\OperatorTok{~}\StringTok{ }\KeywordTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\NormalTok{  ),}
+  \DataTypeTok{data =}\NormalTok{ religion,}
+  \DataTypeTok{chains=}\DecValTok{2}\NormalTok{ , }\DataTypeTok{iter=}\DecValTok{10000}\NormalTok{ , }\DataTypeTok{warmup=}\DecValTok{1000}\NormalTok{,}
+  \DataTypeTok{start=}\KeywordTok{list}\NormalTok{(}\DataTypeTok{no =} \DecValTok{1}\NormalTok{, }\DataTypeTok{a =} \DecValTok{0}\NormalTok{, }\DataTypeTok{d =} \DecValTok{0}\NormalTok{, }\DataTypeTok{sigma=} \FloatTok{.3}\NormalTok{), }\DataTypeTok{log_lik =}  \OtherTok{TRUE}
+\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+\noindent Let's see if the individualized model does better than the
+previous models in light of WAIC which does add penalty for the number
+of parameters.
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{compareBaselineCoefsSeparate<-}\StringTok{ }\KeywordTok{readRDS}\NormalTok{(}\StringTok{"../datasets/compareBaselineCoefsSeparate.rds"}\NormalTok{)}
+\NormalTok{compareBaselineCoefsSeparate}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+##                   WAIC SE dWAIC dSE pWAIC weight
+## religionSeparate -2400 93     0  NA    60      1
+## religionCoefs    -2328 93    72  29    20      0
+## religionBaseline -2283 95   117  37    16      0
+\end{verbatim}
+
+\normalsize
+
+\section{Protected classes in Reddit and Google
+embeddings}\label{protected-classes-in-reddit-and-google-embeddings}
+
+It seems that we do want to prefer this model, despite its relative
+complication. Now, what does it tell us about the protected words? Let's
+visualise the predicted means together with 89\% highest posterior
+density intervals.
+
+\includegraphics[width=14cm]{../images/visReligionReddit.png}
+
+Before we move on, let's perform analogous analyses for the remaining
+types of supposed bias: gender and race (the model building is
+analogous).
+
+\includegraphics[width=14cm]{../images/visGenderReddit.png}
+
+\includegraphics[width=14cm]{../images/visRaceReddit.png}
+
+We first encountered the GoogleNews word embedding model in Bolukbasi et
+al. (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}). They use it for
+their calculations as written in
+\href{https://github.com/tolga-b/debiaswe}{Tolga Code}. We decided to
+compare the results obtained with the use of Reddit L2 model and the
+ones that we got by GoogleNews model. The details of the model can be
+found here \href{https://code.google.com/archive/p/word2vec/}{Google
+News model}. One of the main differences between the models is that
+Reddit word embeggings have 50 dimensions and GoogleNews word embeddings
+have 300 dimensions. As the dimension is increasing, the vector can
+capture much more information but this is not always the case. According
+to some researchers the amount of dimensions should be chosen by taking
+into account the type of corpus and its features. In our case both
+Reddit and GoogleNews models are already trained and ready to use so we
+do not analyse further the choice of their hyperparameters.
+
+\includegraphics[width=14cm]{../images/visReligionGoogle.png}
+
+\includegraphics[width=14cm]{../images/visGenderGoogle.png}
+
+\includegraphics[width=14cm]{../images/visRaceGoogle.png}
+
+\chapter{The role of debiasing}\label{the-role-of-debiasing}
+
+We also applied to the word embeddings hard debiasing method found in
+Manzini et al.
+(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}). As it was
+mentioned before, debiasing consists of two components: identifying the
+bias subspace and then removing this vector subspace from the chosen
+embeddings. The aim is to increase the cosine distance between protected
+words and the set of attributes. After debiasing the Reddit word
+embeddings we calculated the cosine distance again to see if there are
+any significant changes in terms of the similarity between protected
+words and the attributes.
+
+\section{Dataset-level coefficients after
+debiasing}\label{dataset-level-coefficients-after-debiasing}
+
+First, let's look at coefficient estimated for the whole datasets, as
+compared to their estimation prior to debiasing:
+
+\begin{center}
+\begin{figure}[!htb]\centering
+   \begin{minipage}{0.55\textwidth}
+  \includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
+   \end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/debiasedReligionRedditCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/genderCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/debiasedGenderRedditCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/raceCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/debiasedRaceRedditCoeffs.jpeg}
+   \end{minipage}
+\end{figure}
+
+\end{center}
+
+One may observe very slight changes in the estimated coeﬀicients. It
+seems that in religion dataset the changes are the most significant as
+the mean moves towards zero which stands for no similarity between
+words.
+
+\section{Protected classes after
+debiasing}\label{protected-classes-after-debiasing}
+
+Now, perhaps, the effects of debiasing will be better appreciated if we
+look at the level of protected words. After all, the hope is, the
+situation of extremely ill-positioned protected words have improved?
+
+\includegraphics[width=14cm]{../images/visDebReligionReddit.png}
+
+\includegraphics[width=14cm]{../images/visDebGenderReddit.png}
+
+\includegraphics[width=14cm]{../images/visDebRaceReddit.png}
+
+One may notice that in a religion data set most of the distances for
+\texttt{associated} words slightly increased. Some of the distances from
+\texttt{different} class are still smaller than for the
+\texttt{associated} class. The distances for all classes seem in most
+cases to cluster together even after debiasing. The reason may be that
+uncertainty is quite high in \texttt{associated} and \texttt{different}
+class because the number of samples is so small. Additionally in some
+cases distances from \texttt{human} class are smaller than for the
+\texttt{assosiated} class which also makes it less clear how to
+interpret the results, if the distances are really able to catch the
+bias presence.
+
+It is worth paying attention to the gender debiased dataset. The change
+is really minor, still zero out of HPDI range. The \texttt{different}
+attributes still have high similarity value. It seems that the distances
+for gender protected words are still clustered together which suggests
+that this method is not able to catch the complex bias presence. One
+could conclude that co-occurrence of certain terms does not always mean
+direct connection in terms of terms associations. The fact that some of
+stereotypically associated with female attributes still have high
+similarity with male protected words suggest that other techniques
+should be applied to detect and remove the bias.
+
+At the same time, there is a small improvement in race dataset. In some
+cases the distances just cluster together. In general distances for
+\texttt{assosiated} class are still higher than those for the rest of
+classes. One may assume that there reason for so small change is that
+either the metric does not indicate properly biases or that the issue is
+more complex and subtracting subspaces is not enough.
+
+In Manzini et al.
+(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) it is
+mentioned that debiasing with this method that is based on bias
+component removal may be insufficient. The reason is that words
+containing similar biases may still cluster together even after
+debiasing.
+
+\chapter{Discussion and summary}\label{discussion-and-summary}
+
+We propose the use of Bayesian methods to measure uncertainty in bias
+detection. There are a few advantages of this method. One of them is
+that including uncertainty enables one to directly observe the influence
+of sample sizes. Analyzing individual words and connection coefficients,
+one may notice how \texttt{neutral} words have smaller uncertainty
+intervals and \texttt{different} or \texttt{associated} quite the
+opposite. One of the reasons for such outcome is that we used
+approximately 230 neutral words and only between 11-25 (the number
+varies from class to class) stereotypical attributes from Manzini et al.
+(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) article. In
+our approach, we also pay attention to the distribution and details
+regarding anomalous values. With the use of simple visualizations that
+we introduced before, we were able to indicate suspicious cosine
+distance values. Additionally, we compare in details how the cosine
+distance values and uncertainty change after the debiasing. One can
+verify then how the individual vectors changed and if it is what was
+expected. Our analysis with the use of Bayesian method gave us new ideas
+and hypothesis concerning not only the bias detection method but also
+the efficiency of debiasing itself.
+
+We created a summary table for each of the datasets: Reddit, Reddit
+Debiased, and Google word embeddings. Let us first analyse the general
+observations from estimated coefficients mean introduced in 3.6.
+DATASET-LEVEL COEFFICIENTS. For Google embeddings the HPDI for all
+classes coefficients (associated, different, human, and none) has an
+interval that includes zero. This can lead one to a conclusion that the
+impact for associated, different, human and neutral attribute is, when
+averaged, quite similar. This indicates how including the uncertainty
+may change the use and interpretation of Manzini et al.
+(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) MAC metric.
+It seems that if one focuses only on differences between means of means,
+it is too simplistic. In case of Reddit word embeddings the situation is
+similar although HPDI interval is below 0 for Gender class when looking
+at \texttt{associated} and \texttt{different} mean coefficient. This can
+suggest that there is indeed slightly stronger impact of these
+attributes. One should also notice how in general \texttt{associated}
+and \texttt{different} coefficients have quite similar HPDI range, the
+highest observed absolute difference is equal to only 0.1. This suggests
+that again the impact of associated attributes and difference ones is
+not clear at first sight when looking at averaged coefficients. Finally,
+let's compare the HPDI intervals for Reddit and Reddit debiased
+datasets. For Religion and Race dataset there is a minor shift (in
+absolute values the highest change is equal to approximately 0.1) of the
+mean coefficients towards zero. However for Gender dataset there is no
+significant change. This is of course the general look at the data,
+let's now analyze the individual words.
+
+In Reddit table one can observe that for Religion class the cosine
+distance results for the associated attributes are for approximately
+60\% of the words close to human attributes as well. This can suggest
+that in some cases words concerning humans can have higher similarity
+with some protected words independently if they are stereotypical or
+neutral-human words. One should also pay attention to the fact that for
+all of the \texttt{associated} and \texttt{different} attributes in
+Religion class the uncertainty interval overlaps at some point. What is
+even more surprising is that for protected words, such as ``torah'' the
+\texttt{associated} attribute has the the cosine distance slightly over
+1, which means no positive similarity! If the protected words that we
+chose do not have high similarity with harmful associated attributes,
+then one should consider at least three scenarios. The first one is that
+the choice of the protected words and attributes may be corrupted. The
+second one is that the metric is not able to catch the hidden bias
+properly. The third one is that there is actually no bias between the
+words. Regardless of which scenario one considers, it is essential to
+take a look at the individual values before averaging them or
+aggregating in other ways. It seems that using Bayesian method can
+enhance the process of verifying the hypotheses concerning the choice of
+protected words and attributes.
+
+Surprisingly in Gender class one can observe high cosine similarity
+values between some female stereotypical professions and male protected
+words. If a word stereotypically associated with females has low cosine
+distance to male protected words, then one should investigate the issue
+further. The reasons for this unexpected cosine distance may lie in the
+frequency of appearance of the protected words in the raw data. Some of
+the groups (like Muslim people or females) may have less representation
+in the data that is taken as an input for word embeddings. Therefore, if
+we assume that the MAC detects actually co-occurrence only, it makes
+sense that they can have high similarity with associated attributes and
+lower similarity with different ones. At the same time male protected
+words and other religions can have high similarity with different
+attributes because they have greater representation in the dataset in
+general and occur close to much more concepts. Cosine distance seems to
+capture the information on the co-occurrence of words and not on the
+semantic similarity strictly speaking.
 
 In Google table one may notice interesting observations as well.
-Although GoogleNews dataset is larger, trained on different data sources and with the use of more dimensions, some of the results are similar to the ones obtained for the Reddit corpus.
-For Religion case almost all of the values for \texttt{associated} class have an intersection with \texttt{different} class as well.
-In the case of Race it is 100\% of the available words.
-This indicates again that it is not clear how the metric should be used.
-Quite different situation takes place for Gender data where the similarity between \texttt{associated} and \texttt{different} is present mostly for the male protected words.
-This means that male words have high similarity with both male stereotypical professions and female ones.
-However, for females the similarity is high mostly only for the female stereotypical professions.
-This observation could not be make when using MAC metric only as it requires investigating the individual words.
+Although GoogleNews dataset is larger, trained on different data sources
+and with the use of more dimensions, some of the results are similar to
+the ones obtained for the Reddit corpus. For Religion case almost all of
+the values for \texttt{associated} class have an intersection with
+\texttt{different} class as well. In the case of Race it is 100\% of the
+available words. This indicates again that it is not clear how the
+metric should be used. Quite different situation takes place for Gender
+data where the similarity between \texttt{associated} and
+\texttt{different} is present mostly for the male protected words. This
+means that male words have high similarity with both male stereotypical
+professions and female ones. However, for females the similarity is high
+mostly only for the female stereotypical professions. This observation
+could not be make when using MAC metric only as it requires
+investigating the individual words.
 
-When analyzing the Reddit debiased results one should ask a question of the change of the cosine distance values.
-It seems that not all of the protected words are treated equally when performing debiasing.
-In religion data the values for \texttt{associated} class moved towards 0 for most of the words except the ones for Islam religion where still the \texttt{associated} class has quite lower cosine distance than the \texttt{different} one.
-Similar situation takes place in Gender data where female words have still much lower cosine distance values for \texttt{associated} class than for the \texttt{different} one.
-In the case of males it is mostly almost the same interval for \texttt{associated} and \texttt{different}.
-As the Gender data is quite specific as the attributes are not harmful adjectives but (in an ideal world) neutral professions, the aim (if we follow MAC assumptions) should actually be to make the cosine distance same for both female and male protected words.
-However, as we pointed out the situation after debiasing can sometimes be better only for one protected group which is not the desired outcome.
-In the case of Religion data even after debiasing, all of the protected words have intersections between \texttt{associated} and \texttt{different} class and similarity greater than 0.
-As all of the attributes in religion data are negative and harmful stereotypes, one should not aim at making the distance between protected words and \texttt{associated}, and \texttt{different} class the same but rather to move it towards 0.
+When analyzing the Reddit debiased results one should ask a question of
+the change of the cosine distance values. It seems that not all of the
+protected words are treated equally when performing debiasing. In
+religion data the values for \texttt{associated} class moved towards 0
+for most of the words except the ones for Islam religion where still the
+\texttt{associated} class has quite lower cosine distance than the
+\texttt{different} one. Similar situation takes place in Gender data
+where female words have still much lower cosine distance values for
+\texttt{associated} class than for the \texttt{different} one. In the
+case of males it is mostly almost the same interval for
+\texttt{associated} and \texttt{different}. As the Gender data is quite
+specific as the attributes are not harmful adjectives but (in an ideal
+world) neutral professions, the aim (if we follow MAC assumptions)
+should actually be to make the cosine distance same for both female and
+male protected words. However, as we pointed out the situation after
+debiasing can sometimes be better only for one protected group which is
+not the desired outcome. In the case of Religion data even after
+debiasing, all of the protected words have intersections between
+\texttt{associated} and \texttt{different} class and similarity greater
+than 0. As all of the attributes in religion data are negative and
+harmful stereotypes, one should not aim at making the distance between
+protected words and \texttt{associated}, and \texttt{different} class
+the same but rather to move it towards 0.
 
-Let's summarize the results of the bias detection methods analysis.
-As it was presented above, using the bias detection methods without involving the uncertainty may bring the risk of limited insight into the data.
-One should be concerned with the limitations of the metric that evaluates the debiasing by the mean of the averaged mean approach.
-Additionally, one cannot be sure if the bias is still preserved after debiasing.
-The fact that all of the cosine distances for protected words and harmful attributes moved slightly to the right, does not mean that the bias is removed.
-It is shown in articles such as \protect\hyperlink{ref-Gonen2019Lipstick}{Gonen \& Goldberg} (\protect\hyperlink{ref-Gonen2019Lipstick}{2019}), that the bias can hide in the vector geometry and preserve even after applying popular debiasing methods.
-One of the reasons for that may be the fact that relative differences between words may be preserved even after debiasing.
-Another thing is that the choice of word lists used to verify the metric effectiveness is not well justified.
-The sample size is very small which leads to large HPDI.
-Some of the protected words and attributes originate from old articles and it seems that the bias phenomenon is quite dynamic and the data should be as up-to-date as it is possible.
-One could for example research the IAT examples to verify the effectiveness of MAC metric.
-One should remember that there is no clear interpretation for the values obtained with MAC metric from \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}).
-One may assume that if the cosine distance is close to 1 then it is a desired outcome as it means, according to cosine distance assumptions, that there is almost no similarity between the words.
-However what does it mean to be close to 0?
-If the averaged cosine distance is equal to 0.8, then should we still debias it?
-It is unclear what the criteria are.
-On one hand, it seems to be beneficial when the outcome is simplified as it is easier to compare results with one value per set.
-On the other hand, it is prone to misunderstanding of how to interpret the results and what threshold to assume.
+Let's summarize the results of the bias detection methods analysis. As
+it was presented above, using the bias detection methods without
+involving the uncertainty may bring the risk of limited insight into the
+data. One should be concerned with the limitations of the metric that
+evaluates the debiasing by the mean of the averaged mean approach.
+Additionally, one cannot be sure if the bias is still preserved after
+debiasing. The fact that all of the cosine distances for protected words
+and harmful attributes moved slightly to the right, does not mean that
+the bias is removed. It is shown in articles such as Gonen \& Goldberg
+(\protect\hyperlink{ref-Gonen2019Lipstick}{2019}), that the bias can
+hide in the vector geometry and preserve even after applying popular
+debiasing methods. One of the reasons for that may be the fact that
+relative differences between words may be preserved even after
+debiasing. Another thing is that the choice of word lists used to verify
+the metric effectiveness is not well justified. The sample size is very
+small which leads to large HPDI. Some of the protected words and
+attributes originate from old articles and it seems that the bias
+phenomenon is quite dynamic and the data should be as up-to-date as it
+is possible. One could for example research the IAT examples to verify
+the effectiveness of MAC metric. One should remember that there is no
+clear interpretation for the values obtained with MAC metric from
+Manzini et al.
+(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}). One may
+assume that if the cosine distance is close to 1 then it is a desired
+outcome as it means, according to cosine distance assumptions, that
+there is almost no similarity between the words. However what does it
+mean to be close to 0? If the averaged cosine distance is equal to 0.8,
+then should we still debias it? It is unclear what the criteria are. On
+one hand, it seems to be beneficial when the outcome is simplified as it
+is easier to compare results with one value per set. On the other hand,
+it is prone to misunderstanding of how to interpret the results and what
+threshold to assume.
 
-The bottom line is that if we want to take bias seriously, so should we approach the uncertainty involved in our estimations.
-There is no replacement for proper statistical evaluation that does not discard information about the uncertainty involved, larger word lists are needed, and visualization of the results for particular protected classes provides much better guidance than chasing a single metric based on a means of means.
+The bottom line is that if we want to take bias seriously, so should we
+approach the uncertainty involved in our estimations. There is no
+replacement for proper statistical evaluation that does not discard
+information about the uncertainty involved, larger word lists are
+needed, and visualization of the results for particular protected
+classes provides much better guidance than chasing a single metric based
+on a means of means.
 
@@
 \hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\hypertarget{ref-Gonen2019Lipstick}{}%
-Gonen, H., \& Goldberg, Y.
-(2019).
-Lipstick on a pig: Debiasing methods cover up systematic gender biases in word embeddings but do not remove them.
-\emph{CoRR}, \emph{abs/1903.03862}.
-Retrieved from \url{http://arxiv.org/abs/1903.03862}
+\hypertarget{ref-Bolukbasi2016Man}{}
+Bolukbasi, T., Chang, K., Zou, J. Y., Saligrama, V., \& Kalai, A.
+(2016). Man is to computer programmer as woman is to homemaker?
+Debiasing word embeddings. \emph{CoRR}, \emph{abs/1607.06520}. Retrieved
+from \url{http://arxiv.org/abs/1607.06520}
 
-\leavevmode\hypertarget{ref-Manzini2019blackToCriminal}{}%
-Manzini, T., Lim, Y.
-C., Tsvetkov, Y., \& Black, A.
-W.
-(2019).
-Black is to criminal as caucasian is to police: Detecting and removing multiclass bias in word embeddings.
-Retrieved from \url{http://arxiv.org/abs/1904.04047}
+\hypertarget{ref-Gonen2019Lipstick}{}
+Gonen, H., \& Goldberg, Y. (2019). Lipstick on a pig: Debiasing methods
+cover up systematic gender biases in word embeddings but do not remove
+them. \emph{CoRR}, \emph{abs/1903.03862}. Retrieved from
+\url{http://arxiv.org/abs/1903.03862}
 
-\end{CSLReferences}
+\hypertarget{ref-Caliskan2017Semantics}{}
+Islam, A. C., Bryson, J. J., \& Narayanan, A. (2016). Semantics derived
+automatically from language corpora necessarily contain human biases.
+\emph{CoRR}, \emph{abs/1608.07187}. Retrieved from
+\url{http://arxiv.org/abs/1608.07187}
+
+\hypertarget{ref-Manzini2019blackToCriminal}{}
+Manzini, T., Lim, Y. C., Tsvetkov, Y., \& Black, A. W. (2019). Black is
+to criminal as caucasian is to police: Detecting and removing multiclass
+bias in word embeddings.
+
+\hypertarget{ref-Mehrabi2019Survey}{}
+Mehrabi, N., Morstatter, F., Saxena, N., Lerman, K., \& Galstyan, A.
+(2019). A survey on bias and fairness in machine learning. \emph{CoRR},
+\emph{abs/1908.09635}. Retrieved from
+\url{http://arxiv.org/abs/1908.09635}
+
+\hypertarget{ref-Nissim2019Fair}{}
+Nissim, M., Noord, R. van, \& Goot, R. van der. (2019). Fair is better
+than sensational: Man is to doctor as woman is to doctor. \emph{CoRR},
+\emph{abs/1905.09866}. Retrieved from
+\url{http://arxiv.org/abs/1905.09866}
+
+\hypertarget{ref-Venkat2018Curse}{}
+Venkat, N. (2018). The curse of dimensionality: Inside out.
 

### 2021-06-22 thesis/02-cosine.Rmd

@@
 A very vivid way to follow their method of arguing that bias in word embeddings is real is to plot the values of inner product of chosen words.
-The plot below does not originate from the original paper but from [kaggle.com R notebook using data from GloVe: Global Vectors for Word Representation](https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings).
-However, similar visualization may be found there.
+The plot below does not originate from the original paper but from [kaggle.com R notebook using data from GloVe: Global Vectors for Word Representation](https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings).^[[https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings](https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings).] However, similar visualization may be found there.
 Data used to create our plot is as follows.
@@
 
+\pagebreak
+
 \textbf{["judaism", "christianity", "islam"]}
@@
 As the cosine distance increases, it is assumed that the association between protected and stereotypical words decreases.
+
+\newpage
 
@@
 We propose the use of Bayesian method to obtain some understanding of the influence the uncertainty has on the interpretation of final results.
+
+\pagebreak
 

### 2021-06-22 thesis/03-bayesian.Rmd

@@
 In order to obtain this  dataset, we  calculated the cosine distance between each protected word and each word from both the bias-related attribute groups, which were used in the original study, and to neutral and human control attributes which we added  as control groups.
-For instance, for religion, the bias-related predicates (coming from the original study in @Manzini2019blackToCriminal) include muslim bias attributes, jew bias attributes, christian bias attributes (see a list in the APPENDIX).
+For instance, for religion, the bias-related predicates (coming from the original study in @Manzini2019blackToCriminal) include muslim bias attributes, jew bias attributes, christian bias attributes (see a list in the Appendix).
 
@@
 
+\pagebreak 
 \vspace{1mm}
@@
 
-```{r religionTableHead,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%", warning = FALSE, message = FALSE}
+```{r religionTableHead,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%", warning = FALSE, message = FALSE}
 religion <- read.csv("../datasets/religionReddit.csv")[-1]
@@
                       linesep = "",  escape = FALSE, 
-                      caption = "Head of the religion dataset.") %>%      kable_styling(latex_options=c("scale_down"))
+                      caption = "Head of the religion dataset.") %>%  
+                      kable_styling(latex_options=c("scale_down"))
 ```
@@
 \footnotesize
-```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%",message =FALSE, warning = FALSE}
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%",message =FALSE, warning = FALSE}
 library(tidyverse)
@@
 \footnotesize
-```{r tableMuslimActive,echo=TRUE,eval=TRUE,fig.align = "center",cache=FALSE, fig.show = "hold", out.width = "100%"}
+```{r tableMuslimActive,echo=FALSE,eval=TRUE,fig.align = "center",cache=FALSE, fig.show = "hold", out.width = "100%"}
 library(tidyverse)
@@
 \footnotesize
-```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
 library(tidyverse)
@@
 First, we just build a baseline model which estimates cosine distance to the attributes separately for each protected word.
-The underlying idea is that different protected words might in general have different relations to all the attributes and these relations should be our point of departure.
+The underlying idea is that different protected words might in general have different relations to all the attributes and these relations should be our point of departure.^[The construction of the Bayesian models and code for visualisations is due to Rafal Urbaniak.]
 
@@
 
-## Dataset-level coefficients
+## Dataset-level coefficients \label{sec:datasetsLevelCoeffs}
 
@@
 
+We leave a discussion of these results for later, for now let us also inspect the impact of debiasing.
+
 # The role of debiasing

### 2021-06-22 thesis/04-discussion.Rmd

@@
 Our analysis with the use of Bayesian method gave us new ideas and hypothesis concerning not only the bias detection method but also the efficiency of debiasing itself.
+We created  summary tables for each of the datasets: Google word embeddings (Table \@ref(tab:dTabG)), Reddit (Table \@ref(tab:dTabR)), and Reddit Debiased (Table \@ref(tab:dTabRD)).
 
-<!-- General insights - estimated coefficients mean -->
+\vspace{1mm}
+\footnotesize
+```{r dTabG,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+dTabG <- read.csv("../datasets/discussionTableGoogle.csv")
+rownames(dTabG) <-dTabG[,1]
+dTabG <- dTabG[,-1]
+colnames(dTabG) <- c("Religion (percent)","Gender (percent)", "Race (percent)" )
+dTabG  %>%  kable(format = "latex",booktabs=T,
+                      linesep = "",  escape = FALSE,
+                      caption = "Overview statistics  (GoogleNews embedding)")  %>%  
+                      kable_styling(latex_options=c("scale_down"))
+```
+\normalsize
 
-We created a summary table for each of the datasets: Reddit, Reddit Debiased, and Google word embeddings.
-Let us first analyse the general observations from estimated coefficients mean introduced in 3.6.
-Dataset-level coefficients.
+\vspace{1mm}
+\footnotesize
+```{r dTabR,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+dTabR <- read.csv("../datasets/discussionTableReddit.csv")
+rownames(dTabR) <-dTabR[,1]
+dTabR <- dTabR[,-1]
+colnames(dTabR) <- c("Religion (percent)","Gender (percent)", "Race (percent)" )
+dTabR  %>%  kable(format = "latex",booktabs=T,
+                      linesep = "",  escape = FALSE,
+                      caption = "Overview statistics  (Reddit embedding)") %>%  
+                      kable_styling(latex_options=c("scale_down"))
+```
+\normalsize
+
+\vspace{1mm}
+\footnotesize
+```{r dTabRD,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+dTabRD <- read.csv("../datasets/discussionTableRedditDebiased2.csv")
+rownames(dTabRD) <-dTabRD[,1]
+dTabRD <- dTabRD[,-1]
+colnames(dTabRD) <- c("Religion (percent)","Gender (percent)", "Race (percent)" )
+dTabRD  %>%  kable(format = "latex",booktabs=T,
+                      linesep = "",  escape = FALSE,
+                      caption = "Overview statistics  (Reddit embedding, debiased)") %>%  
+                      kable_styling(latex_options=c("scale_down"))
+```
+\normalsize
+
+Let us first analyse the general observations from estimated coefficients mean introduced in Section \ref{sec:datasetsLevelCoeffs} on dataset-level coefficients.
 For Google embeddings the HPDIs for all class coefficients (associated, different, human, and none) include zero.
@@
 
-Bibliography:
-
 <!--

### 2021-06-22 thesis/05-appendix.Rmd

# Appendix {-}

## Original wordlist from [@Manzini2019blackToCriminal] {-}

Definite sets are the protected words, analogy templates assign stereotypical attributes to protected words, test terms were included but are not  used in the original MAC calculations.

### Religion  {-}

\vspace{1mm}
\scriptsize
```{r,echo=TRUE,eval=FALSE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
{
	"definite_sets":[
		["judaism", "christianity", "islam"],
		["jew", "christian", "muslim"],
        ["synagogue", "church", "mosque"],
        ["torah", "bible", "quran"],
        ["rabbi", "priest", "imam"]
	],
	"analogy_templates":
	{
		"attribute": {
			"jew"       : ["greedy", "cheap", "hairy", "liberal"],
    		"christian" : ["judgemental", "conservative", "familial"],
    		"muslim"    : ["violent", "terrorist", "dirty", "uneducated"]
		}
	},
	"eval_targets":[
		["judaism", "christianity", "islam"],
		["jew", "christian", "muslim"],
        ["synagogue", "church", "mosque"],
        ["torah", "bible", "quran"],
        ["rabbi", "priest", "imam"]
	], 
	"testTerms":["chair", "house", "greedy", "terrorist", "dirty", "greedy"]
}
```
\normalsize

### Gender  {-}

\vspace{1mm}
\scriptsize
```{r,echo=TRUE,eval=FALSE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
{
	"definite_sets":[
		["he", "she"],
		["his", "hers"],
		["son", "daughter"],
		["father", "mother"],
		["male", "female"],
		["boy", "girl"],
		["uncle", "aunt"]
	],
	"analogy_templates":
	{
		"role": {
			"man"   : ["manager", "executive", "doctor", "lawyer", "programmer", 
			           "scientist", "soldier", "supervisor", "rancher", "janitor",
			           "firefighter", "officer"],
			"woman" : ["secretary", "nurse", "clerk", "artist", "homemaker", "dancer",
			           "singer", "librarian", "maid", "hairdresser", "stylist",
			           "receptionist", "counselor"]
		}
	}, 
	"eval_targets":[
		["he", "she"],
		["his", "hers"],
		["son", "daughter"],
		["father", "mother"],
		["male", "female"],
		["boy", "girl"],
		["uncle", "aunt"]
	], 
	"testTerms":["chair", "house", "supervisor", "secretary", "loud", "weak"]
}
```
\normalsize

### Race  {-}

\vspace{1mm}
\scriptsize
```{r,echo=TRUE,eval=FALSE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
{
	"definite_sets":[
		["black", "caucasian", "asian"],
		["african", "caucasian", "asian"],
		["black", "white", "asian"],
		["africa", "america", "asia"],
		["africa", "america", "china"],
		["africa", "europe", "asia"]
    ],
	"analogy_templates":
	{
		"role": {
			"caucasian": ["manager", "executive", "redneck", "hillbilly", "leader", "farmer"],
			"asian"    : ["doctor", "engineer", "laborer", "teacher"],
			"black"    : ["slave", "musician", "runner", "criminal", "homeless"]
		}
	}, 
	"eval_targets":[
		["black", "caucasian", "asian"],
		["africa", "america", "asia"]
	], 
	"testTerms":["chair", "house", "smart", "criminal", "executive", "farmer"]
}
```
\normalsize

## Our control groups {-}

## Human neutral attributes {-}

\vspace{1mm}
\scriptsize
```{r,echo=TRUE,eval=FALSE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
human_words = ['wear', 'walk', 'visitor', 'toy', 'tissue', 'throw', 'talk', 
               'speak', 'sleep', 'eye', 'enjoy', 'blogger', 'character', 
               'candidate', 'breakfast', 'supper', 'dinner', 'eat', 'drink',
               "carry", "run", "cast", "ask", "awake", "ear", "nose", "lunch"]
```
\normalsize

## Non-human neutral attributes {-}

\vspace{1mm}
\footnotesize
```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 10), tidy = TRUE)
```
\normalsize

\vspace{1mm}
\scriptsize
```{r,echo=TRUE,eval=FALSE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
neutral_words = ['liquor', 'pow', 'ballpark', 'glitchy', 'billy', 'dallas', 
                 'rip', 'called', 'outlooks', 'viet', 'floater', 'rattlesnake', 'exports',
                 'peruvian', 'recursion', 'shortfall', 'corrected', 'amicable',
                 'solutions', 'diagnostic', 'patently', 'flops', 'approx', 'percents',
                'lox', 'catapults', 'hamburger', 'engulfed', 'households', 'north',
                'snubbed', 'playtest', 'replayability', 'glottal', 'parable', 'gingers',
                'anachronism', 'organizing', 'reach', 'shtick', 'eleventh', 'cpu', 'ranked',
                'irreversibly', 'ponce', 'velociraptor', 'rubber', 'defects', 'puzzle',
                'smasher', 'northside', 'heft', 'observation', 'rectum', 'mystical',
                'telltale', 'remnants', 'inquiry', 'indisputable', 'boatload', 'lessening',
                'uselessness', 'observes', 'fictitious', 'repatriation', 'duh',
                'attic', 'schilling', 'charges', 'chatter', 'pad', 'smurfing',
                'worthiness', 'definitive', 'neat', 'homogenized','lexicon',
                'nationalized', 'earpiece', 'specializations', 'lapse',
                'concludes', 'weaving', 'apprentices', 'fri', 'younglings',
                'militias', 'inscriptions', 'gouda', 'lift', 'laboring', 
                'adaptive', 'lecture', 'hogging', 'thorne', 'fud', 'skews',
                'epistles', 'tagging', 'crud', 'two', 'rebalanced', 'payroll',
                'damned', 'approve', 'reason', 'formally', 'releasing', 'muddled',
                'mineral', 'shied', 'capital', 'nodded', 'escrow', 'unscientific',
                'recognizable', 'entitlement', 'disconnecting', 'marshals',
                'winamp', 'forceful', 'lowes', 'ptr', 'sip', 'pencils', 'stomachs',
                'goff', 'cg', 'backyard', 'uprooting', 'merging',
                'helpful', 'eid', 'trenchcoat', 'airlift', 'frothing',
                'pulls', 'volta', 'guinness', 'viewership',
                'eruption', 'peeves', 'goat', 'goofy', 'disbanding', 
                'relented', 'ratings', 'disputed', 'vitamins', 'singled',
                'hydroxide', 'telegraphed', 'mercantile', 'headache', 
                'muppets', 'petal', 'arrange', 'donovan', 'scrutinized',
                'spoil', 'examiner', 'ironed', 'maia', 'condensation',
                'receipt', 'solider', 'tattooing', 'encoded',
                'compartmentalize', 'lain', 'the', 'gov',
                'printers', 'hiked', 'resentment', 'revisionism', 'tavern',
                'backpacking', 'pestering', 'gassed', 'acknowledges',
                'testimonies', 'parlance', 'hallucinate', 'speeches',
                'engaging', 'solder', 'perceptive', 'microbiology', 'reconnaissance', 'garlic',
                'neutrals', 'width', 'literaly', 'guild', 'despicable', 'dion',
                'option', 'transistors', 'chiropractic', 'tattered', 'consolidating',
                'olds', 'garmin', 'shift', 'granted', 'intramural', 'allie', 'cylinders',
                'wishlist', 'crank', 'wrongly', 'workshop', 'yesterday', 'wooden', 
                'without', 'wheel', 'weather', 'watch', 'wage', 'version', 'usually',
                'twice', 'tomato', 'ticket', 'text', 'switch', 'studio', 'stick',
                'soup', 'sometimes', 'signal', 'prior', 'plant', 'photo',
                'path', 'park', 'near', 'menu', 'latter', 'grass', 'clock']
```
\normalsize

# References {-}

### 2021-06-22 thesis/_book/_main.tex

@@
 \usepackage[unicode=true]{hyperref}
+\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
 \hypersetup{
@@
             pdfauthor={Alicja Dobrzeniecka},
-            pdfborder={0 0 0},
+            colorlinks=true,
+            linkcolor=Maroon,
+            citecolor=Blue,
+            urlcolor=blue,
             breaklinks=true}
@@
 \author{Alicja Dobrzeniecka}
-\date{2021-06-21}
+\date{2021-06-22}
 
@@
 {
+\hypersetup{linkcolor=black}
 \setcounter{tocdepth}{5}
@@
 represented by means of numbers and this is accomplished with word
-embeddings. The word embeddings are usually assigned the values based on
-the context in which the words appear. This means that the input data
-can have an enormous influence on the outcome. The biases seem to have
-their primary source in the way the words are assigned numerical values.
-One can find methods that aim at identifying and measuring hidden biases
-and/or try to remove them by for example modifying the word embeddings
-directly.
+embeddings. Word embeddings are usually assigned the values based on the
+context in which the words appear. This means that the input data can
+have an enormous influence on the outcome. One can find methods that aim
+at identifying and measuring hidden biases and/or try to remove them by,
+for example, modifying the word embeddings directly.
 
@@
 fairness actually are. With the choice of the definition, implications
-into the real-life applications may change as well. Mehrabi, Morstatter,
+for the real-life applications may change as well. Mehrabi, Morstatter,
 Saxena, Lerman, \& Galstyan
@@
 groups of people. Finally, statistical discrimination consists in
-relying on group statistics to judge person belonging to a given group.
-The topic of discrimination is entangled with another concept; fairness.
-It is essential to grasp some concepts of fairness to take them into
-consideration while designing implementation of some machine learning
-model. In Mehrabi et al.
+relying on group statistics to judge a person belonging to a given
+group. The topic of discrimination is entangled with another concept;
+fairness. It is essential to grasp some concepts of fairness to take
+them into consideration while designing implementation of some machine
+learning model. In Mehrabi et al.
 (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) one may notice that
@@
 Saligrama, \& Kalai (\protect\hyperlink{ref-Bolukbasi2016Man}{2016})
-focus on gender biases that may be observable while investigating the
-representation of job occupations and gender in terms of their assigned
-numerical values. The authors apply cosine similarity measurement to
-investigate the phenomenon where (the vectors corresponding to) words
-related to jobs that are stereotypically associated with a given gender
-are in fact in the model situated closer to this gender. They also use
-analogy tasks to evaluate if bias is present in a word embedding model.
-They check analogies by comparing pairs of word vectors. For example
-they search for the word complementing the puzzle: man is to doctor as
-woman is to \ldots{}? First they subtract the word ``man'' from the word
-``woman''\footnote{This is pointwise subtraction of vectors, but for the
-  simplicity we will often talk about words while really meaning
-  vectors. What is meant should be clear from the context.} and then
-they search for the ranked list of other words pairs that have similar
-vector difference. They also include in the formula a threshold to
-ensure that the resulting pairs could not be randomly picked.
+focus on gender biases that may be observable in the representation of
+job occupations and gender in terms of numerical values. The authors
+apply cosine similarity measurement to investigate the phenomenon where
+(the vectors corresponding to) words related to jobs that are
+stereotypically associated with a given gender are in fact in the model
+situated closer to this gender. They also use analogy tasks to evaluate
+if a bias is present in a word embedding model. They check analogies by
+comparing pairs of word vectors. For example they search for the word
+complementing the puzzle: man is to doctor as woman is to \ldots{}?
+First they subtract the word ``man'' from the word ``woman''\footnote{This
+  is pointwise subtraction of vectors, but for the simplicity we will
+  often talk about words while really meaning vectors. What is meant
+  should be clear from the context.} and then they search for the ranked
+list of other words pairs that have similar vector difference. They also
+include in the formula a threshold to ensure that the resulting pairs
+could not be randomly picked.
 
@@
 not make sense to expect the algorithm to return the same profession for
-both woman and man. Therefore, this method seems to be limited in terms
-of bias detection. Other problems are related for example to the choice
-of pairs and words that are used to detect the presence of
-discrimination, as it is often subjective and without proper
-justification. Additionally, the choice of the parameter used in
-Bolukbasi et al. (\protect\hyperlink{ref-Bolukbasi2016Man}{2016})
-formula to ensure that word pairs are not picked by random, is also not
-justified and changing it drastically influences the results.
+both woman and man. Therefore, this method of bias detection seems
+limited. Other problems are related for example to the choice of pairs
+and words that are used to detect the presence of discrimination, as it
+is often subjective and without proper justification. Additionally, the
+choice of the parameter used in Bolukbasi et al.
+(\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) formula to ensure that
+word pairs are not picked by random, is also not justified and changing
+it drastically influences the results.
 
@@
 resemblance between protected words and harmful attributes exists. They
-investigate an approach that enables them to measure the bias for a
-class (such as gender, religion or race) and express the final result
-with a single metric. Their approach will be analyzed in details in the
-next chapter.
+develop an approach that enables them to measure the bias for a class
+(such as gender, religion or race) and express the final result with a
+single metric. Their approach will be analyzed in details in the next
+chapter.
 
@@
 feelings of pleasantness or unpleasantness. In general, people would
-rather associate flowers with feeling pleasant than insects, and this
+rather associate flowers than insects with feeling pleasant, and this
 preference in a sense is a bias or prejudice in some direction. However,
@@
 that are considered to be stereotypical or harmful in some way. One can
-find in this group methods such as euclidean distance or cosine
+find in this group methods such as Euclidean distance or cosine
 similarity (which is equivalent to dot product if the vectors are
@@
 and validate if the model uses the gender or race information to
-complete the task. The currently used methods employing cosine
-similarity, often aggregate the similarity values often aggregated in a
-way that may lead to hasty conclusions. The averaging of values and the
-lack of attention to uncertainty involved in such aggregation may lead
-to the incomplete picture of the situation.
+complete the task. Unfortunately, the currently used methods employing
+cosine similarity, often aggregate the similarity values in a way that
+may lead to hasty conclusions. The averaging of values and the lack of
+attention to uncertainty involved in such aggregation may lead to an
+incomplete picture of the situation.
 
@@
 representations. One of the most straightforward ones is a skip-gram
-model. Given a word the model tries to predict its neighboring words
+model. Given a word, the model tries to predict its neighboring words
 from the sentence. The mathematics behind the process relies on the idea
@@
 greatest similarity. It is also possible to obtain negative cosine
-similarity, if the value approaches -1 it means that the words are
-contrary to each other.
+similarity. If the value approaches -1, this intuitively means that the
+words are contrary to each other.
 
@@
 \href{https://stats.stackexchange.com/questions/198080/proving-that-cosine-distance-function-defined-by-cosine-similarity-between-two-u}{discussion
-in stats.stackexchange.com} in the case of cosine distance it would have
-to fulfill this equation 1+cos-sim(𝐴,𝐶) \textless{} cos-sim(𝐴,𝐵) +
-cos-sim(𝐵,𝐶). If one chooses specific unit vectors it is easy to
-demonstrate that the triangle inequality is not preserved.
+in stats.stackexchange.com}\footnote{\url{https://stats.stackexchange.com/questions/198080/proving-that-cosine-distance-function-defined-by-cosine-similarity-between-two-u}}
+in the case of cosine distance it would have to fulfill this equation
+1+cos-sim(𝐴,𝐶) \textless{} cos-sim(𝐴,𝐵) + cos-sim(𝐵,𝐶). If one chooses
+specific unit vectors it is easy to demonstrate that the triangle
+inequality is not preserved.
 
@@
 similarity between words as the outcome inner product of their
-normalized vectors. They focus on examining what the geometry of word
+normalized vectors. They focus on examining what the geometry of a word
 embedding is in regard to ``he'' and ``she'' words. In other words,
@@
 R notebook using data from GloVe: Global Vectors for Word
-Representation}. However, similar visualization may be found there. Data
-used to create our plot is as follows.
+Representation}.\footnote{\url{https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings}.}
+However, similar visualization may be found there. Data used to create
+our plot is as follows.
 
@@
 different approach towards finding similarities between classes of
-words. The authors claim that texts available online are full of direct
-or indirect human stereotypes. As a result, word embeddings are prone to
-learn and maybe amplify those biases and propagate them further into AI
-models that are used for various applications. Cosine distance is used
-in the article as a measure to first argue for the existence of
-multi-class bias and then to show how through bias mitigation techniques
-the bias may be decreased.
+words. Cosine distance is used in the article as a measure to first
+argue for the existence of multi-class bias and then to show how through
+bias mitigation techniques the bias may be decreased.
 
@@
 
-In the final step there is an evaluation of the results. The cosine
-distances are calculated again but this time using debiased vocabulary.
-After taking the mean of all distances one final value is obtained and
-then it is compared with the average value from the beginning. If the
-cosine distances are on average greater than before then it leads the
-authors to the conclusion that improvement has been achieved. As the
-cosine distance increases, it is assumed that the association between
-protected and stereotypical words decreases.
+In the final step they evaluate the results. The cosine distances are
+calculated again but this time using the debiased vocabulary. After
+taking the mean of all distances one final value is obtained and then it
+is compared with the average value from the beginning. If the cosine
+distances are on average greater than before then it leads the authors
+to the conclusion that improvement has been achieved. As the cosine
+distance increases, it is assumed that the association between protected
+and stereotypical words decreases.
+
+\newpage
 
@@
 issue by providing the tables and visualizations of individual cosine
-distances to make sure that we get the proper insight into the data.
+distances to make sure that we obtain a proper insight into the data.
 
@@
 
-With such a method the uncertainty is not included which makes it even
-more difficult to give reasonable interpretations of the results. We
-propose the use of Bayesian method to obtain some understanding of the
-influence the uncertainty has on the interpretation of final results.
-
-\textbf{4. No word class distinction}
+With such a method the uncertainty involved is not really considered
+which makes it even more difficult to give reasonable interpretations of
+the results. We propose the use of Bayesian method to obtain some
+understanding of the influence the uncertainty has on the interpretation
+of final results.
+
+\pagebreak
+
+\textbf{4. No word class distinction and no control group}
 
@@
 with another one. Not all of the stereotypical words have to be
-considered as harmful for all of the religions. In our analysis, we
-distinguish between stereotypes associated with a given group,
-stereotypes associated with different groups, and control groups:
-neutral words and stereotypes-free human predicates.
+considered as harmful for all of the religions. One should investigate
+the religions separately as some of them may have stronger harmful
+associations that others. One should also include control groups to have
+a way of comparing the stereotypical results with neutral or human-like
+words. Later in the text we will explain in details reasons for
+introducing control groups. In our analysis, we distinguish between
+stereotypes associated with a given group, stereotypes associated with
+different groups, and control groups: neutral words and stereotypes-free
+human predicates.
 
@@
 small changes are meaningful. This is in general the problem of scale
-and the lack of universal intervals. In contrast, proper intervals
-decide whether given cosine similarity is high enough to consider the
-words to be more similar than if we chose them at random. We will use
-highest posterior density intervals.
+and the lack of universal intervals. In contrast, statistical intervals
+will help us decide whether a given cosine similarity is high enough to
+consider the words to be more similar than if we chose them at random.
+We will use highest posterior density intervals, in line with Bayesian
+methodology.
 
@@
 We will use the choice of protected words and stereotypical predicates
-used in REF. This is a decent point of departure, not only we want to
-compare our method to that of REF, but also because this data format is
-fairly general (as contrasted, say, with a set up for binary
-stereotypes). Note also that the method we develop here can fairly
-easily be run for different stereotypization patterns. Let's start with
-explaining the method and its deployment using a dataset obtained for
-the religion-related protected words.
+used in Manzini et al.
+(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}). This is a
+decent point of departure, as not only we want to compare our method to
+that of Manzini et al.
+(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}), but also
+because this data format is fairly general (as contrasted, say, with a
+set up for binary stereotypes). Note also that the method we develop
+here can fairly easily be run for different stereotypization patterns.
+Let's start with explaining the method and its deployment using a
+dataset obtained for the religion-related protected words.
 
@@
 groups. For instance, for religion, the bias-related predicates (coming
-from the original study in REF) include muslim bias attributes, jew bias
-attributes, christian bias attributes (see a list in the APPENDIX).
+from the original study in Manzini et al.
+(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019})) include
+muslim bias attributes, jew bias attributes, christian bias attributes
+(see a list in the Appendix).
 
@@
 
-\vspace{1mm} \footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{religion <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
-\KeywordTok{colnames}\NormalTok{(religion) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
-                        \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
-\CommentTok{#head(religion)}
-\CommentTok{#library(plyr)}
-\NormalTok{religion}\OperatorTok{$}\NormalTok{wordClass <-}\StringTok{ }\KeywordTok{as.factor}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass)}
-\CommentTok{#levels(religion$wordClass)}
-\CommentTok{#religion$wordClass <- mapvalues(religion$wordClass, }
-\CommentTok{#                          from = levels(religion$wordClass),}
-\CommentTok{#                          to = c("christian","human","jewish","muslim","neutral"))}
-\KeywordTok{levels}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
-\KeywordTok{head}\NormalTok{(religion)  }\OperatorTok{%>%}\StringTok{  }\KeywordTok{kable}\NormalTok{(}\DataTypeTok{format =} \StringTok{"latex"}\NormalTok{,}\DataTypeTok{booktabs=}\NormalTok{T,}
-                      \DataTypeTok{linesep =} \StringTok{""}\NormalTok{,  }\DataTypeTok{escape =} \OtherTok{FALSE}\NormalTok{, }
-                      \DataTypeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\OperatorTok{%>%}\StringTok{      }\KeywordTok{kable_styling}\NormalTok{(}\DataTypeTok{latex_options=}\KeywordTok{c}\NormalTok{(}\StringTok{"scale_down"}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
+\pagebreak  \vspace{1mm} \footnotesize
 
@@
 visible for attributes associated with different stereotypes. Moreover,
-the empirical distributions overlap to alarge extent and the means
+the empirical distributions overlap to a large extent and the means
 grouped by connection type do not seem too far from each other. In fact,
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{(tidyverse)}
-\NormalTok{muslimWords <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"imam"}\NormalTok{,}\StringTok{"islam"}\NormalTok{,}\StringTok{"mosque"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"quran"}\NormalTok{)}
-\NormalTok{muslim <-}\StringTok{ }\NormalTok{religion }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(protectedWord }\OperatorTok{%in%}\StringTok{ }\NormalTok{muslimWords)}
-\KeywordTok{ggplot}\NormalTok{(muslim, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{  cosineDistance, }\DataTypeTok{fill =}\NormalTok{ connection, }\DataTypeTok{color =}\NormalTok{ connection))}\OperatorTok{+}
-\StringTok{  }\KeywordTok{geom_density}\NormalTok{(}\DataTypeTok{alpha=}\FloatTok{0.6}\NormalTok{,}\DataTypeTok{size =} \FloatTok{.2}\NormalTok{)}\OperatorTok{+}\StringTok{ }
-\StringTok{  }\KeywordTok{scale_fill_manual}\NormalTok{(}\DataTypeTok{values =} \KeywordTok{c}\NormalTok{(}\StringTok{"orangered4"}\NormalTok{,}\StringTok{"chartreuse4"}\NormalTok{, }\StringTok{"skyblue"}\NormalTok{, }\StringTok{"gray"}\NormalTok{))}\OperatorTok{+}
-\StringTok{  }\KeywordTok{scale_x_continuous}\NormalTok{(}\DataTypeTok{breaks =} \KeywordTok{seq}\NormalTok{(}\FloatTok{0.3}\NormalTok{,}\FloatTok{1.5}\NormalTok{, }\DataTypeTok{by =} \FloatTok{0.1}\NormalTok{))}\OperatorTok{+}\KeywordTok{xlab}\NormalTok{(}\StringTok{"cosine distance"}\NormalTok{)}\OperatorTok{+}
-\StringTok{  }\KeywordTok{scale_color_manual}\NormalTok{(}\DataTypeTok{values =} \KeywordTok{c}\NormalTok{(}\StringTok{"orangered4"}\NormalTok{,}\StringTok{"chartreuse4"}\NormalTok{,}\StringTok{"skyblue"}\NormalTok{,}\StringTok{"gray"}\NormalTok{))}\OperatorTok{+}
-\StringTok{  }\KeywordTok{theme_tufte}\NormalTok{()}\OperatorTok{+}\KeywordTok{ggtitle}\NormalTok{(}\StringTok{"Empirical distribution of distances (muslim)"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
 \begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-7-1} \end{center}
@@
 \vspace{1mm} \footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{(tidyverse)}
-\NormalTok{muslimClass <-}\StringTok{ }\NormalTok{muslim }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(protectedWord }\OperatorTok{==}\StringTok{ "muslim"}\NormalTok{)}
-\NormalTok{neutralSample <-}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\KeywordTok{filter}\NormalTok{(muslimClass,connection }\OperatorTok{==}\StringTok{ "none"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
-\NormalTok{humanSample <-}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\KeywordTok{filter}\NormalTok{(muslimClass,connection }\OperatorTok{==}\StringTok{ "human"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
-\NormalTok{muslimVis <-}\StringTok{ }\NormalTok{muslimClass }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(connection }\OperatorTok{!=}\StringTok{ "none"} \OperatorTok{&}\StringTok{ }\NormalTok{connection }\OperatorTok{!=}\StringTok{"human"}\NormalTok{)}
-\NormalTok{muslimVis <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(muslimVis,neutralSample,humanSample)}
-
-\CommentTok{#we plug in our visualisation script}
-\KeywordTok{source}\NormalTok{(}\StringTok{"../functions/visualisationTools.R"}\NormalTok{)}
-\CommentTok{#two arguments: dataset and protected word}
-\KeywordTok{visualiseProtected}\NormalTok{(muslimVis,}\StringTok{"muslim"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
 
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{(tidyverse)}
-\NormalTok{priestClass <-}\StringTok{ }\NormalTok{religion }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(protectedWord }\OperatorTok{==}\StringTok{ "priest"}\NormalTok{)}
-\NormalTok{neutralSample <-}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\KeywordTok{filter}\NormalTok{(priestClass,connection }\OperatorTok{==}\StringTok{ "none"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
-\NormalTok{humanSample <-}\StringTok{ }\KeywordTok{sample_n}\NormalTok{(}\KeywordTok{filter}\NormalTok{(priestClass,connection }\OperatorTok{==}\StringTok{ "human"}\NormalTok{), }\DecValTok{5}\NormalTok{)}
-\NormalTok{priestVis <-}\StringTok{ }\NormalTok{priestClass }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(connection }\OperatorTok{!=}\StringTok{ "none"} \OperatorTok{&}\StringTok{ }\NormalTok{connection }\OperatorTok{!=}\StringTok{"human"}\NormalTok{)}
-\NormalTok{priestVis <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(priestVis,neutralSample,humanSample)}
-
-\CommentTok{#we plug in our visualisation script}
-\KeywordTok{source}\NormalTok{(}\StringTok{"../functions/visualisationTools.R"}\NormalTok{)}
-\CommentTok{#two arguments: dataset and protected word}
-\KeywordTok{visualiseProtected}\NormalTok{(priestVis,}\StringTok{"priest"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
 \begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-8-1} \end{center}
@@
 words might in general have different relations to all the attributes
-and these relations should be our point of departure.
+and these relations should be our point of departure.\footnote{The
+  construction of the Bayesian models and code for visualisations is due
+  to Rafal Urbaniak.}
 
@@
 
+\normalsize
 The only reason we need it is the evaluation of connection as a
@@
 
-\section{Dataset-level coefficients}\label{dataset-level-coefficients}
+\section{\texorpdfstring{Dataset-level coefficients
+\label{sec:datasetsLevelCoeffs}}{Dataset-level coefficients }}\label{dataset-level-coefficients}
 
@@
 Let's turn then to data analysis that takes a look at protected words
-separately. This time for each combination of a protected word and a
+separately. This time, for each combination of a protected word and a
 connection status we will have a separate mean cosine distance estimate,
@@
 their calculations as written in
-\href{https://github.com/tolga-b/debiaswe}{Tolga Code}. We decided to
+\href{https://github.com/tolga-b/debiaswe}{Tolga Bolukbasi
+code}\footnote{\url{https://github.com/tolga-b/debiaswe}}. We decided to
 compare the results obtained with the use of Reddit L2 model and the
@@
 Reddit word embeggings have 50 dimensions and GoogleNews word embeddings
-have 300 dimensions. As the dimension is increasing, the vector can
-capture much more information but this is not always the case. According
-to some researchers the amount of dimensions should be chosen by taking
+have 300 dimensions. As the dimension increases, the vectors can capture
+much more information, but this is not always the case. According to
+some researchers, the amount of dimensions should be chosen by taking
 into account the type of corpus and its features. In our case both
@@
 \includegraphics[width=14cm]{../images/visRaceGoogle.png}
+
+We leave a discussion of these results for later, for now let us also
+inspect the impact of debiasing.
 
@@
 First, let's look at coefficient estimated for the whole datasets, as
-compared to their estimation prior to debiasing:
+compared to their estimation prior to debiasing. One may observe very
+slight changes in the estimated coeﬀicients. It seems that in religion
+dataset the changes are the most significant as the mean moves towards
+zero which stands for no similarity between words.
 
@@
 \end{center}
-
-One may observe very slight changes in the estimated coeﬀicients. It
-seems that in religion dataset the changes are the most significant as
-the mean moves towards zero which stands for no similarity between
-words.
 
@@
 It is worth paying attention to the gender debiased dataset. The change
-is really minor, still zero out of HPDI range. The \texttt{different}
-attributes still have high similarity value. It seems that the distances
-for gender protected words are still clustered together which suggests
-that this method is not able to catch the complex bias presence. One
-could conclude that co-occurrence of certain terms does not always mean
-direct connection in terms of terms associations. The fact that some of
-stereotypically associated with female attributes still have high
-similarity with male protected words suggest that other techniques
-should be applied to detect and remove the bias.
+is really minor, still with zero out of the HPDI range. The
+\texttt{different} attributes still have high similarity value. It seems
+that the distances for gender protected words are still clustered
+together which suggests that this method is unable to catch the complex
+bias presence. One could conclude that co-occurrence of certain terms
+does not always mean direct connection in terms of terms associations.
+The fact that some of stereotypically associated with female attributes
+still have high similarity with male protected words suggest that other
+techniques should be applied to detect and remove the bias.
 
@@
 
-In Manzini et al.
-(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) it is
-mentioned that debiasing with this method that is based on bias
-component removal may be insufficient. The reason is that words
-containing similar biases may still cluster together even after
-debiasing.
-
 \chapter{Discussion and summary}\label{discussion-and-summary}
@@
 intervals and \texttt{different} or \texttt{associated} quite the
-opposite. One of the reasons for such outcome is that we used
+opposite. One of the reasons for such an outcome is that we used
 approximately 230 neutral words and only between 11-25 (the number
@@
 distance values and uncertainty change after the debiasing. One can
-verify then how the individual vectors changed and if it is what was
-expected. Our analysis with the use of Bayesian method gave us new ideas
-and hypothesis concerning not only the bias detection method but also
-the efficiency of debiasing itself.
-
-We created a summary table for each of the datasets: Reddit, Reddit
-Debiased, and Google word embeddings. Let us first analyse the general
-observations from estimated coefficients mean introduced in 3.6.
-DATASET-LEVEL COEFFICIENTS. For Google embeddings the HPDI for all
-classes coefficients (associated, different, human, and none) has an
-interval that includes zero. This can lead one to a conclusion that the
-impact for associated, different, human and neutral attribute is, when
-averaged, quite similar. This indicates how including the uncertainty
-may change the use and interpretation of Manzini et al.
+investigate then how the individual vectors changed and if it is what
+was expected. Our analysis with the use of Bayesian method gave us new
+ideas and hypothesis concerning not only the bias detection method but
+also the efficiency of debiasing itself. We created summary tables for
+each of the datasets: Google word embeddings (Table \ref{tab:dTabG}),
+Reddit (Table \ref{tab:dTabR}), and Reddit Debiased (Table
+\ref{tab:dTabRD}).
+
+\vspace{1mm} \footnotesize
+
+\begin{table}
+
+\caption{\label{tab:dTabG}Overview statistics  (GoogleNews embedding)}
+\centering
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{llll}
+\toprule
+  & Religion (percent) & Gender (percent) & Race (percent)\\
+\midrule
+A close to N & 66.67 & 14.29 & 80\\
+A close to H & 66.67 & 21.43 & 100\\
+H closer than A & 20 & 21.43 & 50\\
+D similar with A & 93.33 & 42.86 & 100\\
+D closer than A & 33.33 & 21.43 & 30\\
+General HPDI outside of  0 & no & no & no\\
+General D similar to A & slightly not & slightly not & slightly not\\
+\bottomrule
+\end{tabular}}
+\end{table}
+
+\normalsize
+
+\vspace{1mm} \footnotesize
+
+\begin{table}
+
+\caption{\label{tab:dTabR}Overview statistics  (Reddit embedding)}
+\centering
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{llll}
+\toprule
+  & Religion (percent) & Gender (percent) & Race (percent)\\
+\midrule
+A close to N & 66.67 & 0 & 40\\
+A close to H & 60 & 21.43 & 50\\
+H closer than A & 33.33 & 0 & 0\\
+D similar with A & 100 & 92.86 & 90\\
+D closer than A & 33.33 & 42.86 & 30\\
+General HPDI outside of 0 & no & yes & no\\
+General D similar to A & slightly not & yes & slightly not\\
+\bottomrule
+\end{tabular}}
+\end{table}
+
+\normalsize
+
+\vspace{1mm} \footnotesize
+
+\begin{table}
+
+\caption{\label{tab:dTabRD}Overview statistics  (Reddit embedding, debiased)}
+\centering
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{llll}
+\toprule
+  & Religion (percent) & Gender (percent) & Race (percent)\\
+\midrule
+A close to N & 73.33 & 7.14 & 50\\
+A close to H & 46.67 & 21.43 & 70\\
+H closer than A & 66.67 & 0 & 20\\
+D similar with A & 100 & 100 & 80\\
+D closer than A & 66.67 & 50 & 30\\
+General HPDI outside of 0 & no & yes & no\\
+General D similar to A & yes & yes & yes\\
+General H moved closer to A & yes & no & no\\
+General N moved closer & no & yes & yes\\
+\bottomrule
+\end{tabular}}
+\end{table}
+
+\normalsize
+
+Let us first analyse the general observations from estimated
+coefficients mean introduced in Section \ref{sec:datasetsLevelCoeffs} on
+dataset-level coefficients. For Google embeddings the HPDIs for all
+class coefficients (associated, different, human, and none) include
+zero. This can lead one to a conclusion that the impact for associated,
+different, human and neutral attribute is, when averaged, quite similar.
+This indicates how including the uncertainty may change the use and
+interpretation of Manzini et al.
 (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) MAC metric.
 It seems that if one focuses only on differences between means of means,
-it is too simplistic. In case of Reddit word embeddings the situation is
-similar although HPDI interval is below 0 for Gender class when looking
-at \texttt{associated} and \texttt{different} mean coefficient. This can
+this is too simplistic. In case of Reddit word embeddings the situation
+is similar although HPDI is below 0 for Gender class when looking at
+\texttt{associated} and \texttt{different} mean coefficient. This can
 suggest that there is indeed slightly stronger impact of these
@@
 highest observed absolute difference is equal to only 0.1. This suggests
-that again the impact of associated attributes and difference ones is
-not clear at first sight when looking at averaged coefficients. Finally,
-let's compare the HPDI intervals for Reddit and Reddit debiased
-datasets. For Religion and Race dataset there is a minor shift (in
-absolute values the highest change is equal to approximately 0.1) of the
-mean coefficients towards zero. However for Gender dataset there is no
-significant change. This is of course the general look at the data,
-let's now analyze the individual words.
-
-In Reddit table one can observe that for Religion class the cosine
-distance results for the associated attributes are for approximately
-60\% of the words close to human attributes as well. This can suggest
-that in some cases words concerning humans can have higher similarity
-with some protected words independently if they are stereotypical or
+that again the impact of associated attributes and different ones is not
+clear at first sight. Finally, let's compare the HPDIs for Reddit and
+Reddit debiased datasets. For Religion and Race dataset there is a minor
+shift (in absolute values the highest change is equal to approximately
+0.1) of the mean coefficients towards zero. However for Gender dataset
+there is no significant change. This is of course a general look at the
+group coefficients. Let's now analyze the individual words.
+
+In Reddit table one can observe that for Religion the cosine distance
+results for the associated attributes are for approximately 60\% of the
+words close to human attributes as well. This can suggest that in some
+cases words concerning humans can have higher similarity with some
+protected words independently of whether they are stereotypical or
 neutral-human words. One should also pay attention to the fact that for
 all of the \texttt{associated} and \texttt{different} attributes in
-Religion class the uncertainty interval overlaps at some point. What is
-even more surprising is that for protected words, such as ``torah'' the
+Religion, the uncertainty intervals overlap at some point. What is even
+more surprising is that for protected words, such as ``torah'' the
 \texttt{associated} attribute has the the cosine distance slightly over
@@
 the choice of the protected words and attributes may be corrupted. The
-second one is that the metric is not able to catch the hidden bias
+second one is that the metric is unable to catch the hidden bias
 properly. The third one is that there is actually no bias between the
@@
 
-Surprisingly in Gender class one can observe high cosine similarity
-values between some female stereotypical professions and male protected
-words. If a word stereotypically associated with females has low cosine
+Surprisingly in Gender one can observe high cosine similarity values
+between some female stereotypical professions and male protected words.
+If a word stereotypically associated with females has low cosine
 distance to male protected words, then one should investigate the issue
@@
 
-In Google table one may notice interesting observations as well.
-Although GoogleNews dataset is larger, trained on different data sources
-and with the use of more dimensions, some of the results are similar to
-the ones obtained for the Reddit corpus. For Religion case almost all of
-the values for \texttt{associated} class have an intersection with
-\texttt{different} class as well. In the case of Race it is 100\% of the
-available words. This indicates again that it is not clear how the
-metric should be used. Quite different situation takes place for Gender
-data where the similarity between \texttt{associated} and
-\texttt{different} is present mostly for the male protected words. This
-means that male words have high similarity with both male stereotypical
-professions and female ones. However, for females the similarity is high
-mostly only for the female stereotypical professions. This observation
-could not be make when using MAC metric only as it requires
-investigating the individual words.
-
-When analyzing the Reddit debiased results one should ask a question of
-the change of the cosine distance values. It seems that not all of the
+In Google, one may notice interesting phenomena as well. Although the
+GoogleNews dataset is larger, trained on different data sources and with
+the use of more dimensions, some of the results are similar to the ones
+obtained for the Reddit corpus. For Religion almost all of the values
+for \texttt{associated} class intersect with \texttt{different} class as
+well. In the case of Race it is 100\% of the available words. This
+indicates again that it is not clear how the metric should be used.
+Quite a different situation takes place for Gender, where the similarity
+between \texttt{associated} and \texttt{different} is present mostly for
+the male protected words. This means that male words have high
+similarity with both male stereotypical professions and female ones.
+However, for females, the similarity is high mostly only for the female
+stereotypical professions. This observation could not be made when using
+MAC metric only, as it requires investigating the individual words, and
+providing uncertainty,
+
+When analyzing the Reddit debiased results one should investigate the
+change of the cosine distance values. It seems that not all of the
 protected words are treated equally when performing debiasing. In
-religion data the values for \texttt{associated} class moved towards 0
-for most of the words except the ones for Islam religion where still the
+religion, the values for \texttt{associated} class moved towards 0 for
+most of the words except for the ones for Islam, where still the
 \texttt{associated} class has quite lower cosine distance than the
-\texttt{different} one. Similar situation takes place in Gender data
-where female words have still much lower cosine distance values for
+\texttt{different} one. Similar situation takes place in Gender, where
+female words have still much lower cosine distance values for
 \texttt{associated} class than for the \texttt{different} one. In the
-case of males it is mostly almost the same interval for
+case of male protected words it is mostly almost the same interval for
 \texttt{associated} and \texttt{different}. As the Gender data is quite
@@
 male protected words. However, as we pointed out the situation after
-debiasing can sometimes be better only for one protected group which is
-not the desired outcome. In the case of Religion data even after
-debiasing, all of the protected words have intersections between
+debiasing can sometimes be better only for one protected group, which is
+not the desired outcome. In the case of Religion, even after debiasing,
+all of the protected words have intersections between
 \texttt{associated} and \texttt{different} class and similarity greater
@@
 protected words and \texttt{associated}, and \texttt{different} class
-the same but rather to move it towards 0.
+the same but rather to move it towards 1.
 
@@
 data. One should be concerned with the limitations of the metric that
-evaluates the debiasing by the mean of the averaged mean approach.
+evaluates the debiasing by means of the averaged mean approach.
 Additionally, one cannot be sure if the bias is still preserved after
 debiasing. The fact that all of the cosine distances for protected words
-and harmful attributes moved slightly to the right, does not mean that
-the bias is removed. It is shown in articles such as Gonen \& Goldberg
-(\protect\hyperlink{ref-Gonen2019Lipstick}{2019}), that the bias can
-hide in the vector geometry and preserve even after applying popular
+and harmful attributes moved slightly to the right does not mean that
+the bias is removed. Gonen \& Goldberg
+(\protect\hyperlink{ref-Gonen2019Lipstick}{2019}) argue that the bias
+can hide in the vector geometry and preserve even after applying popular
 debiasing methods. One of the reasons for that may be the fact that
@@
 the metric effectiveness is not well justified. The sample size is very
-small which leads to large HPDI. Some of the protected words and
-attributes originate from old articles and it seems that the bias
-phenomenon is quite dynamic and the data should be as up-to-date as it
-is possible. One could for example research the IAT examples to verify
-the effectiveness of MAC metric. One should remember that there is no
-clear interpretation for the values obtained with MAC metric from
-Manzini et al.
+small, which leads to large HPDIs. Some of the protected words and
+attributes originate from somewhat old articles and it seems that the
+bias phenomenon is quite dynamic and the data should be as up-to-date as
+it is possible. Moreover, recall that there is no clear interpretation
+for the values obtained with MAC metric from Manzini et al.
 (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}). One may
@@
 mean to be close to 0? If the averaged cosine distance is equal to 0.8,
-then should we still debias it? It is unclear what the criteria are. On
-one hand, it seems to be beneficial when the outcome is simplified as it
-is easier to compare results with one value per set. On the other hand,
-it is prone to misunderstanding of how to interpret the results and what
+then should we still debias? It is unclear what the criteria are. On one
+hand, it seems to be beneficial when the outcome is simplified as it is
+easier to compare results with one value per set. On the other hand, it
+is prone to misunderstanding of how to interpret the results and what
 threshold to assume.
@@
 
-Bibliography:
+\chapter*{Appendix}\label{appendix}
+\addcontentsline{toc}{chapter}{Appendix}
+
+\section*{\texorpdfstring{Original wordlist from (Manzini et al.,
+\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019})}{Original wordlist from (Manzini et al., 2019)}}\label{original-wordlist-from-manzini2019blacktocriminal}
+\addcontentsline{toc}{section}{Original wordlist from (Manzini et al.,
+\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019})}
+
+Definite sets are the protected words, analogy templates assign
+stereotypical attributes to protected words, test terms were included
+but are not used in the original MAC calculations.
+
+\subsection*{Religion}\label{religion}
+\addcontentsline{toc}{subsection}{Religion}
+
+\vspace{1mm} \scriptsize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{\{}
+    \StringTok{"definite_sets"}\OperatorTok{:}\NormalTok{[}
+\NormalTok{        [}\StringTok{"judaism"}\NormalTok{, }\StringTok{"christianity"}\NormalTok{, }\StringTok{"islam"}\NormalTok{],}
+\NormalTok{        [}\StringTok{"jew"}\NormalTok{, }\StringTok{"christian"}\NormalTok{, }\StringTok{"muslim"}\NormalTok{],}
+\NormalTok{        [}\StringTok{"synagogue"}\NormalTok{, }\StringTok{"church"}\NormalTok{, }\StringTok{"mosque"}\NormalTok{],}
+\NormalTok{        [}\StringTok{"torah"}\NormalTok{, }\StringTok{"bible"}\NormalTok{, }\StringTok{"quran"}\NormalTok{],}
+\NormalTok{        [}\StringTok{"rabbi"}\NormalTok{, }\StringTok{"priest"}\NormalTok{, }\StringTok{"imam"}\NormalTok{]}
+\NormalTok{    ],}
+    \StringTok{"analogy_templates"}\OperatorTok{:}
+\StringTok{    }\NormalTok{\{}
+        \StringTok{"attribute"}\OperatorTok{:}\StringTok{ }\NormalTok{\{}
+            \StringTok{"jew"}       \OperatorTok{:}\StringTok{ }\NormalTok{[}\StringTok{"greedy"}\NormalTok{, }\StringTok{"cheap"}\NormalTok{, }\StringTok{"hairy"}\NormalTok{, }\StringTok{"liberal"}\NormalTok{],}
+            \StringTok{"christian"} \OperatorTok{:}\StringTok{ }\NormalTok{[}\StringTok{"judgemental"}\NormalTok{, }\StringTok{"conservative"}\NormalTok{, }\StringTok{"familial"}\NormalTok{],}
+            \StringTok{"muslim"}    \OperatorTok{:}\StringTok{ }\NormalTok{[}\StringTok{"violent"}\NormalTok{, }\StringTok{"terrorist"}\NormalTok{, }\StringTok{"dirty"}\NormalTok{, }\StringTok{"uneducated"}\NormalTok{]}
+\NormalTok{        \}}
+\NormalTok{    \},}
+    \StringTok{"eval_targets"}\OperatorTok{:}\NormalTok{[}
+\NormalTok{        [}\StringTok{"judaism"}\NormalTok{, }\StringTok{"christianity"}\NormalTok{, }\StringTok{"islam"}\NormalTok{],}
+\NormalTok{        [}\StringTok{"jew"}\NormalTok{, }\StringTok{"christian"}\NormalTok{, }\StringTok{"muslim"}\NormalTok{],}
+\NormalTok{        [}\StringTok{"synagogue"}\NormalTok{, }\StringTok{"church"}\NormalTok{, }\StringTok{"mosque"}\NormalTok{],}
+\NormalTok{        [}\StringTok{"torah"}\NormalTok{, }\StringTok{"bible"}\NormalTok{, }\StringTok{"quran"}\NormalTok{],}
+\NormalTok{        [}\StringTok{"rabbi"}\NormalTok{, }\StringTok{"priest"}\NormalTok{, }\StringTok{"imam"}\NormalTok{]}
+\NormalTok{    ], }
+    \StringTok{"testTerms"}\OperatorTok{:}\NormalTok{[}\StringTok{"chair"}\NormalTok{, }\StringTok{"house"}\NormalTok{, }\StringTok{"greedy"}\NormalTok{, }\StringTok{"terrorist"}\NormalTok{, }\StringTok{"dirty"}\NormalTok{, }\StringTok{"greedy"}\NormalTok{]}
+\NormalTok{\}}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+\subsection*{Gender}\label{gender}
+\addcontentsline{toc}{subsection}{Gender}
+
+\vspace{1mm} \scriptsize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{\{}
+    \StringTok{"definite_sets"}\OperatorTok{:}\NormalTok{[}
+\NormalTok{        [}\StringTok{"he"}\NormalTok{, }\StringTok{"she"}\NormalTok{],}
+\NormalTok{        [}\StringTok{"his"}\NormalTok{, }\StringTok{"hers"}\NormalTok{],}
+\NormalTok{        [}\StringTok{"son"}\NormalTok{, }\StringTok{"daughter"}\NormalTok{],}
+\NormalTok{        [}\StringTok{"father"}\NormalTok{, }\StringTok{"mother"}\NormalTok{],}
+\NormalTok{        [}\StringTok{"male"}\NormalTok{, }\StringTok{"female"}\NormalTok{],}
+\NormalTok{        [}\StringTok{"boy"}\NormalTok{, }\StringTok{"girl"}\NormalTok{],}
+\NormalTok{        [}\StringTok{"uncle"}\NormalTok{, }\StringTok{"aunt"}\NormalTok{]}
+\NormalTok{    ],}
+    \StringTok{"analogy_templates"}\OperatorTok{:}
+\StringTok{    }\NormalTok{\{}
+        \StringTok{"role"}\OperatorTok{:}\StringTok{ }\NormalTok{\{}
+            \StringTok{"man"}   \OperatorTok{:}\StringTok{ }\NormalTok{[}\StringTok{"manager"}\NormalTok{, }\StringTok{"executive"}\NormalTok{, }\StringTok{"doctor"}\NormalTok{, }\StringTok{"lawyer"}\NormalTok{, }\StringTok{"programmer"}\NormalTok{, }
+                       \StringTok{"scientist"}\NormalTok{, }\StringTok{"soldier"}\NormalTok{, }\StringTok{"supervisor"}\NormalTok{, }\StringTok{"rancher"}\NormalTok{, }\StringTok{"janitor"}\NormalTok{,}
+                       \StringTok{"firefighter"}\NormalTok{, }\StringTok{"officer"}\NormalTok{],}
+            \StringTok{"woman"} \OperatorTok{:}\StringTok{ }\NormalTok{[}\StringTok{"secretary"}\NormalTok{, }\StringTok{"nurse"}\NormalTok{, }\StringTok{"clerk"}\NormalTok{, }\StringTok{"artist"}\NormalTok{, }\StringTok{"homemaker"}\NormalTok{, }\StringTok{"dancer"}\NormalTok{,}
+                       \StringTok{"singer"}\NormalTok{, }\StringTok{"librarian"}\NormalTok{, }\StringTok{"maid"}\NormalTok{, }\StringTok{"hairdresser"}\NormalTok{, }\StringTok{"stylist"}\NormalTok{,}
+                       \StringTok{"receptionist"}\NormalTok{, }\StringTok{"counselor"}\NormalTok{]}
+\NormalTok{        \}}
+\NormalTok{    \}, }
+    \StringTok{"eval_targets"}\OperatorTok{:}\NormalTok{[}
+\NormalTok{        [}\StringTok{"he"}\NormalTok{, }\StringTok{"she"}\NormalTok{],}
+\NormalTok{        [}\StringTok{"his"}\NormalTok{, }\StringTok{"hers"}\NormalTok{],}
+\NormalTok{        [}\StringTok{"son"}\NormalTok{, }\StringTok{"daughter"}\NormalTok{],}
+\NormalTok{        [}\StringTok{"father"}\NormalTok{, }\StringTok{"mother"}\NormalTok{],}
+\NormalTok{        [}\StringTok{"male"}\NormalTok{, }\StringTok{"female"}\NormalTok{],}
+\NormalTok{        [}\StringTok{"boy"}\NormalTok{, }\StringTok{"girl"}\NormalTok{],}
+\NormalTok{        [}\StringTok{"uncle"}\NormalTok{, }\StringTok{"aunt"}\NormalTok{]}
+\NormalTok{    ], }
+    \StringTok{"testTerms"}\OperatorTok{:}\NormalTok{[}\StringTok{"chair"}\NormalTok{, }\StringTok{"house"}\NormalTok{, }\StringTok{"supervisor"}\NormalTok{, }\StringTok{"secretary"}\NormalTok{, }\StringTok{"loud"}\NormalTok{, }\StringTok{"weak"}\NormalTok{]}
+\NormalTok{\}}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+\subsection*{Race}\label{race}
+\addcontentsline{toc}{subsection}{Race}
+
+\vspace{1mm} \scriptsize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{\{}
+    \StringTok{"definite_sets"}\OperatorTok{:}\NormalTok{[}
+\NormalTok{        [}\StringTok{"black"}\NormalTok{, }\StringTok{"caucasian"}\NormalTok{, }\StringTok{"asian"}\NormalTok{],}
+\NormalTok{        [}\StringTok{"african"}\NormalTok{, }\StringTok{"caucasian"}\NormalTok{, }\StringTok{"asian"}\NormalTok{],}
+\NormalTok{        [}\StringTok{"black"}\NormalTok{, }\StringTok{"white"}\NormalTok{, }\StringTok{"asian"}\NormalTok{],}
+\NormalTok{        [}\StringTok{"africa"}\NormalTok{, }\StringTok{"america"}\NormalTok{, }\StringTok{"asia"}\NormalTok{],}
+\NormalTok{        [}\StringTok{"africa"}\NormalTok{, }\StringTok{"america"}\NormalTok{, }\StringTok{"china"}\NormalTok{],}
+\NormalTok{        [}\StringTok{"africa"}\NormalTok{, }\StringTok{"europe"}\NormalTok{, }\StringTok{"asia"}\NormalTok{]}
+\NormalTok{    ],}
+    \StringTok{"analogy_templates"}\OperatorTok{:}
+\StringTok{    }\NormalTok{\{}
+        \StringTok{"role"}\OperatorTok{:}\StringTok{ }\NormalTok{\{}
+            \StringTok{"caucasian"}\OperatorTok{:}\StringTok{ }\NormalTok{[}\StringTok{"manager"}\NormalTok{, }\StringTok{"executive"}\NormalTok{, }\StringTok{"redneck"}\NormalTok{, }\StringTok{"hillbilly"}\NormalTok{, }\StringTok{"leader"}\NormalTok{, }\StringTok{"farmer"}\NormalTok{],}
+            \StringTok{"asian"}    \OperatorTok{:}\StringTok{ }\NormalTok{[}\StringTok{"doctor"}\NormalTok{, }\StringTok{"engineer"}\NormalTok{, }\StringTok{"laborer"}\NormalTok{, }\StringTok{"teacher"}\NormalTok{],}
+            \StringTok{"black"}    \OperatorTok{:}\StringTok{ }\NormalTok{[}\StringTok{"slave"}\NormalTok{, }\StringTok{"musician"}\NormalTok{, }\StringTok{"runner"}\NormalTok{, }\StringTok{"criminal"}\NormalTok{, }\StringTok{"homeless"}\NormalTok{]}
+\NormalTok{        \}}
+\NormalTok{    \}, }
+    \StringTok{"eval_targets"}\OperatorTok{:}\NormalTok{[}
+\NormalTok{        [}\StringTok{"black"}\NormalTok{, }\StringTok{"caucasian"}\NormalTok{, }\StringTok{"asian"}\NormalTok{],}
+\NormalTok{        [}\StringTok{"africa"}\NormalTok{, }\StringTok{"america"}\NormalTok{, }\StringTok{"asia"}\NormalTok{]}
+\NormalTok{    ], }
+    \StringTok{"testTerms"}\OperatorTok{:}\NormalTok{[}\StringTok{"chair"}\NormalTok{, }\StringTok{"house"}\NormalTok{, }\StringTok{"smart"}\NormalTok{, }\StringTok{"criminal"}\NormalTok{, }\StringTok{"executive"}\NormalTok{, }\StringTok{"farmer"}\NormalTok{]}
+\NormalTok{\}}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+\section*{Our control groups}\label{our-control-groups}
+\addcontentsline{toc}{section}{Our control groups}
+
+\section*{Human neutral attributes}\label{human-neutral-attributes}
+\addcontentsline{toc}{section}{Human neutral attributes}
+
+\vspace{1mm} \scriptsize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{human_words =}\StringTok{ }\NormalTok{[}\StringTok{'wear'}\NormalTok{, }\StringTok{'walk'}\NormalTok{, }\StringTok{'visitor'}\NormalTok{, }\StringTok{'toy'}\NormalTok{, }\StringTok{'tissue'}\NormalTok{, }\StringTok{'throw'}\NormalTok{, }\StringTok{'talk'}\NormalTok{, }
+               \StringTok{'speak'}\NormalTok{, }\StringTok{'sleep'}\NormalTok{, }\StringTok{'eye'}\NormalTok{, }\StringTok{'enjoy'}\NormalTok{, }\StringTok{'blogger'}\NormalTok{, }\StringTok{'character'}\NormalTok{, }
+               \StringTok{'candidate'}\NormalTok{, }\StringTok{'breakfast'}\NormalTok{, }\StringTok{'supper'}\NormalTok{, }\StringTok{'dinner'}\NormalTok{, }\StringTok{'eat'}\NormalTok{, }\StringTok{'drink'}\NormalTok{,}
+               \StringTok{"carry"}\NormalTok{, }\StringTok{"run"}\NormalTok{, }\StringTok{"cast"}\NormalTok{, }\StringTok{"ask"}\NormalTok{, }\StringTok{"awake"}\NormalTok{, }\StringTok{"ear"}\NormalTok{, }\StringTok{"nose"}\NormalTok{, }\StringTok{"lunch"}\NormalTok{]}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+\section*{Non-human neutral
+attributes}\label{non-human-neutral-attributes}
+\addcontentsline{toc}{section}{Non-human neutral attributes}
+
+\vspace{1mm} \footnotesize
+
+\normalsize
+
+\vspace{1mm} \scriptsize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{neutral_words =}\StringTok{ }\NormalTok{[}\StringTok{'liquor'}\NormalTok{, }\StringTok{'pow'}\NormalTok{, }\StringTok{'ballpark'}\NormalTok{, }\StringTok{'glitchy'}\NormalTok{, }\StringTok{'billy'}\NormalTok{, }\StringTok{'dallas'}\NormalTok{, }
+                 \StringTok{'rip'}\NormalTok{, }\StringTok{'called'}\NormalTok{, }\StringTok{'outlooks'}\NormalTok{, }\StringTok{'viet'}\NormalTok{, }\StringTok{'floater'}\NormalTok{, }\StringTok{'rattlesnake'}\NormalTok{, }\StringTok{'exports'}\NormalTok{,}
+                 \StringTok{'peruvian'}\NormalTok{, }\StringTok{'recursion'}\NormalTok{, }\StringTok{'shortfall'}\NormalTok{, }\StringTok{'corrected'}\NormalTok{, }\StringTok{'amicable'}\NormalTok{,}
+                 \StringTok{'solutions'}\NormalTok{, }\StringTok{'diagnostic'}\NormalTok{, }\StringTok{'patently'}\NormalTok{, }\StringTok{'flops'}\NormalTok{, }\StringTok{'approx'}\NormalTok{, }\StringTok{'percents'}\NormalTok{,}
+                \StringTok{'lox'}\NormalTok{, }\StringTok{'catapults'}\NormalTok{, }\StringTok{'hamburger'}\NormalTok{, }\StringTok{'engulfed'}\NormalTok{, }\StringTok{'households'}\NormalTok{, }\StringTok{'north'}\NormalTok{,}
+                \StringTok{'snubbed'}\NormalTok{, }\StringTok{'playtest'}\NormalTok{, }\StringTok{'replayability'}\NormalTok{, }\StringTok{'glottal'}\NormalTok{, }\StringTok{'parable'}\NormalTok{, }\StringTok{'gingers'}\NormalTok{,}
+                \StringTok{'anachronism'}\NormalTok{, }\StringTok{'organizing'}\NormalTok{, }\StringTok{'reach'}\NormalTok{, }\StringTok{'shtick'}\NormalTok{, }\StringTok{'eleventh'}\NormalTok{, }\StringTok{'cpu'}\NormalTok{, }\StringTok{'ranked'}\NormalTok{,}
+                \StringTok{'irreversibly'}\NormalTok{, }\StringTok{'ponce'}\NormalTok{, }\StringTok{'velociraptor'}\NormalTok{, }\StringTok{'rubber'}\NormalTok{, }\StringTok{'defects'}\NormalTok{, }\StringTok{'puzzle'}\NormalTok{,}
+                \StringTok{'smasher'}\NormalTok{, }\StringTok{'northside'}\NormalTok{, }\StringTok{'heft'}\NormalTok{, }\StringTok{'observation'}\NormalTok{, }\StringTok{'rectum'}\NormalTok{, }\StringTok{'mystical'}\NormalTok{,}
+                \StringTok{'telltale'}\NormalTok{, }\StringTok{'remnants'}\NormalTok{, }\StringTok{'inquiry'}\NormalTok{, }\StringTok{'indisputable'}\NormalTok{, }\StringTok{'boatload'}\NormalTok{, }\StringTok{'lessening'}\NormalTok{,}
+                \StringTok{'uselessness'}\NormalTok{, }\StringTok{'observes'}\NormalTok{, }\StringTok{'fictitious'}\NormalTok{, }\StringTok{'repatriation'}\NormalTok{, }\StringTok{'duh'}\NormalTok{,}
+                \StringTok{'attic'}\NormalTok{, }\StringTok{'schilling'}\NormalTok{, }\StringTok{'charges'}\NormalTok{, }\StringTok{'chatter'}\NormalTok{, }\StringTok{'pad'}\NormalTok{, }\StringTok{'smurfing'}\NormalTok{,}
+                \StringTok{'worthiness'}\NormalTok{, }\StringTok{'definitive'}\NormalTok{, }\StringTok{'neat'}\NormalTok{, }\StringTok{'homogenized'}\NormalTok{,}\StringTok{'lexicon'}\NormalTok{,}
+                \StringTok{'nationalized'}\NormalTok{, }\StringTok{'earpiece'}\NormalTok{, }\StringTok{'specializations'}\NormalTok{, }\StringTok{'lapse'}\NormalTok{,}
+                \StringTok{'concludes'}\NormalTok{, }\StringTok{'weaving'}\NormalTok{, }\StringTok{'apprentices'}\NormalTok{, }\StringTok{'fri'}\NormalTok{, }\StringTok{'younglings'}\NormalTok{,}
+                \StringTok{'militias'}\NormalTok{, }\StringTok{'inscriptions'}\NormalTok{, }\StringTok{'gouda'}\NormalTok{, }\StringTok{'lift'}\NormalTok{, }\StringTok{'laboring'}\NormalTok{, }
+                \StringTok{'adaptive'}\NormalTok{, }\StringTok{'lecture'}\NormalTok{, }\StringTok{'hogging'}\NormalTok{, }\StringTok{'thorne'}\NormalTok{, }\StringTok{'fud'}\NormalTok{, }\StringTok{'skews'}\NormalTok{,}
+                \StringTok{'epistles'}\NormalTok{, }\StringTok{'tagging'}\NormalTok{, }\StringTok{'crud'}\NormalTok{, }\StringTok{'two'}\NormalTok{, }\StringTok{'rebalanced'}\NormalTok{, }\StringTok{'payroll'}\NormalTok{,}
+                \StringTok{'damned'}\NormalTok{, }\StringTok{'approve'}\NormalTok{, }\StringTok{'reason'}\NormalTok{, }\StringTok{'formally'}\NormalTok{, }\StringTok{'releasing'}\NormalTok{, }\StringTok{'muddled'}\NormalTok{,}
+                \StringTok{'mineral'}\NormalTok{, }\StringTok{'shied'}\NormalTok{, }\StringTok{'capital'}\NormalTok{, }\StringTok{'nodded'}\NormalTok{, }\StringTok{'escrow'}\NormalTok{, }\StringTok{'unscientific'}\NormalTok{,}
+                \StringTok{'recognizable'}\NormalTok{, }\StringTok{'entitlement'}\NormalTok{, }\StringTok{'disconnecting'}\NormalTok{, }\StringTok{'marshals'}\NormalTok{,}
+                \StringTok{'winamp'}\NormalTok{, }\StringTok{'forceful'}\NormalTok{, }\StringTok{'lowes'}\NormalTok{, }\StringTok{'ptr'}\NormalTok{, }\StringTok{'sip'}\NormalTok{, }\StringTok{'pencils'}\NormalTok{, }\StringTok{'stomachs'}\NormalTok{,}
+                \StringTok{'goff'}\NormalTok{, }\StringTok{'cg'}\NormalTok{, }\StringTok{'backyard'}\NormalTok{, }\StringTok{'uprooting'}\NormalTok{, }\StringTok{'merging'}\NormalTok{,}
+                \StringTok{'helpful'}\NormalTok{, }\StringTok{'eid'}\NormalTok{, }\StringTok{'trenchcoat'}\NormalTok{, }\StringTok{'airlift'}\NormalTok{, }\StringTok{'frothing'}\NormalTok{,}
+                \StringTok{'pulls'}\NormalTok{, }\StringTok{'volta'}\NormalTok{, }\StringTok{'guinness'}\NormalTok{, }\StringTok{'viewership'}\NormalTok{,}
+                \StringTok{'eruption'}\NormalTok{, }\StringTok{'peeves'}\NormalTok{, }\StringTok{'goat'}\NormalTok{, }\StringTok{'goofy'}\NormalTok{, }\StringTok{'disbanding'}\NormalTok{, }
+                \StringTok{'relented'}\NormalTok{, }\StringTok{'ratings'}\NormalTok{, }\StringTok{'disputed'}\NormalTok{, }\StringTok{'vitamins'}\NormalTok{, }\StringTok{'singled'}\NormalTok{,}
+                \StringTok{'hydroxide'}\NormalTok{, }\StringTok{'telegraphed'}\NormalTok{, }\StringTok{'mercantile'}\NormalTok{, }\StringTok{'headache'}\NormalTok{, }
+                \StringTok{'muppets'}\NormalTok{, }\StringTok{'petal'}\NormalTok{, }\StringTok{'arrange'}\NormalTok{, }\StringTok{'donovan'}\NormalTok{, }\StringTok{'scrutinized'}\NormalTok{,}
+                \StringTok{'spoil'}\NormalTok{, }\StringTok{'examiner'}\NormalTok{, }\StringTok{'ironed'}\NormalTok{, }\StringTok{'maia'}\NormalTok{, }\StringTok{'condensation'}\NormalTok{,}
+                \StringTok{'receipt'}\NormalTok{, }\StringTok{'solider'}\NormalTok{, }\StringTok{'tattooing'}\NormalTok{, }\StringTok{'encoded'}\NormalTok{,}
+                \StringTok{'compartmentalize'}\NormalTok{, }\StringTok{'lain'}\NormalTok{, }\StringTok{'the'}\NormalTok{, }\StringTok{'gov'}\NormalTok{,}
+                \StringTok{'printers'}\NormalTok{, }\StringTok{'hiked'}\NormalTok{, }\StringTok{'resentment'}\NormalTok{, }\StringTok{'revisionism'}\NormalTok{, }\StringTok{'tavern'}\NormalTok{,}
+                \StringTok{'backpacking'}\NormalTok{, }\StringTok{'pestering'}\NormalTok{, }\StringTok{'gassed'}\NormalTok{, }\StringTok{'acknowledges'}\NormalTok{,}
+                \StringTok{'testimonies'}\NormalTok{, }\StringTok{'parlance'}\NormalTok{, }\StringTok{'hallucinate'}\NormalTok{, }\StringTok{'speeches'}\NormalTok{,}
+                \StringTok{'engaging'}\NormalTok{, }\StringTok{'solder'}\NormalTok{, }\StringTok{'perceptive'}\NormalTok{, }\StringTok{'microbiology'}\NormalTok{, }\StringTok{'reconnaissance'}\NormalTok{, }\StringTok{'garlic'}\NormalTok{,}
+                \StringTok{'neutrals'}\NormalTok{, }\StringTok{'width'}\NormalTok{, }\StringTok{'literaly'}\NormalTok{, }\StringTok{'guild'}\NormalTok{, }\StringTok{'despicable'}\NormalTok{, }\StringTok{'dion'}\NormalTok{,}
+                \StringTok{'option'}\NormalTok{, }\StringTok{'transistors'}\NormalTok{, }\StringTok{'chiropractic'}\NormalTok{, }\StringTok{'tattered'}\NormalTok{, }\StringTok{'consolidating'}\NormalTok{,}
+                \StringTok{'olds'}\NormalTok{, }\StringTok{'garmin'}\NormalTok{, }\StringTok{'shift'}\NormalTok{, }\StringTok{'granted'}\NormalTok{, }\StringTok{'intramural'}\NormalTok{, }\StringTok{'allie'}\NormalTok{, }\StringTok{'cylinders'}\NormalTok{,}
+                \StringTok{'wishlist'}\NormalTok{, }\StringTok{'crank'}\NormalTok{, }\StringTok{'wrongly'}\NormalTok{, }\StringTok{'workshop'}\NormalTok{, }\StringTok{'yesterday'}\NormalTok{, }\StringTok{'wooden'}\NormalTok{, }
+                \StringTok{'without'}\NormalTok{, }\StringTok{'wheel'}\NormalTok{, }\StringTok{'weather'}\NormalTok{, }\StringTok{'watch'}\NormalTok{, }\StringTok{'wage'}\NormalTok{, }\StringTok{'version'}\NormalTok{, }\StringTok{'usually'}\NormalTok{,}
+                \StringTok{'twice'}\NormalTok{, }\StringTok{'tomato'}\NormalTok{, }\StringTok{'ticket'}\NormalTok{, }\StringTok{'text'}\NormalTok{, }\StringTok{'switch'}\NormalTok{, }\StringTok{'studio'}\NormalTok{, }\StringTok{'stick'}\NormalTok{,}
+                \StringTok{'soup'}\NormalTok{, }\StringTok{'sometimes'}\NormalTok{, }\StringTok{'signal'}\NormalTok{, }\StringTok{'prior'}\NormalTok{, }\StringTok{'plant'}\NormalTok{, }\StringTok{'photo'}\NormalTok{,}
+                \StringTok{'path'}\NormalTok{, }\StringTok{'park'}\NormalTok{, }\StringTok{'near'}\NormalTok{, }\StringTok{'menu'}\NormalTok{, }\StringTok{'latter'}\NormalTok{, }\StringTok{'grass'}\NormalTok{, }\StringTok{'clock'}\NormalTok{]}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+\chapter*{References}\label{references}
+\addcontentsline{toc}{chapter}{References}
 

### 2021-06-22 thesis/index.Rmd

--- 2021-06-08 thesis/index.Rmd
@@
 documentclass: book
+urlcolor: blue
 classoptions: dvipsnames, usenames

### 2021-06-22 thesis/02-cosine.Rmd
fixing small things

@@
 The triangle inequality formula says that for any triangle, the sum of the lengths of any two sides must be greater than or equal to the length of the remaining side.
-As shown in [discussion in stats.stackexchange.com](https://stats.stackexchange.com/questions/198080/proving-that-cosine-distance-function-defined-by-cosine-similarity-between-two-u)^[https://stats.stackexchange.com/questions/198080/proving-that-cosine-distance-function-defined-by-cosine-similarity-between-two-u] in the case of cosine distance it would have to fulfill this equation 1+cos-sim(𝐴,𝐶) < cos-sim(𝐴,𝐵) + cos-sim(𝐵,𝐶).
+As shown in [discussion in stats.stackexchange.com](https://stats.stackexchange.com/questions/198080/proving-that-cosine-distance-function-defined-by-cosine-similarity-between-two-u)^[https://stats.stackexchange.com/questions/198080/proving-that-cosine-distance-function-defined-by-cosine-similarity-between-two-u] in the case of cosine distance it would have to fulfill this equation $1+\mathsf{cos-sim}(A,C) < \mathsf{cos-sim}(A,B) + \mathsf{cos-sim}(B,C)$.
 If one chooses specific unit vectors it is easy to demonstrate that the triangle inequality is not preserved.
@@
 
+\begin{minipage}{1.0\textwidth}
 \textbf{"jew"       : ["judaism", "jew", "synagogue", "torah", "rabbi"]}
@@
 \textbf{"muslim"    : ["islam", "muslim", "mosque", "quran", "imam"]}
+\end{minipage}
 
@@
 
+\begin{minipage}{1.0\textwidth}
 \textbf{"jew"       : ["greedy", "cheap", "hairy", "liberal"]}
@@
 \textbf{"muslim"       : ["violent", "terrorist", "dirty", "uneducated"]}
+\end{minipage}
 
@@
 
-\pagebreak
-
+\begin{minipage}{1.0\textwidth}
 \textbf{["judaism", "christianity", "islam"]}
@@
 \textbf{["synagogue", "church", "mosque"]}
+\end{minipage}
 
@@
 As the cosine distance increases, it is assumed that the association between protected and stereotypical words decreases.
-
-\newpage
 

### 2021-06-22 thesis/03-bayesian.Rmd
fixing small things

@@
 \footnotesize
-```{r tableMuslimActive,echo=FALSE,eval=TRUE,fig.align = "center",cache=FALSE, fig.show = "hold", out.width = "100%"}
+```{r tableMuslimActive,echo=FALSE,eval=TRUE,fig.align = "center",cache=FALSE, fig.show = "hold", out.width = "100%", warning = FALSE, message = FALSE}
 library(tidyverse)
@@
 \footnotesize
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%", warning = FALSE, message = FALSE}
 library(tidyverse)

### 2021-06-22 thesis/_book/_main.tex
fixing small things

@@
 in the case of cosine distance it would have to fulfill this equation
-1+cos-sim(𝐴,𝐶) \textless{} cos-sim(𝐴,𝐵) + cos-sim(𝐵,𝐶). If one chooses
-specific unit vectors it is easy to demonstrate that the triangle
-inequality is not preserved.
+\(1+\mathsf{cos-sim}(A,C) < \mathsf{cos-sim}(A,B) + \mathsf{cos-sim}(B,C)\).
+If one chooses specific unit vectors it is easy to demonstrate that the
+triangle inequality is not preserved.
 
@@
 and stereotypical words decreases.
-
-\newpage
 

### 2021-06-23 thesis/_book/_main.tex
Adds logs

@@
-\documentclass[12pt,]{book}
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}
+%
+\documentclass[
+  12pt,
+]{book}
+\usepackage{amsmath,amssymb}
 \usepackage{lmodern}
 \usepackage{setspace}
-\setstretch{1.5}
-\usepackage{amssymb,amsmath}
 \usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
 \ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
@@
   linkcolor=Maroon,
+  filecolor=Maroon,
   citecolor=Blue,
   urlcolor=blue,
-            breaklinks=true}
-\urlstyle{same}  % don't use monospace font for urls
-\usepackage[left=3.5cm, right=2.5cm, top=2.5cm, bottom=2.5cm, asymmetric,
-includeheadfoot]{geometry}
+  pdfcreator={LaTeX via pandoc}}
+\urlstyle{same} % disable monospaced font for URLs
+\usepackage[left=3.5cm, right=2.5cm, top=2.5cm, bottom=2.5cm, asymmetric, includeheadfoot]{geometry}
 \usepackage{color}
@@
 {
-\hypersetup{linkcolor=black}
+\hypersetup{linkcolor=}
 \setcounter{tocdepth}{5}
@@
 }
-\chapter{Introduction}\label{introduction}
-
-Natural language processing (NLP) is a subfield of computer science that
-processes and analyzes language in text and speech with the use of
-modern programming methods. It has practical applications in everyday
-life as it concerns tasks such as email filters, smart assistants,
-search results, language translations, text analytics and so on. Models
-used to accomplish these tasks need a lot of data to learn from. This
-data originates from human activities and historical recordings such as
-texts, messages or speeches. It turns out that in the learning process
-these models can learn implicit biases that reflect harmful
-stereotypical thinking still present in modern societies. There are many
-different types of models in NLP depending on a task that they are
-supposed to solve. However, all of them need as an input words
-represented by means of numbers and this is accomplished with word
-embeddings. Word embeddings are usually assigned the values based on the
-context in which the words appear. This means that the input data can
-have an enormous influence on the outcome. One can find methods that aim
-at identifying and measuring hidden biases and/or try to remove them by,
-for example, modifying the word embeddings directly.
-
-One can find various definitions trying to capture what bias and
-fairness actually are. With the choice of the definition, implications
-for the real-life applications may change as well. Mehrabi, Morstatter,
-Saxena, Lerman, \& Galstyan
-(\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) mark out that there
-exist different types of biases such as historical bias, representation
-bias or measurement bias (the list is long). This indicates how complex
-the issue of bias is. Without the proper understanding and awareness of
-the problem, people are prone to unconsciously sustain the bias. Mehrabi
-et al. (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) also
-distinguish different types of discrimination, some of them will be
-briefly described. By protected attributes we mean those qualities,
-traits or characteristics that one should not discriminate against.
-Direct discrimination occurs when protected attributes of individuals
-explicitly result in non-favorable outcomes toward them. Such
-discrimination takes place for example when one uses the information
-about someone's religious beliefs to evaluate whether they are competent
-for the work they apply for. In contrast, in indirect discrimination
-individuals appear to be treated equally, but anyway they end up being
-treated unjustly due to the hidden effects of biases towards their
-protected attributes. An example of such situation is when in a job
-advert there is a requirement of 10 years experience instead of the list
-of a specific type of experience and knowledge. This job advert can
-discriminate indirectly against young people who can have the required
-skills, but not yet so much work experience. Systemic discrimination
-takes place when policies, customs or behaviors that result from certain
-culture or organizational structure lead to discrimination against some
-groups of people. Finally, statistical discrimination consists in
-relying on group statistics to judge a person belonging to a given
-group. The topic of discrimination is entangled with another concept;
-fairness. It is essential to grasp some concepts of fairness to take
-them into consideration while designing implementation of some machine
-learning model. In Mehrabi et al.
-(\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) one may notice that
-depending on the context and application different definitions may be
-applied.
-
-There is a considerable amount of literature available on the topic of
-bias detection and mitigation in NLP models. Bolukbasi, Chang, Zou,
-Saligrama, \& Kalai (\protect\hyperlink{ref-Bolukbasi2016Man}{2016})
-focus on gender biases that may be observable in the representation of
-job occupations and gender in terms of numerical values. The authors
-apply cosine similarity measurement to investigate the phenomenon where
-(the vectors corresponding to) words related to jobs that are
-stereotypically associated with a given gender are in fact in the model
-situated closer to this gender. They also use analogy tasks to evaluate
-if a bias is present in a word embedding model. They check analogies by
-comparing pairs of word vectors. For example they search for the word
-complementing the puzzle: man is to doctor as woman is to \ldots{}?
-First they subtract the word ``man'' from the word ``woman''\footnote{This
-  is pointwise subtraction of vectors, but for the simplicity we will
-  often talk about words while really meaning vectors. What is meant
-  should be clear from the context.} and then they search for the ranked
-list of other words pairs that have similar vector difference. They also
-include in the formula a threshold to ensure that the resulting pairs
-could not be randomly picked.
-
-However, as Nissim, Noord, \& Goot
-(\protect\hyperlink{ref-Nissim2019Fair}{2019}) points out, there are
-some limitations to this approach. In practice, most of analogies
-implementations do not return any input words. This means that it does
-not make sense to expect the algorithm to return the same profession for
-both woman and man. Therefore, this method of bias detection seems
-limited. Other problems are related for example to the choice of pairs
-and words that are used to detect the presence of discrimination, as it
-is often subjective and without proper justification. Additionally, the
-choice of the parameter used in Bolukbasi et al.
-(\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) formula to ensure that
-word pairs are not picked by random, is also not justified and changing
-it drastically influences the results.
-
-Islam, Bryson, \& Narayanan
-(\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) touch upon the
-topic of biases regarding race and gender. They apply knowledge from
-well-known psychological studies such as Implicit Association Test to
-investigate the relation between human stereotypical thinking and
-model-learnt biases to discover a close relationship between these two.
-For the evaluation they use Word Embedding Association Test (WEAT) and
-the Word Embedding Factual Association Test (WEFAT).
-
-Manzini, Lim, Tsvetkov, \& Black
-(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) propose a
-novel way of using cosine similarity to verify if the assumed
-resemblance between protected words and harmful attributes exists. They
-develop an approach that enables them to measure the bias for a class
-(such as gender, religion or race) and express the final result with a
-single metric. Their approach will be analyzed in details in the next
-chapter.
-
-Islam et al. (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) refer
-to Implicit Association Test (Greenwald et al., 1998) that measures the
-strength of associations between concepts or stereotypes by measuring
-human reaction time for special tasks. Humans naturally exhibit some
-biases which do not always cause social concern. One can imagine the
-intuitive associations between for example insects and flowers, and the
-feelings of pleasantness or unpleasantness. In general, people would
-rather associate flowers than insects with feeling pleasant, and this
-preference in a sense is a bias or prejudice in some direction. However,
-this type of preference does not cause an uproar and is a rather morally
-neutral case. This example is rather used to show that the methodology
-used in IAT makes sense. Islam et al.
-(\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) were able to
-obtain similar results to the ones that measured reaction time with the
-use of WEAT metric. Unfortunately, further studies discover other biases
-and prejudices that directly influence the quality of other people's
-lives and therefore they should be taken care of.
-
-The most common bias detection methods in the literature focus on
-comparing the similarity between words from protected groups and those
-that are considered to be stereotypical or harmful in some way. One can
-find in this group methods such as Euclidean distance or cosine
-similarity (which is equivalent to dot product if the vectors are
-normalized). There are also other ways to detect the effects of biases.
-One can investigate the model performance on certain downstream tasks
-and validate if the model uses the gender or race information to
-complete the task. Unfortunately, the currently used methods employing
-cosine similarity, often aggregate the similarity values in a way that
-may lead to hasty conclusions. The averaging of values and the lack of
-attention to uncertainty involved in such aggregation may lead to an
-incomplete picture of the situation.
-
-In the paper we indicate how current methods used to detect biases in
-natural language models are limited, if we investigate them from the
-perspective of Bayesian analysis. Our research enhances the current way
-in which the bias detection is performed to make sure that it is
-methodologically valid. The key hypothesis is that greater understanding
-of data and bias implications can be achieved when Bayesian methods are
-applied to issue.
-
-\chapter{Cosine similarity and bias
-detection}\label{cosine-similarity-and-bias-detection}
-
-\section{Word embeddings}\label{word-embeddings}
-
-To understand what cosine similarity measurement is, one first needs to
-grasp the concept of translating words to a computer-readable form. In
-the field of natural language processsing there are two main types of
-words representation --- localist and distributed. One-hot encoding is
-an example of a method used to achieve a localist representation of
-words. Here, each vector contains information only about a single data
-point. This is achieved by first mapping categorical values (words) to
-integers and then to each integers a binary vector is assigned which
-contains only 0s except for the index of the integer, which is assigned
-1. An example of a localist representation is:
+\setstretch{1.5}
+\hypertarget{introduction}{%
+\chapter{Introduction}\label{introduction}}
+
+Natural language processing (NLP) is a subfield of computer science that processes and analyzes language in text and speech with the use of modern programming methods.
+It has practical applications in everyday life as it concerns tasks such as email filters, smart assistants, search results, language translations, text analytics and so on.
+Models used to accomplish these tasks need a lot of data to learn from.
+This data originates from human activities and historical recordings such as texts, messages or speeches.
+It turns out that in the learning process these models can learn implicit biases that reflect harmful stereotypical thinking still present in modern societies.
+There are many different types of models in NLP depending on a task that they are supposed to solve.
+However, all of them need as an input words represented by means of numbers and this is accomplished with word embeddings.
+Word embeddings are usually assigned the values based on the context in which the words appear.
+This means that the input data can have an enormous influence on the outcome.
+One can find methods that aim at identifying and measuring hidden biases and/or try to remove them by, for example, modifying the word embeddings directly.
+
+One can find various definitions trying to capture what bias and fairness actually are.
+With the choice of the definition, implications for the real-life applications may change as well.
+\protect\hyperlink{ref-Mehrabi2019Survey}{Mehrabi, Morstatter, Saxena, Lerman, \& Galstyan} (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) mark out that there exist different types of biases such as historical bias, representation bias or measurement bias (the list is long).
+This indicates how complex the issue of bias is.
+Without the proper understanding and awareness of the problem, people are prone to unconsciously sustain the bias.
+\protect\hyperlink{ref-Mehrabi2019Survey}{Mehrabi, Morstatter, Saxena, Lerman, \& Galstyan} (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) also distinguish different types of discrimination, some of them will be briefly described.
+By protected attributes we mean those qualities, traits or characteristics that one should not discriminate against.
+Direct discrimination occurs when protected attributes of individuals explicitly result in non-favorable outcomes toward them.
+Such discrimination takes place for example when one uses the information about someone's religious beliefs to evaluate whether they are competent for the work they apply for.
+In contrast, in indirect discrimination individuals appear to be treated equally, but anyway they end up being treated unjustly due to the hidden effects of biases towards their protected attributes.
+An example of such situation is when in a job advert there is a requirement of 10 years experience instead of the list of a specific type of experience and knowledge.
+This job advert can discriminate indirectly against young people who can have the required skills, but not yet so much work experience.
+Systemic discrimination takes place when policies, customs or behaviors that result from certain culture or organizational structure lead to discrimination against some groups of people.
+Finally, statistical discrimination consists in relying on group statistics to judge a person belonging to a given group.
+The topic of discrimination is entangled with another concept; fairness.
+It is essential to grasp some concepts of fairness to take them into consideration while designing implementation of some machine learning model.
+In \protect\hyperlink{ref-Mehrabi2019Survey}{Mehrabi, Morstatter, Saxena, Lerman, \& Galstyan} (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) one may notice that depending on the context and application different definitions may be applied.
+
+There is a considerable amount of literature available on the topic of bias detection and mitigation in NLP models.
+\protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) focus on gender biases that may be observable in the representation of job occupations and gender in terms of numerical
+values.
+The authors apply cosine similarity measurement to investigate the phenomenon where (the vectors corresponding to) words related to jobs that are stereotypically associated with a given gender are in fact in the model situated closer to this gender.
+They also use analogy tasks to evaluate if a bias is present in a word embedding model.
+They check analogies by comparing pairs of word vectors.
+For example they search for the word complementing the puzzle: man is to doctor as woman is to \ldots?
+First they subtract the word ``man'' from the word ``woman''\footnote{This is pointwise subtraction of vectors, but for the simplicity we will often talk about words while really meaning vectors.
+What is meant should be clear from the context.} and then
+they search for the ranked list of other words pairs that have similar vector difference. They also include in the formula a threshold to ensure that the resulting pairs could not be randomly picked.
+
+However, as \protect\hyperlink{ref-Nissim2019Fair}{Nissim, Noord, \& Goot} (\protect\hyperlink{ref-Nissim2019Fair}{2019}) points out, there are some limitations to this approach.
+In practice, most of analogies implementations do not return any input words.
+This means that it does not make sense to expect the algorithm to return the same profession for both woman and man.
+Therefore, this method of bias detection seems limited.
+Other problems are related for example to the choice of pairs and words that are used to detect the presence of discrimination, as it is often subjective and without proper justification.
+Additionally, the choice of the parameter used in \protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) formula to ensure that word pairs are not picked by random, is also not justified and changing it drastically
+influences the results.
+
+\protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) touch upon the topic of biases regarding race and gender.
+They apply knowledge from well-known psychological studies such as Implicit Association Test to investigate the relation between human stereotypical thinking and model-learnt biases to discover a close relationship between these two.
+For the evaluation they use Word Embedding Association Test (WEAT) and the Word Embedding Factual Association Test (WEFAT).
+
+\protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) propose a novel way of using cosine similarity to verify if the assumed resemblance between protected words and harmful attributes exists.
+They develop an approach that enables them to measure the bias for a class (such as gender, religion or race) and express the final result with a single metric.
+Their approach will be analyzed in details in the next chapter.
+
+\protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) refer to Implicit Association Test (Greenwald et al., 1998) that measures the strength of associations between concepts or stereotypes by measuring human reaction time for special tasks.
+Humans naturally exhibit some biases which do not always cause social concern.
+One can imagine the intuitive associations between for example insects and flowers, and the feelings of pleasantness or unpleasantness.
+In general, people would rather associate flowers than insects with feeling pleasant, and this preference in a sense is a bias or
+prejudice in some direction.
+However, this type of preference does not cause an uproar and is a rather morally neutral case.
+This example is rather used to show that the methodology used in IAT makes sense.
+\protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) were able to obtain similar results to the ones that measured reaction time with the use of WEAT metric.
+Unfortunately, further studies discover other biases and prejudices that directly influence the quality of other people's lives and therefore they should be taken care of.
+
+The most common bias detection methods in the literature focus on comparing the similarity between words from protected groups and those that are considered to be stereotypical or harmful in some way.
+One can find in this group methods such as Euclidean distance or cosine similarity (which is equivalent to dot product if the vectors are normalized).
+There are also other ways to detect the effects of biases.
+One can investigate the model performance on certain downstream tasks and validate if the model uses the gender or race information to complete the task.
+Unfortunately, the currently used methods employing cosine similarity, often aggregate the similarity values in a way that may lead to hasty conclusions.
+The averaging of values and the lack of attention to uncertainty involved in such aggregation may lead to an incomplete picture of the situation.
+
+In the paper we indicate how current methods used to detect biases in natural language models are limited, if we investigate them from the perspective of Bayesian analysis.
+Our research enhances the current way in which the bias detection is performed to make sure that it is methodologically valid.
+The key hypothesis is that greater understanding of data and bias implications can be achieved when Bayesian methods are applied to issue.
+
+\hypertarget{cosine-similarity-and-bias-detection}{%
+\chapter{Cosine similarity and bias detection}\label{cosine-similarity-and-bias-detection}}
+
+\hypertarget{word-embeddings}{%
+\section{Word embeddings}\label{word-embeddings}}
+
+To understand what cosine similarity measurement is, one first needs to grasp the concept of translating words to a computer-readable form.
+In the field of natural language processsing there are two main types of words representation --- localist and distributed.
+One-hot encoding is an example of a method used to achieve a localist representation of words.
+Here, each vector contains information only about a single data point.
+This is achieved by first mapping categorical values (words) to integers and then to each integers a binary vector is assigned which contains only 0s except for the index of the integer, which is assigned 1.
+An example of a localist representation is:
 
@@
 
-One of the advantages of using a distributed representation is that one
-is able to represent an enormous number of concepts with a smaller
-number of units. It is also possible to better capture similarities as
-words of similar meanings can have similar numeric vectors.
-
-The numbers occurring in such representations are not random. They are
-learned in a process that uses a very shallow neural network. There are
-various types of techniques used for learning the vectors
-representations. One of the most straightforward ones is a skip-gram
-model. Given a word, the model tries to predict its neighboring words
-from the sentence. The mathematics behind the process relies on the idea
-that the prediction concerns the conditional probability of the adjacent
-words. The algorithm tries to minimize the loss function, which
-penalizes the system for discrepancy with actual co-occurrence
-frequencies in the corpus. One can choose various parameters of the
-model, such as the window size that determines how many surrounding
-words the model should predict. After preparing such a fitted model, one
-takes only the learned weights from a neural network, and uses them as
-vectors in a word embeddings representation.
+One of the advantages of using a distributed representation is that one is able to represent an enormous number of concepts with a smaller number of units.
+It is also possible to better capture similarities as words of similar meanings can have similar numeric vectors.
+
+The numbers occurring in such representations are not random.
+They are learned in a process that uses a very shallow neural network.
+There are various types of techniques used for learning the vectors representations.
+One of the most straightforward ones is a skip-gram model.
+Given a word, the model tries to predict its neighboring words from the sentence.
+The mathematics behind the process relies on the idea that the prediction concerns the conditional probability of the adjacent words.
+The algorithm tries to minimize the loss function, which penalizes the system for discrepancy with actual co-occurrence frequencies in the corpus.
+One can choose various parameters of the model, such as the window size that determines how many surrounding words the model should predict.
+After preparing such a fitted model, one takes only the learned weights from a neural network, and uses them as vectors in a word embeddings representation.
 
 Word embeddings have many applications in natural language processing.
-They are handy in document search and information retrieval. They also
-play their part in improving automatic translations. Well learned word
-representations may also contribute to the improvement of sentiment
-analysis or spam detection.
-
-\section{Cosine similarity and
-distance}\label{cosine-similarity-and-distance}
-
-Cosine similarity is often used as a method of finding out whether
-vector representations for two words suggest that they are similar or
-somehow connected. Cosine similarity is the cosine of the angle between
-two vectors: the result of dividing their inner product (dot product
-usually) by the product of their magnitudes.
+They are handy in document search and information retrieval.
+They also play their part in improving automatic translations.
+Well learned word representations may also contribute to the improvement of sentiment analysis or spam detection.
+
+\hypertarget{cosine-similarity-and-distance}{%
+\section{Cosine similarity and distance}\label{cosine-similarity-and-distance}}
+
+Cosine similarity is often used as a method of finding out whether vector representations for two words suggest that they are similar or somehow connected.
+Cosine similarity is the cosine of the angle between two vectors: the result of dividing their inner product (dot product usually) by the product of their magnitudes.
 
 It is worth mentioning one more point concerning cosine similarity.
-After the vectors are normalized to have length equal to 1, the inner
-product itself (often dot product) is used to measure the similarity,
-because it then equals the cosine similarity.
+After the vectors are normalized to have length equal to 1, the inner product itself (often dot product) is used to measure the similarity, because it then equals the cosine similarity.
 
@@
 \end{align}
-
-Cosine similarity is considered a proper tool for this operation as its
-result has a clear connection to geometry and at least for a low number
-of dimensions may be easily interpreted. Using this scale, one can
-compare vector similarities in a fairly clear manner. When the vectors
-are aligned perpendicularly to each other, their similarity equals 0
-(which is the same as the cosine of 90 degrees). This tells us that the
-similarity between the vectors is small. As the angle between vectors
-decreases, cosine similarity approaches one, which stands for the
-greatest similarity. It is also possible to obtain negative cosine
-similarity. If the value approaches -1, this intuitively means that the
-words are contrary to each other.
-
-One of the limitations of this measure is that it informs us only about
-similarities between vectors in terms of their orientation. However, it
-is often argued that in comparing words in terms of this metric, the
-magnitude of vectors may be treated as irrelevant, as the most important
-information pertains to direction.
-
-In what follows, it is important to distinguish between cosine
-similarity and cosine distance, defined as:
-
+Cosine similarity is considered a proper tool for this operation as its result has a clear connection to geometry and at least for a low number of dimensions may be easily interpreted.
+Using this scale, one can compare vector similarities in a fairly clear manner.
+When the vectors are aligned perpendicularly to each other, their similarity equals 0 (which is the same as the cosine of 90 degrees).
+This tells us that the similarity between the vectors is small.
+As the angle between vectors decreases, cosine similarity approaches one, which stands for the greatest similarity.
+It is also possible to obtain negative cosine similarity.
+If the value approaches -1, this intuitively means that the words are contrary to each other.
+
+One of the limitations of this measure is that it informs us only about similarities between vectors in terms of their orientation.
+However, it is often argued that in comparing words in terms of this metric, the magnitude of vectors may be treated as irrelevant, as the most important information pertains to direction.
+
+In what follows, it is important to distinguish between cosine similarity and cosine distance, defined as:
 \begin{align} \tag{Sim}
@@
 
-The greater the similarity between two vectors, the smaller the distance
-between them. The cosine distance ranges between 0 and 2. If the vectors
-are in an opposite directions, the cosine distance is 2. And if the
-vectors are extremely similar then the cosine distance is very close to
-0.
-
-One should note that cosine distance is not exactly a distance measure,
-as it does not meet triangle inequality requirements. The triangle
-inequality formula says that for any triangle, the sum of the lengths of
-any two sides must be greater than or equal to the length of the
-remaining side. As shown in
-\href{https://stats.stackexchange.com/questions/198080/proving-that-cosine-distance-function-defined-by-cosine-similarity-between-two-u}{discussion
-in stats.stackexchange.com}\footnote{\url{https://stats.stackexchange.com/questions/198080/proving-that-cosine-distance-function-defined-by-cosine-similarity-between-two-u}}
-in the case of cosine distance it would have to fulfill this equation
-\(1+\mathsf{cos-sim}(A,C) < \mathsf{cos-sim}(A,B) + \mathsf{cos-sim}(B,C)\).
-If one chooses specific unit vectors it is easy to demonstrate that the
-triangle inequality is not preserved.
-
-\section{Cosine distance in a one-class bias
-detection}\label{cosine-distance-in-a-one-class-bias-detection}
-
-Bolukbasi et al. (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) define
-similarity between words as the outcome inner product of their
-normalized vectors. They focus on examining what the geometry of a word
-embedding is in regard to ``he'' and ``she'' words. In other words,
-whether the similarity between those concepts and other words reflects
-expected gender stereotypes. They test this hypothesis by investigating
-whether there is a connection between word embeddings representing
-certain professions and words referring to gender. They also evaluate
-whether automatically produced analogies between words reflect the
-stereotypes as well.
-
-Here the gender bias of a word \(w\) is understood as its projection on
-the gender direction
-\(\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})\) (the
-gender direction is the top principal compontent of ten gender pair
-difference vectors). The underlying idea is that no bias is present if
-non-explicitly gender words are in equal distance to both elements in
-all explicitly gender pairs. Given the (ideally) gender neutral words
-\(N\) and the gender direction \(g\) the direct direct gender bias is
-defined as the average distance of the words in \(N\) from \(g\) (\(c\)
-is a parameter determining how strict we want to be):
-
+The greater the similarity between two vectors, the smaller the distance between them.
+The cosine distance ranges between 0 and 2.
+If the vectors are in an opposite directions, the cosine distance is 2.
+And if the vectors are extremely similar then the cosine distance is very close to 0.
+
+One should note that cosine distance is not exactly a distance measure, as it does not meet triangle inequality requirements.
+The triangle inequality formula says that for any triangle, the sum of the lengths of any two sides must be greater than or equal to the length of the remaining side.
+As shown in \href{https://stats.stackexchange.com/questions/198080/proving-that-cosine-distance-function-defined-by-cosine-similarity-between-two-u}{discussion in stats.stackexchange.com}\footnote{\url{https://stats.stackexchange.com/questions/198080/proving-that-cosine-distance-function-defined-by-cosine-similarity-between-two-u}} in the case of cosine distance it would have to fulfill this equation \(1+\mathsf{cos-sim}(A,C) < \mathsf{cos-sim}(A,B) + \mathsf{cos-sim}(B,C)\).
+If one chooses specific unit vectors it is easy to demonstrate that the triangle inequality is not preserved.
+
+\hypertarget{cosine-distance-in-a-one-class-bias-detection}{%
+\section{Cosine distance in a one-class bias detection}\label{cosine-distance-in-a-one-class-bias-detection}}
+
+\protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) define similarity between words as the outcome inner product of their normalized vectors.
+They focus on examining what the geometry of a word embedding is in regard to ``he'' and ``she'' words.
+In other words, whether the similarity between those concepts and other words reflects expected gender stereotypes.
+They test this hypothesis by investigating whether there is a connection between word embeddings representing certain professions and words referring to gender.
+They also evaluate whether automatically produced analogies between words reflect the stereotypes as well.
+
+Here the gender bias of a word \(w\) is understood as its projection on the gender direction \(\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})\) (the gender direction is the top principal compontent of ten gender pair difference vectors).
+The underlying idea is that no bias is present if non-explicitly gender words are in equal distance to both elements in all explicitly gender pairs.
+Given the (ideally) gender neutral words \(N\) and the gender direction \(g\) the direct direct gender bias is defined as the average distance of the words in \(N\) from \(g\) (\(c\) is a parameter determining how strict we want to be):
 \begin{align}
@@
 
-A very vivid way to follow their method of arguing that bias in word
-embeddings is real is to plot the values of inner product of chosen
-words. The plot below does not originate from the original paper but
-from
-\href{https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings}{kaggle.com
-R notebook using data from GloVe: Global Vectors for Word
-Representation}.\footnote{\url{https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings}.}
-However, similar visualization may be found there. Data used to create
-our plot is as follows.
-
-Occupations associated with feminine:
-\textbf{"homemaker", "nurse", "receptionist", "librarian", "socialite", "hairdresser", "nanny", "bookkeeper", "stylist", "housekeeper", "interior designer", "guidance counselor"}
-
-Occupations associated with masculine:
-\textbf{"maestro", "skipper", "protege", "philosopher", "captain", "architect", "financier", "warrior", "broadcaster", "magician", "fighter pilot", "boss"}
-
-\vspace{1mm} \footnotesize
+A very vivid way to follow their method of arguing that bias in word embeddings is real is to plot the values of inner product of chosen words.
+The plot below does not originate from the original paper but from \href{https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings}{kaggle.com R notebook using data from GloVe: Global Vectors for Word Representation}.\footnote{\url{https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings}.} However, similar visualization may be found there.
+Data used to create our plot is as follows.
+
+Occupations associated with feminine: \textbf{"homemaker", "nurse", "receptionist", "librarian", "socialite", "hairdresser", "nanny", "bookkeeper", "stylist", "housekeeper", "interior designer", "guidance counselor"}
+
+Occupations associated with masculine: \textbf{"maestro", "skipper", "protege", "philosopher", "captain", "architect", "financier", "warrior", "broadcaster", "magician", "fighter pilot", "boss"}
+
+\vspace{1mm}
+\footnotesize
 
 \begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-1-1} \end{center}
-
-\normalsize
-
-The points in the plot above result from the calculation of the inner
-product of a chosen vector for a profession word and a vector for a
-gender word (she or he). Inner product of two vectors expresses
-similarity between words. This assumption originates from the geometry
-and properties of a vector space. In the picture one may observe a
-correlation between the occupation and gender. Stereotypical male
-professions are closer to the ``he'' word and stereotypical female
-professions have greater similarity with ``she'' word. One could argue
-that this is a hint that there is some hidden bias information stored in
-the words vectors.
-
-\section{Cosine distance in a multi-class bias
-detection}\label{cosine-distance-in-a-multi-class-bias-detection}
-
-Manzini et al.
-(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) present a
-different approach towards finding similarities between classes of
-words. Cosine distance is used in the article as a measure to first
-argue for the existence of multi-class bias and then to show how through
-bias mitigation techniques the bias may be decreased.
-
-They modify WEAT to a multi-class setting, introducing Mean Average
-Cosine similarity as a measure of bias (in fact, in the paper they
-report distances rather than similarities). Let
-\(T = \{t_1, \dots, t_k\}\) be a class of protected word embeddings, and
-let each \(A_j\in A\) be a set of attributes stereotypically associated
-with a protected word). Then:
-
+\normalsize
+
+The points in the plot above result from the calculation of the inner product of a chosen vector for a profession word and a vector for a gender word (she or he).
+Inner product of two vectors expresses similarity between words.
+This assumption originates from the geometry and properties of a vector space.
+In the picture one may observe a correlation between the occupation and gender.
+Stereotypical male professions are closer to the ``he'' word and stereotypical female professions have greater similarity with ``she'' word.
+One could argue that this is a hint that there is some hidden bias information stored in the words vectors.
+
+\hypertarget{cosine-distance-in-a-multi-class-bias-detection}{%
+\section{Cosine distance in a multi-class bias detection}\label{cosine-distance-in-a-multi-class-bias-detection}}
+
+\protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) present a different approach towards finding similarities between classes of words.
+Cosine distance is used in the article as a measure to first argue for the existence of multi-class bias and then to show how through bias mitigation techniques the bias may be decreased.
+
+They modify WEAT to a multi-class setting, introducing Mean Average Cosine similarity as a measure of bias (in fact, in the paper they report distances rather than similarities).
+Let \(T = \{t_1, \dots, t_k\}\) be a class of protected word embeddings, and let each \(A_j\in A\) be a set of attributes stereotypically associated with a protected word).
+Then:
 \begin{align}
@@
 
-As the code has been provided by the authors we were able to reconstruct
-their results. The main steps in the procedure are as follows. Let us go
-through an example that refers to the process of hard debiasing on
-religious attributes.
-
-First we load word embeddings from
-reddit.US.txt.tok.clean.cleanedforw2v.w2v dataset. The word embeddings
-have only 50 dimensions and the number of individual words from the
-dataset is 44895. The authors assume that the protected group should
-ideally not have high cosine similarity to stereotypical words. The word
-embeddings geometry should not place this group close to harmful
-stereotypes, if it is to be bias-free. For instance, let's look at the
-religion-related words.
+As the code has been provided by the authors we were able to reconstruct their results.
+The main steps in the procedure are as follows.
+Let us go through an example that refers to the process of hard debiasing on religious attributes.
+
+First we load word embeddings from reddit.US.txt.tok.clean.cleanedforw2v.w2v dataset.
+The word embeddings have only 50 dimensions and the number of individual words from the dataset is 44895.
+The authors assume that the protected group should ideally not have high cosine similarity to stereotypical words.
+The word embeddings geometry should not place this group close to harmful stereotypes, if it is to be bias-free.
+For instance, let's look at the religion-related words.
 
@@
 
-We have prepared a table presenting the values of cosine distance for
-each protected word with each attribute (stereotype). The part of the
-results is shown in Table 2.3
+We have prepared a table presenting the values of cosine distance for each protected word with each attribute (stereotype). The part of the results is shown in Table 2.3
 
@@
 \begin{Highlighting}[]
-\NormalTok{religion <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
-\KeywordTok{colnames}\NormalTok{(religion) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
+\NormalTok{religion }\OtherTok{\textless{}{-}} \FunctionTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{]}
+\FunctionTok{colnames}\NormalTok{(religion) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
                         \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
-\NormalTok{religion}\OperatorTok{$}\NormalTok{wordClass <-}\StringTok{ }\KeywordTok{as.factor}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass)}
-\KeywordTok{levels}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
-\KeywordTok{head}\NormalTok{(religion)  }\OperatorTok{%>%}\StringTok{  }\KeywordTok{kable}\NormalTok{(}\DataTypeTok{format =} \StringTok{"latex"}\NormalTok{,}\DataTypeTok{booktabs=}\NormalTok{T,}
-                      \DataTypeTok{linesep =} \StringTok{""}\NormalTok{,  }\DataTypeTok{escape =} \OtherTok{FALSE}\NormalTok{, }
-                      \DataTypeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\OperatorTok{%>%}
-\StringTok{                      }\KeywordTok{kable_styling}\NormalTok{(}\DataTypeTok{latex_options=}\KeywordTok{c}\NormalTok{(}\StringTok{"scale_down"}\NormalTok{))}
+\NormalTok{religion}\SpecialCharTok{$}\NormalTok{wordClass }\OtherTok{\textless{}{-}} \FunctionTok{as.factor}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{wordClass)}
+\FunctionTok{levels}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{wordClass) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
+\FunctionTok{head}\NormalTok{(religion)  }\SpecialCharTok{\%\textgreater{}\%}  \FunctionTok{kable}\NormalTok{(}\AttributeTok{format =} \StringTok{"latex"}\NormalTok{,}\AttributeTok{booktabs=}\NormalTok{T,}
+                      \AttributeTok{linesep =} \StringTok{""}\NormalTok{,  }\AttributeTok{escape =} \ConstantTok{FALSE}\NormalTok{, }
+                      \AttributeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+                      \FunctionTok{kable\_styling}\NormalTok{(}\AttributeTok{latex\_options=}\FunctionTok{c}\NormalTok{(}\StringTok{"scale\_down"}\NormalTok{))}
 \end{Highlighting}
@@
 
-In the article there was no analysis of individual distances, but the
-general look at the means. The authors introduced a metric that tries to
-generalize WEAT in measuring the presence of bias through the
-classification of multi-class bias in groups of words connected with
-gender, religion or race. In the process they first take the mean of
-cosine distances between a given protected word and attributes assigned
-to each stereotype. They do not differentiate between stereotypes
-associated with a word and stereotypes associated with different words
-(in the case of religion, stereotypes characteristic for Christianity
-has also cosine distance measured with for instance, Judaism or Islam).
-Then, after collecting the list of mean cosine distances, they average
-the list to obtain one final value representing the whole group, in this
-example religion, for which the final mean of all mean distances is
-equal to 0.859.
-
-In the article the authors also try to remove biases, as previously
-defined, from a word embedding. First they identify the bias subspace
-using Principal Component Analysis (PCA) which is a technique for
-dimensionality reduction. It is applied here to choose the subspace that
-contains the greatest amount of information. There can be many subspaces
-found in a given group. For example in terms of religion one can
-identify at least a few sets that are to grasp the concept of religion
-in general:
+In the article there was no analysis of individual distances, but the general look at the means.
+The authors introduced a metric that tries to generalize WEAT in measuring the presence of bias through the classification of multi-class bias in groups of words connected with gender, religion or race.
+In the process they first take the mean of cosine distances between a given protected word and attributes assigned to each stereotype.
+They do not differentiate between stereotypes associated with a word and stereotypes associated with different words (in the case of religion, stereotypes characteristic for Christianity has also cosine distance measured with for instance, Judaism or Islam).
+Then, after collecting the list of mean cosine distances, they average the list to obtain one final value representing the whole group, in this example religion, for which the final mean of all mean distances is equal to 0.859.
+
+In the article the authors also try to remove biases, as previously defined, from a word embedding.
+First they identify the bias subspace using Principal Component Analysis (PCA) which is a technique for dimensionality reduction.
+It is applied here to choose the subspace that contains the greatest amount of information.
+There can be many subspaces found in a given group.
+For example in terms of religion one can identify at least a few sets that are to grasp the concept of religion in general:
 
@@
 
-The idea is to find a set that provides enough information to create
-from it a vector representing the concept of religion among words. This
-strategy is based on the idea that different dimensions of vectors
-contain different types of information and in some words in vector
-layers (subspaces) the information about religiousness is implicitly
-conveyed. In some cases this knowledge is useful, but in the case of
-harmful stereotypes one does not want to include the concept of religion
-in stereotypical words.
-
-After finding the bias subspace, they use it to modify the vector values
-individually so that their cosine distances towards certain words are
-changed. In the case of stereotypes the aim is to make the cosine
-distances larger so that the association between protected word and
-stereotype is smaller.
-
-In the final step they evaluate the results. The cosine distances are
-calculated again but this time using the debiased vocabulary. After
-taking the mean of all distances one final value is obtained and then it
-is compared with the average value from the beginning. If the cosine
-distances are on average greater than before then it leads the authors
-to the conclusion that improvement has been achieved. As the cosine
-distance increases, it is assumed that the association between protected
-and stereotypical words decreases.
-
-\section{Limitations of the approach}\label{limitations-of-the-approach}
+The idea is to find a set that provides enough information to create from it a vector representing the concept of religion among words.
+This strategy is based on the idea that different dimensions of vectors contain different types of information and in some words in vector layers (subspaces) the information about religiousness is implicitly conveyed.
+In some cases this knowledge is useful, but in the case of harmful stereotypes one does not want to include the concept of religion in stereotypical words.
+
+After finding the bias subspace, they use it to modify the vector values individually so that their cosine distances towards certain words are changed.
+In the case of stereotypes the aim is to make the cosine distances larger so that the association between protected word and stereotype is smaller.
+
+In the final step they evaluate the results.
+The cosine distances are calculated again but this time using the debiased vocabulary.
+After taking the mean of all distances one final value is obtained and then it is compared with the average value from the beginning.
+If the cosine distances are on average greater than before then it leads the authors to the conclusion that improvement has been achieved.
+As the cosine distance increases, it is assumed that the association between protected and stereotypical words decreases.
+
+\hypertarget{limitations-of-the-approach}{%
+\section{Limitations of the approach}\label{limitations-of-the-approach}}
 
@@
 
-The attributes are taken from different sources, there is no principled
-justification for their choice. From our analysis it will become clear
-that the list is rather uneven.
-
-There is no mention of methodology for deciding on the number of
-attributes necessary to decide a hypothesis on the given size of
-dataset. There are however some ways to estimate sample sizes needed to
-ensure that the results are meaningful if the effect is present. Our
-research will show that the numbers used are rather insufficient.
+The attributes are taken from different sources, there is no principled justification for their choice. From our analysis it will become clear that the list is rather uneven.
+
+There is no mention of methodology for deciding on the number of attributes necessary to decide a hypothesis on the given size of dataset.
+There are however some ways to estimate sample sizes needed to ensure that the results are meaningful if the effect is present.
+Our research will show that the numbers used
+are rather insufficient.
 
@@
 
-The authors use the mean of mean average cosine similarities to measure
-multi-class similarity between protected word and harmful stereotypes.
-They average the results until they obtain one final value to represent
-the mean cosine distance between protected word from a given class and
-the attributes of that class. If one takes closer look at the individual
-values that are taken for the calculations it turns out that there is a
-bunch of outliers and surprisingly dissimilar words. We approach this
-issue by providing the tables and visualizations of individual cosine
-distances to make sure that we obtain a proper insight into the data.
+The authors use the mean of mean average cosine similarities to measure multi-class similarity between protected word and harmful stereotypes.
+They average the results until they obtain one final value to represent the mean cosine distance between protected word from a given class and the attributes of that class.
+If one takes closer look at the individual values that are taken for the calculations it turns out that there is a bunch of outliers and surprisingly dissimilar words.
+We approach this issue by providing the tables and visualizations of individual cosine distances to make sure that we obtain a proper insight into the data.
 
@@
 
-With such a method the uncertainty involved is not really considered
-which makes it even more difficult to give reasonable interpretations of
-the results. We propose the use of Bayesian method to obtain some
-understanding of the influence the uncertainty has on the interpretation
-of final results.
+With such a method the uncertainty involved is not really considered which makes it even more difficult to give reasonable interpretations of the results.
+We propose the use of Bayesian method to obtain some understanding of the influence the uncertainty has on the interpretation of final results.
 
@@
 
-In the original paper, words from all three religions were compared
-against all of the stereotypes, which means that there was no
-distinction between cases in which the stereotype is associated with a
-given religion, as opposed to the situation in which it is associated
-with another one. Not all of the stereotypical words have to be
-considered as harmful for all of the religions. One should investigate
-the religions separately as some of them may have stronger harmful
-associations that others. One should also include control groups to have
-a way of comparing the stereotypical results with neutral or human-like
-words. Later in the text we will explain in details reasons for
-introducing control groups. In our analysis, we distinguish between
-stereotypes associated with a given group, stereotypes associated with
-different groups, and control groups: neutral words and stereotypes-free
-human predicates.
+In the original paper, words from all three religions were compared against all of the stereotypes, which means that there was no distinction between cases in which the stereotype is associated with a given religion, as opposed to the situation in which it is associated with another one.
+Not all of the stereotypical words have to be considered as harmful for all of the religions.
+One should investigate the religions separately as some of them may have stronger harmful associations that others.
+One should also include control groups to have a way of comparing the stereotypical results with neutral or human-like words.
+Later in the text we will explain in details reasons for introducing control groups.
+In our analysis, we distinguish between stereotypes associated with a given group, stereotypes associated with different groups, and control groups: neutral words and stereotypes-free human predicates.
 
@@
 
-Assuming for a moment that the value of multi-class cosine distance is
-correct, one may question the interpretation. Manzini et al.
-(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) summarize the
-averages of cosine distance per group (gender, race, religion). For now
-let us focus now on analyzing the values relating to religious biases.
+Assuming for a moment that the value of multi-class cosine distance is correct, one may question the interpretation.
+\protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) summarize the averages of cosine distance per group (gender, race, religion).
+For now let us focus now on analyzing the values relating to religious biases.
 Here is the relevant fragment of table:
@@
 
-MAC stand for mean average cosine similarity, although in reality the
-the table contains mean distances. What may attract attention is the
-fact that the value of cosine distance in ``Biased'' category is already
-quite high even before debiasing. High cosine distance indicates low
-cosine similarity between values. One could think that average cosine
-similarity equal to approximately 0.141 is not significant enough to
-consider it as bias. However the authors aim to mitigate ``biases'' in
-vectors with such great distance to make it even larger.
-Methodologically the question is, on what basis is this small similarity
-still considered as a proof of the presence of bias, and whether these
-small changes are meaningful. This is in general the problem of scale
-and the lack of universal intervals. In contrast, statistical intervals
-will help us decide whether a given cosine similarity is high enough to
-consider the words to be more similar than if we chose them at random.
-We will use highest posterior density intervals, in line with Bayesian
-methodology.
+MAC stand for mean average cosine similarity, although in reality the the table contains mean distances.
+What may attract attention is the fact that the value of cosine distance in ``Biased'' category is already quite high even before debiasing.
+High cosine distance indicates low cosine similarity between values.
+One could think that average cosine similarity equal to approximately 0.141 is not significant enough to consider it as bias.
+However the authors aim to mitigate ``biases'' in vectors with such great distance to make it even larger.
+Methodologically the question is, on what basis is this small similarity still considered as a proof of the presence of bias, and whether these small changes are meaningful.
+This is in general the problem of scale and the lack of universal intervals.
+In contrast, statistical intervals will help us decide whether a given cosine similarity is high enough to consider the words to be more similar than if we chose them at random.
+We will use highest posterior density intervals, in line with Bayesian methodology.
 
@@
 
-In our case, the curse of dimensionality may take place when there is an
-increase in the volume of data that results in adding extra dimensions
-to the Euclidean space. According to the article
-\href{https://analyticsindiamag.com/curse-of-dimensionality-and-what-beginners-should-do-to-overcome-it/}{Curse
-of dimensionality at analyticsindiamag.com} as the number of features
-increases, it may be harder and harder to obtain useful information from
-the data using the available algorithms. One may notice that more data
-should contribute to greater amount of information, but more information
-also means greater risk of noise and distractions in data. At the same
-time, modern solutions are often adapted to smaller dimensions and their
-results in higher ones are not intuitive, or may be prone to error.
-
-Using cosine similarity in high dimensions in word embeddings may also
-be prone to the curse of dimensionality. According to Venkat
-(\protect\hyperlink{ref-Venkat2018Curse}{2018}) there are reasons to
-consider this phenomenon when searching for word similarities in higher
-dimensions. An experiment is conducted that aims at showing how the
-similarity values and variation change as the number of dimensions
-increases. The hypothesis made in the paper states that two things will
-happen as the number of dimensions increase. First, the effort required
-to measure cosine similarity will be greater, and two, the similarity
-between data will blur out and have less variation. The authors generate
-random points with increasing number of dimensions where each dimension
-of a data point is given a value between 0 and 1. Then they pick one
-vector at random from each dimension class and calculate the cosine
-similarity between the chosen vector and the rest of the data. Then they
-check how the variation of values changes as the number of dimensions
-increases. It seems like the more dimensions there are, the smaller the
-variance and therefore it is less obvious how to interpret the resulting
-cosine similarities. Maybe the scale should be adjusted to the number of
-dimensions and variance so that it still gives us sensible information
-about data. Yet, according to some articles cosine similarity in high
-dimensions is not reliable enough as it may be the case that choosing
-words at random may result in similar values as when picking them
-consciously.
+In our case, the curse of dimensionality may take place when there is an increase in the volume of data that results in adding extra dimensions to the Euclidean space. According to the article
+\href{https://analyticsindiamag.com/curse-of-dimensionality-and-what-beginners-should-do-to-overcome-it/}{Curse of dimensionality at analyticsindiamag.com} as the number of features increases, it may be harder and harder to obtain useful information from the data using the available algorithms.
+One may notice that more data should contribute to greater amount of information, but more information also means greater risk of noise and distractions in data.
+At the same time, modern solutions are often adapted to smaller dimensions and their results in higher ones are not intuitive, or may be prone to error.
+
+Using cosine similarity in high dimensions in word embeddings may also be prone to the curse of dimensionality.
+According to \protect\hyperlink{ref-Venkat2018Curse}{Venkat} (\protect\hyperlink{ref-Venkat2018Curse}{2018}) there are reasons to consider this phenomenon when searching for word similarities in higher dimensions.
+An experiment is conducted that aims at showing how the similarity values and variation change as the number of dimensions increases.
+The hypothesis made in the paper states that two things will happen as the number of dimensions increase.
+First, the effort required to measure cosine similarity will be greater, and two, the similarity between data will blur out and have less variation.
+The authors generate random points with increasing number of dimensions where each dimension of a data point is given a value between 0 and 1.
+Then they pick one vector at random from each dimension class and calculate the cosine similarity between the chosen vector and the rest of the data.
+Then they check how the variation of values changes as the number of dimensions increases.
+It seems like the more dimensions there are, the smaller the variance and therefore it is less obvious how to interpret the resulting cosine similarities.
+Maybe the scale should be adjusted to the number of dimensions and variance so that it still gives us sensible information about data.
+Yet, according to some articles cosine similarity in high dimensions is not reliable enough as it may be the case that choosing words at random may result in similar values as when picking them consciously.
 
@@
 \begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{(ggplot2)}
-\KeywordTok{library}\NormalTok{(ggthemes)}
-\KeywordTok{library}\NormalTok{(rethinking)}
-\KeywordTok{library}\NormalTok{(tidyverse)}
-\KeywordTok{library}\NormalTok{(ggpubr)}
-\KeywordTok{library}\NormalTok{(kableExtra)}
-\KeywordTok{library}\NormalTok{(dplyr)}
-\KeywordTok{library}\NormalTok{(ggExtra)}
-\KeywordTok{library}\NormalTok{(cowplot)}
+\FunctionTok{library}\NormalTok{(ggplot2)}
+\FunctionTok{library}\NormalTok{(ggthemes)}
+\FunctionTok{library}\NormalTok{(rethinking)}
+\FunctionTok{library}\NormalTok{(tidyverse)}
+\FunctionTok{library}\NormalTok{(ggpubr)}
+\FunctionTok{library}\NormalTok{(kableExtra)}
+\FunctionTok{library}\NormalTok{(dplyr)}
+\FunctionTok{library}\NormalTok{(ggExtra)}
+\FunctionTok{library}\NormalTok{(cowplot)}
 \end{Highlighting}
@@
 
-We will use the choice of protected words and stereotypical predicates
-used in Manzini et al.
-(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}). This is a
-decent point of departure, as not only we want to compare our method to
-that of Manzini et al.
-(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}), but also
-because this data format is fairly general (as contrasted, say, with a
-set up for binary stereotypes). Note also that the method we develop
-here can fairly easily be run for different stereotypization patterns.
-Let's start with explaining the method and its deployment using a
-dataset obtained for the religion-related protected words.
-
-Let's load, clean a bit and inspect the head of the religion dataset we
-prepared. In order to obtain this dataset, we calculated the cosine
-distance between each protected word and each word from both the
-bias-related attribute groups, which were used in the original study,
-and to neutral and human control attributes which we added as control
-groups. For instance, for religion, the bias-related predicates (coming
-from the original study in Manzini et al.
-(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019})) include
-muslim bias attributes, jew bias attributes, christian bias attributes
-(see a list in the Appendix).
-
-We decided to add control groups in the form of two classes --- neutral
-words and human-related words. Without a proper control group it is
-quite hard to compare the resulting cosine distances and decide on their
-significance in bias detection. We prepared approximately 230 neutral
-words to double-check the prima-facie neutral hypothesis that their
-cosine similarity to the protected words will oscillate around 0 (that
-is, the distances will be around 1). This provides us with a more
-reliable point of reference. Moreover, we added human attributes that
-are associated with people in general to investigate whether the smaller
-cosine distance between protected words and stereotypes can result
-simply from the fact that the stereotype predicates are associated with
-humans. For two control groups, we have randomly drawn 230 words that do
-not express any property usually attributed to humans, and human related
-attributes.
-
-\pagebreak  \vspace{1mm} \footnotesize
+We will use the choice of protected words and stereotypical predicates used in \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}).
+This is a decent point of departure, as not only we want to compare our method to that of \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}), but also because this data format is fairly general (as contrasted, say, with a set up for binary stereotypes).
+Note also that the method we develop here can fairly easily be run for different stereotypization patterns.
+Let's start with explaining the method and its deployment using a dataset obtained for the religion-related protected words.
+
+Let's load, clean a bit and inspect the head of the religion dataset we prepared.
+In order to obtain this dataset, we calculated the cosine distance between each protected word and each word from both the bias-related attribute groups, which were used in the original study, and to neutral and human control attributes which we added as control groups.
+For instance, for religion, the bias-related predicates (coming from the original study in \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019})) include muslim bias attributes, jew bias attributes, christian bias attributes (see a list in the Appendix).
+
+We decided to add control groups in the form of two classes --- neutral words and human-related words.
+Without a proper control group it is quite hard to compare the resulting cosine distances and decide on their significance in bias detection.
+We prepared approximately 230 neutral words to double-check the prima-facie neutral hypothesis that their cosine similarity to the protected words will oscillate around 0 (that is, the distances will be around 1).
+This provides us with a more reliable point of reference.
+Moreover, we added human attributes that are associated with people in general to investigate whether the smaller cosine distance between protected words and stereotypes can result simply from the fact that the stereotype predicates are associated with humans.
+For two control groups, we have randomly drawn 230 words that do not express any property usually attributed to humans, and human related attributes.
+
+\pagebreak 
+\vspace{1mm}
+\footnotesize
 
@@
 \end{table}
-
-\normalsize
-
-The \texttt{protectedWord} column contains words from a protected class
-that (in a perfect world according to the assumptions of the orignal
-study) should not be associated with harmful stereotypes.
-\texttt{wordToCompare} contains attributes, including stereotypes and
-control group words. For each row we compute the cosine distances
-between a given protected word and a given attribute word.
-\texttt{wordClass} tells us which class an attribute is supposed to be
-stereotypically associated with, that is, whether the word from
-\texttt{wordToCompare} is associated stereotypically with jews,
-christians or muslims, or whether it belongs to a control group.
-\texttt{cosineDistance} is simply a calculation of the cosine distance
-between protected word and atrribute. \texttt{cosineSimilarity} contains
-the result of substracting cosine distance from 1. \texttt{connection}
-contains information about the relation type between a protected word
-and an attribute. If the attribute is e.g.~a harmful jewish stereotype
-and the protected word is also from the judaism group, the connection
-has value \texttt{associated}. If the attribute is still stereotypically
-jewish, but the protected word comes from another religion, the
-connection is labelled as \texttt{different}. If the attribute belongs
-to a neutral group then the connection is labelled as \texttt{none} and
-if an attribute belongs to the \texttt{human} class, then the connection
-is labelled as \texttt{human}.
-
-\section{First look at the empirical
-distributions}\label{first-look-at-the-empirical-distributions}
-
-First let's take a look at the empirical distribution of distances by
-the connection type, initially ignoring the human control class for now.
-
-\vspace{1mm} \footnotesize
+\normalsize
+
+The \texttt{protectedWord} column contains words from a protected class that (in a perfect world according to the assumptions of the orignal study) should not be associated with harmful stereotypes.
+\texttt{wordToCompare} contains attributes, including stereotypes and control group words.
+For each row we compute the cosine distances between a given protected word and a given attribute word.
+\texttt{wordClass} tells us which class an attribute is supposed to be stereotypically associated with, that is, whether the word from \texttt{wordToCompare} is associated stereotypically with jews, christians or muslims, or whether it belongs to a control group.
+\texttt{cosineDistance} is simply a calculation of the cosine distance between protected word and atrribute.
+\texttt{cosineSimilarity} contains the result of substracting cosine distance from 1.
+\texttt{connection} contains information about the relation type between a protected word and an attribute.
+If the attribute is e.g.~a harmful jewish stereotype and the protected word is also from the judaism group, the connection has value \texttt{associated}.
+If the attribute is still stereotypically jewish, but the protected word comes from another religion, the connection is labelled as \texttt{different}.
+If the attribute belongs to a neutral group then the connection is labelled as \texttt{none} and if an attribute belongs to the \texttt{human} class, then the connection is labelled as \texttt{human}.
+
+\hypertarget{first-look-at-the-empirical-distributions}{%
+\section{First look at the empirical distributions}\label{first-look-at-the-empirical-distributions}}
+
+First let's take a look at the empirical distribution of distances by the connection type, initially ignoring the human control class for now.
+
+\vspace{1mm}
+\footnotesize
 
 \begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-3-1} \end{center}
-
-\normalsize
-
-The first impression is that while there is a shift for associated words
-towards smaller cosine distances as compared to the neutral words,
-slightly surprisingly a slightly weaker shift in the same direction is
-visible for attributes associated with different stereotypes. Moreover,
-the empirical distributions overlap to a large extent and the means
-grouped by connection type do not seem too far from each other. In fact,
-as there is a lot of variety in the cosine distances (as we will soon
-see), we need to gauge the uncertainty involved, and to look more
-carefully at individual protected words to get a better idea of how the
-cosine distance distribution changes for different attribute groups and
-different protected classes. Now, let's add the human attributes to the
-picture:
-
-\vspace{1mm} \footnotesize
+\normalsize
+
+The first impression is that while there is a shift for associated words towards smaller cosine distances as compared to the neutral words, slightly surprisingly a slightly weaker shift in the same direction is visible for attributes associated with different stereotypes.
+Moreover, the empirical distributions overlap to a large extent and the means grouped by connection type do not seem too far from each other.
+In fact, as there is a lot of variety in the cosine distances (as we will soon see), we need to gauge the uncertainty involved, and to look more carefully at individual protected words to get a better idea of how the cosine distance distribution changes for different attribute groups and different protected classes.
+Now, let's add the human attributes to the picture:
+
+\vspace{1mm}
+\footnotesize
 
 \begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-4-1} \end{center}
-
-\normalsize
-
-\noindent Notice that the distribution for \texttt{human} (even though
-we did our best not to include in it any stereotype-related atributes)
-is left-skewed, with much overlap with \texttt{associated} and
-\texttt{different}, which illustrates the need to take being associated
-with humans as an important predictor.
-
-Our focus lies in \texttt{connection} as a predictor. Morever, later on
-we'll be interested in looking at the protected words separately, and at
-protected words split by connection. For technical reasons it is useful
-to represent these factors as integer vectors.
-
-\vspace{1mm} \footnotesize
+\normalsize
+
+\noindent Notice that the distribution for \texttt{human} (even though we did our best not to include in it any stereotype-related atributes) is left-skewed, with much overlap with \texttt{associated} and \texttt{different}, which illustrates the need to take being associated with humans as an important predictor.
+
+Our focus lies in \texttt{connection} as a predictor.
+Morever, later on we'll be interested in looking at the protected words separately, and at protected words split by connection.
+For technical reasons it is useful to represent these factors as integer vectors.
+
+\vspace{1mm}
+\footnotesize
 
@@
 \begin{Highlighting}[]
-\NormalTok{religion}\OperatorTok{$}\NormalTok{con <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(religion}\OperatorTok{$}\NormalTok{connection)}
-\NormalTok{religion}\OperatorTok{$}\NormalTok{pw <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(religion}\OperatorTok{$}\NormalTok{protectedWord)}
-\NormalTok{religion}\OperatorTok{$}\NormalTok{pwFactor <-}\StringTok{ }\KeywordTok{factor}\NormalTok{(}\KeywordTok{paste0}\NormalTok{(religion}\OperatorTok{$}\NormalTok{protectedWord, religion}\OperatorTok{$}\NormalTok{connection))}
-\NormalTok{religion}\OperatorTok{$}\NormalTok{pwIndex <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(religion}\OperatorTok{$}\NormalTok{pwFactor)}
+\NormalTok{religion}\SpecialCharTok{$}\NormalTok{con }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{connection)}
 \end{Highlighting}
@@
 
-\normalsize
-
-A short script, \texttt{cleanDataset} to make this faster, so
-equivalently:
-
-\vspace{1mm} \footnotesize
+\begin{verbatim}
+## Warning: NAs introduced by coercion
+\end{verbatim}
 
@@
 \begin{Highlighting}[]
-\KeywordTok{source}\NormalTok{(}\StringTok{"../functions/cleanDataset.R"}\NormalTok{)}
-\NormalTok{religion <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
-\NormalTok{religion <-}\StringTok{ }\KeywordTok{cleanDataset}\NormalTok{(religion,}\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{))}
+\NormalTok{religion}\SpecialCharTok{$}\NormalTok{pw }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{protectedWord)}
 \end{Highlighting}
@@
 
-\normalsize
-
-\section{Looking at the islam-related
-words}\label{looking-at-the-islam-related-words}
-
-For now, let's focus on five protected words related to islam (``imam'',
-``islam'', ``mosque'', ``muslim'', and ``quran''). The word list
-associates with islam four stereotypical attributes (``violent'',
-``terrorist'', ``uneducated'' and ``dirty''). First, we select and plot
-the empirical distributions for these protected words.
-
-\vspace{1mm} \footnotesize
+\begin{verbatim}
+## Warning: NAs introduced by coercion
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{religion}\SpecialCharTok{$}\NormalTok{pwFactor }\OtherTok{\textless{}{-}} \FunctionTok{factor}\NormalTok{(}\FunctionTok{paste0}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{protectedWord, religion}\SpecialCharTok{$}\NormalTok{connection))}
+\NormalTok{religion}\SpecialCharTok{$}\NormalTok{pwIndex }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{pwFactor)}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+A short script, \texttt{cleanDataset} to make this faster, so equivalently:
+
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{source}\NormalTok{(}\StringTok{"../functions/cleanDataset.R"}\NormalTok{)}
+\NormalTok{religion }\OtherTok{\textless{}{-}} \FunctionTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{]}
+\NormalTok{religion }\OtherTok{\textless{}{-}} \FunctionTok{cleanDataset}\NormalTok{(religion,}\FunctionTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Warning in cleanDataset(religion, c("christian", "human", "jewish", "muslim", :
+## NAs introduced by coercion
+
+## Warning in cleanDataset(religion, c("christian", "human", "jewish", "muslim", :
+## NAs introduced by coercion
+\end{verbatim}
+
+\normalsize
+
+\hypertarget{looking-at-the-islam-related-words}{%
+\section{Looking at the islam-related words}\label{looking-at-the-islam-related-words}}
+
+For now, let's focus on five protected words related to islam (``imam,'' ``islam,'' ``mosque,'' ``muslim,'' and ``quran'').
+The word list associates with islam four stereotypical attributes (``violent,'' ``terrorist,'' ``uneducated'' and ``dirty'').
+First, we select and plot the empirical distributions for these protected words.
+
+\vspace{1mm}
+\footnotesize
 
 \begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-7-1} \end{center}
-
-\normalsize
-
-\noindent Once we focus on words related to islam, the associated bias
-seems to be stronger than in the whole dataset. This is a step towards
-illustrating that the distribution of bias is uneven.
-
-Now, say we want to look at a single protected word. Since the dataset
-also contains comparison multiple control neutral and human attributes,
-we randomly select only 5 from \texttt{none} and 5 from \texttt{human}
-control groups of those for the visualisation purposes.
-
-\vspace{1mm} \footnotesize
+\normalsize
+
+\noindent Once we focus on words related to islam, the associated bias seems to be stronger than in the whole dataset. This is a step towards illustrating that the distribution of bias is uneven.
+
+Now, say we want to look at a single protected word.
+Since the dataset also contains comparison multiple control neutral and human attributes, we randomly select only 5 from \texttt{none} and 5 from \texttt{human} control groups of those for the visualisation purposes.
+
+\vspace{1mm}
+\footnotesize
 
 \begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/tableMuslimActive-1} \end{center}
-
-\normalsize
-
-Note that the distance between the grey point and the other points is
-proportional to cosine distance, the non-grey point size is proportional
-to cosine similarity to the protected word, and color groups by the
-connection type. So for \texttt{muslim} it seems that the stereotypes
-coming from the word list are fairly well visible. To give you some
-taste of how uneven the dataset is, compare this to what happens with
-\texttt{priest}.
-
-\vspace{1mm} \footnotesize
+\normalsize
+
+Note that the distance between the grey point and the other points is proportional to cosine distance, the non-grey point size is proportional to cosine similarity to the protected word, and color groups by the connection type.
+So for \texttt{muslim} it seems that the stereotypes coming from the word list are fairly well visible.
+To give you some taste of how uneven the dataset is, compare this to what happens with \texttt{priest}.
+
+\vspace{1mm}
+\footnotesize
 
 \begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-8-1} \end{center}
-
-\normalsize
-
-\noindent Here you can see that some human attributes are closer than
-stereotype attributes, and that there is no clear reason to claim that
-\texttt{associated} attributes are closer than \texttt{different} or
-\texttt{human} attributes. This, again, illustrates the need of
-case-by-case analysis with control groups.
-
-The general idea now is that the word lists provided in different pieces
-of research are just samples of attributes associates with various
-stereotypes and should be treated as such: the uncertainty involved and
-the sample sizes should have clear impact on our estimates.
-
-\section{Bayesian model structure and
-assumptions}\label{bayesian-model-structure-and-assumptions}
-
-We will now think of cosine distance as the output variable, and will
-build a few Bayesian models to compare. First, we just build a baseline
-model which estimates cosine distance to the attributes separately for
-each protected word. The underlying idea is that different protected
-words might in general have different relations to all the attributes
-and these relations should be our point of departure.\footnote{The
-  construction of the Bayesian models and code for visualisations is due
-  to Rafal Urbaniak.}
+\normalsize
+
+\noindent Here you can see that some human attributes are closer than stereotype attributes, and that there is no clear reason to claim that \texttt{associated} attributes are closer than \texttt{different} or \texttt{human} attributes.
+This, again, illustrates the need of case-by-case analysis with control groups.
+
+The general idea now is that the word lists provided in different pieces of research are just samples of attributes associates with various stereotypes and should be treated as such: the uncertainty involved and the sample sizes should have clear impact on our estimates.
+
+\hypertarget{bayesian-model-structure-and-assumptions}{%
+\section{Bayesian model structure and assumptions}\label{bayesian-model-structure-and-assumptions}}
+
+We will now think of cosine distance as the output variable, and
+will build a few Bayesian models to compare.
+First, we just build a baseline model which estimates cosine distance to the attributes separately for each protected word.
+The underlying idea is that different protected words might in general have different relations to all the attributes and these relations should be our point of departure.\footnote{The construction of the Bayesian models and code for visualisations is due to Rafal Urbaniak.}
 
 Here is the intuition behind the mathematical Bayesian model involved.
-Our outcome variable is \texttt{cosine\ difference}, which we take to me
-normally distributed around the predicted mean for a given protected
-word (that is, we assume the residuals are normally distributed). The
-simplest model specification is:
+Our outcome variable is \texttt{cosine\ difference}, which we take to me normally distributed around the predicted mean for a given protected word (that is, we assume the residuals are normally distributed).
+The simplest model specification is:
 
@@
 
-\section{Choosing predictors}\label{choosing-predictors}
-
-Now we can define and compile the baseline model. Its parameters will
-have a posterior distribution obtained using either Hamiltionian Monte
-Carlo methods (STAN) available through the \texttt{rethinking} package.
-
-\vspace{1mm} \footnotesize
+\hypertarget{choosing-predictors}{%
+\section{Choosing predictors}\label{choosing-predictors}}
+
+Now we can define and compile the baseline model.
+Its parameters will have a posterior distribution obtained using either Hamiltionian Monte Carlo methods (STAN) available through the \texttt{rethinking} package.
+
+\vspace{1mm}
+\footnotesize
 
@@
 \begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{(rethinking)}
-\KeywordTok{options}\NormalTok{(}\DataTypeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\OtherTok{TRUE}\NormalTok{ )}
-\NormalTok{religionBaseline <-}\StringTok{ }\KeywordTok{ulam}\NormalTok{(}
-  \KeywordTok{alist}\NormalTok{(}
-\NormalTok{    cosineDistance }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(mu,sigma),}
-\NormalTok{    mu <-}\StringTok{ }\NormalTok{m[pw],}
-\NormalTok{    m[pw] }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
-\NormalTok{    sigma }\OperatorTok{~}\StringTok{ }\KeywordTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\FunctionTok{library}\NormalTok{(rethinking)}
+\FunctionTok{options}\NormalTok{(}\AttributeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\ConstantTok{TRUE}\NormalTok{ )}
+\NormalTok{religionBaseline }\OtherTok{\textless{}{-}} \FunctionTok{ulam}\NormalTok{(}
+  \FunctionTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu }\OtherTok{\textless{}{-}}\NormalTok{ m[pw],}
+\NormalTok{    m[pw] }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
 \NormalTok{  ),}
-  \DataTypeTok{data =}\NormalTok{ religion,}
-  \DataTypeTok{chains=}\DecValTok{2}\NormalTok{ , }\DataTypeTok{iter=}\DecValTok{4000}\NormalTok{ , }\DataTypeTok{warmup=}\DecValTok{1000}\NormalTok{,}
-  \DataTypeTok{start=} \KeywordTok{list}\NormalTok{(}\DataTypeTok{mu =} \DecValTok{1}\NormalTok{, }\DataTypeTok{co =} \DecValTok{0}\NormalTok{, }\DataTypeTok{sigma=} \FloatTok{.3}\NormalTok{),}
-  \DataTypeTok{log_lik =} \OtherTok{TRUE}\NormalTok{, }\DataTypeTok{cores=}\DecValTok{4}
+  \AttributeTok{data =}\NormalTok{ religion,}
+  \AttributeTok{chains=}\DecValTok{2}\NormalTok{ , }\AttributeTok{iter=}\DecValTok{4000}\NormalTok{ , }\AttributeTok{warmup=}\DecValTok{1000}\NormalTok{,}
+  \AttributeTok{start=} \FunctionTok{list}\NormalTok{(}\AttributeTok{mu =} \DecValTok{1}\NormalTok{, }\AttributeTok{co =} \DecValTok{0}\NormalTok{, }\AttributeTok{sigma=}\NormalTok{ .}\DecValTok{3}\NormalTok{),}
+  \AttributeTok{log\_lik =} \ConstantTok{TRUE}\NormalTok{, }\AttributeTok{cores=}\DecValTok{4}
 \NormalTok{)}
-\CommentTok{#saving}
-\CommentTok{#saveRDS(religionBaseline, }
-\CommentTok{#file = "cosineAnalysis/models/religionBaseline.rds")}
+\CommentTok{\#saving}
+\CommentTok{\#saveRDS(religionBaseline, }
+\CommentTok{\#file = "cosineAnalysis/models/religionBaseline.rds")}
 \end{Highlighting}
@@
 \normalsize
-The only reason we need it is the evaluation of connection as a
-predictor. Does including it in o the model improve the situation? To
-investigate this, let's now build a model according to the following
-specification:
+
+The only reason we need it is the evaluation of connection as a predictor.
+Does including it in o the model improve the situation?
+To investigate this, let's now build a model according to the following specification:
 
@@
 
-\noindent The idea now is that each connection type comes with its own
-coefficient \(co\) that has impact on mean distances for protected words
-taken separately.
-
-\vspace{1mm} \footnotesize
+\noindent The idea now is that each connection type comes with its own coefficient \(co\) that has impact on mean distances for protected words taken separately.
+
+\vspace{1mm}
+\footnotesize
 
@@
 \begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{(rethinking)}
-\KeywordTok{options}\NormalTok{(}\DataTypeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\OtherTok{TRUE}\NormalTok{ )}
-\NormalTok{religionCoefs <-}\StringTok{ }\KeywordTok{ulam}\NormalTok{(}
-  \KeywordTok{alist}\NormalTok{(}
-\NormalTok{    cosineDistance }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(mu,sigma),}
-\NormalTok{    mu <-}\StringTok{ }\NormalTok{m[pw] }\OperatorTok{+}\StringTok{ }\NormalTok{co[con],}
-\NormalTok{    m[pw] }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
-\NormalTok{    co[con] }\OperatorTok{~}\KeywordTok{dnorm}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
-\NormalTok{    sigma }\OperatorTok{~}\StringTok{ }\KeywordTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\FunctionTok{library}\NormalTok{(rethinking)}
+\FunctionTok{options}\NormalTok{(}\AttributeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\ConstantTok{TRUE}\NormalTok{ )}
+\NormalTok{religionCoefs }\OtherTok{\textless{}{-}} \FunctionTok{ulam}\NormalTok{(}
+  \FunctionTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu }\OtherTok{\textless{}{-}}\NormalTok{ m[pw] }\SpecialCharTok{+}\NormalTok{ co[con],}
+\NormalTok{    m[pw] }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    co[con] }\SpecialCharTok{\textasciitilde{}}\FunctionTok{dnorm}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
 \NormalTok{  ),}
-  \DataTypeTok{data =}\NormalTok{ religion,}
-  \DataTypeTok{chains=}\DecValTok{2}\NormalTok{ , }\DataTypeTok{iter=}\DecValTok{8000}\NormalTok{ , }\DataTypeTok{warmup=}\DecValTok{1000}\NormalTok{, }
-  \DataTypeTok{log_lik =} \OtherTok{TRUE}
+  \AttributeTok{data =}\NormalTok{ religion,}
+  \AttributeTok{chains=}\DecValTok{2}\NormalTok{ , }\AttributeTok{iter=}\DecValTok{8000}\NormalTok{ , }\AttributeTok{warmup=}\DecValTok{1000}\NormalTok{, }
+  \AttributeTok{log\_lik =} \ConstantTok{TRUE}
 \NormalTok{)}
@@
 
-\section{\texorpdfstring{Dataset-level coefficients
-\label{sec:datasetsLevelCoeffs}}{Dataset-level coefficients }}\label{dataset-level-coefficients}
-
-\noindent Let's plot them together with their highest posterior density
-invervals, for the three topic groups.
+\hypertarget{dataset-level-coefficients}{%
+\section{\texorpdfstring{Dataset-level coefficients \label{sec:datasetsLevelCoeffs}}{Dataset-level coefficients }}\label{dataset-level-coefficients}}
+
+\noindent Let's plot them together with their highest posterior density invervals, for the three topic groups.
 
@@
 
-\noindent What should strike us is that while the mean estimates of the
-coefficients indeed do differ a bit, usually the highest posterior
-density invervals all include zero, and so we do not have strong reasons
-to say that, say, as far as the whole religion dataset is involved,
-being associated indeed is connected with lower cosine distance. A
-second striking observation is that the estimated impact for associated
-stereotypes is quite often not too different from the estimated impact
-of attributes associated with different stereotypes, and both are
-sometimes not too far from the estimated impact for simply human
-attributes. In general, once the uncertainty involved is taken seriously
-by using control groups and statistical uncertainty estimation that does
-not dispose of pointwise data, the picture which focuses only on
-differences between means of means is too simplistic.
-
-But this doesn't mean important differences for some protected words are
-not there. For one thing, if you start with a word list that is very
-uneven, the actually not so bad status of some of the protected words
-might mask a pretty bad situation in which some other protected words
-are. For comparison, let's see what a model focused on words related to
-islam tells us.
-
-\vspace{1mm} \footnotesize
+\noindent What should strike us is that while the mean estimates of the coefficients indeed do differ a bit, usually the highest posterior density invervals all include zero, and so we do not have strong reasons to say that, say, as far as the whole religion dataset is involved, being associated indeed is connected with lower cosine distance.
+A second striking observation is that the estimated impact for associated stereotypes is quite often not too different from the estimated impact of attributes associated with different stereotypes, and both are sometimes not too far from the estimated impact for simply human attributes.
+In general, once the uncertainty involved is taken seriously by using control groups and statistical uncertainty estimation that does not dispose of pointwise data, the picture which focuses only on differences between means of means is too simplistic.
+
+But this doesn't mean important differences for some protected words are not there.
+For one thing, if you start with a word list that is very uneven, the actually not so bad status of some of the protected words might mask a pretty bad situation in which some other protected words are.
+For comparison, let's see what a model focused on words related to islam tells us.
+
+\vspace{1mm}
+\footnotesize
 
@@
 \begin{Highlighting}[]
-\CommentTok{#this is how we build the model}
-\NormalTok{religion <-}\StringTok{ }\KeywordTok{read.csv}\NormalTok{(}\StringTok{"cosineAnalysis/datasets/religionReddit.csv"}\NormalTok{)[}\OperatorTok{-}\DecValTok{1}\NormalTok{]}
-\KeywordTok{colnames}\NormalTok{(religion) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
+\CommentTok{\#this is how we build the model}
+\NormalTok{religion }\OtherTok{\textless{}{-}} \FunctionTok{read.csv}\NormalTok{(}\StringTok{"cosineAnalysis/datasets/religionReddit.csv"}\NormalTok{)[}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{]}
+\FunctionTok{colnames}\NormalTok{(religion) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
                         \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
-\KeywordTok{levels}\NormalTok{(religion}\OperatorTok{$}\NormalTok{wordClass) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
-\NormalTok{muslimWords <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"imam"}\NormalTok{,}\StringTok{"islam"}\NormalTok{,}\StringTok{"mosque"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"quran"}\NormalTok{)}
-\NormalTok{muslim <-}\StringTok{ }\NormalTok{religion }\OperatorTok{%>%}\StringTok{ }\KeywordTok{filter}\NormalTok{(protectedWord }\OperatorTok{%in%}\StringTok{ }\NormalTok{muslimWords)}
-\NormalTok{muslim}\OperatorTok{$}\NormalTok{protectedWord <-}\StringTok{ }\KeywordTok{droplevels}\NormalTok{(muslim}\OperatorTok{$}\NormalTok{protectedWord)}
-\NormalTok{muslim}\OperatorTok{$}\NormalTok{pw <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(muslim}\OperatorTok{$}\NormalTok{protectedWord)}
-\NormalTok{muslim}\OperatorTok{$}\NormalTok{con <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(muslim}\OperatorTok{$}\NormalTok{connection)}
-\NormalTok{muslim}\OperatorTok{$}\NormalTok{pwFactor <-}\StringTok{ }\KeywordTok{factor}\NormalTok{(}\KeywordTok{paste0}\NormalTok{(muslim}\OperatorTok{$}\NormalTok{protectedWord, muslim}\OperatorTok{$}\NormalTok{connection))}
-\NormalTok{muslim}\OperatorTok{$}\NormalTok{pwIndex <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(muslim}\OperatorTok{$}\NormalTok{pwFactor)}
-
-\NormalTok{islamCoefs <-}\StringTok{ }\KeywordTok{ulam}\NormalTok{(}
-  \KeywordTok{alist}\NormalTok{(}
-\NormalTok{    cosineDistance }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(mu,sigma),}
-\NormalTok{    mu <-}\StringTok{ }\NormalTok{m[pw] }\OperatorTok{+}\StringTok{ }\NormalTok{co[con],}
-\NormalTok{    m[pw] }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
-\NormalTok{    co[con] }\OperatorTok{~}\KeywordTok{dnorm}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
-\NormalTok{    sigma }\OperatorTok{~}\StringTok{ }\KeywordTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\FunctionTok{levels}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{wordClass) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
+\NormalTok{muslimWords }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"imam"}\NormalTok{,}\StringTok{"islam"}\NormalTok{,}\StringTok{"mosque"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"quran"}\NormalTok{)}
+\NormalTok{muslim }\OtherTok{\textless{}{-}}\NormalTok{ religion }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(protectedWord }\SpecialCharTok{\%in\%}\NormalTok{ muslimWords)}
+\NormalTok{muslim}\SpecialCharTok{$}\NormalTok{protectedWord }\OtherTok{\textless{}{-}} \FunctionTok{droplevels}\NormalTok{(muslim}\SpecialCharTok{$}\NormalTok{protectedWord)}
+\NormalTok{muslim}\SpecialCharTok{$}\NormalTok{pw }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(muslim}\SpecialCharTok{$}\NormalTok{protectedWord)}
+\NormalTok{muslim}\SpecialCharTok{$}\NormalTok{con }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(muslim}\SpecialCharTok{$}\NormalTok{connection)}
+\NormalTok{muslim}\SpecialCharTok{$}\NormalTok{pwFactor }\OtherTok{\textless{}{-}} \FunctionTok{factor}\NormalTok{(}\FunctionTok{paste0}\NormalTok{(muslim}\SpecialCharTok{$}\NormalTok{protectedWord, muslim}\SpecialCharTok{$}\NormalTok{connection))}
+\NormalTok{muslim}\SpecialCharTok{$}\NormalTok{pwIndex }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(muslim}\SpecialCharTok{$}\NormalTok{pwFactor)}
+
+\NormalTok{islamCoefs }\OtherTok{\textless{}{-}} \FunctionTok{ulam}\NormalTok{(}
+  \FunctionTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu }\OtherTok{\textless{}{-}}\NormalTok{ m[pw] }\SpecialCharTok{+}\NormalTok{ co[con],}
+\NormalTok{    m[pw] }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    co[con] }\SpecialCharTok{\textasciitilde{}}\FunctionTok{dnorm}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
 \NormalTok{  ),}
-  \DataTypeTok{data =}\NormalTok{ muslim,}
-  \DataTypeTok{chains=}\DecValTok{2}\NormalTok{ , }\DataTypeTok{iter=}\DecValTok{10000}\NormalTok{ , }\DataTypeTok{warmup=}\DecValTok{1000}\NormalTok{, }\DataTypeTok{cores =} \DecValTok{4}\NormalTok{,}
-  \DataTypeTok{log_lik =} \OtherTok{TRUE}
+  \AttributeTok{data =}\NormalTok{ muslim,}
+  \AttributeTok{chains=}\DecValTok{2}\NormalTok{ , }\AttributeTok{iter=}\DecValTok{10000}\NormalTok{ , }\AttributeTok{warmup=}\DecValTok{1000}\NormalTok{, }\AttributeTok{cores =} \DecValTok{4}\NormalTok{,}
+  \AttributeTok{log\_lik =} \ConstantTok{TRUE}
 \NormalTok{)}
@@
 
-While muslim words were unusual in the sense that the disparity between
-associated attributes and others is stronger, the evidence is still not
-conclusive. This is because the variation even within islam-related
-words is large enough (and sample sizes sufficiently small) for all the
-highest posterior density invervals to still include zeros.
-
-So, it seems, taking a closer look does seem to make a difference. The
-question is, what happens if we do take a close look at the level of
-protected words?
-
-\chapter{Protected-word level
-analysis}\label{protected-word-level-analysis}
-
-\section{Model structure and
-assumptions}\label{model-structure-and-assumptions}
-
-Let's turn then to data analysis that takes a look at protected words
-separately. This time, for each combination of a protected word and a
-connection status we will have a separate mean cosine distance estimate,
-each coming with its own highest posterior density interval. This means
-we will use indices that are result from all such combinations (and then
-we will split them up in the model precis to build visualisation, feel
-free to look at the \texttt{visualiseStats.R} script for details).
-
-\vspace{1mm} \footnotesize
+While muslim words were unusual in the sense that the disparity between associated attributes and others is stronger, the evidence is still not conclusive.
+This is because the variation even within islam-related words is large enough (and sample sizes sufficiently small) for all the highest posterior density invervals to still include zeros.
+
+So, it seems, taking a closer look does seem to make a difference. The question is, what happens if we do take a close look at the level of protected words?
+
+\hypertarget{protected-word-level-analysis}{%
+\chapter{Protected-word level analysis}\label{protected-word-level-analysis}}
+
+\hypertarget{model-structure-and-assumptions}{%
+\section{Model structure and assumptions}\label{model-structure-and-assumptions}}
+
+Let's turn then to data analysis that takes a look at protected words separately.
+This time, for each combination of a protected word and a connection status we will have a separate mean cosine distance estimate, each coming with its own highest posterior density interval.
+This means we will use indices that are result from all such combinations (and then we will split them up in the model precis to build visualisation, feel free to look at the \texttt{visualiseStats.R} script for details).
+
+\vspace{1mm}
+\footnotesize
 
@@
 \begin{Highlighting}[]
-\KeywordTok{options}\NormalTok{(}\DataTypeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\OtherTok{TRUE}\NormalTok{ ) }\CommentTok{#removes install pop-up request}
-\NormalTok{religion}\OperatorTok{$}\NormalTok{pwFactor <-}\StringTok{ }\KeywordTok{factor}\NormalTok{(}\KeywordTok{paste0}\NormalTok{(religion}\OperatorTok{$}\NormalTok{protectedWord, }\StringTok{"-"}\NormalTok{, religion}\OperatorTok{$}\NormalTok{connection))}
-\NormalTok{religion}\OperatorTok{$}\NormalTok{pwIndex <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(religion}\OperatorTok{$}\NormalTok{pwFactor)}
-
-\NormalTok{religionSeparate <-}\StringTok{ }\KeywordTok{ulam}\NormalTok{(}
-  \KeywordTok{alist}\NormalTok{(}
-\NormalTok{    cosineDistance }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(mu,sigma),}
-\NormalTok{    mu <-}\StringTok{ }\NormalTok{c[pwIndex],}
-\NormalTok{    c[pwIndex] }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
-\NormalTok{    sigma }\OperatorTok{~}\StringTok{ }\KeywordTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\FunctionTok{options}\NormalTok{(}\AttributeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\ConstantTok{TRUE}\NormalTok{ ) }\CommentTok{\#removes install pop{-}up request}
+\NormalTok{religion}\SpecialCharTok{$}\NormalTok{pwFactor }\OtherTok{\textless{}{-}} \FunctionTok{factor}\NormalTok{(}\FunctionTok{paste0}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{protectedWord, }\StringTok{"{-}"}\NormalTok{, religion}\SpecialCharTok{$}\NormalTok{connection))}
+\NormalTok{religion}\SpecialCharTok{$}\NormalTok{pwIndex }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{pwFactor)}
+
+\NormalTok{religionSeparate }\OtherTok{\textless{}{-}} \FunctionTok{ulam}\NormalTok{(}
+  \FunctionTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu }\OtherTok{\textless{}{-}}\NormalTok{ c[pwIndex],}
+\NormalTok{    c[pwIndex] }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
 \NormalTok{  ),}
-  \DataTypeTok{data =}\NormalTok{ religion,}
-  \DataTypeTok{chains=}\DecValTok{2}\NormalTok{ , }\DataTypeTok{iter=}\DecValTok{10000}\NormalTok{ , }\DataTypeTok{warmup=}\DecValTok{1000}\NormalTok{,}
-  \DataTypeTok{start=}\KeywordTok{list}\NormalTok{(}\DataTypeTok{no =} \DecValTok{1}\NormalTok{, }\DataTypeTok{a =} \DecValTok{0}\NormalTok{, }\DataTypeTok{d =} \DecValTok{0}\NormalTok{, }\DataTypeTok{sigma=} \FloatTok{.3}\NormalTok{), }\DataTypeTok{log_lik =}  \OtherTok{TRUE}
+  \AttributeTok{data =}\NormalTok{ religion,}
+  \AttributeTok{chains=}\DecValTok{2}\NormalTok{ , }\AttributeTok{iter=}\DecValTok{10000}\NormalTok{ , }\AttributeTok{warmup=}\DecValTok{1000}\NormalTok{,}
+  \AttributeTok{start=}\FunctionTok{list}\NormalTok{(}\AttributeTok{no =} \DecValTok{1}\NormalTok{, }\AttributeTok{a =} \DecValTok{0}\NormalTok{, }\AttributeTok{d =} \DecValTok{0}\NormalTok{, }\AttributeTok{sigma=}\NormalTok{ .}\DecValTok{3}\NormalTok{), }\AttributeTok{log\_lik =}  \ConstantTok{TRUE}
 \NormalTok{)}
@@
 
-\noindent Let's see if the individualized model does better than the
-previous models in light of WAIC which does add penalty for the number
-of parameters.
-
-\vspace{1mm} \footnotesize
+\noindent Let's see if the individualized model does better than the previous models in light of WAIC which does add penalty for the number of parameters.
+
+\vspace{1mm}
+\footnotesize
 
@@
 \begin{Highlighting}[]
-\NormalTok{compareBaselineCoefsSeparate<-}\StringTok{ }\KeywordTok{readRDS}\NormalTok{(}\StringTok{"../datasets/compareBaselineCoefsSeparate.rds"}\NormalTok{)}
+\NormalTok{compareBaselineCoefsSeparate}\OtherTok{\textless{}{-}} \FunctionTok{readRDS}\NormalTok{(}\StringTok{"../datasets/compareBaselineCoefsSeparate.rds"}\NormalTok{)}
 \NormalTok{compareBaselineCoefsSeparate}
@@
 
-We first encountered the GoogleNews word embedding model in Bolukbasi et
-al. (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}). They use it for
-their calculations as written in
-\href{https://github.com/tolga-b/debiaswe}{Tolga Bolukbasi
-code}\footnote{\url{https://github.com/tolga-b/debiaswe}}. We decided to
-compare the results obtained with the use of Reddit L2 model and the
-ones that we got by GoogleNews model. The details of the model can be
-found here \href{https://code.google.com/archive/p/word2vec/}{Google
-News model}. One of the main differences between the models is that
-Reddit word embeggings have 50 dimensions and GoogleNews word embeddings
-have 300 dimensions. As the dimension increases, the vectors can capture
-much more information, but this is not always the case. According to
-some researchers, the amount of dimensions should be chosen by taking
-into account the type of corpus and its features. In our case both
-Reddit and GoogleNews models are already trained and ready to use so we
-do not analyse further the choice of their hyperparameters.
+We first encountered the GoogleNews word embedding model in \protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}).
+They use it for their calculations as written in \href{https://github.com/tolga-b/debiaswe}{Tolga Bolukbasi code}\footnote{\url{https://github.com/tolga-b/debiaswe}}.
+We decided to compare the results obtained with the use of Reddit L2 model and the ones that we got by GoogleNews model.
+The details of the model can be found here \href{https://code.google.com/archive/p/word2vec/}{Google News model}.
+One of the main differences between the models is that Reddit word embeggings have 50 dimensions and GoogleNews word embeddings have 300 dimensions.
+As the dimension increases, the vectors can capture much more information, but this is not always the case.
+According to some researchers, the amount of dimensions should be chosen by taking into account the type of corpus and its features.
+In our case both Reddit and GoogleNews models are already trained and ready to use so we do not analyse further the choice of their hyperparameters.
 
@@
 
-We leave a discussion of these results for later, for now let us also
-inspect the impact of debiasing.
-
-\chapter{The role of debiasing}\label{the-role-of-debiasing}
-
-We also applied to the word embeddings hard debiasing method found in
-Manzini et al.
-(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}). As it was
-mentioned before, debiasing consists of two components: identifying the
-bias subspace and then removing this vector subspace from the chosen
-embeddings. The aim is to increase the cosine distance between protected
-words and the set of attributes. After debiasing the Reddit word
-embeddings we calculated the cosine distance again to see if there are
-any significant changes in terms of the similarity between protected
-words and the attributes.
-
-\section{Dataset-level coefficients after
-debiasing}\label{dataset-level-coefficients-after-debiasing}
-
-First, let's look at coefficient estimated for the whole datasets, as
-compared to their estimation prior to debiasing. One may observe very
-slight changes in the estimated coeﬀicients. It seems that in religion
-dataset the changes are the most significant as the mean moves towards
-zero which stands for no similarity between words.
+We leave a discussion of these results for later, for now let us also inspect the impact of debiasing.
+
+\hypertarget{the-role-of-debiasing}{%
+\chapter{The role of debiasing}\label{the-role-of-debiasing}}
+
+We also applied to the word embeddings hard debiasing method found in \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}).
+As it was mentioned before, debiasing consists of two components: identifying the bias subspace and then removing this vector subspace from the chosen embeddings.
+The aim is to increase the cosine distance between protected words and the set of attributes.
+After debiasing the Reddit word embeddings we calculated the cosine distance again to see if there are any significant changes in terms of the similarity between protected words and the attributes.
+
+\hypertarget{dataset-level-coefficients-after-debiasing}{%
+\section{Dataset-level coefficients after debiasing}\label{dataset-level-coefficients-after-debiasing}}
+
+First, let's look at coefficient estimated for the whole datasets, as compared to their estimation prior to debiasing.
+One may observe very slight changes in the estimated coeﬀicients.
+It seems that in religion dataset the changes are the most significant as the mean moves towards zero which stands for no similarity between words.
 
@@
 
-One may notice that in a religion data set most of the distances for
-\texttt{associated} words slightly increased. Some of the distances from
-\texttt{different} class are still smaller than for the
-\texttt{associated} class. The distances for all classes seem in most
-cases to cluster together even after debiasing. The reason may be that
-uncertainty is quite high in \texttt{associated} and \texttt{different}
-class because the number of samples is so small. Additionally in some
-cases distances from \texttt{human} class are smaller than for the
-\texttt{assosiated} class which also makes it less clear how to
-interpret the results, if the distances are really able to catch the
-bias presence.
-
-It is worth paying attention to the gender debiased dataset. The change
-is really minor, still with zero out of the HPDI range. The
-\texttt{different} attributes still have high similarity value. It seems
-that the distances for gender protected words are still clustered
-together which suggests that this method is unable to catch the complex
-bias presence. One could conclude that co-occurrence of certain terms
-does not always mean direct connection in terms of terms associations.
-The fact that some of stereotypically associated with female attributes
-still have high similarity with male protected words suggest that other
-techniques should be applied to detect and remove the bias.
-
-At the same time, there is a small improvement in race dataset. In some
-cases the distances just cluster together. In general distances for
-\texttt{assosiated} class are still higher than those for the rest of
-classes. One may assume that there reason for so small change is that
-either the metric does not indicate properly biases or that the issue is
-more complex and subtracting subspaces is not enough.
-
-\chapter{Discussion and summary}\label{discussion-and-summary}
-
-We propose the use of Bayesian methods to measure uncertainty in bias
-detection. There are a few advantages of this method. One of them is
-that including uncertainty enables one to directly observe the influence
-of sample sizes. Analyzing individual words and connection coefficients,
-one may notice how \texttt{neutral} words have smaller uncertainty
-intervals and \texttt{different} or \texttt{associated} quite the
-opposite. One of the reasons for such an outcome is that we used
-approximately 230 neutral words and only between 11-25 (the number
-varies from class to class) stereotypical attributes from Manzini et al.
-(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) article. In
-our approach, we also pay attention to the distribution and details
-regarding anomalous values. With the use of simple visualizations that
-we introduced before, we were able to indicate suspicious cosine
-distance values. Additionally, we compare in details how the cosine
-distance values and uncertainty change after the debiasing. One can
-investigate then how the individual vectors changed and if it is what
-was expected. Our analysis with the use of Bayesian method gave us new
-ideas and hypothesis concerning not only the bias detection method but
-also the efficiency of debiasing itself. We created summary tables for
-each of the datasets: Google word embeddings (Table \ref{tab:dTabG}),
-Reddit (Table \ref{tab:dTabR}), and Reddit Debiased (Table
-\ref{tab:dTabRD}).
-
-\vspace{1mm} \footnotesize
-
+One may notice that in a religion data set most of the distances for \texttt{associated} words slightly increased.
+Some of the distances from \texttt{different} class are still smaller than for the \texttt{associated} class.
+The distances for all classes seem in most cases to cluster together even after debiasing.
+The reason may be that uncertainty is quite high in \texttt{associated} and \texttt{different} class because the number of samples is so small.
+Additionally in some cases distances from \texttt{human} class are smaller than for the \texttt{assosiated} class which also makes it less clear how to interpret the results, if the distances are really able to catch the bias presence.
+
+It is worth paying attention to the gender debiased dataset.
+The change is really minor, still with zero out of the HPDI range.
+The \texttt{different} attributes still have high similarity value.
+It seems that the distances for gender protected words are still clustered together which suggests that this method is unable to catch the complex bias presence.
+One could conclude that co-occurrence of certain terms does not always mean direct connection in terms of terms associations.
+The fact that some of stereotypically associated with female attributes still have high similarity with male protected words suggest that other techniques should be applied to detect and remove the bias.
+
+At the same time, there is a small improvement in race dataset.
+In some cases the distances just cluster together.
+In general distances for \texttt{assosiated} class are still higher than those for the rest of classes.
+One may assume that there reason for so small change is that either the metric does not indicate properly biases or that the issue is more complex and subtracting subspaces is not enough.
+
+\hypertarget{discussion-and-summary}{%
+\chapter{Discussion and summary}\label{discussion-and-summary}}
+
+We propose the use of Bayesian methods to measure uncertainty in bias detection.
+There are a few advantages of this method.
+One of them is that including uncertainty enables one to directly observe the influence of sample sizes.
+Analyzing individual words and connection coefficients, one may notice how \texttt{neutral} words have smaller uncertainty intervals and \texttt{different} or \texttt{associated} quite the opposite.
+One of the reasons for such an outcome is that we used approximately 230 neutral words and only between 11-25 (the number varies from class to class) stereotypical attributes from \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) article.
+In our approach, we also pay attention to the distribution and details regarding anomalous values.
+With the use of simple visualizations that we introduced before, we were able to indicate suspicious cosine distance values.
+Additionally, we compare in details how the cosine distance values and uncertainty change after the debiasing.
+One can investigate then how the individual vectors changed and if it is what was expected.
+Our analysis with the use of Bayesian method gave us new ideas and hypothesis concerning not only the bias detection method but also the efficiency of debiasing itself.
+We created summary tables for each of the datasets: Google word embeddings (Table \ref{tab:dTabG}), Reddit (Table \ref{tab:dTabR}), and Reddit Debiased (Table \ref{tab:dTabRD}).
+
+\vspace{1mm}
+\footnotesize
 \begin{table}
@@
 \end{table}
-
-\normalsize
-
-\vspace{1mm} \footnotesize
-
+\normalsize
+
+\vspace{1mm}
+\footnotesize
 \begin{table}
@@
 \end{table}
-
-\normalsize
-
-Let us first analyse the general observations from estimated
-coefficients mean introduced in Section \ref{sec:datasetsLevelCoeffs} on
-dataset-level coefficients. For Google embeddings the HPDIs for all
-class coefficients (associated, different, human, and none) include
-zero. This can lead one to a conclusion that the impact for associated,
-different, human and neutral attribute is, when averaged, quite similar.
-This indicates how including the uncertainty may change the use and
-interpretation of Manzini et al.
-(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) MAC metric.
-It seems that if one focuses only on differences between means of means,
-this is too simplistic. In case of Reddit word embeddings the situation
-is similar although HPDI is below 0 for Gender class when looking at
-\texttt{associated} and \texttt{different} mean coefficient. This can
-suggest that there is indeed slightly stronger impact of these
-attributes. One should also notice how in general \texttt{associated}
-and \texttt{different} coefficients have quite similar HPDI range, the
-highest observed absolute difference is equal to only 0.1. This suggests
-that again the impact of associated attributes and different ones is not
-clear at first sight. Finally, let's compare the HPDIs for Reddit and
-Reddit debiased datasets. For Religion and Race dataset there is a minor
-shift (in absolute values the highest change is equal to approximately
-0.1) of the mean coefficients towards zero. However for Gender dataset
-there is no significant change. This is of course a general look at the
-group coefficients. Let's now analyze the individual words.
-
-In Reddit table one can observe that for Religion the cosine distance
-results for the associated attributes are for approximately 60\% of the
-words close to human attributes as well. This can suggest that in some
-cases words concerning humans can have higher similarity with some
-protected words independently of whether they are stereotypical or
-neutral-human words. One should also pay attention to the fact that for
-all of the \texttt{associated} and \texttt{different} attributes in
-Religion, the uncertainty intervals overlap at some point. What is even
-more surprising is that for protected words, such as ``torah'' the
-\texttt{associated} attribute has the the cosine distance slightly over
-1, which means no positive similarity! If the protected words that we
-chose do not have high similarity with harmful associated attributes,
-then one should consider at least three scenarios. The first one is that
-the choice of the protected words and attributes may be corrupted. The
-second one is that the metric is unable to catch the hidden bias
-properly. The third one is that there is actually no bias between the
-words. Regardless of which scenario one considers, it is essential to
-take a look at the individual values before averaging them or
-aggregating in other ways. It seems that using Bayesian method can
-enhance the process of verifying the hypotheses concerning the choice of
-protected words and attributes.
-
-Surprisingly in Gender one can observe high cosine similarity values
-between some female stereotypical professions and male protected words.
-If a word stereotypically associated with females has low cosine
-distance to male protected words, then one should investigate the issue
-further. The reasons for this unexpected cosine distance may lie in the
-frequency of appearance of the protected words in the raw data. Some of
-the groups (like Muslim people or females) may have less representation
-in the data that is taken as an input for word embeddings. Therefore, if
-we assume that the MAC detects actually co-occurrence only, it makes
-sense that they can have high similarity with associated attributes and
-lower similarity with different ones. At the same time male protected
-words and other religions can have high similarity with different
-attributes because they have greater representation in the dataset in
-general and occur close to much more concepts. Cosine distance seems to
-capture the information on the co-occurrence of words and not on the
-semantic similarity strictly speaking.
-
-In Google, one may notice interesting phenomena as well. Although the
-GoogleNews dataset is larger, trained on different data sources and with
-the use of more dimensions, some of the results are similar to the ones
-obtained for the Reddit corpus. For Religion almost all of the values
-for \texttt{associated} class intersect with \texttt{different} class as
-well. In the case of Race it is 100\% of the available words. This
-indicates again that it is not clear how the metric should be used.
-Quite a different situation takes place for Gender, where the similarity
-between \texttt{associated} and \texttt{different} is present mostly for
-the male protected words. This means that male words have high
-similarity with both male stereotypical professions and female ones.
-However, for females, the similarity is high mostly only for the female
-stereotypical professions. This observation could not be made when using
-MAC metric only, as it requires investigating the individual words, and
-providing uncertainty,
-
-When analyzing the Reddit debiased results one should investigate the
-change of the cosine distance values. It seems that not all of the
-protected words are treated equally when performing debiasing. In
-religion, the values for \texttt{associated} class moved towards 0 for
-most of the words except for the ones for Islam, where still the
-\texttt{associated} class has quite lower cosine distance than the
-\texttt{different} one. Similar situation takes place in Gender, where
-female words have still much lower cosine distance values for
-\texttt{associated} class than for the \texttt{different} one. In the
-case of male protected words it is mostly almost the same interval for
-\texttt{associated} and \texttt{different}. As the Gender data is quite
-specific as the attributes are not harmful adjectives but (in an ideal
-world) neutral professions, the aim (if we follow MAC assumptions)
-should actually be to make the cosine distance same for both female and
-male protected words. However, as we pointed out the situation after
-debiasing can sometimes be better only for one protected group, which is
-not the desired outcome. In the case of Religion, even after debiasing,
-all of the protected words have intersections between
-\texttt{associated} and \texttt{different} class and similarity greater
-than 0. As all of the attributes in religion data are negative and
-harmful stereotypes, one should not aim at making the distance between
-protected words and \texttt{associated}, and \texttt{different} class
-the same but rather to move it towards 1.
-
-Let's summarize the results of the bias detection methods analysis. As
-it was presented above, using the bias detection methods without
-involving the uncertainty may bring the risk of limited insight into the
-data. One should be concerned with the limitations of the metric that
-evaluates the debiasing by means of the averaged mean approach.
-Additionally, one cannot be sure if the bias is still preserved after
-debiasing. The fact that all of the cosine distances for protected words
-and harmful attributes moved slightly to the right does not mean that
-the bias is removed. Gonen \& Goldberg
-(\protect\hyperlink{ref-Gonen2019Lipstick}{2019}) argue that the bias
-can hide in the vector geometry and preserve even after applying popular
-debiasing methods. One of the reasons for that may be the fact that
-relative differences between words may be preserved even after
-debiasing. Another thing is that the choice of word lists used to verify
-the metric effectiveness is not well justified. The sample size is very
-small, which leads to large HPDIs. Some of the protected words and
-attributes originate from somewhat old articles and it seems that the
-bias phenomenon is quite dynamic and the data should be as up-to-date as
-it is possible. Moreover, recall that there is no clear interpretation
-for the values obtained with MAC metric from Manzini et al.
-(\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}). One may
-assume that if the cosine distance is close to 1 then it is a desired
-outcome as it means, according to cosine distance assumptions, that
-there is almost no similarity between the words. However what does it
-mean to be close to 0? If the averaged cosine distance is equal to 0.8,
-then should we still debias? It is unclear what the criteria are. On one
-hand, it seems to be beneficial when the outcome is simplified as it is
-easier to compare results with one value per set. On the other hand, it
-is prone to misunderstanding of how to interpret the results and what
-threshold to assume.
-
-The bottom line is that if we want to take bias seriously, so should we
-approach the uncertainty involved in our estimations. There is no
-replacement for proper statistical evaluation that does not discard
-information about the uncertainty involved, larger word lists are
-needed, and visualization of the results for particular protected
-classes provides much better guidance than chasing a single metric based
-on a means of means.
-
-\chapter*{Appendix}\label{appendix}
+\normalsize
+
+Let us first analyse the general observations from estimated coefficients mean introduced in Section \ref{sec:datasetsLevelCoeffs} on dataset-level coefficients.
+For Google embeddings the HPDIs for all class coefficients (associated, different, human, and none) include zero.
+This can lead one to a conclusion that the impact for associated, different, human and neutral attribute is, when averaged, quite similar.
+This indicates how including the uncertainty may change the use and interpretation of \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) MAC metric.
+It seems that if one focuses only on differences between means of means, this is too simplistic.
+In case of Reddit word embeddings the situation is similar although HPDI is below 0 for Gender class when looking at \texttt{associated} and \texttt{different} mean coefficient.
+This can suggest that there is indeed slightly stronger impact of these attributes.
+One should also notice how in general \texttt{associated} and \texttt{different} coefficients have quite similar HPDI range, the highest observed absolute difference is equal to only 0.1.
+This suggests that again the impact of associated attributes and different ones is not clear at first sight.
+Finally, let's compare the HPDIs for Reddit and Reddit debiased datasets.
+For Religion and Race dataset there is a minor shift (in absolute values the highest change is equal to approximately 0.1) of the mean coefficients towards zero.
+However for Gender dataset there is no significant change.
+This is of course a general look at the group coefficients.
+Let's now analyze the individual words.
+
+In Reddit table one can observe that for Religion the cosine distance results for the associated attributes are for approximately 60\% of the words close to human attributes as well.
+This can suggest that in some cases words concerning humans can have higher similarity with some protected words independently of whether they are stereotypical or neutral-human words.
+One should also pay attention to the fact that for all of the \texttt{associated} and \texttt{different} attributes in Religion, the uncertainty intervals overlap at some point.
+What is even more surprising is that for protected words, such as ``torah'' the \texttt{associated} attribute has the the cosine distance slightly over 1, which means no positive similarity!
+If the protected words that we chose do not have high similarity with harmful associated attributes, then one should consider at least three scenarios.
+The first one is that the choice of the protected words and attributes may be corrupted.
+The second one is that the metric is unable to catch the hidden bias properly.
+The third one is that there is actually no bias between the words.
+Regardless of which scenario one considers, it is essential to take a look at the individual values before averaging them or aggregating in other ways.
+It seems that using Bayesian method can enhance the process of verifying the hypotheses concerning the choice of protected words and attributes.
+
+Surprisingly in Gender one can observe high cosine similarity values between some female stereotypical professions and male protected words.
+If a word stereotypically associated with females has low cosine distance to male protected words, then one should investigate the issue further.
+The reasons for this unexpected cosine distance may lie in the frequency of appearance of the protected words in the raw data.
+Some of the groups (like Muslim people or females) may have less representation in the data that is taken as an input for word embeddings.
+Therefore, if we assume that the MAC detects actually co-occurrence only, it makes sense that they can have high similarity with associated attributes and lower similarity with different ones.
+At the same time male protected words and other religions can have high similarity with different attributes because they have greater representation in the dataset in general and occur close to much more concepts.
+Cosine distance seems to capture the information on the co-occurrence of words and not on the semantic similarity strictly speaking.
+
+In Google, one may notice interesting phenomena as well.
+Although the GoogleNews dataset is larger, trained on different data sources and with the use of more dimensions, some of the results are similar to the ones obtained for the Reddit corpus.
+For Religion almost all of the values for \texttt{associated} class intersect with \texttt{different} class as well.
+In the case of Race it is 100\% of the available words.
+This indicates again that it is not clear how the metric should be used.
+Quite a different situation takes place for Gender, where the similarity between \texttt{associated} and \texttt{different} is present mostly for the male protected words.
+This means that male words have high similarity with both male stereotypical professions and female ones.
+However, for females, the similarity is high mostly only for the female stereotypical professions.
+This observation could not be made when using MAC metric only, as it requires investigating the individual words, and providing uncertainty,
+
+When analyzing the Reddit debiased results one should investigate the change of the cosine distance values.
+It seems that not all of the protected words are treated equally when performing debiasing.
+In religion, the values for \texttt{associated} class moved towards 0 for most of the words except for the ones for Islam, where still the \texttt{associated} class has quite lower cosine distance than the \texttt{different} one.
+Similar situation takes place in Gender, where female words have still much lower cosine distance values for \texttt{associated} class than for the \texttt{different} one.
+In the case of male protected words it is mostly almost the same interval for \texttt{associated} and \texttt{different}.
+As the Gender data is quite specific as the attributes are not harmful adjectives but (in an ideal world) neutral professions, the aim (if we follow MAC assumptions) should actually be to make the cosine distance same for both female and male protected words.
+However, as we pointed out the situation after debiasing can sometimes be better only for one protected group, which is not the desired outcome.
+In the case of Religion, even after debiasing, all of the protected words have intersections between \texttt{associated} and \texttt{different} class and similarity greater than 0.
+As all of the attributes in religion data are negative and harmful stereotypes, one should not aim at making the distance between protected words and \texttt{associated}, and \texttt{different} class the same but rather to move it towards 1.
+
+Let's summarize the results of the bias detection methods analysis.
+As it was presented above, using the bias detection methods without involving the uncertainty may bring the risk of limited insight into the data.
+One should be concerned with the limitations of the metric that evaluates the debiasing by means of the averaged mean approach.
+Additionally, one cannot be sure if the bias is still preserved after debiasing.
+The fact that all of the cosine distances for protected words and harmful attributes moved slightly to the right does not mean that the bias is removed.
+\protect\hyperlink{ref-Gonen2019Lipstick}{Gonen \& Goldberg} (\protect\hyperlink{ref-Gonen2019Lipstick}{2019}) argue that the bias can hide in the vector geometry and preserve even after applying popular debiasing methods.
+One of the reasons for that may be the fact that relative differences between words may be preserved even after debiasing.
+Another thing is that the choice of word lists used to verify the metric effectiveness is not well justified.
+The sample size is very small, which leads to large HPDIs.
+Some of the protected words and attributes originate from somewhat old articles and it seems that the bias phenomenon is quite dynamic and the data should be as up-to-date as it is possible.
+Moreover, recall that there is no clear interpretation for the values obtained with MAC metric from \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}).
+One may assume that if the cosine distance is close to 1 then it is a desired outcome as it means, according to cosine distance assumptions, that there is almost no similarity between the words.
+However what does it mean to be close to 0?
+If the averaged cosine distance is equal to 0.8, then should we still debias?
+It is unclear what the criteria are.
+On one hand, it seems to be beneficial when the outcome is simplified as it is easier to compare results with one value per set.
+On the other hand, it is prone to misunderstanding of how to interpret the results and what threshold to assume.
+
+The bottom line is that if we want to take bias seriously, so should we approach the uncertainty involved in our estimations.
+There is no replacement for proper statistical evaluation that does not discard information about the uncertainty involved, larger word lists are needed, and visualization of the results for particular protected classes provides much better guidance than chasing a single metric based on a means of means.
+
+\hypertarget{appendix}{%
+\chapter*{Appendix}\label{appendix}}
 \addcontentsline{toc}{chapter}{Appendix}
 
-\section*{\texorpdfstring{Original wordlist from (Manzini et al.,
-\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019})}{Original wordlist from (Manzini et al., 2019)}}\label{original-wordlist-from-manzini2019blacktocriminal}
-\addcontentsline{toc}{section}{Original wordlist from (Manzini et al.,
-\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019})}
-
-Definite sets are the protected words, analogy templates assign
-stereotypical attributes to protected words, test terms were included
-but are not used in the original MAC calculations.
-
-\subsection*{Religion}\label{religion}
+\hypertarget{original-wordlist-from-manzini2019blacktocriminal}{%
+\section*{\texorpdfstring{Original wordlist from (\protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black, 2019})}{Original wordlist from (Manzini, Lim, Tsvetkov, \& Black, 2019)}}\label{original-wordlist-from-manzini2019blacktocriminal}}
+\addcontentsline{toc}{section}{Original wordlist from (\protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black, 2019})}
+
+Definite sets are the protected words, analogy templates assign stereotypical attributes to protected words, test terms were included but are not used in the original MAC calculations.
+
+\hypertarget{religion}{%
+\subsection*{Religion}\label{religion}}
 \addcontentsline{toc}{subsection}{Religion}
 
-\vspace{1mm} \scriptsize
+\vspace{1mm}
+\scriptsize
 
@@
 \NormalTok{\{}
-    \StringTok{"definite_sets"}\OperatorTok{:}\NormalTok{[}
+    \StringTok{"definite\_sets"}\SpecialCharTok{:}\NormalTok{[}
 \NormalTok{        [}\StringTok{"judaism"}\NormalTok{, }\StringTok{"christianity"}\NormalTok{, }\StringTok{"islam"}\NormalTok{],}
@@
 \NormalTok{    ],}
-    \StringTok{"analogy_templates"}\OperatorTok{:}
-\StringTok{    }\NormalTok{\{}
-        \StringTok{"attribute"}\OperatorTok{:}\StringTok{ }\NormalTok{\{}
-            \StringTok{"jew"}       \OperatorTok{:}\StringTok{ }\NormalTok{[}\StringTok{"greedy"}\NormalTok{, }\StringTok{"cheap"}\NormalTok{, }\StringTok{"hairy"}\NormalTok{, }\StringTok{"liberal"}\NormalTok{],}
-            \StringTok{"christian"} \OperatorTok{:}\StringTok{ }\NormalTok{[}\StringTok{"judgemental"}\NormalTok{, }\StringTok{"conservative"}\NormalTok{, }\StringTok{"familial"}\NormalTok{],}
-            \StringTok{"muslim"}    \OperatorTok{:}\StringTok{ }\NormalTok{[}\StringTok{"violent"}\NormalTok{, }\StringTok{"terrorist"}\NormalTok{, }\StringTok{"dirty"}\NormalTok{, }\StringTok{"uneducated"}\NormalTok{]}
+    \StringTok{"analogy\_templates"}\SpecialCharTok{:}
+\NormalTok{    \{}
+        \StringTok{"attribute"}\SpecialCharTok{:}\NormalTok{ \{}
+            \StringTok{"jew"}       \SpecialCharTok{:}\NormalTok{ [}\StringTok{"greedy"}\NormalTok{, }\StringTok{"cheap"}\NormalTok{, }\StringTok{"hairy"}\NormalTok{, }\StringTok{"liberal"}\NormalTok{],}
+            \StringTok{"christian"} \SpecialCharTok{:}\NormalTok{ [}\StringTok{"judgemental"}\NormalTok{, }\StringTok{"conservative"}\NormalTok{, }\StringTok{"familial"}\NormalTok{],}
+            \StringTok{"muslim"}    \SpecialCharTok{:}\NormalTok{ [}\StringTok{"violent"}\NormalTok{, }\StringTok{"terrorist"}\NormalTok{, }\StringTok{"dirty"}\NormalTok{, }\StringTok{"uneducated"}\NormalTok{]}
 \NormalTok{        \}}
 \NormalTok{    \},}
-    \StringTok{"eval_targets"}\OperatorTok{:}\NormalTok{[}
+    \StringTok{"eval\_targets"}\SpecialCharTok{:}\NormalTok{[}
 \NormalTok{        [}\StringTok{"judaism"}\NormalTok{, }\StringTok{"christianity"}\NormalTok{, }\StringTok{"islam"}\NormalTok{],}
@@
 \NormalTok{    ], }
-    \StringTok{"testTerms"}\OperatorTok{:}\NormalTok{[}\StringTok{"chair"}\NormalTok{, }\StringTok{"house"}\NormalTok{, }\StringTok{"greedy"}\NormalTok{, }\StringTok{"terrorist"}\NormalTok{, }\StringTok{"dirty"}\NormalTok{, }\StringTok{"greedy"}\NormalTok{]}
+    \StringTok{"testTerms"}\SpecialCharTok{:}\NormalTok{[}\StringTok{"chair"}\NormalTok{, }\StringTok{"house"}\NormalTok{, }\StringTok{"greedy"}\NormalTok{, }\StringTok{"terrorist"}\NormalTok{, }\StringTok{"dirty"}\NormalTok{, }\StringTok{"greedy"}\NormalTok{]}
 \NormalTok{\}}
@@
 
-\subsection*{Gender}\label{gender}
+\hypertarget{gender}{%
+\subsection*{Gender}\label{gender}}
 \addcontentsline{toc}{subsection}{Gender}
 
-\vspace{1mm} \scriptsize
+\vspace{1mm}
+\scriptsize
 
@@
 \NormalTok{\{}
-    \StringTok{"definite_sets"}\OperatorTok{:}\NormalTok{[}
+    \StringTok{"definite\_sets"}\SpecialCharTok{:}\NormalTok{[}
 \NormalTok{        [}\StringTok{"he"}\NormalTok{, }\StringTok{"she"}\NormalTok{],}
@@
 \NormalTok{    ],}
-    \StringTok{"analogy_templates"}\OperatorTok{:}
-\StringTok{    }\NormalTok{\{}
-        \StringTok{"role"}\OperatorTok{:}\StringTok{ }\NormalTok{\{}
-            \StringTok{"man"}   \OperatorTok{:}\StringTok{ }\NormalTok{[}\StringTok{"manager"}\NormalTok{, }\StringTok{"executive"}\NormalTok{, }\StringTok{"doctor"}\NormalTok{, }\StringTok{"lawyer"}\NormalTok{, }\StringTok{"programmer"}\NormalTok{, }
+    \StringTok{"analogy\_templates"}\SpecialCharTok{:}
+\NormalTok{    \{}
+        \StringTok{"role"}\SpecialCharTok{:}\NormalTok{ \{}
+            \StringTok{"man"}   \SpecialCharTok{:}\NormalTok{ [}\StringTok{"manager"}\NormalTok{, }\StringTok{"executive"}\NormalTok{, }\StringTok{"doctor"}\NormalTok{, }\StringTok{"lawyer"}\NormalTok{, }\StringTok{"programmer"}\NormalTok{, }
                        \StringTok{"scientist"}\NormalTok{, }\StringTok{"soldier"}\NormalTok{, }\StringTok{"supervisor"}\NormalTok{, }\StringTok{"rancher"}\NormalTok{, }\StringTok{"janitor"}\NormalTok{,}
                        \StringTok{"firefighter"}\NormalTok{, }\StringTok{"officer"}\NormalTok{],}
-            \StringTok{"woman"} \OperatorTok{:}\StringTok{ }\NormalTok{[}\StringTok{"secretary"}\NormalTok{, }\StringTok{"nurse"}\NormalTok{, }\StringTok{"clerk"}\NormalTok{, }\StringTok{"artist"}\NormalTok{, }\StringTok{"homemaker"}\NormalTok{, }\StringTok{"dancer"}\NormalTok{,}
+            \StringTok{"woman"} \SpecialCharTok{:}\NormalTok{ [}\StringTok{"secretary"}\NormalTok{, }\StringTok{"nurse"}\NormalTok{, }\StringTok{"clerk"}\NormalTok{, }\StringTok{"artist"}\NormalTok{, }\StringTok{"homemaker"}\NormalTok{, }\StringTok{"dancer"}\NormalTok{,}
                        \StringTok{"singer"}\NormalTok{, }\StringTok{"librarian"}\NormalTok{, }\StringTok{"maid"}\NormalTok{, }\StringTok{"hairdresser"}\NormalTok{, }\StringTok{"stylist"}\NormalTok{,}
@@
 \NormalTok{    \}, }
-    \StringTok{"eval_targets"}\OperatorTok{:}\NormalTok{[}
+    \StringTok{"eval\_targets"}\SpecialCharTok{:}\NormalTok{[}
 \NormalTok{        [}\StringTok{"he"}\NormalTok{, }\StringTok{"she"}\NormalTok{],}
@@
 \NormalTok{    ], }
-    \StringTok{"testTerms"}\OperatorTok{:}\NormalTok{[}\StringTok{"chair"}\NormalTok{, }\StringTok{"house"}\NormalTok{, }\StringTok{"supervisor"}\NormalTok{, }\StringTok{"secretary"}\NormalTok{, }\StringTok{"loud"}\NormalTok{, }\StringTok{"weak"}\NormalTok{]}
+    \StringTok{"testTerms"}\SpecialCharTok{:}\NormalTok{[}\StringTok{"chair"}\NormalTok{, }\StringTok{"house"}\NormalTok{, }\StringTok{"supervisor"}\NormalTok{, }\StringTok{"secretary"}\NormalTok{, }\StringTok{"loud"}\NormalTok{, }\StringTok{"weak"}\NormalTok{]}
 \NormalTok{\}}
@@
 
-\subsection*{Race}\label{race}
+\hypertarget{race}{%
+\subsection*{Race}\label{race}}
 \addcontentsline{toc}{subsection}{Race}
 
-\vspace{1mm} \scriptsize
+\vspace{1mm}
+\scriptsize
 
@@
 \NormalTok{\{}
-    \StringTok{"definite_sets"}\OperatorTok{:}\NormalTok{[}
+    \StringTok{"definite\_sets"}\SpecialCharTok{:}\NormalTok{[}
 \NormalTok{        [}\StringTok{"black"}\NormalTok{, }\StringTok{"caucasian"}\NormalTok{, }\StringTok{"asian"}\NormalTok{],}
@@
 \NormalTok{    ],}
-    \StringTok{"analogy_templates"}\OperatorTok{:}
-\StringTok{    }\NormalTok{\{}
-        \StringTok{"role"}\OperatorTok{:}\StringTok{ }\NormalTok{\{}
-            \StringTok{"caucasian"}\OperatorTok{:}\StringTok{ }\NormalTok{[}\StringTok{"manager"}\NormalTok{, }\StringTok{"executive"}\NormalTok{, }\StringTok{"redneck"}\NormalTok{, }\StringTok{"hillbilly"}\NormalTok{, }\StringTok{"leader"}\NormalTok{, }\StringTok{"farmer"}\NormalTok{],}
-            \StringTok{"asian"}    \OperatorTok{:}\StringTok{ }\NormalTok{[}\StringTok{"doctor"}\NormalTok{, }\StringTok{"engineer"}\NormalTok{, }\StringTok{"laborer"}\NormalTok{, }\StringTok{"teacher"}\NormalTok{],}
-            \StringTok{"black"}    \OperatorTok{:}\StringTok{ }\NormalTok{[}\StringTok{"slave"}\NormalTok{, }\StringTok{"musician"}\NormalTok{, }\StringTok{"runner"}\NormalTok{, }\StringTok{"criminal"}\NormalTok{, }\StringTok{"homeless"}\NormalTok{]}
+    \StringTok{"analogy\_templates"}\SpecialCharTok{:}
+\NormalTok{    \{}
+        \StringTok{"role"}\SpecialCharTok{:}\NormalTok{ \{}
+            \StringTok{"caucasian"}\SpecialCharTok{:}\NormalTok{ [}\StringTok{"manager"}\NormalTok{, }\StringTok{"executive"}\NormalTok{, }\StringTok{"redneck"}\NormalTok{, }\StringTok{"hillbilly"}\NormalTok{, }\StringTok{"leader"}\NormalTok{, }\StringTok{"farmer"}\NormalTok{],}
+            \StringTok{"asian"}    \SpecialCharTok{:}\NormalTok{ [}\StringTok{"doctor"}\NormalTok{, }\StringTok{"engineer"}\NormalTok{, }\StringTok{"laborer"}\NormalTok{, }\StringTok{"teacher"}\NormalTok{],}
+            \StringTok{"black"}    \SpecialCharTok{:}\NormalTok{ [}\StringTok{"slave"}\NormalTok{, }\StringTok{"musician"}\NormalTok{, }\StringTok{"runner"}\NormalTok{, }\StringTok{"criminal"}\NormalTok{, }\StringTok{"homeless"}\NormalTok{]}
 \NormalTok{        \}}
 \NormalTok{    \}, }
-    \StringTok{"eval_targets"}\OperatorTok{:}\NormalTok{[}
+    \StringTok{"eval\_targets"}\SpecialCharTok{:}\NormalTok{[}
 \NormalTok{        [}\StringTok{"black"}\NormalTok{, }\StringTok{"caucasian"}\NormalTok{, }\StringTok{"asian"}\NormalTok{],}
@@
 \NormalTok{    ], }
-    \StringTok{"testTerms"}\OperatorTok{:}\NormalTok{[}\StringTok{"chair"}\NormalTok{, }\StringTok{"house"}\NormalTok{, }\StringTok{"smart"}\NormalTok{, }\StringTok{"criminal"}\NormalTok{, }\StringTok{"executive"}\NormalTok{, }\StringTok{"farmer"}\NormalTok{]}
+    \StringTok{"testTerms"}\SpecialCharTok{:}\NormalTok{[}\StringTok{"chair"}\NormalTok{, }\StringTok{"house"}\NormalTok{, }\StringTok{"smart"}\NormalTok{, }\StringTok{"criminal"}\NormalTok{, }\StringTok{"executive"}\NormalTok{, }\StringTok{"farmer"}\NormalTok{]}
 \NormalTok{\}}
@@
 
-\section*{Our control groups}\label{our-control-groups}
+\hypertarget{our-control-groups}{%
+\section*{Our control groups}\label{our-control-groups}}
 \addcontentsline{toc}{section}{Our control groups}
 
-\section*{Human neutral attributes}\label{human-neutral-attributes}
+\hypertarget{human-neutral-attributes}{%
+\section*{Human neutral attributes}\label{human-neutral-attributes}}
 \addcontentsline{toc}{section}{Human neutral attributes}
 
-\vspace{1mm} \scriptsize
+\vspace{1mm}
+\scriptsize
 
@@
 \begin{Highlighting}[]
-\NormalTok{human_words =}\StringTok{ }\NormalTok{[}\StringTok{'wear'}\NormalTok{, }\StringTok{'walk'}\NormalTok{, }\StringTok{'visitor'}\NormalTok{, }\StringTok{'toy'}\NormalTok{, }\StringTok{'tissue'}\NormalTok{, }\StringTok{'throw'}\NormalTok{, }\StringTok{'talk'}\NormalTok{, }
-               \StringTok{'speak'}\NormalTok{, }\StringTok{'sleep'}\NormalTok{, }\StringTok{'eye'}\NormalTok{, }\StringTok{'enjoy'}\NormalTok{, }\StringTok{'blogger'}\NormalTok{, }\StringTok{'character'}\NormalTok{, }
-               \StringTok{'candidate'}\NormalTok{, }\StringTok{'breakfast'}\NormalTok{, }\StringTok{'supper'}\NormalTok{, }\StringTok{'dinner'}\NormalTok{, }\StringTok{'eat'}\NormalTok{, }\StringTok{'drink'}\NormalTok{,}
+\NormalTok{human\_words }\OtherTok{=}\NormalTok{ [}\StringTok{\textquotesingle{}wear\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}walk\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}visitor\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}toy\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}tissue\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}throw\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}talk\textquotesingle{}}\NormalTok{, }
+               \StringTok{\textquotesingle{}speak\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}sleep\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}eye\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}enjoy\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}blogger\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}character\textquotesingle{}}\NormalTok{, }
+               \StringTok{\textquotesingle{}candidate\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}breakfast\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}supper\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}dinner\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}eat\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}drink\textquotesingle{}}\NormalTok{,}
                \StringTok{"carry"}\NormalTok{, }\StringTok{"run"}\NormalTok{, }\StringTok{"cast"}\NormalTok{, }\StringTok{"ask"}\NormalTok{, }\StringTok{"awake"}\NormalTok{, }\StringTok{"ear"}\NormalTok{, }\StringTok{"nose"}\NormalTok{, }\StringTok{"lunch"}\NormalTok{]}
@@
 
-\section*{Non-human neutral
-attributes}\label{non-human-neutral-attributes}
+\hypertarget{non-human-neutral-attributes}{%
+\section*{Non-human neutral attributes}\label{non-human-neutral-attributes}}
 \addcontentsline{toc}{section}{Non-human neutral attributes}
 
-\vspace{1mm} \footnotesize
-
-\normalsize
-
-\vspace{1mm} \scriptsize
+\vspace{1mm}
+\footnotesize
+
+\normalsize
+
+\vspace{1mm}
+\scriptsize
 
@@
 \begin{Highlighting}[]
-\NormalTok{neutral_words =}\StringTok{ }\NormalTok{[}\StringTok{'liquor'}\NormalTok{, }\StringTok{'pow'}\NormalTok{, }\StringTok{'ballpark'}\NormalTok{, }\StringTok{'glitchy'}\NormalTok{, }\StringTok{'billy'}\NormalTok{, }\StringTok{'dallas'}\NormalTok{, }
-                 \StringTok{'rip'}\NormalTok{, }\StringTok{'called'}\NormalTok{, }\StringTok{'outlooks'}\NormalTok{, }\StringTok{'viet'}\NormalTok{, }\StringTok{'floater'}\NormalTok{, }\StringTok{'rattlesnake'}\NormalTok{, }\StringTok{'exports'}\NormalTok{,}
-                 \StringTok{'peruvian'}\NormalTok{, }\StringTok{'recursion'}\NormalTok{, }\StringTok{'shortfall'}\NormalTok{, }\StringTok{'corrected'}\NormalTok{, }\StringTok{'amicable'}\NormalTok{,}
-                 \StringTok{'solutions'}\NormalTok{, }\StringTok{'diagnostic'}\NormalTok{, }\StringTok{'patently'}\NormalTok{, }\StringTok{'flops'}\NormalTok{, }\StringTok{'approx'}\NormalTok{, }\StringTok{'percents'}\NormalTok{,}
-                \StringTok{'lox'}\NormalTok{, }\StringTok{'catapults'}\NormalTok{, }\StringTok{'hamburger'}\NormalTok{, }\StringTok{'engulfed'}\NormalTok{, }\StringTok{'households'}\NormalTok{, }\StringTok{'north'}\NormalTok{,}
-                \StringTok{'snubbed'}\NormalTok{, }\StringTok{'playtest'}\NormalTok{, }\StringTok{'replayability'}\NormalTok{, }\StringTok{'glottal'}\NormalTok{, }\StringTok{'parable'}\NormalTok{, }\StringTok{'gingers'}\NormalTok{,}
-                \StringTok{'anachronism'}\NormalTok{, }\StringTok{'organizing'}\NormalTok{, }\StringTok{'reach'}\NormalTok{, }\StringTok{'shtick'}\NormalTok{, }\StringTok{'eleventh'}\NormalTok{, }\StringTok{'cpu'}\NormalTok{, }\StringTok{'ranked'}\NormalTok{,}
-                \StringTok{'irreversibly'}\NormalTok{, }\StringTok{'ponce'}\NormalTok{, }\StringTok{'velociraptor'}\NormalTok{, }\StringTok{'rubber'}\NormalTok{, }\StringTok{'defects'}\NormalTok{, }\StringTok{'puzzle'}\NormalTok{,}
-                \StringTok{'smasher'}\NormalTok{, }\StringTok{'northside'}\NormalTok{, }\StringTok{'heft'}\NormalTok{, }\StringTok{'observation'}\NormalTok{, }\StringTok{'rectum'}\NormalTok{, }\StringTok{'mystical'}\NormalTok{,}
-                \StringTok{'telltale'}\NormalTok{, }\StringTok{'remnants'}\NormalTok{, }\StringTok{'inquiry'}\NormalTok{, }\StringTok{'indisputable'}\NormalTok{, }\StringTok{'boatload'}\NormalTok{, }\StringTok{'lessening'}\NormalTok{,}
-                \StringTok{'uselessness'}\NormalTok{, }\StringTok{'observes'}\NormalTok{, }\StringTok{'fictitious'}\NormalTok{, }\StringTok{'repatriation'}\NormalTok{, }\StringTok{'duh'}\NormalTok{,}
-                \StringTok{'attic'}\NormalTok{, }\StringTok{'schilling'}\NormalTok{, }\StringTok{'charges'}\NormalTok{, }\StringTok{'chatter'}\NormalTok{, }\StringTok{'pad'}\NormalTok{, }\StringTok{'smurfing'}\NormalTok{,}
-                \StringTok{'worthiness'}\NormalTok{, }\StringTok{'definitive'}\NormalTok{, }\StringTok{'neat'}\NormalTok{, }\StringTok{'homogenized'}\NormalTok{,}\StringTok{'lexicon'}\NormalTok{,}
-                \StringTok{'nationalized'}\NormalTok{, }\StringTok{'earpiece'}\NormalTok{, }\StringTok{'specializations'}\NormalTok{, }\StringTok{'lapse'}\NormalTok{,}
-                \StringTok{'concludes'}\NormalTok{, }\StringTok{'weaving'}\NormalTok{, }\StringTok{'apprentices'}\NormalTok{, }\StringTok{'fri'}\NormalTok{, }\StringTok{'younglings'}\NormalTok{,}
-                \StringTok{'militias'}\NormalTok{, }\StringTok{'inscriptions'}\NormalTok{, }\StringTok{'gouda'}\NormalTok{, }\StringTok{'lift'}\NormalTok{, }\StringTok{'laboring'}\NormalTok{, }
-                \StringTok{'adaptive'}\NormalTok{, }\StringTok{'lecture'}\NormalTok{, }\StringTok{'hogging'}\NormalTok{, }\StringTok{'thorne'}\NormalTok{, }\StringTok{'fud'}\NormalTok{, }\StringTok{'skews'}\NormalTok{,}
-                \StringTok{'epistles'}\NormalTok{, }\StringTok{'tagging'}\NormalTok{, }\StringTok{'crud'}\NormalTok{, }\StringTok{'two'}\NormalTok{, }\StringTok{'rebalanced'}\NormalTok{, }\StringTok{'payroll'}\NormalTok{,}
-                \StringTok{'damned'}\NormalTok{, }\StringTok{'approve'}\NormalTok{, }\StringTok{'reason'}\NormalTok{, }\StringTok{'formally'}\NormalTok{, }\StringTok{'releasing'}\NormalTok{, }\StringTok{'muddled'}\NormalTok{,}
-                \StringTok{'mineral'}\NormalTok{, }\StringTok{'shied'}\NormalTok{, }\StringTok{'capital'}\NormalTok{, }\StringTok{'nodded'}\NormalTok{, }\StringTok{'escrow'}\NormalTok{, }\StringTok{'unscientific'}\NormalTok{,}
-                \StringTok{'recognizable'}\NormalTok{, }\StringTok{'entitlement'}\NormalTok{, }\StringTok{'disconnecting'}\NormalTok{, }\StringTok{'marshals'}\NormalTok{,}
-                \StringTok{'winamp'}\NormalTok{, }\StringTok{'forceful'}\NormalTok{, }\StringTok{'lowes'}\NormalTok{, }\StringTok{'ptr'}\NormalTok{, }\StringTok{'sip'}\NormalTok{, }\StringTok{'pencils'}\NormalTok{, }\StringTok{'stomachs'}\NormalTok{,}
-                \StringTok{'goff'}\NormalTok{, }\StringTok{'cg'}\NormalTok{, }\StringTok{'backyard'}\NormalTok{, }\StringTok{'uprooting'}\NormalTok{, }\StringTok{'merging'}\NormalTok{,}
-                \StringTok{'helpful'}\NormalTok{, }\StringTok{'eid'}\NormalTok{, }\StringTok{'trenchcoat'}\NormalTok{, }\StringTok{'airlift'}\NormalTok{, }\StringTok{'frothing'}\NormalTok{,}
-                \StringTok{'pulls'}\NormalTok{, }\StringTok{'volta'}\NormalTok{, }\StringTok{'guinness'}\NormalTok{, }\StringTok{'viewership'}\NormalTok{,}
-                \StringTok{'eruption'}\NormalTok{, }\StringTok{'peeves'}\NormalTok{, }\StringTok{'goat'}\NormalTok{, }\StringTok{'goofy'}\NormalTok{, }\StringTok{'disbanding'}\NormalTok{, }
-                \StringTok{'relented'}\NormalTok{, }\StringTok{'ratings'}\NormalTok{, }\StringTok{'disputed'}\NormalTok{, }\StringTok{'vitamins'}\NormalTok{, }\StringTok{'singled'}\NormalTok{,}
-                \StringTok{'hydroxide'}\NormalTok{, }\StringTok{'telegraphed'}\NormalTok{, }\StringTok{'mercantile'}\NormalTok{, }\StringTok{'headache'}\NormalTok{, }
-                \StringTok{'muppets'}\NormalTok{, }\StringTok{'petal'}\NormalTok{, }\StringTok{'arrange'}\NormalTok{, }\StringTok{'donovan'}\NormalTok{, }\StringTok{'scrutinized'}\NormalTok{,}
-                \StringTok{'spoil'}\NormalTok{, }\StringTok{'examiner'}\NormalTok{, }\StringTok{'ironed'}\NormalTok{, }\StringTok{'maia'}\NormalTok{, }\StringTok{'condensation'}\NormalTok{,}
-                \StringTok{'receipt'}\NormalTok{, }\StringTok{'solider'}\NormalTok{, }\StringTok{'tattooing'}\NormalTok{, }\StringTok{'encoded'}\NormalTok{,}
-                \StringTok{'compartmentalize'}\NormalTok{, }\StringTok{'lain'}\NormalTok{, }\StringTok{'the'}\NormalTok{, }\StringTok{'gov'}\NormalTok{,}
-                \StringTok{'printers'}\NormalTok{, }\StringTok{'hiked'}\NormalTok{, }\StringTok{'resentment'}\NormalTok{, }\StringTok{'revisionism'}\NormalTok{, }\StringTok{'tavern'}\NormalTok{,}
-                \StringTok{'backpacking'}\NormalTok{, }\StringTok{'pestering'}\NormalTok{, }\StringTok{'gassed'}\NormalTok{, }\StringTok{'acknowledges'}\NormalTok{,}
-                \StringTok{'testimonies'}\NormalTok{, }\StringTok{'parlance'}\NormalTok{, }\StringTok{'hallucinate'}\NormalTok{, }\StringTok{'speeches'}\NormalTok{,}
-                \StringTok{'engaging'}\NormalTok{, }\StringTok{'solder'}\NormalTok{, }\StringTok{'perceptive'}\NormalTok{, }\StringTok{'microbiology'}\NormalTok{, }\StringTok{'reconnaissance'}\NormalTok{, }\StringTok{'garlic'}\NormalTok{,}
-                \StringTok{'neutrals'}\NormalTok{, }\StringTok{'width'}\NormalTok{, }\StringTok{'literaly'}\NormalTok{, }\StringTok{'guild'}\NormalTok{, }\StringTok{'despicable'}\NormalTok{, }\StringTok{'dion'}\NormalTok{,}
-                \StringTok{'option'}\NormalTok{, }\StringTok{'transistors'}\NormalTok{, }\StringTok{'chiropractic'}\NormalTok{, }\StringTok{'tattered'}\NormalTok{, }\StringTok{'consolidating'}\NormalTok{,}
-                \StringTok{'olds'}\NormalTok{, }\StringTok{'garmin'}\NormalTok{, }\StringTok{'shift'}\NormalTok{, }\StringTok{'granted'}\NormalTok{, }\StringTok{'intramural'}\NormalTok{, }\StringTok{'allie'}\NormalTok{, }\StringTok{'cylinders'}\NormalTok{,}
-                \StringTok{'wishlist'}\NormalTok{, }\StringTok{'crank'}\NormalTok{, }\StringTok{'wrongly'}\NormalTok{, }\StringTok{'workshop'}\NormalTok{, }\StringTok{'yesterday'}\NormalTok{, }\StringTok{'wooden'}\NormalTok{, }
-                \StringTok{'without'}\NormalTok{, }\StringTok{'wheel'}\NormalTok{, }\StringTok{'weather'}\NormalTok{, }\StringTok{'watch'}\NormalTok{, }\StringTok{'wage'}\NormalTok{, }\StringTok{'version'}\NormalTok{, }\StringTok{'usually'}\NormalTok{,}
-                \StringTok{'twice'}\NormalTok{, }\StringTok{'tomato'}\NormalTok{, }\StringTok{'ticket'}\NormalTok{, }\StringTok{'text'}\NormalTok{, }\StringTok{'switch'}\NormalTok{, }\StringTok{'studio'}\NormalTok{, }\StringTok{'stick'}\NormalTok{,}
-                \StringTok{'soup'}\NormalTok{, }\StringTok{'sometimes'}\NormalTok{, }\StringTok{'signal'}\NormalTok{, }\StringTok{'prior'}\NormalTok{, }\StringTok{'plant'}\NormalTok{, }\StringTok{'photo'}\NormalTok{,}
-                \StringTok{'path'}\NormalTok{, }\StringTok{'park'}\NormalTok{, }\StringTok{'near'}\NormalTok{, }\StringTok{'menu'}\NormalTok{, }\StringTok{'latter'}\NormalTok{, }\StringTok{'grass'}\NormalTok{, }\StringTok{'clock'}\NormalTok{]}
+\NormalTok{neutral\_words }\OtherTok{=}\NormalTok{ [}\StringTok{\textquotesingle{}liquor\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}pow\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}ballpark\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}glitchy\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}billy\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}dallas\textquotesingle{}}\NormalTok{, }
+                 \StringTok{\textquotesingle{}rip\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}called\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}outlooks\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}viet\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}floater\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}rattlesnake\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}exports\textquotesingle{}}\NormalTok{,}
+                 \StringTok{\textquotesingle{}peruvian\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}recursion\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}shortfall\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}corrected\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}amicable\textquotesingle{}}\NormalTok{,}
+                 \StringTok{\textquotesingle{}solutions\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}diagnostic\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}patently\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}flops\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}approx\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}percents\textquotesingle{}}\NormalTok{,}
+                \StringTok{\textquotesingle{}lox\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}catapults\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}hamburger\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}engulfed\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}households\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}north\textquotesingle{}}\NormalTok{,}
+                \StringTok{\textquotesingle{}snubbed\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}playtest\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}replayability\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}glottal\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}parable\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}gingers\textquotesingle{}}\NormalTok{,}
+                \StringTok{\textquotesingle{}anachronism\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}organizing\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}reach\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}shtick\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}eleventh\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}cpu\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}ranked\textquotesingle{}}\NormalTok{,}
+                \StringTok{\textquotesingle{}irreversibly\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}ponce\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}velociraptor\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}rubber\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}defects\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}puzzle\textquotesingle{}}\NormalTok{,}
+                \StringTok{\textquotesingle{}smasher\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}northside\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}heft\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}observation\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}rectum\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}mystical\textquotesingle{}}\NormalTok{,}
+                \StringTok{\textquotesingle{}telltale\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}remnants\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}inquiry\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}indisputable\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}boatload\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}lessening\textquotesingle{}}\NormalTok{,}
+                \StringTok{\textquotesingle{}uselessness\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}observes\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}fictitious\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}repatriation\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}duh\textquotesingle{}}\NormalTok{,}
+                \StringTok{\textquotesingle{}attic\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}schilling\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}charges\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}chatter\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}pad\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}smurfing\textquotesingle{}}\NormalTok{,}
+                \StringTok{\textquotesingle{}worthiness\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}definitive\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}neat\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}homogenized\textquotesingle{}}\NormalTok{,}\StringTok{\textquotesingle{}lexicon\textquotesingle{}}\NormalTok{,}
+                \StringTok{\textquotesingle{}nationalized\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}earpiece\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}specializations\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}lapse\textquotesingle{}}\NormalTok{,}
+                \StringTok{\textquotesingle{}concludes\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}weaving\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}apprentices\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}fri\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}younglings\textquotesingle{}}\NormalTok{,}
+                \StringTok{\textquotesingle{}militias\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}inscriptions\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}gouda\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}lift\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}laboring\textquotesingle{}}\NormalTok{, }
+                \StringTok{\textquotesingle{}adaptive\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}lecture\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}hogging\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}thorne\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}fud\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}skews\textquotesingle{}}\NormalTok{,}
+                \StringTok{\textquotesingle{}epistles\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}tagging\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}crud\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}two\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}rebalanced\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}payroll\textquotesingle{}}\NormalTok{,}
+                \StringTok{\textquotesingle{}damned\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}approve\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}reason\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}formally\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}releasing\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}muddled\textquotesingle{}}\NormalTok{,}
+                \StringTok{\textquotesingle{}mineral\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}shied\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}capital\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}nodded\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}escrow\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}unscientific\textquotesingle{}}\NormalTok{,}
+                \StringTok{\textquotesingle{}recognizable\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}entitlement\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}disconnecting\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}marshals\textquotesingle{}}\NormalTok{,}
+                \StringTok{\textquotesingle{}winamp\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}forceful\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}lowes\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}ptr\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}sip\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}pencils\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}stomachs\textquotesingle{}}\NormalTok{,}
+                \StringTok{\textquotesingle{}goff\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}cg\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}backyard\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}uprooting\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}merging\textquotesingle{}}\NormalTok{,}
+                \StringTok{\textquotesingle{}helpful\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}eid\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}trenchcoat\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}airlift\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}frothing\textquotesingle{}}\NormalTok{,}
+                \StringTok{\textquotesingle{}pulls\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}volta\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}guinness\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}viewership\textquotesingle{}}\NormalTok{,}
+                \StringTok{\textquotesingle{}eruption\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}peeves\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}goat\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}goofy\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}disbanding\textquotesingle{}}\NormalTok{, }
+                \StringTok{\textquotesingle{}relented\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}ratings\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}disputed\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}vitamins\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}singled\textquotesingle{}}\NormalTok{,}
+                \StringTok{\textquotesingle{}hydroxide\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}telegraphed\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}mercantile\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}headache\textquotesingle{}}\NormalTok{, }
+                \StringTok{\textquotesingle{}muppets\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}petal\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}arrange\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}donovan\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}scrutinized\textquotesingle{}}\NormalTok{,}
+                \StringTok{\textquotesingle{}spoil\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}examiner\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}ironed\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}maia\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}condensation\textquotesingle{}}\NormalTok{,}
+                \StringTok{\textquotesingle{}receipt\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}solider\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}tattooing\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}encoded\textquotesingle{}}\NormalTok{,}
+                \StringTok{\textquotesingle{}compartmentalize\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}lain\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}the\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}gov\textquotesingle{}}\NormalTok{,}
+                \StringTok{\textquotesingle{}printers\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}hiked\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}resentment\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}revisionism\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}tavern\textquotesingle{}}\NormalTok{,}
+                \StringTok{\textquotesingle{}backpacking\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}pestering\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}gassed\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}acknowledges\textquotesingle{}}\NormalTok{,}
+                \StringTok{\textquotesingle{}testimonies\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}parlance\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}hallucinate\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}speeches\textquotesingle{}}\NormalTok{,}
+                \StringTok{\textquotesingle{}engaging\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}solder\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}perceptive\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}microbiology\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}reconnaissance\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}garlic\textquotesingle{}}\NormalTok{,}
+                \StringTok{\textquotesingle{}neutrals\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}width\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}literaly\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}guild\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}despicable\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}dion\textquotesingle{}}\NormalTok{,}
+                \StringTok{\textquotesingle{}option\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}transistors\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}chiropractic\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}tattered\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}consolidating\textquotesingle{}}\NormalTok{,}
+                \StringTok{\textquotesingle{}olds\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}garmin\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}shift\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}granted\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}intramural\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}allie\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}cylinders\textquotesingle{}}\NormalTok{,}
+                \StringTok{\textquotesingle{}wishlist\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}crank\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}wrongly\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}workshop\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}yesterday\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}wooden\textquotesingle{}}\NormalTok{, }
+                \StringTok{\textquotesingle{}without\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}wheel\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}weather\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}watch\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}wage\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}version\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}usually\textquotesingle{}}\NormalTok{,}
+                \StringTok{\textquotesingle{}twice\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}tomato\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}ticket\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}text\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}switch\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}studio\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}stick\textquotesingle{}}\NormalTok{,}
+                \StringTok{\textquotesingle{}soup\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}sometimes\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}signal\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}prior\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}plant\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}photo\textquotesingle{}}\NormalTok{,}
+                \StringTok{\textquotesingle{}path\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}park\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}near\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}menu\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}latter\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}grass\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}clock\textquotesingle{}}\NormalTok{]}
 \end{Highlighting}
@@
 
-\chapter*{References}\label{references}
+\hypertarget{references}{%
+\chapter*{References}\label{references}}
 \addcontentsline{toc}{chapter}{References}
@@
 \hypertarget{refs}{}
-\hypertarget{ref-Bolukbasi2016Man}{}
-Bolukbasi, T., Chang, K., Zou, J. Y., Saligrama, V., \& Kalai, A.
-(2016). Man is to computer programmer as woman is to homemaker?
-Debiasing word embeddings. \emph{CoRR}, \emph{abs/1607.06520}. Retrieved
-from \url{http://arxiv.org/abs/1607.06520}
-
-\hypertarget{ref-Gonen2019Lipstick}{}
-Gonen, H., \& Goldberg, Y. (2019). Lipstick on a pig: Debiasing methods
-cover up systematic gender biases in word embeddings but do not remove
-them. \emph{CoRR}, \emph{abs/1903.03862}. Retrieved from
-\url{http://arxiv.org/abs/1903.03862}
-
-\hypertarget{ref-Caliskan2017Semantics}{}
-Islam, A. C., Bryson, J. J., \& Narayanan, A. (2016). Semantics derived
-automatically from language corpora necessarily contain human biases.
-\emph{CoRR}, \emph{abs/1608.07187}. Retrieved from
-\url{http://arxiv.org/abs/1608.07187}
-
-\hypertarget{ref-Manzini2019blackToCriminal}{}
-Manzini, T., Lim, Y. C., Tsvetkov, Y., \& Black, A. W. (2019). Black is
-to criminal as caucasian is to police: Detecting and removing multiclass
-bias in word embeddings.
-
-\hypertarget{ref-Mehrabi2019Survey}{}
+\begin{CSLReferences}{1}{0}
+\leavevmode\hypertarget{ref-Bolukbasi2016Man}{}%
+Bolukbasi, T., Chang, K.-W., Zou, J.
+Y., Saligrama, V., \& Kalai, A.
+(2016).
+Man is to computer programmer as woman is to homemaker?
+Debiasing word embeddings.
+\emph{CoRR}, \emph{abs/1607.06520}.
+Retrieved from \url{http://arxiv.org/abs/1607.06520}
+
+\leavevmode\hypertarget{ref-Gonen2019Lipstick}{}%
+Gonen, H., \& Goldberg, Y.
+(2019).
+Lipstick on a pig: Debiasing methods cover up systematic gender biases in word embeddings but do not remove them.
+\emph{CoRR}, \emph{abs/1903.03862}.
+Retrieved from \url{http://arxiv.org/abs/1903.03862}
+
+\leavevmode\hypertarget{ref-Caliskan2017Semantics}{}%
+Islam, A.
+C., Bryson, J.
+J., \& Narayanan, A.
+(2016).
+Semantics derived automatically from language corpora necessarily contain human biases.
+\emph{CoRR}, \emph{abs/1608.07187}.
+Retrieved from \url{http://arxiv.org/abs/1608.07187}
+
+\leavevmode\hypertarget{ref-Manzini2019blackToCriminal}{}%
+Manzini, T., Lim, Y.
+C., Tsvetkov, Y., \& Black, A.
+W.
+(2019).
+Black is to criminal as caucasian is to police: Detecting and removing multiclass bias in word embeddings.
+Retrieved from \url{http://arxiv.org/abs/1904.04047}
+
+\leavevmode\hypertarget{ref-Mehrabi2019Survey}{}%
 Mehrabi, N., Morstatter, F., Saxena, N., Lerman, K., \& Galstyan, A.
-(2019). A survey on bias and fairness in machine learning. \emph{CoRR},
-\emph{abs/1908.09635}. Retrieved from
-\url{http://arxiv.org/abs/1908.09635}
-
-\hypertarget{ref-Nissim2019Fair}{}
-Nissim, M., Noord, R. van, \& Goot, R. van der. (2019). Fair is better
-than sensational: Man is to doctor as woman is to doctor. \emph{CoRR},
-\emph{abs/1905.09866}. Retrieved from
-\url{http://arxiv.org/abs/1905.09866}
-
-\hypertarget{ref-Venkat2018Curse}{}
+(2019).
+A survey on bias and fairness in machine learning.
+\emph{CoRR}, \emph{abs/1908.09635}.
+Retrieved from \url{http://arxiv.org/abs/1908.09635}
+
+\leavevmode\hypertarget{ref-Nissim2019Fair}{}%
+Nissim, M., Noord, R.
+van, \& Goot, R.
+van der.
+(2019).
+Fair is better than sensational: Man is to doctor as woman is to doctor.
+\emph{CoRR}, \emph{abs/1905.09866}.
+Retrieved from \url{http://arxiv.org/abs/1905.09866}
+
+\leavevmode\hypertarget{ref-Venkat2018Curse}{}%
 Venkat, N. (2018). The curse of dimensionality: Inside out.
 
+\end{CSLReferences}
+
 \end{document}

### 2021-06-23 paperDraft/paperDraft.tex

--- 2021-06-10 paperDraft/paperDraft.tex
@@
 
+\tableofcontents
+
 \section{Cosine-based measures of
 bias}\label{cosine-based-measures-of-bias}
+
+\todo{two introductory}
 
@@
 non-explicitly gendered words are in equal distance to both elements in
-all explicitly gendered pairs. Given the (ideally) gender netural words
+all explicitly gender pairs. Given the (ideally) gender neutral words
 \(N\) and the gender direction \(g\) the direct direct gender bias is
@@
 
-\subsection{The problem with means of
+\section{Some methodological
+problems}\label{some-methodological-problems}
+
+\section{The problem with means of
 means}\label{the-problem-with-means-of-means}
@@
 
+\section{Bayesian estimation}\label{bayesian-estimation}
+
+\begin{itemize}
+\item
+  reddit i google, ogolnie i z podzialem na protected words
+\item
+  dokladniej gender, reszta na koniec w zalaczniku
+\end{itemize}
+
+\section{Effects of debiasing}\label{effects-of-debiasing}
+
+\section{Discussion}\label{discussion}
+
 \newpage

### 2021-06-23 thesis/index.Rmd
Adds logs

--- 2021-06-08 thesis/index.Rmd
@@
     extra_dependencies: ["todonotes"]
+
   bookdown::gitbook:
@@
 ```
-
-# Preface {-}
-
-Test
-
-<!-- testsds -->
-
-<!-- \begin{align}  -->
-<!--   f\left(k\right) = \binom{n}{k} p^k\left(1-p\right)^{n-k} -->
-<!--   (\#eq:binom) -->
-<!-- \end{align}  -->
-
-<!-- This is a citation [@diamond90] which uses keys from the bib file listed in the preamble. -->
-
-<!-- Equation \@ref(eq:binom)^[This is a footnote containing a double citation [@diamond90; @dahlmanNakedStat2020; @Zhao2019Gender].] -->
-
-<!-- Note that chapter files are found and compiled automatically, but the file names have to contain chapter numbers first.
-For instance, we used `01-intro.Rmd`, placed in the same folder.
-Observe how we included r code inline.
--->
-
-<!-- ```{r fig-margin, fig.margin=TRUE} -->
-<!-- plot(cars) -->
-<!-- ``` -->

### 2021-06-28 paperDraft/paperDraft.Rmd
paper plan
Adds a few sentences to paper

@@
 
+\tableofcontents
+
 # Cosine-based measures of bias
+
+Modern Natural Language Processing (NLP) models are used to complete various tasks such as providing email filters, smart assistants, search results, language translations, text analytics and so on.
+All of them need as an input words represented by means of numbers which is accomplished with word embeddings.
+It seems that in the learning process these models can learn implicit biases that reflect harmful stereotypical thinking.
+One of the sources of bias in NLP can be located in the way the word embeddings are made.
+There is a considerable amount of literature available on the topic of bias detection and mitigation in NLP models.
 
@@
 The underlying idea is that no bias is present if non-explicitly gendered words are in equal distance to both elements in all explicitly gender pairs.
-Given the (ideally) gender neutral words $N$ and the gender direction $g$ the direct direct gender bias is defined as the average distance of the words in $N$ from $g$ ($c$ is a parameter determining how strict we want to be):
+Given the (ideally) gender neutral words $N$ and the gender direction $g$ the direct gender bias is defined as the average distance of the words in $N$ from $g$ ($c$ is a parameter determining how strict we want to be):
 \begin{align}
@@
 
-## The problem with means of means
+# Some methodological problems
+
+# The problem with means of means
 
@@
 
+# Bayesian estimation
+
+- reddit i google, ogolnie i z podzialem na protected words
+
+- dokladniej gender, reszta na koniec w zalaczniku
+
+# Effects of debiasing
+
+# Discussion
+
 \newpage

### 2021-07-01 paperDraft/paperDraft.tex

--- 2021-06-16 paperDraft/paperDraft.tex
@@
 
+\tableofcontents
+
 \hypertarget{cosine-based-measures-of-bias}{%
@@
 bias}\label{cosine-based-measures-of-bias}}
+
+Modern Natural Language Processing (NLP) models are used to complete
+various tasks such as providing email filters, smart assistants, search
+results, language translations, text analytics and so on. All of them
+need as an input words represented by means of numbers which is
+accomplished with word embeddings. It seems that in the learning process
+these models can learn implicit biases that reflect harmful
+stereotypical thinking. One of the sources of bias in NLP can be located
+in the way the word embeddings are made. There is a considerable amount
+of literature available on the topic of bias detection and mitigation in
+NLP models.
 
@@
 non-explicitly gendered words are in equal distance to both elements in
-all explicitly gendered pairs. Given the (ideally) gender netural words
-\(N\) and the gender direction \(g\) the direct direct gender bias is
-defined as the average distance of the words in \(N\) from \(g\) (\(c\)
-is a parameter determining how strict we want to be): \begin{align}
+all explicitly gender pairs. Given the (ideally) gender neutral words
+\(N\) and the gender direction \(g\) the direct gender bias is defined
+as the average distance of the words in \(N\) from \(g\) (\(c\) is a
+parameter determining how strict we want to be): \begin{align}
 \mathsf{directBias_c(N,g)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},g)\vert^c}{\vert N \vert }
@@
 
+\hypertarget{some-methodological-problems}{%
+\section{Some methodological
+problems}\label{some-methodological-problems}}
+
 \hypertarget{the-problem-with-means-of-means}{%
-\subsection{The problem with means of
+\section{The problem with means of
 means}\label{the-problem-with-means-of-means}}
@@
 
+\hypertarget{bayesian-estimation}{%
+\section{Bayesian estimation}\label{bayesian-estimation}}
+
+\begin{itemize}
+\item
+  reddit i google, ogolnie i z podzialem na protected words
+\item
+  dokladniej gender, reszta na koniec w zalaczniku
+\end{itemize}
+
+\hypertarget{effects-of-debiasing}{%
+\section{Effects of debiasing}\label{effects-of-debiasing}}
+
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
+
 \newpage

### 2021-07-11 paperDraft/paperDraft.Rmd
Starts article

@@
 
-- reddit i google, ogolnie i z podzialem na protected words
+We noticed that in @Manzini2019blackToCriminal words from all three sub-classes (in the case of religion by sub-class we mean Christianity, Judaism, Islam) were compared against all of the stereotypes, which means that there was no distinction between cases in which the stereotype is associated with a given sub-class, as opposed to the situation in which it is associated with another one.
+Not all of the stereotypical words have to be considered as harmful for all of the sub-groups.
+One should investigate the sub-groups separately as some of them may have stronger harmful associations than others.
 
-- dokladniej gender, reszta na koniec w zalaczniku
+We decided to add control groups in the form of two classes — neutral words and human-related words.
+Without a proper control group it is quite hard to compare the resulting cosine distances and decide on their significance in bias detection.
+We prepared approximately 230 neutral words to double-check the prima-facie neutral hypothesis that their cosine similarity to the protected words will oscillate around 0 (that is, the distances will be around 1).
+This provides us with a more reliable point of reference.
+Moreover,
+we added human attributes that are associated with people in general to investigate whether the smaller cosine distance between protected words and stereotypes can result simply from the fact that the stereotype predicates are associated with humans.
+
+One may consider improving the bias detection method presented in @Manzini2019blackToCriminal.
+As the chosen protected words and attributes do not represent the whole biased population but only a sample, it is justified to introduce the uncertainty measure to the method.
+We decided to use Bayesian estimation to make the cosine distance method more methodologically correct.
+
+Our priors distributions for the Bayesian model reflect our knowledge about the data.
+The distribution for mean distance (cosine distance) is set to be between 0 and 2 with the mean equal to 1.
+We assume that most of the cosine distances may be considered neutral and therefore their similarity will be small.
+The distribution for prior for standard deviation is mostly between 0 and 1 with mean around 0.
+Finally the distribution for prior for coefficients is mostly between -1 and 1, with mean around 0.
+One should remember that the cosine distance values have range between 0 and 2 therefore the choice of the prior distributions should reflect this constraint.
+As we want to verify as objectively as possible the assumptions present in @Manzini2019blackToCriminal we decided to use fairly weak and slightly skeptical regularizing priors.
+The models parameters will have a posterior distribution obtained using either Hamiltionian Monte Carlo methods (STAN) available through the rethinking package.
+
+\begin{center}
+\begin{figure}[!htb]\centering
+   \begin{minipage}{0.55\textwidth}
+  \includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
+   \end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/religionGoogleCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/genderCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/genderGoogleCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/raceCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/raceGoogleCoeffs.jpeg}
+   \end{minipage}
+\end{figure}
+
+\end{center}
+
+Let's analyse dataset-level coeﬀicients together with their highest posterior density intervals, for the three topic groups.
+We compare the results for two different word emebeddings corpuses - RedditL2 and GoogleNews.
+For Google embeddings the HPDIs
+for all class coeﬀicients (associated, different, human, and none) include zero.
+This can lead one to a conclusion that the impact for associated, different, human and neutral attribute is, when averaged, quite similar.
+This indicates how including the uncertainty
+may change the use and interpretation of Manzini, Lim, Tsvetkov, & Black (2019) MAC metric. It seems that if one focuses only on differences between means of means, this is too simplistic.
+
+In the case of Reddit word embeddings the situation is similar although HPDI is below 0 for Gender class when looking at `associated` and `different` mean coeﬀicient.
+This can suggest that there is indeed slightly stronger impact of these attributes.
+One should also notice how in general associated and different coeﬀicients have quite similar HPDI range, the highest observed absolute difference is equal to only 0.1.
+This suggests that again the impact of associated attributes and different ones is not clear at first sight.
+Finally, let’s compare the HPDIs for Reddit and Reddit debiased datasets.
+For Religion and Race dataset there is a minor shift (in absolute values the highest change is equal to approximately 0.1) of the mean coeﬀicients towards zero.
+However for Gender dataset there is no significant change.
+This is of course a general look at the group coeﬀicients.
+Let’s now analyze the individual words.
+
+Looking at Reddit individual words one can observe that for Religion the cosine distance results for the associated attributes are for most of the words close to human attributes as well.
+This can suggest that in some cases words concerning humans can have higher
+similarity with some protected words independently of whether they are stereotypical or neutral-human words.
+One should also pay attention to the fact that for all of the associated and different attributes in Religion, the uncertainty intervals overlap at
+some point.
+What is even more surprising is that for protected words, such as “torah” the associated attribute has the the cosine distance slightly over 1, which means no positive similarity!
+If the protected words that we chose do not have high similarity
+with harmful associated attributes, then one should consider at least three scenarios.
+The first one is that the choice of the protected words and attributes may be corrupted.
+The second one is that the metric is unable to catch the hidden bias properly.
+The third one is that there is actually no bias between the words.
+Regardless of which scenario one considers, it is essential to take a look at the individual values before averaging them or aggregating in other ways.
+It seems that using Bayesian method can enhance
+the process of verifying the hypotheses concerning the choice of protected words and attributes.
+
+Surprisingly in Gender one can observe high cosine similarity values between some female stereotypical professions and male protected words.
+If a word stereotypically associated with females has low cosine distance to male protected words, then one should investigate the issue further.
+The reasons for this unexpected cosine distance may lie in the frequency of appearance of the protected words in the raw data.
+Some of the groups (like Muslim people or females) may have less representation in the data that is taken as an input for word embeddings.
+Therefore, if we assume that the MAC detects actually co-occurrence only, it makes sense that they can have high similarity with associated attributes and lower similarity with different ones.
+At the same time male protected words and other religions can have high similarity with different attributes because they have greater representation in the dataset in general and occur close to much more concepts.
+Cosine distance seems to capture the information on the co-occurrence of words and not on the semantic similarity strictly speaking.
+
+In Google, one may notice interesting phenomena as well.
+Although the GoogleNews dataset is larger, trained on different data sources and with the use of more dimensions, some of the results are similar to the ones obtained for the Reddit corpus.
+For Religion
+almost all of the values for associated class intersect with different class as well.
+In the case of Race it is 100% of the available words.
+This indicates again that it is not clear how the metric should be used.
+Quite a different situation takes place for
+Gender, where the similarity between associated and different is present mostly for the male protected words.
+This means that male words have high similarity with both male stereotypical professions and female ones.
+However, for females, the similarity is
+high mostly only for the female stereotypical professions. This observation could not be made when using MAC metric only, as it requires investigating the individual words, and providing uncertainty.
 
 # Effects of debiasing
+
+One may verify if the dataset-level coefficients changed after the debiasing method used in @Manzini2019blackToCriminal.
+
+\begin{center}
+\begin{figure}[!htb]\centering
+   \begin{minipage}{0.55\textwidth}
+  \includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
+   \end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/debiasedReligionRedditCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/genderCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/debiasedGenderRedditCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/raceCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/debiasedRaceRedditCoeffs.jpeg}
+   \end{minipage}
+\end{figure}
+
+\end{center}
+
+When analyzing the Reddit debiased results one should investigate the change of the individual cosine distance values as well.
+It seems that not all of the protected words are treated equally when performing debiasing.
+In religion, the values for associated class moved towards 0 for most of the words except for the ones for Islam, where still the associated class has quite lower cosine distance than the different one.
+Similar situation takes place in Gender, where female words have still much lower cosine distance values for
+associated class than for the different one.
+In the case of male protected words it is mostly almost the same interval for associated and different.
+As the Gender data is quite specific as the attributes are not harmful adjectives but (in an ideal world)
+neutral professions, the aim (if we follow MAC assumptions) should actually be to make the cosine distance same for both female and male protected words.
+However, as we pointed out the situation after debiasing can sometimes be better only for one protected
+group, which is not the desired outcome.
+In the case of Religion, even after debiasing, all of the protected words have intersections between associated and different class and similarity greater than 0.
+As all of the attributes in religion data are negative
+and harmful stereotypes, one should not aim at making the distance between protected words and associated, and different class the same but rather to move it towards 1.
 

### 2021-07-11 paperDraft/paperDraft.tex
Starts article

@@
 
-(\textbf{Manzini2019blackToCriminal?}) modify WEAT to a multi-class
+Manzini, Lim, Tsvetkov, \& Black (2019) modify WEAT to a multi-class
 setting, introducing Mean Average Cosine similarity as a measure of bias
@@
 
-\begin{itemize}
-\item
-  reddit i google, ogolnie i z podzialem na protected words
-\item
-  dokladniej gender, reszta na koniec w zalaczniku
-\end{itemize}
+We noticed that in Manzini, Lim, Tsvetkov, \& Black (2019) words from
+all three sub-classes (in the case of religion by sub-class we mean
+Christianity, Judaism, Islam) were compared against all of the
+stereotypes, which means that there was no distinction between cases in
+which the stereotype is associated with a given sub-class, as opposed to
+the situation in which it is associated with another one. Not all of the
+stereotypical words have to be considered as harmful for all of the
+sub-groups. One should investigate the sub-groups separately as some of
+them may have stronger harmful associations than others.
+
+We decided to add control groups in the form of two classes --- neutral
+words and human-related words. Without a proper control group it is
+quite hard to compare the resulting cosine distances and decide on their
+significance in bias detection. We prepared approximately 230 neutral
+words to double-check the prima-facie neutral hypothesis that their
+cosine similarity to the protected words will oscillate around 0 (that
+is, the distances will be around 1). This provides us with a more
+reliable point of reference. Moreover, we added human attributes that
+are associated with people in general to investigate whether the smaller
+cosine distance between protected words and stereotypes can result
+simply from the fact that the stereotype predicates are associated with
+humans.
+
+One may consider improving the bias detection method presented in
+Manzini, Lim, Tsvetkov, \& Black (2019). As the chosen protected words
+and attributes do not represent the whole biased population but only a
+sample, it is justified to introduce the uncertainty measure to the
+method. We decided to use Bayesian estimation to make the cosine
+distance method more methodologically correct.
+
+Our priors distributions for the Bayesian model reflect our knowledge
+about the data. The distribution for mean distance (cosine distance) is
+set to be between 0 and 2 with the mean equal to 1. We assume that most
+of the cosine distances may be considered neutral and therefore their
+similarity will be small. The distribution for prior for standard
+deviation is mostly between 0 and 1 with mean around 0. Finally the
+distribution for prior for coefficients is mostly between -1 and 1, with
+mean around 0. One should remember that the cosine distance values have
+range between 0 and 2 therefore the choice of the prior distributions
+should reflect this constraint. As we want to verify as objectively as
+possible the assumptions present in Manzini, Lim, Tsvetkov, \& Black
+(2019) we decided to use fairly weak and slightly skeptical regularizing
+priors. The models parameters will have a posterior distribution
+obtained using either Hamiltionian Monte Carlo methods (STAN) available
+through the rethinking package.
+
+\begin{center}
+\begin{figure}[!htb]\centering
+   \begin{minipage}{0.55\textwidth}
+  \includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
+   \end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/religionGoogleCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/genderCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/genderGoogleCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/raceCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/raceGoogleCoeffs.jpeg}
+   \end{minipage}
+\end{figure}
+
+\end{center}
+
+Let's analyse dataset-level coeﬀicients together with their highest
+posterior density intervals, for the three topic groups. We compare the
+results for two different word emebeddings corpuses - RedditL2 and
+GoogleNews. For Google embeddings the HPDIs for all class coeﬀicients
+(associated, different, human, and none) include zero. This can lead one
+to a conclusion that the impact for associated, different, human and
+neutral attribute is, when averaged, quite similar. This indicates how
+including the uncertainty may change the use and interpretation of
+Manzini, Lim, Tsvetkov, \& Black (2019) MAC metric. It seems that if one
+focuses only on differences between means of means, this is too
+simplistic.
+
+In the case of Reddit word embeddings the situation is similar although
+HPDI is below 0 for Gender class when looking at \texttt{associated} and
+\texttt{different} mean coeﬀicient. This can suggest that there is
+indeed slightly stronger impact of these attributes. One should also
+notice how in general associated and different coeﬀicients have quite
+similar HPDI range, the highest observed absolute difference is equal to
+only 0.1. This suggests that again the impact of associated attributes
+and different ones is not clear at first sight. Finally, let's compare
+the HPDIs for Reddit and Reddit debiased datasets. For Religion and Race
+dataset there is a minor shift (in absolute values the highest change is
+equal to approximately 0.1) of the mean coeﬀicients towards zero.
+However for Gender dataset there is no significant change. This is of
+course a general look at the group coeﬀicients. Let's now analyze the
+individual words.
+
+Looking at Reddit individual words one can observe that for Religion the
+cosine distance results for the associated attributes are for most of
+the words close to human attributes as well. This can suggest that in
+some cases words concerning humans can have higher similarity with some
+protected words independently of whether they are stereotypical or
+neutral-human words. One should also pay attention to the fact that for
+all of the associated and different attributes in Religion, the
+uncertainty intervals overlap at some point. What is even more
+surprising is that for protected words, such as ``torah'' the associated
+attribute has the the cosine distance slightly over 1, which means no
+positive similarity! If the protected words that we chose do not have
+high similarity with harmful associated attributes, then one should
+consider at least three scenarios. The first one is that the choice of
+the protected words and attributes may be corrupted. The second one is
+that the metric is unable to catch the hidden bias properly. The third
+one is that there is actually no bias between the words. Regardless of
+which scenario one considers, it is essential to take a look at the
+individual values before averaging them or aggregating in other ways. It
+seems that using Bayesian method can enhance the process of verifying
+the hypotheses concerning the choice of protected words and attributes.
+
+Surprisingly in Gender one can observe high cosine similarity values
+between some female stereotypical professions and male protected words.
+If a word stereotypically associated with females has low cosine
+distance to male protected words, then one should investigate the issue
+further. The reasons for this unexpected cosine distance may lie in the
+frequency of appearance of the protected words in the raw data. Some of
+the groups (like Muslim people or females) may have less representation
+in the data that is taken as an input for word embeddings. Therefore, if
+we assume that the MAC detects actually co-occurrence only, it makes
+sense that they can have high similarity with associated attributes and
+lower similarity with different ones. At the same time male protected
+words and other religions can have high similarity with different
+attributes because they have greater representation in the dataset in
+general and occur close to much more concepts. Cosine distance seems to
+capture the information on the co-occurrence of words and not on the
+semantic similarity strictly speaking.
+
+In Google, one may notice interesting phenomena as well. Although the
+GoogleNews dataset is larger, trained on different data sources and with
+the use of more dimensions, some of the results are similar to the ones
+obtained for the Reddit corpus. For Religion almost all of the values
+for associated class intersect with different class as well. In the case
+of Race it is 100\% of the available words. This indicates again that it
+is not clear how the metric should be used. Quite a different situation
+takes place for Gender, where the similarity between associated and
+different is present mostly for the male protected words. This means
+that male words have high similarity with both male stereotypical
+professions and female ones. However, for females, the similarity is
+high mostly only for the female stereotypical professions. This
+observation could not be made when using MAC metric only, as it requires
+investigating the individual words, and providing uncertainty.
 
@@
 \section{Effects of debiasing}\label{effects-of-debiasing}}
+
+When analyzing the Reddit debiased results one should investigate the
+change of the cosine distance values. It seems that not all of the
+protected words are treated equally when performing debiasing. In
+religion, the values for associated class moved towards 0 for most of
+the words except for the ones for Islam, where still the associated
+class has quite lower cosine distance than the different one. Similar
+situation takes place in Gender, where female words have still much
+lower cosine distance values for associated class than for the different
+one. In the case of male protected words it is mostly almost the same
+interval for associated and different. As the Gender data is quite
+specific as the attributes are not harmful adjectives but (in an ideal
+world) neutral professions, the aim (if we follow MAC assumptions)
+should actually be to make the cosine distance same for both female and
+male protected words. However, as we pointed out the situation after
+debiasing can sometimes be better only for one protected group, which is
+not the desired outcome. In the case of Religion, even after debiasing,
+all of the protected words have intersections between associated and
+different class and similarity greater than 0. As all of the attributes
+in religion data are negative and harmful stereotypes, one should not
+aim at making the distance between protected words and associated, and
+different class the same but rather to move it towards 1.
 
@@
 
+\leavevmode\hypertarget{ref-Manzini2019blackToCriminal}{}%
+Manzini, T., Lim, Y. C., Tsvetkov, Y., \& Black, A. W. (2019). Black is
+to criminal as caucasian is to police: Detecting and removing multiclass
+bias in word embeddings. Retrieved from
+\url{http://arxiv.org/abs/1904.04047}
+
 \leavevmode\hypertarget{ref-Nissim2020fair}{}%

### 2021-07-12 paperDraft/paperDraft.Rmd
Develops article

@@
 # Some methodological problems
+
+\textbf{1. Selection of attributes}
+
+The attributes are taken from different sources, there is no principled justification for their choice. From our analysis it will become clear that the list is rather uneven.
+
+There is no mention of methodology for deciding on the number of attributes necessary to decide a hypothesis on the given size of dataset.
+There are however some ways to estimate sample sizes needed to ensure that the results are meaningful if the effect is present.
+Our research will show that the numbers used
+are rather insufficient.
+
+\textbf{2. No attention to distributions and details}
+
+The authors use the mean of mean average cosine similarities to measure multi-class similarity between protected word and harmful stereotypes.
+They average the results until they obtain one final value to represent the mean cosine distance between protected word from a given class and the attributes of that class.
+If one takes closer look at the individual values that are taken for the calculations it turns out that there is a bunch of outliers and surprisingly dissimilar words.
+We approach this issue by providing the tables and visualizations of individual cosine distances to make sure that we obtain a proper insight into the data.
+
+\textbf{3. Hiding the impact of uncertainty}
+
+With such a method the uncertainty involved is not really considered which makes it even more difficult to give reasonable interpretations of the results.
+We propose the use of Bayesian method to obtain some understanding of the influence the uncertainty has on the interpretation of final results.
+
+\textbf{4. No word class distinction and no control group}
+
+In the original paper, words from all three religions were compared against all of the stereotypes, which means that there was no distinction between cases in which the stereotype is associated with a given religion, as opposed to the situation in which it is associated with another one.
+Not all of the stereotypical words have to be considered as harmful for all of the religions.
+One should investigate the religions separately as some of them may have stronger harmful associations that others.
+One should also include control groups to have a way of comparing the stereotypical results with neutral or human-like words.
+Later in the text we will explain in details reasons for introducing control groups.
+In our analysis, we distinguish between stereotypes associated with a given group, stereotypes associated with different groups, and control groups: neutral words and stereotypes-free human predicates.
+
+\textbf{5. Interpreting the results}
+
+Assuming for a moment that the value of multi-class cosine distance is correct, one may question the interpretation.
+@Manzini2019blackToCriminal summarize the averages of cosine distance per group (gender, race, religion).
+For now let us focus now on analyzing the values relating to religious biases.
+Here is the relevant fragment of table:
+
+Religion Debiasing  | MAC
+------------- | -------------
+Biased        | 0.859
+Hard Debiased | 0.934
+Soft Debiased ($\lambda$ = 0.2) | 0.894
+
+MAC stand for mean average cosine similarity, although in reality the the table contains mean distances.
+What may attract attention is the fact that the value of cosine distance in "Biased" category is already quite high even before debiasing.
+High cosine distance indicates low cosine similarity between values.
+One could think that average cosine similarity equal to approximately 0.141 is not significant enough to consider it as bias.
+However the authors aim to mitigate "biases" in vectors with such great distance to make it even larger.
+Methodologically the question is, on what basis is this small similarity still considered as a proof of the presence of bias, and whether these small changes are meaningful.
+This is in general the problem of scale and the lack of universal intervals.
+In contrast, statistical intervals will help us decide whether a given cosine similarity is high enough to consider the words to be more similar than if we chose them at random.
+We will use highest posterior density intervals, in line with Bayesian methodology.
+
+\textbf{6. The curse of dimensionality}
+
+In our case, the curse of dimensionality may take place when there is an increase in the volume of data that results in adding extra dimensions to the Euclidean space. According to the article 
+[Curse of dimensionality at analyticsindiamag.com](https://analyticsindiamag.com/curse-of-dimensionality-and-what-beginners-should-do-to-overcome-it/) as the number of features increases, it may be harder and harder to obtain useful information from the data using the available algorithms.
+One may notice that more data should contribute to greater amount of information, but more information also means greater risk of noise and distractions in data.
+At the same time, modern solutions are often adapted to smaller dimensions and their results in higher ones are not intuitive, or may be prone to error.
+
+Using cosine similarity in high dimensions in word embeddings may also be prone to the curse of dimensionality.
+According to @Venkat2018Curse there are reasons to consider this phenomenon when searching for word similarities in higher dimensions.
+An experiment is conducted that aims at showing how the similarity values and variation change as the number of dimensions increases.
+The hypothesis made in the paper states that two things will happen as the number of dimensions increase.
+First, the effort required to measure cosine similarity will be greater, and two, the similarity between data will blur out and have less variation.
+The authors generate random points with increasing number of dimensions where each dimension of a data point is given a value between 0 and 1.
+Then they pick one vector at random from each dimension class and calculate the cosine similarity between the chosen vector and the rest of the data.
+Then they check how the variation of values changes as the number of dimensions increases.
+It seems like the more dimensions there are, the smaller the variance and therefore it is less obvious how to interpret the resulting cosine similarities.
+Maybe the scale should be adjusted to the number of dimensions and variance so that it still gives us sensible information about data.
+Yet, according to some articles cosine similarity in high dimensions is not reliable enough as it may be the case that choosing words at random may result in similar values as when picking them consciously.
+
+![curse of dimensionality,  number of dimensions on the x axis,  standard deviation of similarity on the y axis](../images/curseOfDimensionality.png)
 
@@
 One may verify if the dataset-level coefficients changed after the debiasing method used in @Manzini2019blackToCriminal. 
+In some cases there is a minor change to the right, the mean value is closer to 0.
+The `associated` and `differen` attributes are now more overlapping than before.
+However, the uncertainty intervals are still high and therefore the significance of the change cannot be considered as enormous.
+The surprising obervation is that in the case of `Gender` and `Race` groups, the neutral words shifted to the left, further from the neutral position.
 

### 2021-07-12 paperDraft/paperDraft.tex
Develops article

@@
 \newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
 \usepackage{graphicx}
@@
 
+\textbf{1. Selection of attributes}
+
+The attributes are taken from different sources, there is no principled
+justification for their choice. From our analysis it will become clear
+that the list is rather uneven.
+
+There is no mention of methodology for deciding on the number of
+attributes necessary to decide a hypothesis on the given size of
+dataset. There are however some ways to estimate sample sizes needed to
+ensure that the results are meaningful if the effect is present. Our
+research will show that the numbers used are rather insufficient.
+
+\textbf{2. No attention to distributions and details}
+
+The authors use the mean of mean average cosine similarities to measure
+multi-class similarity between protected word and harmful stereotypes.
+They average the results until they obtain one final value to represent
+the mean cosine distance between protected word from a given class and
+the attributes of that class. If one takes closer look at the individual
+values that are taken for the calculations it turns out that there is a
+bunch of outliers and surprisingly dissimilar words. We approach this
+issue by providing the tables and visualizations of individual cosine
+distances to make sure that we obtain a proper insight into the data.
+
+\textbf{3. Hiding the impact of uncertainty}
+
+With such a method the uncertainty involved is not really considered
+which makes it even more difficult to give reasonable interpretations of
+the results. We propose the use of Bayesian method to obtain some
+understanding of the influence the uncertainty has on the interpretation
+of final results.
+
+\textbf{4. No word class distinction and no control group}
+
+In the original paper, words from all three religions were compared
+against all of the stereotypes, which means that there was no
+distinction between cases in which the stereotype is associated with a
+given religion, as opposed to the situation in which it is associated
+with another one. Not all of the stereotypical words have to be
+considered as harmful for all of the religions. One should investigate
+the religions separately as some of them may have stronger harmful
+associations that others. One should also include control groups to have
+a way of comparing the stereotypical results with neutral or human-like
+words. Later in the text we will explain in details reasons for
+introducing control groups. In our analysis, we distinguish between
+stereotypes associated with a given group, stereotypes associated with
+different groups, and control groups: neutral words and stereotypes-free
+human predicates.
+
+\textbf{5. Interpreting the results}
+
+Assuming for a moment that the value of multi-class cosine distance is
+correct, one may question the interpretation. Manzini, Lim, Tsvetkov, \&
+Black (2019) summarize the averages of cosine distance per group
+(gender, race, religion). For now let us focus now on analyzing the
+values relating to religious biases. Here is the relevant fragment of
+table:
+
+\begin{longtable}[]{@{}ll@{}}
+\toprule
+Religion Debiasing & MAC \\
+\midrule
+\endhead
+Biased & 0.859 \\
+Hard Debiased & 0.934 \\
+Soft Debiased (\(\lambda\) = 0.2) & 0.894 \\
+\bottomrule
+\end{longtable}
+
+MAC stand for mean average cosine similarity, although in reality the
+the table contains mean distances. What may attract attention is the
+fact that the value of cosine distance in ``Biased'' category is already
+quite high even before debiasing. High cosine distance indicates low
+cosine similarity between values. One could think that average cosine
+similarity equal to approximately 0.141 is not significant enough to
+consider it as bias. However the authors aim to mitigate ``biases'' in
+vectors with such great distance to make it even larger.
+Methodologically the question is, on what basis is this small similarity
+still considered as a proof of the presence of bias, and whether these
+small changes are meaningful. This is in general the problem of scale
+and the lack of universal intervals. In contrast, statistical intervals
+will help us decide whether a given cosine similarity is high enough to
+consider the words to be more similar than if we chose them at random.
+We will use highest posterior density intervals, in line with Bayesian
+methodology.
+
+\textbf{6. The curse of dimensionality}
+
+In our case, the curse of dimensionality may take place when there is an
+increase in the volume of data that results in adding extra dimensions
+to the Euclidean space. According to the article
+\href{https://analyticsindiamag.com/curse-of-dimensionality-and-what-beginners-should-do-to-overcome-it/}{Curse
+of dimensionality at analyticsindiamag.com} as the number of features
+increases, it may be harder and harder to obtain useful information from
+the data using the available algorithms. One may notice that more data
+should contribute to greater amount of information, but more information
+also means greater risk of noise and distractions in data. At the same
+time, modern solutions are often adapted to smaller dimensions and their
+results in higher ones are not intuitive, or may be prone to error.
+
+Using cosine similarity in high dimensions in word embeddings may also
+be prone to the curse of dimensionality. According to Venkat (2018)
+there are reasons to consider this phenomenon when searching for word
+similarities in higher dimensions. An experiment is conducted that aims
+at showing how the similarity values and variation change as the number
+of dimensions increases. The hypothesis made in the paper states that
+two things will happen as the number of dimensions increase. First, the
+effort required to measure cosine similarity will be greater, and two,
+the similarity between data will blur out and have less variation. The
+authors generate random points with increasing number of dimensions
+where each dimension of a data point is given a value between 0 and 1.
+Then they pick one vector at random from each dimension class and
+calculate the cosine similarity between the chosen vector and the rest
+of the data. Then they check how the variation of values changes as the
+number of dimensions increases. It seems like the more dimensions there
+are, the smaller the variance and therefore it is less obvious how to
+interpret the resulting cosine similarities. Maybe the scale should be
+adjusted to the number of dimensions and variance so that it still gives
+us sensible information about data. Yet, according to some articles
+cosine similarity in high dimensions is not reliable enough as it may be
+the case that choosing words at random may result in similar values as
+when picking them consciously.
+
+\begin{figure}
+\centering
+\includegraphics{../images/curseOfDimensionality.png}
+\caption{curse of dimensionality, number of dimensions on the x axis,
+standard deviation of similarity on the y axis}
+\end{figure}
+
 \hypertarget{the-problem-with-means-of-means}{%
@@
 
+One may verify if the dataset-level coefficients changed after the
+debiasing method used in Manzini, Lim, Tsvetkov, \& Black (2019). In
+some cases there is a minor change to the right, the mean value is
+closer to 0. The \texttt{associated} and \texttt{differen} attributes
+are now more overlapping than before. However, the uncertainty intervals
+are still high and therefore the significance of the change cannot be
+considered as enormous. The surprising obervation is that in the case of
+\texttt{Gender} and \texttt{Race} groups, the neutral words shifted to
+the left, further from the neutral position.
+
+\begin{center}
+\begin{figure}[!htb]\centering
+   \begin{minipage}{0.55\textwidth}
+  \includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
+   \end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/debiasedReligionRedditCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/genderCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/debiasedGenderRedditCoeffs.jpeg}
+   \end{minipage}
+
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/raceCoeffs.jpeg}
+\end{minipage}
+   \begin {minipage}{0.43\textwidth}
+    \includegraphics[width=7cm]{../images/debiasedRaceRedditCoeffs.jpeg}
+   \end{minipage}
+\end{figure}
+
+\end{center}
+
 When analyzing the Reddit debiased results one should investigate the
-change of the cosine distance values. It seems that not all of the
-protected words are treated equally when performing debiasing. In
-religion, the values for associated class moved towards 0 for most of
-the words except for the ones for Islam, where still the associated
-class has quite lower cosine distance than the different one. Similar
-situation takes place in Gender, where female words have still much
-lower cosine distance values for associated class than for the different
-one. In the case of male protected words it is mostly almost the same
-interval for associated and different. As the Gender data is quite
-specific as the attributes are not harmful adjectives but (in an ideal
-world) neutral professions, the aim (if we follow MAC assumptions)
+change of the individual cosine distance values as well. It seems that
+not all of the protected words are treated equally when performing
+debiasing. In religion, the values for associated class moved towards 0
+for most of the words except for the ones for Islam, where still the
+associated class has quite lower cosine distance than the different one.
+Similar situation takes place in Gender, where female words have still
+much lower cosine distance values for associated class than for the
+different one. In the case of male protected words it is mostly almost
+the same interval for associated and different. As the Gender data is
+quite specific as the attributes are not harmful adjectives but (in an
+ideal world) neutral professions, the aim (if we follow MAC assumptions)
 should actually be to make the cosine distance same for both female and

### 2021-07-13 paperDraft/paperDraft.Rmd
small style changes

@@
 ---
-title: "A bayesian method of cosine-based word2vec bias"
+title: "Taking uncertainty in word embedding bias estimation seriously: a  bayesian approach"
 author: "Alicja Dobrzeniecka and Rafal Urbaniak"
@@
 urlcolor: blue
-bibliography: cosineReferences1.bib
-csl: apa-5th-edition.csl
+bibliography: [../references/bibliographyRafal.bib]
+csl: [../references/apa-6th-edition.csl] 
 ---
@@
 
-The attributes are taken from different sources, there is no principled justification for their choice. From our analysis it will become clear that the list is rather uneven.
+Such approaches, however, are not unproblematic. For one thing, the word lists are often fairly small, and the papers do not discuss the impact of word list sizes on the uncertainty involved. 
+ We will argue that the word list sizes are too small to yield strong conclusions, given the variances involved.
 
-There is no mention of methodology for deciding on the number of attributes necessary to decide a hypothesis on the given size of dataset.
-There are however some ways to estimate sample sizes needed to ensure that the results are meaningful if the effect is present.
-Our research will show that the numbers used
-are rather insufficient.
-
-\textbf{2. No attention to distributions and details}
-
-The authors use the mean of mean average cosine similarities to measure multi-class similarity between protected word and harmful stereotypes.
-They average the results until they obtain one final value to represent the mean cosine distance between protected word from a given class and the attributes of that class.
-If one takes closer look at the individual values that are taken for the calculations it turns out that there is a bunch of outliers and surprisingly dissimilar words.
-We approach this issue by providing the tables and visualizations of individual cosine distances to make sure that we obtain a proper insight into the data.
+ Crucially, these approaches use  means of mean average cosine similarities to measure  similarity between protected word and harmful stereotypes.
+If one takes  a closer look at the individual values that are taken for the calculations, it turns out that there are quite a few  outliers and surprisingly dissimilar words.
+This issue will become transparent when  we inspect  the visualizations of individual cosine distances, following the idea that one of the first step to understand data is to look at it.
 
@@
 grid.arrange(t1d2Plot,t2d2Plot, t3d2Plot, t4d2Plot, ncol=2,   
-      top = textGrob("Different means, sd=0.05",gp=gpar(fontsize=15,font=1)))
+      top = textGrob("different means, sd=0.05",gp=gpar(fontsize=15,font=1)))
 ```
@@
 
-Let's analyse dataset-level coeﬀicients together with their highest posterior density intervals, for the three topic groups.
+Let's analyse dataset-level coefficients together with their highest posterior density intervals, for the three topic groups.
 We compare the results for two different word emebeddings corpuses - RedditL2 and GoogleNews.
 For Google embeddings the HPDIs
-for all class coeﬀicients (associated, different, human, and none) include zero.
+for all class coefficients (associated, different, human, and none) include zero.
 This can lead one to a conclusion that the impact for associated, different, human and neutral attribute is, when averaged, quite similar.
@@
 
-In the case of Reddit word embeddings the situation is similar although HPDI is below 0 for Gender class when looking at `associated` and `different` mean coeﬀicient.
+In the case of Reddit word embeddings the situation is similar although HPDI is below 0 for Gender class when looking at `associated` and `different` mean coefficient.
 This can suggest that there is indeed slightly stronger impact of these attributes.
-One should also notice how in general associated and different coeﬀicients have quite similar HPDI range, the highest observed absolute difference is equal to only 0.1.
+One should also notice how in general associated and different coefficients have quite similar HPDI range, the highest observed absolute difference is equal to only 0.1.
 This suggests that again the impact of associated attributes and different ones is not clear at first sight.
 Finally, let’s compare the HPDIs for Reddit and Reddit debiased datasets.
-For Religion and Race dataset there is a minor shift (in absolute values the highest change is equal to approximately 0.1) of the mean coeﬀicients towards zero.
+For Religion and Race dataset there is a minor shift (in absolute values the highest change is equal to approximately 0.1) of the mean coefficients towards zero.
 However for Gender dataset there is no significant change.
-This is of course a general look at the group coeﬀicients.
+This is of course a general look at the group coefficients.
 Let’s now analyze the individual words.

### 2021-07-13 paperDraft/paperDraft.tex
small style changes

@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}
-%
-\documentclass[
-  10pt,
-  dvipsnames,enabledeprecatedfontcommands]{scrartcl}
-\usepackage{amsmath,amssymb}
+\documentclass[10pt,dvipsnames,enabledeprecatedfontcommands]{scrartcl}
 \usepackage{lmodern}
+\usepackage{amssymb,amsmath}
 \usepackage{ifxetex,ifluatex}
+\usepackage{fixltx2e} % provides \textsubscript
 \ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
@@
 }{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
+\usepackage[unicode=true]{hyperref}
+\PassOptionsToPackage{usenames,dvipsnames}{color} % color is loaded by hyperref
 \hypersetup{
-  pdftitle={A bayesian method of cosine-based word2vec bias},
+            pdftitle={Taking uncertainty in word embedding bias estimation seriously: a bayesian approach},
             pdfauthor={Alicja Dobrzeniecka and Rafal Urbaniak},
@@
             linkcolor=Maroon,
-  filecolor=Maroon,
             citecolor=Blue,
             urlcolor=blue,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
+            breaklinks=true}
+\urlstyle{same}  % don't use monospace font for urls
 \usepackage{color}
@@
 \setcounter{secnumdepth}{5}
+% Redefines (sub)paragraphs to behave more like sections
+\ifx\paragraph\undefined\else
+\let\oldparagraph\paragraph
+\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+\let\oldsubparagraph\subparagraph
+\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
+
+% set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
+
 %\documentclass{article}
@@
 %\usepackage{polski}
-
+\usepackage[utf8]{inputenc}
 \usepackage{mathptmx}
@@
 %\bibliographystyle{apalike}
-\ifluatex
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
-  % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-
-\title{A bayesian method of cosine-based word2vec bias}
+
+\title{Taking uncertainty in word embedding bias estimation seriously: a
+bayesian approach}
 \author{Alicja Dobrzeniecka and Rafal Urbaniak}
@@
 
-\hypertarget{cosine-based-measures-of-bias}{%
 \section{Cosine-based measures of
-bias}\label{cosine-based-measures-of-bias}}
+bias}\label{cosine-based-measures-of-bias}
 
@@
 One of the first measures in the discussion has been developed by
-Bolukbasi, Chang, Zou, Saligrama, \& Kalai (2016). There, the gender
-bias of a word \(w\) is understood as its projection on the gender
-direction \(\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})\)
-(the gender direction is the top principal compontent of ten gender pair
+({\textbf{???}}). There, the gender bias of a word \(w\) is understood
+as its projection on the gender direction
+\(\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})\) (the
+gender direction is the top principal compontent of ten gender pair
 difference vectors). The underlying idea is that no bias is present if
@@
 as the average distance of the words in \(N\) from \(g\) (\(c\) is a
-parameter determining how strict we want to be): \begin{align}
+parameter determining how strict we want to be):
+
+\begin{align}
 \mathsf{directBias_c(N,g)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},g)\vert^c}{\vert N \vert }
@@
 
-The use of projections has been ciriticized for instance by Gonen \&
-Goldberg (2019), who point out that while gender-direction might be an
+The use of projections has been ciriticized for instance by
+({\textbf{???}}), who point out that while gender-direction might be an
 indicator of bias, it is only one possible manifestation of it, and
@@
 preserve similarity under debiasing meant to minimize projection-based
-bias.\footnote{Bolukbasi, Chang, Zou, Saligrama, \& Kalai (2016) use
-  also another method which involves analogies and their evaluations by
-  human users on Mechanical Turk. It is discussed and criticized in
-  (Nissim, Noord, \& Goot, 2020).}
-
-To measure bias in word embeddings, Caliskan, Bryson, \& Narayanan
-(2017) proposed the Word Embedding Association Test (WEAT). The idea is
-that the measure of biases between two sets of target words, \(X\) and
-\(Y\), (we call them protected words) should be quantified in terms of
-the cosine similarity between the protected words and attribute words
-coming from two sets of stereotype attribute words, \(A\) and \(B\)
-(we'll call them attributes). For instance, \(X\) might be a set of male
-names, \(Y\) a set of female names, \(A\) might contain stereotypically
-male-related career words, and \(B\) stereotypically female-related
-family words. WEAT is a modification of the Implicit Association Test
-(IAT) (Nosek, Banaji, \& Greenwald, 2002) used in psychology and uses
-almost the same word sets, allowing for a \emph{prima facie} sensible
-comparison with bias in humans. If the person's attitude towards given
-pair of concept is to be interpreted as neutral, there should be no
-noticeable task completion time difference, and the final value from the
-formula should be around 0. Let \(f\) be a similarity measure (usually,
-cosine similarity). The association difference for a term \(t\) is:
+bias.\footnote{({\textbf{???}}) use also another method which involves
+  analogies and their evaluations by human users on Mechanical Turk. It
+  is discussed and criticized in ({\textbf{???}}).}
+
+To measure bias in word embeddings, ({\textbf{???}}) proposed the Word
+Embedding Association Test (WEAT). The idea is that the measure of
+biases between two sets of target words, \(X\) and \(Y\), (we call them
+protected words) should be quantified in terms of the cosine similarity
+between the protected words and attribute words coming from two sets of
+stereotype attribute words, \(A\) and \(B\) (we'll call them
+attributes). For instance, \(X\) might be a set of male names, \(Y\) a
+set of female names, \(A\) might contain stereotypically male-related
+career words, and \(B\) stereotypically female-related family words.
+WEAT is a modification of the Implicit Association Test (IAT)
+({\textbf{???}}) used in psychology and uses almost the same word sets,
+allowing for a \emph{prima facie} sensible comparison with bias in
+humans. If the person's attitude towards given pair of concept is to be
+interpreted as neutral, there should be no noticeable task completion
+time difference, and the final value from the formula should be around
+0. Let \(f\) be a similarity measure (usually, cosine similarity). The
+association difference for a term \(t\) is:
+
 \begin{align}
 s(t,A,B) & = \frac{\sum_{a\in A}f(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}f(t,b)}{\vert B\vert}
-\end{align} \noindent then, the association difference between \(A\) a
-\(B\) is: \begin{align}
+\end{align}
+
+\noindent then, the association difference between \(A\) a \(B\) is:
+
+\begin{align}
 s(X,Y,A,B) & = \sum_{x\in X} s(x,A,B) -  \sum_{y\in Y} s(y,A,B)
-\end{align} \noindent \(s(X,Y,A,B)\) is the statistic used in the
-signifcance test, and the \(p\)-value obtained by bootstrapping: it is
-the frequency of \(s(X_i,Y_i,A,B)>s(X,Y,A,B)\) for all equally sized
-partitions \(X_i, Y_i\) of \(X\cup Y\). The effect size is computed by
-normalizing the difference in means as follows: \begin{align}
+\end{align}
+
+\noindent
+\(s(X,Y,A,B)\) is the statistic used in the signifcance test, and the
+\(p\)-value obtained by bootstrapping: it is the frequency of
+\(s(X_i,Y_i,A,B)>s(X,Y,A,B)\) for all equally sized partitions
+\(X_i, Y_i\) of \(X\cup Y\). The effect size is computed by normalizing
+the difference in means as follows:
+
+\begin{align}
 bias(A,B) & = \frac{
@@
 
-Caliskan, Bryson, \& Narayanan (2017) show that significant
-biases---thus measured--- similar to the ones discovered by IAT can be
-discovered in word embeddings. Lauscher \& Glavas (2019) extended the
-methodology to a multilingual and cross-lingual setting, arguing that
-using Euclidean distance instead of similarity does not make much
-difference, while the bias effects vary greatly across embedding models
-(interestingly, with social media-text trained embeddings being less
-biased than those based on Wikipedia).
-
-A similar methodology is employed by Garg, Schiebinger, Jurafsky, \& Zou
-(2018), who employ word embeddings trained on corpora from different
-decades to study the shifts in various biases. For instance, to compute
-the occupational embeddings bias for women the authors first compute the
-average vector of vector emeddings of words that represent women
-(e.g.~``she,'' ``female''), then calculate the Euclidean distance
-between this mean vector and words for occupations. Then they take the
-mean of these distances and subtract from it the analogously obtained
-mean for the average vector of vector embeddings of words that represent
-men. Formally they take the relative norm distance between \(X\) and
-\(Y\) to be: \begin{align}
+({\textbf{???}}) show that significant biases---thus measured--- similar
+to the ones discovered by IAT can be discovered in word embeddings.
+({\textbf{???}}) extended the methodology to a multilingual and
+cross-lingual setting, arguing that using Euclidean distance instead of
+similarity does not make much difference, while the bias effects vary
+greatly across embedding models (interestingly, with social media-text
+trained embeddings being less biased than those based on Wikipedia).
+
+A similar methodology is employed by ({\textbf{???}}), who employ word
+embeddings trained on corpora from different decades to study the shifts
+in various biases. For instance, to compute the occupational embeddings
+bias for women the authors first compute the average vector of vector
+emeddings of words that represent women (e.g. ``she'', ``female''), then
+calculate the Euclidean distance between this mean vector and words for
+occupations. Then they take the mean of these distances and subtract
+from it the analogously obtained mean for the average vector of vector
+embeddings of words that represent men. Formally they take the relative
+norm distance between \(X\) and \(Y\) to be:
+
+\begin{align}
 \textsf{relative norm distance} & = \sum_{v_m\in M} \vert \vert v_m - v_X\vert \vert_2 - \vert v_m - v_Y\vert \vert_2
-\end{align} \noindent where the norm used is Euclidean, and \(v_X\) and
-\(x_Y\) are average vectors for sets \(X\) and \(Y\) respectively.
+\end{align}
+
+\noindent where the norm used is Euclidean, and \(v_X\) and \(x_Y\) are
+average vectors for sets \(X\) and \(Y\) respectively.
 
@@
 and let each \(A_j\in A\) be a set of attributes stereotypically
-associated with a protected word). Then: \begin{align}
+associated with a protected word). Then:
+
+\begin{align}
 S(t_i, A_j) & = \frac{1}{\vert A_j\vert}\sum_{a\in A_j}\mathsf{cos}(t,a) \\
 MAC(T,A) & = \frac{1}{\vert T \vert \,\vert A\vert}\sum_{t_i \in T }\sum_{A_j \in A} S(t_i,A_j)
-\end{align} That is, for each protected word \$T and each attribute
-class, they first take the mean for this protected word and all
-attributes in a given attribute class, and then take the mean of thus
-obtained means for all the protected words. The t-tests they employ are
-run on average cosines used to calculate MAC.
-
-\hypertarget{some-methodological-problems}{%
+\end{align}
+
+That is, for each protected word \$T and each attribute class, they
+first take the mean for this protected word and all attributes in a
+given attribute class, and then take the mean of thus obtained means for
+all the protected words. The t-tests they employ are run on average
+cosines used to calculate MAC.
+
 \section{Some methodological
-problems}\label{some-methodological-problems}}
+problems}\label{some-methodological-problems}
 
@@
 Assuming for a moment that the value of multi-class cosine distance is
-correct, one may question the interpretation. Manzini, Lim, Tsvetkov, \&
-Black (2019) summarize the averages of cosine distance per group
-(gender, race, religion). For now let us focus now on analyzing the
-values relating to religious biases. Here is the relevant fragment of
-table:
+correct, one may question the interpretation. Manzini et al. (2019)
+summarize the averages of cosine distance per group (gender, race,
+religion). For now let us focus now on analyzing the values relating to
+religious biases. Here is the relevant fragment of table:
 
@@
 
-\hypertarget{the-problem-with-means-of-means}{%
 \section{The problem with means of
-means}\label{the-problem-with-means-of-means}}
+means}\label{the-problem-with-means-of-means}
 
@@
 
-To illustrate let's employ the formulas used by Caliskan, Bryson, \&
-Narayanan (2017) in a simple example. Conceptually, all such tests come
-up with rather short lists of protected words and rather short lists of
-stereotypical attributes. Clearly, these are not complete list. So let's
-treat them as samples from richer pools of stereotypical predicates and
-let's take the uncertainty and variation involved seriously.
+To illustrate let's employ the formulas used by ({\textbf{???}}) in a
+simple example. Conceptually, all such tests come up with rather short
+lists of protected words and rather short lists of stereotypical
+attributes. Clearly, these are not complete list. So let's treat them as
+samples from richer pools of stereotypical predicates and let's take the
+uncertainty and variation involved seriously.
 
@@
 \begin{Highlighting}[]
-\FunctionTok{set.seed}\NormalTok{(}\DecValTok{123}\NormalTok{)}
-\NormalTok{t1 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
-\NormalTok{t2 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
-\NormalTok{t3 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
-\NormalTok{t4 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{123}\NormalTok{)}
+\NormalTok{t1 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\NormalTok{t2 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\NormalTok{t3 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\NormalTok{t4 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
 \end{Highlighting}
@@
 a strong bias here? We wouldn't think so. But now let's run the
-calculations from (Caliskan, Bryson, \& Narayanan, 2017).
-
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{s }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{ (table)\{ }\FunctionTok{mean}\NormalTok{(table}\SpecialCharTok{$}\NormalTok{A) }\SpecialCharTok{{-}} \FunctionTok{mean}\NormalTok{(table}\SpecialCharTok{$}\NormalTok{B)\}}
-\NormalTok{factor }\OtherTok{\textless{}{-}} \FunctionTok{sd}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1),}\FunctionTok{s}\NormalTok{(t2),}\FunctionTok{s}\NormalTok{(t3),}\FunctionTok{s}\NormalTok{(t4)))}
-\NormalTok{numerator }\OtherTok{\textless{}{-}}  \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1),}\FunctionTok{s}\NormalTok{(t2)) }\SpecialCharTok{{-}} \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t3),}\FunctionTok{s}\NormalTok{(t4))}
-\FunctionTok{print}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{factor =}\NormalTok{ factor,}\AttributeTok{numerator =}\NormalTok{ numerator, }\AttributeTok{bias =}\NormalTok{ numerator }\SpecialCharTok{/}\NormalTok{ factor))}
+calculations from ({\textbf{???}}).
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{s <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{ (table)\{ }\KeywordTok{mean}\NormalTok{(table}\OperatorTok{$}\NormalTok{A) }\OperatorTok{-}\StringTok{ }\KeywordTok{mean}\NormalTok{(table}\OperatorTok{$}\NormalTok{B)\}}
+\NormalTok{factor <-}\StringTok{ }\KeywordTok{sd}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1),}\KeywordTok{s}\NormalTok{(t2),}\KeywordTok{s}\NormalTok{(t3),}\KeywordTok{s}\NormalTok{(t4)))}
+\NormalTok{numerator <-}\StringTok{  }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1),}\KeywordTok{s}\NormalTok{(t2)) }\OperatorTok{-}\StringTok{ }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t3),}\KeywordTok{s}\NormalTok{(t4))}
+\KeywordTok{print}\NormalTok{(}\KeywordTok{list}\NormalTok{(}\DataTypeTok{factor =}\NormalTok{ factor,}\DataTypeTok{numerator =}\NormalTok{ numerator, }\DataTypeTok{bias =}\NormalTok{ numerator }\OperatorTok{/}\StringTok{ }\NormalTok{factor))}
 \end{Highlighting}
@@
 calculated effect size is 1.82, whereas the largest effect size reported
-by Caliskan, Bryson, \& Narayanan (2017) is 1.81!
+by ({\textbf{???}}) is 1.81!
 
@@
 
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{biasesNull }\OtherTok{\textless{}{-}} \FunctionTok{numeric}\NormalTok{(}\DecValTok{10000}\NormalTok{)}
-\ControlFlowTok{for}\NormalTok{(i }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\DecValTok{10000}\NormalTok{)\{}
-\NormalTok{t1 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
-\NormalTok{t2 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
-\NormalTok{t3 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
-\NormalTok{t4 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
-
-\NormalTok{factor }\OtherTok{\textless{}{-}} \FunctionTok{sd}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1),}\FunctionTok{s}\NormalTok{(t2),}\FunctionTok{s}\NormalTok{(t3),}\FunctionTok{s}\NormalTok{(t4)))}
-\NormalTok{numerator }\OtherTok{\textless{}{-}}  \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1),}\FunctionTok{s}\NormalTok{(t2)) }\SpecialCharTok{{-}} \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t3),}\FunctionTok{s}\NormalTok{(t4))}
-\NormalTok{biasesNull[i]  }\OtherTok{\textless{}{-}}\NormalTok{ numerator }\SpecialCharTok{/}\NormalTok{ factor}
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{biasesNull <-}\StringTok{ }\KeywordTok{numeric}\NormalTok{(}\DecValTok{10000}\NormalTok{)}
+\ControlFlowTok{for}\NormalTok{(i }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\DecValTok{10000}\NormalTok{)\{}
+\NormalTok{t1 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\NormalTok{t2 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\NormalTok{t3 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\NormalTok{t4 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+
+\NormalTok{factor <-}\StringTok{ }\KeywordTok{sd}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1),}\KeywordTok{s}\NormalTok{(t2),}\KeywordTok{s}\NormalTok{(t3),}\KeywordTok{s}\NormalTok{(t4)))}
+\NormalTok{numerator <-}\StringTok{  }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1),}\KeywordTok{s}\NormalTok{(t2)) }\OperatorTok{-}\StringTok{ }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t3),}\KeywordTok{s}\NormalTok{(t4))}
+\NormalTok{biasesNull[i]  <-}\StringTok{ }\NormalTok{numerator }\OperatorTok{/}\StringTok{ }\NormalTok{factor}
 \NormalTok{\}}
-\FunctionTok{ggplot}\NormalTok{()}\SpecialCharTok{+}\FunctionTok{geom\_histogram}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{biasesNull, }\AttributeTok{y =}\NormalTok{ ..density..), }\AttributeTok{alpha =} \FloatTok{0.6}\NormalTok{, }\AttributeTok{bins=}\DecValTok{50}\NormalTok{)}\SpecialCharTok{+}
-  \FunctionTok{theme\_tufte}\NormalTok{()}\SpecialCharTok{+}\FunctionTok{labs}\NormalTok{(}\AttributeTok{title=}\StringTok{"10k biases for identical means and sd =.05"}\NormalTok{)}\SpecialCharTok{+} \FunctionTok{xlab}\NormalTok{(}\StringTok{"bias"}\NormalTok{)}
+\KeywordTok{ggplot}\NormalTok{()}\OperatorTok{+}\KeywordTok{geom_histogram}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x=}\NormalTok{biasesNull, }\DataTypeTok{y =}\NormalTok{ ..density..), }\DataTypeTok{alpha =} \FloatTok{0.6}\NormalTok{, }\DataTypeTok{bins=}\DecValTok{50}\NormalTok{)}\OperatorTok{+}
+\StringTok{  }\KeywordTok{theme_tufte}\NormalTok{()}\OperatorTok{+}\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title=}\StringTok{"10k biases for identical means and sd =.05"}\NormalTok{)}\OperatorTok{+}\StringTok{ }\KeywordTok{xlab}\NormalTok{(}\StringTok{"bias"}\NormalTok{)}
 \end{Highlighting}
@@
 \begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/unnamed-chunk-4-1} \end{center}
+
 \normalsize
@@
 \begin{Highlighting}[]
-\FunctionTok{set.seed}\NormalTok{(}\DecValTok{124}\NormalTok{)}
-\NormalTok{t1v }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
-\NormalTok{t2v }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
-\NormalTok{t3v }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
-\NormalTok{t4v }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{124}\NormalTok{)}
+\NormalTok{t1v <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
+\NormalTok{t2v <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
+\NormalTok{t3v <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
+\NormalTok{t4v <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
 \end{Highlighting}
@@
 a strong bias here? We wouldn't think so. But now let's run the
-calculations from (Caliskan, Bryson, \& Narayanan, 2017).
-
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{factorV }\OtherTok{\textless{}{-}} \FunctionTok{sd}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1v),}\FunctionTok{s}\NormalTok{(t2v),}\FunctionTok{s}\NormalTok{(t3v),}\FunctionTok{s}\NormalTok{(t4v)))}
-\NormalTok{numeratorV }\OtherTok{\textless{}{-}}  \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1v),}\FunctionTok{s}\NormalTok{(t2v)) }\SpecialCharTok{{-}} \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t3v),}\FunctionTok{s}\NormalTok{(t4v))}
-\FunctionTok{print}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{factor =}\NormalTok{ factorV,}\AttributeTok{numerator =}\NormalTok{ numeratorV, }\AttributeTok{bias =}\NormalTok{ numeratorV }\SpecialCharTok{/}\NormalTok{ factorV))}
+calculations from ({\textbf{???}}).
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{factorV <-}\StringTok{ }\KeywordTok{sd}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1v),}\KeywordTok{s}\NormalTok{(t2v),}\KeywordTok{s}\NormalTok{(t3v),}\KeywordTok{s}\NormalTok{(t4v)))}
+\NormalTok{numeratorV <-}\StringTok{  }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1v),}\KeywordTok{s}\NormalTok{(t2v)) }\OperatorTok{-}\StringTok{ }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t3v),}\KeywordTok{s}\NormalTok{(t4v))}
+\KeywordTok{print}\NormalTok{(}\KeywordTok{list}\NormalTok{(}\DataTypeTok{factor =}\NormalTok{ factorV,}\DataTypeTok{numerator =}\NormalTok{ numeratorV, }\DataTypeTok{bias =}\NormalTok{ numeratorV }\OperatorTok{/}\StringTok{ }\NormalTok{factorV))}
 \end{Highlighting}
@@
 
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{set.seed}\NormalTok{(}\DecValTok{124}\NormalTok{)}
-
-\NormalTok{biasesLowVariance }\OtherTok{\textless{}{-}} \FunctionTok{numeric}\NormalTok{(}\DecValTok{10000}\NormalTok{)}
-\ControlFlowTok{for}\NormalTok{(i }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\DecValTok{10000}\NormalTok{)\{}
-\NormalTok{t1v }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
-\NormalTok{t2v }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
-\NormalTok{t3v }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
-\NormalTok{t4v }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
-
-\NormalTok{factorV }\OtherTok{\textless{}{-}} \FunctionTok{sd}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1v),}\FunctionTok{s}\NormalTok{(t2v),}\FunctionTok{s}\NormalTok{(t3v),}\FunctionTok{s}\NormalTok{(t4v)))}
-
-\NormalTok{numeratorV }\OtherTok{\textless{}{-}}  \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1v),}\FunctionTok{s}\NormalTok{(t2v)) }\SpecialCharTok{{-}} \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t3v),}\FunctionTok{s}\NormalTok{(t4v))}
-
-\NormalTok{biasesLowVariance[i] }\OtherTok{\textless{}{-}}\NormalTok{ numeratorV }\SpecialCharTok{/}\NormalTok{ factorV}
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{124}\NormalTok{)}
+
+\NormalTok{biasesLowVariance <-}\StringTok{ }\KeywordTok{numeric}\NormalTok{(}\DecValTok{10000}\NormalTok{)}
+\ControlFlowTok{for}\NormalTok{(i }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\DecValTok{10000}\NormalTok{)\{}
+\NormalTok{t1v <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
+\NormalTok{t2v <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
+\NormalTok{t3v <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
+\NormalTok{t4v <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
+
+\NormalTok{factorV <-}\StringTok{ }\KeywordTok{sd}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1v),}\KeywordTok{s}\NormalTok{(t2v),}\KeywordTok{s}\NormalTok{(t3v),}\KeywordTok{s}\NormalTok{(t4v)))}
+
+\NormalTok{numeratorV <-}\StringTok{  }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1v),}\KeywordTok{s}\NormalTok{(t2v)) }\OperatorTok{-}\StringTok{ }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t3v),}\KeywordTok{s}\NormalTok{(t4v))}
+
+\NormalTok{biasesLowVariance[i] <-}\StringTok{ }\NormalTok{numeratorV }\OperatorTok{/}\StringTok{ }\NormalTok{factorV}
 \NormalTok{\}}
-\FunctionTok{ggplot}\NormalTok{()}\SpecialCharTok{+}\FunctionTok{geom\_histogram}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{biasesLowVariance, }\AttributeTok{y =}\NormalTok{ ..density..), }\AttributeTok{alpha =} \FloatTok{0.6}\NormalTok{, }\AttributeTok{bins=}\DecValTok{50}\NormalTok{)}\SpecialCharTok{+}
-  \FunctionTok{theme\_tufte}\NormalTok{()}\SpecialCharTok{+}\FunctionTok{labs}\NormalTok{(}\AttributeTok{title=}\StringTok{"10k biases for identical means and sd =.001"}\NormalTok{)}\SpecialCharTok{+} \FunctionTok{xlab}\NormalTok{(}\StringTok{"bias"}\NormalTok{)}
+\KeywordTok{ggplot}\NormalTok{()}\OperatorTok{+}\KeywordTok{geom_histogram}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x=}\NormalTok{biasesLowVariance, }\DataTypeTok{y =}\NormalTok{ ..density..), }\DataTypeTok{alpha =} \FloatTok{0.6}\NormalTok{, }\DataTypeTok{bins=}\DecValTok{50}\NormalTok{)}\OperatorTok{+}
+\StringTok{  }\KeywordTok{theme_tufte}\NormalTok{()}\OperatorTok{+}\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title=}\StringTok{"10k biases for identical means and sd =.001"}\NormalTok{)}\OperatorTok{+}\StringTok{ }\KeywordTok{xlab}\NormalTok{(}\StringTok{"bias"}\NormalTok{)}
 \end{Highlighting}
@@
 \begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/unnamed-chunk-8-1} \end{center}
+
 \normalsize
@@
 
-\vspace{1mm}
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{766}\NormalTok{)}
+\NormalTok{t1d2 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\NormalTok{t2d2 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\NormalTok{t3d2 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\NormalTok{t4d2 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+\vspace{1mm} \footnotesize
+
+\begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/unnamed-chunk-10-1} \end{center}
+
+\normalsize
+
+\noindent  Is this clear difference mirrored in the calculations?
+
 \footnotesize
@@
 \begin{Highlighting}[]
-\FunctionTok{set.seed}\NormalTok{(}\DecValTok{766}\NormalTok{)}
-\NormalTok{t1d2 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
-\NormalTok{t2d2 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
-\NormalTok{t3d2 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
-\NormalTok{t4d2 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\normalsize
-
-\vspace{1mm}
-\footnotesize
-
-\begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/unnamed-chunk-10-1} \end{center}
-\normalsize
-
-\noindent  Is this clear difference mirrored in the calculations?
-
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{factorD2 }\OtherTok{\textless{}{-}} \FunctionTok{sd}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1d2),}\FunctionTok{s}\NormalTok{(t2d2),}\FunctionTok{s}\NormalTok{(t3d2),}\FunctionTok{s}\NormalTok{(t4d2)))}
-\NormalTok{numeratorD2 }\OtherTok{\textless{}{-}}  \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1d2),}\FunctionTok{s}\NormalTok{(t2d2)) }\SpecialCharTok{{-}} \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t3d2),}\FunctionTok{s}\NormalTok{(t4d2))}
-\NormalTok{biasD2 }\OtherTok{\textless{}{-}}\NormalTok{ numeratorD2 }\SpecialCharTok{/}\NormalTok{ factorD2}
+\NormalTok{factorD2 <-}\StringTok{ }\KeywordTok{sd}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1d2),}\KeywordTok{s}\NormalTok{(t2d2),}\KeywordTok{s}\NormalTok{(t3d2),}\KeywordTok{s}\NormalTok{(t4d2)))}
+\NormalTok{numeratorD2 <-}\StringTok{  }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1d2),}\KeywordTok{s}\NormalTok{(t2d2)) }\OperatorTok{-}\StringTok{ }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t3d2),}\KeywordTok{s}\NormalTok{(t4d2))}
+\NormalTok{biasD2 <-}\StringTok{ }\NormalTok{numeratorD2 }\OperatorTok{/}\StringTok{ }\NormalTok{factorD2}
 \NormalTok{biasD2}
@@
 
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{biasesD2 }\OtherTok{\textless{}{-}} \FunctionTok{numeric}\NormalTok{(}\DecValTok{10000}\NormalTok{)}
-\ControlFlowTok{for}\NormalTok{(i }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\DecValTok{10000}\NormalTok{)\{}
-\NormalTok{t1d2 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
-\NormalTok{t2d2 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
-\NormalTok{t3d2 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
-\NormalTok{t4d2 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
-
-\NormalTok{factorD2 }\OtherTok{\textless{}{-}} \FunctionTok{sd}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1d2),}\FunctionTok{s}\NormalTok{(t2d2),}\FunctionTok{s}\NormalTok{(t3d2),}\FunctionTok{s}\NormalTok{(t4d2)))}
-\NormalTok{numeratorD2 }\OtherTok{\textless{}{-}}  \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1d2),}\FunctionTok{s}\NormalTok{(t2d2)) }\SpecialCharTok{{-}} \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t3d2),}\FunctionTok{s}\NormalTok{(t4d2))}
-\NormalTok{biasesD2[i] }\OtherTok{\textless{}{-}}\NormalTok{ numeratorD2}\SpecialCharTok{/}\NormalTok{factorD2}
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{biasesD2 <-}\StringTok{ }\KeywordTok{numeric}\NormalTok{(}\DecValTok{10000}\NormalTok{)}
+\ControlFlowTok{for}\NormalTok{(i }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\DecValTok{10000}\NormalTok{)\{}
+\NormalTok{t1d2 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\NormalTok{t2d2 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\NormalTok{t3d2 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+\NormalTok{t4d2 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
+
+\NormalTok{factorD2 <-}\StringTok{ }\KeywordTok{sd}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1d2),}\KeywordTok{s}\NormalTok{(t2d2),}\KeywordTok{s}\NormalTok{(t3d2),}\KeywordTok{s}\NormalTok{(t4d2)))}
+\NormalTok{numeratorD2 <-}\StringTok{  }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1d2),}\KeywordTok{s}\NormalTok{(t2d2)) }\OperatorTok{-}\StringTok{ }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t3d2),}\KeywordTok{s}\NormalTok{(t4d2))}
+\NormalTok{biasesD2[i] <-}\StringTok{ }\NormalTok{numeratorD2}\OperatorTok{/}\NormalTok{factorD2}
 \NormalTok{\}}
 
-\FunctionTok{ggplot}\NormalTok{()}\SpecialCharTok{+}\FunctionTok{geom\_histogram}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{biasesD2, }\AttributeTok{y =}\NormalTok{ ..density..), }\AttributeTok{alpha =} \FloatTok{0.6}\NormalTok{, }\AttributeTok{bins=}\DecValTok{50}\NormalTok{)}\SpecialCharTok{+}
-  \FunctionTok{theme\_tufte}\NormalTok{()}\SpecialCharTok{+}\FunctionTok{labs}\NormalTok{(}\AttributeTok{title=}\StringTok{"10k biases for different means and sd =.05"}\NormalTok{)}\SpecialCharTok{+} \FunctionTok{xlab}\NormalTok{(}\StringTok{"bias"}\NormalTok{)}
+\KeywordTok{ggplot}\NormalTok{()}\OperatorTok{+}\KeywordTok{geom_histogram}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x=}\NormalTok{biasesD2, }\DataTypeTok{y =}\NormalTok{ ..density..), }\DataTypeTok{alpha =} \FloatTok{0.6}\NormalTok{, }\DataTypeTok{bins=}\DecValTok{50}\NormalTok{)}\OperatorTok{+}
+\StringTok{  }\KeywordTok{theme_tufte}\NormalTok{()}\OperatorTok{+}\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title=}\StringTok{"10k biases for different means and sd =.05"}\NormalTok{)}\OperatorTok{+}\StringTok{ }\KeywordTok{xlab}\NormalTok{(}\StringTok{"bias"}\NormalTok{)}
 \end{Highlighting}
@@
 \begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/unnamed-chunk-12-1} \end{center}
+
 \normalsize
@@
 
-\vspace{1mm}
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{set.seed}\NormalTok{(}\DecValTok{766}\NormalTok{)}
+\NormalTok{t1d1 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
+\NormalTok{t2d1 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
+\NormalTok{t3d1 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
+\NormalTok{t4d1 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+\vspace{1mm} \footnotesize
+
+\begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/unnamed-chunk-14-1} \end{center}
+
+\normalsize
+
+\noindent  Is this clear difference mirrored in the calculations?
+
 \footnotesize
@@
 \begin{Highlighting}[]
-\FunctionTok{set.seed}\NormalTok{(}\DecValTok{766}\NormalTok{)}
-\NormalTok{t1d1 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
-\NormalTok{t2d1 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
-\NormalTok{t3d1 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
-\NormalTok{t4d1 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\normalsize
-
-\vspace{1mm}
-\footnotesize
-
-\begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/unnamed-chunk-14-1} \end{center}
-\normalsize
-
-\noindent  Is this clear difference mirrored in the calculations?
-
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{factorD1 }\OtherTok{\textless{}{-}} \FunctionTok{sd}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1d1),}\FunctionTok{s}\NormalTok{(t2d1),}\FunctionTok{s}\NormalTok{(t3d1),}\FunctionTok{s}\NormalTok{(t4d1)))}
-\NormalTok{numeratorD1 }\OtherTok{\textless{}{-}}  \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1d1),}\FunctionTok{s}\NormalTok{(t2d1)) }\SpecialCharTok{{-}} \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t3d1),}\FunctionTok{s}\NormalTok{(t4d1))}
-\NormalTok{biasD1 }\OtherTok{\textless{}{-}}\NormalTok{ numeratorD1 }\SpecialCharTok{/}\NormalTok{ factorD1}
+\NormalTok{factorD1 <-}\StringTok{ }\KeywordTok{sd}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1d1),}\KeywordTok{s}\NormalTok{(t2d1),}\KeywordTok{s}\NormalTok{(t3d1),}\KeywordTok{s}\NormalTok{(t4d1)))}
+\NormalTok{numeratorD1 <-}\StringTok{  }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1d1),}\KeywordTok{s}\NormalTok{(t2d1)) }\OperatorTok{-}\StringTok{ }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t3d1),}\KeywordTok{s}\NormalTok{(t4d1))}
+\NormalTok{biasD1 <-}\StringTok{ }\NormalTok{numeratorD1 }\OperatorTok{/}\StringTok{ }\NormalTok{factorD1}
 \NormalTok{biasD1}
@@
 
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{biasesD1 }\OtherTok{\textless{}{-}} \FunctionTok{numeric}\NormalTok{(}\DecValTok{10000}\NormalTok{)}
-
-\ControlFlowTok{for}\NormalTok{(i }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\DecValTok{10000}\NormalTok{)\{}
-\NormalTok{t1d1 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
-\NormalTok{t2d1 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
-\NormalTok{t3d1 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
-\NormalTok{t4d1 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
-
-\NormalTok{factorD1 }\OtherTok{\textless{}{-}} \FunctionTok{sd}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1d1),}\FunctionTok{s}\NormalTok{(t2d1),}\FunctionTok{s}\NormalTok{(t3d1),}\FunctionTok{s}\NormalTok{(t4d1)))}
-\NormalTok{numeratorD1 }\OtherTok{\textless{}{-}}  \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1d1),}\FunctionTok{s}\NormalTok{(t2d1)) }\SpecialCharTok{{-}} \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t3d1),}\FunctionTok{s}\NormalTok{(t4d1))}
-\NormalTok{biasesD1[i] }\OtherTok{\textless{}{-}}\NormalTok{ numeratorD1}\SpecialCharTok{/}\NormalTok{factorD1}
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{biasesD1 <-}\StringTok{ }\KeywordTok{numeric}\NormalTok{(}\DecValTok{10000}\NormalTok{)}
+
+\ControlFlowTok{for}\NormalTok{(i }\ControlFlowTok{in} \DecValTok{1}\OperatorTok{:}\DecValTok{10000}\NormalTok{)\{}
+\NormalTok{t1d1 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
+\NormalTok{t2d1 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
+\NormalTok{t3d1 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
+\NormalTok{t4d1 <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{A  =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\DataTypeTok{B =} \KeywordTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
+
+\NormalTok{factorD1 <-}\StringTok{ }\KeywordTok{sd}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1d1),}\KeywordTok{s}\NormalTok{(t2d1),}\KeywordTok{s}\NormalTok{(t3d1),}\KeywordTok{s}\NormalTok{(t4d1)))}
+\NormalTok{numeratorD1 <-}\StringTok{  }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t1d1),}\KeywordTok{s}\NormalTok{(t2d1)) }\OperatorTok{-}\StringTok{ }\KeywordTok{mean}\NormalTok{(}\KeywordTok{s}\NormalTok{(t3d1),}\KeywordTok{s}\NormalTok{(t4d1))}
+\NormalTok{biasesD1[i] <-}\StringTok{ }\NormalTok{numeratorD1}\OperatorTok{/}\NormalTok{factorD1}
 \NormalTok{\}}
 
-\FunctionTok{ggplot}\NormalTok{()}\SpecialCharTok{+}\FunctionTok{geom\_histogram}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{biasesD1, }\AttributeTok{y =}\NormalTok{ ..density..), }\AttributeTok{alpha =} \FloatTok{0.6}\NormalTok{, }\AttributeTok{bins=}\DecValTok{50}\NormalTok{)}\SpecialCharTok{+}
-  \FunctionTok{theme\_tufte}\NormalTok{()}\SpecialCharTok{+}\FunctionTok{labs}\NormalTok{(}\AttributeTok{title=}\StringTok{"10k biases for different means and sd =.001"}\NormalTok{)}\SpecialCharTok{+} \FunctionTok{xlab}\NormalTok{(}\StringTok{"bias"}\NormalTok{)}
+\KeywordTok{ggplot}\NormalTok{()}\OperatorTok{+}\KeywordTok{geom_histogram}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x=}\NormalTok{biasesD1, }\DataTypeTok{y =}\NormalTok{ ..density..), }\DataTypeTok{alpha =} \FloatTok{0.6}\NormalTok{, }\DataTypeTok{bins=}\DecValTok{50}\NormalTok{)}\OperatorTok{+}
+\StringTok{  }\KeywordTok{theme_tufte}\NormalTok{()}\OperatorTok{+}\KeywordTok{labs}\NormalTok{(}\DataTypeTok{title=}\StringTok{"10k biases for different means and sd =.001"}\NormalTok{)}\OperatorTok{+}\StringTok{ }\KeywordTok{xlab}\NormalTok{(}\StringTok{"bias"}\NormalTok{)}
 \end{Highlighting}
@@
 \begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/unnamed-chunk-16-1} \end{center}
+
 \normalsize
@@
 
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{mean}\NormalTok{(biasesD1); }\FunctionTok{mean}\NormalTok{(biasesD2)}
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{mean}\NormalTok{(biasesD1); }\KeywordTok{mean}\NormalTok{(biasesD2)}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\FunctionTok{median}\NormalTok{(biasesD1); }\FunctionTok{median}\NormalTok{(biasesD2)}
+\KeywordTok{median}\NormalTok{(biasesD1); }\KeywordTok{median}\NormalTok{(biasesD2)}
 \end{Highlighting}
@@
 \normalsize
-
 and so the point estimations of bias are sensitive to factors other than
@@
 
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{quantile}\NormalTok{(biasesD2, }\AttributeTok{probs =} \FunctionTok{c}\NormalTok{(}\FloatTok{0.275}\NormalTok{,}\FloatTok{0.975}\NormalTok{))}
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{quantile}\NormalTok{(biasesD2, }\DataTypeTok{probs =} \KeywordTok{c}\NormalTok{(}\FloatTok{0.275}\NormalTok{,}\FloatTok{0.975}\NormalTok{))}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\FunctionTok{quantile}\NormalTok{(biasesD1, }\AttributeTok{probs =} \FunctionTok{c}\NormalTok{(}\FloatTok{0.275}\NormalTok{,}\FloatTok{0.975}\NormalTok{))}
+\KeywordTok{quantile}\NormalTok{(biasesD1, }\DataTypeTok{probs =} \KeywordTok{c}\NormalTok{(}\FloatTok{0.275}\NormalTok{,}\FloatTok{0.975}\NormalTok{))}
 \end{Highlighting}
@@
 
-\hypertarget{bayesian-estimation}{%
-\section{Bayesian estimation}\label{bayesian-estimation}}
-
-We noticed that in Manzini, Lim, Tsvetkov, \& Black (2019) words from
-all three sub-classes (in the case of religion by sub-class we mean
-Christianity, Judaism, Islam) were compared against all of the
-stereotypes, which means that there was no distinction between cases in
-which the stereotype is associated with a given sub-class, as opposed to
-the situation in which it is associated with another one. Not all of the
+\section{Bayesian estimation}\label{bayesian-estimation}
+
+We noticed that in Manzini et al. (2019) words from all three
+sub-classes (in the case of religion by sub-class we mean Christianity,
+Judaism, Islam) were compared against all of the stereotypes, which
+means that there was no distinction between cases in which the
+stereotype is associated with a given sub-class, as opposed to the
+situation in which it is associated with another one. Not all of the
 stereotypical words have to be considered as harmful for all of the
@@
 One may consider improving the bias detection method presented in
-Manzini, Lim, Tsvetkov, \& Black (2019). As the chosen protected words
-and attributes do not represent the whole biased population but only a
-sample, it is justified to introduce the uncertainty measure to the
-method. We decided to use Bayesian estimation to make the cosine
-distance method more methodologically correct.
+Manzini et al. (2019). As the chosen protected words and attributes do
+not represent the whole biased population but only a sample, it is
+justified to introduce the uncertainty measure to the method. We decided
+to use Bayesian estimation to make the cosine distance method more
+methodologically correct.
 
@@
 should reflect this constraint. As we want to verify as objectively as
-possible the assumptions present in Manzini, Lim, Tsvetkov, \& Black
-(2019) we decided to use fairly weak and slightly skeptical regularizing
-priors. The models parameters will have a posterior distribution
-obtained using either Hamiltionian Monte Carlo methods (STAN) available
-through the rethinking package.
+possible the assumptions present in Manzini et al. (2019) we decided to
+use fairly weak and slightly skeptical regularizing priors. The models
+parameters will have a posterior distribution obtained using either
+Hamiltionian Monte Carlo methods (STAN) available through the rethinking
+package.
 
@@
 
-Let's analyse dataset-level coeﬀicients together with their highest
+Let's analyse dataset-level coefficients together with their highest
 posterior density intervals, for the three topic groups. We compare the
 results for two different word emebeddings corpuses - RedditL2 and
-GoogleNews. For Google embeddings the HPDIs for all class coeﬀicients
+GoogleNews. For Google embeddings the HPDIs for all class coefficients
 (associated, different, human, and none) include zero. This can lead one
@@
 HPDI is below 0 for Gender class when looking at \texttt{associated} and
-\texttt{different} mean coeﬀicient. This can suggest that there is
+\texttt{different} mean coefficient. This can suggest that there is
 indeed slightly stronger impact of these attributes. One should also
-notice how in general associated and different coeﬀicients have quite
+notice how in general associated and different coefficients have quite
 similar HPDI range, the highest observed absolute difference is equal to
@@
 dataset there is a minor shift (in absolute values the highest change is
-equal to approximately 0.1) of the mean coeﬀicients towards zero.
+equal to approximately 0.1) of the mean coefficients towards zero.
 However for Gender dataset there is no significant change. This is of
-course a general look at the group coeﬀicients. Let's now analyze the
+course a general look at the group coefficients. Let's now analyze the
 individual words.
@@
 
-\hypertarget{effects-of-debiasing}{%
-\section{Effects of debiasing}\label{effects-of-debiasing}}
+\section{Effects of debiasing}\label{effects-of-debiasing}
 
 One may verify if the dataset-level coefficients changed after the
-debiasing method used in Manzini, Lim, Tsvetkov, \& Black (2019). In
-some cases there is a minor change to the right, the mean value is
-closer to 0. The \texttt{associated} and \texttt{differen} attributes
-are now more overlapping than before. However, the uncertainty intervals
-are still high and therefore the significance of the change cannot be
-considered as enormous. The surprising obervation is that in the case of
+debiasing method used in Manzini et al. (2019). In some cases there is a
+minor change to the right, the mean value is closer to 0. The
+\texttt{associated} and \texttt{differen} attributes are now more
+overlapping than before. However, the uncertainty intervals are still
+high and therefore the significance of the change cannot be considered
+as enormous. The surprising obervation is that in the case of
 \texttt{Gender} and \texttt{Race} groups, the neutral words shifted to
@@
 
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
+\section{Discussion}\label{discussion}
 
@@
 
-\hypertarget{references}{%
-\section*{References}\label{references}}
+\section*{References}\label{references}
 \addcontentsline{toc}{section}{References}
@@
 \hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\hypertarget{ref-bolukbasi2016man}{}%
-Bolukbasi, T., Chang, K.-W., Zou, J., Saligrama, V., \& Kalai, A.
-(2016). Man is to computer programmer as woman is to homemaker?
-Debiasing word embeddings. Retrieved from
-\url{http://arxiv.org/abs/1607.06520}
-
-\leavevmode\hypertarget{ref-Caliskan2017semanticsBiases}{}%
-Caliskan, A., Bryson, J. J., \& Narayanan, A. (2017). Semantics derived
-automatically from language corpora contain human-like biases.
-\emph{Science}, \emph{356}(6334), 183--186. American Association for the
-Advancement of Science ({AAAS}). Retrieved from
-\url{https://doi.org/10.1126/science.aal4230}
-
-\leavevmode\hypertarget{ref-Garg2018years}{}%
-Garg, N., Schiebinger, L., Jurafsky, D., \& Zou, J. (2018). Word
-embeddings quantify 100 years of gender and ethnic stereotypes.
-\emph{Proceedings of the National Academy of Sciences}, \emph{115}(16),
-E3635--E3644. Proceedings of the National Academy of Sciences. Retrieved
-from \url{https://doi.org/10.1073/pnas.1720347115}
-
-\leavevmode\hypertarget{ref-Gonen2019lipstick}{}%
-Gonen, H., \& Goldberg, Y. (2019). Lipstick on a pig: {D}ebiasing
-methods cover up systematic gender biases in word embeddings but do not
-remove them. \emph{Proceedings of the 2019 conference of the north
-{A}merican chapter of the association for computational linguistics:
-Human language technologies, volume 1 (long and short papers)} (pp.
-609--614). Minneapolis, Minnesota: Association for Computational
-Linguistics. Retrieved from
-\url{https://www.aclweb.org/anthology/N19-1061}
-
-\leavevmode\hypertarget{ref-Lauscher2019multidimensional}{}%
-Lauscher, A., \& Glavas, G. (2019). Are we consistently biased?
-Multidimensional analysis of biases in distributional word vectors.
-\emph{CoRR}, \emph{abs/1904.11783}. Retrieved from
-\url{http://arxiv.org/abs/1904.11783}
-
-\leavevmode\hypertarget{ref-Manzini2019blackToCriminal}{}%
-Manzini, T., Lim, Y. C., Tsvetkov, Y., \& Black, A. W. (2019). Black is
-to criminal as caucasian is to police: Detecting and removing multiclass
-bias in word embeddings. Retrieved from
-\url{http://arxiv.org/abs/1904.04047}
-
-\leavevmode\hypertarget{ref-Nissim2020fair}{}%
-Nissim, M., Noord, R. van, \& Goot, R. van der. (2020). Fair is better
-than sensational: Man is to doctor as woman is to doctor.
-\emph{Computational Linguistics}, \emph{46}(2), 487--497. {MIT} Press -
-Journals. Retrieved from \url{https://doi.org/10.1162/coli_a_00379}
-
-\leavevmode\hypertarget{ref-Nosek2002harvesting}{}%
-Nosek, B. A., Banaji, M. R., \& Greenwald, A. G. (2002). Harvesting
-implicit group attitudes and beliefs from a demonstration web site.
-\emph{Group Dynamics: Theory, Research, and Practice}, \emph{6}(1),
-101--115. American Psychological Association ({APA}). Retrieved from
-\url{https://doi.org/10.1037/1089-2699.6.1.101}
-
-\leavevmode\hypertarget{ref-Venkat2018Curse}{}%
-Venkat, N. (2018). The curse of dimensionality: Inside out.
-
-\end{CSLReferences}
+\hypertarget{ref-Manzini2019blackToCriminal}{}
+Manzini, T., Lim, Y. C., Tsvetkov, Y., \& Black, A. W. (2019).
+\emph{Black is to criminal as caucasian is to police: Detecting and
+removing multiclass bias in word embeddings}.
+
+\hypertarget{ref-Venkat2018Curse}{}
+Venkat, N. (2018). \emph{The curse of dimensionality: Inside out}.
+\url{https://doi.org/10.13140/RG.2.2.29631.36006}
 

### 2021-07-21 paperDraft/paperDraft.Rmd

@@
 
-We noticed that in @Manzini2019blackToCriminal words from all three sub-classes (in the case of religion by sub-class we mean Christianity, Judaism, Islam) were compared against all of the stereotypes, which means that there was no distinction between cases in which the stereotype is associated with a given sub-class, as opposed to the situation in which it is associated with another one.
+One thing to notice is that in @Manzini2019blackToCriminal words from all three subclasses (in the case of religion by a subclass we mean Christianity, Judaism, and Islam) were compared against all of the stereotypes, which means that there was no distinction between cases in which the stereotype is associated with a given sub-class, as opposed to the situation in which it is associated with another one.
 Not all of the stereotypical words have to be considered as harmful for all of the sub-groups.
@@
 
-Our priors distributions for the Bayesian model reflect our knowledge about the data.
-The distribution for mean distance (cosine distance) is set to be between 0 and 2 with the mean equal to 1.
-We assume that most of the cosine distances may be considered neutral and therefore their similarity will be small.
-The distribution for prior for standard deviation is mostly between 0 and 1 with mean around 0.
-Finally the distribution for prior for coefficients is mostly between -1 and 1, with mean around 0.
-One should remember that the cosine distance values have range between 0 and 2 therefore the choice of the prior distributions should reflect this constraint.
-As we want to verify as objectively as possible the assumptions present in @Manzini2019blackToCriminal we decided to use fairly weak and slightly skeptical regularizing priors.
-The models parameters will have a posterior distribution obtained using either Hamiltionian Monte Carlo methods (STAN) available through the rethinking package.
+Our prior distributions for the Bayesian model reflect our first stab at what we know or expect before we see the data.
+All our distributions are fairly weak.
+The distribution for mean distance (cosine distance) is set to fit (almost completely) between 0 and 2 with the mean equal to 1, in line with the idea that  most of the cosine distances may be considered neutral and therefore their similarity will be around 0.
+That is, we take it to be DISTRIBUTION.
+The  prior distribution for standard deviation is half-cauchy between 0 and 1, but unlikely to be as large as 1.
+Finally the prior distributions for  the coefficients have most of their mass between -1 and 1, with mean around 0.
+As we want to put ourselves in skeptical shoes looking at   @Manzini2019blackToCriminal and asking whether the data would indeed convince us of the existence of bias, we decided to use slightly skeptical (mean 0), but  fairly weak (sd = ) as our regularizing priors.
+The model parameters will have a posterior distribution obtained using either Hamiltionian Monte Carlo methods (STAN) available through the rethinking package.
 

### 2021-07-21 paperDraft/paperDraft.tex

@@
 
-The attributes are taken from different sources, there is no principled
-justification for their choice. From our analysis it will become clear
-that the list is rather uneven.
-
-There is no mention of methodology for deciding on the number of
-attributes necessary to decide a hypothesis on the given size of
-dataset. There are however some ways to estimate sample sizes needed to
-ensure that the results are meaningful if the effect is present. Our
-research will show that the numbers used are rather insufficient.
-
-\textbf{2. No attention to distributions and details}
-
-The authors use the mean of mean average cosine similarities to measure
-multi-class similarity between protected word and harmful stereotypes.
-They average the results until they obtain one final value to represent
-the mean cosine distance between protected word from a given class and
-the attributes of that class. If one takes closer look at the individual
-values that are taken for the calculations it turns out that there is a
-bunch of outliers and surprisingly dissimilar words. We approach this
-issue by providing the tables and visualizations of individual cosine
-distances to make sure that we obtain a proper insight into the data.
+Such approaches, however, are not unproblematic. For one thing, the word
+lists are often fairly small, and the papers do not discuss the impact
+of word list sizes on the uncertainty involved. We will argue that the
+word list sizes are too small to yield strong conclusions, given the
+variances involved.
+
+Crucially, these approaches use means of mean average cosine
+similarities to measure similarity between protected word and harmful
+stereotypes. If one takes a closer look at the individual values that
+are taken for the calculations, it turns out that there are quite a few
+outliers and surprisingly dissimilar words. This issue will become
+transparent when we inspect the visualizations of individual cosine
+distances, following the idea that one of the first step to understand
+data is to look at it.
 
@@
 
-We noticed that in Manzini et al. (2019) words from all three
-sub-classes (in the case of religion by sub-class we mean Christianity,
-Judaism, Islam) were compared against all of the stereotypes, which
-means that there was no distinction between cases in which the
-stereotype is associated with a given sub-class, as opposed to the
-situation in which it is associated with another one. Not all of the
+One thing to notice is that in Manzini et al. (2019) words from all
+three subclasses (in the case of religion by a subclass we mean
+Christianity, Judaism, and Islam) were compared against all of the
+stereotypes, which means that there was no distinction between cases in
+which the stereotype is associated with a given sub-class, as opposed to
+the situation in which it is associated with another one. Not all of the
 stereotypical words have to be considered as harmful for all of the
@@
 
-Our priors distributions for the Bayesian model reflect our knowledge
-about the data. The distribution for mean distance (cosine distance) is
-set to be between 0 and 2 with the mean equal to 1. We assume that most
-of the cosine distances may be considered neutral and therefore their
-similarity will be small. The distribution for prior for standard
-deviation is mostly between 0 and 1 with mean around 0. Finally the
-distribution for prior for coefficients is mostly between -1 and 1, with
-mean around 0. One should remember that the cosine distance values have
-range between 0 and 2 therefore the choice of the prior distributions
-should reflect this constraint. As we want to verify as objectively as
-possible the assumptions present in Manzini et al. (2019) we decided to
-use fairly weak and slightly skeptical regularizing priors. The models
-parameters will have a posterior distribution obtained using either
-Hamiltionian Monte Carlo methods (STAN) available through the rethinking
-package.
+Our prior distributions for the Bayesian model reflect our first stab at
+what we know or expect before we see the data. All our distributions are
+fairly weak. The distribution for mean distance (cosine distance) is set
+to fit (almost completely) between 0 and 2 with the mean equal to 1, in
+line with the idea that most of the cosine distances may be considered
+neutral and therefore their similarity will be around 0. That is, we
+take it to be DISTRIBUTION. The prior distribution for standard
+deviation is half-cauchy between 0 and 1, but unlikely to be as large as
+1. Finally the prior distributions for the coefficients have most of
+their mass between -1 and 1, with mean around 0. As we want to put
+ourselves in skeptical shoes looking at Manzini et al. (2019) and asking
+whether the data would indeed convince us of the existence of bias, we
+decided to use slightly skeptical (mean 0), but fairly weak (sd = ) as
+our regularizing priors. The model parameters will have a posterior
+distribution obtained using either Hamiltionian Monte Carlo methods
+(STAN) available through the rethinking package.
 

### 2021-07-21 paperDraft/paperDraft.Rmd
Adds small changes

@@
 ---
-title: "Taking uncertainty in word embedding bias estimation seriously: a  bayesian approach"
+title: 'Taking uncertainty in word embedding bias estimation seriously: a  bayesian
+  approach'
 author: "Alicja Dobrzeniecka and Rafal Urbaniak"
@@
   pdf_document:
-    number_sections: true
+    number_sections: yes
     df_print: kable
-    keep_tex: true
+    keep_tex: yes
     includes:
-      in_header:
-        - Rafal_latex6.sty
+      in_header: Rafal_latex6.sty
+  html_document:
+    df_print: paged
 classoption: dvipsnames,enabledeprecatedfontcommands
@@
 urlcolor: blue
-bibliography: [../references/bibliographyRafal.bib]
-csl: [../references/apa-6th-edition.csl] 
+bibliography: ../references/bibliographyRafal.bib
+csl: ../references/apa-6th-edition.csl
 ---
@@
 
-\textbf{1. Selection of attributes}
-
-Such approaches, however, are not unproblematic. For one thing, the word lists are often fairly small, and the papers do not discuss the impact of word list sizes on the uncertainty involved. 
+Such approaches, however, are not unproblematic.
+One may cosinder the selection of attributes for the bias research.
+For one thing, the word lists are often fairly small, and the papers do not discuss the impact of word list sizes on the uncertainty involved.
 We will argue that the word list sizes are too small to yield strong conclusions, given the variances involved.
@@
 
-\textbf{3. Hiding the impact of uncertainty}
-
 With such a method the uncertainty involved is not really considered which makes it even more difficult to give reasonable interpretations of the results.
@@
 
-\textbf{4. No word class distinction and no control group}
-
+We also observed that in the paper there is no word class distinction and no control group.
 In the original paper, words from all three religions were compared against all of the stereotypes, which means that there was no distinction between cases in which the stereotype is associated with a given religion, as opposed to the situation in which it is associated with another one.
@@
 
-\textbf{5. Interpreting the results}
-
+The interpretation of the results is also a challenge.
 Assuming for a moment that the value of multi-class cosine distance is correct, one may question the interpretation.
@@
 
-\textbf{6. The curse of dimensionality}
-
-In our case, the curse of dimensionality may take place when there is an increase in the volume of data that results in adding extra dimensions to the Euclidean space. According to the article 
-[Curse of dimensionality at analyticsindiamag.com](https://analyticsindiamag.com/curse-of-dimensionality-and-what-beginners-should-do-to-overcome-it/) as the number of features increases, it may be harder and harder to obtain useful information from the data using the available algorithms.
-One may notice that more data should contribute to greater amount of information, but more information also means greater risk of noise and distractions in data.
-At the same time, modern solutions are often adapted to smaller dimensions and their results in higher ones are not intuitive, or may be prone to error.
-
-Using cosine similarity in high dimensions in word embeddings may also be prone to the curse of dimensionality.
+One may consider the issue of the curse of dimensionality.
+In our case, the curse of dimensionality may take place when there is an increase in the volume of data that results in adding extra dimensions to the Euclidean space.
+As the number of features increases, it may be harder and harder to obtain useful information from the data using the available algorithms.
+Using cosine similarity in high dimensions in word embeddings may be prone to the curse of dimensionality.
 According to @Venkat2018Curse there are reasons to consider this phenomenon when searching for word similarities in higher dimensions.
@@
 Maybe the scale should be adjusted to the number of dimensions and variance so that it still gives us sensible information about data.
-Yet, according to some articles cosine similarity in high dimensions is not reliable enough as it may be the case that choosing words at random may result in similar values as when picking them consciously.
-
-![curse of dimensionality,  number of dimensions on the x axis,  standard deviation of similarity on the y axis](../images/curseOfDimensionality.png)
 

### 2021-07-21 paperDraft/paperDraft.tex
Adds small changes

--- 2021-07-12 paperDraft/paperDraft.tex
@@
 \hypersetup{
-  pdftitle={A bayesian method of cosine-based word2vec bias},
+  pdftitle={Taking uncertainty in word embedding bias estimation seriously: a bayesian approach},
   pdfauthor={Alicja Dobrzeniecka and Rafal Urbaniak},
@@
 
-\title{A bayesian method of cosine-based word2vec bias}
+\title{Taking uncertainty in word embedding bias estimation seriously: a
+bayesian approach}
 \author{Alicja Dobrzeniecka and Rafal Urbaniak}
@@
 One of the first measures in the discussion has been developed by
-Bolukbasi, Chang, Zou, Saligrama, \& Kalai (2016). There, the gender
-bias of a word \(w\) is understood as its projection on the gender
-direction \(\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})\)
-(the gender direction is the top principal compontent of ten gender pair
+(\textbf{bolukbasi2016man?}). There, the gender bias of a word \(w\) is
+understood as its projection on the gender direction
+\(\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})\) (the
+gender direction is the top principal compontent of ten gender pair
 difference vectors). The underlying idea is that no bias is present if
@@
 
-The use of projections has been ciriticized for instance by Gonen \&
-Goldberg (2019), who point out that while gender-direction might be an
-indicator of bias, it is only one possible manifestation of it, and
-reducing a projection of words might be insufficient. For instance,
-``math'' and ``delicate'' might be in equal distance to both explicitly
-gendered words while being closer to quite different stereotypical
-attribute words. Further, the authors point out that most word pairs
-preserve similarity under debiasing meant to minimize projection-based
-bias.\footnote{Bolukbasi, Chang, Zou, Saligrama, \& Kalai (2016) use
-  also another method which involves analogies and their evaluations by
-  human users on Mechanical Turk. It is discussed and criticized in
-  (Nissim, Noord, \& Goot, 2020).}
-
-To measure bias in word embeddings, Caliskan, Bryson, \& Narayanan
-(2017) proposed the Word Embedding Association Test (WEAT). The idea is
-that the measure of biases between two sets of target words, \(X\) and
-\(Y\), (we call them protected words) should be quantified in terms of
-the cosine similarity between the protected words and attribute words
-coming from two sets of stereotype attribute words, \(A\) and \(B\)
-(we'll call them attributes). For instance, \(X\) might be a set of male
-names, \(Y\) a set of female names, \(A\) might contain stereotypically
-male-related career words, and \(B\) stereotypically female-related
-family words. WEAT is a modification of the Implicit Association Test
-(IAT) (Nosek, Banaji, \& Greenwald, 2002) used in psychology and uses
-almost the same word sets, allowing for a \emph{prima facie} sensible
-comparison with bias in humans. If the person's attitude towards given
-pair of concept is to be interpreted as neutral, there should be no
-noticeable task completion time difference, and the final value from the
-formula should be around 0. Let \(f\) be a similarity measure (usually,
-cosine similarity). The association difference for a term \(t\) is:
+The use of projections has been ciriticized for instance by
+(\textbf{Gonen2019lipstick?}), who point out that while gender-direction
+might be an indicator of bias, it is only one possible manifestation of
+it, and reducing a projection of words might be insufficient. For
+instance, ``math'' and ``delicate'' might be in equal distance to both
+explicitly gendered words while being closer to quite different
+stereotypical attribute words. Further, the authors point out that most
+word pairs preserve similarity under debiasing meant to minimize
+projection-based bias.\footnote{(\textbf{bolukbasi2016man?}) use also
+  another method which involves analogies and their evaluations by human
+  users on Mechanical Turk. It is discussed and criticized in
+  (\textbf{Nissim2020fair?}).}
+
+To measure bias in word embeddings,
+(\textbf{Caliskan2017semanticsBiases?}) proposed the Word Embedding
+Association Test (WEAT). The idea is that the measure of biases between
+two sets of target words, \(X\) and \(Y\), (we call them protected
+words) should be quantified in terms of the cosine similarity between
+the protected words and attribute words coming from two sets of
+stereotype attribute words, \(A\) and \(B\) (we'll call them
+attributes). For instance, \(X\) might be a set of male names, \(Y\) a
+set of female names, \(A\) might contain stereotypically male-related
+career words, and \(B\) stereotypically female-related family words.
+WEAT is a modification of the Implicit Association Test (IAT)
+(\textbf{Nosek2002harvesting?}) used in psychology and uses almost the
+same word sets, allowing for a \emph{prima facie} sensible comparison
+with bias in humans. If the person's attitude towards given pair of
+concept is to be interpreted as neutral, there should be no noticeable
+task completion time difference, and the final value from the formula
+should be around 0. Let \(f\) be a similarity measure (usually, cosine
+similarity). The association difference for a term \(t\) is:
 \begin{align}
@@
 
-Caliskan, Bryson, \& Narayanan (2017) show that significant
+(\textbf{Caliskan2017semanticsBiases?}) show that significant
 biases---thus measured--- similar to the ones discovered by IAT can be
-discovered in word embeddings. Lauscher \& Glavas (2019) extended the
-methodology to a multilingual and cross-lingual setting, arguing that
-using Euclidean distance instead of similarity does not make much
-difference, while the bias effects vary greatly across embedding models
-(interestingly, with social media-text trained embeddings being less
-biased than those based on Wikipedia).
-
-A similar methodology is employed by Garg, Schiebinger, Jurafsky, \& Zou
-(2018), who employ word embeddings trained on corpora from different
-decades to study the shifts in various biases. For instance, to compute
-the occupational embeddings bias for women the authors first compute the
+discovered in word embeddings. (\textbf{Lauscher2019multidimensional?})
+extended the methodology to a multilingual and cross-lingual setting,
+arguing that using Euclidean distance instead of similarity does not
+make much difference, while the bias effects vary greatly across
+embedding models (interestingly, with social media-text trained
+embeddings being less biased than those based on Wikipedia).
+
+A similar methodology is employed by (\textbf{Garg2018years?}), who
+employ word embeddings trained on corpora from different decades to
+study the shifts in various biases. For instance, to compute the
+occupational embeddings bias for women the authors first compute the
 average vector of vector emeddings of words that represent women
@@
 
-\textbf{1. Selection of attributes}
-
-The attributes are taken from different sources, there is no principled
-justification for their choice. From our analysis it will become clear
-that the list is rather uneven.
-
-There is no mention of methodology for deciding on the number of
-attributes necessary to decide a hypothesis on the given size of
-dataset. There are however some ways to estimate sample sizes needed to
-ensure that the results are meaningful if the effect is present. Our
-research will show that the numbers used are rather insufficient.
-
-\textbf{2. No attention to distributions and details}
-
-The authors use the mean of mean average cosine similarities to measure
-multi-class similarity between protected word and harmful stereotypes.
-They average the results until they obtain one final value to represent
-the mean cosine distance between protected word from a given class and
-the attributes of that class. If one takes closer look at the individual
-values that are taken for the calculations it turns out that there is a
-bunch of outliers and surprisingly dissimilar words. We approach this
-issue by providing the tables and visualizations of individual cosine
-distances to make sure that we obtain a proper insight into the data.
-
-\textbf{3. Hiding the impact of uncertainty}
+Such approaches, however, are not unproblematic. One may cosinder the
+selection of attributes for the bias research. For one thing, the word
+lists are often fairly small, and the papers do not discuss the impact
+of word list sizes on the uncertainty involved. We will argue that the
+word list sizes are too small to yield strong conclusions, given the
+variances involved.
+
+Crucially, these approaches use means of mean average cosine
+similarities to measure similarity between protected word and harmful
+stereotypes. If one takes a closer look at the individual values that
+are taken for the calculations, it turns out that there are quite a few
+outliers and surprisingly dissimilar words. This issue will become
+transparent when we inspect the visualizations of individual cosine
+distances, following the idea that one of the first step to understand
+data is to look at it.
 
@@
 
-\textbf{4. No word class distinction and no control group}
-
-In the original paper, words from all three religions were compared
-against all of the stereotypes, which means that there was no
-distinction between cases in which the stereotype is associated with a
-given religion, as opposed to the situation in which it is associated
-with another one. Not all of the stereotypical words have to be
-considered as harmful for all of the religions. One should investigate
-the religions separately as some of them may have stronger harmful
-associations that others. One should also include control groups to have
-a way of comparing the stereotypical results with neutral or human-like
-words. Later in the text we will explain in details reasons for
-introducing control groups. In our analysis, we distinguish between
+We also observed that in the paper there is no word class distinction
+and no control group. In the original paper, words from all three
+religions were compared against all of the stereotypes, which means that
+there was no distinction between cases in which the stereotype is
+associated with a given religion, as opposed to the situation in which
+it is associated with another one. Not all of the stereotypical words
+have to be considered as harmful for all of the religions. One should
+investigate the religions separately as some of them may have stronger
+harmful associations that others. One should also include control groups
+to have a way of comparing the stereotypical results with neutral or
+human-like words. Later in the text we will explain in details reasons
+for introducing control groups. In our analysis, we distinguish between
 stereotypes associated with a given group, stereotypes associated with
@@
 
-\textbf{5. Interpreting the results}
-
-Assuming for a moment that the value of multi-class cosine distance is
-correct, one may question the interpretation. Manzini, Lim, Tsvetkov, \&
-Black (2019) summarize the averages of cosine distance per group
-(gender, race, religion). For now let us focus now on analyzing the
-values relating to religious biases. Here is the relevant fragment of
-table:
+The interpretation of the results is also a challenge. Assuming for a
+moment that the value of multi-class cosine distance is correct, one may
+question the interpretation. Manzini, Lim, Tsvetkov, \& Black (2019)
+summarize the averages of cosine distance per group (gender, race,
+religion). For now let us focus now on analyzing the values relating to
+religious biases. Here is the relevant fragment of table:
 
@@
 
-\textbf{6. The curse of dimensionality}
-
-In our case, the curse of dimensionality may take place when there is an
-increase in the volume of data that results in adding extra dimensions
-to the Euclidean space. According to the article
-\href{https://analyticsindiamag.com/curse-of-dimensionality-and-what-beginners-should-do-to-overcome-it/}{Curse
-of dimensionality at analyticsindiamag.com} as the number of features
-increases, it may be harder and harder to obtain useful information from
-the data using the available algorithms. One may notice that more data
-should contribute to greater amount of information, but more information
-also means greater risk of noise and distractions in data. At the same
-time, modern solutions are often adapted to smaller dimensions and their
-results in higher ones are not intuitive, or may be prone to error.
-
-Using cosine similarity in high dimensions in word embeddings may also
-be prone to the curse of dimensionality. According to Venkat (2018)
-there are reasons to consider this phenomenon when searching for word
-similarities in higher dimensions. An experiment is conducted that aims
-at showing how the similarity values and variation change as the number
-of dimensions increases. The hypothesis made in the paper states that
-two things will happen as the number of dimensions increase. First, the
-effort required to measure cosine similarity will be greater, and two,
-the similarity between data will blur out and have less variation. The
-authors generate random points with increasing number of dimensions
-where each dimension of a data point is given a value between 0 and 1.
-Then they pick one vector at random from each dimension class and
-calculate the cosine similarity between the chosen vector and the rest
-of the data. Then they check how the variation of values changes as the
-number of dimensions increases. It seems like the more dimensions there
-are, the smaller the variance and therefore it is less obvious how to
-interpret the resulting cosine similarities. Maybe the scale should be
-adjusted to the number of dimensions and variance so that it still gives
-us sensible information about data. Yet, according to some articles
-cosine similarity in high dimensions is not reliable enough as it may be
-the case that choosing words at random may result in similar values as
-when picking them consciously.
-
-\begin{figure}
-\centering
-\includegraphics{../images/curseOfDimensionality.png}
-\caption{curse of dimensionality, number of dimensions on the x axis,
-standard deviation of similarity on the y axis}
-\end{figure}
+One may consider the issue of the curse of dimensionality. In our case,
+the curse of dimensionality may take place when there is an increase in
+the volume of data that results in adding extra dimensions to the
+Euclidean space. As the number of features increases, it may be harder
+and harder to obtain useful information from the data using the
+available algorithms. Using cosine similarity in high dimensions in word
+embeddings may be prone to the curse of dimensionality. According to
+Venkat (2018) there are reasons to consider this phenomenon when
+searching for word similarities in higher dimensions. An experiment is
+conducted that aims at showing how the similarity values and variation
+change as the number of dimensions increases. The hypothesis made in the
+paper states that two things will happen as the number of dimensions
+increase. First, the effort required to measure cosine similarity will
+be greater, and two, the similarity between data will blur out and have
+less variation. The authors generate random points with increasing
+number of dimensions where each dimension of a data point is given a
+value between 0 and 1. Then they pick one vector at random from each
+dimension class and calculate the cosine similarity between the chosen
+vector and the rest of the data. Then they check how the variation of
+values changes as the number of dimensions increases. It seems like the
+more dimensions there are, the smaller the variance and therefore it is
+less obvious how to interpret the resulting cosine similarities. Maybe
+the scale should be adjusted to the number of dimensions and variance so
+that it still gives us sensible information about data.
 
@@
 
-To illustrate let's employ the formulas used by Caliskan, Bryson, \&
-Narayanan (2017) in a simple example. Conceptually, all such tests come
-up with rather short lists of protected words and rather short lists of
-stereotypical attributes. Clearly, these are not complete list. So let's
-treat them as samples from richer pools of stereotypical predicates and
-let's take the uncertainty and variation involved seriously.
+To illustrate let's employ the formulas used by
+(\textbf{Caliskan2017semanticsBiases?}) in a simple example.
+Conceptually, all such tests come up with rather short lists of
+protected words and rather short lists of stereotypical attributes.
+Clearly, these are not complete list. So let's treat them as samples
+from richer pools of stereotypical predicates and let's take the
+uncertainty and variation involved seriously.
 
@@
 a strong bias here? We wouldn't think so. But now let's run the
-calculations from (Caliskan, Bryson, \& Narayanan, 2017).
+calculations from (\textbf{Caliskan2017semanticsBiases?}).
 
@@
 calculated effect size is 1.82, whereas the largest effect size reported
-by Caliskan, Bryson, \& Narayanan (2017) is 1.81!
+by (\textbf{Caliskan2017semanticsBiases?}) is 1.81!
 
@@
 
-We noticed that in Manzini, Lim, Tsvetkov, \& Black (2019) words from
-all three sub-classes (in the case of religion by sub-class we mean
-Christianity, Judaism, Islam) were compared against all of the
-stereotypes, which means that there was no distinction between cases in
-which the stereotype is associated with a given sub-class, as opposed to
-the situation in which it is associated with another one. Not all of the
-stereotypical words have to be considered as harmful for all of the
+One thing to notice is that in Manzini, Lim, Tsvetkov, \& Black (2019)
+words from all three subclasses (in the case of religion by a subclass
+we mean Christianity, Judaism, and Islam) were compared against all of
+the stereotypes, which means that there was no distinction between cases
+in which the stereotype is associated with a given sub-class, as opposed
+to the situation in which it is associated with another one. Not all of
+the stereotypical words have to be considered as harmful for all of the
 sub-groups. One should investigate the sub-groups separately as some of
@@
 
-Our priors distributions for the Bayesian model reflect our knowledge
-about the data. The distribution for mean distance (cosine distance) is
-set to be between 0 and 2 with the mean equal to 1. We assume that most
-of the cosine distances may be considered neutral and therefore their
-similarity will be small. The distribution for prior for standard
-deviation is mostly between 0 and 1 with mean around 0. Finally the
-distribution for prior for coefficients is mostly between -1 and 1, with
-mean around 0. One should remember that the cosine distance values have
-range between 0 and 2 therefore the choice of the prior distributions
-should reflect this constraint. As we want to verify as objectively as
-possible the assumptions present in Manzini, Lim, Tsvetkov, \& Black
-(2019) we decided to use fairly weak and slightly skeptical regularizing
-priors. The models parameters will have a posterior distribution
-obtained using either Hamiltionian Monte Carlo methods (STAN) available
-through the rethinking package.
+Our prior distributions for the Bayesian model reflect our first stab at
+what we know or expect before we see the data. All our distributions are
+fairly weak. The distribution for mean distance (cosine distance) is set
+to fit (almost completely) between 0 and 2 with the mean equal to 1, in
+line with the idea that most of the cosine distances may be considered
+neutral and therefore their similarity will be around 0. That is, we
+take it to be DISTRIBUTION. The prior distribution for standard
+deviation is half-cauchy between 0 and 1, but unlikely to be as large as
+1. Finally the prior distributions for the coefficients have most of
+their mass between -1 and 1, with mean around 0. As we want to put
+ourselves in skeptical shoes looking at Manzini, Lim, Tsvetkov, \& Black
+(2019) and asking whether the data would indeed convince us of the
+existence of bias, we decided to use slightly skeptical (mean 0), but
+fairly weak (sd = ) as our regularizing priors. The model parameters
+will have a posterior distribution obtained using either Hamiltionian
+Monte Carlo methods (STAN) available through the rethinking package.
 
@@
 \begin{CSLReferences}{1}{0}
-\leavevmode\hypertarget{ref-bolukbasi2016man}{}%
-Bolukbasi, T., Chang, K.-W., Zou, J., Saligrama, V., \& Kalai, A.
-(2016). Man is to computer programmer as woman is to homemaker?
-Debiasing word embeddings. Retrieved from
-\url{http://arxiv.org/abs/1607.06520}
-
-\leavevmode\hypertarget{ref-Caliskan2017semanticsBiases}{}%
-Caliskan, A., Bryson, J. J., \& Narayanan, A. (2017). Semantics derived
-automatically from language corpora contain human-like biases.
-\emph{Science}, \emph{356}(6334), 183--186. American Association for the
-Advancement of Science ({AAAS}). Retrieved from
-\url{https://doi.org/10.1126/science.aal4230}
-
-\leavevmode\hypertarget{ref-Garg2018years}{}%
-Garg, N., Schiebinger, L., Jurafsky, D., \& Zou, J. (2018). Word
-embeddings quantify 100 years of gender and ethnic stereotypes.
-\emph{Proceedings of the National Academy of Sciences}, \emph{115}(16),
-E3635--E3644. Proceedings of the National Academy of Sciences. Retrieved
-from \url{https://doi.org/10.1073/pnas.1720347115}
-
-\leavevmode\hypertarget{ref-Gonen2019lipstick}{}%
-Gonen, H., \& Goldberg, Y. (2019). Lipstick on a pig: {D}ebiasing
-methods cover up systematic gender biases in word embeddings but do not
-remove them. \emph{Proceedings of the 2019 conference of the north
-{A}merican chapter of the association for computational linguistics:
-Human language technologies, volume 1 (long and short papers)} (pp.
-609--614). Minneapolis, Minnesota: Association for Computational
-Linguistics. Retrieved from
-\url{https://www.aclweb.org/anthology/N19-1061}
-
-\leavevmode\hypertarget{ref-Lauscher2019multidimensional}{}%
-Lauscher, A., \& Glavas, G. (2019). Are we consistently biased?
-Multidimensional analysis of biases in distributional word vectors.
-\emph{CoRR}, \emph{abs/1904.11783}. Retrieved from
-\url{http://arxiv.org/abs/1904.11783}
-
 \leavevmode\hypertarget{ref-Manzini2019blackToCriminal}{}%
-Manzini, T., Lim, Y. C., Tsvetkov, Y., \& Black, A. W. (2019). Black is
-to criminal as caucasian is to police: Detecting and removing multiclass
-bias in word embeddings. Retrieved from
+Manzini, T., Lim, Y. C., Tsvetkov, Y., \& Black, A. W. (2019).
+\emph{Black is to criminal as caucasian is to police: Detecting and
+removing multiclass bias in word embeddings}. Retrieved from
 \url{http://arxiv.org/abs/1904.04047}
 
-\leavevmode\hypertarget{ref-Nissim2020fair}{}%
-Nissim, M., Noord, R. van, \& Goot, R. van der. (2020). Fair is better
-than sensational: Man is to doctor as woman is to doctor.
-\emph{Computational Linguistics}, \emph{46}(2), 487--497. {MIT} Press -
-Journals. Retrieved from \url{https://doi.org/10.1162/coli_a_00379}
-
-\leavevmode\hypertarget{ref-Nosek2002harvesting}{}%
-Nosek, B. A., Banaji, M. R., \& Greenwald, A. G. (2002). Harvesting
-implicit group attitudes and beliefs from a demonstration web site.
-\emph{Group Dynamics: Theory, Research, and Practice}, \emph{6}(1),
-101--115. American Psychological Association ({APA}). Retrieved from
-\url{https://doi.org/10.1037/1089-2699.6.1.101}
-
 \leavevmode\hypertarget{ref-Venkat2018Curse}{}%
-Venkat, N. (2018). The curse of dimensionality: Inside out.
+Venkat, N. (2018). \emph{The curse of dimensionality: Inside out}.
+\url{https://doi.org/10.13140/RG.2.2.29631.36006}
 

### 2021-07-27 presentationESSLLI21/presentationESSLLI.Rmd
added presentation template

---
title: "\\large A bayesian method of cosine-based word2vec bias estimation"
author: "Alicja Dobrzeniecka \\& Rafal  Urbaniak \\footnotesize \\newline (LoPSE research group, University of  Gdansk)" 
date:  "ExpSem2021, ESSLLI"
fontsize: 10pt
output:  
  beamer_presentation:
    theme: Rafal_beamerSly1
    keep_tex: yes
    slide_level: 2
classoption: x11names, dvipsnames, bibspacing,natbib
urlcolor: blue
bibliography: [../references/bibliographyRafal.bib]
csl: [../references/apa-6th-edition.csl] 
link-citations: yes
---

```{r setup, include=FALSE}
library(tidyverse)
library(gridExtra)
library(kableExtra)
library(magrittr)
library(ggplot2)
library(ggpubr)
library(ggExtra)
library(ggthemes)
library(latex2exp)
th <- theme_tufte(base_size = 20)
options(kableExtra.latex.load_packages = FALSE)
```

\frametitle{The gist of the project}

\pause

### Classical probabilism in epistemology (and in applications)

Represent uncertainty by a single probabilistic measure (point estimates).

\vspace{-3mm}

\begin{align*}
\pr(\mathsf{recovery}\vert \mathsf{treatment}) & = .6\\
\end{align*}

\vspace{-6mm}

\pause

### Imprecise probabilism

Evidence (e.g. expert evidence) sometimes fails to warrant such precision.

\vspace{-3mm}

\begin{align*}
\pr(\mathsf{recovery}\vert \mathsf{treatment}) & \in [.4,.6]\\
\end{align*}

\vspace{-6mm}

\pause

### Epistemological difficulties of imprecise probabilism

- hard to model the weight of and combine evidence.

- inertia of agnosticism.

- growth of uncertainty  in light of new evidence.

## The gist of the project

### My novel contribution

- Meet the desiderata motivating imprecision.

- Avoid these difficulties.

- Do so by using higher-order probabilities.

```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "70%"}
p <- ggplot(data = data.frame(probability = seq(0,1, by = .25)),mapping = aes(x = probability))+th +ylab("density")+scale_x_continuous(breaks = seq(0,1, by = .1))

agentPrior <- function(p) dbeta(p,28,28)

priorPlot <- p + stat_function(fun=agentPrior)+ggtitle("Uncertainty about recovery")+xlab("probability of recovery if treated")

priorPlot
```

## Motivations for imprecision  \large (straighforward considerations)

\pause

### Radical lack of information

$\pr(\mathsf{It \,\, will \,\,  snow\,\,  in\,\,  Boston\,\,  on \,\, New\,\,  Year's\,\,  Eve\,\,  2026}) =$ ?

\pause

<!-- ### Mere bounds in reasoning -->

<!-- - Say $\pr(A) = .5$, $\pr(B) = .3$, but you don't know if they're independent. -->

<!-- \vspace{-2mm} -->

<!-- - We can only calculate bounds for $\pr(A \et B)$, $\pr(A \vee B)$, etc. -->

<!-- \pause  -->

### Weight of evidence

- You've seen 4/10 patients recover. $\pr(\mathsf{recovery})=.4$?

\pause
\vspace{-2mm}

- You've seen 40/100 patients recover. $\pr(\mathsf{recovery})=.4$?

\footnotesize (You could try to use classical statistical tools, such as confidence intervals,\linebreak  but they have their own problems.)

## Motivations for imprecision \large (uniformity and agnosticism)

### Agnosticism preservation requirement

 Agnosticism about uniquely connected variables should be the same.

\vspace{-5mm}

 \[X \mbox{ and } Y=X+2 \mbox{ or } Z = X^2 \,\,\, \mbox{ for $10>X>0$.}\]

\pause

### The trouble

```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "60%"}
p <- ggplot(data = data.frame(value = 0:10), 
            mapping = aes(x = value))

unifX <-  function(value) punif(value,min = 0, max = 10)

p +   stat_function(fun=unifX)+
  ylab("density")+theme_tufte(base_size = 25)+labs(title = "Uniform cumulative probability for X in (0,10)")+xlab("X")+ylab("Cumulative probability")+scale_x_continuous(breaks = 0:10)+
  geom_segment(aes(x = 5, y = 0, xend = 5, yend = .5), lty = 2, alpha = .2, size = .4, color = "skyblue")+
  geom_segment(aes(x = 0, y = .5, xend = 5, yend = .5), lty = 2, alpha = .2, size = .4, color = "skyblue")

```

\vspace{-9mm}

\begin{align*}\pr(X<5) = .5 = \pr(X^2 < 25)
\end{align*}

## Motivations for imprecision \large (uniformity and agnosticism)

### The trouble

```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "60%", message = FALSE, warning = FALSE}
p <- ggplot(data = data.frame(value = 0:10), 
            mapping = aes(x = value))

unifY <-  function(value) punif(sqrt(value),min = 0, max = 10)

p +   stat_function(fun=unifY)+
  ylab("cumulative probability")+theme_tufte(base_size = 25)+
  labs(title = TeX("Resulting cumulative probability for $X^2$"))+scale_x_continuous(breaks = 0:100)+xlab(TeX("X^2"))+xlim(c(0,100))+
  geom_segment(aes(x = 25, y = 0, xend = 25, yend = .5), lty = 2, alpha = .2, size = .4, color = "skyblue")+
  geom_segment(aes(x = 0, y = .5, xend = 25, yend = .5), lty = 2, alpha = .2, size = .4, color = "skyblue")

```

## Motivations for imprecision \large (expert evidence)

### Weather forecasters

\vspace{-1mm}

\begin{tabular}{ll}
Al: & $\pr(\mathsf{It \,\,  will\,\,  rain\,\,  tomorrow})=.6$\\
Bert: & $\pr(\mathsf{It \,\,  will\,\,  rain\,\,  tomorrow})=.8$
\end{tabular}

\pause

### Linear pooling

\vspace{-4mm}

\begin{align*}
\pr(\mathsf{It \,\,  will\,\,  rain\,\,  tomorrow}) & = 
\alpha \times .6 + \beta \times .8 \,\,\,\,\,  (where \,\, \alpha + \beta = 1.)
\end{align*}

\pause

### Mathematical impossibilities

- Experts's agreement on independence is not preserved by linear pooling. [@List2011Group]

- You can't at the same time hold the following: [@Gallow2018No]

\vspace{-7mm}

\begin{align}
\pr(A=B) & <1\\
\pr(r\vert A=a) & = a \\
\pr(r\vert B=b) & = b \\
\forall a,b \, \pr(r \vert A =a, B = b) & =  \alpha a + \beta b
\end{align}

## Imprecise probabilism and the motivations

### The main claim

\justify Represent uncertainty by  a set of probability measures.

\vspace{1mm}
\scriptsize

[@Levi1974;@Grdenfors1982;@Jeffrey1983-JEFBWA-2;@walley1991statistical;@Kaplan1996;@Joyce2005;@Grove2009;@elkin2017imprecise]

\pause

### Snowing in Boston?

\vspace{-1mm}

Say, (.5,1).

\vspace{1mm}
\scriptsize

[@Mahtani2019Imprecise]

\pause

### Weight of evidence?

\vspace{-1mm}

The more evidence you have, the more narrow the interval.

\vspace{1mm}
\scriptsize

[@Kaplan1996;@Joyce2005;@Sturgeon2008-STURAT]

\pause

<!-- ## Imprecise probabilism and the motivations -->

### Agnosticisim?

\vspace{-1mm}

[0,1] interval instead of a single uniform distribution.

\pause

### Multiple sources?

\vspace{-1mm}

Just put all those probabilities in one set.

## Epistemological challenges

\pause

### Radical uncertainty and weight

Say one experiment observed 40/100  and  another 8/15 recoveries.

Simply putting the frequencies in a set won't do justice to the evidence.

\scriptsize 
\vspace{1mm}

[@Bradley2019Imprecise]

\pause

### Dilation

  \justify Whenever your roomate goes to the track he flips a fair coin. If it comes up heads he bets on  Speedy. Otherwise he places no bets. He comes home from the track grinning.  You're now more uncertain about  $\pr(\mathsf{heads})$ than before.

\scriptsize

\vspace{1mm}
 [@walley1991statistical;@seidenfeld1993;@White2009-WHIESA;@Topey2012;@pedersen2014demystifying;@Hart2015;@Bradley2016]

## Epistemological challenges

### Belief inertia

Say you're fully agnostic about $H$: $P(H) = [0,1]$.

\begin{center}
```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "70%", warning=FALSE, message = FALSE, dpi = 800}
p <- ggplot(data = data.frame(value = 1:7), 
            mapping = aes(x = value))+ theme_tufte(base_size = 24)+
  theme(axis.line.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.x=element_blank(),
        panel.grid.minor.x=element_blank(),
        panel.grid.major.x=element_blank())+
  ylab("probability")+xlim(c(1.8,6.2))

p + geom_segment(aes(x = 2, y = 0, xend = 2, yend = 1), 
                 lineend = "square")+
   geom_segment(aes(x = 6, y = 0, xend = 6, yend = 1), 
                 lineend = "square")+
   annotate("text", x = 1.8, y = 0, label = TeX("$p_F$"), size = 7 )+
  annotate("text", x = 1.8, y = 1, label = TeX("$p_T$"), size = 7)+
  annotate("text", x = 1.8, y = .4, label = TeX("$p_{1}$"), size = 7)+
  annotate("text", x = 1.8, y = .6, label = TeX("$p_{2}$"), size = 7)+
  geom_segment(aes(x = 1.95, y = 0, xend = 2.05, yend = 0), 
             lineend = "square")+
  geom_segment(aes(x = 1.95, y = 0.4, xend = 2.05, yend = 0.4), 
               lineend = "square")+
  geom_segment(aes(x = 1.95, y = 0.6, xend = 2.05, yend = 0.6), 
               lineend = "square")+
  geom_segment(aes(x = 1.95, y = 1, xend = 2.05, yend = 1), 
               lineend = "square") + #now the second one
  annotate("text", x = 6.2, y = 0, label = TeX("$p_F'$"), size = 7)+
  annotate("text", x = 6.2, y = 1, label = TeX("$p_T'$"), size = 7)+
  annotate("text", x = 6.2, y = .2, label = TeX("$p_{1}'$"), size = 7)+
  annotate("text", x = 6.2, y = .5, label = TeX("$p_{2}'$"), size = 7)+
  geom_segment(aes(x = 5.95, y = 0, xend = 6.05, yend = 0), 
               lineend = "square")+
  geom_segment(aes(x = 5.95, y = 0.2, xend = 6.05, yend = 0.2), 
               lineend = "square")+
  geom_segment(aes(x = 5.95, y = 0.5, xend = 6.05, yend = 0.5), 
               lineend = "square")+
  geom_segment(aes(x = 5.95, y = 1, xend = 6.05, yend = 1), 
               lineend = "square")+ 
                 geom_segment(aes(x = 2.1, xend = 5.9, y = 0,
                yend = 0), arrow = arrow(length = unit(0.08, "inches")),
                color = "red", size = 0.6)+ 
  geom_segment(aes(x = 2.1, xend = 5.9, y = 1,
                   yend = 1), arrow = arrow(length = unit(0.08, "inches")),
               color = "red", size = 0.6)+ 
  geom_segment(aes(x = 2.1, xend = 5.9, y = 0.4,
                   yend = .2), arrow = arrow(length = unit(0.08, "inches")),
               color = "red", size = 0.6)+ 
  geom_segment(aes(x = 2.1, xend = 5.9, y = 0.6,
                   yend = .5), arrow = arrow(length = unit(0.08, "inches")),
               color = "red", size = 0.6)+ 
  geom_segment(aes(x = 2.2, xend = 5.8, y = 0.1, yend = 0.06), arrow = arrow(length = unit(0.06, "inches")),
               color = "skyblue", size = 0.3)+ 
  geom_segment(aes(x = 2.2, xend = 5.8, y = 0.2, yend = 0.1), arrow = arrow(length = unit(0.06, "inches")),
               color = "skyblue", size = 0.3)+ 
  geom_segment(aes(x = 2.2, xend = 5.8, y = 0.3, yend = 0.14), arrow = arrow(length = unit(0.06, "inches")),
               color = "skyblue", size = 0.3)+
  geom_segment(aes(x = 2.2, xend = 5.8, y = 0.45, yend = 0.28), arrow = arrow(length = unit(0.06, "inches")),
                                                            color = "skyblue", size = 0.3)+ 
  geom_segment(aes(x = 2.2, xend = 5.8, y = 0.5, yend = 0.36), arrow = arrow(length = unit(0.06, "inches")),
               color = "skyblue", size = 0.3)+ 
  geom_segment(aes(x = 2.2, xend = 5.8, y = 0.55, yend = 0.45), arrow = arrow(length = unit(0.06, "inches")),
               color = "skyblue", size = 0.3)+
  geom_segment(aes(x = 2.2, xend = 5.8, y = 0.65, yend = 0.55), arrow = arrow(length = unit(0.06, "inches")),
               color = "skyblue", size = 0.3)+ 
  geom_segment(aes(x = 2.2, xend = 5.8, y = 0.7, yend = 0.65), arrow = arrow(length = unit(0.06, "inches")),
               color = "skyblue", size = 0.3)+ 
  geom_segment(aes(x = 2.2, xend = 5.8, y = 0.75, yend = 0.72), arrow = arrow(length = unit(0.06, "inches")),
               color = "skyblue", size = 0.3)+ 
  geom_segment(aes(x = 2.2, xend = 5.8, y = 0.79, yend = 0.8), arrow = arrow(length = unit(0.06, "inches")),
               color = "skyblue", size = 0.3)+ 
  geom_segment(aes(x = 2.2, xend = 5.8, y = 0.84, yend = 0.86), arrow = arrow(length = unit(0.06, "inches")),
               color = "skyblue", size = 0.3)+ 
  geom_segment(aes(x = 2.2, xend = 5.8, y = 0.9, yend = 0.93), arrow = arrow(length = unit(0.06, "inches")),
               color = "skyblue", size = 0.3)
```
\end{center}

<!-- \begin{center} -->
<!-- \includegraphics[width = 6cm]{inertia.jpg} -->
<!-- \end{center} -->

Then $\pr(H \vert E) =[0,1]$, no matter what the evidence says.

\vspace{1mm}
\scriptsize

[@Joyce2010,@Dodd2012;@Rinard2013;@Bradley2014;@Vallinder2017;@Moss2020]

## Epistemological challenges

### Mystery coins

- A fair draw of one of two biased coins, with biases $.4, and .6$.

\pause

\vspace{-2mm}

- Imprecise reaction: $\{.4, .6\}$.

\pause

\vspace{-2mm}

- Precise probabilism: expected bias of $.5$.

\pause

### Question

As far as accuracy is concerned, why prefer the imprecise way?

Say the actual bias is .4. A first stab:

\vspace{-3mm}

\begin{align*}
\mathsf{inaccuracy(imprecise)} & = \frac{0 + .2}{2} = .1\\
\mathsf{inaccuracy(precise)} & =  .1\\
\end{align*}

\vspace{-5mm}
\scriptsize

[@Seidenfeld2012forecasting; @MayoWilson2016; @schoenfield2017AccuracyRationalityImprecise]

## Higher-order approach

### The mystery coin with  $(.2,.8)$ and 1 head in five tosses

```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "90%"}
theta <- c(.4, .6) #possible parameter values
priorTheta <- c(.2,.8) #priors assigned to them
data <- c(1,0,0,0,0) #1 heads with five tosses
z = sum(data) # number of 1's in Data
N = length(data) #number of tosses

# Compute the Bernoulli likelihood at each value of Theta:
pDataIfTheta <- dbinom(z,N,theta)
# Compute the probability of th evidence and 
# the posterior via Bayes' rule:
pData <- sum(pDataIfTheta * priorTheta)
pThetaIfData <- pDataIfTheta * priorTheta / pData

#put together
table <- data.frame(theta,  priorTheta,
                     pData,  pDataIfTheta, pThetaIfData)

#plot of the prior
prior <- ggplot(table, aes(x = theta, y = priorTheta))+  
  geom_bar(stat="identity", alpha = .8)+ 
  scale_x_continuous(breaks = theta)+th+ylim(c(0,1))+ggtitle("Prior")

#plot of the likelihood
likelihood <-  ggplot(table, aes(x = theta, y = pDataIfTheta))+
  geom_bar(stat="identity", alpha = .8)+ 
  scale_x_continuous(breaks = theta)+
  th+ylim(c(0,1))+ggtitle("Likelihood")

#plot of the posterior
posterior <- ggplot(table, aes(x = theta, y = pThetaIfData))+  #set up
  geom_bar(stat="identity", alpha = .8)+  #barplot layer
  scale_x_continuous(breaks = theta)+
  th+ylim(c(0,1))+ggtitle("Posterior")

#convert the table to graphics, transpose for readability
tableGG <- ggtexttable(t(round(table,3)))

#plot together
ggarrange(prior, likelihood, posterior, tableGG)
```

## Higher-order approach

### Two experts, uniform prior

```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "90%"}
p <- ggplot(data = data.frame(probability = seq(0,1, by = .25)),mapping = aes(x = probability))+th +ylab("density")+scale_x_continuous(breaks = seq(0,1, by = .25))+ylim(c(0,10))

agentPrior <- function(p) dbeta(p,1,1) 
al <- function(p) dbeta(p,5,20) 
bert <- function(p) dbeta(p,8,12) 
posterior <- function(p) dbeta(p,19,62)

priorPlot <- p + stat_function(fun=agentPrior)+ggtitle("Agent\'s prior")
alPlot <- p + stat_function(fun=al)+ggtitle("If Al says .2")

bertPlot <- p + stat_function(fun=bert)+ggtitle("If Bert says .4")
posterior <- p + stat_function(fun=posterior)+ggtitle("Agent's posterior")

ggarrange(priorPlot, alPlot, bertPlot, posterior)
```

## The key conjecture

### With higher-order methods, it will be possible to:

- match the representation of uncertainty with the evidence,

- combine various sources of uncertainty without running into the known problems,

- use existing mathematics to make sense of accuracy in such contexts,

- capture reasoning with radical uncertainty without dilation and belief inertia.

## Tasks (discussion)

\footnotesize

- Relate this representation to weight of evidence.

\vspace{-2mm}

- Go beyond grid approximation and beta representation.

- Represent  agnosticism so that:

    - agnosticism  preservation holds, and

    - inertia fails.

\vspace{-2mm}

- Represent opinion aggregation more generally, compare to known limitations and features of already studied opinion pooling methods.

\vspace{-2mm}

- Investigate dilation-like phenomena in this context.

\vspace{-2mm}

- Study known distance functions between probabilistic measures as candidates for key elements of accuracy measures.

\vspace{-2mm}

- Explain away the apparent contrast between evidence-grounding and accuracy considerations [@easwaranAndFitelson2015].

\begin{center}
\Large Thank you for your attention!
\end{center}

## References
\tiny

### 2021-07-27 presentationESSLLI21/presentationESSLLI.tex
added presentation template

\documentclass[10pt,ignorenonframetext,x11names, dvipsnames, bibspacing,natbib]{beamer}
\setbeamertemplate{caption}[numbered]
\setbeamertemplate{caption label separator}{: }
\setbeamercolor{caption name}{fg=normal text.fg}
\beamertemplatenavigationsymbolsempty
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
\usetheme[]{Rafal_beamerSly1}
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\newif\ifbibliography
\hypersetup{
            pdftitle={A bayesian method of cosine-based word2vec bias estimation},
            pdfauthor={Alicja Dobrzeniecka \& Rafal Urbaniak (LoPSE research group, University of Gdansk)},
            colorlinks=true,
            linkcolor=Maroon,
            citecolor=Blue,
            urlcolor=blue,
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls

% Prevent slide breaks in the middle of a paragraph:
\widowpenalties 1 10000
\raggedbottom

\AtBeginPart{
  \let\insertpartnumber\relax
  \let\partname\relax
  \frame{\partpage}
}
\AtBeginSection{
  \ifbibliography
  \else
    \let\insertsectionnumber\relax
    \let\sectionname\relax
    \frame{\sectionpage}
  \fi
}
\AtBeginSubsection{
  \let\insertsubsectionnumber\relax
  \let\subsectionname\relax
  \frame{\subsectionpage}
}

\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}

\title{\large A bayesian method of cosine-based word2vec bias estimation}
\author{Alicja Dobrzeniecka \& Rafal Urbaniak \footnotesize \newline (LoPSE
research group, University of Gdansk)}
\date{ExpSem2021, ESSLLI}

\begin{document}
\frame{\titlepage}

\begin{frame}

\frametitle{The gist of the project}

\pause

\begin{block}{Classical probabilism in epistemology (and in
applications)}

Represent uncertainty by a single probabilistic measure (point
estimates).

\vspace{-3mm}

\begin{align*}
\pr(\mathsf{recovery}\vert \mathsf{treatment}) & = .6\\
\end{align*}

\vspace{-6mm}

\pause

\end{block}

\begin{block}{Imprecise probabilism}

Evidence (e.g.~expert evidence) sometimes fails to warrant such
precision.

\vspace{-3mm}

\begin{align*}
\pr(\mathsf{recovery}\vert \mathsf{treatment}) & \in [.4,.6]\\
\end{align*}

\vspace{-6mm}

\pause

\end{block}

\begin{block}{Epistemological difficulties of imprecise probabilism}

\begin{itemize}
\item
  hard to model the weight of and combine evidence.
\item
  inertia of agnosticism.
\item
  growth of uncertainty in light of new evidence.
\end{itemize}

\end{block}

\end{frame}

\begin{frame}{The gist of the project}

\begin{block}{My novel contribution}

\begin{itemize}
\item
  Meet the desiderata motivating imprecision.
\item
  Avoid these difficulties.
\item
  Do so by using higher-order probabilities.
\end{itemize}

\begin{center}\includegraphics[width=0.7\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-1-1} \end{center}

\end{block}

\end{frame}

\begin{frame}{Motivations for imprecision \large (straighforward
considerations)}

\pause

\begin{block}{Radical lack of information}

\(\pr(\mathsf{It \,\, will \,\, snow\,\, in\,\, Boston\,\, on \,\, New\,\, Year's\,\, Eve\,\, 2026}) =\)
?

\pause

\end{block}

\begin{block}{Weight of evidence}

\begin{itemize}
\tightlist
\item
  You've seen 4/10 patients recover. \(\pr(\mathsf{recovery})=.4\)?
\end{itemize}

\pause
\vspace{-2mm}

\begin{itemize}
\tightlist
\item
  You've seen 40/100 patients recover. \(\pr(\mathsf{recovery})=.4\)?
\end{itemize}

\footnotesize (You could try to use classical statistical tools, such as
confidence intervals,\linebreak  but they have their own problems.)

\end{block}

\end{frame}

\begin{frame}{Motivations for imprecision \large (uniformity and
agnosticism)}

\begin{block}{Agnosticism preservation requirement}

Agnosticism about uniquely connected variables should be the same.

\vspace{-5mm}

\[X \mbox{ and } Y=X+2 \mbox{ or } Z = X^2 \,\,\, \mbox{ for $10>X>0$.}\]

\pause

\end{block}

\begin{block}{The trouble}

\begin{center}\includegraphics[width=0.6\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-2-1} \end{center}

\vspace{-9mm}

\begin{align*}\pr(X<5) = .5 = \pr(X^2 < 25)
\end{align*}

\end{block}

\end{frame}

\begin{frame}{Motivations for imprecision \large (uniformity and
agnosticism)}

\begin{block}{The trouble}

\begin{center}\includegraphics[width=0.6\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-3-1} \end{center}

\end{block}

\end{frame}

\begin{frame}{Motivations for imprecision \large (expert evidence)}

\begin{block}{Weather forecasters}

\vspace{-1mm}

\begin{tabular}{ll}
Al: & $\pr(\mathsf{It \,\,  will\,\,  rain\,\,  tomorrow})=.6$\\
Bert: & $\pr(\mathsf{It \,\,  will\,\,  rain\,\,  tomorrow})=.8$
\end{tabular}

\pause

\end{block}

\begin{block}{Linear pooling}

\vspace{-4mm}

\begin{align*}
\pr(\mathsf{It \,\,  will\,\,  rain\,\,  tomorrow}) & = 
\alpha \times .6 + \beta \times .8 \,\,\,\,\,  (where \,\, \alpha + \beta = 1.)
\end{align*}

\pause

\end{block}

\begin{block}{Mathematical impossibilities}

\begin{itemize}
\item
  Experts's agreement on independence is not preserved by linear
  pooling. ({\textbf{???}})
\item
  You can't at the same time hold the following: ({\textbf{???}})
\end{itemize}

\vspace{-7mm}

\begin{align}
\pr(A=B) & <1\\
\pr(r\vert A=a) & = a \\
\pr(r\vert B=b) & = b \\
\forall a,b \, \pr(r \vert A =a, B = b) & =  \alpha a + \beta b
\end{align}

\end{block}

\end{frame}

\begin{frame}{Imprecise probabilism and the motivations}

\begin{block}{The main claim}

\justify Represent uncertainty by a set of probability measures.

\vspace{1mm} \scriptsize

({\textbf{???}}, {\textbf{???}}, {\textbf{???}}, {\textbf{???}},
{\textbf{???}}, {\textbf{???}}, {\textbf{???}}; Walley,
\protect\hyperlink{ref-walley1991statistical}{1991})

\pause

\end{block}

\begin{block}{Snowing in Boston?}

\vspace{-1mm}

Say, (.5,1).

\vspace{1mm} \scriptsize

({\textbf{???}})

\pause

\end{block}

\begin{block}{Weight of evidence?}

\vspace{-1mm}

The more evidence you have, the more narrow the interval.

\vspace{1mm} \scriptsize

({\textbf{???}}, {\textbf{???}}, {\textbf{???}})

\pause

\end{block}

\begin{block}{Agnosticisim?}

\vspace{-1mm}

{[}0,1{]} interval instead of a single uniform distribution.

\pause

\end{block}

\begin{block}{Multiple sources?}

\vspace{-1mm}

Just put all those probabilities in one set.

\end{block}

\end{frame}

\begin{frame}{Epistemological challenges}

\pause

\begin{block}{Radical uncertainty and weight}

Say one experiment observed 40/100 and another 8/15 recoveries.

Simply putting the frequencies in a set won't do justice to the
evidence.

\scriptsize  \vspace{1mm}

({\textbf{???}})

\pause

\end{block}

\begin{block}{Dilation}

\justify Whenever your roomate goes to the track he flips a fair coin.
If it comes up heads he bets on Speedy. Otherwise he places no bets. He
comes home from the track grinning. You're now more uncertain about
\(\pr(\mathsf{heads})\) than before.

\scriptsize

\vspace{1mm} ({\textbf{???}}, {\textbf{???}}, {\textbf{???}},
{\textbf{???}}, {\textbf{???}}, {\textbf{???}}; Walley,
\protect\hyperlink{ref-walley1991statistical}{1991})

\end{block}

\end{frame}

\begin{frame}{Epistemological challenges}

\begin{block}{Belief inertia}

Say you're fully agnostic about \(H\): \(P(H) = [0,1]\).

\begin{center}

\begin{center}\includegraphics[width=0.7\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-4-1} \end{center}
\end{center}

Then \(\pr(H \vert E) =[0,1]\), no matter what the evidence says.

\vspace{1mm} \scriptsize

({\textbf{???}},({\textbf{???}}), {\textbf{???}}, {\textbf{???}},
{\textbf{???}}, {\textbf{???}})

\end{block}

\end{frame}

\begin{frame}{Epistemological challenges}

\begin{block}{Mystery coins}

\begin{itemize}
\tightlist
\item
  A fair draw of one of two biased coins, with biases \(.4, and .6\).
\end{itemize}

\pause

\vspace{-2mm}

\begin{itemize}
\tightlist
\item
  Imprecise reaction: \(\{.4, .6\}\).
\end{itemize}

\pause

\vspace{-2mm}

\begin{itemize}
\tightlist
\item
  Precise probabilism: expected bias of \(.5\).
\end{itemize}

\pause

\end{block}

\begin{block}{Question}

As far as accuracy is concerned, why prefer the imprecise way?

Say the actual bias is .4. A first stab:

\vspace{-3mm}

\begin{align*}
\mathsf{inaccuracy(imprecise)} & = \frac{0 + .2}{2} = .1\\
\mathsf{inaccuracy(precise)} & =  .1\\
\end{align*}

\vspace{-5mm} \scriptsize

({\textbf{???}}, {\textbf{???}}; Schoenfield,
\protect\hyperlink{ref-schoenfield2017AccuracyRationalityImprecise}{2017})

\end{block}

\end{frame}

\begin{frame}{Higher-order approach}

\begin{block}{The mystery coin with \((.2,.8)\) and 1 head in five
tosses}

\begin{center}\includegraphics[width=0.9\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-5-1} \end{center}

\end{block}

\end{frame}

\begin{frame}{Higher-order approach}

\begin{block}{Two experts, uniform prior}

\begin{center}\includegraphics[width=0.9\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-6-1} \end{center}

\end{block}

\end{frame}

\begin{frame}{The key conjecture}

\begin{block}{With higher-order methods, it will be possible to:}

\begin{itemize}
\item
  match the representation of uncertainty with the evidence,
\item
  combine various sources of uncertainty without running into the known
  problems,
\item
  use existing mathematics to make sense of accuracy in such contexts,
\item
  capture reasoning with radical uncertainty without dilation and belief
  inertia.
\end{itemize}

\end{block}

\end{frame}

\begin{frame}{Tasks (discussion)}

\footnotesize

\begin{itemize}
\tightlist
\item
  Relate this representation to weight of evidence.
\end{itemize}

\vspace{-2mm}

\begin{itemize}
\item
  Go beyond grid approximation and beta representation.
\item
  Represent agnosticism so that:

  \begin{itemize}
  \item
    agnosticism preservation holds, and
  \item
    inertia fails.
  \end{itemize}
\end{itemize}

\vspace{-2mm}

\begin{itemize}
\tightlist
\item
  Represent opinion aggregation more generally, compare to known
  limitations and features of already studied opinion pooling methods.
\end{itemize}

\vspace{-2mm}

\begin{itemize}
\tightlist
\item
  Investigate dilation-like phenomena in this context.
\end{itemize}

\vspace{-2mm}

\begin{itemize}
\tightlist
\item
  Study known distance functions between probabilistic measures as
  candidates for key elements of accuracy measures.
\end{itemize}

\vspace{-2mm}

\begin{itemize}
\tightlist
\item
  Explain away the apparent contrast between evidence-grounding and
  accuracy considerations (Easwaran \& Fitelson,
  \protect\hyperlink{ref-easwaranAndFitelson2015}{2015}).
\end{itemize}

\begin{center}
\Large Thank you for your attention!
\end{center}

\end{frame}

\begin{frame}{References}

\tiny

\hypertarget{refs}{}
\hypertarget{ref-easwaranAndFitelson2015}{}
Easwaran, K., \& Fitelson, B. (2015). Accuracy, coherence, and evidence.
In T. S. Gendler \& J. Hawthorne (Eds.), \emph{Oxford studies in
epistemology} (Vol. 5). Oxford University Press.

\hypertarget{ref-schoenfield2017AccuracyRationalityImprecise}{}
Schoenfield, M. (2017). The Accuracy and Rationality of Imprecise
Credences: The Accuracy and Rationality of Imprecise Credences.
\emph{Noûs}, \emph{51}(4), 667--685.
\url{https://doi.org/10.1111/nous.12105}

\hypertarget{ref-walley1991statistical}{}
Walley, P. (1991). \emph{Statistical reasoning with imprecise
probabilities}. Chapman; Hall London.

\end{frame}

\end{document}

### 2021-07-30 paperDraft/paperDraft.Rmd
started presentation

@@
 \noindent
-$s(X,Y,A,B)$ is the statistic used in the signifcance test, and the $p$-value obtained by bootstrapping: it is the frequency of $s(X_i,Y_i,A,B)>s(X,Y,A,B)$ for all equally sized partitions $X_i, Y_i$ of $X\cup Y$.
+$s(X,Y,A,B)$ is the statistic used in the significance test, and the $p$-value obtained by bootstrapping: it is the frequency of $s(X_i,Y_i,A,B)>s(X,Y,A,B)$ for all equally sized partitions $X_i, Y_i$ of $X\cup Y$.
 The effect size is computed by normalizing the difference in means as follows:
@@
 # The problem with means of means
+
+<!-- A <- c(1,5,10) -->
+<!-- B <- c(6, 8, 15) -->
+
+<!-- c(A,B) -->
+
+<!-- sd(c(A,B)) -->
+
+<!-- mA <- mean(A) -->
+<!-- mB <- mean(B) -->
+
+<!-- sd(c(mA,mB)) -->
 

### 2021-07-30 presentationESSLLI21/presentationESSLLI.Rmd
started presentation

@@
 ---
-title: "\\large A bayesian method of cosine-based word2vec bias estimation"
+title: "\\large Taking uncertainty in word embedding bias estimation seriously \\ a  Bayesian
+  approach"
 author: "Alicja Dobrzeniecka \\& Rafal  Urbaniak \\footnotesize \\newline (LoPSE research group, University of  Gdansk)" 
@@
 urlcolor: blue
-bibliography: [../references/bibliographyRafal.bib]
-csl: [../references/apa-6th-edition.csl] 
+bibliography: [../references/cosineReferences1.bib]
+csl: ../references/apa-6th-edition.csl 
 link-citations: yes
@@
 
-\frametitle{The gist of the project}
+## Cosine-based measures of bias
+
+### Word embeddings
+
+- Representation of words (or words in contexts) with vectors of real numbers
+
+- Built to predict the probability of co-occurence
+
+word | 1 | 2 | 3 | 4 | ... 
+|---|---|---|---|---|---
+woman | 0.456 | 0.267 | 0.675 | 0.131| ...
+man | 0.451 | 0.897 | 0.472 | 0.088| ...
+
+## Cosine-based measures of bias
+
+### Cosine similarity \& distance
+
+<!-- - Dot product for normalized vectors -->
+
+<!-- often used in discussion of bias -->
+
+\vspace{-4mm}
+
+\begin{align} \tag{Sim}
+\mathsf{cosineSimilarity}(A,B) & = \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert}
+\\
+\tag{Distance}
+\mathsf{cosineDistance}(A,B) &  = 1 - \mathsf{cosineSimilarity}(A,B)
+\end{align}
+
+-  Geometric interpretation: direction (not length)
+
+-  $\mathsf{cosineDistance}\in (0, 2)$
+
+- Naive interpretation: proximity corresponds to semantic similarity (e.g. no triangle inequality)
+
+<!-- - Note cosine distance fails to meet triangle inequality -->
+
+## Cosine-based measures of bias
+
+### The worry
+
+In the learning process these models can learn implicit biases that reflect harmful stereotypical thinking
 
@@
 
-### Classical probabilism in epistemology (and in applications)
+### Cosine-based bias: basic intuition
 
-Represent uncertainty by a single probabilistic measure (point estimates).
+Words belonging to an intuitively harmful stereotype are cosine-close to each other
 
-\vspace{-3mm}
+## Cosine-based measures of bias
+
+### Stereotypical lists
+
+\footnotesize
+
+- feminine occupations: "homemaker", "nurse", "receptionist", "librarian", etc.
+
+- masculine occupations:  "maestro",  "captain", "architect", etc.
+
+\normalsize
+
+### Visual example
+
+\vspace{1mm}
+\footnotesize
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "60%"}
+genderProjections <-  read.csv("../datasets/genderProjections.csv")[,-1]
+colnames(genderProjections)[4] <- "Stereotype" 
+levels(genderProjections$Stereotype) <- c("male occupation", "female occupation")
+ggplot(genderProjections, aes(x=he,y=she, color = Stereotype, shape = Stereotype))+geom_point(size = 6, alpha = 0.75)+theme_tufte(base_size = 27)+labs(title ="GloVe  on Wikipedia 2014 and Gigaword 5th ed.")+scale_color_manual(values =c("orangered4","chartreuse4"))+
+  theme(legend.position = c(0.8,.15)) +
+    scale_shape_manual(values=c(15, 19)) 
+```
+\normalsize
+
+## Cosine-based measures of bias
+
+### Example: direct bias
+
+- The gender bias of a word $w$ is its projection on the gender direction $\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})$
+
+<!-- (the gender direction is the top principal compontent of ten gender pair difference vectors).  -->
+
+-  Given the (ideally) gender neutral words $N$ and the gender direction $g$ the direct gender bias is:
+<!-- ($c$ is a parameter determining how strict we want to be): -->
+
+\vspace{-2mm}
+
+\begin{align}
+\mathsf{directBias_c(N,g)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},g)\vert^c}{\vert N \vert }
+\end{align}
+
+\footnotesize
+
+[@Bolukbasi2016Man]
+
+## Cosine-based measures of bias
+
+### Example: Word Embedding Association Test (WEAT)
 
 \begin{align*}
-\pr(\mathsf{recovery}\vert \mathsf{treatment}) & = .6\\
+s(t,A,B) & = \frac{\sum_{a\in A}f(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}f(t,b)}{\vert B\vert}
+\\
+WEAT(A,B) & = \frac{
+\mu\left(\{s(x,A,B)\}_{x\in X}\right) -\mu\left(\{s(y,A,B)\}_{y\in Y}\right) 
+}{
+\sigma\left(\{s(w,A,B)\}_{w\in X\cup Y}\right)
+}
 \end{align*}
 
-\vspace{-6mm}
+-  $t$ is a term,  $A, B$ are sets of stereotype attribute words, $X, $Y$ are protected group words
+
+- For instance, $X$ might be a set of male names, $Y$ a set of female names, $A$ might contain stereotypically male-related career words, and $B$ stereotypically female-related family words
+
+- $s$-values are used as datapoints in statistical significance tests
+
+\footnotesize
+
+[@Caliskan2017semanticsBiases] with extensions in [@Lauscher2019multidimensional] and applications in [@Garg2018years]
+
+## Cosine-based measures of bias
+
+### Our main target: Mean Average Cosine Similarity (MAC)
+
+<!-- - more general, multi-class setting -->
+
+\begin{align*}
+S(t_i, A_j) & = \frac{1}{\vert A_j\vert}\sum_{a\in A_j}\mathsf{cos}(t,a) \\
+MAC(T,A) & = \frac{1}{\vert T \vert \,\vert A\vert}\sum_{t_i \in T }\sum_{A_j \in A} S(t_i,A_j)
+\end{align*}
+
+- $T = \{t_1, \dots, t_k\}$ is a class of protected words
+
+- each $A_j\in A$ is a set of attributes stereotypically associated with a protected word
+
+<!-- - For each protected word $T$ and each attribute class, they first take the mean for this protected word and all attributes in a given attribute class, and then take the mean of thus obtained means for all the protected words.
+-->
+
+- The t-tests they employ are run on average cosines used to calculate MAC
+
+\footnotesize
+
+[@Manzini2019blackToCriminal]
+
+## Cosine-based measures of bias
+
+### Our main target: Mean Average Cosine Similarity (MAC)
+
+\footnotesize 
+```{r religionTableHeadEarly,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+religion <- read.csv("../datasets/religionReddit.csv")[-1]
+colnames(religion) <- c("protectedWord","wordToCompare","wordClass",
+                        "cosineDistance","cosineSimilarity","connection")
+religion$wordClass <- as.factor(religion$wordClass)
+levels(religion$wordClass) <- c("christian","human","jewish","muslim","neutral")
+religionSmall <- religion[religion$wordClass != "neutral" & religion$wordClass != "human" ,]
+set.seed(127)
+religionSmall <- religionSmall[sample(1:nrow(religionSmall),6),-c(3,6)]
+rownames(religionSmall) <- c()
+religionSmall  %>%  kable(format = "latex",booktabs=T,
+                      linesep = "",  escape = FALSE, 
+                      caption = "Rows the religion dataset.") %>%
+                      kable_styling(latex_options=c("scale_down"))
+```
+\normalsize
+
+## Cosine-based measures of bias
+
+### Known challenges
+
+- Gender-direction might be an indicator of bias, but is insufficient. After debiasing other non-gendered words can remain in biased relations  [@Gonen2019lipstick]
+
+- Methods which involve analogies and their evaluations by human users on Mechanical Turk are unreliable [@Nissim2020fair]
+
+<!-- dodac? -->
+
+## Some methodological problems
+
+###  Word list choice is unprincipled
+
+We run with it for comparison
 
@@
 
-### Imprecise probabilism
+### No design considerations to sample size
 
-Evidence (e.g. expert evidence) sometimes fails to warrant such precision.
+We investigate the uncertainty that arises from raw sample sizes
 
-\vspace{-3mm}
+## Some methodological problems
 
-\begin{align*}
-\pr(\mathsf{recovery}\vert \mathsf{treatment}) & \in [.4,.6]\\
-\end{align*}
+### No word class distinction and no control group
 
-\vspace{-6mm}
+We make the subclasses clear, add human neutral predicates and neutral predicates for control
+
+\footnotesize 
+```{r religionTableHeadLate,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+religion <- read.csv("../datasets/religionReddit.csv")[-1]
+colnames(religion) <- c("protectedWord","wordToCompare","wordClass",
+                        "cosineDistance","cosineSimilarity","connection")
+religion$wordClass <- as.factor(religion$wordClass)
+levels(religion$wordClass) <- c("christian","human","jewish","muslim","neutral")
+religionSmall <- religion[religion$wordClass != "neutral" & religion$wordClass != "human" ,]
+religionControl <- religion[religion$wordClass == "neutral" | religion$wordClass == "human" ,]
+set.seed(123154196)
+religionDisplayA <- religionSmall[sample(1:nrow(religionSmall),4),]
+religionDisplayB <- religionControl[sample(1:nrow(religionControl),4),]
+religionDisplay <- rbind(religionDisplayA,religionDisplayB)
+religionDisplay$cosineDistance <- round(religionDisplay$cosineDistance,3)
+religionDisplay$cosineSimilarity <- round(religionDisplay$cosineSimilarity,3)
+rownames(religionDisplay) <- c()
+religionDisplay  %>%  kable(format = "latex",booktabs=T,
+                      linesep = "",  escape = FALSE, 
+                      caption = "Rows from extended religion dataset.") %>%
+                      kable_styling(latex_options=c("scale_down"))
+```
+\normalsize
+
+## Some methodological problems
+
+### Outliers and surprisingly dissimilar words
+
+We study those by  visualizations and uncertainty estimates
 
@@
 
-### Epistemological difficulties of imprecise probabilism
+### No principled interpretation
 
-- hard to model the weight of and combine evidence.
+Religion Debiasing  | MAC (distance)
+------------- | -------------
+Biased        | 0.859
+Hard Debiased | 0.934
+Soft Debiased ($\lambda$ = 0.2) | 0.894
 
-- inertia of agnosticism.
+What values are sufficient for the presence of bias and what differences are sign of real improvement? Low $p$-values are not high effect indicators!
 
-- growth of uncertainty  in light of new evidence.
+We compare HPDIs.
 
-## The gist of the project
+<!--  What may attract attention is the fact that the value of cosine distance in "Biased" category is already quite high even before debiasing.
+High cosine distance indicates low cosine similarity between values.
+One could think that average cosine similarity equal to approximately 0.141 is not significant enough to consider it as bias.
+However the authors aim to mitigate "biases" in vectors with such great distance to make it even larger.
+Methodologically the question is, on what basis is this small similarity still considered as a proof of the presence of bias, and whether these small changes are meaningful.
+This is in general the problem of scale and the lack of universal intervals.
+In contrast, statistical intervals will help us decide whether a given cosine similarity is high enough to consider the words to be more similar than if we chose them at random.
+We will use highest posterior density intervals, in line with Bayesian methodology.
+-->
 
-### My novel contribution
+## The problem with pre-averaging
 
-- Meet the desiderata motivating imprecision.
+##  Bayesian estimation
 
-- Avoid these difficulties.
-
-- Do so by using higher-order probabilities.
-
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "70%"}
-p <- ggplot(data = data.frame(probability = seq(0,1, by = .25)),mapping = aes(x = probability))+th +ylab("density")+scale_x_continuous(breaks = seq(0,1, by = .1))
-
-agentPrior <- function(p) dbeta(p,28,28)
-
-priorPlot <- p + stat_function(fun=agentPrior)+ggtitle("Uncertainty about recovery")+xlab("probability of recovery if treated")
-
-priorPlot
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "105%",message =FALSE, warning = FALSE}
+library(tidyverse)
+muslimWords <- c("imam","islam","mosque","muslim","quran")
+muslim <- religion %>% filter(protectedWord %in% muslimWords)
+muslimClass <- muslim %>% filter(protectedWord == "muslim")
+neutralSample <- sample_n(filter(muslimClass,connection == "none"), 5)
+humanSample <- sample_n(filter(muslimClass,connection == "human"), 5)
+muslimVis <- muslimClass %>% filter(connection != "none" & connection !="human")
+muslimVis <- rbind(muslimVis,neutralSample,humanSample)
+#we plug in our visualisation script
+source("../functions/visualisationTools.R")
+#two arguments: dataset and protected word
+visualiseProtected(muslimVis,"muslim")+theme_tufte(base_size = 20)
 ```
 
-## Motivations for imprecision  \large (straighforward considerations)
+##  Bayesian estimation
 
-\pause
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "105%"}
+library(tidyverse)
+priestClass <- religion %>% filter(protectedWord == "priest")
+neutralSample <- sample_n(filter(priestClass,connection == "none"), 5)
+humanSample <- sample_n(filter(priestClass,connection == "human"), 5)
+priestVis <- priestClass %>% filter(connection != "none" & connection !="human")
+priestVis <- rbind(priestVis,neutralSample,humanSample)
 
-### Radical lack of information
+#we plug in our visualisation script
+source("../functions/visualisationTools.R")
+#two arguments: dataset and protected word
+visualiseProtected(priestVis,"priest")+theme_tufte(base_size = 20)
+```
 
-$\pr(\mathsf{It \,\, will \,\,  snow\,\,  in\,\,  Boston\,\,  on \,\, New\,\,  Year's\,\,  Eve\,\,  2026}) =$ ?
+## Effects of debiasing
 
-\pause
+### Further work
 
-<!-- ### Mere bounds in reasoning -->
+- downstream tasks
 
-<!-- - Say $\pr(A) = .5$, $\pr(B) = .3$, but you don't know if they're independent. -->
-
-<!-- \vspace{-2mm} -->
-
-<!-- - We can only calculate bounds for $\pr(A \et B)$, $\pr(A \vee B)$, etc. -->
+<!-- \frametitle{The gist of the project} -->
 
@@
 
-### Weight of evidence
+<!-- ### Classical probabilism in epistemology (and in applications) -->
 
-- You've seen 4/10 patients recover. $\pr(\mathsf{recovery})=.4$?
+<!-- Represent uncertainty by a single probabilistic measure (point estimates). -->
 
-\pause
-\vspace{-2mm}
+<!-- \vspace{-3mm} -->
 
-- You've seen 40/100 patients recover. $\pr(\mathsf{recovery})=.4$?
+<!-- \begin{align*} -->
+<!-- \pr(\mathsf{recovery}\vert \mathsf{treatment}) & = .6\\ -->
+<!-- \end{align*} -->
 
-\footnotesize (You could try to use classical statistical tools, such as confidence intervals,\linebreak  but they have their own problems.)
+<!-- \vspace{-6mm} -->
 
-## Motivations for imprecision \large (uniformity and agnosticism)
+<!-- \pause -->
 
-### Agnosticism preservation requirement
+<!-- ### Imprecise probabilism -->
 
- Agnosticism about uniquely connected variables should be the same.
+<!-- Evidence (e.g. expert evidence) sometimes fails to warrant such precision. -->
 
-\vspace{-5mm}
+<!-- \vspace{-3mm} -->
 
- \[X \mbox{ and } Y=X+2 \mbox{ or } Z = X^2 \,\,\, \mbox{ for $10>X>0$.}\]
+<!-- \begin{align*} -->
+<!-- \pr(\mathsf{recovery}\vert \mathsf{treatment}) & \in [.4,.6]\\ -->
+<!-- \end{align*} -->
 
-\pause
+<!-- \vspace{-6mm} -->
 
-### The trouble
+<!-- \pause  -->
 
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "60%"}
-p <- ggplot(data = data.frame(value = 0:10), 
-            mapping = aes(x = value))
+<!-- ### Epistemological difficulties of imprecise probabilism -->
 
-unifX <-  function(value) punif(value,min = 0, max = 10)
+<!-- - hard to model the weight of and combine evidence. -->
 
-p +   stat_function(fun=unifX)+
-  ylab("density")+theme_tufte(base_size = 25)+labs(title = "Uniform cumulative probability for X in (0,10)")+xlab("X")+ylab("Cumulative probability")+scale_x_continuous(breaks = 0:10)+
-  geom_segment(aes(x = 5, y = 0, xend = 5, yend = .5), lty = 2, alpha = .2, size = .4, color = "skyblue")+
-  geom_segment(aes(x = 0, y = .5, xend = 5, yend = .5), lty = 2, alpha = .2, size = .4, color = "skyblue")
+<!-- - inertia of agnosticism. -->
 
-```
+<!-- - growth of uncertainty  in light of new evidence. -->
 
-\vspace{-9mm}
+<!-- ## The gist of the project -->
 
-\begin{align*}\pr(X<5) = .5 = \pr(X^2 < 25)
-\end{align*}
+<!-- ### My novel contribution -->
 
-## Motivations for imprecision \large (uniformity and agnosticism)
+<!-- - Meet the desiderata motivating imprecision. -->
 
-### The trouble
+<!-- - Avoid these difficulties. -->
 
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "60%", message = FALSE, warning = FALSE}
-p <- ggplot(data = data.frame(value = 0:10), 
-            mapping = aes(x = value))
+<!-- - Do so by using higher-order probabilities.  -->
 
-unifY <-  function(value) punif(sqrt(value),min = 0, max = 10)
+<!-- ```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "70%"} -->
+<!-- p <- ggplot(data = data.frame(probability = seq(0,1, by = .25)),mapping = aes(x = probability))+th +ylab("density")+scale_x_continuous(breaks = seq(0,1, by = .1)) -->
 
-p +   stat_function(fun=unifY)+
-  ylab("cumulative probability")+theme_tufte(base_size = 25)+
-  labs(title = TeX("Resulting cumulative probability for $X^2$"))+scale_x_continuous(breaks = 0:100)+xlab(TeX("X^2"))+xlim(c(0,100))+
-  geom_segment(aes(x = 25, y = 0, xend = 25, yend = .5), lty = 2, alpha = .2, size = .4, color = "skyblue")+
-  geom_segment(aes(x = 0, y = .5, xend = 25, yend = .5), lty = 2, alpha = .2, size = .4, color = "skyblue")
+<!-- agentPrior <- function(p) dbeta(p,28,28)  -->
 
-```
+<!-- priorPlot <- p + stat_function(fun=agentPrior)+ggtitle("Uncertainty about recovery")+xlab("probability of recovery if treated") -->
 
-## Motivations for imprecision \large (expert evidence)
+<!-- priorPlot -->
+<!-- ``` -->
 
-### Weather forecasters
+<!-- ## Motivations for imprecision  \large (straighforward considerations) -->
 
-\vspace{-1mm}
+<!-- \pause -->
 
-\begin{tabular}{ll}
-Al: & $\pr(\mathsf{It \,\,  will\,\,  rain\,\,  tomorrow})=.6$\\
-Bert: & $\pr(\mathsf{It \,\,  will\,\,  rain\,\,  tomorrow})=.8$
-\end{tabular}
+<!-- ### Radical lack of information -->
 
-\pause
+<!-- $\pr(\mathsf{It \,\, will \,\,  snow\,\,  in\,\,  Boston\,\,  on \,\, New\,\,  Year's\,\,  Eve\,\,  2026}) =$ ? -->
 
-### Linear pooling
+<!-- \pause -->
 
-\vspace{-4mm}
+<!-- <!-- ### Mere bounds in reasoning -->
 
-\begin{align*}
-\pr(\mathsf{It \,\,  will\,\,  rain\,\,  tomorrow}) & = 
-\alpha \times .6 + \beta \times .8 \,\,\,\,\,  (where \,\, \alpha + \beta = 1.)
-\end{align*}
+<!-- <!-- - Say $\pr(A) = .5$, $\pr(B) = .3$, but you don't know if they're independent. -->
 
-\pause
+<!-- <!-- \vspace{-2mm} -->
 
-### Mathematical impossibilities
+<!-- <!-- - We can only calculate bounds for $\pr(A \et B)$, $\pr(A \vee B)$, etc. -->
 
-- Experts's agreement on independence is not preserved by linear pooling. [@List2011Group]
+<!-- <!-- \pause  -->
 
-- You can't at the same time hold the following: [@Gallow2018No]
+<!-- ### Weight of evidence -->
 
-\vspace{-7mm}
+<!-- - You've seen 4/10 patients recover. $\pr(\mathsf{recovery})=.4$? -->
 
-\begin{align}
-\pr(A=B) & <1\\
-\pr(r\vert A=a) & = a \\
-\pr(r\vert B=b) & = b \\
-\forall a,b \, \pr(r \vert A =a, B = b) & =  \alpha a + \beta b
-\end{align}
+<!-- \pause -->
+<!-- \vspace{-2mm} -->
 
-## Imprecise probabilism and the motivations
+<!-- - You've seen 40/100 patients recover. $\pr(\mathsf{recovery})=.4$? -->
 
-### The main claim
+<!-- \footnotesize (You could try to use classical statistical tools, such as confidence intervals,\linebreak  but they have their own problems.) -->
 
-\justify Represent uncertainty by  a set of probability measures.
+<!-- ## Motivations for imprecision \large (uniformity and agnosticism) -->
 
-\vspace{1mm}
-\scriptsize
+<!-- ### Agnosticism preservation requirement -->
 
-[@Levi1974;@Grdenfors1982;@Jeffrey1983-JEFBWA-2;@walley1991statistical;@Kaplan1996;@Joyce2005;@Grove2009;@elkin2017imprecise]
+<!--  Agnosticism about uniquely connected variables should be the same.   -->
 
-\pause
+<!-- \vspace{-5mm} -->
 
-### Snowing in Boston?
+<!--  \[X \mbox{ and } Y=X+2 \mbox{ or } Z = X^2 \,\,\, \mbox{ for $10>X>0$.}\]  -->
 
-\vspace{-1mm}
+<!-- \pause -->
 
-Say, (.5,1).
+<!-- ### The trouble -->
 
-\vspace{1mm}
-\scriptsize
+<!-- ```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "60%"} -->
+<!-- p <- ggplot(data = data.frame(value = 0:10),  -->
+<!--             mapping = aes(x = value))  -->
 
-[@Mahtani2019Imprecise]
+<!-- unifX <-  function(value) punif(value,min = 0, max = 10)  -->
 
-\pause
+<!-- p +   stat_function(fun=unifX)+ -->
+<!--   ylab("density")+theme_tufte(base_size = 25)+labs(title = "Uniform cumulative probability for X in (0,10)")+xlab("X")+ylab("Cumulative probability")+scale_x_continuous(breaks = 0:10)+ -->
+<!--   geom_segment(aes(x = 5, y = 0, xend = 5, yend = .5), lty = 2, alpha = .2, size = .4, color = "skyblue")+ -->
+<!--   geom_segment(aes(x = 0, y = .5, xend = 5, yend = .5), lty = 2, alpha = .2, size = .4, color = "skyblue") -->
 
-### Weight of evidence?
+<!-- ``` -->
 
-\vspace{-1mm}
+<!-- \vspace{-9mm} -->
 
-The more evidence you have, the more narrow the interval.
+<!-- \begin{align*}\pr(X<5) = .5 = \pr(X^2 < 25) -->
+<!-- \end{align*} -->
 
-\vspace{1mm}
-\scriptsize
+<!-- ## Motivations for imprecision \large (uniformity and agnosticism) -->
 
-[@Kaplan1996;@Joyce2005;@Sturgeon2008-STURAT]
+<!-- ### The trouble -->
 
-\pause
+<!-- ```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "60%", message = FALSE, warning = FALSE} -->
+<!-- p <- ggplot(data = data.frame(value = 0:10),  -->
+<!--             mapping = aes(x = value))  -->
+
+<!-- unifY <-  function(value) punif(sqrt(value),min = 0, max = 10)  -->
+
+<!-- p +   stat_function(fun=unifY)+ -->
+<!--   ylab("cumulative probability")+theme_tufte(base_size = 25)+ -->
+<!--   labs(title = TeX("Resulting cumulative probability for $X^2$"))+scale_x_continuous(breaks = 0:100)+xlab(TeX("X^2"))+xlim(c(0,100))+ -->
+<!--   geom_segment(aes(x = 25, y = 0, xend = 25, yend = .5), lty = 2, alpha = .2, size = .4, color = "skyblue")+ -->
+<!--   geom_segment(aes(x = 0, y = .5, xend = 25, yend = .5), lty = 2, alpha = .2, size = .4, color = "skyblue") -->
+
+<!-- ``` -->
+
+<!-- ## Motivations for imprecision \large (expert evidence) -->
+
+<!-- ### Weather forecasters -->
+
+<!-- \vspace{-1mm} -->
+
+<!-- \begin{tabular}{ll} -->
+<!-- Al: & $\pr(\mathsf{It \,\,  will\,\,  rain\,\,  tomorrow})=.6$\\ -->
+<!-- Bert: & $\pr(\mathsf{It \,\,  will\,\,  rain\,\,  tomorrow})=.8$ -->
+<!-- \end{tabular} -->
+
+<!-- \pause -->
+
+<!-- ### Linear pooling -->
+
+<!-- \vspace{-4mm} -->
+
+<!-- \begin{align*} -->
+<!-- \pr(\mathsf{It \,\,  will\,\,  rain\,\,  tomorrow}) & =  -->
+<!-- \alpha \times .6 + \beta \times .8 \,\,\,\,\,  (where \,\, \alpha + \beta = 1.) -->
+<!-- \end{align*} -->
+
+<!-- \pause -->
+
+<!-- ### Mathematical impossibilities -->
+
+<!-- - Experts's agreement on independence is not preserved by linear pooling. [@List2011Group] -->
+
+<!-- - You can't at the same time hold the following: [@Gallow2018No] -->
+
+<!-- \vspace{-7mm} -->
+
+<!-- \begin{align} -->
+<!-- \pr(A=B) & <1\\ -->
+<!-- \pr(r\vert A=a) & = a \\ -->
+<!-- \pr(r\vert B=b) & = b \\ -->
+<!-- \forall a,b \, \pr(r \vert A =a, B = b) & =  \alpha a + \beta b -->
+<!-- \end{align} -->
 
@@
 
-### Agnosticisim?
+<!-- ### The main claim -->
 
-\vspace{-1mm}
+<!-- \justify Represent uncertainty by  a set of probability measures. -->
 
-[0,1] interval instead of a single uniform distribution.
+<!-- \vspace{1mm} -->
+<!-- \scriptsize  -->
 
-\pause
+<!-- [@Levi1974;@Grdenfors1982;@Jeffrey1983-JEFBWA-2;@walley1991statistical;@Kaplan1996;@Joyce2005;@Grove2009;@elkin2017imprecise] -->
 
-### Multiple sources?
+<!-- \pause -->
 
-\vspace{-1mm}
+<!-- ### Snowing in Boston?  -->
 
-Just put all those probabilities in one set.
+<!-- \vspace{-1mm} -->
 
-## Epistemological challenges
+<!-- Say, (.5,1). -->
 
-\pause
+<!-- \vspace{1mm} -->
+<!-- \scriptsize  -->
 
-### Radical uncertainty and weight
+<!-- [@Mahtani2019Imprecise] -->
 
-Say one experiment observed 40/100  and  another 8/15 recoveries.
+<!-- \pause -->
 
-Simply putting the frequencies in a set won't do justice to the evidence.
+<!-- ### Weight of evidence?  -->
 
-\scriptsize 
-\vspace{1mm}
+<!-- \vspace{-1mm} -->
 
-[@Bradley2019Imprecise]
+<!-- The more evidence you have, the more narrow the interval. -->
 
-\pause
+<!-- \vspace{1mm} -->
+<!-- \scriptsize  -->
 
-### Dilation
+<!-- [@Kaplan1996;@Joyce2005;@Sturgeon2008-STURAT] -->
 
-  \justify Whenever your roomate goes to the track he flips a fair coin.
+<!-- \pause -->
+
+<!-- <!-- ## Imprecise probabilism and the motivations -->
+
+<!-- ### Agnosticisim? -->
+
+<!-- \vspace{-1mm} -->
+
+<!-- [0,1] interval instead of a single uniform distribution. -->
+
+<!-- \pause -->
+
+<!-- ### Multiple sources?  -->
+
+<!-- \vspace{-1mm} -->
+
+<!-- Just put all those probabilities in one set. -->
+
+<!-- ## Epistemological challenges -->
+
+<!-- \pause -->
+
+<!-- ### Radical uncertainty and weight -->
+
+<!-- Say one experiment observed 40/100  and  another 8/15 recoveries.  -->
+
+<!-- Simply putting the frequencies in a set won't do justice to the evidence.  -->
+
+<!-- \scriptsize  -->
+<!-- \vspace{1mm} -->
+
+<!-- [@Bradley2019Imprecise] -->
+
+<!-- \pause -->
+
+<!-- ### Dilation  -->
+
+<!--   \justify Whenever your roomate goes to the track he flips a fair coin.
 If it comes up heads he bets on  Speedy.
@@
 You're now more uncertain about  $\pr(\mathsf{heads})$ than before.
+-->
 
-\scriptsize
+<!-- \scriptsize  -->
 
-\vspace{1mm}
- [@walley1991statistical;@seidenfeld1993;@White2009-WHIESA;@Topey2012;@pedersen2014demystifying;@Hart2015;@Bradley2016]
+<!-- \vspace{1mm} -->
+<!--  [@walley1991statistical;@seidenfeld1993;@White2009-WHIESA;@Topey2012;@pedersen2014demystifying;@Hart2015;@Bradley2016] -->
 
-## Epistemological challenges
+<!-- ## Epistemological challenges -->
 
-### Belief inertia
+<!-- ### Belief inertia -->
 
-Say you're fully agnostic about $H$: $P(H) = [0,1]$.
-
-\begin{center}
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "70%", warning=FALSE, message = FALSE, dpi = 800}
-p <- ggplot(data = data.frame(value = 1:7), 
-            mapping = aes(x = value))+ theme_tufte(base_size = 24)+
-  theme(axis.line.x=element_blank(),
-        axis.text.x=element_blank(),
-        axis.ticks.x=element_blank(),
-        axis.title.x=element_blank(),
-        panel.grid.minor.x=element_blank(),
-        panel.grid.major.x=element_blank())+
-  ylab("probability")+xlim(c(1.8,6.2))
-
-p + geom_segment(aes(x = 2, y = 0, xend = 2, yend = 1), 
-                 lineend = "square")+
-   geom_segment(aes(x = 6, y = 0, xend = 6, yend = 1), 
-                 lineend = "square")+
-   annotate("text", x = 1.8, y = 0, label = TeX("$p_F$"), size = 7 )+
-  annotate("text", x = 1.8, y = 1, label = TeX("$p_T$"), size = 7)+
-  annotate("text", x = 1.8, y = .4, label = TeX("$p_{1}$"), size = 7)+
-  annotate("text", x = 1.8, y = .6, label = TeX("$p_{2}$"), size = 7)+
-  geom_segment(aes(x = 1.95, y = 0, xend = 2.05, yend = 0), 
-             lineend = "square")+
-  geom_segment(aes(x = 1.95, y = 0.4, xend = 2.05, yend = 0.4), 
-               lineend = "square")+
-  geom_segment(aes(x = 1.95, y = 0.6, xend = 2.05, yend = 0.6), 
-               lineend = "square")+
-  geom_segment(aes(x = 1.95, y = 1, xend = 2.05, yend = 1), 
-               lineend = "square") + #now the second one
-  annotate("text", x = 6.2, y = 0, label = TeX("$p_F'$"), size = 7)+
-  annotate("text", x = 6.2, y = 1, label = TeX("$p_T'$"), size = 7)+
-  annotate("text", x = 6.2, y = .2, label = TeX("$p_{1}'$"), size = 7)+
-  annotate("text", x = 6.2, y = .5, label = TeX("$p_{2}'$"), size = 7)+
-  geom_segment(aes(x = 5.95, y = 0, xend = 6.05, yend = 0), 
-               lineend = "square")+
-  geom_segment(aes(x = 5.95, y = 0.2, xend = 6.05, yend = 0.2), 
-               lineend = "square")+
-  geom_segment(aes(x = 5.95, y = 0.5, xend = 6.05, yend = 0.5), 
-               lineend = "square")+
-  geom_segment(aes(x = 5.95, y = 1, xend = 6.05, yend = 1), 
-               lineend = "square")+ 
-                 geom_segment(aes(x = 2.1, xend = 5.9, y = 0,
-                yend = 0), arrow = arrow(length = unit(0.08, "inches")),
-                color = "red", size = 0.6)+ 
-  geom_segment(aes(x = 2.1, xend = 5.9, y = 1,
-                   yend = 1), arrow = arrow(length = unit(0.08, "inches")),
-               color = "red", size = 0.6)+ 
-  geom_segment(aes(x = 2.1, xend = 5.9, y = 0.4,
-                   yend = .2), arrow = arrow(length = unit(0.08, "inches")),
-               color = "red", size = 0.6)+ 
-  geom_segment(aes(x = 2.1, xend = 5.9, y = 0.6,
-                   yend = .5), arrow = arrow(length = unit(0.08, "inches")),
-               color = "red", size = 0.6)+ 
-  geom_segment(aes(x = 2.2, xend = 5.8, y = 0.1, yend = 0.06), arrow = arrow(length = unit(0.06, "inches")),
-               color = "skyblue", size = 0.3)+ 
-  geom_segment(aes(x = 2.2, xend = 5.8, y = 0.2, yend = 0.1), arrow = arrow(length = unit(0.06, "inches")),
-               color = "skyblue", size = 0.3)+ 
-  geom_segment(aes(x = 2.2, xend = 5.8, y = 0.3, yend = 0.14), arrow = arrow(length = unit(0.06, "inches")),
-               color = "skyblue", size = 0.3)+
-  geom_segment(aes(x = 2.2, xend = 5.8, y = 0.45, yend = 0.28), arrow = arrow(length = unit(0.06, "inches")),
-                                                            color = "skyblue", size = 0.3)+ 
-  geom_segment(aes(x = 2.2, xend = 5.8, y = 0.5, yend = 0.36), arrow = arrow(length = unit(0.06, "inches")),
-               color = "skyblue", size = 0.3)+ 
-  geom_segment(aes(x = 2.2, xend = 5.8, y = 0.55, yend = 0.45), arrow = arrow(length = unit(0.06, "inches")),
-               color = "skyblue", size = 0.3)+
-  geom_segment(aes(x = 2.2, xend = 5.8, y = 0.65, yend = 0.55), arrow = arrow(length = unit(0.06, "inches")),
-               color = "skyblue", size = 0.3)+ 
-  geom_segment(aes(x = 2.2, xend = 5.8, y = 0.7, yend = 0.65), arrow = arrow(length = unit(0.06, "inches")),
-               color = "skyblue", size = 0.3)+ 
-  geom_segment(aes(x = 2.2, xend = 5.8, y = 0.75, yend = 0.72), arrow = arrow(length = unit(0.06, "inches")),
-               color = "skyblue", size = 0.3)+ 
-  geom_segment(aes(x = 2.2, xend = 5.8, y = 0.79, yend = 0.8), arrow = arrow(length = unit(0.06, "inches")),
-               color = "skyblue", size = 0.3)+ 
-  geom_segment(aes(x = 2.2, xend = 5.8, y = 0.84, yend = 0.86), arrow = arrow(length = unit(0.06, "inches")),
-               color = "skyblue", size = 0.3)+ 
-  geom_segment(aes(x = 2.2, xend = 5.8, y = 0.9, yend = 0.93), arrow = arrow(length = unit(0.06, "inches")),
-               color = "skyblue", size = 0.3)
-```
-\end{center}
+<!-- Say you're fully agnostic about $H$: $P(H) = [0,1]$.  -->
 
 <!-- \begin{center} -->
-<!-- \includegraphics[width = 6cm]{inertia.jpg} -->
+<!-- ```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "70%", warning=FALSE, message = FALSE, dpi = 800} -->
+<!-- p <- ggplot(data = data.frame(value = 1:7),  -->
+<!--             mapping = aes(x = value))+ theme_tufte(base_size = 24)+ -->
+<!--   theme(axis.line.x=element_blank(), -->
+<!--         axis.text.x=element_blank(), -->
+<!--         axis.ticks.x=element_blank(), -->
+<!--         axis.title.x=element_blank(), -->
+<!--         panel.grid.minor.x=element_blank(), -->
+<!--         panel.grid.major.x=element_blank())+ -->
+<!--   ylab("probability")+xlim(c(1.8,6.2)) -->
+
+<!-- p + geom_segment(aes(x = 2, y = 0, xend = 2, yend = 1),  -->
+<!--                  lineend = "square")+ -->
+<!--    geom_segment(aes(x = 6, y = 0, xend = 6, yend = 1),  -->
+<!--                  lineend = "square")+ -->
+<!--    annotate("text", x = 1.8, y = 0, label = TeX("$p_F$"), size = 7 )+ -->
+<!--   annotate("text", x = 1.8, y = 1, label = TeX("$p_T$"), size = 7)+ -->
+<!--   annotate("text", x = 1.8, y = .4, label = TeX("$p_{1}$"), size = 7)+ -->
+<!--   annotate("text", x = 1.8, y = .6, label = TeX("$p_{2}$"), size = 7)+ -->
+<!--   geom_segment(aes(x = 1.95, y = 0, xend = 2.05, yend = 0),  -->
+<!--              lineend = "square")+ -->
+<!--   geom_segment(aes(x = 1.95, y = 0.4, xend = 2.05, yend = 0.4),  -->
+<!--                lineend = "square")+ -->
+<!--   geom_segment(aes(x = 1.95, y = 0.6, xend = 2.05, yend = 0.6),  -->
+<!--                lineend = "square")+ -->
+<!--   geom_segment(aes(x = 1.95, y = 1, xend = 2.05, yend = 1),  -->
+<!--                lineend = "square") + #now the second one -->
+<!--   annotate("text", x = 6.2, y = 0, label = TeX("$p_F'$"), size = 7)+ -->
+<!--   annotate("text", x = 6.2, y = 1, label = TeX("$p_T'$"), size = 7)+ -->
+<!--   annotate("text", x = 6.2, y = .2, label = TeX("$p_{1}'$"), size = 7)+ -->
+<!--   annotate("text", x = 6.2, y = .5, label = TeX("$p_{2}'$"), size = 7)+ -->
+<!--   geom_segment(aes(x = 5.95, y = 0, xend = 6.05, yend = 0),  -->
+<!--                lineend = "square")+ -->
+<!--   geom_segment(aes(x = 5.95, y = 0.2, xend = 6.05, yend = 0.2),  -->
+<!--                lineend = "square")+ -->
+<!--   geom_segment(aes(x = 5.95, y = 0.5, xend = 6.05, yend = 0.5),  -->
+<!--                lineend = "square")+ -->
+<!--   geom_segment(aes(x = 5.95, y = 1, xend = 6.05, yend = 1),  -->
+<!--                lineend = "square")+  -->
+<!--                  geom_segment(aes(x = 2.1, xend = 5.9, y = 0, -->
+<!--                 yend = 0), arrow = arrow(length = unit(0.08, "inches")), -->
+<!--                 color = "red", size = 0.6)+  -->
+<!--   geom_segment(aes(x = 2.1, xend = 5.9, y = 1, -->
+<!--                    yend = 1), arrow = arrow(length = unit(0.08, "inches")), -->
+<!--                color = "red", size = 0.6)+  -->
+<!--   geom_segment(aes(x = 2.1, xend = 5.9, y = 0.4, -->
+<!--                    yend = .2), arrow = arrow(length = unit(0.08, "inches")), -->
+<!--                color = "red", size = 0.6)+  -->
+<!--   geom_segment(aes(x = 2.1, xend = 5.9, y = 0.6, -->
+<!--                    yend = .5), arrow = arrow(length = unit(0.08, "inches")), -->
+<!--                color = "red", size = 0.6)+  -->
+<!--   geom_segment(aes(x = 2.2, xend = 5.8, y = 0.1, yend = 0.06), arrow = arrow(length = unit(0.06, "inches")), -->
+<!--                color = "skyblue", size = 0.3)+  -->
+<!--   geom_segment(aes(x = 2.2, xend = 5.8, y = 0.2, yend = 0.1), arrow = arrow(length = unit(0.06, "inches")), -->
+<!--                color = "skyblue", size = 0.3)+  -->
+<!--   geom_segment(aes(x = 2.2, xend = 5.8, y = 0.3, yend = 0.14), arrow = arrow(length = unit(0.06, "inches")), -->
+<!--                color = "skyblue", size = 0.3)+ -->
+<!--   geom_segment(aes(x = 2.2, xend = 5.8, y = 0.45, yend = 0.28), arrow = arrow(length = unit(0.06, "inches")), -->
+<!--                                                             color = "skyblue", size = 0.3)+  -->
+<!--   geom_segment(aes(x = 2.2, xend = 5.8, y = 0.5, yend = 0.36), arrow = arrow(length = unit(0.06, "inches")), -->
+<!--                color = "skyblue", size = 0.3)+  -->
+<!--   geom_segment(aes(x = 2.2, xend = 5.8, y = 0.55, yend = 0.45), arrow = arrow(length = unit(0.06, "inches")), -->
+<!--                color = "skyblue", size = 0.3)+ -->
+<!--   geom_segment(aes(x = 2.2, xend = 5.8, y = 0.65, yend = 0.55), arrow = arrow(length = unit(0.06, "inches")), -->
+<!--                color = "skyblue", size = 0.3)+  -->
+<!--   geom_segment(aes(x = 2.2, xend = 5.8, y = 0.7, yend = 0.65), arrow = arrow(length = unit(0.06, "inches")), -->
+<!--                color = "skyblue", size = 0.3)+  -->
+<!--   geom_segment(aes(x = 2.2, xend = 5.8, y = 0.75, yend = 0.72), arrow = arrow(length = unit(0.06, "inches")), -->
+<!--                color = "skyblue", size = 0.3)+  -->
+<!--   geom_segment(aes(x = 2.2, xend = 5.8, y = 0.79, yend = 0.8), arrow = arrow(length = unit(0.06, "inches")), -->
+<!--                color = "skyblue", size = 0.3)+  -->
+<!--   geom_segment(aes(x = 2.2, xend = 5.8, y = 0.84, yend = 0.86), arrow = arrow(length = unit(0.06, "inches")), -->
+<!--                color = "skyblue", size = 0.3)+  -->
+<!--   geom_segment(aes(x = 2.2, xend = 5.8, y = 0.9, yend = 0.93), arrow = arrow(length = unit(0.06, "inches")), -->
+<!--                color = "skyblue", size = 0.3) -->
+<!-- ``` -->
 <!-- \end{center} -->
 
-Then $\pr(H \vert E) =[0,1]$, no matter what the evidence says.
+<!-- <!-- \begin{center} --> 
+<!-- <!-- \includegraphics[width = 6cm]{inertia.jpg} --> 
+<!-- <!-- \end{center} -->
 
-\vspace{1mm}
-\scriptsize
+<!-- Then $\pr(H \vert E) =[0,1]$, no matter what the evidence says.  -->
 
-[@Joyce2010,@Dodd2012;@Rinard2013;@Bradley2014;@Vallinder2017;@Moss2020]
+<!-- \vspace{1mm} -->
+<!-- \scriptsize -->
 
-## Epistemological challenges
+<!-- [@Joyce2010,@Dodd2012;@Rinard2013;@Bradley2014;@Vallinder2017;@Moss2020] -->
 
-### Mystery coins
+<!-- ## Epistemological challenges -->
 
-- A fair draw of one of two biased coins, with biases $.4, and .6$.
+<!-- ### Mystery coins -->
 
-\pause
+<!-- - A fair draw of one of two biased coins, with biases $.4, and .6$. -->
 
-\vspace{-2mm}
+<!-- \pause -->
 
-- Imprecise reaction: $\{.4, .6\}$.
+<!-- \vspace{-2mm} -->
 
-\pause
+<!-- - Imprecise reaction: $\{.4, .6\}$.  -->
 
-\vspace{-2mm}
+<!-- \pause -->
 
-- Precise probabilism: expected bias of $.5$.
+<!-- \vspace{-2mm} -->
 
-\pause
+<!-- - Precise probabilism: expected bias of $.5$. -->
 
-### Question
+<!-- \pause -->
 
-As far as accuracy is concerned, why prefer the imprecise way?
+<!-- ### Question -->
 
-Say the actual bias is .4. A first stab:
+<!-- As far as accuracy is concerned, why prefer the imprecise way? -->
 
-\vspace{-3mm}
+<!-- Say the actual bias is .4. A first stab: -->
 
-\begin{align*}
-\mathsf{inaccuracy(imprecise)} & = \frac{0 + .2}{2} = .1\\
-\mathsf{inaccuracy(precise)} & =  .1\\
-\end{align*}
+<!-- \vspace{-3mm} -->
 
-\vspace{-5mm}
-\scriptsize
+<!-- \begin{align*} -->
+<!-- \mathsf{inaccuracy(imprecise)} & = \frac{0 + .2}{2} = .1\\ -->
+<!-- \mathsf{inaccuracy(precise)} & =  .1\\ -->
+<!-- \end{align*} -->
 
-[@Seidenfeld2012forecasting; @MayoWilson2016; @schoenfield2017AccuracyRationalityImprecise]
+<!-- \vspace{-5mm} -->
+<!-- \scriptsize -->
 
-## Higher-order approach
+<!-- [@Seidenfeld2012forecasting; @MayoWilson2016; @schoenfield2017AccuracyRationalityImprecise] -->
 
-### The mystery coin with  $(.2,.8)$ and 1 head in five tosses
+<!-- ## Higher-order approach -->
 
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "90%"}
-theta <- c(.4, .6) #possible parameter values
-priorTheta <- c(.2,.8) #priors assigned to them
-data <- c(1,0,0,0,0) #1 heads with five tosses
-z = sum(data) # number of 1's in Data
-N = length(data) #number of tosses
+<!-- ### The mystery coin with  $(.2,.8)$ and 1 head in five tosses -->
 
-# Compute the Bernoulli likelihood at each value of Theta:
-pDataIfTheta <- dbinom(z,N,theta)
-# Compute the probability of th evidence and 
-# the posterior via Bayes' rule:
-pData <- sum(pDataIfTheta * priorTheta)
-pThetaIfData <- pDataIfTheta * priorTheta / pData
+<!-- ```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "90%"} -->
+<!-- theta <- c(.4, .6) #possible parameter values -->
+<!-- priorTheta <- c(.2,.8) #priors assigned to them -->
+<!-- data <- c(1,0,0,0,0) #1 heads with five tosses -->
+<!-- z = sum(data) # number of 1's in Data -->
+<!-- N = length(data) #number of tosses -->
 
-#put together
-table <- data.frame(theta,  priorTheta,
-                     pData,  pDataIfTheta, pThetaIfData)
+<!-- # Compute the Bernoulli likelihood at each value of Theta: -->
+<!-- pDataIfTheta <- dbinom(z,N,theta) -->
+<!-- # Compute the probability of th evidence and  -->
+<!-- # the posterior via Bayes' rule: -->
+<!-- pData <- sum(pDataIfTheta * priorTheta) -->
+<!-- pThetaIfData <- pDataIfTheta * priorTheta / pData -->
 
-#plot of the prior
-prior <- ggplot(table, aes(x = theta, y = priorTheta))+  
-  geom_bar(stat="identity", alpha = .8)+ 
-  scale_x_continuous(breaks = theta)+th+ylim(c(0,1))+ggtitle("Prior")
+<!-- #put together -->
+<!-- table <- data.frame(theta,  priorTheta, -->
+<!--                      pData,  pDataIfTheta, pThetaIfData) -->
 
-#plot of the likelihood
-likelihood <-  ggplot(table, aes(x = theta, y = pDataIfTheta))+
-  geom_bar(stat="identity", alpha = .8)+ 
-  scale_x_continuous(breaks = theta)+
-  th+ylim(c(0,1))+ggtitle("Likelihood")
+<!-- #plot of the prior -->
+<!-- prior <- ggplot(table, aes(x = theta, y = priorTheta))+   -->
+<!--   geom_bar(stat="identity", alpha = .8)+  -->
+<!--   scale_x_continuous(breaks = theta)+th+ylim(c(0,1))+ggtitle("Prior") -->
 
-#plot of the posterior
-posterior <- ggplot(table, aes(x = theta, y = pThetaIfData))+  #set up
-  geom_bar(stat="identity", alpha = .8)+  #barplot layer
-  scale_x_continuous(breaks = theta)+
-  th+ylim(c(0,1))+ggtitle("Posterior")
+<!-- #plot of the likelihood -->
+<!-- likelihood <-  ggplot(table, aes(x = theta, y = pDataIfTheta))+ -->
+<!--   geom_bar(stat="identity", alpha = .8)+  -->
+<!--   scale_x_continuous(breaks = theta)+ -->
+<!--   th+ylim(c(0,1))+ggtitle("Likelihood") -->
 
-#convert the table to graphics, transpose for readability
-tableGG <- ggtexttable(t(round(table,3)))
+<!-- #plot of the posterior -->
+<!-- posterior <- ggplot(table, aes(x = theta, y = pThetaIfData))+  #set up -->
+<!--   geom_bar(stat="identity", alpha = .8)+  #barplot layer -->
+<!--   scale_x_continuous(breaks = theta)+ -->
+<!--   th+ylim(c(0,1))+ggtitle("Posterior") -->
 
-#plot together
-ggarrange(prior, likelihood, posterior, tableGG)
-```
+<!-- #convert the table to graphics, transpose for readability -->
+<!-- tableGG <- ggtexttable(t(round(table,3))) -->
 
-## Higher-order approach
+<!-- #plot together -->
+<!-- ggarrange(prior, likelihood, posterior, tableGG) -->
+<!-- ``` -->
 
-### Two experts, uniform prior
+<!-- ## Higher-order approach -->
 
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "90%"}
-p <- ggplot(data = data.frame(probability = seq(0,1, by = .25)),mapping = aes(x = probability))+th +ylab("density")+scale_x_continuous(breaks = seq(0,1, by = .25))+ylim(c(0,10))
+<!-- ### Two experts, uniform prior -->
 
-agentPrior <- function(p) dbeta(p,1,1) 
-al <- function(p) dbeta(p,5,20) 
-bert <- function(p) dbeta(p,8,12) 
-posterior <- function(p) dbeta(p,19,62)
+<!-- ```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "90%"} -->
+<!-- p <- ggplot(data = data.frame(probability = seq(0,1, by = .25)),mapping = aes(x = probability))+th +ylab("density")+scale_x_continuous(breaks = seq(0,1, by = .25))+ylim(c(0,10)) -->
 
-priorPlot <- p + stat_function(fun=agentPrior)+ggtitle("Agent\'s prior")
-alPlot <- p + stat_function(fun=al)+ggtitle("If Al says .2")
+<!-- agentPrior <- function(p) dbeta(p,1,1)  -->
+<!-- al <- function(p) dbeta(p,5,20)  -->
+<!-- bert <- function(p) dbeta(p,8,12)  -->
+<!-- posterior <- function(p) dbeta(p,19,62)  -->
 
-bertPlot <- p + stat_function(fun=bert)+ggtitle("If Bert says .4")
-posterior <- p + stat_function(fun=posterior)+ggtitle("Agent's posterior")
+<!-- priorPlot <- p + stat_function(fun=agentPrior)+ggtitle("Agent\'s prior") -->
+<!-- alPlot <- p + stat_function(fun=al)+ggtitle("If Al says .2") -->
 
-ggarrange(priorPlot, alPlot, bertPlot, posterior)
-```
+<!-- bertPlot <- p + stat_function(fun=bert)+ggtitle("If Bert says .4") -->
+<!-- posterior <- p + stat_function(fun=posterior)+ggtitle("Agent's posterior") -->
 
-## The key conjecture
+<!-- ggarrange(priorPlot, alPlot, bertPlot, posterior) -->
+<!-- ``` -->
 
-### With higher-order methods, it will be possible to:
+<!-- ## The key conjecture -->
 
-- match the representation of uncertainty with the evidence,
+<!-- ### With higher-order methods, it will be possible to: -->
 
-- combine various sources of uncertainty without running into the known problems,
+<!-- - match the representation of uncertainty with the evidence, -->
 
-- use existing mathematics to make sense of accuracy in such contexts,
+<!-- - combine various sources of uncertainty without running into the known problems, -->
 
-- capture reasoning with radical uncertainty without dilation and belief inertia.
+<!-- - use existing mathematics to make sense of accuracy in such contexts, -->
 
-## Tasks (discussion)
+<!-- - capture reasoning with radical uncertainty without dilation and belief inertia. -->
 
-\footnotesize
+<!-- ## Tasks (discussion) -->
 
-- Relate this representation to weight of evidence.
+<!-- \footnotesize  -->
 
-\vspace{-2mm}
+<!-- - Relate this representation to weight of evidence. -->
 
-- Go beyond grid approximation and beta representation.
+<!-- \vspace{-2mm} -->
 
-- Represent  agnosticism so that:
+<!-- - Go beyond grid approximation and beta representation. -->
 
-    - agnosticism  preservation holds, and
+<!-- - Represent  agnosticism so that: -->
 
-    - inertia fails.
+<!--     - agnosticism  preservation holds, and -->
 
-\vspace{-2mm}
+<!--     - inertia fails.  -->
 
-- Represent opinion aggregation more generally, compare to known limitations and features of already studied opinion pooling methods.
+<!-- \vspace{-2mm} -->
 
-\vspace{-2mm}
+<!-- - Represent opinion aggregation more generally, compare to known limitations and features of already studied opinion pooling methods.  -->
 
-- Investigate dilation-like phenomena in this context.
+<!-- \vspace{-2mm} -->
 
-\vspace{-2mm}
+<!-- - Investigate dilation-like phenomena in this context.  -->
 
-- Study known distance functions between probabilistic measures as candidates for key elements of accuracy measures.
+<!-- \vspace{-2mm} -->
 
-\vspace{-2mm}
+<!-- - Study known distance functions between probabilistic measures as candidates for key elements of accuracy measures. -->
 
-- Explain away the apparent contrast between evidence-grounding and accuracy considerations [@easwaranAndFitelson2015].
+<!-- \vspace{-2mm} -->
 
-\begin{center}
-\Large Thank you for your attention!
-\end{center}
+<!-- - Explain away the apparent contrast between evidence-grounding and accuracy considerations [@easwaranAndFitelson2015]. -->
+
+<!-- \begin{center} -->
+<!-- \Large Thank you for your attention! -->
+<!-- \end{center} -->
 

### 2021-07-30 presentationESSLLI21/presentationESSLLI.tex
started presentation

@@
 \hypersetup{
-            pdftitle={A bayesian method of cosine-based word2vec bias estimation},
+            pdftitle={Taking uncertainty in word embedding bias estimation seriously ~a Bayesian approach},
             pdfauthor={Alicja Dobrzeniecka \& Rafal Urbaniak (LoPSE research group, University of Gdansk)},
@@
 \urlstyle{same}  % don't use monospace font for urls
+\usepackage{longtable,booktabs}
+\usepackage{caption}
+% These lines are needed to make table captions work with longtable:
+\makeatletter
+\def\fnum@table{\tablename~\thetable}
+\makeatother
 
@@
 
-\title{\large A bayesian method of cosine-based word2vec bias estimation}
+\title{\large Taking uncertainty in word embedding bias estimation seriously ~a
+Bayesian approach}
 \author{Alicja Dobrzeniecka \& Rafal Urbaniak \footnotesize \newline (LoPSE
@@
 
-\begin{frame}
+\begin{frame}{Cosine-based measures of bias}
 
-\frametitle{The gist of the project}
+\begin{block}{Word embeddings}
 
-\pause
+\begin{itemize}
+\item
+  Representation of words (or words in contexts) with vectors of real
+  numbers
+\item
+  Built to predict the probability of co-occurence
+\end{itemize}
 
-\begin{block}{Classical probabilism in epistemology (and in
-applications)}
-
-Represent uncertainty by a single probabilistic measure (point
-estimates).
-
-\vspace{-3mm}
-
-\begin{align*}
-\pr(\mathsf{recovery}\vert \mathsf{treatment}) & = .6\\
-\end{align*}
-
-\vspace{-6mm}
-
-\pause
+\begin{longtable}[]{@{}llllll@{}}
+\toprule
+word & 1 & 2 & 3 & 4 & \ldots{}\tabularnewline
+\midrule
+\endhead
+woman & 0.456 & 0.267 & 0.675 & 0.131 & \ldots{}\tabularnewline
+man & 0.451 & 0.897 & 0.472 & 0.088 & \ldots{}\tabularnewline
+\bottomrule
+\end{longtable}
 
@@
 
-\begin{block}{Imprecise probabilism}
+\end{frame}
 
-Evidence (e.g.~expert evidence) sometimes fails to warrant such
-precision.
+\begin{frame}{Cosine-based measures of bias}
 
-\vspace{-3mm}
+\begin{block}{Cosine similarity \& distance}
 
-\begin{align*}
-\pr(\mathsf{recovery}\vert \mathsf{treatment}) & \in [.4,.6]\\
-\end{align*}
+\vspace{-4mm}
 
-\vspace{-6mm}
-
-\pause
-
-\end{block}
-
-\begin{block}{Epistemological difficulties of imprecise probabilism}
+\begin{align} \tag{Sim}
+\mathsf{cosineSimilarity}(A,B) & = \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert}
+\\
+\tag{Distance}
+\mathsf{cosineDistance}(A,B) &  = 1 - \mathsf{cosineSimilarity}(A,B)
+\end{align}
 
@@
 \item
-  hard to model the weight of and combine evidence.
+  Geometric interpretation: direction (not length)
 \item
-  inertia of agnosticism.
+  \(\mathsf{cosineDistance}\in (0, 2)\)
 \item
-  growth of uncertainty in light of new evidence.
+  Naive interpretation: proximity corresponds to semantic similarity
+  (e.g.~no triangle inequality)
 \end{itemize}
@@
 
-\begin{frame}{The gist of the project}
+\begin{frame}{Cosine-based measures of bias}
 
-\begin{block}{My novel contribution}
+\begin{block}{The worry}
 
-\begin{itemize}
-\item
-  Meet the desiderata motivating imprecision.
-\item
-  Avoid these difficulties.
-\item
-  Do so by using higher-order probabilities.
-\end{itemize}
+In the learning process these models can learn implicit biases that
+reflect harmful stereotypical thinking
 
-\begin{center}\includegraphics[width=0.7\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-1-1} \end{center}
+\pause
+
+\end{block}
+
+\begin{block}{Cosine-based bias: basic intuition}
+
+Words belonging to an intuitively harmful stereotype are cosine-close to
+each other
 
@@
 
-\begin{frame}{Motivations for imprecision \large (straighforward
-considerations)}
+\begin{frame}{Cosine-based measures of bias}
 
-\pause
+\begin{block}{Stereotypical lists}
 
-\begin{block}{Radical lack of information}
+\footnotesize
 
-\(\pr(\mathsf{It \,\, will \,\, snow\,\, in\,\, Boston\,\, on \,\, New\,\, Year's\,\, Eve\,\, 2026}) =\)
-?
+\begin{itemize}
+\item
+  feminine occupations: ``homemaker'', ``nurse'', ``receptionist'',
+  ``librarian'', etc.
+\item
+  masculine occupations: ``maestro'', ``captain'', ``architect'', etc.
+\end{itemize}
 
-\pause
+\normalsize
 
@@
 
-\begin{block}{Weight of evidence}
+\begin{block}{Visual example}
 
-\begin{itemize}
-\tightlist
-\item
-  You've seen 4/10 patients recover. \(\pr(\mathsf{recovery})=.4\)?
-\end{itemize}
+\vspace{1mm} \footnotesize
 
-\pause
-\vspace{-2mm}
+\begin{center}\includegraphics[width=0.6\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-1-1} \end{center}
 
-\begin{itemize}
-\tightlist
-\item
-  You've seen 40/100 patients recover. \(\pr(\mathsf{recovery})=.4\)?
-\end{itemize}
-
-\footnotesize (You could try to use classical statistical tools, such as
-confidence intervals,\linebreak  but they have their own problems.)
+\normalsize
 
@@
 
-\begin{frame}{Motivations for imprecision \large (uniformity and
-agnosticism)}
+\begin{frame}{Cosine-based measures of bias}
 
-\begin{block}{Agnosticism preservation requirement}
+\begin{block}{Example: direct bias}
 
-Agnosticism about uniquely connected variables should be the same.
+\begin{itemize}
+\tightlist
+\item
+  The gender bias of a word \(w\) is its projection on the gender
+  direction
+  \(\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})\)
+\end{itemize}
 
-\vspace{-5mm}
+\begin{itemize}
+\tightlist
+\item
+  Given the (ideally) gender neutral words \(N\) and the gender
+  direction \(g\) the direct gender bias is: 
+\end{itemize}
 
-\[X \mbox{ and } Y=X+2 \mbox{ or } Z = X^2 \,\,\, \mbox{ for $10>X>0$.}\]
+\vspace{-2mm}
 
-\pause
+\begin{align}
+\mathsf{directBias_c(N,g)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},g)\vert^c}{\vert N \vert }
+\end{align}
 
-\end{block}
+\footnotesize
 
-\begin{block}{The trouble}
-
-\begin{center}\includegraphics[width=0.6\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-2-1} \end{center}
-
-\vspace{-9mm}
-
-\begin{align*}\pr(X<5) = .5 = \pr(X^2 < 25)
-\end{align*}
+(Bolukbasi, Chang, Zou, Saligrama, \& Kalai,
+\protect\hyperlink{ref-Bolukbasi2016Man}{2016})
 
@@
 
-\begin{frame}{Motivations for imprecision \large (uniformity and
-agnosticism)}
+\begin{frame}{Cosine-based measures of bias}
 
-\begin{block}{The trouble}
+\begin{block}{Example: Word Embedding Association Test (WEAT)}
 
-\begin{center}\includegraphics[width=0.6\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-3-1} \end{center}
+\begin{align*}
+s(t,A,B) & = \frac{\sum_{a\in A}f(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}f(t,b)}{\vert B\vert}
+\\
+WEAT(A,B) & = \frac{
+\mu\left(\{s(x,A,B)\}_{x\in X}\right) -\mu\left(\{s(y,A,B)\}_{y\in Y}\right) 
+}{
+\sigma\left(\{s(w,A,B)\}_{w\in X\cup Y}\right)
+}
+\end{align*}
+
+\begin{itemize}
+\item
+  \(t\) is a term, \(A, B\) are sets of stereotype attribute words, \$X,
+  \(Y\) are protected group words
+\item
+  For instance, \(X\) might be a set of male names, \(Y\) a set of
+  female names, \(A\) might contain stereotypically male-related career
+  words, and \(B\) stereotypically female-related family words
+\item
+  \(s\)-values are used as datapoints in statistical significance tests
+\end{itemize}
+
+\footnotesize
+
+(Caliskan, Bryson, \& Narayanan,
+\protect\hyperlink{ref-Caliskan2017semanticsBiases}{2017}) with
+extensions in (Lauscher \& Glavas,
+\protect\hyperlink{ref-Lauscher2019multidimensional}{2019}) and
+applications in (Garg, Schiebinger, Jurafsky, \& Zou,
+\protect\hyperlink{ref-Garg2018years}{2018})
 
@@
 
-\begin{frame}{Motivations for imprecision \large (expert evidence)}
+\begin{frame}{Cosine-based measures of bias}
 
-\begin{block}{Weather forecasters}
-
-\vspace{-1mm}
-
-\begin{tabular}{ll}
-Al: & $\pr(\mathsf{It \,\,  will\,\,  rain\,\,  tomorrow})=.6$\\
-Bert: & $\pr(\mathsf{It \,\,  will\,\,  rain\,\,  tomorrow})=.8$
-\end{tabular}
-
-\pause
-
-\end{block}
-
-\begin{block}{Linear pooling}
-
-\vspace{-4mm}
+\begin{block}{Our main target: Mean Average Cosine Similarity (MAC)}
 
 \begin{align*}
-\pr(\mathsf{It \,\,  will\,\,  rain\,\,  tomorrow}) & = 
-\alpha \times .6 + \beta \times .8 \,\,\,\,\,  (where \,\, \alpha + \beta = 1.)
+S(t_i, A_j) & = \frac{1}{\vert A_j\vert}\sum_{a\in A_j}\mathsf{cos}(t,a) \\
+MAC(T,A) & = \frac{1}{\vert T \vert \,\vert A\vert}\sum_{t_i \in T }\sum_{A_j \in A} S(t_i,A_j)
 \end{align*}
-
-\pause
-
-\end{block}
-
-\begin{block}{Mathematical impossibilities}
 
@@
 \item
-  Experts's agreement on independence is not preserved by linear
-  pooling. ({\textbf{???}})
+  \(T = \{t_1, \dots, t_k\}\) is a class of protected words
 \item
-  You can't at the same time hold the following: ({\textbf{???}})
+  each \(A_j\in A\) is a set of attributes stereotypically associated
+  with a protected word
 \end{itemize}
 
-\vspace{-7mm}
+\begin{itemize}
+\tightlist
+\item
+  The t-tests they employ are run on average cosines used to calculate
+  MAC
+\end{itemize}
 
-\begin{align}
-\pr(A=B) & <1\\
-\pr(r\vert A=a) & = a \\
-\pr(r\vert B=b) & = b \\
-\forall a,b \, \pr(r \vert A =a, B = b) & =  \alpha a + \beta b
-\end{align}
+\footnotesize
+
+(Manzini, Lim, Tsvetkov, \& Black,
+\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019})
 
@@
 
-\begin{frame}{Imprecise probabilism and the motivations}
+\begin{frame}{Cosine-based measures of bias}
 
-\begin{block}{The main claim}
+\begin{block}{Our main target: Mean Average Cosine Similarity (MAC)}
 
-\justify Represent uncertainty by a set of probability measures.
+\footnotesize
 
-\vspace{1mm} \scriptsize
+\begin{table}
 
-({\textbf{???}}, {\textbf{???}}, {\textbf{???}}, {\textbf{???}},
-{\textbf{???}}, {\textbf{???}}, {\textbf{???}}; Walley,
-\protect\hyperlink{ref-walley1991statistical}{1991})
+\caption{\label{tab:religionTableHeadEarly}Rows the religion dataset.}
+\centering
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{llrr}
+\toprule
+protectedWord & wordToCompare & cosineDistance & cosineSimilarity\\
+\midrule
+jew & greedy & 0.6947042 & 0.3052958\\
+rabbi & greedy & 1.0306175 & -0.0306175\\
+rabbi & conservative & 0.7175887 & 0.2824113\\
+christian & uneducated & 0.5081939 & 0.4918061\\
+christianity & cheap & 1.2816164 & -0.2816164\\
+muslim & terrorist & 0.2726106 & 0.7273894\\
+\bottomrule
+\end{tabular}}
+\end{table}
 
-\pause
-
-\end{block}
-
-\begin{block}{Snowing in Boston?}
-
-\vspace{-1mm}
-
-Say, (.5,1).
-
-\vspace{1mm} \scriptsize
-
-({\textbf{???}})
-
-\pause
-
-\end{block}
-
-\begin{block}{Weight of evidence?}
-
-\vspace{-1mm}
-
-The more evidence you have, the more narrow the interval.
-
-\vspace{1mm} \scriptsize
-
-({\textbf{???}}, {\textbf{???}}, {\textbf{???}})
-
-\pause
-
-\end{block}
-
-\begin{block}{Agnosticisim?}
-
-\vspace{-1mm}
-
-{[}0,1{]} interval instead of a single uniform distribution.
-
-\pause
-
-\end{block}
-
-\begin{block}{Multiple sources?}
-
-\vspace{-1mm}
-
-Just put all those probabilities in one set.
+\normalsize
 
@@
 
-\begin{frame}{Epistemological challenges}
+\begin{frame}{Cosine-based measures of bias}
 
-\pause
-
-\begin{block}{Radical uncertainty and weight}
-
-Say one experiment observed 40/100 and another 8/15 recoveries.
-
-Simply putting the frequencies in a set won't do justice to the
-evidence.
-
-\scriptsize  \vspace{1mm}
-
-({\textbf{???}})
-
-\pause
-
-\end{block}
-
-\begin{block}{Dilation}
-
-\justify Whenever your roomate goes to the track he flips a fair coin.
-If it comes up heads he bets on Speedy. Otherwise he places no bets. He
-comes home from the track grinning. You're now more uncertain about
-\(\pr(\mathsf{heads})\) than before.
-
-\scriptsize
-
-\vspace{1mm} ({\textbf{???}}, {\textbf{???}}, {\textbf{???}},
-{\textbf{???}}, {\textbf{???}}, {\textbf{???}}; Walley,
-\protect\hyperlink{ref-walley1991statistical}{1991})
-
-\end{block}
-
-\end{frame}
-
-\begin{frame}{Epistemological challenges}
-
-\begin{block}{Belief inertia}
-
-Say you're fully agnostic about \(H\): \(P(H) = [0,1]\).
-
-\begin{center}
-
-\begin{center}\includegraphics[width=0.7\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-4-1} \end{center}
-\end{center}
-
-Then \(\pr(H \vert E) =[0,1]\), no matter what the evidence says.
-
-\vspace{1mm} \scriptsize
-
-({\textbf{???}},({\textbf{???}}), {\textbf{???}}, {\textbf{???}},
-{\textbf{???}}, {\textbf{???}})
-
-\end{block}
-
-\end{frame}
-
-\begin{frame}{Epistemological challenges}
-
-\begin{block}{Mystery coins}
-
-\begin{itemize}
-\tightlist
-\item
-  A fair draw of one of two biased coins, with biases \(.4, and .6\).
-\end{itemize}
-
-\pause
-
-\vspace{-2mm}
-
-\begin{itemize}
-\tightlist
-\item
-  Imprecise reaction: \(\{.4, .6\}\).
-\end{itemize}
-
-\pause
-
-\vspace{-2mm}
-
-\begin{itemize}
-\tightlist
-\item
-  Precise probabilism: expected bias of \(.5\).
-\end{itemize}
-
-\pause
-
-\end{block}
-
-\begin{block}{Question}
-
-As far as accuracy is concerned, why prefer the imprecise way?
-
-Say the actual bias is .4. A first stab:
-
-\vspace{-3mm}
-
-\begin{align*}
-\mathsf{inaccuracy(imprecise)} & = \frac{0 + .2}{2} = .1\\
-\mathsf{inaccuracy(precise)} & =  .1\\
-\end{align*}
-
-\vspace{-5mm} \scriptsize
-
-({\textbf{???}}, {\textbf{???}}; Schoenfield,
-\protect\hyperlink{ref-schoenfield2017AccuracyRationalityImprecise}{2017})
-
-\end{block}
-
-\end{frame}
-
-\begin{frame}{Higher-order approach}
-
-\begin{block}{The mystery coin with \((.2,.8)\) and 1 head in five
-tosses}
-
-\begin{center}\includegraphics[width=0.9\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-5-1} \end{center}
-
-\end{block}
-
-\end{frame}
-
-\begin{frame}{Higher-order approach}
-
-\begin{block}{Two experts, uniform prior}
-
-\begin{center}\includegraphics[width=0.9\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-6-1} \end{center}
-
-\end{block}
-
-\end{frame}
-
-\begin{frame}{The key conjecture}
-
-\begin{block}{With higher-order methods, it will be possible to:}
+\begin{block}{Known challenges}
 
@@
 \item
-  match the representation of uncertainty with the evidence,
+  Gender-direction might be an indicator of bias, but is insufficient.
+  After debiasing other non-gendered words can remain in biased
+  relations (Gonen \& Goldberg,
+  \protect\hyperlink{ref-Gonen2019lipstick}{2019})
 \item
-  combine various sources of uncertainty without running into the known
-  problems,
-\item
-  use existing mathematics to make sense of accuracy in such contexts,
-\item
-  capture reasoning with radical uncertainty without dilation and belief
-  inertia.
+  Methods which involve analogies and their evaluations by human users
+  on Mechanical Turk are unreliable (Nissim, Noord, \& Goot,
+  \protect\hyperlink{ref-Nissim2020fair}{2020})
 \end{itemize}
@@
 
-\begin{frame}{Tasks (discussion)}
+\begin{frame}{Some methodological problems}
+
+\begin{block}{Word list choice is unprincipled}
+
+We run with it for comparison
+
+\pause
+
+\end{block}
+
+\begin{block}{No design considerations to sample size}
+
+We investigate the uncertainty that arises from raw sample sizes
+
+\end{block}
+
+\end{frame}
+
+\begin{frame}{Some methodological problems}
+
+\begin{block}{No word class distinction and no control group}
+
+We make the subclasses clear, add human neutral predicates and neutral
+predicates for control
 
 \footnotesize
+
+\begin{table}
+
+\caption{\label{tab:religionTableHeadLate}Rows from extended religion dataset.}
+\centering
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{lllrrl}
+\toprule
+protectedWord & wordToCompare & wordClass & cosineDistance & cosineSimilarity & connection\\
+\midrule
+torah & hairy & jewish & 1.170 & -0.170 & associated\\
+christian & dirty & muslim & 0.949 & 0.051 & different\\
+judaism & cheap & jewish & 1.232 & -0.232 & associated\\
+christianity & familial & christian & 0.645 & 0.355 & associated\\
+mosque & approve & neutral & 0.995 & 0.005 & none\\
+imam & carry & human & 0.993 & 0.007 & human\\
+mosque & merging & neutral & 0.868 & 0.132 & none\\
+muslim & nationalized & neutral & 0.870 & 0.130 & none\\
+\bottomrule
+\end{tabular}}
+\end{table}
+
+\normalsize
+
+\end{block}
+
+\end{frame}
+
+\begin{frame}{Some methodological problems}
+
+\begin{block}{Outliers and surprisingly dissimilar words}
+
+We study those by visualizations and uncertainty estimates
+
+\pause
+
+\end{block}
+
+\begin{block}{No principled interpretation}
+
+\begin{longtable}[]{@{}ll@{}}
+\toprule
+Religion Debiasing & MAC (distance)\tabularnewline
+\midrule
+\endhead
+Biased & 0.859\tabularnewline
+Hard Debiased & 0.934\tabularnewline
+Soft Debiased (\(\lambda\) = 0.2) & 0.894\tabularnewline
+\bottomrule
+\end{longtable}
+
+What values are sufficient for the presence of bias and what differences
+are sign of real improvement? Low \(p\)-values are not high effect
+indicators!
+
+We compare HPDIs.
+
+\end{block}
+
+\end{frame}
+
+\begin{frame}{The problem with pre-averaging}
+
+\end{frame}
+
+\begin{frame}{Bayesian estimation}
+
+\begin{center}\includegraphics[width=1.05\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-2-1} \end{center}
+
+\end{frame}
+
+\begin{frame}{Bayesian estimation}
+
+\begin{center}\includegraphics[width=1.05\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-3-1} \end{center}
+
+\end{frame}
+
+\begin{frame}{Effects of debiasing}
+
+\begin{block}{Further work}
 
@@
 \item
-  Relate this representation to weight of evidence.
+  downstream tasks
 \end{itemize}
 
-\vspace{-2mm}
-
-\begin{itemize}
-\item
-  Go beyond grid approximation and beta representation.
-\item
-  Represent agnosticism so that:
-
-  \begin{itemize}
-  \item
-    agnosticism preservation holds, and
-  \item
-    inertia fails.
-  \end{itemize}
-\end{itemize}
-
-\vspace{-2mm}
-
-\begin{itemize}
-\tightlist
-\item
-  Represent opinion aggregation more generally, compare to known
-  limitations and features of already studied opinion pooling methods.
-\end{itemize}
-
-\vspace{-2mm}
-
-\begin{itemize}
-\tightlist
-\item
-  Investigate dilation-like phenomena in this context.
-\end{itemize}
-
-\vspace{-2mm}
-
-\begin{itemize}
-\tightlist
-\item
-  Study known distance functions between probabilistic measures as
-  candidates for key elements of accuracy measures.
-\end{itemize}
-
-\vspace{-2mm}
-
-\begin{itemize}
-\tightlist
-\item
-  Explain away the apparent contrast between evidence-grounding and
-  accuracy considerations (Easwaran \& Fitelson,
-  \protect\hyperlink{ref-easwaranAndFitelson2015}{2015}).
-\end{itemize}
-
-\begin{center}
-\Large Thank you for your attention!
-\end{center}
+\end{block}
 
@@
 \hypertarget{refs}{}
-\hypertarget{ref-easwaranAndFitelson2015}{}
-Easwaran, K., \& Fitelson, B. (2015). Accuracy, coherence, and evidence.
-In T. S. Gendler \& J. Hawthorne (Eds.), \emph{Oxford studies in
-epistemology} (Vol. 5). Oxford University Press.
+\hypertarget{ref-Bolukbasi2016Man}{}
+Bolukbasi, T., Chang, K., Zou, J. Y., Saligrama, V., \& Kalai, A.
+(2016). Man is to computer programmer as woman is to homemaker?
+Debiasing word embeddings. \emph{CoRR}, \emph{abs/1607.06520}. Retrieved
+from \url{http://arxiv.org/abs/1607.06520}
 
-\hypertarget{ref-schoenfield2017AccuracyRationalityImprecise}{}
-Schoenfield, M. (2017). The Accuracy and Rationality of Imprecise
-Credences: The Accuracy and Rationality of Imprecise Credences.
-\emph{Noûs}, \emph{51}(4), 667--685.
-\url{https://doi.org/10.1111/nous.12105}
+\hypertarget{ref-Caliskan2017semanticsBiases}{}
+Caliskan, A., Bryson, J. J., \& Narayanan, A. (2017). Semantics derived
+automatically from language corpora contain human-like biases.
+\emph{Science}, \emph{356}(6334), 183--186.
+\url{https://doi.org/10.1126/science.aal4230}
 
-\hypertarget{ref-walley1991statistical}{}
-Walley, P. (1991). \emph{Statistical reasoning with imprecise
-probabilities}. Chapman; Hall London.
+\hypertarget{ref-Garg2018years}{}
+Garg, N., Schiebinger, L., Jurafsky, D., \& Zou, J. (2018). Word
+embeddings quantify 100 years of gender and ethnic stereotypes.
+\emph{Proceedings of the National Academy of Sciences}, \emph{115}(16),
+E3635--E3644. \url{https://doi.org/10.1073/pnas.1720347115}
+
+\hypertarget{ref-Gonen2019lipstick}{}
+Gonen, H., \& Goldberg, Y. (2019). Lipstick on a pig: Debiasing methods
+cover up systematic gender biases in word embeddings but do not remove
+them. \emph{Proceedings of the 2019 Conference of the North American
+Chapter of the Association for Computational Linguistics: Human Language
+Technologies, Volume 1 (Long and Short Papers)}, 609--614. Minneapolis,
+Minnesota: Association for Computational Linguistics.
+\url{https://doi.org/10.18653/v1/N19-1061}
+
+\hypertarget{ref-Lauscher2019multidimensional}{}
+Lauscher, A., \& Glavas, G. (2019). Are we consistently biased?
+Multidimensional analysis of biases in distributional word vectors.
+\emph{CoRR}, \emph{abs/1904.11783}. Retrieved from
+\url{http://arxiv.org/abs/1904.11783}
+
+\hypertarget{ref-Manzini2019blackToCriminal}{}
+Manzini, T., Lim, Y. C., Tsvetkov, Y., \& Black, A. W. (2019).
+\emph{Black is to criminal as caucasian is to police: Detecting and
+removing multiclass bias in word embeddings}.
+
+\hypertarget{ref-Nissim2020fair}{}
+Nissim, M., Noord, R. van, \& Goot, R. van der. (2020). Fair is better
+than sensational: Man is to doctor as woman is to doctor.
+\emph{Computational Linguistics}, \emph{46}(2), 487--497.
+\url{https://doi.org/10.1162/coli_a_00379}
 

### 2021-08-01 presentationESSLLI21/presentationESSLLI.tex
Modifies presentation

@@
-\documentclass[10pt,ignorenonframetext,x11names, dvipsnames, bibspacing,natbib]{beamer}
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}
+%
+\documentclass[
+  10pt,
+  ignorenonframetext,
+  x11names, dvipsnames, bibspacing,natbib]{beamer}
+\usepackage{pgfpages}
 \setbeamertemplate{caption}[numbered]
@@
 \beamertemplatenavigationsymbolsempty
+% Prevent slide breaks in the middle of a paragraph
+\widowpenalties 1 10000
+\raggedbottom
+\setbeamertemplate{part page}{
+  \centering
+  \begin{beamercolorbox}[sep=16pt,center]{part title}
+    \usebeamerfont{part title}\insertpart\par
+  \end{beamercolorbox}
+}
+\setbeamertemplate{section page}{
+  \centering
+  \begin{beamercolorbox}[sep=12pt,center]{part title}
+    \usebeamerfont{section title}\insertsection\par
+  \end{beamercolorbox}
+}
+\setbeamertemplate{subsection page}{
+  \centering
+  \begin{beamercolorbox}[sep=8pt,center]{part title}
+    \usebeamerfont{subsection title}\insertsubsection\par
+  \end{beamercolorbox}
+}
+\AtBeginPart{
+  \frame{\partpage}
+}
+\AtBeginSection{
+  \ifbibliography
+  \else
+    \frame{\sectionpage}
+  \fi
+}
+\AtBeginSubsection{
+  \frame{\subsectionpage}
+}
+\usepackage{amsmath,amssymb}
 \usepackage{lmodern}
-\usepackage{amssymb,amsmath}
 \usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
 \ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
@@
   \usepackage[utf8]{inputenc}
-\else % if luatex or xelatex
-  \ifxetex
-    \usepackage{mathspec}
-  \else
-    \usepackage{fontspec}
-  \fi
-  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
+  \usepackage{unicode-math}
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
 \fi
 \usetheme[]{Rafal_beamerSly1}
-% use upquote if available, for straight quotes in verbatim environments
+% Use upquote if available, for straight quotes in verbatim environments
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-% use microtype if available
-\IfFileExists{microtype.sty}{%
-\usepackage{microtype}
+\IfFileExists{microtype.sty}{% use microtype if available
+  \usepackage[]{microtype}
   \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
 }{}
-\newif\ifbibliography
+\makeatletter
+\@ifundefined{KOMAClassName}{% if non-KOMA class
+  \IfFileExists{parskip.sty}{%
+    \usepackage{parskip}
+  }{% else
+    \setlength{\parindent}{0pt}
+    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
+}{% if KOMA class
+  \KOMAoptions{parskip=half}}
+\makeatother
+\usepackage{xcolor}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
 \hypersetup{
@@
   linkcolor=Maroon,
+  filecolor=Maroon,
   citecolor=Blue,
   urlcolor=blue,
-            breaklinks=true}
-\urlstyle{same}  % don't use monospace font for urls
-\usepackage{longtable,booktabs}
+  pdfcreator={LaTeX via pandoc}}
+\urlstyle{same} % disable monospaced font for URLs
+\newif\ifbibliography
+\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
 \usepackage{caption}
-% These lines are needed to make table captions work with longtable:
+% Make caption package work with longtable
 \makeatletter
@@
 \makeatother
-
-% Prevent slide breaks in the middle of a paragraph:
-\widowpenalties 1 10000
-\raggedbottom
-
-\AtBeginPart{
-  \let\insertpartnumber\relax
-  \let\partname\relax
-  \frame{\partpage}
-}
-\AtBeginSection{
-  \ifbibliography
-  \else
-    \let\insertsectionnumber\relax
-    \let\sectionname\relax
-    \frame{\sectionpage}
-  \fi
-}
-\AtBeginSubsection{
-  \let\insertsubsectionnumber\relax
-  \let\subsectionname\relax
-  \frame{\subsectionpage}
-}
-
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
 \setlength{\emergencystretch}{3em} % prevent overfull lines
@@
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{0}
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+\ifluatex
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
+  % set entry spacing
+  \ifnum #2 > 0
+  \setlength{\parskip}{#2\baselineskip}
+  \fi
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
-\title{\large Taking uncertainty in word embedding bias estimation seriously ~a
-Bayesian approach}
-\author{Alicja Dobrzeniecka \& Rafal Urbaniak \footnotesize \newline (LoPSE
-research group, University of Gdansk)}
+\title{\large Taking uncertainty in word embedding bias estimation
+seriously ~a Bayesian approach}
+\author{Alicja Dobrzeniecka \& Rafal Urbaniak
+\footnotesize \newline (LoPSE research group, University of Gdansk)}
 \date{ExpSem2021, ESSLLI}
@@
 \begin{frame}{Cosine-based measures of bias}
-
+\protect\hypertarget{cosine-based-measures-of-bias}{}
 \begin{block}{Word embeddings}
-
+\protect\hypertarget{word-embeddings}{}
 \begin{itemize}
@@
 \toprule
-word & 1 & 2 & 3 & 4 & \ldots{}\tabularnewline
+word & 1 & 2 & 3 & 4 & \ldots{} \\
 \midrule
 \endhead
-woman & 0.456 & 0.267 & 0.675 & 0.131 & \ldots{}\tabularnewline
-man & 0.451 & 0.897 & 0.472 & 0.088 & \ldots{}\tabularnewline
+woman & 0.456 & 0.267 & 0.675 & 0.131 & \ldots{} \\
+man & 0.451 & 0.897 & 0.472 & 0.088 & \ldots{} \\
 \bottomrule
 \end{longtable}
-
 \end{block}
-
 \end{frame}
@@
 \begin{frame}{Cosine-based measures of bias}
-
+\protect\hypertarget{cosine-based-measures-of-bias-1}{}
 \begin{block}{Cosine similarity \& distance}
-
+\protect\hypertarget{cosine-similarity-distance}{}
 \vspace{-4mm}
@@
 \end{itemize}
-
 \end{block}
-
 \end{frame}
@@
 \begin{frame}{Cosine-based measures of bias}
-
+\protect\hypertarget{cosine-based-measures-of-bias-2}{}
 \begin{block}{The worry}
-
+\protect\hypertarget{the-worry}{}
 In the learning process these models can learn implicit biases that
@@
 \pause
-
 \end{block}
@@
 \begin{block}{Cosine-based bias: basic intuition}
-
+\protect\hypertarget{cosine-based-bias-basic-intuition}{}
 Words belonging to an intuitively harmful stereotype are cosine-close to
 each other
-
 \end{block}
-
 \end{frame}
@@
 \begin{frame}{Cosine-based measures of bias}
-
+\protect\hypertarget{cosine-based-measures-of-bias-3}{}
 \begin{block}{Stereotypical lists}
-
+\protect\hypertarget{stereotypical-lists}{}
 \footnotesize
@@
 \item
-  feminine occupations: ``homemaker'', ``nurse'', ``receptionist'',
-  ``librarian'', etc.
+  feminine occupations: ``homemaker,'' ``nurse,'' ``receptionist,''
+  ``librarian,'' etc.
 \item
-  masculine occupations: ``maestro'', ``captain'', ``architect'', etc.
+  masculine occupations: ``maestro,'' ``captain,'' ``architect,'' etc.
 \end{itemize}
@@
 \normalsize
-
 \end{block}
@@
 \begin{block}{Visual example}
-
-\vspace{1mm} \footnotesize
+\protect\hypertarget{visual-example}{}
+\vspace{1mm}
+\footnotesize
 
 \begin{center}\includegraphics[width=0.6\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-1-1} \end{center}
-
 \normalsize
-
 \end{block}
-
 \end{frame}
@@
 \begin{frame}{Cosine-based measures of bias}
-
+\protect\hypertarget{cosine-based-measures-of-bias-4}{}
 \begin{block}{Example: direct bias}
-
+\protect\hypertarget{example-direct-bias}{}
 \begin{itemize}
@@
 
-(Bolukbasi, Chang, Zou, Saligrama, \& Kalai,
-\protect\hyperlink{ref-Bolukbasi2016Man}{2016})
-
+(\protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou,
+Saligrama, \& Kalai, 2016})
 \end{block}
-
 \end{frame}
@@
 \begin{frame}{Cosine-based measures of bias}
-
+\protect\hypertarget{cosine-based-measures-of-bias-5}{}
 \begin{block}{Example: Word Embedding Association Test (WEAT)}
-
+\protect\hypertarget{example-word-embedding-association-test-weat}{}
 \begin{align*}
@@
 
-(Caliskan, Bryson, \& Narayanan,
-\protect\hyperlink{ref-Caliskan2017semanticsBiases}{2017}) with
-extensions in (Lauscher \& Glavas,
-\protect\hyperlink{ref-Lauscher2019multidimensional}{2019}) and
-applications in (Garg, Schiebinger, Jurafsky, \& Zou,
-\protect\hyperlink{ref-Garg2018years}{2018})
-
+(\protect\hyperlink{ref-Caliskan2017semanticsBiases}{Caliskan, Bryson,
+\& Narayanan, 2017}) with extensions in
+(\protect\hyperlink{ref-Lauscher2019multidimensional}{Lauscher \&
+Glavas, 2019}) and applications in
+(\protect\hyperlink{ref-Garg2018years}{Garg, Schiebinger, Jurafsky, \&
+Zou, 2018})
 \end{block}
-
 \end{frame}
@@
 \begin{frame}{Cosine-based measures of bias}
-
+\protect\hypertarget{cosine-based-measures-of-bias-6}{}
 \begin{block}{Our main target: Mean Average Cosine Similarity (MAC)}
-
+\protect\hypertarget{our-main-target-mean-average-cosine-similarity-mac}{}
 \begin{align*}
@@
 
-(Manzini, Lim, Tsvetkov, \& Black,
-\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019})
-
+(\protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim,
+Tsvetkov, \& Black, 2019})
 \end{block}
-
 \end{frame}
@@
 \begin{frame}{Cosine-based measures of bias}
-
+\protect\hypertarget{cosine-based-measures-of-bias-7}{}
 \begin{block}{Our main target: Mean Average Cosine Similarity (MAC)}
-
+\protect\hypertarget{our-main-target-mean-average-cosine-similarity-mac-1}{}
 \footnotesize 
-
 \begin{table}
@@
 \end{table}
-
 \normalsize
-
 \end{block}
-
 \end{frame}
@@
 \begin{frame}{Cosine-based measures of bias}
-
+\protect\hypertarget{cosine-based-measures-of-bias-8}{}
 \begin{block}{Known challenges}
-
+\protect\hypertarget{known-challenges}{}
 \begin{itemize}
@@
   After debiasing other non-gendered words can remain in biased
-  relations (Gonen \& Goldberg,
-  \protect\hyperlink{ref-Gonen2019lipstick}{2019})
+  relations (\protect\hyperlink{ref-Gonen2019lipstick}{Gonen \&
+  Goldberg, 2019})
 \item
   Methods which involve analogies and their evaluations by human users
-  on Mechanical Turk are unreliable (Nissim, Noord, \& Goot,
-  \protect\hyperlink{ref-Nissim2020fair}{2020})
+  on Mechanical Turk are unreliable
+  (\protect\hyperlink{ref-Nissim2020fair}{Nissim, Noord, \& Goot, 2020})
 \end{itemize}
-
 \end{block}
-
 \end{frame}
@@
 \begin{frame}{Some methodological problems}
-
+\protect\hypertarget{some-methodological-problems}{}
 \begin{block}{Word list choice is unprincipled}
-
+\protect\hypertarget{word-list-choice-is-unprincipled}{}
 We run with it for comparison
@@
 \begin{block}{No design considerations to sample size}
-
+\protect\hypertarget{no-design-considerations-to-sample-size}{}
 We investigate the uncertainty that arises from raw sample sizes
-
 \end{block}
-
 \end{frame}
@@
 \begin{frame}{Some methodological problems}
-
+\protect\hypertarget{some-methodological-problems-1}{}
 \begin{block}{No word class distinction and no control group}
-
+\protect\hypertarget{no-word-class-distinction-and-no-control-group}{}
 We make the subclasses clear, add human neutral predicates and neutral
@@
 \footnotesize 
-
 \begin{table}
@@
 \begin{frame}{Some methodological problems}
-
+\protect\hypertarget{some-methodological-problems-2}{}
 \begin{block}{Outliers and surprisingly dissimilar words}
-
+\protect\hypertarget{outliers-and-surprisingly-dissimilar-words}{}
 We study those by visualizations and uncertainty estimates
@@
 \begin{block}{No principled interpretation}
-
+\protect\hypertarget{no-principled-interpretation}{}
 \begin{longtable}[]{@{}ll@{}}
 \toprule
-Religion Debiasing & MAC (distance)\tabularnewline
+Religion Debiasing & MAC (distance) \\
 \midrule
 \endhead
-Biased & 0.859\tabularnewline
-Hard Debiased & 0.934\tabularnewline
-Soft Debiased (\(\lambda\) = 0.2) & 0.894\tabularnewline
+Biased & 0.859 \\
+Hard Debiased & 0.934 \\
+Soft Debiased (\(\lambda\) = 0.2) & 0.894 \\
 \bottomrule
@@
 We compare HPDIs.
-
 \end{block}
-
 \end{frame}
@@
 \begin{frame}{The problem with pre-averaging}
-
+\protect\hypertarget{the-problem-with-pre-averaging}{}
+\begin{itemize}
+\tightlist
+\item
+  It throws away information about sample sizes
+\item
+  It removes variation which may result in false confidence
+\end{itemize}
 \end{frame}
 
-\begin{frame}{Bayesian estimation}
-
+\begin{frame}{Cosine distance and word connection visualization}
+\protect\hypertarget{cosine-distance-and-word-connection-visualization}{}
+\begin{block}{Analysis of word ``muslim''}
+\protect\hypertarget{analysis-of-word-muslim}{}
 \begin{center}\includegraphics[width=1.05\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-2-1} \end{center}
-
+\end{block}
 \end{frame}
 
-\begin{frame}{Bayesian estimation}
-
+\begin{frame}{Cosine distance and word connection visualization}
+\protect\hypertarget{cosine-distance-and-word-connection-visualization-1}{}
+\begin{block}{Analysis of word ``priest''}
+\protect\hypertarget{analysis-of-word-priest}{}
 \begin{center}\includegraphics[width=1.05\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-3-1} \end{center}
-
+\end{block}
 \end{frame}
 
-\begin{frame}{Effects of debiasing}
+\begin{frame}{Bayesian model}
+\protect\hypertarget{bayesian-model}{}
+\end{frame}
 
+\begin{frame}{HPDIs}
+\protect\hypertarget{hpdis}{}
 \begin{block}{Further work}
-
+\protect\hypertarget{further-work}{}
 \begin{itemize}
@@
 \begin{frame}{References}
-
+\protect\hypertarget{references}{}
 \tiny
@@
 \hypertarget{refs}{}
-\hypertarget{ref-Bolukbasi2016Man}{}
-Bolukbasi, T., Chang, K., Zou, J. Y., Saligrama, V., \& Kalai, A.
+\begin{CSLReferences}{1}{0}
+\leavevmode\hypertarget{ref-Bolukbasi2016Man}{}%
+Bolukbasi, T., Chang, K.-W., Zou, J. Y., Saligrama, V., \& Kalai, A.
 (2016). Man is to computer programmer as woman is to homemaker?
@@
 
-\hypertarget{ref-Caliskan2017semanticsBiases}{}
+\leavevmode\hypertarget{ref-Caliskan2017semanticsBiases}{}%
 Caliskan, A., Bryson, J. J., \& Narayanan, A. (2017). Semantics derived
@@
 
-\hypertarget{ref-Garg2018years}{}
+\leavevmode\hypertarget{ref-Garg2018years}{}%
 Garg, N., Schiebinger, L., Jurafsky, D., \& Zou, J. (2018). Word
@@
 
-\hypertarget{ref-Gonen2019lipstick}{}
-Gonen, H., \& Goldberg, Y. (2019). Lipstick on a pig: Debiasing methods
-cover up systematic gender biases in word embeddings but do not remove
-them. \emph{Proceedings of the 2019 Conference of the North American
-Chapter of the Association for Computational Linguistics: Human Language
-Technologies, Volume 1 (Long and Short Papers)}, 609--614. Minneapolis,
-Minnesota: Association for Computational Linguistics.
-\url{https://doi.org/10.18653/v1/N19-1061}
+\leavevmode\hypertarget{ref-Gonen2019lipstick}{}%
+Gonen, H., \& Goldberg, Y. (2019). Lipstick on a pig: {D}ebiasing
+methods cover up systematic gender biases in word embeddings but do not
+remove them. \emph{Proceedings of the 2019 Conference of the North
+{A}merican Chapter of the Association for Computational Linguistics:
+Human Language Technologies, Volume 1 (Long and Short Papers)},
+609--614. Minneapolis, Minnesota: Association for Computational
+Linguistics. \url{https://doi.org/10.18653/v1/N19-1061}
 
-\hypertarget{ref-Lauscher2019multidimensional}{}
+\leavevmode\hypertarget{ref-Lauscher2019multidimensional}{}%
 Lauscher, A., \& Glavas, G. (2019). Are we consistently biased?
@@
 
-\hypertarget{ref-Manzini2019blackToCriminal}{}
+\leavevmode\hypertarget{ref-Manzini2019blackToCriminal}{}%
 Manzini, T., Lim, Y. C., Tsvetkov, Y., \& Black, A. W. (2019).
 \emph{Black is to criminal as caucasian is to police: Detecting and
-removing multiclass bias in word embeddings}.
+removing multiclass bias in word embeddings}. Retrieved from
+\url{http://arxiv.org/abs/1904.04047}
 
-\hypertarget{ref-Nissim2020fair}{}
+\leavevmode\hypertarget{ref-Nissim2020fair}{}%
 Nissim, M., Noord, R. van, \& Goot, R. van der. (2020). Fair is better
@@
 
+\end{CSLReferences}
 \end{frame}

### 2021-08-02 presentationESSLLI21/presentationESSLLI.Rmd
Modifies presentation
Working on the presentation

@@
 ---
-title: "\\large Taking uncertainty in word embedding bias estimation seriously \\ a  Bayesian
+title: "\\large Taking uncertainty in word embedding bias estimation seriously \\- a  Bayesian
   approach"
@@
 
-- Representation of words (or words in contexts) with vectors of real numbers
+- Representation of words with vectors of real numbers
 
 - Built to predict the probability of co-occurence
+
+<!-- check if the statement of co-occurence always right-->
 
@@
 
-## Cosine-based measures of bias
+## Cosine-based measures of bias - visual example
 
@@
 \normalsize
-
-### Visual example
 
@@
 
-##  Bayesian estimation
+- It throws away information about sample sizes
+- It removes variation which may result in false confidence
+
+##  Cosine distance and word connection visualization
+
+### Analysis of a word "muslim"
 
@@
 
-##  Bayesian estimation
+## Cosine distance and word connection visualization
+
+### Analysis of a word "priest"
 
@@
 
-## Effects of debiasing
+## Advantages of including uncertainty
 
-### Further work
+- It enables one to directly observe the influence of sample sizes
+- It may influence risk assessment and decision making
+- ...
 
+## Bayesian model
+
+### Choosing priors
+
+```{r priorsVis,echo=FALSE,eval=TRUE,fig.align = "center",cache=FALSE, out.width = "100%", fig.caption = "Regularizing priors for the bayesian models."}
+library(ggpubr)
+sim <- seq(-1,1,by = 0.001)
+dis <- 1-sim
+
+p <- ggplot(data = data.frame(distance = dis), 
+            mapping = aes(x = distance))+theme_tufte()
+
+none <- function(x) dnorm(x,1,.5)
+cau <- function(x) dcauchy(x,0,1)
+par <- function(x) dnorm(x,0,1)
+
+noneG <- p + stat_function(fun = none)+xlim(c(-0.5,2.5))+ggtitle("Prior for mean distances")
+parG <- p + stat_function(fun = par)+xlim(c(-2,2))+xlab("distance change")+ggtitle("Prior for coefficients")
+cauG <- p + stat_function(fun = cau)+xlim(c(0,2))+ggtitle("Prior for standard deviation")+xlab("distance change")
+ggarrange(noneG,cauG, parG, ncol = 1)
+```
+
+##  Dataset-level HPDIs coefficients
+
+### Religion coefficients
+
+\begin{center}
+\begin{figure}[!htb]\centering
+   \begin{minipage}{0.55\textwidth}
+  \includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
+   \end{minipage}
+  \end{figure}
+
+\end{center}
+
+##  Dataset-level HPDIs coefficients
+
+### Gender coefficients
+
+\begin{center}
+\begin{figure}[!htb]\centering
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/genderCoeffs.jpeg}
+\end{minipage}
+
+\end{figure}
+
+\end{center}
+
+## Uncertainty included in bias detection
+
+### Closer look at Gender class
+
+\includegraphics[width=8cm]{../images/visGenderReddit.png}
+
+## Summary of the research
+
+-
+
+## Further work
+
+- including contrasts in Bayesian model
 - downstream tasks
-
-<!-- \frametitle{The gist of the project} -->
-
-<!-- \pause -->
-
-<!-- ### Classical probabilism in epistemology (and in applications) -->
-
-<!-- Represent uncertainty by a single probabilistic measure (point estimates). -->
-
-<!-- \vspace{-3mm} -->
-
-<!-- \begin{align*} -->
-<!-- \pr(\mathsf{recovery}\vert \mathsf{treatment}) & = .6\\ -->
-<!-- \end{align*} -->
-
-<!-- \vspace{-6mm} -->
-
-<!-- \pause -->
-
-<!-- ### Imprecise probabilism -->
-
-<!-- Evidence (e.g. expert evidence) sometimes fails to warrant such precision. -->
-
-<!-- \vspace{-3mm} -->
-
-<!-- \begin{align*} -->
-<!-- \pr(\mathsf{recovery}\vert \mathsf{treatment}) & \in [.4,.6]\\ -->
-<!-- \end{align*} -->
-
-<!-- \vspace{-6mm} -->
-
-<!-- \pause  -->
-
-<!-- ### Epistemological difficulties of imprecise probabilism -->
-
-<!-- - hard to model the weight of and combine evidence. -->
-
-<!-- - inertia of agnosticism. -->
-
-<!-- - growth of uncertainty  in light of new evidence. -->
-
-<!-- ## The gist of the project -->
-
-<!-- ### My novel contribution -->
-
-<!-- - Meet the desiderata motivating imprecision. -->
-
-<!-- - Avoid these difficulties. -->
-
-<!-- - Do so by using higher-order probabilities.  -->
-
-<!-- ```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "70%"} -->
-<!-- p <- ggplot(data = data.frame(probability = seq(0,1, by = .25)),mapping = aes(x = probability))+th +ylab("density")+scale_x_continuous(breaks = seq(0,1, by = .1)) -->
-
-<!-- agentPrior <- function(p) dbeta(p,28,28)  -->
-
-<!-- priorPlot <- p + stat_function(fun=agentPrior)+ggtitle("Uncertainty about recovery")+xlab("probability of recovery if treated") -->
-
-<!-- priorPlot -->
-<!-- ``` -->
-
-<!-- ## Motivations for imprecision  \large (straighforward considerations) -->
-
-<!-- \pause -->
-
-<!-- ### Radical lack of information -->
-
-<!-- $\pr(\mathsf{It \,\, will \,\,  snow\,\,  in\,\,  Boston\,\,  on \,\, New\,\,  Year's\,\,  Eve\,\,  2026}) =$ ? -->
-
-<!-- \pause -->
-
-<!-- <!-- ### Mere bounds in reasoning -->
-
-<!-- <!-- - Say $\pr(A) = .5$, $\pr(B) = .3$, but you don't know if they're independent. -->
-
-<!-- <!-- \vspace{-2mm} -->
-
-<!-- <!-- - We can only calculate bounds for $\pr(A \et B)$, $\pr(A \vee B)$, etc. -->
-
-<!-- <!-- \pause  -->
-
-<!-- ### Weight of evidence -->
-
-<!-- - You've seen 4/10 patients recover. $\pr(\mathsf{recovery})=.4$? -->
-
-<!-- \pause -->
-<!-- \vspace{-2mm} -->
-
-<!-- - You've seen 40/100 patients recover. $\pr(\mathsf{recovery})=.4$? -->
-
-<!-- \footnotesize (You could try to use classical statistical tools, such as confidence intervals,\linebreak  but they have their own problems.) -->
-
-<!-- ## Motivations for imprecision \large (uniformity and agnosticism) -->
-
-<!-- ### Agnosticism preservation requirement -->
-
-<!--  Agnosticism about uniquely connected variables should be the same.   -->
-
-<!-- \vspace{-5mm} -->
-
-<!--  \[X \mbox{ and } Y=X+2 \mbox{ or } Z = X^2 \,\,\, \mbox{ for $10>X>0$.}\]  -->
-
-<!-- \pause -->
-
-<!-- ### The trouble -->
-
-<!-- ```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "60%"} -->
-<!-- p <- ggplot(data = data.frame(value = 0:10),  -->
-<!--             mapping = aes(x = value))  -->
-
-<!-- unifX <-  function(value) punif(value,min = 0, max = 10)  -->
-
-<!-- p +   stat_function(fun=unifX)+ -->
-<!--   ylab("density")+theme_tufte(base_size = 25)+labs(title = "Uniform cumulative probability for X in (0,10)")+xlab("X")+ylab("Cumulative probability")+scale_x_continuous(breaks = 0:10)+ -->
-<!--   geom_segment(aes(x = 5, y = 0, xend = 5, yend = .5), lty = 2, alpha = .2, size = .4, color = "skyblue")+ -->
-<!--   geom_segment(aes(x = 0, y = .5, xend = 5, yend = .5), lty = 2, alpha = .2, size = .4, color = "skyblue") -->
-
-<!-- ``` -->
-
-<!-- \vspace{-9mm} -->
-
-<!-- \begin{align*}\pr(X<5) = .5 = \pr(X^2 < 25) -->
-<!-- \end{align*} -->
-
-<!-- ## Motivations for imprecision \large (uniformity and agnosticism) -->
-
-<!-- ### The trouble -->
-
-<!-- ```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "60%", message = FALSE, warning = FALSE} -->
-<!-- p <- ggplot(data = data.frame(value = 0:10),  -->
-<!--             mapping = aes(x = value))  -->
-
-<!-- unifY <-  function(value) punif(sqrt(value),min = 0, max = 10)  -->
-
-<!-- p +   stat_function(fun=unifY)+ -->
-<!--   ylab("cumulative probability")+theme_tufte(base_size = 25)+ -->
-<!--   labs(title = TeX("Resulting cumulative probability for $X^2$"))+scale_x_continuous(breaks = 0:100)+xlab(TeX("X^2"))+xlim(c(0,100))+ -->
-<!--   geom_segment(aes(x = 25, y = 0, xend = 25, yend = .5), lty = 2, alpha = .2, size = .4, color = "skyblue")+ -->
-<!--   geom_segment(aes(x = 0, y = .5, xend = 25, yend = .5), lty = 2, alpha = .2, size = .4, color = "skyblue") -->
-
-<!-- ``` -->
-
-<!-- ## Motivations for imprecision \large (expert evidence) -->
-
-<!-- ### Weather forecasters -->
-
-<!-- \vspace{-1mm} -->
-
-<!-- \begin{tabular}{ll} -->
-<!-- Al: & $\pr(\mathsf{It \,\,  will\,\,  rain\,\,  tomorrow})=.6$\\ -->
-<!-- Bert: & $\pr(\mathsf{It \,\,  will\,\,  rain\,\,  tomorrow})=.8$ -->
-<!-- \end{tabular} -->
-
-<!-- \pause -->
-
-<!-- ### Linear pooling -->
-
-<!-- \vspace{-4mm} -->
-
-<!-- \begin{align*} -->
-<!-- \pr(\mathsf{It \,\,  will\,\,  rain\,\,  tomorrow}) & =  -->
-<!-- \alpha \times .6 + \beta \times .8 \,\,\,\,\,  (where \,\, \alpha + \beta = 1.) -->
-<!-- \end{align*} -->
-
-<!-- \pause -->
-
-<!-- ### Mathematical impossibilities -->
-
-<!-- - Experts's agreement on independence is not preserved by linear pooling. [@List2011Group] -->
-
-<!-- - You can't at the same time hold the following: [@Gallow2018No] -->
-
-<!-- \vspace{-7mm} -->
-
-<!-- \begin{align} -->
-<!-- \pr(A=B) & <1\\ -->
-<!-- \pr(r\vert A=a) & = a \\ -->
-<!-- \pr(r\vert B=b) & = b \\ -->
-<!-- \forall a,b \, \pr(r \vert A =a, B = b) & =  \alpha a + \beta b -->
-<!-- \end{align} -->
-
-<!-- ## Imprecise probabilism and the motivations -->
-
-<!-- ### The main claim -->
-
-<!-- \justify Represent uncertainty by  a set of probability measures. -->
-
-<!-- \vspace{1mm} -->
-<!-- \scriptsize  -->
-
-<!-- [@Levi1974;@Grdenfors1982;@Jeffrey1983-JEFBWA-2;@walley1991statistical;@Kaplan1996;@Joyce2005;@Grove2009;@elkin2017imprecise] -->
-
-<!-- \pause -->
-
-<!-- ### Snowing in Boston?  -->
-
-<!-- \vspace{-1mm} -->
-
-<!-- Say, (.5,1). -->
-
-<!-- \vspace{1mm} -->
-<!-- \scriptsize  -->
-
-<!-- [@Mahtani2019Imprecise] -->
-
-<!-- \pause -->
-
-<!-- ### Weight of evidence?  -->
-
-<!-- \vspace{-1mm} -->
-
-<!-- The more evidence you have, the more narrow the interval. -->
-
-<!-- \vspace{1mm} -->
-<!-- \scriptsize  -->
-
-<!-- [@Kaplan1996;@Joyce2005;@Sturgeon2008-STURAT] -->
-
-<!-- \pause -->
-
-<!-- <!-- ## Imprecise probabilism and the motivations -->
-
-<!-- ### Agnosticisim? -->
-
-<!-- \vspace{-1mm} -->
-
-<!-- [0,1] interval instead of a single uniform distribution. -->
-
-<!-- \pause -->
-
-<!-- ### Multiple sources?  -->
-
-<!-- \vspace{-1mm} -->
-
-<!-- Just put all those probabilities in one set. -->
-
-<!-- ## Epistemological challenges -->
-
-<!-- \pause -->
-
-<!-- ### Radical uncertainty and weight -->
-
-<!-- Say one experiment observed 40/100  and  another 8/15 recoveries.  -->
-
-<!-- Simply putting the frequencies in a set won't do justice to the evidence.  -->
-
-<!-- \scriptsize  -->
-<!-- \vspace{1mm} -->
-
-<!-- [@Bradley2019Imprecise] -->
-
-<!-- \pause -->
-
-<!-- ### Dilation  -->
-
-<!--   \justify Whenever your roomate goes to the track he flips a fair coin.
-If it comes up heads he bets on  Speedy.
-Otherwise he places no bets.
-He comes home from the track grinning.
-You're now more uncertain about  $\pr(\mathsf{heads})$ than before.
--->
-
-<!-- \scriptsize  -->
-
-<!-- \vspace{1mm} -->
-<!--  [@walley1991statistical;@seidenfeld1993;@White2009-WHIESA;@Topey2012;@pedersen2014demystifying;@Hart2015;@Bradley2016] -->
-
-<!-- ## Epistemological challenges -->
-
-<!-- ### Belief inertia -->
-
-<!-- Say you're fully agnostic about $H$: $P(H) = [0,1]$.  -->
-
-<!-- \begin{center} -->
-<!-- ```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "70%", warning=FALSE, message = FALSE, dpi = 800} -->
-<!-- p <- ggplot(data = data.frame(value = 1:7),  -->
-<!--             mapping = aes(x = value))+ theme_tufte(base_size = 24)+ -->
-<!--   theme(axis.line.x=element_blank(), -->
-<!--         axis.text.x=element_blank(), -->
-<!--         axis.ticks.x=element_blank(), -->
-<!--         axis.title.x=element_blank(), -->
-<!--         panel.grid.minor.x=element_blank(), -->
-<!--         panel.grid.major.x=element_blank())+ -->
-<!--   ylab("probability")+xlim(c(1.8,6.2)) -->
-
-<!-- p + geom_segment(aes(x = 2, y = 0, xend = 2, yend = 1),  -->
-<!--                  lineend = "square")+ -->
-<!--    geom_segment(aes(x = 6, y = 0, xend = 6, yend = 1),  -->
-<!--                  lineend = "square")+ -->
-<!--    annotate("text", x = 1.8, y = 0, label = TeX("$p_F$"), size = 7 )+ -->
-<!--   annotate("text", x = 1.8, y = 1, label = TeX("$p_T$"), size = 7)+ -->
-<!--   annotate("text", x = 1.8, y = .4, label = TeX("$p_{1}$"), size = 7)+ -->
-<!--   annotate("text", x = 1.8, y = .6, label = TeX("$p_{2}$"), size = 7)+ -->
-<!--   geom_segment(aes(x = 1.95, y = 0, xend = 2.05, yend = 0),  -->
-<!--              lineend = "square")+ -->
-<!--   geom_segment(aes(x = 1.95, y = 0.4, xend = 2.05, yend = 0.4),  -->
-<!--                lineend = "square")+ -->
-<!--   geom_segment(aes(x = 1.95, y = 0.6, xend = 2.05, yend = 0.6),  -->
-<!--                lineend = "square")+ -->
-<!--   geom_segment(aes(x = 1.95, y = 1, xend = 2.05, yend = 1),  -->
-<!--                lineend = "square") + #now the second one -->
-<!--   annotate("text", x = 6.2, y = 0, label = TeX("$p_F'$"), size = 7)+ -->
-<!--   annotate("text", x = 6.2, y = 1, label = TeX("$p_T'$"), size = 7)+ -->
-<!--   annotate("text", x = 6.2, y = .2, label = TeX("$p_{1}'$"), size = 7)+ -->
-<!--   annotate("text", x = 6.2, y = .5, label = TeX("$p_{2}'$"), size = 7)+ -->
-<!--   geom_segment(aes(x = 5.95, y = 0, xend = 6.05, yend = 0),  -->
-<!--                lineend = "square")+ -->
-<!--   geom_segment(aes(x = 5.95, y = 0.2, xend = 6.05, yend = 0.2),  -->
-<!--                lineend = "square")+ -->
-<!--   geom_segment(aes(x = 5.95, y = 0.5, xend = 6.05, yend = 0.5),  -->
-<!--                lineend = "square")+ -->
-<!--   geom_segment(aes(x = 5.95, y = 1, xend = 6.05, yend = 1),  -->
-<!--                lineend = "square")+  -->
-<!--                  geom_segment(aes(x = 2.1, xend = 5.9, y = 0, -->
-<!--                 yend = 0), arrow = arrow(length = unit(0.08, "inches")), -->
-<!--                 color = "red", size = 0.6)+  -->
-<!--   geom_segment(aes(x = 2.1, xend = 5.9, y = 1, -->
-<!--                    yend = 1), arrow = arrow(length = unit(0.08, "inches")), -->
-<!--                color = "red", size = 0.6)+  -->
-<!--   geom_segment(aes(x = 2.1, xend = 5.9, y = 0.4, -->
-<!--                    yend = .2), arrow = arrow(length = unit(0.08, "inches")), -->
-<!--                color = "red", size = 0.6)+  -->
-<!--   geom_segment(aes(x = 2.1, xend = 5.9, y = 0.6, -->
-<!--                    yend = .5), arrow = arrow(length = unit(0.08, "inches")), -->
-<!--                color = "red", size = 0.6)+  -->
-<!--   geom_segment(aes(x = 2.2, xend = 5.8, y = 0.1, yend = 0.06), arrow = arrow(length = unit(0.06, "inches")), -->
-<!--                color = "skyblue", size = 0.3)+  -->
-<!--   geom_segment(aes(x = 2.2, xend = 5.8, y = 0.2, yend = 0.1), arrow = arrow(length = unit(0.06, "inches")), -->
-<!--                color = "skyblue", size = 0.3)+  -->
-<!--   geom_segment(aes(x = 2.2, xend = 5.8, y = 0.3, yend = 0.14), arrow = arrow(length = unit(0.06, "inches")), -->
-<!--                color = "skyblue", size = 0.3)+ -->
-<!--   geom_segment(aes(x = 2.2, xend = 5.8, y = 0.45, yend = 0.28), arrow = arrow(length = unit(0.06, "inches")), -->
-<!--                                                             color = "skyblue", size = 0.3)+  -->
-<!--   geom_segment(aes(x = 2.2, xend = 5.8, y = 0.5, yend = 0.36), arrow = arrow(length = unit(0.06, "inches")), -->
-<!--                color = "skyblue", size = 0.3)+  -->
-<!--   geom_segment(aes(x = 2.2, xend = 5.8, y = 0.55, yend = 0.45), arrow = arrow(length = unit(0.06, "inches")), -->
-<!--                color = "skyblue", size = 0.3)+ -->
-<!--   geom_segment(aes(x = 2.2, xend = 5.8, y = 0.65, yend = 0.55), arrow = arrow(length = unit(0.06, "inches")), -->
-<!--                color = "skyblue", size = 0.3)+  -->
-<!--   geom_segment(aes(x = 2.2, xend = 5.8, y = 0.7, yend = 0.65), arrow = arrow(length = unit(0.06, "inches")), -->
-<!--                color = "skyblue", size = 0.3)+  -->
-<!--   geom_segment(aes(x = 2.2, xend = 5.8, y = 0.75, yend = 0.72), arrow = arrow(length = unit(0.06, "inches")), -->
-<!--                color = "skyblue", size = 0.3)+  -->
-<!--   geom_segment(aes(x = 2.2, xend = 5.8, y = 0.79, yend = 0.8), arrow = arrow(length = unit(0.06, "inches")), -->
-<!--                color = "skyblue", size = 0.3)+  -->
-<!--   geom_segment(aes(x = 2.2, xend = 5.8, y = 0.84, yend = 0.86), arrow = arrow(length = unit(0.06, "inches")), -->
-<!--                color = "skyblue", size = 0.3)+  -->
-<!--   geom_segment(aes(x = 2.2, xend = 5.8, y = 0.9, yend = 0.93), arrow = arrow(length = unit(0.06, "inches")), -->
-<!--                color = "skyblue", size = 0.3) -->
-<!-- ``` -->
-<!-- \end{center} -->
-
-<!-- <!-- \begin{center} --> 
-<!-- <!-- \includegraphics[width = 6cm]{inertia.jpg} --> 
-<!-- <!-- \end{center} -->
-
-<!-- Then $\pr(H \vert E) =[0,1]$, no matter what the evidence says.  -->
-
-<!-- \vspace{1mm} -->
-<!-- \scriptsize -->
-
-<!-- [@Joyce2010,@Dodd2012;@Rinard2013;@Bradley2014;@Vallinder2017;@Moss2020] -->
-
-<!-- ## Epistemological challenges -->
-
-<!-- ### Mystery coins -->
-
-<!-- - A fair draw of one of two biased coins, with biases $.4, and .6$. -->
-
-<!-- \pause -->
-
-<!-- \vspace{-2mm} -->
-
-<!-- - Imprecise reaction: $\{.4, .6\}$.  -->
-
-<!-- \pause -->
-
-<!-- \vspace{-2mm} -->
-
-<!-- - Precise probabilism: expected bias of $.5$. -->
-
-<!-- \pause -->
-
-<!-- ### Question -->
-
-<!-- As far as accuracy is concerned, why prefer the imprecise way? -->
-
-<!-- Say the actual bias is .4. A first stab: -->
-
-<!-- \vspace{-3mm} -->
-
-<!-- \begin{align*} -->
-<!-- \mathsf{inaccuracy(imprecise)} & = \frac{0 + .2}{2} = .1\\ -->
-<!-- \mathsf{inaccuracy(precise)} & =  .1\\ -->
-<!-- \end{align*} -->
-
-<!-- \vspace{-5mm} -->
-<!-- \scriptsize -->
-
-<!-- [@Seidenfeld2012forecasting; @MayoWilson2016; @schoenfield2017AccuracyRationalityImprecise] -->
-
-<!-- ## Higher-order approach -->
-
-<!-- ### The mystery coin with  $(.2,.8)$ and 1 head in five tosses -->
-
-<!-- ```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "90%"} -->
-<!-- theta <- c(.4, .6) #possible parameter values -->
-<!-- priorTheta <- c(.2,.8) #priors assigned to them -->
-<!-- data <- c(1,0,0,0,0) #1 heads with five tosses -->
-<!-- z = sum(data) # number of 1's in Data -->
-<!-- N = length(data) #number of tosses -->
-
-<!-- # Compute the Bernoulli likelihood at each value of Theta: -->
-<!-- pDataIfTheta <- dbinom(z,N,theta) -->
-<!-- # Compute the probability of th evidence and  -->
-<!-- # the posterior via Bayes' rule: -->
-<!-- pData <- sum(pDataIfTheta * priorTheta) -->
-<!-- pThetaIfData <- pDataIfTheta * priorTheta / pData -->
-
-<!-- #put together -->
-<!-- table <- data.frame(theta,  priorTheta, -->
-<!--                      pData,  pDataIfTheta, pThetaIfData) -->
-
-<!-- #plot of the prior -->
-<!-- prior <- ggplot(table, aes(x = theta, y = priorTheta))+   -->
-<!--   geom_bar(stat="identity", alpha = .8)+  -->
-<!--   scale_x_continuous(breaks = theta)+th+ylim(c(0,1))+ggtitle("Prior") -->
-
-<!-- #plot of the likelihood -->
-<!-- likelihood <-  ggplot(table, aes(x = theta, y = pDataIfTheta))+ -->
-<!--   geom_bar(stat="identity", alpha = .8)+  -->
-<!--   scale_x_continuous(breaks = theta)+ -->
-<!--   th+ylim(c(0,1))+ggtitle("Likelihood") -->
-
-<!-- #plot of the posterior -->
-<!-- posterior <- ggplot(table, aes(x = theta, y = pThetaIfData))+  #set up -->
-<!--   geom_bar(stat="identity", alpha = .8)+  #barplot layer -->
-<!--   scale_x_continuous(breaks = theta)+ -->
-<!--   th+ylim(c(0,1))+ggtitle("Posterior") -->
-
-<!-- #convert the table to graphics, transpose for readability -->
-<!-- tableGG <- ggtexttable(t(round(table,3))) -->
-
-<!-- #plot together -->
-<!-- ggarrange(prior, likelihood, posterior, tableGG) -->
-<!-- ``` -->
-
-<!-- ## Higher-order approach -->
-
-<!-- ### Two experts, uniform prior -->
-
-<!-- ```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "90%"} -->
-<!-- p <- ggplot(data = data.frame(probability = seq(0,1, by = .25)),mapping = aes(x = probability))+th +ylab("density")+scale_x_continuous(breaks = seq(0,1, by = .25))+ylim(c(0,10)) -->
-
-<!-- agentPrior <- function(p) dbeta(p,1,1)  -->
-<!-- al <- function(p) dbeta(p,5,20)  -->
-<!-- bert <- function(p) dbeta(p,8,12)  -->
-<!-- posterior <- function(p) dbeta(p,19,62)  -->
-
-<!-- priorPlot <- p + stat_function(fun=agentPrior)+ggtitle("Agent\'s prior") -->
-<!-- alPlot <- p + stat_function(fun=al)+ggtitle("If Al says .2") -->
-
-<!-- bertPlot <- p + stat_function(fun=bert)+ggtitle("If Bert says .4") -->
-<!-- posterior <- p + stat_function(fun=posterior)+ggtitle("Agent's posterior") -->
-
-<!-- ggarrange(priorPlot, alPlot, bertPlot, posterior) -->
-<!-- ``` -->
-
-<!-- ## The key conjecture -->
-
-<!-- ### With higher-order methods, it will be possible to: -->
-
-<!-- - match the representation of uncertainty with the evidence, -->
-
-<!-- - combine various sources of uncertainty without running into the known problems, -->
-
-<!-- - use existing mathematics to make sense of accuracy in such contexts, -->
-
-<!-- - capture reasoning with radical uncertainty without dilation and belief inertia. -->
-
-<!-- ## Tasks (discussion) -->
-
-<!-- \footnotesize  -->
-
-<!-- - Relate this representation to weight of evidence. -->
-
-<!-- \vspace{-2mm} -->
-
-<!-- - Go beyond grid approximation and beta representation. -->
-
-<!-- - Represent  agnosticism so that: -->
-
-<!--     - agnosticism  preservation holds, and -->
-
-<!--     - inertia fails.  -->
-
-<!-- \vspace{-2mm} -->
-
-<!-- - Represent opinion aggregation more generally, compare to known limitations and features of already studied opinion pooling methods.  -->
-
-<!-- \vspace{-2mm} -->
-
-<!-- - Investigate dilation-like phenomena in this context.  -->
-
-<!-- \vspace{-2mm} -->
-
-<!-- - Study known distance functions between probabilistic measures as candidates for key elements of accuracy measures. -->
-
-<!-- \vspace{-2mm} -->
-
-<!-- - Explain away the apparent contrast between evidence-grounding and accuracy considerations [@easwaranAndFitelson2015]. -->
-
-<!-- \begin{center} -->
-<!-- \Large Thank you for your attention! -->
-<!-- \end{center} -->
+- applying uncertainty to WEAT metric
+- testing AIT dataset
 

### 2021-08-02 presentationESSLLI21/presentationESSLLI.tex
Working on the presentation

@@
 \hypersetup{
-  pdftitle={Taking uncertainty in word embedding bias estimation seriously ~a Bayesian approach},
+  pdftitle={Taking uncertainty in word embedding bias estimation seriously - a Bayesian approach},
   pdfauthor={Alicja Dobrzeniecka \& Rafal Urbaniak (LoPSE research group, University of Gdansk)},
@@
 \title{\large Taking uncertainty in word embedding bias estimation
-seriously ~a Bayesian approach}
+seriously - a Bayesian approach}
 \author{Alicja Dobrzeniecka \& Rafal Urbaniak
@@
 \item
-  Representation of words (or words in contexts) with vectors of real
-  numbers
+  Representation of words with vectors of real numbers
 \item
@@
 
-\begin{frame}{Cosine-based measures of bias}
-\protect\hypertarget{cosine-based-measures-of-bias-3}{}
+\begin{frame}{Cosine-based measures of bias - visual example}
+\protect\hypertarget{cosine-based-measures-of-bias---visual-example}{}
 \begin{block}{Stereotypical lists}
@@
 \normalsize
-\end{block}
 
-\begin{block}{Visual example}
-\protect\hypertarget{visual-example}{}
 \vspace{1mm}
@@
 \begin{frame}{Cosine-based measures of bias}
-\protect\hypertarget{cosine-based-measures-of-bias-4}{}
+\protect\hypertarget{cosine-based-measures-of-bias-3}{}
 \begin{block}{Example: direct bias}
@@
 \begin{frame}{Cosine-based measures of bias}
-\protect\hypertarget{cosine-based-measures-of-bias-5}{}
+\protect\hypertarget{cosine-based-measures-of-bias-4}{}
 \begin{block}{Example: Word Embedding Association Test (WEAT)}
@@
 \begin{frame}{Cosine-based measures of bias}
-\protect\hypertarget{cosine-based-measures-of-bias-6}{}
+\protect\hypertarget{cosine-based-measures-of-bias-5}{}
 \begin{block}{Our main target: Mean Average Cosine Similarity (MAC)}
@@
 \begin{frame}{Cosine-based measures of bias}
-\protect\hypertarget{cosine-based-measures-of-bias-7}{}
+\protect\hypertarget{cosine-based-measures-of-bias-6}{}
 \begin{block}{Our main target: Mean Average Cosine Similarity (MAC)}
@@
 \begin{frame}{Cosine-based measures of bias}
-\protect\hypertarget{cosine-based-measures-of-bias-8}{}
+\protect\hypertarget{cosine-based-measures-of-bias-7}{}
 \begin{block}{Known challenges}
@@
 \protect\hypertarget{cosine-distance-and-word-connection-visualization}{}
-\begin{block}{Analysis of word ``muslim''}
-\protect\hypertarget{analysis-of-word-muslim}{}
+\begin{block}{Analysis of a word ``muslim''}
+\protect\hypertarget{analysis-of-a-word-muslim}{}
 \begin{center}\includegraphics[width=1.05\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-2-1} \end{center}
@@
 \protect\hypertarget{cosine-distance-and-word-connection-visualization-1}{}
-\begin{block}{Analysis of word ``priest''}
-\protect\hypertarget{analysis-of-word-priest}{}
+\begin{block}{Analysis of a word ``priest''}
+\protect\hypertarget{analysis-of-a-word-priest}{}
 \begin{center}\includegraphics[width=1.05\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-3-1} \end{center}
@@
 
+\begin{frame}{Advantages of including uncertainty}
+\protect\hypertarget{advantages-of-including-uncertainty}{}
+\begin{itemize}
+\tightlist
+\item
+  It enables one to directly observe the influence of sample sizes
+\item
+  It may influence risk assessment and decision making
+\item
+  \ldots{}
+\end{itemize}
+\end{frame}
+
 \begin{frame}{Bayesian model}
 \protect\hypertarget{bayesian-model}{}
+\begin{block}{Choosing priors}
+\protect\hypertarget{choosing-priors}{}
+\begin{center}\includegraphics[width=1\linewidth]{presentationESSLLI_files/figure-beamer/priorsVis-1} \end{center}
+\end{block}
 \end{frame}
 
-\begin{frame}{HPDIs}
-\protect\hypertarget{hpdis}{}
-\begin{block}{Further work}
+\begin{frame}{Dataset-level HPDIs coefficients}
+\protect\hypertarget{dataset-level-hpdis-coefficients}{}
+\begin{block}{Religion coefficients}
+\protect\hypertarget{religion-coefficients}{}
+\begin{center}
+\begin{figure}[!htb]\centering
+   \begin{minipage}{0.55\textwidth}
+  \includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
+   \end{minipage}
+  \end{figure}
+
+\end{center}
+\end{block}
+\end{frame}
+
+\begin{frame}{Dataset-level HPDIs coefficients}
+\protect\hypertarget{dataset-level-hpdis-coefficients-1}{}
+\begin{block}{Gender coefficients}
+\protect\hypertarget{gender-coefficients}{}
+\begin{center}
+\begin{figure}[!htb]\centering
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7cm]{../images/genderCoeffs.jpeg}
+\end{minipage}
+
+\end{figure}
+
+\end{center}
+\end{block}
+\end{frame}
+
+\begin{frame}{Uncertainty included in bias detection}
+\protect\hypertarget{uncertainty-included-in-bias-detection}{}
+\begin{block}{Closer look at Gender class}
+\protect\hypertarget{closer-look-at-gender-class}{}
+\includegraphics[width=8cm]{../images/visGenderReddit.png}
+\end{block}
+\end{frame}
+
+\begin{frame}{Summary of the research}
+\protect\hypertarget{summary-of-the-research}{}
+\begin{itemize}
+\tightlist
+\item
+\end{itemize}
+\end{frame}
+
+\begin{frame}{Further work}
 \protect\hypertarget{further-work}{}
@@
 \item
+  including contrasts in Bayesian model
+\item
   downstream tasks
+\item
+  applying uncertainty to WEAT metric
+\item
+  testing AIT dataset
 \end{itemize}
-\end{block}
 \end{frame}

### 2021-08-02 thesis/_book/_main.tex
Working on the presentation

@@
 \author{Alicja Dobrzeniecka}
-\date{2021-06-22}
+\date{2021-08-02}
 
@@
 
-Natural language processing (NLP) is a subfield of computer science that processes and analyzes language in text and speech with the use of modern programming methods.
-It has practical applications in everyday life as it concerns tasks such as email filters, smart assistants, search results, language translations, text analytics and so on.
-Models used to accomplish these tasks need a lot of data to learn from.
-This data originates from human activities and historical recordings such as texts, messages or speeches.
-It turns out that in the learning process these models can learn implicit biases that reflect harmful stereotypical thinking still present in modern societies.
-There are many different types of models in NLP depending on a task that they are supposed to solve.
-However, all of them need as an input words represented by means of numbers and this is accomplished with word embeddings.
-Word embeddings are usually assigned the values based on the context in which the words appear.
-This means that the input data can have an enormous influence on the outcome.
-One can find methods that aim at identifying and measuring hidden biases and/or try to remove them by, for example, modifying the word embeddings directly.
-
-One can find various definitions trying to capture what bias and fairness actually are.
-With the choice of the definition, implications for the real-life applications may change as well.
-\protect\hyperlink{ref-Mehrabi2019Survey}{Mehrabi, Morstatter, Saxena, Lerman, \& Galstyan} (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) mark out that there exist different types of biases such as historical bias, representation bias or measurement bias (the list is long).
-This indicates how complex the issue of bias is.
-Without the proper understanding and awareness of the problem, people are prone to unconsciously sustain the bias.
-\protect\hyperlink{ref-Mehrabi2019Survey}{Mehrabi, Morstatter, Saxena, Lerman, \& Galstyan} (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) also distinguish different types of discrimination, some of them will be briefly described.
-By protected attributes we mean those qualities, traits or characteristics that one should not discriminate against.
-Direct discrimination occurs when protected attributes of individuals explicitly result in non-favorable outcomes toward them.
-Such discrimination takes place for example when one uses the information about someone's religious beliefs to evaluate whether they are competent for the work they apply for.
-In contrast, in indirect discrimination individuals appear to be treated equally, but anyway they end up being treated unjustly due to the hidden effects of biases towards their protected attributes.
-An example of such situation is when in a job advert there is a requirement of 10 years experience instead of the list of a specific type of experience and knowledge.
-This job advert can discriminate indirectly against young people who can have the required skills, but not yet so much work experience.
-Systemic discrimination takes place when policies, customs or behaviors that result from certain culture or organizational structure lead to discrimination against some groups of people.
-Finally, statistical discrimination consists in relying on group statistics to judge a person belonging to a given group.
-The topic of discrimination is entangled with another concept; fairness.
-It is essential to grasp some concepts of fairness to take them into consideration while designing implementation of some machine learning model.
-In \protect\hyperlink{ref-Mehrabi2019Survey}{Mehrabi, Morstatter, Saxena, Lerman, \& Galstyan} (\protect\hyperlink{ref-Mehrabi2019Survey}{2019}) one may notice that depending on the context and application different definitions may be applied.
-
-There is a considerable amount of literature available on the topic of bias detection and mitigation in NLP models.
-\protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) focus on gender biases that may be observable in the representation of job occupations and gender in terms of numerical
-values.
-The authors apply cosine similarity measurement to investigate the phenomenon where (the vectors corresponding to) words related to jobs that are stereotypically associated with a given gender are in fact in the model situated closer to this gender.
-They also use analogy tasks to evaluate if a bias is present in a word embedding model.
-They check analogies by comparing pairs of word vectors.
-For example they search for the word complementing the puzzle: man is to doctor as woman is to \ldots?
-First they subtract the word ``man'' from the word ``woman''\footnote{This is pointwise subtraction of vectors, but for the simplicity we will often talk about words while really meaning vectors.
-What is meant should be clear from the context.} and then
-they search for the ranked list of other words pairs that have similar vector difference. They also include in the formula a threshold to ensure that the resulting pairs could not be randomly picked.
-
-However, as \protect\hyperlink{ref-Nissim2019Fair}{Nissim, Noord, \& Goot} (\protect\hyperlink{ref-Nissim2019Fair}{2019}) points out, there are some limitations to this approach.
-In practice, most of analogies implementations do not return any input words.
-This means that it does not make sense to expect the algorithm to return the same profession for both woman and man.
-Therefore, this method of bias detection seems limited.
-Other problems are related for example to the choice of pairs and words that are used to detect the presence of discrimination, as it is often subjective and without proper justification.
-Additionally, the choice of the parameter used in \protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) formula to ensure that word pairs are not picked by random, is also not justified and changing it drastically
-influences the results.
-
-\protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) touch upon the topic of biases regarding race and gender.
-They apply knowledge from well-known psychological studies such as Implicit Association Test to investigate the relation between human stereotypical thinking and model-learnt biases to discover a close relationship between these two.
-For the evaluation they use Word Embedding Association Test (WEAT) and the Word Embedding Factual Association Test (WEFAT).
-
-\protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) propose a novel way of using cosine similarity to verify if the assumed resemblance between protected words and harmful attributes exists.
-They develop an approach that enables them to measure the bias for a class (such as gender, religion or race) and express the final result with a single metric.
-Their approach will be analyzed in details in the next chapter.
-
-\protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) refer to Implicit Association Test (Greenwald et al., 1998) that measures the strength of associations between concepts or stereotypes by measuring human reaction time for special tasks.
-Humans naturally exhibit some biases which do not always cause social concern.
-One can imagine the intuitive associations between for example insects and flowers, and the feelings of pleasantness or unpleasantness.
-In general, people would rather associate flowers than insects with feeling pleasant, and this preference in a sense is a bias or
-prejudice in some direction.
-However, this type of preference does not cause an uproar and is a rather morally neutral case.
-This example is rather used to show that the methodology used in IAT makes sense.
-\protect\hyperlink{ref-Caliskan2017Semantics}{Islam, Bryson, \& Narayanan} (\protect\hyperlink{ref-Caliskan2017Semantics}{2016}) were able to obtain similar results to the ones that measured reaction time with the use of WEAT metric.
-Unfortunately, further studies discover other biases and prejudices that directly influence the quality of other people's lives and therefore they should be taken care of.
-
-The most common bias detection methods in the literature focus on comparing the similarity between words from protected groups and those that are considered to be stereotypical or harmful in some way.
-One can find in this group methods such as Euclidean distance or cosine similarity (which is equivalent to dot product if the vectors are normalized).
-There are also other ways to detect the effects of biases.
-One can investigate the model performance on certain downstream tasks and validate if the model uses the gender or race information to complete the task.
-Unfortunately, the currently used methods employing cosine similarity, often aggregate the similarity values in a way that may lead to hasty conclusions.
-The averaging of values and the lack of attention to uncertainty involved in such aggregation may lead to an incomplete picture of the situation.
-
-In the paper we indicate how current methods used to detect biases in natural language models are limited, if we investigate them from the perspective of Bayesian analysis.
-Our research enhances the current way in which the bias detection is performed to make sure that it is methodologically valid.
-The key hypothesis is that greater understanding of data and bias implications can be achieved when Bayesian methods are applied to issue.
+Placeholder
 
@@
 
+Placeholder
+
 \hypertarget{word-embeddings}{%
@@
 
-To understand what cosine similarity measurement is, one first needs to grasp the concept of translating words to a computer-readable form.
-In the field of natural language processsing there are two main types of words representation --- localist and distributed.
-One-hot encoding is an example of a method used to achieve a localist representation of words.
-Here, each vector contains information only about a single data point.
-This is achieved by first mapping categorical values (words) to integers and then to each integers a binary vector is assigned which contains only 0s except for the index of the integer, which is assigned 1.
-An example of a localist representation is:
-
-\begin{longtable}[]{@{}llllll@{}}
-\toprule
-word & 1 & 2 & 3 & 4 & 5 \\
-\midrule
-\endhead
-woman & 1 & 0 & 0 & 0 & 0 \\
-man & 0 & 1 & 0 & 0 & 0 \\
-girl & 0 & 0 & 1 & 0 & 0 \\
-boy & 0 & 0 & 0 & 1 & 0 \\
-monarch & 0 & 0 & 0 & 0 & 1 \\
-\bottomrule
-\end{longtable}
-
-In the example above it is clear that the length of the vectors increases with the number of words in a vocabulary.
-It is not a very computationally efficient representation.
-It has other flaws as well.
-For example, it is unable to capture the resemblance between words appearing in similar contexts.
-
-In contrast to the localist representation, a distributed representation returns vectors that contain continuous values instead of discrete 1s and 0s.
-Word embeddings are a class of various techniques that allow one to represent words as distributed vectors.
-Such learned representations of text have certain properties.
-At least prima facie, they store similar (or at least co-occurring) words close to each other in a vector space.
-An example of distributed representation is:
-
-\begin{longtable}[]{@{}lllll@{}}
-\toprule
-word & 1 & 2 & 3 & 4 \\
-\midrule
-\endhead
-woman & 0.456 & 0.267 & 0.675 & 0.131 \\
-man & 0.451 & 0.897 & 0.472 & 0.088 \\
-girl & 0.604 & 0.262 & 0.414 & 0.706 \\
-boy & 0.279 & 0.172 & 0.475 & 0.010 \\
-monarch & 0.565 & 0.678 & 0.463 & 0.975 \\
-\bottomrule
-\end{longtable}
-
-One of the advantages of using a distributed representation is that one is able to represent an enormous number of concepts with a smaller number of units.
-It is also possible to better capture similarities as words of similar meanings can have similar numeric vectors.
-
-The numbers occurring in such representations are not random.
-They are learned in a process that uses a very shallow neural network.
-There are various types of techniques used for learning the vectors representations.
-One of the most straightforward ones is a skip-gram model.
-Given a word, the model tries to predict its neighboring words from the sentence.
-The mathematics behind the process relies on the idea that the prediction concerns the conditional probability of the adjacent words.
-The algorithm tries to minimize the loss function, which penalizes the system for discrepancy with actual co-occurrence frequencies in the corpus.
-One can choose various parameters of the model, such as the window size that determines how many surrounding words the model should predict.
-After preparing such a fitted model, one takes only the learned weights from a neural network, and uses them as vectors in a word embeddings representation.
-
-Word embeddings have many applications in natural language processing.
-They are handy in document search and information retrieval.
-They also play their part in improving automatic translations.
-Well learned word representations may also contribute to the improvement of sentiment analysis or spam detection.
-
 \hypertarget{cosine-similarity-and-distance}{%
@@
 
-Cosine similarity is often used as a method of finding out whether vector representations for two words suggest that they are similar or somehow connected.
-Cosine similarity is the cosine of the angle between two vectors: the result of dividing their inner product (dot product usually) by the product of their magnitudes.
-
-It is worth mentioning one more point concerning cosine similarity.
-After the vectors are normalized to have length equal to 1, the inner product itself (often dot product) is used to measure the similarity, because it then equals the cosine similarity.
-
-\begin{align} \tag{Sim}
-\mathsf{cosineSimilarity}(A,B) & = \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert}
-\end{align}
-Cosine similarity is considered a proper tool for this operation as its result has a clear connection to geometry and at least for a low number of dimensions may be easily interpreted.
-Using this scale, one can compare vector similarities in a fairly clear manner.
-When the vectors are aligned perpendicularly to each other, their similarity equals 0 (which is the same as the cosine of 90 degrees).
-This tells us that the similarity between the vectors is small.
-As the angle between vectors decreases, cosine similarity approaches one, which stands for the greatest similarity.
-It is also possible to obtain negative cosine similarity.
-If the value approaches -1, this intuitively means that the words are contrary to each other.
-
-One of the limitations of this measure is that it informs us only about similarities between vectors in terms of their orientation.
-However, it is often argued that in comparing words in terms of this metric, the magnitude of vectors may be treated as irrelevant, as the most important information pertains to direction.
-
-In what follows, it is important to distinguish between cosine similarity and cosine distance, defined as:
-\begin{align} \tag{Sim}
-\mathsf{cosineDistance}(A,B) &  = 1 - \mathsf{cosineSimilarity}(A,B)\\
- &  = 1 - \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert} \nonumber
-\end{align}
-
-The greater the similarity between two vectors, the smaller the distance between them.
-The cosine distance ranges between 0 and 2.
-If the vectors are in an opposite directions, the cosine distance is 2.
-And if the vectors are extremely similar then the cosine distance is very close to 0.
-
-One should note that cosine distance is not exactly a distance measure, as it does not meet triangle inequality requirements.
-The triangle inequality formula says that for any triangle, the sum of the lengths of any two sides must be greater than or equal to the length of the remaining side.
-As shown in \href{https://stats.stackexchange.com/questions/198080/proving-that-cosine-distance-function-defined-by-cosine-similarity-between-two-u}{discussion in stats.stackexchange.com}\footnote{\url{https://stats.stackexchange.com/questions/198080/proving-that-cosine-distance-function-defined-by-cosine-similarity-between-two-u}} in the case of cosine distance it would have to fulfill this equation \(1+\mathsf{cos-sim}(A,C) < \mathsf{cos-sim}(A,B) + \mathsf{cos-sim}(B,C)\).
-If one chooses specific unit vectors it is easy to demonstrate that the triangle inequality is not preserved.
-
 \hypertarget{cosine-distance-in-a-one-class-bias-detection}{%
@@
 
-\protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou, Saligrama, \& Kalai} (\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) define similarity between words as the outcome inner product of their normalized vectors.
-They focus on examining what the geometry of a word embedding is in regard to ``he'' and ``she'' words.
-In other words, whether the similarity between those concepts and other words reflects expected gender stereotypes.
-They test this hypothesis by investigating whether there is a connection between word embeddings representing certain professions and words referring to gender.
-They also evaluate whether automatically produced analogies between words reflect the stereotypes as well.
-
-Here the gender bias of a word \(w\) is understood as its projection on the gender direction \(\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})\) (the gender direction is the top principal compontent of ten gender pair difference vectors).
-The underlying idea is that no bias is present if non-explicitly gender words are in equal distance to both elements in all explicitly gender pairs.
-Given the (ideally) gender neutral words \(N\) and the gender direction \(g\) the direct direct gender bias is defined as the average distance of the words in \(N\) from \(g\) (\(c\) is a parameter determining how strict we want to be):
-\begin{align}
-\mathsf{directBias_c(N,g)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},g)\vert^c}{\vert N \vert }
-\end{align}
-
-A very vivid way to follow their method of arguing that bias in word embeddings is real is to plot the values of inner product of chosen words.
-The plot below does not originate from the original paper but from \href{https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings}{kaggle.com R notebook using data from GloVe: Global Vectors for Word Representation}.\footnote{\url{https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings}.} However, similar visualization may be found there.
-Data used to create our plot is as follows.
-
-Occupations associated with feminine: \textbf{"homemaker", "nurse", "receptionist", "librarian", "socialite", "hairdresser", "nanny", "bookkeeper", "stylist", "housekeeper", "interior designer", "guidance counselor"}
-
-Occupations associated with masculine: \textbf{"maestro", "skipper", "protege", "philosopher", "captain", "architect", "financier", "warrior", "broadcaster", "magician", "fighter pilot", "boss"}
-
-\vspace{1mm}
-\footnotesize
-
-\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-1-1} \end{center}
-\normalsize
-
-The points in the plot above result from the calculation of the inner product of a chosen vector for a profession word and a vector for a gender word (she or he).
-Inner product of two vectors expresses similarity between words.
-This assumption originates from the geometry and properties of a vector space.
-In the picture one may observe a correlation between the occupation and gender.
-Stereotypical male professions are closer to the ``he'' word and stereotypical female professions have greater similarity with ``she'' word.
-One could argue that this is a hint that there is some hidden bias information stored in the words vectors.
-
 \hypertarget{cosine-distance-in-a-multi-class-bias-detection}{%
@@
 
-\protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) present a different approach towards finding similarities between classes of words.
-Cosine distance is used in the article as a measure to first argue for the existence of multi-class bias and then to show how through bias mitigation techniques the bias may be decreased.
-
-They modify WEAT to a multi-class setting, introducing Mean Average Cosine similarity as a measure of bias (in fact, in the paper they report distances rather than similarities).
-Let \(T = \{t_1, \dots, t_k\}\) be a class of protected word embeddings, and let each \(A_j\in A\) be a set of attributes stereotypically associated with a protected word).
-Then:
-\begin{align}
-S(t_i, A_j) & = \frac{1}{\vert A_j\vert}\sum_{a\in A_j}\mathsf{cos}(t,a) \\
-MAC(T,A) & = \frac{1}{\vert T \vert \,\vert A\vert}\sum_{t_i \in T }\sum_{A_j \in A} S(t_i,A_j)
-\end{align}
-
-As the code has been provided by the authors we were able to reconstruct their results.
-The main steps in the procedure are as follows.
-Let us go through an example that refers to the process of hard debiasing on religious attributes.
-
-First we load word embeddings from reddit.US.txt.tok.clean.cleanedforw2v.w2v dataset.
-The word embeddings have only 50 dimensions and the number of individual words from the dataset is 44895.
-The authors assume that the protected group should ideally not have high cosine similarity to stereotypical words.
-The word embeddings geometry should not place this group close to harmful stereotypes, if it is to be bias-free.
-For instance, let's look at the religion-related words.
-
-Protected words by religion type:
-
-\begin{minipage}{1.0\textwidth}
-\textbf{"jew"       : ["judaism", "jew", "synagogue", "torah", "rabbi"]}
-
-\textbf{"christian" : ["christianity", "christian", "church", "bible", "priest"]}
-
-\textbf{"muslim"    : ["islam", "muslim", "mosque", "quran", "imam"]}
-\end{minipage}
-
-Stereotypical words by religion type:
-
-\begin{minipage}{1.0\textwidth}
-\textbf{"jew"       : ["greedy", "cheap", "hairy", "liberal"]}
-
-\textbf{"christian"       : ["judgemental", "conservative", "familial"]}
-
-\textbf{"muslim"       : ["violent", "terrorist", "dirty", "uneducated"]}
-\end{minipage}
-
-We have prepared a table presenting the values of cosine distance for each protected word with each attribute (stereotype). The part of the results is shown in Table 2.3
-
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{religion }\OtherTok{\textless{}{-}} \FunctionTok{read.csv}\NormalTok{(}\StringTok{"../datasets/religionReddit.csv"}\NormalTok{)[}\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{]}
-\FunctionTok{colnames}\NormalTok{(religion) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"protectedWord"}\NormalTok{,}\StringTok{"wordToCompare"}\NormalTok{,}\StringTok{"wordClass"}\NormalTok{,}
-                        \StringTok{"cosineDistance"}\NormalTok{,}\StringTok{"cosineSimilarity"}\NormalTok{,}\StringTok{"connection"}\NormalTok{)}
-\NormalTok{religion}\SpecialCharTok{$}\NormalTok{wordClass }\OtherTok{\textless{}{-}} \FunctionTok{as.factor}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{wordClass)}
-\FunctionTok{levels}\NormalTok{(religion}\SpecialCharTok{$}\NormalTok{wordClass) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"christian"}\NormalTok{,}\StringTok{"human"}\NormalTok{,}\StringTok{"jewish"}\NormalTok{,}\StringTok{"muslim"}\NormalTok{,}\StringTok{"neutral"}\NormalTok{)}
-\FunctionTok{head}\NormalTok{(religion)  }\SpecialCharTok{\%\textgreater{}\%}  \FunctionTok{kable}\NormalTok{(}\AttributeTok{format =} \StringTok{"latex"}\NormalTok{,}\AttributeTok{booktabs=}\NormalTok{T,}
-                      \AttributeTok{linesep =} \StringTok{""}\NormalTok{,  }\AttributeTok{escape =} \ConstantTok{FALSE}\NormalTok{, }
-                      \AttributeTok{caption =} \StringTok{"Head of the religion dataset."}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
-                      \FunctionTok{kable\_styling}\NormalTok{(}\AttributeTok{latex\_options=}\FunctionTok{c}\NormalTok{(}\StringTok{"scale\_down"}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{table}
-
-\caption{\label{tab:religionTableHeadEarly}Head of the religion dataset.}
-\centering
-\resizebox{\linewidth}{!}{
-\begin{tabular}[t]{lllrrl}
-\toprule
-protectedWord & wordToCompare & wordClass & cosineDistance & cosineSimilarity & connection\\
-\midrule
-judaism & violent & muslim & 0.7141939 & 0.2858061 & different\\
-judaism & terrorist & muslim & 0.7461333 & 0.2538667 & different\\
-judaism & dirty & muslim & 1.2002599 & -0.2002599 & different\\
-judaism & uneducated & muslim & 0.7885469 & 0.2114531 & different\\
-judaism & greedy & jewish & 1.0026172 & -0.0026172 & associated\\
-judaism & cheap & jewish & 1.2323229 & -0.2323229 & associated\\
-\bottomrule
-\end{tabular}}
-\end{table}
-\normalsize
-
-\pagebreak
-
-In the article there was no analysis of individual distances, but the general look at the means.
-The authors introduced a metric that tries to generalize WEAT in measuring the presence of bias through the classification of multi-class bias in groups of words connected with gender, religion or race.
-In the process they first take the mean of cosine distances between a given protected word and attributes assigned to each stereotype.
-They do not differentiate between stereotypes associated with a word and stereotypes associated with different words (in the case of religion, stereotypes characteristic for Christianity has also cosine distance measured with for instance, Judaism or Islam).
-Then, after collecting the list of mean cosine distances, they average the list to obtain one final value representing the whole group, in this example religion, for which the final mean of all mean distances is equal to 0.859.
-
-In the article the authors also try to remove biases, as previously defined, from a word embedding.
-First they identify the bias subspace using Principal Component Analysis (PCA) which is a technique for dimensionality reduction.
-It is applied here to choose the subspace that contains the greatest amount of information.
-There can be many subspaces found in a given group.
-For example in terms of religion one can identify at least a few sets that are to grasp the concept of religion in general:
-
-\begin{minipage}{1.0\textwidth}
-\textbf{["judaism", "christianity", "islam"]}
-
-\textbf{["jew", "christian", "muslim"]}
-
-\textbf{["synagogue", "church", "mosque"]}
-\end{minipage}
-
-The idea is to find a set that provides enough information to create from it a vector representing the concept of religion among words.
-This strategy is based on the idea that different dimensions of vectors contain different types of information and in some words in vector layers (subspaces) the information about religiousness is implicitly conveyed.
-In some cases this knowledge is useful, but in the case of harmful stereotypes one does not want to include the concept of religion in stereotypical words.
-
-After finding the bias subspace, they use it to modify the vector values individually so that their cosine distances towards certain words are changed.
-In the case of stereotypes the aim is to make the cosine distances larger so that the association between protected word and stereotype is smaller.
-
-In the final step they evaluate the results.
-The cosine distances are calculated again but this time using the debiased vocabulary.
-After taking the mean of all distances one final value is obtained and then it is compared with the average value from the beginning.
-If the cosine distances are on average greater than before then it leads the authors to the conclusion that improvement has been achieved.
-As the cosine distance increases, it is assumed that the association between protected and stereotypical words decreases.
-
 \hypertarget{limitations-of-the-approach}{%
 \section{Limitations of the approach}\label{limitations-of-the-approach}}
-
-\textbf{1. Selection of attributes}
-
-The attributes are taken from different sources, there is no principled justification for their choice. From our analysis it will become clear that the list is rather uneven.
-
-There is no mention of methodology for deciding on the number of attributes necessary to decide a hypothesis on the given size of dataset.
-There are however some ways to estimate sample sizes needed to ensure that the results are meaningful if the effect is present.
-Our research will show that the numbers used
-are rather insufficient.
-
-\textbf{2. No attention to distributions and details}
-
-The authors use the mean of mean average cosine similarities to measure multi-class similarity between protected word and harmful stereotypes.
-They average the results until they obtain one final value to represent the mean cosine distance between protected word from a given class and the attributes of that class.
-If one takes closer look at the individual values that are taken for the calculations it turns out that there is a bunch of outliers and surprisingly dissimilar words.
-We approach this issue by providing the tables and visualizations of individual cosine distances to make sure that we obtain a proper insight into the data.
-
-\textbf{3. Hiding the impact of uncertainty}
-
-With such a method the uncertainty involved is not really considered which makes it even more difficult to give reasonable interpretations of the results.
-We propose the use of Bayesian method to obtain some understanding of the influence the uncertainty has on the interpretation of final results.
-
-\pagebreak
-
-\textbf{4. No word class distinction and no control group}
-
-In the original paper, words from all three religions were compared against all of the stereotypes, which means that there was no distinction between cases in which the stereotype is associated with a given religion, as opposed to the situation in which it is associated with another one.
-Not all of the stereotypical words have to be considered as harmful for all of the religions.
-One should investigate the religions separately as some of them may have stronger harmful associations that others.
-One should also include control groups to have a way of comparing the stereotypical results with neutral or human-like words.
-Later in the text we will explain in details reasons for introducing control groups.
-In our analysis, we distinguish between stereotypes associated with a given group, stereotypes associated with different groups, and control groups: neutral words and stereotypes-free human predicates.
-
-\textbf{5. Interpreting the results}
-
-Assuming for a moment that the value of multi-class cosine distance is correct, one may question the interpretation.
-\protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) summarize the averages of cosine distance per group (gender, race, religion).
-For now let us focus now on analyzing the values relating to religious biases.
-Here is the relevant fragment of table:
-
-\begin{longtable}[]{@{}ll@{}}
-\toprule
-Religion Debiasing & MAC \\
-\midrule
-\endhead
-Biased & 0.859 \\
-Hard Debiased & 0.934 \\
-Soft Debiased (\(\lambda\) = 0.2) & 0.894 \\
-\bottomrule
-\end{longtable}
-
-MAC stand for mean average cosine similarity, although in reality the the table contains mean distances.
-What may attract attention is the fact that the value of cosine distance in ``Biased'' category is already quite high even before debiasing.
-High cosine distance indicates low cosine similarity between values.
-One could think that average cosine similarity equal to approximately 0.141 is not significant enough to consider it as bias.
-However the authors aim to mitigate ``biases'' in vectors with such great distance to make it even larger.
-Methodologically the question is, on what basis is this small similarity still considered as a proof of the presence of bias, and whether these small changes are meaningful.
-This is in general the problem of scale and the lack of universal intervals.
-In contrast, statistical intervals will help us decide whether a given cosine similarity is high enough to consider the words to be more similar than if we chose them at random.
-We will use highest posterior density intervals, in line with Bayesian methodology.
-
-\textbf{6. The curse of dimensionality}
-
-In our case, the curse of dimensionality may take place when there is an increase in the volume of data that results in adding extra dimensions to the Euclidean space. According to the article
-\href{https://analyticsindiamag.com/curse-of-dimensionality-and-what-beginners-should-do-to-overcome-it/}{Curse of dimensionality at analyticsindiamag.com} as the number of features increases, it may be harder and harder to obtain useful information from the data using the available algorithms.
-One may notice that more data should contribute to greater amount of information, but more information also means greater risk of noise and distractions in data.
-At the same time, modern solutions are often adapted to smaller dimensions and their results in higher ones are not intuitive, or may be prone to error.
-
-Using cosine similarity in high dimensions in word embeddings may also be prone to the curse of dimensionality.
-According to \protect\hyperlink{ref-Venkat2018Curse}{Venkat} (\protect\hyperlink{ref-Venkat2018Curse}{2018}) there are reasons to consider this phenomenon when searching for word similarities in higher dimensions.
-An experiment is conducted that aims at showing how the similarity values and variation change as the number of dimensions increases.
-The hypothesis made in the paper states that two things will happen as the number of dimensions increase.
-First, the effort required to measure cosine similarity will be greater, and two, the similarity between data will blur out and have less variation.
-The authors generate random points with increasing number of dimensions where each dimension of a data point is given a value between 0 and 1.
-Then they pick one vector at random from each dimension class and calculate the cosine similarity between the chosen vector and the rest of the data.
-Then they check how the variation of values changes as the number of dimensions increases.
-It seems like the more dimensions there are, the smaller the variance and therefore it is less obvious how to interpret the resulting cosine similarities.
-Maybe the scale should be adjusted to the number of dimensions and variance so that it still gives us sensible information about data.
-Yet, according to some articles cosine similarity in high dimensions is not reliable enough as it may be the case that choosing words at random may result in similar values as when picking them consciously.
-
-\begin{figure}
-\centering
-\includegraphics{../images/curseOfDimensionality.png}
-\caption{curse of dimensionality, number of dimensions on the x axis, standard deviation of similarity on the y axis}
-\end{figure}
 
@@
 
-\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-3-1} \end{center}
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-2-1} \end{center}
 \normalsize
@@
 
-\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-4-1} \end{center}
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-3-1} \end{center}
 \normalsize
@@
 
-\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-7-1} \end{center}
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-6-1} \end{center}
 \normalsize
@@
 
-\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-8-1} \end{center}
+\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-7-1} \end{center}
 \normalsize
@@
 
-We propose the use of Bayesian methods to measure uncertainty in bias detection.
-There are a few advantages of this method.
-One of them is that including uncertainty enables one to directly observe the influence of sample sizes.
-Analyzing individual words and connection coefficients, one may notice how \texttt{neutral} words have smaller uncertainty intervals and \texttt{different} or \texttt{associated} quite the opposite.
-One of the reasons for such an outcome is that we used approximately 230 neutral words and only between 11-25 (the number varies from class to class) stereotypical attributes from \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) article.
-In our approach, we also pay attention to the distribution and details regarding anomalous values.
-With the use of simple visualizations that we introduced before, we were able to indicate suspicious cosine distance values.
-Additionally, we compare in details how the cosine distance values and uncertainty change after the debiasing.
-One can investigate then how the individual vectors changed and if it is what was expected.
-Our analysis with the use of Bayesian method gave us new ideas and hypothesis concerning not only the bias detection method but also the efficiency of debiasing itself.
-We created summary tables for each of the datasets: Google word embeddings (Table \ref{tab:dTabG}), Reddit (Table \ref{tab:dTabR}), and Reddit Debiased (Table \ref{tab:dTabRD}).
-
-\vspace{1mm}
-\footnotesize
-\begin{table}
-
-\caption{\label{tab:dTabG}Overview statistics  (GoogleNews embedding)}
-\centering
-\resizebox{\linewidth}{!}{
-\begin{tabular}[t]{llll}
-\toprule
-  & Religion (percent) & Gender (percent) & Race (percent)\\
-\midrule
-A close to N & 66.67 & 14.29 & 80\\
-A close to H & 66.67 & 21.43 & 100\\
-H closer than A & 20 & 21.43 & 50\\
-D similar with A & 93.33 & 42.86 & 100\\
-D closer than A & 33.33 & 21.43 & 30\\
-General HPDI outside of  0 & no & no & no\\
-General D similar to A & slightly not & slightly not & slightly not\\
-\bottomrule
-\end{tabular}}
-\end{table}
-\normalsize
-
-\vspace{1mm}
-\footnotesize
-\begin{table}
-
-\caption{\label{tab:dTabR}Overview statistics  (Reddit embedding)}
-\centering
-\resizebox{\linewidth}{!}{
-\begin{tabular}[t]{llll}
-\toprule
-  & Religion (percent) & Gender (percent) & Race (percent)\\
-\midrule
-A close to N & 66.67 & 0 & 40\\
-A close to H & 60 & 21.43 & 50\\
-H closer than A & 33.33 & 0 & 0\\
-D similar with A & 100 & 92.86 & 90\\
-D closer than A & 33.33 & 42.86 & 30\\
-General HPDI outside of 0 & no & yes & no\\
-General D similar to A & slightly not & yes & slightly not\\
-\bottomrule
-\end{tabular}}
-\end{table}
-\normalsize
-
-\vspace{1mm}
-\footnotesize
-\begin{table}
-
-\caption{\label{tab:dTabRD}Overview statistics  (Reddit embedding, debiased)}
-\centering
-\resizebox{\linewidth}{!}{
-\begin{tabular}[t]{llll}
-\toprule
-  & Religion (percent) & Gender (percent) & Race (percent)\\
-\midrule
-A close to N & 73.33 & 7.14 & 50\\
-A close to H & 46.67 & 21.43 & 70\\
-H closer than A & 66.67 & 0 & 20\\
-D similar with A & 100 & 100 & 80\\
-D closer than A & 66.67 & 50 & 30\\
-General HPDI outside of 0 & no & yes & no\\
-General D similar to A & yes & yes & yes\\
-General H moved closer to A & yes & no & no\\
-General N moved closer & no & yes & yes\\
-\bottomrule
-\end{tabular}}
-\end{table}
-\normalsize
-
-Let us first analyse the general observations from estimated coefficients mean introduced in Section \ref{sec:datasetsLevelCoeffs} on dataset-level coefficients.
-For Google embeddings the HPDIs for all class coefficients (associated, different, human, and none) include zero.
-This can lead one to a conclusion that the impact for associated, different, human and neutral attribute is, when averaged, quite similar.
-This indicates how including the uncertainty may change the use and interpretation of \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}) MAC metric.
-It seems that if one focuses only on differences between means of means, this is too simplistic.
-In case of Reddit word embeddings the situation is similar although HPDI is below 0 for Gender class when looking at \texttt{associated} and \texttt{different} mean coefficient.
-This can suggest that there is indeed slightly stronger impact of these attributes.
-One should also notice how in general \texttt{associated} and \texttt{different} coefficients have quite similar HPDI range, the highest observed absolute difference is equal to only 0.1.
-This suggests that again the impact of associated attributes and different ones is not clear at first sight.
-Finally, let's compare the HPDIs for Reddit and Reddit debiased datasets.
-For Religion and Race dataset there is a minor shift (in absolute values the highest change is equal to approximately 0.1) of the mean coefficients towards zero.
-However for Gender dataset there is no significant change.
-This is of course a general look at the group coefficients.
-Let's now analyze the individual words.
-
-In Reddit table one can observe that for Religion the cosine distance results for the associated attributes are for approximately 60\% of the words close to human attributes as well.
-This can suggest that in some cases words concerning humans can have higher similarity with some protected words independently of whether they are stereotypical or neutral-human words.
-One should also pay attention to the fact that for all of the \texttt{associated} and \texttt{different} attributes in Religion, the uncertainty intervals overlap at some point.
-What is even more surprising is that for protected words, such as ``torah'' the \texttt{associated} attribute has the the cosine distance slightly over 1, which means no positive similarity!
-If the protected words that we chose do not have high similarity with harmful associated attributes, then one should consider at least three scenarios.
-The first one is that the choice of the protected words and attributes may be corrupted.
-The second one is that the metric is unable to catch the hidden bias properly.
-The third one is that there is actually no bias between the words.
-Regardless of which scenario one considers, it is essential to take a look at the individual values before averaging them or aggregating in other ways.
-It seems that using Bayesian method can enhance the process of verifying the hypotheses concerning the choice of protected words and attributes.
-
-Surprisingly in Gender one can observe high cosine similarity values between some female stereotypical professions and male protected words.
-If a word stereotypically associated with females has low cosine distance to male protected words, then one should investigate the issue further.
-The reasons for this unexpected cosine distance may lie in the frequency of appearance of the protected words in the raw data.
-Some of the groups (like Muslim people or females) may have less representation in the data that is taken as an input for word embeddings.
-Therefore, if we assume that the MAC detects actually co-occurrence only, it makes sense that they can have high similarity with associated attributes and lower similarity with different ones.
-At the same time male protected words and other religions can have high similarity with different attributes because they have greater representation in the dataset in general and occur close to much more concepts.
-Cosine distance seems to capture the information on the co-occurrence of words and not on the semantic similarity strictly speaking.
-
-In Google, one may notice interesting phenomena as well.
-Although the GoogleNews dataset is larger, trained on different data sources and with the use of more dimensions, some of the results are similar to the ones obtained for the Reddit corpus.
-For Religion almost all of the values for \texttt{associated} class intersect with \texttt{different} class as well.
-In the case of Race it is 100\% of the available words.
-This indicates again that it is not clear how the metric should be used.
-Quite a different situation takes place for Gender, where the similarity between \texttt{associated} and \texttt{different} is present mostly for the male protected words.
-This means that male words have high similarity with both male stereotypical professions and female ones.
-However, for females, the similarity is high mostly only for the female stereotypical professions.
-This observation could not be made when using MAC metric only, as it requires investigating the individual words, and providing uncertainty,
-
-When analyzing the Reddit debiased results one should investigate the change of the cosine distance values.
-It seems that not all of the protected words are treated equally when performing debiasing.
-In religion, the values for \texttt{associated} class moved towards 0 for most of the words except for the ones for Islam, where still the \texttt{associated} class has quite lower cosine distance than the \texttt{different} one.
-Similar situation takes place in Gender, where female words have still much lower cosine distance values for \texttt{associated} class than for the \texttt{different} one.
-In the case of male protected words it is mostly almost the same interval for \texttt{associated} and \texttt{different}.
-As the Gender data is quite specific as the attributes are not harmful adjectives but (in an ideal world) neutral professions, the aim (if we follow MAC assumptions) should actually be to make the cosine distance same for both female and male protected words.
-However, as we pointed out the situation after debiasing can sometimes be better only for one protected group, which is not the desired outcome.
-In the case of Religion, even after debiasing, all of the protected words have intersections between \texttt{associated} and \texttt{different} class and similarity greater than 0.
-As all of the attributes in religion data are negative and harmful stereotypes, one should not aim at making the distance between protected words and \texttt{associated}, and \texttt{different} class the same but rather to move it towards 1.
-
-Let's summarize the results of the bias detection methods analysis.
-As it was presented above, using the bias detection methods without involving the uncertainty may bring the risk of limited insight into the data.
-One should be concerned with the limitations of the metric that evaluates the debiasing by means of the averaged mean approach.
-Additionally, one cannot be sure if the bias is still preserved after debiasing.
-The fact that all of the cosine distances for protected words and harmful attributes moved slightly to the right does not mean that the bias is removed.
-\protect\hyperlink{ref-Gonen2019Lipstick}{Gonen \& Goldberg} (\protect\hyperlink{ref-Gonen2019Lipstick}{2019}) argue that the bias can hide in the vector geometry and preserve even after applying popular debiasing methods.
-One of the reasons for that may be the fact that relative differences between words may be preserved even after debiasing.
-Another thing is that the choice of word lists used to verify the metric effectiveness is not well justified.
-The sample size is very small, which leads to large HPDIs.
-Some of the protected words and attributes originate from somewhat old articles and it seems that the bias phenomenon is quite dynamic and the data should be as up-to-date as it is possible.
-Moreover, recall that there is no clear interpretation for the values obtained with MAC metric from \protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim, Tsvetkov, \& Black} (\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019}).
-One may assume that if the cosine distance is close to 1 then it is a desired outcome as it means, according to cosine distance assumptions, that there is almost no similarity between the words.
-However what does it mean to be close to 0?
-If the averaged cosine distance is equal to 0.8, then should we still debias?
-It is unclear what the criteria are.
-On one hand, it seems to be beneficial when the outcome is simplified as it is easier to compare results with one value per set.
-On the other hand, it is prone to misunderstanding of how to interpret the results and what threshold to assume.
-
-The bottom line is that if we want to take bias seriously, so should we approach the uncertainty involved in our estimations.
-There is no replacement for proper statistical evaluation that does not discard information about the uncertainty involved, larger word lists are needed, and visualization of the results for particular protected classes provides much better guidance than chasing a single metric based on a means of means.
+Placeholder
 
@@
 
+Placeholder
+
 \hypertarget{original-wordlist-from-manzini2019blacktocriminal}{%
@@
 
-Definite sets are the protected words, analogy templates assign stereotypical attributes to protected words, test terms were included but are not used in the original MAC calculations.
-
 \hypertarget{religion}{%
@@
 
-\vspace{1mm}
-\scriptsize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{\{}
-    \StringTok{"definite\_sets"}\SpecialCharTok{:}\NormalTok{[}
-\NormalTok{        [}\StringTok{"judaism"}\NormalTok{, }\StringTok{"christianity"}\NormalTok{, }\StringTok{"islam"}\NormalTok{],}
-\NormalTok{        [}\StringTok{"jew"}\NormalTok{, }\StringTok{"christian"}\NormalTok{, }\StringTok{"muslim"}\NormalTok{],}
-\NormalTok{        [}\StringTok{"synagogue"}\NormalTok{, }\StringTok{"church"}\NormalTok{, }\StringTok{"mosque"}\NormalTok{],}
-\NormalTok{        [}\StringTok{"torah"}\NormalTok{, }\StringTok{"bible"}\NormalTok{, }\StringTok{"quran"}\NormalTok{],}
-\NormalTok{        [}\StringTok{"rabbi"}\NormalTok{, }\StringTok{"priest"}\NormalTok{, }\StringTok{"imam"}\NormalTok{]}
-\NormalTok{    ],}
-    \StringTok{"analogy\_templates"}\SpecialCharTok{:}
-\NormalTok{    \{}
-        \StringTok{"attribute"}\SpecialCharTok{:}\NormalTok{ \{}
-            \StringTok{"jew"}       \SpecialCharTok{:}\NormalTok{ [}\StringTok{"greedy"}\NormalTok{, }\StringTok{"cheap"}\NormalTok{, }\StringTok{"hairy"}\NormalTok{, }\StringTok{"liberal"}\NormalTok{],}
-            \StringTok{"christian"} \SpecialCharTok{:}\NormalTok{ [}\StringTok{"judgemental"}\NormalTok{, }\StringTok{"conservative"}\NormalTok{, }\StringTok{"familial"}\NormalTok{],}
-            \StringTok{"muslim"}    \SpecialCharTok{:}\NormalTok{ [}\StringTok{"violent"}\NormalTok{, }\StringTok{"terrorist"}\NormalTok{, }\StringTok{"dirty"}\NormalTok{, }\StringTok{"uneducated"}\NormalTok{]}
-\NormalTok{        \}}
-\NormalTok{    \},}
-    \StringTok{"eval\_targets"}\SpecialCharTok{:}\NormalTok{[}
-\NormalTok{        [}\StringTok{"judaism"}\NormalTok{, }\StringTok{"christianity"}\NormalTok{, }\StringTok{"islam"}\NormalTok{],}
-\NormalTok{        [}\StringTok{"jew"}\NormalTok{, }\StringTok{"christian"}\NormalTok{, }\StringTok{"muslim"}\NormalTok{],}
-\NormalTok{        [}\StringTok{"synagogue"}\NormalTok{, }\StringTok{"church"}\NormalTok{, }\StringTok{"mosque"}\NormalTok{],}
-\NormalTok{        [}\StringTok{"torah"}\NormalTok{, }\StringTok{"bible"}\NormalTok{, }\StringTok{"quran"}\NormalTok{],}
-\NormalTok{        [}\StringTok{"rabbi"}\NormalTok{, }\StringTok{"priest"}\NormalTok{, }\StringTok{"imam"}\NormalTok{]}
-\NormalTok{    ], }
-    \StringTok{"testTerms"}\SpecialCharTok{:}\NormalTok{[}\StringTok{"chair"}\NormalTok{, }\StringTok{"house"}\NormalTok{, }\StringTok{"greedy"}\NormalTok{, }\StringTok{"terrorist"}\NormalTok{, }\StringTok{"dirty"}\NormalTok{, }\StringTok{"greedy"}\NormalTok{]}
-\NormalTok{\}}
-\end{Highlighting}
-\end{Shaded}
-
-\normalsize
-
 \hypertarget{gender}{%
@@
 
-\vspace{1mm}
-\scriptsize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{\{}
-    \StringTok{"definite\_sets"}\SpecialCharTok{:}\NormalTok{[}
-\NormalTok{        [}\StringTok{"he"}\NormalTok{, }\StringTok{"she"}\NormalTok{],}
-\NormalTok{        [}\StringTok{"his"}\NormalTok{, }\StringTok{"hers"}\NormalTok{],}
-\NormalTok{        [}\StringTok{"son"}\NormalTok{, }\StringTok{"daughter"}\NormalTok{],}
-\NormalTok{        [}\StringTok{"father"}\NormalTok{, }\StringTok{"mother"}\NormalTok{],}
-\NormalTok{        [}\StringTok{"male"}\NormalTok{, }\StringTok{"female"}\NormalTok{],}
-\NormalTok{        [}\StringTok{"boy"}\NormalTok{, }\StringTok{"girl"}\NormalTok{],}
-\NormalTok{        [}\StringTok{"uncle"}\NormalTok{, }\StringTok{"aunt"}\NormalTok{]}
-\NormalTok{    ],}
-    \StringTok{"analogy\_templates"}\SpecialCharTok{:}
-\NormalTok{    \{}
-        \StringTok{"role"}\SpecialCharTok{:}\NormalTok{ \{}
-            \StringTok{"man"}   \SpecialCharTok{:}\NormalTok{ [}\StringTok{"manager"}\NormalTok{, }\StringTok{"executive"}\NormalTok{, }\StringTok{"doctor"}\NormalTok{, }\StringTok{"lawyer"}\NormalTok{, }\StringTok{"programmer"}\NormalTok{, }
-                       \StringTok{"scientist"}\NormalTok{, }\StringTok{"soldier"}\NormalTok{, }\StringTok{"supervisor"}\NormalTok{, }\StringTok{"rancher"}\NormalTok{, }\StringTok{"janitor"}\NormalTok{,}
-                       \StringTok{"firefighter"}\NormalTok{, }\StringTok{"officer"}\NormalTok{],}
-            \StringTok{"woman"} \SpecialCharTok{:}\NormalTok{ [}\StringTok{"secretary"}\NormalTok{, }\StringTok{"nurse"}\NormalTok{, }\StringTok{"clerk"}\NormalTok{, }\StringTok{"artist"}\NormalTok{, }\StringTok{"homemaker"}\NormalTok{, }\StringTok{"dancer"}\NormalTok{,}
-                       \StringTok{"singer"}\NormalTok{, }\StringTok{"librarian"}\NormalTok{, }\StringTok{"maid"}\NormalTok{, }\StringTok{"hairdresser"}\NormalTok{, }\StringTok{"stylist"}\NormalTok{,}
-                       \StringTok{"receptionist"}\NormalTok{, }\StringTok{"counselor"}\NormalTok{]}
-\NormalTok{        \}}
-\NormalTok{    \}, }
-    \StringTok{"eval\_targets"}\SpecialCharTok{:}\NormalTok{[}
-\NormalTok{        [}\StringTok{"he"}\NormalTok{, }\StringTok{"she"}\NormalTok{],}
-\NormalTok{        [}\StringTok{"his"}\NormalTok{, }\StringTok{"hers"}\NormalTok{],}
-\NormalTok{        [}\StringTok{"son"}\NormalTok{, }\StringTok{"daughter"}\NormalTok{],}
-\NormalTok{        [}\StringTok{"father"}\NormalTok{, }\StringTok{"mother"}\NormalTok{],}
-\NormalTok{        [}\StringTok{"male"}\NormalTok{, }\StringTok{"female"}\NormalTok{],}
-\NormalTok{        [}\StringTok{"boy"}\NormalTok{, }\StringTok{"girl"}\NormalTok{],}
-\NormalTok{        [}\StringTok{"uncle"}\NormalTok{, }\StringTok{"aunt"}\NormalTok{]}
-\NormalTok{    ], }
-    \StringTok{"testTerms"}\SpecialCharTok{:}\NormalTok{[}\StringTok{"chair"}\NormalTok{, }\StringTok{"house"}\NormalTok{, }\StringTok{"supervisor"}\NormalTok{, }\StringTok{"secretary"}\NormalTok{, }\StringTok{"loud"}\NormalTok{, }\StringTok{"weak"}\NormalTok{]}
-\NormalTok{\}}
-\end{Highlighting}
-\end{Shaded}
-
-\normalsize
-
 \hypertarget{race}{%
@@
 
-\vspace{1mm}
-\scriptsize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{\{}
-    \StringTok{"definite\_sets"}\SpecialCharTok{:}\NormalTok{[}
-\NormalTok{        [}\StringTok{"black"}\NormalTok{, }\StringTok{"caucasian"}\NormalTok{, }\StringTok{"asian"}\NormalTok{],}
-\NormalTok{        [}\StringTok{"african"}\NormalTok{, }\StringTok{"caucasian"}\NormalTok{, }\StringTok{"asian"}\NormalTok{],}
-\NormalTok{        [}\StringTok{"black"}\NormalTok{, }\StringTok{"white"}\NormalTok{, }\StringTok{"asian"}\NormalTok{],}
-\NormalTok{        [}\StringTok{"africa"}\NormalTok{, }\StringTok{"america"}\NormalTok{, }\StringTok{"asia"}\NormalTok{],}
-\NormalTok{        [}\StringTok{"africa"}\NormalTok{, }\StringTok{"america"}\NormalTok{, }\StringTok{"china"}\NormalTok{],}
-\NormalTok{        [}\StringTok{"africa"}\NormalTok{, }\StringTok{"europe"}\NormalTok{, }\StringTok{"asia"}\NormalTok{]}
-\NormalTok{    ],}
-    \StringTok{"analogy\_templates"}\SpecialCharTok{:}
-\NormalTok{    \{}
-        \StringTok{"role"}\SpecialCharTok{:}\NormalTok{ \{}
-            \StringTok{"caucasian"}\SpecialCharTok{:}\NormalTok{ [}\StringTok{"manager"}\NormalTok{, }\StringTok{"executive"}\NormalTok{, }\StringTok{"redneck"}\NormalTok{, }\StringTok{"hillbilly"}\NormalTok{, }\StringTok{"leader"}\NormalTok{, }\StringTok{"farmer"}\NormalTok{],}
-            \StringTok{"asian"}    \SpecialCharTok{:}\NormalTok{ [}\StringTok{"doctor"}\NormalTok{, }\StringTok{"engineer"}\NormalTok{, }\StringTok{"laborer"}\NormalTok{, }\StringTok{"teacher"}\NormalTok{],}
-            \StringTok{"black"}    \SpecialCharTok{:}\NormalTok{ [}\StringTok{"slave"}\NormalTok{, }\StringTok{"musician"}\NormalTok{, }\StringTok{"runner"}\NormalTok{, }\StringTok{"criminal"}\NormalTok{, }\StringTok{"homeless"}\NormalTok{]}
-\NormalTok{        \}}
-\NormalTok{    \}, }
-    \StringTok{"eval\_targets"}\SpecialCharTok{:}\NormalTok{[}
-\NormalTok{        [}\StringTok{"black"}\NormalTok{, }\StringTok{"caucasian"}\NormalTok{, }\StringTok{"asian"}\NormalTok{],}
-\NormalTok{        [}\StringTok{"africa"}\NormalTok{, }\StringTok{"america"}\NormalTok{, }\StringTok{"asia"}\NormalTok{]}
-\NormalTok{    ], }
-    \StringTok{"testTerms"}\SpecialCharTok{:}\NormalTok{[}\StringTok{"chair"}\NormalTok{, }\StringTok{"house"}\NormalTok{, }\StringTok{"smart"}\NormalTok{, }\StringTok{"criminal"}\NormalTok{, }\StringTok{"executive"}\NormalTok{, }\StringTok{"farmer"}\NormalTok{]}
-\NormalTok{\}}
-\end{Highlighting}
-\end{Shaded}
-
-\normalsize
-
 \hypertarget{our-control-groups}{%
@@
 
-\vspace{1mm}
-\scriptsize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{human\_words }\OtherTok{=}\NormalTok{ [}\StringTok{\textquotesingle{}wear\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}walk\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}visitor\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}toy\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}tissue\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}throw\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}talk\textquotesingle{}}\NormalTok{, }
-               \StringTok{\textquotesingle{}speak\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}sleep\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}eye\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}enjoy\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}blogger\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}character\textquotesingle{}}\NormalTok{, }
-               \StringTok{\textquotesingle{}candidate\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}breakfast\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}supper\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}dinner\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}eat\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}drink\textquotesingle{}}\NormalTok{,}
-               \StringTok{"carry"}\NormalTok{, }\StringTok{"run"}\NormalTok{, }\StringTok{"cast"}\NormalTok{, }\StringTok{"ask"}\NormalTok{, }\StringTok{"awake"}\NormalTok{, }\StringTok{"ear"}\NormalTok{, }\StringTok{"nose"}\NormalTok{, }\StringTok{"lunch"}\NormalTok{]}
-\end{Highlighting}
-\end{Shaded}
-
-\normalsize
-
 \hypertarget{non-human-neutral-attributes}{%
@@
 \addcontentsline{toc}{section}{Non-human neutral attributes}
-
-\vspace{1mm}
-\footnotesize
-
-\normalsize
-
-\vspace{1mm}
-\scriptsize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{neutral\_words }\OtherTok{=}\NormalTok{ [}\StringTok{\textquotesingle{}liquor\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}pow\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}ballpark\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}glitchy\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}billy\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}dallas\textquotesingle{}}\NormalTok{, }
-                 \StringTok{\textquotesingle{}rip\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}called\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}outlooks\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}viet\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}floater\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}rattlesnake\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}exports\textquotesingle{}}\NormalTok{,}
-                 \StringTok{\textquotesingle{}peruvian\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}recursion\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}shortfall\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}corrected\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}amicable\textquotesingle{}}\NormalTok{,}
-                 \StringTok{\textquotesingle{}solutions\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}diagnostic\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}patently\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}flops\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}approx\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}percents\textquotesingle{}}\NormalTok{,}
-                \StringTok{\textquotesingle{}lox\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}catapults\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}hamburger\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}engulfed\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}households\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}north\textquotesingle{}}\NormalTok{,}
-                \StringTok{\textquotesingle{}snubbed\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}playtest\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}replayability\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}glottal\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}parable\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}gingers\textquotesingle{}}\NormalTok{,}
-                \StringTok{\textquotesingle{}anachronism\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}organizing\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}reach\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}shtick\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}eleventh\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}cpu\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}ranked\textquotesingle{}}\NormalTok{,}
-                \StringTok{\textquotesingle{}irreversibly\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}ponce\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}velociraptor\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}rubber\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}defects\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}puzzle\textquotesingle{}}\NormalTok{,}
-                \StringTok{\textquotesingle{}smasher\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}northside\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}heft\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}observation\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}rectum\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}mystical\textquotesingle{}}\NormalTok{,}
-                \StringTok{\textquotesingle{}telltale\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}remnants\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}inquiry\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}indisputable\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}boatload\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}lessening\textquotesingle{}}\NormalTok{,}
-                \StringTok{\textquotesingle{}uselessness\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}observes\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}fictitious\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}repatriation\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}duh\textquotesingle{}}\NormalTok{,}
-                \StringTok{\textquotesingle{}attic\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}schilling\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}charges\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}chatter\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}pad\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}smurfing\textquotesingle{}}\NormalTok{,}
-                \StringTok{\textquotesingle{}worthiness\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}definitive\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}neat\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}homogenized\textquotesingle{}}\NormalTok{,}\StringTok{\textquotesingle{}lexicon\textquotesingle{}}\NormalTok{,}
-                \StringTok{\textquotesingle{}nationalized\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}earpiece\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}specializations\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}lapse\textquotesingle{}}\NormalTok{,}
-                \StringTok{\textquotesingle{}concludes\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}weaving\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}apprentices\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}fri\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}younglings\textquotesingle{}}\NormalTok{,}
-                \StringTok{\textquotesingle{}militias\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}inscriptions\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}gouda\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}lift\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}laboring\textquotesingle{}}\NormalTok{, }
-                \StringTok{\textquotesingle{}adaptive\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}lecture\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}hogging\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}thorne\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}fud\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}skews\textquotesingle{}}\NormalTok{,}
-                \StringTok{\textquotesingle{}epistles\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}tagging\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}crud\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}two\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}rebalanced\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}payroll\textquotesingle{}}\NormalTok{,}
-                \StringTok{\textquotesingle{}damned\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}approve\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}reason\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}formally\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}releasing\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}muddled\textquotesingle{}}\NormalTok{,}
-                \StringTok{\textquotesingle{}mineral\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}shied\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}capital\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}nodded\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}escrow\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}unscientific\textquotesingle{}}\NormalTok{,}
-                \StringTok{\textquotesingle{}recognizable\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}entitlement\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}disconnecting\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}marshals\textquotesingle{}}\NormalTok{,}
-                \StringTok{\textquotesingle{}winamp\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}forceful\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}lowes\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}ptr\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}sip\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}pencils\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}stomachs\textquotesingle{}}\NormalTok{,}
-                \StringTok{\textquotesingle{}goff\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}cg\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}backyard\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}uprooting\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}merging\textquotesingle{}}\NormalTok{,}
-                \StringTok{\textquotesingle{}helpful\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}eid\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}trenchcoat\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}airlift\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}frothing\textquotesingle{}}\NormalTok{,}
-                \StringTok{\textquotesingle{}pulls\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}volta\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}guinness\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}viewership\textquotesingle{}}\NormalTok{,}
-                \StringTok{\textquotesingle{}eruption\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}peeves\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}goat\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}goofy\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}disbanding\textquotesingle{}}\NormalTok{, }
-                \StringTok{\textquotesingle{}relented\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}ratings\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}disputed\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}vitamins\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}singled\textquotesingle{}}\NormalTok{,}
-                \StringTok{\textquotesingle{}hydroxide\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}telegraphed\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}mercantile\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}headache\textquotesingle{}}\NormalTok{, }
-                \StringTok{\textquotesingle{}muppets\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}petal\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}arrange\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}donovan\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}scrutinized\textquotesingle{}}\NormalTok{,}
-                \StringTok{\textquotesingle{}spoil\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}examiner\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}ironed\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}maia\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}condensation\textquotesingle{}}\NormalTok{,}
-                \StringTok{\textquotesingle{}receipt\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}solider\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}tattooing\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}encoded\textquotesingle{}}\NormalTok{,}
-                \StringTok{\textquotesingle{}compartmentalize\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}lain\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}the\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}gov\textquotesingle{}}\NormalTok{,}
-                \StringTok{\textquotesingle{}printers\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}hiked\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}resentment\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}revisionism\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}tavern\textquotesingle{}}\NormalTok{,}
-                \StringTok{\textquotesingle{}backpacking\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}pestering\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}gassed\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}acknowledges\textquotesingle{}}\NormalTok{,}
-                \StringTok{\textquotesingle{}testimonies\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}parlance\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}hallucinate\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}speeches\textquotesingle{}}\NormalTok{,}
-                \StringTok{\textquotesingle{}engaging\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}solder\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}perceptive\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}microbiology\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}reconnaissance\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}garlic\textquotesingle{}}\NormalTok{,}
-                \StringTok{\textquotesingle{}neutrals\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}width\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}literaly\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}guild\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}despicable\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}dion\textquotesingle{}}\NormalTok{,}
-                \StringTok{\textquotesingle{}option\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}transistors\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}chiropractic\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}tattered\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}consolidating\textquotesingle{}}\NormalTok{,}
-                \StringTok{\textquotesingle{}olds\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}garmin\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}shift\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}granted\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}intramural\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}allie\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}cylinders\textquotesingle{}}\NormalTok{,}
-                \StringTok{\textquotesingle{}wishlist\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}crank\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}wrongly\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}workshop\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}yesterday\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}wooden\textquotesingle{}}\NormalTok{, }
-                \StringTok{\textquotesingle{}without\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}wheel\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}weather\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}watch\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}wage\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}version\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}usually\textquotesingle{}}\NormalTok{,}
-                \StringTok{\textquotesingle{}twice\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}tomato\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}ticket\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}text\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}switch\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}studio\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}stick\textquotesingle{}}\NormalTok{,}
-                \StringTok{\textquotesingle{}soup\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}sometimes\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}signal\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}prior\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}plant\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}photo\textquotesingle{}}\NormalTok{,}
-                \StringTok{\textquotesingle{}path\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}park\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}near\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}menu\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}latter\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}grass\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}clock\textquotesingle{}}\NormalTok{]}
-\end{Highlighting}
-\end{Shaded}
-
-\normalsize
-
-\hypertarget{references}{%
-\chapter*{References}\label{references}}
-\addcontentsline{toc}{chapter}{References}
 
@@
 
-\leavevmode\hypertarget{ref-Gonen2019Lipstick}{}%
-Gonen, H., \& Goldberg, Y.
-(2019).
-Lipstick on a pig: Debiasing methods cover up systematic gender biases in word embeddings but do not remove them.
-\emph{CoRR}, \emph{abs/1903.03862}.
-Retrieved from \url{http://arxiv.org/abs/1903.03862}
-
-\leavevmode\hypertarget{ref-Caliskan2017Semantics}{}%
-Islam, A.
-C., Bryson, J.
-J., \& Narayanan, A.
-(2016).
-Semantics derived automatically from language corpora necessarily contain human biases.
-\emph{CoRR}, \emph{abs/1608.07187}.
-Retrieved from \url{http://arxiv.org/abs/1608.07187}
-
 \leavevmode\hypertarget{ref-Manzini2019blackToCriminal}{}%
@@
 
-\leavevmode\hypertarget{ref-Mehrabi2019Survey}{}%
-Mehrabi, N., Morstatter, F., Saxena, N., Lerman, K., \& Galstyan, A.
-(2019).
-A survey on bias and fairness in machine learning.
-\emph{CoRR}, \emph{abs/1908.09635}.
-Retrieved from \url{http://arxiv.org/abs/1908.09635}
-
-\leavevmode\hypertarget{ref-Nissim2019Fair}{}%
-Nissim, M., Noord, R.
-van, \& Goot, R.
-van der.
-(2019).
-Fair is better than sensational: Man is to doctor as woman is to doctor.
-\emph{CoRR}, \emph{abs/1905.09866}.
-Retrieved from \url{http://arxiv.org/abs/1905.09866}
-
-\leavevmode\hypertarget{ref-Venkat2018Curse}{}%
-Venkat, N. (2018). The curse of dimensionality: Inside out.
-
 \end{CSLReferences}

### 2021-08-02 presentationESSLLI21/presentationESSLLI.Rmd
Adds presentation changes

@@
 
-### Closer look at Gender class
+### Closer look at Religion class
 
-\includegraphics[width=8cm]{../images/visGenderReddit.png}
+<!-- add religion class -->
 
-## Summary of the research
+## Observations
 
+- HPDIs for dataset level coefficients around 0 (Gender class is an exception)
+<!-- he impact for associated, different, human and neutral attribute is, when averaged, quite similar -->
 -
@@
 
-- including contrasts in Bayesian model
-- downstream tasks
-- applying uncertainty to WEAT metric
-- testing AIT dataset
+- Including contrasts in Bayesian model
+- Downstream tasks
+- Applying uncertainty to WEAT metric
+- Testing AIT dataset
 

### 2021-08-03 paperDraft/paperDraft.Rmd
r

@@
 s <- function (table){ mean(table$A) - mean(table$B)}
+whole <- rbind(t1,t2,t3,t4)
+rawsd <- sd(c(whole$A,whole$B))
 factor <- sd(c(s(t1),s(t2),s(t3),s(t4)))
+rawsd/factor
 numerator <-  mean(s(t1),s(t2)) - mean(s(t3),s(t4))

### 2021-08-03 presentationESSLLI21/presentationESSLLI.Rmd
Adds small fixes
r

--- 2021-08-02 presentationESSLLI21/presentationESSLLI.Rmd
@@
 ---
-title: "\\large Taking uncertainty in word embedding bias estimation seriously \\- a  Bayesian
-  approach"
+title: "\\Large Taking uncertainty seriously  \\newline \\normalsize A  Bayesian
+  approach to  word embedding bias estimation "
 author: "Alicja Dobrzeniecka \\& Rafal  Urbaniak \\footnotesize \\newline (LoPSE research group, University of  Gdansk)" 
@@
 
-## Cosine-based measures of bias - visual example
+## Cosine-based measures of bias
 
-### Stereotypical lists
+### A visual example
 
@@
 
-- masculine occupations:  "maestro",  "captain", "architect", etc.
+- masculine occupations:  "maestro",  "captain", "architect", "boss", etc.
 
@@
 
+<!-- ## Cosine-based measures of bias -->
+
+<!-- ### Example: direct bias -->
+
+<!-- - The gender bias of a word $w$ is its projection on the gender direction $\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})$ -->
+
+<!-- <!-- (the gender direction is the top principal compontent of ten gender pair difference vectors).  --> -->
+
+<!-- -  Given the (ideally) gender neutral words $N$ and the gender direction $g$ the direct gender bias is: -->
+<!-- <!-- ($c$ is a parameter determining how strict we want to be): --> -->
+
+<!-- \vspace{-2mm} -->
+
+<!-- \begin{align} -->
+<!-- \mathsf{directBias_c(N,g)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},g)\vert^c}{\vert N \vert } -->
+<!-- \end{align} -->
+
+<!-- \footnotesize  -->
+
+<!-- [@Bolukbasi2016Man] -->
+
 ## Cosine-based measures of bias
 
-### Example: direct bias
+### Example: Word Embedding Association Test (WEAT)
 
-- The gender bias of a word $w$ is its projection on the gender direction $\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})$
+\vspace{1mm}
+\footnotesize
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "55%"}
+words <- c("man", "he", "boss", "architect","nurse","receptionist","woman","she")
+x <- c(.1,.1,.5,.5,.7,.7,1.2,1.2)
+y <- c(.4,.3,.3,.2,.5,.4,.4,.3)
+weat <- data.frame(words,x,y)
 
-<!-- (the gender direction is the top principal compontent of ten gender pair difference vectors).  -->
+ggplot(weat, aes(x=x,y=y, label = words))+geom_point()+geom_text(check_overlap = TRUE,
+                                                                 nudge_y = 0.018, size = 7 )+
+    geom_segment(data=weat, mapping=aes(x=.1, y=.3, xend=.5, yend=.2), , linetype =2, arrow=arrow(), size=.05, color="blue") +
+  geom_segment(data=weat, mapping=aes(x=.1, y=.3, xend=.5, yend=.3), linetype =2, arrow=arrow(), size=.05, color="blue")+
+  geom_segment(data=weat, mapping=aes(x=.1, y=.3, xend=.7, yend=.4), linetype =2, arrow=arrow(), size=.05, color="orangered")+
+  geom_segment(data=weat, mapping=aes(x=.1, y=.3, xend=.7, yend=.5), linetype =2, arrow=arrow(), size=.05, color="orangered")+
+    annotate("text", x = c(.32,.32,.32,.32), y = c(.23,.29,.35,.39), 
+             label = c(".7", ".6","1.1", "1.2") , color="black", 
+             size=7,  fontface="bold")+
+    annotate("text", x = c(.1,.5,.7,1.2), y = c(.55,.55,.55,.55), 
+           label = c("X", "A","B", "Y") , color=c("black", "blue","orangered","black"), 
+           size=10,  fontface="bold")+xlim(0,1.3)+
+  theme_void(base_size = 25)
+```
+\normalsize
 
--  Given the (ideally) gender neutral words $N$ and the gender direction $g$ the direct gender bias is:
-<!-- ($c$ is a parameter determining how strict we want to be): -->
-
-\vspace{-2mm}
-
-\begin{align}
-\mathsf{directBias_c(N,g)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},g)\vert^c}{\vert N \vert }
-\end{align}
+\pause
 
@@
 
-[@Bolukbasi2016Man]
+- $s_1 = s(he,A,B) =  \frac{.6+.7}{2}  - \frac{1.2+1.1}{2} = 0.65-1.15= -0.5$
+
+- $s_2 = s(man,A,B) = -.3$, \linebreak  $s_3 = s(woman,A,B) = .2, s_4 = s(she, A, B) = .3$
+
+\vspace{-4mm}
+
+\normalsize 
+\begin{align*}
+\mathsf{WEAT}(A,B) & = \frac{\nicefrac{(s_1+s_2)}{2} - \nicefrac{(s_3+s_4)}{2}}{sd(\{s_1,s_2,s_3,s_4\})} \approx - 1.95
+\end{align*}
 
@@
 
--  $t$ is a term,  $A, B$ are sets of stereotype attribute words, $X, $Y$ are protected group words
+-  $t$ is a term,  $A, B$ are sets of stereotype attribute words, $X$, $Y$ are protected group words
 
@@
 [@Caliskan2017semanticsBiases] with extensions in [@Lauscher2019multidimensional] and applications in [@Garg2018years]
+
+## Cosine-based measures of bias
+
+### Our main target: Mean Average Cosine Similarity (MAC)
+
+\vspace{1mm}
+\footnotesize
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "50%"}
+words <- c("jew", "muslim", "christian", "terrorist","dirty","conservative","familiar","greedy","cheap")
+x <- c(.1,.1,.1,.4,.4,.7,.7,1.1,1.1)
+y <- c(.4,.3,.2,.1,.2,.3,.4,.1,.2)
+mac <- data.frame(words,x)
+
+ggplot(mac, aes(x=x,y=y, label = words))+geom_point()+geom_text(check_overlap = TRUE,
+                                                                 nudge_y = 0.02, size = 9 )+
+  geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=.4, yend=.1), , linetype =2, arrow=arrow(), size=.03, color="blue")+geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=.4, yend=.2), , linetype =4, arrow=arrow(), size=.0, color="blue") + geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=.7, yend=.3), linetype =2, arrow=arrow(), size=.05, color="orangered")+ geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=.7, yend=.4), linetype =2, arrow=arrow(), size=.05, color="orangered")+ geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=1.1, yend=.1), linetype =3, arrow=arrow(), size=.05, color="chartreuse4")+ geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=1.1, yend=.2), linetype =2, arrow=arrow(), size=.05, color="chartreuse4")+ annotate("text", x = c(.1,.4,.7,1.2), y = c(.45,.45,.45,.45),  label = c("T", "A[1]","A[2]", "A[3]") ,parse = TRUE, color=c("black", "blue","orangered","chartreuse4"),            size=10,  fontface="bold")+xlim(0,1.3)+theme_void(base_size = 25)
+
+```
+\normalsize
+
+\vspace{-4mm}
+
+\footnotesize 
+\begin{align*}
+ s_1 & = s(muslim,A_1) = \frac{cos(muslim,dirty)+cos(muslim,terrorist)}{2}\\
+s_2 & = s(muslim,A_2) = \frac{cos(muslim,familiar)+cos(muslim,conservative)}{2}\\
+& \vdots \end{align*}
+
+\vspace{-6mm}
+
+\normalsize 
+\begin{align*}
+ MAC(T,A) & = \mathsf{mean}(\{s_i \vert i \in 1, \dots, k\})
+\end{align*}
 
@@
 
+- Sensitive to the choice of protected words, capitalization, distance measures and embedding methods [@zhang2020robustness]
+
 <!-- dodac? -->
@@
 
-We investigate the uncertainty that arises from raw sample sizes
+We statistically gauge the uncertainty that arises from raw sample sizes
 
@@
 
-We make the subclasses clear, add human neutral predicates and neutral predicates for control
+We make the subclasses clear, add human neutral predicates and neutral predicates for control. We used L2-Reddit corpus and GoogleNews (we present the results for Reddit for brevity).
 
@@
 
+### Key conceptual issues
+
 - It throws away information about sample sizes
-- It removes variation which may result in false confidence
+- It ignores variation in the raw data, which leads to false confidence
 
-##  Cosine distance and word connection visualization
+### Our simulation
 
-### Analysis of a word "muslim"
+Suppose all similarities for two classes are randomly drawn from the same distribution, $\mathsf{Normal}(0,.05)$, you still can get really high WEAT!
 
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "105%",message =FALSE, warning = FALSE}
+## The problem with pre-averaging
+
+### One simulation
+
+\vspace{1mm}
+\footnotesize
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "70%"}
+set.seed(123)
+t1 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
+t2 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
+t3 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
+t4 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
+xlimits <- c(-.3,.3)
+
+t1Plot <- ggplot(t1)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t1 in X")+xlim(xlimits)
+
+t2Plot <- ggplot(t2)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t2 in X")+xlim(xlimits)
+
+t3Plot <- ggplot(t3)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t3 in Y")+xlim(xlimits)
+
+t4Plot <- ggplot(t4)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t4 in Y")+xlim(xlimits)
+
+grid.arrange(t1Plot,t2Plot, t3Plot, t4Plot, ncol=2)
+```
+\normalsize
+
+\vspace{1mm}
+\footnotesize
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+s <- function (table){ mean(table$A) - mean(table$B)}
+whole <- rbind(t1,t2,t3,t4)
+rawsd <- sd(c(whole$A,whole$B))
+factor <- sd(c(s(t1),s(t2),s(t3),s(t4)))
+numerator <-  mean(s(t1),s(t2)) - mean(s(t3),s(t4))
+```
+\normalsize
+\pause
+
+\footnotesize
+
+\vspace{-4mm}
+
+- Raw sd in data is `r round(rawsd,3)`
+- The sd of means is `r round(factor,3)`
+- The WEAT score is `r round(numerator/factor,3)`
+- The largest effect size reported by @Caliskan2017semanticsBiases is 1.81!
+
+## The problem with pre-averaging
+
+### 50k simulations (same parameters)
+
+\vspace{1mm}
+\footnotesize
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "80%"}
+set.seed(123)
+biasesNull <- numeric(50000)
+for(i in 1:50000){
+t1 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
+t2 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
+t3 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
+t4 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
+
+factor <- sd(c(s(t1),s(t2),s(t3),s(t4)))
+numerator <-  mean(s(t1),s(t2)) - mean(s(t3),s(t4))
+biasesNull[i]  <- numerator / factor
+}
+
+ggplot()+geom_histogram(aes(x=biasesNull, y = ..density..), alpha = 0.6, bins=50)+
+  theme_tufte(base_size = 15)+labs(title="50k biases for identical means and sd =.05")+ xlab("bias")
+```
+\normalsize
+
+##   Cosine distance and  connection type
+
+### Distances for  "muslim"
+
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "80%",message =FALSE, warning = FALSE}
 library(tidyverse)
@@
 #two arguments: dataset and protected word
-visualiseProtected(muslimVis,"muslim")+theme_tufte(base_size = 20)
+visualiseProtected(muslimVis,"muslim")+theme_void(base_size = 20)
 ```
 
-## Cosine distance and word connection visualization
+## Cosine distance and  connection type
 
-### Analysis of a word "priest"
+### Distances for  "priest"
 
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "105%"}
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "80%"}
 library(tidyverse)
@@
 #two arguments: dataset and protected word
-visualiseProtected(priestVis,"priest")+theme_tufte(base_size = 20)
+visualiseProtected(priestVis,"priest")+theme_void(base_size = 20)
 ```
 
-## Advantages of including uncertainty
+## Advantages of the Bayesian way
 
-- It enables one to directly observe the influence of sample sizes
-- It may influence risk assessment and decision making
-- ...
+- Direct impact of sample sizes
+- Straightforward interpretation in terms of posterior probabilities
+- Freedom to choose granularity level
+- More honest risk assessment and decision making
 
@@
 p <- ggplot(data = data.frame(distance = dis), 
-            mapping = aes(x = distance))+theme_tufte()
+            mapping = aes(x = distance))+theme_tufte(base_size = 15)
 
@@
 
-##  Dataset-level HPDIs coefficients
+## Bayesian model architecture
 
-### Religion coefficients
+\vspace{1mm}
+\footnotesize
+```{r,echo=TRUE,eval=FALSE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+library(rethinking)
+options(buildtools.check = function(action) TRUE )
+religionCoefs <- ulam(
+  alist(
+    cosineDistance ~ dnorm(mu,sigma),
+    mu <- m + co[con],
+    m ~ dnorm(1,.5),
+    co[con] ~ dnorm(0,.5),
+    sigma ~ dcauchy(0,1)
+  ),
+  data = religion,
+  chains=2 , iter=8000 , warmup=1000, 
+  log_lik = TRUE
+)
+```
+\normalsize
+
+##   Dataset-level coefficients
+
+### Religion with 89\%-compatibility intervals (HPDI)
 
@@
    \begin{minipage}{0.55\textwidth}
-  \includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
+  \includegraphics[width=6cm]{../images/religionCoeffs.jpeg}
+   \end{minipage} \begin{minipage}{0.4\textwidth}\footnotesize
+   \begin{itemize}
+   \item All HPDIs overlap with 0 
+   \item Differences between classes are relatively small
+   \item Coefficients for Race are similar
+   \end{itemize}
    \end{minipage}
@@
 
-##  Dataset-level HPDIs coefficients
+##  Dataset-level  coefficients
 
-### Gender coefficients
+### Gender with 89\%-compatibility intervals (HPDI)
 
@@
   \begin{minipage}{0.55\textwidth}
-\includegraphics[width=7cm]{../images/genderCoeffs.jpeg}
+\includegraphics[width=6cm]{../images/genderCoeffs.jpeg}
 \end{minipage}
-
+\begin{minipage}{0.4\textwidth}\footnotesize
+   \begin{itemize}
+   \item Associated and different are away from 0
+   \item But they were supposed to be opposites and are very close to each other (co-occurrence?)
+   \item Differences between classes are still relatively small
+   \end{itemize}
+   \end{minipage}
 \end{figure}
@@
 
-## Uncertainty included in bias detection
+## Bayesian model architecture
 
-### Closer look at Gender class
+\vspace{1mm}
+\footnotesize
+```{r,echo=TRUE,eval=FALSE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+library(rethinking)
+options(buildtools.check = function(action) TRUE )
+religionCoefs <- ulam(
+  alist(
+    cosineDistance ~ dnorm(mu,sigma),
+    mu <- m[pw] + co[con],
+    m[pw] ~ dnorm(1,.5),
+    co[con] ~ dnorm(0,.5),
+    sigma ~ dcauchy(0,1)
+  ),
+  data = religion,
+  chains=2 , iter=8000 , warmup=1000, 
+  log_lik = TRUE
+)
+```
+\normalsize
 
-\includegraphics[width=8cm]{../images/visGenderReddit.png}
+## Word-level coefficients
 
-## Summary of the research
+\begin{figure}[!htb]\centering
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7.5cm]{../images/visReligionReddit.png}
+\end{minipage}
+\begin{minipage}{0.4\textwidth}\footnotesize
 
--
+\vspace{-4cm}
+
+   \begin{itemize}
+   \item Most intervals overlap with control groups
+   \item Often not too much diffence between associated and different
+   \end{itemize}
+   \end{minipage}
+\end{figure}
+
+## Word-level coefficients
+
+\begin{figure}[!htb]\centering
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7.5cm]{../images/visGenderReddit.png}
+\end{minipage}
+\begin{minipage}{0.4\textwidth}\footnotesize
+
+\vspace{-4cm}
+
+   \begin{itemize}
+   \item Male attributes strong co-occurrence with female attributes
+   \item Sometimes different is stronger than associated
+   \item Almost no overlap with control groups
+   \end{itemize}
+   \end{minipage}
+\end{figure}
+
+## Word-level coefficients
+
+\begin{figure}[!htb]\centering
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7.5cm]{../images/visRaceReddit.png}
+\end{minipage}
+\begin{minipage}{0.4\textwidth}\footnotesize
+
+\vspace{-4cm}
+
+   \begin{itemize}
+   \item A lot of variation between races
+
+   \item Often not much difference between associated and different
+   \end{itemize}
+   \end{minipage}
+\end{figure}
 
@@
 
-- including contrasts in Bayesian model
-- downstream tasks
-- applying uncertainty to WEAT metric
-- testing AIT dataset
+- Including contrasts in Bayesian calculation
+- Performance cross-validation in comparison to other methods (regular linear regression, KNN, \dots?)
+- Downstream tasks and connection with intrinsic evaluation
+- Testing data from the original Implicit Association Test (IAT) 
+- Applying uncertainty to WEAT and better word lists
+- Looking at other similarity measures
 

### 2021-08-03 presentationESSLLI21/presentationESSLLI.tex
Adds presentation changes
Adds small fixes
r

--- 2021-07-30 presentationESSLLI21/presentationESSLLI.tex
@@
 \hypersetup{
-            pdftitle={Taking uncertainty in word embedding bias estimation seriously ~a Bayesian approach},
+            pdftitle={Taking uncertainty seriously A Bayesian approach to word embedding bias estimation},
             pdfauthor={Alicja Dobrzeniecka \& Rafal Urbaniak (LoPSE research group, University of Gdansk)},
@@
 
-\title{\large Taking uncertainty in word embedding bias estimation seriously ~a
-Bayesian approach}
+\title{\Large Taking uncertainty seriously \newline \normalsize A Bayesian
+approach to word embedding bias estimation}
 \author{Alicja Dobrzeniecka \& Rafal Urbaniak \footnotesize \newline (LoPSE
@@
 
-\begin{block}{Stereotypical lists}
+\begin{block}{A visual example}
 
@@
 \item
-  masculine occupations: ``maestro'', ``captain'', ``architect'', etc.
+  masculine occupations: ``maestro'', ``captain'', ``architect'',
+  ``boss'', etc.
 \end{itemize}
@@
 \normalsize
-
-\end{block}
-
-\begin{block}{Visual example}
 
@@
 
+--\textgreater{}
+
+--\textgreater{}
+
 \end{block}
@@
 
-\begin{block}{Example: direct bias}
+\begin{block}{Example: Word Embedding Association Test (WEAT)}
 
-\begin{itemize}
-\tightlist
-\item
-  The gender bias of a word \(w\) is its projection on the gender
-  direction
-  \(\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})\)
-\end{itemize}
+\vspace{1mm} \footnotesize
 
-\begin{itemize}
-\tightlist
-\item
-  Given the (ideally) gender neutral words \(N\) and the gender
-  direction \(g\) the direct gender bias is: 
-\end{itemize}
+\begin{center}\includegraphics[width=0.55\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-2-1} \end{center}
 
-\vspace{-2mm}
+\normalsize
 
-\begin{align}
-\mathsf{directBias_c(N,g)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},g)\vert^c}{\vert N \vert }
-\end{align}
+\pause
 
@@
 
-(Bolukbasi, Chang, Zou, Saligrama, \& Kalai,
-\protect\hyperlink{ref-Bolukbasi2016Man}{2016})
+\begin{itemize}
+\item
+  \(s_1 = s(he,A,B) = \frac{.6+.7}{2} - \frac{1.2+1.1}{2} = 0.65-1.15= -0.5\)
+\item
+  \(s_2 = s(man,A,B) = -.3\),
+  \linebreak  \(s_3 = s(woman,A,B) = .2, s_4 = s(she, A, B) = .3\)
+\end{itemize}
+
+\vspace{-4mm}
+
+\normalsize
+
+\begin{align*}
+\mathsf{WEAT}(A,B) & = \frac{\nicefrac{(s_1+s_2)}{2} - \nicefrac{(s_3+s_4)}{2}}{sd(\{s_1,s_2,s_3,s_4\})} \approx - 1.95
+\end{align*}
 
@@
 \item
-  \(t\) is a term, \(A, B\) are sets of stereotype attribute words, \$X,
-  \(Y\) are protected group words
+  \(t\) is a term, \(A, B\) are sets of stereotype attribute words,
+  \(X\), \(Y\) are protected group words
 \item
@@
 \protect\hyperlink{ref-Garg2018years}{2018})
+
+\end{block}
+
+\end{frame}
+
+\begin{frame}{Cosine-based measures of bias}
+
+\begin{block}{Our main target: Mean Average Cosine Similarity (MAC)}
+
+\vspace{1mm} \footnotesize
+
+\begin{center}\includegraphics[width=0.5\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-3-1} \end{center}
+
+\normalsize
+
+\vspace{-4mm}
+
+\footnotesize
+
+\begin{align*}
+ s_1 & = s(muslim,A_1) = \frac{cos(muslim,dirty)+cos(muslim,terrorist)}{2}\\
+s_2 & = s(muslim,A_2) = \frac{cos(muslim,familiar)+cos(muslim,conservative)}{2}\\
+& \vdots \end{align*}
+
+\vspace{-6mm}
+
+\normalsize
+
+\begin{align*}
+ MAC(T,A) & = \mathsf{mean}(\{s_i \vert i \in 1, \dots, k\})
+\end{align*}
 
@@
   \protect\hyperlink{ref-Nissim2020fair}{2020})
+\item
+  Sensitive to the choice of protected words, capitalization, distance
+  measures and embedding methods (Zhang, Sneyd, \& Stevenson,
+  \protect\hyperlink{ref-zhang2020robustness}{2020})
 \end{itemize}
@@
 We make the subclasses clear, add human neutral predicates and neutral
-predicates for control
+predicates for control. We used L2-Reddit corpus and GoogleNews (we
+present the results for Reddit for brevity).
 
@@
 
-\end{frame}
-
-\begin{frame}{Bayesian estimation}
-
-\begin{center}\includegraphics[width=1.05\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-2-1} \end{center}
-
-\end{frame}
-
-\begin{frame}{Bayesian estimation}
-
-\begin{center}\includegraphics[width=1.05\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-3-1} \end{center}
-
-\end{frame}
-
-\begin{frame}{Effects of debiasing}
-
-\begin{block}{Further work}
+\begin{block}{Key conceptual issues}
 
@@
 \item
-  downstream tasks
+  It throws away information about sample sizes
+\item
+  It ignores variation in the raw data, which leads to false confidence
 \end{itemize}
@@
 \end{block}
+
+\begin{block}{Our simulation}
+
+Suppose all similarities for two classes are randomly drawn from the
+same distribution, \(\mathsf{Normal}(0,.05)\), you still can get really
+high WEAT!
+
+\end{block}
+
+\end{frame}
+
+\begin{frame}{The problem with pre-averaging}
+
+\begin{block}{One simulation}
+
+\vspace{1mm} \footnotesize
+
+\begin{center}\includegraphics[width=0.7\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-4-1} \end{center}
+
+\normalsize
+
+\vspace{1mm} \footnotesize
+
+\normalsize
+\pause
+
+\footnotesize
+
+\vspace{-4mm}
+
+\begin{itemize}
+\tightlist
+\item
+  Raw sd in data is 0.045
+\item
+  The sd of means is 0.023
+\item
+  The WEAT score is 1.825
+\item
+  The largest effect size reported by Caliskan et al.
+  (\protect\hyperlink{ref-Caliskan2017semanticsBiases}{2017}) is 1.81!
+\end{itemize}
+
+\end{block}
+
+\end{frame}
+
+\begin{frame}{The problem with pre-averaging}
+
+\begin{block}{50k simulations (same parameters)}
+
+\vspace{1mm} \footnotesize
+
+\begin{center}\includegraphics[width=0.8\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-6-1} \end{center}
+
+\normalsize
+
+\end{block}
+
+\end{frame}
+
+\begin{frame}{Cosine distance and connection type}
+
+\begin{block}{Distances for ``muslim''}
+
+\begin{center}\includegraphics[width=0.8\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-7-1} \end{center}
+
+\end{block}
+
+\end{frame}
+
+\begin{frame}{Cosine distance and connection type}
+
+\begin{block}{Distances for ``priest''}
+
+\begin{center}\includegraphics[width=0.8\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-8-1} \end{center}
+
+\end{block}
+
+\end{frame}
+
+\begin{frame}{Advantages of the Bayesian way}
+
+\begin{itemize}
+\tightlist
+\item
+  Direct impact of sample sizes
+\item
+  Straightforward interpretation in terms of posterior probabilities
+\item
+  Freedom to choose granularity level
+\item
+  More honest risk assessment and decision making
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}{Bayesian model}
+
+\begin{block}{Choosing priors}
+
+\begin{center}\includegraphics[width=1\linewidth]{presentationESSLLI_files/figure-beamer/priorsVis-1} \end{center}
+
+\end{block}
+
+\end{frame}
+
+\begin{frame}[fragile]{Bayesian model architecture}
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(rethinking)}
+\KeywordTok{options}\NormalTok{(}\DataTypeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\OtherTok{TRUE}\NormalTok{ )}
+\NormalTok{religionCoefs <-}\StringTok{ }\KeywordTok{ulam}\NormalTok{(}
+  \KeywordTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu <-}\StringTok{ }\NormalTok{m }\OperatorTok{+}\StringTok{ }\NormalTok{co[con],}
+\NormalTok{    m }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    co[con] }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\OperatorTok{~}\StringTok{ }\KeywordTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\NormalTok{  ),}
+  \DataTypeTok{data =}\NormalTok{ religion,}
+  \DataTypeTok{chains=}\DecValTok{2}\NormalTok{ , }\DataTypeTok{iter=}\DecValTok{8000}\NormalTok{ , }\DataTypeTok{warmup=}\DecValTok{1000}\NormalTok{, }
+  \DataTypeTok{log_lik =} \OtherTok{TRUE}
+\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+\end{frame}
+
+\begin{frame}{Dataset-level coefficients}
+
+\begin{block}{Religion with 89\%-compatibility intervals (HPDI)}
+
+\begin{center}
+\begin{figure}[!htb]\centering
+   \begin{minipage}{0.55\textwidth}
+  \includegraphics[width=6cm]{../images/religionCoeffs.jpeg}
+   \end{minipage} \begin{minipage}{0.4\textwidth}\footnotesize
+   \begin{itemize}
+   \item All HPDIs overlap with 0 
+   \item Differences between classes are relatively small
+   \item Coefficients for Race are similar
+   \end{itemize}
+   \end{minipage}
+  \end{figure}
+
+\end{center}
+
+\end{block}
+
+\end{frame}
+
+\begin{frame}{Dataset-level coefficients}
+
+\begin{block}{Gender with 89\%-compatibility intervals (HPDI)}
+
+\begin{center}
+\begin{figure}[!htb]\centering
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=6cm]{../images/genderCoeffs.jpeg}
+\end{minipage}
+\begin{minipage}{0.4\textwidth}\footnotesize
+   \begin{itemize}
+   \item Associated and different are away from 0
+   \item But they were supposed to be opposites and are very close to each other (co-occurrence?)
+   \item Differences between classes are still relatively small
+   \end{itemize}
+   \end{minipage}
+\end{figure}
+
+\end{center}
+
+\end{block}
+
+\end{frame}
+
+\begin{frame}[fragile]{Bayesian model architecture}
+
+\vspace{1mm} \footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(rethinking)}
+\KeywordTok{options}\NormalTok{(}\DataTypeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\OtherTok{TRUE}\NormalTok{ )}
+\NormalTok{religionCoefs <-}\StringTok{ }\KeywordTok{ulam}\NormalTok{(}
+  \KeywordTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu <-}\StringTok{ }\NormalTok{m[pw] }\OperatorTok{+}\StringTok{ }\NormalTok{co[con],}
+\NormalTok{    m[pw] }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    co[con] }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\OperatorTok{~}\StringTok{ }\KeywordTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\NormalTok{  ),}
+  \DataTypeTok{data =}\NormalTok{ religion,}
+  \DataTypeTok{chains=}\DecValTok{2}\NormalTok{ , }\DataTypeTok{iter=}\DecValTok{8000}\NormalTok{ , }\DataTypeTok{warmup=}\DecValTok{1000}\NormalTok{, }
+  \DataTypeTok{log_lik =} \OtherTok{TRUE}
+\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+
+\end{frame}
+
+\begin{frame}{Word-level coefficients}
+
+\begin{figure}[!htb]\centering
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7.5cm]{../images/visReligionReddit.png}
+\end{minipage}
+\begin{minipage}{0.4\textwidth}\footnotesize
+
+\vspace{-4cm}
+
+   \begin{itemize}
+   \item Most intervals overlap with control groups
+   \item Often not too much diffence between associated and different
+   \end{itemize}
+   \end{minipage}
+\end{figure}
+
+\end{frame}
+
+\begin{frame}{Word-level coefficients}
+
+\begin{figure}[!htb]\centering
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7.5cm]{../images/visGenderReddit.png}
+\end{minipage}
+\begin{minipage}{0.4\textwidth}\footnotesize
+
+\vspace{-4cm}
+
+   \begin{itemize}
+   \item Male attributes strong co-occurrence with female attributes
+   \item Sometimes different is stronger than associated
+   \item Almost no overlap with control groups
+   \end{itemize}
+   \end{minipage}
+\end{figure}
+
+\end{frame}
+
+\begin{frame}{Word-level coefficients}
+
+\begin{figure}[!htb]\centering
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7.5cm]{../images/visRaceReddit.png}
+\end{minipage}
+\begin{minipage}{0.4\textwidth}\footnotesize
+
+\vspace{-4cm}
+
+   \begin{itemize}
+   \item A lot of variation between races
+
+   \item Often not much difference between associated and different
+   \end{itemize}
+   \end{minipage}
+\end{figure}
+
+\end{frame}
+
+\begin{frame}{Further work}
+
+\begin{itemize}
+\tightlist
+\item
+  Including contrasts in Bayesian calculation
+\item
+  Performance cross-validation in comparison to other methods (regular
+  linear regression, KNN, \dots?)
+\item
+  Downstream tasks and connection with intrinsic evaluation
+\item
+  Testing data from the original Implicit Association Test (IAT)
+\item
+  Applying uncertainty to WEAT and better word lists
+\item
+  Looking at other similarity measures
+\end{itemize}
 
@@
 \hypertarget{refs}{}
-\hypertarget{ref-Bolukbasi2016Man}{}
-Bolukbasi, T., Chang, K., Zou, J. Y., Saligrama, V., \& Kalai, A.
-(2016). Man is to computer programmer as woman is to homemaker?
-Debiasing word embeddings. \emph{CoRR}, \emph{abs/1607.06520}. Retrieved
-from \url{http://arxiv.org/abs/1607.06520}
-
 \hypertarget{ref-Caliskan2017semanticsBiases}{}
@@
 
+\hypertarget{ref-zhang2020robustness}{}
+Zhang, H., Sneyd, A., \& Stevenson, M. (2020). \emph{Robustness and
+reliability of gender bias assessment in word embeddings: The role of
+base pairs}.
+
 \end{frame}

### 2021-08-05 presentationESSLLI21/presentationESSLLI.Rmd
Adds current presentation version

@@
 
+## The plan of the presentation
+
+- Introduction to the topic of bias in word embeddings
+- Analysis of methodological problems with MAC implementation
+- Explanation of pre-averaging limitations in methods such as WEAT and MAC
+- Our research and Bayesian approach
+
 ## Cosine-based measures of bias
@@
 
-- Built to predict the probability of co-occurence
+- Often built to predict the probability of co-occurence
 
@@
 
-- Naive interpretation: proximity corresponds to semantic similarity (e.g. no triangle inequality)
+- Naive interpretation: proximity corresponds to semantic similarity
 
@@
 
-In the learning process these models can learn implicit biases that reflect harmful stereotypical thinking
+In the learning process word embeddings can learn implicit biases that reflect harmful stereotypical thinking
 
@@
     annotate("text", x = c(.32,.32,.32,.32), y = c(.23,.29,.35,.39), 
-             label = c(".7", ".6","1.1", "1.2") , color="black", 
+             label = c(".7", ".6",".1", ".2") , color="black", 
              size=7,  fontface="bold")+
@@
 
-- $s_1 = s(he,A,B) =  \frac{.6+.7}{2}  - \frac{1.2+1.1}{2} = 0.65-1.15= -0.5$
+- $s_1 = s(he,A,B) =  \frac{.6+.7}{2}  - \frac{.2+.1}{2} = .65-.15= .5$
 
-- $s_2 = s(man,A,B) = -.3$, \linebreak  $s_3 = s(woman,A,B) = .2, s_4 = s(she, A, B) = .3$
+- $s_2 = s(man,A,B) = .3$, \linebreak  $s_3 = s(woman,A,B) = -.6, s_4 = s(she, A, B) = -.3$
 
@@
 \begin{align*}
-\mathsf{WEAT}(A,B) & = \frac{\nicefrac{(s_1+s_2)}{2} - \nicefrac{(s_3+s_4)}{2}}{sd(\{s_1,s_2,s_3,s_4\})} \approx - 1.95
+\mathsf{WEAT}(A,B) & = \frac{\nicefrac{(s_1+s_2)}{2} - \nicefrac{(s_3+s_4)}{2}}{sd(\{s_1,s_2,s_3,s_4\})} \approx 1.93
 \end{align*}
@@
 
+## Summary
+
+- Introduced to the topic of bias in word embeddings
+- Analysed methodological problems with MAC implementation
+- Explained of pre-averaging limitations in methods such as WEAT and MAC
+- Proposed Bayesian approach
+
 ## Further work
@@
 - Including contrasts in Bayesian calculation
-- Performance cross-validation in comparison to other methods (regular linear regression, KNN, \dots?)
+- Performance cross-validation in comparison to other methods (regular linear regression, KNN, \dots)
 - Downstream tasks and connection with intrinsic evaluation

### 2021-08-05 presentationESSLLI21/presentationESSLLI.tex
Adds current presentation version

--- 2021-08-01 presentationESSLLI21/presentationESSLLI.tex
@@
 \hypersetup{
-  pdftitle={Taking uncertainty in word embedding bias estimation seriously ~a Bayesian approach},
+  pdftitle={Taking uncertainty seriously A Bayesian approach to word embedding bias estimation},
   pdfauthor={Alicja Dobrzeniecka \& Rafal Urbaniak (LoPSE research group, University of Gdansk)},
@@
 \newif\ifbibliography
+\usepackage{color}
+\usepackage{fancyvrb}
+\newcommand{\VerbBar}{|}
+\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+% Add ',fontsize=\small' for more characters per line
+\usepackage{framed}
+\definecolor{shadecolor}{RGB}{248,248,248}
+\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\BuiltInTok}[1]{#1}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
+\newcommand{\ExtensionTok}[1]{#1}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ImportTok}[1]{#1}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\NormalTok}[1]{#1}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\RegionMarkerTok}[1]{#1}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
 \usepackage{longtable,booktabs,array}
@@
 
-\title{\large Taking uncertainty in word embedding bias estimation
-seriously ~a Bayesian approach}
+\title{\Large Taking uncertainty seriously \newline \normalsize A
+Bayesian approach to word embedding bias estimation}
 \author{Alicja Dobrzeniecka \& Rafal Urbaniak
@@
 
+\begin{frame}{The plan of the presentation}
+\protect\hypertarget{the-plan-of-the-presentation}{}
+\begin{itemize}
+\tightlist
+\item
+  Introduction to the topic of bias in word embeddings
+\item
+  Analysis of methodological problems with MAC implementation
+\item
+  Explanation of pre-averaging limitations in methods such as WEAT and
+  MAC
+\item
+  Our research and Bayesian approach
+\end{itemize}
+\end{frame}
+
 \begin{frame}{Cosine-based measures of bias}
@@
 \item
-  Representation of words (or words in contexts) with vectors of real
-  numbers
+  Representation of words with vectors of real numbers
 \item
-  Built to predict the probability of co-occurence
+  Often built to predict the probability of co-occurence
 \end{itemize}
@@
   Naive interpretation: proximity corresponds to semantic similarity
-  (e.g.~no triangle inequality)
 \end{itemize}
@@
 \protect\hypertarget{the-worry}{}
-In the learning process these models can learn implicit biases that
+In the learning process word embeddings can learn implicit biases that
 reflect harmful stereotypical thinking
@@
 \protect\hypertarget{cosine-based-measures-of-bias-3}{}
-\begin{block}{Stereotypical lists}
-\protect\hypertarget{stereotypical-lists}{}
+\begin{block}{A visual example}
+\protect\hypertarget{a-visual-example}{}
 \footnotesize
@@
 \item
-  masculine occupations: ``maestro,'' ``captain,'' ``architect,'' etc.
+  masculine occupations: ``maestro,'' ``captain,'' ``architect,''
+  ``boss,'' etc.
 \end{itemize}
@@
 \normalsize
+
+--\textgreater{}
+
+--\textgreater{}
 \end{block}
@@
 \protect\hypertarget{cosine-based-measures-of-bias-4}{}
-\begin{block}{Example: direct bias}
-\protect\hypertarget{example-direct-bias}{}
-\begin{itemize}
-\tightlist
-\item
-  The gender bias of a word \(w\) is its projection on the gender
-  direction
-  \(\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})\)
-\end{itemize}
+\begin{block}{Example: Word Embedding Association Test (WEAT)}
+\protect\hypertarget{example-word-embedding-association-test-weat}{}
+\vspace{1mm}
+\footnotesize
 
-\begin{itemize}
-\tightlist
-\item
-  Given the (ideally) gender neutral words \(N\) and the gender
-  direction \(g\) the direct gender bias is:
-\end{itemize}
+\begin{center}\includegraphics[width=0.55\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-2-1} \end{center}
+\normalsize
 
-\vspace{-2mm}
-
-\begin{align}
-\mathsf{directBias_c(N,g)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},g)\vert^c}{\vert N \vert }
-\end{align}
+\pause
 
@@
 
-(\protect\hyperlink{ref-Bolukbasi2016Man}{Bolukbasi, Chang, Zou,
-Saligrama, \& Kalai, 2016})
+\begin{itemize}
+\item
+  \(s_1 = s(he,A,B) = \frac{.6+.7}{2} - \frac{.2+.1}{2} = .65-.15= .5\)
+\item
+  \(s_2 = s(man,A,B) = .3\),
+  \linebreak  \(s_3 = s(woman,A,B) = -.6, s_4 = s(she, A, B) = -.3\)
+\end{itemize}
+
+\vspace{-4mm}
+
+\normalsize
+
+\begin{align*}
+\mathsf{WEAT}(A,B) & = \frac{\nicefrac{(s_1+s_2)}{2} - \nicefrac{(s_3+s_4)}{2}}{sd(\{s_1,s_2,s_3,s_4\})} \approx 1.93
+\end{align*}
 \end{block}
@@
 \begin{block}{Example: Word Embedding Association Test (WEAT)}
-\protect\hypertarget{example-word-embedding-association-test-weat}{}
+\protect\hypertarget{example-word-embedding-association-test-weat-1}{}
 \begin{align*}
@@
 \protect\hypertarget{our-main-target-mean-average-cosine-similarity-mac}{}
+\vspace{1mm}
+\footnotesize
+
+\begin{center}\includegraphics[width=0.5\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-3-1} \end{center}
+\normalsize
+
+\vspace{-4mm}
+
+\footnotesize
+
+\begin{align*}
+ s_1 & = s(muslim,A_1) = \frac{cos(muslim,dirty)+cos(muslim,terrorist)}{2}\\
+s_2 & = s(muslim,A_2) = \frac{cos(muslim,familiar)+cos(muslim,conservative)}{2}\\
+& \vdots \end{align*}
+
+\vspace{-6mm}
+
+\normalsize
+
+\begin{align*}
+ MAC(T,A) & = \mathsf{mean}(\{s_i \vert i \in 1, \dots, k\})
+\end{align*}
+\end{block}
+\end{frame}
+
+\begin{frame}{Cosine-based measures of bias}
+\protect\hypertarget{cosine-based-measures-of-bias-7}{}
+\begin{block}{Our main target: Mean Average Cosine Similarity (MAC)}
+\protect\hypertarget{our-main-target-mean-average-cosine-similarity-mac-1}{}
 \begin{align*}
@@
 \begin{frame}{Cosine-based measures of bias}
-\protect\hypertarget{cosine-based-measures-of-bias-7}{}
+\protect\hypertarget{cosine-based-measures-of-bias-8}{}
 \begin{block}{Our main target: Mean Average Cosine Similarity (MAC)}
-\protect\hypertarget{our-main-target-mean-average-cosine-similarity-mac-1}{}
+\protect\hypertarget{our-main-target-mean-average-cosine-similarity-mac-2}{}
 \footnotesize 
@@
 \begin{frame}{Cosine-based measures of bias}
-\protect\hypertarget{cosine-based-measures-of-bias-8}{}
+\protect\hypertarget{cosine-based-measures-of-bias-9}{}
 \begin{block}{Known challenges}
@@
   (\protect\hyperlink{ref-Nissim2020fair}{Nissim, Noord, \& Goot, 2020})
+\item
+  Sensitive to the choice of protected words, capitalization, distance
+  measures and embedding methods
+  (\protect\hyperlink{ref-zhang2020robustness}{Zhang, Sneyd, \&
+  Stevenson, 2020})
 \end{itemize}
@@
 \protect\hypertarget{no-design-considerations-to-sample-size}{}
-We investigate the uncertainty that arises from raw sample sizes
+We statistically gauge the uncertainty that arises from raw sample sizes
 \end{block}
@@
 \protect\hypertarget{the-problem-with-pre-averaging}{}
+\begin{block}{Key conceptual issues}
+\protect\hypertarget{key-conceptual-issues}{}
 \begin{itemize}
@@
 \item
-  It removes variation which may result in false confidence
+  It ignores variation in the raw data, which leads to false confidence
 \end{itemize}
-\end{frame}
+\end{block}
 
-\begin{frame}{Cosine distance and word connection visualization}
-\protect\hypertarget{cosine-distance-and-word-connection-visualization}{}
-\begin{block}{Analysis of word ``muslim''}
-\protect\hypertarget{analysis-of-word-muslim}{}
-\begin{center}\includegraphics[width=1.05\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-2-1} \end{center}
+\begin{block}{Our simulation}
+\protect\hypertarget{our-simulation}{}
+Suppose all similarities for two classes are randomly drawn from the
+same distribution, \(\mathsf{Normal}(0,.05)\), you still can get really
+high WEAT!
 \end{block}
@@
 
-\begin{frame}{Cosine distance and word connection visualization}
-\protect\hypertarget{cosine-distance-and-word-connection-visualization-1}{}
-\begin{block}{Analysis of word ``priest''}
-\protect\hypertarget{analysis-of-word-priest}{}
-\begin{center}\includegraphics[width=1.05\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-3-1} \end{center}
+\begin{frame}{The problem with pre-averaging}
+\protect\hypertarget{the-problem-with-pre-averaging-1}{}
+\begin{block}{One simulation}
+\protect\hypertarget{one-simulation}{}
+\vspace{1mm}
+\footnotesize
+
+\begin{center}\includegraphics[width=0.7\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-4-1} \end{center}
+\normalsize
+
+\vspace{1mm}
+\footnotesize
+
+\normalsize
+\pause
+
+\footnotesize
+
+\vspace{-4mm}
+
+\begin{itemize}
+\tightlist
+\item
+  Raw sd in data is 0.045
+\item
+  The sd of means is 0.023
+\item
+  The WEAT score is 1.825
+\item
+  The largest effect size reported by
+  \protect\hyperlink{ref-Caliskan2017semanticsBiases}{Caliskan, Bryson,
+  \& Narayanan}
+  (\protect\hyperlink{ref-Caliskan2017semanticsBiases}{2017}) is 1.81!
+\end{itemize}
 \end{block}
+\end{frame}
+
+\begin{frame}{The problem with pre-averaging}
+\protect\hypertarget{the-problem-with-pre-averaging-2}{}
+\begin{block}{50k simulations (same parameters)}
+\protect\hypertarget{k-simulations-same-parameters}{}
+\vspace{1mm}
+\footnotesize
+
+\begin{center}\includegraphics[width=0.8\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-6-1} \end{center}
+\normalsize
+\end{block}
+\end{frame}
+
+\begin{frame}{Cosine distance and connection type}
+\protect\hypertarget{cosine-distance-and-connection-type}{}
+\begin{block}{Distances for ``muslim''}
+\protect\hypertarget{distances-for-muslim}{}
+\begin{center}\includegraphics[width=0.8\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-7-1} \end{center}
+\end{block}
+\end{frame}
+
+\begin{frame}{Cosine distance and connection type}
+\protect\hypertarget{cosine-distance-and-connection-type-1}{}
+\begin{block}{Distances for ``priest''}
+\protect\hypertarget{distances-for-priest}{}
+\begin{center}\includegraphics[width=0.8\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-8-1} \end{center}
+\end{block}
+\end{frame}
+
+\begin{frame}{Advantages of the Bayesian way}
+\protect\hypertarget{advantages-of-the-bayesian-way}{}
+\begin{itemize}
+\tightlist
+\item
+  Direct impact of sample sizes
+\item
+  Straightforward interpretation in terms of posterior probabilities
+\item
+  Freedom to choose granularity level
+\item
+  More honest risk assessment and decision making
+\end{itemize}
 \end{frame}
@@
 \protect\hypertarget{bayesian-model}{}
+\begin{block}{Choosing priors}
+\protect\hypertarget{choosing-priors}{}
+\begin{center}\includegraphics[width=1\linewidth]{presentationESSLLI_files/figure-beamer/priorsVis-1} \end{center}
+\end{block}
 \end{frame}
 
-\begin{frame}{HPDIs}
-\protect\hypertarget{hpdis}{}
-\begin{block}{Further work}
+\begin{frame}[fragile]{Bayesian model architecture}
+\protect\hypertarget{bayesian-model-architecture}{}
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{library}\NormalTok{(rethinking)}
+\FunctionTok{options}\NormalTok{(}\AttributeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\ConstantTok{TRUE}\NormalTok{ )}
+\NormalTok{religionCoefs }\OtherTok{\textless{}{-}} \FunctionTok{ulam}\NormalTok{(}
+  \FunctionTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu }\OtherTok{\textless{}{-}}\NormalTok{ m }\SpecialCharTok{+}\NormalTok{ co[con],}
+\NormalTok{    m }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    co[con] }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\NormalTok{  ),}
+  \AttributeTok{data =}\NormalTok{ religion,}
+  \AttributeTok{chains=}\DecValTok{2}\NormalTok{ , }\AttributeTok{iter=}\DecValTok{8000}\NormalTok{ , }\AttributeTok{warmup=}\DecValTok{1000}\NormalTok{, }
+  \AttributeTok{log\_lik =} \ConstantTok{TRUE}
+\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+\end{frame}
+
+\begin{frame}{Dataset-level coefficients}
+\protect\hypertarget{dataset-level-coefficients}{}
+\begin{block}{Religion with 89\%-compatibility intervals (HPDI)}
+\protect\hypertarget{religion-with-89-compatibility-intervals-hpdi}{}
+\begin{center}
+\begin{figure}[!htb]\centering
+   \begin{minipage}{0.55\textwidth}
+  \includegraphics[width=6cm]{../images/religionCoeffs.jpeg}
+   \end{minipage} \begin{minipage}{0.4\textwidth}\footnotesize
+   \begin{itemize}
+   \item All HPDIs overlap with 0 
+   \item Differences between classes are relatively small
+   \item Coefficients for Race are similar
+   \end{itemize}
+   \end{minipage}
+  \end{figure}
+
+\end{center}
+\end{block}
+\end{frame}
+
+\begin{frame}{Dataset-level coefficients}
+\protect\hypertarget{dataset-level-coefficients-1}{}
+\begin{block}{Gender with 89\%-compatibility intervals (HPDI)}
+\protect\hypertarget{gender-with-89-compatibility-intervals-hpdi}{}
+\begin{center}
+\begin{figure}[!htb]\centering
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=6cm]{../images/genderCoeffs.jpeg}
+\end{minipage}
+\begin{minipage}{0.4\textwidth}\footnotesize
+   \begin{itemize}
+   \item Associated and different are away from 0
+   \item But they were supposed to be opposites and are very close to each other (co-occurrence?)
+   \item Differences between classes are still relatively small
+   \end{itemize}
+   \end{minipage}
+\end{figure}
+
+\end{center}
+\end{block}
+\end{frame}
+
+\begin{frame}[fragile]{Bayesian model architecture}
+\protect\hypertarget{bayesian-model-architecture-1}{}
+\vspace{1mm}
+\footnotesize
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{library}\NormalTok{(rethinking)}
+\FunctionTok{options}\NormalTok{(}\AttributeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\ConstantTok{TRUE}\NormalTok{ )}
+\NormalTok{religionCoefs }\OtherTok{\textless{}{-}} \FunctionTok{ulam}\NormalTok{(}
+  \FunctionTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu }\OtherTok{\textless{}{-}}\NormalTok{ m[pw] }\SpecialCharTok{+}\NormalTok{ co[con],}
+\NormalTok{    m[pw] }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    co[con] }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\NormalTok{  ),}
+  \AttributeTok{data =}\NormalTok{ religion,}
+  \AttributeTok{chains=}\DecValTok{2}\NormalTok{ , }\AttributeTok{iter=}\DecValTok{8000}\NormalTok{ , }\AttributeTok{warmup=}\DecValTok{1000}\NormalTok{, }
+  \AttributeTok{log\_lik =} \ConstantTok{TRUE}
+\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\normalsize
+\end{frame}
+
+\begin{frame}{Word-level coefficients}
+\protect\hypertarget{word-level-coefficients}{}
+\begin{figure}[!htb]\centering
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7.5cm]{../images/visReligionReddit.png}
+\end{minipage}
+\begin{minipage}{0.4\textwidth}\footnotesize
+
+\vspace{-4cm}
+
+   \begin{itemize}
+   \item Most intervals overlap with control groups
+   \item Often not too much diffence between associated and different
+   \end{itemize}
+   \end{minipage}
+\end{figure}
+\end{frame}
+
+\begin{frame}{Word-level coefficients}
+\protect\hypertarget{word-level-coefficients-1}{}
+\begin{figure}[!htb]\centering
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7.5cm]{../images/visGenderReddit.png}
+\end{minipage}
+\begin{minipage}{0.4\textwidth}\footnotesize
+
+\vspace{-4cm}
+
+   \begin{itemize}
+   \item Male attributes strong co-occurrence with female attributes
+   \item Sometimes different is stronger than associated
+   \item Almost no overlap with control groups
+   \end{itemize}
+   \end{minipage}
+\end{figure}
+\end{frame}
+
+\begin{frame}{Word-level coefficients}
+\protect\hypertarget{word-level-coefficients-2}{}
+\begin{figure}[!htb]\centering
+  \begin{minipage}{0.55\textwidth}
+\includegraphics[width=7.5cm]{../images/visRaceReddit.png}
+\end{minipage}
+\begin{minipage}{0.4\textwidth}\footnotesize
+
+\vspace{-4cm}
+
+   \begin{itemize}
+   \item A lot of variation between races
+
+   \item Often not much difference between associated and different
+   \end{itemize}
+   \end{minipage}
+\end{figure}
+\end{frame}
+
+\begin{frame}{Summary}
+\protect\hypertarget{summary}{}
+\begin{itemize}
+\tightlist
+\item
+  Introduced to the topic of bias in word embeddings
+\item
+  Analysed methodological problems with MAC implementation
+\item
+  Explained of pre-averaging limitations in methods such as WEAT and MAC
+\item
+  Proposed Bayesian approach
+\end{itemize}
+\end{frame}
+
+\begin{frame}{Further work}
 \protect\hypertarget{further-work}{}
@@
 \item
-  downstream tasks
+  Including contrasts in Bayesian calculation
+\item
+  Performance cross-validation in comparison to other methods (regular
+  linear regression, KNN, \dots)
+\item
+  Downstream tasks and connection with intrinsic evaluation
+\item
+  Testing data from the original Implicit Association Test (IAT)
+\item
+  Applying uncertainty to WEAT and better word lists
+\item
+  Looking at other similarity measures
 \end{itemize}
-\end{block}
 \end{frame}
@@
 \begin{CSLReferences}{1}{0}
-\leavevmode\hypertarget{ref-Bolukbasi2016Man}{}%
-Bolukbasi, T., Chang, K.-W., Zou, J. Y., Saligrama, V., \& Kalai, A.
-(2016). Man is to computer programmer as woman is to homemaker?
-Debiasing word embeddings. \emph{CoRR}, \emph{abs/1607.06520}. Retrieved
-from \url{http://arxiv.org/abs/1607.06520}
-
 \leavevmode\hypertarget{ref-Caliskan2017semanticsBiases}{}%
@@
 
+\leavevmode\hypertarget{ref-zhang2020robustness}{}%
+Zhang, H., Sneyd, A., \& Stevenson, M. (2020). \emph{Robustness and
+reliability of gender bias assessment in word embeddings: The role of
+base pairs}. Retrieved from \url{http://arxiv.org/abs/2010.02847}
+
 \end{CSLReferences}

### 2021-08-05 presentationESSLLI21/presentationESSLLI.Rmd
Adds up-to-date version

@@
 ---
-title: "\\Large Taking uncertainty seriously  \\newline \\normalsize A  Bayesian
-  approach to  word embedding bias estimation "
-author: "Alicja Dobrzeniecka \\& Rafal  Urbaniak \\footnotesize \\newline (LoPSE research group, University of  Gdansk)" 
+title: '\Large Taking uncertainty seriously  \newline \normalsize A  Bayesian approach
+  to  word embedding bias estimation '
+author: Alicja Dobrzeniecka \& Rafal  Urbaniak \footnotesize \newline (LoPSE research
+  group, University of  Gdansk)
 date: "ExpSem2021, ESSLLI"
-fontsize: 10pt
 output:
@@
     slide_level: 2
+  slidy_presentation: default
+fontsize: 10pt
 classoption: x11names, dvipsnames, bibspacing,natbib
 urlcolor: blue
-bibliography: [../references/cosineReferences1.bib]
+bibliography: ../references/cosineReferences1.bib
 csl: ../references/apa-6th-edition.csl
@@
 - Introduction to the topic of bias in word embeddings
-- Analysis of methodological problems with MAC implementation
-- Explanation of pre-averaging limitations in methods such as WEAT and MAC
-- Our research and Bayesian approach
+- WEAT and MAC methods
+- Methodological problems with MAC implementation
+- Limitations of pre-averaging in bias detection methods
+- Including uncertainty with Bayesian approach
 
@@
 
-<!-- <!-- (the gender direction is the top principal compontent of ten gender pair difference vectors).  --> -->
+<!-- <!-- (the gender direction is the top principal compontent of ten gender pair difference vectors).  -->
 
 <!-- -  Given the (ideally) gender neutral words $N$ and the gender direction $g$ the direct gender bias is: -->
-<!-- <!-- ($c$ is a parameter determining how strict we want to be): --> -->
+<!-- <!-- ($c$ is a parameter determining how strict we want to be): -->
 
@@
 
-\pause
+## Some methodological problems
+
+### Distances for  "muslim"
+
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "80%",message =FALSE, warning = FALSE}
+library(tidyverse)
+muslimWords <- c("imam","islam","mosque","muslim","quran")
+muslim <- religion %>% filter(protectedWord %in% muslimWords)
+muslimClass <- muslim %>% filter(protectedWord == "muslim")
+neutralSample <- sample_n(filter(muslimClass,connection == "none"), 5)
+humanSample <- sample_n(filter(muslimClass,connection == "human"), 5)
+muslimVis <- muslimClass %>% filter(connection != "none" & connection !="human")
+muslimVis <- rbind(muslimVis,neutralSample,humanSample)
+#we plug in our visualisation script
+source("../functions/visualisationTools.R")
+#two arguments: dataset and protected word
+visualiseProtected(muslimVis,"muslim")+theme_void(base_size = 20)
+```
+
+## Some methodological problems
+
+### Distances for  "priest"
+
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "80%"}
+library(tidyverse)
+priestClass <- religion %>% filter(protectedWord == "priest")
+neutralSample <- sample_n(filter(priestClass,connection == "none"), 5)
+humanSample <- sample_n(filter(priestClass,connection == "human"), 5)
+priestVis <- priestClass %>% filter(connection != "none" & connection !="human")
+priestVis <- rbind(priestVis,neutralSample,humanSample)
+
+#we plug in our visualisation script
+source("../functions/visualisationTools.R")
+#two arguments: dataset and protected word
+visualiseProtected(priestVis,"priest")+theme_void(base_size = 20)
+```
+
+## Some methodological problems
 
@@
 - It ignores variation in the raw data, which leads to false confidence
+
+\pause
 
@@
 \normalsize
-
-##   Cosine distance and  connection type
-
-### Distances for  "muslim"
-
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "80%",message =FALSE, warning = FALSE}
-library(tidyverse)
-muslimWords <- c("imam","islam","mosque","muslim","quran")
-muslim <- religion %>% filter(protectedWord %in% muslimWords)
-muslimClass <- muslim %>% filter(protectedWord == "muslim")
-neutralSample <- sample_n(filter(muslimClass,connection == "none"), 5)
-humanSample <- sample_n(filter(muslimClass,connection == "human"), 5)
-muslimVis <- muslimClass %>% filter(connection != "none" & connection !="human")
-muslimVis <- rbind(muslimVis,neutralSample,humanSample)
-#we plug in our visualisation script
-source("../functions/visualisationTools.R")
-#two arguments: dataset and protected word
-visualiseProtected(muslimVis,"muslim")+theme_void(base_size = 20)
-```
-
-## Cosine distance and  connection type
-
-### Distances for  "priest"
-
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "80%"}
-library(tidyverse)
-priestClass <- religion %>% filter(protectedWord == "priest")
-neutralSample <- sample_n(filter(priestClass,connection == "none"), 5)
-humanSample <- sample_n(filter(priestClass,connection == "human"), 5)
-priestVis <- priestClass %>% filter(connection != "none" & connection !="human")
-priestVis <- rbind(priestVis,neutralSample,humanSample)
-
-#we plug in our visualisation script
-source("../functions/visualisationTools.R")
-#two arguments: dataset and protected word
-visualiseProtected(priestVis,"priest")+theme_void(base_size = 20)
-```
 
@@
 
-- Introduced to the topic of bias in word embeddings
-- Analysed methodological problems with MAC implementation
-- Explained of pre-averaging limitations in methods such as WEAT and MAC
-- Proposed Bayesian approach
+- Introduction to the topic of bias in word embeddings
+- WEAT and MAC methods
+- Methodological problems with MAC implementation
+- Limitations of pre-averaging in bias detection methods
+- Including uncertainty with Bayesian approach
 

### 2021-08-05 presentationESSLLI21/presentationESSLLI.tex
Adds up-to-date version

@@
 \item
-  Analysis of methodological problems with MAC implementation
+  WEAT and MAC methods
 \item
-  Explanation of pre-averaging limitations in methods such as WEAT and
-  MAC
+  Methodological problems with MAC implementation
 \item
-  Our research and Bayesian approach
+  Limitations of pre-averaging in biad detection methods
+\item
+  Bayesian approach
 \end{itemize}
@@
 \normalsize
-
---\textgreater{}
-
---\textgreater{}
 \end{block}
@@
 We study those by visualizations and uncertainty estimates
+\end{block}
+\end{frame}
 
-\pause
+\begin{frame}{Some methodological problems}
+\protect\hypertarget{some-methodological-problems-3}{}
+\begin{block}{Distances for ``muslim''}
+\protect\hypertarget{distances-for-muslim}{}
+\begin{center}\includegraphics[width=0.8\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-4-1} \end{center}
 \end{block}
+\end{frame}
 
+\begin{frame}{Some methodological problems}
+\protect\hypertarget{some-methodological-problems-4}{}
+\begin{block}{Distances for ``priest''}
+\protect\hypertarget{distances-for-priest}{}
+\begin{center}\includegraphics[width=0.8\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-5-1} \end{center}
+\end{block}
+\end{frame}
+
+\begin{frame}{Some methodological problems}
+\protect\hypertarget{some-methodological-problems-5}{}
 \begin{block}{No principled interpretation}
@@
 \end{itemize}
+
+\pause
 \end{block}
@@
 
-\begin{center}\includegraphics[width=0.7\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-4-1} \end{center}
+\begin{center}\includegraphics[width=0.7\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-6-1} \end{center}
 \normalsize
@@
 
-\begin{center}\includegraphics[width=0.8\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-6-1} \end{center}
+\begin{center}\includegraphics[width=0.8\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-8-1} \end{center}
 \normalsize
-\end{block}
-\end{frame}
-
-\begin{frame}{Cosine distance and connection type}
-\protect\hypertarget{cosine-distance-and-connection-type}{}
-\begin{block}{Distances for ``muslim''}
-\protect\hypertarget{distances-for-muslim}{}
-\begin{center}\includegraphics[width=0.8\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-7-1} \end{center}
-\end{block}
-\end{frame}
-
-\begin{frame}{Cosine distance and connection type}
-\protect\hypertarget{cosine-distance-and-connection-type-1}{}
-\begin{block}{Distances for ``priest''}
-\protect\hypertarget{distances-for-priest}{}
-\begin{center}\includegraphics[width=0.8\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-8-1} \end{center}
 \end{block}

### 2021-08-06 presentationESSLLI21/presentationESSLLI.Rmd
Final presentation

@@
 
-## The plan of the presentation
+## Presentation plan
 
-- Introduction to the topic of bias in word embeddings
+- Bias in word embeddings
 - WEAT and MAC methods
-- Methodological problems with MAC implementation
+- Methodological problems
 - Limitations of pre-averaging in bias detection methods
-- Including uncertainty with Bayesian approach
+- Accounting for uncertainty with Bayesian approach
 
@@
 
-In the learning process word embeddings can learn implicit biases that reflect harmful stereotypical thinking
+Word embeddings can learn implicit harmful biases
 
@@
 
-Words belonging to an intuitively harmful stereotype are cosine-close to each other
+Stereotypically connected words are cosine-close
 
@@
 
+\vspace{2mm}
 [@Caliskan2017semanticsBiases] with extensions in [@Lauscher2019multidimensional] and applications in [@Garg2018years]
@@
                       linesep = "",  escape = FALSE, 
-                      caption = "Rows the religion dataset.") %>%
+                      caption = "A few rows from the religion dataset") %>%
                       kable_styling(latex_options=c("scale_down"))
@@
 
-- Gender-direction might be an indicator of bias, but is insufficient. After debiasing other non-gendered words can remain in biased relations  [@Gonen2019lipstick]
+- Gender-direction: insufficient indicator of bias [@Gonen2019lipstick]
 
-- Methods which involve analogies and their evaluations by human users on Mechanical Turk are unreliable [@Nissim2020fair]
+- Use of analogies: unreliable [@Nissim2020fair]
 
-- Sensitive to the choice of protected words, capitalization, distance measures and embedding methods [@zhang2020robustness]
-
-<!-- dodac? -->
+- High sensitivity to irrelevant factors [@zhang2020robustness]
 
@@
 
-## Summary
+## Thank you!
 
-- Introduction to the topic of bias in word embeddings
+### Summary
+
+- Bias in word embeddings
 - WEAT and MAC methods
-- Methodological problems with MAC implementation
+- Methodological problems
 - Limitations of pre-averaging in bias detection methods
-- Including uncertainty with Bayesian approach
+- Accounting for uncertainty with Bayesian approach
 
-## Further work
+### Further work
 

### 2021-08-06 presentationESSLLI21/presentationESSLLI.tex
Final presentation

@@
 
-\begin{frame}{The plan of the presentation}
-\protect\hypertarget{the-plan-of-the-presentation}{}
+\begin{frame}{Presentation plan}
+\protect\hypertarget{presentation-plan}{}
 \begin{itemize}
@@
 \item
-  Introduction to the topic of bias in word embeddings
+  Bias in word embeddings
 \item
@@
 \item
-  Methodological problems with MAC implementation
+  Methodological problems
 \item
-  Limitations of pre-averaging in biad detection methods
+  Limitations of pre-averaging in bias detection methods
 \item
-  Bayesian approach
+  Accounting for uncertainty with Bayesian approach
 \end{itemize}
@@
 \protect\hypertarget{the-worry}{}
-In the learning process word embeddings can learn implicit biases that
-reflect harmful stereotypical thinking
+Word embeddings can learn implicit harmful biases
 
@@
 \protect\hypertarget{cosine-based-bias-basic-intuition}{}
-Words belonging to an intuitively harmful stereotype are cosine-close to
-each other
+Stereotypically connected words are cosine-close
 \end{block}
@@
 
+\vspace{2mm}
+
 (\protect\hyperlink{ref-Caliskan2017semanticsBiases}{Caliskan, Bryson,
@@
 
-\caption{\label{tab:religionTableHeadEarly}Rows the religion dataset.}
+\caption{\label{tab:religionTableHeadEarly}A few rows from the religion dataset}
 \centering
@@
 \item
-  Gender-direction might be an indicator of bias, but is insufficient.
-  After debiasing other non-gendered words can remain in biased
-  relations (\protect\hyperlink{ref-Gonen2019lipstick}{Gonen \&
-  Goldberg, 2019})
+  Gender-direction: insufficient indicator of bias
+  (\protect\hyperlink{ref-Gonen2019lipstick}{Gonen \& Goldberg, 2019})
 \item
-  Methods which involve analogies and their evaluations by human users
-  on Mechanical Turk are unreliable
+  Use of analogies: unreliable
   (\protect\hyperlink{ref-Nissim2020fair}{Nissim, Noord, \& Goot, 2020})
 \item
-  Sensitive to the choice of protected words, capitalization, distance
-  measures and embedding methods
+  High sensitivity to irrelevant factors
   (\protect\hyperlink{ref-zhang2020robustness}{Zhang, Sneyd, \&
@@
 
-\begin{frame}{Summary}
+\begin{frame}{Thank you!}
+\protect\hypertarget{thank-you}{}
+\begin{block}{Summary}
 \protect\hypertarget{summary}{}
@@
 \item
-  Introduced to the topic of bias in word embeddings
+  Bias in word embeddings
 \item
-  Analysed methodological problems with MAC implementation
+  WEAT and MAC methods
 \item
-  Explained of pre-averaging limitations in methods such as WEAT and MAC
+  Methodological problems
 \item
-  Proposed Bayesian approach
+  Limitations of pre-averaging in bias detection methods
+\item
+  Accounting for uncertainty with Bayesian approach
 \end{itemize}
-\end{frame}
+\end{block}
 
-\begin{frame}{Further work}
+\begin{block}{Further work}
 \protect\hypertarget{further-work}{}
@@
 \end{itemize}
+\end{block}
 \end{frame}

### 2021-08-08 presentationESSLLI21/presentationESSLLI.Rmd
final details

@@
 
+\vspace{2mm}
+
 \footnotesize
@@
 
-- Gender-direction: insufficient indicator of bias [@Gonen2019lipstick]
+- Gender-direction: insufficient indicator of
+    bias  \footnotesize  [@Gonen2019lipstick]
 
-- Use of analogies: unreliable [@Nissim2020fair]
+\normalsize
 
-- High sensitivity to irrelevant factors [@zhang2020robustness]
+- Use of analogies: unreliable \footnotesize  [@Nissim2020fair] \normalsize
+
+- High sensitivity to irrelevant factors \footnotesize  [@zhang2020robustness] \normalsize
 
@@
 
-What values are sufficient for the presence of bias and what differences are sign of real improvement? Low $p$-values are not high effect indicators!
+- What values are sufficient for the presence of bias and what differences are sign of real improvement?
 
-We compare HPDIs.
+- Low $p$-values are not high effect indicators!
+
+- We compare HPDIs.
 
@@
 
-Suppose all similarities for two classes are randomly drawn from the same distribution, $\mathsf{Normal}(0,.05)$, you still can get really high WEAT!
+Suppose all similarities for two classes are randomly drawn from the same distribution, $\mathsf{Normal}(\mu = 0, \sigma = .05)$, you still can get a really high WEAT!
 
@@
 
-\vspace{-4mm}
+\vspace{-2mm}
 
@@
    \begin{itemize}
-   \item Male attributes strong co-occurrence with female attributes
-   \item Sometimes different is stronger than associated
+   \item Male attributes: strong co-occurrence with female attributes
+   \item Sometimes \textsf{different} is closer than \textsf{associated}
    \item Almost no overlap with control groups

### 2021-08-08 presentationESSLLI21/presentationESSLLI.tex
final details

@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
 \PassOptionsToPackage{hyphens}{url}
@@
 %
-\documentclass[
-  10pt,
-  ignorenonframetext,
-  x11names, dvipsnames, bibspacing,natbib]{beamer}
+\documentclass[10pt,ignorenonframetext,x11names, dvipsnames, bibspacing,natbib]{beamer}
 \usepackage{pgfpages}
@@
 \beamertemplatenavigationsymbolsempty
-% Prevent slide breaks in the middle of a paragraph
+% Prevent slide breaks in the middle of a paragraph:
 \widowpenalties 1 10000
@@
 }
-\usepackage{amsmath,amssymb}
 \usepackage{lmodern}
+\usepackage{amssymb,amsmath}
 \usepackage{ifxetex,ifluatex}
+\usepackage{fixltx2e} % provides \textsubscript
 \ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
@@
   \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
+  \usepackage{textcomp} % provides euro and other symbols
+\else % if luatex or xelatex
   \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
 \fi
 \usetheme[]{Rafal_beamerSly1}
-% Use upquote if available, for straight quotes in verbatim environments
+% use upquote if available, for straight quotes in verbatim environments
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
+% use microtype if available
+\IfFileExists{microtype.sty}{%
 \usepackage[]{microtype}
@@
 }{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
 \IfFileExists{parskip.sty}{%
@@
 \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
+\setlength{\parskip}{6pt plus 2pt minus 1pt}
+}
 \usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\usepackage{hyperref}
 \hypersetup{
@@
             urlcolor=blue,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
+            breaklinks=true}
+\urlstyle{same}  % don't use monospace font for urls
 \newif\ifbibliography
@@
 \newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\usepackage{longtable,booktabs,array}
-\usepackage{calc} % for calculating minipage widths
+\usepackage{longtable,booktabs}
 \usepackage{caption}
-% Make caption package work with longtable
+% These lines are needed to make table captions work with longtable:
 \makeatletter
@@
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-\ifluatex
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
-  % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+\setcounter{secnumdepth}{0}
 
-\title{\Large Taking uncertainty seriously \newline \normalsize A
-Bayesian approach to word embedding bias estimation}
-\author{Alicja Dobrzeniecka \& Rafal Urbaniak
-\footnotesize \newline (LoPSE research group, University of Gdansk)}
+% set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
+
+\title{\Large Taking uncertainty seriously \newline \normalsize A Bayesian
+approach to word embedding bias estimation}
+\author{Alicja Dobrzeniecka \& Rafal Urbaniak \footnotesize \newline (LoPSE
+research group, University of Gdansk)}
 \date{ExpSem2021, ESSLLI}
@@
 \protect\hypertarget{presentation-plan}{}
+
 \begin{itemize}
@@
 \end{itemize}
+
 \end{frame}
@@
 \protect\hypertarget{cosine-based-measures-of-bias}{}
+
 \begin{block}{Word embeddings}
-\protect\hypertarget{word-embeddings}{}
+
 \begin{itemize}
@@
 \toprule
-word & 1 & 2 & 3 & 4 & \ldots{} \\
+word & 1 & 2 & 3 & 4 & \ldots{}\tabularnewline
 \midrule
 \endhead
-woman & 0.456 & 0.267 & 0.675 & 0.131 & \ldots{} \\
-man & 0.451 & 0.897 & 0.472 & 0.088 & \ldots{} \\
+woman & 0.456 & 0.267 & 0.675 & 0.131 & \ldots{}\tabularnewline
+man & 0.451 & 0.897 & 0.472 & 0.088 & \ldots{}\tabularnewline
 \bottomrule
 \end{longtable}
+
 \end{block}
+
 \end{frame}
@@
 \protect\hypertarget{cosine-based-measures-of-bias-1}{}
+
 \begin{block}{Cosine similarity \& distance}
-\protect\hypertarget{cosine-similarity-distance}{}
+
 \vspace{-4mm}
@@
 \end{itemize}
+
 \end{block}
+
 \end{frame}
@@
 \protect\hypertarget{cosine-based-measures-of-bias-2}{}
+
 \begin{block}{The worry}
-\protect\hypertarget{the-worry}{}
+
 Word embeddings can learn implicit harmful biases
@@
 \pause
+
 \end{block}
@@
 \begin{block}{Cosine-based bias: basic intuition}
-\protect\hypertarget{cosine-based-bias-basic-intuition}{}
+
 Stereotypically connected words are cosine-close
+
 \end{block}
+
 \end{frame}
@@
 \protect\hypertarget{cosine-based-measures-of-bias-3}{}
+
 \begin{block}{A visual example}
-\protect\hypertarget{a-visual-example}{}
+
 \footnotesize
@@
 \item
-  feminine occupations: ``homemaker,'' ``nurse,'' ``receptionist,''
-  ``librarian,'' etc.
+  feminine occupations: ``homemaker'', ``nurse'', ``receptionist'',
+  ``librarian'', etc.
 \item
-  masculine occupations: ``maestro,'' ``captain,'' ``architect,''
-  ``boss,'' etc.
+  masculine occupations: ``maestro'', ``captain'', ``architect'',
+  ``boss'', etc.
 \end{itemize}
@@
 \normalsize
+
 \end{block}
+
 \end{frame}
@@
 \protect\hypertarget{cosine-based-measures-of-bias-4}{}
+
 \begin{block}{Example: Word Embedding Association Test (WEAT)}
-\protect\hypertarget{example-word-embedding-association-test-weat}{}
+
 \vspace{1mm}
@@
 \end{align*}
+
 \end{block}
+
 \end{frame}
@@
 \protect\hypertarget{cosine-based-measures-of-bias-5}{}
+
 \begin{block}{Example: Word Embedding Association Test (WEAT)}
-\protect\hypertarget{example-word-embedding-association-test-weat-1}{}
+
 \begin{align*}
@@
 
-(\protect\hyperlink{ref-Caliskan2017semanticsBiases}{Caliskan, Bryson,
-\& Narayanan, 2017}) with extensions in
-(\protect\hyperlink{ref-Lauscher2019multidimensional}{Lauscher \&
-Glavas, 2019}) and applications in
-(\protect\hyperlink{ref-Garg2018years}{Garg, Schiebinger, Jurafsky, \&
-Zou, 2018})
+(Caliskan, Bryson, \& Narayanan,
+\protect\hyperlink{ref-Caliskan2017semanticsBiases}{2017}) with
+extensions in (Lauscher \& Glavas,
+\protect\hyperlink{ref-Lauscher2019multidimensional}{2019}) and
+applications in (Garg, Schiebinger, Jurafsky, \& Zou,
+\protect\hyperlink{ref-Garg2018years}{2018})
+
 \end{block}
+
 \end{frame}
@@
 \protect\hypertarget{cosine-based-measures-of-bias-6}{}
+
 \begin{block}{Our main target: Mean Average Cosine Similarity (MAC)}
-\protect\hypertarget{our-main-target-mean-average-cosine-similarity-mac}{}
+
 \vspace{1mm}
@@
 \protect\hypertarget{cosine-based-measures-of-bias-7}{}
+
 \begin{block}{Our main target: Mean Average Cosine Similarity (MAC)}
-\protect\hypertarget{our-main-target-mean-average-cosine-similarity-mac-1}{}
+
 \begin{align*}
@@
 
+\vspace{2mm}
+
 \footnotesize
 
-(\protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim,
-Tsvetkov, \& Black, 2019})
+(Manzini, Lim, Tsvetkov, \& Black,
+\protect\hyperlink{ref-Manzini2019blackToCriminal}{2019})
+
 \end{block}
+
 \end{frame}
@@
 \protect\hypertarget{cosine-based-measures-of-bias-8}{}
+
 \begin{block}{Our main target: Mean Average Cosine Similarity (MAC)}
-\protect\hypertarget{our-main-target-mean-average-cosine-similarity-mac-2}{}
+
 \footnotesize 
@@
 \protect\hypertarget{cosine-based-measures-of-bias-9}{}
+
 \begin{block}{Known challenges}
-\protect\hypertarget{known-challenges}{}
+
+\begin{itemize}
+\tightlist
+\item
+  Gender-direction: insufficient indicator of bias \footnotesize  (Gonen
+  \& Goldberg, \protect\hyperlink{ref-Gonen2019lipstick}{2019})
+\end{itemize}
+
+\normalsize
+
 \begin{itemize}
 \item
-  Gender-direction: insufficient indicator of bias
-  (\protect\hyperlink{ref-Gonen2019lipstick}{Gonen \& Goldberg, 2019})
+  Use of analogies: unreliable \footnotesize  (Nissim, Noord, \& Goot,
+  \protect\hyperlink{ref-Nissim2020fair}{2020}) \normalsize
 \item
-  Use of analogies: unreliable
-  (\protect\hyperlink{ref-Nissim2020fair}{Nissim, Noord, \& Goot, 2020})
-\item
-  High sensitivity to irrelevant factors
-  (\protect\hyperlink{ref-zhang2020robustness}{Zhang, Sneyd, \&
-  Stevenson, 2020})
+  High sensitivity to irrelevant factors \footnotesize  (Zhang, Sneyd,
+  \& Stevenson, \protect\hyperlink{ref-zhang2020robustness}{2020})
+  \normalsize
 \end{itemize}
+
 \end{block}
+
 \end{frame}
@@
 \protect\hypertarget{some-methodological-problems}{}
+
 \begin{block}{Word list choice is unprincipled}
-\protect\hypertarget{word-list-choice-is-unprincipled}{}
+
 We run with it for comparison
@@
 \begin{block}{No design considerations to sample size}
-\protect\hypertarget{no-design-considerations-to-sample-size}{}
+
 We statistically gauge the uncertainty that arises from raw sample sizes
+
 \end{block}
+
 \end{frame}
@@
 \protect\hypertarget{some-methodological-problems-1}{}
+
 \begin{block}{No word class distinction and no control group}
-\protect\hypertarget{no-word-class-distinction-and-no-control-group}{}
+
 We make the subclasses clear, add human neutral predicates and neutral
@@
 \protect\hypertarget{some-methodological-problems-2}{}
+
 \begin{block}{Outliers and surprisingly dissimilar words}
-\protect\hypertarget{outliers-and-surprisingly-dissimilar-words}{}
+
 We study those by visualizations and uncertainty estimates
+
 \end{block}
+
 \end{frame}
@@
 \protect\hypertarget{some-methodological-problems-3}{}
+
 \begin{block}{Distances for ``muslim''}
-\protect\hypertarget{distances-for-muslim}{}
+
 \begin{center}\includegraphics[width=0.8\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-4-1} \end{center}
+
 \end{block}
+
 \end{frame}
@@
 \protect\hypertarget{some-methodological-problems-4}{}
+
 \begin{block}{Distances for ``priest''}
-\protect\hypertarget{distances-for-priest}{}
+
 \begin{center}\includegraphics[width=0.8\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-5-1} \end{center}
+
 \end{block}
+
 \end{frame}
@@
 \protect\hypertarget{some-methodological-problems-5}{}
+
 \begin{block}{No principled interpretation}
-\protect\hypertarget{no-principled-interpretation}{}
+
 \begin{longtable}[]{@{}ll@{}}
 \toprule
-Religion Debiasing & MAC (distance) \\
+Religion Debiasing & MAC (distance)\tabularnewline
 \midrule
 \endhead
-Biased & 0.859 \\
-Hard Debiased & 0.934 \\
-Soft Debiased (\(\lambda\) = 0.2) & 0.894 \\
+Biased & 0.859\tabularnewline
+Hard Debiased & 0.934\tabularnewline
+Soft Debiased (\(\lambda\) = 0.2) & 0.894\tabularnewline
 \bottomrule
@@
 
-What values are sufficient for the presence of bias and what differences
-are sign of real improvement? Low \(p\)-values are not high effect
-indicators!
+\begin{itemize}
+\item
+  What values are sufficient for the presence of bias and what
+  differences are sign of real improvement?
+\item
+  Low \(p\)-values are not high effect indicators!
+\item
+  We compare HPDIs.
+\end{itemize}
 
-We compare HPDIs.
 \end{block}
+
 \end{frame}
@@
 \protect\hypertarget{the-problem-with-pre-averaging}{}
+
 \begin{block}{Key conceptual issues}
-\protect\hypertarget{key-conceptual-issues}{}
+
 \begin{itemize}
@@
 \begin{block}{Our simulation}
-\protect\hypertarget{our-simulation}{}
+
 Suppose all similarities for two classes are randomly drawn from the
-same distribution, \(\mathsf{Normal}(0,.05)\), you still can get really
-high WEAT!
+same distribution, \(\mathsf{Normal}(\mu = 0, \sigma = .05)\), you still
+can get a really high WEAT!
+
 \end{block}
+
 \end{frame}
@@
 \protect\hypertarget{the-problem-with-pre-averaging-1}{}
+
 \begin{block}{One simulation}
-\protect\hypertarget{one-simulation}{}
+
 \vspace{1mm}
@@
 \item
-  The largest effect size reported by
-  \protect\hyperlink{ref-Caliskan2017semanticsBiases}{Caliskan, Bryson,
-  \& Narayanan}
+  The largest effect size reported by Caliskan et al.
   (\protect\hyperlink{ref-Caliskan2017semanticsBiases}{2017}) is 1.81!
 \end{itemize}
+
 \end{block}
+
 \end{frame}
@@
 \protect\hypertarget{the-problem-with-pre-averaging-2}{}
+
 \begin{block}{50k simulations (same parameters)}
-\protect\hypertarget{k-simulations-same-parameters}{}
+
 \vspace{1mm}
@@
 \protect\hypertarget{advantages-of-the-bayesian-way}{}
+
 \begin{itemize}
@@
 \protect\hypertarget{bayesian-model}{}
+
 \begin{block}{Choosing priors}
-\protect\hypertarget{choosing-priors}{}
+
 \begin{center}\includegraphics[width=1\linewidth]{presentationESSLLI_files/figure-beamer/priorsVis-1} \end{center}
+
 \end{block}
+
 \end{frame}
@@
 \protect\hypertarget{bayesian-model-architecture}{}
+
 \vspace{1mm}
@@
 \begin{Highlighting}[]
-\FunctionTok{library}\NormalTok{(rethinking)}
-\FunctionTok{options}\NormalTok{(}\AttributeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\ConstantTok{TRUE}\NormalTok{ )}
-\NormalTok{religionCoefs }\OtherTok{\textless{}{-}} \FunctionTok{ulam}\NormalTok{(}
-  \FunctionTok{alist}\NormalTok{(}
-\NormalTok{    cosineDistance }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(mu,sigma),}
-\NormalTok{    mu }\OtherTok{\textless{}{-}}\NormalTok{ m }\SpecialCharTok{+}\NormalTok{ co[con],}
-\NormalTok{    m }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
-\NormalTok{    co[con] }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
-\NormalTok{    sigma }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\KeywordTok{library}\NormalTok{(rethinking)}
+\KeywordTok{options}\NormalTok{(}\DataTypeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\OtherTok{TRUE}\NormalTok{ )}
+\NormalTok{religionCoefs <-}\StringTok{ }\KeywordTok{ulam}\NormalTok{(}
+  \KeywordTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu <-}\StringTok{ }\NormalTok{m }\OperatorTok{+}\StringTok{ }\NormalTok{co[con],}
+\NormalTok{    m }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    co[con] }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\OperatorTok{~}\StringTok{ }\KeywordTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
 \NormalTok{  ),}
-  \AttributeTok{data =}\NormalTok{ religion,}
-  \AttributeTok{chains=}\DecValTok{2}\NormalTok{ , }\AttributeTok{iter=}\DecValTok{8000}\NormalTok{ , }\AttributeTok{warmup=}\DecValTok{1000}\NormalTok{, }
-  \AttributeTok{log\_lik =} \ConstantTok{TRUE}
+  \DataTypeTok{data =}\NormalTok{ religion,}
+  \DataTypeTok{chains=}\DecValTok{2}\NormalTok{ , }\DataTypeTok{iter=}\DecValTok{8000}\NormalTok{ , }\DataTypeTok{warmup=}\DecValTok{1000}\NormalTok{, }
+  \DataTypeTok{log_lik =} \OtherTok{TRUE}
 \NormalTok{)}
@@
 \normalsize
+
 \end{frame}
@@
 \protect\hypertarget{dataset-level-coefficients}{}
+
 \begin{block}{Religion with 89\%-compatibility intervals (HPDI)}
-\protect\hypertarget{religion-with-89-compatibility-intervals-hpdi}{}
+
 \begin{center}
@@
 \end{center}
+
 \end{block}
+
 \end{frame}
@@
 \protect\hypertarget{dataset-level-coefficients-1}{}
+
 \begin{block}{Gender with 89\%-compatibility intervals (HPDI)}
-\protect\hypertarget{gender-with-89-compatibility-intervals-hpdi}{}
+
 \begin{center}
@@
 \protect\hypertarget{bayesian-model-architecture-1}{}
+
 \vspace{1mm}
@@
 \begin{Highlighting}[]
-\FunctionTok{library}\NormalTok{(rethinking)}
-\FunctionTok{options}\NormalTok{(}\AttributeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\ConstantTok{TRUE}\NormalTok{ )}
-\NormalTok{religionCoefs }\OtherTok{\textless{}{-}} \FunctionTok{ulam}\NormalTok{(}
-  \FunctionTok{alist}\NormalTok{(}
-\NormalTok{    cosineDistance }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(mu,sigma),}
-\NormalTok{    mu }\OtherTok{\textless{}{-}}\NormalTok{ m[pw] }\SpecialCharTok{+}\NormalTok{ co[con],}
-\NormalTok{    m[pw] }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
-\NormalTok{    co[con] }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
-\NormalTok{    sigma }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
+\KeywordTok{library}\NormalTok{(rethinking)}
+\KeywordTok{options}\NormalTok{(}\DataTypeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\OtherTok{TRUE}\NormalTok{ )}
+\NormalTok{religionCoefs <-}\StringTok{ }\KeywordTok{ulam}\NormalTok{(}
+  \KeywordTok{alist}\NormalTok{(}
+\NormalTok{    cosineDistance }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{    mu <-}\StringTok{ }\NormalTok{m[pw] }\OperatorTok{+}\StringTok{ }\NormalTok{co[con],}
+\NormalTok{    m[pw] }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    co[con] }\OperatorTok{~}\StringTok{ }\KeywordTok{dnorm}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
+\NormalTok{    sigma }\OperatorTok{~}\StringTok{ }\KeywordTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
 \NormalTok{  ),}
-  \AttributeTok{data =}\NormalTok{ religion,}
-  \AttributeTok{chains=}\DecValTok{2}\NormalTok{ , }\AttributeTok{iter=}\DecValTok{8000}\NormalTok{ , }\AttributeTok{warmup=}\DecValTok{1000}\NormalTok{, }
-  \AttributeTok{log\_lik =} \ConstantTok{TRUE}
+  \DataTypeTok{data =}\NormalTok{ religion,}
+  \DataTypeTok{chains=}\DecValTok{2}\NormalTok{ , }\DataTypeTok{iter=}\DecValTok{8000}\NormalTok{ , }\DataTypeTok{warmup=}\DecValTok{1000}\NormalTok{, }
+  \DataTypeTok{log_lik =} \OtherTok{TRUE}
 \NormalTok{)}
@@
 \protect\hypertarget{word-level-coefficients}{}
+
 \begin{figure}[!htb]\centering
@@
 \end{figure}
+
 \end{frame}
@@
 \protect\hypertarget{word-level-coefficients-1}{}
+
 \begin{figure}[!htb]\centering
@@
 \protect\hypertarget{word-level-coefficients-2}{}
+
 \begin{figure}[!htb]\centering
@@
 \protect\hypertarget{thank-you}{}
+
 \begin{block}{Summary}
-\protect\hypertarget{summary}{}
+
 \begin{itemize}
@@
 \end{itemize}
+
 \end{block}
@@
 \begin{block}{Further work}
-\protect\hypertarget{further-work}{}
+
 \begin{itemize}
@@
 \protect\hypertarget{references}{}
+
 \tiny
@@
 \hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
 \leavevmode\hypertarget{ref-Caliskan2017semanticsBiases}{}%
@@
 \leavevmode\hypertarget{ref-Gonen2019lipstick}{}%
-Gonen, H., \& Goldberg, Y. (2019). Lipstick on a pig: {D}ebiasing
-methods cover up systematic gender biases in word embeddings but do not
-remove them. \emph{Proceedings of the 2019 Conference of the North
-{A}merican Chapter of the Association for Computational Linguistics:
-Human Language Technologies, Volume 1 (Long and Short Papers)},
-609--614. Minneapolis, Minnesota: Association for Computational
-Linguistics. \url{https://doi.org/10.18653/v1/N19-1061}
+Gonen, H., \& Goldberg, Y. (2019). Lipstick on a pig: Debiasing methods
+cover up systematic gender biases in word embeddings but do not remove
+them. \emph{Proceedings of the 2019 Conference of the North American
+Chapter of the Association for Computational Linguistics: Human Language
+Technologies, Volume 1 (Long and Short Papers)}, 609--614. Minneapolis,
+Minnesota: Association for Computational Linguistics.
+\url{https://doi.org/10.18653/v1/N19-1061}
 
@@
 
-\end{CSLReferences}
 \end{frame}

### 2021-08-09 presentationESSLLI21/presentationESSLLI.tex
Adds all

--- 2021-08-06 presentationESSLLI21/presentationESSLLI.tex
@@
 
-\begin{block}{Cosine-based bias: basic intuition}
-\protect\hypertarget{cosine-based-bias-basic-intuition}{}
+\begin{block}{The basic intuition}
+\protect\hypertarget{the-basic-intuition}{}
 Stereotypically connected words are cosine-close
@@
 \begin{itemize}
+\tightlist
 \item
   Gender-direction: insufficient indicator of bias
-  (\protect\hyperlink{ref-Gonen2019lipstick}{Gonen \& Goldberg, 2019})
+  \footnotesize  (\protect\hyperlink{ref-Gonen2019lipstick}{Gonen \&
+  Goldberg, 2019})
+\end{itemize}
+
+\normalsize
+
+\begin{itemize}
 \item
   Use of analogies: unreliable
-  (\protect\hyperlink{ref-Nissim2020fair}{Nissim, Noord, \& Goot, 2020})
+  \footnotesize  (\protect\hyperlink{ref-Nissim2020fair}{Nissim, Noord,
+  \& Goot, 2020}) \normalsize
 \item
   High sensitivity to irrelevant factors
-  (\protect\hyperlink{ref-zhang2020robustness}{Zhang, Sneyd, \&
-  Stevenson, 2020})
+  \footnotesize  (\protect\hyperlink{ref-zhang2020robustness}{Zhang,
+  Sneyd, \& Stevenson, 2020}) \normalsize
 \end{itemize}
@@
 
-What values are sufficient for the presence of bias and what differences
-are sign of real improvement? Low \(p\)-values are not high effect
-indicators!
-
+\begin{itemize}
+\item
+  What values are sufficient for the presence of bias and what
+  differences are sign of real improvement?
+\item
+  Low \(p\)-values are not high effect indicators!
+\item
   We compare HPDIs.
+\end{itemize}
 \end{block}
@@
 Suppose all similarities for two classes are randomly drawn from the
-same distribution, \(\mathsf{Normal}(0,.05)\), you still can get really
-high WEAT!
+same distribution, \(\mathsf{Normal}(\mu = 0, \sigma = .05)\), you still
+can get a really high WEAT!
 \end{block}

### 2021-08-10 presentationESSLLI21/presentationESSLLI.Rmd
Adds all
Updates
Adds current

@@
 
-### Cosine-based bias: basic intuition
+### The basic intuition
 
@@
 
-- feminine occupations: "homemaker", "nurse", "receptionist", "librarian", etc.
+- "feminine" occupations: "homemaker", "nurse", "receptionist", "librarian", etc.
 
-- masculine occupations:  "maestro",  "captain", "architect", "boss", etc.
+- "masculine" occupations:  "maestro",  "captain", "architect", "boss", etc.
 
@@
    \item Most intervals overlap with control groups
-   \item Often not too much diffence between associated and different
+   \item Often not too much difference between associated and different
    \end{itemize}

### 2021-12-10 paperDraft/paperDraft2.Rmd
setting up paper draft 2

--- 2021-07-30 paperDraft/paperDraft.Rmd
@@
     includes:
-      in_header: Rafal_latex6.sty
+      in_header: Rafal_latex7.sty
   html_document:
@@
 urlcolor: blue
-bibliography: ../references/bibliographyRafal.bib
+bibliography: ../references/cosineReferences1.bib
 csl: ../references/apa-6th-edition.csl
@@
 
-\tableofcontents
+# Cosine-based measures of bias
 
-# Cosine-based measures of bias
+\raf{fix bibliography}
+
+\ali{proofread and think if we should add more}
 
@@
 A similar methodology is employed by @Garg2018years, who employ word embeddings trained on corpora from different decades to study the shifts in various biases.
-For instance, to compute the occupational embeddings bias for women the authors first compute the average vector of vector emeddings of words that represent women (e.g.
+For instance, to compute the occupational embeddings bias for women the authors first compute the average vector of vector embeddings of words that represent women (e.g.
 "she", "female"), then calculate the Euclidean distance between this mean vector and words for occupations.
@@
 
+\todo{odleglosc nie musi uchwytywac relacji semantycznej zeby oceniac bias}
+
+\todo{zalozenie, bardziej ze jezeli odleglosci przekladaja sie na downstream, to warto patrzec na odleglosci, nawet bez glebszej filozoficznej interpretacji ich}
+
 # Some methodological problems
@@
 Such approaches, however, are not unproblematic.
-One may cosinder the selection of attributes for the bias research.
+One may consider the selection of attributes for the bias research.
 For one thing, the word lists are often fairly small, and the papers do not discuss the impact of word list sizes on the uncertainty involved.
@@
 
-# The problem with means of means
+\todo{MAC: avoid looking at only one vector}
 
-<!-- A <- c(1,5,10) -->
-<!-- B <- c(6, 8, 15) -->
+\todo{effect sizes}
 
-<!-- c(A,B) -->
-
-<!-- sd(c(A,B)) -->
-
-<!-- mA <- mean(A) -->
-<!-- mB <- mean(B) -->
-
-<!-- sd(c(mA,mB)) -->
-
-The measures described all calculate means of means and their authors run statistical tests on sets of means.
-This, however, is problematic for two related reasons.
-One, by pre-averaging data we throw away information about sample sizes.
-For the former point, think about proportions: 10 out of 20 and 2 out of 4 give the same mean, but you would obtain more information by making the former observation rather than by making the latter.
-Two, when we pre-average, we remove variation, and so pre-averaging  tend to manufacture false confidence.
-We will have more to say about whether this happens in the case of applications of MAC, for now let's go over a simple example to make the conceptual point clear.
-
-To illustrate let's employ the formulas used by @Caliskan2017semanticsBiases in a simple example.
-Conceptually, all such tests come up with rather short lists of protected words and rather short lists of stereotypical attributes.
-Clearly, these are not complete list.
-So let's treat them as  samples from richer pools of stereotypical predicates and let's take the uncertainty and variation involved seriously.
-
-Consider a simple situation in which there are two protected classes, $X=\{t_1,t_2\}$ and $Y=\{t_3,t_4\}$ and two five-element attribute sets $A$ and $B$.
-
-First, we play around with a scenario in which all the protected terms are on average equally unsimilar to both sets ($\mu =0$) with standard deviation of $.05$.
-Let's randomly draw similarity scores and plot the results with group means plotted as vertical lines.
-
-\footnotesize
-```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
-set.seed(123)
-t1 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
-t2 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
-t3 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
-t4 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
-```
-\normalsize
-
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
-xlimits <- c(-.3,.3)
-
-t1Plot <- ggplot(t1)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t1 in X")+xlim(xlimits)
-
-t2Plot <- ggplot(t2)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t2 in X")+xlim(xlimits)
-
-t3Plot <- ggplot(t3)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t3 in Y")+xlim(xlimits)
-
-t4Plot <- ggplot(t4)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t4 in Y")+xlim(xlimits)
-
-grid.arrange(t1Plot,t2Plot, t3Plot, t4Plot, ncol=2)
-```
-
-\noindent When you look at the datapoints, do you have the impression of a strong bias here? We wouldn't think so.  But now let's run the calculations from [@Caliskan2017semanticsBiases].
-
-\vspace{1mm}
-\footnotesize
-```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
-s <- function (table){ mean(table$A) - mean(table$B)}
-factor <- sd(c(s(t1),s(t2),s(t3),s(t4)))
-numerator <-  mean(s(t1),s(t2)) - mean(s(t3),s(t4))
-print(list(factor = factor,numerator = numerator, bias = numerator / factor))
-```
-\normalsize
-
-\noindent This should make us pause.
-We know these were points randomly drawn from distributions where there is no difference in means.
-Yet, the calculated effect size is 1.82, whereas the largest effect size reported by @Caliskan2017semanticsBiases is 1.81!
-
-Interestingly, if we repeat the drawing 10000 times, each time calculating the  bias, it turns out that with this variance and sample size, pretty much anything can happen.
-
-\vspace{1mm}
-\footnotesize
-```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
-
-biasesNull <- numeric(10000)
-for(i in 1:10000){
-t1 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
-t2 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
-t3 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
-t4 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
-
-factor <- sd(c(s(t1),s(t2),s(t3),s(t4)))
-numerator <-  mean(s(t1),s(t2)) - mean(s(t3),s(t4))
-biasesNull[i]  <- numerator / factor
-}
-ggplot()+geom_histogram(aes(x=biasesNull, y = ..density..), alpha = 0.6, bins=50)+
-  theme_tufte()+labs(title="10k biases for identical means and sd =.05")+ xlab("bias")
-```
-\normalsize
-
-Now, let's simulate a situation where the means are identical but the standard deviation is much smaller, .001.
-
-\footnotesize
-```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
-set.seed(124)
-t1v <- data.frame(A  = rnorm(5,0,0.001), B = rnorm(5,0,0.001))
-t2v <- data.frame(A  = rnorm(5,0,0.001), B = rnorm(5,0,0.001))
-t3v <- data.frame(A  = rnorm(5,0,0.001), B = rnorm(5,0,0.001))
-t4v <- data.frame(A  = rnorm(5,0,0.001), B = rnorm(5,0,0.001))
-```
-\normalsize
-
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
-xlimits <- c(-.3,.3)
-
-t1vPlot <- ggplot(t1v)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t1 in X")+xlim(xlimits)
-
-t2vPlot <- ggplot(t2v)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t2 in X")+xlim(xlimits)
-
-t3vPlot <- ggplot(t3v)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t3 in Y")+xlim(xlimits)
-
-t4vPlot <- ggplot(t4v)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t4 in Y")+xlim(xlimits)
-
-grid.arrange(t1vPlot,t2vPlot, t3vPlot, t4vPlot, ncol=2)
-```
-
-\noindent When you look at the datapoints, do you have the impression of a strong bias here? We wouldn't think so.  But now let's run the calculations from [@Caliskan2017semanticsBiases].
-
-\vspace{1mm}
-\footnotesize
-```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
-factorV <- sd(c(s(t1v),s(t2v),s(t3v),s(t4v)))
-numeratorV <-  mean(s(t1v),s(t2v)) - mean(s(t3v),s(t4v))
-print(list(factor = factorV,numerator = numeratorV, bias = numeratorV / factorV))
-```
-\normalsize
-
-While the numerator and the factors changed a lot, the bias actually increased.
-One reason bias increases is that once the standard deviation goes down, so does the factor used in the calculation of bias.
-
-Again, to see whether this metric would provide us with meaningful information, let's simulate 10000 drawings.
-
-\vspace{1mm}
-\footnotesize
-```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
-set.seed(124)
-
-biasesLowVariance <- numeric(10000)
-for(i in 1:10000){
-t1v <- data.frame(A  = rnorm(5,0,0.001), B = rnorm(5,0,0.001))
-t2v <- data.frame(A  = rnorm(5,0,0.001), B = rnorm(5,0,0.001))
-t3v <- data.frame(A  = rnorm(5,0,0.001), B = rnorm(5,0,0.001))
-t4v <- data.frame(A  = rnorm(5,0,0.001), B = rnorm(5,0,0.001))
-
-factorV <- sd(c(s(t1v),s(t2v),s(t3v),s(t4v)))
-
-numeratorV <-  mean(s(t1v),s(t2v)) - mean(s(t3v),s(t4v))
-
-biasesLowVariance[i] <- numeratorV / factorV
-}
-ggplot()+geom_histogram(aes(x=biasesLowVariance, y = ..density..), alpha = 0.6, bins=50)+
-  theme_tufte()+labs(title="10k biases for identical means and sd =.001")+ xlab("bias")
-```
-\normalsize
-
-Again, not so informative.
-Now, let's try sampling from distributions where there in fact is a difference in means.
-Terms from $X$ are on average .1 similar to $A$ (and still $0$ to $B$), while terms from $Y$ are $.1$ similar to $B$ (and 0 to $A$).
-The standard deviation is 0.05 in all the cases.
-There is a  clear difference between $X$ and $Y$ and quick visual inspection should tell us so.
-
-\vspace{1mm}
-\footnotesize
-```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=FALSE, fig.show = "hold", out.width = "100%"}
-set.seed(766)
-t1d2 <- data.frame(A  = rnorm(5,.1,0.05), B = rnorm(5,0,0.05))
-t2d2 <- data.frame(A  = rnorm(5,.1,0.05), B = rnorm(5,0,0.05))
-t3d2 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,.1,0.05))
-t4d2 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,.1,0.05))
-```
-\normalsize
-
-\vspace{1mm}
-\footnotesize
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
-library(grid)
-library(gridExtra)
-t1d2Plot <- ggplot(t1d2)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t1 in X")+xlim(xlimits)
-
-t2d2Plot <- ggplot(t2d2)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t2 in X")+xlim(xlimits)
-
-t3d2Plot <- ggplot(t3d2)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t3 in Y")+xlim(xlimits)
-
-t4d2Plot <- ggplot(t4d2)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t4 in Y")+xlim(xlimits)
-
-grid.arrange(t1d2Plot,t2d2Plot, t3d2Plot, t4d2Plot, ncol=2,   
-      top = textGrob("different means, sd=0.05",gp=gpar(fontsize=15,font=1)))
-```
-\normalsize
-
-\noindent  Is this clear difference mirrored in the calculations?
-
-\footnotesize
-```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=FALSE, fig.show = "hold", out.width = "100%"}
-factorD2 <- sd(c(s(t1d2),s(t2d2),s(t3d2),s(t4d2)))
-numeratorD2 <-  mean(s(t1d2),s(t2d2)) - mean(s(t3d2),s(t4d2))
-biasD2 <- numeratorD2 / factorD2
-biasD2
-```
-\normalsize
-
-\noindent The absolute value of the effect size is smaller than in the null case with the same standard deviation.  Let's simulate 10000 drawings:
-
-\vspace{1mm}
-\footnotesize
-```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
-biasesD2 <- numeric(10000)
-for(i in 1:10000){
-t1d2 <- data.frame(A  = rnorm(5,.1,0.05), B = rnorm(5,0,0.05))
-t2d2 <- data.frame(A  = rnorm(5,.1,0.05), B = rnorm(5,0,0.05))
-t3d2 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,.1,0.05))
-t4d2 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,.1,0.05))
-
-factorD2 <- sd(c(s(t1d2),s(t2d2),s(t3d2),s(t4d2)))
-numeratorD2 <-  mean(s(t1d2),s(t2d2)) - mean(s(t3d2),s(t4d2))
-biasesD2[i] <- numeratorD2/factorD2
-}
-
-ggplot()+geom_histogram(aes(x=biasesD2, y = ..density..), alpha = 0.6, bins=50)+
-  theme_tufte()+labs(title="10k biases for different means and sd =.05")+ xlab("bias")
-```
-\normalsize
-
-\noindent Now suppose we keep the means the same but increase the standard deviation to .1.
-
-\vspace{1mm}
-\footnotesize
-```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=FALSE, fig.show = "hold", out.width = "100%"}
-set.seed(766)
-t1d1 <- data.frame(A  = rnorm(5,.1,0.1), B = rnorm(5,0,0.1))
-t2d1 <- data.frame(A  = rnorm(5,.1,0.1), B = rnorm(5,0,0.1))
-t3d1 <- data.frame(A  = rnorm(5,0,0.1), B = rnorm(5,.1,0.1))
-t4d1 <- data.frame(A  = rnorm(5,0,0.1), B = rnorm(5,.1,0.1))
-```
-\normalsize
-
-\vspace{1mm}
-\footnotesize
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
-t1d1Plot <- ggplot(t1d1)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t1 in X")+xlim(xlimits)
-
-t2d1Plot <- ggplot(t2d1)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t2 in X")+xlim(xlimits)
-
-t3d1Plot <- ggplot(t3d1)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t3 in Y")+xlim(xlimits)
-
-t4d1Plot <- ggplot(t4d1)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t4 in Y")+xlim(xlimits)
-
-grid.arrange(t1d1Plot,t2d1Plot, t3d1Plot, t4d1Plot, ncol=2)
-```
-\normalsize
-
-\noindent  Is this clear difference mirrored in the calculations?
-
-\footnotesize
-```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=FALSE, fig.show = "hold", out.width = "100%"}
-factorD1 <- sd(c(s(t1d1),s(t2d1),s(t3d1),s(t4d1)))
-numeratorD1 <-  mean(s(t1d1),s(t2d1)) - mean(s(t3d1),s(t4d1))
-biasD1 <- numeratorD1 / factorD1
-biasD1
-```
-\normalsize
-
-\noindent The absolute value of the effect size is smaller than in the previous case!
-
-\vspace{1mm}
-\footnotesize
-```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
-biasesD1 <- numeric(10000)
-
-for(i in 1:10000){
-t1d1 <- data.frame(A  = rnorm(5,.1,0.1), B = rnorm(5,0,0.1))
-t2d1 <- data.frame(A  = rnorm(5,.1,0.1), B = rnorm(5,0,0.1))
-t3d1 <- data.frame(A  = rnorm(5,0,0.1), B = rnorm(5,.1,0.1))
-t4d1 <- data.frame(A  = rnorm(5,0,0.1), B = rnorm(5,.1,0.1))
-
-factorD1 <- sd(c(s(t1d1),s(t2d1),s(t3d1),s(t4d1)))
-numeratorD1 <-  mean(s(t1d1),s(t2d1)) - mean(s(t3d1),s(t4d1))
-biasesD1[i] <- numeratorD1/factorD1
-}
-
-ggplot()+geom_histogram(aes(x=biasesD1, y = ..density..), alpha = 0.6, bins=50)+
-  theme_tufte()+labs(title="10k biases for different means and sd =.001")+ xlab("bias")
-```
-\normalsize
-
-This is is a bit better, but still quite some uncertainty is involved, far from what systematically low mean-based p-values reported in the papers might suggest.
-Of course, this is a bit of a caricature, as our word lists were short (four protected words and 10 attributes).
-But the word lists used in the actually papers are not much longer.
-In such a set-up the key observations are as follows:
-
-- Seemingly high bias measures might arise even if the underlying processes actually have the same mean.
-
-- Even if the mean remains the same, non-negligible changes can result from a shift in the standard deviation of the original process, and the change might go in the opposite direction than a visualisation of datapoints might suggest, because with the decrease of standard deviation, the factor decreases and the bias increases.
-
-- Even if the underlying means are the same, but the variation is different, the bias metric in the long run could tend toward a different value:
-
-- The lack of control group in the paper and our analysis indicates that without neutral baseline it is difficult
-to interpret the effectiveness of the metric.
-
-\vspace{1mm}
-\footnotesize
-```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
-mean(biasesD1); mean(biasesD2)
-median(biasesD1); median(biasesD2)
-```
-
-\normalsize
-and so the point estimations of bias are sensitive to factors other than the underlying process means.
-
-- Even if there is a difference in means, the bias metric can be lower, and the uncertainty about it needs to be gauged.
-
-- The uncertainty resulting from including the raw datapoint variance into considerations is more extensive than the one suggested by the low p-values obtained from taking means as datapoints.
-
-\vspace{1mm}
-\footnotesize
-```{r,echo=TRUE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
-quantile(biasesD2, probs = c(0.275,0.975))
-quantile(biasesD1, probs = c(0.275,0.975))
-```
-\normalsize
-
-# Bayesian estimation
-
-One thing to notice is that in @Manzini2019blackToCriminal words from all three subclasses (in the case of religion by a subclass we mean Christianity, Judaism, and Islam) were compared against all of the stereotypes, which means that there was no distinction between cases in which the stereotype is associated with a given sub-class, as opposed to the situation in which it is associated with another one.
-Not all of the stereotypical words have to be considered as harmful for all of the sub-groups.
-One should investigate the sub-groups separately as some of them may have stronger harmful associations than others.
-
-We decided to add control groups in the form of two classes — neutral words and human-related words.
-Without a proper control group it is quite hard to compare the resulting cosine distances and decide on their significance in bias detection.
-We prepared approximately 230 neutral words to double-check the prima-facie neutral hypothesis that their cosine similarity to the protected words will oscillate around 0 (that is, the distances will be around 1).
-This provides us with a more reliable point of reference.
-Moreover,
-we added human attributes that are associated with people in general to investigate whether the smaller cosine distance between protected words and stereotypes can result simply from the fact that the stereotype predicates are associated with humans.
-
-One may consider improving the bias detection method presented in @Manzini2019blackToCriminal.
-As the chosen protected words and attributes do not represent the whole biased population but only a sample, it is justified to introduce the uncertainty measure to the method.
-We decided to use Bayesian estimation to make the cosine distance method more methodologically correct.
-
-Our prior distributions for the Bayesian model reflect our first stab at what we know or expect before we see the data.
-All our distributions are fairly weak.
-The distribution for mean distance (cosine distance) is set to fit (almost completely) between 0 and 2 with the mean equal to 1, in line with the idea that  most of the cosine distances may be considered neutral and therefore their similarity will be around 0.
-That is, we take it to be DISTRIBUTION.
-The  prior distribution for standard deviation is half-cauchy between 0 and 1, but unlikely to be as large as 1.
-Finally the prior distributions for  the coefficients have most of their mass between -1 and 1, with mean around 0.
-As we want to put ourselves in skeptical shoes looking at   @Manzini2019blackToCriminal and asking whether the data would indeed convince us of the existence of bias, we decided to use slightly skeptical (mean 0), but  fairly weak (sd = ) as our regularizing priors.
-The model parameters will have a posterior distribution obtained using either Hamiltionian Monte Carlo methods (STAN) available through the rethinking package.
-
-\begin{center}
-\begin{figure}[!htb]\centering
-   \begin{minipage}{0.55\textwidth}
-  \includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
-   \end{minipage}
-   \begin {minipage}{0.43\textwidth}
-    \includegraphics[width=7cm]{../images/religionGoogleCoeffs.jpeg}
-   \end{minipage}
-
-  \begin{minipage}{0.55\textwidth}
-\includegraphics[width=7cm]{../images/genderCoeffs.jpeg}
-\end{minipage}
-   \begin {minipage}{0.43\textwidth}
-    \includegraphics[width=7cm]{../images/genderGoogleCoeffs.jpeg}
-   \end{minipage}
-
-  \begin{minipage}{0.55\textwidth}
-\includegraphics[width=7cm]{../images/raceCoeffs.jpeg}
-\end{minipage}
-   \begin {minipage}{0.43\textwidth}
-    \includegraphics[width=7cm]{../images/raceGoogleCoeffs.jpeg}
-   \end{minipage}
-\end{figure}
-
-\end{center}
-
-Let's analyse dataset-level coefficients together with their highest posterior density intervals, for the three topic groups.
-We compare the results for two different word emebeddings corpuses - RedditL2 and GoogleNews.
-For Google embeddings the HPDIs
-for all class coefficients (associated, different, human, and none) include zero.
-This can lead one to a conclusion that the impact for associated, different, human and neutral attribute is, when averaged, quite similar.
-This indicates how including the uncertainty
-may change the use and interpretation of Manzini, Lim, Tsvetkov, & Black (2019) MAC metric. It seems that if one focuses only on differences between means of means, this is too simplistic.
-
-In the case of Reddit word embeddings the situation is similar although HPDI is below 0 for Gender class when looking at `associated` and `different` mean coefficient.
-This can suggest that there is indeed slightly stronger impact of these attributes.
-One should also notice how in general associated and different coefficients have quite similar HPDI range, the highest observed absolute difference is equal to only 0.1.
-This suggests that again the impact of associated attributes and different ones is not clear at first sight.
-Finally, let’s compare the HPDIs for Reddit and Reddit debiased datasets.
-For Religion and Race dataset there is a minor shift (in absolute values the highest change is equal to approximately 0.1) of the mean coefficients towards zero.
-However for Gender dataset there is no significant change.
-This is of course a general look at the group coefficients.
-Let’s now analyze the individual words.
-
-Looking at Reddit individual words one can observe that for Religion the cosine distance results for the associated attributes are for most of the words close to human attributes as well.
-This can suggest that in some cases words concerning humans can have higher
-similarity with some protected words independently of whether they are stereotypical or neutral-human words.
-One should also pay attention to the fact that for all of the associated and different attributes in Religion, the uncertainty intervals overlap at
-some point.
-What is even more surprising is that for protected words, such as “torah” the associated attribute has the the cosine distance slightly over 1, which means no positive similarity!
-If the protected words that we chose do not have high similarity
-with harmful associated attributes, then one should consider at least three scenarios.
-The first one is that the choice of the protected words and attributes may be corrupted.
-The second one is that the metric is unable to catch the hidden bias properly.
-The third one is that there is actually no bias between the words.
-Regardless of which scenario one considers, it is essential to take a look at the individual values before averaging them or aggregating in other ways.
-It seems that using Bayesian method can enhance
-the process of verifying the hypotheses concerning the choice of protected words and attributes.
-
-Surprisingly in Gender one can observe high cosine similarity values between some female stereotypical professions and male protected words.
-If a word stereotypically associated with females has low cosine distance to male protected words, then one should investigate the issue further.
-The reasons for this unexpected cosine distance may lie in the frequency of appearance of the protected words in the raw data.
-Some of the groups (like Muslim people or females) may have less representation in the data that is taken as an input for word embeddings.
-Therefore, if we assume that the MAC detects actually co-occurrence only, it makes sense that they can have high similarity with associated attributes and lower similarity with different ones.
-At the same time male protected words and other religions can have high similarity with different attributes because they have greater representation in the dataset in general and occur close to much more concepts.
-Cosine distance seems to capture the information on the co-occurrence of words and not on the semantic similarity strictly speaking.
-
-In Google, one may notice interesting phenomena as well.
-Although the GoogleNews dataset is larger, trained on different data sources and with the use of more dimensions, some of the results are similar to the ones obtained for the Reddit corpus.
-For Religion
-almost all of the values for associated class intersect with different class as well.
-In the case of Race it is 100% of the available words.
-This indicates again that it is not clear how the metric should be used.
-Quite a different situation takes place for
-Gender, where the similarity between associated and different is present mostly for the male protected words.
-This means that male words have high similarity with both male stereotypical professions and female ones.
-However, for females, the similarity is
-high mostly only for the female stereotypical professions. This observation could not be made when using MAC metric only, as it requires investigating the individual words, and providing uncertainty.
-
-# Effects of debiasing
-
-One may verify if the dataset-level coefficients changed after the debiasing method used in @Manzini2019blackToCriminal. 
-In some cases there is a minor change to the right, the mean value is closer to 0.
-The `associated` and `differen` attributes are now more overlapping than before.
-However, the uncertainty intervals are still high and therefore the significance of the change cannot be considered as enormous.
-The surprising obervation is that in the case of `Gender` and `Race` groups, the neutral words shifted to the left, further from the neutral position.
-
-\begin{center}
-\begin{figure}[!htb]\centering
-   \begin{minipage}{0.55\textwidth}
-  \includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
-   \end{minipage}
-   \begin {minipage}{0.43\textwidth}
-    \includegraphics[width=7cm]{../images/debiasedReligionRedditCoeffs.jpeg}
-   \end{minipage}
-
-  \begin{minipage}{0.55\textwidth}
-\includegraphics[width=7cm]{../images/genderCoeffs.jpeg}
-\end{minipage}
-   \begin {minipage}{0.43\textwidth}
-    \includegraphics[width=7cm]{../images/debiasedGenderRedditCoeffs.jpeg}
-   \end{minipage}
-
-  \begin{minipage}{0.55\textwidth}
-\includegraphics[width=7cm]{../images/raceCoeffs.jpeg}
-\end{minipage}
-   \begin {minipage}{0.43\textwidth}
-    \includegraphics[width=7cm]{../images/debiasedRaceRedditCoeffs.jpeg}
-   \end{minipage}
-\end{figure}
-
-\end{center}
-
-When analyzing the Reddit debiased results one should investigate the change of the individual cosine distance values as well.
-It seems that not all of the protected words are treated equally when performing debiasing.
-In religion, the values for associated class moved towards 0 for most of the words except for the ones for Islam, where still the associated class has quite lower cosine distance than the different one.
-Similar situation takes place in Gender, where female words have still much lower cosine distance values for
-associated class than for the different one.
-In the case of male protected words it is mostly almost the same interval for associated and different.
-As the Gender data is quite specific as the attributes are not harmful adjectives but (in an ideal world)
-neutral professions, the aim (if we follow MAC assumptions) should actually be to make the cosine distance same for both female and male protected words.
-However, as we pointed out the situation after debiasing can sometimes be better only for one protected
-group, which is not the desired outcome.
-In the case of Religion, even after debiasing, all of the protected words have intersections between associated and different class and similarity greater than 0.
-As all of the attributes in religion data are negative
-and harmful stereotypes, one should not aim at making the distance between protected words and associated, and different class the same but rather to move it towards 1.
-
-# Discussion
-
-\newpage
-
-\noindent \huge  \textbf{Appendix} \normalsize
-
-\noindent \Large \textbf{Word lists, including human and neutral predicates} \normalsize
+# Library for MAC
 
 # References {-}
-
-\vspace{-3mm}

### 2021-12-10 paperDraft/paperDraft2.tex
setting up paper draft 2

--- 2021-07-21 paperDraft/paperDraft.tex
@@
 \PassOptionsToPackage{hyphens}{url}
-\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}
+\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
 %
@@
   dvipsnames,enabledeprecatedfontcommands]{scrartcl}
+\title{Taking uncertainty in word embedding bias estimation seriously: a
+bayesian approach}
+\author{Alicja Dobrzeniecka and Rafal Urbaniak}
+\date{}
+
 \usepackage{amsmath,amssymb}
 \usepackage{lmodern}
-\usepackage{ifxetex,ifluatex}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+\usepackage{iftex}
+\ifPDFTeX
   \usepackage[T1]{fontenc}
@@
   colorlinks=true,
-  linkcolor=Maroon,
-  filecolor=Maroon,
-  citecolor=Blue,
-  urlcolor=blue,
+  linkcolor={Maroon},
+  filecolor={Maroon},
+  citecolor={Blue},
+  urlcolor={blue},
   pdfcreator={LaTeX via pandoc}}
 \urlstyle{same} % disable monospaced font for URLs
-\usepackage{color}
-\usepackage{fancyvrb}
-\newcommand{\VerbBar}{|}
-\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
-\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
-% Add ',fontsize=\small' for more characters per line
-\usepackage{framed}
-\definecolor{shadecolor}{RGB}{248,248,248}
-\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\BuiltInTok}[1]{#1}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
-\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
-\newcommand{\ExtensionTok}[1]{#1}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ImportTok}[1]{#1}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\NormalTok}[1]{#1}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\RegionMarkerTok}[1]{#1}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\usepackage{longtable,booktabs,array}
-\usepackage{calc} % for calculating minipage widths
-% Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-\makeatother
-% Allow footnotes in longtable head/foot
-\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
-\makesavenoteenv{longtable}
 \usepackage{graphicx}
@@
 \setcounter{secnumdepth}{5}
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
+  \fi
+  % set entry spacing
+  \setlength{\parskip}{#2\cslentryspacingunit}
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 %\documentclass{article}
@@
 \usepackage{booktabs}
-
+%\usepackage[left]{showlabels}
 \usepackage{multirow}
+\usepackage{subcaption}
 
@@
 \usepackage[textsize=footnotesize]{todonotes}
+\newcommand{\ali}[1]{\todo[color=gray!40]{\textbf{Alicja:} #1}}
+\newcommand{\mar}[1]{\todo[color=blue!40]{#1}}
+\newcommand{\raf}[1]{\todo[color=olive!40]{#1}}
+
 %\linespread{1.5}
+\newcommand{\indep}{\!\perp \!\!\! \perp\!}
 
@@
 %language
-\usepackage{times}
+%\usepackage{times}
+\usepackage{mathptmx}
+\usepackage[scaled=0.86]{helvet}
 \usepackage{t1enc}
@@
 %\usepackage{polski}
-\usepackage[utf8]{inputenc}
-\usepackage{mathptmx}
-\usepackage[scaled=0.88]{helvet}
 
@@
 \newcommand{\exi}[1]{\exists\, {_{#1}}}
-\newcommand{\pr}[1]{\mathsf{P}(#1)}
+\newcommand{\pr}[1]{\ensuremath{\mathsf{P}(#1)}}
 \newcommand{\cost}{\mathsf{cost}}
+\newcommand{\benefit}{\mathsf{benefit}}
+\newcommand{\ut}{\mathsf{ut}}
 
@@
 \newtheorem{theorem}{Theorem}
+\newtheorem{corollary}{Corollary}[fact]
 
@@
 	}
+
 %--------------------
@@
 %\bibliographystyle{apalike}
-\ifluatex
+
+\usepackage{tikz}
+\usetikzlibrary{positioning,shapes,arrows}
+
+\ifLuaTeX
   \usepackage{selnolig}  % disable illegal ligatures
 \fi
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
-  % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-
-\title{Taking uncertainty in word embedding bias estimation seriously: a
-bayesian approach}
-\author{Alicja Dobrzeniecka and Rafal Urbaniak}
-\date{}
 
@@
 
-\tableofcontents
-
 \hypertarget{cosine-based-measures-of-bias}{%
@@
 bias}\label{cosine-based-measures-of-bias}}
+
+\raf{fix bibliography}
 
@@
 One of the first measures in the discussion has been developed by
-(\textbf{bolukbasi2016man?}). There, the gender bias of a word \(w\) is
-understood as its projection on the gender direction
-\(\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})\) (the
-gender direction is the top principal compontent of ten gender pair
+Bolukbasi, Chang, Zou, Saligrama, \& Kalai (2016). There, the gender
+bias of a word \(w\) is understood as its projection on the gender
+direction \(\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})\)
+(the gender direction is the top principal compontent of ten gender pair
 difference vectors). The underlying idea is that no bias is present if
@@
 
-The use of projections has been ciriticized for instance by
-(\textbf{Gonen2019lipstick?}), who point out that while gender-direction
-might be an indicator of bias, it is only one possible manifestation of
-it, and reducing a projection of words might be insufficient. For
-instance, ``math'' and ``delicate'' might be in equal distance to both
-explicitly gendered words while being closer to quite different
-stereotypical attribute words. Further, the authors point out that most
-word pairs preserve similarity under debiasing meant to minimize
-projection-based bias.\footnote{(\textbf{bolukbasi2016man?}) use also
-  another method which involves analogies and their evaluations by human
-  users on Mechanical Turk. It is discussed and criticized in
-  (\textbf{Nissim2020fair?}).}
+The use of projections has been ciriticized for instance by Gonen \&
+Goldberg (2019), who point out that while gender-direction might be an
+indicator of bias, it is only one possible manifestation of it, and
+reducing a projection of words might be insufficient. For instance,
+``math'' and ``delicate'' might be in equal distance to both explicitly
+gendered words while being closer to quite different stereotypical
+attribute words. Further, the authors point out that most word pairs
+preserve similarity under debiasing meant to minimize projection-based
+bias.\footnote{Bolukbasi et al. (2016) use also another method which
+  involves analogies and their evaluations by human users on Mechanical
+  Turk. It is discussed and criticized in (Nissim, Noord, \& Goot,
+  2020).}
 
-To measure bias in word embeddings,
-(\textbf{Caliskan2017semanticsBiases?}) proposed the Word Embedding
-Association Test (WEAT). The idea is that the measure of biases between
-two sets of target words, \(X\) and \(Y\), (we call them protected
-words) should be quantified in terms of the cosine similarity between
-the protected words and attribute words coming from two sets of
-stereotype attribute words, \(A\) and \(B\) (we'll call them
-attributes). For instance, \(X\) might be a set of male names, \(Y\) a
-set of female names, \(A\) might contain stereotypically male-related
-career words, and \(B\) stereotypically female-related family words.
-WEAT is a modification of the Implicit Association Test (IAT)
-(\textbf{Nosek2002harvesting?}) used in psychology and uses almost the
-same word sets, allowing for a \emph{prima facie} sensible comparison
-with bias in humans. If the person's attitude towards given pair of
-concept is to be interpreted as neutral, there should be no noticeable
-task completion time difference, and the final value from the formula
-should be around 0. Let \(f\) be a similarity measure (usually, cosine
-similarity). The association difference for a term \(t\) is:
+To measure bias in word embeddings, Caliskan, Bryson, \& Narayanan
+(2017) proposed the Word Embedding Association Test (WEAT). The idea is
+that the measure of biases between two sets of target words, \(X\) and
+\(Y\), (we call them protected words) should be quantified in terms of
+the cosine similarity between the protected words and attribute words
+coming from two sets of stereotype attribute words, \(A\) and \(B\)
+(we'll call them attributes). For instance, \(X\) might be a set of male
+names, \(Y\) a set of female names, \(A\) might contain stereotypically
+male-related career words, and \(B\) stereotypically female-related
+family words. WEAT is a modification of the Implicit Association Test
+(IAT) (Nosek, Banaji, \& Greenwald, 2002) used in psychology and uses
+almost the same word sets, allowing for a \emph{prima facie} sensible
+comparison with bias in humans. If the person's attitude towards given
+pair of concept is to be interpreted as neutral, there should be no
+noticeable task completion time difference, and the final value from the
+formula should be around 0. Let \(f\) be a similarity measure (usually,
+cosine similarity). The association difference for a term \(t\) is:
 \begin{align}
@@
 \end{align} \noindent \(s(X,Y,A,B)\) is the statistic used in the
-signifcance test, and the \(p\)-value obtained by bootstrapping: it is
+significance test, and the \(p\)-value obtained by bootstrapping: it is
 the frequency of \(s(X_i,Y_i,A,B)>s(X,Y,A,B)\) for all equally sized
@@
 
-(\textbf{Caliskan2017semanticsBiases?}) show that significant
-biases---thus measured--- similar to the ones discovered by IAT can be
-discovered in word embeddings. (\textbf{Lauscher2019multidimensional?})
-extended the methodology to a multilingual and cross-lingual setting,
-arguing that using Euclidean distance instead of similarity does not
-make much difference, while the bias effects vary greatly across
-embedding models (interestingly, with social media-text trained
-embeddings being less biased than those based on Wikipedia).
+Caliskan et al. (2017) show that significant biases---thus measured---
+similar to the ones discovered by IAT can be discovered in word
+embeddings. Lauscher \& Glavas (2019) extended the methodology to a
+multilingual and cross-lingual setting, arguing that using Euclidean
+distance instead of similarity does not make much difference, while the
+bias effects vary greatly across embedding models (interestingly, with
+social media-text trained embeddings being less biased than those based
+on Wikipedia).
 
-A similar methodology is employed by (\textbf{Garg2018years?}), who
-employ word embeddings trained on corpora from different decades to
-study the shifts in various biases. For instance, to compute the
-occupational embeddings bias for women the authors first compute the
-average vector of vector emeddings of words that represent women
+A similar methodology is employed by Garg, Schiebinger, Jurafsky, \& Zou
+(2018), who employ word embeddings trained on corpora from different
+decades to study the shifts in various biases. For instance, to compute
+the occupational embeddings bias for women the authors first compute the
+average vector of vector embeddings of words that represent women
 (e.g.~``she,'' ``female''), then calculate the Euclidean distance
@@
 
-\hypertarget{some-methodological-problems}{%
-\section{Some methodological
-problems}\label{some-methodological-problems}}
-
-Such approaches, however, are not unproblematic. One may cosinder the
-selection of attributes for the bias research. For one thing, the word
-lists are often fairly small, and the papers do not discuss the impact
-of word list sizes on the uncertainty involved. We will argue that the
-word list sizes are too small to yield strong conclusions, given the
-variances involved.
-
-Crucially, these approaches use means of mean average cosine
-similarities to measure similarity between protected word and harmful
-stereotypes. If one takes a closer look at the individual values that
-are taken for the calculations, it turns out that there are quite a few
-outliers and surprisingly dissimilar words. This issue will become
-transparent when we inspect the visualizations of individual cosine
-distances, following the idea that one of the first step to understand
-data is to look at it.
-
-With such a method the uncertainty involved is not really considered
-which makes it even more difficult to give reasonable interpretations of
-the results. We propose the use of Bayesian method to obtain some
-understanding of the influence the uncertainty has on the interpretation
-of final results.
-
-We also observed that in the paper there is no word class distinction
-and no control group. In the original paper, words from all three
-religions were compared against all of the stereotypes, which means that
-there was no distinction between cases in which the stereotype is
-associated with a given religion, as opposed to the situation in which
-it is associated with another one. Not all of the stereotypical words
-have to be considered as harmful for all of the religions. One should
-investigate the religions separately as some of them may have stronger
-harmful associations that others. One should also include control groups
-to have a way of comparing the stereotypical results with neutral or
-human-like words. Later in the text we will explain in details reasons
-for introducing control groups. In our analysis, we distinguish between
-stereotypes associated with a given group, stereotypes associated with
-different groups, and control groups: neutral words and stereotypes-free
-human predicates.
-
-The interpretation of the results is also a challenge. Assuming for a
-moment that the value of multi-class cosine distance is correct, one may
-question the interpretation. Manzini, Lim, Tsvetkov, \& Black (2019)
-summarize the averages of cosine distance per group (gender, race,
-religion). For now let us focus now on analyzing the values relating to
-religious biases. Here is the relevant fragment of table:
-
-\begin{longtable}[]{@{}ll@{}}
-\toprule
-Religion Debiasing & MAC \\
-\midrule
-\endhead
-Biased & 0.859 \\
-Hard Debiased & 0.934 \\
-Soft Debiased (\(\lambda\) = 0.2) & 0.894 \\
-\bottomrule
-\end{longtable}
-
-MAC stand for mean average cosine similarity, although in reality the
-the table contains mean distances. What may attract attention is the
-fact that the value of cosine distance in ``Biased'' category is already
-quite high even before debiasing. High cosine distance indicates low
-cosine similarity between values. One could think that average cosine
-similarity equal to approximately 0.141 is not significant enough to
-consider it as bias. However the authors aim to mitigate ``biases'' in
-vectors with such great distance to make it even larger.
-Methodologically the question is, on what basis is this small similarity
-still considered as a proof of the presence of bias, and whether these
-small changes are meaningful. This is in general the problem of scale
-and the lack of universal intervals. In contrast, statistical intervals
-will help us decide whether a given cosine similarity is high enough to
-consider the words to be more similar than if we chose them at random.
-We will use highest posterior density intervals, in line with Bayesian
-methodology.
-
-One may consider the issue of the curse of dimensionality. In our case,
-the curse of dimensionality may take place when there is an increase in
-the volume of data that results in adding extra dimensions to the
-Euclidean space. As the number of features increases, it may be harder
-and harder to obtain useful information from the data using the
-available algorithms. Using cosine similarity in high dimensions in word
-embeddings may be prone to the curse of dimensionality. According to
-Venkat (2018) there are reasons to consider this phenomenon when
-searching for word similarities in higher dimensions. An experiment is
-conducted that aims at showing how the similarity values and variation
-change as the number of dimensions increases. The hypothesis made in the
-paper states that two things will happen as the number of dimensions
-increase. First, the effort required to measure cosine similarity will
-be greater, and two, the similarity between data will blur out and have
-less variation. The authors generate random points with increasing
-number of dimensions where each dimension of a data point is given a
-value between 0 and 1. Then they pick one vector at random from each
-dimension class and calculate the cosine similarity between the chosen
-vector and the rest of the data. Then they check how the variation of
-values changes as the number of dimensions increases. It seems like the
-more dimensions there are, the smaller the variance and therefore it is
-less obvious how to interpret the resulting cosine similarities. Maybe
-the scale should be adjusted to the number of dimensions and variance so
-that it still gives us sensible information about data.
-
-\hypertarget{the-problem-with-means-of-means}{%
-\section{The problem with means of
-means}\label{the-problem-with-means-of-means}}
-
-The measures described all calculate means of means and their authors
-run statistical tests on sets of means. This, however, is problematic
-for two related reasons. One, by pre-averaging data we throw away
-information about sample sizes. For the former point, think about
-proportions: 10 out of 20 and 2 out of 4 give the same mean, but you
-would obtain more information by making the former observation rather
-than by making the latter. Two, when we pre-average, we remove
-variation, and so pre-averaging tend to manufacture false confidence. We
-will have more to say about whether this happens in the case of
-applications of MAC, for now let's go over a simple example to make the
-conceptual point clear.
-
-To illustrate let's employ the formulas used by
-(\textbf{Caliskan2017semanticsBiases?}) in a simple example.
-Conceptually, all such tests come up with rather short lists of
-protected words and rather short lists of stereotypical attributes.
-Clearly, these are not complete list. So let's treat them as samples
-from richer pools of stereotypical predicates and let's take the
-uncertainty and variation involved seriously.
-
-Consider a simple situation in which there are two protected classes,
-\(X=\{t_1,t_2\}\) and \(Y=\{t_3,t_4\}\) and two five-element attribute
-sets \(A\) and \(B\).
-
-First, we play around with a scenario in which all the protected terms
-are on average equally unsimilar to both sets (\(\mu =0\)) with standard
-deviation of \(.05\). Let's randomly draw similarity scores and plot the
-results with group means plotted as vertical lines.
-
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{set.seed}\NormalTok{(}\DecValTok{123}\NormalTok{)}
-\NormalTok{t1 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
-\NormalTok{t2 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
-\NormalTok{t3 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
-\NormalTok{t4 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\normalsize
-
-\begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/unnamed-chunk-2-1} \end{center}
-
-\noindent When you look at the datapoints, do you have the impression of
-a strong bias here? We wouldn't think so. But now let's run the
-calculations from (\textbf{Caliskan2017semanticsBiases?}).
-
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{s }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{ (table)\{ }\FunctionTok{mean}\NormalTok{(table}\SpecialCharTok{$}\NormalTok{A) }\SpecialCharTok{{-}} \FunctionTok{mean}\NormalTok{(table}\SpecialCharTok{$}\NormalTok{B)\}}
-\NormalTok{factor }\OtherTok{\textless{}{-}} \FunctionTok{sd}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1),}\FunctionTok{s}\NormalTok{(t2),}\FunctionTok{s}\NormalTok{(t3),}\FunctionTok{s}\NormalTok{(t4)))}
-\NormalTok{numerator }\OtherTok{\textless{}{-}}  \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1),}\FunctionTok{s}\NormalTok{(t2)) }\SpecialCharTok{{-}} \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t3),}\FunctionTok{s}\NormalTok{(t4))}
-\FunctionTok{print}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{factor =}\NormalTok{ factor,}\AttributeTok{numerator =}\NormalTok{ numerator, }\AttributeTok{bias =}\NormalTok{ numerator }\SpecialCharTok{/}\NormalTok{ factor))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## $factor
-## [1] 0.02342637
-## 
-## $numerator
-## [1] 0.04275325
-## 
-## $bias
-## [1] 1.825005
-\end{verbatim}
-
-\normalsize
-
-\noindent This should make us pause. We know these were points randomly
-drawn from distributions where there is no difference in means. Yet, the
-calculated effect size is 1.82, whereas the largest effect size reported
-by (\textbf{Caliskan2017semanticsBiases?}) is 1.81!
-
-Interestingly, if we repeat the drawing 10000 times, each time
-calculating the bias, it turns out that with this variance and sample
-size, pretty much anything can happen.
-
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{biasesNull }\OtherTok{\textless{}{-}} \FunctionTok{numeric}\NormalTok{(}\DecValTok{10000}\NormalTok{)}
-\ControlFlowTok{for}\NormalTok{(i }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\DecValTok{10000}\NormalTok{)\{}
-\NormalTok{t1 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
-\NormalTok{t2 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
-\NormalTok{t3 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
-\NormalTok{t4 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
-
-\NormalTok{factor }\OtherTok{\textless{}{-}} \FunctionTok{sd}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1),}\FunctionTok{s}\NormalTok{(t2),}\FunctionTok{s}\NormalTok{(t3),}\FunctionTok{s}\NormalTok{(t4)))}
-\NormalTok{numerator }\OtherTok{\textless{}{-}}  \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1),}\FunctionTok{s}\NormalTok{(t2)) }\SpecialCharTok{{-}} \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t3),}\FunctionTok{s}\NormalTok{(t4))}
-\NormalTok{biasesNull[i]  }\OtherTok{\textless{}{-}}\NormalTok{ numerator }\SpecialCharTok{/}\NormalTok{ factor}
-\NormalTok{\}}
-\FunctionTok{ggplot}\NormalTok{()}\SpecialCharTok{+}\FunctionTok{geom\_histogram}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{biasesNull, }\AttributeTok{y =}\NormalTok{ ..density..), }\AttributeTok{alpha =} \FloatTok{0.6}\NormalTok{, }\AttributeTok{bins=}\DecValTok{50}\NormalTok{)}\SpecialCharTok{+}
-  \FunctionTok{theme\_tufte}\NormalTok{()}\SpecialCharTok{+}\FunctionTok{labs}\NormalTok{(}\AttributeTok{title=}\StringTok{"10k biases for identical means and sd =.05"}\NormalTok{)}\SpecialCharTok{+} \FunctionTok{xlab}\NormalTok{(}\StringTok{"bias"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/unnamed-chunk-4-1} \end{center}
-\normalsize
-
-Now, let's simulate a situation where the means are identical but the
-standard deviation is much smaller, .001.
-
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{set.seed}\NormalTok{(}\DecValTok{124}\NormalTok{)}
-\NormalTok{t1v }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
-\NormalTok{t2v }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
-\NormalTok{t3v }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
-\NormalTok{t4v }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\normalsize
-
-\begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/unnamed-chunk-6-1} \end{center}
-
-\noindent When you look at the datapoints, do you have the impression of
-a strong bias here? We wouldn't think so. But now let's run the
-calculations from (\textbf{Caliskan2017semanticsBiases?}).
-
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{factorV }\OtherTok{\textless{}{-}} \FunctionTok{sd}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1v),}\FunctionTok{s}\NormalTok{(t2v),}\FunctionTok{s}\NormalTok{(t3v),}\FunctionTok{s}\NormalTok{(t4v)))}
-\NormalTok{numeratorV }\OtherTok{\textless{}{-}}  \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1v),}\FunctionTok{s}\NormalTok{(t2v)) }\SpecialCharTok{{-}} \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t3v),}\FunctionTok{s}\NormalTok{(t4v))}
-\FunctionTok{print}\NormalTok{(}\FunctionTok{list}\NormalTok{(}\AttributeTok{factor =}\NormalTok{ factorV,}\AttributeTok{numerator =}\NormalTok{ numeratorV, }\AttributeTok{bias =}\NormalTok{ numeratorV }\SpecialCharTok{/}\NormalTok{ factorV))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## $factor
-## [1] 0.0006402666
-## 
-## $numerator
-## [1] -0.001237637
-## 
-## $bias
-## [1] -1.933003
-\end{verbatim}
-
-\normalsize
-
-While the numerator and the factors changed a lot, the bias actually
-increased. One reason bias increases is that once the standard deviation
-goes down, so does the factor used in the calculation of bias.
-
-Again, to see whether this metric would provide us with meaningful
-information, let's simulate 10000 drawings.
-
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{set.seed}\NormalTok{(}\DecValTok{124}\NormalTok{)}
-
-\NormalTok{biasesLowVariance }\OtherTok{\textless{}{-}} \FunctionTok{numeric}\NormalTok{(}\DecValTok{10000}\NormalTok{)}
-\ControlFlowTok{for}\NormalTok{(i }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\DecValTok{10000}\NormalTok{)\{}
-\NormalTok{t1v }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
-\NormalTok{t2v }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
-\NormalTok{t3v }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
-\NormalTok{t4v }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.001}\NormalTok{))}
-
-\NormalTok{factorV }\OtherTok{\textless{}{-}} \FunctionTok{sd}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1v),}\FunctionTok{s}\NormalTok{(t2v),}\FunctionTok{s}\NormalTok{(t3v),}\FunctionTok{s}\NormalTok{(t4v)))}
-
-\NormalTok{numeratorV }\OtherTok{\textless{}{-}}  \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1v),}\FunctionTok{s}\NormalTok{(t2v)) }\SpecialCharTok{{-}} \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t3v),}\FunctionTok{s}\NormalTok{(t4v))}
-
-\NormalTok{biasesLowVariance[i] }\OtherTok{\textless{}{-}}\NormalTok{ numeratorV }\SpecialCharTok{/}\NormalTok{ factorV}
-\NormalTok{\}}
-\FunctionTok{ggplot}\NormalTok{()}\SpecialCharTok{+}\FunctionTok{geom\_histogram}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{biasesLowVariance, }\AttributeTok{y =}\NormalTok{ ..density..), }\AttributeTok{alpha =} \FloatTok{0.6}\NormalTok{, }\AttributeTok{bins=}\DecValTok{50}\NormalTok{)}\SpecialCharTok{+}
-  \FunctionTok{theme\_tufte}\NormalTok{()}\SpecialCharTok{+}\FunctionTok{labs}\NormalTok{(}\AttributeTok{title=}\StringTok{"10k biases for identical means and sd =.001"}\NormalTok{)}\SpecialCharTok{+} \FunctionTok{xlab}\NormalTok{(}\StringTok{"bias"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/unnamed-chunk-8-1} \end{center}
-\normalsize
-
-Again, not so informative. Now, let's try sampling from distributions
-where there in fact is a difference in means. Terms from \(X\) are on
-average .1 similar to \(A\) (and still \(0\) to \(B\)), while terms from
-\(Y\) are \(.1\) similar to \(B\) (and 0 to \(A\)). The standard
-deviation is 0.05 in all the cases. There is a clear difference between
-\(X\) and \(Y\) and quick visual inspection should tell us so.
-
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{set.seed}\NormalTok{(}\DecValTok{766}\NormalTok{)}
-\NormalTok{t1d2 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
-\NormalTok{t2d2 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
-\NormalTok{t3d2 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
-\NormalTok{t4d2 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\normalsize
-
-\vspace{1mm}
-\footnotesize
-
-\begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/unnamed-chunk-10-1} \end{center}
-\normalsize
-
-\noindent  Is this clear difference mirrored in the calculations?
-
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{factorD2 }\OtherTok{\textless{}{-}} \FunctionTok{sd}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1d2),}\FunctionTok{s}\NormalTok{(t2d2),}\FunctionTok{s}\NormalTok{(t3d2),}\FunctionTok{s}\NormalTok{(t4d2)))}
-\NormalTok{numeratorD2 }\OtherTok{\textless{}{-}}  \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1d2),}\FunctionTok{s}\NormalTok{(t2d2)) }\SpecialCharTok{{-}} \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t3d2),}\FunctionTok{s}\NormalTok{(t4d2))}
-\NormalTok{biasD2 }\OtherTok{\textless{}{-}}\NormalTok{ numeratorD2 }\SpecialCharTok{/}\NormalTok{ factorD2}
-\NormalTok{biasD2}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## [1] 1.490014
-\end{verbatim}
-
-\normalsize
-
-\noindent The absolute value of the effect size is smaller than in the
-null case with the same standard deviation. Let's simulate 10000
-drawings:
-
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{biasesD2 }\OtherTok{\textless{}{-}} \FunctionTok{numeric}\NormalTok{(}\DecValTok{10000}\NormalTok{)}
-\ControlFlowTok{for}\NormalTok{(i }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\DecValTok{10000}\NormalTok{)\{}
-\NormalTok{t1d2 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
-\NormalTok{t2d2 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
-\NormalTok{t3d2 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
-\NormalTok{t4d2 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.05}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.05}\NormalTok{))}
-
-\NormalTok{factorD2 }\OtherTok{\textless{}{-}} \FunctionTok{sd}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1d2),}\FunctionTok{s}\NormalTok{(t2d2),}\FunctionTok{s}\NormalTok{(t3d2),}\FunctionTok{s}\NormalTok{(t4d2)))}
-\NormalTok{numeratorD2 }\OtherTok{\textless{}{-}}  \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1d2),}\FunctionTok{s}\NormalTok{(t2d2)) }\SpecialCharTok{{-}} \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t3d2),}\FunctionTok{s}\NormalTok{(t4d2))}
-\NormalTok{biasesD2[i] }\OtherTok{\textless{}{-}}\NormalTok{ numeratorD2}\SpecialCharTok{/}\NormalTok{factorD2}
-\NormalTok{\}}
-
-\FunctionTok{ggplot}\NormalTok{()}\SpecialCharTok{+}\FunctionTok{geom\_histogram}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{biasesD2, }\AttributeTok{y =}\NormalTok{ ..density..), }\AttributeTok{alpha =} \FloatTok{0.6}\NormalTok{, }\AttributeTok{bins=}\DecValTok{50}\NormalTok{)}\SpecialCharTok{+}
-  \FunctionTok{theme\_tufte}\NormalTok{()}\SpecialCharTok{+}\FunctionTok{labs}\NormalTok{(}\AttributeTok{title=}\StringTok{"10k biases for different means and sd =.05"}\NormalTok{)}\SpecialCharTok{+} \FunctionTok{xlab}\NormalTok{(}\StringTok{"bias"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/unnamed-chunk-12-1} \end{center}
-\normalsize
-
-\noindent Now suppose we keep the means the same but increase the
-standard deviation to .1.
-
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{set.seed}\NormalTok{(}\DecValTok{766}\NormalTok{)}
-\NormalTok{t1d1 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
-\NormalTok{t2d1 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
-\NormalTok{t3d1 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
-\NormalTok{t4d1 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\normalsize
-
-\vspace{1mm}
-\footnotesize
-
-\begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/unnamed-chunk-14-1} \end{center}
-\normalsize
-
-\noindent  Is this clear difference mirrored in the calculations?
-
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{factorD1 }\OtherTok{\textless{}{-}} \FunctionTok{sd}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1d1),}\FunctionTok{s}\NormalTok{(t2d1),}\FunctionTok{s}\NormalTok{(t3d1),}\FunctionTok{s}\NormalTok{(t4d1)))}
-\NormalTok{numeratorD1 }\OtherTok{\textless{}{-}}  \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1d1),}\FunctionTok{s}\NormalTok{(t2d1)) }\SpecialCharTok{{-}} \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t3d1),}\FunctionTok{s}\NormalTok{(t4d1))}
-\NormalTok{biasD1 }\OtherTok{\textless{}{-}}\NormalTok{ numeratorD1 }\SpecialCharTok{/}\NormalTok{ factorD1}
-\NormalTok{biasD1}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## [1] 1.223023
-\end{verbatim}
-
-\normalsize
-
-\noindent The absolute value of the effect size is smaller than in the
-previous case!
-
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{biasesD1 }\OtherTok{\textless{}{-}} \FunctionTok{numeric}\NormalTok{(}\DecValTok{10000}\NormalTok{)}
-
-\ControlFlowTok{for}\NormalTok{(i }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\DecValTok{10000}\NormalTok{)\{}
-\NormalTok{t1d1 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
-\NormalTok{t2d1 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
-\NormalTok{t3d1 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
-\NormalTok{t4d1 }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{A  =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,}\DecValTok{0}\NormalTok{,}\FloatTok{0.1}\NormalTok{), }\AttributeTok{B =} \FunctionTok{rnorm}\NormalTok{(}\DecValTok{5}\NormalTok{,.}\DecValTok{1}\NormalTok{,}\FloatTok{0.1}\NormalTok{))}
-
-\NormalTok{factorD1 }\OtherTok{\textless{}{-}} \FunctionTok{sd}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1d1),}\FunctionTok{s}\NormalTok{(t2d1),}\FunctionTok{s}\NormalTok{(t3d1),}\FunctionTok{s}\NormalTok{(t4d1)))}
-\NormalTok{numeratorD1 }\OtherTok{\textless{}{-}}  \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t1d1),}\FunctionTok{s}\NormalTok{(t2d1)) }\SpecialCharTok{{-}} \FunctionTok{mean}\NormalTok{(}\FunctionTok{s}\NormalTok{(t3d1),}\FunctionTok{s}\NormalTok{(t4d1))}
-\NormalTok{biasesD1[i] }\OtherTok{\textless{}{-}}\NormalTok{ numeratorD1}\SpecialCharTok{/}\NormalTok{factorD1}
-\NormalTok{\}}
-
-\FunctionTok{ggplot}\NormalTok{()}\SpecialCharTok{+}\FunctionTok{geom\_histogram}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x=}\NormalTok{biasesD1, }\AttributeTok{y =}\NormalTok{ ..density..), }\AttributeTok{alpha =} \FloatTok{0.6}\NormalTok{, }\AttributeTok{bins=}\DecValTok{50}\NormalTok{)}\SpecialCharTok{+}
-  \FunctionTok{theme\_tufte}\NormalTok{()}\SpecialCharTok{+}\FunctionTok{labs}\NormalTok{(}\AttributeTok{title=}\StringTok{"10k biases for different means and sd =.001"}\NormalTok{)}\SpecialCharTok{+} \FunctionTok{xlab}\NormalTok{(}\StringTok{"bias"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{center}\includegraphics[width=1\linewidth]{paperDraft_files/figure-latex/unnamed-chunk-16-1} \end{center}
-\normalsize
-
-This is is a bit better, but still quite some uncertainty is involved,
-far from what systematically low mean-based p-values reported in the
-papers might suggest. Of course, this is a bit of a caricature, as our
-word lists were short (four protected words and 10 attributes). But the
-word lists used in the actually papers are not much longer. In such a
-set-up the key observations are as follows:
-
-\begin{itemize}
-\item
-  Seemingly high bias measures might arise even if the underlying
-  processes actually have the same mean.
-\item
-  Even if the mean remains the same, non-negligible changes can result
-  from a shift in the standard deviation of the original process, and
-  the change might go in the opposite direction than a visualisation of
-  datapoints might suggest, because with the decrease of standard
-  deviation, the factor decreases and the bias increases.
-\item
-  Even if the underlying means are the same, but the variation is
-  different, the bias metric in the long run could tend toward a
-  different value:
-\item
-  The lack of control group in the paper and our analysis indicates that
-  without neutral baseline it is difficult to interpret the
-  effectiveness of the metric.
-\end{itemize}
-
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{mean}\NormalTok{(biasesD1); }\FunctionTok{mean}\NormalTok{(biasesD2)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## [1] 1.563319
-\end{verbatim}
-
-\begin{verbatim}
-## [1] 1.690664
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{median}\NormalTok{(biasesD1); }\FunctionTok{median}\NormalTok{(biasesD2)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## [1] 1.638136
-\end{verbatim}
-
-\begin{verbatim}
-## [1] 1.707511
-\end{verbatim}
-
-\normalsize
-
-and so the point estimations of bias are sensitive to factors other than
-the underlying process means.
-
-\begin{itemize}
-\item
-  Even if there is a difference in means, the bias metric can be lower,
-  and the uncertainty about it needs to be gauged.
-\item
-  The uncertainty resulting from including the raw datapoint variance
-  into considerations is more extensive than the one suggested by the
-  low p-values obtained from taking means as datapoints.
-\end{itemize}
-
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{quantile}\NormalTok{(biasesD2, }\AttributeTok{probs =} \FunctionTok{c}\NormalTok{(}\FloatTok{0.275}\NormalTok{,}\FloatTok{0.975}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-##    27.5%    97.5% 
-## 1.539149 2.166488
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{quantile}\NormalTok{(biasesD1, }\AttributeTok{probs =} \FunctionTok{c}\NormalTok{(}\FloatTok{0.275}\NormalTok{,}\FloatTok{0.975}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-##    27.5%    97.5% 
-## 1.300648 2.356815
-\end{verbatim}
-
-\normalsize
-
-\hypertarget{bayesian-estimation}{%
-\section{Bayesian estimation}\label{bayesian-estimation}}
-
-One thing to notice is that in Manzini, Lim, Tsvetkov, \& Black (2019)
-words from all three subclasses (in the case of religion by a subclass
-we mean Christianity, Judaism, and Islam) were compared against all of
-the stereotypes, which means that there was no distinction between cases
-in which the stereotype is associated with a given sub-class, as opposed
-to the situation in which it is associated with another one. Not all of
-the stereotypical words have to be considered as harmful for all of the
-sub-groups. One should investigate the sub-groups separately as some of
-them may have stronger harmful associations than others.
-
-We decided to add control groups in the form of two classes --- neutral
-words and human-related words. Without a proper control group it is
-quite hard to compare the resulting cosine distances and decide on their
-significance in bias detection. We prepared approximately 230 neutral
-words to double-check the prima-facie neutral hypothesis that their
-cosine similarity to the protected words will oscillate around 0 (that
-is, the distances will be around 1). This provides us with a more
-reliable point of reference. Moreover, we added human attributes that
-are associated with people in general to investigate whether the smaller
-cosine distance between protected words and stereotypes can result
-simply from the fact that the stereotype predicates are associated with
-humans.
-
-One may consider improving the bias detection method presented in
-Manzini, Lim, Tsvetkov, \& Black (2019). As the chosen protected words
-and attributes do not represent the whole biased population but only a
-sample, it is justified to introduce the uncertainty measure to the
-method. We decided to use Bayesian estimation to make the cosine
-distance method more methodologically correct.
-
-Our prior distributions for the Bayesian model reflect our first stab at
-what we know or expect before we see the data. All our distributions are
-fairly weak. The distribution for mean distance (cosine distance) is set
-to fit (almost completely) between 0 and 2 with the mean equal to 1, in
-line with the idea that most of the cosine distances may be considered
-neutral and therefore their similarity will be around 0. That is, we
-take it to be DISTRIBUTION. The prior distribution for standard
-deviation is half-cauchy between 0 and 1, but unlikely to be as large as
-1. Finally the prior distributions for the coefficients have most of
-their mass between -1 and 1, with mean around 0. As we want to put
-ourselves in skeptical shoes looking at Manzini, Lim, Tsvetkov, \& Black
-(2019) and asking whether the data would indeed convince us of the
-existence of bias, we decided to use slightly skeptical (mean 0), but
-fairly weak (sd = ) as our regularizing priors. The model parameters
-will have a posterior distribution obtained using either Hamiltionian
-Monte Carlo methods (STAN) available through the rethinking package.
-
-\begin{center}
-\begin{figure}[!htb]\centering
-   \begin{minipage}{0.55\textwidth}
-  \includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
-   \end{minipage}
-   \begin {minipage}{0.43\textwidth}
-    \includegraphics[width=7cm]{../images/religionGoogleCoeffs.jpeg}
-   \end{minipage}
-
-  \begin{minipage}{0.55\textwidth}
-\includegraphics[width=7cm]{../images/genderCoeffs.jpeg}
-\end{minipage}
-   \begin {minipage}{0.43\textwidth}
-    \includegraphics[width=7cm]{../images/genderGoogleCoeffs.jpeg}
-   \end{minipage}
-
-  \begin{minipage}{0.55\textwidth}
-\includegraphics[width=7cm]{../images/raceCoeffs.jpeg}
-\end{minipage}
-   \begin {minipage}{0.43\textwidth}
-    \includegraphics[width=7cm]{../images/raceGoogleCoeffs.jpeg}
-   \end{minipage}
-\end{figure}
-
-\end{center}
-
-Let's analyse dataset-level coefficients together with their highest
-posterior density intervals, for the three topic groups. We compare the
-results for two different word emebeddings corpuses - RedditL2 and
-GoogleNews. For Google embeddings the HPDIs for all class coefficients
-(associated, different, human, and none) include zero. This can lead one
-to a conclusion that the impact for associated, different, human and
-neutral attribute is, when averaged, quite similar. This indicates how
-including the uncertainty may change the use and interpretation of
-Manzini, Lim, Tsvetkov, \& Black (2019) MAC metric. It seems that if one
-focuses only on differences between means of means, this is too
-simplistic.
-
-In the case of Reddit word embeddings the situation is similar although
-HPDI is below 0 for Gender class when looking at \texttt{associated} and
-\texttt{different} mean coefficient. This can suggest that there is
-indeed slightly stronger impact of these attributes. One should also
-notice how in general associated and different coefficients have quite
-similar HPDI range, the highest observed absolute difference is equal to
-only 0.1. This suggests that again the impact of associated attributes
-and different ones is not clear at first sight. Finally, let's compare
-the HPDIs for Reddit and Reddit debiased datasets. For Religion and Race
-dataset there is a minor shift (in absolute values the highest change is
-equal to approximately 0.1) of the mean coefficients towards zero.
-However for Gender dataset there is no significant change. This is of
-course a general look at the group coefficients. Let's now analyze the
-individual words.
-
-Looking at Reddit individual words one can observe that for Religion the
-cosine distance results for the associated attributes are for most of
-the words close to human attributes as well. This can suggest that in
-some cases words concerning humans can have higher similarity with some
-protected words independently of whether they are stereotypical or
-neutral-human words. One should also pay attention to the fact that for
-all of the associated and different attributes in Religion, the
-uncertainty intervals overlap at some point. What is even more
-surprising is that for protected words, such as ``torah'' the associated
-attribute has the the cosine distance slightly over 1, which means no
-positive similarity! If the protected words that we chose do not have
-high similarity with harmful associated attributes, then one should
-consider at least three scenarios. The first one is that the choice of
-the protected words and attributes may be corrupted. The second one is
-that the metric is unable to catch the hidden bias properly. The third
-one is that there is actually no bias between the words. Regardless of
-which scenario one considers, it is essential to take a look at the
-individual values before averaging them or aggregating in other ways. It
-seems that using Bayesian method can enhance the process of verifying
-the hypotheses concerning the choice of protected words and attributes.
-
-Surprisingly in Gender one can observe high cosine similarity values
-between some female stereotypical professions and male protected words.
-If a word stereotypically associated with females has low cosine
-distance to male protected words, then one should investigate the issue
-further. The reasons for this unexpected cosine distance may lie in the
-frequency of appearance of the protected words in the raw data. Some of
-the groups (like Muslim people or females) may have less representation
-in the data that is taken as an input for word embeddings. Therefore, if
-we assume that the MAC detects actually co-occurrence only, it makes
-sense that they can have high similarity with associated attributes and
-lower similarity with different ones. At the same time male protected
-words and other religions can have high similarity with different
-attributes because they have greater representation in the dataset in
-general and occur close to much more concepts. Cosine distance seems to
-capture the information on the co-occurrence of words and not on the
-semantic similarity strictly speaking.
-
-In Google, one may notice interesting phenomena as well. Although the
-GoogleNews dataset is larger, trained on different data sources and with
-the use of more dimensions, some of the results are similar to the ones
-obtained for the Reddit corpus. For Religion almost all of the values
-for associated class intersect with different class as well. In the case
-of Race it is 100\% of the available words. This indicates again that it
-is not clear how the metric should be used. Quite a different situation
-takes place for Gender, where the similarity between associated and
-different is present mostly for the male protected words. This means
-that male words have high similarity with both male stereotypical
-professions and female ones. However, for females, the similarity is
-high mostly only for the female stereotypical professions. This
-observation could not be made when using MAC metric only, as it requires
-investigating the individual words, and providing uncertainty.
-
-\hypertarget{effects-of-debiasing}{%
-\section{Effects of debiasing}\label{effects-of-debiasing}}
-
-One may verify if the dataset-level coefficients changed after the
-debiasing method used in Manzini, Lim, Tsvetkov, \& Black (2019). In
-some cases there is a minor change to the right, the mean value is
-closer to 0. The \texttt{associated} and \texttt{differen} attributes
-are now more overlapping than before. However, the uncertainty intervals
-are still high and therefore the significance of the change cannot be
-considered as enormous. The surprising obervation is that in the case of
-\texttt{Gender} and \texttt{Race} groups, the neutral words shifted to
-the left, further from the neutral position.
-
-\begin{center}
-\begin{figure}[!htb]\centering
-   \begin{minipage}{0.55\textwidth}
-  \includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
-   \end{minipage}
-   \begin {minipage}{0.43\textwidth}
-    \includegraphics[width=7cm]{../images/debiasedReligionRedditCoeffs.jpeg}
-   \end{minipage}
-
-  \begin{minipage}{0.55\textwidth}
-\includegraphics[width=7cm]{../images/genderCoeffs.jpeg}
-\end{minipage}
-   \begin {minipage}{0.43\textwidth}
-    \includegraphics[width=7cm]{../images/debiasedGenderRedditCoeffs.jpeg}
-   \end{minipage}
-
-  \begin{minipage}{0.55\textwidth}
-\includegraphics[width=7cm]{../images/raceCoeffs.jpeg}
-\end{minipage}
-   \begin {minipage}{0.43\textwidth}
-    \includegraphics[width=7cm]{../images/debiasedRaceRedditCoeffs.jpeg}
-   \end{minipage}
-\end{figure}
-
-\end{center}
-
-When analyzing the Reddit debiased results one should investigate the
-change of the individual cosine distance values as well. It seems that
-not all of the protected words are treated equally when performing
-debiasing. In religion, the values for associated class moved towards 0
-for most of the words except for the ones for Islam, where still the
-associated class has quite lower cosine distance than the different one.
-Similar situation takes place in Gender, where female words have still
-much lower cosine distance values for associated class than for the
-different one. In the case of male protected words it is mostly almost
-the same interval for associated and different. As the Gender data is
-quite specific as the attributes are not harmful adjectives but (in an
-ideal world) neutral professions, the aim (if we follow MAC assumptions)
-should actually be to make the cosine distance same for both female and
-male protected words. However, as we pointed out the situation after
-debiasing can sometimes be better only for one protected group, which is
-not the desired outcome. In the case of Religion, even after debiasing,
-all of the protected words have intersections between associated and
-different class and similarity greater than 0. As all of the attributes
-in religion data are negative and harmful stereotypes, one should not
-aim at making the distance between protected words and associated, and
-different class the same but rather to move it towards 1.
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-\newpage
-
-\noindent \huge  \textbf{Appendix} \normalsize
-
-\noindent \Large \textbf{Word lists, including human and neutral predicates}
-\normalsize
-
 \hypertarget{references}{%
@@
 
-\vspace{-3mm}
-
 \hypertarget{refs}{}
 \begin{CSLReferences}{1}{0}
-\leavevmode\hypertarget{ref-Manzini2019blackToCriminal}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-bolukbasi2016man}{}}%
+Bolukbasi, T., Chang, K.-W., Zou, J., Saligrama, V., \& Kalai, A.
+(2016). \emph{Man is to computer programmer as woman is to homemaker?
+Debiasing word embeddings}. Retrieved from
+\url{https://arxiv.org/abs/1607.06520}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Caliskan2017semanticsBiases}{}}%
+Caliskan, A., Bryson, J. J., \& Narayanan, A. (2017). Semantics derived
+automatically from language corpora contain human-like biases.
+\emph{Science}, \emph{356}(6334), 183--186.
+\url{https://doi.org/10.1126/science.aal4230}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Garg2018years}{}}%
+Garg, N., Schiebinger, L., Jurafsky, D., \& Zou, J. (2018). Word
+embeddings quantify 100 years of gender and ethnic stereotypes.
+\emph{Proceedings of the National Academy of Sciences}, \emph{115}(16),
+E3635--E3644. \url{https://doi.org/10.1073/pnas.1720347115}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Gonen2019lipstick}{}}%
+Gonen, H., \& Goldberg, Y. (2019). Lipstick on a pig: {D}ebiasing
+methods cover up systematic gender biases in word embeddings but do not
+remove them. \emph{Proceedings of the 2019 Conference of the North
+{A}merican Chapter of the Association for Computational Linguistics:
+Human Language Technologies, Volume 1 (Long and Short Papers)},
+609--614. Minneapolis, Minnesota: Association for Computational
+Linguistics. \url{https://doi.org/10.18653/v1/N19-1061}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Lauscher2019multidimensional}{}}%
+Lauscher, A., \& Glavas, G. (2019). Are we consistently biased?
+Multidimensional analysis of biases in distributional word vectors.
+\emph{CoRR}, \emph{abs/1904.11783}. Retrieved from
+\url{http://arxiv.org/abs/1904.11783}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Manzini2019blackToCriminal}{}}%
 Manzini, T., Lim, Y. C., Tsvetkov, Y., \& Black, A. W. (2019).
@@
 removing multiclass bias in word embeddings}. Retrieved from
-\url{http://arxiv.org/abs/1904.04047}
+\url{https://arxiv.org/abs/1904.04047}
 
-\leavevmode\hypertarget{ref-Venkat2018Curse}{}%
-Venkat, N. (2018). \emph{The curse of dimensionality: Inside out}.
-\url{https://doi.org/10.13140/RG.2.2.29631.36006}
+\leavevmode\vadjust pre{\hypertarget{ref-Nissim2020fair}{}}%
+Nissim, M., Noord, R. van, \& Goot, R. van der. (2020). Fair is better
+than sensational: Man is to doctor as woman is to doctor.
+\emph{Computational Linguistics}, \emph{46}(2), 487--497.
+\url{https://doi.org/10.1162/coli_a_00379}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Nosek2002harvesting}{}}%
+Nosek, B. A., Banaji, M. R., \& Greenwald, A. G. (2002). Harvesting
+implicit group attitudes and beliefs from a demonstration web site.
+\emph{Group Dynamics: Theory, Research, and Practice}, \emph{6}(1),
+101--115. \url{https://doi.org/10.1037/1089-2699.6.1.101}
 

### 2022-01-03 paperDraft/paperDraft3.tex
paper 4 started

--- 2021-12-10 paperDraft/paperDraft2.tex
@@
   10pt,
-  dvipsnames,enabledeprecatedfontcommands]{scrartcl}
+  dvipsnames,enabledeprecatedfontcommands, twocolumn]{scrartcl}
 \title{Taking uncertainty in word embedding bias estimation seriously: a
@@
 \urlstyle{same} % disable monospaced font for URLs
+\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
 \usepackage{graphicx}
@@
 \usepackage{graphicx}
-\usepackage{longtable}
+%\usepackage{longtable}
+\usepackage{supertabular, booktabs}
 \usepackage{ragged2e}
@@
 
+\ali{proofread and think if we should add more}
+
 Modern Natural Language Processing (NLP) models are used to complete
@@
 
+\todo{odleglosc nie musi uchwytywac relacji semantycznej zeby oceniac bias}
+
+\todo{zalozenie, bardziej ze jezeli odleglosci przekladaja sie na downstream, to warto patrzec na odleglosci, nawet bez glebszej filozoficznej interpretacji ich}
+
+\hypertarget{some-methodological-problems}{%
+\section{Some methodological
+problems}\label{some-methodological-problems}}
+
+Such approaches, however, are not unproblematic. One may consider the
+selection of attributes for the bias research. For one thing, the word
+lists are often fairly small, and the papers do not discuss the impact
+of word list sizes on the uncertainty involved. We will argue that the
+word list sizes are too small to yield strong conclusions, given the
+variances involved.
+
+Crucially, these approaches use means of mean average cosine
+similarities to measure similarity between protected word and harmful
+stereotypes. If one takes a closer look at the individual values that
+are taken for the calculations, it turns out that there are quite a few
+outliers and surprisingly dissimilar words. This issue will become
+transparent when we inspect the visualizations of individual cosine
+distances, following the idea that one of the first step to understand
+data is to look at it.
+
+With such a method the uncertainty involved is not really considered
+which makes it even more difficult to give reasonable interpretations of
+the results. We propose the use of Bayesian method to obtain some
+understanding of the influence the uncertainty has on the interpretation
+of final results.
+
+We also observed that in the paper there is no word class distinction
+and no control group. In the original paper, words from all three
+religions were compared against all of the stereotypes, which means that
+there was no distinction between cases in which the stereotype is
+associated with a given religion, as opposed to the situation in which
+it is associated with another one. Not all of the stereotypical words
+have to be considered as harmful for all of the religions. One should
+investigate the religions separately as some of them may have stronger
+harmful associations that others. One should also include control groups
+to have a way of comparing the stereotypical results with neutral or
+human-like words. Later in the text we will explain in details reasons
+for introducing control groups. In our analysis, we distinguish between
+stereotypes associated with a given group, stereotypes associated with
+different groups, and control groups: neutral words and stereotypes-free
+human predicates.
+
+The interpretation of the results is also a challenge. Assuming for a
+moment that the value of multi-class cosine distance is correct, one may
+question the interpretation. Manzini et al. (2019) summarize the
+averages of cosine distance per group (gender, race, religion). For now
+let us focus now on analyzing the values relating to religious biases.
+Here is the relevant fragment of table:
+
+\begin{longtable}[]{@{}ll@{}}
+\toprule
+Religion Debiasing & MAC \\
+\midrule
+\endhead
+Biased & 0.859 \\
+Hard Debiased & 0.934 \\
+Soft Debiased (\(\lambda\) = 0.2) & 0.894 \\
+\bottomrule
+\end{longtable}
+
+MAC stand for mean average cosine similarity, although in reality the
+the table contains mean distances. What may attract attention is the
+fact that the value of cosine distance in ``Biased'' category is already
+quite high even before debiasing. High cosine distance indicates low
+cosine similarity between values. One could think that average cosine
+similarity equal to approximately 0.141 is not significant enough to
+consider it as bias. However the authors aim to mitigate ``biases'' in
+vectors with such great distance to make it even larger.
+Methodologically the question is, on what basis is this small similarity
+still considered as a proof of the presence of bias, and whether these
+small changes are meaningful. This is in general the problem of scale
+and the lack of universal intervals. In contrast, statistical intervals
+will help us decide whether a given cosine similarity is high enough to
+consider the words to be more similar than if we chose them at random.
+We will use highest posterior density intervals, in line with Bayesian
+methodology.
+
+One may consider the issue of the curse of dimensionality. In our case,
+the curse of dimensionality may take place when there is an increase in
+the volume of data that results in adding extra dimensions to the
+Euclidean space. As the number of features increases, it may be harder
+and harder to obtain useful information from the data using the
+available algorithms. Using cosine similarity in high dimensions in word
+embeddings may be prone to the curse of dimensionality. According to
+(\textbf{Venkat2018Curse?}) there are reasons to consider this
+phenomenon when searching for word similarities in higher dimensions. An
+experiment is conducted that aims at showing how the similarity values
+and variation change as the number of dimensions increases. The
+hypothesis made in the paper states that two things will happen as the
+number of dimensions increase. First, the effort required to measure
+cosine similarity will be greater, and two, the similarity between data
+will blur out and have less variation. The authors generate random
+points with increasing number of dimensions where each dimension of a
+data point is given a value between 0 and 1. Then they pick one vector
+at random from each dimension class and calculate the cosine similarity
+between the chosen vector and the rest of the data. Then they check how
+the variation of values changes as the number of dimensions increases.
+It seems like the more dimensions there are, the smaller the variance
+and therefore it is less obvious how to interpret the resulting cosine
+similarities. Maybe the scale should be adjusted to the number of
+dimensions and variance so that it still gives us sensible information
+about data.
+
+\todo{MAC: avoid looking at only one vector}
+
+\todo{effect sizes}
+
+\hypertarget{library-for-mac}{%
+\section{Library for MAC}\label{library-for-mac}}
+
 \hypertarget{references}{%

### 2022-01-03 paperDraft/paperDraft4.Rmd
paper 4 started

---
title: 'Taking uncertainty in word embedding bias estimation seriously: a  bayesian
  approach'
author: "Alicja Dobrzeniecka and Rafal Urbaniak"
output:
  pdf_document:
    number_sections: yes
    df_print: kable
    keep_tex: yes
    includes:
      in_header: Rafal_latex7.sty
  html_document:
    df_print: paged
classoption: dvipsnames,enabledeprecatedfontcommands, twocolumn
fontsize: 10pt
documentclass: scrartcl
urlcolor: blue
bibliography: ../references/cosineReferences1.bib
csl: ../references/transactions-on-computational-logic.csl
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = '../')

library(ggplot2)
library(ggthemes)
library(gridExtra)
```

# Cosine-based measures of bias

\raf{fix bibliography}

\ali{proofread and think if we should add more}

Modern Natural Language Processing (NLP) models are used to complete various tasks such as providing email filters, smart assistants, search results, language translations, text analytics and so on. All of them need as an input words represented by means of numbers which is accomplished with word embeddings, in which particular lexical units are repesented as vectors of real numbers. It has been suggested [@Bolukbasi2016man; @Caliskan2017semanticsBiases; @Gonen2019lipstick; @Lauscher2019multidimensional; @Garg2018years; @Manzini2019blackToCriminal]  that in the learning process  such models can learn implicit biases that reflect harmful stereotypical thinking. A large chunk of  the literature on the topic  focuses on  the geometry of the learnt word embeddings;  in particular, unusually low (cosine) proximity of words belonging, intuitively, to a stereotype, is taken as a sign that  this stereotype has been built into a given embedding. This methodology will be in the focus of our paper.

One of the first measures in the discussion has been developed by @Bolukbasi2016man. First, the gender direction $gd$ in  is obtained by taking the differences of the vectors corresponding to ten different gendered pairs (such as $\overrightarrow{she} - \overrightarrow{he}$ or $\overrightarrow{girl} - \overrightarrow{boy}$) and then identifying their principal component (which is the vector obtained by projecting the data points on their  linear combination in a way that maximizes the variance of the projections).^[In the notebook associated with the paper, the authors simply use $\overrightarrow{she} - \overrightarrow{he}$ as a gender direction, though.]   The gender bias of a word $w$ is understood as its projection on the gender direction: $\vec{w} \cdot gd$.    Given the supposedly gender neutral words $N$^[We follow the methodology in assuming that there is a class of words that ideally should be  identified as neutral, such as \emph{ballpark, solution, lecture, science, book} can be identified. We will have a bit to say about this assumption when we describe our dataset construction.] \todo{get back to this!} and the gender direction $gd$    the direct gender bias is defined as the average cosine similarity of the words in $N$ from $gd$ ($c$ is a parameter determining how strict we want to be):

\footnotesize
\begin{align}
\mathsf{directBias_c(N,gd)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},gd)\vert^c}{\vert N \vert }
\end{align}
\normalsize 
The use of projections has been criticized for instance by @Gonen2019lipstick, who point out that while the distance to the  gender direction might be an indicator of bias, it is only one possible manifestation of it, and reducing the cosine distance to such a projection  might be insufficient. For instance, "math" and "delicate" might be in equal distance to a pair of  opposed explicitly gendered words, while being closer to quite different stereotypical attribute words. Further, it is observed in [@Gonen2019lipstick]  that most word pairs preserve similarity under debiasing meant to minimize projection-based bias.^[In [@Bolukbasi2016man]  another method which involves analogies and their evaluations by human users on Mechanical Turk is also used. We do not discuss this method in this paper, see its  criticism in [@Nissim2020fair].]

A measure of bias in word embeddings which does not employ gender directions, the Word Embedding Association Test (WEAT), has been proposed in  [@Caliskan2017semanticsBiases]. The idea is that the   bias between two sets of target words, $X$ and $Y$ (we call them protected words), should be quantified in terms of the cosine similarity  between the protected words and attribute words coming from  two sets of stereotype attribute words, $A$ and $B$ (we'll call them attributes). For instance, $X$ might be a set of male names, $Y$ a set of female names, $A$ might contain stereotypically male-related, and $B$ stereotypically female-related career words. The association difference for a term $t$ is:

\vspace{-2mm}

\footnotesize
\begin{align}
\mathsf{s}(t,A,B) & = \frac{\sum_{a\in A}\mathsf{cos}(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}\mathsf{cos}(t,b)}{\vert B\vert}
\end{align}
\normalsize
\noindent then, the association difference between $A$ a $B$ is:

\vspace{-2mm}

\footnotesize

\begin{align}
\mathsf{s}(X,Y,A,B) & = \sum_{x\in X} \mathsf{s}(x,A,B) -  \sum_{y\in Y} \mathsf{s}(y,A,B)
\end{align}
\normalsize

The effect size is computed by normalizing the difference in means as follows:^[ WEAT is a modification  of the Implicit Association Test (IAT) [@Nosek2002harvesting] used in psychology, and it uses almost the same word sets, allowing for a \emph{prima facie} sensible comparison with bias in humans.  @Caliskan2017semanticsBiases argue that significant biases---thus measured--- similar to the ones discovered by IAT can be discovered in word embeddings. @Lauscher2019multidimensional extended the methodology to a multilingual and cross-lingual setting, arguing that using Euclidean distance instead of similarity does not make much difference, while the bias effects vary greatly across embedding models (interestingly, with social media-text trained embeddings being less biased than those based on Wikipedia).]$^{, }$ ^[
A similar methodology is employed in [@Garg2018years]. The authors employ word embeddings trained on corpora from different decades to study the shifts in various biases. For instance, to compute the occupational embeddings bias for women the authors first compute the average vector of vector embeddings of words that represent women (e.g. \emph{she}, and \emph{female}), then calculate the Euclidean distance between this mean vector and words for occupations. Then they take the mean of these distances and subtract from it the analogously obtained mean for the average vector of vector embeddings of words that represent men. Formally they take the relative norm distance between $X$ and $Y$ to be:

\vspace{-2mm}

\footnotesize

\begin{align}
\textsf{relative norm distance} & = \sum_{v_m\in M} \vert \vert v_m - v_X\vert \vert_2 - \vert v_m - v_Y\vert \vert_2
\end{align}

\normalsize

\noindent where the norm used is Euclidean, and $v_X$ and $x_Y$ are average vectors for sets $X$ and $Y$ respectively. 
]

\vspace{-2mm}

\footnotesize

\begin{align}
\mathsf{weat}(A,B) & = \frac{
\mu(\{\mathsf{s}(x,A,B)\}_{x\in X}) -\mu(\{\mathsf{s}(y,A,B)\}_{y\in Y}) 
}{
\sigma(\{\mathsf{s}(w,A,B)\}_{w\in X\cup Y})
}
\end{align}

\normalsize

WEAT, however, has been developed to investigate biases corresponding to a pair of supposedly opposing stereotypes, and so the question arises as to how generalize the measure to contexts in which biases with respect to more than two stereotypical groups are to be measured. Such a generalization can be found in [@Manzini2019blackToCriminal]. The authors introduce Mean Average Cosine similarity as a measure of bias (strictly speaking, in the paper  they report cosine distances
^[By the cosine distance in the literature the authors mean 1-cosine similarity; note however that this terminology is slightly misleading, as mathematically it is not a distance measure, because it does not satisfy the triangle inequality, as generally
$dist(A,C) \not \leq dist(A,B)+ dist(B,C)$; we'll keep using this mainstream terminology though.]  rather than similarities). Let $T = \{t_1, \dots, t_k\}$ be a class of protected word embeddings, and let each $A_j\in A$ be a set of attributes stereotypically associated with a protected word). Then:

\vspace{-2mm }
\footnotesize

\begin{align}
\mathsf{s}(t_i, A_j) & = \frac{1}{\vert A_j\vert}\sum_{a\in A_j}\mathsf{cos}(t,a) \\
\mathsf{mac}(T,A) & = \frac{1}{\vert T \vert \,\vert A\vert}\sum_{t_i \in T }\sum_{A_j \in A} \mathsf{s}(t_i,A_j)
\end{align}

\normalsize

\noindent That is, for each protected word $T$ and each attribute class, they first take the mean for this protected word and all attributes in a given attribute class, and then take the mean of thus obtained means for all the protected words.

Having introduced the measures,  first, we will introduce a selection of general problems with this approach, and then we will move on to more specific but important problems related to the statistical significance of such measurement. We will focus on WEAT and MAC, as we want to put issues with the use of projections aside. WEAT will be useful in our criticism of the statistical methods involved, as it is simpler, so the explanation and visualizations will be more transparent (and \emph{mutatis mutandis} this criticism will apply to MAC as well), and MAC will be useful in the development of our alternative method as its range of applicability is the widest.

# Non-statistical problems with cosine-based measures of bias

One issue to consider is  the selection of attributes for bias measurement. The word lists used in the literature are often fairly small (5-50)\todo{Check Manzani}. While the papers do employ statistical tests to measure the uncertainty involved, we will later on argue that these method are not  proper for the goal at hand and  show that a more appropriate use of statistical methods leads to  estimates of uncertainty that are rather epistemologically pessimistic.  \todo{sensitivity to choice of words}

Let's think about MAC, using the  case of religion-related stereotypes.\todo{crossref to a list of words an explanation} In the original paper,  words from all three religions were compared against all of the stereotypes. One reason this is problematic is that  no distinction between cases in which the stereotype is associated with a given religion, as opposed to the situation in which it is associated with another one, is made. This is problematic, as  not all of the stereotypical words have to be considered as harmful for all of the religions. One should investigate the religions separately as some of them may have stronger harmful associations that others.

The interpretation of the results is also a challenge. In [@Manzini2019blackToCriminal] we can find  summaries of  average cosine distances per group (such as gender, race, or religion). For instance, for religion, here is the relevant fragment of table:\todo{add p-values in the table}

\todo{kable the table}

<!-- Religion Debiasing  | MAC -->
<!-- ------------- | ------------- -->
<!-- Biased        | 0.859 -->
<!-- Hard Debiased | 0.934 -->
<!-- Soft Debiased ($\lambda$ = 0.2) | 0.894 -->

\noindent (MAC stand for mean average cosine similarity, although in reality the the table contains mean cosine distances). What may attract attention is the fact that the value of cosine distance in "Biased" category is already quite high (i.e. close to 1) even before debiasing. High cosine distance indicates low cosine similarity between values. One could think that average cosine similarity equal to approximately 0.141 is not significant enough to consider it as bias. However,  the authors still aim to mitigate such "bias"  to make the distance even larger. Methodologically the question is, on what basis is this small similarity still considered as a proof of the presence of bias, and whether these small changes are meaningful.

The underlying problem here is that in the paper there is no  control group.  One should also include control groups to have a way of comparing the  results for a supposedly stereotyped group with the results for sets of neutral or human-related neutral words. In our approach later on, we distinguish between stereotypes associated with a given group, stereotypes associated with different groups, and introduce control groups: neutral words and stereotype-free human predicates.

\todo{bliskosc znaczeniowa a ko-okurencja}

\todo{comparison should be made to actual frequencies, to separate bias caused by cosine similarty to upstream bias}

\todo{unclear czy to sie przeklada na downstream effect}

<!-- One may consider the issue of the curse of dimensionality. In our case, the curse of dimensionality may take place when there is an increase in the volume of data that results in adding extra dimensions to the Euclidean space. As the number of features increases, it may be harder and harder to obtain useful information from the data using the available algorithms. Using cosine similarity in high dimensions in word embeddings may be prone to the curse of dimensionality. According to @Venkat2018Curse there are reasons to consider this phenomenon when searching for word similarities in higher dimensions. An experiment is conducted that aims at showing how the similarity values and variation change as the number of dimensions increases. The hypothesis made in the paper states that two things will happen as the number of dimensions increase. First, the effort required to measure cosine similarity will be greater, and two, the similarity between data will blur out and have less variation. The authors generate random points with increasing number of dimensions where each dimension of a data point is given a value between 0 and 1. Then they pick one vector at random from each dimension class and calculate the cosine similarity between the chosen vector and the rest of the data. Then they check how the variation of values changes as the number of dimensions increases. It seems like the more dimensions there are, the smaller the variance and therefore it is less obvious how to interpret the resulting cosine similarities. Maybe the scale should be adjusted to the number of dimensions and variance so that it still gives us sensible information about data. -->

# Statistical issues

In contrast, statistical intervals will help us decide whether a given cosine similarity is high enough to consider the words to be more similar than if we chose them at random. We will use highest posterior density intervals, in line with Bayesian methodology.

Crucially, these approaches use  means of mean average cosine similarities to measure  similarity between protected word and harmful stereotypes. If one takes  a closer look at the individual values that are taken for the calculations, it turns out that there are quite a few  outliers and surprisingly dissimilar words. This issue will become transparent when  we inspect  the visualizations of individual cosine distances, following the idea that one of the first step to understand data is to look at it.

With such a method the uncertainty involved is not really considered which makes it even more difficult to give reasonable interpretations of the results. We propose the use of Bayesian method to obtain some understanding of the influence the uncertainty has on the interpretation of final results.

\todo{odleglosc nie musi uchwytywac relacji semantycznej zeby oceniac bias}

\todo{zalozenie, bardziej ze jezeli odleglosci przekladaja sie na downstream, to warto patrzec na odleglosci, nawet bez glebszej filozoficznej interpretacji ich}

\noindent
$s(X,Y,A,B)$ is the statistic used in the significance test, and the $p$-value is obtained by bootstrapping: it is the frequency of $s(X_i,Y_i,A,B)>s(X,Y,A,B)$ for all equally sized partitions $X_i, Y_i$ of $X\cup Y$. The effect size is computed by normalizing the difference in means as follows:

\vspace{-2mm}

\footnotesize

\begin{align}
bias(A,B) & = \frac{
\mu(\{s(x,A,B)\}_{x\in X}) -\mu(\{s(y,A,B)\}_{y\in Y}) 
}{
\sigma(\{s(w,A,B)\}_{w\in X\cup Y})
}
\end{align}

\normalsize

The t-tests they employ are run on average cosines used to calculate MAC.

### 2022-01-03 paperDraft/paperDraft4.tex
paper 4 started

--- 2021-12-10 paperDraft/paperDraft2.tex
@@
 
+\usepackage{sectsty}
+\sectionfont{\fontsize{11}{11}\selectfont}
+
 % \setbeamertemplate{blocks}[rounded][shadow=true]
@@
 need as an input words represented by means of numbers which is
-accomplished with word embeddings. It seems that in the learning process
-these models can learn implicit biases that reflect harmful
-stereotypical thinking. One of the sources of bias in NLP can be located
-in the way the word embeddings are made. There is a considerable amount
-of literature available on the topic of bias detection and mitigation in
-NLP models.
+accomplished with word embeddings, in which particular lexical units are
+repesented as vectors of real numbers. It has been suggested {[}1--6{]}
+that in the learning process such models can learn implicit biases that
+reflect harmful stereotypical thinking. A large chunk of the literature
+on the topic focuses on the geometry of the learnt word embeddings; in
+particular, unusually low (cosine) proximity of words belonging,
+intuitively, to a stereotype, is taken as a sign that this stereotype
+has been built into a given embedding. This methodology will be in the
+focus of our paper.
 
 One of the first measures in the discussion has been developed by
-Bolukbasi, Chang, Zou, Saligrama, \& Kalai (2016). There, the gender
-bias of a word \(w\) is understood as its projection on the gender
-direction \(\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})\)
-(the gender direction is the top principal compontent of ten gender pair
-difference vectors). The underlying idea is that no bias is present if
-non-explicitly gendered words are in equal distance to both elements in
-all explicitly gender pairs. Given the (ideally) gender neutral words
-\(N\) and the gender direction \(g\) the direct gender bias is defined
-as the average distance of the words in \(N\) from \(g\) (\(c\) is a
-parameter determining how strict we want to be): \begin{align}
-\mathsf{directBias_c(N,g)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},g)\vert^c}{\vert N \vert }
+{[}1{]}. First, the gender direction \(gd\) in is obtained by taking the
+differences of the vectors corresponding to ten different gendered pairs
+(such as \(\overrightarrow{she} - \overrightarrow{he}\) or
+\(\overrightarrow{girl} - \overrightarrow{boy}\)) and then identifying
+their principal component (which is the vector obtained by projecting
+the data points on their linear combination in a way that maximizes the
+variance of the projections).\footnote{In the notebook associated with
+  the paper, the authors simply use
+  \(\overrightarrow{she} - \overrightarrow{he}\) as a gender direction,
+  though.} The gender bias of a word \(w\) is understood as its
+projection on the gender direction: \(\vec{w} \cdot gd\). Given the
+supposedly gender neutral words \(N\)\footnote{We follow the methodology
+  in assuming that there is a class of words that ideally should be
+  identified as neutral, such as
+  \emph{ballpark, solution, lecture, science, book} can be identified.
+  We will have a bit to say about this assumption when we describe our
+  dataset construction.} \todo{get back to this!} and the gender
+direction \(gd\) the direct gender bias is defined as the average cosine
+similarity of the words in \(N\) from \(gd\) (\(c\) is a parameter
+determining how strict we want to be):
+
+\footnotesize
+
+\begin{align}
+\mathsf{directBias_c(N,gd)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},gd)\vert^c}{\vert N \vert }
+\end{align} \normalsize  The use of projections has been criticized for
+instance by {[}4{]}, who point out that while the distance to the gender
+direction might be an indicator of bias, it is only one possible
+manifestation of it, and reducing the cosine distance to such a
+projection might be insufficient. For instance, ``math'' and
+``delicate'' might be in equal distance to a pair of opposed explicitly
+gendered words, while being closer to quite different stereotypical
+attribute words. Further, it is observed in {[}4{]} that most word pairs
+preserve similarity under debiasing meant to minimize projection-based
+bias.\footnote{In {[}1{]} another method which involves analogies and
+  their evaluations by human users on Mechanical Turk is also used. We
+  do not discuss this method in this paper, see its criticism in
+  {[}7{]}.}
+
+A measure of bias in word embeddings which does not employ gender
+directions, the Word Embedding Association Test (WEAT), has been
+proposed in {[}2{]}. The idea is that the bias between two sets of
+target words, \(X\) and \(Y\) (we call them protected words), should be
+quantified in terms of the cosine similarity between the protected words
+and attribute words coming from two sets of stereotype attribute words,
+\(A\) and \(B\) (we'll call them attributes). For instance, \(X\) might
+be a set of male names, \(Y\) a set of female names, \(A\) might contain
+stereotypically male-related, and \(B\) stereotypically female-related
+career words. The association difference for a term \(t\) is:
+
+\vspace{-2mm}
+
+\footnotesize
+
+\begin{align}
+\mathsf{s}(t,A,B) & = \frac{\sum_{a\in A}\mathsf{cos}(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}\mathsf{cos}(t,b)}{\vert B\vert}
+\end{align} \normalsize \noindent then, the association difference
+between \(A\) a \(B\) is:
+
+\vspace{-2mm}
+
+\footnotesize
+
+\begin{align}
+\mathsf{s}(X,Y,A,B) & = \sum_{x\in X} \mathsf{s}(x,A,B) -  \sum_{y\in Y} \mathsf{s}(y,A,B)
+\end{align} \normalsize
+
+The effect size is computed by normalizing the difference in means as
+follows:\footnote{ WEAT is a modification of the Implicit Association
+  Test (IAT) {[}8{]} used in psychology, and it uses almost the same
+  word sets, allowing for a \emph{prima facie} sensible comparison with
+  bias in humans. {[}2{]} argue that significant biases---thus
+  measured--- similar to the ones discovered by IAT can be discovered in
+  word embeddings. {[}5{]} extended the methodology to a multilingual
+  and cross-lingual setting, arguing that using Euclidean distance
+  instead of similarity does not make much difference, while the bias
+  effects vary greatly across embedding models (interestingly, with
+  social media-text trained embeddings being less biased than those
+  based on Wikipedia).}\(^{, }\)\footnote{ A similar methodology is
+  employed in {[}3{]}. The authors employ word embeddings trained on
+  corpora from different decades to study the shifts in various biases.
+  For instance, to compute the occupational embeddings bias for women
+  the authors first compute the average vector of vector embeddings of
+  words that represent women (e.g.~\emph{she}, and \emph{female}), then
+  calculate the Euclidean distance between this mean vector and words
+  for occupations. Then they take the mean of these distances and
+  subtract from it the analogously obtained mean for the average vector
+  of vector embeddings of words that represent men. Formally they take
+  the relative norm distance between \(X\) and \(Y\) to be:}
+
+\vspace{-2mm}
+
+\footnotesize
+
+\begin{align}
+\mathsf{weat}(A,B) & = \frac{
+\mu(\{\mathsf{s}(x,A,B)\}_{x\in X}) -\mu(\{\mathsf{s}(y,A,B)\}_{y\in Y}) 
+}{
+\sigma(\{\mathsf{s}(w,A,B)\}_{w\in X\cup Y})
+}
 \end{align}
 
-The use of projections has been ciriticized for instance by Gonen \&
-Goldberg (2019), who point out that while gender-direction might be an
-indicator of bias, it is only one possible manifestation of it, and
-reducing a projection of words might be insufficient. For instance,
-``math'' and ``delicate'' might be in equal distance to both explicitly
-gendered words while being closer to quite different stereotypical
-attribute words. Further, the authors point out that most word pairs
-preserve similarity under debiasing meant to minimize projection-based
-bias.\footnote{Bolukbasi et al. (2016) use also another method which
-  involves analogies and their evaluations by human users on Mechanical
-  Turk. It is discussed and criticized in (Nissim, Noord, \& Goot,
-  2020).}
+\normalsize
 
-To measure bias in word embeddings, Caliskan, Bryson, \& Narayanan
-(2017) proposed the Word Embedding Association Test (WEAT). The idea is
-that the measure of biases between two sets of target words, \(X\) and
-\(Y\), (we call them protected words) should be quantified in terms of
-the cosine similarity between the protected words and attribute words
-coming from two sets of stereotype attribute words, \(A\) and \(B\)
-(we'll call them attributes). For instance, \(X\) might be a set of male
-names, \(Y\) a set of female names, \(A\) might contain stereotypically
-male-related career words, and \(B\) stereotypically female-related
-family words. WEAT is a modification of the Implicit Association Test
-(IAT) (Nosek, Banaji, \& Greenwald, 2002) used in psychology and uses
-almost the same word sets, allowing for a \emph{prima facie} sensible
-comparison with bias in humans. If the person's attitude towards given
-pair of concept is to be interpreted as neutral, there should be no
-noticeable task completion time difference, and the final value from the
-formula should be around 0. Let \(f\) be a similarity measure (usually,
-cosine similarity). The association difference for a term \(t\) is:
+WEAT, however, has been developed to investigate biases corresponding to
+a pair of supposedly opposing stereotypes, and so the question arises as
+to how generalize the measure to contexts in which biases with respect
+to more than two stereotypical groups are to be measured. Such a
+generalization can be found in {[}6{]}. The authors introduce Mean
+Average Cosine similarity as a measure of bias (strictly speaking, in
+the paper they report cosine distances\footnote{By the cosine distance
+  in the literature the authors mean 1-cosine similarity; note however
+  that this terminology is slightly misleading, as mathematically it is
+  not a distance measure, because it does not satisfy the triangle
+  inequality, as generally \(dist(A,C) \not \leq dist(A,B)+ dist(B,C)\);
+  we'll keep using this mainstream terminology though.} rather than
+similarities). Let \(T = \{t_1, \dots, t_k\}\) be a class of protected
+word embeddings, and let each \(A_j\in A\) be a set of attributes
+stereotypically associated with a protected word). Then:
+
+\vspace{-2mm }
+\footnotesize
+
 \begin{align}
-s(t,A,B) & = \frac{\sum_{a\in A}f(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}f(t,b)}{\vert B\vert}
-\end{align} \noindent then, the association difference between \(A\) a
-\(B\) is: \begin{align}
-s(X,Y,A,B) & = \sum_{x\in X} s(x,A,B) -  \sum_{y\in Y} s(y,A,B)
-\end{align} \noindent \(s(X,Y,A,B)\) is the statistic used in the
-significance test, and the \(p\)-value obtained by bootstrapping: it is
-the frequency of \(s(X_i,Y_i,A,B)>s(X,Y,A,B)\) for all equally sized
-partitions \(X_i, Y_i\) of \(X\cup Y\). The effect size is computed by
-normalizing the difference in means as follows: \begin{align}
+\mathsf{s}(t_i, A_j) & = \frac{1}{\vert A_j\vert}\sum_{a\in A_j}\mathsf{cos}(t,a) \\
+\mathsf{mac}(T,A) & = \frac{1}{\vert T \vert \,\vert A\vert}\sum_{t_i \in T }\sum_{A_j \in A} \mathsf{s}(t_i,A_j)
+\end{align}
+
+\normalsize
+
+\noindent That is, for each protected word \(T\) and each attribute
+class, they first take the mean for this protected word and all
+attributes in a given attribute class, and then take the mean of thus
+obtained means for all the protected words.
+
+Having introduced the measures, first, we will introduce a selection of
+general problems with this approach, and then we will move on to more
+specific but important problems related to the statistical significance
+of such measurement. We will focus on WEAT and MAC, as we want to put
+issues with the use of projections aside. WEAT will be useful in our
+criticism of the statistical methods involved, as it is simpler, so the
+explanation and visualizations will be more transparent (and
+\emph{mutatis mutandis} this criticism will apply to MAC as well), and
+MAC will be useful in the development of our alternative method as its
+range of applicability is the widest.
+
+\hypertarget{non-statistical-problems-with-cosine-based-measures-of-bias}{%
+\section{Non-statistical problems with cosine-based measures of
+bias}\label{non-statistical-problems-with-cosine-based-measures-of-bias}}
+
+One issue to consider is the selection of attributes for bias
+measurement. The word lists used in the literature are often fairly
+small (5-50)\todo{Check Manzani}. While the papers do employ statistical
+tests to measure the uncertainty involved, we will later on argue that
+these method are not proper for the goal at hand and show that a more
+appropriate use of statistical methods leads to estimates of uncertainty
+that are rather epistemologically pessimistic.
+\todo{sensitivity to choice of words}
+
+Let's think about MAC, using the case of religion-related
+stereotypes.\todo{crossref to a list of words an explanation} In the
+original paper, words from all three religions were compared against all
+of the stereotypes. One reason this is problematic is that no
+distinction between cases in which the stereotype is associated with a
+given religion, as opposed to the situation in which it is associated
+with another one, is made. This is problematic, as not all of the
+stereotypical words have to be considered as harmful for all of the
+religions. One should investigate the religions separately as some of
+them may have stronger harmful associations that others.
+
+The interpretation of the results is also a challenge. In {[}6{]} we can
+find summaries of average cosine distances per group (such as gender,
+race, or religion). For instance, for religion, here is the relevant
+fragment of table:\todo{add p-values in the table}
+
+\todo{kable the table}
+
+\noindent (MAC stand for mean average cosine similarity, although in
+reality the the table contains mean cosine distances). What may attract
+attention is the fact that the value of cosine distance in ``Biased''
+category is already quite high (i.e.~close to 1) even before debiasing.
+High cosine distance indicates low cosine similarity between values. One
+could think that average cosine similarity equal to approximately 0.141
+is not significant enough to consider it as bias. However, the authors
+still aim to mitigate such ``bias'' to make the distance even larger.
+Methodologically the question is, on what basis is this small similarity
+still considered as a proof of the presence of bias, and whether these
+small changes are meaningful.
+
+The underlying problem here is that in the paper there is no control
+group. One should also include control groups to have a way of comparing
+the results for a supposedly stereotyped group with the results for sets
+of neutral or human-related neutral words. In our approach later on, we
+distinguish between stereotypes associated with a given group,
+stereotypes associated with different groups, and introduce control
+groups: neutral words and stereotype-free human predicates.
+
+\todo{bliskosc znaczeniowa a ko-okurencja}
+
+\todo{comparison should be made to actual frequencies, to separate bias caused by cosine similarty to upstream bias}
+
+\todo{unclear czy to sie przeklada na downstream effect}
+
+\hypertarget{statistical-issues}{%
+\section{Statistical issues}\label{statistical-issues}}
+
+In contrast, statistical intervals will help us decide whether a given
+cosine similarity is high enough to consider the words to be more
+similar than if we chose them at random. We will use highest posterior
+density intervals, in line with Bayesian methodology.
+
+Crucially, these approaches use means of mean average cosine
+similarities to measure similarity between protected word and harmful
+stereotypes. If one takes a closer look at the individual values that
+are taken for the calculations, it turns out that there are quite a few
+outliers and surprisingly dissimilar words. This issue will become
+transparent when we inspect the visualizations of individual cosine
+distances, following the idea that one of the first step to understand
+data is to look at it.
+
+With such a method the uncertainty involved is not really considered
+which makes it even more difficult to give reasonable interpretations of
+the results. We propose the use of Bayesian method to obtain some
+understanding of the influence the uncertainty has on the interpretation
+of final results.
+
+\todo{odleglosc nie musi uchwytywac relacji semantycznej zeby oceniac bias}
+
+\todo{zalozenie, bardziej ze jezeli odleglosci przekladaja sie na downstream, to warto patrzec na odleglosci, nawet bez glebszej filozoficznej interpretacji ich}
+
+\noindent \(s(X,Y,A,B)\) is the statistic used in the significance test,
+and the \(p\)-value is obtained by bootstrapping: it is the frequency of
+\(s(X_i,Y_i,A,B)>s(X,Y,A,B)\) for all equally sized partitions
+\(X_i, Y_i\) of \(X\cup Y\). The effect size is computed by normalizing
+the difference in means as follows:
+
+\vspace{-2mm}
+
+\footnotesize
+
+\begin{align}
 bias(A,B) & = \frac{
@@
 
-Caliskan et al. (2017) show that significant biases---thus measured---
-similar to the ones discovered by IAT can be discovered in word
-embeddings. Lauscher \& Glavas (2019) extended the methodology to a
-multilingual and cross-lingual setting, arguing that using Euclidean
-distance instead of similarity does not make much difference, while the
-bias effects vary greatly across embedding models (interestingly, with
-social media-text trained embeddings being less biased than those based
-on Wikipedia).
+\normalsize
 
-A similar methodology is employed by Garg, Schiebinger, Jurafsky, \& Zou
-(2018), who employ word embeddings trained on corpora from different
-decades to study the shifts in various biases. For instance, to compute
-the occupational embeddings bias for women the authors first compute the
-average vector of vector embeddings of words that represent women
-(e.g.~``she,'' ``female''), then calculate the Euclidean distance
-between this mean vector and words for occupations. Then they take the
-mean of these distances and subtract from it the analogously obtained
-mean for the average vector of vector embeddings of words that represent
-men. Formally they take the relative norm distance between \(X\) and
-\(Y\) to be: \begin{align}
-\textsf{relative norm distance} & = \sum_{v_m\in M} \vert \vert v_m - v_X\vert \vert_2 - \vert v_m - v_Y\vert \vert_2
-\end{align} \noindent where the norm used is Euclidean, and \(v_X\) and
-\(x_Y\) are average vectors for sets \(X\) and \(Y\) respectively.
-
-Manzini, Lim, Tsvetkov, \& Black (2019) modify WEAT to a multi-class
-setting, introducing Mean Average Cosine similarity as a measure of bias
-(in fact, in the paper they report distances rather than similarities).
-Let \(T = \{t_1, \dots, t_k\}\) be a class of protected word embeddings,
-and let each \(A_j\in A\) be a set of attributes stereotypically
-associated with a protected word). Then: \begin{align}
-S(t_i, A_j) & = \frac{1}{\vert A_j\vert}\sum_{a\in A_j}\mathsf{cos}(t,a) \\
-MAC(T,A) & = \frac{1}{\vert T \vert \,\vert A\vert}\sum_{t_i \in T }\sum_{A_j \in A} S(t_i,A_j)
-\end{align} That is, for each protected word \$T and each attribute
-class, they first take the mean for this protected word and all
-attributes in a given attribute class, and then take the mean of thus
-obtained means for all the protected words. The t-tests they employ are
-run on average cosines used to calculate MAC.
-
-\hypertarget{references}{%
-\section*{References}\label{references}}
-\addcontentsline{toc}{section}{References}
+The t-tests they employ are run on average cosines used to calculate
+MAC.
 
 \hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-bolukbasi2016man}{}}%
-Bolukbasi, T., Chang, K.-W., Zou, J., Saligrama, V., \& Kalai, A.
-(2016). \emph{Man is to computer programmer as woman is to homemaker?
-Debiasing word embeddings}. Retrieved from
-\url{https://arxiv.org/abs/1607.06520}
+\begin{CSLReferences}{0}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-Bolukbasi2016man}{}}%
+\CSLLeftMargin{{[}1{]} }
+\CSLRightInline{Tolga Bolukbasi, Kai-Wei Chang, James Y. Zou, Venkatesh
+Saligrama, and Adam Kalai. 2016. Man is to computer programmer as woman
+is to homemaker? Debiasing word embeddings. \emph{CoRR} abs/1607.06520,
+(2016). Retrieved from \url{http://arxiv.org/abs/1607.06520}}
 
 \leavevmode\vadjust pre{\hypertarget{ref-Caliskan2017semanticsBiases}{}}%
-Caliskan, A., Bryson, J. J., \& Narayanan, A. (2017). Semantics derived
-automatically from language corpora contain human-like biases.
-\emph{Science}, \emph{356}(6334), 183--186.
-\url{https://doi.org/10.1126/science.aal4230}
+\CSLLeftMargin{{[}2{]} }
+\CSLRightInline{Aylin Caliskan, Joanna J. Bryson, and Arvind Narayanan.
+2017. Semantics derived automatically from language corpora contain
+human-like biases. \emph{Science} 356, 6334 (April 2017), 183--186.
+DOI:https://doi.org/\href{https://doi.org/10.1126/science.aal4230}{10.1126/science.aal4230}}
 
 \leavevmode\vadjust pre{\hypertarget{ref-Garg2018years}{}}%
-Garg, N., Schiebinger, L., Jurafsky, D., \& Zou, J. (2018). Word
-embeddings quantify 100 years of gender and ethnic stereotypes.
-\emph{Proceedings of the National Academy of Sciences}, \emph{115}(16),
-E3635--E3644. \url{https://doi.org/10.1073/pnas.1720347115}
+\CSLLeftMargin{{[}3{]} }
+\CSLRightInline{Nikhil Garg, Londa Schiebinger, Dan Jurafsky, and James
+Zou. 2018. Word embeddings quantify 100 years of gender and ethnic
+stereotypes. \emph{Proceedings of the National Academy of Sciences} 115,
+16 (April 2018), E3635--E3644.
+DOI:https://doi.org/\href{https://doi.org/10.1073/pnas.1720347115}{10.1073/pnas.1720347115}}
 
 \leavevmode\vadjust pre{\hypertarget{ref-Gonen2019lipstick}{}}%
-Gonen, H., \& Goldberg, Y. (2019). Lipstick on a pig: {D}ebiasing
-methods cover up systematic gender biases in word embeddings but do not
-remove them. \emph{Proceedings of the 2019 Conference of the North
-{A}merican Chapter of the Association for Computational Linguistics:
-Human Language Technologies, Volume 1 (Long and Short Papers)},
-609--614. Minneapolis, Minnesota: Association for Computational
-Linguistics. \url{https://doi.org/10.18653/v1/N19-1061}
+\CSLLeftMargin{{[}4{]} }
+\CSLRightInline{Hila Gonen and Yoav Goldberg. 2019. Lipstick on a pig:
+{D}ebiasing methods cover up systematic gender biases in word embeddings
+but do not remove them. In \emph{Proceedings of the 2019 conference of
+the north {A}merican chapter of the association for computational
+linguistics: Human language technologies, volume 1 (long and short
+papers)}, Association for Computational Linguistics, Minneapolis,
+Minnesota, 609--614.
+DOI:https://doi.org/\href{https://doi.org/10.18653/v1/N19-1061}{10.18653/v1/N19-1061}}
 
 \leavevmode\vadjust pre{\hypertarget{ref-Lauscher2019multidimensional}{}}%
-Lauscher, A., \& Glavas, G. (2019). Are we consistently biased?
-Multidimensional analysis of biases in distributional word vectors.
-\emph{CoRR}, \emph{abs/1904.11783}. Retrieved from
-\url{http://arxiv.org/abs/1904.11783}
+\CSLLeftMargin{{[}5{]} }
+\CSLRightInline{Anne Lauscher and Goran Glavas. 2019. Are we
+consistently biased? Multidimensional analysis of biases in
+distributional word vectors. \emph{CoRR} abs/1904.11783, (2019).
+Retrieved from \url{http://arxiv.org/abs/1904.11783}}
 
 \leavevmode\vadjust pre{\hypertarget{ref-Manzini2019blackToCriminal}{}}%
-Manzini, T., Lim, Y. C., Tsvetkov, Y., \& Black, A. W. (2019).
-\emph{Black is to criminal as caucasian is to police: Detecting and
-removing multiclass bias in word embeddings}. Retrieved from
-\url{https://arxiv.org/abs/1904.04047}
+\CSLLeftMargin{{[}6{]} }
+\CSLRightInline{Thomas Manzini, Yao Chong Lim, Yulia Tsvetkov, and Alan
+W Black. 2019. Black is to criminal as caucasian is to police: Detecting
+and removing multiclass bias in word embeddings. Retrieved from
+\url{https://arxiv.org/abs/1904.04047}}
 
 \leavevmode\vadjust pre{\hypertarget{ref-Nissim2020fair}{}}%
-Nissim, M., Noord, R. van, \& Goot, R. van der. (2020). Fair is better
-than sensational: Man is to doctor as woman is to doctor.
-\emph{Computational Linguistics}, \emph{46}(2), 487--497.
-\url{https://doi.org/10.1162/coli_a_00379}
+\CSLLeftMargin{{[}7{]} }
+\CSLRightInline{Malvina Nissim, Rik van Noord, and Rob van der Goot.
+2020. Fair is better than sensational: Man is to doctor as woman is to
+doctor. \emph{Computational Linguistics} 46, 2 (June 2020), 487--497.
+DOI:https://doi.org/\href{https://doi.org/10.1162/coli_a_00379}{10.1162/coli\_a\_00379}}
 
 \leavevmode\vadjust pre{\hypertarget{ref-Nosek2002harvesting}{}}%
-Nosek, B. A., Banaji, M. R., \& Greenwald, A. G. (2002). Harvesting
-implicit group attitudes and beliefs from a demonstration web site.
-\emph{Group Dynamics: Theory, Research, and Practice}, \emph{6}(1),
-101--115. \url{https://doi.org/10.1037/1089-2699.6.1.101}
+\CSLLeftMargin{{[}8{]} }
+\CSLRightInline{Brian A. Nosek, Mahzarin R. Banaji, and Anthony G.
+Greenwald. 2002. Harvesting implicit group attitudes and beliefs from a
+demonstration web site. \emph{Group Dynamics: Theory, Research, and
+Practice} 6, 1 (2002), 101--115.
+DOI:https://doi.org/\href{https://doi.org/10.1037/1089-2699.6.1.101}{10.1037/1089-2699.6.1.101}}
 

### 2022-02-27 cosineAnalysisNew/controlCleaning.R
added folder with old scripts

@@ -1,0 +1,140 @@
+controlReddit <- read.csv("cosineAnalysis/datasets/controlReddit.csv")[-1]
+controlGoogle <- read.csv("cosineAnalysis/datasets/controlGoogle.csv")[-1]
+
+
+controlRedditClean <- cleanDataset(controlReddit, c("asian", "black", "caucasian", "christian", "human", "jew", "man", "muslim", "neutral", "woman"))
+
+controlGoogleClean <- cleanDataset(controlGoogle, c("asian", "black", "caucasian", "christian", "human", "jew", "man", "muslim", "neutral", "woman"))
+
+humans <- controlRedditClean %>% filter(connection =="human")
+humanWords <- droplevels(unique(humans$protectedWord))
+
+
+reddit <- controlRedditClean[,1:(length(controlRedditClean)-4)]
+google <- controlGoogleClean[,1:(length(controlGoogleClean)-4)]
+
+
+reddit$humanProtected <- ifelse(reddit$protectedWord %in% humanWords, 1, 0)
+google$humanProtected <- ifelse(reddit$protectedWord %in% humanWords, 1, 0)
+
+
+reddit$humanPredicate <- ifelse(reddit$wordClass == "human", 1, 0)
+reddit$activePredicate <- ifelse(reddit$wordClass != "neutral" & reddit$wordClass != "human", 1, 0)
+reddit$asianPredicate <- ifelse(reddit$wordClass == "asian", 1, 0)
+reddit$blackPredicate <- ifelse(reddit$wordClass == "black", 1, 0)
+reddit$caucasianPredicate <- ifelse(reddit$wordClass == "caucasian", 1, 0)
+reddit$christianPredicate <- ifelse(reddit$wordClass == "christian", 1, 0)
+reddit$jewPredicate <- ifelse(reddit$wordClass == "jew", 1, 0)
+reddit$manPredicate <- ifelse(reddit$wordClass == "man", 1, 0)
+reddit$muslimPredicate <- ifelse(reddit$wordClass == "muslim", 1, 0)
+reddit$womanPredicate <- ifelse(reddit$wordClass == "woman", 1, 0)
+
+
+google$humanPredicate <- ifelse(google$wordClass == "human", 1, 0)
[…]

### 2022-02-27 cosineAnalysisNew/hierarchicalModels.R

@@ -1,0 +1,59 @@
+
+library(rethinking)
+source("cosineAnalysis/functions/cleanDataset.R")
+
+
+religion <- read.csv("cosineAnalysis/datasets/religionReddit.csv")[-1]
+
+#religion <- cleanDataset(religion,c("christian","human","jewish","muslim","neutral"))
+
+
+
+
+
+religion.adh.st <- map2stan(
+  alist(
+    cosineDistance ~ dnorm(mu,sigma),
+    mu <- no + a * associated  + d * different + h * human,
+    no ~ dnorm(1,.5),
+    a ~ dnorm(0,1),
+    d ~ dnorm(0,1),
+    h ~ dnorm(0,1),
+    sigma ~ dcauchy(0,1)
+  ),
+  data = religion,
+  chains=2 , iter=4000 , warmup=1000,
+  start=list(no = 1, a = 0, d = 0, sigma= .3)
+)
+
+
+
+
+
+
+religion$pw <- coerce_index(religion$protectedWord)
+
+
+
+religion.Separate <- map2stan(
+  alist(
+    cosineDistance ~ dnorm(mu,sigma),
+    mu <- no + a[pw] * associated,
+    a[pw] ~ dnorm(0,1),
+    sigma ~ dcauchy(0,1),
+    no ~ dnorm(1,.5)
+  ),
+  data = religion,
+  chains=2 , iter=4000 , warmup=1000,
+  start=list(no = 1, a = 0, d = 0, sigma= .3)
+)
+
+
+
+
+
+#+ a * associated,
+# d[pw] ~ dnorm(0,1),
+# h[pw] ~ dnorm(0,1),
+# + d * different + h * human,
+#,

### 2022-02-27 cosineAnalysisNew/functions/cleanDataset.R
adding old functions

@@ -1,0 +1,21 @@
+#gender <- read.csv("cosineAnalysis/datasets/genderOriginal.csv")[,-1]
+#genderClean <- cleanDataset(gender, c("man","neutral","woman"))
+
+
+
+cleanDataset <- function(dataset, levelsWordClass){
+
+    colnames(dataset) <- c("protectedWord","wordToCompare","wordClass",
+                      "cosineDistance","cosineSimilarity","connection")
+
+    levels(dataset$wordClass) <- levelsWordClass
+
+    dataset$con <- as.integer(dataset$connection)
+    dataset$pw <- as.integer(dataset$protectedWord)
+    dataset$pwFactor <- factor(paste0(dataset$protectedWord, "-", dataset$connection))
+    dataset$pwIndex <- as.integer(dataset$pwFactor)
+#    dataset$associated <- ifelse(dataset$connection == "associated", 1, 0)
+#    dataset$different <- ifelse(dataset$connection == "different", 1, 0)
+#    dataset$human <- ifelse(dataset$connection == "human", 1, 0)
+    return(dataset)
+    }

### 2022-02-27 cosineAnalysisNew/functions/visualiseStats.R

@@ -1,0 +1,75 @@
+library(dplyr)
+library(forcats)
+
+library(stringr)
+
+#setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
+#load("cosineAnalysis/datasets/religionOriginalTable.RData")  
+
+
+#dataset <- religionOriginalTable
+
+
+visualiseTable <- function(dataset, orderVector = a){
+parameterPlot <-   dataset %>% mutate(protectedClass = fct_reorder(protectedClass, a)) %>%
+  ggplot()+
+  geom_point(aes(x = protectedClass,y=a,color = "associated"))+
+  geom_segment(aes(x=protectedClass,xend=protectedClass,y=alo,yend=ahigh), alpha = 0.3,
+               color = "chartreuse4")+
+  geom_segment(data = dataset[dataset$protectedClass == "full dataset",],aes(x=protectedClass, xend = protectedClass,y=min(dataset$alo),yend = max(dataset$ahigh)), size = 9, color = "red",
+                alpha = 0.1)+geom_hline(yintercept = 0, size = 0.15, lty = 2)+
+  geom_point(aes(x=protectedClass, y = d,color = "different"),position = position_nudge(0.15))+
+  geom_segment(aes(x=protectedClass,xend=protectedClass,y=dlo,yend=dhigh, color ="different"), alpha = 0.3,
+                position = position_nudge(0.15))+
+  geom_point(aes(x=protectedClass, y = h,color = "human"),position = position_nudge(-0.15) )+
+  geom_segment(aes(x=protectedClass,xend=protectedClass,y=hlo,yend=hhigh, color = "human"), alpha = 0.3,    position = position_nudge(-0.15))+
+  scale_color_manual(values = c("chartreuse4","orangered4","grey"))+
+  coord_flip()+theme_tufte()+xlab("protected class")+ylab("class coefficient")+ 
+  labs(color = "connection")
+print(parameterPlot)
+}
+
+
+visualisePWestimates <- function(data, model){
+  pr <- as.data.frame(precis(model, depth = 2))
+  pr$tags <- c(levels(data$pwFactor), "sigma")
+  tags <- pr$tags
+  pr <- cbind(pr,str_split_fixed(tags, "-", 2))
+  pr <- pr[-nrow(pr),]
+  colnames(pr) <- c("mean", "sd", "low", "high", "neff", "rhat", "tags", "pw", "connection")
+  dataset <- pr
+  
[…]

### 2022-02-27 presentations/presentationBoston/presentationBoston.Rmd
getting started with the presentation

--- 2021-08-10 presentationESSLLI21/presentationESSLLI.Rmd
@@
   group, University of  Gdansk)
-date: "ExpSem2021, ESSLLI"
+date: "Boston, April Fools' Day"
 output:
@@
 urlcolor: blue
-bibliography: ../references/cosineReferences1.bib
-csl: ../references/apa-6th-edition.csl
+bibliography: ../../references/cosineReferences1.bib
+csl: ../../references/apa-6th-edition.csl
 link-citations: yes
@@
 ```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "60%"}
-genderProjections <-  read.csv("../datasets/genderProjections.csv")[,-1]
+genderProjections <-  read.csv("datasets/genderProjections.csv")[,-1]
 colnames(genderProjections)[4] <- "Stereotype" 
@@
 ```{r religionTableHeadEarly,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
-religion <- read.csv("../datasets/religionReddit.csv")[-1]
+religion <- read.csv("datasets/religionReddit.csv")[-1]
 colnames(religion) <- c("protectedWord","wordToCompare","wordClass",
@@
 ```{r religionTableHeadLate,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
-religion <- read.csv("../datasets/religionReddit.csv")[-1]
+religion <- read.csv("datasets/religionReddit.csv")[-1]
 colnames(religion) <- c("protectedWord","wordToCompare","wordClass",
@@
 #we plug in our visualisation script
-source("../functions/visualisationTools.R")
+source("../../functions/visualisationTools.R")
 #two arguments: dataset and protected word
@@
    \begin{minipage}{0.55\textwidth}
-  \includegraphics[width=6cm]{../images/religionCoeffs.jpeg}
+  \includegraphics[width=6cm]{../../images/religionCoeffs.jpeg}
    \end{minipage} \begin{minipage}{0.4\textwidth}\footnotesize
@@
   \begin{minipage}{0.55\textwidth}
-\includegraphics[width=6cm]{../images/genderCoeffs.jpeg}
+\includegraphics[width=6cm]{../../images/genderCoeffs.jpeg}
 \end{minipage}
@@
   \begin{minipage}{0.55\textwidth}
-\includegraphics[width=7.5cm]{../images/visReligionReddit.png}
+\includegraphics[width=7.5cm]{../../images/visReligionReddit.png}
 \end{minipage}
@@
   \begin{minipage}{0.55\textwidth}
-\includegraphics[width=7.5cm]{../images/visGenderReddit.png}
+\includegraphics[width=7.5cm]{../../images/visGenderReddit.png}
 \end{minipage}
@@
   \begin{minipage}{0.55\textwidth}
-\includegraphics[width=7.5cm]{../images/visRaceReddit.png}
+\includegraphics[width=7.5cm]{../../images/visRaceReddit.png}
 \end{minipage}

### 2022-02-27 functions/rafalTestsModels.R

@@ -1,0 +1,69 @@
+library(rethinking)
+library(tidyverse)
+library(stringr)
+library(ggthemes)
+
+
+#clean dataset
+cleanDataset <- function(dataset, levelsWordClass){
+  
+  colnames(dataset) <- c("protectedWord","wordToCompare","wordClass",
+                         "cosineDistance","cosineSimilarity","connection")
+  levels(dataset$wordClass) <- levelsWordClass
+  
+  dataset$conFactor <- factor(paste0(dataset$connection))
+  dataset$con <- as.integer(dataset$conFactor)
+  
+  dataset$pw <- as.factor(dataset$protectedWord)
+  dataset$pw <- as.integer(dataset$pw)
+  
+  dataset$pwConFactor <- factor(paste0(dataset$protectedWord, "-", dataset$connection))
+  dataset$pwConIndex <- as.integer(dataset$pwConFactor)
+  #   dataset$associated <- ifelse(dataset$connection == "associated", 1, 0)
+  #   dataset$different <- ifelse(dataset$connection == "different", 1, 0)
+  #   dataset$human <- ifelse(dataset$connection == "human", 1, 0)
+  return(dataset)
+}
+
+
+#build model
+buildModel <- function(dataset, mm = 1, sm =.5, mco =0 , sco =.5){
+  options(buildtools.check = function(action) TRUE )
+  d <- list()
+  d$cosineDistance <- data2$cosineDistance
+  d$pw <- data2$pw
+  d$con <- data2$con
+  model <- ulam(
+    alist(
+      cosineDistance ~ dnorm(mu,sigma),
+      mu <- m[pw] + co[con],
+      m[pw] ~ dnorm(1,.5),
+      co[con] ~dnorm(0,.5),
+      sigma ~ dcauchy(0,1)
+    ),
+    data = d,
+    chains=2 , iter=80000 , warmup=1000, 
+    log_lik = TRUE
+  )
+  return(model)
+}
+
+
+
+#test on a small dataset
+
+religionReddit <- read.csv("./datasets/religionReddit.csv")[-1]
+religionReddit <- cleanDataset(religionReddit,c("christian","human","jewish","muslim","neutral"))
+
+head(religionReddit)
+
+set.seed(123)
+data2 <- sample_n(religionReddit, 300)
+data2$pw
+
+model <- buildModel(data2)
+
+precis(model, depth = 2)
+
+
+

### 2022-03-01 functions/caliskanExtended.R
Adds new simulation

@@ -60,6 +60,7 @@ numerator <-  mean(s(t1),s(t2),s(t3),s(t4),s(t5),s(t6),s(t7),s(t8)) - mean(s(t9)
 print(list(factor = factor,numerator = numerator, bias = numerator / factor))
 
 
+# SIMULATION FOR SAME MEANS AND SMALL STD
 biasesD1 <- numeric(10000)
 
 for(i in 1:10000){
@@ -79,10 +80,6 @@ for(i in 1:10000){
   t14 <- data.frame(A  = rnorm(8,0,0.05), B = rnorm(8,0,0.05))
   t15 <- data.frame(A  = rnorm(8,0,0.05), B = rnorm(8,0,0.05))
   t16 <- data.frame(A  = rnorm(8,0,0.05), B = rnorm(8,0,0.05))
-  t1d1 <- data.frame(A  = rnorm(5,.1,0.1), B = rnorm(5,0,0.1))
-  t2d1 <- data.frame(A  = rnorm(5,.1,0.1), B = rnorm(5,0,0.1))
-  t3d1 <- data.frame(A  = rnorm(5,0,0.1), B = rnorm(5,.1,0.1))
-  t4d1 <- data.frame(A  = rnorm(5,0,0.1), B = rnorm(5,.1,0.1))
   
   factorD1 <- sd(c(s(t1),s(t2),s(t3),s(t4),s(t5),s(t6),s(t7),s(t8),s(t9),s(t10),s(t11),s(t12),s(t13),s(t14),s(t15),s(t16)))
   numeratorD1 <-  mean(s(t1),s(t2),s(t3),s(t4),s(t5),s(t6),s(t7),s(t8)) - mean(s(t9),s(t10),s(t11),s(t12),s(t13),s(t14),s(t15),s(t16))
@@ -90,15 +87,67 @@ for(i in 1:10000){
 }
 
 ggplot()+geom_histogram(aes(x=biasesD1, y = ..density..), alpha = 0.6, bins=50)+
-  theme_tufte()+labs(title="10k biases for different means and sd =.05")+ xlab("bias")
-
-
-f = ggplot()+geom_histogram(aes(x=biasesD1, y = ..density..), alpha = 0.6, bins=50)+
-  theme_tufte()+labs(title="10k biases for different means and sd =.05")+ xlab("bias")
[…]

### 2022-03-01 presentations/presentationBoston/presentationBoston.tex

--- 2021-08-09 presentationESSLLI21/presentationESSLLI.tex
@@
 \footnotesize \newline (LoPSE research group, University of Gdansk)}
-\date{ExpSem2021, ESSLLI}
+\date{Boston, April Fools' Day}
 
@@
 \item
-  feminine occupations: ``homemaker,'' ``nurse,'' ``receptionist,''
+  ``feminine'' occupations: ``homemaker,'' ``nurse,'' ``receptionist,''
   ``librarian,'' etc.
 \item
-  masculine occupations: ``maestro,'' ``captain,'' ``architect,''
+  ``masculine'' occupations: ``maestro,'' ``captain,'' ``architect,''
   ``boss,'' etc.
@@
 
-\begin{center}\includegraphics[width=0.6\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-1-1} \end{center}
+\begin{center}\includegraphics[width=0.6\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-1-1} \end{center}
 \normalsize
@@
 
-\begin{center}\includegraphics[width=0.55\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-2-1} \end{center}
+\begin{center}\includegraphics[width=0.55\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-2-1} \end{center}
 \normalsize
@@
 
-\begin{center}\includegraphics[width=0.5\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-3-1} \end{center}
+\begin{center}\includegraphics[width=0.5\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-3-1} \end{center}
 \normalsize
@@
 \protect\hypertarget{distances-for-muslim}{}
-\begin{center}\includegraphics[width=0.8\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-4-1} \end{center}
+\begin{center}\includegraphics[width=0.8\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-4-1} \end{center}
 \end{block}
@@
 \protect\hypertarget{distances-for-priest}{}
-\begin{center}\includegraphics[width=0.8\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-5-1} \end{center}
+\begin{center}\includegraphics[width=0.8\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-5-1} \end{center}
 \end{block}
@@
 
-\begin{center}\includegraphics[width=0.7\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-6-1} \end{center}
+\begin{center}\includegraphics[width=0.7\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-6-1} \end{center}
 \normalsize
@@
 
-\begin{center}\includegraphics[width=0.8\linewidth]{presentationESSLLI_files/figure-beamer/unnamed-chunk-8-1} \end{center}
+\begin{center}\includegraphics[width=0.8\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-8-1} \end{center}
 \normalsize
@@
 \protect\hypertarget{choosing-priors}{}
-\begin{center}\includegraphics[width=1\linewidth]{presentationESSLLI_files/figure-beamer/priorsVis-1} \end{center}
+\begin{center}\includegraphics[width=1\linewidth]{presentationBoston_files/figure-beamer/priorsVis-1} \end{center}
 \end{block}

### 2022-03-01 functions/caliskanExtended.R
Adds new comments

@@ -12,27 +12,27 @@ library(ggthemes)
 library(gridExtra)
 library(grid)
 
-# BASIC SIMULATION -> samples=5, mean=0, st=0.05
+# BASIC SIMULATION -> samples=5, mean=0, st=0.08
 
 # X = {t1,t2,t3,t4,t5,t6,t7,t8}, Y = {t9,t10,t11,t12,t13,t14,t15,t16}
 
 set.seed(123456)
-t1 <- data.frame(A  = rnorm(8,0,0.05), B = rnorm(8,0,0.05))
-t2 <- data.frame(A  = rnorm(8,0,0.05), B = rnorm(8,0,0.05))
-t3 <- data.frame(A  = rnorm(8,0,0.05), B = rnorm(8,0,0.05))
-t4 <- data.frame(A  = rnorm(8,0,0.05), B = rnorm(8,0,0.05))
-t5 <- data.frame(A  = rnorm(8,0,0.05), B = rnorm(8,0,0.05))
-t6 <- data.frame(A  = rnorm(8,0,0.05), B = rnorm(8,0,0.05))
-t7 <- data.frame(A  = rnorm(8,0,0.05), B = rnorm(8,0,0.05))
-t8 <- data.frame(A  = rnorm(8,0,0.05), B = rnorm(8,0,0.05))
-t9 <- data.frame(A  = rnorm(8,0,0.05), B = rnorm(8,0,0.05))
-t10 <- data.frame(A  = rnorm(8,0,0.05), B = rnorm(8,0,0.05))
-t11 <- data.frame(A  = rnorm(8,0,0.05), B = rnorm(8,0,0.05))
-t12 <- data.frame(A  = rnorm(8,0,0.05), B = rnorm(8,0,0.05))
-t13 <- data.frame(A  = rnorm(8,0,0.05), B = rnorm(8,0,0.05))
-t14 <- data.frame(A  = rnorm(8,0,0.05), B = rnorm(8,0,0.05))
-t15 <- data.frame(A  = rnorm(8,0,0.05), B = rnorm(8,0,0.05))
-t16 <- data.frame(A  = rnorm(8,0,0.05), B = rnorm(8,0,0.05))
+t1 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
+t2 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
[…]

### 2022-03-06 presentations/presentationBoston/presentationBoston.Rmd
Adds weat simulations

@@
 
-## The problem with pre-averaging
+## The problem with pre-averaging on real set-up
 
-### 50k simulations (same parameters)
+### One simulation on real set-up (no similarities and small std)
+
+\vspace{1mm}
+\footnotesize
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "70%"}
+set.seed(123456)
+t1 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
+t2 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
+t3 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
+t4 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
+t5 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
+t6 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
+t7 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
+t8 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
+t9 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
+t10 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
+t11 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
+t12 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
+t13 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
+t14 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
+t15 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
+t16 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
+
+xlimits <- c(-.3,.3)
+
+t1Plot <- ggplot(t1)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t1 in X")+xlim(xlimits)
+
+t2Plot <- ggplot(t2)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t2 in X")+xlim(xlimits)
+
+t8Plot <- ggplot(t8)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t3 in Y")+xlim(xlimits)
+
+t9Plot <- ggplot(t9)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t4 in Y")+xlim(xlimits)
+
+grid.arrange(t1Plot,t2Plot, t8Plot, t9Plot, ncol=2)
+```
+
+## The problem with pre-averaging on real set-up
+
+### 10k simulations (same parameters)
 
@@
 ```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "80%"}
+
 set.seed(123)
-biasesNull <- numeric(50000)
-for(i in 1:50000){
-t1 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
-t2 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
-t3 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
-t4 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
+biasesD0 <- numeric(10000)
 
-factor <- sd(c(s(t1),s(t2),s(t3),s(t4)))
-numerator <-  mean(s(t1),s(t2)) - mean(s(t3),s(t4))
-biasesNull[i]  <- numerator / factor
+for(i in 1:10000){
+  t1 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
+  t2 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
+  t3 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
+  t4 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
+  t5 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
+  t6 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
+  t7 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
+  t8 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
+  t9 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
+  t10 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
+  t11 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
+  t12 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
+  t13 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
+  t14 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
+  t15 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
+  t16 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
+
+  factorD0 <- sd(c(s(t1),s(t2),s(t3),s(t4),s(t5),s(t6),s(t7),s(t8),s(t9),s(t10),s(t11),s(t12),s(t13),s(t14),s(t15),s(t16)))
+  numeratorD0 <-  mean(s(t1),s(t2),s(t3),s(t4),s(t5),s(t6),s(t7),s(t8)) - mean(s(t9),s(t10),s(t11),s(t12),s(t13),s(t14),s(t15),s(t16))
+  biasesD0[i] <- numeratorD0/factorD0
 }
 
-ggplot()+geom_histogram(aes(x=biasesNull, y = ..density..), alpha = 0.6, bins=50)+
-  theme_tufte(base_size = 15)+labs(title="50k biases for identical means and sd =.05")+ xlab("bias")
+ggplot()+geom_histogram(aes(x=biasesD0, y = ..density..), alpha = 0.6, bins=50)+
+  theme_tufte()+labs(title="10k biases for same means and sd =.08")+ xlab("bias")
+
+```
+\vspace{1mm}
+\footnotesize
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+quantile <- quantile(biasesD0, c(0.025, 0.975))
+absOver182 <- mean(abs(biasesD0)>=1.81)
 ```
 \normalsize
+\pause
+
+\footnotesize
+
+\vspace{-2mm}
+
+- The 95% of the data is in range `r round(quantile, 3)`
+- `r round(absOver182*100, 2)`% of the absolute values are above 1.81
+
+## The problem with pre-averaging on real set-up
+
+### 10k simulations with similarity equal to 0.1
+
+\vspace{1mm}
+\footnotesize
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "80%"}
+set.seed(123456)
+biasesD1 <- numeric(10000)
+
+for(i in 1:10000){
+  t1 <- data.frame(A  = rnorm(8,.1,0.08), B = rnorm(8,0,0.08))
+  t2 <- data.frame(A  = rnorm(8,.1,0.08), B = rnorm(8,0,0.08))
+  t3 <- data.frame(A  = rnorm(8,.1,0.08), B = rnorm(8,0,0.08))
+  t4 <- data.frame(A  = rnorm(8,.1,0.08), B = rnorm(8,0,0.08))
+  t5 <- data.frame(A  = rnorm(8,.1,0.08), B = rnorm(8,0,0.08))
+  t6 <- data.frame(A  = rnorm(8,.1,0.08), B = rnorm(8,0,0.08))
+  t7 <- data.frame(A  = rnorm(8,.1,0.08), B = rnorm(8,0,0.08))
+  t8 <- data.frame(A  = rnorm(8,.1,0.08), B = rnorm(8,0,0.08))
+  t9 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,.1,0.08))
+  t10 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,.1,0.08))
+  t11 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,.1,0.08))
+  t12 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,.1,0.08))
+  t13 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,.1,0.08))
+  t14 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,.1,0.08))
+  t15 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,.1,0.08))
+  t16 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,.1,0.08))
+
+  factorD1 <- sd(c(s(t1),s(t2),s(t3),s(t4),s(t5),s(t6),s(t7),s(t8),s(t9),s(t10),s(t11),s(t12),s(t13),s(t14),s(t15),s(t16)))
+  numeratorD1 <-  mean(s(t1),s(t2),s(t3),s(t4),s(t5),s(t6),s(t7),s(t8)) - mean(s(t9),s(t10),s(t11),s(t12),s(t13),s(t14),s(t15),s(t16))
+  biasesD1[i] <- numeratorD1/factorD1
+}
+
+ggplot()+geom_histogram(aes(x=biasesD1, y = ..density..), alpha = 0.6, bins=50)+
+  theme_tufte()+labs(title="10k biases for different means 0.1 and sd =.08")+ xlab("bias")
+
+```
+\vspace{1mm}
+\footnotesize
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+quantiles1 <- quantile(biasesD1, c(0.025, 0.975))
+absMean1 <- mean(abs(biasesD1)>=1.81)
+```
+\normalsize
+\pause
+
+\footnotesize
+
+\vspace{-2mm}
+
+- The 95% of the data is in range `r round(quantiles1, 3)`
+- `r round(absMean1*100, 2)`% of the absolute values are above 1.81
+
+## The problem with pre-averaging on real set-up
+
+### 10k simulations with similarity equal to 0.4
+
+\vspace{1mm}
+\footnotesize
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "80%"}
+set.seed(123456)
+biasesD4 <- numeric(10000)
+
+for(i in 1:10000){
+  t1 <- data.frame(A  = rnorm(8,.4,0.08), B = rnorm(8,0,0.08))
+  t2 <- data.frame(A  = rnorm(8,.4,0.08), B = rnorm(8,0,0.08))
+  t3 <- data.frame(A  = rnorm(8,.4,0.08), B = rnorm(8,0,0.08))
+  t4 <- data.frame(A  = rnorm(8,.4,0.08), B = rnorm(8,0,0.08))
+  t5 <- data.frame(A  = rnorm(8,.4,0.08), B = rnorm(8,0,0.08))
+  t6 <- data.frame(A  = rnorm(8,.4,0.08), B = rnorm(8,0,0.08))
+  t7 <- data.frame(A  = rnorm(8,.4,0.08), B = rnorm(8,0,0.08))
+  t8 <- data.frame(A  = rnorm(8,.4,0.08), B = rnorm(8,0,0.08))
+  t9 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,.4,0.08))
+  t10 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,.4,0.08))
+  t11 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,.4,0.08))
+  t12 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,.4,0.08))
+  t13 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,.4,0.08))
+  t14 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,.4,0.08))
+  t15 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,.4,0.08))
+  t16 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,.4,0.08))
+
+  factorD4 <- sd(c(s(t1),s(t2),s(t3),s(t4),s(t5),s(t6),s(t7),s(t8),s(t9),s(t10),s(t11),s(t12),s(t13),s(t14),s(t15),s(t16)))
+  numeratorD4 <-  mean(s(t1),s(t2),s(t3),s(t4),s(t5),s(t6),s(t7),s(t8)) - mean(s(t9),s(t10),s(t11),s(t12),s(t13),s(t14),s(t15),s(t16))
+  biasesD4[i] <- numeratorD4/factorD4
+}
+
+ggplot()+geom_histogram(aes(x=biasesD4, y = ..density..), alpha = 0.6, bins=50)+
+  theme_tufte()+labs(title="10k biases for different means 0.4 and sd =.08")+ xlab("bias")
+
+```
+\vspace{1mm}
+\footnotesize
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
+quantiles4 <- quantile(biasesD4, c(0.025, 0.975))
+absMean4 <- mean(abs(biasesD4)>=1.81)
+```
+\normalsize
+\pause
+
+\footnotesize
+
+\vspace{-2mm}
+
+- The 95% of the data is in range `r round(quantiles4, 3)`
+- `r round(absMean4*100, 2)`% of the absolute values are above 1.81
 

### 2022-03-06 presentations/presentationBoston/presentationBoston.tex
Adds weat simulations

@@
 
-\begin{frame}{The problem with pre-averaging}
-\protect\hypertarget{the-problem-with-pre-averaging-2}{}
-\begin{block}{50k simulations (same parameters)}
+\begin{frame}{The problem with pre-averaging on real set-up}
+\protect\hypertarget{the-problem-with-pre-averaging-on-real-set-up}{}
+\begin{block}{One simulation on real set-up (no similarities and small
+std)}
+\protect\hypertarget{one-simulation-on-real-set-up-no-similarities-and-small-std}{}
+\vspace{1mm}
+\footnotesize
+
+\begin{center}\includegraphics[width=0.7\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-8-1} \end{center}
+\end{block}
+\end{frame}
+
+\begin{frame}{The problem with pre-averaging on real set-up}
+\protect\hypertarget{the-problem-with-pre-averaging-on-real-set-up-1}{}
+\begin{block}{10k simulations (same parameters)}
 \protect\hypertarget{k-simulations-same-parameters}{}
@@
 
-\begin{center}\includegraphics[width=0.8\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-8-1} \end{center}
+\begin{center}\includegraphics[width=0.8\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-9-1} \end{center}
+\vspace{1mm}
+\footnotesize
+
 \normalsize
+\pause
+
+\footnotesize
+
+\vspace{-2mm}
+
+\begin{itemize}
+\tightlist
+\item
+  The 95\% of the data is in range -2.763, 2.698
+\item
+  21.38\% of the absolute values are above 1.81
+\end{itemize}
+\end{block}
+\end{frame}
+
+\begin{frame}{The problem with pre-averaging on real set-up}
+\protect\hypertarget{the-problem-with-pre-averaging-on-real-set-up-2}{}
+\begin{block}{10k simulations with similarity equal to 0.1}
+\protect\hypertarget{k-simulations-with-similarity-equal-to-0.1}{}
+\vspace{1mm}
+\footnotesize
+
+\begin{center}\includegraphics[width=0.8\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-11-1} \end{center}
+\vspace{1mm}
+\footnotesize
+
+\normalsize
+\pause
+
+\footnotesize
+
+\vspace{-2mm}
+
+\begin{itemize}
+\tightlist
+\item
+  The 95\% of the data is in range 0.851, 2.764
+\item
+  51.3\% of the absolute values are above 1.81
+\end{itemize}
+\end{block}
+\end{frame}
+
+\begin{frame}{The problem with pre-averaging on real set-up}
+\protect\hypertarget{the-problem-with-pre-averaging-on-real-set-up-3}{}
+\begin{block}{10k simulations with similarity equal to 0.4}
+\protect\hypertarget{k-simulations-with-similarity-equal-to-0.4}{}
+\vspace{1mm}
+\footnotesize
+
+\begin{center}\includegraphics[width=0.8\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-13-1} \end{center}
+\vspace{1mm}
+\footnotesize
+
+\normalsize
+\pause
+
+\footnotesize
+
+\vspace{-2mm}
+
+\begin{itemize}
+\tightlist
+\item
+  The 95\% of the data is in range 1.679, 2.185
+\item
+  82.9\% of the absolute values are above 1.81
+\end{itemize}
 \end{block}

### 2022-03-06 functions/caliskanExtended.R

@@ -38,8 +38,8 @@ xlimits <- c(-.3,.3)
 
 t1Plot <- ggplot(t1)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t1 in X")+xlim(xlimits)
 t2Plot <- ggplot(t2)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t2 in X")+xlim(xlimits)
-t3Plot <- ggplot(t3)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t3 in Y")+xlim(xlimits)
[…]

### 2022-03-08 presentations/presentationBoston/presentationBoston.Rmd
presentation Boston added cosine etc.

@@
 fontsize: 10pt
-classoption: x11names, dvipsnames, bibspacing,natbib
+classoption: x11names, dvipsnames, bibspacing,natbib, table
 urlcolor: blue
@@
 library(ggthemes)
+library(ggforce)
 library(latex2exp)
@@
 
-## Cosine-based measures of bias
-
-### Word embeddings
-
-- Representation of words with vectors of real numbers
-
-- Often built to predict the probability of co-occurence
-
-<!-- check if the statement of co-occurence always right-->
+## Word2vec
+
+### Question
+
+How to sensibly represent words with numbers?
+
+\pause
+
+### One-hot encoding
+
+Well, you could use 30k binary vectors with a slot for each lexical unit\dots
+
+\pause
+
+\dots \dots but this would be inefficient and wouldn't capture any relations between words.
+
+## Word2vec
+
+\begin{center}
+ \includegraphics[width = 9cm]{images/perceptron1.png}
+\end{center}
+
+\vspace{-3mm}
+\tiny \hfill \color{gray}Illustration: M. Mitchell \color{black}
+
+\footnotesize
+
+\begin{block}{Rosenblatt's perceptron}
+\begin{itemize}
+\item Inputs (pixel intensities) with weights
+\item Nodes with activation levels from 0-1
+\item (Perhaps) 0-1 output based on a threshold
+\end{itemize}
+\end{block}
+
+## Word2vec
+
+\begin{center}
+ \includegraphics[width = 9cm]{images/perceptron1.png}
+\end{center}
+
+\vspace{-3mm}
+\tiny \hfill \color{gray}Illustration: M. Mitchell \color{black}
+
+\footnotesize
+
+\begin{block}{Learning}
+\begin{itemize}
+\item Start with random weights
+\item Test on a case:
+\begin{itemize}
+\item If right, don't change weights.
+\item If wrong, change weights a bit, with focus on the ones more responsible for the judgment:
+\begin{align*}
+w_j & \leftarrow w_j = \overbrace{\eta}^{\text{learning rate}}(\underbrace{t}_{\text{correct output}} - \overbrace{y}^{\text{actual output}})\underbrace{x_j}_{\text{actual input}}
+\end{align*}
+\end{itemize}
+\end{itemize}
+
+\end{block}
+
+## Word2vec
+
+\begin{columns}
+\column{0.45\linewidth}
+
+\begin{center}
+ \includegraphics[height = 6cm, width = 5cm]{images/perceptron2.png}
+\end{center}
+
+\vspace{-3mm}
+\tiny \hfill \color{gray}Illustration: M. Mitchell \color{black}
+
+\column{0.5\linewidth}
+
+\footnotesize
+
+\begin{itemize}
+\item Each hidden unit takes a weighted sum of 324 inputs and passes on its activation level as input to outer layer units.
+
+\item Activation levels of outer layers are interpreted as network's levels of confidence in a classification problem.
+
+\item Learning: back-propagation (gradient descent: approximate  the direction of steepest descent in the error surface w.r.t to weights, modify accordingly).
+\end{itemize}
+
+\end{columns}
+
+## Word2vec
+
+### Distributional semantics
+
+\begin{itemize}
+\item "You shall know a word by the company it keeps" (John Firth, 1957)
+
+\item "the degree of semantic similarity between two linguistic expressions $A$ and $B$ is a function of the similarity of the linguistic contexts in which $A$ and $B$ can appear." (A. Lenci, 2008)
+
+\end{itemize}
+
+\pause
+
+\begin{center}
+ \includegraphics[height = 5cm, width = 5cm]{images/similarity1.png}
+\end{center}
+
+\vspace{-3mm}
+\tiny \hfill \color{gray}Illustration: M. Mitchell \color{black}
+
+## Word2vec
+
+### Google and Mikolov
+
+\emph{Efficient Estimation of Word Representation in Vector Space}, 2013
+
+Let's train a neural network and use vectors of weights!
+
+\begin{center}
+ \includegraphics[height = 5cm, width = 6cm]{images/word2vec1.png}
+\end{center}
+
+\vspace{-3mm}
+\tiny \hfill \color{gray}Illustration: M. Mitchell \color{black}
+
+## Word2vec
+
+### Google and Mikolov
+
+\emph{Efficient Estimation of Word Representation in Vector Space}, 2013
+
+Let's train a neural network and use vectors of weights!
+
+\begin{center}
+ \includegraphics[height = 3cm, width = 7cm]{images/word2vec2.png}
+\end{center}
+
+\vspace{-3mm}
+\tiny \hfill \color{gray}Illustration: M. Mitchell \color{black}
+
+## Word2vec
 
@@
 
-## Cosine-based measures of bias
-
-### Cosine similarity \& distance
+### Question
+
+How is this supposed to capture semantic relations?
+
+\pause
+
+### General idea
+
+Similarity in vector direction.
+
+## Cosine similarity
+
+\begin{columns}
+\column{0.45\linewidth}
+
+```{r cosine1,echo=FALSE,eval=TRUE,fig.align = "center", cache=TRUE, fig.show = "hold", out.width = "100%",   message = FALSE, warning = FALSE, results = FALSE}
+a <- c(1,2)
+b <- c(3,2.5)
+
+ggplot()+
+  xlim(c(0,4.5))+ylim(c(0,4.5))+
+  geom_segment(aes(x = 0, y = 0, xend = a[1], yend = a[2]),  
+               arrow = arrow(length = unit(0.02, "npc")))+
+  annotate(geom = "text", label  = "a = [1,2]", x = 1.2, y = 2.4, size = 7)+
+  geom_segment(aes(x = 0, y = 0, xend = b[1], yend = b[2]), arrow = arrow(length = unit(0.02, "npc")))+   annotate(geom = "text", label  = "b = [3,2.5]", x = 3.2, y = 2.9, size = 7)+
+  theme_tufte(base_size = 24)
+
+```
+
+\column{0.5\linewidth}
+
+\footnotesize
+
+\begin{block}{Vectors}
+\begin{align*}
+a  & = [1,2]\\
+b  &= [3,2]
+\end{align*}
+
+\end{block}
+\pause
+
+\begin{block}{Dot product}
+
+\begin{align*}
+a \cdot b & = a_1 b_1 + a_2 b_2\\
+a \cdot a & = a_1^2 + a_2 ^ 2 \\
+\lVert a\rVert & = \sqrt(a \cdot a)
+\end{align*}
+
+\end{block}
+
+\end{columns}
+
+## Cosine similarity
+
+\begin{columns}
+\column{0.45\linewidth}
+
+```{r cosine2,echo=FALSE,eval=TRUE,fig.align = "center", cache=TRUE, fig.show = "hold", out.width = "100%",   message = FALSE, warning = FALSE, results = FALSE}
+a <- c(1,2)
+b <- c(3,2.5)
+bma <- b - a
+
+ggplot()+
+  xlim(c(0,4.5))+ylim(c(0,4.5))+
+  geom_segment(aes(x = 0, y = 0, xend = a[1], yend = a[2]),  
+               arrow = arrow(length = unit(0.02, "npc")))+
+  annotate(geom = "text", label  = "a = [1,2]", x = 1.2, y = 2.4, size = 7)+
+  geom_segment(aes(x = 0, y = 0, xend = b[1], yend = b[2]), arrow = arrow(length = unit(0.02, "npc")))+   annotate(geom = "text", label  = "b = [3,2.5]", x = 3.2, y = 2.9, size = 7)+
+  theme_tufte(base_size = 24)+
+  geom_segment(aes(x = 0, y = 0, xend = bma[1], yend = bma[2]),
+      arrow = arrow(length = unit(0.02, "npc")), color = "grey")+ 
+  annotate(geom = "text", label  = "b - a = [2,.5]", x = 2.2, y = .9, size = 7)+
+  theme_tufte(base_size = 24)
+
+```
+
+\column{0.5\linewidth}
+
+\footnotesize
+
+\begin{block}{Vectors}
+
+\begin{align*}
+a  & = [1,2]\\
+b  &= [4,4]
+\end{align*}
+
+\end{block}
+
+\begin{block}{Dot product}
+
+\begin{align*}
+a \cdot b & = a_1 b_1 + a_2 b_2\\
+a \cdot a & = a_1^2 + a_2 ^ 2 \\
+\lVert a\rVert & = \sqrt(a \cdot a)
+\end{align*}
+
+\end{block}
+
+\begin{block}{Vector difference}
+
+\begin{align*}
+b - a & = [b_1- a_1, b_2 - a_2 ]
+\end{align*}
+
+\end{block}
+
+\end{columns}
+
+## Cosine similarity
+
+\begin{columns}
+\column{0.45\linewidth}
+
+```{r cosine3,echo=FALSE,eval=TRUE,fig.align = "center", cache=TRUE, fig.show = "hold", out.width = "100%",   message = FALSE, warning = FALSE, results = FALSE}
+a <- c(1,2)
+b <- c(3,2.5)
+
+ggplot()+
+  xlim(c(0,4.5))+ylim(c(0,4.5))+
+  geom_segment(aes(x = 0, y = 0, xend = a[1], yend = a[2]),  
+               arrow = arrow(length = unit(0.02, "npc")))+
+  annotate(geom = "text", label  = "a = [1,2]", x = 1.2, y = 2.4, size = 7)+
+  geom_segment(aes(x = 0, y = 0, xend = b[1], yend = b[2]), arrow = arrow(length = unit(0.02, "npc")))+   annotate(geom = "text", label  = "b = [3,2.5]", x = 3.2, y = 2.9, size = 7)+
+  theme_tufte(base_size = 24)+
+  geom_segment(aes(x = a[1], y = a[2], xend = b[1], yend = b[2]),
+      arrow = arrow(length = unit(0.02, "npc")), color = "grey")+ 
+  annotate(geom = "text", label  = "b - a = [2,.5] (shifted)", x = 2, y = 2.7, size = 7)+
+  theme_tufte(base_size = 24)
+
+```
+
+\column{0.5\linewidth}
+
+\footnotesize
+
+\begin{block}{Vectors}
+
+\begin{align*}
+a  & = [1,2]\\
+b  &= [4,4]
+\end{align*}
+
+\end{block}
+
+\begin{block}{Dot product}
+
+\begin{align*}
+a \cdot b & = a_1 b_1 + a_2 b_2\\
+a \cdot a & = a_1^2 + a_2 ^ 2 \\
+\lVert a\rVert & = \sqrt(a \cdot a)
+\end{align*}
+
+\end{block}
+
+\begin{block}{Vector difference}
+
+\begin{align*}
+b - a & = [b_1- a_1, b_2 - a_2 ]
+\end{align*}
+
+\end{block}
+
+\end{columns}
+
+## Cosine similarity
+
+\begin{columns}
+\column{0.45\linewidth}
+
+```{r cosine4,echo=FALSE,eval=TRUE,fig.align = "center", cache=TRUE, fig.show = "hold", out.width = "100%",   message = FALSE, warning = FALSE, results = FALSE}
+a <- c(1,2)
+b <- c(3,2.5)
+#atan2(a[1] , a[2])
+#atan2(b[1] , b[2])
+
+ggplot()+
+  xlim(c(0,4.5))+ylim(c(0,4.5))+
+  geom_segment(aes(x = 0, y = 0, xend = a[1], yend = a[2]),  
+               arrow = arrow(length = unit(0.02, "npc")))+
+  annotate(geom = "text", label  = "a = [1,2]", x = 1.2, y = 2.4, size = 7)+
+  geom_segment(aes(x = 0, y = 0, xend = b[1], yend = b[2]), arrow = arrow(length = unit(0.02, "npc")))+   annotate(geom = "text", label  = "b = [3,2.5]", x = 3.2, y = 2.9, size = 7)+
+  theme_tufte(base_size = 24)+
+  geom_segment(aes(x = a[1], y = a[2], xend = b[1], yend = b[2]),
+      arrow = arrow(length = unit(0.02, "npc")), color = "grey")+ 
+  annotate(geom = "text", label  = "b - a = [2,.5] (shifted)", x = 2, y = 2.7, size = 7)+
+  theme_tufte(base_size = 24)+
+  annotate(geom = "text", label  = expression(theta), x = .6, y = .78, size = 7)+
+  geom_arc(aes(x0 = 0, y0 = 0, r = 1.3, 
+               start = .463, end = .876))
+
+```
+
+\column{0.5\linewidth}
+
+\footnotesize
+
+\begin{block}{Angle}
+
+\begin{align*}
+\lVert b - a \rVert^2 &  = \lVert  b \rVert^2 + \lVert a \rVert^2 - 
+  2 \lVert b \rVert   \lvert  a \rVert  \cos \theta  \\
+b \cdot a  & = \lVert b\rVert \lVert  a \rVert \cos \theta \\ 
+\cos \theta & = \frac{b \cdot a}{\lVert b\rVert  \lVert  a \rVert }
+\end{align*}
+
+\end{block}
+
+\pause
+
+\begin{block}{Orthogonality}
+\begin{align*}
+\cos ( 90^{\circ}) & = 0 \\ 
+\frac{b \cdot a}{\lVert  b\rVert \lVert a \rVert} & = 0\\ 
+ b \cdot a & = 0  
+\end{align*}
+\end{block}
+
+\end{columns}
+
+## Cosine similarity \& distance
 
@@
 \begin{align} \tag{Sim}
-\mathsf{cosineSimilarity}(A,B) & = \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert}
+\mathsf{cosineSimilarity}(A,B) & = \frac{A \cdot B}{\lVert  A \rVert \,\lVert B \rVert}
 \\
@@
 
-### Known challenges
+### General challenges
 
@@
 
-We run with it for comparison
+- We run with it for comparison.
 
@@
 
-We statistically gauge the uncertainty that arises from raw sample sizes
+- ETHAYARAJH IS YOUR criticizes WEAT uses Bernstein bounds and argues that we would need a bias specific dataset of size at least 11903 to claim that the system is biased (three times larger than WinoBias).
+
+\pause
+
+- We show progress can be made with more sensitive Bayesian methods.
+
+### The form of the definition is suspicious
+
+ETHAYARAJH UNDERSTANDING shows if there are two target words only WEAT is always maximal in one direction.
+
+\pause
+
+- We show the problem runs deeper and stems from pre-averaging, and we statistically gauge the uncertainty that arises from raw sample sizes.
 
@@
 
-We study those by  visualizations and uncertainty estimates
+We study those by  visualizations and uncertainty estimates.
 
@@
 
-### Our simulation
-
-Suppose all similarities for two classes are randomly drawn from the same distribution, $\mathsf{Normal}(\mu = 0, \sigma = .05)$, you still can get a really high WEAT!
+### Our simulations
+
+Suppose all similarities for two classes are randomly drawn from the same distribution, $\mathsf{Normal}(\mu = 0, \sigma = .08)$, you still can get a really high WEAT!
 
@@
 
-### One simulation
+### Simple case: two pws, four terms
 
@@
 set.seed(123)
-t1 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
-t2 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
-t3 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
-t4 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
-xlimits <- c(-.3,.3)
+t1 <- data.frame(A  = rnorm(5,0,0.08), B = rnorm(5,0,0.08))
+t2 <- data.frame(A  = rnorm(5,0,0.08), B = rnorm(5,0,0.08))
+t3 <- data.frame(A  = rnorm(5,0,0.08), B = rnorm(5,0,0.08))
+t4 <- data.frame(A  = rnorm(5,0,0.08), B = rnorm(5,0,0.08))
+xlimits <- c(-1,1)
 
@@
 
-## The problem with pre-averaging on real set-up
-
-### One simulation on real set-up (no similarities and small std)
+## The problem with pre-averaging on realistic set-up
+
+### Simulation with realistic set-up (16 predicates)
 
@@
 
-xlimits <- c(-.3,.3)
+xlimits <- c(-1,1)
 
@@
 grid.arrange(t1Plot,t2Plot, t8Plot, t9Plot, ncol=2)
-```
-
-## The problem with pre-averaging on real set-up
+
+whole <- rbind(t1,t2,t3,t4)
+rawsd <- sd(c(whole$A,whole$B))
+factor <- sd(c(s(t1),s(t2),s(t3),s(t4)))
+numerator <-  mean(s(t1),s(t2)) - mean(s(t3),s(t4))
+```
+
+\footnotesize
+
+\vspace{-2mm}
+
+ Raw sd: `r round(rawsd,3)`, sd of means: `r round(factor,3)`, WEAT:  `r round(numerator/factor,3)`.
+
+## The problem with pre-averaging on realistic set-up
 
@@
 
-- The 95% of the data is in range `r round(quantile, 3)`
+- 95% of the scores are in range `r round(quantile, 3)`
 - `r round(absOver182*100, 2)`% of the absolute values are above 1.81
 
-## The problem with pre-averaging on real set-up
-
-### 10k simulations with similarity equal to 0.1
+## The problem with pre-averaging on realistic set-up
+
+### 10k simulations with mean similarity  0.1
 
@@
 \normalsize
-\pause
 
@@
 
-- The 95% of the data is in range `r round(quantiles1, 3)`
+- 95% of the scores are in range `r round(quantiles1, 3)`
 - `r round(absMean1*100, 2)`% of the absolute values are above 1.81
 
-## The problem with pre-averaging on real set-up
-
-### 10k simulations with similarity equal to 0.4
+## The problem with pre-averaging on realistic set-up
+
+### 10k simulations with mean similarity  0.4
 
@@
 
-- The 95% of the data is in range `r round(quantiles4, 3)`
+- 95% of the scores are in range `r round(quantiles4, 3)`
 - `r round(absMean4*100, 2)`% of the absolute values are above 1.81

### 2022-03-08 presentations/presentationBoston/presentationBoston.tex
presentation Boston added cosine etc.

@@
   ignorenonframetext,
-  x11names, dvipsnames, bibspacing,natbib]{beamer}
+  x11names, dvipsnames, bibspacing,natbib, table]{beamer}
 \usepackage{pgfpages}
@@
 
-\begin{frame}{Cosine-based measures of bias}
-\protect\hypertarget{cosine-based-measures-of-bias}{}
-\begin{block}{Word embeddings}
-\protect\hypertarget{word-embeddings}{}
-\begin{itemize}
-\item
-  Representation of words with vectors of real numbers
-\item
-  Often built to predict the probability of co-occurence
-\end{itemize}
-
+\begin{frame}{Word2vec}
+\protect\hypertarget{word2vec}{}
+\begin{block}{Question}
+\protect\hypertarget{question}{}
+How to sensibly represent words with numbers?
+
+\pause
+\end{block}
+
+\begin{block}{One-hot encoding}
+\protect\hypertarget{one-hot-encoding}{}
+Well, you could use 30k binary vectors with a slot for each lexical
+unit\dots
+
+\pause
+
+\dots \dots but this would be inefficient and wouldn't capture any
+relations between words.
+\end{block}
+\end{frame}
+
+\begin{frame}{Word2vec}
+\protect\hypertarget{word2vec-1}{}
+\begin{center}
+ \includegraphics[width = 9cm]{images/perceptron1.png}
+\end{center}
+
+\vspace{-3mm}
+
+\tiny \hfill \color{gray}Illustration: M. Mitchell \color{black}
+
+\footnotesize
+
+\begin{block}{Rosenblatt's perceptron}
+\begin{itemize}
+\item Inputs (pixel intensities) with weights
+\item Nodes with activation levels from 0-1
+\item (Perhaps) 0-1 output based on a threshold
+\end{itemize}
+\end{block}
+\end{frame}
+
+\begin{frame}{Word2vec}
+\protect\hypertarget{word2vec-2}{}
+\begin{center}
+ \includegraphics[width = 9cm]{images/perceptron1.png}
+\end{center}
+
+\vspace{-3mm}
+
+\tiny \hfill \color{gray}Illustration: M. Mitchell \color{black}
+
+\footnotesize
+
+\begin{block}{Learning}
+\begin{itemize}
+\item Start with random weights
+\item Test on a case:
+\begin{itemize}
+\item If right, don't change weights.
+\item If wrong, change weights a bit, with focus on the ones more responsible for the judgment:
+\begin{align*}
+w_j & \leftarrow w_j = \overbrace{\eta}^{\text{learning rate}}(\underbrace{t}_{\text{correct output}} - \overbrace{y}^{\text{actual output}})\underbrace{x_j}_{\text{actual input}}
+\end{align*}
+\end{itemize}
+\end{itemize}
+
+\end{block}
+\end{frame}
+
+\begin{frame}{Word2vec}
+\protect\hypertarget{word2vec-3}{}
+\begin{columns}
+\column{0.45\linewidth}
+
+\begin{center}
+ \includegraphics[height = 6cm, width = 5cm]{images/perceptron2.png}
+\end{center}
+
+\vspace{-3mm}
+\tiny \hfill \color{gray}Illustration: M. Mitchell \color{black}
+
+\column{0.5\linewidth}
+
+\footnotesize
+
+\begin{itemize}
+\item Each hidden unit takes a weighted sum of 324 inputs and passes on its activation level as input to outer layer units.
+
+\item Activation levels of outer layers are interpreted as network's levels of confidence in a classification problem.
+
+\item Learning: back-propagation (gradient descent: approximate  the direction of steepest descent in the error surface w.r.t to weights, modify accordingly).
+\end{itemize}
+
+\end{columns}
+\end{frame}
+
+\begin{frame}{Word2vec}
+\protect\hypertarget{word2vec-4}{}
+\begin{block}{Distributional semantics}
+\protect\hypertarget{distributional-semantics}{}
+\begin{itemize}
+\item "You shall know a word by the company it keeps" (John Firth, 1957)
+
+\item "the degree of semantic similarity between two linguistic expressions $A$ and $B$ is a function of the similarity of the linguistic contexts in which $A$ and $B$ can appear." (A. Lenci, 2008)
+
+\end{itemize}
+
+\pause
+
+\begin{center}
+ \includegraphics[height = 5cm, width = 5cm]{images/similarity1.png}
+\end{center}
+
+\vspace{-3mm}
+
+\tiny \hfill \color{gray}Illustration: M. Mitchell \color{black}
+\end{block}
+\end{frame}
+
+\begin{frame}{Word2vec}
+\protect\hypertarget{word2vec-5}{}
+\begin{block}{Google and Mikolov}
+\protect\hypertarget{google-and-mikolov}{}
+\emph{Efficient Estimation of Word Representation in Vector Space}, 2013
+
+Let's train a neural network and use vectors of weights!
+
+\begin{center}
+ \includegraphics[height = 5cm, width = 6cm]{images/word2vec1.png}
+\end{center}
+
+\vspace{-3mm}
+
+\tiny \hfill \color{gray}Illustration: M. Mitchell \color{black}
+\end{block}
+\end{frame}
+
+\begin{frame}{Word2vec}
+\protect\hypertarget{word2vec-6}{}
+\begin{block}{Google and Mikolov}
+\protect\hypertarget{google-and-mikolov-1}{}
+\emph{Efficient Estimation of Word Representation in Vector Space}, 2013
+
+Let's train a neural network and use vectors of weights!
+
+\begin{center}
+ \includegraphics[height = 3cm, width = 7cm]{images/word2vec2.png}
+\end{center}
+
+\vspace{-3mm}
+
+\tiny \hfill \color{gray}Illustration: M. Mitchell \color{black}
+\end{block}
+\end{frame}
+
+\begin{frame}{Word2vec}
+\protect\hypertarget{word2vec-7}{}
 \begin{longtable}[]{@{}llllll@{}}
@@
 \end{longtable}
-\end{block}
-\end{frame}
-
-\begin{frame}{Cosine-based measures of bias}
-\protect\hypertarget{cosine-based-measures-of-bias-1}{}
-\begin{block}{Cosine similarity \& distance}
+
+\begin{block}{Question}
+\protect\hypertarget{question-1}{}
+How is this supposed to capture semantic relations?
+
+\pause
+\end{block}
+
+\begin{block}{General idea}
+\protect\hypertarget{general-idea}{}
+Similarity in vector direction.
+\end{block}
+\end{frame}
+
+\begin{frame}{Cosine similarity}
+\protect\hypertarget{cosine-similarity}{}
+\begin{columns}
+\column{0.45\linewidth}
+
+\begin{center}\includegraphics[width=1\linewidth]{presentationBoston_files/figure-beamer/cosine1-1} \end{center}
+
+\column{0.5\linewidth}
+
+\footnotesize
+
+\begin{block}{Vectors}
+\begin{align*}
+a  & = [1,2]\\
+b  &= [3,2]
+\end{align*}
+
+\end{block}
+\pause
+
+\begin{block}{Dot product}
+
+\begin{align*}
+a \cdot b & = a_1 b_1 + a_2 b_2\\
+a \cdot a & = a_1^2 + a_2 ^ 2 \\
+\lVert a\rVert & = \sqrt(a \cdot a)
+\end{align*}
+
+\end{block}
+
+\end{columns}
+\end{frame}
+
+\begin{frame}{Cosine similarity}
+\protect\hypertarget{cosine-similarity-1}{}
+\begin{columns}
+\column{0.45\linewidth}
+
+\begin{center}\includegraphics[width=1\linewidth]{presentationBoston_files/figure-beamer/cosine2-1} \end{center}
+
+\column{0.5\linewidth}
+
+\footnotesize
+
+\begin{block}{Vectors}
+
+\begin{align*}
+a  & = [1,2]\\
+b  &= [4,4]
+\end{align*}
+
+\end{block}
+
+\begin{block}{Dot product}
+
+\begin{align*}
+a \cdot b & = a_1 b_1 + a_2 b_2\\
+a \cdot a & = a_1^2 + a_2 ^ 2 \\
+\lVert a\rVert & = \sqrt(a \cdot a)
+\end{align*}
+
+\end{block}
+
+\begin{block}{Vector difference}
+
+\begin{align*}
+b - a & = [b_1- a_1, b_2 - a_2 ]
+\end{align*}
+
+\end{block}
+
+\end{columns}
+\end{frame}
+
+\begin{frame}{Cosine similarity}
+\protect\hypertarget{cosine-similarity-2}{}
+\begin{columns}
+\column{0.45\linewidth}
+
+\begin{center}\includegraphics[width=1\linewidth]{presentationBoston_files/figure-beamer/cosine3-1} \end{center}
+
+\column{0.5\linewidth}
+
+\footnotesize
+
+\begin{block}{Vectors}
+
+\begin{align*}
+a  & = [1,2]\\
+b  &= [4,4]
+\end{align*}
+
+\end{block}
+
+\begin{block}{Dot product}
+
+\begin{align*}
+a \cdot b & = a_1 b_1 + a_2 b_2\\
+a \cdot a & = a_1^2 + a_2 ^ 2 \\
+\lVert a\rVert & = \sqrt(a \cdot a)
+\end{align*}
+
+\end{block}
+
+\begin{block}{Vector difference}
+
+\begin{align*}
+b - a & = [b_1- a_1, b_2 - a_2 ]
+\end{align*}
+
+\end{block}
+
+\end{columns}
+\end{frame}
+
+\begin{frame}{Cosine similarity}
+\protect\hypertarget{cosine-similarity-3}{}
+\begin{columns}
+\column{0.45\linewidth}
+
+\begin{center}\includegraphics[width=1\linewidth]{presentationBoston_files/figure-beamer/cosine4-1} \end{center}
+
+\column{0.5\linewidth}
+
+\footnotesize
+
+\begin{block}{Angle}
+
+\begin{align*}
+\lVert b - a \rVert^2 &  = \lVert  b \rVert^2 + \lVert a \rVert^2 - 
+  2 \lVert b \rVert   \lvert  a \rVert  \cos \theta  \\
+b \cdot a  & = \lVert b\rVert \lVert  a \rVert \cos \theta \\ 
+\cos \theta & = \frac{b \cdot a}{\lVert b\rVert  \lVert  a \rVert }
+\end{align*}
+
+\end{block}
+
+\pause
+
+\begin{block}{Orthogonality}
+\begin{align*}
+\cos ( 90^{\circ}) & = 0 \\ 
+\frac{b \cdot a}{\lVert  b\rVert \lVert a \rVert} & = 0\\ 
+ b \cdot a & = 0  
+\end{align*}
+\end{block}
+
+\end{columns}
+\end{frame}
+
+\begin{frame}{Cosine similarity \& distance}
 \protect\hypertarget{cosine-similarity-distance}{}
@@
 \end{itemize}
-\end{block}
 \end{frame}
@@
 \begin{frame}{Cosine-based measures of bias}
-\protect\hypertarget{cosine-based-measures-of-bias-2}{}
+\protect\hypertarget{cosine-based-measures-of-bias}{}
 \begin{block}{The worry}
@@
 \begin{frame}{Cosine-based measures of bias}
-\protect\hypertarget{cosine-based-measures-of-bias-3}{}
+\protect\hypertarget{cosine-based-measures-of-bias-1}{}
 \begin{block}{A visual example}
@@
 \begin{frame}{Cosine-based measures of bias}
-\protect\hypertarget{cosine-based-measures-of-bias-4}{}
+\protect\hypertarget{cosine-based-measures-of-bias-2}{}
 \begin{block}{Example: Word Embedding Association Test (WEAT)}
@@
 \begin{frame}{Cosine-based measures of bias}
-\protect\hypertarget{cosine-based-measures-of-bias-5}{}
+\protect\hypertarget{cosine-based-measures-of-bias-3}{}
 \begin{block}{Example: Word Embedding Association Test (WEAT)}
@@
 \begin{frame}{Cosine-based measures of bias}
-\protect\hypertarget{cosine-based-measures-of-bias-6}{}
+\protect\hypertarget{cosine-based-measures-of-bias-4}{}
 \begin{block}{Our main target: Mean Average Cosine Similarity (MAC)}
@@
 \begin{frame}{Cosine-based measures of bias}
-\protect\hypertarget{cosine-based-measures-of-bias-7}{}
+\protect\hypertarget{cosine-based-measures-of-bias-5}{}
 \begin{block}{Our main target: Mean Average Cosine Similarity (MAC)}
@@
 \begin{frame}{Cosine-based measures of bias}
-\protect\hypertarget{cosine-based-measures-of-bias-8}{}
+\protect\hypertarget{cosine-based-measures-of-bias-6}{}
 \begin{block}{Our main target: Mean Average Cosine Similarity (MAC)}
@@
 \begin{frame}{Cosine-based measures of bias}
-\protect\hypertarget{cosine-based-measures-of-bias-9}{}
-\begin{block}{Known challenges}
-\protect\hypertarget{known-challenges}{}
+\protect\hypertarget{cosine-based-measures-of-bias-7}{}
+\begin{block}{General challenges}
+\protect\hypertarget{general-challenges}{}
 \begin{itemize}
@@
 \protect\hypertarget{word-list-choice-is-unprincipled}{}
-We run with it for comparison
+\begin{itemize}
+\tightlist
+\item
+  We run with it for comparison.
+\end{itemize}
 
@@
 \protect\hypertarget{no-design-considerations-to-sample-size}{}
-We statistically gauge the uncertainty that arises from raw sample sizes
+\begin{itemize}
+\tightlist
+\item
+  ETHAYARAJH IS YOUR criticizes WEAT uses Bernstein bounds and argues
+  that we would need a bias specific dataset of size at least 11903 to
+  claim that the system is biased (three times larger than WinoBias).
+\end{itemize}
+
+\pause
+
+\begin{itemize}
+\tightlist
+\item
+  We show progress can be made with more sensitive Bayesian methods.
+\end{itemize}
+\end{block}
+
+\begin{block}{The form of the definition is suspicious}
+\protect\hypertarget{the-form-of-the-definition-is-suspicious}{}
+ETHAYARAJH UNDERSTANDING shows if there are two target words only WEAT
+is always maximal in one direction.
+
+\pause
+
+\begin{itemize}
+\tightlist
+\item
+  We show the problem runs deeper and stems from pre-averaging, and we
+  statistically gauge the uncertainty that arises from raw sample sizes.
+\end{itemize}
 \end{block}
@@
 \protect\hypertarget{outliers-and-surprisingly-dissimilar-words}{}
-We study those by visualizations and uncertainty estimates
+We study those by visualizations and uncertainty estimates.
 \end{block}
@@
 
-\begin{block}{Our simulation}
-\protect\hypertarget{our-simulation}{}
+\begin{block}{Our simulations}
+\protect\hypertarget{our-simulations}{}
 Suppose all similarities for two classes are randomly drawn from the
-same distribution, \(\mathsf{Normal}(\mu = 0, \sigma = .05)\), you still
+same distribution, \(\mathsf{Normal}(\mu = 0, \sigma = .08)\), you still
 can get a really high WEAT!
@@
 \protect\hypertarget{the-problem-with-pre-averaging-1}{}
-\begin{block}{One simulation}
-\protect\hypertarget{one-simulation}{}
+\begin{block}{Simple case: two pws, four terms}
+\protect\hypertarget{simple-case-two-pws-four-terms}{}
 \vspace{1mm}
@@
 \item
-  The WEAT score is 1.825
+  The WEAT score is 1.825\\
 \item
@@
 
-\begin{frame}{The problem with pre-averaging on real set-up}
-\protect\hypertarget{the-problem-with-pre-averaging-on-real-set-up}{}
-\begin{block}{One simulation on real set-up (no similarities and small
-std)}
-\protect\hypertarget{one-simulation-on-real-set-up-no-similarities-and-small-std}{}
+\begin{frame}{The problem with pre-averaging on realistic set-up}
+\protect\hypertarget{the-problem-with-pre-averaging-on-realistic-set-up}{}
+\begin{block}{Simulation with realistic set-up (16 predicates)}
+\protect\hypertarget{simulation-with-realistic-set-up-16-predicates}{}
 \vspace{1mm}
@@
 \begin{center}\includegraphics[width=0.7\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-8-1} \end{center}
-\end{block}
-\end{frame}
-
-\begin{frame}{The problem with pre-averaging on real set-up}
-\protect\hypertarget{the-problem-with-pre-averaging-on-real-set-up-1}{}
+
+\footnotesize
+
+\vspace{-2mm}
+
+Raw sd: 0.082, sd of means: 0.031, WEAT: 2.337.
+\end{block}
+\end{frame}
+
+\begin{frame}{The problem with pre-averaging on realistic set-up}
+\protect\hypertarget{the-problem-with-pre-averaging-on-realistic-set-up-1}{}
 \begin{block}{10k simulations (same parameters)}
@@
 \item
-  The 95\% of the data is in range -2.763, 2.698
+  95\% of the scores are in range -2.763, 2.698
 \item
@@
 
-\begin{frame}{The problem with pre-averaging on real set-up}
-\protect\hypertarget{the-problem-with-pre-averaging-on-real-set-up-2}{}
-\begin{block}{10k simulations with similarity equal to 0.1}
-\protect\hypertarget{k-simulations-with-similarity-equal-to-0.1}{}
+\begin{frame}{The problem with pre-averaging on realistic set-up}
+\protect\hypertarget{the-problem-with-pre-averaging-on-realistic-set-up-2}{}
+\begin{block}{10k simulations with mean similarity 0.1}
+\protect\hypertarget{k-simulations-with-mean-similarity-0.1}{}
 \vspace{1mm}
@@
 \item
-  The 95\% of the data is in range 0.851, 2.764
+  95\% of the scores are in range 0.851, 2.764
 \item
@@
 
-\begin{frame}{The problem with pre-averaging on real set-up}
-\protect\hypertarget{the-problem-with-pre-averaging-on-real-set-up-3}{}
-\begin{block}{10k simulations with similarity equal to 0.4}
-\protect\hypertarget{k-simulations-with-similarity-equal-to-0.4}{}
+\begin{frame}{The problem with pre-averaging on realistic set-up}
+\protect\hypertarget{the-problem-with-pre-averaging-on-realistic-set-up-3}{}
+\begin{block}{10k simulations with mean similarity 0.4}
+\protect\hypertarget{k-simulations-with-mean-similarity-0.4}{}
 \vspace{1mm}
@@
 \item
-  The 95\% of the data is in range 1.679, 2.185
+  95\% of the scores are in range 1.679, 2.185
 \item

### 2022-03-08 presentations/presentationBoston/tryOut.Rmd
presentation Boston added cosine etc.

--- 2022-02-27 presentations/presentationBoston/presentationBoston.Rmd
@@
 fontsize: 10pt
-classoption: x11names, dvipsnames, bibspacing,natbib
+classoption: x11names, dvipsnames, bibspacing, natbib, table
 urlcolor: blue
@@
 library(ggpubr)
+library(ggforce)
 library(ggExtra)
@@
 
-## Presentation plan
+## Cosine similarity
 
-- Bias in word embeddings
-- WEAT and MAC methods
-- Methodological problems
-- Limitations of pre-averaging in bias detection methods
-- Accounting for uncertainty with Bayesian approach
+\begin{columns}
+\column{0.45\linewidth}
 
-## Cosine-based measures of bias
+```{r cosine1,echo=FALSE,eval=TRUE,fig.align = "center", cache=TRUE, fig.show = "hold", out.width = "100%",   message = FALSE, warning = FALSE, results = FALSE}
+a <- c(1,2)
+b <- c(3,2.5)
 
-### Word embeddings
+ggplot()+
+  xlim(c(0,4.5))+ylim(c(0,4.5))+
+  geom_segment(aes(x = 0, y = 0, xend = a[1], yend = a[2]),  
+               arrow = arrow(length = unit(0.02, "npc")))+
+  annotate(geom = "text", label  = "a = [1,2]", x = 1.2, y = 2.4, size = 7)+
+  geom_segment(aes(x = 0, y = 0, xend = b[1], yend = b[2]), arrow = arrow(length = unit(0.02, "npc")))+   annotate(geom = "text", label  = "b = [3,2.5]", x = 3.2, y = 2.9, size = 7)+
+  theme_tufte(base_size = 24)
 
-- Representation of words with vectors of real numbers
+```
 
-- Often built to predict the probability of co-occurence
+\column{0.5\linewidth}
 
-<!-- check if the statement of co-occurence always right-->
+\footnotesize
 
-word | 1 | 2 | 3 | 4 | ... 
-|---|---|---|---|---|---
-woman | 0.456 | 0.267 | 0.675 | 0.131| ...
-man | 0.451 | 0.897 | 0.472 | 0.088| ...
+\begin{block}{Vectors}
+\begin{align*}
+a  & = [1,2]\\
+b  &= [3,2]
+\end{align*}
 
-## Cosine-based measures of bias
+\end{block}
+\pause
 
-### Cosine similarity \& distance
+\begin{block}{Dot product}
 
-<!-- - Dot product for normalized vectors -->
+\begin{align*}
+a \cdot b & = a_1 b_1 + a_2 b_2\\
+a \cdot a & = a_1^2 + a_2 ^ 2 \\
+\lVert a\rVert & = \sqrt(a \cdot a)
+\end{align*}
 
-<!-- often used in discussion of bias -->
+\end{block}
 
-\vspace{-4mm}
+\end{columns}
 
-\begin{align} \tag{Sim}
-\mathsf{cosineSimilarity}(A,B) & = \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert}
-\\
-\tag{Distance}
-\mathsf{cosineDistance}(A,B) &  = 1 - \mathsf{cosineSimilarity}(A,B)
-\end{align}
+## Cosine similarity
 
--  Geometric interpretation: direction (not length)
+\begin{columns}
+\column{0.45\linewidth}
 
--  $\mathsf{cosineDistance}\in (0, 2)$
+```{r cosine2,echo=FALSE,eval=TRUE,fig.align = "center", cache=TRUE, fig.show = "hold", out.width = "100%",   message = FALSE, warning = FALSE, results = FALSE}
+a <- c(1,2)
+b <- c(3,2.5)
+bma <- b - a
 
-- Naive interpretation: proximity corresponds to semantic similarity
+ggplot()+
+  xlim(c(0,4.5))+ylim(c(0,4.5))+
+  geom_segment(aes(x = 0, y = 0, xend = a[1], yend = a[2]),  
+               arrow = arrow(length = unit(0.02, "npc")))+
+  annotate(geom = "text", label  = "a = [1,2]", x = 1.2, y = 2.4, size = 7)+
+  geom_segment(aes(x = 0, y = 0, xend = b[1], yend = b[2]), arrow = arrow(length = unit(0.02, "npc")))+   annotate(geom = "text", label  = "b = [3,2.5]", x = 3.2, y = 2.9, size = 7)+
+  theme_tufte(base_size = 24)+
+  geom_segment(aes(x = 0, y = 0, xend = bma[1], yend = bma[2]),
+      arrow = arrow(length = unit(0.02, "npc")), color = "grey")+ 
+  annotate(geom = "text", label  = "b - a = [2,.5]", x = 2.2, y = .9, size = 7)+
+  theme_tufte(base_size = 24)
 
-<!-- - Note cosine distance fails to meet triangle inequality -->
+```
 
-## Cosine-based measures of bias
+\column{0.5\linewidth}
 
-### The worry
+\footnotesize
 
-Word embeddings can learn implicit harmful biases
+\begin{block}{Vectors}
+
+\begin{align*}
+a  & = [1,2]\\
+b  &= [4,4]
+\end{align*}
+
+\end{block}
+
+\begin{block}{Dot product}
+
+\begin{align*}
+a \cdot b & = a_1 b_1 + a_2 b_2\\
+a \cdot a & = a_1^2 + a_2 ^ 2 \\
+\lVert a\rVert & = \sqrt(a \cdot a)
+\end{align*}
+
+\end{block}
+
+\begin{block}{Vector difference}
+
+\begin{align*}
+b - a & = [b_1- a_1, b_2 - a_2 ]
+\end{align*}
+
+\end{block}
+
+\end{columns}
+
+## Cosine similarity
+
+\begin{columns}
+\column{0.45\linewidth}
+
+```{r cosine3,echo=FALSE,eval=TRUE,fig.align = "center", cache=TRUE, fig.show = "hold", out.width = "100%",   message = FALSE, warning = FALSE, results = FALSE}
+a <- c(1,2)
+b <- c(3,2.5)
+
+ggplot()+
+  xlim(c(0,4.5))+ylim(c(0,4.5))+
+  geom_segment(aes(x = 0, y = 0, xend = a[1], yend = a[2]),  
+               arrow = arrow(length = unit(0.02, "npc")))+
+  annotate(geom = "text", label  = "a = [1,2]", x = 1.2, y = 2.4, size = 7)+
+  geom_segment(aes(x = 0, y = 0, xend = b[1], yend = b[2]), arrow = arrow(length = unit(0.02, "npc")))+   annotate(geom = "text", label  = "b = [3,2.5]", x = 3.2, y = 2.9, size = 7)+
+  theme_tufte(base_size = 24)+
+  geom_segment(aes(x = a[1], y = a[2], xend = b[1], yend = b[2]),
+      arrow = arrow(length = unit(0.02, "npc")), color = "grey")+ 
+  annotate(geom = "text", label  = "b - a = [2,.5] (shifted)", x = 2, y = 2.7, size = 7)+
+  theme_tufte(base_size = 24)
+
+```
+
+\column{0.5\linewidth}
+
+\footnotesize
+
+\begin{block}{Vectors}
+
+\begin{align*}
+a  & = [1,2]\\
+b  &= [4,4]
+\end{align*}
+
+\end{block}
+
+\begin{block}{Dot product}
+
+\begin{align*}
+a \cdot b & = a_1 b_1 + a_2 b_2\\
+a \cdot a & = a_1^2 + a_2 ^ 2 \\
+\lVert a\rVert & = \sqrt(a \cdot a)
+\end{align*}
+
+\end{block}
+
+\begin{block}{Vector difference}
+
+\begin{align*}
+b - a & = [b_1- a_1, b_2 - a_2 ]
+\end{align*}
+
+\end{block}
+
+\end{columns}
+
+## Cosine similarity
+
+\begin{columns}
+\column{0.45\linewidth}
+
+```{r cosine4,echo=FALSE,eval=TRUE,fig.align = "center", cache=TRUE, fig.show = "hold", out.width = "100%",   message = FALSE, warning = FALSE, results = FALSE}
+a <- c(1,2)
+b <- c(3,2.5)
+#atan2(a[1] , a[2])
+#atan2(b[1] , b[2])
+
+ggplot()+
+  xlim(c(0,4.5))+ylim(c(0,4.5))+
+  geom_segment(aes(x = 0, y = 0, xend = a[1], yend = a[2]),  
+               arrow = arrow(length = unit(0.02, "npc")))+
+  annotate(geom = "text", label  = "a = [1,2]", x = 1.2, y = 2.4, size = 7)+
+  geom_segment(aes(x = 0, y = 0, xend = b[1], yend = b[2]), arrow = arrow(length = unit(0.02, "npc")))+   annotate(geom = "text", label  = "b = [3,2.5]", x = 3.2, y = 2.9, size = 7)+
+  theme_tufte(base_size = 24)+
+  geom_segment(aes(x = a[1], y = a[2], xend = b[1], yend = b[2]),
+      arrow = arrow(length = unit(0.02, "npc")), color = "grey")+ 
+  annotate(geom = "text", label  = "b - a = [2,.5] (shifted)", x = 2, y = 2.7, size = 7)+
+  theme_tufte(base_size = 24)+
+  annotate(geom = "text", label  = expression(theta), x = .6, y = .78, size = 7)+
+  geom_arc(aes(x0 = 0, y0 = 0, r = 1.3, 
+               start = .463, end = .876))
+
+```
+
+\column{0.5\linewidth}
+
+\footnotesize
+
+\begin{block}{Angle}
+
+\begin{align*}
+\lVert b - a \rVert^2 &  = \lVert  b \rVert^2 + \lVert a \rVert^2 - 
+  2 \lVert b \rVert   \lvert  a \rVert  \cos \theta  \\
+b \cdot a  & = \lVert b\rVert \lVert  a \rVert \cos \theta \\ 
+\cos \theta & = \frac{b \cdot a}{\lVert b\rVert  \lVert  a \rVert }
+\end{align*}
+
+\end{block}
 
@@
 
-### The basic intuition
+\begin{block}{Orthogonality}
+\begin{align*}
+\cos ( 90^{\circ}) & = 0 \\ 
+\frac{b \cdot a}{\lVert  b\rVert \lVert a \rVert} & = 0\\ 
+ b \cdot a & = 0  
+\end{align*}
+\end{block}
 
-Stereotypically connected words are cosine-close
-
-## Cosine-based measures of bias
-
-### A visual example
-
-\footnotesize
-
-- "feminine" occupations: "homemaker", "nurse", "receptionist", "librarian", etc.
-
-- "masculine" occupations:  "maestro",  "captain", "architect", "boss", etc.
-
-\normalsize
-
-\vspace{1mm}
-\footnotesize
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "60%"}
-genderProjections <-  read.csv("datasets/genderProjections.csv")[,-1]
-colnames(genderProjections)[4] <- "Stereotype" 
-levels(genderProjections$Stereotype) <- c("male occupation", "female occupation")
-ggplot(genderProjections, aes(x=he,y=she, color = Stereotype, shape = Stereotype))+geom_point(size = 6, alpha = 0.75)+theme_tufte(base_size = 27)+labs(title ="GloVe  on Wikipedia 2014 and Gigaword 5th ed.")+scale_color_manual(values =c("orangered4","chartreuse4"))+
-  theme(legend.position = c(0.8,.15)) +
-    scale_shape_manual(values=c(15, 19)) 
-```
-\normalsize
-
-<!-- ## Cosine-based measures of bias -->
-
-<!-- ### Example: direct bias -->
-
-<!-- - The gender bias of a word $w$ is its projection on the gender direction $\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})$ -->
-
-<!-- <!-- (the gender direction is the top principal compontent of ten gender pair difference vectors).  -->
-
-<!-- -  Given the (ideally) gender neutral words $N$ and the gender direction $g$ the direct gender bias is: -->
-<!-- <!-- ($c$ is a parameter determining how strict we want to be): -->
-
-<!-- \vspace{-2mm} -->
-
-<!-- \begin{align} -->
-<!-- \mathsf{directBias_c(N,g)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},g)\vert^c}{\vert N \vert } -->
-<!-- \end{align} -->
-
-<!-- \footnotesize  -->
-
-<!-- [@Bolukbasi2016Man] -->
-
-## Cosine-based measures of bias
-
-### Example: Word Embedding Association Test (WEAT)
-
-\vspace{1mm}
-\footnotesize
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "55%"}
-words <- c("man", "he", "boss", "architect","nurse","receptionist","woman","she")
-x <- c(.1,.1,.5,.5,.7,.7,1.2,1.2)
-y <- c(.4,.3,.3,.2,.5,.4,.4,.3)
-weat <- data.frame(words,x,y)
-
-ggplot(weat, aes(x=x,y=y, label = words))+geom_point()+geom_text(check_overlap = TRUE,
-                                                                 nudge_y = 0.018, size = 7 )+
-    geom_segment(data=weat, mapping=aes(x=.1, y=.3, xend=.5, yend=.2), , linetype =2, arrow=arrow(), size=.05, color="blue") +
-  geom_segment(data=weat, mapping=aes(x=.1, y=.3, xend=.5, yend=.3), linetype =2, arrow=arrow(), size=.05, color="blue")+
-  geom_segment(data=weat, mapping=aes(x=.1, y=.3, xend=.7, yend=.4), linetype =2, arrow=arrow(), size=.05, color="orangered")+
-  geom_segment(data=weat, mapping=aes(x=.1, y=.3, xend=.7, yend=.5), linetype =2, arrow=arrow(), size=.05, color="orangered")+
-    annotate("text", x = c(.32,.32,.32,.32), y = c(.23,.29,.35,.39), 
-             label = c(".7", ".6",".1", ".2") , color="black", 
-             size=7,  fontface="bold")+
-    annotate("text", x = c(.1,.5,.7,1.2), y = c(.55,.55,.55,.55), 
-           label = c("X", "A","B", "Y") , color=c("black", "blue","orangered","black"), 
-           size=10,  fontface="bold")+xlim(0,1.3)+
-  theme_void(base_size = 25)
-```
-\normalsize
-
-\pause
-
-\footnotesize
-
-- $s_1 = s(he,A,B) =  \frac{.6+.7}{2}  - \frac{.2+.1}{2} = .65-.15= .5$
-
-- $s_2 = s(man,A,B) = .3$, \linebreak  $s_3 = s(woman,A,B) = -.6, s_4 = s(she, A, B) = -.3$
-
-\vspace{-4mm}
-
-\normalsize 
-\begin{align*}
-\mathsf{WEAT}(A,B) & = \frac{\nicefrac{(s_1+s_2)}{2} - \nicefrac{(s_3+s_4)}{2}}{sd(\{s_1,s_2,s_3,s_4\})} \approx 1.93
-\end{align*}
-
-## Cosine-based measures of bias
-
-### Example: Word Embedding Association Test (WEAT)
-
-\begin{align*}
-s(t,A,B) & = \frac{\sum_{a\in A}f(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}f(t,b)}{\vert B\vert}
-\\
-WEAT(A,B) & = \frac{
-\mu\left(\{s(x,A,B)\}_{x\in X}\right) -\mu\left(\{s(y,A,B)\}_{y\in Y}\right) 
-}{
-\sigma\left(\{s(w,A,B)\}_{w\in X\cup Y}\right)
-}
-\end{align*}
-
--  $t$ is a term,  $A, B$ are sets of stereotype attribute words, $X$, $Y$ are protected group words
-
-- For instance, $X$ might be a set of male names, $Y$ a set of female names, $A$ might contain stereotypically male-related career words, and $B$ stereotypically female-related family words
-
-- $s$-values are used as datapoints in statistical significance tests
-
-\footnotesize
-
-\vspace{2mm}
-[@Caliskan2017semanticsBiases] with extensions in [@Lauscher2019multidimensional] and applications in [@Garg2018years]
-
-## Cosine-based measures of bias
-
-### Our main target: Mean Average Cosine Similarity (MAC)
-
-\vspace{1mm}
-\footnotesize
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "50%"}
-words <- c("jew", "muslim", "christian", "terrorist","dirty","conservative","familiar","greedy","cheap")
-x <- c(.1,.1,.1,.4,.4,.7,.7,1.1,1.1)
-y <- c(.4,.3,.2,.1,.2,.3,.4,.1,.2)
-mac <- data.frame(words,x)
-
-ggplot(mac, aes(x=x,y=y, label = words))+geom_point()+geom_text(check_overlap = TRUE,
-                                                                 nudge_y = 0.02, size = 9 )+
-  geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=.4, yend=.1), , linetype =2, arrow=arrow(), size=.03, color="blue")+geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=.4, yend=.2), , linetype =4, arrow=arrow(), size=.0, color="blue") + geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=.7, yend=.3), linetype =2, arrow=arrow(), size=.05, color="orangered")+ geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=.7, yend=.4), linetype =2, arrow=arrow(), size=.05, color="orangered")+ geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=1.1, yend=.1), linetype =3, arrow=arrow(), size=.05, color="chartreuse4")+ geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=1.1, yend=.2), linetype =2, arrow=arrow(), size=.05, color="chartreuse4")+ annotate("text", x = c(.1,.4,.7,1.2), y = c(.45,.45,.45,.45),  label = c("T", "A[1]","A[2]", "A[3]") ,parse = TRUE, color=c("black", "blue","orangered","chartreuse4"),            size=10,  fontface="bold")+xlim(0,1.3)+theme_void(base_size = 25)
-
-```
-\normalsize
-
-\vspace{-4mm}
-
-\footnotesize 
-\begin{align*}
- s_1 & = s(muslim,A_1) = \frac{cos(muslim,dirty)+cos(muslim,terrorist)}{2}\\
-s_2 & = s(muslim,A_2) = \frac{cos(muslim,familiar)+cos(muslim,conservative)}{2}\\
-& \vdots \end{align*}
-
-\vspace{-6mm}
-
-\normalsize 
-\begin{align*}
- MAC(T,A) & = \mathsf{mean}(\{s_i \vert i \in 1, \dots, k\})
-\end{align*}
-
-## Cosine-based measures of bias
-
-### Our main target: Mean Average Cosine Similarity (MAC)
-
-<!-- - more general, multi-class setting -->
-
-\begin{align*}
-S(t_i, A_j) & = \frac{1}{\vert A_j\vert}\sum_{a\in A_j}\mathsf{cos}(t,a) \\
-MAC(T,A) & = \frac{1}{\vert T \vert \,\vert A\vert}\sum_{t_i \in T }\sum_{A_j \in A} S(t_i,A_j)
-\end{align*}
-
-- $T = \{t_1, \dots, t_k\}$ is a class of protected words
-
-- each $A_j\in A$ is a set of attributes stereotypically associated with a protected word
-
-<!-- - For each protected word $T$ and each attribute class, they first take the mean for this protected word and all attributes in a given attribute class, and then take the mean of thus obtained means for all the protected words.
--->
-
-- The t-tests they employ are run on average cosines used to calculate MAC
-
-\vspace{2mm}
-
-\footnotesize
-
-[@Manzini2019blackToCriminal]
-
-## Cosine-based measures of bias
-
-### Our main target: Mean Average Cosine Similarity (MAC)
-
-\footnotesize 
-```{r religionTableHeadEarly,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
-religion <- read.csv("datasets/religionReddit.csv")[-1]
-colnames(religion) <- c("protectedWord","wordToCompare","wordClass",
-                        "cosineDistance","cosineSimilarity","connection")
-religion$wordClass <- as.factor(religion$wordClass)
-levels(religion$wordClass) <- c("christian","human","jewish","muslim","neutral")
-religionSmall <- religion[religion$wordClass != "neutral" & religion$wordClass != "human" ,]
-set.seed(127)
-religionSmall <- religionSmall[sample(1:nrow(religionSmall),6),-c(3,6)]
-rownames(religionSmall) <- c()
-religionSmall  %>%  kable(format = "latex",booktabs=T,
-                      linesep = "",  escape = FALSE, 
-                      caption = "A few rows from the religion dataset") %>%
-                      kable_styling(latex_options=c("scale_down"))
-```
-\normalsize
-
-## Cosine-based measures of bias
-
-### Known challenges
-
-- Gender-direction: insufficient indicator of
-    bias  \footnotesize  [@Gonen2019lipstick]
-
-\normalsize
-
-- Use of analogies: unreliable \footnotesize  [@Nissim2020fair] \normalsize
-
-- High sensitivity to irrelevant factors \footnotesize  [@zhang2020robustness] \normalsize
-
-## Some methodological problems
-
-###  Word list choice is unprincipled
-
-We run with it for comparison
-
-\pause
-
-### No design considerations to sample size
-
-We statistically gauge the uncertainty that arises from raw sample sizes
-
-## Some methodological problems
-
-### No word class distinction and no control group
-
-We make the subclasses clear, add human neutral predicates and neutral predicates for control. We used L2-Reddit corpus and GoogleNews (we present the results for Reddit for brevity).
-
-\footnotesize 
-```{r religionTableHeadLate,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
-religion <- read.csv("datasets/religionReddit.csv")[-1]
-colnames(religion) <- c("protectedWord","wordToCompare","wordClass",
-                        "cosineDistance","cosineSimilarity","connection")
-religion$wordClass <- as.factor(religion$wordClass)
-levels(religion$wordClass) <- c("christian","human","jewish","muslim","neutral")
-religionSmall <- religion[religion$wordClass != "neutral" & religion$wordClass != "human" ,]
-religionControl <- religion[religion$wordClass == "neutral" | religion$wordClass == "human" ,]
-set.seed(123154196)
-religionDisplayA <- religionSmall[sample(1:nrow(religionSmall),4),]
-religionDisplayB <- religionControl[sample(1:nrow(religionControl),4),]
-religionDisplay <- rbind(religionDisplayA,religionDisplayB)
-religionDisplay$cosineDistance <- round(religionDisplay$cosineDistance,3)
-religionDisplay$cosineSimilarity <- round(religionDisplay$cosineSimilarity,3)
-rownames(religionDisplay) <- c()
-religionDisplay  %>%  kable(format = "latex",booktabs=T,
-                      linesep = "",  escape = FALSE, 
-                      caption = "Rows from extended religion dataset.") %>%
-                      kable_styling(latex_options=c("scale_down"))
-```
-\normalsize
-
-## Some methodological problems
-
-### Outliers and surprisingly dissimilar words
-
-We study those by  visualizations and uncertainty estimates
-
-## Some methodological problems
-
-### Distances for  "muslim"
-
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "80%",message =FALSE, warning = FALSE}
-library(tidyverse)
-muslimWords <- c("imam","islam","mosque","muslim","quran")
-muslim <- religion %>% filter(protectedWord %in% muslimWords)
-muslimClass <- muslim %>% filter(protectedWord == "muslim")
-neutralSample <- sample_n(filter(muslimClass,connection == "none"), 5)
-humanSample <- sample_n(filter(muslimClass,connection == "human"), 5)
-muslimVis <- muslimClass %>% filter(connection != "none" & connection !="human")
-muslimVis <- rbind(muslimVis,neutralSample,humanSample)
-#we plug in our visualisation script
-source("../../functions/visualisationTools.R")
-#two arguments: dataset and protected word
-visualiseProtected(muslimVis,"muslim")+theme_void(base_size = 20)
-```
-
-## Some methodological problems
-
-### Distances for  "priest"
-
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "80%"}
-library(tidyverse)
-priestClass <- religion %>% filter(protectedWord == "priest")
-neutralSample <- sample_n(filter(priestClass,connection == "none"), 5)
-humanSample <- sample_n(filter(priestClass,connection == "human"), 5)
-priestVis <- priestClass %>% filter(connection != "none" & connection !="human")
-priestVis <- rbind(priestVis,neutralSample,humanSample)
-
-#we plug in our visualisation script
-source("../../functions/visualisationTools.R")
-#two arguments: dataset and protected word
-visualiseProtected(priestVis,"priest")+theme_void(base_size = 20)
-```
-
-## Some methodological problems
-
-### No principled interpretation
-
-Religion Debiasing  | MAC (distance)
-------------- | -------------
-Biased        | 0.859
-Hard Debiased | 0.934
-Soft Debiased ($\lambda$ = 0.2) | 0.894
-
-- What values are sufficient for the presence of bias and what differences are sign of real improvement?
-
-- Low $p$-values are not high effect indicators!
-
-- We compare HPDIs.
-
-<!--  What may attract attention is the fact that the value of cosine distance in "Biased" category is already quite high even before debiasing.
-High cosine distance indicates low cosine similarity between values.
-One could think that average cosine similarity equal to approximately 0.141 is not significant enough to consider it as bias.
-However the authors aim to mitigate "biases" in vectors with such great distance to make it even larger.
-Methodologically the question is, on what basis is this small similarity still considered as a proof of the presence of bias, and whether these small changes are meaningful.
-This is in general the problem of scale and the lack of universal intervals.
-In contrast, statistical intervals will help us decide whether a given cosine similarity is high enough to consider the words to be more similar than if we chose them at random.
-We will use highest posterior density intervals, in line with Bayesian methodology.
--->
-
-## The problem with pre-averaging
-
-### Key conceptual issues
-
-- It throws away information about sample sizes
-- It ignores variation in the raw data, which leads to false confidence
-
-\pause
-
-### Our simulation
-
-Suppose all similarities for two classes are randomly drawn from the same distribution, $\mathsf{Normal}(\mu = 0, \sigma = .05)$, you still can get a really high WEAT!
-
-## The problem with pre-averaging
-
-### One simulation
-
-\vspace{1mm}
-\footnotesize
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "70%"}
-set.seed(123)
-t1 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
-t2 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
-t3 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
-t4 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
-xlimits <- c(-.3,.3)
-
-t1Plot <- ggplot(t1)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t1 in X")+xlim(xlimits)
-
-t2Plot <- ggplot(t2)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t2 in X")+xlim(xlimits)
-
-t3Plot <- ggplot(t3)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t3 in Y")+xlim(xlimits)
-
-t4Plot <- ggplot(t4)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t4 in Y")+xlim(xlimits)
-
-grid.arrange(t1Plot,t2Plot, t3Plot, t4Plot, ncol=2)
-```
-\normalsize
-
-\vspace{1mm}
-\footnotesize
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
-s <- function (table){ mean(table$A) - mean(table$B)}
-whole <- rbind(t1,t2,t3,t4)
-rawsd <- sd(c(whole$A,whole$B))
-factor <- sd(c(s(t1),s(t2),s(t3),s(t4)))
-numerator <-  mean(s(t1),s(t2)) - mean(s(t3),s(t4))
-```
-\normalsize
-\pause
-
-\footnotesize
-
-\vspace{-2mm}
-
-- Raw sd in data is `r round(rawsd,3)`
-- The sd of means is `r round(factor,3)`
-- The WEAT score is `r round(numerator/factor,3)`
-- The largest effect size reported by @Caliskan2017semanticsBiases is 1.81!
-
-## The problem with pre-averaging
-
-### 50k simulations (same parameters)
-
-\vspace{1mm}
-\footnotesize
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "80%"}
-set.seed(123)
-biasesNull <- numeric(50000)
-for(i in 1:50000){
-t1 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
-t2 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
-t3 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
-t4 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
-
-factor <- sd(c(s(t1),s(t2),s(t3),s(t4)))
-numerator <-  mean(s(t1),s(t2)) - mean(s(t3),s(t4))
-biasesNull[i]  <- numerator / factor
-}
-
-ggplot()+geom_histogram(aes(x=biasesNull, y = ..density..), alpha = 0.6, bins=50)+
-  theme_tufte(base_size = 15)+labs(title="50k biases for identical means and sd =.05")+ xlab("bias")
-```
-\normalsize
-
-## Advantages of the Bayesian way
-
-- Direct impact of sample sizes
-- Straightforward interpretation in terms of posterior probabilities
-- Freedom to choose granularity level
-- More honest risk assessment and decision making
-
-## Bayesian model
-
-### Choosing priors
-
-```{r priorsVis,echo=FALSE,eval=TRUE,fig.align = "center",cache=FALSE, out.width = "100%", fig.caption = "Regularizing priors for the bayesian models."}
-library(ggpubr)
-sim <- seq(-1,1,by = 0.001)
-dis <- 1-sim
-
-p <- ggplot(data = data.frame(distance = dis), 
-            mapping = aes(x = distance))+theme_tufte(base_size = 15)
-
-none <- function(x) dnorm(x,1,.5)
-cau <- function(x) dcauchy(x,0,1)
-par <- function(x) dnorm(x,0,1)
-
-noneG <- p + stat_function(fun = none)+xlim(c(-0.5,2.5))+ggtitle("Prior for mean distances")
-parG <- p + stat_function(fun = par)+xlim(c(-2,2))+xlab("distance change")+ggtitle("Prior for coefficients")
-cauG <- p + stat_function(fun = cau)+xlim(c(0,2))+ggtitle("Prior for standard deviation")+xlab("distance change")
-ggarrange(noneG,cauG, parG, ncol = 1)
-```
-
-## Bayesian model architecture
-
-\vspace{1mm}
-\footnotesize
-```{r,echo=TRUE,eval=FALSE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
-library(rethinking)
-options(buildtools.check = function(action) TRUE )
-religionCoefs <- ulam(
-  alist(
-    cosineDistance ~ dnorm(mu,sigma),
-    mu <- m + co[con],
-    m ~ dnorm(1,.5),
-    co[con] ~ dnorm(0,.5),
-    sigma ~ dcauchy(0,1)
-  ),
-  data = religion,
-  chains=2 , iter=8000 , warmup=1000, 
-  log_lik = TRUE
-)
-```
-\normalsize
-
-##   Dataset-level coefficients
-
-### Religion with 89\%-compatibility intervals (HPDI)
-
-\begin{center}
-\begin{figure}[!htb]\centering
-   \begin{minipage}{0.55\textwidth}
-  \includegraphics[width=6cm]{../../images/religionCoeffs.jpeg}
-   \end{minipage} \begin{minipage}{0.4\textwidth}\footnotesize
-   \begin{itemize}
-   \item All HPDIs overlap with 0 
-   \item Differences between classes are relatively small
-   \item Coefficients for Race are similar
-   \end{itemize}
-   \end{minipage}
-  \end{figure}
-
-\end{center}
-
-##  Dataset-level  coefficients
-
-### Gender with 89\%-compatibility intervals (HPDI)
-
-\begin{center}
-\begin{figure}[!htb]\centering
-  \begin{minipage}{0.55\textwidth}
-\includegraphics[width=6cm]{../../images/genderCoeffs.jpeg}
-\end{minipage}
-\begin{minipage}{0.4\textwidth}\footnotesize
-   \begin{itemize}
-   \item Associated and different are away from 0
-   \item But they were supposed to be opposites and are very close to each other (co-occurrence?)
-   \item Differences between classes are still relatively small
-   \end{itemize}
-   \end{minipage}
-\end{figure}
-
-\end{center}
-
-## Bayesian model architecture
-
-\vspace{1mm}
-\footnotesize
-```{r,echo=TRUE,eval=FALSE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
-library(rethinking)
-options(buildtools.check = function(action) TRUE )
-religionCoefs <- ulam(
-  alist(
-    cosineDistance ~ dnorm(mu,sigma),
-    mu <- m[pw] + co[con],
-    m[pw] ~ dnorm(1,.5),
-    co[con] ~ dnorm(0,.5),
-    sigma ~ dcauchy(0,1)
-  ),
-  data = religion,
-  chains=2 , iter=8000 , warmup=1000, 
-  log_lik = TRUE
-)
-```
-\normalsize
-
-## Word-level coefficients
-
-\begin{figure}[!htb]\centering
-  \begin{minipage}{0.55\textwidth}
-\includegraphics[width=7.5cm]{../../images/visReligionReddit.png}
-\end{minipage}
-\begin{minipage}{0.4\textwidth}\footnotesize
-
-\vspace{-4cm}
-
-   \begin{itemize}
-   \item Most intervals overlap with control groups
-   \item Often not too much difference between associated and different
-   \end{itemize}
-   \end{minipage}
-\end{figure}
-
-## Word-level coefficients
-
-\begin{figure}[!htb]\centering
-  \begin{minipage}{0.55\textwidth}
-\includegraphics[width=7.5cm]{../../images/visGenderReddit.png}
-\end{minipage}
-\begin{minipage}{0.4\textwidth}\footnotesize
-
-\vspace{-4cm}
-
-   \begin{itemize}
-   \item Male attributes: strong co-occurrence with female attributes
-   \item Sometimes \textsf{different} is closer than \textsf{associated}
-   \item Almost no overlap with control groups
-   \end{itemize}
-   \end{minipage}
-\end{figure}
-
-## Word-level coefficients
-
-\begin{figure}[!htb]\centering
-  \begin{minipage}{0.55\textwidth}
-\includegraphics[width=7.5cm]{../../images/visRaceReddit.png}
-\end{minipage}
-\begin{minipage}{0.4\textwidth}\footnotesize
-
-\vspace{-4cm}
-
-   \begin{itemize}
-   \item A lot of variation between races
-
-   \item Often not much difference between associated and different
-   \end{itemize}
-   \end{minipage}
-\end{figure}
-
-## Thank you!
-
-### Summary
-
-- Bias in word embeddings
-- WEAT and MAC methods
-- Methodological problems
-- Limitations of pre-averaging in bias detection methods
-- Accounting for uncertainty with Bayesian approach
-
-### Further work
-
-- Including contrasts in Bayesian calculation
-- Performance cross-validation in comparison to other methods (regular linear regression, KNN, \dots)
-- Downstream tasks and connection with intrinsic evaluation
-- Testing data from the original Implicit Association Test (IAT) 
-- Applying uncertainty to WEAT and better word lists
-- Looking at other similarity measures
-
-## References
-\tiny
+\end{columns}

### 2022-03-08 presentations/presentationBoston/tryOut2.Rmd
presentation Boston added cosine etc.

--- 2022-02-27 presentations/presentationBoston/presentationBoston.Rmd
@@
 
-## Presentation plan
+## Word2vec
 
-- Bias in word embeddings
-- WEAT and MAC methods
-- Methodological problems
-- Limitations of pre-averaging in bias detection methods
-- Accounting for uncertainty with Bayesian approach
+### Question
 
-## Cosine-based measures of bias
-
-### Word embeddings
-
-- Representation of words with vectors of real numbers
-
-- Often built to predict the probability of co-occurence
-
-<!-- check if the statement of co-occurence always right-->
-
-word | 1 | 2 | 3 | 4 | ... 
-|---|---|---|---|---|---
-woman | 0.456 | 0.267 | 0.675 | 0.131| ...
-man | 0.451 | 0.897 | 0.472 | 0.088| ...
-
-## Cosine-based measures of bias
-
-### Cosine similarity \& distance
-
-<!-- - Dot product for normalized vectors -->
-
-<!-- often used in discussion of bias -->
-
-\vspace{-4mm}
-
-\begin{align} \tag{Sim}
-\mathsf{cosineSimilarity}(A,B) & = \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert}
-\\
-\tag{Distance}
-\mathsf{cosineDistance}(A,B) &  = 1 - \mathsf{cosineSimilarity}(A,B)
-\end{align}
-
--  Geometric interpretation: direction (not length)
-
--  $\mathsf{cosineDistance}\in (0, 2)$
-
-- Naive interpretation: proximity corresponds to semantic similarity
-
-<!-- - Note cosine distance fails to meet triangle inequality -->
-
-## Cosine-based measures of bias
-
-### The worry
-
-Word embeddings can learn implicit harmful biases
+How to sensibly represent words with numbers?
 
@@
 
-### The basic intuition
+### One-hot encoding
 
-Stereotypically connected words are cosine-close
+Well, you could use 30k binary vectors with a slot for each lexical unit\dots
 
-## Cosine-based measures of bias
+\pause
 
-### A visual example
+\dots \dots but this would be inefficient and wouldn't capture any relations between words.
+
+## Word2vec
+
+\begin{center}
+ \includegraphics[width = 9cm]{images/perceptron1.png}
+\end{center}
+
+\vspace{-3mm}
+\tiny \hfill \color{gray}Illustration: M. Mitchell \color{black}
 
@@
 
-- "feminine" occupations: "homemaker", "nurse", "receptionist", "librarian", etc.
+\begin{block}{Rosenblatt's perceptron}
+\begin{itemize}
+\item Inputs (pixel intensities) with weights
+\item Nodes with activation levels from 0-1
+\item (Perhaps) 0-1 output based on a threshold
+\end{itemize}
+\end{block}
 
-- "masculine" occupations:  "maestro",  "captain", "architect", "boss", etc.
+## Word2vec
 
-\normalsize
+\begin{center}
+ \includegraphics[width = 9cm]{images/perceptron1.png}
+\end{center}
 
-\vspace{1mm}
+\vspace{-3mm}
+\tiny \hfill \color{gray}Illustration: M. Mitchell \color{black}
+
 \footnotesize
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "60%"}
-genderProjections <-  read.csv("datasets/genderProjections.csv")[,-1]
-colnames(genderProjections)[4] <- "Stereotype" 
-levels(genderProjections$Stereotype) <- c("male occupation", "female occupation")
-ggplot(genderProjections, aes(x=he,y=she, color = Stereotype, shape = Stereotype))+geom_point(size = 6, alpha = 0.75)+theme_tufte(base_size = 27)+labs(title ="GloVe  on Wikipedia 2014 and Gigaword 5th ed.")+scale_color_manual(values =c("orangered4","chartreuse4"))+
-  theme(legend.position = c(0.8,.15)) +
-    scale_shape_manual(values=c(15, 19)) 
-```
-\normalsize
 
-<!-- ## Cosine-based measures of bias -->
+\begin{block}{Learning}
+\begin{itemize}
+\item Start with random weights
+\item Test on a case:
+\begin{itemize}
+\item If right, don't change weights.
+\item If wrong, change weights a bit, with focus on the ones more responsible for the judgment:
+\begin{align*}
+w_j & \leftarrow w_j = \overbrace{\eta}^{\text{learning rate}}(\underbrace{t}_{\text{correct output}} - \overbrace{y}^{\text{actual output}})\underbrace{x_j}_{\text{actual input}}
+\end{align*}
+\end{itemize}
+\end{itemize}
 
-<!-- ### Example: direct bias -->
+\end{block}
 
-<!-- - The gender bias of a word $w$ is its projection on the gender direction $\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})$ -->
+## Word2vec
 
-<!-- <!-- (the gender direction is the top principal compontent of ten gender pair difference vectors).  -->
+\begin{columns}
+\column{0.45\linewidth}
 
-<!-- -  Given the (ideally) gender neutral words $N$ and the gender direction $g$ the direct gender bias is: -->
-<!-- <!-- ($c$ is a parameter determining how strict we want to be): -->
+\begin{center}
+ \includegraphics[height = 6cm, width = 5cm]{images/perceptron2.png}
+\end{center}
 
-<!-- \vspace{-2mm} -->
+\vspace{-3mm}
+\tiny \hfill \color{gray}Illustration: M. Mitchell \color{black}
 
-<!-- \begin{align} -->
-<!-- \mathsf{directBias_c(N,g)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},g)\vert^c}{\vert N \vert } -->
-<!-- \end{align} -->
+\column{0.5\linewidth}
 
-<!-- \footnotesize  -->
+\footnotesize
 
-<!-- [@Bolukbasi2016Man] -->
+\begin{itemize}
+\item Each hidden unit takes a weighted sum of 324 inputs and passes on its activation level as input to outer layer units.
 
-## Cosine-based measures of bias
+\item Activation levels of outer layers are interpreted as network's levels of confidence in a classification problem.
 
-### Example: Word Embedding Association Test (WEAT)
+\item Learning: back-propagation (gradient descent: approximate  the direction of steepest descent in the error surface w.r.t to weights, modify accordingly).
+\end{itemize}
 
-\vspace{1mm}
-\footnotesize
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "55%"}
-words <- c("man", "he", "boss", "architect","nurse","receptionist","woman","she")
-x <- c(.1,.1,.5,.5,.7,.7,1.2,1.2)
-y <- c(.4,.3,.3,.2,.5,.4,.4,.3)
-weat <- data.frame(words,x,y)
+\end{columns}
 
-ggplot(weat, aes(x=x,y=y, label = words))+geom_point()+geom_text(check_overlap = TRUE,
-                                                                 nudge_y = 0.018, size = 7 )+
-    geom_segment(data=weat, mapping=aes(x=.1, y=.3, xend=.5, yend=.2), , linetype =2, arrow=arrow(), size=.05, color="blue") +
-  geom_segment(data=weat, mapping=aes(x=.1, y=.3, xend=.5, yend=.3), linetype =2, arrow=arrow(), size=.05, color="blue")+
-  geom_segment(data=weat, mapping=aes(x=.1, y=.3, xend=.7, yend=.4), linetype =2, arrow=arrow(), size=.05, color="orangered")+
-  geom_segment(data=weat, mapping=aes(x=.1, y=.3, xend=.7, yend=.5), linetype =2, arrow=arrow(), size=.05, color="orangered")+
-    annotate("text", x = c(.32,.32,.32,.32), y = c(.23,.29,.35,.39), 
-             label = c(".7", ".6",".1", ".2") , color="black", 
-             size=7,  fontface="bold")+
-    annotate("text", x = c(.1,.5,.7,1.2), y = c(.55,.55,.55,.55), 
-           label = c("X", "A","B", "Y") , color=c("black", "blue","orangered","black"), 
-           size=10,  fontface="bold")+xlim(0,1.3)+
-  theme_void(base_size = 25)
-```
-\normalsize
+## Word2vec
+
+### Distributional semantics
+
+\begin{itemize}
+\item "You shall know a word by the company it keeps" (John Firth, 1957)
+
+\item "the degree of semantic similarity between two linguistic expressions $A$ and $B$ is a function of the similarity of the linguistic contexts in which $A$ and $B$ can appear." (A. Lenci, 2008)
+
+\end{itemize}
 
@@
 
-\footnotesize
+\begin{center}
+ \includegraphics[height = 5cm, width = 5cm]{images/similarity1.png}
+\end{center}
 
-- $s_1 = s(he,A,B) =  \frac{.6+.7}{2}  - \frac{.2+.1}{2} = .65-.15= .5$
+\vspace{-3mm}
+\tiny \hfill \color{gray}Illustration: M. Mitchell \color{black}
 
-- $s_2 = s(man,A,B) = .3$, \linebreak  $s_3 = s(woman,A,B) = -.6, s_4 = s(she, A, B) = -.3$
+## Word2vec
 
-\vspace{-4mm}
+### Google and Mikolov
 
-\normalsize 
-\begin{align*}
-\mathsf{WEAT}(A,B) & = \frac{\nicefrac{(s_1+s_2)}{2} - \nicefrac{(s_3+s_4)}{2}}{sd(\{s_1,s_2,s_3,s_4\})} \approx 1.93
-\end{align*}
+\emph{Efficient Estimation of Word Representation in Vector Space}, 2013
 
-## Cosine-based measures of bias
+Let's train a neural network and use vectors of weights!
 
-### Example: Word Embedding Association Test (WEAT)
+\begin{center}
+ \includegraphics[height = 5cm, width = 6cm]{images/word2vec1.png}
+\end{center}
 
-\begin{align*}
-s(t,A,B) & = \frac{\sum_{a\in A}f(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}f(t,b)}{\vert B\vert}
-\\
-WEAT(A,B) & = \frac{
-\mu\left(\{s(x,A,B)\}_{x\in X}\right) -\mu\left(\{s(y,A,B)\}_{y\in Y}\right) 
-}{
-\sigma\left(\{s(w,A,B)\}_{w\in X\cup Y}\right)
-}
-\end{align*}
+\vspace{-3mm}
+\tiny \hfill \color{gray}Illustration: M. Mitchell \color{black}
 
--  $t$ is a term,  $A, B$ are sets of stereotype attribute words, $X$, $Y$ are protected group words
+## Word2vec
 
-- For instance, $X$ might be a set of male names, $Y$ a set of female names, $A$ might contain stereotypically male-related career words, and $B$ stereotypically female-related family words
+### Google and Mikolov
 
-- $s$-values are used as datapoints in statistical significance tests
+\emph{Efficient Estimation of Word Representation in Vector Space}, 2013
 
-\footnotesize
+Let's train a neural network and use vectors of weights!
 
-\vspace{2mm}
-[@Caliskan2017semanticsBiases] with extensions in [@Lauscher2019multidimensional] and applications in [@Garg2018years]
+\begin{center}
+ \includegraphics[height = 3cm, width = 7cm]{images/word2vec2.png}
+\end{center}
 
-## Cosine-based measures of bias
+\vspace{-3mm}
+\tiny \hfill \color{gray}Illustration: M. Mitchell \color{black}
 
-### Our main target: Mean Average Cosine Similarity (MAC)
+## Word2vec
 
-\vspace{1mm}
-\footnotesize
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "50%"}
-words <- c("jew", "muslim", "christian", "terrorist","dirty","conservative","familiar","greedy","cheap")
-x <- c(.1,.1,.1,.4,.4,.7,.7,1.1,1.1)
-y <- c(.4,.3,.2,.1,.2,.3,.4,.1,.2)
-mac <- data.frame(words,x)
+### Nearest words
 
-ggplot(mac, aes(x=x,y=y, label = words))+geom_point()+geom_text(check_overlap = TRUE,
-                                                                 nudge_y = 0.02, size = 9 )+
-  geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=.4, yend=.1), , linetype =2, arrow=arrow(), size=.03, color="blue")+geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=.4, yend=.2), , linetype =4, arrow=arrow(), size=.0, color="blue") + geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=.7, yend=.3), linetype =2, arrow=arrow(), size=.05, color="orangered")+ geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=.7, yend=.4), linetype =2, arrow=arrow(), size=.05, color="orangered")+ geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=1.1, yend=.1), linetype =3, arrow=arrow(), size=.05, color="chartreuse4")+ geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=1.1, yend=.2), linetype =2, arrow=arrow(), size=.05, color="chartreuse4")+ annotate("text", x = c(.1,.4,.7,1.2), y = c(.45,.45,.45,.45),  label = c("T", "A[1]","A[2]", "A[3]") ,parse = TRUE, color=c("black", "blue","orangered","chartreuse4"),            size=10,  fontface="bold")+xlim(0,1.3)+theme_void(base_size = 25)
+\begin{itemize}
+\item \textbf{philosophy}: philosophies, credo, ethos, principles, ethic, tenets, mantra, ideology, mindset, worldview
 
-```
-\normalsize
-
-\vspace{-4mm}
-
-\footnotesize 
-\begin{align*}
- s_1 & = s(muslim,A_1) = \frac{cos(muslim,dirty)+cos(muslim,terrorist)}{2}\\
-s_2 & = s(muslim,A_2) = \frac{cos(muslim,familiar)+cos(muslim,conservative)}{2}\\
-& \vdots \end{align*}
-
-\vspace{-6mm}
-
-\normalsize 
-\begin{align*}
- MAC(T,A) & = \mathsf{mean}(\{s_i \vert i \in 1, \dots, k\})
-\end{align*}
-
-## Cosine-based measures of bias
-
-### Our main target: Mean Average Cosine Similarity (MAC)
-
-<!-- - more general, multi-class setting -->
-
-\begin{align*}
-S(t_i, A_j) & = \frac{1}{\vert A_j\vert}\sum_{a\in A_j}\mathsf{cos}(t,a) \\
-MAC(T,A) & = \frac{1}{\vert T \vert \,\vert A\vert}\sum_{t_i \in T }\sum_{A_j \in A} S(t_i,A_j)
-\end{align*}
-
-- $T = \{t_1, \dots, t_k\}$ is a class of protected words
-
-- each $A_j\in A$ is a set of attributes stereotypically associated with a protected word
-
-<!-- - For each protected word $T$ and each attribute class, they first take the mean for this protected word and all attributes in a given attribute class, and then take the mean of thus obtained means for all the protected words.
--->
-
-- The t-tests they employ are run on average cosines used to calculate MAC
-
-\vspace{2mm}
-
-\footnotesize
-
-[@Manzini2019blackToCriminal]
-
-## Cosine-based measures of bias
-
-### Our main target: Mean Average Cosine Similarity (MAC)
-
-\footnotesize 
-```{r religionTableHeadEarly,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
-religion <- read.csv("datasets/religionReddit.csv")[-1]
-colnames(religion) <- c("protectedWord","wordToCompare","wordClass",
-                        "cosineDistance","cosineSimilarity","connection")
-religion$wordClass <- as.factor(religion$wordClass)
-levels(religion$wordClass) <- c("christian","human","jewish","muslim","neutral")
-religionSmall <- religion[religion$wordClass != "neutral" & religion$wordClass != "human" ,]
-set.seed(127)
-religionSmall <- religionSmall[sample(1:nrow(religionSmall),6),-c(3,6)]
-rownames(religionSmall) <- c()
-religionSmall  %>%  kable(format = "latex",booktabs=T,
-                      linesep = "",  escape = FALSE, 
-                      caption = "A few rows from the religion dataset") %>%
-                      kable_styling(latex_options=c("scale_down"))
-```
-\normalsize
-
-## Cosine-based measures of bias
-
-### Known challenges
-
-- Gender-direction: insufficient indicator of
-    bias  \footnotesize  [@Gonen2019lipstick]
-
-\normalsize
-
-- Use of analogies: unreliable \footnotesize  [@Nissim2020fair] \normalsize
-
-- High sensitivity to irrelevant factors \footnotesize  [@zhang2020robustness] \normalsize
-
-## Some methodological problems
-
-###  Word list choice is unprincipled
-
-We run with it for comparison
+\item \textbf{sandwich}: sandwiches, burger, chicken sandwich, cheeseburger, burrito, burgers, pizza, turkey sandwich, hamburger, burritos
+\end{itemize}
 
@@
 
-### No design considerations to sample size
+### Some similarities from \emph{philosophy}
 
-We statistically gauge the uncertainty that arises from raw sample sizes
-
-## Some methodological problems
-
-### No word class distinction and no control group
-
-We make the subclasses clear, add human neutral predicates and neutral predicates for control. We used L2-Reddit corpus and GoogleNews (we present the results for Reddit for brevity).
-
-\footnotesize 
-```{r religionTableHeadLate,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
-religion <- read.csv("datasets/religionReddit.csv")[-1]
-colnames(religion) <- c("protectedWord","wordToCompare","wordClass",
-                        "cosineDistance","cosineSimilarity","connection")
-religion$wordClass <- as.factor(religion$wordClass)
-levels(religion$wordClass) <- c("christian","human","jewish","muslim","neutral")
-religionSmall <- religion[religion$wordClass != "neutral" & religion$wordClass != "human" ,]
-religionControl <- religion[religion$wordClass == "neutral" | religion$wordClass == "human" ,]
-set.seed(123154196)
-religionDisplayA <- religionSmall[sample(1:nrow(religionSmall),4),]
-religionDisplayB <- religionControl[sample(1:nrow(religionControl),4),]
-religionDisplay <- rbind(religionDisplayA,religionDisplayB)
-religionDisplay$cosineDistance <- round(religionDisplay$cosineDistance,3)
-religionDisplay$cosineSimilarity <- round(religionDisplay$cosineSimilarity,3)
-rownames(religionDisplay) <- c()
-religionDisplay  %>%  kable(format = "latex",booktabs=T,
-                      linesep = "",  escape = FALSE, 
-                      caption = "Rows from extended religion dataset.") %>%
-                      kable_styling(latex_options=c("scale_down"))
-```
-\normalsize
-
-## Some methodological problems
-
-### Outliers and surprisingly dissimilar words
-
-We study those by  visualizations and uncertainty estimates
-
-## Some methodological problems
-
-### Distances for  "muslim"
-
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "80%",message =FALSE, warning = FALSE}
-library(tidyverse)
-muslimWords <- c("imam","islam","mosque","muslim","quran")
-muslim <- religion %>% filter(protectedWord %in% muslimWords)
-muslimClass <- muslim %>% filter(protectedWord == "muslim")
-neutralSample <- sample_n(filter(muslimClass,connection == "none"), 5)
-humanSample <- sample_n(filter(muslimClass,connection == "human"), 5)
-muslimVis <- muslimClass %>% filter(connection != "none" & connection !="human")
-muslimVis <- rbind(muslimVis,neutralSample,humanSample)
-#we plug in our visualisation script
-source("../../functions/visualisationTools.R")
-#two arguments: dataset and protected word
-visualiseProtected(muslimVis,"muslim")+theme_void(base_size = 20)
-```
-
-## Some methodological problems
-
-### Distances for  "priest"
-
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "80%"}
-library(tidyverse)
-priestClass <- religion %>% filter(protectedWord == "priest")
-neutralSample <- sample_n(filter(priestClass,connection == "none"), 5)
-humanSample <- sample_n(filter(priestClass,connection == "human"), 5)
-priestVis <- priestClass %>% filter(connection != "none" & connection !="human")
-priestVis <- rbind(priestVis,neutralSample,humanSample)
-
-#we plug in our visualisation script
-source("../../functions/visualisationTools.R")
-#two arguments: dataset and protected word
-visualiseProtected(priestVis,"priest")+theme_void(base_size = 20)
-```
-
-## Some methodological problems
-
-### No principled interpretation
-
-Religion Debiasing  | MAC (distance)
-------------- | -------------
-Biased        | 0.859
-Hard Debiased | 0.934
-Soft Debiased ($\lambda$ = 0.2) | 0.894
-
-- What values are sufficient for the presence of bias and what differences are sign of real improvement?
-
-- Low $p$-values are not high effect indicators!
-
-- We compare HPDIs.
-
-<!--  What may attract attention is the fact that the value of cosine distance in "Biased" category is already quite high even before debiasing.
-High cosine distance indicates low cosine similarity between values.
-One could think that average cosine similarity equal to approximately 0.141 is not significant enough to consider it as bias.
-However the authors aim to mitigate "biases" in vectors with such great distance to make it even larger.
-Methodologically the question is, on what basis is this small similarity still considered as a proof of the presence of bias, and whether these small changes are meaningful.
-This is in general the problem of scale and the lack of universal intervals.
-In contrast, statistical intervals will help us decide whether a given cosine similarity is high enough to consider the words to be more similar than if we chose them at random.
-We will use highest posterior density intervals, in line with Bayesian methodology.
--->
-
-## The problem with pre-averaging
-
-### Key conceptual issues
-
-- It throws away information about sample sizes
-- It ignores variation in the raw data, which leads to false confidence
-
-\pause
-
-### Our simulation
-
-Suppose all similarities for two classes are randomly drawn from the same distribution, $\mathsf{Normal}(\mu = 0, \sigma = .05)$, you still can get a really high WEAT!
-
-## The problem with pre-averaging
-
-### One simulation
-
-\vspace{1mm}
-\footnotesize
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "70%"}
-set.seed(123)
-t1 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
-t2 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
-t3 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
-t4 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
-xlimits <- c(-.3,.3)
-
-t1Plot <- ggplot(t1)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t1 in X")+xlim(xlimits)
-
-t2Plot <- ggplot(t2)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t2 in X")+xlim(xlimits)
-
-t3Plot <- ggplot(t3)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t3 in Y")+xlim(xlimits)
-
-t4Plot <- ggplot(t4)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t4 in Y")+xlim(xlimits)
-
-grid.arrange(t1Plot,t2Plot, t3Plot, t4Plot, ncol=2)
-```
-\normalsize
-
-\vspace{1mm}
-\footnotesize
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
-s <- function (table){ mean(table$A) - mean(table$B)}
-whole <- rbind(t1,t2,t3,t4)
-rawsd <- sd(c(whole$A,whole$B))
-factor <- sd(c(s(t1),s(t2),s(t3),s(t4)))
-numerator <-  mean(s(t1),s(t2)) - mean(s(t3),s(t4))
-```
-\normalsize
-\pause
-
-\footnotesize
-
-\vspace{-2mm}
-
-- Raw sd in data is `r round(rawsd,3)`
-- The sd of means is `r round(factor,3)`
-- The WEAT score is `r round(numerator/factor,3)`
-- The largest effect size reported by @Caliskan2017semanticsBiases is 1.81!
-
-## The problem with pre-averaging
-
-### 50k simulations (same parameters)
-
-\vspace{1mm}
-\footnotesize
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "80%"}
-set.seed(123)
-biasesNull <- numeric(50000)
-for(i in 1:50000){
-t1 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
-t2 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
-t3 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
-t4 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,0,0.05))
-
-factor <- sd(c(s(t1),s(t2),s(t3),s(t4)))
-numerator <-  mean(s(t1),s(t2)) - mean(s(t3),s(t4))
-biasesNull[i]  <- numerator / factor
-}
-
-ggplot()+geom_histogram(aes(x=biasesNull, y = ..density..), alpha = 0.6, bins=50)+
-  theme_tufte(base_size = 15)+labs(title="50k biases for identical means and sd =.05")+ xlab("bias")
-```
-\normalsize
-
-## Advantages of the Bayesian way
-
-- Direct impact of sample sizes
-- Straightforward interpretation in terms of posterior probabilities
-- Freedom to choose granularity level
-- More honest risk assessment and decision making
-
-## Bayesian model
-
-### Choosing priors
-
-```{r priorsVis,echo=FALSE,eval=TRUE,fig.align = "center",cache=FALSE, out.width = "100%", fig.caption = "Regularizing priors for the bayesian models."}
-library(ggpubr)
-sim <- seq(-1,1,by = 0.001)
-dis <- 1-sim
-
-p <- ggplot(data = data.frame(distance = dis), 
-            mapping = aes(x = distance))+theme_tufte(base_size = 15)
-
-none <- function(x) dnorm(x,1,.5)
-cau <- function(x) dcauchy(x,0,1)
-par <- function(x) dnorm(x,0,1)
-
-noneG <- p + stat_function(fun = none)+xlim(c(-0.5,2.5))+ggtitle("Prior for mean distances")
-parG <- p + stat_function(fun = par)+xlim(c(-2,2))+xlab("distance change")+ggtitle("Prior for coefficients")
-cauG <- p + stat_function(fun = cau)+xlim(c(0,2))+ggtitle("Prior for standard deviation")+xlab("distance change")
-ggarrange(noneG,cauG, parG, ncol = 1)
-```
-
-## Bayesian model architecture
-
-\vspace{1mm}
-\footnotesize
-```{r,echo=TRUE,eval=FALSE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
-library(rethinking)
-options(buildtools.check = function(action) TRUE )
-religionCoefs <- ulam(
-  alist(
-    cosineDistance ~ dnorm(mu,sigma),
-    mu <- m + co[con],
-    m ~ dnorm(1,.5),
-    co[con] ~ dnorm(0,.5),
-    sigma ~ dcauchy(0,1)
-  ),
-  data = religion,
-  chains=2 , iter=8000 , warmup=1000, 
-  log_lik = TRUE
-)
-```
-\normalsize
-
-##   Dataset-level coefficients
-
-### Religion with 89\%-compatibility intervals (HPDI)
-
-\begin{center}
-\begin{figure}[!htb]\centering
-   \begin{minipage}{0.55\textwidth}
-  \includegraphics[width=6cm]{../../images/religionCoeffs.jpeg}
-   \end{minipage} \begin{minipage}{0.4\textwidth}\footnotesize
-   \begin{itemize}
-   \item All HPDIs overlap with 0 
-   \item Differences between classes are relatively small
-   \item Coefficients for Race are similar
-   \end{itemize}
-   \end{minipage}
-  \end{figure}
-
-\end{center}
-
-##  Dataset-level  coefficients
-
-### Gender with 89\%-compatibility intervals (HPDI)
-
-\begin{center}
-\begin{figure}[!htb]\centering
-  \begin{minipage}{0.55\textwidth}
-\includegraphics[width=6cm]{../../images/genderCoeffs.jpeg}
-\end{minipage}
-\begin{minipage}{0.4\textwidth}\footnotesize
-   \begin{itemize}
-   \item Associated and different are away from 0
-   \item But they were supposed to be opposites and are very close to each other (co-occurrence?)
-   \item Differences between classes are still relatively small
-   \end{itemize}
-   \end{minipage}
-\end{figure}
-
-\end{center}
-
-## Bayesian model architecture
-
-\vspace{1mm}
-\footnotesize
-```{r,echo=TRUE,eval=FALSE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
-library(rethinking)
-options(buildtools.check = function(action) TRUE )
-religionCoefs <- ulam(
-  alist(
-    cosineDistance ~ dnorm(mu,sigma),
-    mu <- m[pw] + co[con],
-    m[pw] ~ dnorm(1,.5),
-    co[con] ~ dnorm(0,.5),
-    sigma ~ dcauchy(0,1)
-  ),
-  data = religion,
-  chains=2 , iter=8000 , warmup=1000, 
-  log_lik = TRUE
-)
-```
-\normalsize
-
-## Word-level coefficients
-
-\begin{figure}[!htb]\centering
-  \begin{minipage}{0.55\textwidth}
-\includegraphics[width=7.5cm]{../../images/visReligionReddit.png}
-\end{minipage}
-\begin{minipage}{0.4\textwidth}\footnotesize
-
-\vspace{-4cm}
-
-   \begin{itemize}
-   \item Most intervals overlap with control groups
-   \item Often not too much difference between associated and different
-   \end{itemize}
-   \end{minipage}
-\end{figure}
-
-## Word-level coefficients
-
-\begin{figure}[!htb]\centering
-  \begin{minipage}{0.55\textwidth}
-\includegraphics[width=7.5cm]{../../images/visGenderReddit.png}
-\end{minipage}
-\begin{minipage}{0.4\textwidth}\footnotesize
-
-\vspace{-4cm}
-
-   \begin{itemize}
-   \item Male attributes: strong co-occurrence with female attributes
-   \item Sometimes \textsf{different} is closer than \textsf{associated}
-   \item Almost no overlap with control groups
-   \end{itemize}
-   \end{minipage}
-\end{figure}
-
-## Word-level coefficients
-
-\begin{figure}[!htb]\centering
-  \begin{minipage}{0.55\textwidth}
-\includegraphics[width=7.5cm]{../../images/visRaceReddit.png}
-\end{minipage}
-\begin{minipage}{0.4\textwidth}\footnotesize
-
-\vspace{-4cm}
-
-   \begin{itemize}
-   \item A lot of variation between races
-
-   \item Often not much difference between associated and different
-   \end{itemize}
-   \end{minipage}
-\end{figure}
-
-## Thank you!
-
-### Summary
-
-- Bias in word embeddings
-- WEAT and MAC methods
-- Methodological problems
-- Limitations of pre-averaging in bias detection methods
-- Accounting for uncertainty with Bayesian approach
-
-### Further work
-
-- Including contrasts in Bayesian calculation
-- Performance cross-validation in comparison to other methods (regular linear regression, KNN, \dots)
-- Downstream tasks and connection with intrinsic evaluation
-- Testing data from the original Implicit Association Test (IAT) 
-- Applying uncertainty to WEAT and better word lists
-- Looking at other similarity measures
-
-## References
-\tiny
+Logic (.47), Nietzsche (.32), Hegel (.32), analytic (.13), burger (.08), continental (.04), Russell (.04)

### 2022-03-08 presentations/presentationBoston/tryOut2.tex
presentation Boston added cosine etc.

--- 2022-03-01 presentations/presentationBoston/presentationBoston.tex
@@
   ignorenonframetext,
-  x11names, dvipsnames, bibspacing,natbib]{beamer}
+  x11names, dvipsnames, bibspacing, natbib, table]{beamer}
 \usepackage{pgfpages}
@@
 \newif\ifbibliography
-\usepackage{color}
-\usepackage{fancyvrb}
-\newcommand{\VerbBar}{|}
-\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
-\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
-% Add ',fontsize=\small' for more characters per line
-\usepackage{framed}
-\definecolor{shadecolor}{RGB}{248,248,248}
-\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\BuiltInTok}[1]{#1}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
-\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
-\newcommand{\ExtensionTok}[1]{#1}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ImportTok}[1]{#1}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\NormalTok}[1]{#1}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\RegionMarkerTok}[1]{#1}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\usepackage{longtable,booktabs,array}
-\usepackage{calc} % for calculating minipage widths
-\usepackage{caption}
-% Make caption package work with longtable
-\makeatletter
-\def\fnum@table{\tablename~\thetable}
-\makeatother
 \setlength{\emergencystretch}{3em} % prevent overfull lines
@@
 
-\begin{frame}{Presentation plan}
-\protect\hypertarget{presentation-plan}{}
-\begin{itemize}
-\tightlist
-\item
-  Bias in word embeddings
-\item
-  WEAT and MAC methods
-\item
-  Methodological problems
-\item
-  Limitations of pre-averaging in bias detection methods
-\item
-  Accounting for uncertainty with Bayesian approach
-\end{itemize}
-\end{frame}
+\begin{frame}{Word2vec}
+\protect\hypertarget{word2vec}{}
+\begin{block}{Question}
+\protect\hypertarget{question}{}
+How to sensibly represent words with numbers?
 
-\begin{frame}{Cosine-based measures of bias}
-\protect\hypertarget{cosine-based-measures-of-bias}{}
-\begin{block}{Word embeddings}
-\protect\hypertarget{word-embeddings}{}
-\begin{itemize}
-\item
-  Representation of words with vectors of real numbers
-\item
-  Often built to predict the probability of co-occurence
-\end{itemize}
+\pause
+\end{block}
 
-\begin{longtable}[]{@{}llllll@{}}
-\toprule
-word & 1 & 2 & 3 & 4 & \ldots{} \\
-\midrule
-\endhead
-woman & 0.456 & 0.267 & 0.675 & 0.131 & \ldots{} \\
-man & 0.451 & 0.897 & 0.472 & 0.088 & \ldots{} \\
-\bottomrule
-\end{longtable}
+\begin{block}{One-hot encoding}
+\protect\hypertarget{one-hot-encoding}{}
+Well, you could use 30k binary vectors with a slot for each lexical
+unit\dots
+
+\pause
+
+\dots \dots but this would be inefficient and wouldn't capture any
+relations between words.
 \end{block}
@@
 
-\begin{frame}{Cosine-based measures of bias}
-\protect\hypertarget{cosine-based-measures-of-bias-1}{}
-\begin{block}{Cosine similarity \& distance}
-\protect\hypertarget{cosine-similarity-distance}{}
-\vspace{-4mm}
+\begin{frame}{Word2vec}
+\protect\hypertarget{word2vec-1}{}
+\begin{center}
+ \includegraphics[width = 9cm]{images/perceptron1.png}
+\end{center}
 
-\begin{align} \tag{Sim}
-\mathsf{cosineSimilarity}(A,B) & = \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert}
-\\
-\tag{Distance}
-\mathsf{cosineDistance}(A,B) &  = 1 - \mathsf{cosineSimilarity}(A,B)
-\end{align}
+\vspace{-3mm}
 
+\tiny \hfill \color{gray}Illustration: M. Mitchell \color{black}
+
+\footnotesize
+
+\begin{block}{Rosenblatt's perceptron}
 \begin{itemize}
-\item
-  Geometric interpretation: direction (not length)
-\item
-  \(\mathsf{cosineDistance}\in (0, 2)\)
-\item
-  Naive interpretation: proximity corresponds to semantic similarity
+\item Inputs (pixel intensities) with weights
+\item Nodes with activation levels from 0-1
+\item (Perhaps) 0-1 output based on a threshold
 \end{itemize}
@@
 
-\begin{frame}{Cosine-based measures of bias}
-\protect\hypertarget{cosine-based-measures-of-bias-2}{}
-\begin{block}{The worry}
-\protect\hypertarget{the-worry}{}
-Word embeddings can learn implicit harmful biases
+\begin{frame}{Word2vec}
+\protect\hypertarget{word2vec-2}{}
+\begin{center}
+ \includegraphics[width = 9cm]{images/perceptron1.png}
+\end{center}
 
-\pause
-\end{block}
+\vspace{-3mm}
 
-\begin{block}{The basic intuition}
-\protect\hypertarget{the-basic-intuition}{}
-Stereotypically connected words are cosine-close
+\tiny \hfill \color{gray}Illustration: M. Mitchell \color{black}
+
+\footnotesize
+
+\begin{block}{Learning}
+\begin{itemize}
+\item Start with random weights
+\item Test on a case:
+\begin{itemize}
+\item If right, don't change weights.
+\item If wrong, change weights a bit, with focus on the ones more responsible for the judgment:
+\begin{align*}
+w_j & \leftarrow w_j = \overbrace{\eta}^{\text{learning rate}}(\underbrace{t}_{\text{correct output}} - \overbrace{y}^{\text{actual output}})\underbrace{x_j}_{\text{actual input}}
+\end{align*}
+\end{itemize}
+\end{itemize}
+
 \end{block}
@@
 
-\begin{frame}{Cosine-based measures of bias}
-\protect\hypertarget{cosine-based-measures-of-bias-3}{}
-\begin{block}{A visual example}
-\protect\hypertarget{a-visual-example}{}
-\footnotesize
+\begin{frame}{Word2vec}
+\protect\hypertarget{word2vec-3}{}
+\begin{columns}
+\column{0.45\linewidth}
 
-\begin{itemize}
-\item
-  ``feminine'' occupations: ``homemaker,'' ``nurse,'' ``receptionist,''
-  ``librarian,'' etc.
-\item
-  ``masculine'' occupations: ``maestro,'' ``captain,'' ``architect,''
-  ``boss,'' etc.
-\end{itemize}
+\begin{center}
+ \includegraphics[height = 6cm, width = 5cm]{images/perceptron2.png}
+\end{center}
 
-\normalsize
+\vspace{-3mm}
+\tiny \hfill \color{gray}Illustration: M. Mitchell \color{black}
 
-\vspace{1mm}
-\footnotesize
-
-\begin{center}\includegraphics[width=0.6\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-1-1} \end{center}
-\normalsize
-\end{block}
-\end{frame}
-
-\begin{frame}{Cosine-based measures of bias}
-\protect\hypertarget{cosine-based-measures-of-bias-4}{}
-\begin{block}{Example: Word Embedding Association Test (WEAT)}
-\protect\hypertarget{example-word-embedding-association-test-weat}{}
-\vspace{1mm}
-\footnotesize
-
-\begin{center}\includegraphics[width=0.55\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-2-1} \end{center}
-\normalsize
-
-\pause
+\column{0.5\linewidth}
 
@@
 \begin{itemize}
-\item
-  \(s_1 = s(he,A,B) = \frac{.6+.7}{2} - \frac{.2+.1}{2} = .65-.15= .5\)
-\item
-  \(s_2 = s(man,A,B) = .3\),
-  \linebreak  \(s_3 = s(woman,A,B) = -.6, s_4 = s(she, A, B) = -.3\)
+\item Each hidden unit takes a weighted sum of 324 inputs and passes on its activation level as input to outer layer units.
+
+\item Activation levels of outer layers are interpreted as network's levels of confidence in a classification problem.
+
+\item Learning: back-propagation (gradient descent: approximate  the direction of steepest descent in the error surface w.r.t to weights, modify accordingly).
 \end{itemize}
 
-\vspace{-4mm}
+\end{columns}
+\end{frame}
 
-\normalsize
+\begin{frame}{Word2vec}
+\protect\hypertarget{word2vec-4}{}
+\begin{block}{Distributional semantics}
+\protect\hypertarget{distributional-semantics}{}
+\begin{itemize}
+\item "You shall know a word by the company it keeps" (John Firth, 1957)
 
-\begin{align*}
-\mathsf{WEAT}(A,B) & = \frac{\nicefrac{(s_1+s_2)}{2} - \nicefrac{(s_3+s_4)}{2}}{sd(\{s_1,s_2,s_3,s_4\})} \approx 1.93
-\end{align*}
+\item "the degree of semantic similarity between two linguistic expressions $A$ and $B$ is a function of the similarity of the linguistic contexts in which $A$ and $B$ can appear." (A. Lenci, 2008)
+
+\end{itemize}
+
+\pause
+
+\begin{center}
+ \includegraphics[height = 5cm, width = 5cm]{images/similarity1.png}
+\end{center}
+
+\vspace{-3mm}
+
+\tiny \hfill \color{gray}Illustration: M. Mitchell \color{black}
 \end{block}
@@
 
-\begin{frame}{Cosine-based measures of bias}
-\protect\hypertarget{cosine-based-measures-of-bias-5}{}
-\begin{block}{Example: Word Embedding Association Test (WEAT)}
-\protect\hypertarget{example-word-embedding-association-test-weat-1}{}
-\begin{align*}
-s(t,A,B) & = \frac{\sum_{a\in A}f(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}f(t,b)}{\vert B\vert}
-\\
-WEAT(A,B) & = \frac{
-\mu\left(\{s(x,A,B)\}_{x\in X}\right) -\mu\left(\{s(y,A,B)\}_{y\in Y}\right) 
-}{
-\sigma\left(\{s(w,A,B)\}_{w\in X\cup Y}\right)
-}
-\end{align*}
+\begin{frame}{Word2vec}
+\protect\hypertarget{word2vec-5}{}
+\begin{block}{Google and Mikolov}
+\protect\hypertarget{google-and-mikolov}{}
+\emph{Efficient Estimation of Word Representation in Vector Space}, 2013
 
-\begin{itemize}
-\item
-  \(t\) is a term, \(A, B\) are sets of stereotype attribute words,
-  \(X\), \(Y\) are protected group words
-\item
-  For instance, \(X\) might be a set of male names, \(Y\) a set of
-  female names, \(A\) might contain stereotypically male-related career
-  words, and \(B\) stereotypically female-related family words
-\item
-  \(s\)-values are used as datapoints in statistical significance tests
-\end{itemize}
+Let's train a neural network and use vectors of weights!
 
-\footnotesize
+\begin{center}
+ \includegraphics[height = 5cm, width = 6cm]{images/word2vec1.png}
+\end{center}
 
-\vspace{2mm}
+\vspace{-3mm}
 
-(\protect\hyperlink{ref-Caliskan2017semanticsBiases}{Caliskan, Bryson,
-\& Narayanan, 2017}) with extensions in
-(\protect\hyperlink{ref-Lauscher2019multidimensional}{Lauscher \&
-Glavas, 2019}) and applications in
-(\protect\hyperlink{ref-Garg2018years}{Garg, Schiebinger, Jurafsky, \&
-Zou, 2018})
+\tiny \hfill \color{gray}Illustration: M. Mitchell \color{black}
 \end{block}
@@
 
-\begin{frame}{Cosine-based measures of bias}
-\protect\hypertarget{cosine-based-measures-of-bias-6}{}
-\begin{block}{Our main target: Mean Average Cosine Similarity (MAC)}
-\protect\hypertarget{our-main-target-mean-average-cosine-similarity-mac}{}
-\vspace{1mm}
-\footnotesize
+\begin{frame}{Word2vec}
+\protect\hypertarget{word2vec-6}{}
+\begin{block}{Google and Mikolov}
+\protect\hypertarget{google-and-mikolov-1}{}
+\emph{Efficient Estimation of Word Representation in Vector Space}, 2013
 
-\begin{center}\includegraphics[width=0.5\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-3-1} \end{center}
-\normalsize
+Let's train a neural network and use vectors of weights!
 
-\vspace{-4mm}
+\begin{center}
+ \includegraphics[height = 3cm, width = 7cm]{images/word2vec2.png}
+\end{center}
 
-\footnotesize
+\vspace{-3mm}
 
-\begin{align*}
- s_1 & = s(muslim,A_1) = \frac{cos(muslim,dirty)+cos(muslim,terrorist)}{2}\\
-s_2 & = s(muslim,A_2) = \frac{cos(muslim,familiar)+cos(muslim,conservative)}{2}\\
-& \vdots \end{align*}
-
-\vspace{-6mm}
-
-\normalsize
-
-\begin{align*}
- MAC(T,A) & = \mathsf{mean}(\{s_i \vert i \in 1, \dots, k\})
-\end{align*}
+\tiny \hfill \color{gray}Illustration: M. Mitchell \color{black}
 \end{block}
@@
 
-\begin{frame}{Cosine-based measures of bias}
-\protect\hypertarget{cosine-based-measures-of-bias-7}{}
-\begin{block}{Our main target: Mean Average Cosine Similarity (MAC)}
-\protect\hypertarget{our-main-target-mean-average-cosine-similarity-mac-1}{}
-\begin{align*}
-S(t_i, A_j) & = \frac{1}{\vert A_j\vert}\sum_{a\in A_j}\mathsf{cos}(t,a) \\
-MAC(T,A) & = \frac{1}{\vert T \vert \,\vert A\vert}\sum_{t_i \in T }\sum_{A_j \in A} S(t_i,A_j)
-\end{align*}
+\begin{frame}{Word2vec}
+\protect\hypertarget{word2vec-7}{}
+\begin{block}{Nearest words}
+\protect\hypertarget{nearest-words}{}
+\begin{itemize}
+\item \textbf{philosophy}: philosophies, credo, ethos, principles, ethic, tenets, mantra, ideology, mindset, worldview
 
-\begin{itemize}
-\item
-  \(T = \{t_1, \dots, t_k\}\) is a class of protected words
-\item
-  each \(A_j\in A\) is a set of attributes stereotypically associated
-  with a protected word
-\end{itemize}
-
-\begin{itemize}
-\tightlist
-\item
-  The t-tests they employ are run on average cosines used to calculate
-  MAC
-\end{itemize}
-
-\vspace{2mm}
-
-\footnotesize
-
-(\protect\hyperlink{ref-Manzini2019blackToCriminal}{Manzini, Lim,
-Tsvetkov, \& Black, 2019})
-\end{block}
-\end{frame}
-
-\begin{frame}{Cosine-based measures of bias}
-\protect\hypertarget{cosine-based-measures-of-bias-8}{}
-\begin{block}{Our main target: Mean Average Cosine Similarity (MAC)}
-\protect\hypertarget{our-main-target-mean-average-cosine-similarity-mac-2}{}
-\footnotesize 
-\begin{table}
-
-\caption{\label{tab:religionTableHeadEarly}A few rows from the religion dataset}
-\centering
-\resizebox{\linewidth}{!}{
-\begin{tabular}[t]{llrr}
-\toprule
-protectedWord & wordToCompare & cosineDistance & cosineSimilarity\\
-\midrule
-jew & greedy & 0.6947042 & 0.3052958\\
-rabbi & greedy & 1.0306175 & -0.0306175\\
-rabbi & conservative & 0.7175887 & 0.2824113\\
-christian & uneducated & 0.5081939 & 0.4918061\\
-christianity & cheap & 1.2816164 & -0.2816164\\
-muslim & terrorist & 0.2726106 & 0.7273894\\
-\bottomrule
-\end{tabular}}
-\end{table}
-\normalsize
-\end{block}
-\end{frame}
-
-\begin{frame}{Cosine-based measures of bias}
-\protect\hypertarget{cosine-based-measures-of-bias-9}{}
-\begin{block}{Known challenges}
-\protect\hypertarget{known-challenges}{}
-\begin{itemize}
-\tightlist
-\item
-  Gender-direction: insufficient indicator of bias
-  \footnotesize  (\protect\hyperlink{ref-Gonen2019lipstick}{Gonen \&
-  Goldberg, 2019})
-\end{itemize}
-
-\normalsize
-
-\begin{itemize}
-\item
-  Use of analogies: unreliable
-  \footnotesize  (\protect\hyperlink{ref-Nissim2020fair}{Nissim, Noord,
-  \& Goot, 2020}) \normalsize
-\item
-  High sensitivity to irrelevant factors
-  \footnotesize  (\protect\hyperlink{ref-zhang2020robustness}{Zhang,
-  Sneyd, \& Stevenson, 2020}) \normalsize
-\end{itemize}
-\end{block}
-\end{frame}
-
-\begin{frame}{Some methodological problems}
-\protect\hypertarget{some-methodological-problems}{}
-\begin{block}{Word list choice is unprincipled}
-\protect\hypertarget{word-list-choice-is-unprincipled}{}
-We run with it for comparison
-
-\pause
-\end{block}
-
-\begin{block}{No design considerations to sample size}
-\protect\hypertarget{no-design-considerations-to-sample-size}{}
-We statistically gauge the uncertainty that arises from raw sample sizes
-\end{block}
-\end{frame}
-
-\begin{frame}{Some methodological problems}
-\protect\hypertarget{some-methodological-problems-1}{}
-\begin{block}{No word class distinction and no control group}
-\protect\hypertarget{no-word-class-distinction-and-no-control-group}{}
-We make the subclasses clear, add human neutral predicates and neutral
-predicates for control. We used L2-Reddit corpus and GoogleNews (we
-present the results for Reddit for brevity).
-
-\footnotesize 
-\begin{table}
-
-\caption{\label{tab:religionTableHeadLate}Rows from extended religion dataset.}
-\centering
-\resizebox{\linewidth}{!}{
-\begin{tabular}[t]{lllrrl}
-\toprule
-protectedWord & wordToCompare & wordClass & cosineDistance & cosineSimilarity & connection\\
-\midrule
-torah & hairy & jewish & 1.170 & -0.170 & associated\\
-christian & dirty & muslim & 0.949 & 0.051 & different\\
-judaism & cheap & jewish & 1.232 & -0.232 & associated\\
-christianity & familial & christian & 0.645 & 0.355 & associated\\
-mosque & approve & neutral & 0.995 & 0.005 & none\\
-imam & carry & human & 0.993 & 0.007 & human\\
-mosque & merging & neutral & 0.868 & 0.132 & none\\
-muslim & nationalized & neutral & 0.870 & 0.130 & none\\
-\bottomrule
-\end{tabular}}
-\end{table}
-\normalsize
-\end{block}
-\end{frame}
-
-\begin{frame}{Some methodological problems}
-\protect\hypertarget{some-methodological-problems-2}{}
-\begin{block}{Outliers and surprisingly dissimilar words}
-\protect\hypertarget{outliers-and-surprisingly-dissimilar-words}{}
-We study those by visualizations and uncertainty estimates
-\end{block}
-\end{frame}
-
-\begin{frame}{Some methodological problems}
-\protect\hypertarget{some-methodological-problems-3}{}
-\begin{block}{Distances for ``muslim''}
-\protect\hypertarget{distances-for-muslim}{}
-\begin{center}\includegraphics[width=0.8\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-4-1} \end{center}
-\end{block}
-\end{frame}
-
-\begin{frame}{Some methodological problems}
-\protect\hypertarget{some-methodological-problems-4}{}
-\begin{block}{Distances for ``priest''}
-\protect\hypertarget{distances-for-priest}{}
-\begin{center}\includegraphics[width=0.8\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-5-1} \end{center}
-\end{block}
-\end{frame}
-
-\begin{frame}{Some methodological problems}
-\protect\hypertarget{some-methodological-problems-5}{}
-\begin{block}{No principled interpretation}
-\protect\hypertarget{no-principled-interpretation}{}
-\begin{longtable}[]{@{}ll@{}}
-\toprule
-Religion Debiasing & MAC (distance) \\
-\midrule
-\endhead
-Biased & 0.859 \\
-Hard Debiased & 0.934 \\
-Soft Debiased (\(\lambda\) = 0.2) & 0.894 \\
-\bottomrule
-\end{longtable}
-
-\begin{itemize}
-\item
-  What values are sufficient for the presence of bias and what
-  differences are sign of real improvement?
-\item
-  Low \(p\)-values are not high effect indicators!
-\item
-  We compare HPDIs.
-\end{itemize}
-\end{block}
-\end{frame}
-
-\begin{frame}{The problem with pre-averaging}
-\protect\hypertarget{the-problem-with-pre-averaging}{}
-\begin{block}{Key conceptual issues}
-\protect\hypertarget{key-conceptual-issues}{}
-\begin{itemize}
-\tightlist
-\item
-  It throws away information about sample sizes
-\item
-  It ignores variation in the raw data, which leads to false confidence
+\item \textbf{sandwich}: sandwiches, burger, chicken sandwich, cheeseburger, burrito, burgers, pizza, turkey sandwich, hamburger, burritos
 \end{itemize}
@@
 
-\begin{block}{Our simulation}
-\protect\hypertarget{our-simulation}{}
-Suppose all similarities for two classes are randomly drawn from the
-same distribution, \(\mathsf{Normal}(\mu = 0, \sigma = .05)\), you still
-can get a really high WEAT!
+\begin{block}{Some similarities from \emph{philosophy}}
+\protect\hypertarget{some-similarities-from}{}
+Logic (.47), Nietzsche (.32), Hegel (.32), analytic (.13), burger (.08),
+continental (.04), Russell (.04)
 \end{block}
@@
 
-\begin{frame}{The problem with pre-averaging}
-\protect\hypertarget{the-problem-with-pre-averaging-1}{}
-\begin{block}{One simulation}
-\protect\hypertarget{one-simulation}{}
-\vspace{1mm}
-\footnotesize
-
-\begin{center}\includegraphics[width=0.7\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-6-1} \end{center}
-\normalsize
-
-\vspace{1mm}
-\footnotesize
-
-\normalsize
-\pause
-
-\footnotesize
-
-\vspace{-2mm}
-
-\begin{itemize}
-\tightlist
-\item
-  Raw sd in data is 0.045
-\item
-  The sd of means is 0.023
-\item
-  The WEAT score is 1.825
-\item
-  The largest effect size reported by
-  \protect\hyperlink{ref-Caliskan2017semanticsBiases}{Caliskan, Bryson,
-  \& Narayanan}
-  (\protect\hyperlink{ref-Caliskan2017semanticsBiases}{2017}) is 1.81!
-\end{itemize}
-\end{block}
-\end{frame}
-
-\begin{frame}{The problem with pre-averaging}
-\protect\hypertarget{the-problem-with-pre-averaging-2}{}
-\begin{block}{50k simulations (same parameters)}
-\protect\hypertarget{k-simulations-same-parameters}{}
-\vspace{1mm}
-\footnotesize
-
-\begin{center}\includegraphics[width=0.8\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-8-1} \end{center}
-\normalsize
-\end{block}
-\end{frame}
-
-\begin{frame}{Advantages of the Bayesian way}
-\protect\hypertarget{advantages-of-the-bayesian-way}{}
-\begin{itemize}
-\tightlist
-\item
-  Direct impact of sample sizes
-\item
-  Straightforward interpretation in terms of posterior probabilities
-\item
-  Freedom to choose granularity level
-\item
-  More honest risk assessment and decision making
-\end{itemize}
-\end{frame}
-
-\begin{frame}{Bayesian model}
-\protect\hypertarget{bayesian-model}{}
-\begin{block}{Choosing priors}
-\protect\hypertarget{choosing-priors}{}
-\begin{center}\includegraphics[width=1\linewidth]{presentationBoston_files/figure-beamer/priorsVis-1} \end{center}
-\end{block}
-\end{frame}
-
-\begin{frame}[fragile]{Bayesian model architecture}
-\protect\hypertarget{bayesian-model-architecture}{}
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{library}\NormalTok{(rethinking)}
-\FunctionTok{options}\NormalTok{(}\AttributeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\ConstantTok{TRUE}\NormalTok{ )}
-\NormalTok{religionCoefs }\OtherTok{\textless{}{-}} \FunctionTok{ulam}\NormalTok{(}
-  \FunctionTok{alist}\NormalTok{(}
-\NormalTok{    cosineDistance }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(mu,sigma),}
-\NormalTok{    mu }\OtherTok{\textless{}{-}}\NormalTok{ m }\SpecialCharTok{+}\NormalTok{ co[con],}
-\NormalTok{    m }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
-\NormalTok{    co[con] }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
-\NormalTok{    sigma }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
-\NormalTok{  ),}
-  \AttributeTok{data =}\NormalTok{ religion,}
-  \AttributeTok{chains=}\DecValTok{2}\NormalTok{ , }\AttributeTok{iter=}\DecValTok{8000}\NormalTok{ , }\AttributeTok{warmup=}\DecValTok{1000}\NormalTok{, }
-  \AttributeTok{log\_lik =} \ConstantTok{TRUE}
-\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\normalsize
-\end{frame}
-
-\begin{frame}{Dataset-level coefficients}
-\protect\hypertarget{dataset-level-coefficients}{}
-\begin{block}{Religion with 89\%-compatibility intervals (HPDI)}
-\protect\hypertarget{religion-with-89-compatibility-intervals-hpdi}{}
-\begin{center}
-\begin{figure}[!htb]\centering
-   \begin{minipage}{0.55\textwidth}
-  \includegraphics[width=6cm]{../../images/religionCoeffs.jpeg}
-   \end{minipage} \begin{minipage}{0.4\textwidth}\footnotesize
-   \begin{itemize}
-   \item All HPDIs overlap with 0 
-   \item Differences between classes are relatively small
-   \item Coefficients for Race are similar
-   \end{itemize}
-   \end{minipage}
-  \end{figure}
-
-\end{center}
-\end{block}
-\end{frame}
-
-\begin{frame}{Dataset-level coefficients}
-\protect\hypertarget{dataset-level-coefficients-1}{}
-\begin{block}{Gender with 89\%-compatibility intervals (HPDI)}
-\protect\hypertarget{gender-with-89-compatibility-intervals-hpdi}{}
-\begin{center}
-\begin{figure}[!htb]\centering
-  \begin{minipage}{0.55\textwidth}
-\includegraphics[width=6cm]{../../images/genderCoeffs.jpeg}
-\end{minipage}
-\begin{minipage}{0.4\textwidth}\footnotesize
-   \begin{itemize}
-   \item Associated and different are away from 0
-   \item But they were supposed to be opposites and are very close to each other (co-occurrence?)
-   \item Differences between classes are still relatively small
-   \end{itemize}
-   \end{minipage}
-\end{figure}
-
-\end{center}
-\end{block}
-\end{frame}
-
-\begin{frame}[fragile]{Bayesian model architecture}
-\protect\hypertarget{bayesian-model-architecture-1}{}
-\vspace{1mm}
-\footnotesize
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{library}\NormalTok{(rethinking)}
-\FunctionTok{options}\NormalTok{(}\AttributeTok{buildtools.check =} \ControlFlowTok{function}\NormalTok{(action) }\ConstantTok{TRUE}\NormalTok{ )}
-\NormalTok{religionCoefs }\OtherTok{\textless{}{-}} \FunctionTok{ulam}\NormalTok{(}
-  \FunctionTok{alist}\NormalTok{(}
-\NormalTok{    cosineDistance }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(mu,sigma),}
-\NormalTok{    mu }\OtherTok{\textless{}{-}}\NormalTok{ m[pw] }\SpecialCharTok{+}\NormalTok{ co[con],}
-\NormalTok{    m[pw] }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
-\NormalTok{    co[con] }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{0}\NormalTok{,.}\DecValTok{5}\NormalTok{),}
-\NormalTok{    sigma }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dcauchy}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{1}\NormalTok{)}
-\NormalTok{  ),}
-  \AttributeTok{data =}\NormalTok{ religion,}
-  \AttributeTok{chains=}\DecValTok{2}\NormalTok{ , }\AttributeTok{iter=}\DecValTok{8000}\NormalTok{ , }\AttributeTok{warmup=}\DecValTok{1000}\NormalTok{, }
-  \AttributeTok{log\_lik =} \ConstantTok{TRUE}
-\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\normalsize
-\end{frame}
-
-\begin{frame}{Word-level coefficients}
-\protect\hypertarget{word-level-coefficients}{}
-\begin{figure}[!htb]\centering
-  \begin{minipage}{0.55\textwidth}
-\includegraphics[width=7.5cm]{../../images/visReligionReddit.png}
-\end{minipage}
-\begin{minipage}{0.4\textwidth}\footnotesize
-
-\vspace{-4cm}
-
-   \begin{itemize}
-   \item Most intervals overlap with control groups
-   \item Often not too much difference between associated and different
-   \end{itemize}
-   \end{minipage}
-\end{figure}
-\end{frame}
-
-\begin{frame}{Word-level coefficients}
-\protect\hypertarget{word-level-coefficients-1}{}
-\begin{figure}[!htb]\centering
-  \begin{minipage}{0.55\textwidth}
-\includegraphics[width=7.5cm]{../../images/visGenderReddit.png}
-\end{minipage}
-\begin{minipage}{0.4\textwidth}\footnotesize
-
-\vspace{-4cm}
-
-   \begin{itemize}
-   \item Male attributes: strong co-occurrence with female attributes
-   \item Sometimes \textsf{different} is closer than \textsf{associated}
-   \item Almost no overlap with control groups
-   \end{itemize}
-   \end{minipage}
-\end{figure}
-\end{frame}
-
-\begin{frame}{Word-level coefficients}
-\protect\hypertarget{word-level-coefficients-2}{}
-\begin{figure}[!htb]\centering
-  \begin{minipage}{0.55\textwidth}
-\includegraphics[width=7.5cm]{../../images/visRaceReddit.png}
-\end{minipage}
-\begin{minipage}{0.4\textwidth}\footnotesize
-
-\vspace{-4cm}
-
-   \begin{itemize}
-   \item A lot of variation between races
-
-   \item Often not much difference between associated and different
-   \end{itemize}
-   \end{minipage}
-\end{figure}
-\end{frame}
-
-\begin{frame}{Thank you!}
-\protect\hypertarget{thank-you}{}
-\begin{block}{Summary}
-\protect\hypertarget{summary}{}
-\begin{itemize}
-\tightlist
-\item
-  Bias in word embeddings
-\item
-  WEAT and MAC methods
-\item
-  Methodological problems
-\item
-  Limitations of pre-averaging in bias detection methods
-\item
-  Accounting for uncertainty with Bayesian approach
-\end{itemize}
-\end{block}
-
-\begin{block}{Further work}
-\protect\hypertarget{further-work}{}
-\begin{itemize}
-\tightlist
-\item
-  Including contrasts in Bayesian calculation
-\item
-  Performance cross-validation in comparison to other methods (regular
-  linear regression, KNN, \dots)
-\item
-  Downstream tasks and connection with intrinsic evaluation
-\item
-  Testing data from the original Implicit Association Test (IAT)
-\item
-  Applying uncertainty to WEAT and better word lists
-\item
-  Looking at other similarity measures
-\end{itemize}
-\end{block}
-\end{frame}
-
-\begin{frame}{References}
-\protect\hypertarget{references}{}
-\tiny
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\hypertarget{ref-Caliskan2017semanticsBiases}{}%
-Caliskan, A., Bryson, J. J., \& Narayanan, A. (2017). Semantics derived
-automatically from language corpora contain human-like biases.
-\emph{Science}, \emph{356}(6334), 183--186.
-\url{https://doi.org/10.1126/science.aal4230}
-
-\leavevmode\hypertarget{ref-Garg2018years}{}%
-Garg, N., Schiebinger, L., Jurafsky, D., \& Zou, J. (2018). Word
-embeddings quantify 100 years of gender and ethnic stereotypes.
-\emph{Proceedings of the National Academy of Sciences}, \emph{115}(16),
-E3635--E3644. \url{https://doi.org/10.1073/pnas.1720347115}
-
-\leavevmode\hypertarget{ref-Gonen2019lipstick}{}%
-Gonen, H., \& Goldberg, Y. (2019). Lipstick on a pig: {D}ebiasing
-methods cover up systematic gender biases in word embeddings but do not
-remove them. \emph{Proceedings of the 2019 Conference of the North
-{A}merican Chapter of the Association for Computational Linguistics:
-Human Language Technologies, Volume 1 (Long and Short Papers)},
-609--614. Minneapolis, Minnesota: Association for Computational
-Linguistics. \url{https://doi.org/10.18653/v1/N19-1061}
-
-\leavevmode\hypertarget{ref-Lauscher2019multidimensional}{}%
-Lauscher, A., \& Glavas, G. (2019). Are we consistently biased?
-Multidimensional analysis of biases in distributional word vectors.
-\emph{CoRR}, \emph{abs/1904.11783}. Retrieved from
-\url{http://arxiv.org/abs/1904.11783}
-
-\leavevmode\hypertarget{ref-Manzini2019blackToCriminal}{}%
-Manzini, T., Lim, Y. C., Tsvetkov, Y., \& Black, A. W. (2019).
-\emph{Black is to criminal as caucasian is to police: Detecting and
-removing multiclass bias in word embeddings}. Retrieved from
-\url{http://arxiv.org/abs/1904.04047}
-
-\leavevmode\hypertarget{ref-Nissim2020fair}{}%
-Nissim, M., Noord, R. van, \& Goot, R. van der. (2020). Fair is better
-than sensational: Man is to doctor as woman is to doctor.
-\emph{Computational Linguistics}, \emph{46}(2), 487--497.
-\url{https://doi.org/10.1162/coli_a_00379}
-
-\leavevmode\hypertarget{ref-zhang2020robustness}{}%
-Zhang, H., Sneyd, A., \& Stevenson, M. (2020). \emph{Robustness and
-reliability of gender bias assessment in word embeddings: The role of
-base pairs}. Retrieved from \url{http://arxiv.org/abs/2010.02847}
-
-\end{CSLReferences}
-\end{frame}
-
 \end{document}

### 2022-03-08 presentations/presentationBoston/cosineGraphs.R

@@ -1,0 +1,114 @@
+library(ggplot2)
+library(ggforce)
+library(ggthemes)
+
+
+#say you start with two 2-dimensional vectors, with origin at (0,0)
+
+a <- c(1,2)
+b <- c(3,4)
+
+#The dot product of vectors of the same length is a1b1 + a2b2 + ...
+#it is a number
+
+ab <- a %*% b
+
+#in particular for length 2, $a \dot a$ is going to be the same as $a_1^2 + a_2^2$
+a %*% a
+
+1^2 + 2^2
+
+#now think about a:
+
+ggplot()+ geom_segment(aes(x = 0, y = 0, xend = a[1], yend = a[2]))+
+  xlim(c(0,6))+ylim(c(0,6))+
+  annotate(geom = "text", label  = "a = [1,2]", x = 1.1, y = 2.1)
+
+
+#It's length is $\sqrt{a_1^2 + a_2^2}= \sqrt{a \dot a} = \sqrt{5}$
+# \vert \vert v \vert \vert & = \sqrt(v \times v)
+
+#normalized v? Scalar multiple by $\nicefrac{1}{\vert \vert v\vert \vert}
+
+lengthA <- sqrt(a %*% a)
+
+aNormalized <- a/lengthA
+aNormalized
+sqrt(aNormalized %*% aNormalized)
+
+ggplot()+
+  xlim(c(0,6))+ylim(c(0,6))+geom_segment(aes(x = 0, y = 0, xend = a[1], yend = a[2]))+
+  annotate(geom = "text", label  = "a = [1,2]", x = 1.1, y = 2.1)+
+  geom_segment(aes(x = 0, y = 0, xend = aNormalized[1], yend = aNormalized[2]), color = "grey")+ 
+  annotate(geom = "text", label  = "a normalized = [.447,.894]", x = 1.3, y = .994)
+
+
+#distance between vectors is the length of the difference
+#$d(u,v) = \vert \vert u - v \vert \vert$
+ 
+ggplot()+
+  xlim(c(0,6))+ylim(c(0,6))+
+  geom_segment(aes(x = 0, y = 0, xend = a[1], yend = a[2]),  arrow = arrow(length = unit(0.02, "npc")))+
+  annotate(geom = "text", label  = "a = [1,2]", x = 1.1, y = 2.1)+
[…]

### 2022-03-11 presentations/presentationBoston/presentationBoston.Rmd
Adds new table

@@
 
-Religion Debiasing  | MAC (distance)
-------------- | -------------
-Biased        | 0.859
-Hard Debiased | 0.934
-Soft Debiased ($\lambda$ = 0.2) | 0.894
+Category | Biased | Hard Debiased | Soft Debiased | Diff |
+-------| ------| -------| ----- | --------
+Religion | 0.859 | 0.934 | 0.894| 0.075 |
+Race | 0.892 | 0.925| 0.985 | 0.033 |
+Gender | 0.623 | 0.700 | 0.747 | 0.077 |
 

### 2022-03-11 presentations/presentationBoston/presentationBoston.Rmd
Adds lists of human and neutral words

@@
 
+### Our lists of neutral and human words
+
+- neutral words = [liquor, pow, ballpark, glitchy, billy, dallas, rip, called, outlooks, viet, floater, rattlesnake, exports,
+peruvian, recursion, shortfall, corrected, amicable, solutions, diagnostic, patently, flops, approx, percents, lox, catapults,
+hamburger, engulfed, households, north, snubbed, playtest] \newline
+
+- human words = [switch, studio, stick, soup, sometimes, signal, prior, plant, photo, path, park, near, menu, latter, grass, clock, 
+wear, walk, visitor, toy, tissue, throw, talk, speak, sleep, eye, enjoy, blogger, character, candidate, breakfast, supper, dinner,
+eat, drink, carry, run, cast, ask, awake, ear, nose, lunch]
+
+## Some methodological problems
+
 ### Outliers and surprisingly dissimilar words

### 2022-03-11 presentations/presentationBoston/presentationBoston.tex
Adds new table
Adds lists of human and neutral words

@@
 \protect\hypertarget{some-methodological-problems-2}{}
+\begin{block}{Our lists of neutral and human words}
+\protect\hypertarget{our-lists-of-neutral-and-human-words}{}
+\begin{itemize}
+\item
+  neutral words = {[}liquor, pow, ballpark, glitchy, billy, dallas, rip,
+  called, outlooks, viet, floater, rattlesnake, exports, peruvian,
+  recursion, shortfall, corrected, amicable, solutions, diagnostic,
+  patently, flops, approx, percents, lox, catapults, hamburger,
+  engulfed, households, north, snubbed, playtest{]} \newline
+\item
+  human words = {[}switch, studio, stick, soup, sometimes, signal,
+  prior, plant, photo, path, park, near, menu, latter, grass, clock,
+  wear, walk, visitor, toy, tissue, throw, talk, speak, sleep, eye,
+  enjoy, blogger, character, candidate, breakfast, supper, dinner, eat,
+  drink, carry, run, cast, ask, awake, ear, nose, lunch{]}
+\end{itemize}
+\end{block}
+\end{frame}
+
+\begin{frame}{Some methodological problems}
+\protect\hypertarget{some-methodological-problems-3}{}
 \begin{block}{Outliers and surprisingly dissimilar words}
@@
 \begin{frame}{Some methodological problems}
-\protect\hypertarget{some-methodological-problems-3}{}
+\protect\hypertarget{some-methodological-problems-4}{}
 \begin{block}{Distances for ``muslim''}
@@
 \begin{frame}{Some methodological problems}
-\protect\hypertarget{some-methodological-problems-4}{}
+\protect\hypertarget{some-methodological-problems-5}{}
 \begin{block}{Distances for ``priest''}
@@
 \begin{frame}{Some methodological problems}
-\protect\hypertarget{some-methodological-problems-5}{}
+\protect\hypertarget{some-methodological-problems-6}{}
 \begin{block}{No principled interpretation}
 \protect\hypertarget{no-principled-interpretation}{}
-\begin{longtable}[]{@{}ll@{}}
+\begin{longtable}[]{@{}lllll@{}}
 \toprule
-Religion Debiasing & MAC (distance) \\
+Category & Biased & Hard Debiased & Soft Debiased & Diff \\
 \midrule
 \endhead
-Biased & 0.859 \\
-Hard Debiased & 0.934 \\
-Soft Debiased (\(\lambda\) = 0.2) & 0.894 \\
+Religion & 0.859 & 0.934 & 0.894 & 0.075 \\
+Race & 0.892 & 0.925 & 0.985 & 0.033 \\
+Gender & 0.623 & 0.700 & 0.747 & 0.077 \\
 \bottomrule
@@
 \item
-  Raw sd in data is 0.045
-\item
-  The sd of means is 0.023
+  Raw sd in data is 0.072
+\item
+  The sd of means is 0.037
 \item

### 2022-03-11 functions/cosineDistribution.R

@@ -1,0 +1,1 @@
+#  cosine similarity histogram for all groups (what percentage is greater than -+0.1) 

### 2022-03-14 functions/cosineDistribution.R
Adds cosine distribution tests

@@ -1,1 +1,115 @@
-#  cosine similarity histogram for all groups (what percentage is greater than -+0.1) 
+#  cosine similarity histogram for all groups
+
+library(ggplot2)
+library(tidyverse)
+library(gridExtra)
+library(kableExtra)
+library(magrittr)
+library(ggplot2)
+library(ggpubr)
+library(ggExtra)
+library(ggthemes)
+library(ggforce)
+library(latex2exp)
+
+religionReddit <- read.csv("./datasets/religionReddit.csv")[-1]
+raceReddit <- read.csv("./datasets/raceReddit.csv")[-1]
+genderReddit <- read.csv("./datasets/genderReddit.csv")[-1]
+
+
+# --------------------------- RELIGION DATASET ------------------------------------------------
+
+cosAssReligion <- religionReddit['cosine_similarity'][which(religionReddit$connection == 'associated'),]
+cosDiffReligion <- religionReddit['cosine_similarity'][which(religionReddit$connection == 'different'),]
+cosNoneReligion <- religionReddit['cosine_similarity'][which(religionReddit$connection == 'none'),]
+
+# ------------ Associated
+par(mfrow=c(1,3))
+ggplot()+geom_histogram(aes(x=cosAssReligion, y = ..density..), alpha = 0.8, bins=20, color="lightgreen", fill="lightgreen")+
+  theme_tufte()+labs(title="Cosine similarity distribution")+ xlab("bias") + geom_vline(xintercept = .1) + geom_vline(xintercept = -.1) 
+
+(sum(abs(cosAssReligion)>0.1)) / length(cosAssReligion)
+
+# ------------ Different
+
+ggplot()+geom_histogram(aes(x=cosDiffReligion, y = ..density..), alpha = 0.8, bins=20, color="pink", fill="pink")+
+  theme_tufte()+labs(title="Cosine similarity distribution")+ xlab("bias") + geom_vline(xintercept = .1) + geom_vline(xintercept = -.1) 
+
+(sum(abs(cosDiffReligion)>0.1)) / length(cosDiffReligion)
+
+# ------------- Neutral and Human
+
+ggplot()+geom_histogram(aes(x=cosNoneReligion, y = ..density..), alpha = 0.8, bins=20, color="lightblue", fill="lightblue")+
[…]

### 2022-03-15 presentations/presentationBoston/presentationBoston.Rmd
small corrections, added examples

--- 2022-03-11 presentations/presentationBoston/presentationBoston.Rmd
@@
 author: Alicja Dobrzeniecka \& Rafal  Urbaniak \footnotesize \newline (LoPSE research
-  group, University of  Gdansk)
+  group, University of  Gdansk, Vrij Universiteit Amsterdam)
 date: "Boston, April Fools' Day"
@@
 library(latex2exp)
+library(forcats)
 th <- theme_tufte(base_size = 20)
@@
 a  & = [1,2]\\
-b  &= [3,2]
+b  &= [3,2.5]
 \end{align*}
@@
 a \cdot a & = a_1^2 + a_2 ^ 2 \\
-\lVert a\rVert & = \sqrt(a \cdot a)
+\lVert a\rVert & = \sqrt{(a \cdot a)}
 \end{align*}
@@
 a  & = [1,2]\\
-b  &= [4,4]
+b  &= [3,2.5]
 \end{align*}
@@
 
--  Geometric interpretation: direction (not length)
+- Naive interpretation: proximity corresponds to semantic similarity
+
+-  Geometric interpretation: direction $\mathsf{cos} \in (-1,1)$
+
+    - $1$: maximally smilar 
+    - $-1$: opposites 
+    - $0$: dissimilar
 
@@
 
-- Naive interpretation: proximity corresponds to semantic similarity
-
 <!-- - Note cosine distance fails to meet triangle inequality -->
+
+## Cosine similarity \& distance
+
+```{r philosophy,echo=FALSE,eval=TRUE,fig.align = "center", cache=TRUE, fig.show = "hold", out.width = "80%",   message = FALSE, warning = FALSE, results = FALSE}
+words <- c("logic", "Nietzsche", "analytic", "Hegel", "Russell", "continental", "burger", "sandwich")
+distances <- c(.47, .32, .13, .29, .08, .08, .04, .04)
+philosophy <- data.frame(words, distances)
+ggplot(philosophy, aes(x = fct_reorder(words, distances), y = distances))+
+  geom_point(size = 3)+coord_flip()+theme_tufte(base_size = 20)+ggtitle("Similarity to <philosophy>, examples")+ylab("")+xlab("similarity")
+```
+
+\pause
+
+### The only "jobs" in top-tens
+
+- Man: robber (.55)
+
+- Woman: policewoman (.6)
 
@@
 
-- ETHAYARAJH IS YOUR criticizes WEAT uses Bernstein bounds and argues that we would need a bias specific dataset of size at least 11903 to claim that the system is biased (three times larger than WinoBias).
+- @Ethayarajh2020measuring criticizes WEAT uses Bernstein bounds and argues that we would need a bias specific dataset of size at least 11903 to claim that the system is biased (three times larger than WinoBias).
 
@@
 
-ETHAYARAJH UNDERSTANDING shows if there are two target words only WEAT is always maximal in one direction.
+- @Ethayarajh2019understanding show that if there are two target words only WEAT is always maximal in one direction.
 
@@
 
+### Our neutral words (examples, full list size = ADD)
+
+liquor, pow, ballpark, glitchy, billy, dallas, rip, called, outlooks, viet, floater, rattlesnake, exports,
+peruvian, recursion, shortfall, corrected, amicable, solutions, diagnostic, patently, flops, approx, percents, lox, catapults,
+hamburger, engulfed, households, north, snubbed, playtest
+
+### Our human-related words (examples, full list size = ADD)
+
+switch, studio, stick, soup, sometimes, signal, prior, plant, photo, path, park, near, menu, latter, grass, clock,  wear, walk, visitor, toy, tissue, throw, talk, speak, sleep, eye, enjoy, blogger, character, candidate, breakfast, supper, dinner,
+eat, drink, carry, run, cast, ask, awake, ear, nose, lunch
+
+## Some methodological problems
+
 ### Outliers and surprisingly dissimilar words
@@
 
-t1Plot <- ggplot(t1)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t1 in X")+xlim(xlimits)
-
-t2Plot <- ggplot(t2)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t2 in X")+xlim(xlimits)
-
-t3Plot <- ggplot(t3)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t3 in Y")+xlim(xlimits)
-
-t4Plot <- ggplot(t4)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t4 in Y")+xlim(xlimits)
+t1Plot <- ggplot(t1)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "orangered", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "orangered") +ggtitle("Similarities for term t1 in X")+xlim(xlimits)
+
+t2Plot <- ggplot(t2)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "orangered", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "orangered") +ggtitle("Similarities for term t2 in X")+xlim(xlimits)
+
+t3Plot <- ggplot(t3)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "orangered", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "orangered") +ggtitle("Similarities for term t3 in Y")+xlim(xlimits)
+
+t4Plot <- ggplot(t4)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "orangered", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "orangered") +ggtitle("Similarities for term t4 in Y")+xlim(xlimits)
 
@@
 
-t1Plot <- ggplot(t1)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t1 in X")+xlim(xlimits)
-
-t2Plot <- ggplot(t2)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t2 in X")+xlim(xlimits)
-
-t8Plot <- ggplot(t8)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t3 in Y")+xlim(xlimits)
-
-t9Plot <- ggplot(t9)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t4 in Y")+xlim(xlimits)
+t1Plot <- ggplot(t1)+geom_jitter(aes(y = 0, x = A), width  = 0, height = .02,  size = 1, alpha = 0.5)+geom_jitter(aes(y = 0.1, x = B), col = "orangered", width  = 0, height = .02,  size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "orangered") +ggtitle("Similarities for term t1 in X")+xlim(xlimits)
+
+t2Plot <- ggplot(t2)+geom_jitter(aes(y = 0, x = A), width  = 0, height = .02,  size = 1, alpha = 0.5)+geom_jitter(aes(y = 0.1, x = B),  width  = 0, height = .02, col = "orangered", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "orangered") +ggtitle("Similarities for term t2 in X")+xlim(xlimits)
+
+t8Plot <- ggplot(t8)+geom_jitter(aes(y = 0, x = A), width  = 0, height = .02,  size = 1, alpha = 0.5)+geom_jitter(aes(y = 0.1, x = B), width  = 0, height = .02,  col = "orangered", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "orangered") +ggtitle("Similarities for term t3 in Y")+xlim(xlimits)
+
+t9Plot <- ggplot(t9)+geom_jitter(aes(y = 0, x = A), width  = 0, height = .02,  size = 1, alpha = 0.5)+geom_jitter(aes(y = 0.1, x = B), width  = 0, height = .02,  col = "orangered", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "orangered") +ggtitle("Similarities for term t4 in Y")+xlim(xlimits)
 

### 2022-03-15 presentations/presentationBoston/presentationBoston.tex
small corrections, added examples

--- 2022-03-08 presentations/presentationBoston/presentationBoston.tex
@@
   pdftitle={Taking uncertainty seriously A Bayesian approach to word embedding bias estimation},
-  pdfauthor={Alicja Dobrzeniecka \& Rafal Urbaniak (LoPSE research group, University of Gdansk)},
+  pdfauthor={Alicja Dobrzeniecka \& Rafal Urbaniak (LoPSE research group, University of Gdansk, Vrij Universiteit Amsterdam)},
   colorlinks=true,
@@
 \author{Alicja Dobrzeniecka \& Rafal Urbaniak
-\footnotesize \newline (LoPSE research group, University of Gdansk)}
+\footnotesize \newline (LoPSE research group, University of Gdansk, Vrij
+Universiteit Amsterdam)}
 \date{Boston, April Fools' Day}
@@
 \item
-  Geometric interpretation: direction (not length)
+  Naive interpretation: proximity corresponds to semantic similarity
+\item
+  Geometric interpretation: direction \(\mathsf{cos} \in (-1,1)\)
+
+  \begin{itemize}
+  \tightlist
+  \item
+    \(1\): maximally smilar
+  \item
+    \(-1\): opposites
+  \item
+    \(0\): dissimilar
+  \end{itemize}
 \item
   \(\mathsf{cosineDistance}\in (0, 2)\)
-\item
-  Naive interpretation: proximity corresponds to semantic similarity
-\end{itemize}
+\end{itemize}
+\end{frame}
+
+\begin{frame}{Cosine similarity \& distance}
+\protect\hypertarget{cosine-similarity-distance-1}{}
+\begin{center}\includegraphics[width=0.8\linewidth]{presentationBoston_files/figure-beamer/philosophy-1} \end{center}
+
+\pause
+
+\begin{block}{The only ``jobs'' in top-tens}
+\protect\hypertarget{the-only-jobs-in-top-tens}{}
+\begin{itemize}
+\item
+  Man: robber (.55)
+\item
+  Woman: policewoman (.6)
+\end{itemize}
+\end{block}
 \end{frame}
@@
 \item
-  ETHAYARAJH IS YOUR criticizes WEAT uses Bernstein bounds and argues
-  that we would need a bias specific dataset of size at least 11903 to
-  claim that the system is biased (three times larger than WinoBias).
+  \protect\hyperlink{ref-Ethayarajh2020measuring}{Ethayarajh}
+  (\protect\hyperlink{ref-Ethayarajh2020measuring}{2020}) criticizes
+  WEAT uses Bernstein bounds and argues that we would need a bias
+  specific dataset of size at least 11903 to claim that the system is
+  biased (three times larger than WinoBias).
 \end{itemize}
@@
 \protect\hypertarget{the-form-of-the-definition-is-suspicious}{}
-ETHAYARAJH UNDERSTANDING shows if there are two target words only WEAT
-is always maximal in one direction.
+\begin{itemize}
+\tightlist
+\item
+  \protect\hyperlink{ref-Ethayarajh2019understanding}{Ethayarajh,
+  Duvenaud, \& Hirst}
+  (\protect\hyperlink{ref-Ethayarajh2019understanding}{2019}) show that
+  if there are two target words only WEAT is always maximal in one
+  direction.
+\end{itemize}
 
@@
 \protect\hypertarget{some-methodological-problems-2}{}
+\begin{block}{Our neutral words (examples, full list size = ADD)}
+\protect\hypertarget{our-neutral-words-examples-full-list-size-add}{}
+liquor, pow, ballpark, glitchy, billy, dallas, rip, called, outlooks,
+viet, floater, rattlesnake, exports, peruvian, recursion, shortfall,
+corrected, amicable, solutions, diagnostic, patently, flops, approx,
+percents, lox, catapults, hamburger, engulfed, households, north,
+snubbed, playtest
+\end{block}
+
+\begin{block}{Our human-related words (examples, full list size = ADD)}
+\protect\hypertarget{our-human-related-words-examples-full-list-size-add}{}
+switch, studio, stick, soup, sometimes, signal, prior, plant, photo,
+path, park, near, menu, latter, grass, clock, wear, walk, visitor, toy,
+tissue, throw, talk, speak, sleep, eye, enjoy, blogger, character,
+candidate, breakfast, supper, dinner, eat, drink, carry, run, cast, ask,
+awake, ear, nose, lunch
+\end{block}
+\end{frame}
+
+\begin{frame}{Some methodological problems}
+\protect\hypertarget{some-methodological-problems-3}{}
 \begin{block}{Outliers and surprisingly dissimilar words}
@@
 
+\leavevmode\hypertarget{ref-Ethayarajh2020measuring}{}%
+Ethayarajh, K. (2020). \emph{Is your classifier actually biased?
+Measuring fairness under uncertainty with bernstein bounds}.
+
+\leavevmode\hypertarget{ref-Ethayarajh2019understanding}{}%
+Ethayarajh, K., Duvenaud, D., \& Hirst, G. (2019). Understanding
+undesirable word embedding associations. \emph{Proceedings of the 57th
+Annual Meeting of the Association for Computational Linguistics},
+1696--1705.
+
 \leavevmode\hypertarget{ref-Garg2018years}{}%

### 2022-03-15 functions/visualisationTools.R

@@ -35,8 +35,8 @@ furthestDistance <- bunch$cosineDistance[nrow(bunch)]
 
 plot <- ggplot(bunch) + geom_point(aes(x=0, y=0), colour="black", size =10,alpha=0.02)+
           geom_point(aes(x=xaxis,y=yaxis, color = connection, size = cosineSimilarity), alpha = 0.3)+ 
-  geom_segment(aes(x= 0, y = 0, xend = closestx,yend=closesty),alpha= 0.01, size = 0.15)+
-  geom_segment(aes(x= 0, y = 0, xend = furthestx,yend=furthesty),alpha=0.01,size = 0.15)+
+  geom_segment(aes(x= 0, y = 0, xend = closestx,yend=closesty),alpha= 0.1, size = 0.2)+
+  geom_segment(aes(x= 0, y = 0, xend = furthestx,yend=furthesty),alpha=0.1,size = 0.2)+
   annotate("text", x = 0.5*closestx, y = 0.5*closesty-0.05, label = round(closestDistance,3), size = 3)+
   annotate("text", x = 1.5*furthestx, y = 0.5*furthesty, label = round(furthestDistance,3), size = 3)+
   annotate("text", x = -0.02, y = -0.13, label = protected, size = 4)+

### 2022-03-15 presentations/presentationBoston/philosophyWords.R

@@ -1,0 +1,17 @@
+words <- c("logic", "Nietzsche", "analytic", "Hegel", "Russell", "continental", "burger", "sandwich" 
+           )
+
+distances <- c(.47, .32, .13, .29, .08, .08, .04, .04)
+
+philosophy <- data.frame(words, distances)
+
+
+library(ggplot2)
+library(ggthemes)
+library(forcats)
+
+
+fct_reorder(words, distances)
+ggplot(philosophy, aes(x = fct_reorder(words, distances), y = distances))+
+  geom_point(size = 3)+coord_flip()+theme_tufte(base_size = 7)
+

### 2022-03-16 presentations/presentationBoston/presentationBoston.Rmd
Small typo

@@
 author: Alicja Dobrzeniecka \& Rafal  Urbaniak \footnotesize \newline (LoPSE research
-  group, University of  Gdansk, Vrij Universiteit Amsterdam)
+  group, University of  Gdansk, Vrije Universiteit Amsterdam)
 date: "Boston, April Fools' Day"

### 2022-03-16 presentations/presentationBoston/presentationBoston.tex
Adds notes

@@
   pdftitle={Taking uncertainty seriously A Bayesian approach to word embedding bias estimation},
-  pdfauthor={Alicja Dobrzeniecka \& Rafal Urbaniak (LoPSE research group, University of Gdansk, Vrij Universiteit Amsterdam)},
+  pdfauthor={Alicja Dobrzeniecka \& Rafal Urbaniak (LoPSE research group, University of Gdansk, Vrije Universiteit Amsterdam)},
   colorlinks=true,
@@
 \author{Alicja Dobrzeniecka \& Rafal Urbaniak
-\footnotesize \newline (LoPSE research group, University of Gdansk, Vrij
-Universiteit Amsterdam)}
+\footnotesize \newline (LoPSE research group, University of Gdansk,
+Vrije Universiteit Amsterdam)}
 \date{Boston, April Fools' Day}

### 2022-03-20 presentations/presentationBoston/presentationBoston.Rmd
Adds word lists sizes

@@
 
-### Our neutral words (examples, full list size = ADD)
+### Our neutral words (examples, full list size = 242)
 
@@
 
-### Our human-related words (examples, full list size = ADD)
-
-switch, studio, stick, soup, sometimes, signal, prior, plant, photo, path, park, near, menu, latter, grass, clock,  wear, walk, visitor, toy, tissue, throw, talk, speak, sleep, eye, enjoy, blogger, character, candidate, breakfast, supper, dinner,
-eat, drink, carry, run, cast, ask, awake, ear, nose, lunch
+### Our human-related words (examples, full list size = 27)
+
+wear, walk, visitor, toy, tissue, throw, talk, speak, sleep, eye, enjoy, blogger, character, candidate, breakfast, supper, dinner, eat,
+drink, carry, run, cast, ask, awake, ear, nose, lunch
 

### 2022-03-20 functions/cosineDistribution.R

@@ -13,9 +13,9 @@ library(ggforce)
 library(latex2exp)
 
 getwd()
-religionReddit <- read.csv("../datasets/religionReddit.csv")[-1]
-raceReddit <- read.csv("../datasets/raceReddit.csv")[-1]
-genderReddit <- read.csv("../datasets/genderReddit.csv")[-1]
+religionReddit <- read.csv("datasets/religionReddit.csv")[-1]
+raceReddit <- read.csv("datasets/raceReddit.csv")[-1]
+genderReddit <- read.csv("datasets/genderReddit.csv")[-1]
 
 
 # --------------------------- RELIGION DATASET ------------------------------------------------

### 2022-03-20 presentations/presentationBoston/presentationBoston.Rmd
Fixes table

@@
 
-Category | Biased | Hard Debiased | Soft Debiased | Diff |
+Category | Biased | Hard Debiased | Diff 
 -------| ------| -------| ----- | --------
-Religion | 0.859 | 0.934 | 0.894| 0.075 |
-Race | 0.892 | 0.925| 0.985 | 0.033 |
-Gender | 0.623 | 0.700 | 0.747 | 0.077 |
+Religion | 0.859 | 0.934 | 0,141 
+Race | 0.892 | 0.925| 0,108 
+Gender | 0.623 | 0.700 | 0,377
 

### 2022-03-20 presentations/presentationBoston/presentationBoston.tex
Fixes table

@@
 \protect\hypertarget{some-methodological-problems-2}{}
-\begin{block}{Our neutral words (examples, full list size = ADD)}
-\protect\hypertarget{our-neutral-words-examples-full-list-size-add}{}
+\begin{block}{Our neutral words (examples, full list size = 242)}
+\protect\hypertarget{our-neutral-words-examples-full-list-size-242}{}
 liquor, pow, ballpark, glitchy, billy, dallas, rip, called, outlooks,
@@
 
-\begin{block}{Our human-related words (examples, full list size = ADD)}
-\protect\hypertarget{our-human-related-words-examples-full-list-size-add}{}
-switch, studio, stick, soup, sometimes, signal, prior, plant, photo,
-path, park, near, menu, latter, grass, clock, wear, walk, visitor, toy,
-tissue, throw, talk, speak, sleep, eye, enjoy, blogger, character,
-candidate, breakfast, supper, dinner, eat, drink, carry, run, cast, ask,
-awake, ear, nose, lunch
+\begin{block}{Our human-related words (examples, full list size = 27)}
+\protect\hypertarget{our-human-related-words-examples-full-list-size-27}{}
+wear, walk, visitor, toy, tissue, throw, talk, speak, sleep, eye, enjoy,
+blogger, character, candidate, breakfast, supper, dinner, eat, drink,
+carry, run, cast, ask, awake, ear, nose, lunch
 \end{block}
@@
 \toprule
-Category & Biased & Hard Debiased & Soft Debiased & Diff \\
+Category & Biased & Hard Debiased & Diff & \\
 \midrule
 \endhead
-Religion & 0.859 & 0.934 & 0.894 & 0.075 \\
-Race & 0.892 & 0.925 & 0.985 & 0.033 \\
-Gender & 0.623 & 0.700 & 0.747 & 0.077 \\
+Religion & 0.859 & 0.934 & 0,141 & \\
+Race & 0.892 & 0.925 & 0,108 & \\
+Gender & 0.623 & 0.700 & 0,377 & \\
 \bottomrule

### 2022-03-20 presentations/presentationBoston/presentationBoston.Rmd
Adds the rest of violins

@@
 
+## Some methodological problems
+
+### No principled interpretation - visualization 1
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "90%"}
+library(ggplot2)
+library(tidyverse)
+library(gridExtra)
+library(kableExtra)
+library(magrittr)
+library(ggplot2)
+library(ggpubr)
+library(ggExtra)
+library(ggthemes)
+library(ggforce)
+library(latex2exp)
+
+religionReddit <- read.csv("../../datasets/religionReddit.csv")[-1]
+
+cosAssReligion <- religionReddit['cosine_similarity'][which(religionReddit$connection == 'associated'),]
+cosDiffReligion <- religionReddit['cosine_similarity'][which(religionReddit$connection == 'different'),]
+cosNoneReligion <- religionReddit['cosine_similarity'][which(religionReddit$connection == 'none'),]
+cosHumReligion <- religionReddit['cosine_similarity'][which(religionReddit$connection == 'human'),]
+
+ggplot(religionReddit, aes(x = connection, y = cosine_similarity, fill = connection, color = connection))+
+  geom_violin(alpha = .5)+coord_flip()+
+  geom_vline(xintercept = .14) + geom_hline(yintercept = -.14, color = "grey") +
+  ylab("similarity")+theme_tufte()+ geom_hline(yintercept = .14, color = "grey")+
+  ggtitle("Religion: empirical distribution of cosine similarity")+
+  annotate(geom = "label", x = 4, y = 0, label = "56%")+
+  annotate(geom = "label", x = 3, y = 0, label = "55%")+
+  annotate(geom = "label", x = 2, y = 0, label = "43%")+
+  annotate(geom = "label", x = 1, y = 0, label = "36%")
+```
+
+## Some methodological problems
+
+### No principled interpretation - visualization 2
+
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "90%"}
+library(ggplot2)
+library(tidyverse)
+library(gridExtra)
+library(kableExtra)
+library(magrittr)
+library(ggplot2)
+library(ggpubr)
+library(ggExtra)
+library(ggthemes)
+library(ggforce)
+library(latex2exp)
+
+raceReddit <- read.csv("../../datasets/raceReddit.csv")[-1]
+
+cosAssReligion <- raceReddit['cosine_similarity'][which(raceReddit$connection == 'associated'),]
+cosDiffReligion <- raceReddit['cosine_similarity'][which(raceReddit$connection == 'different'),]
+cosNoneReligion <- raceReddit['cosine_similarity'][which(raceReddit$connection == 'none'),]
+cosHumReligion <- raceReddit['cosine_similarity'][which(raceReddit$connection == 'human'),]
+
+ggplot(raceReddit, aes(x = connection, y = cosine_similarity, fill = connection, color = connection))+
+  geom_violin(alpha = .5)+coord_flip()+
+  geom_vline(xintercept = .11) + geom_hline(yintercept = -.11, color = "grey") +
+  ylab("similarity")+theme_tufte()+ geom_hline(yintercept = .11, color = "grey")+
+  ggtitle("Race: empirical distribution of cosine similarity")+
+  annotate(geom = "label", x = 4, y = 0, label = "46%")+
+  annotate(geom = "label", x = 3, y = 0, label = "67%")+
+  annotate(geom = "label", x = 2, y = 0, label = "39%")+
+  annotate(geom = "label", x = 1, y = 0, label = "33%")
+```
+
+## Some methodological problems
+
+### No principled interpretation - visualization 3
+
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "90%"}
+library(ggplot2)
+library(tidyverse)
+library(gridExtra)
+library(kableExtra)
+library(magrittr)
+library(ggplot2)
+library(ggpubr)
+library(ggExtra)
+library(ggthemes)
+library(ggforce)
+library(latex2exp)
+
+genderReddit <- read.csv("../../datasets/genderReddit.csv")[-1]
+
+cosAssReligion <- genderReddit['cosine_similarity'][which(genderReddit$connection == 'associated'),]
+cosDiffReligion <- genderReddit['cosine_similarity'][which(genderReddit$connection == 'different'),]
+cosNoneReligion <- genderReddit['cosine_similarity'][which(genderReddit$connection == 'none'),]
+cosHumReligion <- genderReddit['cosine_similarity'][which(genderReddit$connection == 'human'),]
+
+ggplot(genderReddit, aes(x = connection, y = cosine_similarity, fill = connection, color = connection))+
+  geom_violin(alpha = .5)+coord_flip()+
+  geom_vline(xintercept = .38) + geom_hline(yintercept = -.38, color = "grey") +
+  ylab("similarity")+theme_tufte()+ geom_hline(yintercept = .38, color = "grey")+
+  ggtitle("Gender: empirical distribution of cosine similarity")+
+  annotate(geom = "label", x = 4, y = 0, label = "96%")+
+  annotate(geom = "label", x = 3, y = 0, label = "87%")+
+  annotate(geom = "label", x = 2, y = 0, label = "53%")+
+  annotate(geom = "label", x = 1, y = 0, label = "47%")
+```
+
 ## The problem with pre-averaging

### 2022-03-20 presentations/presentationBoston/presentationBoston.tex
Adds the rest of violins

@@
 
+\begin{frame}{Some methodological problems}
+\protect\hypertarget{some-methodological-problems-7}{}
+\begin{block}{No principled interpretation - visualization 1}
+\protect\hypertarget{no-principled-interpretation---visualization-1}{}
+\begin{center}\includegraphics[width=0.9\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-6-1} \end{center}
+\end{block}
+\end{frame}
+
+\begin{frame}{Some methodological problems}
+\protect\hypertarget{some-methodological-problems-8}{}
+\begin{block}{No principled interpretation - visualization 2}
+\protect\hypertarget{no-principled-interpretation---visualization-2}{}
+\begin{center}\includegraphics[width=0.9\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-7-1} \end{center}
+\end{block}
+\end{frame}
+
+\begin{frame}{Some methodological problems}
+\protect\hypertarget{some-methodological-problems-9}{}
+\begin{block}{No principled interpretation - visualization 3}
+\protect\hypertarget{no-principled-interpretation---visualization-3}{}
+\begin{center}\includegraphics[width=0.9\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-8-1} \end{center}
+\end{block}
+\end{frame}
+
 \begin{frame}{The problem with pre-averaging}
@@
 
-\begin{center}\includegraphics[width=0.7\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-6-1} \end{center}
+\begin{center}\includegraphics[width=0.7\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-9-1} \end{center}
 \normalsize
@@
 
-\begin{center}\includegraphics[width=0.7\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-8-1} \end{center}
+\begin{center}\includegraphics[width=0.7\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-11-1} \end{center}
 
@@
 
-\begin{center}\includegraphics[width=0.8\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-9-1} \end{center}
+\begin{center}\includegraphics[width=0.8\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-12-1} \end{center}
 \vspace{1mm}
@@
 
-\begin{center}\includegraphics[width=0.8\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-11-1} \end{center}
+\begin{center}\includegraphics[width=0.8\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-14-1} \end{center}
 \vspace{1mm}
@@
 
-\begin{center}\includegraphics[width=0.8\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-13-1} \end{center}
+\begin{center}\includegraphics[width=0.8\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-16-1} \end{center}
 \vspace{1mm}

### 2022-03-20 functions/cosineDistribution2.R

@@ -130,7 +130,7 @@ ggplot()+geom_histogram(aes(x=cosAssGender, y = ..density..), alpha = 0.8, bins=
   theme_tufte()+labs(title="Cosine similarity distribution")+ xlab("similarity") +
   geom_vline(xintercept = .38) + geom_vline(xintercept = -.38) 
 
-(sum(abs(cosAssGender)>0.38)) / length(cosAssGender)
+1-(sum(abs(cosAssGender)>0.38)) / length(cosAssGender)
 
 
 # ------------ Different
@@ -138,7 +138,7 @@ ggplot()+geom_histogram(aes(x=cosAssGender, y = ..density..), alpha = 0.8, bins=
 ggplot()+geom_histogram(aes(x=cosDiffGender, y = ..density..), alpha = 0.8, bins=20, color="pink", fill="pink")+
   theme_tufte()+labs(title="Cosine similarity distribution")+ xlab("similarity") + geom_vline(xintercept = .38) + geom_vline(xintercept = -.38) 
 
-(sum(abs(cosDiffGender)>0.38)) / length(cosDiffGender)
+1- (sum(abs(cosDiffGender)>0.38)) / length(cosDiffGender)
 
 
 # ------------- Neutral 
@@ -146,7 +146,7 @@ ggplot()+geom_histogram(aes(x=cosDiffGender, y = ..density..), alpha = 0.8, bins
 ggplot()+geom_histogram(aes(x=cosNoneGender, y = ..density..), alpha = 0.8, bins=20, color="lightblue", fill="lightblue")+
   theme_tufte()+labs(title="Cosine similarity distribution")+ xlab("similarity") + geom_vline(xintercept = .38) + geom_vline(xintercept = -.38) 
 
-(sum(abs(cosNoneGender)>0.38)) / length(cosNoneGender)
+1- (sum(abs(cosNoneGender)>0.38)) / length(cosNoneGender)
 
 
 
@@ -155,7 +155,7 @@ ggplot()+geom_histogram(aes(x=cosNoneGender, y = ..density..), alpha = 0.8, bins
 ggplot()+geom_histogram(aes(x=cosHumGender, y = ..density..), alpha = 0.8, bins=20, color="lightblue", fill="lightblue")+
[…]

### 2022-03-21 presentations/presentationBoston/presentationBoston.Rmd
Adds final version

--- 2022-03-20 presentations/presentationBoston/presentationBoston.Rmd
@@
 
+- Word embeddings 
+- Cosine similarity
 - Bias in word embeddings
-- WEAT and MAC methods
+- WEAT and MAC measures
 - Methodological problems
-- Limitations of pre-averaging in bias detection methods
-- Accounting for uncertainty with Bayesian approach
-
-## Word2vec
+- Bad consequences of pre-averaging
+- Our Bayesian alternative
+
+## Word embeddings
 
@@
 
-\dots \dots but this would be inefficient and wouldn't capture any relations between words.
-
-## Word2vec
+\dots but this would be inefficient and wouldn't capture any relations between words.
+
+## Word embeddings
 
@@
 
-## Word2vec
+## Word embeddings
 
@@
 \begin{center}
- \includegraphics[height = 6cm, width = 5cm]{images/perceptron2.png}
+ \includegraphics[height = 5.5cm, width = 5cm]{images/perceptron2.png}
 \end{center}
 
-\vspace{-3mm}
+\vspace{2mm}
 \tiny \hfill \color{gray}Illustration: M. Mitchell \color{black}
@@
 
-\item "the degree of semantic similarity between two linguistic expressions $A$ and $B$ is a function of the similarity of the linguistic contexts in which $A$ and $B$ can appear." (A. Lenci, 2008)
+\item "\dots the degree of semantic similarity between two linguistic expressions $A$ and $B$ is a function of the similarity of the linguistic contexts in which $A$ and $B$ can appear." (A.
+Lenci, 2008)
 
@@
 
+\vspace{2mm}
+
 \begin{center}
- \includegraphics[height = 3cm, width = 7cm]{images/word2vec2.png}
+ \includegraphics[height = 3.5cm, width = 7cm]{images/word2vec2.png}
 \end{center}
@@
 
-```{r cosine1,echo=FALSE,eval=TRUE,fig.align = "center", cache=TRUE, fig.show = "hold", out.width = "100%",   message = FALSE, warning = FALSE, results = FALSE}
+```{r cosine1,echo=FALSE,eval=TRUE,fig.align = "center", cache=TRUE, fig.show = "hold", out.width = "115%",   message = FALSE, warning = FALSE, results = FALSE}
 a <- c(1,2)
@@
   geom_segment(aes(x = 0, y = 0, xend = b[1], yend = b[2]), arrow = arrow(length = unit(0.02, "npc")))+   annotate(geom = "text", label  = "b = [3,2.5]", x = 3.2, y = 2.9, size = 7)+
-  theme_tufte(base_size = 24)
+  theme_tufte(base_size = 30)
 
@@
 
-```{r cosine2,echo=FALSE,eval=TRUE,fig.align = "center", cache=TRUE, fig.show = "hold", out.width = "100%",   message = FALSE, warning = FALSE, results = FALSE}
+```{r cosine2,echo=FALSE,eval=TRUE,fig.align = "center", cache=TRUE, fig.show = "hold", out.width = "115%",   message = FALSE, warning = FALSE, results = FALSE}
 a <- c(1,2)
@@
   annotate(geom = "text", label  = "b - a = [2,.5]", x = 2.2, y = .9, size = 7)+
-  theme_tufte(base_size = 24)
+  theme_tufte(base_size = 30)
 
@@
 
-```{r cosine3,echo=FALSE,eval=TRUE,fig.align = "center", cache=TRUE, fig.show = "hold", out.width = "100%",   message = FALSE, warning = FALSE, results = FALSE}
+```{r cosine3,echo=FALSE,eval=TRUE,fig.align = "center", cache=TRUE, fig.show = "hold", out.width = "115%",   message = FALSE, warning = FALSE, results = FALSE}
 a <- c(1,2)
@@
   annotate(geom = "text", label  = "b - a = [2,.5] (shifted)", x = 2, y = 2.7, size = 7)+
-  theme_tufte(base_size = 24)
+  theme_tufte(base_size = 30)
 
@@
 
-```{r cosine4,echo=FALSE,eval=TRUE,fig.align = "center", cache=TRUE, fig.show = "hold", out.width = "100%",   message = FALSE, warning = FALSE, results = FALSE}
+```{r cosine4,echo=FALSE,eval=TRUE,fig.align = "center", cache=TRUE, fig.show = "hold", out.width = "115%",   message = FALSE, warning = FALSE, results = FALSE}
 a <- c(1,2)
@@
   annotate(geom = "text", label  = "b - a = [2,.5] (shifted)", x = 2, y = 2.7, size = 7)+
-  theme_tufte(base_size = 24)+
+  theme_tufte(base_size = 30)+
   annotate(geom = "text", label  = expression(theta), x = .6, y = .78, size = 7)+
@@
     - $-1$: opposites 
-    - $0$: dissimilar
+    - $0$: lack of similarity
 
@@
 
-- @Ethayarajh2020measuring criticizes WEAT uses Bernstein bounds and argues that we would need a bias specific dataset of size at least 11903 to claim that the system is biased (three times larger than WinoBias).
+- @Ethayarajh2020measuring uses Bernstein bounds to criticize WEAT, and argues that we would need a bias specific dataset of size at least 11903 to claim that the system is biased (three times larger than WinoBias).
 
@@
 
-We make the subclasses clear, add human neutral predicates and neutral predicates for control. We used L2-Reddit corpus and GoogleNews (we present the results for Reddit for brevity).
+We make the subclasses clear, add human neutral predicates and neutral predicates for control.
+We used L2-Reddit corpus and GoogleNews \linebreak \footnotesize (we present the results for Reddit for brevity).
+\normalsize
 
@@
 
-### Distances for  "muslim"
-
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "80%",message =FALSE, warning = FALSE}
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%",message =FALSE, warning = FALSE}
 library(tidyverse)
@@
 
-### Distances for  "priest"
-
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "80%"}
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
 library(tidyverse)
@@
 -------| ------| -------| ----- | --------
-Religion | 0.859 | 0.934 | 0,141 
-Race | 0.892 | 0.925| 0,108 
-Gender | 0.623 | 0.700 | 0,377
+Religion | 0.859 | 0.934 | 0.141 
+Race | 0.892 | 0.925| 0.108 
+Gender | 0.623 | 0.700 | 0.377
 
@@
 
+## Some methodological problems
+
+### No principled interpretation 
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "90%"}
+library(ggplot2)
+library(tidyverse)
+library(gridExtra)
+library(kableExtra)
+library(magrittr)
+library(ggplot2)
+library(ggpubr)
+library(ggExtra)
+library(ggthemes)
+library(ggforce)
+library(latex2exp)
+
+religionReddit <- read.csv("../../datasets/religionReddit.csv")[-1]
+
+cosAssReligion <- religionReddit['cosine_similarity'][which(religionReddit$connection == 'associated'),]
+cosDiffReligion <- religionReddit['cosine_similarity'][which(religionReddit$connection == 'different'),]
+cosNoneReligion <- religionReddit['cosine_similarity'][which(religionReddit$connection == 'none'),]
+cosHumReligion <- religionReddit['cosine_similarity'][which(religionReddit$connection == 'human'),]
+
+ggplot(religionReddit, aes(x = connection, y = cosine_similarity, fill = connection, color = connection))+
+  geom_violin(alpha = .5)+coord_flip()+
+  geom_vline(xintercept = .14) + geom_hline(yintercept = -.14, color = "grey") +
+  ylab("similarity")+theme_tufte(base_size=20)+ geom_hline(yintercept = .14, color = "grey")+
+  ggtitle("Religion, empirical distribution (threshold=.14)")+
+  annotate(geom = "label", x = 4, y = 0, label = "56%")+
+  annotate(geom = "label", x = 3, y = 0, label = "55%")+
+  annotate(geom = "label", x = 2, y = 0, label = "43%")+
+  annotate(geom = "label", x = 1, y = 0, label = "36%")
+```
+
+## Some methodological problems
+
+### No principled interpretation
+
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "90%"}
+library(ggplot2)
+library(tidyverse)
+library(gridExtra)
+library(kableExtra)
+library(magrittr)
+library(ggplot2)
+library(ggpubr)
+library(ggExtra)
+library(ggthemes)
+library(ggforce)
+library(latex2exp)
+
+raceReddit <- read.csv("../../datasets/raceReddit.csv")[-1]
+
+cosAssReligion <- raceReddit['cosine_similarity'][which(raceReddit$connection == 'associated'),]
+cosDiffReligion <- raceReddit['cosine_similarity'][which(raceReddit$connection == 'different'),]
+cosNoneReligion <- raceReddit['cosine_similarity'][which(raceReddit$connection == 'none'),]
+cosHumReligion <- raceReddit['cosine_similarity'][which(raceReddit$connection == 'human'),]
+
+ggplot(raceReddit, aes(x = connection, y = cosine_similarity, fill = connection, color = connection))+
+  geom_violin(alpha = .5)+coord_flip()+
+  geom_vline(xintercept = .11) + geom_hline(yintercept = -.11, color = "grey") +
+  ylab("similarity")+theme_tufte(base_size=20)+ geom_hline(yintercept = .11, color = "grey")+
+  ggtitle("Race, empirical distribution (threshold=.11)")+
+  annotate(geom = "label", x = 4, y = 0, label = "46%")+
+  annotate(geom = "label", x = 3, y = 0, label = "67%")+
+  annotate(geom = "label", x = 2, y = 0, label = "39%")+
+  annotate(geom = "label", x = 1, y = 0, label = "33%")
+```
+
+## Some methodological problems
+
+### No principled interpretation
+
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "90%"}
+library(ggplot2)
+library(tidyverse)
+library(gridExtra)
+library(kableExtra)
+library(magrittr)
+library(ggplot2)
+library(ggpubr)
+library(ggExtra)
+library(ggthemes)
+library(ggforce)
+library(latex2exp)
+
+genderReddit <- read.csv("../../datasets/genderReddit.csv")[-1]
+
+cosAssReligion <- genderReddit['cosine_similarity'][which(genderReddit$connection == 'associated'),]
+cosDiffReligion <- genderReddit['cosine_similarity'][which(genderReddit$connection == 'different'),]
+cosNoneReligion <- genderReddit['cosine_similarity'][which(genderReddit$connection == 'none'),]
+cosHumReligion <- genderReddit['cosine_similarity'][which(genderReddit$connection == 'human'),]
+
+ggplot(genderReddit, aes(x = connection, y = cosine_similarity, fill = connection, color = connection))+
+  geom_violin(alpha = .5)+coord_flip()+
+  geom_vline(xintercept = .38) + geom_hline(yintercept = -.38, color = "grey") +
+  ylab("similarity")+theme_tufte(base_size=20)+ geom_hline(yintercept = .38, color = "grey")+
+  ggtitle("Gender, empirical distribution (threshold=.38)")+
+  annotate(geom = "label", x = 4, y = 0, label = "96%")+
+  annotate(geom = "label", x = 3, y = 0, label = "87%")+
+  annotate(geom = "label", x = 2, y = 0, label = "53%")+
+  annotate(geom = "label", x = 1, y = 0, label = "47%")
+```
+
 ## The problem with pre-averaging
@@
 
-### Simple case: two pws, four terms
-
 \vspace{1mm}
 \footnotesize
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "70%"}
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "95%"}
 set.seed(123)
@@
 
-- Raw sd in data is `r round(rawsd,3)`
-- The sd of means is `r round(factor,3)`
-- The WEAT score is `r round(numerator/factor,3)`   
-- The largest effect size reported by @Caliskan2017semanticsBiases is 1.81!
+- Raw sd in data is `r round(rawsd,3)`, the sd of means is `r round(factor,3)`
+- The WEAT score is `r round(numerator/factor,3)`, the largest effect size reported by @Caliskan2017semanticsBiases is 1.81!
 
@@
 
-### Simulation with realistic set-up (16 predicates)
-
 \vspace{1mm}
 \footnotesize
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "70%"}
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "95%"}
 set.seed(123456)
@@
 
-### 10k simulations (same parameters)
-
 \vspace{1mm}
 \footnotesize
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "80%"}
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "90%"}
 
@@
 ggplot()+geom_histogram(aes(x=biasesD0, y = ..density..), alpha = 0.6, bins=50)+
-  theme_tufte()+labs(title="10k biases for same means and sd =.08")+ xlab("bias")
+  theme_tufte(base_size=18)+labs(title="10k biases for same means and sd =.08")+ xlab("bias")
 
@@
 
-### 10k simulations with mean similarity  0.1
-
 \vspace{1mm}
 \footnotesize
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "80%"}
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "90%"}
 set.seed(123456)
@@
 ggplot()+geom_histogram(aes(x=biasesD1, y = ..density..), alpha = 0.6, bins=50)+
-  theme_tufte()+labs(title="10k biases for different means 0.1 and sd =.08")+ xlab("bias")
+  theme_tufte(base_size=18)+labs(title="10k biases for different means 0.1 and sd =.08")+ xlab("bias")
 
@@
 
-### 10k simulations with mean similarity  0.4
-
 \vspace{1mm}
 \footnotesize
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "80%"}
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "90%"}
 set.seed(123456)
@@
 ggplot()+geom_histogram(aes(x=biasesD4, y = ..density..), alpha = 0.6, bins=50)+
-  theme_tufte()+labs(title="10k biases for different means 0.4 and sd =.08")+ xlab("bias")
+  theme_tufte(base_size=18)+labs(title="10k biases for different means 0.4 and sd =.08")+ xlab("bias")
 
@@
 
-- Bias in word embeddings
-- WEAT and MAC methods
-- Methodological problems
-- Limitations of pre-averaging in bias detection methods
-- Accounting for uncertainty with Bayesian approach
-
-### Further work
-
-- Including contrasts in Bayesian calculation
-- Performance cross-validation in comparison to other methods (regular linear regression, KNN, \dots)
-- Downstream tasks and connection with intrinsic evaluation
-- Testing data from the original Implicit Association Test (IAT) 
-- Applying uncertainty to WEAT and better word lists
-- Looking at other similarity measures
+\pause
+
+- The existing measures of bias by pre-averaging oversimplify the picture and lead to false confidence
+
+\pause
+
+- Our simulation results show that the mainstream interpretations are too hasty
+
+\pause
+
+- Bayesian analysis with proper control groups suggests that:
+
+  - The word lists in use are too short
+  - Associated and different groups behave similarly, which undermines the semantic interpretation of cosine similarity
+  - Analysis at different levels of generality shows that there is more variety than one may initially think
+
+\pause
+
+### Please remember about the feedback!
+
+<!-- ### Further work -->
+
+<!-- - Including contrasts in Bayesian calculation -->
+<!-- - Performance cross-validation in comparison to other methods (regular linear regression, KNN, \dots) -->
+<!-- - Downstream tasks and connection with intrinsic evaluation -->
+<!-- - Testing data from the original Implicit Association Test (IAT)  -->
+<!-- - Applying uncertainty to WEAT and better word lists -->
+<!-- - Looking at other similarity measures -->
 

### 2022-03-21 presentations/presentationBoston/presentationBoston.tex
Adds final version

@@
 \item
+  Word embeddings
+\item
+  Cosine similarity
+\item
   Bias in word embeddings
 \item
-  WEAT and MAC methods
+  WEAT and MAC measures
 \item
@@
 \item
-  Limitations of pre-averaging in bias detection methods
-\item
-  Accounting for uncertainty with Bayesian approach
-\end{itemize}
-\end{frame}
-
-\begin{frame}{Word2vec}
-\protect\hypertarget{word2vec}{}
+  Bad consequences of pre-averaging
+\item
+  Our Bayesian alternative
+\end{itemize}
+\end{frame}
+
+\begin{frame}{Word embeddings}
+\protect\hypertarget{word-embeddings}{}
 \begin{block}{Question}
@@
 
-\dots \dots but this would be inefficient and wouldn't capture any
-relations between words.
-\end{block}
-\end{frame}
-
-\begin{frame}{Word2vec}
-\protect\hypertarget{word2vec-1}{}
+\dots but this would be inefficient and wouldn't capture any relations
+between words.
+\end{block}
+\end{frame}
+
+\begin{frame}{Word embeddings}
+\protect\hypertarget{word-embeddings-1}{}
 \begin{center}
@@
 
-\begin{frame}{Word2vec}
-\protect\hypertarget{word2vec-2}{}
+\begin{frame}{Word embeddings}
+\protect\hypertarget{word-embeddings-2}{}
 \begin{center}
@@
 
-\begin{frame}{Word2vec}
-\protect\hypertarget{word2vec-3}{}
+\begin{frame}{Word embeddings}
+\protect\hypertarget{word-embeddings-3}{}
 \begin{columns}
@@
 
-\begin{frame}{Word2vec}
-\protect\hypertarget{word2vec-4}{}
+\begin{frame}{Word embeddings}
+\protect\hypertarget{word-embeddings-4}{}
 \begin{block}{Distributional semantics}
@@
 
-\begin{frame}{Word2vec}
-\protect\hypertarget{word2vec-5}{}
+\begin{frame}{Word embeddings}
+\protect\hypertarget{word-embeddings-5}{}
 \begin{block}{Google and Mikolov}
@@
 
-\begin{frame}{Word2vec}
-\protect\hypertarget{word2vec-6}{}
+\begin{frame}{Word embeddings}
+\protect\hypertarget{word-embeddings-6}{}
 \begin{block}{Google and Mikolov}
@@
 
-\begin{frame}{Word2vec}
-\protect\hypertarget{word2vec-7}{}
+\begin{frame}{Word embeddings}
+\protect\hypertarget{word-embeddings-7}{}
 \begin{longtable}[]{@{}llllll@{}}
@@
 
-\begin{center}\includegraphics[width=1\linewidth]{presentationBoston_files/figure-beamer/cosine1-1} \end{center}
+\begin{center}\includegraphics[width=1.15\linewidth]{presentationBoston_files/figure-beamer/cosine1-1} \end{center}
 
@@
 
-\begin{center}\includegraphics[width=1\linewidth]{presentationBoston_files/figure-beamer/cosine2-1} \end{center}
+\begin{center}\includegraphics[width=1.15\linewidth]{presentationBoston_files/figure-beamer/cosine2-1} \end{center}
 
@@
 
-\begin{center}\includegraphics[width=1\linewidth]{presentationBoston_files/figure-beamer/cosine3-1} \end{center}
+\begin{center}\includegraphics[width=1.15\linewidth]{presentationBoston_files/figure-beamer/cosine3-1} \end{center}
 
@@
 
-\begin{center}\includegraphics[width=1\linewidth]{presentationBoston_files/figure-beamer/cosine4-1} \end{center}
+\begin{center}\includegraphics[width=1.15\linewidth]{presentationBoston_files/figure-beamer/cosine4-1} \end{center}
 
@@
   \item
-    \(0\): dissimilar
+    \(0\): lack of similarity
   \end{itemize}
@@
   \protect\hyperlink{ref-Ethayarajh2020measuring}{Ethayarajh}
-  (\protect\hyperlink{ref-Ethayarajh2020measuring}{2020}) criticizes
-  WEAT uses Bernstein bounds and argues that we would need a bias
+  (\protect\hyperlink{ref-Ethayarajh2020measuring}{2020}) uses Bernstein
+  bounds to criticize WEAT, and argues that we would need a bias
   specific dataset of size at least 11903 to claim that the system is
@@
 We make the subclasses clear, add human neutral predicates and neutral
-predicates for control. We used L2-Reddit corpus and GoogleNews (we
-present the results for Reddit for brevity).
+predicates for control. We used L2-Reddit corpus and GoogleNews
+\linebreak \footnotesize (we present the results for Reddit for
+brevity). \normalsize
 
@@
 \protect\hypertarget{some-methodological-problems-4}{}
-\begin{block}{Distances for ``muslim''}
-\protect\hypertarget{distances-for-muslim}{}
-\begin{center}\includegraphics[width=0.8\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-4-1} \end{center}
-\end{block}
+\begin{center}\includegraphics[width=1\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-4-1} \end{center}
 \end{frame}
@@
 \protect\hypertarget{some-methodological-problems-5}{}
-\begin{block}{Distances for ``priest''}
-\protect\hypertarget{distances-for-priest}{}
-\begin{center}\includegraphics[width=0.8\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-5-1} \end{center}
-\end{block}
+\begin{center}\includegraphics[width=1\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-5-1} \end{center}
 \end{frame}
@@
 \endhead
-Religion & 0.859 & 0.934 & 0,141 & \\
-Race & 0.892 & 0.925 & 0,108 & \\
-Gender & 0.623 & 0.700 & 0,377 & \\
+Religion & 0.859 & 0.934 & 0.141 & \\
+Race & 0.892 & 0.925 & 0.108 & \\
+Gender & 0.623 & 0.700 & 0.377 & \\
 \bottomrule
@@
 \protect\hypertarget{some-methodological-problems-7}{}
-\begin{block}{No principled interpretation - visualization 1}
-\protect\hypertarget{no-principled-interpretation---visualization-1}{}
+\begin{block}{No principled interpretation}
+\protect\hypertarget{no-principled-interpretation-1}{}
 \begin{center}\includegraphics[width=0.9\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-6-1} \end{center}
@@
 \protect\hypertarget{some-methodological-problems-8}{}
-\begin{block}{No principled interpretation - visualization 2}
-\protect\hypertarget{no-principled-interpretation---visualization-2}{}
+\begin{block}{No principled interpretation}
+\protect\hypertarget{no-principled-interpretation-2}{}
 \begin{center}\includegraphics[width=0.9\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-7-1} \end{center}
@@
 \protect\hypertarget{some-methodological-problems-9}{}
-\begin{block}{No principled interpretation - visualization 3}
-\protect\hypertarget{no-principled-interpretation---visualization-3}{}
+\begin{block}{No principled interpretation}
+\protect\hypertarget{no-principled-interpretation-3}{}
 \begin{center}\includegraphics[width=0.9\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-8-1} \end{center}
@@
 \protect\hypertarget{the-problem-with-pre-averaging-1}{}
-\begin{block}{Simple case: two pws, four terms}
-\protect\hypertarget{simple-case-two-pws-four-terms}{}
 \vspace{1mm}
@@
 
-\begin{center}\includegraphics[width=0.7\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-9-1} \end{center}
+\begin{center}\includegraphics[width=0.95\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-9-1} \end{center}
 \normalsize
@@
 \item
-  Raw sd in data is 0.072
-\item
-  The sd of means is 0.037
-\item
-  The WEAT score is 1.825\\
-\item
-  The largest effect size reported by
+  Raw sd in data is 0.072, the sd of means is 0.037
+\item
+  The WEAT score is 1.825, the largest effect size reported by
   \protect\hyperlink{ref-Caliskan2017semanticsBiases}{Caliskan, Bryson,
@@
 \protect\hypertarget{the-problem-with-pre-averaging-on-realistic-set-up}{}
-\begin{block}{Simulation with realistic set-up (16 predicates)}
-\protect\hypertarget{simulation-with-realistic-set-up-16-predicates}{}
 \vspace{1mm}
@@
 
-\begin{center}\includegraphics[width=0.7\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-11-1} \end{center}
+\begin{center}\includegraphics[width=0.95\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-11-1} \end{center}
 
@@
 Raw sd: 0.082, sd of means: 0.031, WEAT: 2.337.
-\end{block}
 \end{frame}
@@
 \protect\hypertarget{the-problem-with-pre-averaging-on-realistic-set-up-1}{}
-\begin{block}{10k simulations (same parameters)}
-\protect\hypertarget{k-simulations-same-parameters}{}
 \vspace{1mm}
@@
 
-\begin{center}\includegraphics[width=0.8\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-12-1} \end{center}
+\begin{center}\includegraphics[width=0.9\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-12-1} \end{center}
 \vspace{1mm}
@@
 \protect\hypertarget{the-problem-with-pre-averaging-on-realistic-set-up-2}{}
-\begin{block}{10k simulations with mean similarity 0.1}
-\protect\hypertarget{k-simulations-with-mean-similarity-0.1}{}
 \vspace{1mm}
@@
 
-\begin{center}\includegraphics[width=0.8\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-14-1} \end{center}
+\begin{center}\includegraphics[width=0.9\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-14-1} \end{center}
 \vspace{1mm}
@@
 \protect\hypertarget{the-problem-with-pre-averaging-on-realistic-set-up-3}{}
-\begin{block}{10k simulations with mean similarity 0.4}
-\protect\hypertarget{k-simulations-with-mean-similarity-0.4}{}
 \vspace{1mm}
@@
 
-\begin{center}\includegraphics[width=0.8\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-16-1} \end{center}
+\begin{center}\includegraphics[width=0.9\linewidth]{presentationBoston_files/figure-beamer/unnamed-chunk-16-1} \end{center}
 \vspace{1mm}
@@
 \protect\hypertarget{summary}{}
-\begin{itemize}
-\tightlist
-\item
-  Bias in word embeddings
-\item
-  WEAT and MAC methods
-\item
-  Methodological problems
-\item
-  Limitations of pre-averaging in bias detection methods
-\item
-  Accounting for uncertainty with Bayesian approach
-\end{itemize}
-\end{block}
-
-\begin{block}{Further work}
-\protect\hypertarget{further-work}{}
-\begin{itemize}
-\tightlist
-\item
-  Including contrasts in Bayesian calculation
-\item
-  Performance cross-validation in comparison to other methods (regular
-  linear regression, KNN, \dots)
-\item
-  Downstream tasks and connection with intrinsic evaluation
-\item
-  Testing data from the original Implicit Association Test (IAT)
-\item
-  Applying uncertainty to WEAT and better word lists
-\item
-  Looking at other similarity measures
-\end{itemize}
+\pause
+
+\begin{itemize}
+\tightlist
+\item
+  The existing measures of bias by pre-averaging oversimplify the
+  picture and lead to false confidence
+\end{itemize}
+
+\pause
+
+\begin{itemize}
+\tightlist
+\item
+  Our simulation results show that the mainstream interpretations are
+  too hasty
+\end{itemize}
+
+\pause
+
+\begin{itemize}
+\item
+  Bayesian analysis with proper control groups suggests that:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    The word lists in use are too short
+  \item
+    Associated and different groups behave similarly, which undermines
+    the semantic interpretation of cosine similarity
+  \item
+    Analysis at different levels of generality shows that there is more
+    variety than one may initially think
+  \end{itemize}
+\end{itemize}
+
+\pause
+\end{block}
+
+\begin{block}{Please remember about the feedback!}
+\protect\hypertarget{please-remember-about-the-feedback}{}
 \end{block}

### 2022-04-01 abstractESSLLI21/abstractESSLLI1.Rmd
Adds bitly

@@
 \begin{center}
-\Large \textbf{A bayesian method of cosine-based word2vec bias estimation}
+\large \textbf{Taking uncertainty seriously}
+
+\textbf{A  Bayesian approach to  word embedding bias estimation }
 \end{center}
@@
 
-A considerable amount of literature exists on  bias detection and mitigation in NLP models, especially word2vec embeddings, which represent words as  vectors of real numbers (see e.g.
-@bolukbasi2016man and @Gonen2019lipstick and references therein).
-The most common method used compares  cosine similarity between words from protected groups and attributes that are considered to be stereotypical or harmful in some way, and this method will be in our focus.
+A considerable amount of literature exists on  bias detection and mitigation in NLP models, especially word embeddings, which represent words as  vectors of real numbers (see e.g.
+@Bolukbasi2016man and @Gonen2019lipstick and references therein).
+The most common methods used compare cosine similarity or distance between words from protected groups and attributes that are considered to be stereotypical or harmful in some way.
+Many of them rely on pre-averaging, which we find problematic --- two such methods will be in our focus.
+We will argue that the existing measures of bias by pre-averaging oversimplify the picture and lead to false confidence.
+Our simulation results show that the mainstream interpretations are too hasty.
+Our Bayesian analysis with proper control groups suggests that (1) the word lists in use are too short, (2) associated and different groups behave similarly, which undermines the semantic interpretation of cosine similarity, and (3) our analysis at different levels of generality shows that there is more variety than one may initially think.
 
@@
 @Lauscher2019multidimensional extend the methodology to a multilingual and cross-lingual setting; a similar methodology is employed by @Garg2018years, who use word embeddings trained on corpora from different decades.
-@Manzini2019blackToCriminal modify WEAT to a multi-class setting, introducing Mean Average Cosine (MAC) similarity as a measure of bias (in fact, in the paper  they report distances rather than similarities).
+@Manzini2019blackToCriminal modify WEAT to a multi-class setting, introducing Mean Average Cosine (MAC) similarity distance as a measure of bias.
 Let $T = \{t_1, \dots, t_k\}$ be a class of protected word embeddings, and let each $A_j\in A$ be a set of attributes stereotypically associated with a protected word).
@@
 
-\noindent
-\begin{tabular}{ll}
-\begin{minipage}[c]{0.35\linewidth}
-\scalebox{0.65}{
+\begin{center}
 \begin{tabular}{cccccc}\toprule
@@
 \end{tabular}
-}
-\end{minipage}& \begin{minipage}[c]{0.6\linewidth}
+\end{center}
 
-We have randomly drawn protected classes, $X=\{t_1,t_2\}$ and $Y=\{t_3,t_4\}$, and two five-element attribute sets $A$ and $B$.
-One simulation draws normally distributed values  for two situations in which the underlying mean similarities are in fact equal to zero, in two other cases  the means are different, with a range of choices of common standard deviation  accross the cases (code available upon request).
+To illustrate, we have randomly drawn protected classes, $X=\{t_1,t_2\}$ and $Y=\{t_3,t_4\}$, and two five-element attribute sets $A$ and $B$.
+One simulation draws normally distributed values  for two situations in which the underlying mean similarities are in fact equal to zero, in two other cases  the means are different, with a range of choices of common standard deviation  across the cases (see the table).
+The following observations are worthwhile.
+For points randomly drawn from distributions where there is no difference in means the calculated effect size can easily be  1.82, whereas the largest effect size reported by by Caliskan is 1.81.
+For samples from distributions where the means are different, the (absolute) effect sizes can easily be lower than in the first two simulations.
+Moreover, 10k simulations with more realistic word list sizes (normal distributions with equal means, standard deviation=0.08, 16 protected words and 16 predicates) show that high scores are easy to get even if there is no difference in the generating model (Figure 2, code available upon request).
+This indicates that the low p-value reported in the papers are highly misleading.
 
-\end{minipage}
-\end{tabular}
-
-\vspace{1mm}
-
-The following observations are worthwhile.
-(1) For points randomly drawn from distributions where there is no difference in means the calculated effect size can easily be  1.82, whereas the largest effect size reported by @Caliskan2017semanticsBiases is 1.81.
-(2) For samples from distributions where the means are different, the (absolute) effect sizes can easily be lower than in the first two simulations.
-As Figure 2.
-illustrates,  quite some uncertainty is involved, far more than what systematically low mean-based p-values reported in the papers might suggest.
-Part of the problem is random variation unaccounted for in the original approach (see Figures 3 and 4 for an example), and part of the problem is that    non-negligible changes in effect size can result from a shift in the standard deviation of the original process, because with the decrease of standard deviation the numerator in (WEAT) decreases.
-
-To improve on the situation, we  build  Bayesian models to estimate the biases involved using the raw datapoints, actually using control groups,  distinguishing the connection types,  and taking the uncertainty involved seriously.
-For the general impact of being associated, we build  models using hamiltonian Monte Carlo (STAN) according to the following specification (linear categorical model and the priors):
+To improve on the situation, we  build  Bayesian models to estimate the biases involved using the raw data points, actually using control groups,  distinguishing the connection types,  and taking the uncertainty involved seriously.
+For the general impact of being associated, we build  models using Hamiltonian Monte Carlo (STAN) according to the following specification (linear categorical model and the priors):
 
@@
 
-\begin{center}
-\begin{figure}[!htb]\centering
-   \begin{minipage}{0.48\textwidth}
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
-ggplot(religion, aes(x =  cosineDistance, fill = connection, color = connection))+geom_density(alpha=0.6,size = .2)+theme_tufte()+labs(title  = "Figure 1.", subtitle = "Empirical distribution of cosine distances (religion, Reddit)")+ scale_fill_manual(values = c("orangered4","chartreuse4", "skyblue", "gray"))+scale_x_continuous(breaks = seq(0.3,1.5, by = 0.1))+xlab("cosine distance")+ scale_color_manual(values = c("orangered4","chartreuse4","skyblue","gray"))
-```
-   \end{minipage}
-   \begin {minipage}{0.48\textwidth}
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
-s <- function (table){ mean(table$A) - mean(table$B)}
-biasesD2 <- numeric(10000)
-for(i in 1:10000){
-t1d2 <- data.frame(A  = rnorm(5,.1,0.05), B = rnorm(5,0,0.05))
-t2d2 <- data.frame(A  = rnorm(5,.1,0.05), B = rnorm(5,0,0.05))
-t3d2 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,.1,0.05))
-t4d2 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,.1,0.05))
+\includegraphics[width=8cm]{violinReligion.pdf}
+\includegraphics[width=8cm]{simulations.pdf}
 
-factorD2 <- sd(c(s(t1d2),s(t2d2),s(t3d2),s(t4d2)))
-numeratorD2 <-  mean(s(t1d2),s(t2d2)) - mean(s(t3d2),s(t4d2))
-biasesD2[i] <- numeratorD2/factorD2
-}
+\includegraphics[width=8cm]{muslim.pdf}
+\includegraphics[width=8cm]{priest.pdf}
 
-ggplot()+geom_histogram(aes(x=biasesD2, y = ..density..), alpha = 0.6, bins=50)+
-  theme_tufte()+labs(title="Figure 2.", subtitle = "10k biases for different means  and sd =.05 (case 3 in the table)")+ xlab("bias")
-```
-   \end{minipage}
+\includegraphics[width=8cm]{../images/religionCoeffs.jpeg}
+\includegraphics[width=8cm]{../images/visReligionReddit.png}
 
-  \begin{minipage}{0.48\textwidth}
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%",message =FALSE, warning = FALSE}
-library(tidyverse)
-muslimWords <- c("imam","islam","mosque","muslim","quran")
-muslim <- religion %>% filter(protectedWord %in% muslimWords)
-muslimClass <- muslim %>% filter(protectedWord == "muslim")
-neutralSample <- sample_n(filter(muslimClass,connection == "none"), 5)
-humanSample <- sample_n(filter(muslimClass,connection == "human"), 5)
-muslimVis <- muslimClass %>% filter(connection != "none" & connection !="human")
-muslimVis <- rbind(muslimVis,neutralSample,humanSample)
-#we plug in our visualisation script
-source("../functions/visualisationTools.R")
-#two arguments: dataset and protected word
-visualiseProtected(muslimVis,"muslim")+labs(title = "Figure 3.",
-subtitle = "Cosine distances for active attributes (musim, Reddit)")
-```
-\end{minipage}
-   \begin {minipage}{0.48\textwidth}
-```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
-library(tidyverse)
-priestClass <- religion %>% filter(protectedWord == "priest")
-neutralSample <- sample_n(filter(priestClass,connection == "none"), 5)
-humanSample <- sample_n(filter(priestClass,connection == "human"), 5)
-priestVis <- priestClass %>% filter(connection != "none" & connection !="human")
-priestVis <- rbind(priestVis,neutralSample,humanSample)
+<!-- ```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"} -->
+<!-- ggplot(religion, aes(x =  cosineDistance, fill = connection, color = connection))+geom_density(alpha=0.6,size = .2)+theme_tufte()+labs(title  = "Figure 1.", subtitle = "Empirical distribution of cosine distances (religion, Reddit)")+ scale_fill_manual(values = c("orangered4","chartreuse4", "skyblue", "gray"))+scale_x_continuous(breaks = seq(0.3,1.5, by = 0.1))+xlab("cosine distance")+ scale_color_manual(values = c("orangered4","chartreuse4","skyblue","gray")) -->
+<!-- ``` -->
 
-#we plug in our visualisation script
-source("../functions/visualisationTools.R")
-#two arguments: dataset and protected word
-visualiseProtected(priestVis,"priest")+labs(title = "Figure 4.",
-subtitle = "Cosine distances for active attributes (priest, Reddit)")
-```
-   \end{minipage}
+<!-- ```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"} -->
+<!-- s <- function (table){ mean(table$A) - mean(table$B)} -->
+<!-- biasesD2 <- numeric(10000) -->
+<!-- for(i in 1:10000){ -->
+<!-- t1d2 <- data.frame(A  = rnorm(5,.1,0.05), B = rnorm(5,0,0.05)) -->
+<!-- t2d2 <- data.frame(A  = rnorm(5,.1,0.05), B = rnorm(5,0,0.05)) -->
+<!-- t3d2 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,.1,0.05)) -->
+<!-- t4d2 <- data.frame(A  = rnorm(5,0,0.05), B = rnorm(5,.1,0.05)) -->
 
-  \begin{minipage}{0.38\textwidth}
-\includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
-\end{minipage}
-   \begin {minipage}{0.58\textwidth}
-\includegraphics{../images/visReligionReddit.png}
-   \end{minipage}
-\end{figure}
+<!-- factorD2 <- sd(c(s(t1d2),s(t2d2),s(t3d2),s(t4d2))) -->
+<!-- numeratorD2 <-  mean(s(t1d2),s(t2d2)) - mean(s(t3d2),s(t4d2)) -->
+<!-- biasesD2[i] <- numeratorD2/factorD2 -->
+<!-- } -->
 
-\end{center}
+<!-- ggplot()+geom_histogram(aes(x=biasesD2, y = ..density..), alpha = 0.6, bins=50)+ -->
+<!--   theme_tufte()+labs(title="Figure 2.", subtitle = "10k biases for different means  and sd =.05 (case 3 in the table)")+ xlab("bias") -->
+<!-- ``` -->
 
+<!-- # ```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%",message =FALSE, warning = FALSE} -->
+<!-- # library(tidyverse) -->
+<!-- # muslimWords <- c("imam","islam","mosque","muslim","quran") -->
+<!-- # muslim <- religion %>% filter(protectedWord %in% muslimWords) -->
+<!-- # muslimClass <- muslim %>% filter(protectedWord == "muslim") -->
+<!-- # neutralSample <- sample_n(filter(muslimClass,connection == "none"), 5) -->
+<!-- # humanSample <- sample_n(filter(muslimClass,connection == "human"), 5) -->
+<!-- # muslimVis <- muslimClass %>% filter(connection != "none" & connection !="human") -->
+<!-- # muslimVis <- rbind(muslimVis,neutralSample,humanSample) -->
+<!-- # #we plug in our visualisation script -->
+<!-- # source("../functions/visualisationTools.R") -->
+<!-- # #two arguments: dataset and protected word -->
+<!-- # visualiseProtected(muslimVis,"muslim")+labs(title = "Figure 3.", -->
+<!-- # subtitle = "Cosine distances for active attributes (musim, Reddit)") -->
+<!-- # ``` -->
+
+ <!-- ```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"} -->
+<!-- library(tidyverse) -->
+<!-- priestClass <- religion %>% filter(protectedWord == "priest") -->
+<!-- neutralSample <- sample_n(filter(priestClass,connection == "none"), 5) -->
+<!-- humanSample <- sample_n(filter(priestClass,connection == "human"), 5) -->
+<!-- priestVis <- priestClass %>% filter(connection != "none" & connection !="human") -->
+<!-- priestVis <- rbind(priestVis,neutralSample,humanSample) -->
+
+<!-- #we plug in our visualisation script -->
+<!-- source("../functions/visualisationTools.R") -->
+<!-- #two arguments: dataset and protected word -->
+<!-- visualiseProtected(priestVis,"priest")+labs(title = "Figure 4.", -->
+<!-- subtitle = "Cosine distances for active attributes (priest, Reddit)") -->
+<!-- ``` -->
+
+\pagebreak
+## References {-}
 \scriptsize
-
-\vspace{-4mm}

### 2022-04-01 abstractESSLLI21/abstractESSLLI1.tex
Adds bitly

@@
 \usepackage{multicol}
-
+\usepackage{subcaption}
 \usepackage{graphicx}
@@
 \usepackage{geometry}
- \geometry{a4paper,left=20mm,top=15mm,bottom = 20mm, right = 20mm}
+ \geometry{a4paper,left=35mm,top=20mm,bottom = 20mm, right = 35mm}
 
@@
 A considerable amount of literature exists on bias detection and
-mitigation in NLP models, especially word2vec embeddings, which
-represent words as vectors of real numbers (see e.g. {[}1{]} and {[}4{]}
-and references therein). The most common method used compares cosine
-similarity between words from protected groups and attributes that are
-considered to be stereotypical or harmful in some way, and this method
-will be in our focus.
+mitigation in NLP models, especially word embeddings, which represent
+words as vectors of real numbers (see e.g. {[}1{]} and {[}4{]} and
+references therein). The most common methods used compare cosine
+similarity or distance between words from protected groups and
+attributes that are considered to be stereotypical or harmful in some
+way. Many of them rely on pre-averaging, which we find problematic ---
+two such methods will be in our focus. We will argue that the existing
+measures of bias by pre-averaging oversimplify the picture and lead to
+false confidence. Our simulation results show that the mainstream
+interpretations are too hasty. Our Bayesian analysis with proper control
+groups suggests that (1) the word lists in use are too short, (2)
+associated and different groups behave similarly, which undermines the
+semantic interpretation of cosine similarity, and (3) our analysis at
+different levels of generality shows that there is more variety than one
+may initially think.
 
@@
 } \tag{WEAT}
-\end{align}
-
-\normalsize
-\noindent {[}2{]} show that significant biases---thus measured---
-similar to the ones discovered by IAT can be discovered in word
-embeddings. {[}5{]} extend the methodology to a multilingual and
-cross-lingual setting; a similar methodology is employed by {[}3{]}, who
-use word embeddings trained on corpora from different decades. {[}6{]}
-modify WEAT to a multi-class setting, introducing Mean Average Cosine
-(MAC) similarity as a measure of bias (in fact, in the paper they report
-distances rather than similarities). Let \(T = \{t_1, \dots, t_k\}\) be
-a class of protected word embeddings, and let each \(A_j\in A\) be a set
-of attributes stereotypically associated with a protected word). Then:
+\end{align} \normalsize \noindent {[}2{]} show that significant
+biases---thus measured--- similar to the ones discovered by IAT can be
+discovered in word embeddings. {[}5{]} extend the methodology to a
+multilingual and cross-lingual setting; a similar methodology is
+employed by {[}3{]}, who use word embeddings trained on corpora from
+different decades. {[}6{]} modify WEAT to a multi-class setting,
+introducing Mean Average Cosine (MAC) similarity distance as a measure
+of bias. Let \(T = \{t_1, \dots, t_k\}\) be a class of protected word
+embeddings, and let each \(A_j\in A\) be a set of attributes
+stereotypically associated with a protected word). Then:
 
@@
 
-\noindent
-
-\begin{tabular}{ll}
-\begin{minipage}[c]{0.35\linewidth}
-\scalebox{0.65}{
+\begin{center}
 \begin{tabular}{cccccc}\toprule
@@
 \end{tabular}
-}
-\end{minipage}& \begin{minipage}[c]{0.6\linewidth}
+\end{center}
 
-We have randomly drawn protected classes, $X=\{t_1,t_2\}$ and $Y=\{t_3,t_4\}$, and two five-element attribute sets $A$ and $B$.
-One simulation draws normally distributed values  for two situations in which the underlying mean similarities are in fact equal to zero, in two other cases  the means are different, with a range of choices of common standard deviation  accross the cases (code available upon request).
-
-\end{minipage}
-\end{tabular}
-
-\vspace{1mm}
-
-The following observations are worthwhile. (1) For points randomly drawn
-from distributions where there is no difference in means the calculated
-effect size can easily be 1.82, whereas the largest effect size reported
-by {[}2{]} is 1.81. (2) For samples from distributions where the means
-are different, the (absolute) effect sizes can easily be lower than in
-the first two simulations. As Figure 2. illustrates, quite some
-uncertainty is involved, far more than what systematically low
-mean-based p-values reported in the papers might suggest. Part of the
-problem is random variation unaccounted for in the original approach
-(see Figures 3 and 4 for an example), and part of the problem is that
-non-negligible changes in effect size can result from a shift in the
-standard deviation of the original process, because with the decrease of
-standard deviation the numerator in (WEAT) decreases.
+To illustrate, we have randomly drawn protected classes,
+\(X=\{t_1,t_2\}\) and \(Y=\{t_3,t_4\}\), and two five-element attribute
+sets \(A\) and \(B\). One simulation draws normally distributed values
+for two situations in which the underlying mean similarities are in fact
+equal to zero, in two other cases the means are different, with a range
+of choices of common standard deviation across the cases (see the
+table). The following observations are worthwhile. For points randomly
+drawn from distributions where there is no difference in means the
+calculated effect size can easily be 1.82, whereas the largest effect
+size reported by by Caliskan is 1.81. For samples from distributions
+where the means are different, the (absolute) effect sizes can easily be
+lower than in the first two simulations. Moreover, 10k simulations with
+more realistic word list sizes (normal distributions with equal means,
+standard deviation=0.08, 16 protected words and 16 predicates) show that
+high scores are easy to get even if there is no difference in the
+generating model (Figure 2, code available upon request). This indicates
+that the low p-value reported in the papers are highly misleading.
 
 To improve on the situation, we build Bayesian models to estimate the
-biases involved using the raw datapoints, actually using control groups,
-distinguishing the connection types, and taking the uncertainty involved
-seriously. For the general impact of being associated, we build models
-using hamiltonian Monte Carlo (STAN) according to the following
+biases involved using the raw data points, actually using control
+groups, distinguishing the connection types, and taking the uncertainty
+involved seriously. For the general impact of being associated, we build
+models using Hamiltonian Monte Carlo (STAN) according to the following
 specification (linear categorical model and the priors):
@@
 
-\begin{center}
-\begin{figure}[!htb]\centering
-   \begin{minipage}{0.48\textwidth}
+\includegraphics[width=8cm]{violinReligion.pdf}
+\includegraphics[width=8cm]{simulations.pdf}
 
-\begin{center}\includegraphics[width=1\linewidth]{abstractESSLLI1_files/figure-latex/unnamed-chunk-1-1} \end{center}
-   \end{minipage}
-   \begin {minipage}{0.48\textwidth}
+\includegraphics[width=8cm]{muslim.pdf}
+\includegraphics[width=8cm]{priest.pdf}
 
-\begin{center}\includegraphics[width=1\linewidth]{abstractESSLLI1_files/figure-latex/unnamed-chunk-2-1} \end{center}
-   \end{minipage}
+\includegraphics[width=8cm]{../images/religionCoeffs.jpeg}
+\includegraphics[width=8cm]{../images/visReligionReddit.png}
 
-  \begin{minipage}{0.48\textwidth}
+\pagebreak
 
-\begin{center}\includegraphics[width=1\linewidth]{abstractESSLLI1_files/figure-latex/unnamed-chunk-3-1} \end{center}
-\end{minipage}
-   \begin {minipage}{0.48\textwidth}
-
-\begin{center}\includegraphics[width=1\linewidth]{abstractESSLLI1_files/figure-latex/unnamed-chunk-4-1} \end{center}
-   \end{minipage}
-
-  \begin{minipage}{0.38\textwidth}
-\includegraphics[width=7cm]{../images/religionCoeffs.jpeg}
-\end{minipage}
-   \begin {minipage}{0.58\textwidth}
-\includegraphics{../images/visReligionReddit.png}
-   \end{minipage}
-\end{figure}
-
-\end{center}
+\hypertarget{references}{%
+\subsection*{References}\label{references}}
+\addcontentsline{toc}{subsection}{References}
 
@@
 
-\vspace{-4mm}
-
 \hypertarget{refs}{}
 \begin{CSLReferences}{0}{0}
-\leavevmode\hypertarget{ref-bolukbasi2016man}{}%
+\leavevmode\hypertarget{ref-Bolukbasi2016man}{}%
 \CSLLeftMargin{{[}1{]} }
-\CSLRightInline{Tolga Bolukbasi, Kai-Wei Chang, James Zou, Venkatesh
+\CSLRightInline{Tolga Bolukbasi, Kai-Wei Chang, James Y. Zou, Venkatesh
 Saligrama, and Adam Kalai. 2016. Man is to computer programmer as woman
-is to homemaker? Debiasing word embeddings. Retrieved from
-\url{http://arxiv.org/abs/1607.06520}}
+is to homemaker? Debiasing word embeddings. \emph{CoRR} abs/1607.06520,
+(2016). Retrieved from \url{http://arxiv.org/abs/1607.06520}}
 

### 2022-04-25 paperDraft/outdated/paperDraft3.Rmd
added model for jobs and cleaned the paperDraft folder.

--- 2021-12-10 paperDraft/paperDraft2.Rmd
@@
     df_print: paged
-classoption: dvipsnames,enabledeprecatedfontcommands
+classoption: dvipsnames,enabledeprecatedfontcommands, twocolumn
 fontsize: 10pt

### 2022-04-25 modelBuilding/jobsModel.R

@@ -1,0 +1,45 @@
+library(tidyr)
+library(rethinking)
+library(ggthemes)
+
+genderProjections <-  read.csv("datasets/genderProjections.csv")[,-1]
+head(genderProjections)
+colnames(genderProjections)[4] <- "stereotype" 
+head(genderProjections)
+levels(genderProjections$stereotype) <- c("male", "female")
+
+ggplot(genderProjections, aes(x=he,y=she, color = stereotype, shape = stereotype))+geom_point(size = 6, alpha = 0.75)+theme_tufte(base_size = 27)+labs(title ="GloVe  on Wikipedia 2014 and Gigaword 5th ed.")+scale_color_manual(values =c("orangered4","chartreuse4"))+ 
+  theme(legend.position = c(0.8,.15)) +
+  scale_shape_manual(values=c(15, 19)) 
+
+genderLong <- gather(genderProjections, gender, similarity, c("he", "she"),factor_key=TRUE)
+genderLong$stereotype <- as.factor(genderLong$stereotype)
+genderLong$stereotypeID <- as.integer(genderLong$stereotype)
+genderLong$genderID <- as.integer(genderLong$gender)
+
+genderLong$treatmentID <- (genderLong$stereotypeID == genderLong$genderID)+1
+#note:
+#1: opposite
+#2: same
+
+
+model <- quap(
+  alist(
+    similarity ~ dnorm(mu,sigma),
+    mu <- m[treatmentID],
+    m[treatmentID] ~ dnorm(10,5),
+    sigma ~ dexp(1)
+  ),
+  data = genderLong
+)
+
+#postcheck(model)
+
+
+precis(model, depth = 2, pars = "m")
+
+plot(precis(model, depth = 2, pars="m"), labels = c("opposite", "same") )
+
+
+
+

### 2022-04-27 paperDraft/metaTechnicalDetails.Rmd
Commits old fiels

---
title: ""
author: ""
output:
  pdf_document:
    number_sections: true
    df_print: kable 
    keep_tex: true
    includes:
      in_header:
        - Rafal_latex6.sty
classoption: dvipsnames,enabledeprecatedfontcommands
fontsize: 12pt
documentclass: scrartcl
urlcolor: blue
csl: transactions-on-computational-logic.csl
---

# Recreating Code for MAC Manzini

Word embeddings sources:
\newline

- the same as the authors - Reddit-L2 corpus (reddit.l2.tar.gz) from http://cl.haifa.ac.il/projects/L2/ 
\newline

- our idea - GoogleNews corpus (GoogleNews-vectors-negative300.bin.gz.) from  https://code.google.com/archive/p/word2vec/

Getting cosine similarity and cosine distance:
\newline

Description of the precedure:

TODO: Run code with fresh downloaded corpora

# Adding control groups

## Human words assumptions and procedure of creation

TODO

## Neutral words assumptions and procedure of creation

TODO

### 2022-06-22 paperDraft/paperDraft4.tex

@@
 \PassOptionsToPackage{hyphens}{url}
-\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
+\PassOptionsToPackage{dvipsnames,svgnames*,x11names*}{xcolor}
 %
@@
   dvipsnames,enabledeprecatedfontcommands, twocolumn]{scrartcl}
-\title{Taking uncertainty in word embedding bias estimation seriously: a
-bayesian approach}
-\author{Alicja Dobrzeniecka and Rafal Urbaniak}
-\date{}
-
 \usepackage{amsmath,amssymb}
 \usepackage{lmodern}
-\usepackage{iftex}
-\ifPDFTeX
+\usepackage{ifxetex,ifluatex}
+\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
   \usepackage[T1]{fontenc}
@@
   colorlinks=true,
-  linkcolor={Maroon},
-  filecolor={Maroon},
-  citecolor={Blue},
-  urlcolor={blue},
+  linkcolor=Maroon,
+  filecolor=Maroon,
+  citecolor=Blue,
+  urlcolor=blue,
   pdfcreator={LaTeX via pandoc}}
@@
 \setcounter{secnumdepth}{5}
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
-  \fi
-  % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 %\documentclass{article}
@@
 
-\ifLuaTeX
+\ifluatex
   \usepackage{selnolig}  % disable illegal ligatures
 \fi
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
+  % set entry spacing
+  \ifnum #2 > 0
+  \setlength{\parskip}{#2\baselineskip}
+  \fi
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+
+\title{Taking uncertainty in word embedding bias estimation seriously: a
+bayesian approach}
+\author{Alicja Dobrzeniecka and Rafal Urbaniak}
+\date{}
 
@@
 
-\raf{fix bibliography}
+\hypertarget{what-is-bias}{%
+\subsection{What is bias?}\label{what-is-bias}}
 
-\ali{proofread and think if we should add more}
+\hypertarget{word-embeddings}{%
+\subsection{Word embeddings}\label{word-embeddings}}
 
@@
 focus of our paper.
+
+\hypertarget{semantical-interpretation}{%
+\subsection{Semantical interpretation}\label{semantical-interpretation}}
+
+\hypertarget{general-challenges}{%
+\subsection{General challenges}\label{general-challenges}}
+
+\hypertarget{weat-and-mac-measures}{%
+\subsection{WEAT and MAC measures}\label{weat-and-mac-measures}}
 
@@
   social media-text trained embeddings being less biased than those
-  based on Wikipedia).}\(^{, }\)\footnote{ A similar methodology is
+  based on Wikipedia).}\(^{, }\) \^{}{[} A similar methodology is
 employed in {[}3{]}. The authors employ word embeddings trained on
 corpora from different decades to study the shifts in various biases.
-  For instance, to compute the occupational embeddings bias for women
-  the authors first compute the average vector of vector embeddings of
-  words that represent women (e.g.~\emph{she}, and \emph{female}), then
-  calculate the Euclidean distance between this mean vector and words
-  for occupations. Then they take the mean of these distances and
-  subtract from it the analogously obtained mean for the average vector
-  of vector embeddings of words that represent men. Formally they take
-  the relative norm distance between \(X\) and \(Y\) to be:}
+For instance, to compute the occupational embeddings bias for women the
+authors first compute the average vector of vector embeddings of words
+that represent women (e.g.~\emph{she}, and \emph{female}), then
+calculate the Euclidean distance between this mean vector and words for
+occupations. Then they take the mean of these distances and subtract
+from it the analogously obtained mean for the average vector of vector
+embeddings of words that represent men. Formally they take the relative
+norm distance between \(X\) and \(Y\) to be:
+
+\vspace{-2mm}
+
+\footnotesize
+
+\begin{align}
+\textsf{relative norm distance} & = \sum_{v_m\in M} \vert \vert v_m - v_X\vert \vert_2 - \vert v_m - v_Y\vert \vert_2
+\end{align}
+
+\normalsize
+
+\noindent where the norm used is Euclidean, and \(v_X\) and \(x_Y\) are
+average vectors for sets \(X\) and \(Y\) respectively. {]}
 
@@
 \hypertarget{non-statistical-problems-with-cosine-based-measures-of-bias}{%
-\section{Non-statistical problems with cosine-based measures of
+\subsection{Non-statistical problems with cosine-based measures of
 bias}\label{non-statistical-problems-with-cosine-based-measures-of-bias}}
@@
 \hypertarget{statistical-issues}{%
-\section{Statistical issues}\label{statistical-issues}}
+\subsection{Statistical issues}\label{statistical-issues}}
 
@@
 
+\hypertarget{pre-averaging-and-manufactured-certainty}{%
+\section{Pre-averaging and manufactured
+certainty}\label{pre-averaging-and-manufactured-certainty}}
+
+\hypertarget{general-problem-with-pre-averaging}{%
+\subsection{General problem with
+pre-averaging}\label{general-problem-with-pre-averaging}}
+
+\hypertarget{simulations-for-weat}{%
+\subsection{Simulations for WEAT}\label{simulations-for-weat}}
+
+\hypertarget{simulations-for-mac}{%
+\subsection{Simulations for MAC}\label{simulations-for-mac}}
+
+\hypertarget{bayesian-method}{%
+\section{Bayesian method}\label{bayesian-method}}
+
+\hypertarget{bernstein-approach}{%
+\subsection{Bernstein approach}\label{bernstein-approach}}
+
+\hypertarget{chasing-metrics}{%
+\subsection{Chasing metrics}\label{chasing-metrics}}
+
+\hypertarget{bayesian-method-introduction}{%
+\subsection{Bayesian method
+introduction}\label{bayesian-method-introduction}}
+
+\hypertarget{existing-applications-to-nlp-and-perhaps-to-bias}{%
+\subsection{Existing applications to NLP and perhaps to
+bias}\label{existing-applications-to-nlp-and-perhaps-to-bias}}
+
+\hypertarget{model}{%
+\subsection{Model}\label{model}}
+
+\hypertarget{results}{%
+\section{Results}\label{results}}
+
+\hypertarget{weat}{%
+\subsection{WEAT}\label{weat}}
+
+\hypertarget{mac}{%
+\subsection{MAC}\label{mac}}
+
+\hypertarget{discussion-and-summary}{%
+\section{Discussion and summary}\label{discussion-and-summary}}
+
+\hypertarget{appendix}{%
+\section{APPENDIX}\label{appendix}}
+
+\hypertarget{examples-of-weat-and-mac-calculations}{%
+\subsection{Examples of WEAT and MAC
+calculations}\label{examples-of-weat-and-mac-calculations}}
+
+\hypertarget{word-lists}{%
+\subsection*{Word lists}\label{word-lists}}
+\addcontentsline{toc}{subsection}{Word lists}
+
 \hypertarget{refs}{}
 \begin{CSLReferences}{0}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-Bolukbasi2016man}{}}%
+\leavevmode\hypertarget{ref-Bolukbasi2016man}{}%
 \CSLLeftMargin{{[}1{]} }
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Caliskan2017semanticsBiases}{}}%
+\leavevmode\hypertarget{ref-Caliskan2017semanticsBiases}{}%
 \CSLLeftMargin{{[}2{]} }
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Garg2018years}{}}%
+\leavevmode\hypertarget{ref-Garg2018years}{}%
 \CSLLeftMargin{{[}3{]} }
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Gonen2019lipstick}{}}%
+\leavevmode\hypertarget{ref-Gonen2019lipstick}{}%
 \CSLLeftMargin{{[}4{]} }
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Lauscher2019multidimensional}{}}%
+\leavevmode\hypertarget{ref-Lauscher2019multidimensional}{}%
 \CSLLeftMargin{{[}5{]} }
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Manzini2019blackToCriminal}{}}%
+\leavevmode\hypertarget{ref-Manzini2019blackToCriminal}{}%
 \CSLLeftMargin{{[}6{]} }
@@
 and removing multiclass bias in word embeddings. Retrieved from
-\url{https://arxiv.org/abs/1904.04047}}
+\url{http://arxiv.org/abs/1904.04047}}
 
-\leavevmode\vadjust pre{\hypertarget{ref-Nissim2020fair}{}}%
+\leavevmode\hypertarget{ref-Nissim2020fair}{}%
 \CSLLeftMargin{{[}7{]} }
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Nosek2002harvesting}{}}%
+\leavevmode\hypertarget{ref-Nosek2002harvesting}{}%
 \CSLLeftMargin{{[}8{]} }

### 2022-07-22 paperDraft/notes/bayesianMethod.tex
bayesian appendix started

\appendix
\section{Appendix: Why Bayesian methods?}

As referees found it useful to include an explanation of why classical statistical methods have not been used in the paper, here is a brief rationale.

Suppose you are interested in how many referees of a certain class of journals prefer Bayesian statistical methods over the classical approach (call this property $\beta$). You randomly draw 24 such referees  and it turns out 7 of them actually do. Let us work with the assumption that the probability of $\beta$ is the same for each, and that whether a referee has property $\beta$ is independent of whether other referees do. What should you make of this?

  First, let us see what can be made of this from the classical perspective of the null hypothesis significance testing (NHST) paradigm.

First, prior to looking at the data you need to formulate a null hypothesis. Say yours is that the probability that a referee will prefer Bayesian methods, $\theta$, is $.5$. Then you expect about half of the sample---12---to have $\beta$. If the number is far greater or smaller than that, you are supposed to reject the null. Specifically, in our case we need to calculate the probability that the number is as extreme or more extreme than 7 on the assumption that $\theta =.5$, called the $p$-value, and reject the null just in case this probability is less than a certain low value, say $5\%$.

Crucially, the outcome depends on defining a space of all possible outcomes. For instance, if the intention of the experiment was to draw 24 referees and count the $\beta$s, the space contains all binary sequences of length 24. Then, we use the binomial probability distribution on which the probability of obtaining $z$ $\beta$s out of $N$ observations given the null fixed at $\theta$ is:
  \begin{align*}
\mathsf{P}(z\vert N, \theta) & = {N\choose  z} \theta^z (1  - \theta)^{N-z}
\end{align*}
\noindent With $\theta_{null}=.5$ the binomial distribution is symmetric so using a two-tailed test with rejection threshold at $5\%$ we can reject the null if the probability of at most seven $\beta$s is not greater than $2.5\%$. In our case, $\mathsf{P}(7\vert 24, \theta =.5)\approx.32$  so, from the perspective of NHST we fail to reject the null and we have not learned much.

Note however that we could have obtained the same data with a different intention: that of sampling until we observe 7 $\beta$s. In that case, the space of all possible outcomes contains binary sequences of varying lengths whose last element is a 7th success. (Other sampling intentions are also possible). The probability of getting a sequence of length 24 with the intention to stop at 7 is negative binomial:
  \begin{align*}
\mathsf{P}(N\vert z, \theta) & = {N- 1 \choose z - 1}\theta^{z-1}(1-\theta)^{N - z}\theta
\end{align*}
\noindent with this stopping intention, the $p$-value is $.017$, which is less than $2.5\%$, so we are supposed to reject the null.

This shows there is no such a thing as the unique $p$-value for a data set even if the null hypothesis is kept fixed.

Moreover, the $p$-value for any particular comparison depends on whether we intended to make multiple comparisons. Suppose we in fact want to study two groups of referees (say, from two different groups of journals). As in NHST we want to monitor the overall false alarm rate, we are supposed to focus on the probability of a false alarm in at least one group. If the null hypotheses and sample sizes are the same, and we use the binomial distribution, then the $p$ value for the first group even without making any observation from the second group is now $.063$. Let this sink in: what we should do with the null hypothesis for the first group depends on imaginary possibilities related to the second group.

In light of the above problems, and other issues with $p$-values (roughly speaking, related to the fact that focusing on $p$-values is partially responsible for the replicability crisis), some recommend using estimation and classical confidence intervals (CIs) instead.

This does not help much for various reasons.  One mathematically correct conceptualization of CIs is that they are ranges of parameter values that would not be rejected by the given data set. The $95\%$ CI consists of values of $\theta$ that would not be rejected by a two-tailed significance test with false alarm rate $\alpha = 5\%$. For this reason, CIs depend on whatever $p$-values depend on, and so still depend on stopping intention. For instance, if we use the binomial distribution in the referees example, the CI is $[.126,.511]$. If we use the negative binomial distribution it is $[.123,.484]$, and for the two-group study, it is $[.110, .539]$. Moreover, CIs are notoriously susceptible to misinterpretation. People tend to mistakenly:

  \begin{itemize}
\item   think that CI indicates some probability distribution over $\theta$, which it does not. Values in the middle of a CI are not more believable than the values at its limits,

\item   think that  if the probability that a random interval contains the true value is $x\%$, then the plausibility or probability that a particular observed interval contains the true value is also $x\%$.

\item    interpret confidence intervals as indicating the precision of the estimate,

\item  think that values inside the confidence interval are more likely than those outside.

\end{itemize}

\noindent The literature on misinterpretations of CIs is vast and we encourage the interested reader to search the web for papers on this topic.

Another problem is that NHST pays no attention to prior knowledge. On this approach what we should do when we inspect referees' preferences is exactly the same as what we should do if we were flipping long nails defining success as the nail landing balanced on its head with its pointy tail sticking up, even though there are clearly more reasons to think the outcome of 7 successes with nails is more of a sheer luck, not an evidence that we should not reject the null.

From the perspective of Bayesian analysis there is no intention dependence. All that the data contributes is the likelihood, the product of the probabilities of the individual outcomes. The likelihood is:
    $\prod_{i = 1}^{N} \theta^z(1- \theta)^{N-z}$. It is then combined with the prior probability to obtain a posterior probability mass function describing how probable particular values of $\theta$ are. For instance, if you start completely uninformed with $\mathsf{beta}(1,1)$, the posterior distribution is $\mathsf{beta}(8,18)$. Given that most referees are senior scholars and that Bayesian methods made its strides in college education only fairly recently, the case is more like that of nails, and we should not start with a completely uninformed prior. If you implement this knowledge starting rather with $\mathsf{beta}(2, 3)$, the posterior density is $\mathsf{beta}(9,20)$.
% \begin{figure}[H]
\begin{center}
\includegraphics[width=1\linewidth]{imagesFinal/bayesianDensities} 
\end{center}
% \caption{Two potential priors and posteriors in a Bayesian analysis of the referees' data.}
% \label{fig:referees}
% \end{figure}

In contrast with CIs, the posterior distributions (and their highest posterior density intervals HPDIs, the narrowest intervals containing a certain ratio of the area under the curve) are easily interpretable and have direct relevance for the question at hand. Given the slightly informed prior, for instance, in our case the posterior density over $\theta$ in the referees example has its 95\% HPDI at  $[.16, .48]$ and it is the case that the real $\theta$ with probability $95\%$ belongs to this interval.

The reader might be worried that priors are subjective and mysterious. But usually, they are much less so than the intricacies and theoretical choices made in classical analysis.  Prior beliefs can be explicitly stated, debated, or often found in previous research. In particular since in our research we intend to convince someone mildly skeptical of the results (and want to avoid model  over-fitting), it makes sense to start with priors slightly centered around 0, but easily admitting reasonable ranges of impact, and this is what we have done.

Interestingly, Bayesian analysis is not susceptible to all the difficulties with multiple comparisons that plague NHST. What is estimated is a single posterior density, even if it involves multiple predictors with multiple levels. For this reason, with Bayesian methods one can look at the data as much as one wants, stopping whenever one wants and updating the distribution with new evidence whenever one wants. In the  particular case of our study, as we look at quite a few different groups, trying to implement ANOVA tests with post-hoc comparison would lead to a disastrous maze of statistical tests and nearly meaningless multiple-comparison-corrected $p$-values (or CI intervals). This has been avoided by using Bayesian methods.

### 2022-07-25 paperDraft/paperDraft4.Rmd
Updates paper structure
small structure changes

@@
 
-\raf{fix bibliography}
+## Word embeddings and their bias
 
-\ali{proofread and think if we should add more}
+- akapit filozoficzny, z odniesieniem do value free
 
@@
 This methodology will be in the focus of our paper.
+
+- undesirable bias with respect to a certain groups words if stereotypically connected words (for stereotypes that we don't want to be used or relied on in downstream tasks) are systematically closer to each other.
++ EXAMPLE
+
+## Semantical interpretation
+
+## General challenges
+
+## WEAT and MAC measures
 
@@
 
-# Non-statistical problems with cosine-based measures of bias
+## Non-statistical problems with cosine-based measures of bias
 
@@
 
-# Statistical issues
+## Statistical issues
 
@@
 The t-tests they employ are run on average cosines used to calculate MAC.
+
+# Pre-averaging and manufactured certainty
+
+## General problem with pre-averaging
+
+## Simulations for WEAT
+
+## Simulations for MAC
+
+# Bayesian method
+
+## Bernstein approach
+
+## Chasing metrics
+
+## Bayesian method introduction
+
+## Existing applications to NLP and perhaps to bias
+
+## Model
+
+# Results
+
+## WEAT
+
+## MAC
+
+# Discussion and summary
+
+# APPENDIX
+
+## Examples of WEAT and MAC calculations
+
+## Word lists

### 2022-07-25 paperDraft/paperDraft4.tex
small structure changes

--- 2022-01-03 paperDraft/paperDraft4.tex
@@
 
-\raf{fix bibliography}
+\hypertarget{word-embeddings-and-their-bias}{%
+\subsection{Word embeddings and their
+bias}\label{word-embeddings-and-their-bias}}
 
-\ali{proofread and think if we should add more}
+\begin{itemize}
+\tightlist
+\item
+  akapit filozoficzny, z odniesieniem do value free
+\end{itemize}
 
@@
 focus of our paper.
+
+\begin{itemize}
+\tightlist
+\item
+  undesirable bias with respect to a certain groups words if
+  stereotypically connected words (for stereotypes that we don't want to
+  be used or relied on in downstream tasks) are systematically closer to
+  each other. + EXAMPLE
+\end{itemize}
+
+\hypertarget{semantical-interpretation}{%
+\subsection{Semantical interpretation}\label{semantical-interpretation}}
+
+\hypertarget{general-challenges}{%
+\subsection{General challenges}\label{general-challenges}}
+
+\hypertarget{weat-and-mac-measures}{%
+\subsection{WEAT and MAC measures}\label{weat-and-mac-measures}}
 
@@
 MAC.
+
+\hypertarget{pre-averaging-and-manufactured-certainty}{%
+\section{Pre-averaging and manufactured
+certainty}\label{pre-averaging-and-manufactured-certainty}}
+
+\hypertarget{general-problem-with-pre-averaging}{%
+\subsection{General problem with
+pre-averaging}\label{general-problem-with-pre-averaging}}
+
+\hypertarget{simulations-for-weat}{%
+\subsection{Simulations for WEAT}\label{simulations-for-weat}}
+
+\hypertarget{simulations-for-mac}{%
+\subsection{Simulations for MAC}\label{simulations-for-mac}}
+
+\hypertarget{bayesian-method}{%
+\section{Bayesian method}\label{bayesian-method}}
+
+\hypertarget{bernstein-approach}{%
+\subsection{Bernstein approach}\label{bernstein-approach}}
+
+\hypertarget{chasing-metrics}{%
+\subsection{Chasing metrics}\label{chasing-metrics}}
+
+\hypertarget{bayesian-method-introduction}{%
+\subsection{Bayesian method
+introduction}\label{bayesian-method-introduction}}
+
+\hypertarget{existing-applications-to-nlp-and-perhaps-to-bias}{%
+\subsection{Existing applications to NLP and perhaps to
+bias}\label{existing-applications-to-nlp-and-perhaps-to-bias}}
+
+\hypertarget{model}{%
+\subsection{Model}\label{model}}
+
+\hypertarget{results}{%
+\section{Results}\label{results}}
+
+\hypertarget{weat}{%
+\subsection{WEAT}\label{weat}}
+
+\hypertarget{mac}{%
+\subsection{MAC}\label{mac}}
+
+\hypertarget{discussion-and-summary}{%
+\section{Discussion and summary}\label{discussion-and-summary}}
+
+\hypertarget{appendix}{%
+\section{APPENDIX}\label{appendix}}
+
+\hypertarget{examples-of-weat-and-mac-calculations}{%
+\subsection{Examples of WEAT and MAC
+calculations}\label{examples-of-weat-and-mac-calculations}}
+
+\hypertarget{word-lists}{%
+\subsection*{Word lists}\label{word-lists}}
+\addcontentsline{toc}{subsection}{Word lists}
 

### 2022-07-25 paperDraft/paperDraft4.Rmd
small paper structure changes

@@
 Modern Natural Language Processing (NLP) models are used to complete various tasks such as providing email filters, smart assistants, search results, language translations, text analytics and so on.
-All of them need as an input words represented by means of numbers which is accomplished with word embeddings, in which particular lexical units are repesented as vectors of real numbers.
+All of them need as an input words represented by means of numbers which is accomplished with word embeddings, in which particular lexical units are represented as vectors of real numbers.
+
+- troche o tym jak sa skonstruowane (optimizing for co-ocurrence frequency)
+
+- przejrzec jak to jest wprowadzane w kilku innych artykulach, zwl. fair is better than sensational
+
+- troche wiecej o tej konstrukcji i intuicji semantycznych, ale to nie super dobry pomysl
+
+- undesirable bias with respect to a certain groups words if stereotypically connected words (for stereotypes that we don't want to be used or relied on in downstream tasks) are systematically closer to each other.
++ EXAMPLES
+
 It has been suggested [@Bolukbasi2016man; @Caliskan2017semanticsBiases; @Gonen2019lipstick; @Lauscher2019multidimensional; @Garg2018years; @Manzini2019blackToCriminal]  that in the learning process  such models can learn implicit biases that reflect harmful stereotypical thinking.
@@
 
-- undesirable bias with respect to a certain groups words if stereotypically connected words (for stereotypes that we don't want to be used or relied on in downstream tasks) are systematically closer to each other.
-+ EXAMPLE
-
-## Semantical interpretation
-
 ## General challenges
 
-## WEAT and MAC measures
+## WEAT and MAC
 
@@
 
-## Non-statistical problems with cosine-based measures of bias
+## Methodological problems with cosine-based measures of bias
 
@@
 
-## Statistical issues
+## Metrics that pre-average are a bad guide
 

### 2022-07-28 paperDraft/paperDraft4.tex
results added to v 7

@@
   10pt,
-  dvipsnames,enabledeprecatedfontcommands, twocolumn]{scrartcl}
+  dvipsnames,enabledeprecatedfontcommands]{scrartcl}
 \title{Taking uncertainty in word embedding bias estimation seriously: a
@@
 accomplished with word embeddings, in which particular lexical units are
-repesented as vectors of real numbers. It has been suggested {[}1--6{]}
-that in the learning process such models can learn implicit biases that
-reflect harmful stereotypical thinking. A large chunk of the literature
-on the topic focuses on the geometry of the learnt word embeddings; in
-particular, unusually low (cosine) proximity of words belonging,
-intuitively, to a stereotype, is taken as a sign that this stereotype
-has been built into a given embedding. This methodology will be in the
-focus of our paper.
+represented as vectors of real numbers.
 
 \begin{itemize}
-\tightlist
+\item
+  troche o tym jak sa skonstruowane (optimizing for co-ocurrence
+  frequency)
+\item
+  przejrzec jak to jest wprowadzane w kilku innych artykulach, zwl. fair
+  is better than sensational
+\item
+  troche wiecej o tej konstrukcji i intuicji semantycznych, ale to nie
+  super dobry pomysl
 \item
@@
   be used or relied on in downstream tasks) are systematically closer to
-  each other. + EXAMPLE
+  each other. + EXAMPLES
 \end{itemize}
 
-\hypertarget{semantical-interpretation}{%
-\subsection{Semantical interpretation}\label{semantical-interpretation}}
+It has been suggested {[}1--6{]} that in the learning process such
+models can learn implicit biases that reflect harmful stereotypical
+thinking. A large chunk of the literature on the topic focuses on the
+geometry of the learnt word embeddings; in particular, unusually low
+(cosine) proximity of words belonging, intuitively, to a stereotype, is
+taken as a sign that this stereotype has been built into a given
+embedding. This methodology will be in the focus of our paper.
 
@@
 
-\hypertarget{weat-and-mac-measures}{%
-\subsection{WEAT and MAC measures}\label{weat-and-mac-measures}}
+\hypertarget{weat-and-mac}{%
+\subsection{WEAT and MAC}\label{weat-and-mac}}
 
@@
 
-\hypertarget{non-statistical-problems-with-cosine-based-measures-of-bias}{%
-\subsection{Non-statistical problems with cosine-based measures of
-bias}\label{non-statistical-problems-with-cosine-based-measures-of-bias}}
+\hypertarget{methodological-problems-with-cosine-based-measures-of-bias}{%
+\subsection{Methodological problems with cosine-based measures of
+bias}\label{methodological-problems-with-cosine-based-measures-of-bias}}
 
@@
 
-\hypertarget{statistical-issues}{%
-\subsection{Statistical issues}\label{statistical-issues}}
+\hypertarget{metrics-that-pre-average-are-a-bad-guide}{%
+\subsection{Metrics that pre-average are a bad
+guide}\label{metrics-that-pre-average-are-a-bad-guide}}
 
@@
 
-\hypertarget{weat}{%
-\subsection{WEAT}\label{weat}}
+\vspace{1mm}
+\footnotesize
 
-\hypertarget{mac}{%
-\subsection{MAC}\label{mac}}
+\begin{figure}
+
+\begin{center}\includegraphics[width=1.1\linewidth]{paperDraft4_files/figure-latex/unnamed-chunk-1-1} \end{center}
+\caption{dsds}
+\label{fig:weat7google}
+\end{figure}
 

### 2022-07-28 paperDraft/paperDraft5.Rmd
results added to v 7

--- 2022-07-25 paperDraft/paperDraft4.Rmd
@@
     df_print: paged
-classoption: dvipsnames,enabledeprecatedfontcommands, twocolumn
+classoption: dvipsnames,enabledeprecatedfontcommands
 fontsize: 10pt
@@
 
-## WEAT
+\vspace{1mm}
+\footnotesize
+```{r generate plots,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%", message = FALSE, warning = FALSE, results = FALSE}
+source("functions/cleanDataset.R")
+source("functions/plotFromPrecis.R")
 
-## MAC
+weat7Google <- cleanDataset(read.csv("./datasets/macWeatDatasets/weat_7_google.csv")[,-1])
+precisWeat7Google <- readRDS("resultsDFs/Weat7GoogleDF.rds")
+resultsWeat7Google <- plotFromPrecis(precis = precisWeat7Google,
+                                     dataset = weat7Google,
+                                     list = "Weat 7",
+                                     embedding = "Google")
+
+weat7Glove <- cleanDataset(read.csv("./datasets/macWeatDatasets/weat_7_glove.csv")[,-1])
+precisWeat7Glove <- readRDS("resultsDFs/Weat7GloveDF.rds")
+resultsWeat7Glove <- plotFromPrecis(precis = precisWeat7Glove,
+                                     dataset = weat7Glove,
+                                     list = "Weat 7",
+                                     embedding = "Glove")
+
+weat7Reddit <- cleanDataset(read.csv("./datasets/macWeatDatasets/weat_7_reddit.csv")[,-1])
+precisWeat7Reddit <- readRDS("resultsDFs/Weat7RedditDF.rds")
+resultsWeat7Reddit <- plotFromPrecis(precis = precisWeat7Reddit,
+                                    dataset = weat7Reddit,
+                                    list = "Weat 7",
+                                    embedding = "Reddit", ylims = c(.5,1.3))
+
+weat1Google <- cleanDataset(read.csv("./datasets/macWeatDatasets/weat_1_google.csv")[,-1])
+precisWeat1Google <- readRDS("resultsDFs/Weat1GoogleDF.rds")
+resultsWeat1Google <- plotFromPrecis(precis = precisWeat1Google,
+                                     dataset = weat1Google,
+                                     list = "Weat 1",
+                                     embedding = "Google")
+
+weat1Glove <- cleanDataset(read.csv("./datasets/macWeatDatasets/weat_1_glove.csv")[,-1])
+precisWeat1Glove <- readRDS("resultsDFs/Weat1GloveDF.rds")
+resultsWeat1Glove <- plotFromPrecis(precis = precisWeat1Glove,
+                                     dataset = weat1Glove,
+                                     list = "Weat 1",
+                                     embedding = "Glove")
+
+weat1Reddit <- cleanDataset(read.csv("./datasets/macWeatDatasets/weat_1_reddit.csv")[,-1])
+precisWeat1Reddit <- readRDS("resultsDFs/Weat1RedditDF.rds")
+resultsWeat1Reddit <- plotFromPrecis(precis = precisWeat1Reddit,
+                                     dataset = weat1Reddit,
+                                     list = "Weat 1",
+                                     embedding = "Reddit")
+
+religionGoogle <- cleanDataset(read.csv("./datasets/macWeatDatasets/religion_group_google_dataset.csv")[,-1])
+precisReligionGoogle <- readRDS("resultsDFs/ReligionGoogleDF.rds")
+resultsReligionGoogle <- plotFromPrecis(precis = precisReligionGoogle,
+                                     dataset = religionGoogle,
+                                     list = "Religion",
+                                     embedding = "Google")
+
+religionGlove <- cleanDataset(read.csv("./datasets/macWeatDatasets/religion_group_glove_dataset.csv")[,-1])
+precisReligionGlove <- readRDS("resultsDFs/ReligionGloveDF.rds")
+resultsReligionGlove <- plotFromPrecis(precis = precisReligionGlove,
+                                        dataset = religionGlove,
+                                        list = "Religion",
+                                        embedding = "Glove")
+
+religionReddit <- cleanDataset(read.csv("./datasets/macWeatDatasets/religion_group_reddit_dataset.csv")[,-1])
+precisReligionReddit <- readRDS("resultsDFs/ReligionRedditDF.rds")
+resultsReligionReddit <- plotFromPrecis(precis = precisReligionReddit,
+                                       dataset = religionReddit,
+                                       list = "Religion",
+                                       embedding = "Reddit")
+
+raceGoogle <-  cleanDataset(read.csv("./datasets/macWeatDatasets/race_group_google_dataset.csv")[,-1])
+precisRaceGoogle <- readRDS("resultsDFs/RaceGoogleDF.rds")
+resultsRaceGoogle <- plotFromPrecis(precis = precisRaceGoogle,
+                                        dataset = raceGoogle,
+                                        list = "Race",
+                                        embedding = "Google")
+
+raceGlove <-  cleanDataset(read.csv("./datasets/macWeatDatasets/race_group_glove_dataset.csv")[,-1])
+precisRaceGlove <- readRDS("resultsDFs/RaceGloveDF.rds")
+resultsRaceGlove <- plotFromPrecis(precis = precisRaceGlove,
+                                    dataset = raceGlove,
+                                    list = "Race",
+                                    embedding = "Glove")
+
+raceReddit <-  cleanDataset(read.csv("./datasets/macWeatDatasets/race_group_reddit_dataset.csv")[,-1])
+precisRaceReddit <- readRDS("resultsDFs/RaceRedditDF.rds")
+resultsRaceReddit <- plotFromPrecis(precis = precisRaceReddit,
+                                    dataset = raceReddit,
+                                    list = "Race",
+                                    embedding = "Reddit")
+
+genderGoogle <- cleanDataset(read.csv("./datasets/macWeatDatasets/gender_group_google_dataset.csv")[,-1])
+precisGenderGoogle <- readRDS("resultsDFs/GenderGoogleDF.rds")
+resultsGenderGoogle <- plotFromPrecis(precis = precisGenderGoogle,
+                                    dataset = genderGoogle,
+                                    list = "Gender",
+                                    embedding = "Google")
+
+genderGlove <- cleanDataset(read.csv("./datasets/macWeatDatasets/gender_group_glove_dataset.csv")[,-1])
+precisGenderGlove <- readRDS("resultsDFs/GenderGloveDF.rds")
+resultsGenderGlove <- plotFromPrecis(precis = precisGenderGlove,
+                                      dataset = genderGlove,
+                                      list = "Gender",
+                                      embedding = "Glove")
+
+genderReddit <- cleanDataset(read.csv("./datasets/macWeatDatasets/gender_group_reddit_dataset.csv")[,-1])
+precisGenderReddit <- readRDS("resultsDFs/GenderRedditDF.rds")
+
+resultsGenderReddit <- plotFromPrecis(precis = precisGenderReddit,
+                                     dataset = genderReddit,
+                                     list = "Gender",
+                                     embedding = "Reddit",  ylims = c(.4,1.2))
+
+```
+
+\begin{figure}
+
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "110%", dpi = 300}
+resultsWeat7Google$plotJoint
+```
+\caption{dsds}
+\label{fig:weat7google}
+\end{figure}
 

### 2022-07-28 paperDraft/paperDraft6.tex
results added to v 7

--- 2022-07-28 paperDraft/paperDraft4.tex
@@
 
-\begin{center}\includegraphics[width=1.1\linewidth]{paperDraft4_files/figure-latex/unnamed-chunk-1-1} \end{center}
+\begin{center}\includegraphics[width=1.1\linewidth]{paperDraft6_files/figure-latex/unnamed-chunk-1-1} \end{center}
 \caption{dsds}
 \label{fig:weat7google}
+\end{figure}
+
+\begin{figure}
+
+\begin{center}\includegraphics[width=1.1\linewidth]{paperDraft6_files/figure-latex/unnamed-chunk-2-1} \end{center}
+\caption{dsds}
+\label{fig:weat7glove}
+\end{figure}
+
+\begin{figure}
+
+\begin{center}\includegraphics[width=1.1\linewidth]{paperDraft6_files/figure-latex/unnamed-chunk-3-1} \end{center}
+\caption{dsds}
+\label{fig:weat7reddit}
+\end{figure}
+
+\begin{figure}
+
+\begin{center}\includegraphics[width=1.1\linewidth]{paperDraft6_files/figure-latex/unnamed-chunk-4-1} \end{center}
+\caption{dsds}
+\label{fig:weat1google}
+\end{figure}
+
+\begin{figure}
+
+\begin{center}\includegraphics[width=1.1\linewidth]{paperDraft6_files/figure-latex/unnamed-chunk-5-1} \end{center}
+\caption{dsds}
+\label{fig:weat1glove}
+\end{figure}
+
+\begin{figure}
+
+\begin{center}\includegraphics[width=1.1\linewidth]{paperDraft6_files/figure-latex/unnamed-chunk-6-1} \end{center}
+\caption{dsds}
+\label{fig:weat1reddit}
+\end{figure}
+
+\begin{figure}
+
+\begin{center}\includegraphics[width=1.1\linewidth]{paperDraft6_files/figure-latex/unnamed-chunk-7-1} \end{center}
+\caption{dsds}
+\label{fig:religionGoogle}
+\end{figure}
+
+\begin{figure}
+
+\begin{center}\includegraphics[width=1.1\linewidth]{paperDraft6_files/figure-latex/unnamed-chunk-8-1} \end{center}
+\caption{dsds}
+\label{fig:religionGlove}
+\end{figure}
+
+\begin{figure}
+
+\begin{center}\includegraphics[width=1.1\linewidth]{paperDraft6_files/figure-latex/unnamed-chunk-9-1} \end{center}
+\caption{dsds}
+\label{fig:religionReddit}
+\end{figure}
+
+\begin{figure}
+
+\begin{center}\includegraphics[width=1.1\linewidth]{paperDraft6_files/figure-latex/unnamed-chunk-10-1} \end{center}
+\caption{dsds}
+\label{fig:raceGoogle}
+\end{figure}
+
+\begin{figure}
+
+\begin{center}\includegraphics[width=1.1\linewidth]{paperDraft6_files/figure-latex/unnamed-chunk-11-1} \end{center}
+\caption{dsds}
+\label{fig:raceGlove}
+\end{figure}
+
+\begin{figure}
+
+\begin{center}\includegraphics[width=1.1\linewidth]{paperDraft6_files/figure-latex/unnamed-chunk-12-1} \end{center}
+\caption{dsds}
+\label{fig:raceReddit}
+\end{figure}
+
+\begin{figure}
+
+\begin{center}\includegraphics[width=1.1\linewidth]{paperDraft6_files/figure-latex/unnamed-chunk-13-1} \end{center}
+\caption{dsds}
+\label{fig:genderGoogle}
+\end{figure}
+
+\begin{figure}
+
+\begin{center}\includegraphics[width=1.1\linewidth]{paperDraft6_files/figure-latex/unnamed-chunk-14-1} \end{center}
+\caption{dsds}
+\label{fig:genderGlove}
+\end{figure}
+
+\begin{figure}
+
+\begin{center}\includegraphics[width=1.1\linewidth]{paperDraft6_files/figure-latex/unnamed-chunk-15-1} \end{center}
+\caption{dsds}
+\label{fig:genderReddit}
 \end{figure}

### 2022-07-28 paperDraft/paperDraft7.Rmd
results added to v 7

--- 2022-07-25 paperDraft/paperDraft4.Rmd
@@
 
-## WEAT
+\vspace{1mm}
+\footnotesize
+```{r generate plots,echo=FALSE,eval=TRUE,fig.align = "center",cache=FALSE, fig.show = "hold", out.width = "100%", message = FALSE, warning = FALSE, results = FALSE}
+source("functions/cleanDataset.R")
+source("functions/plotFromPrecis.R")
 
-## MAC
+weat7Google <- cleanDataset(read.csv("./datasets/macWeatDatasets/weat_7_google.csv")[,-1])
+precisWeat7Google <- readRDS("resultsDFs/Weat7GoogleDF.rds")
+resultsWeat7Google <- plotFromPrecis(precis = precisWeat7Google,
+                                     dataset = weat7Google,
+                                     list = "Weat 7",
+                                     embedding = "Google",
+                                     ylims = c(.75,1.1))
+
+weat7Glove <- cleanDataset(read.csv("./datasets/macWeatDatasets/weat_7_glove.csv")[,-1])
+precisWeat7Glove <- readRDS("resultsDFs/Weat7GloveDF.rds")
+resultsWeat7Glove <- plotFromPrecis(precis = precisWeat7Glove,
+                                     dataset = weat7Glove,
+                                     list = "Weat 7",
+                                     embedding = "Glove")
+
+weat7Reddit <- cleanDataset(read.csv("./datasets/macWeatDatasets/weat_7_reddit.csv")[,-1])
+precisWeat7Reddit <- readRDS("resultsDFs/Weat7RedditDF.rds")
+resultsWeat7Reddit <- plotFromPrecis(precis = precisWeat7Reddit,
+                                    dataset = weat7Reddit,
+                                    list = "Weat 7",
+                                    embedding = "Reddit", ylims = c(.5,1.3))
+
+weat1Google <- cleanDataset(read.csv("./datasets/macWeatDatasets/weat_1_google.csv")[,-1])
+precisWeat1Google <- readRDS("resultsDFs/Weat1GoogleDF.rds")
+resultsWeat1Google <- plotFromPrecis(precis = precisWeat1Google,
+                                     dataset = weat1Google,
+                                     list = "Weat 1",
+                                     embedding = "Google",
+                                     ylims = c(.8,1.05))
+
+weat1Glove <- cleanDataset(read.csv("./datasets/macWeatDatasets/weat_1_glove.csv")[,-1])
+precisWeat1Glove <- readRDS("resultsDFs/Weat1GloveDF.rds")
+resultsWeat1Glove <- plotFromPrecis(precis = precisWeat1Glove,
+                                     dataset = weat1Glove,
+                                     list = "Weat 1",
+                                     embedding = "Glove",
+                                     ylims = c(.75,1.2))
+
+weat1Reddit <- cleanDataset(read.csv("./datasets/macWeatDatasets/weat_1_reddit.csv")[,-1])
+precisWeat1Reddit <- readRDS("resultsDFs/Weat1RedditDF.rds")
+resultsWeat1Reddit <- plotFromPrecis(precis = precisWeat1Reddit,
+                                     dataset = weat1Reddit,
+                                     list = "Weat 1",
+                                     embedding = "Reddit",
+                                      ylims = c(.5,1.1))
+
+religionGoogle <- cleanDataset(read.csv("./datasets/macWeatDatasets/religion_group_google_dataset.csv")[,-1])
+precisReligionGoogle <- readRDS("resultsDFs/ReligionGoogleDF.rds")
+resultsReligionGoogle <- plotFromPrecis(precis = precisReligionGoogle,
+                                     dataset = religionGoogle,
+                                     list = "Religion",
+                                     embedding = "Google",
+                                      ylims = c(.65,1.05))
+
+religionGlove <- cleanDataset(read.csv("./datasets/macWeatDatasets/religion_group_glove_dataset.csv")[,-1])
+precisReligionGlove <- readRDS("resultsDFs/ReligionGloveDF.rds")
+resultsReligionGlove <- plotFromPrecis(precis = precisReligionGlove,
+                                        dataset = religionGlove,
+                                        list = "Religion",
+                                        embedding = "Glove",
+                                        ylims = c(.7,1.1))
+
+religionReddit <- cleanDataset(read.csv("./datasets/macWeatDatasets/religion_group_reddit_dataset.csv")[,-1])
+precisReligionReddit <- readRDS("resultsDFs/ReligionRedditDF.rds")
+resultsReligionReddit <- plotFromPrecis(precis = precisReligionReddit,
+                                       dataset = religionReddit,
+                                       list = "Religion",
+                                       embedding = "Reddit")
+
+raceGoogle <-  cleanDataset(read.csv("./datasets/macWeatDatasets/race_group_google_dataset.csv")[,-1])
+precisRaceGoogle <- readRDS("resultsDFs/RaceGoogleDF.rds")
+resultsRaceGoogle <- plotFromPrecis(precis = precisRaceGoogle,
+                                        dataset = raceGoogle,
+                                        list = "Race",
+                                        embedding = "Google",
+                                     ylims = c(.8,1.1))
+
+raceGlove <-  cleanDataset(read.csv("./datasets/macWeatDatasets/race_group_glove_dataset.csv")[,-1])
+precisRaceGlove <- readRDS("resultsDFs/RaceGloveDF.rds")
+resultsRaceGlove <- plotFromPrecis(precis = precisRaceGlove,
+                                    dataset = raceGlove,
+                                    list = "Race",
+                                    embedding = "Glove",
+                                    ylims = c(.7,1.1))
+
+raceReddit <-  cleanDataset(read.csv("./datasets/macWeatDatasets/race_group_reddit_dataset.csv")[,-1])
+precisRaceReddit <- readRDS("resultsDFs/RaceRedditDF.rds")
+resultsRaceReddit <- plotFromPrecis(precis = precisRaceReddit,
+                                    dataset = raceReddit,
+                                    list = "Race",
+                                    embedding = "Reddit")
+
+genderGoogle <- cleanDataset(read.csv("./datasets/macWeatDatasets/gender_group_google_dataset.csv")[,-1])
+precisGenderGoogle <- readRDS("resultsDFs/GenderGoogleDF.rds")
+resultsGenderGoogle <- plotFromPrecis(precis = precisGenderGoogle,
+                                    dataset = genderGoogle,
+                                    list = "Gender",
+                                    embedding = "Google",
+                                     ylims = c(.6,1.05))
+
+genderGlove <- cleanDataset(read.csv("./datasets/macWeatDatasets/gender_group_glove_dataset.csv")[,-1])
+precisGenderGlove <- readRDS("resultsDFs/GenderGloveDF.rds")
+resultsGenderGlove <- plotFromPrecis(precis = precisGenderGlove,
+                                      dataset = genderGlove,
+                                      list = "Gender",
+                                      embedding = "Glove",
+                                      ylims = c(.5,1.1))
+
+genderReddit <- cleanDataset(read.csv("./datasets/macWeatDatasets/gender_group_reddit_dataset.csv")[,-1])
+precisGenderReddit <- readRDS("resultsDFs/GenderRedditDF.rds")
+
+resultsGenderReddit <- plotFromPrecis(precis = precisGenderReddit,
+                                     dataset = genderReddit,
+                                     list = "Gender",
+                                     embedding = "Reddit",  ylims = c(.4,1.1))
+
+```
+
+\begin{figure}
+
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "110%", dpi = 300}
+resultsWeat7Google$plotJoint
+```
+\caption{dsds}
+\label{fig:weat7google}
+\end{figure}
+
+\begin{figure}
+
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "110%", dpi = 300}
+resultsWeat7Glove$plotJoint
+```
+\caption{dsds}
+\label{fig:weat7glove}
+\end{figure}
+
+\begin{figure}
+
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "110%", dpi = 300}
+resultsWeat7Reddit$plotJoint
+```
+\caption{dsds}
+\label{fig:weat7reddit}
+\end{figure}
+
+\begin{figure}
+
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "110%", dpi = 300}
+resultsWeat1Google$plotJoint
+```
+\caption{dsds}
+\label{fig:weat1google}
+\end{figure}
+
+\begin{figure}
+
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "110%", dpi = 300}
+resultsWeat1Glove$plotJoint
+```
+\caption{dsds}
+\label{fig:weat1glove}
+\end{figure}
+
+\begin{figure}
+
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "110%", dpi = 300}
+resultsWeat1Reddit$plotJoint
+```
+\caption{dsds}
+\label{fig:weat1reddit}
+\end{figure}
+
+\begin{figure}
+
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "110%", dpi = 300}
+resultsReligionGoogle$plotJoint
+```
+\caption{dsds}
+\label{fig:religionGoogle}
+\end{figure}
+
+\begin{figure}
+
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "110%", dpi = 300}
+resultsReligionGlove$plotJoint
+```
+\caption{dsds}
+\label{fig:religionGlove}
+\end{figure}
+
+\begin{figure}
+
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "110%", dpi = 300}
+resultsReligionReddit$plotJoint
+```
+\caption{dsds}
+\label{fig:religionReddit}
+\end{figure}
+
+\begin{figure}
+
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "110%", dpi = 300}
+resultsRaceGoogle$plotJoint
+```
+\caption{dsds}
+\label{fig:raceGoogle}
+\end{figure}
+
+\begin{figure}
+
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "110%", dpi = 300}
+resultsRaceGlove$plotJoint
+```
+\caption{dsds}
+\label{fig:raceGlove}
+\end{figure}
+
+\begin{figure}
+
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "110%", dpi = 300}
+resultsRaceReddit$plotJoint
+```
+\caption{dsds}
+\label{fig:raceReddit}
+\end{figure}
+
+\begin{figure}
+
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "110%", dpi = 300}
+resultsGenderGoogle$plotJoint
+```
+\caption{dsds}
+\label{fig:genderGoogle}
+\end{figure}
+
+\begin{figure}
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "110%", dpi = 300}
+resultsGenderGlove$plotJoint
+```
+\caption{dsds}
+\label{fig:genderGlove}
+\end{figure}
+
+\begin{figure}
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "110%", dpi = 300}
+resultsGenderReddit$plotJoint
+```
+\caption{dsds}
+\label{fig:genderReddit}
+\end{figure}
 

### 2022-07-28 paperDraft/paperDraft7.tex
results added to v 7

--- 2022-07-28 paperDraft/paperDraft4.tex
@@
 
-\begin{center}\includegraphics[width=1.1\linewidth]{paperDraft4_files/figure-latex/unnamed-chunk-1-1} \end{center}
+\begin{center}\includegraphics[width=1.1\linewidth]{paperDraft7_files/figure-latex/unnamed-chunk-1-1} \end{center}
 \caption{dsds}
 \label{fig:weat7google}
+\end{figure}
+
+\begin{figure}
+
+\begin{center}\includegraphics[width=1.1\linewidth]{paperDraft7_files/figure-latex/unnamed-chunk-2-1} \end{center}
+\caption{dsds}
+\label{fig:weat7glove}
+\end{figure}
+
+\begin{figure}
+
+\begin{center}\includegraphics[width=1.1\linewidth]{paperDraft7_files/figure-latex/unnamed-chunk-3-1} \end{center}
+\caption{dsds}
+\label{fig:weat7reddit}
+\end{figure}
+
+\begin{figure}
+
+\begin{center}\includegraphics[width=1.1\linewidth]{paperDraft7_files/figure-latex/unnamed-chunk-4-1} \end{center}
+\caption{dsds}
+\label{fig:weat1google}
+\end{figure}
+
+\begin{figure}
+
+\begin{center}\includegraphics[width=1.1\linewidth]{paperDraft7_files/figure-latex/unnamed-chunk-5-1} \end{center}
+\caption{dsds}
+\label{fig:weat1glove}
+\end{figure}
+
+\begin{figure}
+
+\begin{center}\includegraphics[width=1.1\linewidth]{paperDraft7_files/figure-latex/unnamed-chunk-6-1} \end{center}
+\caption{dsds}
+\label{fig:weat1reddit}
+\end{figure}
+
+\begin{figure}
+
+\begin{center}\includegraphics[width=1.1\linewidth]{paperDraft7_files/figure-latex/unnamed-chunk-7-1} \end{center}
+\caption{dsds}
+\label{fig:religionGoogle}
+\end{figure}
+
+\begin{figure}
+
+\begin{center}\includegraphics[width=1.1\linewidth]{paperDraft7_files/figure-latex/unnamed-chunk-8-1} \end{center}
+\caption{dsds}
+\label{fig:religionGlove}
+\end{figure}
+
+\begin{figure}
+
+\begin{center}\includegraphics[width=1.1\linewidth]{paperDraft7_files/figure-latex/unnamed-chunk-9-1} \end{center}
+\caption{dsds}
+\label{fig:religionReddit}
+\end{figure}
+
+\begin{figure}
+
+\begin{center}\includegraphics[width=1.1\linewidth]{paperDraft7_files/figure-latex/unnamed-chunk-10-1} \end{center}
+\caption{dsds}
+\label{fig:raceGoogle}
+\end{figure}
+
+\begin{figure}
+
+\begin{center}\includegraphics[width=1.1\linewidth]{paperDraft7_files/figure-latex/unnamed-chunk-11-1} \end{center}
+\caption{dsds}
+\label{fig:raceGlove}
+\end{figure}
+
+\begin{figure}
+
+\begin{center}\includegraphics[width=1.1\linewidth]{paperDraft7_files/figure-latex/unnamed-chunk-12-1} \end{center}
+\caption{dsds}
+\label{fig:raceReddit}
+\end{figure}
+
+\begin{figure}
+
+\begin{center}\includegraphics[width=1.1\linewidth]{paperDraft7_files/figure-latex/unnamed-chunk-13-1} \end{center}
+\caption{dsds}
+\label{fig:genderGoogle}
+\end{figure}
+
+\begin{figure}
+
+\begin{center}\includegraphics[width=1.1\linewidth]{paperDraft7_files/figure-latex/unnamed-chunk-14-1} \end{center}
+\caption{dsds}
+\label{fig:genderGlove}
+\end{figure}
+
+\begin{figure}
+
+\begin{center}\includegraphics[width=1.1\linewidth]{paperDraft7_files/figure-latex/unnamed-chunk-15-1} \end{center}
+\caption{dsds}
+\label{fig:genderReddit}
 \end{figure}

### 2022-07-28 functions/plotFromPrecis.R

@@ -5,16 +5,16 @@ library(tidyverse)
 library(gridExtra)
 library(grid)
 
-precis
+# precis
 
 
 plotFromPrecis <- function (precis, dataset, ylims = c(.5,1.2), list, embedding){
   
   
-  precis = precisWeat7Google
-  dataset = weat7Google
-  list = "Weat 7"
-  embedding = "Google"
+  # precis = precisWeat7Google
+  # dataset = weat7Google
+  # list = "Weat 7"
+  # embedding = "Google"
   
   
   results <- list()
@@ -73,8 +73,8 @@ plotFromPrecis <- function (precis, dataset, ylims = c(.5,1.2), list, embedding)
   
   
   
-  grobIndividual <- ggplotGrob(results$plotIndividual+theme_tufte(base_size = 7) + theme(legend.position='none'))
-  grobOverall <- ggplotGrob(results$plotOverall+theme_tufte(base_size = 7)+ theme(legend.position = c(0.9, 0.7)))
+  grobIndividual <- ggplotGrob(results$plotIndividual+theme_tufte(base_size = 6) + theme(legend.position='none'))
+  grobOverall <- ggplotGrob(results$plotOverall+theme_tufte(base_size = 6)+ theme(legend.position = c(0.9, 0.7)))
   
   
   results$plotJoint <- ggplot(data.frame(a=1)) + xlim(1, 20) + ylim(10, 60)+theme_void()+

### 2022-07-28 paperDraft/paperDraft4.Rmd
Updates chapters alicja

@@
 
+### What is Bayesian statistics?
+
+Bayesian statistics is a theory which enables one to assign epistemological uncertainty with the use of probability.
+It is a system that allows to express a degree of belief in a certain event.
+Bayesian methods use existing prior beliefs together with real data to receive posterior beliefs.
+Bayesian inference together with probability tools may be used to quantify the uncertainty in the inferences.
+
+### How does it differ from the classical statistics?
+
+One could ask what is the difference between Bayesian statistics and the classical, frequentist approach.
+Broadly speaking the Frequentist statistics looks, as the name suggests, at the frequency of an event in a long run and based on that assigns its probability.
+In Bayesian statistics the approach is more epistemological as one may include his/her prior beliefs on the likelihood of an event and update it accordingly to the available data which results in the posterior probability of an event.
+What is more in a classical approach, the parameters are fixed whereas in Bayesian statistics the parameters are assumed to be random variables to which one assigns a probability distribution.
+In contrast with CIs, the posterior distributions (and their highest posterior density intervals HPDIs, the narrowest intervals containing a certain ratio of the area under the curve) are easily interpretable and have direct relevance for the question at hand.
+
+### How does the Bayesian model work?
+
+Referring to the \textit{Statistical rethinking} one may distinguish three steps that are necessary to design Bayesian model:
+1. Data story: Think about the narrative on how the data can emerge. 
+2. Update: Let the model learn from the data.
+3. Evaluate: Verify whether it is a good representation of the phenomenon.
+
+A Bayesian model consists of the variables and the distributions that define these variables.
+The priors that one assigns provide the likelihood of each parameter before any data has been seen.
+Then using real data one may apply numerical techniques like Markov chain Monte Carlo to calculate the posterior distribution.
+
+### Interpreting model's results
+
+The standard output from the model's learning may contain the following pieces of information.
+
+WAIC -
+
+### What are hierarchical models?
+
+Hierarchical models are also known as multilevel models, random effects, varying effects, or mixed effects models.
+In \textit{Statistical rethinking} one may read that the central device of the multilevel models is to learn the strength of the prior from the data itself.
+Multilevel models may be regarded as adaptive regularization, meaning that the model tunes itself to get the most optimal skepticism.
+It is worth mentioning that with these types of models adding more parameters can sometimes lead to the effect where effective number of parameters (the penalty term in WAIC) is actually reduced.
+The overfitting may be more dependent on how the parameters are related to one another than with how many parameters one inputs into the model.
+
+### Coding example from our model
+
+To build a model one may use original \textit{Stan} which is a state-of-the-art platform for statistical modeling and high-performance statistical computation (https://mc-stan.org/).
+The simpler option is to use \textit{Rethinking} package from the \textit{Statistical rethinking} course.
+With one of the functions \textit{ulam} one may build RStan models from the formulas in an easy manner.
+
+Explain the model?
+
+```{r, eval=FALSE, echo=TRUE}
+buildModel <- function(dataset){
+  options(buildtools.check =
+  function(action) TRUE )
+
+  modelResult <- ulam(
+    alist(
+      distance ~ dnorm(mu,sigma),
+      mu <- d[pwi] * different
+      + a[pwi] * associated
+      + h[pwi] * human
+      + n[pwi] * none,
+
+      d[pwi] ~ dnorm(dbar, dsigmabar),
+      a[pwi] ~ dnorm(abar, asigmabar),
+      h[pwi] ~ dnorm(hbar, hsigmabar),
+      n[pwi] ~ dnorm(nbar, nsigmabar),
+
+      dbar ~ dnorm(1,.3),
+      abar ~ dnorm(1,.3),
+      hbar ~ dnorm(1,.3),
+      nbar ~ dnorm(1,.3),
+      dsigmabar ~ dexp(2),
+      asigmabar ~ dexp(2),
+      hsigmabar ~ dexp(2),
+      nsigmabar ~ dexp(2),
+      sigma <- s[connection],
+      s[connection] ~  dexp(2)
+    ),
+
+    data = dataset, chains=2 ,
+    iter=8000 , warmup=1000,
+    log_lik = TRUE, cores = 4
+  )
+}
+```
+
+### Limitations
+
+Each method has its own limitations and Bayesian statistics is no exception.
+One may think about a few limitations with this approach.
+One of them is that there is no final and correct way to choose a prior.
+At the same time prior beliefs can be explicitly stated, debated, or often found in previous research which may give good justification for a given set of priors.
+Finally, Bayesian method may be computationally expensive which may be a limiting factor when considering to use it in a more real-time environment.
+
 ## Existing applications to NLP and perhaps to bias

### 2022-07-28 paperDraft/paperDraft4.tex
Updates chapters alicja

--- 2022-06-22 paperDraft/paperDraft4.tex
@@
 \urlstyle{same} % disable monospaced font for URLs
+\usepackage{color}
+\usepackage{fancyvrb}
+\newcommand{\VerbBar}{|}
+\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+% Add ',fontsize=\small' for more characters per line
+\usepackage{framed}
+\definecolor{shadecolor}{RGB}{248,248,248}
+\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\BuiltInTok}[1]{#1}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
+\newcommand{\ExtensionTok}[1]{#1}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ImportTok}[1]{#1}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\NormalTok}[1]{#1}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\RegionMarkerTok}[1]{#1}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
 \usepackage{graphicx}
@@
 
-\hypertarget{what-is-bias}{%
-\subsection{What is bias?}\label{what-is-bias}}
+\hypertarget{word-embeddings-and-their-bias}{%
+\subsection{Word embeddings and their
+bias}\label{word-embeddings-and-their-bias}}
 
-\hypertarget{word-embeddings}{%
-\subsection{Word embeddings}\label{word-embeddings}}
+\begin{itemize}
+\tightlist
+\item
+  akapit filozoficzny, z odniesieniem do value free
+\end{itemize}
 
@@
 accomplished with word embeddings, in which particular lexical units are
-repesented as vectors of real numbers. It has been suggested {[}1--6{]}
-that in the learning process such models can learn implicit biases that
-reflect harmful stereotypical thinking. A large chunk of the literature
-on the topic focuses on the geometry of the learnt word embeddings; in
-particular, unusually low (cosine) proximity of words belonging,
-intuitively, to a stereotype, is taken as a sign that this stereotype
-has been built into a given embedding. This methodology will be in the
-focus of our paper.
+represented as vectors of real numbers.
 
-\hypertarget{semantical-interpretation}{%
-\subsection{Semantical interpretation}\label{semantical-interpretation}}
+\begin{itemize}
+\item
+  troche o tym jak sa skonstruowane (optimizing for co-ocurrence
+  frequency)
+\item
+  przejrzec jak to jest wprowadzane w kilku innych artykulach, zwl. fair
+  is better than sensational
+\item
+  troche wiecej o tej konstrukcji i intuicji semantycznych, ale to nie
+  super dobry pomysl
+\item
+  undesirable bias with respect to a certain groups words if
+  stereotypically connected words (for stereotypes that we don't want to
+  be used or relied on in downstream tasks) are systematically closer to
+  each other. + EXAMPLES
+\end{itemize}
+
+It has been suggested {[}1--6{]} that in the learning process such
+models can learn implicit biases that reflect harmful stereotypical
+thinking. A large chunk of the literature on the topic focuses on the
+geometry of the learnt word embeddings; in particular, unusually low
+(cosine) proximity of words belonging, intuitively, to a stereotype, is
+taken as a sign that this stereotype has been built into a given
+embedding. This methodology will be in the focus of our paper.
 
@@
 introduction}\label{bayesian-method-introduction}}
+
+\hypertarget{what-is-bayesian-statistics}{%
+\subsubsection{What is Bayesian
+statistics?}\label{what-is-bayesian-statistics}}
+
+Bayesian statistics is a theory which enables one to assign
+epistemological uncertainty with the use of probability. It is a system
+that allows to express a degree of belief in a certain event. Bayesian
+methods use existing prior beliefs together with real data to receive
+posterior beliefs. Bayesian inference together with probability tools
+may be used to quantify the uncertainty in the inferences.
+
+\hypertarget{how-does-it-differ-from-the-classical-statistics}{%
+\subsubsection{How does it differ from the classical
+statistics?}\label{how-does-it-differ-from-the-classical-statistics}}
+
+One could ask what is the difference between Bayesian statistics and the
+classical, frequentist approach. Broadly speaking the Frequentist
+statistics looks, as the name suggests, at the frequency of an event in
+a long run and based on that assigns its probability. In Bayesian
+statistics the approach is more epistemological as one may include
+his/her prior beliefs on the likelihood of an event and update it
+accordingly to the available data which results in the posterior
+probability of an event. What is more in a classical approach, the
+parameters are fixed whereas in Bayesian statistics the parameters are
+assumed to be random variables to which one assigns a probability
+distribution. In contrast with CIs, the posterior distributions (and
+their highest posterior density intervals HPDIs, the narrowest intervals
+containing a certain ratio of the area under the curve) are easily
+interpretable and have direct relevance for the question at hand.
+
+\hypertarget{how-does-the-bayesian-model-work}{%
+\subsubsection{How does the Bayesian model
+work?}\label{how-does-the-bayesian-model-work}}
+
+Referring to the \textit{Statistical rethinking} one may distinguish
+three steps that are necessary to design Bayesian model: 1. Data story:
+Think about the narrative on how the data can emerge. 2. Update: Let the
+model learn from the data. 3. Evaluate: Verify whether it is a good
+representation of the phenomenon.
+
+A Bayesian model consists of the variables and the distributions that
+define these variables. The priors that one assigns provide the
+likelihood of each parameter before any data has been seen. Then using
+real data one may apply numerical techniques like Markov chain Monte
+Carlo to calculate the posterior distribution.
+
+\hypertarget{interpreting-models-results}{%
+\subsubsection{Interpreting model's
+results}\label{interpreting-models-results}}
+
+The standard output from the model's learning may contain the following
+pieces of information.
+
+WAIC -
+
+\hypertarget{what-are-hierarchical-models}{%
+\subsubsection{What are hierarchical
+models?}\label{what-are-hierarchical-models}}
+
+Hierarchical models are also known as multilevel models, random effects,
+varying effects, or mixed effects models. In
+\textit{Statistical rethinking} one may read that the central device of
+the multilevel models is to learn the strength of the prior from the
+data itself. Multilevel models may be regarded as adaptive
+regularization, meaning that the model tunes itself to get the most
+optimal skepticism. It is worth mentioning that with these types of
+models adding more parameters can sometimes lead to the effect where
+effective number of parameters (the penalty term in WAIC) is actually
+reduced. The overfitting may be more dependent on how the parameters are
+related to one another than with how many parameters one inputs into the
+model.
+
+\hypertarget{coding-example-from-our-model}{%
+\subsubsection{Coding example from our
+model}\label{coding-example-from-our-model}}
+
+To build a model one may use original \textit{Stan} which is a
+state-of-the-art platform for statistical modeling and high-performance
+statistical computation (\url{https://mc-stan.org/}). The simpler option
+is to use \textit{Rethinking} package from the
+\textit{Statistical rethinking} course. With one of the functions
+\textit{ulam} one may build RStan models from the formulas in an easy
+manner.
+
+Explain the model?
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{buildModel }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(dataset)\{}
+  \FunctionTok{options}\NormalTok{(}\AttributeTok{buildtools.check =}
+  \ControlFlowTok{function}\NormalTok{(action) }\ConstantTok{TRUE}\NormalTok{ )}
+
+\NormalTok{  modelResult }\OtherTok{\textless{}{-}} \FunctionTok{ulam}\NormalTok{(}
+    \FunctionTok{alist}\NormalTok{(}
+\NormalTok{      distance }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(mu,sigma),}
+\NormalTok{      mu }\OtherTok{\textless{}{-}}\NormalTok{ d[pwi] }\SpecialCharTok{*}\NormalTok{ different}
+      \SpecialCharTok{+}\NormalTok{ a[pwi] }\SpecialCharTok{*}\NormalTok{ associated}
+      \SpecialCharTok{+}\NormalTok{ h[pwi] }\SpecialCharTok{*}\NormalTok{ human}
+      \SpecialCharTok{+}\NormalTok{ n[pwi] }\SpecialCharTok{*}\NormalTok{ none,}
+
+\NormalTok{      d[pwi] }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(dbar, dsigmabar),}
+\NormalTok{      a[pwi] }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(abar, asigmabar),}
+\NormalTok{      h[pwi] }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(hbar, hsigmabar),}
+\NormalTok{      n[pwi] }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(nbar, nsigmabar),}
+
+\NormalTok{      dbar }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{3}\NormalTok{),}
+\NormalTok{      abar }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{3}\NormalTok{),}
+\NormalTok{      hbar }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{3}\NormalTok{),}
+\NormalTok{      nbar }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dnorm}\NormalTok{(}\DecValTok{1}\NormalTok{,.}\DecValTok{3}\NormalTok{),}
+\NormalTok{      dsigmabar }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dexp}\NormalTok{(}\DecValTok{2}\NormalTok{),}
+\NormalTok{      asigmabar }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dexp}\NormalTok{(}\DecValTok{2}\NormalTok{),}
+\NormalTok{      hsigmabar }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dexp}\NormalTok{(}\DecValTok{2}\NormalTok{),}
+\NormalTok{      nsigmabar }\SpecialCharTok{\textasciitilde{}} \FunctionTok{dexp}\NormalTok{(}\DecValTok{2}\NormalTok{),}
+\NormalTok{      sigma }\OtherTok{\textless{}{-}}\NormalTok{ s[connection],}
+\NormalTok{      s[connection] }\SpecialCharTok{\textasciitilde{}}  \FunctionTok{dexp}\NormalTok{(}\DecValTok{2}\NormalTok{)}
+\NormalTok{    ),}
+
+    \AttributeTok{data =}\NormalTok{ dataset, }\AttributeTok{chains=}\DecValTok{2}\NormalTok{ ,}
+    \AttributeTok{iter=}\DecValTok{8000}\NormalTok{ , }\AttributeTok{warmup=}\DecValTok{1000}\NormalTok{,}
+    \AttributeTok{log\_lik =} \ConstantTok{TRUE}\NormalTok{, }\AttributeTok{cores =} \DecValTok{4}
+\NormalTok{  )}
+\NormalTok{\}}
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{limitations}{%
+\subsubsection{Limitations}\label{limitations}}
+
+Each method has its own limitations and Bayesian statistics is no
+exception. One may think about a few limitations with this approach. One
+of them is that there is no final and correct way to choose a prior. At
+the same time prior beliefs can be explicitly stated, debated, or often
+found in previous research which may give good justification for a given
+set of priors. Finally, Bayesian method may be computationally expensive
+which may be a limiting factor when considering to use it in a more
+real-time environment.
 

### 2022-07-28 functions/caliskanExtended.R

@@ -25,6 +25,7 @@ t5 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
 t6 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
 t7 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
 t8 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
+
 t9 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
 t10 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
 t11 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))

### 2022-07-28 functions/macSimulations.R

@@ -1,0 +1,115 @@
+# 2 protected groups in a class, eg. gender -> man, woman
+# calculate the average inside the group and then average all groups
+
+
+knitr::opts_knit$set(root.dir = '../')
+library(ggplot2)
+library(ggthemes)
+library(gridExtra)
+library(grid)
+
+# BASIC SIMULATION -> samples=5, mean=0, st=0.08
+
+# X = {t1,t2,t3,t4,t5,t6,t7,t8}, Y = {t9,t10,t11,t12,t13,t14,t15,t16}
+
+
+# does it hide existing difference between the class and attribute?
+
+# visualize systematic difference
+# calculate simulation
+# save p-value (check also p-value when no difference) -> two variants
+
+set.seed(123456)
+# X
+t1 <- data.frame(A  = rnorm(8,1,0.08), B = rnorm(8,1,0.08))
+t2 <- data.frame(A  = rnorm(8,1,0.08), B = rnorm(8,1,0.08))
+t3 <- data.frame(A  = rnorm(8,1,0.08), B = rnorm(8,1,0.08))
+t4 <- data.frame(A  = rnorm(8,1,0.08), B = rnorm(8,1,0.08))
+t5 <- data.frame(A  = rnorm(8,1,0.08), B = rnorm(8,1,0.08))
+t6 <- data.frame(A  = rnorm(8,1,0.08), B = rnorm(8,1,0.08))
+t7 <- data.frame(A  = rnorm(8,1,0.08), B = rnorm(8,1,0.08))
+
+# Y
+t8 <- data.frame(A  = rnorm(8,1,0.08), B = rnorm(8,1,0.08))
+t9 <- data.frame(A  = rnorm(8,1,0.08), B = rnorm(8,1,0.08))
+t10 <- data.frame(A  = rnorm(8,1,0.08), B = rnorm(8,1,0.08))
+t11 <- data.frame(A  = rnorm(8,1,0.08), B = rnorm(8,1,0.08))
+t12 <- data.frame(A  = rnorm(8,1,0.08), B = rnorm(8,1,0.08))
+t13 <- data.frame(A  = rnorm(8,1,0.08), B = rnorm(8,1,0.08))
+t14 <- data.frame(A  = rnorm(8,1,0.08), B = rnorm(8,1,0.08))
+
+
[…]

### 2022-07-28 paperDraft/paperDraft6.Rmd

--- 2022-07-28 paperDraft/paperDraft5.Rmd
@@
 \footnotesize
-```{r generate plots,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%", message = FALSE, warning = FALSE, results = FALSE}
+```{r generate plots,echo=FALSE,eval=TRUE,fig.align = "center",cache=FALSE, fig.show = "hold", out.width = "100%", message = FALSE, warning = FALSE, results = FALSE}
 source("functions/cleanDataset.R")
@@
                                      list = "Weat 7",
-                                     embedding = "Google")
+                                     embedding = "Google",
+                                     ylims = c(.75,1.1))
 
@@
 
+\begin{figure}
+
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "110%", dpi = 300}
+resultsWeat7Glove$plotJoint
+```
+\caption{dsds}
+\label{fig:weat7glove}
+\end{figure}
+
+\begin{figure}
+
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "110%", dpi = 300}
+resultsWeat7Reddit$plotJoint
+```
+\caption{dsds}
+\label{fig:weat7reddit}
+\end{figure}
+
+\begin{figure}
+
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "110%", dpi = 300}
+resultsWeat1Google$plotJoint
+```
+\caption{dsds}
+\label{fig:weat1google}
+\end{figure}
+
+\begin{figure}
+
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "110%", dpi = 300}
+resultsWeat1Glove$plotJoint
+```
+\caption{dsds}
+\label{fig:weat1glove}
+\end{figure}
+
+\begin{figure}
+
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "110%", dpi = 300}
+resultsWeat1Reddit$plotJoint
+```
+\caption{dsds}
+\label{fig:weat1reddit}
+\end{figure}
+
+\begin{figure}
+
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "110%", dpi = 300}
+resultsReligionGoogle$plotJoint
+```
+\caption{dsds}
+\label{fig:religionGoogle}
+\end{figure}
+
+\begin{figure}
+
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "110%", dpi = 300}
+resultsReligionGlove$plotJoint
+```
+\caption{dsds}
+\label{fig:religionGlove}
+\end{figure}
+
+\begin{figure}
+
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "110%", dpi = 300}
+resultsReligionReddit$plotJoint
+```
+\caption{dsds}
+\label{fig:religionReddit}
+\end{figure}
+
+\begin{figure}
+
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "110%", dpi = 300}
+resultsRaceGoogle$plotJoint
+```
+\caption{dsds}
+\label{fig:raceGoogle}
+\end{figure}
+
+\begin{figure}
+
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "110%", dpi = 300}
+resultsRaceGlove$plotJoint
+```
+\caption{dsds}
+\label{fig:raceGlove}
+\end{figure}
+
+\begin{figure}
+
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "110%", dpi = 300}
+resultsRaceReddit$plotJoint
+```
+\caption{dsds}
+\label{fig:raceReddit}
+\end{figure}
+
+\begin{figure}
+
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "110%", dpi = 300}
+resultsGenderGoogle$plotJoint
+```
+\caption{dsds}
+\label{fig:genderGoogle}
+\end{figure}
+
+\begin{figure}
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "110%", dpi = 300}
+resultsGenderGlove$plotJoint
+```
+\caption{dsds}
+\label{fig:genderGlove}
+\end{figure}
+
+\begin{figure}
+```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "110%", dpi = 300}
+resultsGenderReddit$plotJoint
+```
+\caption{dsds}
+\label{fig:genderReddit}
+\end{figure}
+
 # Discussion and summary

### 2022-07-29 paperDraft/paperDraft4.Rmd
Updates draft

@@
 
-- akapit filozoficzny, z odniesieniem do value free
-
 Modern Natural Language Processing (NLP) models are used to complete various tasks such as providing email filters, smart assistants, search results, language translations, text analytics and so on.
 All of them need as an input words represented by means of numbers which is accomplished with word embeddings, in which particular lexical units are represented as vectors of real numbers.
+They are necessary to perform machine learning tasks that require the use of natural language.
+One of the ways to learn these vectors representations is to use simple neural network that based on the millions of real sentences examples assigns the numbers based on their co-occurrence and context.
+Using co-occurence results in a vector space that contains similar words closer to each other and dissimilar further away.
+It has been suggested [@Bolukbasi2016man; @Caliskan2017semanticsBiases; @Gonen2019lipstick; @Lauscher2019multidimensional; @Garg2018years; @Manzini2019blackToCriminal]  that in the learning process  such models can learn implicit biases that reflect harmful stereotypical thinking (eg.
+that the word \textit{she} is much closer in the vector space to the word \textit{cooking} than the word \textit{he}).
+This phenomenon may be undesired when using these vectors for further downstream tasks (eg.
+web search, recommendations, etc.).\todo{find downstream tasks and bias examples from papers} Before we move to more technical details regarding this topic, it is worth looking first at the theoretical background which may help to better grasp the concept of bias within machine learning algorithms.
 
-- troche o tym jak sa skonstruowane (optimizing for co-ocurrence frequency)
+In \todo{cite Are algorithms Value-Free} the author focuses on the general issue, namely whether algorithms can in principle be value-free.
+The author draws attention to the observation that in both scientific research and machine learning domain, induction plays a crucial role in the decision-making process.
+Induction is essential for both of these domains as "if ever considered literally just the facts, we would never be able to draw inductive conclusions" \todo{cite Gabrielle}.
+Induction comes with certain risks which are 1) the problem with justification, and 2) the risk of getting things wrong.
+During the process of designing machine learning process some decisions have to always be made.
+As the author states, there is no such thing as the justification in the abstract.
+\todo{cite Gabrielle} We may justify certain design decisions, data choice, etc.
+only by referring to the world's being certain way.
+Therefore if the induction is justified by contingent features that the world exhibits, which are intrinsically dependent on our current set of values, then our values indeed are the source of the justified inductive arguments.
+The other issue with induction is that as the machine learning algorithm always has access to a limited amount of information, it is prone to make mistakes from time to time, even if it's general accuracy is very high.
+What does it all mean in the context of bias within word embeddings?
+It seems that the presence of bias, value-laden algorithms is unavoidable due to the very nature of induction.
 
-- przejrzec jak to jest wprowadzane w kilku innych artykulach, zwl. fair is better than sensational
-
-- troche wiecej o tej konstrukcji i intuicji semantycznych, ale to nie super dobry pomysl
-
-- undesirable bias with respect to a certain groups words if stereotypically connected words (for stereotypes that we don't want to be used or relied on in downstream tasks) are systematically closer to each other.
-+ EXAMPLES
-
-It has been suggested [@Bolukbasi2016man; @Caliskan2017semanticsBiases; @Gonen2019lipstick; @Lauscher2019multidimensional; @Garg2018years; @Manzini2019blackToCriminal]  that in the learning process  such models can learn implicit biases that reflect harmful stereotypical thinking.
-A large chunk of  the literature on the topic  focuses on  the geometry of the learnt word embeddings;  in particular, unusually low (cosine) proximity of words belonging, intuitively, to a stereotype, is taken as a sign that  this stereotype has been built into a given embedding.
-This methodology will be in the focus of our paper.
+Assuming that the algorithms indeed due to the nature of their decision-making cannot fulfill the ideal of being value-free, it is even more justified to focus on ways how to properly detect harmful bias to limit its negative impact by the awareness of its presence.
+The aim of this paper is to contribute to this idea that using proper statistical tools may improve our understanding of what is happening inside machine learning models and especially within word embeddings.
 

### 2022-07-29 paperDraft/paperDraft4.tex
Updates draft

@@
   10pt,
-  dvipsnames,enabledeprecatedfontcommands, twocolumn]{scrartcl}
+  dvipsnames,enabledeprecatedfontcommands]{scrartcl}
 \usepackage{amsmath,amssymb}
@@
 
-\begin{itemize}
-\tightlist
-\item
-  akapit filozoficzny, z odniesieniem do value free
-\end{itemize}
-
 Modern Natural Language Processing (NLP) models are used to complete
@@
 accomplished with word embeddings, in which particular lexical units are
-represented as vectors of real numbers.
+represented as vectors of real numbers. They are necessary to perform
+machine learning tasks that require the use of natural language. One of
+the ways to learn these vectors representations is to use simple neural
+network that based on the millions of real sentences examples assigns
+the numbers based on their co-occurrence and context. Using co-occurence
+results in a vector space that contains similar words closer to each
+other and dissimilar further away. It has been suggested {[}1--6{]} that
+in the learning process such models can learn implicit biases that
+reflect harmful stereotypical thinking (eg. that the word \textit{she}
+is much closer in the vector space to the word \textit{cooking} than the
+word \textit{he}). This phenomenon may be undesired when using these
+vectors for further downstream tasks (eg. web search, recommendations,
+etc.).\todo{find downstream tasks and bias examples from papers} Before
+we move to more technical details regarding this topic, it is worth
+looking first at the theoretical background which may help to better
+grasp the concept of bias within machine learning algorithms.
 
-\begin{itemize}
-\item
-  troche o tym jak sa skonstruowane (optimizing for co-ocurrence
-  frequency)
-\item
-  przejrzec jak to jest wprowadzane w kilku innych artykulach, zwl. fair
-  is better than sensational
-\item
-  troche wiecej o tej konstrukcji i intuicji semantycznych, ale to nie
-  super dobry pomysl
-\item
-  undesirable bias with respect to a certain groups words if
-  stereotypically connected words (for stereotypes that we don't want to
-  be used or relied on in downstream tasks) are systematically closer to
-  each other. + EXAMPLES
-\end{itemize}
+In \todo{cite Are algorithms Value-Free} the author focuses on the
+general issue, namely whether algorithms can in principle be value-free.
+The author draws attention to the observation that in both scientific
+research and machine learning domain, induction plays a crucial role in
+the decision-making process. Induction is essential for both of these
+domains as ``if ever considered literally just the facts, we would never
+be able to draw inductive conclusions'' \todo{cite Gabrielle}. Induction
+comes with certain risks which are 1) the problem with justification,
+and 2) the risk of getting things wrong. During the process of designing
+machine learning process some decisions have to always be made. As the
+author states, there is no such thing as the justification in the
+abstract. \todo{cite Gabrielle} We may justify certain design decisions,
+data choice, etc. only by referring to the world's being certain way.
+Therefore if the induction is justified by contingent features that the
+world exhibits, which are intrinsically dependent on our current set of
+values, then our values indeed are the source of the justified inductive
+arguments. The other issue with induction is that as the machine
+learning algorithm always has access to a limited amount of information,
+it is prone to make mistakes from time to time, even if it's general
+accuracy is very high. What does it all mean in the context of bias
+within word embeddings? It seems that the presence of bias, value-laden
+algorithms is unavoidable due to the very nature of induction.
 
-It has been suggested {[}1--6{]} that in the learning process such
-models can learn implicit biases that reflect harmful stereotypical
-thinking. A large chunk of the literature on the topic focuses on the
-geometry of the learnt word embeddings; in particular, unusually low
-(cosine) proximity of words belonging, intuitively, to a stereotype, is
-taken as a sign that this stereotype has been built into a given
-embedding. This methodology will be in the focus of our paper.
+Assuming that the algorithms indeed due to the nature of their
+decision-making cannot fulfill the ideal of being value-free, it is even
+more justified to focus on ways how to properly detect harmful bias to
+limit its negative impact by the awareness of its presence. The aim of
+this paper is to contribute to this idea that using proper statistical
+tools may improve our understanding of what is happening inside machine
+learning models and especially within word embeddings.
 
@@
 
-\hypertarget{weat}{%
-\subsection{WEAT}\label{weat}}
+\vspace{1mm}
+\footnotesize
 
-\hypertarget{mac}{%
-\subsection{MAC}\label{mac}}
+\begin{figure}
+
+\begin{center}\includegraphics[width=1.1\linewidth]{paperDraft4_files/figure-latex/unnamed-chunk-2-1} \end{center}
+\caption{dsds}
+\label{fig:weat7google}
+\end{figure}
 
